### 2017-12-18 DraftExec.Rmd
beginnings of a draft report

---
title: "Efficiency"
author: "Aidan Morrison"
date: "11/12/2017"
output: pdf_document
---

---
references:

- id: garland2005
  title: Elementary Physics of Reactor Control Module - Fission Product Poisoning
  author:
  - family: Garland
    given: Bill
  volume: 3
  URL: 'http://www.nuceng.ca/br_space/2017-09_4d03_6d03/misc_files/xenon_poisoning.pdf'
  publisher: Nuclear Engineering (Originally from McMaster University)
  type: article-journal
  issued:
    year: 2005

- id: carlton2007
  title: Marine Propellers and Propulsion
  author:
  - family: Carlton
    given: John
  URL: 'http://www.nuceng.ca/br_space/2017-09_4d03_6d03/misc_files/xenon_poisoning.pdf'
  publisher: Elsevier
  issued:
    year: 2007

- id: liang2014
  title: The economics of slow steaming
  author:
  - family: Liang
    given: Lee Hong
  URL: 'http://www.seatrade-maritime.com/news/americas/the-economics-of-slow-steaming.html'
  publisher: Seatrade Maritime News
  issued:
    year: 2014
    month: 10
    day: 7

- id: sanguri2012
  title: The Guide to Slow Steaming On Ships
  author:
  - family: Sanguri
    given: Mohit
  URL: 'https://www.marineinsight.com/wp-content/uploads/2013/01/The-guide-to-slow-steaming-on-ships.pdf'
  publisher: Marine Insight
  issued:
    year: 2012
    month: 12

- id: WNA2017
  title: Nuclear Powered Ships
  author:
  - family: 
    given: 
  URL: 'http://www.world-nuclear.org/information-library/non-power-nuclear-applications/transport/nuclear-powered-ships.aspx#ECSArticleLink0'
  publisher: World Nuclear Association
  issued:
    year: 2017

- id: WNA2009
  title: Chernobyl Accident Appendix 1
  author:
  - family: 
    given: 
  URL: 'http://www.world-nuclear.org/information-library/safety-and-security/safety-of-plants/appendices/chernobyl-accident-appendix-1-sequence-of-events.aspx'
  publisher: World Nuclear Association
  issued:
    year: 2009

- id: patrick2012
  title: The Driving Factor in the SEA 1000 choice
  author:
  - family: Patrick
    given: Rex
  URL: 'https://corporate.siemens.com.au/content/dam/internet/siemens-com-au/root/aunz-defence-solutions/apdr-october-2015-issue-future-submarine.pdf'
  publisher: Asia-Pacific Defence Reporter
  issued:
    year: 2015
    month: 10

- id: hamilton1997
  title: Jet Torque
  author:
  - family: Waterjets
    given: Hamilton
  volume: 8
  URL: ''
  publisher: Hamilton Waterjets
  issued:
    year: 1997
    month: 3
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Contents

# Introduction

The purpose of this paper is to investigate the suitability of the pump-jet selection for Australia's future fleet of submarines.  In particular, the question of the relative efficiency, and quietness of pump-jets in comparison to open propellers is of particular importance.  Much has been made of the significance of the pump-jet in the DCNS (now Naval Group) bid for the submarine, to the extent that it has been claimed that the pump-jet rendered propellers obsolete.

Necessarily, this topic depends upon a considerable amount of physics, particularly around hydrodynamics, fluid mechanics, and turbomachinery, all of which might be difficult or daunting topics to usefully inform a public policy debate. However, in the case of a $50 billion dollar military acquisition, neglecting to engage with the technical matters that are so crucial to such a decision would be deeply foolish.  To that end, this paper has a very specific ambition.  It doesn't seek to advance the technical field with any new research or insight.  Instead, it aims to make as much of the relvant physics as possible to this particular question comprehensible in laymans terms, and connect the essential concepts as directly as possible with the most important conclusions of public import.

As such, it will tend to make frequent use of diagrams, illustrations, as well as simple prose, and relatively infrequently use mathematical equations, (unless of crucial importance), though some effort is made to point to references where the fuller and more formal derivations of these relationships can be found.

It should be noted that the technical fields involved here are vast and rapidly evolving.  The author makes no claim to be a world authority on the entirety of the topics which are touched. It is quite possible that there are other effects and phenomena which could be relevant which aren't mentioned here, and it's also true that additional advances in the state of the art are being made continually, and might not be known about in public literature.

However, the nature of scientific discovery and technical progress is that many of best discoveries are actually built-upon and confirmed, rather than swept aside, by later developments.  As such, plenty of foundational principals, in particular the conservation laws of energy and momentum and essential principals of mechanics and thermodynamics are just as true, right, and relevant as they were hundreds of years ago.  Consequently, whilst we cannot always know the precise degree of advance in a particular field of engineering, we can still confidently know some of the fundamental bounds and constraints that will be inherent in that field.

In this paper, since we cannot know the precise details of the particular state of the art in largely classified military programs, I will attempt to be particularly clear about those things for which we can have great confidence (often broad principals or relationships which determine trade-offs) and those things which we can't (the precise degrees or points and sensitivities).  To that end, in this paper I've underaken modelling based upon the broad principals which we can have confidence, and conducted sensitivity testing based on a range of plausible values which seem realistic for those things about which we can have less confidence. I have also, in discussion with experts and through reading the available literature, attempted where possible to identify the most plausible of the uncertain values, in order to advance discussion.

With the support of the principal sponsor of this paper, the relevant models may be found in a user-friendly format as a web-app to allow their further interrogation and testing for other plausible scenarios.  In addition, for increased transparency, the relevant code and equations underlying are available on github.

# Executive Summary

#### The difference between nuclear and conventional for speed
This paper was commissioned in order to investigate whether pumpjets could plausibly be as efficient, or more efficient than a suitably designed open propeller for a conventional submarine. A crucial input to the investigation is a rough understanding of what the speeds of operation are likely to be for a submarine.  Whereas nuclear submarines are reportedly capable of reaching speeds in excess of 30kt, and might transit long distances at such high speeds, conventional sumbarines are not thought of as being able to reach speeds far above 20kt in a sprint, and can only sustain speeds of 8-10kt for long-distance transits.  Moreover, on patrol, a large portion of thier work is done at very low speeds, typically thought to be in the range of 2-4kt.

#### The importance of low speed operation for diesel-electric submarines
The efficiency of the propulsion system at such low speeds is of great significance for a conventional submarine, since it must rely on batteries or other air-independent propulsion sources for power when entirely submerged.  Consequently, excessive energy consumption results in greatly reduced dived endurances.  Since a submarine's position is vastly more likely to be discovered when it is on the surface operating its diesel engines, this ability to remain submerged for a long time is crucially important for combat operations, and in transiting through sensitive or contested areas.  For nuclear submarines, which posess a practically infinite supply of energy from the nuclear reactor, efficiency at low speeds is of no concern.  In fact, dispersing additional energy, (provided it can be done quietly) is probably advantageous for a nuclear submarine, since it will allow the nuclear reactor to avoid running at very low power levels, where the stability of the reactor is reduced.

#### What is cavitation
Pumpjets have been widely adopted by navies operating nuclear submarines.  The principal advantage of a pumpjet relevant to submarines is their ability to avoid problems associated with cavitation, which is known to occur for propellers attempting to operate at high speeds.  Cavitation, which is the rapid expansion and collapse of a bubble or void in the water, is particularly problematic for submarines, since it results in the creation of a great deal of noise which could be detected by an enemy.

#### How does a pump-jet work
A fundamental requirement of a pump-jet in order to avoid this cavitation is that the the working parts of the jet (the rotating blades inside it, known as the impeller) operate at a higher pressure than propeller blades operating in open water would.  This means that the blades can turn at a lower speed relative to the water they are connected with, allowing a less violent action, which induces less cavitation.  In this way, the jet does it's work less by directly acelerating the water, but by raising its pressure.  This raised pressure is converted back to movement, which produces thrust, as it exits the jet and returns to the same pressure as the surrounding environment.

#### The necessity of drag, and decelleration, induced by the duct
The role of the shroud (or tunnel, mantel, duct) around the jet is to allow the pressure to be raised around the impeller, in a way that is not possible for an open propeller.  By a fundamental requirements of physics, this actually requires that the water's incoming speed be **reduced** before it reaches the impeller. (The kinetic energy embodied in movement is converted into potential energy, or pressure.)  Whilst the duct narrowing at the nozzle also necessarily accelerates the water, (as the additinal pressure imparted by the impeller is converted back to kinetic energy) it's an essential feature of all pump-jets that the water flow is decelerated at the point of reaching the impeller. Consequently, an elementary form of the pump-jet is also termed the 'decelerating duct' applied to a propeller.  Put simply, the water has to slow down to go fast again.

#### What happens to a jet at low speed, in simple terms
The problems for pump-jets arise when the water is already going slow, and you don't want it to go that much faster.  This is the case when a jet designed for high-speed attempt to operate at a dramatically reduced speed.  (Not to be confused with a vessel with very little water speed working very hard, as might be the case for a tug or barge.)  In this case, the slowing down and speeding up results in an unnecssary additional step which reduces greatly the efficiency of propulsive system. Or, put another way, overcomign the resistance of moving water through the shroud becomes much greater relative to the total thrus produced by the jet.

#### Why this isn't easily noticeable in normal circumstances
It is for this reason that waterjets of any kind are known to have an efficiency curve that falls off towards zero as the net thrust they produce also diminishes to zero (which for a given vessel will correspond to water speed).  This doesn't mean that they don't still work at low speeds and produce some thrust.  All it means is that far more power per unit of thrust will be required than might be at other speeds, as a higher fraction of energy is expended producing the turblence (random, round-and-round movement) inside and around the jet shroud than goes towards direct front-to-back acceleration of the water column, which produces thrust.

#### What the submarine requirement tells us about the duct
Whist estimating the exact shape and level of the efficiency curve for a particular pump-jet and propeller is impossible without detailed knowledge of their design, the over all trends of their shapes in the extremes can be known from well-established principals.  Moreover, the particular demands of a pump-jet suited to a nuclear submarine, (eliminating all cavitation in as wide a range of operating circumstances as possible) considerably narrows the plausible range concerned.  In order to minimise cavitation, the degree of pressure elvation at the impeller would be relatively high, or the total surfaces of the impeller blades much aslo become larger.  Both of these design requirements necessitate changes to the duct or blade designs which would be in tension with overall efficiency, and most pronounced at the lowest of operating speeds.

#### Quantitative Conclusions
In the production of this paper I have developed a computational model which maps the impact of different efficiency curves directly to the dived range and endurance of a submarine. The most plausible scenarios I find include the reduction of dived endrance and range between 20% and 50%, or effectively halving time and distance that submarine may remain submerged for during combat operations at speeds around 3-5kt.

#### The acoustic advantage of jets at higher speed
My review of the literature confirms that a pumpjet may produce a significant acoustic advantage in circumstances where an open propeller would experience any degree of cavitation. Indeed, it has been remarked by naval researchers that pump-jets could be designed which would not cavitate past the point when the body they propel experienced caviation.   As such, it seems perfectly plausible that pump-jets confer one advantage on a submarine, in that they can accelerate to a higher speed without cavitation occurring.  This is known as a higher 'tactical silent speed'.

#### Turbulence and flow separation at low speed
However, at much lower speeds such as patrol speeds (2-4kt) it is most likely that a propeller will be able to operate well below the point of any cavitation inception, and would consequently also be extremely quiet.  Moreover, a jet will necessarily incur substantially larger degrees and types of turbulence in order to produce net thrust in this regime.  These likely include discontinuities and instability in the flow entering the duct and passing through the impeller and stator (flow separation) as the water is acelerated sufficient to be slowed, and then re-accelerated.  In certain circumstances where resonances might arise these could be highly adverse to acoustic performance.  However, assuming that by careful design such resonances can be eliminated (which we assume has been achieved) these effects would not result in any cavitation, and would only increase the noise attributable to turbulence in solid water, which is far less than for any cavitation.

#### The acoustic question at low speed
But in either case, given the necessarily raised levels of turblence generated by a jet than a propeller at very low speeds, the claim that the jet is quieter in this regime must rely entirely upon shielding effects from the shroud.  These could be substantial in a directions perpendicular to the direction of travel, but would be much smaller when viewed from the aft or forward directions.  It should also be noted that many underwater acoustic environments, sound tennds to reflect and bend in different directions as it propagates. As such, the claim that pump-jets are universally quieter than propellers should be treated with caution.  In a variety of circumstances, including most operations at patrol speed, it may not always be true.

#### The inevitable trade-off
However, as a direct and necessary result of this higher tactical silent speed, some disadvantages will be incurred, owing to the additional drag induced by the shroud which prevents cavitation at the impeller.  These are a lower dived endurance, a lower dived range, a lower overall endurance, a lower overall range, and a worse indescretion ratio.

#### The contradiction between stated requirement and chosen technology
Given the substantial emphasis which was placed on overall range and endurance of the submarine, it is difficult to understand how one particular unique requirement would be elevated so high above the other strategic and tactical advantages afforded by propellers.

# Speed and Drag - Why very slow is very very $(very)^2$ economical

Perhaps the most important relationship to understand is the relationship between speed and drag as it pertains to submarines.  This is immportant because it sets out the fundamental framework as it applies to any submarine, regardless of propulsion type.  (It's also a pretty important for planes, cars, missiles, torpedoes, and basically everything else.)

Drag is the resistance that a fluid (air or water, in our case) gives to a body that is passing through it. Quite simply, it's a force that acts in the opposite direction.  There are multiple sources of drag for different types of scenarios.  For scenarios where an object is in contact with two different types of fluid (like a ship, on the ocean) or when the fluid doesn't really have contact with all of the object (supersonic flight, and supercavitating torpedoes) some more complex physics applies. For a fuller discussion of types of drag, see [@carlton2007]  But in the case of a submarine, which does it's business completely immersed in the ocean, the relevant physics is dominated the skin friction on the hull, which follows a very simple rule and relationship.  The amount of drag ($F_D$) an objet experiences increases directly in proportion to the surface area $A$.  For any given object of a certain (unchanging shape) there will be a constant coefficent ($C_D$) which reflects how aerodynamic or hydrodymanic the shape is.  The drag is also directly proportional to the density of the fluid being moved through, $\rho$.  (Air creates roughly one-thousandth the drag as water does on any given object at a given speed, since it's roughly one thousand times less dense.)

But the most sensitive factor in this relationship is the speed at which the object moves through the liquid.  The drag increases not with the speed ($v$), but with the **square** of the speed . This means if the speed doubles, the drag increases by four.  If the speed triples, the drag increases by a factor of nine.

$$F_D = \frac{1}{2}\rho v^2C_DA$$

(Technical aside: This law applies wherever the the flow over the surface is turbulent. It is true that for very small objects, or very viscous fluids, or very slow movements a different apples called Stokes Equation, in the case wher Reynolds numbers are less than 1.  Given that sea-water is not particularly viscous, and sumbarines not particularly small, Reynolds numbers are liekly to be much much greater than 1 (one or two thousand), even when moving at only one or two knots.  Since it is unlikely that a significant proportion of the flow over the hull will be laminar, we'll use the drag equation in all modeling going forward when considering drag on the hull).

It's crucial to understand, however, that drag is only a force, and doesn't directly inform us about how much energy is consumed, until we multiply it by the **distance** over which it is applied, not the time for which it is applied.  To think about it simply, gravity exerts a force on you downwards all the time.  But you don't expend any energy overcoming it when sitting still.  If you climb stairs, the amount of gravitational potential you attain depends on how high up you climb, not how long you spend on the ladder or stair-case.

This has a significant consequence for propulsion, since the amount of power (energy expended in a given time) required for thrust scales with the drag, multiplied by the distance covered in a given time.  As such, power required (and fuel/battery consumption) scales with velocity **cubed** rather than velocity squared.  This means that if you double your speed of travel, fuel consumption for a given period of time will increase by a factor of eight.  Due to the increased speed of travel, the fuel required to cover a given distance will only quadruple.  Hence the range that can be covered with a given amount of energy (assuming propulsion dominates energy requirements) tends to scale with the inverse of velocity squared, whereas the endurance (amount of time that can be spent travelling) scales with the inverse of velocity cubed.

This has a profound impact on the operation and engineering of maritime vessels.  Reducing velocity has such a substantial reduction on hull drag that going slower is almost always a reasonable means of conserving fuel overall.  To the extent that time is non-critical, slower is always much much better.  It's for this reason that during the financial crisis, many cargo shipping companies adopted the practice of slow steaming [@liang2014] in order to conserve fuel, despite this causing a range of possible new engineering issues which need to be accounted for in order to operate the engines at lower than normal power for sustained periods [@sanguri2012 pp. 8-10]. The fuel saving from operating even 30% slower means that around half the total energy is required for a given journey.  Other incremental costs and inefficiences from operating 'off-design' are frequently outweight by such a dramatic reduction in overall power demand.

This effect is of profound importance to understanding the operations of conventional submarines, which have extremely constrained energy stores when operating under the surface. Diesel fuel when burned has an energy density of approximately 45MJ/kg.  In contrast, a lead-acid battery might have energy densities in the range of 0.08-0.14MJ/kg. With something like 400 times as much energy per kg embarked in diesel, operating on batteries imposes an extreme demand for economy on propulsive power.  Happily for submarines, the square law for drag, and cubed law for power, allow an almost commensurate reduction in power demands to take place by slowing down to very slow speeds when submerged. Hypothetically, a conventional submarine might have a total range of 10,000nm from using it's diesel payload at 8kt.  If an equal weight of lead-acid batteries as fuel were carried, the submarine could only travel about 25nm when submerged at the same speed in a single charge.  But by travelling at 4kt, that quickly increases to 100nm, and 400nm at 2kt (neglecting hotel load here for simplicity).

It's worthwhile pointing out that extreme demand for economy imposed by the poor energy density of batteries distinguishes submarines quite remarkably from other types of boat or ship design.  In every other application, the propulsion system is designed around a particular speed and loading condition which it is optimally efficient for, and a band of plausible variation around this.  For example, a cargo ship designed for operation at 22kt will have it's propeller and engine (propulsion system) very carefully designed around these speeds and the plausible levels of loading at those speeds. A top-speed which might be somewhat higher than 22kt would be calculated, but would be relatively unlikely to be used for any purpose other than an emergency.  Lower speeds might be considered 15-18kt for the purposes of slow-steaming.  However, the relative efficiency at 3kt is of little concern, since the drag at that speed will be under 2% of what it would be at 22kt, and the power required around 0.25%.  Even if the efficiency was substantially worse, or better (by a factor of two or three even) the impact on the overall economics of the operation would be negligible when compared to marginal improvements at the design speed.

The same consideration applies to other types of vessels, which might use water-jets.  Fast pilot-boats, for instance, might spend quite a considerable time manouvering in and out of harbour or alongside at very low waterspeeds. However, their overall fuel consumption is still likely to be dominated by the fast section of their trip to a ship.  If the in-harbour water-speed was 8kt, but 32kt was the optimum speed for the open-water section of their journey, the fuel requirement for a given distance could still be over ten times greater at 32kt. (Aside, the physics described above would say 16 times, but some reduction due to the reduction of surface area from planing might offset this.)  Consequently the owner/operator would be at least ten times as concerned about efficiency at 32kt as at 8kt, (unless the harbour transit was much much longer than the open water section), and have negligible concern about effiencies at 2kt or 3kt, which would rarely amount to more than a couple of percent of total propulsion energy demand.

In stark contrast, a submarine operator might have less than 1% of the energy store available for the submerged movements at 2-3kt, and consequently be substantially **more** concerned about efficiency in this regime, particularly since this would be the regime in which all of their combat operations would be conducted.

This imposes a constraint on the way in which literature on jets, propellers, and propulsion systems in general is read, since almost all of it is deeply concerned with how a particular system would operate at one particular design speed, and choosing the optimum system for that design speed and load.  Considerable attention is given to how a propeller would work in off-design conditions relating to changes in load, such as in cases when a ship is more lightly laden, or a tug-boat is pushing a different sized ship.  There is, however, a relative scarcity of attention given to the performance of a propulsion system operating at a dramatically different speed, with the same load. It also corresponds to scenarios where engineers and manufacturers seem from discussion to be far less confident in making quantitative claims about the performance of their own systems.  (They are perfecly happy to make qualitative claims, "it would be alright", "nothing very bad would happen" etc.)

Given that submarines almost by definition (since their hull shape doesn't change, and they must carry ballast to make up for any under-loading) are essentially always operating with the same load, but have to operate over an extremely wide variety of different speeds (a factor of four or more in variance) with severe efficiency concern at all of them, they represent a truely unique engineering question, seldom discussed in commercial applications.  Consequently, there are relevantly few pieces of literature which address the present question directly. (Happily, not zero)  However, there are plenty of instances where the relevant physics and trade-offs are explained in depth to address related, but not identical questions.

# The difference between nuclear and conventional propulsion

Nuclear submarines vary quite remarkably from conventional submarines because of the means by which they generate their power.  Because of the extremely high energy density in enriched uranium or plutonium, the reactors on board nuclear submarines generate an abundant supply of energy.  Most submarine reactors are reportedly capable of generating between 25 to 50 megawatts (MW) of power, though Russian submarines have hundreds of megawatts of power available [@WNA2017].

In contrast, the Collins Class Submarine's main motor is rated at less than 6MW, with designs for the future submarine appearing to be only slightly larger [@patrick2012].  It is fair to say that in terms of maximum power ouput, nuclear submarines could have something like 5-10 times as much power at their disposal, and their **peak** output.

The difference between the peak-powers of the submarines significantly understates how different the designs are, because the difference in the total amount of energy stored and available for use in a voyage or dive is vastly greater.  A nuclear submarine might have literally millions of times more energy at its disposal, it is practically unlimited for all intents and purposes on a given voyage. Consequently, wheras a nuclear submarine might regularly conduct transits at or near it's peak power, a diesel electric submarine would probably transit at less than half the speed which it could manage at a sprint, (and use less than a quarter of the energy, as discussed earlier) and on patrol it might be operating at a tenth of the maximum speed, and use maybe just 1% of as much power on propulsion. In this situation, the amount of power that is drawn for lights, CO2 scrubbing, washing, cooking, heating, as well as the electronics driving the combat systems (the 'Hotel Load') might well become significant, and even be as larger or larger than what is required for propulsion.

Consequently, the difference between the power output a conventional and a nuclear submarine might regularly operate at could be even quite a bit larger than the maximum amount of energy that they can deliver to their propulsion systems.

Perhaps more significantly for power-design of nuclear submarines is a little-discussed phenomenon called Xenon poisoning, which affects nuclear reactors when they shut down or lower their power significantly. When Uranium or Plutonium atoms split (or fission) into two smaller isotopes atoms, a variety of different radio nucleides (unstable variants of atoms) are produced.  Two of these are Xenon-135 and Iodine 135.  Iodine is produced much more often, and decays into Xenon-135, which a half-life of about nine hours.  This Xenon has a very special, perhaps unique role in reactors, since they very easily captures the free neutrons which cause the continued fission reactions in Uranium or Plutonium which drive the reactor.  As such, Xenon is known as reactor 'poison' since it can kill the reactor's reactivity in very high doses.

This high neutron-capture from Xenon means that it has a duplicitous relationship with the reactor's power level.  At high levels of reactor power, lots of Xenon-135 and Iodine-135 are produced by the fission process.  The high presence of Xenon reduced the reactor's reactivity.  On the other hand, there are lots of neutrons available to 'burn off' the Xenon (which absobs the neturons to become a different isotope).  This keeps the total level of Xenon in check.

The situation becomes much more complicated when the reactor undergoes a sudden change in power output.  If the power is lowered dramatically and suddenly, the production of Xenon continues quite rapidly for some time due to the decay of the large stock of Iodine-135.  With less neutron flux available to 'burn off' the Xenon, the Xenon levels spike, and push down the reactor's reactivity.  Unless the reactor is quickly raised back to relatively high power (~60%) quite quickly (an hour or less), the Xenon levels become so high that the reactor will have to be shut down, otherwise extreme (and dangerous) measures would be required in order to keep the reactor going.  (This is essentially what lead to the Chernobyl Explosion [@WNA2009].)  A fuller discussion of Xenon poisoning can be found in [@garland2005], which demonstrates key concepts related to the poisoning effect shown below.

![Xenon poisoning effect following shutdown from @garland2005](/Users/aidanmorrison/submarines/XenonPoison.png){width=500px}

Consequently, nuclear reactors aren't well suited to rapid fluctuations in power, particularly dramatic reductions in power, as these can lead to instability in the reactor core.  If the reactor is shut down in order to avoid such dangerous circumstances, it generally cannot be started again until Xenon levels have fallen again, which can take a couple of days.  Obviously this is never desirable for a military vessel, and hence is avoided at almost all costs.

This effect has a dramatic impact on nuclear submarine design, since much of their design is oriented around being able to comfortably disperse large amounts of excess power, rather than with conserving it.  For the propulsion system, this actually makes having an inefficient propulsor at low speeds a considerable advantage.  Since the reactor will likely need to dispose of excess power, particularly during ramp-down, an inefficient propulsion system actually provides a useful power sink.  Since any excess power will have to be disposed of by some other means (normally by pumping more water to remove the power as the heat) inefficiency at low speeds has no penalty, and probably a marginal benefit, since it will reduce overall demand for additional systems.    Provided the excess turbulence inside the pumpjet isn't too noisy, wasting energy through the propulsor is useful.

It should also be noted that a nuclear reactor's aversion to sudden reductions in power would also have a substantial impact on the design of a submarine's combat system, and it's demand on the Hotel Load. For the same reason, a high Hotel Load, or power-hungry Combat System, could actually be advantageous, as it helps to set an elevated 'floor' for power requirements, reducing the scale of fluctuations in overall power demand from the reactor due to changes in propulsion speed.

# Some essential concepts

### Conservation of Energy

This is perhaps one of the most fundamental and well-established principals in physics.  The essential idea is that energy can move or change in form, but it isn't ever created or destroyed.  Machines, plants and animals all derive their energy from a particular other source, which can be measured and evalutated to establish the limits of energy available.  Plants collect energy from sunlight falling on their leaves. Humans (as well as combustion engines) capture the chemical potential energy in organic matter, and release it by combining it with oxygen.  Hydro-electric power plants turn the gravitational potential energy of water stored at a height into electricity.

Conservation of energy has a particularly relevant embodiment in fluid flows, which is given it's primary expression in Bernoullis equation.  It says that the total energy in a connected body of fluid is constant, though it can change in form between kinetic energy (movement), gravitational potential energy (it being elevated) the heat energy in the fluid, and the pressure of the fluid.

In different fluids, the dynamics of how energy moves between one and another change.  For example, in gases, heating up a confined piece of gas will increase it's pressure, or if it is unconfined, increase its volume. This is particularly important for understanding gas turbines. In the case of a liquid, however, all the molecules are in close contact, and hence can't increase in volume or pressure substantially except by the creation of steam. As such, in the absence of large amounts of cavitation, the terms in the relationship which are most important for our consideration are the relationships between pressure, and kinetic energy, and gravity.

In the case of most waterjets which propel surface vessels, the water is lifted from the bottom of the hull at the intake to the pump, which is generally incorporated inside the hull. This increase in gravitational potential coincides with a slowing down of the water relative to the vessel-speed. Since the jet in such vessels is generally ejected at the same height as the pump, this potential is never regained, and is technically a loss, however at high speeds such a loss is small relative to the total power ouput.

#### Venturi Effect and Flow Diffusion
In the case of submarines and torpedoes the water doesn't undertake a change in height, since intake and nozzle are generally all in line with the central axis of the submarine or torpedo.  As a consequence, the key relationship in Bernoullis equations is the relationship between the liquids velocity, and its pressure.  The consequence is that when water moves through a pipe (duct, or shroud) it's pressure is inversely related to the square of its velocity.  This means that when a liquid is forced to to travel through a narrowing pipe, it's pressure necessarily decreases as the velocity increases.  This is a simple embodiment of the Venturi effect. The inverse process is where a pipe increases in volume, and the flow is forced to slow down in order to fill the wider area, and the pressure correspondingly increases.  This is a process called 'diffusion', and is important to achieving high pressure levels in many types of water pumps, including those which will be particularly relevant for pumpjets for watercraft propulsion [@hamilton1997].

#### Energy waste and efficiency
The conservation of energy also has useful implications for how the efficiency of systems is thought about.  In particular, because energy is conserved, identifying inefficiencies in a system necessarily involves identifying where energy goes to doing tasks which aren't useful for the intended purpose.  A perfectly efficient system won't do any work that isn't for the intended purpose.  In the case of analysing the efficiency of propulsion systems, the relevant 'work' is almost always related to moving water backwards to produce thrust.  Moving water in directions other than backwards, including random turbulent flows which wind producing heat rather than thust, are two examples of wasted energy.  Noise in the water also reflects energy which is wasted.

### Boundary Layer
When water or any fluid flows with some speed relative to another solid surface nearby, there is some layer adjacent to the surface in which the speed of the fluid is diminished relative to the main flow.  At a microscopic level, there are some molecules of the fluid on the surface which will be effectively static relative to the surface.  The layer of fluid that joins the gap between the static surface, and the part of the flow which is moving at the full flow speed, is called the boundary layer.  Exactly how thick the boundary layer is, and how the fluid moves in the boundary layer, is extremely important for consideration of efficiency of fluid flows over and around solid surfaces.

### Turbulent and Laminar Flow

### Pressure, Thrust, and Momentum

### Cavitation

### Flow separation

# What tech are we talking about? The core relationships between pumps, jets, ducts, and propellers, and all the different things we call them, their parts, and combinations

# Propulsive performance of the pumpjet, or decelerating ducted propeller

## Efficiency Relationships, and various curves

## Practical and recent evidence of efficiency relationships

# Results of numerical simulation

# Acoustic Considerations

# Conclusions

# References

### 2017-12-18 R/calibrator.R

@@ -37,7 +37,7 @@ calibrator <- function(df, system = "jet", method = "hotel match", hotel = 150, 
 
   eff. <- df.$eff[index.]
 
-  const <- speed.^3/(power.*eff.)
+  const <- power.*eff./(speed.^3)
 
   const
 

### 2017-12-18 scripts/basic.R

@@ -30,12 +30,12 @@ eff.jet <- c(new.eff.jet$value)
 
 
 #######  NUMBERS TO ADJUST BY USE IN APP ##########
-hotel <- 150
+hotel <- 200
 total.batt <- 500 #Tonnes,
 batt.dens <- 0.15
 
 #######
-patrol <- 2.5
+patrol <- 4.5
 
 # There are a few different ways of providing a reference point for the power consumption for a given propulsion type.
 # 1. Nominate a speed at which power consumption for propulsion is equal to hotel load, as per Andrew Davies article
@@ -65,7 +65,7 @@ df <- df %>%
 
 ### find the constant here
 
-const <- calibrator(df = df, system = "jet", method = "max power")
+const <- calibrator(df = df, system = "jet", method = "hotel match", patrol = 1.5)
 
 df <- df %>%
   mutate(power.mob.req = const*speed^3,

### 2017-12-18 DraftExec.Rmd
not finished

@@
 date: "11/12/2017"
-output: pdf_document
+output: bookdown::pdf_document2 
 ---
@@
 knitr::opts_chunk$set(echo = TRUE)
+
+# Figure numbering functions ====
+
+fig <- local({
+    i <- 0
+    list(
+        cap=function(refName, text, center=FALSE, col="black", inline=FALSE) {
+            i <<- i + 1
+            ref[[refName]] <<- i
+            css_ctr <- ""
+            if (center) css_ctr <- "text-align:center; display:inline-block; width:100%;"
+            cap_txt <- paste0("<span>", text, "</span>")
+            anchor <- paste0("<a name=\"", refName, "\"></a>")
+            if (inline) {
+                paste0(anchor, cap_txt)    
+            } else {
+                list(anchor=anchor, cap_txt=cap_txt)
+            }
+        },
+
+        ref=function(refName, link=FALSE, checkRef=TRUE) {
+
+            ## This function puts in a cross reference to a caption. You refer to the
+            ## caption with the refName that was passed to fig$cap() (not the code chunk name).
+            ## The cross reference can be hyperlinked.
+
+            if (checkRef && !refName %in% names(ref)) stop(paste0("fig$ref() error: ", refName, " not found"))
+            if (link) {
+                paste0("<A HREF=\"#", refName, "\">Figure ", ref[[refName]], "</A>")
+            } else {
+                paste0("Figure ", ref[[refName]])
+            }
+        },
+
+        ref_all=function(){
+            ## For debugging
+            ref
+        })
+})
+
+library(knitr)
+knit_hooks$set(plot = function(x, options) {
+    sty <- ""
+    if (options$fig.align == 'default') {
+        sty <- ""
+    } else {
+        sty <- paste0(" style=\"text-align:", options$fig.align, ";\"")
+    }
+
+    if (is.list(options$fig.cap)) {
+        ## options$fig.cap is a list returned by the function fig$cap()
+        str_caption <- options$fig.cap$cap_txt
+        str_anchr <- options$fig.cap$anchor
+    } else {
+        ## options$fig.cap is a character object (hard coded, no anchor)
+        str_caption <- options$fig.cap
+        str_anchr <- ""
+    }
+
+    paste('<figure', sty, '>', str_anchr, '<img src="',
+        opts_knit$get('base.url'), paste(x, collapse = '.'),
+        '"><figcaption>', str_caption, '</figcaption></figure>',
+        sep = '')
+
+})
+
+## This chucnk will read through *this* Rmd file, and attempt to extract all of the 
+## labels (not caption text) used for Figure captions. These labels are used
+## as anchors, so scanning through the document now will allow us to create cross references
+## before the caption actually appears.
+
+## Get the name of this Rmd file
+rmdFn <- knitr::current_input()  # filename of input document
+
+## Read lines and close connection
+rmdCon <- file(rmdFn, open = "r")
+rmdLines <- readLines(rmdCon)
+close(rmdCon)
+
+## Pull out all occurences of at least one back tick, followed 
+## by any number of characters, followed by fig$cap (all on one line)
+figscap_idx <- grep("`+(.*)fig\\$cap", rmdLines)
+rmdLines <- rmdLines[figscap_idx]
+
+## Get rid of everything up until the start of the caption label
+## This presumes the caption label is the first argument of fig$cap()
+## E.g., fig.cap = fig$cap("my_label", ...)
+rmdLinesSansPre <- sub("(.*)fig\\$cap(.*?)[\"']", "", rmdLines)
+
+## Identify everything up until the first quote
+match_data <- regexpr("(.*?)[\"']", rmdLinesSansPre)
+
+## Reduce the length by one, because we're not interested in the final quote
+attr(match_data, "match.length") <- attr(match_data, "match.length") - 1
+
+## Extract
+fig_labels <- regmatches(rmdLinesSansPre, match_data, invert=FALSE)
+
+if (length(fig_labels) > 0) {
+
+    ## Test for duplicates
+    if (anyDuplicated(fig_labels) > 0) stop("Duplicate caption labels detected")
+
+    ## Create a named list of Figure numbers
+    ref <- as.list(1:length(fig_labels))
+    names(ref) <- fig_labels
+}
+
 ```
@@
 But the most sensitive factor in this relationship is the speed at which the object moves through the liquid.
-The drag increases not with the speed ($v$), but with the **square** of the speed .
+The drag increases not with the speed ($v$), but with the **square** of the speed.
 This means if the speed doubles, the drag increases by four.
@@
 $$F_D = \frac{1}{2}\rho v^2C_DA$$
+`r fig$cap("xenon1", "Xenon poisoning effect following shutdown from @garland2005")`
 
@@
 Unless the reactor is quickly raised back to relatively high power (~60%) quite quickly (an hour or less), the Xenon levels become so high that the reactor will have to be shut down, otherwise extreme (and dangerous) measures would be required in order to keep the reactor going.
-(This is essentially what lead to the Chernobyl Explosion [@WNA2009].)  A fuller discussion of Xenon poisoning can be found in [@garland2005], which demonstrates key concepts related to the poisoning effect shown below.
+(This is essentially what lead to the Chernobyl Explosion [@WNA2009].)  A fuller discussion of Xenon poisoning can be found in [@garland2005], which demonstrates key concepts related to the poisoning effect shown in `r fig$ref("xenon")`.
+And you can see that in the [Introduction].
 
-![Xenon poisoning effect following shutdown from @garland2005](/Users/aidanmorrison/submarines/XenonPoison.png){width=500px}
+![`r fig$cap("xenon", "Xenon poisoning effect following shutdown from")`](XenonPoison.png){width=500px}
+
+\pagebreak
+\begin{figure}
+\includegraphics[width=0.9\textwidth]{XenonPoison.png}
+\caption{A prototype of the Job Information dialog  @garland2005}
+\label{fig:jobInformationDialog}
+\end{figure}
 

### 2017-12-18 test.Rmd
not finished

--- 2017-12-18 DraftExec.Rmd
@@
 ---
-title: "Efficiency"
-author: "Aidan Morrison"
+title: "Test"
+author: "R"
 date: "11/12/2017"
@@
     year: 2005
-
-- id: carlton2007
-  title: Marine Propellers and Propulsion
-  author:
-  - family: Carlton
-    given: John
-  URL: 'http://www.nuceng.ca/br_space/2017-09_4d03_6d03/misc_files/xenon_poisoning.pdf'
-  publisher: Elsevier
-  issued:
-    year: 2007
-
-- id: liang2014
-  title: The economics of slow steaming
-  author:
-  - family: Liang
-    given: Lee Hong
-  URL: 'http://www.seatrade-maritime.com/news/americas/the-economics-of-slow-steaming.html'
-  publisher: Seatrade Maritime News
-  issued:
-    year: 2014
-    month: 10
-    day: 7
-
-- id: sanguri2012
-  title: The Guide to Slow Steaming On Ships
-  author:
-  - family: Sanguri
-    given: Mohit
-  URL: 'https://www.marineinsight.com/wp-content/uploads/2013/01/The-guide-to-slow-steaming-on-ships.pdf'
-  publisher: Marine Insight
-  issued:
-    year: 2012
-    month: 12
-
-- id: WNA2017
-  title: Nuclear Powered Ships
-  author:
-  - family: 
-    given: 
-  URL: 'http://www.world-nuclear.org/information-library/non-power-nuclear-applications/transport/nuclear-powered-ships.aspx#ECSArticleLink0'
-  publisher: World Nuclear Association
-  issued:
-    year: 2017
-
-- id: WNA2009
-  title: Chernobyl Accident Appendix 1
-  author:
-  - family: 
-    given: 
-  URL: 'http://www.world-nuclear.org/information-library/safety-and-security/safety-of-plants/appendices/chernobyl-accident-appendix-1-sequence-of-events.aspx'
-  publisher: World Nuclear Association
-  issued:
-    year: 2009
-
-- id: patrick2012
-  title: The Driving Factor in the SEA 1000 choice
-  author:
-  - family: Patrick
-    given: Rex
-  URL: 'https://corporate.siemens.com.au/content/dam/internet/siemens-com-au/root/aunz-defence-solutions/apdr-october-2015-issue-future-submarine.pdf'
-  publisher: Asia-Pacific Defence Reporter
-  issued:
-    year: 2015
-    month: 10
-
-- id: hamilton1997
-  title: Jet Torque
-  author:
-  - family: Waterjets
-    given: Hamilton
-  volume: 8
-  URL: ''
-  publisher: Hamilton Waterjets
-  issued:
-    year: 1997
-    month: 3
 ---
@@
 ```{r setup, include=FALSE}
+library(knitr)
 knitr::opts_chunk$set(echo = TRUE)
+
+## This chucnk will read through *this* Rmd file, and attempt to extract all of the 
+## labels (not caption text) used for Figure captions. These labels are used
+## as anchors, so scanning through the document now will allow us to create cross references
+## before the caption actually appears.
+
+## Get the name of this Rmd file
+rmdFn <- knitr::current_input()  # filename of input document
+
+## Read lines and close connection
+rmdCon <- file(rmdFn, open = "r")
+rmdLines <- readLines(rmdCon)
+close(rmdCon)
+
+## Pull out all occurences of at least one back tick, followed 
+## by any number of characters, followed by fig$cap (all on one line)
+figscap_idx <- grep("`+(.*)fig\\$cap", rmdLines)
+rmdLines <- rmdLines[figscap_idx]
+
+## Get rid of everything up until the start of the caption label
+## This presumes the caption label is the first argument of fig$cap()
+## E.g., fig.cap = fig$cap("my_label", ...)
+rmdLinesSansPre <- sub("(.*)fig\\$cap(.*?)[\"']", "", rmdLines)
+
+## Identify everything up until the first quote
+match_data <- regexpr("(.*?)[\"']", rmdLinesSansPre)
+
+## Reduce the length by one, because we're not interested in the final quote
+attr(match_data, "match.length") <- attr(match_data, "match.length") - 1
+
+## Extract
+fig_labels <- regmatches(rmdLinesSansPre, match_data, invert=FALSE)
+
+if (length(fig_labels) > 0) {
+
+    ## Test for duplicates
+    if (anyDuplicated(fig_labels) > 0) stop("Duplicate caption labels detected")
+
+    ## Create a named list of Figure numbers
+    ref <- as.list(1:length(fig_labels))
+    names(ref) <- fig_labels
+}
+
 ```
@@
 # Introduction
-
-The purpose of this paper is to investigate the suitability of the pump-jet selection for Australia's future fleet of submarines.
-In particular, the question of the relative efficiency, and quietness of pump-jets in comparison to open propellers is of particular importance.
-Much has been made of the significance of the pump-jet in the DCNS (now Naval Group) bid for the submarine, to the extent that it has been claimed that the pump-jet rendered propellers obsolete.
-
-Necessarily, this topic depends upon a considerable amount of physics, particularly around hydrodynamics, fluid mechanics, and turbomachinery, all of which might be difficult or daunting topics to usefully inform a public policy debate.
-However, in the case of a $50 billion dollar military acquisition, neglecting to engage with the technical matters that are so crucial to such a decision would be deeply foolish.
-To that end, this paper has a very specific ambition.
-It doesn't seek to advance the technical field with any new research or insight.
-Instead, it aims to make as much of the relvant physics as possible to this particular question comprehensible in laymans terms, and connect the essential concepts as directly as possible with the most important conclusions of public import.
-
-As such, it will tend to make frequent use of diagrams, illustrations, as well as simple prose, and relatively infrequently use mathematical equations, (unless of crucial importance), though some effort is made to point to references where the fuller and more formal derivations of these relationships can be found.
-
-It should be noted that the technical fields involved here are vast and rapidly evolving.
-The author makes no claim to be a world authority on the entirety of the topics which are touched.
-It is quite possible that there are other effects and phenomena which could be relevant which aren't mentioned here, and it's also true that additional advances in the state of the art are being made continually, and might not be known about in public literature.
-
-However, the nature of scientific discovery and technical progress is that many of best discoveries are actually built-upon and confirmed, rather than swept aside, by later developments.
-As such, plenty of foundational principals, in particular the conservation laws of energy and momentum and essential principals of mechanics and thermodynamics are just as true, right, and relevant as they were hundreds of years ago.
-Consequently, whilst we cannot always know the precise degree of advance in a particular field of engineering, we can still confidently know some of the fundamental bounds and constraints that will be inherent in that field.
-
-In this paper, since we cannot know the precise details of the particular state of the art in largely classified military programs, I will attempt to be particularly clear about those things for which we can have great confidence (often broad principals or relationships which determine trade-offs) and those things which we can't (the precise degrees or points and sensitivities).
-To that end, in this paper I've underaken modelling based upon the broad principals which we can have confidence, and conducted sensitivity testing based on a range of plausible values which seem realistic for those things about which we can have less confidence.
-I have also, in discussion with experts and through reading the available literature, attempted where possible to identify the most plausible of the uncertain values, in order to advance discussion.
-
-With the support of the principal sponsor of this paper, the relevant models may be found in a user-friendly format as a web-app to allow their further interrogation and testing for other plausible scenarios.
-In addition, for increased transparency, the relevant code and equations underlying are available on github.
-
-# Executive Summary
-
-#### The difference between nuclear and conventional for speed
-This paper was commissioned in order to investigate whether pumpjets could plausibly be as efficient, or more efficient than a suitably designed open propeller for a conventional submarine.
-A crucial input to the investigation is a rough understanding of what the speeds of operation are likely to be for a submarine.
-Whereas nuclear submarines are reportedly capable of reaching speeds in excess of 30kt, and might transit long distances at such high speeds, conventional sumbarines are not thought of as being able to reach speeds far above 20kt in a sprint, and can only sustain speeds of 8-10kt for long-distance transits.
-Moreover, on patrol, a large portion of thier work is done at very low speeds, typically thought to be in the range of 2-4kt.
-
-#### The importance of low speed operation for diesel-electric submarines
-The efficiency of the propulsion system at such low speeds is of great significance for a conventional submarine, since it must rely on batteries or other air-independent propulsion sources for power when entirely submerged.
-Consequently, excessive energy consumption results in greatly reduced dived endurances.
-Since a submarine's position is vastly more likely to be discovered when it is on the surface operating its diesel engines, this ability to remain submerged for a long time is crucially important for combat operations, and in transiting through sensitive or contested areas.
-For nuclear submarines, which posess a practically infinite supply of energy from the nuclear reactor, efficiency at low speeds is of no concern.
-In fact, dispersing additional energy, (provided it can be done quietly) is probably advantageous for a nuclear submarine, since it will allow the nuclear reactor to avoid running at very low power levels, where the stability of the reactor is reduced.
-
-#### What is cavitation
-Pumpjets have been widely adopted by navies operating nuclear submarines.
-The principal advantage of a pumpjet relevant to submarines is their ability to avoid problems associated with cavitation, which is known to occur for propellers attempting to operate at high speeds.
-Cavitation, which is the rapid expansion and collapse of a bubble or void in the water, is particularly problematic for submarines, since it results in the creation of a great deal of noise which could be detected by an enemy.
-
-#### How does a pump-jet work
-A fundamental requirement of a pump-jet in order to avoid this cavitation is that the the working parts of the jet (the rotating blades inside it, known as the impeller) operate at a higher pressure than propeller blades operating in open water would.
-This means that the blades can turn at a lower speed relative to the water they are connected with, allowing a less violent action, which induces less cavitation.
-In this way, the jet does it's work less by directly acelerating the water, but by raising its pressure.
-This raised pressure is converted back to movement, which produces thrust, as it exits the jet and returns to the same pressure as the surrounding environment.
-
-#### The necessity of drag, and decelleration, induced by the duct
-The role of the shroud (or tunnel, mantel, duct) around the jet is to allow the pressure to be raised around the impeller, in a way that is not possible for an open propeller.
-By a fundamental requirements of physics, this actually requires that the water's incoming speed be **reduced** before it reaches the impeller.
-(The kinetic energy embodied in movement is converted into potential energy, or pressure.)  Whilst the duct narrowing at the nozzle also necessarily accelerates the water, (as the additinal pressure imparted by the impeller is converted back to kinetic energy) it's an essential feature of all pump-jets that the water flow is decelerated at the point of reaching the impeller.
-Consequently, an elementary form of the pump-jet is also termed the 'decelerating duct' applied to a propeller.
-Put simply, the water has to slow down to go fast again.
-
-#### What happens to a jet at low speed, in simple terms
-The problems for pump-jets arise when the water is already going slow, and you don't want it to go that much faster.
-This is the case when a jet designed for high-speed attempt to operate at a dramatically reduced speed.
-(Not to be confused with a vessel with very little water speed working very hard, as might be the case for a tug or barge.)  In this case, the slowing down and speeding up results in an unnecssary additional step which reduces greatly the efficiency of propulsive system.
-Or, put another way, overcomign the resistance of moving water through the shroud becomes much greater relative to the total thrus produced by the jet.
-
-#### Why this isn't easily noticeable in normal circumstances
-It is for this reason that waterjets of any kind are known to have an efficiency curve that falls off towards zero as the net thrust they produce also diminishes to zero (which for a given vessel will correspond to water speed).
-This doesn't mean that they don't still work at low speeds and produce some thrust.
-All it means is that far more power per unit of thrust will be required than might be at other speeds, as a higher fraction of energy is expended producing the turblence (random, round-and-round movement) inside and around the jet shroud than goes towards direct front-to-back acceleration of the water column, which produces thrust.
-
-#### What the submarine requirement tells us about the duct
-Whist estimating the exact shape and level of the efficiency curve for a particular pump-jet and propeller is impossible without detailed knowledge of their design, the over all trends of their shapes in the extremes can be known from well-established principals.
-Moreover, the particular demands of a pump-jet suited to a nuclear submarine, (eliminating all cavitation in as wide a range of operating circumstances as possible) considerably narrows the plausible range concerned.
-In order to minimise cavitation, the degree of pressure elvation at the impeller would be relatively high, or the total surfaces of the impeller blades much aslo become larger.
-Both of these design requirements necessitate changes to the duct or blade designs which would be in tension with overall efficiency, and most pronounced at the lowest of operating speeds.
-
-#### Quantitative Conclusions
-In the production of this paper I have developed a computational model which maps the impact of different efficiency curves directly to the dived range and endurance of a submarine.
-The most plausible scenarios I find include the reduction of dived endrance and range between 20% and 50%, or effectively halving time and distance that submarine may remain submerged for during combat operations at speeds around 3-5kt.
-
-#### The acoustic advantage of jets at higher speed
-My review of the literature confirms that a pumpjet may produce a significant acoustic advantage in circumstances where an open propeller would experience any degree of cavitation.
-Indeed, it has been remarked by naval researchers that pump-jets could be designed which would not cavitate past the point when the body they propel experienced caviation.
-As such, it seems perfectly plausible that pump-jets confer one advantage on a submarine, in that they can accelerate to a higher speed without cavitation occurring.
-This is known as a higher 'tactical silent speed'.
-
-#### Turbulence and flow separation at low speed
-However, at much lower speeds such as patrol speeds (2-4kt) it is most likely that a propeller will be able to operate well below the point of any cavitation inception, and would consequently also be extremely quiet.
-Moreover, a jet will necessarily incur substantially larger degrees and types of turbulence in order to produce net thrust in this regime.
-These likely include discontinuities and instability in the flow entering the duct and passing through the impeller and stator (flow separation) as the water is acelerated sufficient to be slowed, and then re-accelerated.
-In certain circumstances where resonances might arise these could be highly adverse to acoustic performance.
-However, assuming that by careful design such resonances can be eliminated (which we assume has been achieved) these effects would not result in any cavitation, and would only increase the noise attributable to turbulence in solid water, which is far less than for any cavitation.
-
-#### The acoustic question at low speed
-But in either case, given the necessarily raised levels of turblence generated by a jet than a propeller at very low speeds, the claim that the jet is quieter in this regime must rely entirely upon shielding effects from the shroud.
-These could be substantial in a directions perpendicular to the direction of travel, but would be much smaller when viewed from the aft or forward directions.
-It should also be noted that many underwater acoustic environments, sound tennds to reflect and bend in different directions as it propagates.
-As such, the claim that pump-jets are universally quieter than propellers should be treated with caution.
-In a variety of circumstances, including most operations at patrol speed, it may not always be true.
-
-#### The inevitable trade-off
-However, as a direct and necessary result of this higher tactical silent speed, some disadvantages will be incurred, owing to the additional drag induced by the shroud which prevents cavitation at the impeller.
-These are a lower dived endurance, a lower dived range, a lower overall endurance, a lower overall range, and a worse indescretion ratio.
-
-#### The contradiction between stated requirement and chosen technology
-Given the substantial emphasis which was placed on overall range and endurance of the submarine, it is difficult to understand how one particular unique requirement would be elevated so high above the other strategic and tactical advantages afforded by propellers.
-
-# Speed and Drag - Why very slow is very very $(very)^2$ economical
-
-Perhaps the most important relationship to understand is the relationship between speed and drag as it pertains to submarines.
-This is immportant because it sets out the fundamental framework as it applies to any submarine, regardless of propulsion type.
-(It's also a pretty important for planes, cars, missiles, torpedoes, and basically everything else.)
 
@@
 
-It's crucial to understand, however, that drag is only a force, and doesn't directly inform us about how much energy is consumed, until we multiply it by the **distance** over which it is applied, not the time for which it is applied.
-To think about it simply, gravity exerts a force on you downwards all the time.
-But you don't expend any energy overcoming it when sitting still.
-If you climb stairs, the amount of gravitational potential you attain depends on how high up you climb, not how long you spend on the ladder or stair-case.
-
-This has a significant consequence for propulsion, since the amount of power (energy expended in a given time) required for thrust scales with the drag, multiplied by the distance covered in a given time.
-As such, power required (and fuel/battery consumption) scales with velocity **cubed** rather than velocity squared.
-This means that if you double your speed of travel, fuel consumption for a given period of time will increase by a factor of eight.
-Due to the increased speed of travel, the fuel required to cover a given distance will only quadruple.
-Hence the range that can be covered with a given amount of energy (assuming propulsion dominates energy requirements) tends to scale with the inverse of velocity squared, whereas the endurance (amount of time that can be spent travelling) scales with the inverse of velocity cubed.
-
-This has a profound impact on the operation and engineering of maritime vessels.
-Reducing velocity has such a substantial reduction on hull drag that going slower is almost always a reasonable means of conserving fuel overall.
-To the extent that time is non-critical, slower is always much much better.
-It's for this reason that during the financial crisis, many cargo shipping companies adopted the practice of slow steaming [@liang2014] in order to conserve fuel, despite this causing a range of possible new engineering issues which need to be accounted for in order to operate the engines at lower than normal power for sustained periods [@sanguri2012 pp.
-8-10].
-The fuel saving from operating even 30% slower means that around half the total energy is required for a given journey.
-Other incremental costs and inefficiences from operating 'off-design' are frequently outweight by such a dramatic reduction in overall power demand.
-
-This effect is of profound importance to understanding the operations of conventional submarines, which have extremely constrained energy stores when operating under the surface.
-Diesel fuel when burned has an energy density of approximately 45MJ/kg.
-In contrast, a lead-acid battery might have energy densities in the range of 0.08-0.14MJ/kg.
-With something like 400 times as much energy per kg embarked in diesel, operating on batteries imposes an extreme demand for economy on propulsive power.
-Happily for submarines, the square law for drag, and cubed law for power, allow an almost commensurate reduction in power demands to take place by slowing down to very slow speeds when submerged.
-Hypothetically, a conventional submarine might have a total range of 10,000nm from using it's diesel payload at 8kt.
-If an equal weight of lead-acid batteries as fuel were carried, the submarine could only travel about 25nm when submerged at the same speed in a single charge.
-But by travelling at 4kt, that quickly increases to 100nm, and 400nm at 2kt (neglecting hotel load here for simplicity).
-
-It's worthwhile pointing out that extreme demand for economy imposed by the poor energy density of batteries distinguishes submarines quite remarkably from other types of boat or ship design.
-In every other application, the propulsion system is designed around a particular speed and loading condition which it is optimally efficient for, and a band of plausible variation around this.
-For example, a cargo ship designed for operation at 22kt will have it's propeller and engine (propulsion system) very carefully designed around these speeds and the plausible levels of loading at those speeds.
-A top-speed which might be somewhat higher than 22kt would be calculated, but would be relatively unlikely to be used for any purpose other than an emergency.
-Lower speeds might be considered 15-18kt for the purposes of slow-steaming.
-However, the relative efficiency at 3kt is of little concern, since the drag at that speed will be under 2% of what it would be at 22kt, and the power required around 0.25%.
-Even if the efficiency was substantially worse, or better (by a factor of two or three even) the impact on the overall economics of the operation would be negligible when compared to marginal improvements at the design speed.
-
-The same consideration applies to other types of vessels, which might use water-jets.
-Fast pilot-boats, for instance, might spend quite a considerable time manouvering in and out of harbour or alongside at very low waterspeeds.
-However, their overall fuel consumption is still likely to be dominated by the fast section of their trip to a ship.
-If the in-harbour water-speed was 8kt, but 32kt was the optimum speed for the open-water section of their journey, the fuel requirement for a given distance could still be over ten times greater at 32kt.
-(Aside, the physics described above would say 16 times, but some reduction due to the reduction of surface area from planing might offset this.)  Consequently the owner/operator would be at least ten times as concerned about efficiency at 32kt as at 8kt, (unless the harbour transit was much much longer than the open water section), and have negligible concern about effiencies at 2kt or 3kt, which would rarely amount to more than a couple of percent of total propulsion energy demand.
-
-In stark contrast, a submarine operator might have less than 1% of the energy store available for the submerged movements at 2-3kt, and consequently be substantially **more** concerned about efficiency in this regime, particularly since this would be the regime in which all of their combat operations would be conducted.
-
-This imposes a constraint on the way in which literature on jets, propellers, and propulsion systems in general is read, since almost all of it is deeply concerned with how a particular system would operate at one particular design speed, and choosing the optimum system for that design speed and load.
-Considerable attention is given to how a propeller would work in off-design conditions relating to changes in load, such as in cases when a ship is more lightly laden, or a tug-boat is pushing a different sized ship.
-There is, however, a relative scarcity of attention given to the performance of a propulsion system operating at a dramatically different speed, with the same load.
-It also corresponds to scenarios where engineers and manufacturers seem from discussion to be far less confident in making quantitative claims about the performance of their own systems.
-(They are perfecly happy to make qualitative claims, "it would be alright", "nothing very bad would happen" etc.)
-
-Given that submarines almost by definition (since their hull shape doesn't change, and they must carry ballast to make up for any under-loading) are essentially always operating with the same load, but have to operate over an extremely wide variety of different speeds (a factor of four or more in variance) with severe efficiency concern at all of them, they represent a truely unique engineering question, seldom discussed in commercial applications.
-Consequently, there are relevantly few pieces of literature which address the present question directly.
-(Happily, not zero)  However, there are plenty of instances where the relevant physics and trade-offs are explained in depth to address related, but not identical questions.
-
-# The difference between nuclear and conventional propulsion
-
-Nuclear submarines vary quite remarkably from conventional submarines because of the means by which they generate their power.
-Because of the extremely high energy density in enriched uranium or plutonium, the reactors on board nuclear submarines generate an abundant supply of energy.
-Most submarine reactors are reportedly capable of generating between 25 to 50 megawatts (MW) of power, though Russian submarines have hundreds of megawatts of power available [@WNA2017].
-
-In contrast, the Collins Class Submarine's main motor is rated at less than 6MW, with designs for the future submarine appearing to be only slightly larger [@patrick2012].
-It is fair to say that in terms of maximum power ouput, nuclear submarines could have something like 5-10 times as much power at their disposal, and their **peak** output.
-
-The difference between the peak-powers of the submarines significantly understates how different the designs are, because the difference in the total amount of energy stored and available for use in a voyage or dive is vastly greater.
-A nuclear submarine might have literally millions of times more energy at its disposal, it is practically unlimited for all intents and purposes on a given voyage.
-Consequently, wheras a nuclear submarine might regularly conduct transits at or near it's peak power, a diesel electric submarine would probably transit at less than half the speed which it could manage at a sprint, (and use less than a quarter of the energy, as discussed earlier) and on patrol it might be operating at a tenth of the maximum speed, and use maybe just 1% of as much power on propulsion.
-In this situation, the amount of power that is drawn for lights, CO2 scrubbing, washing, cooking, heating, as well as the electronics driving the combat systems (the 'Hotel Load') might well become significant, and even be as larger or larger than what is required for propulsion.
-
-Consequently, the difference between the power output a conventional and a nuclear submarine might regularly operate at could be even quite a bit larger than the maximum amount of energy that they can deliver to their propulsion systems.
-
-Perhaps more significantly for power-design of nuclear submarines is a little-discussed phenomenon called Xenon poisoning, which affects nuclear reactors when they shut down or lower their power significantly.
-When Uranium or Plutonium atoms split (or fission) into two smaller isotopes atoms, a variety of different radio nucleides (unstable variants of atoms) are produced.
-Two of these are Xenon-135 and Iodine 135.
-Iodine is produced much more often, and decays into Xenon-135, which a half-life of about nine hours.
-This Xenon has a very special, perhaps unique role in reactors, since they very easily captures the free neutrons which cause the continued fission reactions in Uranium or Plutonium which drive the reactor.
-As such, Xenon is known as reactor 'poison' since it can kill the reactor's reactivity in very high doses.
-
-This high neutron-capture from Xenon means that it has a duplicitous relationship with the reactor's power level.
-At high levels of reactor power, lots of Xenon-135 and Iodine-135 are produced by the fission process.
-The high presence of Xenon reduced the reactor's reactivity.
-On the other hand, there are lots of neutrons available to 'burn off' the Xenon (which absobs the neturons to become a different isotope).
-This keeps the total level of Xenon in check.
-
 The situation becomes much more complicated when the reactor undergoes a sudden change in power output.
@@
 Unless the reactor is quickly raised back to relatively high power (~60%) quite quickly (an hour or less), the Xenon levels become so high that the reactor will have to be shut down, otherwise extreme (and dangerous) measures would be required in order to keep the reactor going.
-(This is essentially what lead to the Chernobyl Explosion [@WNA2009].)  A fuller discussion of Xenon poisoning can be found in [@garland2005], which demonstrates key concepts related to the poisoning effect shown below.
+(This is essentially what lead to the Chernobyl Explosion [@WNA2009].)  A fuller discussion of Xenon poisoning can be found in [@garland2005], which demonstrates key concepts related to the poisoning effect shown in Figure  x.
 
-![Xenon poisoning effect following shutdown from @garland2005](/Users/aidanmorrison/submarines/XenonPoison.png){width=500px}
+![Xenon poisoning effect following shutdown from @garland2005](XenonPoison.png){width=500px}
+
+`r fig$cap("inline_cap", "Here is a sample inline caption", center=TRUE, inline=TRUE)`
+
+<!-- \begin{figure} -->
+<!-- \includegraphics[width=0.9\textwidth]{XenonPoison.png} -->
+<!-- \caption{Xenon poisoning effect following shutdown}   -->
+<!-- \label{fig:xenon} -->
+<!-- \end{figure} -->
+<!-- \ref{fig:xenon} -->
+<!-- \pageref{fig:xenon}  -->
+
+<!-- ![Xenon poisoning effect following shutdown from @garland2005](XenonPoison.png){width=500px} -->
 
@@
 Obviously this is never desirable for a military vessel, and hence is avoided at almost all costs.
-
-This effect has a dramatic impact on nuclear submarine design, since much of their design is oriented around being able to comfortably disperse large amounts of excess power, rather than with conserving it.
-For the propulsion system, this actually makes having an inefficient propulsor at low speeds a considerable advantage.
-Since the reactor will likely need to dispose of excess power, particularly during ramp-down, an inefficient propulsion system actually provides a useful power sink.
-Since any excess power will have to be disposed of by some other means (normally by pumping more water to remove the power as the heat) inefficiency at low speeds has no penalty, and probably a marginal benefit, since it will reduce overall demand for additional systems.
-Provided the excess turbulence inside the pumpjet isn't too noisy, wasting energy through the propulsor is useful.
-
-It should also be noted that a nuclear reactor's aversion to sudden reductions in power would also have a substantial impact on the design of a submarine's combat system, and it's demand on the Hotel Load.
-For the same reason, a high Hotel Load, or power-hungry Combat System, could actually be advantageous, as it helps to set an elevated 'floor' for power requirements, reducing the scale of fluctuations in overall power demand from the reactor due to changes in propulsion speed.
-
-# Some essential concepts
-
-### Conservation of Energy
-
-This is perhaps one of the most fundamental and well-established principals in physics.
-The essential idea is that energy can move or change in form, but it isn't ever created or destroyed.
-Machines, plants and animals all derive their energy from a particular other source, which can be measured and evalutated to establish the limits of energy available.
-Plants collect energy from sunlight falling on their leaves.
-Humans (as well as combustion engines) capture the chemical potential energy in organic matter, and release it by combining it with oxygen.
-Hydro-electric power plants turn the gravitational potential energy of water stored at a height into electricity.
-
-Conservation of energy has a particularly relevant embodiment in fluid flows, which is given it's primary expression in Bernoullis equation.
-It says that the total energy in a connected body of fluid is constant, though it can change in form between kinetic energy (movement), gravitational potential energy (it being elevated) the heat energy in the fluid, and the pressure of the fluid.
-
-In different fluids, the dynamics of how energy moves between one and another change.
-For example, in gases, heating up a confined piece of gas will increase it's pressure, or if it is unconfined, increase its volume.
-This is particularly important for understanding gas turbines.
-In the case of a liquid, however, all the molecules are in close contact, and hence can't increase in volume or pressure substantially except by the creation of steam.
-As such, in the absence of large amounts of cavitation, the terms in the relationship which are most important for our consideration are the relationships between pressure, and kinetic energy, and gravity.
-
-In the case of most waterjets which propel surface vessels, the water is lifted from the bottom of the hull at the intake to the pump, which is generally incorporated inside the hull.
-This increase in gravitational potential coincides with a slowing down of the water relative to the vessel-speed.
-Since the jet in such vessels is generally ejected at the same height as the pump, this potential is never regained, and is technically a loss, however at high speeds such a loss is small relative to the total power ouput.
-
-#### Venturi Effect and Flow Diffusion
-In the case of submarines and torpedoes the water doesn't undertake a change in height, since intake and nozzle are generally all in line with the central axis of the submarine or torpedo.
-As a consequence, the key relationship in Bernoullis equations is the relationship between the liquids velocity, and its pressure.
-The consequence is that when water moves through a pipe (duct, or shroud) it's pressure is inversely related to the square of its velocity.
-This means that when a liquid is forced to to travel through a narrowing pipe, it's pressure necessarily decreases as the velocity increases.
-This is a simple embodiment of the Venturi effect.
-The inverse process is where a pipe increases in volume, and the flow is forced to slow down in order to fill the wider area, and the pressure correspondingly increases.
-This is a process called 'diffusion', and is important to achieving high pressure levels in many types of water pumps, including those which will be particularly relevant for pumpjets for watercraft propulsion [@hamilton1997].
-
-#### Energy waste and efficiency
-The conservation of energy also has useful implications for how the efficiency of systems is thought about.
-In particular, because energy is conserved, identifying inefficiencies in a system necessarily involves identifying where energy goes to doing tasks which aren't useful for the intended purpose.
-A perfectly efficient system won't do any work that isn't for the intended purpose.
-In the case of analysing the efficiency of propulsion systems, the relevant 'work' is almost always related to moving water backwards to produce thrust.
-Moving water in directions other than backwards, including random turbulent flows which wind producing heat rather than thust, are two examples of wasted energy.
-Noise in the water also reflects energy which is wasted.
-
-### Boundary Layer
-When water or any fluid flows with some speed relative to another solid surface nearby, there is some layer adjacent to the surface in which the speed of the fluid is diminished relative to the main flow.
-At a microscopic level, there are some molecules of the fluid on the surface which will be effectively static relative to the surface.
-The layer of fluid that joins the gap between the static surface, and the part of the flow which is moving at the full flow speed, is called the boundary layer.
-Exactly how thick the boundary layer is, and how the fluid moves in the boundary layer, is extremely important for consideration of efficiency of fluid flows over and around solid surfaces.
-
-### Turbulent and Laminar Flow
-
-### Pressure, Thrust, and Momentum
-
-### Cavitation
-
-### Flow separation
-
-# What tech are we talking about? The core relationships between pumps, jets, ducts, and propellers, and all the different things we call them, their parts, and combinations
-
-# Propulsive performance of the pumpjet, or decelerating ducted propeller
-
-## Efficiency Relationships, and various curves
-
-## Practical and recent evidence of efficiency relationships
-
-# Results of numerical simulation
-
-# Acoustic Considerations
-
-# Conclusions
-
-# References

### 2017-12-27 DraftExec.Rmd
stash

--- 2017-12-18 DraftExec.Rmd
@@
 knitr::opts_chunk$set(echo = TRUE)
+library(ggplot2)
+library(knitr)
+
 ```
@@
 $$F_D = \frac{1}{2}\rho v^2C_DA$$
+
+\begin{equation} \label{eq:1}
+\sum_{i=0}^{\infty} a_i x^i
+\end{equation}
 
@@
 The high presence of Xenon reduced the reactor's reactivity.
-On the other hand, there are lots of neutrons available to 'burn off' the Xenon (which absobs the neturons to become a different isotope).
+On the other hand, there are lots of neutrons available to 'burn off' the ass Xenon (which absobs the neturons to become a different isotope).
 This keeps the total level of Xenon in check.
@@
 Unless the reactor is quickly raised back to relatively high power (~60%) quite quickly (an hour or less), the Xenon levels become so high that the reactor will have to be shut down, otherwise extreme (and dangerous) measures would be required in order to keep the reactor going.
-(This is essentially what lead to the Chernobyl Explosion [@WNA2009].)  A fuller discussion of Xenon poisoning can be found in [@garland2005], which demonstrates key concepts related to the poisoning effect shown below.
+(This is essentially what lead to the Chernobyl Explosion [@WNA2009].)  A fuller discussion of Xenon poisoning can be found in [@garland2005], which demonstrates key concepts related to the poisoning effect shown in \ref{fig:ex2}.
+And you can see that in the [Introduction].
 
-![Xenon poisoning effect following shutdown from @garland2005](/Users/aidanmorrison/submarines/XenonPoison.png){width=500px}
+The situation becomes much more complicated when the reactor undergoes a sudden change in power output.
+
+\begin{figure}
+\includegraphics[width=0.9\textwidth]{XenonPoison.png}
+\caption{A prototype of the Job Information dialog  }
+\label{fig:ex2}
+\end{figure}
 
@@
 Obviously this is never desirable for a military vessel, and hence is avoided at almost all costs.
+
+\begin{figure}
+```{r, echo=F}
+kable(head(mtcars))
+```
+\caption{A table example}
+\label{fig:tbl1}
+\end{figure}
 

### 2017-12-27 test.Rmd
stash

@@
 ---
-title: "Test"
-author: "R"
-date: "11/12/2017"
+title: "test"
+author: "Ross"
+date: "19/12/2017"
 output: pdf_document
@@
 
----
-references:
-
-- id: garland2005
-  title: Elementary Physics of Reactor Control Module - Fission Product Poisoning
-  author:
-  - family: Garland
-    given: Bill
-  volume: 3
-  URL: 'http://www.nuceng.ca/br_space/2017-09_4d03_6d03/misc_files/xenon_poisoning.pdf'
-  publisher: Nuclear Engineering (Originally from McMaster University)
-  type: article-journal
-  issued:
-    year: 2005
----
-
 ```{r setup, include=FALSE}
-library(knitr)
 knitr::opts_chunk$set(echo = TRUE)
-
-## This chucnk will read through *this* Rmd file, and attempt to extract all of the 
-## labels (not caption text) used for Figure captions. These labels are used
-## as anchors, so scanning through the document now will allow us to create cross references
-## before the caption actually appears.
-
-## Get the name of this Rmd file
-rmdFn <- knitr::current_input()  # filename of input document
-
-## Read lines and close connection
-rmdCon <- file(rmdFn, open = "r")
-rmdLines <- readLines(rmdCon)
-close(rmdCon)
-
-## Pull out all occurences of at least one back tick, followed 
-## by any number of characters, followed by fig$cap (all on one line)
-figscap_idx <- grep("`+(.*)fig\\$cap", rmdLines)
-rmdLines <- rmdLines[figscap_idx]
-
-## Get rid of everything up until the start of the caption label
-## This presumes the caption label is the first argument of fig$cap()
-## E.g., fig.cap = fig$cap("my_label", ...)
-rmdLinesSansPre <- sub("(.*)fig\\$cap(.*?)[\"']", "", rmdLines)
-
-## Identify everything up until the first quote
-match_data <- regexpr("(.*?)[\"']", rmdLinesSansPre)
-
-## Reduce the length by one, because we're not interested in the final quote
-attr(match_data, "match.length") <- attr(match_data, "match.length") - 1
-
-## Extract
-fig_labels <- regmatches(rmdLinesSansPre, match_data, invert=FALSE)
-
-if (length(fig_labels) > 0) {
-
-    ## Test for duplicates
-    if (anyDuplicated(fig_labels) > 0) stop("Duplicate caption labels detected")
-
-    ## Create a named list of Figure numbers
-    ref <- as.list(1:length(fig_labels))
-    names(ref) <- fig_labels
-}
-
 ```
 
-# Contents
+\documentclass[12pt, english, oneside]{amsart}
+\begin{document}
 
-# Introduction
+Figure \ref{fig:plot} doesn't have it's label.
 
-Drag is the resistance that a fluid (air or water, in our case) gives to a body that is passing through it.
-Quite simply, it's a force that acts in the opposite direction.
-There are multiple sources of drag for different types of scenarios.
-For scenarios where an object is in contact with two different types of fluid (like a ship, on the ocean) or when the fluid doesn't really have contact with all of the object (supersonic flight, and supercavitating torpedoes) some more complex physics applies.
-For a fuller discussion of types of drag, see [@carlton2007]  But in the case of a submarine, which does it's business completely immersed in the ocean, the relevant physics is dominated the skin friction on the hull, which follows a very simple rule and relationship.
-The amount of drag ($F_D$) an objet experiences increases directly in proportion to the surface area $A$.
-For any given object of a certain (unchanging shape) there will be a constant coefficent ($C_D$) which reflects how aerodynamic or hydrodymanic the shape is.
-The drag is also directly proportional to the density of the fluid being moved through, $\rho$.
-(Air creates roughly one-thousandth the drag as water does on any given object at a given speed, since it's roughly one thousand times less dense.)
+<<plot>>=
+plot(x=0, y=0)
+@
 
-But the most sensitive factor in this relationship is the speed at which the object moves through the liquid.
-The drag increases not with the speed ($v$), but with the **square** of the speed .
-This means if the speed doubles, the drag increases by four.
-If the speed triples, the drag increases by a factor of nine.
+Figure \ref{fig:plot2} has its label.
 
-$$F_D = \frac{1}{2}\rho v^2C_DA$$
+\begin{figure}
+\caption{\label{fig:plot2}}
+<<>>=
+plot(x=1,y=1)
+@
+\end{figure}
 
-(Technical aside: This law applies wherever the the flow over the surface is turbulent.
-It is true that for very small objects, or very viscous fluids, or very slow movements a different apples called Stokes Equation, in the case wher Reynolds numbers are less than 1.
-Given that sea-water is not particularly viscous, and sumbarines not particularly small, Reynolds numbers are liekly to be much much greater than 1 (one or two thousand), even when moving at only one or two knots.
-Since it is unlikely that a significant proportion of the flow over the hull will be laminar, we'll use the drag equation in all modeling going forward when considering drag on the hull).
-
-The situation becomes much more complicated when the reactor undergoes a sudden change in power output.
-If the power is lowered dramatically and suddenly, the production of Xenon continues quite rapidly for some time due to the decay of the large stock of Iodine-135.
-With less neutron flux available to 'burn off' the Xenon, the Xenon levels spike, and push down the reactor's reactivity.
-Unless the reactor is quickly raised back to relatively high power (~60%) quite quickly (an hour or less), the Xenon levels become so high that the reactor will have to be shut down, otherwise extreme (and dangerous) measures would be required in order to keep the reactor going.
-(This is essentially what lead to the Chernobyl Explosion [@WNA2009].)  A fuller discussion of Xenon poisoning can be found in [@garland2005], which demonstrates key concepts related to the poisoning effect shown in Figure  x.
-
-![Xenon poisoning effect following shutdown from @garland2005](XenonPoison.png){width=500px}
-
-`r fig$cap("inline_cap", "Here is a sample inline caption", center=TRUE, inline=TRUE)`
-
-<!-- \begin{figure} -->
-<!-- \includegraphics[width=0.9\textwidth]{XenonPoison.png} -->
-<!-- \caption{Xenon poisoning effect following shutdown}   -->
-<!-- \label{fig:xenon} -->
-<!-- \end{figure} -->
-<!-- \ref{fig:xenon} -->
-<!-- \pageref{fig:xenon}  -->
-
-<!-- ![Xenon poisoning effect following shutdown from @garland2005](XenonPoison.png){width=500px} -->
-
-Consequently, nuclear reactors aren't well suited to rapid fluctuations in power, particularly dramatic reductions in power, as these can lead to instability in the reactor core.
-If the reactor is shut down in order to avoid such dangerous circumstances, it generally cannot be started again until Xenon levels have fallen again, which can take a couple of days.
-Obviously this is never desirable for a military vessel, and hence is avoided at almost all costs.
+\end{document}

### 2017-12-27 test.Rnw
stash

--- 2017-12-18 DraftExec.Rmd
@@
----
-title: "Efficiency"
-author: "Aidan Morrison"
-date: "11/12/2017"
-output: pdf_document
----
+\documentclass{article}
 
----
-references:
-
-- id: garland2005
-  title: Elementary Physics of Reactor Control Module - Fission Product Poisoning
-  author:
-  - family: Garland
-    given: Bill
-  volume: 3
-  URL: 'http://www.nuceng.ca/br_space/2017-09_4d03_6d03/misc_files/xenon_poisoning.pdf'
-  publisher: Nuclear Engineering (Originally from McMaster University)
-  type: article-journal
-  issued:
-    year: 2005
-
-- id: carlton2007
-  title: Marine Propellers and Propulsion
-  author:
-  - family: Carlton
-    given: John
-  URL: 'http://www.nuceng.ca/br_space/2017-09_4d03_6d03/misc_files/xenon_poisoning.pdf'
-  publisher: Elsevier
-  issued:
-    year: 2007
-
-- id: liang2014
-  title: The economics of slow steaming
-  author:
-  - family: Liang
-    given: Lee Hong
-  URL: 'http://www.seatrade-maritime.com/news/americas/the-economics-of-slow-steaming.html'
-  publisher: Seatrade Maritime News
-  issued:
-    year: 2014
-    month: 10
-    day: 7
-
-- id: sanguri2012
-  title: The Guide to Slow Steaming On Ships
-  author:
-  - family: Sanguri
-    given: Mohit
-  URL: 'https://www.marineinsight.com/wp-content/uploads/2013/01/The-guide-to-slow-steaming-on-ships.pdf'
-  publisher: Marine Insight
-  issued:
-    year: 2012
-    month: 12
-
-- id: WNA2017
-  title: Nuclear Powered Ships
-  author:
-  - family: 
-    given: 
-  URL: 'http://www.world-nuclear.org/information-library/non-power-nuclear-applications/transport/nuclear-powered-ships.aspx#ECSArticleLink0'
-  publisher: World Nuclear Association
-  issued:
-    year: 2017
-
-- id: WNA2009
-  title: Chernobyl Accident Appendix 1
-  author:
-  - family: 
-    given: 
-  URL: 'http://www.world-nuclear.org/information-library/safety-and-security/safety-of-plants/appendices/chernobyl-accident-appendix-1-sequence-of-events.aspx'
-  publisher: World Nuclear Association
-  issued:
-    year: 2009
-
-- id: patrick2012
-  title: The Driving Factor in the SEA 1000 choice
-  author:
-  - family: Patrick
-    given: Rex
-  URL: 'https://corporate.siemens.com.au/content/dam/internet/siemens-com-au/root/aunz-defence-solutions/apdr-october-2015-issue-future-submarine.pdf'
-  publisher: Asia-Pacific Defence Reporter
-  issued:
-    year: 2015
-    month: 10
-
-- id: hamilton1997
-  title: Jet Torque
-  author:
-  - family: Waterjets
-    given: Hamilton
-  volume: 8
-  URL: ''
-  publisher: Hamilton Waterjets
-  issued:
-    year: 1997
-    month: 3
----
-
-```{r setup, include=FALSE}
-knitr::opts_chunk$set(echo = TRUE)
-```
-
-# Contents
-
-# Introduction
+\begin{document}
+\SweaveOpts{concordance=TRUE}
 
@@
 
-Necessarily, this topic depends upon a considerable amount of physics, particularly around hydrodynamics, fluid mechanics, and turbomachinery, all of which might be difficult or daunting topics to usefully inform a public policy debate.
-However, in the case of a $50 billion dollar military acquisition, neglecting to engage with the technical matters that are so crucial to such a decision would be deeply foolish.
-To that end, this paper has a very specific ambition.
-It doesn't seek to advance the technical field with any new research or insight.
-Instead, it aims to make as much of the relvant physics as possible to this particular question comprehensible in laymans terms, and connect the essential concepts as directly as possible with the most important conclusions of public import.
+I cannot spel properly
 
-As such, it will tend to make frequent use of diagrams, illustrations, as well as simple prose, and relatively infrequently use mathematical equations, (unless of crucial importance), though some effort is made to point to references where the fuller and more formal derivations of these relationships can be found.
-
-It should be noted that the technical fields involved here are vast and rapidly evolving.
-The author makes no claim to be a world authority on the entirety of the topics which are touched.
-It is quite possible that there are other effects and phenomena which could be relevant which aren't mentioned here, and it's also true that additional advances in the state of the art are being made continually, and might not be known about in public literature.
-
-However, the nature of scientific discovery and technical progress is that many of best discoveries are actually built-upon and confirmed, rather than swept aside, by later developments.
-As such, plenty of foundational principals, in particular the conservation laws of energy and momentum and essential principals of mechanics and thermodynamics are just as true, right, and relevant as they were hundreds of years ago.
-Consequently, whilst we cannot always know the precise degree of advance in a particular field of engineering, we can still confidently know some of the fundamental bounds and constraints that will be inherent in that field.
-
-In this paper, since we cannot know the precise details of the particular state of the art in largely classified military programs, I will attempt to be particularly clear about those things for which we can have great confidence (often broad principals or relationships which determine trade-offs) and those things which we can't (the precise degrees or points and sensitivities).
-To that end, in this paper I've underaken modelling based upon the broad principals which we can have confidence, and conducted sensitivity testing based on a range of plausible values which seem realistic for those things about which we can have less confidence.
-I have also, in discussion with experts and through reading the available literature, attempted where possible to identify the most plausible of the uncertain values, in order to advance discussion.
-
-With the support of the principal sponsor of this paper, the relevant models may be found in a user-friendly format as a web-app to allow their further interrogation and testing for other plausible scenarios.
-In addition, for increased transparency, the relevant code and equations underlying are available on github.
-
-# Executive Summary
-
-#### The difference between nuclear and conventional for speed
-This paper was commissioned in order to investigate whether pumpjets could plausibly be as efficient, or more efficient than a suitably designed open propeller for a conventional submarine.
-A crucial input to the investigation is a rough understanding of what the speeds of operation are likely to be for a submarine.
-Whereas nuclear submarines are reportedly capable of reaching speeds in excess of 30kt, and might transit long distances at such high speeds, conventional sumbarines are not thought of as being able to reach speeds far above 20kt in a sprint, and can only sustain speeds of 8-10kt for long-distance transits.
-Moreover, on patrol, a large portion of thier work is done at very low speeds, typically thought to be in the range of 2-4kt.
-
-#### The importance of low speed operation for diesel-electric submarines
-The efficiency of the propulsion system at such low speeds is of great significance for a conventional submarine, since it must rely on batteries or other air-independent propulsion sources for power when entirely submerged.
-Consequently, excessive energy consumption results in greatly reduced dived endurances.
-Since a submarine's position is vastly more likely to be discovered when it is on the surface operating its diesel engines, this ability to remain submerged for a long time is crucially important for combat operations, and in transiting through sensitive or contested areas.
-For nuclear submarines, which posess a practically infinite supply of energy from the nuclear reactor, efficiency at low speeds is of no concern.
-In fact, dispersing additional energy, (provided it can be done quietly) is probably advantageous for a nuclear submarine, since it will allow the nuclear reactor to avoid running at very low power levels, where the stability of the reactor is reduced.
-
-#### What is cavitation
-Pumpjets have been widely adopted by navies operating nuclear submarines.
-The principal advantage of a pumpjet relevant to submarines is their ability to avoid problems associated with cavitation, which is known to occur for propellers attempting to operate at high speeds.
-Cavitation, which is the rapid expansion and collapse of a bubble or void in the water, is particularly problematic for submarines, since it results in the creation of a great deal of noise which could be detected by an enemy.
-
-#### How does a pump-jet work
-A fundamental requirement of a pump-jet in order to avoid this cavitation is that the the working parts of the jet (the rotating blades inside it, known as the impeller) operate at a higher pressure than propeller blades operating in open water would.
-This means that the blades can turn at a lower speed relative to the water they are connected with, allowing a less violent action, which induces less cavitation.
-In this way, the jet does it's work less by directly acelerating the water, but by raising its pressure.
-This raised pressure is converted back to movement, which produces thrust, as it exits the jet and returns to the same pressure as the surrounding environment.
-
-#### The necessity of drag, and decelleration, induced by the duct
-The role of the shroud (or tunnel, mantel, duct) around the jet is to allow the pressure to be raised around the impeller, in a way that is not possible for an open propeller.
-By a fundamental requirements of physics, this actually requires that the water's incoming speed be **reduced** before it reaches the impeller.
-(The kinetic energy embodied in movement is converted into potential energy, or pressure.)  Whilst the duct narrowing at the nozzle also necessarily accelerates the water, (as the additinal pressure imparted by the impeller is converted back to kinetic energy) it's an essential feature of all pump-jets that the water flow is decelerated at the point of reaching the impeller.
-Consequently, an elementary form of the pump-jet is also termed the 'decelerating duct' applied to a propeller.
-Put simply, the water has to slow down to go fast again.
-
-#### What happens to a jet at low speed, in simple terms
-The problems for pump-jets arise when the water is already going slow, and you don't want it to go that much faster.
-This is the case when a jet designed for high-speed attempt to operate at a dramatically reduced speed.
-(Not to be confused with a vessel with very little water speed working very hard, as might be the case for a tug or barge.)  In this case, the slowing down and speeding up results in an unnecssary additional step which reduces greatly the efficiency of propulsive system.
-Or, put another way, overcomign the resistance of moving water through the shroud becomes much greater relative to the total thrus produced by the jet.
-
-#### Why this isn't easily noticeable in normal circumstances
-It is for this reason that waterjets of any kind are known to have an efficiency curve that falls off towards zero as the net thrust they produce also diminishes to zero (which for a given vessel will correspond to water speed).
-This doesn't mean that they don't still work at low speeds and produce some thrust.
-All it means is that far more power per unit of thrust will be required than might be at other speeds, as a higher fraction of energy is expended producing the turblence (random, round-and-round movement) inside and around the jet shroud than goes towards direct front-to-back acceleration of the water column, which produces thrust.
-
-#### What the submarine requirement tells us about the duct
-Whist estimating the exact shape and level of the efficiency curve for a particular pump-jet and propeller is impossible without detailed knowledge of their design, the over all trends of their shapes in the extremes can be known from well-established principals.
-Moreover, the particular demands of a pump-jet suited to a nuclear submarine, (eliminating all cavitation in as wide a range of operating circumstances as possible) considerably narrows the plausible range concerned.
-In order to minimise cavitation, the degree of pressure elvation at the impeller would be relatively high, or the total surfaces of the impeller blades much aslo become larger.
-Both of these design requirements necessitate changes to the duct or blade designs which would be in tension with overall efficiency, and most pronounced at the lowest of operating speeds.
-
-#### Quantitative Conclusions
-In the production of this paper I have developed a computational model which maps the impact of different efficiency curves directly to the dived range and endurance of a submarine.
-The most plausible scenarios I find include the reduction of dived endrance and range between 20% and 50%, or effectively halving time and distance that submarine may remain submerged for during combat operations at speeds around 3-5kt.
-
-#### The acoustic advantage of jets at higher speed
-My review of the literature confirms that a pumpjet may produce a significant acoustic advantage in circumstances where an open propeller would experience any degree of cavitation.
-Indeed, it has been remarked by naval researchers that pump-jets could be designed which would not cavitate past the point when the body they propel experienced caviation.
-As such, it seems perfectly plausible that pump-jets confer one advantage on a submarine, in that they can accelerate to a higher speed without cavitation occurring.
-This is known as a higher 'tactical silent speed'.
-
-#### Turbulence and flow separation at low speed
-However, at much lower speeds such as patrol speeds (2-4kt) it is most likely that a propeller will be able to operate well below the point of any cavitation inception, and would consequently also be extremely quiet.
-Moreover, a jet will necessarily incur substantially larger degrees and types of turbulence in order to produce net thrust in this regime.
-These likely include discontinuities and instability in the flow entering the duct and passing through the impeller and stator (flow separation) as the water is acelerated sufficient to be slowed, and then re-accelerated.
-In certain circumstances where resonances might arise these could be highly adverse to acoustic performance.
-However, assuming that by careful design such resonances can be eliminated (which we assume has been achieved) these effects would not result in any cavitation, and would only increase the noise attributable to turbulence in solid water, which is far less than for any cavitation.
-
-#### The acoustic question at low speed
-But in either case, given the necessarily raised levels of turblence generated by a jet than a propeller at very low speeds, the claim that the jet is quieter in this regime must rely entirely upon shielding effects from the shroud.
-These could be substantial in a directions perpendicular to the direction of travel, but would be much smaller when viewed from the aft or forward directions.
-It should also be noted that many underwater acoustic environments, sound tennds to reflect and bend in different directions as it propagates.
-As such, the claim that pump-jets are universally quieter than propellers should be treated with caution.
-In a variety of circumstances, including most operations at patrol speed, it may not always be true.
-
-#### The inevitable trade-off
-However, as a direct and necessary result of this higher tactical silent speed, some disadvantages will be incurred, owing to the additional drag induced by the shroud which prevents cavitation at the impeller.
-These are a lower dived endurance, a lower dived range, a lower overall endurance, a lower overall range, and a worse indescretion ratio.
-
-#### The contradiction between stated requirement and chosen technology
-Given the substantial emphasis which was placed on overall range and endurance of the submarine, it is difficult to understand how one particular unique requirement would be elevated so high above the other strategic and tactical advantages afforded by propellers.
-
-# Speed and Drag - Why very slow is very very $(very)^2$ economical
-
-Perhaps the most important relationship to understand is the relationship between speed and drag as it pertains to submarines.
-This is immportant because it sets out the fundamental framework as it applies to any submarine, regardless of propulsion type.
-(It's also a pretty important for planes, cars, missiles, torpedoes, and basically everything else.)
-
-Drag is the resistance that a fluid (air or water, in our case) gives to a body that is passing through it.
-Quite simply, it's a force that acts in the opposite direction.
-There are multiple sources of drag for different types of scenarios.
-For scenarios where an object is in contact with two different types of fluid (like a ship, on the ocean) or when the fluid doesn't really have contact with all of the object (supersonic flight, and supercavitating torpedoes) some more complex physics applies.
-For a fuller discussion of types of drag, see [@carlton2007]  But in the case of a submarine, which does it's business completely immersed in the ocean, the relevant physics is dominated the skin friction on the hull, which follows a very simple rule and relationship.
-The amount of drag ($F_D$) an objet experiences increases directly in proportion to the surface area $A$.
-For any given object of a certain (unchanging shape) there will be a constant coefficent ($C_D$) which reflects how aerodynamic or hydrodymanic the shape is.
-The drag is also directly proportional to the density of the fluid being moved through, $\rho$.
-(Air creates roughly one-thousandth the drag as water does on any given object at a given speed, since it's roughly one thousand times less dense.)
-
-But the most sensitive factor in this relationship is the speed at which the object moves through the liquid.
-The drag increases not with the speed ($v$), but with the **square** of the speed .
-This means if the speed doubles, the drag increases by four.
-If the speed triples, the drag increases by a factor of nine.
-
-$$F_D = \frac{1}{2}\rho v^2C_DA$$
-
-(Technical aside: This law applies wherever the the flow over the surface is turbulent.
-It is true that for very small objects, or very viscous fluids, or very slow movements a different apples called Stokes Equation, in the case wher Reynolds numbers are less than 1.
-Given that sea-water is not particularly viscous, and sumbarines not particularly small, Reynolds numbers are liekly to be much much greater than 1 (one or two thousand), even when moving at only one or two knots.
-Since it is unlikely that a significant proportion of the flow over the hull will be laminar, we'll use the drag equation in all modeling going forward when considering drag on the hull).
-
-It's crucial to understand, however, that drag is only a force, and doesn't directly inform us about how much energy is consumed, until we multiply it by the **distance** over which it is applied, not the time for which it is applied.
-To think about it simply, gravity exerts a force on you downwards all the time.
-But you don't expend any energy overcoming it when sitting still.
-If you climb stairs, the amount of gravitational potential you attain depends on how high up you climb, not how long you spend on the ladder or stair-case.
-
-This has a significant consequence for propulsion, since the amount of power (energy expended in a given time) required for thrust scales with the drag, multiplied by the distance covered in a given time.
-As such, power required (and fuel/battery consumption) scales with velocity **cubed** rather than velocity squared.
-This means that if you double your speed of travel, fuel consumption for a given period of time will increase by a factor of eight.
-Due to the increased speed of travel, the fuel required to cover a given distance will only quadruple.
-Hence the range that can be covered with a given amount of energy (assuming propulsion dominates energy requirements) tends to scale with the inverse of velocity squared, whereas the endurance (amount of time that can be spent travelling) scales with the inverse of velocity cubed.
-
-This has a profound impact on the operation and engineering of maritime vessels.
-Reducing velocity has such a substantial reduction on hull drag that going slower is almost always a reasonable means of conserving fuel overall.
-To the extent that time is non-critical, slower is always much much better.
-It's for this reason that during the financial crisis, many cargo shipping companies adopted the practice of slow steaming [@liang2014] in order to conserve fuel, despite this causing a range of possible new engineering issues which need to be accounted for in order to operate the engines at lower than normal power for sustained periods [@sanguri2012 pp.
-8-10].
-The fuel saving from operating even 30% slower means that around half the total energy is required for a given journey.
-Other incremental costs and inefficiences from operating 'off-design' are frequently outweight by such a dramatic reduction in overall power demand.
-
-This effect is of profound importance to understanding the operations of conventional submarines, which have extremely constrained energy stores when operating under the surface.
-Diesel fuel when burned has an energy density of approximately 45MJ/kg.
-In contrast, a lead-acid battery might have energy densities in the range of 0.08-0.14MJ/kg.
-With something like 400 times as much energy per kg embarked in diesel, operating on batteries imposes an extreme demand for economy on propulsive power.
-Happily for submarines, the square law for drag, and cubed law for power, allow an almost commensurate reduction in power demands to take place by slowing down to very slow speeds when submerged.
-Hypothetically, a conventional submarine might have a total range of 10,000nm from using it's diesel payload at 8kt.
-If an equal weight of lead-acid batteries as fuel were carried, the submarine could only travel about 25nm when submerged at the same speed in a single charge.
-But by travelling at 4kt, that quickly increases to 100nm, and 400nm at 2kt (neglecting hotel load here for simplicity).
-
-It's worthwhile pointing out that extreme demand for economy imposed by the poor energy density of batteries distinguishes submarines quite remarkably from other types of boat or ship design.
-In every other application, the propulsion system is designed around a particular speed and loading condition which it is optimally efficient for, and a band of plausible variation around this.
-For example, a cargo ship designed for operation at 22kt will have it's propeller and engine (propulsion system) very carefully designed around these speeds and the plausible levels of loading at those speeds.
-A top-speed which might be somewhat higher than 22kt would be calculated, but would be relatively unlikely to be used for any purpose other than an emergency.
-Lower speeds might be considered 15-18kt for the purposes of slow-steaming.
-However, the relative efficiency at 3kt is of little concern, since the drag at that speed will be under 2% of what it would be at 22kt, and the power required around 0.25%.
-Even if the efficiency was substantially worse, or better (by a factor of two or three even) the impact on the overall economics of the operation would be negligible when compared to marginal improvements at the design speed.
-
-The same consideration applies to other types of vessels, which might use water-jets.
-Fast pilot-boats, for instance, might spend quite a considerable time manouvering in and out of harbour or alongside at very low waterspeeds.
-However, their overall fuel consumption is still likely to be dominated by the fast section of their trip to a ship.
-If the in-harbour water-speed was 8kt, but 32kt was the optimum speed for the open-water section of their journey, the fuel requirement for a given distance could still be over ten times greater at 32kt.
-(Aside, the physics described above would say 16 times, but some reduction due to the reduction of surface area from planing might offset this.)  Consequently the owner/operator would be at least ten times as concerned about efficiency at 32kt as at 8kt, (unless the harbour transit was much much longer than the open water section), and have negligible concern about effiencies at 2kt or 3kt, which would rarely amount to more than a couple of percent of total propulsion energy demand.
-
-In stark contrast, a submarine operator might have less than 1% of the energy store available for the submerged movements at 2-3kt, and consequently be substantially **more** concerned about efficiency in this regime, particularly since this would be the regime in which all of their combat operations would be conducted.
-
-This imposes a constraint on the way in which literature on jets, propellers, and propulsion systems in general is read, since almost all of it is deeply concerned with how a particular system would operate at one particular design speed, and choosing the optimum system for that design speed and load.
-Considerable attention is given to how a propeller would work in off-design conditions relating to changes in load, such as in cases when a ship is more lightly laden, or a tug-boat is pushing a different sized ship.
-There is, however, a relative scarcity of attention given to the performance of a propulsion system operating at a dramatically different speed, with the same load.
-It also corresponds to scenarios where engineers and manufacturers seem from discussion to be far less confident in making quantitative claims about the performance of their own systems.
-(They are perfecly happy to make qualitative claims, "it would be alright", "nothing very bad would happen" etc.)
-
-Given that submarines almost by definition (since their hull shape doesn't change, and they must carry ballast to make up for any under-loading) are essentially always operating with the same load, but have to operate over an extremely wide variety of different speeds (a factor of four or more in variance) with severe efficiency concern at all of them, they represent a truely unique engineering question, seldom discussed in commercial applications.
-Consequently, there are relevantly few pieces of literature which address the present question directly.
-(Happily, not zero)  However, there are plenty of instances where the relevant physics and trade-offs are explained in depth to address related, but not identical questions.
-
-# The difference between nuclear and conventional propulsion
-
-Nuclear submarines vary quite remarkably from conventional submarines because of the means by which they generate their power.
-Because of the extremely high energy density in enriched uranium or plutonium, the reactors on board nuclear submarines generate an abundant supply of energy.
-Most submarine reactors are reportedly capable of generating between 25 to 50 megawatts (MW) of power, though Russian submarines have hundreds of megawatts of power available [@WNA2017].
-
-In contrast, the Collins Class Submarine's main motor is rated at less than 6MW, with designs for the future submarine appearing to be only slightly larger [@patrick2012].
-It is fair to say that in terms of maximum power ouput, nuclear submarines could have something like 5-10 times as much power at their disposal, and their **peak** output.
-
-The difference between the peak-powers of the submarines significantly understates how different the designs are, because the difference in the total amount of energy stored and available for use in a voyage or dive is vastly greater.
-A nuclear submarine might have literally millions of times more energy at its disposal, it is practically unlimited for all intents and purposes on a given voyage.
-Consequently, wheras a nuclear submarine might regularly conduct transits at or near it's peak power, a diesel electric submarine would probably transit at less than half the speed which it could manage at a sprint, (and use less than a quarter of the energy, as discussed earlier) and on patrol it might be operating at a tenth of the maximum speed, and use maybe just 1% of as much power on propulsion.
-In this situation, the amount of power that is drawn for lights, CO2 scrubbing, washing, cooking, heating, as well as the electronics driving the combat systems (the 'Hotel Load') might well become significant, and even be as larger or larger than what is required for propulsion.
-
-Consequently, the difference between the power output a conventional and a nuclear submarine might regularly operate at could be even quite a bit larger than the maximum amount of energy that they can deliver to their propulsion systems.
-
-Perhaps more significantly for power-design of nuclear submarines is a little-discussed phenomenon called Xenon poisoning, which affects nuclear reactors when they shut down or lower their power significantly.
-When Uranium or Plutonium atoms split (or fission) into two smaller isotopes atoms, a variety of different radio nucleides (unstable variants of atoms) are produced.
-Two of these are Xenon-135 and Iodine 135.
-Iodine is produced much more often, and decays into Xenon-135, which a half-life of about nine hours.
-This Xenon has a very special, perhaps unique role in reactors, since they very easily captures the free neutrons which cause the continued fission reactions in Uranium or Plutonium which drive the reactor.
-As such, Xenon is known as reactor 'poison' since it can kill the reactor's reactivity in very high doses.
-
-This high neutron-capture from Xenon means that it has a duplicitous relationship with the reactor's power level.
-At high levels of reactor power, lots of Xenon-135 and Iodine-135 are produced by the fission process.
-The high presence of Xenon reduced the reactor's reactivity.
-On the other hand, there are lots of neutrons available to 'burn off' the Xenon (which absobs the neturons to become a different isotope).
-This keeps the total level of Xenon in check.
-
-The situation becomes much more complicated when the reactor undergoes a sudden change in power output.
-If the power is lowered dramatically and suddenly, the production of Xenon continues quite rapidly for some time due to the decay of the large stock of Iodine-135.
-With less neutron flux available to 'burn off' the Xenon, the Xenon levels spike, and push down the reactor's reactivity.
-Unless the reactor is quickly raised back to relatively high power (~60%) quite quickly (an hour or less), the Xenon levels become so high that the reactor will have to be shut down, otherwise extreme (and dangerous) measures would be required in order to keep the reactor going.
-(This is essentially what lead to the Chernobyl Explosion [@WNA2009].)  A fuller discussion of Xenon poisoning can be found in [@garland2005], which demonstrates key concepts related to the poisoning effect shown below.
-
-![Xenon poisoning effect following shutdown from @garland2005](/Users/aidanmorrison/submarines/XenonPoison.png){width=500px}
-
-Consequently, nuclear reactors aren't well suited to rapid fluctuations in power, particularly dramatic reductions in power, as these can lead to instability in the reactor core.
-If the reactor is shut down in order to avoid such dangerous circumstances, it generally cannot be started again until Xenon levels have fallen again, which can take a couple of days.
-Obviously this is never desirable for a military vessel, and hence is avoided at almost all costs.
-
-This effect has a dramatic impact on nuclear submarine design, since much of their design is oriented around being able to comfortably disperse large amounts of excess power, rather than with conserving it.
-For the propulsion system, this actually makes having an inefficient propulsor at low speeds a considerable advantage.
-Since the reactor will likely need to dispose of excess power, particularly during ramp-down, an inefficient propulsion system actually provides a useful power sink.
-Since any excess power will have to be disposed of by some other means (normally by pumping more water to remove the power as the heat) inefficiency at low speeds has no penalty, and probably a marginal benefit, since it will reduce overall demand for additional systems.
-Provided the excess turbulence inside the pumpjet isn't too noisy, wasting energy through the propulsor is useful.
-
-It should also be noted that a nuclear reactor's aversion to sudden reductions in power would also have a substantial impact on the design of a submarine's combat system, and it's demand on the Hotel Load.
-For the same reason, a high Hotel Load, or power-hungry Combat System, could actually be advantageous, as it helps to set an elevated 'floor' for power requirements, reducing the scale of fluctuations in overall power demand from the reactor due to changes in propulsion speed.
-
-# Some essential concepts
-
-### Conservation of Energy
-
-This is perhaps one of the most fundamental and well-established principals in physics.
-The essential idea is that energy can move or change in form, but it isn't ever created or destroyed.
-Machines, plants and animals all derive their energy from a particular other source, which can be measured and evalutated to establish the limits of energy available.
-Plants collect energy from sunlight falling on their leaves.
-Humans (as well as combustion engines) capture the chemical potential energy in organic matter, and release it by combining it with oxygen.
-Hydro-electric power plants turn the gravitational potential energy of water stored at a height into electricity.
-
-Conservation of energy has a particularly relevant embodiment in fluid flows, which is given it's primary expression in Bernoullis equation.
-It says that the total energy in a connected body of fluid is constant, though it can change in form between kinetic energy (movement), gravitational potential energy (it being elevated) the heat energy in the fluid, and the pressure of the fluid.
-
-In different fluids, the dynamics of how energy moves between one and another change.
-For example, in gases, heating up a confined piece of gas will increase it's pressure, or if it is unconfined, increase its volume.
-This is particularly important for understanding gas turbines.
-In the case of a liquid, however, all the molecules are in close contact, and hence can't increase in volume or pressure substantially except by the creation of steam.
-As such, in the absence of large amounts of cavitation, the terms in the relationship which are most important for our consideration are the relationships between pressure, and kinetic energy, and gravity.
-
-In the case of most waterjets which propel surface vessels, the water is lifted from the bottom of the hull at the intake to the pump, which is generally incorporated inside the hull.
-This increase in gravitational potential coincides with a slowing down of the water relative to the vessel-speed.
-Since the jet in such vessels is generally ejected at the same height as the pump, this potential is never regained, and is technically a loss, however at high speeds such a loss is small relative to the total power ouput.
-
-#### Venturi Effect and Flow Diffusion
-In the case of submarines and torpedoes the water doesn't undertake a change in height, since intake and nozzle are generally all in line with the central axis of the submarine or torpedo.
-As a consequence, the key relationship in Bernoullis equations is the relationship between the liquids velocity, and its pressure.
-The consequence is that when water moves through a pipe (duct, or shroud) it's pressure is inversely related to the square of its velocity.
-This means that when a liquid is forced to to travel through a narrowing pipe, it's pressure necessarily decreases as the velocity increases.
-This is a simple embodiment of the Venturi effect.
-The inverse process is where a pipe increases in volume, and the flow is forced to slow down in order to fill the wider area, and the pressure correspondingly increases.
-This is a process called 'diffusion', and is important to achieving high pressure levels in many types of water pumps, including those which will be particularly relevant for pumpjets for watercraft propulsion [@hamilton1997].
-
-#### Energy waste and efficiency
-The conservation of energy also has useful implications for how the efficiency of systems is thought about.
-In particular, because energy is conserved, identifying inefficiencies in a system necessarily involves identifying where energy goes to doing tasks which aren't useful for the intended purpose.
-A perfectly efficient system won't do any work that isn't for the intended purpose.
-In the case of analysing the efficiency of propulsion systems, the relevant 'work' is almost always related to moving water backwards to produce thrust.
-Moving water in directions other than backwards, including random turbulent flows which wind producing heat rather than thust, are two examples of wasted energy.
-Noise in the water also reflects energy which is wasted.
-
-### Boundary Layer
-When water or any fluid flows with some speed relative to another solid surface nearby, there is some layer adjacent to the surface in which the speed of the fluid is diminished relative to the main flow.
-At a microscopic level, there are some molecules of the fluid on the surface which will be effectively static relative to the surface.
-The layer of fluid that joins the gap between the static surface, and the part of the flow which is moving at the full flow speed, is called the boundary layer.
-Exactly how thick the boundary layer is, and how the fluid moves in the boundary layer, is extremely important for consideration of efficiency of fluid flows over and around solid surfaces.
-
-### Turbulent and Laminar Flow
-
-### Pressure, Thrust, and Momentum
-
-### Cavitation
-
-### Flow separation
-
-# What tech are we talking about? The core relationships between pumps, jets, ducts, and propellers, and all the different things we call them, their parts, and combinations
-
-# Propulsive performance of the pumpjet, or decelerating ducted propeller
-
-## Efficiency Relationships, and various curves
-
-## Practical and recent evidence of efficiency relationships
-
-# Results of numerical simulation
-
-# Acoustic Considerations
-
-# Conclusions
-
-# References
+\end{document}

### 2017-12-29 DraftExec.Rmd
Big Push, need to migrate

--- 2017-12-18 DraftExec.Rmd
@@
     year: 2005
+
+- id: fujisawa1995
+  title: Measurement of Basic Performances for Waterjet Propulsion Systems in Water Tunnel
+  author:
+  - family: Fujisawa
+    given: Nobuyuki
+  volume: 2
+  URL: 'downloads.hindawi.com/journals/ijrm/1995/936068.pdf'
+  publisher: International Journal of Rotating Machinery
+  type: article-journal
+  issued:
+    year: 1995
 
@@
 
+- id: NPdotnet2017
+  title: Laminar Flow
+  author:
+  - family: 
+    given: 
+  URL: 'http://www.nuclear-power.net/nuclear-engineering/fluid-dynamics/laminar-flow-viscous/'
+  publisher: nuclear-power.net
+  issued:
+    year: 2017
+
 - id: patrick2012
@@
 
+- id: patrick2017
+  title: Senate Estimates Committee 
+  author:
+  - family: Patrick
+    given: Rex
+  URL: 'http://parlview.aph.gov.au/mediaPlayer.php?videoID=379261&operation_mode=parlview#/4%20from%20around%201201'
+  publisher: Australian Parliament House
+  issued:
+    year: 2017
+    month: 10
+
 - id: hamilton1997
@@
   volume: 8
-  URL: ''
+  URL: 'http://www.hamiltonmarine.co.nz/includes/files_cms/file/JetTorque%2008.pdf'
   publisher: Hamilton Waterjets
@@
     month: 3
+
+- id: mollard2011
+  title: Ship Resistance and Propulsion
+  author:
+  - family: Mollard
+    given: Anthony F.
+  author:
+  - family: Turnock
+    given: Stephen R.
+  author:
+  - family: Hudson
+    given: Dominic A.
+  URL: 'http://dl.kashti.ir/ENBOOKS/NEW/Ship%20resistance%20and%20propulsion%20_%20F.Molland%20(2011).pdf'
+  publisher: Cambridge University Press
+  issued:
+    year: 2011
+
+- id: shin2015
+  title: Methods for Cavitation Prediction on Tip-Modified Propellers in Ship Wake Fields
+  author:
+  - family: Shin
+    given: Keun Woo
+  - family: Regener
+    given: Pelle Bo
+  - family: Andersen
+    given: Poul
+  URL: 'http://www.marinepropulsors.com/proceedings/2015/WA2-1.pdf'
+  publisher: Fourth International Symposium on Marine Propulsors
+  issued:
+    year: 2015
+    month: 6
+
+- id: moulijn2015
+  title: Application of Various Computational Methods for Predict the Performance of Cavitation of Ducted Propellers
+  author:
+  - family: Moulijn
+    given: Joost
+  URL: 'www.marinepropulsors.com/proceedings/2015/WB3-1.pdf'
+  publisher: Fourth International Symposium on Marine Propulsors
+  issued:
+    year: 2015
+    month: 6
+
+- id: lewis1988
+  title: Principals of Naval Architecture
+  author:
+  - family: Van Manen
+    given: J.D.
+  - family: Van Oossanen
+    given: P.
+  volume: 2
+  URL: 'http://opac.vimaru.edu.vn/edata/EBook/Principles%20of%20Naval%20achitecture%202.pdf'
+  publisher: The Society of Naval Architects and Marine Engineers
+  editor: 
+  - family: Lewis
+    given: Edward V.
+  issued:
+    year: 1988
+
+- id: li2013
+  title: Dynamic Characteristics of Rotating Stall in Mixed Flow Pump
+  author:
+  - family: Li
+    given: Xiaojun
+  - family: Yuan
+    given: Shouqi
+  - family: Pan
+    given: Zhongyong
+  - family: Li
+    given: Yi
+  - family: Liu
+    given: Wei
+  URL: 'https://www.hindawi.com/journals/jam/2013/104629/'
+  publisher: Hindawi Publishing Corporation Journal of Applied Mathematics
+  issued:
+    year: 2013
+    month: 8
+
+- id: kamal2015
+  title: Powering for Medium Speed Wave-Piercing Catamarans comparing Waterjet and Screw Propeller Performance using Model Testing
+  author:
+  - family: Kamal
+    given: Iwan Mustaffa
+  - family: Zürcher
+    given: Konrad
+  - family: Bose
+    given: Neil
+  - family: Binns
+    given: Jonathan
+  - family: Chai
+    given: Shuhong
+  - family: Davidson
+    given: Gary
+  URL: 'http://www.marinepropulsors.com/proceedings/2015/MA3-2.pdf'
+  publisher: Fourth International Symposium on Marine Propulsors
+  issued:
+    year: 2015
+    month: 6
+
+- id: giles2010
+  title: The Advanced WaterJet Propulsor Performance and Effect on Ship Design
+  author:
+  - family: Giles
+    given: Will
+  - family: Dinham-Peren
+    given: Tom
+  - family: Amaratunga
+    given: Shane
+  - family: Vrijdag
+    given: Arthur
+  - family: Partridge
+    given: Richard
+  URL: 'https://www.bmtdsl.co.uk/media/6097875/BMTDSL-The-Advanced-Waterjet-Confpaper-INEC-May10.pdf'
+  publisher: Tenth International Naval Engineering Conference
+  issued:
+    year: 2010
+    month: 5
+
+- id: haimov2010
+  title: Ducted Propellers. A Solution for Better Propulsion of Ships. Calculations and Practice.
+  author:
+  - family: Haimov
+    given: H.
+  - family: Bobo
+    given: M.J.
+  - family: Vicario
+    given: J.
+  - family: del Corral
+    given: J.
+  URL: 'http://www.cehipar.es/__files/users/publicaciones/217.pdf'
+  publisher: Ministerio de Defensa
+  issued:
+    year: 2010
+    month: 5
+
+- id: abdel2010
+  title: Design of a Multi-Component Propulsor
+  author:
+  - family: Abdel-Maksoud
+    given: Moustafa
+  - family: Steden
+    given: Max
+  - family: Hundemer
+    given: Jochen
+  publisher: 28th Symposium of Naval Hydrodynamics
+  issued:
+    year: 2010
+    month: 9
+
+- id: bakker2006
+  title: Boundary Layers and Separation
+  author:
+  - family: Bakker
+  URL: 'http://www.bakker.org/dartmouth06/engs150/11-bl.pdf'
+  publisher: Dartmouth College
+  issued:
+    year: 2002
+
+- id: mcbride1979
+  title: Design and analysis of turbomachinery in an incompressible, steady flow using the streamline curvature method
+  author:
+  - family: McBride
+    given: Mark W.
+  URL: 'www.dtic.mil/dtic/tr/fulltext/u2/a071107.pdf'
+  publisher: Naval Sea Systems Command
+  issued:
+    year: 1979
+    month: 7
+
+- id: wislicenus1986
+  title: Preliminary Design of Turbopumps and Related Machinery
+  author:
+  - family: Wislicenus
+    given: George F.
+  URL: 'https://ntrs.nasa.gov/archive/nasa/casi.ntrs.nasa.gov/19870008232.pdf'
+  publisher: NASA
+  issued:
+    year: 1986
+    month: 10
+
+- id: wislicenus1973
+  title: Hydrodynamic Design Principles of Pumps and Ducting for Waterjet Propulsion
+  author:
+  - family: Wislicenus
+    given: George F.
+  URL: 'http://www.dtic.mil/dtic/tr/fulltext/u2/775620.pdf'
+  publisher: Naval Ship Research & Development Center
+  issued:
+    year: 1973
+    month: 6
+
+- id: brandau1967
+  title: Apects of performance evaluation of waterjet propulsion systems and a critical review of the state-of-the-art
+  author:
+  - family: Brandau
+    given: John H.
+  URL: 'http://www.dtic.mil/dtic/tr/fulltext/u2/827069.pdf'
+  publisher: Naval Ship Research & Development Center
+  issued:
+    year: 1967
+    month: 10
+
+- id: henderson1964
+  title: A Method for the Design of Pumpjets
+  author:
+  - family: Henderson
+    given: R.E.
+  - family: McMahon
+    given: J.F.
+  - family: Wislicenus
+    given: G.F.
+  URL: 'http://www.dtic.mil/dtic/tr/fulltext/u2/439631.pdf'
+  publisher: Ordinance Research Laboratory, Department of Navy
+  issued:
+    year: 1964
+    month: 5
+
+- id: MAN2017
+  title: Basic Principals of Ship Propulsion
+  author:
+  - family: MAN Diesel & Turbo
+  URL: 'https://marine.man.eu/docs/librariesprovider6/propeller-aftship/basic-principles-of-propulsion.pdf?sfvrsn=0'
+  publisher: MAN Diesel & Turbo
+  issued:
+    year: 2017
+
+- id: oosterveld1970
+  title: Wake Adapted Ducted Propellers
+  author:
+  - family: Oosterveld
+    given: IR.M.W.C.
+  URL: 'http://www.marin.nl/web/Publications/Publication-items/Wake-Adapted-Ducted-Propellers.htm'
+  publisher: Netherlands Ship Model Basin Wageningen - Netherlands
+  issued:
+    year: 1970
+
+- id: davies2017
+  title: Pump up the pump jet?
+  author:
+  - family: Davies
+    given: Andrew
+  URL: 'https://www.aspistrategist.org.au/pump-up-the-pump-jet/'
+  publisher: Australian Strategic Policy Institute
+  issued:
+    year: 2017
+    month: 11
+
+- id: stanford2017
+  title: Australia's Future Submarine Getting This Key Capability Right
+  author:
+  - family: Jon
+    given: Stanford
+  URL: 'http://www.insighteconomics.com.au/reports/2017_Insight_Economics_Submarine_Report.pdf'
+  publisher: Insight Economics
+  issued:
+    year: 2017
+    month: 09
+
 ---
@@
 
-### Conservation of Energy
+## Conservation of Energy
 
@@
 
-#### Venturi Effect and Flow Diffusion
+### Venturi Effect 
 In the case of submarines and torpedoes the water doesn't undertake a change in height, since intake and nozzle are generally all in line with the central axis of the submarine or torpedo.
@@
 This is a simple embodiment of the Venturi effect.
+
+### Flow Diffusion
 The inverse process is where a pipe increases in volume, and the flow is forced to slow down in order to fill the wider area, and the pressure correspondingly increases.
@@
 
-#### Energy waste and efficiency
+### Energy waste and efficiency
 The conservation of energy also has useful implications for how the efficiency of systems is thought about.
@@
 
-### Boundary Layer
+## Boundary Layer
 When water or any fluid flows with some speed relative to another solid surface nearby, there is some layer adjacent to the surface in which the speed of the fluid is diminished relative to the main flow.
@@
 Exactly how thick the boundary layer is, and how the fluid moves in the boundary layer, is extremely important for consideration of efficiency of fluid flows over and around solid surfaces.
+In particular, a boundary layer can be either turbulent or laminar in nature, and can transition to turbulent flow after a short distnace of laminar flow, as shown.
 
-### Turbulent and Laminar Flow
+![The development of a boundary layer as shown in @mollard2011](/Users/aidanmorrison/submarines/BoundaryLayer.png){width=500px}
 
-### Pressure, Thrust, and Momentum
+## Turbulent and Laminar Flow
+The way that fluids move relative to a surface can one of either two methods.
+In 'laminar flow' all the fluid moves in one direction in an smooth and orderly manner, with very little mixing between the layers of fluid travelling at different speeds.
 
-### Cavitation
+The alternative is 'turbulent flow', where the fluid moves around in unpredictable swirls and circles as well as moving overall in an predominant direction.
+There is considerable mixing between all the different layers, and the average speed remains relatively constant in the flow, with the exception of the flow immediately adjacent to the wall, or in the boundary layer.
+Becuase turbulence involves a lot of movement that is not overall in one productive direction and contributing to thrust, it necessarily leads to some loss of energy from the overall thrust.
+Once the swirls and movements become smaller and smaller, the energy winds up simply as heat in the fluid.
 
-### Flow separation
+![A simple comparison of laminar and turbulent flows from @NPdotnet2017](/Users/aidanmorrison/submarines/LaminarTurbulent.png)
 
-# What tech are we talking about? The core relationships between pumps, jets, ducts, and propellers, and all the different things we call them, their parts, and combinations
+The likely point of transition between a flow being turbulent and laminar can be given by unitless number called the Reynolds Number, which is determined by a fluids velocity, viscoscity, density, and a characteristic length-scale over which the flow occurring.
+A fuller discussion of these important concepts can be found in [@NPdotnet2017].
 
-# Propulsive performance of the pumpjet, or decelerating ducted propeller
+Turbulent flows in gasses can be extremely noisy, since those fluids are compressible.
+Consequently, the rapid and intense ciruclar movements result in a lot of oscillatory compressions against the surrounding air.
+Consequently things like jet engines, vacuum cleaners, hand-dryers, and other devices that create rapid movements in gasses tend to be quite loud, including at some distance.
 
-## Efficiency Relationships, and various curves
+In essentially incompressible fluids, the random changes in pressure and velocity which are involved in laminar flow tend not to generate nearly as much noise in the surrounding liquid outside the flow.
+Because the fluid is incompressible, all of the random 'round and round' movements don't amount to much 'in and out' movement, which is what creates the pressure waves which result in propagated noise.
+Consequently, whilst turbulent flows do inevitably generate some noise, the amount of noise generated is dramatically greater in the presence of cavitation, where a void opens up in the water, and vastly more 'in and out' movement is able to occur.
 
-## Practical and recent evidence of efficiency relationships
+## Cavitation
+Cavitation is the rapid expansion and collapse of a void, or bubble in water.
+Put technically, caviation occurs when the local static pressure (pressure in the rest frame of the fluid) falls below the vapour pressure of the fluid (pressure at which the liquid will start to boil).
+
+Imagine what happens when something moves very fast through water.
+The front side of the object pushes the water forward, but on the back side the water has to push in to fill the space left behind.
+If there’s not enough pressure to push the solid water in fast enough, a gap opens up, with just a few gaseous water molecules (steam) inside the cavity.
+(This is also described as water boiling at low pressure.)  However the gap doesn’t stay around for long.
+Soon the water catches up, and the bubble implodes with a pop, leaving only tiny bubbles as a result, which you can see in the wake of almost any boat or ship moving at speed.
+Whilst some energy turns to heat (the remaining steam in the tiny bubbles) quite a bit is propagated away as a sound-wave generated by the implosion.
+
+A watching water boil in a glass kettle gives a quick and intuitive insight into the occurrence of cavitation.
+Quickly after the kettle starts heating, a considerable noise can be heard, which corresponds to the commencement of cavitation on heating element.
+At some local point for a moment in time, there is enough energy for the water present to boil.
+It tends to be on rough surfaces or in the presence of some impurity that cavitation will occur first (in the blackened part of the surface in this case.)  However, whilst the bubbles on the bottom can be seen plainly on the bottom, they collapse almost straight away again, and leave only a tiny bubble of stable steam ciculating in the water.
+It is the rapid expansion and collapse of these bubbles that causes the noise of a kettle, long before it has boiled.
+
+![Cavitation bubbles form and collapse creating noise and some tiny bubbles long before boiling occurs](/Users/aidanmorrison/submarines/EarlyCavitation.JPG){width=500px}
+
+It is only when the water is all at a much higher temperature that the bubbles remain their full size for long enough to detatch and rise all the way to the surface, a process we typically think of as boiling water.
+It is worthwhile noting that the sound emitted at this stage is much softer and lower than the early onset of cavitation.
+Larger bubbles result in lower frequencies of sound being emitted.
+Even at higher temperature, the rougher parts of the surface provide the points where all the cavitation originate.
+
+![Only at very high temperatures do the bubbles endure at full-size in the water and reach the surface](/Users/aidanmorrison/submarines/BoilingCavitation.JPG){width=500px}
+
+Cavitation is extremely important in the study of ship propulsion, since it's occurrence in particular circumstances can lead to substantial losses of efficiency, as well as damage to the propeller and related appendages.
+It is particularly important for submarines, since the expansion and collapse of these bubbles tends to lead to the creation of noise which is often far larger, and more distincly characteristic than other turbulent disturbences in the water when no cavitation is present.
+Consequently, the onset of cavitation can be thought of as a distinct threshold in terms of the acoustic performance of propulsive system.
+
+Despite cavitation necessarily representing some energy being wasted generating unwanted noise in the water, some cavitation inevitably occurs around most propulsion systems operating at full power.
+In plenty of cases, the consequences for efficiency are relatively small, as they tend to be dominated by other efficiency considerations in imparting thrust.
+Put another way, the savings that can be gained by minimising wastage to turblence can outweigh the losses incurred by having some cavitation occur.
+In most cases, propellers are designed to work with a certain extent of cavitation for optimum efficiency under working loads [@shin2015].
+Cavitation also occurs to a considerable extent within the jets of most high-speed surface vessels, but doesn't necessarily have a particularly bad effect on their overall efficiency at their intended speed levels.
+It is generally only within very specialised military circumstances when the absolute avoidance of cavitation supercedes other concerns for efficiency, and require the elimination of all cavitation entirely [@mollard2011], [@lewis1988].
+These circumstances include the design of submarines and torpedoes.
+
+In certain special cases, particularly supercavitating propellers or surface-piercing drives, high levels of cavitation can be higly advantageous for efficiency.
+However these represent specialised designs, generally only for very high speed vessels, where acoustic concerns are negligible, and are of little concern for this particular endeavour.
+
+Cavitation can occur in a number of different ways at different points and forms on or around the propeller.
+It is common for cavitation to occur first near the out extremities of the propeller, where the blades are moving at the highest velocity relative to the water-flow.
+Cavitation tends to spread across the back of the blades in a sheet (sheet cavitation), since the back side of the blades are generally the areas of most depressed pressure, though it can also occur in places on the back side of the blades, where the dynamics of the water moving around the blade can produce points of near the edges of significantly reduced pressure.
+On surface vessels cavitation tends to occur most when the blades are closes to the surface, where the pressure is lowest.
+
+![Cavitation frequently occurs first at the extremities, then spreads inwards across the blad face, as shown in [@mollard2011]](/Users/aidanmorrison/submarines/IncreasingCavitation.png){width=500px}
+
+![Different types of cavitation can occur on different parts of the propulsion system as shown in [@carlton2007]](/Users/aidanmorrison/submarines/CavitationTypes.png){width=500px}
+
+## Pressure, Thrust, and Momentum change
+It is worthwhile spending just a moment to touch on the relationship between a few different concepts which are closely related.
+Thrust is a force, measured in Newtons, which is generally used to refer to the forces acting on a vessel or object by its propulsion system to move it forward in the water.
+Generating thrust is consequently a key objective of any propulsion system.
+
+Pressure is a measure of force divided by an area over which it is applied.
+Pressures are necessarily applied to some extent over all of the surfaces of any object, including due to the atmostpheric pressure of air.
+In considering the efficiency of propulsion systems for solid bodies, thrust is always the consequence of some **net** pressure difference on different sides of an object.
+For instance, on a propeller blade, the rear-facing side of the blade is the side that pushes the water backwards, and hence has a raised pressure on its surface.
+Being generally backwards facing this acts to push the propeller forwards.
+This pressure acts equally to push water backwards, as well as to push the propeller forwards.
+Importantly, the forward-facing side of the propeller will expereince a lowered pressure, which also serves to pull water backwards, and suck the propeller forwards.
+It is the sum of all the pressure differences on the propeller that result in a net force.
+
+Importantly in hydrodynamics, it is important to consider the possibility of other forces being generated other surfaces which may have an impact on the movement of the vessel.
+In the presence of the hull, water being drawn by the propeller can cause pressure changes on other surface of the vessel which may assist or impede producing a net positive thrust.
+Submerged jets represent a particular case in point.
+A famous home experiment suggests that one compare the thrust (backwards pressure) experienced by someone holding a garden hose creating a jet of water when it is in the air, as opposed to submerged in a pool of water.
+When submerged, the backwards pressue appears dramatically reduced.
+In fact, this phenomenon is due to a the water-jet drawing adjacent water with it under-water, the movement of which exerts a negative pressure on the outside of the nozzle, conteracting the pressure exerted on the inside of the nozzle.
+This effect represents one of the important distinctions between the jets which release their water below or above the water-line.
+
+It is also an important law of physics that force or thrust is equal to the rate of change of momentum.
+Put simply, the speed at which one speeds something up, or the increasing amount of something which begins to move, is always equivalent in magnitude to the thrust that is produced.
+Since momentum is a vector, this simple law requires that the momentum change has to net out to one particular direction in order to produce a net thrust.
+Circular movements necessarily are cancelling, and produce no thrust.
+It is useful to recall however that this principal is simply another complete and way of measuring the consequences of the net pressure differences discussed earlier.
+The total rate of change of momentum in the water-column which is induced as a conseqence of the pressure differentials on all sides of the propeller must necessarily be equivalent to the total net force that is produced.
+Consequently literature seeking to optimise propulsion might discuss analysis referring to both of these effects or phenomena, and neither are in conflict.
+
+Full and proper derivations of many of the thrust equations for different propulsion systems from these principals can be found in plenty of authoritative works, including [@lewis1988], [@carlton2007] and [@mollard2011].
+
+## Flow separation
+Flow separation refers to a circumstance when the flow over a surface separates entirely from the surface, and a new eddy or votex is formed where the flow actually moves in the opposite direction to the dominant flow.
+It is of particular significance in hydrodynamics because it can lead to substantial losses of efficiency, as energy is diverted into the kinetic energy of the eddy or vortex, which is unproductive.
+It tends to occur near the surface of aerofoils with high angles of attack, or in the blades of turbomachinery including diffuser or mixed-flow pumps and ducted propellers operating in off-design conditions, including low flow rates [@li2013, @bakker2006, mcbride1979, wislicenus1986].
+
+![Flow separation occurring over the top of an aerofoil in a wind tunnel.
+Image courtesy of Deutsches Zentrum fuer Luft- und Raumfahrt e.
+V.
+(DLR)](/Users/aidanmorrison/submarines/SeparatedAerofoil.jpg){width=500px}
+
+In order for flow separation to occur, a fluid must be moving against an adverse pressure gradient, which simply means that the pressure is increasing in the direction of movement, which means the flow is slowing down as it moves.
+At the boundary layer, where the flow velocity is already reduced due to drag forces experienced near the surface, the adverse pressure gradient can be enough to reverse the flow altogether.
+In this case, the fluid flows in reverse near the surface and forms a vortex, and the main flow becomes separated form the surface by the votex.
+
+![Flow separation involves flow reversal near a surface when a fluid is moving against an adverse pressure gradient, as shown in [@mollard2011]](/Users/aidanmorrison/submarines/FlowSeparation.png){width=500px}
+
+In addition to the necessary loss of efficiency, flow separation can often tend to result in unsteady flows, with the vortices periodically being shed into the flow [@mollard2011 pp.
+480].
+If this occurs ahead of the blades of a propeller or impeller, such disturbances to the flow can lead to instances of cavitation when otherwise a steady homogenous flow might be well below the cavitation inception point, and can be consequential for the acoustic performance of the system, particulaly if such shedding resonates with characteristic frequencies of any of the machinery.
+
+# Basics of Ducted Propellers and Pumpjets
+It's easy to be somewhat confused by the many different names which seem to be associated with related, or similar systems.
+Between pumps, pumpjets, water-jets ducted or shrouded propellers, or impellers, there is plenty of grounds for some confusion  In this section, I aim to clarify in simple terms what the important differences between distinctly different systems are, and where some terms are used somewhat interchangeably without doing any great violence to the concepts underlying.
+
+## Propellers
+
+Perhaps the best starting point is the most basic, and oldest of the systems which we're considering: the basic screw propeller.
+A propeller uses a number of tilted blades attached to a central hub to sweep around disc in the water, and accelerate a column of water passing through the disc.
+Due to the slant of the blades, water on the back sides of the blades is pushed astern, and reduced pressure on the front faces pulls more water from ahead to replace it.
+As such, it accelerates a column of water, which necessarily is contracted in circumference after the point of acceleration.
+
+![Propellers generate thrust by accelerating a column of water, as shown in [@mollard2011 pp 247]](/Users/aidanmorrison/submarines/PropellerAction.png){width=500px}
+
+A comprehensive description of the development of the propeller can be found in [@carlton2007].
+Here I won't elaborate on beyond describing some essential features and characteristics with which one ought to be familiar, primarily for the purpose of comparing different propellers and their evolution into impellers of different designs.
+
+Most propellers will have between three and seven blades.
+In general they are shaped as an aerofoil, with the convex side being upstream, just as the convex side of a plane wing is above, in order to generate lower pressure and lift as it moves through the air.
+The blades tend to be twised so that they have a higher angle of attack closer to the hub, and lesser closer to the extremities, so that those faster moving sections push their respective parts of the water column at an overall similar speed.
+The are often also thinner towards the extremities, and will be swept backwards as if dragged by their rotation through the water (skew) and also dragged by the ship's movement throug the water (rake).
+Whilst the degree of all these characteristics is highly variable for different applications, these are a few of the commonly referred to characteristics which can be varied in order to optimise performance for any given application.
+
+![Propeller geometry including skew and rake, as shown in [@mollard2011 pp 262]](/Users/aidanmorrison/submarines/SkewRake.png){width=500px}
+
+### Pitch
+Perhaps one of the most important characteristics of a propeller is its pitch.
+The pitch represents the distance that a blade section would travel forward if it carved its way around a helix through one full rotation.
+It is intuitive to think of as something akin to the angle-of-attack of the propeller blade to the water, however this is technically misleading as the movement of the water-column incoming to the blade, as well as the speed of rotation, also have a significant bearing on what the actual angle of attack of the blade ends up being.
+Whilst technically pitch is a distance (measured in meters), it is often expressed and used in formula as a ratio of the diameter of the propeller, to give it a unitless equivalent, as is common in the description of propellers and propulsion systems.
+
+![A blade section of a propeller traces a helical line around a cylinder as it is rotated [@carlton2007]](/Users/aidanmorrison/submarines/HelicalLine.png)
+
+![The definition of pitch is the distance traveled by a blade section along a cylinder, as given in [@carlton2007]](/Users/aidanmorrison/submarines/PitchDefinition.png){width=500px}
+
+The selection of the exact blade selection that is selected to define pitch must be specified as being at some fraction of the radius from the centre.
+This distance is calculated to be the 'moment mean' or a technical derived effective average, which tends to lie between 0.6R and 0.7R.
+A thorough discussion of pitch can be found in [@carlton2007 pp 35-37].
+
+Pitch is of particular importance to the discussion the efficiency of propeller and propulsor design because its optimal choice tends to vary considerably with the different loads which a vessel is intended to operate at, which can also relate to a vessel's design speed.
+It is for this reason that a considerable number of vessels actually have variable pitch propellers.
+Whilst considerably more complex, expensive, and heavier than a traditional fixed-pitch propeller, the ability to vary the pitch of a propeller assists considerably, particularly when the amount of load (resistance) a vessel is expected to face varies considerably.
+Controllable pitch propellers also have advangates in terms of manouverability, since the pitch can be reversed and hence reverse thrust can be produced, without requiring the direction of the power coming from the engines to be reversed.
+This has advantages for ferries which undertake frequent docking manouvres in confined spaces [@MAN2017, @lewis1988].
+
+![Controllable Pitch Propellers (CP Propellers) allow the pitch of the blades to be varied. Image credit: [@MAN2017]](/Users/aidanmorrison/submarines/CPPropeller.png){width=500px}
+
+### Pitch and the Advance Ratio
+Pitch is also an extremely important concept to understand because it relates closely to the way that a propeller's 'open-water efficiency' $\eta_O$ is expressed in charts, as well as the advance ratio $J$ that usually comprises the x-axis in such charts.
+
+![A typical chart showing open water efficiency of a propeller, in this case a Wageningen B5-75. Image: MARIN](/Users/aidanmorrison/submarines/OpenWater.png){width=500px}
+
+This chart also shows two other closely related dimensionless coefficients, the torque and thrust coefficients ($K_Q$ and $K_T$) which we shall not dwell on here.
+However, it is important to note that a number of different lines are expressed on this sort of chart, which represent different versions of the same propeller, with only the pitch (expressed in its dimensionless form as $P/D$) being different in each case.
+The x-axis, $J$, also warrants a clear definition:
+
+$$J = \frac{V_a}{nD}$$
+Where $V_a$ is the speed of advance of the propeller, or the speed at which water arrives at the front-side of the propeller, and $n$ represents the rate of the propeller's rotation (revolutions per second) and $D$ is the propeller's diameter.
+For fullness of understanding it is also worthwhile to note here that $V_a$ does not correspond directly to the vessel's speed of travel, though they are related.
+To some extent, a vessel moving through water will always draw water in its wake along with it, which means that in the wake (where the propeller tends to operate to some extent or other) the water incident on the propeller will arrive at a lower speed than that implied by the ship speed [@MAN2017 pp 15].
+Another key distinction for the consideration of jets later on, is that the speed of advance does not necessarily incporate the movement of water due to the movement induced by the propeller (or duct) itself, which is a distinct effect to that of the wake field.
+In [@lewis1988 pp 213] $V_A$ is used specifically to denote the undisturbed stream velocity, since the role of certain ducts can be to accelerate or decelerate the stream prior to the water arriving at the blades of the propeller.
+
+One may also note from the previous figure that for each line representing a given pitch, the efficiency falls precipitously towards zero around the point where $J$ reaches $P/D$.
+This is because this extreme of the curve represents a circumstance where the propeller is rotating, but not actually moving any water, since it is moving through the water at about that speed which corresponds to the exact helical path defined by the pitch being traced.
+This represents circumstances where a propeller encounters effectively zero load.
+This might be the case when a eases it's power, and the momentum of the ship temporarily keeps the ship moving without the propeller exerting much or any net thrust, or where alternative propoulsion sources (sails, or other propellers) are keeping a speed higher than what the propeller actually has push backwards on water to maintain.
+
+A closely related concept here is the slip ratio, which is a measure of the difference between the pitch (the distance the blade section would travel if it traced a perfect helix for one rotation) and the actual distance that the blade did advance through the water.
+This is neatly defined and shown in a diagram in [@MAN2017].
+
+![The slip ratio of a propeller. Image credit: [@MAN2017]](/Users/aidanmorrison/submarines/SlipRatio.png){width=500px}
+
+It can be seen from the diagram of the open-water efficiencies that there is some desirable non-zero slip ratio for a propeller.
+The other end of the spectrum, where efficiencies also fall off toward zero, represent circumstances where the propeller is simply spinning in the same place in the water.
+In this case the slip ratio is one.
+This represents a circumstance known as 'bollard pull' where a vessel is simply exerting force against an immovable object or infinite load.
+Whilst the very low efficiencies here could represent a case of a propeller that is simply too small for the task it is trying to achieve, it is actually a significant and important circumstance to consider for propellers on working vessels such as tugs or trawlers, which exert large forces at very low water-speeds.
+Measurement of the thrust generated in bollard pull, and the power required to achieve this thrust, is still an important metric for such vessels, but the definition of 'useful work' for normal propulsion requieres the thust to be excerted over some distance travelled.
+
+## Ducted Propellers
+A ducted propeller is a propeller in any kind of tubular shroud, casing or nozzle.
+Ducted propellers fall into broadly two categories, one of which is referred to frequently and interchangeably as a pump-jet.
+The two types are called the accelerating type (also known as a Kort Nozzle, after a company that made them popular in the 20th Century) and the decelerating type.
+The names describe the different effects that each type of nozzle have on the flow speed of the water at the point where it meets the propeller (which is also called an impeller, particularly when associated with decelerating ducts, or pumpjets).
+
+A detailed description of the essential differences between the two can be easily demonstrated by diagram, which show a simple comparison an acelerating and decelerating duct, which have the shape of the aerofoil section which comprises the duct inverted.
+(Aside: It should be noted that some designs attempt to strike a balance between efficiency and cavitation performance of accelerating and decelerating ducts for certain purposes, and in such cases the duct shape might not be so clearly contrasted [@abdel2010].
+But for the purposes of explanation, as well as optimisation for military purposes, the distinction between the two will generally be quite clear.)
+
+![The two types of ducted propeller, accelerating (left) and decelerating (right) as shown in [@mollard2011 pp 249]](/Users/aidanmorrison/submarines/AccelDecel.png){width=500px}
+
+A consideration of what happens to the flow as it moves through each type of duct helps to demonstrate how the acceleration and deceleration effects are achieved.
+In the case of an accelerating nozzle, the area through which the flow must pass contracts prior to the water arriving at the impeller.
+In the decellerating case, the available area contracts after the water has passed the impeller.
+In some cases, a degree of diffusion is actually incorporated ahead of the impeller, whereby the area actually of the flow actually expands ahead of the impeller, which slows down the flow and increases its pressure according to Benoulli's principal [@wislicenus1973].
+
+![The streamlines of flow through different duct types [@lewis1988 pp 214]](/Users/aidanmorrison/submarines/DuctStreamlines.png){width=500px}
+
+### The Accelerating Duct
+Accelerating ducts are most commonly used by vessels which have to operate at high loadings at very low speed, such as tugs, barges, or trawlers.
+In those circumstances, relatively small vessels need to exert considerable forces against substantial loads, at very low water-speeds.
+Also, in these circumstances, the practical size of a propeller might be constrained, since operating an optimally sized propeller might increase the complexity of machiner required to deliver the power to so low a hub in the water, and it might also be difficult to arrange for the thrust to be directed substantially against the centre of drag of the vessel.
+
+The effect of the nozzle in these circumstances actually increase the optimal efficiency of the propeller, when operating at very low water-speeds, since the pressure reduction on the inside of the nozzle induced by the movement of the water through the nozzle, actually produces a net thrust forward, since the interior face at the opening of the nozzle is necessarily angled forward.
+This has the effect of distributing the suction force across a larger area, which now includes the interior of the nozzle.
+An equivalent way of describing the same effect is that the nozzle causes the water to be accelerated more gradually, over a longer distance, by inducing it to move more quickly as it passes through to the entrance.
+As a broad generalisation, more gradual or gentle actions tend to involve lower overall losses to turbulence, and hence the efficiency of the propeller can be increased.
+
+A further means by which the efficiency of a propeller can be increased is through the reduction of the turbulence created by the blade-tips passing through the open (static) water at high speed.
+Since the two sides of the blades have low and high pressure on them respectively, at the tips where the two faces meet, there there is a tendancy for water to 'spill' from the high pressure side the the low pressure side, creating a vortex which leads to unwanted turbulence.
+It is also around this vortex that blade tip cavitation tends to occur.
+By limiting the movement of water around the outside of the blade tip by having a tightly fitted shroud, losses due to this particulary type of tubulence, and the onset of this type of cavitation can be diminished.
+
+However, it is important to note that the existence of an accelerating duct by no means eliminates all cavitation.
+In fact, given that the static pressure at the impeller is necessarily reduced by an accleratint duct, the propensity for cavitation other surface of the propeller is generally increased, and cavitation can and does still occur in ducted propellers near the blade tips [@moulijn2015].
+
+![An example of the cavitation occurring at the blade tips of a ducted propeller. Image credit [@moulijn2015]](/Users/aidanmorrison/submarines/CavitatingDuct.png){width=500px}
+
+A detailed explanation of the means by which the ideal efficiency of propeller can be improved can be found in [@lewis1988 pp 213-222], as well as other texts on marine propulsion, such as [@carlton2007].
+A substantial work exploring the efficiencies of a range of accelerating and decelerating ducts can also be found in [@oosterveld1970].
+A common feature of these extensive studies also show that for ducted propellers, including accelerating ducted propellers, the advantage in terms of efficiency is overall restricted to low waterspeeds.
+At higher waterspeeds, the drag induced by the water-movement over the exterior of the duct begins to increase substantially, and the net thrust on the duct becomes negative.
+When the exterior of the duct experience relatively little water velocity, but the interior experiences a great deal more, the greatest advantages of an accelerating duct are realised.
+Furthermore, in general these models do not incorporate the effect of flow separation, which is acknowledged as a possibility on both the exterior and interior of ducts, which can dramatically reduce efficiency performance [@oosterveld1970 pp 20, @lewis1988 pp 214].
+
+![Accelerating ducts produce a net positive thrust at low advance ratios, but a net negative thrust at higher advance ratios as the drag on the nozzle exterior increases.
+Image credit:Open-water test results of Ka 3-65 screw series with nozzle no.
+19A [@oosterveld1970]](/Users/aidanmorrison/submarines/DuctedEfficiency.png){width=500px}
+
+### The Decelerating Duct or Pumpjet
+Decelerating ducts necessarily produce a negative thrust, since the process of raising the pressure of the water about the impeller (equivalent to slowing it down) necessitate reducing its kinetic energy and consequently momentum.
+In a wide range of literature introducing the fundamentals of ducted propellers and water-jets, or discussing their more modern development there is universal acknowledgement that the choice of a decelerating duct is generally made for the avoidance of cavitation for specialised (generally military) purposes, rather than for the achievement of increased efficiency [@carlton2007, @oosterveld1970, @mollard2011, @haimov2010, @lewis1988].
+
+In order to achieve a reduction in cavitation, however, care must be taken in the design of a decelerating duct.
+Since the negative thrust produced by the duct must necessitate additional work being done by the impeller, there can be circumstances where the additional loading of the impeller out-weighs the impacts of elevating the pressure around it.
+[@oosterveld1970 pp 24-25] provides a more thorough derivation of the circumstances in which this can be reliably achieved, which generally include larger blade area ratios or with more rows of rotors for higher loading, or circumstances where the loading of the propeller is quite low.
+This creates an inevitable and necessary tension between optimisation of a duct for propulsive efficiency and reduction of cavitation, which will be returned to in subsequent section.
+
+It is perhaps most relevant at this point to further clarify the distinction between the different names for related or similar systems.
+In [@carlton2007 p 17] pump jet is described as an "interesting development of the classical ducted propeller".
+Oosterveld, in his thesis on ducted propellers also states with regard to the decelerating duct "This ducted propeller system is the so called pumpjet" [@oosterveld1970 pp 8].
+Other authors give separate treatement to ducted propellers and pump jets, though the systems they describe are essentially similar in terms of composition, advantages, and physics [@lewis1988 pp 228].
+
+![In [@carlton2007] the pump jet is described as a development of a classical ducted propeller, with the noticeable addition of a stator](/Users/aidanmorrison/submarines/PumpjetOutline.png){width=500px}
+
+In other definitive literature on the design of pumpjets [@henderson1964] the difference between a pumpjet and a propeller is a described as being that "the stream of flow through the pumpjet is made to depart from the "natural" or free-stream sufrace that bounds the flow though a standard propeller."  According to the diagrams and the subsequent description it is clear that the definition embraces quite exactly the distinction between an accelerating duct, and a decelerating duct, namely that the flow is slowed at the point it encounters the impeller, where as in a free situation the free stream would contract at this point.
+
+The addition of some new part such as stator blades (to straighten out the flow) and other possible features which make the propeller system more like pump or turbomachinery components might be associated with systems more commonly referred to as pumpjets than ducted propellers.
+The key concepts behind the operation of these systems, however, remains essentially similar in their nature and design intent.
+
+### Waterjets
+The use of the word waterjet, as opposed to pumpjet, signifies that the jet expelled from the shroud or pump is situated above the water, which is desirable in very fast-moving surface vessels which attempt to minimise their contact with the waters surface, including high-speed planing hull vessels, including pleasure craft.
+The physics of the two systems, however, are closely related, with the very noticeable exception of the absence of flow considerations for water on the exterior of the duct considerations for waterjets.
+However, the physics relating to the impeller and stator blades or (turbomachinery) designed to add energy to the flow are generally extremely closely linked, though a larger variety of diffeent pump types can be chosen for waterjets which might differ more significantly from the natural evolution of a propeller to an axial-flow pump.
+Many of the essential equations which govern the efficiencies and cavitation performance of pumpjets and waterjets are of the same or extremely similar form, as can be seen in detailed discussions given to both, by similar authors [@wislicenus1973, @henderson1964].
+There are, however, plenty of instances in recent literature where this distinction regarding the ejection of the jet above and below the waterline is not maintained, and waterjets and pumpjets seem to be used interchangeably [@abdel2010].
+
+![A comparison of a pumpjet and waterjet as given in [@wislicenus1973]](/Users/aidanmorrison/submarines/PumpjetWaterjet.png){width=500px}
+
+# Constraints on the efficiency of pumpjets at low speed
+
+## Recent Commentary
+There has been already some attention given to the probable efficiencies of pump jets at low speed.
+Recently Andrew Davies published a piece in the The Strategist (ASPI) arguing that the pump jet may actually save fuel at transit speeds (around 8kt) and overall improve the endurance of the submarine, despite likely having lower efficiency at low speeds such as patrol speed[@davies2017].
+This argument relies pump-jet efficiency still overtaking that of propellers at what is considered a relatively low speed for many applications.
+As Andrew acknowledges, further quantification of the likely efficiency is required, as the particular chart which he cites is not quantified.
+This section aims to identify as far as is reasonably possible from public literature how the efficiency of a likely pump-jet for a submarine might perform in quantitative terms, relative to a propeller.
+
+There has also been questions put in parliamentary hearings regarding the pump jet [@patrick2017] and responses from public officials to such questions indicate that some of the literature provided to support such lines of questions was dated, and that the state of the art may have advanced considerably since this time.
+Whilst the argument that the state of the art is much advanced is obviously true, it is the case with marine propulsion, as with most sciences, that new developments tend to confirm, rather than annul, many of the workings that have been previously undertaken.
+To the extent that modern literature still supports the overall constraints and bounds provided by some of the foundational earlier works, a great deal of useful information can still be obtained from earlier works.
+In particular, the foundational theories, often derived from well-established principals of conservation of momentum, energy, as applied by the likes of Bernoulli and Euler, are as well respected and deeply depended upon today as they were centuries earlier when they were first arrived at.
+
+## Modern literature and results
+In addition, more modern works, including those which involve experimental validation of theoretical results also tend to show that restrictions on the efficiency of waterjets are enduring, and give us a clear starting-point for consideration of their probable orders of magnitude, and likely transition points indicating at what speeds pumpjets might become advantageously more efficient.
+
+A 1995 Japanese experimental study on the performance of a waterjet demonstrated a close experimental fit to a theoretical relationship which sees the propulsive efficiency of the waterjet fall to zero with craft velocity, with significant declines in efficiency commencing just under 10m/s, or somewhere around 18kt [@fujisawa1995].
+
+![A graph from a Japanese study of waterjet efficiency shows efficiency rapidly declining below craft velocities of 10m/s, in accordance with theory.
+[@fujisawa1995]](/Users/aidanmorrison/submarines/EfficiencyJapan.png){width=500px}
+
+In 2015 an Australian team including staff from the Australian Maritime College and Incat published results of an experimental study comparing the use of waterjets and propellers for medium-speed ferries operating in the 20-30kt speed range [@kamal2015].
+This study explicitly set out to establish the cross-over point for where propellers become more efficient than jets, since Incat was interested in pioneering highly efficient medium-speed vessels, after having considerable success in very high speed catamarans.
+It is again acknowledged as given that at lower speeds propellers are more efficient than waterjets.
+The study found that propellers were considerably more efficient over the entire speed range tested, often by a factor of two in terms of transport efficiency.
+
+![An Australian experimental study aimed at determining the crossover point for the efficiency of jets and propellers in the 20-30kt range revealed much higher power was required across the entire speed range.
+[@kamal2015]](/Users/aidanmorrison/submarines/EfficiencyTransport.png){width=500px}
+
+Both of these previous studies confirm the general and widespread assumption that jets are known to be more efficient at high speeds, and propellers at lower speeds.
+They also suggest that differences in efficiency could be very large, of a factor of two or more, including at speeds well over a submarine's transit speed.
+However, care must be taken making precise quantitative comparisons, since such experiements involved jets which expelled into air (waterjets) as opposed to remaining under water (pumpjets).
+Pumpjets would experience drag on the exterior of the shroud/duct, which waterjets would not, however waterjets are likely to have longer intake ducts, which might involve larger duct losses than pumpjets.
+Given that these factors might affect efficiency, and the overall design of the nozzle considerably, more direct comparison with fully submerged jets should be sought to make more direct comparisons.
+
+At IMARESTS 10th International Naval Engineering Conference, a paper was presented by BMT Defence and Rolls Royce [@giles2010], outlining a proposal for 'fully submerged waterjets' to be used on future surface ships in order to achieve better acoustic performance, particulalry for an anti-submarine warfare role.
+The jets comprise a mixed-flow pump coupled to an electric motor, and have been both extensively modelled and also tesed in demonstrator vessels.
+These may be a closer representation of the performance characteristics achievable by submerged pumpjets, since shape of the intake and shroud more closely resembles the shape of a pumpjet that might be found on a submarine.
+Importantly, this work focuses specifically on the performance of the jets as a function of ship-speed, including identifying possible crossover points.
+
+![The 'fully submerged waterjet' proposed by BMT and Rolls Royce for military surface craft may more closely represent the duct shape used on submarines than waterjets that release above the surface.
+[@giles2010]](/Users/aidanmorrison/submarines/BMTShape.png){width=500px}
+
+The paper makes no effort to justify the selection of choice of jets on the efficiency of the propulsion system, instead focussing on improved acoustic performance, and potential machinery space-saving advantages.
+The the vessel under consideration would require 250 tonnes more fuel with jets find than with propellers, the jets would require more power to reach equal waterspeed to a conventionally propelled craft right up to 30 kt.
+The paper states: "This is an encouraging result for a first iteration as the AWJ-21™ arrangement has not been fully optimised and, hence, there is further prospect of reducing the shaft power for this form.
+If the power could be reduced by 5% (e.g.
+by fur ther improving the waterjet/hull fairing), then the AWJ-21™ vessel would be better at all speeds above 25 knots." [@giles2010]
+
+![Studies on submerged waterjets for military surface craft currently show that jets require more power than conventional propellers right up to 30kt [@giles2010]](/Users/aidanmorrison/submarines/BMTPowering.png){width=500px}
+
+In paper also provides specific curves for the efficiency of the propulsor as a function of vessel speed, which demonstrated that the waterjet efficiency was lower than that of conventional propellers throughout the speed range, but declined particularly precipitously below speeds of 10kt.
+At 5kt, the propeller was more than twice as efficient as the waterjet.
+
+![Studies on submerged waterjets for military surface craft currently show that the efficiency of jets decline sharply below 10kt.
+[@giles2010]](/Users/aidanmorrison/submarines/BMTEfficiency.png){width=500px}
+
+It appears from recent examples in the literature that there is no evidence of jets approaching the same efficiency as propellers at speeds as low as 10kt.
+Indeed, there seems to be considerable evidence that jets of most kinds suffer substantial efficiency penalties relative to propellers in this regime, and that this is understood and accepted as an inevitable design trade-off to be made.
+In the case of a submarine, which might spend nearly 100% of its time travelling at 10kt or less, this disadvantage (as opposed to less than 35%, as assumed for a surface ship) would become extremely substantial, which will be quantified in later sections.
+
+Moreover, recent literature seems to confirm that efficiency of pumpjets fall off precipitously towards zero the closer that the vessel speed comes to zero.
+This stronly suggests the existence of some fundamental constraint in the physics of the operation of jets, which recent advances in technology seems not to have overcome.
+To more firmly establish the extent to which adaptation or plausible improvments in technology could or could not change this trend, it is worthwhile considering the theoretical limits that are presented by the physics.
+This should further limit the range of error that might be made in estimating realistic bounds of performance for an unknown system.
+
+## Theoretical Limitations
+
+In this section I will appraise some of the available literature, including some of the most canonical works in the design of pumpjets which is now declassified from US Navy funded research (and still extremely frequently cited by contemporary researchers).
+Before doing so, I'll attempt to summarise in simple language the mose essential conclusions in simple language.
+
+My broad conclusion is that the optimisation of a system for minimising cavitation, and hence improving acoustic performance, is a separate and independent design requirement to that of optimising a system for propulsive efficiency.
+These two design requirements can be found either to be in tension, or alignment, in different regimes.
+At high speeds, where cavitation tends to have severe impacts on propulsive efficiency of open propellers, and high mass-flow (which leads to good efficiency) can be achieved through compact ducts and pumps, pumpjets offer very good performance on both metrics.
+Hence, in this regime the two design requirements tend to align relatively well in favour of pump-jets.
+However, in very low-speed, low-load environments, where cavitation effects are negligible, and larger areas need to be engaged achieve good mass-flow and efficiency, the two design requirements align relatively well in favour of an open propeller.
+In a variety of intermediate regimes, the two requirements find themselves in tension.
+
+For this particular consideration, two questions are of paricular importance.
+The first is where the clear transitions in the regimes will tend to lie, with respect to the probable requirements of a conventional submarine.
+The second is what drives the tension between those design requirements in the intermediate regimes, and whether it is a fundamental or inescapble tension in the laws of physics, or simply a technical hurdle which imminent improvements in the state of the art are likely to overcome by engineering improvements or innovation.
+
+The answer to the first question, as discussed in earlier sections, seems to be that the operating speeds of a submarine tend to lie far below the probable levels where pumpjets are even remotely comparable to waterjets for efficiency.
+In later sections I will address further the acoustic comparisons in the speed ranges at 10kt and below.
+
+The answer to the second question seems to be that there the tension between acoustic and propulsive performance at intermediate speeds and loadings is driven by foundational physical requirements.
+That is not to say that optimal means of balancing those requirements will not still be found, and improved, as the art advances.
+However, it does mean that significantly adapting using known variables that can be changed to favour one design requirement are unlikely to yeild large improvements without some trade-off being made to the other.
+
+The reason for this trade-off occurring can be quite simply attributed to the nature of the shroud, or duct, which surrounds the the rotors of the propulsor, and the effect it has on the flow.
+If the duct acts to increase the pressure at the rotor, (which reduces cavitation) it necessarily must slow the flow at/around the rotor, which necessitates produceing a negative thrust on the duct.
+The system as a whole produces a positive thrust, becuase the impeller or rotor pushes a larger positive thrust, but the duct necessarily produces a negative thrust.
+At the speeds get lower, and the thrust required gets much much lower, this negative thrust becomes proportionally more significant, as described later.
+
+Put another way, the work required to slow the water down, then speed it up again, becomes a gradually greater penalty the the slower the water is coming in, and the less you actually want to speed it up.
+If the water is slowed enough by the duct to effectively stop, then the impeller must do the work of speeding it up (sucking it into the duct) before the duct slows it down, before the impeller raises its pressure, and the pressure accelerates it out of the nozzle afterwards.
+Compared to simply accelerating the water in one simple push, as a propeller blade does in an unconstrained flow, the efficiency penalty is very large.
+
+## The impact of duct loss on an ideal propeller
+
+Derivations of the 'ideal' efficiencies are frequently made for propellers, and jets using basic conservation laws (momentum theory) which consider simply acceleration of some column of water as actuated by a disc representing the impeller or propeller.
+A fuller description of these equations can be found in  most standard texts [@lewis1988 pp 131].
+These methods are computationally simple, and allow simple upper-bounds to be be set for efficiencies achievable for given designs under certain conditions, without considering the portential impact of other more complex effects such as skin friction over the blades, potential flow separation, or cavitation.
+As such, they can be easily relied upon to constrain the overall plausible range of values that can be achieved from universally applicable inputs, such as the velocity of water going into a duct, and the velocity of water exiting a nozzle.
+
+Such methods are used in [@wislicenus1973 pp8-14] to construct an 'ideal' jet efficiency that applies to waterjets and jetpumps.
+Whilst I won't repeat the full expansion here, but a fundamentally important early conclusion of the derivation is the distinction between 'useful work' and the total energy expended.
+since the rate of change in momentum is known to be necessarily equal to the force generated, and increasing momentum change (hence thrust) can be achieved either accelerating more mass, or accelerating mass to higher speeds.
+Given that energy embodied in a flow is proportional to it's velocity squared, but its momentum only to its velocity, a necessary and inevitable consequence is that some energy is always wasted in accelerating a fluid to a higher velocity than its surrounds.
+
+This is expressed in the equation which represents 'ideal jet efficiency', which is directly equivalent to 'ideal propeller efficiency' using momentum theory of propeller action:
+
+$$\eta_j = \frac{1}{1+\frac{\Delta V}{2V_0}} $$
+
+The inevitable consequence of this relationshiop is that efficency approaches unity when $\Delta V$ (the change in water of the water between entering and exiting the system) relative to $V_0$ approaches zero.
+Intuitive embodiments of this relationship are (equivalently) that efficiency is maximum when the least force is exerted on any given bit of water, which might be when negligibly small work is being done (extreme low speed) or when such a large mass of water is acted on that it barely needs to be pushed at all.
+This relationship would lead to the conclusion that infinitely large propellers or jets are always desirable from an efficiency perspective.
+This of course isn't practical to build, and also doesn't hold when the effects of the drag on the propeller blades, or duct, are taken into account.
+Infinitely large systems would increase the total weight, as well as having significant surface draw when moving through the water, all of which are neglected in this 'ideal' consideration.
+
+In [@wislicenus1973 pp8-14], Wislicenus expands the relationship in the case of the jet to incorporate terms to represent some of the inevitable adjustments that would occur in the case of pump jets and waterjets, particulalry the duct head loss ($K$) (fraction of energy required to push the water through the duct), as well as the additional drag on the outside of the propulsor ($\Delta T$).
+This leads to an adjusted efficiency relationship, which still neglects any losses due to pump efficiency, or change in height of the jet (which is added for consideration for waterjets, but not relevant here).
+
+$$ \eta_j = \frac{1}{1 + \frac{\Delta V}{2V_0} + K\frac{V_0}{2 \Delta V}}(1 - \frac{\Delta T}{T}) $$
+
+The conclusion of the analysis, and the plotting of these results for certain possible values of $K$ and $\Delta T$ is: "Its outstanding characteristic is that this efficiency approaches zero rather than unity for $\Delta V / V_0 = 0$, even for small duct losses".
+
+![Ideal jet efficiencies sharply appoach zero at low values of $\Delta V / V_0$, where as ideal propeller efficiencies without a duct $K = 0$ approach unity.
+Credit: [@wislicenus1973]](/Users/aidanmorrison/submarines/PumpjetEfficiency.png){width=500px}
+
+Wislicenus also acknowledges [@wislicenus1973] another author, Brandau, who makes a different assumption about the relationship between the duct loss and water velocity [@brandau1967] in his derivation (along with another thorough discussion of potential and ideal efficiencies), who produces a different peak efficiency point, but retains a similar overall shape featuring efficiency falling very rapidly to zero at low values of $\Delta V / V_0$, and contrasting this with the ideal propeller, which approaches unity in this low regime.
+
+It's important to note at this point that whilst charts such as these are generally used with dimensionless ratios on the axes (such as $\Delta V / V_0$) to allow the most general solutions to be used to solve for any given particular vessel size, speed, or loading requirement which might be required, these  confusing to many readers unfamiliar with hydrodynamics.
+For the avoidance of any confusion, for a particular fixed vessel size and drag, and fixed propulsion system design smaller values of $\Delta V / V_0$ will correspond to to the propulsion system exerting less thrust, and also moving at a lower speed under self-propulsion, since the thust required (proportional to $\Delta V$ when operating at steady state) will decline faster than the vessel speed or water speed at intake, since the drag on the vessel scales with the square of the vessel speed.
+
+![Alternative derivations of ideal jet efficiency also show sharply declining efficiency for low values of $\Delta V / V_0$ Credit: [@brandau1967]](/Users/aidanmorrison/submarines/EfficiencyBrandau.png){width=500px}
+
+The structure of these overall curves is found to be overall quite similar for both pumpjets and waterjets, though the probably values for the losses incurred in the intake duct, or on the exterior of the shroud, are obviously different [@wislicenus1973 pp13-14].
+Similar efficiency shapes can be found for discussions focussing explicitly on underwater pumpjets [@henderson1964, pp13].
+It should be noted that the theoretical efficiency derived by these general equations is consistent both in derivation and results with those produced by other authors [@lewis1988 pp227].
+
+Care should be taken, however, to note that different graphs for efficiency can be produced which have different variables on the x axis.
+For example, [@mollard2011 pp 247] a graph of ideal efficiency of a pump jet is shown using the speed of advance divided by the jet speed, or $V_1/V_2$.
+Given that the x axis used in the previous charts used $\Delta V / V_0$, which would be related to $V_2-V_1$ in this case, this has the effect of reversing the effective direction of the chart.
+In this case, for a fixed jet and vessel design, high-speed operations will be towards the left of the chart, and the precipitous decline in efficiency will occur as speed declines to zero, along with thrust, toward the right of the chart, as opposed to the left in earlier cases.
+
+![Other modern reference texts might demonstrate pump efficiencies with different x-axes, effectively reversing the way the graph should be understood for a single system operating at different speeds.
+Credit: [@mollard2011]](/Users/aidanmorrison/submarines/EfficiencyMollard.png){width=500px}
+
+Similarly, it is very important to note that the very frequently used x-axis unit $J = V_a/nD$ when showing efficiency curves for both propellers and ducted propellers (such as used in earlier figures) are of a similar orientation to the chart shown in the figure above.
+The left side of the chart will correspond to the highest number of rotations for a given amount of progress through the water.
+For vessel of fixed load, and a propeller of fixed geomtry, this will correspond to a propeller facing the higest drag force due to the vessel moving at the highest speed.
+
+Readers may observe that the efficiency for all propellers also decline to zero at the other end of the spectrum, most sharply at a point where $J$ equals or exceeds the pitch ratio of the propeller.
+As discussed earlier, these correspond to cases where the propeller is slicing along its helical path without exerting any force to move water backwards, or when the slip ratio is zero.
+In this circumstance, the rotation of the propeller would not push any net water backwards, and simply act as a paddle-wheel needlessly spinning water.
+Such circumstances are unlikely to occur to a significant degree unless a vessel was decelerating, or experienced some other force to assist its propulsion.
+It is possible that at extremely low speeds a propeller could somewhat descend to the right of optimal advance ratio, when very dramatically reduced load was encounterd because of the extreme low speed, as might be suggested by the slight reduction in propulsive efficiency of the propeller below 5kt in the BMT study.
+The conclusion if design at no higher speeds was required that using a smaller propeller area, which accelerated slightly thicker slices of water with each blade, would re-optimise efficiency.
+These losses of efficiency, however, could never supercede those of a decelerating ducted propeller or jet, which has the shroud which necessarily imposes an **additional** drag to the watercolumn in excess of any other resistance faced by the rotor.
+
+It should be clear from the above analysis that the efficiency of a pumpjet falling towards zero with a vessels waterspeed is an enduring and foundational result of basic physics.
+The addition of negative thrust on a decelerating duct, no matter how small, necessitates that efficiency must decline at some point with with blade loading, in contrast with an ideal propeller which rises to unity.
+This explains the overall shape given by the the mysterious unquantified chart cited by Andrew Davies [@davies2017] and Access Economics [@stanford2017] shown below, and also matches the shapes plotted by more recent studies as shown above, and also the shape presented in other modern publications studying the use of pumpjets [@giles2010] and waterjets [@fujisawa1995].
+
+![The mysterious efficiency comparison chart which has been doing the rounds in the Australian debate probably quite reasonably reflects theoretical physical limits of 'ideal' pumpjets and propellers.
+Credit: [@stanford2017]](/Users/aidanmorrison/submarines/PumpUp.png){width=500px}
+
+In terms of establishing plausible peaks and cross-over points, the study undertaken in [@giles2010] probably provides a sufficient indication as to the current state of the art in the relevant technologies.
+The study was undertaken with the backing of military customers with support and testing faciliteis (the US Navy), and involved two major companies who are world leaders in maritime engineering (BMT) and waterjet manufacture (Rolls Royce).
+Even if furthe technical advances were possible
+
+## The sensitivity to duct size
+
+## The
 

### 2017-12-30 DraftExec-knitr.Rnw
test this

--- 2017-12-27 DraftExec.Rmd
@@
----
-title: "Efficiency"
-author: "Aidan Morrison"
-date: "11/12/2017"
-output: bookdown::pdf_document2 
----
+\documentclass{article}
 
----
-references:
+\usepackage[backend=bibtex, sorting=none, citestyle=authoryear]{biblatex}
 
-- id: garland2005
-  title: Elementary Physics of Reactor Control Module - Fission Product Poisoning
-  author:
-  - family: Garland
-    given: Bill
-  volume: 3
-  URL: 'http://www.nuceng.ca/br_space/2017-09_4d03_6d03/misc_files/xenon_poisoning.pdf'
-  publisher: Nuclear Engineering (Originally from McMaster University)
-  type: article-journal
-  issued:
-    year: 2005
+\bibliography{referencess}
 
-- id: carlton2007
-  title: Marine Propellers and Propulsion
-  author:
-  - family: Carlton
-    given: John
-  URL: 'http://www.nuceng.ca/br_space/2017-09_4d03_6d03/misc_files/xenon_poisoning.pdf'
-  publisher: Elsevier
-  issued:
-    year: 2007
+\title{Efficiency}
+\author{Aidan Morrison}
 
-- id: liang2014
-  title: The economics of slow steaming
-  author:
-  - family: Liang
-    given: Lee Hong
-  URL: 'http://www.seatrade-maritime.com/news/americas/the-economics-of-slow-steaming.html'
-  publisher: Seatrade Maritime News
-  issued:
-    year: 2014
-    month: 10
-    day: 7
+\begin{document}
 
-- id: sanguri2012
-  title: The Guide to Slow Steaming On Ships
-  author:
-  - family: Sanguri
-    given: Mohit
-  URL: 'https://www.marineinsight.com/wp-content/uploads/2013/01/The-guide-to-slow-steaming-on-ships.pdf'
-  publisher: Marine Insight
-  issued:
-    year: 2012
-    month: 12
+<<include=FALSE>>=
+library(knitr)
+opts_chunk$set(
+concordance=TRUE
+)
+@
 
-- id: WNA2017
-  title: Nuclear Powered Ships
-  author:
-  - family: 
-    given: 
-  URL: 'http://www.world-nuclear.org/information-library/non-power-nuclear-applications/transport/nuclear-powered-ships.aspx#ECSArticleLink0'
-  publisher: World Nuclear Association
-  issued:
-    year: 2017
+<<setup, include=FALSE, cache=FALSE, echo=FALSE>>=
+opts_chunk$set(fig.path='figures/plots-', fig.align='center', fig.show='hold', eval=T, echo=F)
+options(replace.assign=TRUE,width=80)
 
-- id: WNA2009
-  title: Chernobyl Accident Appendix 1
-  author:
-  - family: 
-    given: 
-  URL: 'http://www.world-nuclear.org/information-library/safety-and-security/safety-of-plants/appendices/chernobyl-accident-appendix-1-sequence-of-events.aspx'
-  publisher: World Nuclear Association
-  issued:
-    year: 2009
+library(tidyverse)
+@
 
-- id: patrick2012
-  title: The Driving Factor in the SEA 1000 choice
-  author:
-  - family: Patrick
-    given: Rex
-  URL: 'https://corporate.siemens.com.au/content/dam/internet/siemens-com-au/root/aunz-defence-solutions/apdr-october-2015-issue-future-submarine.pdf'
-  publisher: Asia-Pacific Defence Reporter
-  issued:
-    year: 2015
-    month: 10
+\maketitle
 
-- id: hamilton1997
-  title: Jet Torque
-  author:
-  - family: Waterjets
-    given: Hamilton
-  volume: 8
-  URL: ''
-  publisher: Hamilton Waterjets
-  issued:
-    year: 1997
-    month: 3
----
+\section{Contents}
 
-```{r setup, include=FALSE}
-knitr::opts_chunk$set(echo = TRUE)
-library(ggplot2)
-library(knitr)
+Will need to find the Latex command...
 
-```
-
-# Contents
-
-# Introduction
+\section{Introduction}
 
@@
 Necessarily, this topic depends upon a considerable amount of physics, particularly around hydrodynamics, fluid mechanics, and turbomachinery, all of which might be difficult or daunting topics to usefully inform a public policy debate.
-However, in the case of a $50 billion dollar military acquisition, neglecting to engage with the technical matters that are so crucial to such a decision would be deeply foolish.
+However, in the case of a \$50 billion dollar military acquisition, neglecting to engage with the technical matters that are so crucial to such a decision would be deeply foolish.
 To that end, this paper has a very specific ambition.
@@
 
-# Executive Summary
+\subsection{The necessity of drag, and decelleration, induced by the duct}
 
-#### The difference between nuclear and conventional for speed
-This paper was commissioned in order to investigate whether pumpjets could plausibly be as efficient, or more efficient than a suitably designed open propeller for a conventional submarine.
-A crucial input to the investigation is a rough understanding of what the speeds of operation are likely to be for a submarine.
-Whereas nuclear submarines are reportedly capable of reaching speeds in excess of 30kt, and might transit long distances at such high speeds, conventional sumbarines are not thought of as being able to reach speeds far above 20kt in a sprint, and can only sustain speeds of 8-10kt for long-distance transits.
-Moreover, on patrol, a large portion of thier work is done at very low speeds, typically thought to be in the range of 2-4kt.
-
-#### The importance of low speed operation for diesel-electric submarines
-The efficiency of the propulsion system at such low speeds is of great significance for a conventional submarine, since it must rely on batteries or other air-independent propulsion sources for power when entirely submerged.
-Consequently, excessive energy consumption results in greatly reduced dived endurances.
-Since a submarine's position is vastly more likely to be discovered when it is on the surface operating its diesel engines, this ability to remain submerged for a long time is crucially important for combat operations, and in transiting through sensitive or contested areas.
-For nuclear submarines, which posess a practically infinite supply of energy from the nuclear reactor, efficiency at low speeds is of no concern.
-In fact, dispersing additional energy, (provided it can be done quietly) is probably advantageous for a nuclear submarine, since it will allow the nuclear reactor to avoid running at very low power levels, where the stability of the reactor is reduced.
-
-#### What is cavitation
-Pumpjets have been widely adopted by navies operating nuclear submarines.
-The principal advantage of a pumpjet relevant to submarines is their ability to avoid problems associated with cavitation, which is known to occur for propellers attempting to operate at high speeds.
-Cavitation, which is the rapid expansion and collapse of a bubble or void in the water, is particularly problematic for submarines, since it results in the creation of a great deal of noise which could be detected by an enemy.
-
-#### How does a pump-jet work
-A fundamental requirement of a pump-jet in order to avoid this cavitation is that the the working parts of the jet (the rotating blades inside it, known as the impeller) operate at a higher pressure than propeller blades operating in open water would.
-This means that the blades can turn at a lower speed relative to the water they are connected with, allowing a less violent action, which induces less cavitation.
-In this way, the jet does it's work less by directly acelerating the water, but by raising its pressure.
-This raised pressure is converted back to movement, which produces thrust, as it exits the jet and returns to the same pressure as the surrounding environment.
-
-#### The necessity of drag, and decelleration, induced by the duct
-The role of the shroud (or tunnel, mantel, duct) around the jet is to allow the pressure to be raised around the impeller, in a way that is not possible for an open propeller.
-By a fundamental requirements of physics, this actually requires that the water's incoming speed be **reduced** before it reaches the impeller.
-(The kinetic energy embodied in movement is converted into potential energy, or pressure.)  Whilst the duct narrowing at the nozzle also necessarily accelerates the water, (as the additinal pressure imparted by the impeller is converted back to kinetic energy) it's an essential feature of all pump-jets that the water flow is decelerated at the point of reaching the impeller.
-Consequently, an elementary form of the pump-jet is also termed the 'decelerating duct' applied to a propeller.
-Put simply, the water has to slow down to go fast again.
-
-#### What happens to a jet at low speed, in simple terms
-The problems for pump-jets arise when the water is already going slow, and you don't want it to go that much faster.
-This is the case when a jet designed for high-speed attempt to operate at a dramatically reduced speed.
-(Not to be confused with a vessel with very little water speed working very hard, as might be the case for a tug or barge.)  In this case, the slowing down and speeding up results in an unnecssary additional step which reduces greatly the efficiency of propulsive system.
-Or, put another way, overcomign the resistance of moving water through the shroud becomes much greater relative to the total thrus produced by the jet.
-
-#### Why this isn't easily noticeable in normal circumstances
-It is for this reason that waterjets of any kind are known to have an efficiency curve that falls off towards zero as the net thrust they produce also diminishes to zero (which for a given vessel will correspond to water speed).
-This doesn't mean that they don't still work at low speeds and produce some thrust.
-All it means is that far more power per unit of thrust will be required than might be at other speeds, as a higher fraction of energy is expended producing the turblence (random, round-and-round movement) inside and around the jet shroud than goes towards direct front-to-back acceleration of the water column, which produces thrust.
-
-#### What the submarine requirement tells us about the duct
-Whist estimating the exact shape and level of the efficiency curve for a particular pump-jet and propeller is impossible without detailed knowledge of their design, the over all trends of their shapes in the extremes can be known from well-established principals.
-Moreover, the particular demands of a pump-jet suited to a nuclear submarine, (eliminating all cavitation in as wide a range of operating circumstances as possible) considerably narrows the plausible range concerned.
-In order to minimise cavitation, the degree of pressure elvation at the impeller would be relatively high, or the total surfaces of the impeller blades much aslo become larger.
-Both of these design requirements necessitate changes to the duct or blade designs which would be in tension with overall efficiency, and most pronounced at the lowest of operating speeds.
-
-#### Quantitative Conclusions
-In the production of this paper I have developed a computational model which maps the impact of different efficiency curves directly to the dived range and endurance of a submarine.
-The most plausible scenarios I find include the reduction of dived endrance and range between 20% and 50%, or effectively halving time and distance that submarine may remain submerged for during combat operations at speeds around 3-5kt.
-
-#### The acoustic advantage of jets at higher speed
-My review of the literature confirms that a pumpjet may produce a significant acoustic advantage in circumstances where an open propeller would experience any degree of cavitation.
-Indeed, it has been remarked by naval researchers that pump-jets could be designed which would not cavitate past the point when the body they propel experienced caviation.
-As such, it seems perfectly plausible that pump-jets confer one advantage on a submarine, in that they can accelerate to a higher speed without cavitation occurring.
-This is known as a higher 'tactical silent speed'.
-
-#### Turbulence and flow separation at low speed
-However, at much lower speeds such as patrol speeds (2-4kt) it is most likely that a propeller will be able to operate well below the point of any cavitation inception, and would consequently also be extremely quiet.
-Moreover, a jet will necessarily incur substantially larger degrees and types of turbulence in order to produce net thrust in this regime.
-These likely include discontinuities and instability in the flow entering the duct and passing through the impeller and stator (flow separation) as the water is acelerated sufficient to be slowed, and then re-accelerated.
-In certain circumstances where resonances might arise these could be highly adverse to acoustic performance.
-However, assuming that by careful design such resonances can be eliminated (which we assume has been achieved) these effects would not result in any cavitation, and would only increase the noise attributable to turbulence in solid water, which is far less than for any cavitation.
-
-#### The acoustic question at low speed
-But in either case, given the necessarily raised levels of turblence generated by a jet than a propeller at very low speeds, the claim that the jet is quieter in this regime must rely entirely upon shielding effects from the shroud.
-These could be substantial in a directions perpendicular to the direction of travel, but would be much smaller when viewed from the aft or forward directions.
-It should also be noted that many underwater acoustic environments, sound tennds to reflect and bend in different directions as it propagates.
-As such, the claim that pump-jets are universally quieter than propellers should be treated with caution.
-In a variety of circumstances, including most operations at patrol speed, it may not always be true.
-
-#### The inevitable trade-off
-However, as a direct and necessary result of this higher tactical silent speed, some disadvantages will be incurred, owing to the additional drag induced by the shroud which prevents cavitation at the impeller.
-These are a lower dived endurance, a lower dived range, a lower overall endurance, a lower overall range, and a worse indescretion ratio.
-
-#### The contradiction between stated requirement and chosen technology
-Given the substantial emphasis which was placed on overall range and endurance of the submarine, it is difficult to understand how one particular unique requirement would be elevated so high above the other strategic and tactical advantages afforded by propellers.
-
-# Speed and Drag - Why very slow is very very $(very)^2$ economical
-
-Perhaps the most important relationship to understand is the relationship between speed and drag as it pertains to submarines.
-This is immportant because it sets out the fundamental framework as it applies to any submarine, regardless of propulsion type.
-(It's also a pretty important for planes, cars, missiles, torpedoes, and basically everything else.)
-
-Drag is the resistance that a fluid (air or water, in our case) gives to a body that is passing through it.
-Quite simply, it's a force that acts in the opposite direction.
-There are multiple sources of drag for different types of scenarios.
-For scenarios where an object is in contact with two different types of fluid (like a ship, on the ocean) or when the fluid doesn't really have contact with all of the object (supersonic flight, and supercavitating torpedoes) some more complex physics applies.
-For a fuller discussion of types of drag, see [@carlton2007]  But in the case of a submarine, which does it's business completely immersed in the ocean, the relevant physics is dominated the skin friction on the hull, which follows a very simple rule and relationship.
-The amount of drag ($F_D$) an objet experiences increases directly in proportion to the surface area $A$.
-For any given object of a certain (unchanging shape) there will be a constant coefficent ($C_D$) which reflects how aerodynamic or hydrodymanic the shape is.
-The drag is also directly proportional to the density of the fluid being moved through, $\rho$.
-(Air creates roughly one-thousandth the drag as water does on any given object at a given speed, since it's roughly one thousand times less dense.)
-
-But the most sensitive factor in this relationship is the speed at which the object moves through the liquid.
-The drag increases not with the speed ($v$), but with the **square** of the speed.
-This means if the speed doubles, the drag increases by four.
-If the speed triples, the drag increases by a factor of nine.
-
-$$F_D = \frac{1}{2}\rho v^2C_DA$$
-
-\begin{equation} \label{eq:1}
-\sum_{i=0}^{\infty} a_i x^i
+\begin{equation}
+\label{eq:1}
+F_D = \frac{1}{2}\rho v^2C_DA
 \end{equation}
-
-(Technical aside: This law applies wherever the the flow over the surface is turbulent.
-It is true that for very small objects, or very viscous fluids, or very slow movements a different apples called Stokes Equation, in the case wher Reynolds numbers are less than 1.
-Given that sea-water is not particularly viscous, and sumbarines not particularly small, Reynolds numbers are liekly to be much much greater than 1 (one or two thousand), even when moving at only one or two knots.
-Since it is unlikely that a significant proportion of the flow over the hull will be laminar, we'll use the drag equation in all modeling going forward when considering drag on the hull).
-
-It's crucial to understand, however, that drag is only a force, and doesn't directly inform us about how much energy is consumed, until we multiply it by the **distance** over which it is applied, not the time for which it is applied.
-To think about it simply, gravity exerts a force on you downwards all the time.
-But you don't expend any energy overcoming it when sitting still.
-If you climb stairs, the amount of gravitational potential you attain depends on how high up you climb, not how long you spend on the ladder or stair-case.
-
-This has a significant consequence for propulsion, since the amount of power (energy expended in a given time) required for thrust scales with the drag, multiplied by the distance covered in a given time.
-As such, power required (and fuel/battery consumption) scales with velocity **cubed** rather than velocity squared.
-This means that if you double your speed of travel, fuel consumption for a given period of time will increase by a factor of eight.
-Due to the increased speed of travel, the fuel required to cover a given distance will only quadruple.
-Hence the range that can be covered with a given amount of energy (assuming propulsion dominates energy requirements) tends to scale with the inverse of velocity squared, whereas the endurance (amount of time that can be spent travelling) scales with the inverse of velocity cubed.
-
-This has a profound impact on the operation and engineering of maritime vessels.
-Reducing velocity has such a substantial reduction on hull drag that going slower is almost always a reasonable means of conserving fuel overall.
-To the extent that time is non-critical, slower is always much much better.
-It's for this reason that during the financial crisis, many cargo shipping companies adopted the practice of slow steaming [@liang2014] in order to conserve fuel, despite this causing a range of possible new engineering issues which need to be accounted for in order to operate the engines at lower than normal power for sustained periods [@sanguri2012 pp.
-8-10].
-The fuel saving from operating even 30% slower means that around half the total energy is required for a given journey.
-Other incremental costs and inefficiences from operating 'off-design' are frequently outweight by such a dramatic reduction in overall power demand.
-
-This effect is of profound importance to understanding the operations of conventional submarines, which have extremely constrained energy stores when operating under the surface.
-Diesel fuel when burned has an energy density of approximately 45MJ/kg.
-In contrast, a lead-acid battery might have energy densities in the range of 0.08-0.14MJ/kg.
-With something like 400 times as much energy per kg embarked in diesel, operating on batteries imposes an extreme demand for economy on propulsive power.
-Happily for submarines, the square law for drag, and cubed law for power, allow an almost commensurate reduction in power demands to take place by slowing down to very slow speeds when submerged.
-Hypothetically, a conventional submarine might have a total range of 10,000nm from using it's diesel payload at 8kt.
-If an equal weight of lead-acid batteries as fuel were carried, the submarine could only travel about 25nm when submerged at the same speed in a single charge.
-But by travelling at 4kt, that quickly increases to 100nm, and 400nm at 2kt (neglecting hotel load here for simplicity).
-
-It's worthwhile pointing out that extreme demand for economy imposed by the poor energy density of batteries distinguishes submarines quite remarkably from other types of boat or ship design.
-In every other application, the propulsion system is designed around a particular speed and loading condition which it is optimally efficient for, and a band of plausible variation around this.
-For example, a cargo ship designed for operation at 22kt will have it's propeller and engine (propulsion system) very carefully designed around these speeds and the plausible levels of loading at those speeds.
-A top-speed which might be somewhat higher than 22kt would be calculated, but would be relatively unlikely to be used for any purpose other than an emergency.
-Lower speeds might be considered 15-18kt for the purposes of slow-steaming.
-However, the relative efficiency at 3kt is of little concern, since the drag at that speed will be under 2% of what it would be at 22kt, and the power required around 0.25%.
-Even if the efficiency was substantially worse, or better (by a factor of two or three even) the impact on the overall economics of the operation would be negligible when compared to marginal improvements at the design speed.
-
-The same consideration applies to other types of vessels, which might use water-jets.
-Fast pilot-boats, for instance, might spend quite a considerable time manouvering in and out of harbour or alongside at very low waterspeeds.
-However, their overall fuel consumption is still likely to be dominated by the fast section of their trip to a ship.
-If the in-harbour water-speed was 8kt, but 32kt was the optimum speed for the open-water section of their journey, the fuel requirement for a given distance could still be over ten times greater at 32kt.
-(Aside, the physics described above would say 16 times, but some reduction due to the reduction of surface area from planing might offset this.)  Consequently the owner/operator would be at least ten times as concerned about efficiency at 32kt as at 8kt, (unless the harbour transit was much much longer than the open water section), and have negligible concern about effiencies at 2kt or 3kt, which would rarely amount to more than a couple of percent of total propulsion energy demand.
-
-In stark contrast, a submarine operator might have less than 1% of the energy store available for the submerged movements at 2-3kt, and consequently be substantially **more** concerned about efficiency in this regime, particularly since this would be the regime in which all of their combat operations would be conducted.
-
-This imposes a constraint on the way in which literature on jets, propellers, and propulsion systems in general is read, since almost all of it is deeply concerned with how a particular system would operate at one particular design speed, and choosing the optimum system for that design speed and load.
-Considerable attention is given to how a propeller would work in off-design conditions relating to changes in load, such as in cases when a ship is more lightly laden, or a tug-boat is pushing a different sized ship.
-There is, however, a relative scarcity of attention given to the performance of a propulsion system operating at a dramatically different speed, with the same load.
-It also corresponds to scenarios where engineers and manufacturers seem from discussion to be far less confident in making quantitative claims about the performance of their own systems.
-(They are perfecly happy to make qualitative claims, "it would be alright", "nothing very bad would happen" etc.)
-
-Given that submarines almost by definition (since their hull shape doesn't change, and they must carry ballast to make up for any under-loading) are essentially always operating with the same load, but have to operate over an extremely wide variety of different speeds (a factor of four or more in variance) with severe efficiency concern at all of them, they represent a truely unique engineering question, seldom discussed in commercial applications.
-Consequently, there are relevantly few pieces of literature which address the present question directly.
-(Happily, not zero)  However, there are plenty of instances where the relevant physics and trade-offs are explained in depth to address related, but not identical questions.
-
-# The difference between nuclear and conventional propulsion
-
-Nuclear submarines vary quite remarkably from conventional submarines because of the means by which they generate their power.
-Because of the extremely high energy density in enriched uranium or plutonium, the reactors on board nuclear submarines generate an abundant supply of energy.
-Most submarine reactors are reportedly capable of generating between 25 to 50 megawatts (MW) of power, though Russian submarines have hundreds of megawatts of power available [@WNA2017].
-
-In contrast, the Collins Class Submarine's main motor is rated at less than 6MW, with designs for the future submarine appearing to be only slightly larger [@patrick2012].
-It is fair to say that in terms of maximum power ouput, nuclear submarines could have something like 5-10 times as much power at their disposal, and their **peak** output.
-
-The difference between the peak-powers of the submarines significantly understates how different the designs are, because the difference in the total amount of energy stored and available for use in a voyage or dive is vastly greater.
-A nuclear submarine might have literally millions of times more energy at its disposal, it is practically unlimited for all intents and purposes on a given voyage.
-Consequently, wheras a nuclear submarine might regularly conduct transits at or near it's peak power, a diesel electric submarine would probably transit at less than half the speed which it could manage at a sprint, (and use less than a quarter of the energy, as discussed earlier) and on patrol it might be operating at a tenth of the maximum speed, and use maybe just 1% of as much power on propulsion.
-In this situation, the amount of power that is drawn for lights, CO2 scrubbing, washing, cooking, heating, as well as the electronics driving the combat systems (the 'Hotel Load') might well become significant, and even be as larger or larger than what is required for propulsion.
-
-Consequently, the difference between the power output a conventional and a nuclear submarine might regularly operate at could be even quite a bit larger than the maximum amount of energy that they can deliver to their propulsion systems.
-
-Perhaps more significantly for power-design of nuclear submarines is a little-discussed phenomenon called Xenon poisoning, which affects nuclear reactors when they shut down or lower their power significantly.
-When Uranium or Plutonium atoms split (or fission) into two smaller isotopes atoms, a variety of different radio nucleides (unstable variants of atoms) are produced.
-Two of these are Xenon-135 and Iodine 135.
-Iodine is produced much more often, and decays into Xenon-135, which a half-life of about nine hours.
-This Xenon has a very special, perhaps unique role in reactors, since they very easily captures the free neutrons which cause the continued fission reactions in Uranium or Plutonium which drive the reactor.
-As such, Xenon is known as reactor 'poison' since it can kill the reactor's reactivity in very high doses.
-
-This high neutron-capture from Xenon means that it has a duplicitous relationship with the reactor's power level.
-At high levels of reactor power, lots of Xenon-135 and Iodine-135 are produced by the fission process.
-The high presence of Xenon reduced the reactor's reactivity.
-On the other hand, there are lots of neutrons available to 'burn off' the ass Xenon (which absobs the neturons to become a different isotope).
-This keeps the total level of Xenon in check.
-
-The situation becomes much more complicated when the reactor undergoes a sudden change in power output.
-If the power is lowered dramatically and suddenly, the production of Xenon continues quite rapidly for some time due to the decay of the large stock of Iodine-135.
-With less neutron flux available to 'burn off' the Xenon, the Xenon levels spike, and push down the reactor's reactivity.
-Unless the reactor is quickly raised back to relatively high power (~60%) quite quickly (an hour or less), the Xenon levels become so high that the reactor will have to be shut down, otherwise extreme (and dangerous) measures would be required in order to keep the reactor going.
-(This is essentially what lead to the Chernobyl Explosion [@WNA2009].)  A fuller discussion of Xenon poisoning can be found in [@garland2005], which demonstrates key concepts related to the poisoning effect shown in \ref{fig:ex2}.
-And you can see that in the [Introduction].
-
-The situation becomes much more complicated when the reactor undergoes a sudden change in power output.
 
@@
 \includegraphics[width=0.9\textwidth]{XenonPoison.png}
-\caption{A prototype of the Job Information dialog  }
+\caption{A prototype of the Job Information dialog \parencite{garland2005}}
+\label{fig:xenon}
+\end{figure}
+
+In Figure \ref{fig:ex2} we can see a sample plot of the mtcars data.
+
+\begin{figure}
+<<echo = FALSE>>=
+ggplot(data=iris, aes(x = Sepal.Length, y = Sepal.Width)) +
+  geom_point(aes(color=Species, shape=Species)) +
+  xlab("Sepal Length") +  ylab("Sepal Width") +
+  ggtitle("Sepal Length-Width")
+@
+\caption{An example plot}
 \label{fig:ex2}
@@
 
-Consequently, nuclear reactors aren't well suited to rapid fluctuations in power, particularly dramatic reductions in power, as these can lead to instability in the reactor core.
-If the reactor is shut down in order to avoid such dangerous circumstances, it generally cannot be started again until Xenon levels have fallen again, which can take a couple of days.
-Obviously this is never desirable for a military vessel, and hence is avoided at almost all costs.
+\printbibliography
 
-\begin{figure}
-```{r, echo=F}
-kable(head(mtcars))
-```
-\caption{A table example}
-\label{fig:tbl1}
-\end{figure}
-
-This effect has a dramatic impact on nuclear submarine design, since much of their design is oriented around being able to comfortably disperse large amounts of excess power, rather than with conserving it.
-For the propulsion system, this actually makes having an inefficient propulsor at low speeds a considerable advantage.
-Since the reactor will likely need to dispose of excess power, particularly during ramp-down, an inefficient propulsion system actually provides a useful power sink.
-Since any excess power will have to be disposed of by some other means (normally by pumping more water to remove the power as the heat) inefficiency at low speeds has no penalty, and probably a marginal benefit, since it will reduce overall demand for additional systems.
-Provided the excess turbulence inside the pumpjet isn't too noisy, wasting energy through the propulsor is useful.
-
-It should also be noted that a nuclear reactor's aversion to sudden reductions in power would also have a substantial impact on the design of a submarine's combat system, and it's demand on the Hotel Load.
-For the same reason, a high Hotel Load, or power-hungry Combat System, could actually be advantageous, as it helps to set an elevated 'floor' for power requirements, reducing the scale of fluctuations in overall power demand from the reactor due to changes in propulsion speed.
-
-# Some essential concepts
-
-### Conservation of Energy
-
-This is perhaps one of the most fundamental and well-established principals in physics.
-The essential idea is that energy can move or change in form, but it isn't ever created or destroyed.
-Machines, plants and animals all derive their energy from a particular other source, which can be measured and evalutated to establish the limits of energy available.
-Plants collect energy from sunlight falling on their leaves.
-Humans (as well as combustion engines) capture the chemical potential energy in organic matter, and release it by combining it with oxygen.
-Hydro-electric power plants turn the gravitational potential energy of water stored at a height into electricity.
-
-Conservation of energy has a particularly relevant embodiment in fluid flows, which is given it's primary expression in Bernoullis equation.
-It says that the total energy in a connected body of fluid is constant, though it can change in form between kinetic energy (movement), gravitational potential energy (it being elevated) the heat energy in the fluid, and the pressure of the fluid.
-
-In different fluids, the dynamics of how energy moves between one and another change.
-For example, in gases, heating up a confined piece of gas will increase it's pressure, or if it is unconfined, increase its volume.
-This is particularly important for understanding gas turbines.
-In the case of a liquid, however, all the molecules are in close contact, and hence can't increase in volume or pressure substantially except by the creation of steam.
-As such, in the absence of large amounts of cavitation, the terms in the relationship which are most important for our consideration are the relationships between pressure, and kinetic energy, and gravity.
-
-In the case of most waterjets which propel surface vessels, the water is lifted from the bottom of the hull at the intake to the pump, which is generally incorporated inside the hull.
-This increase in gravitational potential coincides with a slowing down of the water relative to the vessel-speed.
-Since the jet in such vessels is generally ejected at the same height as the pump, this potential is never regained, and is technically a loss, however at high speeds such a loss is small relative to the total power ouput.
-
-#### Venturi Effect and Flow Diffusion
-In the case of submarines and torpedoes the water doesn't undertake a change in height, since intake and nozzle are generally all in line with the central axis of the submarine or torpedo.
-As a consequence, the key relationship in Bernoullis equations is the relationship between the liquids velocity, and its pressure.
-The consequence is that when water moves through a pipe (duct, or shroud) it's pressure is inversely related to the square of its velocity.
-This means that when a liquid is forced to to travel through a narrowing pipe, it's pressure necessarily decreases as the velocity increases.
-This is a simple embodiment of the Venturi effect.
-The inverse process is where a pipe increases in volume, and the flow is forced to slow down in order to fill the wider area, and the pressure correspondingly increases.
-This is a process called 'diffusion', and is important to achieving high pressure levels in many types of water pumps, including those which will be particularly relevant for pumpjets for watercraft propulsion [@hamilton1997].
-
-#### Energy waste and efficiency
-The conservation of energy also has useful implications for how the efficiency of systems is thought about.
-In particular, because energy is conserved, identifying inefficiencies in a system necessarily involves identifying where energy goes to doing tasks which aren't useful for the intended purpose.
-A perfectly efficient system won't do any work that isn't for the intended purpose.
-In the case of analysing the efficiency of propulsion systems, the relevant 'work' is almost always related to moving water backwards to produce thrust.
-Moving water in directions other than backwards, including random turbulent flows which wind producing heat rather than thust, are two examples of wasted energy.
-Noise in the water also reflects energy which is wasted.
-
-### Boundary Layer
-When water or any fluid flows with some speed relative to another solid surface nearby, there is some layer adjacent to the surface in which the speed of the fluid is diminished relative to the main flow.
-At a microscopic level, there are some molecules of the fluid on the surface which will be effectively static relative to the surface.
-The layer of fluid that joins the gap between the static surface, and the part of the flow which is moving at the full flow speed, is called the boundary layer.
-Exactly how thick the boundary layer is, and how the fluid moves in the boundary layer, is extremely important for consideration of efficiency of fluid flows over and around solid surfaces.
-
-### Turbulent and Laminar Flow
-
-### Pressure, Thrust, and Momentum
-
-### Cavitation
-
-### Flow separation
-
-# What tech are we talking about? The core relationships between pumps, jets, ducts, and propellers, and all the different things we call them, their parts, and combinations
-
-# Propulsive performance of the pumpjet, or decelerating ducted propeller
-
-## Efficiency Relationships, and various curves
-
-## Practical and recent evidence of efficiency relationships
-
-# Results of numerical simulation
-
-# Acoustic Considerations
-
-# Conclusions
-
-# References
+\end{document}

### 2017-12-30 DraftExec.Rnw
test this

--- 2017-12-27 DraftExec.Rmd
@@
----
-title: "Efficiency"
-author: "Aidan Morrison"
-date: "11/12/2017"
-output: bookdown::pdf_document2 
----
+\documentclass{article}
 
----
-references:
+\usepackage[backend=bibtex, sorting=none, citestyle=authoryear]{biblatex}
 
-- id: garland2005
-  title: Elementary Physics of Reactor Control Module - Fission Product Poisoning
-  author:
-  - family: Garland
-    given: Bill
-  volume: 3
-  URL: 'http://www.nuceng.ca/br_space/2017-09_4d03_6d03/misc_files/xenon_poisoning.pdf'
-  publisher: Nuclear Engineering (Originally from McMaster University)
-  type: article-journal
-  issued:
-    year: 2005
+\bibliography{references}
 
-- id: carlton2007
-  title: Marine Propellers and Propulsion
-  author:
-  - family: Carlton
-    given: John
-  URL: 'http://www.nuceng.ca/br_space/2017-09_4d03_6d03/misc_files/xenon_poisoning.pdf'
-  publisher: Elsevier
-  issued:
-    year: 2007
+\title{Efficiency}
+\author{Aidan Morrison}
 
-- id: liang2014
-  title: The economics of slow steaming
-  author:
-  - family: Liang
-    given: Lee Hong
-  URL: 'http://www.seatrade-maritime.com/news/americas/the-economics-of-slow-steaming.html'
-  publisher: Seatrade Maritime News
-  issued:
-    year: 2014
-    month: 10
-    day: 7
+\begin{document}
 
-- id: sanguri2012
-  title: The Guide to Slow Steaming On Ships
-  author:
-  - family: Sanguri
-    given: Mohit
-  URL: 'https://www.marineinsight.com/wp-content/uploads/2013/01/The-guide-to-slow-steaming-on-ships.pdf'
-  publisher: Marine Insight
-  issued:
-    year: 2012
-    month: 12
+\maketitle
 
-- id: WNA2017
-  title: Nuclear Powered Ships
-  author:
-  - family: 
-    given: 
-  URL: 'http://www.world-nuclear.org/information-library/non-power-nuclear-applications/transport/nuclear-powered-ships.aspx#ECSArticleLink0'
-  publisher: World Nuclear Association
-  issued:
-    year: 2017
+<<setup, include=FALSE, cache=FALSE, echo=FALSE>>=
+opts_chunk$set(fig.path='figures/plots-', fig.align='center', fig.show='hold', eval=T, echo=F)
+options(replace.assign=TRUE,width=80)
+library(tidyverse)
+@
 
-- id: WNA2009
-  title: Chernobyl Accident Appendix 1
-  author:
-  - family: 
-    given: 
-  URL: 'http://www.world-nuclear.org/information-library/safety-and-security/safety-of-plants/appendices/chernobyl-accident-appendix-1-sequence-of-events.aspx'
-  publisher: World Nuclear Association
-  issued:
-    year: 2009
+\tableofcontents
 
-- id: patrick2012
-  title: The Driving Factor in the SEA 1000 choice
-  author:
-  - family: Patrick
-    given: Rex
-  URL: 'https://corporate.siemens.com.au/content/dam/internet/siemens-com-au/root/aunz-defence-solutions/apdr-october-2015-issue-future-submarine.pdf'
-  publisher: Asia-Pacific Defence Reporter
-  issued:
-    year: 2015
-    month: 10
-
-- id: hamilton1997
-  title: Jet Torque
-  author:
-  - family: Waterjets
-    given: Hamilton
-  volume: 8
-  URL: ''
-  publisher: Hamilton Waterjets
-  issued:
-    year: 1997
-    month: 3
----
-
-```{r setup, include=FALSE}
-knitr::opts_chunk$set(echo = TRUE)
-library(ggplot2)
-library(knitr)
-
-```
-
-# Contents
-
-# Introduction
+\section{Introduction}
 
@@
 
-# Executive Summary
+\section{Executive Summary}
 
-#### The difference between nuclear and conventional for speed
+\subsection{The difference between nuclear and conventional for speed}
+
 This paper was commissioned in order to investigate whether pumpjets could plausibly be as efficient, or more efficient than a suitably designed open propeller for a conventional submarine.
@@
 
-#### The importance of low speed operation for diesel-electric submarines
+\subsection{The importance of low speed operation for diesel-electric submarines}
+
 The efficiency of the propulsion system at such low speeds is of great significance for a conventional submarine, since it must rely on batteries or other air-independent propulsion sources for power when entirely submerged.
@@
 
-#### What is cavitation
+\subsection{What is cavitation}
+
 Pumpjets have been widely adopted by navies operating nuclear submarines.
@@
 
-#### How does a pump-jet work
+\subsection{How does a pump-jet work}
+
 A fundamental requirement of a pump-jet in order to avoid this cavitation is that the the working parts of the jet (the rotating blades inside it, known as the impeller) operate at a higher pressure than propeller blades operating in open water would.
@@
 
-#### The necessity of drag, and decelleration, induced by the duct
+\subsection{The necessity of drag, and decelleration, induced by the duct}
+
 The role of the shroud (or tunnel, mantel, duct) around the jet is to allow the pressure to be raised around the impeller, in a way that is not possible for an open propeller.
-By a fundamental requirements of physics, this actually requires that the water's incoming speed be **reduced** before it reaches the impeller.
+By a fundamental requirements of physics, this actually requires that the water's incoming speed be \textbf{reduced} before it reaches the impeller.
 (The kinetic energy embodied in movement is converted into potential energy, or pressure.)  Whilst the duct narrowing at the nozzle also necessarily accelerates the water, (as the additinal pressure imparted by the impeller is converted back to kinetic energy) it's an essential feature of all pump-jets that the water flow is decelerated at the point of reaching the impeller.
@@
 
-#### What happens to a jet at low speed, in simple terms
+\subsection{What happens to a jet at low speed, in simple terms}
+
 The problems for pump-jets arise when the water is already going slow, and you don't want it to go that much faster.
@@
 
-#### Why this isn't easily noticeable in normal circumstances
+\subsection{Why this isn't easily noticeable in normal circumstances}
+
 It is for this reason that waterjets of any kind are known to have an efficiency curve that falls off towards zero as the net thrust they produce also diminishes to zero (which for a given vessel will correspond to water speed).
@@
 
-#### What the submarine requirement tells us about the duct
+\subsection{What the submarine requirement tells us about the duct}
+
 Whist estimating the exact shape and level of the efficiency curve for a particular pump-jet and propeller is impossible without detailed knowledge of their design, the over all trends of their shapes in the extremes can be known from well-established principals.
@@
 
-#### Quantitative Conclusions
+\subsection{Quantitative Conclusions}
+
 In the production of this paper I have developed a computational model which maps the impact of different efficiency curves directly to the dived range and endurance of a submarine.
-The most plausible scenarios I find include the reduction of dived endrance and range between 20% and 50%, or effectively halving time and distance that submarine may remain submerged for during combat operations at speeds around 3-5kt.
+The most plausible scenarios I find include the reduction of dived endrance and range between 20\% and 50\%, or effectively halving time and distance that submarine may remain submerged for during combat operations at speeds around 3-5kt.
 
-#### The acoustic advantage of jets at higher speed
+\subsection{The acoustic advantage of jets at higher speed}
+
 My review of the literature confirms that a pumpjet may produce a significant acoustic advantage in circumstances where an open propeller would experience any degree of cavitation.
@@
 
-#### Turbulence and flow separation at low speed
+\subsection{Turbulence and flow separation at low speed}
+
 However, at much lower speeds such as patrol speeds (2-4kt) it is most likely that a propeller will be able to operate well below the point of any cavitation inception, and would consequently also be extremely quiet.
@@
 
-#### The acoustic question at low speed
+\subsection{The acoustic question at low speed}
+
 But in either case, given the necessarily raised levels of turblence generated by a jet than a propeller at very low speeds, the claim that the jet is quieter in this regime must rely entirely upon shielding effects from the shroud.
@@
 
-#### The inevitable trade-off
+\subsection{The inevitable trade-off}
+
 However, as a direct and necessary result of this higher tactical silent speed, some disadvantages will be incurred, owing to the additional drag induced by the shroud which prevents cavitation at the impeller.
@@
 
-#### The contradiction between stated requirement and chosen technology
+\subsection{The contradiction between stated requirement and chosen technology}
+
 Given the substantial emphasis which was placed on overall range and endurance of the submarine, it is difficult to understand how one particular unique requirement would be elevated so high above the other strategic and tactical advantages afforded by propellers.
 
-# Speed and Drag - Why very slow is very very $(very)^2$ economical
+\subsection{Speed and Drag - Why very slow is very very $(very)^2$ economical}
 
@@
 But the most sensitive factor in this relationship is the speed at which the object moves through the liquid.
-The drag increases not with the speed ($v$), but with the **square** of the speed.
+The drag increases not with the speed ($v$), but with the \textbf{square} of the speed .
 This means if the speed doubles, the drag increases by four.
@@
 
-$$F_D = \frac{1}{2}\rho v^2C_DA$$
-
-\begin{equation} \label{eq:1}
-\sum_{i=0}^{\infty} a_i x^i
+\begin{equation}
+\label{eq:1}
+F_D = \frac{1}{2}\rho v^2C_DA
 \end{equation}
 
-(Technical aside: This law applies wherever the the flow over the surface is turbulent.
+\footnote{Technical aside: This law applies wherever the the flow over the surface is turbulent.
 It is true that for very small objects, or very viscous fluids, or very slow movements a different apples called Stokes Equation, in the case wher Reynolds numbers are less than 1.
 Given that sea-water is not particularly viscous, and sumbarines not particularly small, Reynolds numbers are liekly to be much much greater than 1 (one or two thousand), even when moving at only one or two knots.
-Since it is unlikely that a significant proportion of the flow over the hull will be laminar, we'll use the drag equation in all modeling going forward when considering drag on the hull).
+Since it is unlikely that a significant proportion of the flow over the hull will be laminar, we'll use the drag equation in all modeling going forward when considering drag on the hull.}
 
-It's crucial to understand, however, that drag is only a force, and doesn't directly inform us about how much energy is consumed, until we multiply it by the **distance** over which it is applied, not the time for which it is applied.
+It's crucial to understand, however, that drag is only a force, and doesn't directly inform us about how much energy is consumed, until we multiply it by the \textbf{distance} over which it is applied, not the time for which it is applied.
 To think about it simply, gravity exerts a force on you downwards all the time.
@@
 This has a significant consequence for propulsion, since the amount of power (energy expended in a given time) required for thrust scales with the drag, multiplied by the distance covered in a given time.
-As such, power required (and fuel/battery consumption) scales with velocity **cubed** rather than velocity squared.
+As such, power required (and fuel/battery consumption) scales with velocity \textbf{cubed} rather than velocity squared.
 This means that if you double your speed of travel, fuel consumption for a given period of time will increase by a factor of eight.
@@
 8-10].
-The fuel saving from operating even 30% slower means that around half the total energy is required for a given journey.
+The fuel saving from operating even 30\% slower means that around half the total energy is required for a given journey.
 Other incremental costs and inefficiences from operating 'off-design' are frequently outweight by such a dramatic reduction in overall power demand.
@@
 Lower speeds might be considered 15-18kt for the purposes of slow-steaming.
-However, the relative efficiency at 3kt is of little concern, since the drag at that speed will be under 2% of what it would be at 22kt, and the power required around 0.25%.
+However, the relative efficiency at 3kt is of little concern, since the drag at that speed will be under 2\% of what it would be at 22kt, and the power required around 0.25\%.
 Even if the efficiency was substantially worse, or better (by a factor of two or three even) the impact on the overall economics of the operation would be negligible when compared to marginal improvements at the design speed.
@@
 
-In stark contrast, a submarine operator might have less than 1% of the energy store available for the submerged movements at 2-3kt, and consequently be substantially **more** concerned about efficiency in this regime, particularly since this would be the regime in which all of their combat operations would be conducted.
+In stark contrast, a submarine operator might have less than 1\% of the energy store available for the submerged movements at 2-3kt, and consequently be substantially \textbf{more} concerned about efficiency in this regime, particularly since this would be the regime in which all of their combat operations would be conducted.
 
@@
 
-# The difference between nuclear and conventional propulsion
+\section{The difference between nuclear and conventional propulsion}
 
@@
 Because of the extremely high energy density in enriched uranium or plutonium, the reactors on board nuclear submarines generate an abundant supply of energy.
-Most submarine reactors are reportedly capable of generating between 25 to 50 megawatts (MW) of power, though Russian submarines have hundreds of megawatts of power available [@WNA2017].
+Most submarine reactors are reportedly capable of generating between 25 to 50 megawatts (MW) of power, though Russian submarines have hundreds of megawatts of power available \parencite{WNA2017}.
 
-In contrast, the Collins Class Submarine's main motor is rated at less than 6MW, with designs for the future submarine appearing to be only slightly larger [@patrick2012].
-It is fair to say that in terms of maximum power ouput, nuclear submarines could have something like 5-10 times as much power at their disposal, and their **peak** output.
+In contrast, the Collins Class Submarine's main motor is rated at less than 6MW, with designs for the future submarine appearing to be only slightly larger \parencite{patrick2012}.
+It is fair to say that in terms of maximum power ouput, nuclear submarines could have something like 5-10 times as much power at their disposal, and their \textbf{peak} output.
 
@@
 A nuclear submarine might have literally millions of times more energy at its disposal, it is practically unlimited for all intents and purposes on a given voyage.
-Consequently, wheras a nuclear submarine might regularly conduct transits at or near it's peak power, a diesel electric submarine would probably transit at less than half the speed which it could manage at a sprint, (and use less than a quarter of the energy, as discussed earlier) and on patrol it might be operating at a tenth of the maximum speed, and use maybe just 1% of as much power on propulsion.
+Consequently, wheras a nuclear submarine might regularly conduct transits at or near it's peak power, a diesel electric submarine would probably transit at less than half the speed which it could manage at a sprint, (and use less than a quarter of the energy, as discussed earlier) and on patrol it might be operating at a tenth of the maximum speed, and use maybe just 1\% of as much power on propulsion.
 In this situation, the amount of power that is drawn for lights, CO2 scrubbing, washing, cooking, heating, as well as the electronics driving the combat systems (the 'Hotel Load') might well become significant, and even be as larger or larger than what is required for propulsion.
@@
 The high presence of Xenon reduced the reactor's reactivity.
-On the other hand, there are lots of neutrons available to 'burn off' the ass Xenon (which absobs the neturons to become a different isotope).
+On the other hand, there are lots of neutrons available to 'burn off' the Xenon (which absobs the neturons to become a different isotope).
 This keeps the total level of Xenon in check.
@@
 With less neutron flux available to 'burn off' the Xenon, the Xenon levels spike, and push down the reactor's reactivity.
-Unless the reactor is quickly raised back to relatively high power (~60%) quite quickly (an hour or less), the Xenon levels become so high that the reactor will have to be shut down, otherwise extreme (and dangerous) measures would be required in order to keep the reactor going.
-(This is essentially what lead to the Chernobyl Explosion [@WNA2009].)  A fuller discussion of Xenon poisoning can be found in [@garland2005], which demonstrates key concepts related to the poisoning effect shown in \ref{fig:ex2}.
-And you can see that in the [Introduction].
-
-The situation becomes much more complicated when the reactor undergoes a sudden change in power output.
+Unless the reactor is quickly raised back to relatively high power (~60\%) quite quickly (an hour or less), the Xenon levels become so high that the reactor will have to be shut down, otherwise extreme (and dangerous) measures would be required in order to keep the reactor going.
+(This is essentially what lead to the Chernobyl Explosion \parencite{WNA2009}.)  A fuller discussion of Xenon poisoning can be found in \citetitle{garland2005}  by \citeauthor{garland2005}, which demonstrates key concepts related to the poisoning effect shown in Figure \ref{fig:XenonPoison.png}.
 
 \begin{figure}
-\includegraphics[width=0.9\textwidth]{XenonPoison.png}
-\caption{A prototype of the Job Information dialog  }
-\label{fig:ex2}
+\includegraphics[width=\textwidth]{XenonPoison.png}
+\caption{Xenon poisoning effect following shutdown \parencite{garland2005}}
+\label{fig:XenonPoison.png}
 \end{figure}
@@
 Obviously this is never desirable for a military vessel, and hence is avoided at almost all costs.
-
-\begin{figure}
-```{r, echo=F}
-kable(head(mtcars))
-```
-\caption{A table example}
-\label{fig:tbl1}
-\end{figure}
 
@@
 
-# Some essential concepts
+\section{Some essential concepts}
 
-### Conservation of Energy
+\subsection{Conservation of Energy}
 
@@
 
-#### Venturi Effect and Flow Diffusion
+\subsubsection{Venturi Effect}
+
 In the case of submarines and torpedoes the water doesn't undertake a change in height, since intake and nozzle are generally all in line with the central axis of the submarine or torpedo.
@@
 This is a simple embodiment of the Venturi effect.
+
+\subsubsection{Flow Diffusion}
+
 The inverse process is where a pipe increases in volume, and the flow is forced to slow down in order to fill the wider area, and the pressure correspondingly increases.
-This is a process called 'diffusion', and is important to achieving high pressure levels in many types of water pumps, including those which will be particularly relevant for pumpjets for watercraft propulsion [@hamilton1997].
+This is a process called 'diffusion', and is important to achieving high pressure levels in many types of water pumps, including those which will be particularly relevant for pumpjets for watercraft propulsion \parencite{hamilton1997}.
 
-#### Energy waste and efficiency
+\subsubsection{Energy waste and efficiency}
 The conservation of energy also has useful implications for how the efficiency of systems is thought about.
@@
 
-### Boundary Layer
+\subsection{Boundary Layer}
+
 When water or any fluid flows with some speed relative to another solid surface nearby, there is some layer adjacent to the surface in which the speed of the fluid is diminished relative to the main flow.
@@
 Exactly how thick the boundary layer is, and how the fluid moves in the boundary layer, is extremely important for consideration of efficiency of fluid flows over and around solid surfaces.
+In particular, a boundary layer can be either turbulent or laminar in nature, and can transition to turbulent flow after a short distnace of laminar flow, as shown in Figure \ref{fig:BoundaryLayer.png}.
 
-### Turbulent and Laminar Flow
+\begin{figure}
+\includegraphics[width=\textwidth]{BoundaryLayer.png}
+\caption{The development of a boundary layer \parencite{mollard2011}}
+\label{fig:BoundaryLayer.png}
+\end{figure}
 
-### Pressure, Thrust, and Momentum
+\subsection{Turbulent and Laminar Flow}
 
-### Cavitation
+The way that fluids move relative to a surface can one of either two methods.
+In 'laminar flow' all the fluid moves in one direction in an smooth and orderly manner, with very little mixing between the layers of fluid travelling at different speeds.
 
-### Flow separation
+The alternative is 'turbulent flow', where the fluid moves around in unpredictable swirls and circles as well as moving overall in an predominant direction.
+There is considerable mixing between all the different layers, and the average speed remains relatively constant in the flow, with the exception of the flow immediately adjacent to the wall, or in the boundary layer.
+Becuase turbulence involves a lot of movement that is not overall in one productive direction and contributing to thrust, it necessarily leads to some loss of energy from the overall thrust.
+Once the swirls and movements become smaller and smaller, the energy winds up simply as heat in the fluid.
 
-# What tech are we talking about? The core relationships between pumps, jets, ducts, and propellers, and all the different things we call them, their parts, and combinations
+\begin{figure}
+\includegraphics[width=\textwidth]{LaminarTurbulent.png}
+\caption{A simple comparison of laminar and turbulent flows from \citetitle{NPdotnet2017} \citeyear{NPdotnet2017}}
+\label{fig:LaminarTurbulent.png}
+\end{figure}
 
-# Propulsive performance of the pumpjet, or decelerating ducted propeller
+The likely point of transition between a flow being turbulent and laminar can be given by unitless number called the Reynolds Number, which is determined by a fluids velocity, viscoscity, density, and a characteristic length-scale over which the flow occurring.
+A fuller discussion of these important concepts can be found in \citetitle{NPdotnet2017} \citeyear{NPdotnet2017}.
 
-## Efficiency Relationships, and various curves
+Turbulent flows in gasses can be extremely noisy, since those fluids are compressible.
+Consequently, the rapid and intense ciruclar movements result in a lot of oscillatory compressions against the surrounding air.
+Consequently things like jet engines, vacuum cleaners, hand-dryers, and other devices that create rapid movements in gasses tend to be quite loud, including at some distance.
 
-## Practical and recent evidence of efficiency relationships
+In essentially incompressible fluids, the random changes in pressure and velocity which are involved in laminar flow tend not to generate nearly as much noise in the surrounding liquid outside the flow.
+Because the fluid is incompressible, all of the random 'round and round' movements don't amount to much 'in and out' movement, which is what creates the pressure waves which result in propagated noise.
+Consequently, whilst turbulent flows do inevitably generate some noise, the amount of noise generated is dramatically greater in the presence of cavitation, where a void opens up in the water, and vastly more 'in and out' movement is able to occur.
 
-# Results of numerical simulation
+\subsection{Cavitation}
 
-# Acoustic Considerations
+Cavitation is the rapid expansion and collapse of a void, or bubble in water.
+Put technically, caviation occurs when the local static pressure (pressure in the rest frame of the fluid) falls below the vapour pressure of the fluid (pressure at which the liquid will start to boil).
 
-# Conclusions
+Imagine what happens when something moves very fast through water.
+The front side of the object pushes the water forward, but on the back side the water has to push in to fill the space left behind.
+If there’s not enough pressure to push the solid water in fast enough, a gap opens up, with just a few gaseous water molecules (steam) inside the cavity.
+(This is also described as water boiling at low pressure.)  However the gap doesn’t stay around for long.
+Soon the water catches up, and the bubble implodes with a pop, leaving only tiny bubbles as a result, which you can see in the wake of almost any boat or ship moving at speed.
+Whilst some energy turns to heat (the remaining steam in the tiny bubbles) quite a bit is propagated away as a sound-wave generated by the implosion.
 
-# References
+A watching water boil in a glass kettle gives a quick and intuitive insight into the occurrence of cavitation.
+Quickly after the kettle starts heating, a considerable noise can be heard, which corresponds to the commencement of cavitation on heating element.
+At some local point for a moment in time, there is enough energy for the water present to boil.
+It tends to be on rough surfaces or in the presence of some impurity that cavitation will occur first (in the blackened part of the surface in this case.)  However, whilst the bubbles on the bottom can be seen plainly on the bottom, they collapse almost straight away again, and leave only a tiny bubble of stable steam ciculating in the water.
+It is the rapid expansion and collapse of these bubbles that causes the noise of a kettle, long before it has boiled.
+
+\begin{figure}
+\includegraphics[width=\textwidth]{EarlyCavitation.JPG}
+\caption{Cavitation bubbles form and collapse creating noise and some tiny bubbles long before boiling occurs}
+\label{fig:EarlyCavitation.JPG}
+\end{figure}
+
+It is only when the water is all at a much higher temperature that the bubbles remain their full size for long enough to detatch and rise all the way to the surface, a process we typically think of as boiling water.
+It is worthwhile noting that the sound emitted at this stage is much softer and lower than the early onset of cavitation.
+Larger bubbles result in lower frequencies of sound being emitted.
+Even at higher temperature, the rougher parts of the surface provide the points where all the cavitation originate.
+
+\begin{figure}
+\includegraphics[width=\textwidth]{BoilingCavitation.JPG}
+\caption{Only at very high temperatures do the bubbles endure at full-size in the water and reach the surface}
+\label{fig:BoilingCavitation.JPG}
+\end{figure}
+
+Cavitation is extremely important in the study of ship propulsion, since it's occurrence in particular circumstances can lead to substantial losses of efficiency, as well as damage to the propeller and related appendages.
+It is particularly important for submarines, since the expansion and collapse of these bubbles tends to lead to the creation of noise which is often far larger, and more distincly characteristic than other turbulent disturbences in the water when no cavitation is present.
+Consequently, the onset of cavitation can be thought of as a distinct threshold in terms of the acoustic performance of propulsive system.
+
+Despite cavitation necessarily representing some energy being wasted generating unwanted noise in the water, some cavitation inevitably occurs around most propulsion systems operating at full power.
+In plenty of cases, the consequences for efficiency are relatively small, as they tend to be dominated by other efficiency considerations in imparting thrust.
+Put another way, the savings that can be gained by minimising wastage to turblence can outweigh the losses incurred by having some cavitation occur.
+In most cases, propellers are designed to work with a certain extent of cavitation for optimum efficiency under working loads \parencite{shin2015}.
+Cavitation also occurs to a considerable extent within the jets of most high-speed surface vessels, but doesn't necessarily have a particularly bad effect on their overall efficiency at their intended speed levels.
+It is generally only within very specialised military circumstances when the absolute avoidance of cavitation supercedes other concerns for efficiency, and require the elimination of all cavitation entirely \parencite{mollard2011}, \parencite{lewis1988}.
+These circumstances include the design of submarines and torpedoes.
+
+In certain special cases, particularly supercavitating propellers or surface-piercing drives, high levels of cavitation can be higly advantageous for efficiency.
+However these represent specialised designs, generally only for very high speed vessels, where acoustic concerns are negligible, and are of little concern for this particular endeavour.
+
+Cavitation can occur in a number of different ways at different points and forms on or around the propeller.
+It is common for cavitation to occur first near the out extremities of the propeller, where the blades are moving at the highest velocity relative to the water-flow.
+Cavitation tends to spread across the back of the blades in a sheet (sheet cavitation), since the back side of the blades are generally the areas of most depressed pressure, though it can also occur in places on the back side of the blades, where the dynamics of the water moving around the blade can produce points of near the edges of significantly reduced pressure.
+On surface vessels cavitation tends to occur most when the blades are closes to the surface, where the pressure is lowest.
+
+\begin{figure}
+\includegraphics[width=0.7\textwidth]{IncreasingCavitation.png}
+\caption{Cavitation frequently occurs first at the extremities, then spreads inwards across the blad face, as shown in \parencite{mollard2011}}
+\label{fig:IncreasingCavitation.png}
+\end{figure}
+
+\begin{figure}
+\includegraphics[width=0.7\textwidth]{CavitationTypes.png}
+\caption{Different types of cavitation can occur on different parts of the propulsion system as shown in \parencite{carlton2007}}
+\label{fig:CavitationTypes.png}
+\end{figure}
+
+\subsection{Pressure, Thrust, and Momentum change}
+
+It is worthwhile spending just a moment to touch on the relationship between a few different concepts which are closely related.
+Thrust is a force, measured in Newtons, which is generally used to refer to the forces acting on a vessel or object by its propulsion system to move it forward in the water.
+Generating thrust is consequently a key objective of any propulsion system.
+
+Pressure is a measure of force divided by an area over which it is applied.
+Pressures are necessarily applied to some extent over all of the surfaces of any object, including due to the atmostpheric pressure of air.
+In considering the efficiency of propulsion systems for solid bodies, thrust is always the consequence of some \textbf{net} pressure difference on different sides of an object.
+For instance, on a propeller blade, the rear-facing side of the blade is the side that pushes the water backwards, and hence has a raised pressure on its surface.
+Being generally backwards facing this acts to push the propeller forwards.
+This pressure acts equally to push water backwards, as well as to push the propeller forwards.
+Importantly, the forward-facing side of the propeller will expereince a lowered pressure, which also serves to pull water backwards, and suck the propeller forwards.
+It is the sum of all the pressure differences on the propeller that result in a net force.
+
+Importantly in hydrodynamics, it is important to consider the possibility of other forces being generated other surfaces which may have an impact on the movement of the vessel.
+In the presence of the hull, water being drawn by the propeller can cause pressure changes on other surface of the vessel which may assist or impede producing a net positive thrust.
+Submerged jets represent a particular case in point.
+A famous home experiment suggests that one compare the thrust (backwards pressure) experienced by someone holding a garden hose creating a jet of water when it is in the air, as opposed to submerged in a pool of water.
+When submerged, the backwards pressue appears dramatically reduced.
+In fact, this phenomenon is due to a the water-jet drawing adjacent water with it under-water, the movement of which exerts a negative pressure on the outside of the nozzle, conteracting the pressure exerted on the inside of the nozzle.
+This effect represents one of the important distinctions between the jets which release their water below or above the water-line.
+
+It is also an important law of physics that force or thrust is equal to the rate of change of momentum.
+Put simply, the speed at which one speeds something up, or the increasing amount of something which begins to move, is always equivalent in magnitude to the thrust that is produced.
+Since momentum is a vector, this simple law requires that the momentum change has to net out to one particular direction in order to produce a net thrust.
+Circular movements necessarily are cancelling, and produce no thrust.
+It is useful to recall however that this principal is simply another complete and way of measuring the consequences of the net pressure differences discussed earlier.
+The total rate of change of momentum in the water-column which is induced as a conseqence of the pressure differentials on all sides of the propeller must necessarily be equivalent to the total net force that is produced.
+Consequently literature seeking to optimise propulsion might discuss analysis referring to both of these effects or phenomena, and neither are in conflict.
+
+Full and proper derivations of many of the thrust equations for different propulsion systems from these principals can be found in plenty of authoritative works, including \parencite{lewis1988}, \parencite{carlton2007} and \parencite{mollard2011}.
+
+\subsection{Flow separation}
+
+Flow separation refers to a circumstance when the flow over a surface separates entirely from the surface, and a new eddy or votex is formed where the flow actually moves in the opposite direction to the dominant flow.
+It is of particular significance in hydrodynamics because it can lead to substantial losses of efficiency, as energy is diverted into the kinetic energy of the eddy or vortex, which is unproductive.
+It tends to occur near the surface of aerofoils with high angles of attack, or in the blades of turbomachinery including diffuser or mixed-flow pumps and ducted propellers operating in off-design conditions, including low flow rates \parencite{li2013}  \parencite{bakker2006} \parencite{mcbride1979} \parencite{wislicenus1986}.
+
+\begin{figure}
+\includegraphics[width=\textwidth]{SeparatedAerofoil.jpg}
+\caption{Flow separation occurring over the top of an aerofoil in a wind tunnel. Image courtesy of Deutsches Zentrum fuer Luft- und Raumfahrt e. V. (DLR)}
+\label{fig:SeparatedAerofoil.jpg}
+\end{figure}
+
+In order for flow separation to occur, a fluid must be moving against an adverse pressure gradient, which simply means that the pressure is increasing in the direction of movement, which means the flow is slowing down as it moves.
+At the boundary layer, where the flow velocity is already reduced due to drag forces experienced near the surface, the adverse pressure gradient can be enough to reverse the flow altogether.
+In this case, the fluid flows in reverse near the surface and forms a vortex, and the main flow becomes separated form the surface by the votex.
+
+\begin{figure}
+\includegraphics[width=\textwidth]{FlowSeparation.png}
+\caption{Flow separation involves flow reversal near a surface when a fluid is moving against an adverse pressure gradient, as shown in \parencite{mollard2011}}
+\label{fig:FlowSeparation.png}
+\end{figure}
+
+\section{end}
+
+In Figure \ref{fig:ex2} we can see a sample plot of the mtcars data.
+
+\begin{figure}
+<<fig=TRUE, echo=FALSE>>=
+ggplot(data=iris, aes(x = Sepal.Length, y = Sepal.Width)) +
+  geom_point(aes(color=Species, shape=Species)) +
+  xlab("Sepal Length") +  ylab("Sepal Width") +
+  ggtitle("Sepal Length-Width")
+@
+\caption{An example plot}
+\label{fig:ex2}
+\end{figure}
+
+\printbibliography
+
+\end{document}

### 2017-12-30 DraftExec.tex
test this

--- 2017-12-30 DraftExec.Rnw
@@
-\documentclass{article}
+\documentclass{article}\usepackage[]{graphicx}\usepackage[]{color}
+%% maxwidth is the original width if it is less than linewidth
+%% otherwise use linewidth (to make sure the graphics do not exceed the margin)
+\makeatletter
+\def\maxwidth{ %
+  \ifdim\Gin@nat@width>\linewidth
+    \linewidth
+  \else
+    \Gin@nat@width
+  \fi
+}
+\makeatother
+
+\definecolor{fgcolor}{rgb}{0.345, 0.345, 0.345}
+\newcommand{\hlnum}[1]{\textcolor[rgb]{0.686,0.059,0.569}{#1}}%
+\newcommand{\hlstr}[1]{\textcolor[rgb]{0.192,0.494,0.8}{#1}}%
+\newcommand{\hlcom}[1]{\textcolor[rgb]{0.678,0.584,0.686}{\textit{#1}}}%
+\newcommand{\hlopt}[1]{\textcolor[rgb]{0,0,0}{#1}}%
+\newcommand{\hlstd}[1]{\textcolor[rgb]{0.345,0.345,0.345}{#1}}%
+\newcommand{\hlkwa}[1]{\textcolor[rgb]{0.161,0.373,0.58}{\textbf{#1}}}%
+\newcommand{\hlkwb}[1]{\textcolor[rgb]{0.69,0.353,0.396}{#1}}%
+\newcommand{\hlkwc}[1]{\textcolor[rgb]{0.333,0.667,0.333}{#1}}%
+\newcommand{\hlkwd}[1]{\textcolor[rgb]{0.737,0.353,0.396}{\textbf{#1}}}%
+\let\hlipl\hlkwb
+
+\usepackage{framed}
+\makeatletter
+\newenvironment{kframe}{%
+ \def\at@end@of@kframe{}%
+ \ifinner\ifhmode%
+  \def\at@end@of@kframe{\end{minipage}}%
+  \begin{minipage}{\columnwidth}%
+ \fi\fi%
+ \def\FrameCommand##1{\hskip\@totalleftmargin \hskip-\fboxsep
+ \colorbox{shadecolor}{##1}\hskip-\fboxsep
+     % There is no \\@totalrightmargin, so:
+     \hskip-\linewidth \hskip-\@totalleftmargin \hskip\columnwidth}%
+ \MakeFramed {\advance\hsize-\width
+   \@totalleftmargin\z@ \linewidth\hsize
+   \@setminipage}}%
+ {\par\unskip\endMakeFramed%
+ \at@end@of@kframe}
+\makeatother
+
+\definecolor{shadecolor}{rgb}{.97, .97, .97}
+\definecolor{messagecolor}{rgb}{0, 0, 0}
+\definecolor{warningcolor}{rgb}{1, 0, 1}
+\definecolor{errorcolor}{rgb}{1, 0, 0}
+\newenvironment{knitrout}{}{} % an empty environment to be redefined in TeX
+
+\usepackage{alltt}
 
@@
 \author{Aidan Morrison}
-
+\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
 \begin{document}
@@
 \maketitle
-
-<<setup, include=FALSE, cache=FALSE, echo=FALSE>>=
-opts_chunk$set(fig.path='figures/plots-', fig.align='center', fig.show='hold', eval=T, echo=F)
-options(replace.assign=TRUE,width=80)
-library(tidyverse)
-@
 
@@
 In contrast, the Collins Class Submarine's main motor is rated at less than 6MW, with designs for the future submarine appearing to be only slightly larger \parencite{patrick2012}.
-It is fair to say that in terms of maximum power ouput, nuclear submarines could have something like 5-10 times as much power at their disposal, and their \textbf{peak} output.
+It is fair to say that in terms of maximum power ouput, nuclear submarines could have something like 5-10 times as much power at their disposal, and their **peak** output.
 
@@
 \begin{figure}
-\includegraphics[width=\textwidth]{XenonPoison.png}
+\includegraphics[width=0.9\textwidth]{XenonPoison.png}
 \caption{Xenon poisoning effect following shutdown \parencite{garland2005}}
@@
 \begin{figure}
-\includegraphics[width=\textwidth]{BoundaryLayer.png}
+\includegraphics[width=0.9\textwidth]{BoundaryLayer.png}
 \caption{The development of a boundary layer \parencite{mollard2011}}
@@
 \begin{figure}
-\includegraphics[width=\textwidth]{LaminarTurbulent.png}
+\includegraphics[width=0.9\textwidth]{LaminarTurbulent.png}
 \caption{A simple comparison of laminar and turbulent flows from \citetitle{NPdotnet2017} \citeyear{NPdotnet2017}}
@@
 \begin{figure}
-\includegraphics[width=\textwidth]{EarlyCavitation.JPG}
+\includegraphics[width=0.9\textwidth]{EarlyCavitation.JPG}
 \caption{Cavitation bubbles form and collapse creating noise and some tiny bubbles long before boiling occurs}
@@
 \begin{figure}
-\includegraphics[width=\textwidth]{BoilingCavitation.JPG}
+\includegraphics[width=0.9\textwidth]{BoilingCavitation.JPG}
 \caption{Only at very high temperatures do the bubbles endure at full-size in the water and reach the surface}
@@
 
-\begin{figure}
-\includegraphics[width=\textwidth]{FlowSeparation.png}
-\caption{Flow separation involves flow reversal near a surface when a fluid is moving against an adverse pressure gradient, as shown in \parencite{mollard2011}}
-\label{fig:FlowSeparation.png}
-\end{figure}
-
 \section{end}
@@
 \begin{figure}
-<<fig=TRUE, echo=FALSE>>=
-ggplot(data=iris, aes(x = Sepal.Length, y = Sepal.Width)) +
-  geom_point(aes(color=Species, shape=Species)) +
-  xlab("Sepal Length") +  ylab("Sepal Width") +
-  ggtitle("Sepal Length-Width")
-@
+\begin{knitrout}
+\definecolor{shadecolor}{rgb}{0.969, 0.969, 0.969}\color{fgcolor}
+
+{\centering \includegraphics[width=\maxwidth]{figures/plots-unnamed-chunk-1-1}
+
+}
+
+\end{knitrout}
 \caption{An example plot}

### 2017-12-30 DraftExec.Rnw
little pic reference

@@
 
-\title{Efficiency}
+\title{Efficiency and acoustic considerations for pumpjets in the context of conventionally powered submarines}
 \author{Aidan Morrison}
@@
 
-It is only when the water is all at a much higher temperature that the bubbles remain their full size for long enough to detatch and rise all the way to the surface, a process we typically think of as boiling water.
+It is only when the water is all at a much higher temperature that the bubbles remain their full size for long enough to detatch and rise all the way to the surface, a process we typically think of as boiling water, as shown in Figure ~\ref{fig:BoilingCavitation.JPG}.
 It is worthwhile noting that the sound emitted at this stage is much softer and lower than the early onset of cavitation.

### 2017-12-30 DraftExec.tex
little pic reference

@@
 In contrast, the Collins Class Submarine's main motor is rated at less than 6MW, with designs for the future submarine appearing to be only slightly larger \parencite{patrick2012}.
-It is fair to say that in terms of maximum power ouput, nuclear submarines could have something like 5-10 times as much power at their disposal, and their **peak** output.
+It is fair to say that in terms of maximum power ouput, nuclear submarines could have something like 5-10 times as much power at their disposal, and their \textbf{peak} output.
 
@@
 \begin{figure}
-\includegraphics[width=0.9\textwidth]{XenonPoison.png}
+\includegraphics[width=\textwidth]{XenonPoison.png}
 \caption{Xenon poisoning effect following shutdown \parencite{garland2005}}
@@
 \begin{figure}
-\includegraphics[width=0.9\textwidth]{BoundaryLayer.png}
+\includegraphics[width=\textwidth]{BoundaryLayer.png}
 \caption{The development of a boundary layer \parencite{mollard2011}}
@@
 \begin{figure}
-\includegraphics[width=0.9\textwidth]{LaminarTurbulent.png}
+\includegraphics[width=\textwidth]{LaminarTurbulent.png}
 \caption{A simple comparison of laminar and turbulent flows from \citetitle{NPdotnet2017} \citeyear{NPdotnet2017}}
@@
 \begin{figure}
-\includegraphics[width=0.9\textwidth]{EarlyCavitation.JPG}
+\includegraphics[width=\textwidth]{EarlyCavitation.JPG}
 \caption{Cavitation bubbles form and collapse creating noise and some tiny bubbles long before boiling occurs}
@@
 \begin{figure}
-\includegraphics[width=0.9\textwidth]{BoilingCavitation.JPG}
+\includegraphics[width=\textwidth]{BoilingCavitation.JPG}
 \caption{Only at very high temperatures do the bubbles endure at full-size in the water and reach the surface}
@@
 
+\begin{figure}
+\includegraphics[width=\textwidth]{FlowSeparation.png}
+\caption{Flow separation involves flow reversal near a surface when a fluid is moving against an adverse pressure gradient, as shown in \parencite{mollard2011}}
+\label{fig:FlowSeparation.png}
+\end{figure}
+
 \section{end}

### 2017-12-30 DraftExec.Rnw
last effort that got a problem rendering

@@
 
-\subsection{Speed and Drag - Why very slow is very very $(very)^2$ economical}
+\section{Speed and Drag - Why very slow is very very $(very)^2$ economical}
 
@@
 For scenarios where an object is in contact with two different types of fluid (like a ship, on the ocean) or when the fluid doesn't really have contact with all of the object (supersonic flight, and supercavitating torpedoes) some more complex physics applies.
-For a fuller discussion of types of drag, see [@carlton2007]  But in the case of a submarine, which does it's business completely immersed in the ocean, the relevant physics is dominated the skin friction on the hull, which follows a very simple rule and relationship.
+For a fuller discussion of types of drag, see \parencite{carlton2007}  But in the case of a submarine, which does it's business completely immersed in the ocean, the relevant physics is dominated the skin friction on the hull, which follows a very simple rule and relationship.
 The amount of drag ($F_D$) an objet experiences increases directly in proportion to the surface area $A$.
@@
 
-\section{end}
+In addition to the necessary loss of efficiency, flow separation can often tend to result in unsteady flows, with the vortices periodically being shed into the flow [@mollard2011 pp.
+480].
+If this occurs ahead of the blades of a propeller or impeller, such disturbances to the flow can lead to instances of cavitation when otherwise a steady homogenous flow might be well below the cavitation inception point, and can be consequential for the acoustic performance of the system, particulaly if such shedding resonates with characteristic frequencies of any of the machinery.
+
+\section{Basics of Ducted Propellers and Pumpjets}
+
+It's easy to be somewhat confused by the many different names which seem to be associated with related, or similar systems.
+Between pumps, pumpjets, water-jets ducted or shrouded propellers, or impellers, there is plenty of grounds for some confusion  In this section, I aim to clarify in simple terms what the important differences between distinctly different systems are, and where some terms are used somewhat interchangeably without doing any great violence to the concepts underlying.
+
+\subsection{Propellers}
+
+Perhaps the best starting point is the most basic, and oldest of the systems which we're considering: the basic screw propeller.
+A propeller uses a number of tilted blades attached to a central hub to sweep around disc in the water, and accelerate a column of water passing through the disc.
+Due to the slant of the blades, water on the back sides of the blades is pushed astern, and reduced pressure on the front faces pulls more water from ahead to replace it.
+As such, it accelerates a column of water, which necessarily is contracted in circumference after the point of acceleration.
+
+\begin{figure}
+\includegraphics[width=\textwidth]{PropellerAction.png}
+\caption{Propellers generate thrust by accelerating a column of water, as shown in \parencite{mollard2011} page 247.}
+\label{fig:PropellerAction.png}
+\end{figure}
+
+A comprehensive description of the development of the propeller can be found in \parencite{@carlton2007}.
+Here I won't elaborate on beyond describing some essential features and characteristics with which one ought to be familiar, primarily for the purpose of comparing different propellers and their evolution into impellers of different designs.
+
+Most propellers will have between three and seven blades.
+In general they are shaped as an aerofoil, with the convex side being upstream, just as the convex side of a plane wing is above, in order to generate lower pressure and lift as it moves through the air.
+The blades tend to be twised so that they have a higher angle of attack closer to the hub, and lesser closer to the extremities, so that those faster moving sections push their respective parts of the water column at an overall similar speed.
+The are often also thinner towards the extremities, and will be swept backwards as if dragged by their rotation through the water (skew) and also dragged by the ship's movement throug the water (rake), as seen in Figure ~\ref{fig:SkewRake.png}.
+Whilst the degree of all these characteristics is highly variable for different applications, these are a few of the commonly referred to characteristics which can be varied in order to optimise performance for any given application.
+
+\begin{figure}
+\includegraphics[width=\textwidth]{SkewRake.png}
+\caption{Propeller geometry including skew and rake, as shown in as shown in \parencite{mollard2011} page 262.}
+\label{fig:SkewRake.png}
+\end{figure}
+
+\subsubsection{Pitch}
+Perhaps one of the most important characteristics of a propeller is its pitch.
+The pitch represents the distance that a blade section would travel forward if it carved its way around a helix through one full rotation.
+It is intuitive to think of as something akin to the angle-of-attack of the propeller blade to the water, however this is technically misleading as the movement of the water-column incoming to the blade, as well as the speed of rotation, also have a significant bearing on what the actual angle of attack of the blade ends up being.
+Whilst technically pitch is a distance (measured in meters), it is often expressed and used in formula as a ratio of the diameter of the propeller, to give it a unitless equivalent, as is common in the description of propellers and propulsion systems.
+
+\begin{figure}
+\includegraphics[width=\textwidth]{HelicalLine.png}
+\caption{A blade section of a propeller traces a helical line around a cylinder as it is rotated as shown in \parencite{carlton2007}}
+\label{fig:HelicalLine.png}
+\end{figure}
+
+\begin{figure}
+\includegraphics[width=\textwidth]{PitchDefinition.png}
+\caption{The definition of pitch is the distance traveled by a blade section along a cylinder, as shown in \parencite{carlton2007}}
+\label{fig:PitchDefinition.png}
+\end{figure}
+
+The selection of the exact blade selection that is selected to define pitch must be specified as being at some fraction of the radius from the centre.
+This distance is calculated to be the 'moment mean' or a technical derived effective average, which tends to lie between 0.6R and 0.7R.
+A thorough discussion of pitch can be found in \parencite{carlton2007} pages 35-37.
+
+Pitch is of particular importance to the discussion the efficiency of propeller and propulsor design because its optimal choice tends to vary considerably with the different loads which a vessel is intended to operate at, which can also relate to a vessel's design speed.
+It is for this reason that a considerable number of vessels actually have variable pitch propellers.
+Whilst considerably more complex, expensive, and heavier than a traditional fixed-pitch propeller, the ability to vary the pitch of a propeller assists considerably, particularly when the amount of load (resistance) a vessel is expected to face varies considerably.
+Controllable pitch propellers also have advangates in terms of manouverability, since the pitch can be reversed and hence reverse thrust can be produced, without requiring the direction of the power coming from the engines to be reversed.
+This has advantages for ferries which undertake frequent docking manouvres in confined spaces \parencite{MAN2017} \parencite{lewis1988}.
 

### 2018-01-01 DraftExec.Rnw
migrated! Some Citations missing

@@
 
-\title{Efficiency and acoustic considerations for pumpjets in the context of conventionally powered submarines}
+\title{Efficiency}
 \author{Aidan Morrison}
@@
 \subsection{The difference between nuclear and conventional for speed}
-
 This paper was commissioned in order to investigate whether pumpjets could plausibly be as efficient, or more efficient than a suitably designed open propeller for a conventional submarine.
@@
 \subsection{The importance of low speed operation for diesel-electric submarines}
-
 The efficiency of the propulsion system at such low speeds is of great significance for a conventional submarine, since it must rely on batteries or other air-independent propulsion sources for power when entirely submerged.
@@
 \subsection{What is cavitation}
-
 Pumpjets have been widely adopted by navies operating nuclear submarines.
@@
 \subsection{How does a pump-jet work}
-
 A fundamental requirement of a pump-jet in order to avoid this cavitation is that the the working parts of the jet (the rotating blades inside it, known as the impeller) operate at a higher pressure than propeller blades operating in open water would.
@@
 \subsection{The necessity of drag, and decelleration, induced by the duct}
-
 The role of the shroud (or tunnel, mantel, duct) around the jet is to allow the pressure to be raised around the impeller, in a way that is not possible for an open propeller.
@@
 \subsection{What happens to a jet at low speed, in simple terms}
-
 The problems for pump-jets arise when the water is already going slow, and you don't want it to go that much faster.
@@
 \subsection{Why this isn't easily noticeable in normal circumstances}
-
 It is for this reason that waterjets of any kind are known to have an efficiency curve that falls off towards zero as the net thrust they produce also diminishes to zero (which for a given vessel will correspond to water speed).
@@
 \subsection{What the submarine requirement tells us about the duct}
-
 Whist estimating the exact shape and level of the efficiency curve for a particular pump-jet and propeller is impossible without detailed knowledge of their design, the over all trends of their shapes in the extremes can be known from well-established principals.
@@
 \subsection{Quantitative Conclusions}
-
 In the production of this paper I have developed a computational model which maps the impact of different efficiency curves directly to the dived range and endurance of a submarine.
@@
 \subsection{The acoustic advantage of jets at higher speed}
-
 My review of the literature confirms that a pumpjet may produce a significant acoustic advantage in circumstances where an open propeller would experience any degree of cavitation.
@@
 \subsection{Turbulence and flow separation at low speed}
-
 However, at much lower speeds such as patrol speeds (2-4kt) it is most likely that a propeller will be able to operate well below the point of any cavitation inception, and would consequently also be extremely quiet.
@@
 \subsection{The acoustic question at low speed}
-
 But in either case, given the necessarily raised levels of turblence generated by a jet than a propeller at very low speeds, the claim that the jet is quieter in this regime must rely entirely upon shielding effects from the shroud.
@@
 \subsection{The inevitable trade-off}
-
 However, as a direct and necessary result of this higher tactical silent speed, some disadvantages will be incurred, owing to the additional drag induced by the shroud which prevents cavitation at the impeller.
@@
 \subsection{The contradiction between stated requirement and chosen technology}
-
 Given the substantial emphasis which was placed on overall range and endurance of the submarine, it is difficult to understand how one particular unique requirement would be elevated so high above the other strategic and tactical advantages afforded by propellers.
@@
 For scenarios where an object is in contact with two different types of fluid (like a ship, on the ocean) or when the fluid doesn't really have contact with all of the object (supersonic flight, and supercavitating torpedoes) some more complex physics applies.
-For a fuller discussion of types of drag, see \parencite{carlton2007}  But in the case of a submarine, which does it's business completely immersed in the ocean, the relevant physics is dominated the skin friction on the hull, which follows a very simple rule and relationship.
+For a fuller discussion of types of drag, see \cite{carlton2007}  But in the case of a submarine, which does it's business completely immersed in the ocean, the relevant physics is dominated the skin friction on the hull, which follows a very simple rule and relationship.
 The amount of drag ($F_D$) an objet experiences increases directly in proportion to the surface area $A$.
@@
 To the extent that time is non-critical, slower is always much much better.
-It's for this reason that during the financial crisis, many cargo shipping companies adopted the practice of slow steaming [@liang2014] in order to conserve fuel, despite this causing a range of possible new engineering issues which need to be accounted for in order to operate the engines at lower than normal power for sustained periods [@sanguri2012 pp.
-8-10].
+It's for this reason that during the financial crisis, many cargo shipping companies adopted the practice of slow steaming \parencite{liang2014} in order to conserve fuel, despite this causing a range of possible new engineering issues which need to be accounted for in order to operate the engines at lower than normal power for sustained periods \parencite[8-10]{sanguri2012}.
 The fuel saving from operating even 30\% slower means that around half the total energy is required for a given journey.
@@
 In stark contrast, a submarine operator might have less than 1\% of the energy store available for the submerged movements at 2-3kt, and consequently be substantially \textbf{more} concerned about efficiency in this regime, particularly since this would be the regime in which all of their combat operations would be conducted.
-
+%
 This imposes a constraint on the way in which literature on jets, propellers, and propulsion systems in general is read, since almost all of it is deeply concerned with how a particular system would operate at one particular design speed, and choosing the optimum system for that design speed and load.
@@
 Unless the reactor is quickly raised back to relatively high power (~60\%) quite quickly (an hour or less), the Xenon levels become so high that the reactor will have to be shut down, otherwise extreme (and dangerous) measures would be required in order to keep the reactor going.
-(This is essentially what lead to the Chernobyl Explosion \parencite{WNA2009}.)  A fuller discussion of Xenon poisoning can be found in \citetitle{garland2005}  by \citeauthor{garland2005}, which demonstrates key concepts related to the poisoning effect shown in Figure \ref{fig:XenonPoison.png}.
+(This is essentially what lead to the Chernobyl Explosion \parencite{WNA2009}.)  A fuller discussion of Xenon poisoning can be found in \cite{garland2005}, which demonstrates key concepts related to the poisoning effect shown in Figure \ref{fig:XenonPoison.png}.
 
@@
 \includegraphics[width=\textwidth]{XenonPoison.png}
-\caption{Xenon poisoning effect following shutdown \parencite{garland2005}}
+\caption{Xenon poisoning effect following shutdown from \parencite{garland2005}}
 \label{fig:XenonPoison.png}
@@
 \subsubsection{Venturi Effect}
-
 In the case of submarines and torpedoes the water doesn't undertake a change in height, since intake and nozzle are generally all in line with the central axis of the submarine or torpedo.
@@
 \subsubsection{Flow Diffusion}
-
 The inverse process is where a pipe increases in volume, and the flow is forced to slow down in order to fill the wider area, and the pressure correspondingly increases.
@@
 \subsection{Boundary Layer}
-
 When water or any fluid flows with some speed relative to another solid surface nearby, there is some layer adjacent to the surface in which the speed of the fluid is diminished relative to the main flow.
@@
 \includegraphics[width=\textwidth]{BoundaryLayer.png}
-\caption{The development of a boundary layer \parencite{mollard2011}}
+\caption{The development of a boundary layer as shown in \parencite{mollard2011}}
 \label{fig:BoundaryLayer.png}
@@
 \subsection{Turbulent and Laminar Flow}
-
 The way that fluids move relative to a surface can one of either two methods.
@@
 \includegraphics[width=\textwidth]{LaminarTurbulent.png}
-\caption{A simple comparison of laminar and turbulent flows from \citetitle{NPdotnet2017} \citeyear{NPdotnet2017}}
+\caption{A simple comparison of laminar and turbulent flows from \parencite{NPdotnet2017}}
 \label{fig:LaminarTurbulent.png}
@@
 The likely point of transition between a flow being turbulent and laminar can be given by unitless number called the Reynolds Number, which is determined by a fluids velocity, viscoscity, density, and a characteristic length-scale over which the flow occurring.
-A fuller discussion of these important concepts can be found in \citetitle{NPdotnet2017} \citeyear{NPdotnet2017}.
+A fuller discussion of these important concepts can be found in \cite{NPdotnet2017}.
 
@@
 \subsection{Cavitation}
-
 Cavitation is the rapid expansion and collapse of a void, or bubble in water.
@@
 
-It is only when the water is all at a much higher temperature that the bubbles remain their full size for long enough to detatch and rise all the way to the surface, a process we typically think of as boiling water, as shown in Figure ~\ref{fig:BoilingCavitation.JPG}.
+It is only when the water is all at a much higher temperature that the bubbles remain their full size for long enough to detatch and rise all the way to the surface, a process we typically think of as boiling water.
 It is worthwhile noting that the sound emitted at this stage is much softer and lower than the early onset of cavitation.
@@
 \subsection{Pressure, Thrust, and Momentum change}
-
 It is worthwhile spending just a moment to touch on the relationship between a few different concepts which are closely related.
@@
 
-Full and proper derivations of many of the thrust equations for different propulsion systems from these principals can be found in plenty of authoritative works, including \parencite{lewis1988}, \parencite{carlton2007} and \parencite{mollard2011}.
+Full and proper derivations of many of the thrust equations for different propulsion systems from these principals can be found in plenty of authoritative works, including \cite{lewis1988}, \cite{carlton2007} and \cite{mollard2011}.
 
 \subsection{Flow separation}
-
 Flow separation refers to a circumstance when the flow over a surface separates entirely from the surface, and a new eddy or votex is formed where the flow actually moves in the opposite direction to the dominant flow.
 It is of particular significance in hydrodynamics because it can lead to substantial losses of efficiency, as energy is diverted into the kinetic energy of the eddy or vortex, which is unproductive.
-It tends to occur near the surface of aerofoils with high angles of attack, or in the blades of turbomachinery including diffuser or mixed-flow pumps and ducted propellers operating in off-design conditions, including low flow rates \parencite{li2013}  \parencite{bakker2006} \parencite{mcbride1979} \parencite{wislicenus1986}.
+It tends to occur near the surface of aerofoils with high angles of attack, or in the blades of turbomachinery including diffuser or mixed-flow pumps and ducted propellers operating in off-design conditions, including low flow rates \parencite{li2013}, parencite{bakker2006}, parencite{mcbride1979}, parencite{wislicenus1986}.
 
@@
 
-In addition to the necessary loss of efficiency, flow separation can often tend to result in unsteady flows, with the vortices periodically being shed into the flow [@mollard2011 pp.
-480].
+In addition to the necessary loss of efficiency, flow separation can often tend to result in unsteady flows, with the vortices periodically being shed into the flow \parencite[480]{mollard2011}.
 If this occurs ahead of the blades of a propeller or impeller, such disturbances to the flow can lead to instances of cavitation when otherwise a steady homogenous flow might be well below the cavitation inception point, and can be consequential for the acoustic performance of the system, particulaly if such shedding resonates with characteristic frequencies of any of the machinery.
@@
 \section{Basics of Ducted Propellers and Pumpjets}
-
 It's easy to be somewhat confused by the many different names which seem to be associated with related, or similar systems.
@@
 \subsection{Propellers}
-
 Perhaps the best starting point is the most basic, and oldest of the systems which we're considering: the basic screw propeller.
@@
 \includegraphics[width=\textwidth]{PropellerAction.png}
-\caption{Propellers generate thrust by accelerating a column of water, as shown in \parencite{mollard2011} page 247.}
+\caption{Propellers generate thrust by accelerating a column of water, as shown in  \parencite[247]{mollard2011}}
 \label{fig:PropellerAction.png}
@@
 
-A comprehensive description of the development of the propeller can be found in \parencite{@carlton2007}.
+A comprehensive description of the development of the propeller can be found in \cite{carlton2007}.
 Here I won't elaborate on beyond describing some essential features and characteristics with which one ought to be familiar, primarily for the purpose of comparing different propellers and their evolution into impellers of different designs.
@@
 The blades tend to be twised so that they have a higher angle of attack closer to the hub, and lesser closer to the extremities, so that those faster moving sections push their respective parts of the water column at an overall similar speed.
-The are often also thinner towards the extremities, and will be swept backwards as if dragged by their rotation through the water (skew) and also dragged by the ship's movement throug the water (rake), as seen in Figure ~\ref{fig:SkewRake.png}.
+The are often also thinner towards the extremities, and will be swept backwards as if dragged by their rotation through the water (skew) and also dragged by the ship's movement throug the water (rake).
 Whilst the degree of all these characteristics is highly variable for different applications, these are a few of the commonly referred to characteristics which can be varied in order to optimise performance for any given application.
@@
 \includegraphics[width=\textwidth]{SkewRake.png}
-\caption{Propeller geometry including skew and rake, as shown in as shown in \parencite{mollard2011} page 262.}
+\caption{The development of a boundary layer \parencite[262]{mollard2011}}
 \label{fig:SkewRake.png}
@@
 \includegraphics[width=\textwidth]{HelicalLine.png}
-\caption{A blade section of a propeller traces a helical line around a cylinder as it is rotated as shown in \parencite{carlton2007}}
+\caption{A blade section of a propeller traces a helical line around a cylinder as it is rotated \parencite{carlton2007}}
 \label{fig:HelicalLine.png}
@@
 \includegraphics[width=\textwidth]{PitchDefinition.png}
-\caption{The definition of pitch is the distance traveled by a blade section along a cylinder, as shown in \parencite{carlton2007}}
+\caption{The definition of pitch is the distance traveled by a blade section along a cylinder, as given in \parencite{carlton2007}}
 \label{fig:PitchDefinition.png}
@@
 This distance is calculated to be the 'moment mean' or a technical derived effective average, which tends to lie between 0.6R and 0.7R.
-A thorough discussion of pitch can be found in \parencite{carlton2007} pages 35-37.
+A thorough discussion of pitch can be found in \cite[35-37]{carlton2007}.
 
@@
 Controllable pitch propellers also have advangates in terms of manouverability, since the pitch can be reversed and hence reverse thrust can be produced, without requiring the direction of the power coming from the engines to be reversed.
-This has advantages for ferries which undertake frequent docking manouvres in confined spaces \parencite{MAN2017} \parencite{lewis1988}.
-
-In Figure \ref{fig:ex2} we can see a sample plot of the mtcars data.
+This has advantages for ferries which undertake frequent docking manouvres in confined spaces (\cite{MAN2017}, \cite{lewis1988}).
 
 \begin{figure}
-<<fig=TRUE, echo=FALSE>>=
-ggplot(data=iris, aes(x = Sepal.Length, y = Sepal.Width)) +
-  geom_point(aes(color=Species, shape=Species)) +
-  xlab("Sepal Length") +  ylab("Sepal Width") +
-  ggtitle("Sepal Length-Width")
-@
-\caption{An example plot}
-\label{fig:ex2}
+\includegraphics[width=\textwidth]{CPPropeller.png}
+\caption{Controllable Pitch Propellers (CP Propellers) allow the pitch of the blades to be varied. Image credit: \parencite{mollard2011}}
+\label{fig:CPPropeller.png}
 \end{figure}
 
-\printbibliography
+\subsubsection{Pitch and the Advance Ratio}
+Pitch is also an extremely important concept to understand because it relates closely to the way that a propeller's 'open-water efficiency' $\eta_O$ is expressed in charts, as well as the advance ratio $J$ that usually comprises the x-axis in such charts.
+
+\begin{figure}
+\includegraphics[width=\textwidth]{OpenWater.png}
+\caption{A typical chart showing open water efficiency of a propeller, in this case a Wageningen B5-75. Image: MARIN}
+\label{fig:OpenWater.png}
+\end{figure}
+
+This chart also shows two other closely related dimensionless coefficients, the torque and thrust coefficients ($K_Q$ and $K_T$) which we shall not dwell on here.
+However, it is important to note that a number of different lines are expressed on this sort of chart, which represent different versions of the same propeller, with only the pitch (expressed in its dimensionless form as $P/D$) being different in each case.
+The x-axis, $J$, also warrants a clear definition:
+
+\begin{equation}
+\label{eq:2}
+J = \frac{V_a}{nD}
+\end{equation}
+
+Where $V_a$ is the speed of advance of the propeller, or the speed at which water arrives at the front-side of the propeller, and $n$ represents the rate of the propeller's rotation (revolutions per second) and $D$ is the propeller's diameter.
+For fullness of understanding it is also worthwhile to note here that $V_a$ does not correspond directly to the vessel's speed of travel, though they are related.
+To some extent, a vessel moving through water will always draw water in its wake along with it, which means that in the wake (where the propeller tends to operate to some extent or other) the water incident on the propeller will arrive at a lower speed than that implied by the ship speed \parencite[15]{MAN2017}.
+Another key distinction for the consideration of jets later on, is that the speed of advance does not necessarily incporate the movement of water due to the movement induced by the propeller (or duct) itself, which is a distinct effect to that of the wake field.
+In \cite[213]{lewis1988} $V_A$ is used specifically to denote the undisturbed stream velocity, since the role of certain ducts can be to accelerate or decelerate the stream prior to the water arriving at the blades of the propeller.
+
+One may also note from the previous figure that for each line representing a given pitch, the efficiency falls precipitously towards zero around the point where $J$ reaches $P/D$.
+This is because this extreme of the curve represents a circumstance where the propeller is rotating, but not actually moving any water, since it is moving through the water at about that speed which corresponds to the exact helical path defined by the pitch being traced.
+This represents circumstances where a propeller encounters effectively zero load.
+This might be the case when a eases it's power, and the momentum of the ship temporarily keeps the ship moving without the propeller exerting much or any net thrust, or where alternative propoulsion sources (sails, or other propellers) are keeping a speed higher than what the propeller actually has push backwards on water to maintain.
+
+A closely related concept here is the slip ratio, which is a measure of the difference between the pitch (the distance the blade section would travel if it traced a perfect helix for one rotation) and the actual distance that the blade did advance through the water.
+This is neatly defined and shown in a diagram in \ref{fig:SlipRatio.png}.
+
+\begin{figure}
+\includegraphics[width=\textwidth]{SlipRatio.png}
+\caption{The slip ratio of a propeller. Image credit: \cite{MAN2017}}
+\label{fig:SlipRatio.png}
+\end{figure}
+
+It can be seen from the diagram of the open-water efficiencies that there is some desirable non-zero slip ratio for a propeller.
+The other end of the spectrum, where efficiencies also fall off toward zero, represent circumstances where the propeller is simply spinning in the same place in the water.
+In this case the slip ratio is one.
+This represents a circumstance known as 'bollard pull' where a vessel is simply exerting force against an immovable object or infinite load.
+Whilst the very low efficiencies here could represent a case of a propeller that is simply too small for the task it is trying to achieve, it is actually a significant and important circumstance to consider for propellers on working vessels such as tugs or trawlers, which exert large forces at very low water-speeds.
+Measurement of the thrust generated in bollard pull, and the power required to achieve this thrust, is still an important metric for such vessels, but the definition of 'useful work' for normal propulsion requieres the thust to be excerted over some distance travelled.
+
+\subsection{Ducted Propellers}
+A ducted propeller is a propeller in any kind of tubular shroud, casing or nozzle.
+Ducted propellers fall into broadly two categories, one of which is referred to frequently and interchangeably as a pump-jet.
+The two types are called the accelerating type (also known as a Kort Nozzle, after a company that made them popular in the 20th Century) and the decelerating type.
+The names describe the different effects that each type of nozzle have on the flow speed of the water at the point where it meets the propeller (which is also called an impeller, particularly when associated with decelerating ducts, or pumpjets).
+
+A detailed description of the essential differences between the two can be easily demonstrated by diagram, which show a simple comparison an acelerating and decelerating duct, which have the shape of the aerofoil section which comprises the duct inverted.
+(Aside: It should be noted that some designs attempt to strike a balance between efficiency and cavitation performance of accelerating and decelerating ducts for certain purposes, and in such cases the duct shape might not be so clearly contrasted \parencite{abdel2010}.
+But for the purposes of explanation, as well as optimisation for military purposes, the distinction between the two will generally be quite clear.)
+
+\begin{figure}
+\includegraphics[width=\textwidth]{AccelDecel.png}
+\caption{he two types of ducted propeller, accelerating (left) and decelerating (right) as shown in \parencite[249]{mollard2011}}
+\label{fig:AccelDecel.png}
+\end{figure}
+
+A consideration of what happens to the flow as it moves through each type of duct helps to demonstrate how the acceleration and deceleration effects are achieved.
+In the case of an accelerating nozzle, the area through which the flow must pass contracts prior to the water arriving at the impeller.
+In the decellerating case, the available area contracts after the water has passed the impeller.
+In some cases, a degree of diffusion is actually incorporated ahead of the impeller, whereby the area actually of the flow actually expands ahead of the impeller, which slows down the flow and increases its pressure according to Benoulli's principal \parencite{wislicenus1973}.
+
+\begin{figure}
+\includegraphics[width=\textwidth]{DuctStreamlines.png}
+\caption{The streamlines of flow through different duct types \parencite[214]{lewis1988}}
+\label{fig:DuctStreamlines.png}
+\end{figure}
+
+\subsubsection{The Accelerating Duct}
+Accelerating ducts are most commonly used by vessels which have to operate at high loadings at very low speed, such as tugs, barges, or trawlers.
+In those circumstances, relatively small vessels need to exert considerable forces against substantial loads, at very low water-speeds.
+Also, in these circumstances, the practical size of a propeller might be constrained, since operating an optimally sized propeller might increase the complexity of machiner required to deliver the power to so low a hub in the water, and it might also be difficult to arrange for the thrust to be directed substantially against the centre of drag of the vessel.
+
+The effect of the nozzle in these circumstances actually increase the optimal efficiency of the propeller, when operating at very low water-speeds, since the pressure reduction on the inside of the nozzle induced by the movement of the water through the nozzle, actually produces a net thrust forward, since the interior face at the opening of the nozzle is necessarily angled forward.
+This has the effect of distributing the suction force across a larger area, which now includes the interior of the nozzle.
+An equivalent way of describing the same effect is that the nozzle causes the water to be accelerated more gradually, over a longer distance, by inducing it to move more quickly as it passes through to the entrance.
+As a broad generalisation, more gradual or gentle actions tend to involve lower overall losses to turbulence, and hence the efficiency of the propeller can be increased.
+
+A further means by which the efficiency of a propeller can be increased is through the reduction of the turbulence created by the blade-tips passing through the open (static) water at high speed.
+Since the two sides of the blades have low and high pressure on them respectively, at the tips where the two faces meet, there there is a tendancy for water to 'spill' from the high pressure side the the low pressure side, creating a vortex which leads to unwanted turbulence.
+It is also around this vortex that blade tip cavitation tends to occur.
+By limiting the movement of water around the outside of the blade tip by having a tightly fitted shroud, losses due to this particulary type of tubulence, and the onset of this type of cavitation can be diminished.
+
+However, it is important to note that the existence of an accelerating duct by no means eliminates all cavitation.
+In fact, given that the static pressure at the impeller is necessarily reduced by an accleratint duct, the propensity for cavitation other surface of the propeller is generally increased, and cavitation can and does still occur in ducted propellers near the blade tips \parencite{moulijn2015}.
+
+\begin{figure}
+\includegraphics[width=\textwidth]{CavitatingDuct.png}
+\caption{An example of the cavitation occurring at the blade tips of a ducted propeller. Image credit \cite{moulijn2015}}
+\label{fig:CavitatingDuct.png}
+\end{figure}
+
+A detailed explanation of the means by which the ideal efficiency of propeller can be improved can be found in \cite[213-222]{lewis1988}, as well as other texts on marine propulsion, such as \cite{carlton2007}.
+A substantial work exploring the efficiencies of a range of accelerating and decelerating ducts can also be found in \cite{oosterveld1970}.
+A common feature of these extensive studies also show that for ducted propellers, including accelerating ducted propellers, the advantage in terms of efficiency is overall restricted to low waterspeeds.
+At higher waterspeeds, the drag induced by the water-movement over the exterior of the duct begins to increase substantially, and the net thrust on the duct becomes negative.
+When the exterior of the duct experience relatively little water velocity, but the interior experiences a great deal more, the greatest advantages of an accelerating duct are realised.
+Furthermore, in general these models do not incorporate the effect of flow separation, which is acknowledged as a possibility on both the exterior and interior of ducts, which can dramatically reduce efficiency performance (\cite[20]{oosterveld1970}, \cite[214]{lewis1988}).
+
+\begin{figure}
+\includegraphics[width=\textwidth]{DuctedEfficiency.png}
+\caption{Accelerating ducts produce a net positive thrust at low advance ratios, but a net negative thrust at higher advance ratios as the drag on the nozzle exterior increases.
+Image credit:Open-water test results of Ka 3-65 screw series with nozzle no.
+19A \parencite{oosterveld1970}}
+\label{fig:DuctedEfficiency.png}
+\end{figure}
+
+\subsubsection{The Decelerating Duct or Pumpjet}
+Decelerating ducts necessarily produce a negative thrust, since the process of raising the pressure of the water about the impeller (equivalent to slowing it down) necessitate reducing its kinetic energy and consequently momentum.
+In a wide range of literature introducing the fundamentals of ducted propellers and water-jets, or discussing their more modern development there is universal acknowledgement that the choice of a decelerating duct is generally made for the avoidance of cavitation for specialised (generally military) purposes, rather than for the achievement of increased efficiency (\cite{carlton2007}, \cite{oosterveld1970}, \cite{mollard2011}, \cite{haimov2010}, \cite{lewis1988}).
+
+In order to achieve a reduction in cavitation, however, care must be taken in the design of a decelerating duct.
+Since the negative thrust produced by the duct must necessitate additional work being done by the impeller, there can be circumstances where the additional loading of the impeller out-weighs the impacts of elevating the pressure around it.
+\cite[24-25]{oosterveld1970} provides a more thorough derivation of the circumstances in which this can be reliably achieved, which generally include larger blade area ratios or with more rows of rotors for higher loading, or circumstances where the loading of the propeller is quite low.
+This creates an inevitable and necessary tension between optimisation of a duct for propulsive efficiency and reduction of cavitation, which will be returned to in subsequent section.
+
+It is perhaps most relevant at this point to further clarify the distinction between the different names for related or similar systems.
+In \cite[17]{carlton2007} pump jet is described as an "interesting development of the classical ducted propeller".
+Oosterveld, in his thesis on ducted propellers also states with regard to the decelerating duct "This ducted propeller system is the so called pumpjet" \parencite[8]{oosterveld1970}.
+Other authors give separate treatement to ducted propellers and pump jets, though the systems they describe are essentially similar in terms of composition, advantages, and physics \parencite[288]{lewis1988}.
+
+\begin{figure}
+\includegraphics[width=\textwidth]{PumpjetOutline.png}
+\caption{In \cite{carlton2007} the pump jet is described as a development of a classical ducted propeller, with the noticeable addition of a stator}
+\label{fig:PumpjetOutline.png}
+\end{figure}
+
+In other definitive literature on the design of pumpjets \parencite{henderson1964} the difference between a pumpjet and a propeller is a described as being that "the stream of flow through the pumpjet is made to depart from the "natural" or free-stream sufrace that bounds the flow though a standard propeller."  According to the diagrams and the subsequent description it is clear that the definition embraces quite exactly the distinction between an accelerating duct, and a decelerating duct, namely that the flow is slowed at the point it encounters the impeller, where as in a free situation the free stream would contract at this point.
+
+The addition of some new part such as stator blades (to straighten out the flow) and other possible features which make the propeller system more like pump or turbomachinery components might be associated with systems more commonly referred to as pumpjets than ducted propellers.
+The key concepts behind the operation of these systems, however, remains essentially similar in their nature and design intent.
+
+\subsubsection{Waterjets}
+The use of the word waterjet, as opposed to pumpjet, signifies that the jet expelled from the shroud or pump is situated above the water, which is desirable in very fast-moving surface vessels which attempt to minimise their contact with the waters surface, including high-speed planing hull vessels, including pleasure craft.
+The physics of the two systems, however, are closely related, with the very noticeable exception of the absence of flow considerations for water on the exterior of the duct considerations for waterjets.
+However, the physics relating to the impeller and stator blades or (turbomachinery) designed to add energy to the flow are generally extremely closely linked, though a larger variety of diffeent pump types can be chosen for waterjets which might differ more significantly from the natural evolution of a propeller to an axial-flow pump.
+Many of the essential equations which govern the efficiencies and cavitation performance of pumpjets and waterjets are of the same or extremely similar form, as can be seen in detailed discussions given to both, by similar authors (\cite{wislicenus1973}, \cite{henderson1964}).
+There are, however, plenty of instances in recent literature where this distinction regarding the ejection of the jet above and below the waterline is not maintained, and waterjets and pumpjets seem to be used interchangeably \parencite{abdel2010}.
+
+\begin{figure}
+\includegraphics[width=\textwidth]{PumpjetWaterjet.png}
+\caption{A comparison of a pumpjet and waterjet as given in \cite{wislicenus1973}}
+\label{fig:PumpjetWaterjet.png}
+\end{figure}
+
+\section{Constraints on the efficiency of pumpjets at low speed}
+
+\subsection{Recent Commentary}
+There has been already some attention given to the probable efficiencies of pump jets at low speed.
+Recently Andrew Davies published a piece in the The Strategist (ASPI) arguing that the pump jet may actually save fuel at transit speeds (around 8kt) and overall improve the endurance of the submarine, despite likely having lower efficiency at low speeds such as patrol speed \parencite{davies2017}.
+This argument relies pump-jet efficiency still overtaking that of propellers at what is considered a relatively low speed for many applications.
+As Andrew acknowledges, further quantification of the likely efficiency is required, as the particular chart which he cites is not quantified.
+This section aims to identify as far as is reasonably possible from public literature how the efficiency of a likely pump-jet for a submarine might perform in quantitative terms, relative to a propeller.
+
+There has also been questions put in parliamentary hearings regarding the pump jet \parencite{patrick2017} and responses from public officials to such questions indicate that some of the literature provided to support such lines of questions was dated, and that the state of the art may have advanced considerably since this time.
+Whilst the argument that the state of the art is much advanced is obviously true, it is the case with marine propulsion, as with most sciences, that new developments tend to confirm, rather than annul, many of the workings that have been previously undertaken.
+To the extent that modern literature still supports the overall constraints and bounds provided by some of the foundational earlier works, a great deal of useful information can still be obtained from earlier works.
+In particular, the foundational theories, often derived from well-established principals of conservation of momentum, energy, as applied by the likes of Bernoulli and Euler, are as well respected and deeply depended upon today as they were centuries earlier when they were first arrived at.
+
+\subsection{Modern literature and results}
+In addition, more modern works, including those which involve experimental validation of theoretical results also tend to show that restrictions on the efficiency of waterjets are enduring, and give us a clear starting-point for consideration of their probable orders of magnitude, and likely transition points indicating at what speeds pumpjets might become advantageously more efficient.
+
+A 1995 Japanese experimental study on the performance of a waterjet demonstrated a close experimental fit to a theoretical relationship which sees the propulsive efficiency of the waterjet fall to zero with craft velocity, with significant declines in efficiency commencing just under 10m/s, or somewhere around 18kt \parencite{fujisawa1995}.
+
+\begin{figure}
+\includegraphics[width=\textwidth]{EfficiencyJapan.png}
+\caption{A graph from a Japanese study of waterjet efficiency shows efficiency rapidly declining below craft velocities of 10m/s, in accordance with theory. \parencite{fujisawa1995}}
+\label{fig:EfficiencyJapan.png}
+\end{figure}
+
+In 2015 an Australian team including staff from the Australian Maritime College and Incat published results of an experimental study comparing the use of waterjets and propellers for medium-speed ferries operating in the 20-30kt speed range \parencite{kamal2015}.
+This study explicitly set out to establish the cross-over point for where propellers become more efficient than jets, since Incat was interested in pioneering highly efficient medium-speed vessels, after having considerable success in very high speed catamarans.
+It is again acknowledged as given that at lower speeds propellers are more efficient than waterjets.
+The study found that propellers were considerably more efficient over the entire speed range tested, often by a factor of two in terms of transport efficiency.
+
+\begin{figure}
+\includegraphics[width=\textwidth]{EfficiencyTransport.png}
+\caption{An Australian experimental study aimed at determining the crossover point for the efficiency of jets and propellers in the 20-30kt range revealed much higher power was required across the entire speed range \parencite{kamal2015}}
+\label{fig:EfficiencyTransport.png}
+\end{figure}
+
+Both of these previous studies confirm the general and widespread assumption that jets are known to be more efficient at high speeds, and propellers at lower speeds.
+They also suggest that differences in efficiency could be very large, of a factor of two or more, including at speeds well over a submarine's transit speed.
+However, care must be taken making precise quantitative comparisons, since such experiements involved jets which expelled into air (waterjets) as opposed to remaining under water (pumpjets).
+Pumpjets would experience drag on the exterior of the shroud/duct, which waterjets would not, however waterjets are likely to have longer intake ducts, which might involve larger duct losses than pumpjets.
+Given that these factors might affect efficiency, and the overall design of the nozzle considerably, more direct comparison with fully submerged jets should be sought to make more direct comparisons.
+
+At IMARESTS 10th International Naval Engineering Conference, a paper was presented by BMT Defence and Rolls Royce \parencite{giles2010}, outlining a proposal for 'fully submerged waterjets' to be used on future surface ships in order to achieve better acoustic performance, particulalry for an anti-submarine warfare role.
+The jets comprise a mixed-flow pump coupled to an electric motor, and have been both extensively modelled and also tesed in demonstrator vessels.
+These may be a closer representation of the performance characteristics achievable by submerged pumpjets, since shape of the intake and shroud more closely resembles the shape of a pumpjet that might be found on a submarine.
+Importantly, this work focuses specifically on the performance of the jets as a function of ship-speed, including identifying possible crossover points.
+
+\begin{figure}
+\includegraphics[width=\textwidth]{BMTShape.png}
+\caption{The 'fully submerged waterjet' proposed by BMT and Rolls Royce for military surface craft may more closely represent the duct shape used on submarines than waterjets that release above the surface.
+\parencite{giles2010}}
+\label{fig:BMTShape.png}
+\end{figure}
+
+The paper makes no effort to justify the selection of choice of jets on the efficiency of the propulsion system, instead focussing on improved acoustic performance, and potential machinery space-saving advantages.
+The the vessel under consideration would require 250 tonnes more fuel with jets find than with propellers, the jets would require more power to reach equal waterspeed to a conventionally propelled craft right up to 30 kt.
+The paper states: "This is an encouraging result for a first iteration as the AWJ-21™ arrangement has not been fully optimised and, hence, there is further prospect of reducing the shaft power for this form.
+If the power could be reduced by 5\% (e.g.
+by fur ther improving the waterjet/hull fairing), then the AWJ-21™ vessel would be better at all speeds above 25 knots." \parencite{giles2010}
+
+\begin{figure}
+\includegraphics[width=\textwidth]{BMTPowering.png}
+\caption{Studies on submerged waterjets for military surface craft currently show that jets require more power than conventional propellers right up to 30kt \parencite{giles2010}}
+\label{fig:BMTPowering.png}
+\end{figure}
+
+In paper also provides specific curves for the efficiency of the propulsor as a function of vessel speed, which demonstrated that the waterjet efficiency was lower than that of conventional propellers throughout the speed range, but declined particularly precipitously below speeds of 10kt.
+At 5kt, the propeller was more than twice as efficient as the waterjet.
+
+\begin{figure}
+\includegraphics[width=\textwidth]{BMTEfficiency.png}
+\caption{Studies on submerged waterjets for military surface craft currently show that the efficiency of jets decline sharply below 10kt \parencite{giles2010}}
+\label{fig:BMTEfficiency.png}
+\end{figure}
+
+It appears from recent examples in the literature that there is no evidence of jets approaching the same efficiency as propellers at speeds as low as 10kt.
+Indeed, there seems to be considerable evidence that jets of most kinds suffer substantial efficiency penalties relative to propellers in this regime, and that this is understood and accepted as an inevitable design trade-off to be made.
+In the case of a submarine, which might spend nearly 100\% of its time travelling at 10kt or less, this disadvantage (as opposed to less than 35\%, as assumed for a surface ship) would become extremely substantial, which will be quantified in later sections.
+
+Moreover, recent literature seems to confirm that efficiency of pumpjets fall off precipitously towards zero the closer that the vessel speed comes to zero.
+This stronly suggests the existence of some fundamental constraint in the physics of the operation of jets, which recent advances in technology seems not to have overcome.
+To more firmly establish the extent to which adaptation or plausible improvments in technology could or could not change this trend, it is worthwhile considering the theoretical limits that are presented by the physics.
+This should further limit the range of error that might be made in estimating realistic bounds of performance for an unknown system.
+
+\subsection{Theoretical Limitations}
+In this section I will appraise some of the available literature, including some of the most canonical works in the design of pumpjets which is now declassified from US Navy funded research (and still extremely frequently cited by contemporary researchers).
+Before doing so, I'll attempt to summarise in simple language the mose essential conclusions in simple language.
+
+My broad conclusion is that the optimisation of a system for minimising cavitation, and hence improving acoustic performance, is a separate and independent design requirement to that of optimising a system for propulsive efficiency.
+These two design requirements can be found either to be in tension, or alignment, in different regimes.
+At high speeds, where cavitation tends to have severe impacts on propulsive efficiency of open propellers, and high mass-flow (which leads to good efficiency) can be achieved through compact ducts and pumps, pumpjets offer very good performance on both metrics.
+Hence, in this regime the two design requirements tend to align relatively well in favour of pump-jets.
+However, in very low-speed, low-load environments, where cavitation effects are negligible, and larger areas need to be engaged achieve good mass-flow and efficiency, the two design requirements align relatively well in favour of an open propeller.
+In a variety of intermediate regimes, the two requirements find themselves in tension.
+
+For this particular consideration, two questions are of paricular importance.
+The first is where the clear transitions in the regimes will tend to lie, with respect to the probable requirements of a conventional submarine.
+The second is what drives the tension between those design requirements in the intermediate regimes, and whether it is a fundamental or inescapble tension in the laws of physics, or simply a technical hurdle which imminent improvements in the state of the art are likely to overcome by engineering improvements or innovation.
+
+The answer to the first question, as discussed in earlier sections, seems to be that the operating speeds of a submarine tend to lie far below the probable levels where pumpjets are even remotely comparable to waterjets for efficiency.
+In later sections I will address further the acoustic comparisons in the speed ranges at 10kt and below.
+
+The answer to the second question seems to be that there the tension between acoustic and propulsive performance at intermediate speeds and loadings is driven by foundational physical requirements.
+That is not to say that optimal means of balancing those requirements will not still be found, and improved, as the art advances.
+However, it does mean that significantly adapting using known variables that can be changed to favour one design requirement are unlikely to yeild large improvements without some trade-off being made to the other.
+
+The reason for this trade-off occurring can be quite simply attributed to the nature of the shroud, or duct, which surrounds the the rotors of the propulsor, and the effect it has on the flow.
+If the duct acts to increase the pressure at the rotor, (which reduces cavitation) it necessarily must slow the flow at/around the rotor, which necessitates produceing a negative thrust on the duct.
+The system as a whole produces a positive thrust, becuase the impeller or rotor pushes a larger positive thrust, but the duct necessarily produces a negative thrust.
+At the speeds get lower, and the thrust required gets much much lower, this negative thrust becomes proportionally more significant, as described later.
+
+Put another way, the work required to slow the water down, then speed it up again, becomes a gradually greater penalty the the slower the water is coming in, and the less you actually want to speed it up.
+If the water is slowed enough by the duct to effectively stop, then the impeller must do the work of speeding it up (sucking it into the duct) before the duct slows it down, before the impeller raises its pressure, and the pressure accelerates it out of the nozzle afterwards.
+Compared to simply accelerating the water in one simple push, as a propeller blade does in an unconstrained flow, the efficiency penalty is very large.
+
+\subsection{The impact of duct loss on an ideal propeller}
+
+Derivations of the 'ideal' efficiencies are frequently made for propellers, and jets using basic conservation laws (momentum theory) which consider simply acceleration of some column of water as actuated by a disc representing the impeller or propeller.
+A fuller description of these equations can be found in  most standard texts \parencite[131]{lewis1988}.
+These methods are computationally simple, and allow simple upper-bounds to be be set for efficiencies achievable for given designs under certain conditions, without considering the portential impact of other more complex effects such as skin friction over the blades, potential flow separation, or cavitation.
+As such, they can be easily relied upon to constrain the overall plausible range of values that can be achieved from universally applicable inputs, such as the velocity of water going into a duct, and the velocity of water exiting a nozzle.
+
+Such methods are used in \cite[8-14]{wislicenus1973} to construct an 'ideal' jet efficiency that applies to waterjets and jetpumps.
+Whilst I won't repeat the full expansion here, but a fundamentally important early conclusion of the derivation is the distinction between 'useful work' and the total energy expended.
+since the rate of change in momentum is known to be necessarily equal to the force generated, and increasing momentum change (hence thrust) can be achieved either accelerating more mass, or accelerating mass to higher speeds.
+Given that energy embodied in a flow is proportional to it's velocity squared, but its momentum only to its velocity, a necessary and inevitable consequence is that some energy is always wasted in accelerating a fluid to a higher velocity than its surrounds.
+
+This is expressed in the equation which represents 'ideal jet efficiency', which is directly equivalent to 'ideal propeller efficiency' using momentum theory of propeller action:
+
+\begin{equation}
+\label{eq:3}
+\eta_j = \frac{1}{1+\frac{\Delta V}{2V_0}}
+\end{equation}
+
+The inevitable consequence of this relationshiop is that efficency approaches unity when $\Delta V$ (the change in water of the water between entering and exiting the system) relative to $V_0$ approaches zero.
+Intuitive embodiments of this relationship are (equivalently) that efficiency is maximum when the least force is exerted on any given bit of water, which might be when negligibly small work is being done (extreme low speed) or when such a large mass of water is acted on that it barely needs to be pushed at all.
+This relationship would lead to the conclusion that infinitely large propellers or jets are always desirable from an efficiency perspective.
+This of course isn't practical to build, and also doesn't hold when the effects of the drag on the propeller blades, or duct, are taken into account.
+Infinitely large systems would increase the total weight, as well as having significant surface draw when moving through the water, all of which are neglected in this 'ideal' consideration.
+
+In \cite[8-14]{wislicenus1973}, Wislicenus expands the relationship in the case of the jet to incorporate terms to represent some of the inevitable adjustments that would occur in the case of pump jets and waterjets, particulalry the duct head loss ($K$) (fraction of energy required to push the water through the duct), as well as the additional drag on the outside of the propulsor ($\Delta T$).
+This leads to an adjusted efficiency relationship, which still neglects any losses due to pump efficiency, or change in height of the jet (which is added for consideration for waterjets, but not relevant here).
+
+\begin{equation}
+\label{eq:4}
+\eta_j = \frac{1}{1 + \frac{\Delta V}{2V_0} + K\frac{V_0}{2 \Delta V}}(1 - \frac{\Delta T}{T})
+\end{equation}
+
+The conclusion of the analysis, and the plotting of these results for certain possible values of $K$ and $\Delta T$ is: "Its outstanding characteristic is that this efficiency approaches zero rather than unity for $\Delta V / V_0 = 0$, even for small duct losses".
+
+\begin{figure}
+\includegraphics[width=\textwidth]{PumpjetEfficiency.png}
+\caption{Ideal jet efficiencies sharply appoach zero at low values of $\Delta V / V_0$, where as ideal propeller efficiencies without a duct $K = 0$ approach unity. Credit: \cite{wislicenus1973}}
+\label{fig:PumpjetEfficiency.png}
+\end{figure}
+
+Wislicenus also acknowledges \cite{wislicenus1973} another author, Brandau, who makes a different assumption about the relationship between the duct loss and water velocity \parencite{brandau1967} in his derivation (along with another thorough discussion of potential and ideal efficiencies), who produces a different peak efficiency point, but retains a similar overall shape featuring efficiency falling very rapidly to zero at low values of $\Delta V / V_0$, and contrasting this with the ideal propeller, which approaches unity in this low regime.
+
+It's important to note at this point that whilst charts such as these are generally used with dimensionless ratios on the axes (such as $\Delta V / V_0$) to allow the most general solutions to be used to solve for any given particular vessel size, speed, or loading requirement which might be required, these  confusing to many readers unfamiliar with hydrodynamics.
+For the avoidance of any confusion, for a particular fixed vessel size and drag, and fixed propulsion system design smaller values of $\Delta V / V_0$ will correspond to to the propulsion system exerting less thrust, and also moving at a lower speed under self-propulsion, since the thust required (proportional to $\Delta V$ when operating at steady state) will decline faster than the vessel speed or water speed at intake, since the drag on the vessel scales with the square of the vessel speed.
+
+\begin{figure}
+\includegraphics[width=\textwidth]{EfficiencyBrandau.png}
+\caption{Alternative derivations of ideal jet efficiency also show sharply declining efficiency for low values of $\Delta V / V_0$ Credit: \cite{brandau1967}}
+\label{fig:EfficiencyBrandau.png}
+\end{figure}
+
+The structure of these overall curves is found to be overall quite similar for both pumpjets and waterjets, though the probably values for the losses incurred in the intake duct, or on the exterior of the shroud, are obviously different \parencite[13-14]{wislicenus1973}.
+Similar efficiency shapes can be found for discussions focussing explicitly on underwater pumpjets \parencite[13]{henderson1964}.
+It should be noted that the theoretical efficiency derived by these general equations is consistent both in derivation and results with those produced by other authors \parencite[227]{lewis1988}.
+
+Care should be taken, however, to note that different graphs for efficiency can be produced which have different variables on the x axis.
+For example, \cite[247]{mollard2011} a graph of ideal efficiency of a pump jet is shown using the speed of advance divided by the jet speed, or $V_1/V_2$.
+Given that the x axis used in the previous charts used $\Delta V / V_0$, which would be related to $V_2-V_1$ in this case, this has the effect of reversing the effective direction of the chart.
+In this case, for a fixed jet and vessel design, high-speed operations will be towards the left of the chart, and the precipitous decline in efficiency will occur as speed declines to zero, along with thrust, toward the right of the chart, as opposed to the left in earlier cases.
+
+\begin{figure}
+\includegraphics[width=\textwidth]{EfficiencyMollard.png}
+\caption{Other modern reference texts might demonstrate jet or propeller efficiencies with different x-axes, effectively reversing the way the graph should be understood for a single system operating at different speeds.
+Credit:  \cite{mollard2011}}
+\label{fig:EfficiencyMollard.png}
+\end{figure}
+
+\begin{figure}
+\includegraphics[width=\textwidth]{EfficiencyBulten.png}
+\caption{Another graph representing efficiency as a function of the ratio of jet velocity to incoming velocity, which shows efficiency declining at low thrust (to the right).
+The impact of increasing duct losses can be clearly seen, as opposed to the 'ideal' propeller without a duct.
+Credit: \cite{bulten2006numerical}}
+\label{fig:EfficiencyBulten.png}
+\end{figure}
+
+Similarly, it is very important to note that the very frequently used x-axis unit $J = V_a/nD$ when showing efficiency curves for both propellers and ducted propellers (such as used in earlier figures) are of a similar orientation to the chart shown in the figure above.
+The left side of the chart will correspond to the highest number of rotations for a given amount of progress through the water.
+For vessel of fixed load, and a propeller of fixed geomtry, this will correspond to a propeller facing the higest drag force due to the vessel moving at the highest speed.
+
+Readers may observe that the efficiency for all propellers also decline to zero at the other end of the spectrum, most sharply at a point where $J$ equals or exceeds the pitch ratio of the propeller.
+As discussed earlier, these correspond to cases where the propeller is slicing along its helical path without exerting any force to move water backwards, or when the slip ratio is zero.
+In this circumstance, the rotation of the propeller would not push any net water backwards, and simply act as a paddle-wheel needlessly spinning water.
+Such circumstances are unlikely to occur to a significant degree unless a vessel was decelerating, or experienced some other force to assist its propulsion.
+It is possible that at extremely low speeds a propeller could somewhat descend to the right of optimal advance ratio, when very dramatically reduced load was encounterd because of the extreme low speed, as might be suggested by the slight reduction in propulsive efficiency of the propeller below 5kt in the BMT study.
+The conclusion if design at no higher speeds was required that using a smaller propeller area, which accelerated slightly thicker slices of water with each blade, would re-optimise efficiency.
+These losses of efficiency, however, could never supercede those of a decelerating ducted propeller or jet, which has the shroud which necessarily imposes an \textbf{additional} drag to the watercolumn in excess of any other resistance faced by the rotor.
+This effect can be seen clearly even or an accelerating duct, as shown in \ref{fig:PropDuct.png}.
+
+\begin{figure}
+\includegraphics[width=\textwidth]{PropDuct.png}
+\caption{Even the addition of an accelerating ducts substantially reduces efficiency of propellers at very low loadings, or for constant loads, low speeds.
+Maximum attainable efficiency is also reduced.
+Their benefit lies in the bulging of the curve at low advance ratios (higher loads).
+Credit: \cite[6]{willemsen2013}}
+\label{fig:PropDuct.png}
+\end{figure}
+
+It should be clear from the above analysis that the efficiency of a pumpjet falling towards zero with a vessels waterspeed is an enduring and foundational result of basic physics.
+The addition of negative thrust on a decelerating duct, no matter how small, necessitates that efficiency must decline at some point with with blade loading, in contrast with an ideal propeller which rises to unity.
+This explains the overall shape given by the the mysterious unquantified chart cited by Andrew Davies \parencite{davies2017} and Access Economics \parencite{stanford2017} shown \ref{fig:PumpUp.png}, and also matches the shapes plotted by more recent studies as shown above, and also the shape presented in other modern publications studying the use of pumpjets \parencite{giles2010} and waterjets \parencite{fujisawa1995}.
+
+\begin{figure}
+\includegraphics[width=\textwidth]{PumpUp.png}
+\caption{The mysterious efficiency comparison chart which has been doing the rounds in the Australian debate probably quite reasonably reflects theoretical physical limits of 'ideal' pumpjets and propellers.
+Credit: \cite{stanford2017}}
+\label{fig:PumpUp.png}
+\end{figure}
+
+In terms of establishing plausible peaks and cross-over points, the study undertaken in \cite{giles2010} probably provides a sufficient indication as to the current state of the art in the relevant technologies.
+Whilst this effort focussed on a surface vessel rather than a   The study was undertaken with the backing of military customers with support and testing facilities (the US Navy), and involved two major companies who are world leaders in maritime engineering (BMT) and waterjet manufacture (Rolls Royce).
+Even if further technical advances through design refinement were possible, as the authors alude to, the cross-over point in terms of efficiency might be brought down to 25kt.
+The prospect that the crossover point lies anywhere near the transit speed of a convetional submarine (8-10kt) appears utterly implausible.
+In discussions with leading world experts involved in the development of advanced ducted propellers (including those involved in citations) I was told that a pumpjet could not be more efficient than a propeller below 18kt.
+
+With the optimum efficiency of a jet probably being no greater than 60\% or 70\% at its peak efficiency according to many theoretical and experimental studies \parencite{lu2015}, which would quite probably lie considerably higher in terms of waterspeed, these facts, along with the inevitable decline of efficiency towards zero as waterspeed also declines, a suitably constrained set of plausible efficiency curves may be investigated numerically, as will be done later in later sections.
+
+\subsection{An assessment of scope for improving the efficiency of a pumpjet at low speeds}
+Before continuing to assess the impact of the likely efficiency curves of pumpjets relative to propellers, it is worthwhile considering what the known literature about the theory of pumps, jets, and turbomachinery might tell us about the potential for a radical advance which make might substantially improve the shape of the curve at very low speeds.
+Whilst the field of propulsor design is no doubt complex, and I will not attempt to be exhaustive in detailing all of the potential design considerations which might make an improvement.
+However, the pursuit of reasonably comprehensive confidence is aided by the the scale of the gap which probably needs to be closed to make pumpjets comparably efficient to propellers at the relevant speeds.
+Minor adjustments and the improvement of small or marginal effects, of which there might be many, will clearly not help in shifting the efficiency peak of a jet from over 25kt to under 10kt.
+In other words, this gap will not be closed or noticeably reduced by 'tuning'.
+Quite substantial redesign of core characteristics (like mass flow) would be required, which would necessitate a cycle of redesign and tuning virtually all other aspect of the entire jet, as illustrated by numerous guides to the design of such systems (\cite{bruce1974}, \cite{furuya1988}, \cite{mccormick1963design}, \cite{henderson1964}).
+As such, I will only consider those parameters which could plausibly be altered quite substantially, where the consideration of larger changes make the likely limits or trade-offs quite clear.
+
+\subsubsection{Increase mass-flow by widening area of intake}
+Perhaps the most obvious means by which one might attain substantial increases in efficiency is to increase substantially the mass flow rate.
+This is derived from a fundamental relationship, as expressed earlier in the representation of ideal propeller and jet efficiency (prior to the consideration of any drag or other losses) which was given in the earlier equation.
+Since excessive velocity differences result in more energy expended than momentum changed, as given by momentum theory and basic mechanics, doing work on a large body of water, rather than working harder on more of it, is generally desirable from an efficiency perspective.
+Considering what the appropriate mass-flow should be at design speeds is generally the first thing that is specified to occur in all design processes for pumpjets, and good example of which is in \cite[6-7]{henderson1964}.
+
+As Henderson et al.
+note, the key constraint of this variable is that increasing the diameter of the pumpjet necessarily leads to an increase in the surface (including the external suface) of the shroud.
+The increased drag that this induces must be deducted from the thrust that the pumpjet produces, or effectively increasing $\Delta T$ as given in earlier equations.
+In this respect, there is "no free lunch" to be had in expanding the size of the intake.
+As Henderson et al.
+conclude, for this reason the diameter of a pumpjet tends to be about 15\% or 20\% smaller than that of a corresponding propeller, whereas the diameter to have equivalent mass-flow, highlighting the tension set up in pumpjet design with their earlier observation that "if the cavitation resistance of the blades is to be improved with no reduction in thrust and the same rate of flow through the propulsor, the pump jet rotor must be larger than that of the propeller." \parencite[1]{henderson1964}
+
+It should also noted that in order to substantially increase the opening area of the propeller, the difference between the velocity of movement at the blades nearer to the root (close to the hub) and the extremities of the blades increases substantially.
+In order for equivalent work to be done on all of the water distributed radially across area of the impeller, quite a different blade shape might be required, in order to provide a higher deflection of the water close to the root.
+In these areas, there is increase probability of flow separation occurring, where the velocity of movement of the water across the surface is insufficient to overcome the pressure gradient required to move over the blades.
+(In a pump, the pressure necessarily increases as work is done on it.)  Or, since there isn't enough pressure exerted on the whole of the water column to overcome the resistance imposed by the duct, not all of the water moves forward, and some starts to move backwards.
+This leads to flow separation, or stall occurring close to the blade roots, which dramatically diminishes the efficiency of the system.
+A further discussion of this effect can be found in \cite[15,27]{henderson1964},  \cite[807,8012]{bruce1974},  \cite[185]{wislicenus1973}, \cite[60]{mcbride1979} and many other related texts.
+This is a general problem that affects a range of axial-flow turbomachinery, particulalry operating in off-design conditions, and will present a general concern for the design of a new pumpjet \cite[185]{wislicenus1986}, \cite{li2013}.
+
+\begin{figure}
+\includegraphics[width=\textwidth]{stall.png}
+\caption{Flow separation, or stall, can occur when the flow over the blades of a pump slows too much, as shown in \cite{li2013}}
+\label{fig:stall.png}
+\end{figure}
+
+\begin{figure}
+\includegraphics[width=\textwidth]{backflow.png}
+\caption{Stall results in water flowing backwards near the hub of the blades at low flow rates, as shown in \cite{li2014rotating}}
+\label{fig:backflow.png}
+\end{figure}
+
+It is possible to limit the variation of deflection between the blade tips and roots by simply expanding the size of the hub.
+However, this necessarily reduces the effective frontal area that is available to receive the mass-flow, which necessitates further expansion of the outer diameter, and a very substantial increase in the weight of the system.
+In addition, expansion of the hub will necessarily increase the total surface area relative to the mass flow, including by increasing the amount of tapering required after the stators in order to prevent risk of cavitation occuring around the hub.
+
+Increasing the frontal area of the intake also is necessarily linked to the advance ratio of jet.
+Since low advance ratios correspond to higher shaft speed for a given forward velocity, increasing the thrust produced for a given rotation (by widening the jet area) necessarily results in higher advance ratios, since more work is done on the water for a given rotation.
+However, this also necessarily results in poorer cavitation performance of the jet, since higher blade velocities relative to the water flow result in necessarily increased propensity for cavitation.
+A detailed discussion of this can be found in \cite{gearhart1966selection}, but similar tensions can be readily found in most other pieces of literature, including \cite{henderson1964}, \cite{wislicenus1973}.
+
+Given the fundamental nature of the flow-rate and intake area to the design of pump-jets, and the great lenght of time for which they have been understood and discussed as a crucial design input for a waterjet, it is altogether implausible that some large increase in intake area can be achieved without incurring necessary and unavoidable tradeoffs.
+
+\subsubsection{Reduce degree of diffusion (i.e. switch to accelerating duct) to reduce negative thrust from duct}
+A further means of reducing the negative thrust produced by a duct may involve significantly reducing the degree of diffusion, or compression that occurs in advance of the rotor, since the process of diffusion necessarily slows the water and produces negative thrust.
+However, this amounts to transitioning towards an accelerating duct, or Kort Nozzle.
+Whilst nozzles of this type have been known to improve the cavitation characteristics of a propeller by enclosing the tips (which often tend to cavitate first), their ultimate function is to actually decrease the pressure at the blades of the propeller.
+Consequently, they necessarily increase the propensity of other types of cavitation to occur across the blade-face, including bubble, cloud and sheet cavitation.
+As such, these types of propellers are not considered for the military purposes which require an absolute minimum of noise to be created.
+Transitioning to an accelerating duct signify a shift away from a pump-jet, and necessarily be in direct tension with the need to minimise cavitation for acoustic considerations.
+
+In addition, whilst it is often demonstrated that for a given propeller under certain load conditions (high loading), the addition of an accelerating duct can considerably improve efficiency, all of these situations have generally been in the context of vessels which are designed for a relatively high thrust, low waterspeed situation, when propeller size is necessarily somewhat constrained.
+This is the case fishing vessels, or for tug boats, which need to impart significant thrust to a vessel dozens or hundreds of times their own size.
+Mounting a propeller the equivalent size of the vessels they mean to push, however, would be excessively heavy, arranging a mount and power drive that would keep the entire propeller submerged, and not have the effect of flipping the boat backwards when pushing hard, would be effectively impossible, without proportionally increasing the size of the tug.
+Furthermore, in most circumstances these propellers are still designed to operate with some cavitation, even if the extent of it is reduced by the presence of the duct \parencite{haimov2010}.
+
+In other words, the use of highly loaded propellers in such applications is actually particular to design constraints that are unlikely to be present for a submarine, which can position the propeller at the tail of its body, directly behind its centre of mass and drag, and consequenlty have freedom about the optimal size of the blades.
+It is well known that a duct cannot improve the performance of a low-loaded propeller which would be the situation for a slow-moving vessel with unconstrained propeller size.
+Authoritative references in the literature include \cite{oosterveld1970}, who calculates the minimum thrust coefficient $C_T$ (a measure of loading) for an accelerating duct to improve a propellers performance to be between 1 and 2, as shown in @AcceleratingIdeal.
+
+\begin{figure}
+\includegraphics[width=\textwidth]{DuctedEfficiency.png}
+\caption{Calculations of 'ideal' efficiencies of accelerating ducts (neglecting drag) in \cite{oosterveld1970} show that a duct can only improve the efficiency of a propeller if it is loaded above a certain degree, and that longer ducts produce better improvements in 'ideal' efficiencies}
+\label{fig:DuctedEfficiency.png}
+\end{figure}
+
+A parallel design change in the switch away from the degree of deceleration in the duct would be the reduction of the nozzle contraction after the rotor.
+For waterjets and pumpjets of all sorts, based on fundamental disc actuator theory, the degree of this contraction of the flow area has also been long identified as a means to increase pressure at the rotor, and hence decrease cavitation, but at the direct expense of efficiency \parencite[14]{van1996waterjet}.
+
+Numerous efforts to find suitably compromised duct shapes exist, but even advanced modern designs still exhibit declining efficiency at low speeds, as well as some cavitation during operations at design loads, which still tend to imply relatively high speeds, for example 27kt in \ref{fig:FineOptimised.png} .
+Consideration of accelerating duct designs for a military submarine would be extremely unlikely, and be a profound departure from any existing pumpjet designs.
+
+\begin{figure}
+\includegraphics[width=\textwidth]{FineOptimised.png}
+\caption{Extensive work has been done exploring compromise designs for decelerating ducted propellers (pumpjets), including with very modest deceleration in the intake, limited shroud taper, short duct lengths, or essentially incporate some of the advantages of accelerating ducts.
+Credit: \cite{abdel2010}}
+\label{fig:FineOptimised.png}
+\end{figure}
+
+\begin{figure}
+\includegraphics[width=\textwidth]{FineEfficiency.png}
+\caption{In experimental testing such designs still exhibit sharply falling effiency at high advance ratios, or low speed. Credit:  \cite{abdel2010}}
+\label{fig:FineEfficiency.png}
+\end{figure}
+
+\begin{figure}
+\includegraphics[width=\textwidth]{FineCavitation.png}
+\caption{TAt optimally efficient conditions (corresponding to 27kt) such designs still exhibit some cavitation. Credit: \cite{abdel2010}}
+\label{fig:FineCavitation.png}
+\end{figure}
+
+\subsubsection{Reduce shroud length to decrease drag on duct}
+The reduction of the length of the shroud in order to reduce drag presents another possible means to lower the duct resistance, as shown and discussed in \cite[21-25]{oosterveld1970}.
+The fundamental difficulty with greatly shortening a duct, however, is that it requires that any changes in the shape and with of the water column be accomplished much shorter distance for the duct to achieve its desired effect, which means that generally it has to achieve less of it.
+This means that extremely very low duct lenght achieve less of an efficiency improvement to the ideal efficiency of a propeller, in those circumstances where they can achieve an improvement  (accelerating ducts), as discussed above.
+
+If more aggressive shape changes are attemped in a limited space, the deflections of flow and pressure gradients become much more extreme.
+This leads to a far greater propensity for flow separation to occur.
+This can happen inside the duct ahead of the stator, particulalry when any degree of diffusion occurs in order to elevate static pressure at the blades, or any kind of direction change is required as in the case of waterjets.
+Unlike waterjets, pumpjets also have the potential for separation on the exterior surface, which is an additional concern in the design of duct \cite[13]{henderson1964}.
+Given the role of the duct is to change the speed of the flow at the impeller relative to what would occur in its absence (in particular to decelerate it) separation might tend to occur on different surfaces in different operating conditions, depending on the relative pressure gradients and velocity ratios they experience at different speeds and loads.
+This substantially complicates the challenge of designing a duct that performs well across a wide speed range.
+This is illustrated for the case of a waterjet in \cite[20]{bulten2006numerical}.
+Equivalent complexities occur in pumpjets operating in low speed ranges (2-6kt, where flow separation occurs ahead of the impeller, inside the shroud.
+In the case of an accelerating duct, this phenomenon is also observed at high advance ratios, or the equivalent of low speeds and low loads, as shown and discussed in some detail in \cite{willemsen2013}.
+
+\begin{figure}
+\includegraphics[width=\textwidth]{LowIVR.png}
+\caption{A an intake may experience risks of flow separation even at low speeds (Inlet Velocity Ratio) on some surfaces due to certain flow deflections being more severe.
+Source: \cite{bulten2006numerical}}
+\label{fig:LowIVR.png}
+\end{figure}
+
+\begin{figure}
+\includegraphics[width=\textwidth]{HighIVR.png}
+\caption{A an intake may experience risk of flow separation on different surfaces at high speeds (Inlet Velocity Ratio) where other flow deflections are more severe.
+Source: \cite{bulten2006numerical}}
+\label{fig:HighIVR.png}
+\end{figure}
+
+\begin{figure}
+\includegraphics[width=\textwidth]{DuctSeparation.png}
+\caption{Flow separation occurs at on the outside of an accelerating duct at high $J$ values, which corresponds to low loads, or speeds.
+For high $J$ values, the pressure reduction on the inside of the duct, creating positive duct thrust, can be seen.
+Source: \cite[73]{willemsen2013}}
+\label{fig:DuctSeparation.png}
+\end{figure}
+
+\begin{figure}
+\includegraphics[width=\textwidth]{NozzleDrag.png}
+\caption{Dramatically reducing the length of a duct relative to its diameter ($L/D$) as shown in \cite{oosterveld1970} can reduce its drag, which has the most significant impact at low thrusts or speeds.}
+\label{fig:NozzleDrag.png}
+\end{figure}
+
+In addition, the substantial reduction in the length of the shroud, both ahead of and behind the rotor, will also have the substantial effect of diminishing any acoustic sheilding which the duct may achieve.
+
+\subsubsection{Summary remarks on potential for redesign of pumpjet for low-speed conditions}
+Overall, while the parameter-space for the alteration of marine-propulsors is large, the parameters which would allow for very dramatic rather than minor changes to propulsor efficiency at low speed are relatively few.
+The major ones include increasing the width of the intake, decreasing the length of the duct, and decreasing the pressure inrease at the blades achieved by the duct.
+In short, it involves evolving the pumpjet back towards a propeller.
+None of these changes could be effected to a substantial degree without some inevitable trade-off in terms of the cavitation performance of the system at higher speeds.
+
+\section{Results of numerical simulation}
+
+\section{Acoustic Considerations}
+It is generally considered that the existence of cavitation of any sort will tend to dominate the acoustic signature of any vessel that doesn't produce any machinery noise by relying on all-electric drive as a conventional submarine would when dived (\cite{giles2010}, \cite{harvie1965construction}).
+Its quite possible that during the period when a submarine is snorting, or running its diesel engines, these sources of noise would be a more substantial and distinctive acoustic signature than small amounts of cavitation on the propulsion system, if it were to occur, owing in particular to the low freqency of diesel engines.
+
+Other sources of noise hydrodynamic noise, such as the turbulence produced in the boundary layer of the flow over the hull, or parts of the propulsor, will still contribute to overall noise and some signature, but it will tend to be vastly smaller in magnitude.
+This owes to the essentially incompressible nature of water.
+Without compression, even very relatively intense votices or tubulent flows will not propagate very much noise, if the water remains perfectly solid.
+All of the 'round and round' can't achieve much 'in and out', which is what results in propagated noise.
+This obviously doesn't hold true in compressible fluids, such as air, as one would observe noticing the intense noise produced by hand-dryers, aeroplanes, jet engines etc.
+It also doesn't hold true within a turbulent flow underwater.
+If you were to put a sensor inside a turbulent flow, the rapid movements would result in continual and significant pressure fluctuations, hence noise.
+But at a distance, these fluctuations tend to substantially cancel out.
+The larger the scale of the eddies and vortices, the further noise might propagate, and also the lower the frequencies of resultant noise might be, which makes noise propagate further.
+Likewise, larger caviation voids will tend to emit lower frequency sound aswell, as can be observed from the boiling of a kettle.
+
+However, when dived, it might be reasonably assumed that the complete elimination of all sources of cavitition will be the primary objective of acoustic signature management.
+This is neatly put in \cite[88]{gearhart1966}: "The occurence of cavitation in a propulsor results in adegradation of the propulsor operating characteristics if extensive amounts of cavitation exist, and in significant noise when minute amounts occur.
+As a result, the performance requirements of a submersible weapon system specify that the occurence or inception of cavitation be avoided below certain operating depths." One might consider that a submarines 'tactical silent speed' would most probably correspond to the speed at which it can operate safely below the point of any cavitation inception.
+
+It is crucial to understand that increasing depth necessarily increases the static pressure throughout a propulsion system.
+As such, the speed at which cavitation starts to occur, will necessarily be a function of depth.
+Approximately every 10m of water adds approximately the equivalent of an atmosphere of pressure.
+As such, the static pressure a system experiences at 10m depth will be half that experienced at 30m (since the air in the atmosphere comprises the first atmosphere of pressure).
+Consequently, diving deeper will necessarily increase the speeds one can achieve without any cavitation occurring \ref{fig:InceptionDepth.png}.
+
+\begin{figure}
+\includegraphics[width=\textwidth]{InceptionDepth.png}
+\caption{Increasing in depth improves the speed achievable before cavitation begins to occur, as shown in the case of a torpedo \parencite{suryanarayana2010cavitation}}
+\label{fig:InceptionDepth.png}
+\end{figure}
+
+It is also important to understand that the addition of a duct will necessarily to flow separation occuring on the inside of the duct opening at very low speeds which correspond to high advance ratios when load remains constant.
+Flow separation tends to lead to vortex shedding, which means that inconsistent flows (vortices) are then ingested into the blades of the impeller.
+This effect is well known and document.
+A good example is shown on the exterior of an \textbf{accelerating} duct at $J = .8$ in \cite{willemsen2013}.
+
+\begin{figure}
+\includegraphics[width=\textwidth]{VortexShedding.png}
+\caption{Vortex shedding following a flow separation leads to eddies being released unsteadily into the main flow stream, as shown here on the exterior of an accelerating duct.
+In the case of decelerating duct (pumpjet), duct shape is inverted, leading to the vortices being shed into the impeller blades.
+\parencite{willemsen2013}}
+\label{fig:VortexShedding.png}
+\end{figure}
+
+Whilst at very low power levels, or at good depths, it is quite plausible that this doesn't result in any cavitation, which would normally be a likely occurrence if the blades are operating closer to their design power.
+As such,
+
+\section{Conclusions}
+
+% \printbibliography
 

### 2018-01-02 DraftExec.Rmd
test this

@@
     year: 2011
+
+- id: bruce1974
+  title: The Design of Pumpjets for Hydrodynamic Propulsion
+  author:
+  - family: Bruce
+    given: Edgar P.
+  author:
+  - family: Gearhart
+    given: Walter S.
+  author:
+  - family: Ross
+    given: John R.
+  - family: Treaster
+    given: Allen L.
+  URL: 'https://ntrs.nasa.gov/archive/nasa/casi.ntrs.nasa.gov/19750003135.pdf'
+  publisher: Ordnance Research Laboratory, The Pennsylvania State University
+  issued:
+    year: 1974
+    month: 01
+
+  - id: furuya1988
+  title: A new pumpjet design theory
+  author:
+  - family: Furuya
+    given: Oritsugu
+  author:
+  - family: Chiang
+    given: Wen-Li
+  URL: 'http://www.dtic.mil/dtic/tr/fulltext/u2/a201353.pdf'
+  publisher: Office of Naval Research
+  issued:
+    year: 1988
+    month: 6
 
@@
 
+- id: lu2015
+  title: CFD prediction and simulation of a pumpjet propulsor
+  author:
+  - family: Lu
+    given: Lin
+  - family: Pan
+    given: Guang
+  - family: Sahoo
+    given: Prasanta K.
+  URL: 'http://www.sciencedirect.com/science/article/pii/S2092678216000042'
+  publisher: International Journal of Naval Architecture and Ocean Engineering
+  issued:
+    year: 2015
+    month: 10
+
 - id: giles2010
@@
 
+  - id: willemsen2013
+  title: Improving Potential Flow Predictions for Ducted Propellers
+  author:
+  - family: J.C
+    given: Willemsen
+  URL: 'http://www.refresco.org/wp-content/uploads/2015/03/2013-Msc_Thesis_ChrisWillemsen.pdff'
+  publisher: Marin
+  issued:
+    year: 2013
+    month: 12
+
 ---
@@
 
-![Other modern reference texts might demonstrate pump efficiencies with different x-axes, effectively reversing the way the graph should be understood for a single system operating at different speeds.
+![Other modern reference texts might demonstrate jet or propeller efficiencies with different x-axes, effectively reversing the way the graph should be understood for a single system operating at different speeds.
 Credit: [@mollard2011]](/Users/aidanmorrison/submarines/EfficiencyMollard.png){width=500px}
+
+![Another graph representing efficiency as a function of the ratio of jet velocity to incoming velocity, which shows efficiency declining at low thrust (to the right).
+The impact of increasing duct losses can be clearly seen, as opposed to the 'ideal' propeller without a duct.
+Credit: [bulten2006numerical]](/Users/aidanmorrison/submarines/EfficiencyBulten.png){width=500px}
 
@@
 These losses of efficiency, however, could never supercede those of a decelerating ducted propeller or jet, which has the shroud which necessarily imposes an **additional** drag to the watercolumn in excess of any other resistance faced by the rotor.
+This effect can be seen clearly even or an accelerating duct, as shown in [@willemsen2013].
+
+![Even the addition of an accelerating ducts substantially reduces efficiency of propellers at very low loadings, or for constant loads, low speeds.
+Maximum attainable efficiency is also reduced.
+Their benefit lies in the bulging of the curve at low advance ratios (higher loads).
+Credit: [@willemsen2013, pp 6]](/Users/aidanmorrison/submarines/PropDuct.png){width=500px}
 
@@
 In terms of establishing plausible peaks and cross-over points, the study undertaken in [@giles2010] probably provides a sufficient indication as to the current state of the art in the relevant technologies.
-The study was undertaken with the backing of military customers with support and testing faciliteis (the US Navy), and involved two major companies who are world leaders in maritime engineering (BMT) and waterjet manufacture (Rolls Royce).
-Even if furthe technical advances were possible
-
-## The sensitivity to duct size
-
-## The
+Whilst this effort focussed on a surface vessel rather than a   The study was undertaken with the backing of military customers with support and testing facilities (the US Navy), and involved two major companies who are world leaders in maritime engineering (BMT) and waterjet manufacture (Rolls Royce).
+Even if further technical advances through design refinement were possible, as the authors alude to, the cross-over point in terms of efficiency might be brought down to 25kt.
+The prospect that the crossover point lies anywhere near the transit speed of a convetional submarine (8-10kt) appears utterly implausible.
+In discussions with leading world experts involved in the development of advanced ducted propellers (including those involved in citations) I was told that a pumpjet could not be more efficient than a propeller below 18kt.
+
+With the optimum efficiency of a jet probably being no greater than 60% or 70% at its peak efficiency according to many theoretical and experimental studies [@lu2015], which would quite probably lie considerably higher in terms of waterspeed, these facts, along with the inevitable decline of efficiency towards zero as waterspeed also declines, a suitably constrained set of plausible efficiency curves may be investigated numerically, as will be done later in later sections.
+
+## An assessment of scope for improving the efficiency of a pumpjet at low speeds
+
+Before continuing to assess the impact of the likely efficiency curves of pumpjets relative to propellers, it is worthwhile considering what the known literature about the theory of pumps, jets, and turbomachinery might tell us about the potential for a radical advance which make might substantially improve the shape of the curve at very low speeds.
+Whilst the field of propulsor design is no doubt complex, and I will not attempt to be exhaustive in detailing all of the potential design considerations which might make an improvement.
+However, the pursuit of reasonably comprehensive confidence is aided by the the scale of the gap which probably needs to be closed to make pumpjets comparably efficient to propellers at the relevant speeds.
+Minor adjustments and the improvement of small or marginal effects, of which there might be many, will clearly not help in shifting the efficiency peak of a jet from over 25kt to under 10kt.
+In other words, this gap will not be closed or noticeably reduced by 'tuning'.
+Quite substantial redesign of core characteristics (like mass flow) would be required, which would necessitate a cycle of redesign and tuning virtually all other aspect of the entire jet, as illustrated by numerous guides to the design of such systems [@bruce1974, @furuya1988, @mccormick1963design, @henderson1964] .
+As such, I will only consider those parameters which could plausibly be altered quite substantially, where the consideration of larger changes make the likely limits or trade-offs quite clear.
+
+### Increase mass-flow by widening area of intake
+Perhaps the most obvious means by which one might attain substantial increases in efficiency is to increase substantially the mass flow rate.
+This is derived from a fundamental relationship, as expressed earlier in the representation of ideal propeller and jet efficiency (prior to the consideration of any drag or other losses) which was given in the earlier equation.
+Since excessive velocity differences result in more energy expended than momentum changed, as given by momentum theory and basic mechanics, doing work on a large body of water, rather than working harder on more of it, is generally desirable from an efficiency perspective.
+Considering what the appropriate mass-flow should be at design speeds is generally the first thing that is specified to occur in all design processes for pumpjets, and good example of which is in [@henderson1964 pp6-7].
+
+As Henderson et al.
+note, the key constraint of this variable is that increasing the diameter of the pumpjet necessarily leads to an increase in the surface (including the external suface) of the shroud.
+The increased drag that this induces must be deducted from the thrust that the pumpjet produces, or effectively increasing $\Delta T$ as given in earlier equations.
+In this respect, there is "no free lunch" to be had in expanding the size of the intake.
+As Henderson et al.
+conclude, for this reason the diameter of a pumpjet tends to be about 15% or 20% smaller than that of a corresponding propeller, whereas the diameter to have equivalent mass-flow, highlighting the tension set up in pumpjet design with their earlier observation that "if the cavitation resistance of the blades is to be improved with no reduction in thrust and the same rate of flow through the propulsor, the pump jet rotor must be larger than that of the propeller." [@henderson1964 pp 1]
+
+It should also noted that in order to substantially increase the opening area of the propeller, the difference between the velocity of movement at the blades nearer to the root (close to the hub) and the extremities of the blades increases substantially.
+In order for equivalent work to be done on all of the water distributed radially across area of the impeller, quite a different blade shape might be required, in order to provide a higher deflection of the water close to the root.
+In these areas, there is increase probability of flow separation occurring, where the velocity of movement of the water across the surface is insufficient to overcome the pressure gradient required to move over the blades.
+(In a pump, the pressure necessarily increases as work is done on it.)  Or, since there isn't enough pressure exerted on the whole of the water column to overcome the resistance imposed by the duct, not all of the water moves forward, and some starts to move backwards.
+This leads to flow separation, or stall occurring close to the blade roots, which dramatically diminishes the efficiency of the system.
+A further discussion of this effect can be found in [@henderson1964 pp 15, 27],  [@bruce1974 pp807, 8012, @wislenicus1973 pp 185, mcbride1979 pp 60]] and many other related texts.
+This is a general problem that affects a range of axial-flow turbomachinery, particulalry operating in off-design conditions, and will present a general concern for the design of a new pumpjet [@wislicenus1986 pp 185, li2013].
+
+![Flow separation, or stall, can occur when the flow over the blades of a pump slows too much, as shown in [@li2013]](/Users/aidanmorrison/submarines/stall.png){width=500px}
+
+![Stall results in water flowing backwards near the hub of the blades at low flow rates, as shown in [@li2014rotating]](/Users/aidanmorrison/submarines/backflow.png){width=500px}
+
+It is possible to limit the variation of deflection between the blade tips and roots by simply expanding the size of the hub.
+However, this necessarily reduces the effective frontal area that is available to receive the mass-flow, which necessitates further expansion of the outer diameter, and a very substantial increase in the weight of the system.
+In addition, expansion of the hub will necessarily increase the total surface area relative to the mass flow, including by increasing the amount of tapering required after the stators in order to prevent risk of cavitation occuring around the hub.
+
+Increasing the frontal area of the intake also is necessarily linked to the advance ratio of jet.
+Since low advance ratios correspond to higher shaft speed for a given forward velocity, increasing the thrust produced for a given rotation (by widening the jet area) necessarily results in higher advance ratios, since more work is done on the water for a given rotation.
+However, this also necessarily results in poorer cavitation performance of the jet, since higher blade velocities relative to the water flow result in necessarily increased propensity for cavitation.
+A detailed discussion of this can be found in [@gearhart1966selection], but similar tensions can be readily found in most other pieces of literature, including [@henderson1964, @wislenicus1973].
+
+Given the fundamental nature of the flow-rate and intake area to the design of pump-jets, and the great lenght of time for which they have been understood and discussed as a crucial design input for a waterjet, it is altogether implausible that some large increase in intake area can be achieved without incurring necessary and unavoidable tradeoffs.
+
+### Reduce degree of diffusion (i.e. switch to accelerating duct) to reduce negative thrust from duct
+A further means of reducing the negative thrust produced by a duct may involve significantly reducing the degree of diffusion, or compression that occurs in advance of the rotor, since the process of diffusion necessarily slows the water and produces negative thrust.
+However, this amounts to transitioning towards an accelerating duct, or Kort Nozzle.
+Whilst nozzles of this type have been known to improve the cavitation characteristics of a propeller by enclosing the tips (which often tend to cavitate first), their ultimate function is to actually decrease the pressure at the blades of the propeller.
+Consequently, they necessarily increase the propensity of other types of cavitation to occur across the blade-face, including bubble, cloud and sheet cavitation.
+As such, these types of propellers are not considered for the military purposes which require an absolute minimum of noise to be created.
+Transitioning to an accelerating duct signify a shift away from a pump-jet, and necessarily be in direct tension with the need to minimise cavitation for acoustic considerations.
+
+In addition, whilst it is often demonstrated that for a given propeller under certain load conditions (high loading), the addition of an accelerating duct can considerably improve efficiency, all of these situations have generally been in the context of vessels which are designed for a relatively high thrust, low waterspeed situation, when propeller size is necessarily somewhat constrained.
+This is the case fishing vessels, or for tug boats, which need to impart significant thrust to a vessel dozens or hundreds of times their own size.
+Mounting a propeller the equivalent size of the vessels they mean to push, however, would be excessively heavy, arranging a mount and power drive that would keep the entire propeller submerged, and not have the effect of flipping the boat backwards when pushing hard, would be effectively impossible, without proportionally increasing the size of the tug.
+Furthermore, in most circumstances these propellers are still designed to operate with some cavitation, even if the extent of it is reduced by the presence of the duct [@haimov2010].
+
+In other words, the use of highly loaded propellers in such applications is actually particular to design constraints that are unlikely to be present for a submarine, which can position the propeller at the tail of its body, directly behind its centre of mass and drag, and consequenlty have freedom about the optimal size of the blades.
+It is well known that a duct cannot improve the performance of a low-loaded propeller which would be the situation for a slow-moving vessel with unconstrained propeller size.
+Authoritative references in the literature include [@oosterveld1970], who calculates the minimum thrust coefficient $C_T$ (a measure of loading) for an accelerating duct to improve a propellers performance to be between 1 and 2, as shown in @AcceleratingIdeal.
+
+![Calculations of 'ideal' efficiencies of accelerating ducts (neglecting drag) in [@oosterveld1970] show that a duct can only improve the efficiency of a propeller if it is loaded above a certain degree, and that longer ducts produce better improvements in 'ideal' efficiencies](/Users/aidanmorrison/submarines/AcceleratingIdeal.png){width=500px}
+
+A parallel design change in the switch away from the degree of deceleration in the duct would be the reduction of the nozzle contraction after the rotor.
+For waterjets and pumpjets of all sorts, based on fundamental disc actuator theory, the degree of this contraction of the flow area has also been long identified as a means to increase pressure at the rotor, and hence decrease cavitation, but at the direct expense of efficiency [@van1996waterjet pp 14].
+
+Numerous efforts to find suitably compromised duct shapes exist, but even advanced modern designs still exhibit declining efficiency at low speeds, as well as some cavitation during operations at design loads, which still tend to imply relatively high speeds, for example 27kt in [@abdel2010].
+Consideration of accelerating duct designs for a military submarine would be extremely unlikely, and be a profound departure from any existing pumpjet designs.
+
+![Extensive work has been done exploring compromise designs for decelerating ducted propellers (pumpjets), including with very modest deceleration in the intake, limited shroud taper, short duct lengths, or essentially incporate some of the advantages of accelerating ducts.
+Credit: [@abdel2010]](/Users/aidanmorrison/submarines/FineOptimised.png){width=500px}
+
+![In experimental testing such designs still exhibit sharply falling effiency at high advance ratios, or low speed.
+Credit: [@abdel2010]](/Users/aidanmorrison/submarines/FineEfficiency.png){width=500px}
+
+![At optimally efficient conditions (corresponding to 27kt) such designs still exhibit some cavitation. Credit: [@abdel2010]](/Users/aidanmorrison/submarines/FineCavitation.png){width=500px}
+
+### Reduce shroud length to decrease drag on duct
+The reduction of the length of the shroud in order to reduce drag presents another possible means to lower the duct resistance, as shown and discussed in [@oosterveld1970 pp 21-25].
+The fundamental difficulty with greatly shortening a duct, however, is that it requires that any changes in the shape and with of the water column be accomplished much shorter distance for the duct to achieve its desired effect, which means that generally it has to achieve less of it.
+This means that extremely very low duct lenght achieve less of an efficiency improvement to the ideal efficiency of a propeller, in those circumstances where they can achieve an improvement  (accelerating ducts), as discussed above.
+
+If more aggressive shape changes are attemped in a limited space, the deflections of flow and pressure gradients become much more extreme.
+This leads to a far greater propensity for flow separation to occur.
+This can happen inside the duct ahead of the stator, particulalry when any degree of diffusion occurs in order to elevate static pressure at the blades, or any kind of direction change is required as in the case of waterjets.
+Unlike waterjets, pumpjets also have the potential for separation on the exterior surface, which is an additional concern in the design of duct [@henderson1964 pp 13].
+Given the role of the duct is to change the speed of the flow at the impeller relative to what would occur in its absence (in particular to decelerate it) separation might tend to occur on different surfaces in different operating conditions, depending on the relative pressure gradients and velocity ratios they experience at different speeds and loads.
+This substantially complicates the challenge of designing a duct that performs well across a wide speed range.
+This is illustrated for the case of a waterjet in [@bulten2006numerical pp 20].
+Equivalent complexities occur in pumpjets operating in low speed ranges (2-6kt, where flow separation occurs ahead of the impeller, inside the shroud.
+In the case of an accelerating duct, this phenomenon is also observed at high advance ratios, or the equivalent of low speeds and low loads, as shown and discussed in some detail in [@willemsen2013].
+
+![A an intake may experience risks of flow separation even at low speeds (Inlet Velocity Ratio) on some surfaces due to certain flow deflections being more severe.
+Source: [@bulten2006numerical]](/Users/aidanmorrison/submarines/LowIVR.png){width=500px}
+
+![A an intake may experience risk of flow separation on different surfaces at high speeds (Inlet Velocity Ratio) where other flow deflections are more severe.
+Source: [@bulten2006numerical]](/Users/aidanmorrison/submarines/HighIVR.png){width=500px}
+
+![Flow separation occurs at on the outside of an accelerating duct at high $J$ values, which corresponds to low loads, or speeds.
+For high $J$ values, the pressure reduction on the inside of the duct, creating positive duct thrust, can be seen.
+Source: [@willemsen2013, pp 73]](/Users/aidanmorrison/submarines/DuctSeparation.png){width=500px}
+
+![Dramatically reducing the length of a duct relative to its diameter ($L/D$) as shown in [@oosterveld1970] can reduce its drag, which has the most significant impact at low thrusts or speeds.](/Users/aidanmorrison/submarines/NozzleDrag.png){width=500px}
+
+In addition, the substantial reduction in the length of the shroud, both ahead of and behind the rotor, will also have the substantial effect of diminishing any acoustic sheilding which the duct may achieve.
+
+### Summary remarks on potential for redesign of pumpjet for low-speed conditions
+Overall, while the parameter-space for the alteration of marine-propulsors is large, the parameters which would allow for very dramatic rather than minor changes to propulsor efficiency at low speed are relatively few.
+The major ones include increasing the width of the intake, decreasing the length of the duct, and decreasing the pressure inrease at the blades achieved by the duct.
+In short, it involves evolving the pumpjet back towards a propeller.
+None of these changes could be effected to a substantial degree without some inevitable trade-off in terms of the cavitation performance of the system at higher speeds.
 
@@
 # Acoustic Considerations
+It is generally considered that the existence of cavitation of any sort will tend to dominate the acoustic signature of any vessel that doesn't produce any machinery noise by relying on all-electric drive as a conventional submarine would when dived [@giles2010, @harvie1965construction].
+Its quite possible that during the period when a submarine is snorting, or running its diesel engines, these sources of noise would be a more substantial and distinctive acoustic signature than small amounts of cavitation on the propulsion system, if it were to occur, owing in particular to the low freqency of diesel engines.
+
+Other sources of noise hydrodynamic noise, such as the turbulence produced in the boundary layer of the flow over the hull, or parts of the propulsor, will still contribute to overall noise and some signature, but it will tend to be vastly smaller in magnitude.
+This owes to the essentially incompressible nature of water.
+Without compression, even very relatively intense votices or tubulent flows will not propagate very much noise, if the water remains perfectly solid.
+All of the 'round and round' can't achieve much 'in and out', which is what results in propagated noise.
+This obviously doesn't hold true in compressible fluids, such as air, as one would observe noticing the intense noise produced by hand-dryers, aeroplanes, jet engines etc.
+It also doesn't hold true within a turbulent flow underwater.
+If you were to put a sensor inside a turbulent flow, the rapid movements would result in continual and significant pressure fluctuations, hence noise.
+But at a distance, these fluctuations tend to substantially cancel out.
+The larger the scale of the eddies and vortices, the further noise might propagate, and also the lower the frequencies of resultant noise might be, which makes noise propagate further.
+Likewise, larger caviation voids will tend to emit lower frequency sound aswell, as can be observed from the boiling of a kettle.
+
+However, when dived, it might be reasonably assumed that the complete elimination of all sources of cavitition will be the primary objective of acoustic signature management.
+This is neatly put in [@gearhart1966, pp 88]: "The occurence of cavitation in a propulsor results in adegradation of the propulsor operating characteristics if extensive amounts of cavitation exist, and in significant noise when minute amounts occur.
+As a result, the performance requirements of a submersible weapon system specify that the occurence or inception of cavitation be avoided below certain operating depths." One might consider that a submarines 'tactical silent speed' would most probably correspond to the speed at which it can operate safely below the point of any cavitation inception.
+
+It is crucial to understand that increasing depth necessarily increases the static pressure throughout a propulsion system.
+As such, the speed at which cavitation starts to occur, will necessarily be a function of depth.
+Approximately every 10m of water adds approximately the equivalent of an atmosphere of pressure.
+As such, the static pressure a system experiences at 10m depth will be half that experienced at 30m (since the air in the atmosphere comprises the first atmosphere of pressure).
+Consequently, diving deeper will necessarily increase the speeds one can achieve without any cavitation occurring [@InceptionDepth].
+
+![Increasing in depth improves the speed achievable before cavitation begins to occur, as shown in the case of a torpedo [@suryanarayana2010cavitation]](/Users/aidanmorrison/submarines/InceptionDepth.png){width=500px}
+
+It is also important to understand that the addition of a duct will necessarily to flow separation occuring on the inside of the duct opening at very low speeds which correspond to high advance ratios when load remains constant.
+Flow separation tends to lead to vortex shedding, which means that inconsistent flows (vortices) are then ingested into the blades of the impeller.
+This effect is well known and document.
+A good example is shown on the exterior of an **accelerating** duct at $J = .8$ in [@willemsen2013].
+
+![Vortex shedding following a flow separation leads to eddies being released unsteadily into the main flow stream, as shown here on the exterior of an accelerating duct.
+In the case of decelerating duct (pumpjet), duct shape is inverted, leading to the vortices being shed into the impeller blades.
+[@willemsen2013]](/Users/aidanmorrison/submarines/VortexShedding.png){width=500px}
+
+Whilst at very low power levels, or at good depths, it is quite plausible that this doesn't result in any cavitation, which would normally be a likely occurrence if the blades are operating closer to their design power.
+As such,
 

### 2018-01-02 DraftExec.tex

@@
 Controllable pitch propellers also have advangates in terms of manouverability, since the pitch can be reversed and hence reverse thrust can be produced, without requiring the direction of the power coming from the engines to be reversed.
-This has advantages for ferries which undertake frequent docking manouvres in confined spaces \parencite{MAN2017} \parencite{lewis1988}.
-
-In Figure \ref{fig:ex2} we can see a sample plot of the mtcars data.
+This has advantages for ferries which undertake frequent docking manouvres in confined spaces (\cite{MAN2017}, \cite{lewis1988}).
 
 \begin{figure}
-\begin{knitrout}
-\definecolor{shadecolor}{rgb}{0.969, 0.969, 0.969}\color{fgcolor}
-
-{\centering \includegraphics[width=\maxwidth]{figures/plots-unnamed-chunk-1-1}
-
-}
-
-\end{knitrout}
-\caption{An example plot}
-\label{fig:ex2}
+\includegraphics[width=\textwidth]{CPPropeller.png}
+\caption{Controllable Pitch Propellers (CP Propellers) allow the pitch of the blades to be varied. Image credit: \parencite{mollard2011}}
+\label{fig:CPPropeller.png}
 \end{figure}
 
-\printbibliography
+\subsubsection{Pitch and the Advance Ratio}
+Pitch is also an extremely important concept to understand because it relates closely to the way that a propeller's 'open-water efficiency' $\eta_O$ is expressed in charts, as well as the advance ratio $J$ that usually comprises the x-axis in such charts.
+
+\begin{figure}
+\includegraphics[width=\textwidth]{OpenWater.png}
+\caption{A typical chart showing open water efficiency of a propeller, in this case a Wageningen B5-75. Image: MARIN}
+\label{fig:OpenWater.png}
+\end{figure}
+
+This chart also shows two other closely related dimensionless coefficients, the torque and thrust coefficients ($K_Q$ and $K_T$) which we shall not dwell on here.
+However, it is important to note that a number of different lines are expressed on this sort of chart, which represent different versions of the same propeller, with only the pitch (expressed in its dimensionless form as $P/D$) being different in each case.
+The x-axis, $J$, also warrants a clear definition:
+
+\begin{equation}
+\label{eq:2}
+J = \frac{V_a}{nD}
+\end{equation}
+
+Where $V_a$ is the speed of advance of the propeller, or the speed at which water arrives at the front-side of the propeller, and $n$ represents the rate of the propeller's rotation (revolutions per second) and $D$ is the propeller's diameter.
+For fullness of understanding it is also worthwhile to note here that $V_a$ does not correspond directly to the vessel's speed of travel, though they are related.
+To some extent, a vessel moving through water will always draw water in its wake along with it, which means that in the wake (where the propeller tends to operate to some extent or other) the water incident on the propeller will arrive at a lower speed than that implied by the ship speed \parencite[15]{MAN2017}.
+Another key distinction for the consideration of jets later on, is that the speed of advance does not necessarily incporate the movement of water due to the movement induced by the propeller (or duct) itself, which is a distinct effect to that of the wake field.
+In \cite[213]{lewis1988} $V_A$ is used specifically to denote the undisturbed stream velocity, since the role of certain ducts can be to accelerate or decelerate the stream prior to the water arriving at the blades of the propeller.
+
+One may also note from the previous figure that for each line representing a given pitch, the efficiency falls precipitously towards zero around the point where $J$ reaches $P/D$.
+This is because this extreme of the curve represents a circumstance where the propeller is rotating, but not actually moving any water, since it is moving through the water at about that speed which corresponds to the exact helical path defined by the pitch being traced.
+This represents circumstances where a propeller encounters effectively zero load.
+This might be the case when a eases it's power, and the momentum of the ship temporarily keeps the ship moving without the propeller exerting much or any net thrust, or where alternative propoulsion sources (sails, or other propellers) are keeping a speed higher than what the propeller actually has push backwards on water to maintain.
+
+A closely related concept here is the slip ratio, which is a measure of the difference between the pitch (the distance the blade section would travel if it traced a perfect helix for one rotation) and the actual distance that the blade did advance through the water.
+This is neatly defined and shown in a diagram in \ref{fig:SlipRatio.png}.
+
+\begin{figure}
+\includegraphics[width=\textwidth]{SlipRatio.png}
+\caption{The slip ratio of a propeller. Image credit: \cite{MAN2017}}
+\label{fig:SlipRatio.png}
+\end{figure}
+
+It can be seen from the diagram of the open-water efficiencies that there is some desirable non-zero slip ratio for a propeller.
+The other end of the spectrum, where efficiencies also fall off toward zero, represent circumstances where the propeller is simply spinning in the same place in the water.
+In this case the slip ratio is one.
+This represents a circumstance known as 'bollard pull' where a vessel is simply exerting force against an immovable object or infinite load.
+Whilst the very low efficiencies here could represent a case of a propeller that is simply too small for the task it is trying to achieve, it is actually a significant and important circumstance to consider for propellers on working vessels such as tugs or trawlers, which exert large forces at very low water-speeds.
+Measurement of the thrust generated in bollard pull, and the power required to achieve this thrust, is still an important metric for such vessels, but the definition of 'useful work' for normal propulsion requieres the thust to be excerted over some distance travelled.
+
+\subsection{Ducted Propellers}
+A ducted propeller is a propeller in any kind of tubular shroud, casing or nozzle.
+Ducted propellers fall into broadly two categories, one of which is referred to frequently and interchangeably as a pump-jet.
+The two types are called the accelerating type (also known as a Kort Nozzle, after a company that made them popular in the 20th Century) and the decelerating type.
+The names describe the different effects that each type of nozzle have on the flow speed of the water at the point where it meets the propeller (which is also called an impeller, particularly when associated with decelerating ducts, or pumpjets).
+
+A detailed description of the essential differences between the two can be easily demonstrated by diagram, which show a simple comparison an acelerating and decelerating duct, which have the shape of the aerofoil section which comprises the duct inverted.
+(Aside: It should be noted that some designs attempt to strike a balance between efficiency and cavitation performance of accelerating and decelerating ducts for certain purposes, and in such cases the duct shape might not be so clearly contrasted \parencite{abdel2010}.
+But for the purposes of explanation, as well as optimisation for military purposes, the distinction between the two will generally be quite clear.)
+
+\begin{figure}
+\includegraphics[width=\textwidth]{AccelDecel.png}
+\caption{he two types of ducted propeller, accelerating (left) and decelerating (right) as shown in \parencite[249]{mollard2011}}
+\label{fig:AccelDecel.png}
+\end{figure}
+
+A consideration of what happens to the flow as it moves through each type of duct helps to demonstrate how the acceleration and deceleration effects are achieved.
+In the case of an accelerating nozzle, the area through which the flow must pass contracts prior to the water arriving at the impeller.
+In the decellerating case, the available area contracts after the water has passed the impeller.
+In some cases, a degree of diffusion is actually incorporated ahead of the impeller, whereby the area actually of the flow actually expands ahead of the impeller, which slows down the flow and increases its pressure according to Benoulli's principal \parencite{wislicenus1973}.
+
+\begin{figure}
+\includegraphics[width=\textwidth]{DuctStreamlines.png}
+\caption{The streamlines of flow through different duct types \parencite[214]{lewis1988}}
+\label{fig:DuctStreamlines.png}
+\end{figure}
+
+\subsubsection{The Accelerating Duct}
+Accelerating ducts are most commonly used by vessels which have to operate at high loadings at very low speed, such as tugs, barges, or trawlers.
+In those circumstances, relatively small vessels need to exert considerable forces against substantial loads, at very low water-speeds.
+Also, in these circumstances, the practical size of a propeller might be constrained, since operating an optimally sized propeller might increase the complexity of machiner required to deliver the power to so low a hub in the water, and it might also be difficult to arrange for the thrust to be directed substantially against the centre of drag of the vessel.
+
+The effect of the nozzle in these circumstances actually increase the optimal efficiency of the propeller, when operating at very low water-speeds, since the pressure reduction on the inside of the nozzle induced by the movement of the water through the nozzle, actually produces a net thrust forward, since the interior face at the opening of the nozzle is necessarily angled forward.
+This has the effect of distributing the suction force across a larger area, which now includes the interior of the nozzle.
+An equivalent way of describing the same effect is that the nozzle causes the water to be accelerated more gradually, over a longer distance, by inducing it to move more quickly as it passes through to the entrance.
+As a broad generalisation, more gradual or gentle actions tend to involve lower overall losses to turbulence, and hence the efficiency of the propeller can be increased.
+
+A further means by which the efficiency of a propeller can be increased is through the reduction of the turbulence created by the blade-tips passing through the open (static) water at high speed.
+Since the two sides of the blades have low and high pressure on them respectively, at the tips where the two faces meet, there there is a tendancy for water to 'spill' from the high pressure side the the low pressure side, creating a vortex which leads to unwanted turbulence.
+It is also around this vortex that blade tip cavitation tends to occur.
+By limiting the movement of water around the outside of the blade tip by having a tightly fitted shroud, losses due to this particulary type of tubulence, and the onset of this type of cavitation can be diminished.
+
+However, it is important to note that the existence of an accelerating duct by no means eliminates all cavitation.
+In fact, given that the static pressure at the impeller is necessarily reduced by an accleratint duct, the propensity for cavitation other surface of the propeller is generally increased, and cavitation can and does still occur in ducted propellers near the blade tips \parencite{moulijn2015}.
+
+\begin{figure}
+\includegraphics[width=\textwidth]{CavitatingDuct.png}
+\caption{An example of the cavitation occurring at the blade tips of a ducted propeller. Image credit \cite{moulijn2015}}
+\label{fig:CavitatingDuct.png}
+\end{figure}
+
+A detailed explanation of the means by which the ideal efficiency of propeller can be improved can be found in \cite[213-222]{lewis1988}, as well as other texts on marine propulsion, such as \cite{carlton2007}.
+A substantial work exploring the efficiencies of a range of accelerating and decelerating ducts can also be found in \cite{oosterveld1970}.
+A common feature of these extensive studies also show that for ducted propellers, including accelerating ducted propellers, the advantage in terms of efficiency is overall restricted to low waterspeeds.
+At higher waterspeeds, the drag induced by the water-movement over the exterior of the duct begins to increase substantially, and the net thrust on the duct becomes negative.
+When the exterior of the duct experience relatively little water velocity, but the interior experiences a great deal more, the greatest advantages of an accelerating duct are realised.
+Furthermore, in general these models do not incorporate the effect of flow separation, which is acknowledged as a possibility on both the exterior and interior of ducts, which can dramatically reduce efficiency performance (\cite[20]{oosterveld1970}, \cite[214]{lewis1988}).
+
+\begin{figure}
+\includegraphics[width=\textwidth]{DuctedEfficiency.png}
+\caption{Accelerating ducts produce a net positive thrust at low advance ratios, but a net negative thrust at higher advance ratios as the drag on the nozzle exterior increases.
+Image credit:Open-water test results of Ka 3-65 screw series with nozzle no.
+19A \parencite{oosterveld1970}}
+\label{fig:DuctedEfficiency.png}
+\end{figure}
+
+\subsubsection{The Decelerating Duct or Pumpjet}
+Decelerating ducts necessarily produce a negative thrust, since the process of raising the pressure of the water about the impeller (equivalent to slowing it down) necessitate reducing its kinetic energy and consequently momentum.
+In a wide range of literature introducing the fundamentals of ducted propellers and water-jets, or discussing their more modern development there is universal acknowledgement that the choice of a decelerating duct is generally made for the avoidance of cavitation for specialised (generally military) purposes, rather than for the achievement of increased efficiency (\cite{carlton2007}, \cite{oosterveld1970}, \cite{mollard2011}, \cite{haimov2010}, \cite{lewis1988}).
+
+In order to achieve a reduction in cavitation, however, care must be taken in the design of a decelerating duct.
+Since the negative thrust produced by the duct must necessitate additional work being done by the impeller, there can be circumstances where the additional loading of the impeller out-weighs the impacts of elevating the pressure around it.
+\cite[24-25]{oosterveld1970} provides a more thorough derivation of the circumstances in which this can be reliably achieved, which generally include larger blade area ratios or with more rows of rotors for higher loading, or circumstances where the loading of the propeller is quite low.
+This creates an inevitable and necessary tension between optimisation of a duct for propulsive efficiency and reduction of cavitation, which will be returned to in subsequent section.
+
+It is perhaps most relevant at this point to further clarify the distinction between the different names for related or similar systems.
+In \cite[17]{carlton2007} pump jet is described as an "interesting development of the classical ducted propeller".
+Oosterveld, in his thesis on ducted propellers also states with regard to the decelerating duct "This ducted propeller system is the so called pumpjet" \parencite[8]{oosterveld1970}.
+Other authors give separate treatement to ducted propellers and pump jets, though the systems they describe are essentially similar in terms of composition, advantages, and physics \parencite[288]{lewis1988}.
+
+\begin{figure}
+\includegraphics[width=\textwidth]{PumpjetOutline.png}
+\caption{In \cite{carlton2007} the pump jet is described as a development of a classical ducted propeller, with the noticeable addition of a stator}
+\label{fig:PumpjetOutline.png}
+\end{figure}
+
+In other definitive literature on the design of pumpjets \parencite{henderson1964} the difference between a pumpjet and a propeller is a described as being that "the stream of flow through the pumpjet is made to depart from the "natural" or free-stream sufrace that bounds the flow though a standard propeller."  According to the diagrams and the subsequent description it is clear that the definition embraces quite exactly the distinction between an accelerating duct, and a decelerating duct, namely that the flow is slowed at the point it encounters the impeller, where as in a free situation the free stream would contract at this point.
+
+The addition of some new part such as stator blades (to straighten out the flow) and other possible features which make the propeller system more like pump or turbomachinery components might be associated with systems more commonly referred to as pumpjets than ducted propellers.
+The key concepts behind the operation of these systems, however, remains essentially similar in their nature and design intent.
+
+\subsubsection{Waterjets}
+The use of the word waterjet, as opposed to pumpjet, signifies that the jet expelled from the shroud or pump is situated above the water, which is desirable in very fast-moving surface vessels which attempt to minimise their contact with the waters surface, including high-speed planing hull vessels, including pleasure craft.
+The physics of the two systems, however, are closely related, with the very noticeable exception of the absence of flow considerations for water on the exterior of the duct considerations for waterjets.
+However, the physics relating to the impeller and stator blades or (turbomachinery) designed to add energy to the flow are generally extremely closely linked, though a larger variety of diffeent pump types can be chosen for waterjets which might differ more significantly from the natural evolution of a propeller to an axial-flow pump.
+Many of the essential equations which govern the efficiencies and cavitation performance of pumpjets and waterjets are of the same or extremely similar form, as can be seen in detailed discussions given to both, by similar authors (\cite{wislicenus1973}, \cite{henderson1964}).
+There are, however, plenty of instances in recent literature where this distinction regarding the ejection of the jet above and below the waterline is not maintained, and waterjets and pumpjets seem to be used interchangeably \parencite{abdel2010}.
+
+\begin{figure}
+\includegraphics[width=\textwidth]{PumpjetWaterjet.png}
+\caption{A comparison of a pumpjet and waterjet as given in \cite{wislicenus1973}}
+\label{fig:PumpjetWaterjet.png}
+\end{figure}
+
+\section{Constraints on the efficiency of pumpjets at low speed}
+
+\subsection{Recent Commentary}
+There has been already some attention given to the probable efficiencies of pump jets at low speed.
+Recently Andrew Davies published a piece in the The Strategist (ASPI) arguing that the pump jet may actually save fuel at transit speeds (around 8kt) and overall improve the endurance of the submarine, despite likely having lower efficiency at low speeds such as patrol speed \parencite{davies2017}.
+This argument relies pump-jet efficiency still overtaking that of propellers at what is considered a relatively low speed for many applications.
+As Andrew acknowledges, further quantification of the likely efficiency is required, as the particular chart which he cites is not quantified.
+This section aims to identify as far as is reasonably possible from public literature how the efficiency of a likely pump-jet for a submarine might perform in quantitative terms, relative to a propeller.
+
+There has also been questions put in parliamentary hearings regarding the pump jet \parencite{patrick2017} and responses from public officials to such questions indicate that some of the literature provided to support such lines of questions was dated, and that the state of the art may have advanced considerably since this time.
+Whilst the argument that the state of the art is much advanced is obviously true, it is the case with marine propulsion, as with most sciences, that new developments tend to confirm, rather than annul, many of the workings that have been previously undertaken.
+To the extent that modern literature still supports the overall constraints and bounds provided by some of the foundational earlier works, a great deal of useful information can still be obtained from earlier works.
+In particular, the foundational theories, often derived from well-established principals of conservation of momentum, energy, as applied by the likes of Bernoulli and Euler, are as well respected and deeply depended upon today as they were centuries earlier when they were first arrived at.
+
+\subsection{Modern literature and results}
+In addition, more modern works, including those which involve experimental validation of theoretical results also tend to show that restrictions on the efficiency of waterjets are enduring, and give us a clear starting-point for consideration of their probable orders of magnitude, and likely transition points indicating at what speeds pumpjets might become advantageously more efficient.
+
+A 1995 Japanese experimental study on the performance of a waterjet demonstrated a close experimental fit to a theoretical relationship which sees the propulsive efficiency of the waterjet fall to zero with craft velocity, with significant declines in efficiency commencing just under 10m/s, or somewhere around 18kt \parencite{fujisawa1995}.
+
+\begin{figure}
+\includegraphics[width=\textwidth]{EfficiencyJapan.png}
+\caption{A graph from a Japanese study of waterjet efficiency shows efficiency rapidly declining below craft velocities of 10m/s, in accordance with theory. \parencite{fujisawa1995}}
+\label{fig:EfficiencyJapan.png}
+\end{figure}
+
+In 2015 an Australian team including staff from the Australian Maritime College and Incat published results of an experimental study comparing the use of waterjets and propellers for medium-speed ferries operating in the 20-30kt speed range \parencite{kamal2015}.
+This study explicitly set out to establish the cross-over point for where propellers become more efficient than jets, since Incat was interested in pioneering highly efficient medium-speed vessels, after having considerable success in very high speed catamarans.
+It is again acknowledged as given that at lower speeds propellers are more efficient than waterjets.
+The study found that propellers were considerably more efficient over the entire speed range tested, often by a factor of two in terms of transport efficiency.
+
+\begin{figure}
+\includegraphics[width=\textwidth]{EfficiencyTransport.png}
+\caption{An Australian experimental study aimed at determining the crossover point for the efficiency of jets and propellers in the 20-30kt range revealed much higher power was required across the entire speed range \parencite{kamal2015}}
+\label{fig:EfficiencyTransport.png}
+\end{figure}
+
+Both of these previous studies confirm the general and widespread assumption that jets are known to be more efficient at high speeds, and propellers at lower speeds.
+They also suggest that differences in efficiency could be very large, of a factor of two or more, including at speeds well over a submarine's transit speed.
+However, care must be taken making precise quantitative comparisons, since such experiements involved jets which expelled into air (waterjets) as opposed to remaining under water (pumpjets).
+Pumpjets would experience drag on the exterior of the shroud/duct, which waterjets would not, however waterjets are likely to have longer intake ducts, which might involve larger duct losses than pumpjets.
+Given that these factors might affect efficiency, and the overall design of the nozzle considerably, more direct comparison with fully submerged jets should be sought to make more direct comparisons.
+
+At IMARESTS 10th International Naval Engineering Conference, a paper was presented by BMT Defence and Rolls Royce \parencite{giles2010}, outlining a proposal for 'fully submerged waterjets' to be used on future surface ships in order to achieve better acoustic performance, particulalry for an anti-submarine warfare role.
+The jets comprise a mixed-flow pump coupled to an electric motor, and have been both extensively modelled and also tesed in demonstrator vessels.
+These may be a closer representation of the performance characteristics achievable by submerged pumpjets, since shape of the intake and shroud more closely resembles the shape of a pumpjet that might be found on a submarine.
+Importantly, this work focuses specifically on the performance of the jets as a function of ship-speed, including identifying possible crossover points.
+
+\begin{figure}
+\includegraphics[width=\textwidth]{BMTShape.png}
+\caption{The 'fully submerged waterjet' proposed by BMT and Rolls Royce for military surface craft may more closely represent the duct shape used on submarines than waterjets that release above the surface.
+\parencite{giles2010}}
+\label{fig:BMTShape.png}
+\end{figure}
+
+The paper makes no effort to justify the selection of choice of jets on the efficiency of the propulsion system, instead focussing on improved acoustic performance, and potential machinery space-saving advantages.
+The the vessel under consideration would require 250 tonnes more fuel with jets find than with propellers, the jets would require more power to reach equal waterspeed to a conventionally propelled craft right up to 30 kt.
+The paper states: "This is an encouraging result for a first iteration as the AWJ-21™ arrangement has not been fully optimised and, hence, there is further prospect of reducing the shaft power for this form.
+If the power could be reduced by 5\% (e.g.
+by fur ther improving the waterjet/hull fairing), then the AWJ-21™ vessel would be better at all speeds above 25 knots." \parencite{giles2010}
+
+\begin{figure}
+\includegraphics[width=\textwidth]{BMTPowering.png}
+\caption{Studies on submerged waterjets for military surface craft currently show that jets require more power than conventional propellers right up to 30kt \parencite{giles2010}}
+\label{fig:BMTPowering.png}
+\end{figure}
+
+In paper also provides specific curves for the efficiency of the propulsor as a function of vessel speed, which demonstrated that the waterjet efficiency was lower than that of conventional propellers throughout the speed range, but declined particularly precipitously below speeds of 10kt.
+At 5kt, the propeller was more than twice as efficient as the waterjet.
+
+\begin{figure}
+\includegraphics[width=\textwidth]{BMTEfficiency.png}
+\caption{Studies on submerged waterjets for military surface craft currently show that the efficiency of jets decline sharply below 10kt \parencite{giles2010}}
+\label{fig:BMTEfficiency.png}
+\end{figure}
+
+It appears from recent examples in the literature that there is no evidence of jets approaching the same efficiency as propellers at speeds as low as 10kt.
+Indeed, there seems to be considerable evidence that jets of most kinds suffer substantial efficiency penalties relative to propellers in this regime, and that this is understood and accepted as an inevitable design trade-off to be made.
+In the case of a submarine, which might spend nearly 100\% of its time travelling at 10kt or less, this disadvantage (as opposed to less than 35\%, as assumed for a surface ship) would become extremely substantial, which will be quantified in later sections.
+
+Moreover, recent literature seems to confirm that efficiency of pumpjets fall off precipitously towards zero the closer that the vessel speed comes to zero.
+This stronly suggests the existence of some fundamental constraint in the physics of the operation of jets, which recent advances in technology seems not to have overcome.
+To more firmly establish the extent to which adaptation or plausible improvments in technology could or could not change this trend, it is worthwhile considering the theoretical limits that are presented by the physics.
+This should further limit the range of error that might be made in estimating realistic bounds of performance for an unknown system.
+
+\subsection{Theoretical Limitations}
+In this section I will appraise some of the available literature, including some of the most canonical works in the design of pumpjets which is now declassified from US Navy funded research (and still extremely frequently cited by contemporary researchers).
+Before doing so, I'll attempt to summarise in simple language the mose essential conclusions in simple language.
+
+My broad conclusion is that the optimisation of a system for minimising cavitation, and hence improving acoustic performance, is a separate and independent design requirement to that of optimising a system for propulsive efficiency.
+These two design requirements can be found either to be in tension, or alignment, in different regimes.
+At high speeds, where cavitation tends to have severe impacts on propulsive efficiency of open propellers, and high mass-flow (which leads to good efficiency) can be achieved through compact ducts and pumps, pumpjets offer very good performance on both metrics.
+Hence, in this regime the two design requirements tend to align relatively well in favour of pump-jets.
+However, in very low-speed, low-load environments, where cavitation effects are negligible, and larger areas need to be engaged achieve good mass-flow and efficiency, the two design requirements align relatively well in favour of an open propeller.
+In a variety of intermediate regimes, the two requirements find themselves in tension.
+
+For this particular consideration, two questions are of paricular importance.
+The first is where the clear transitions in the regimes will tend to lie, with respect to the probable requirements of a conventional submarine.
+The second is what drives the tension between those design requirements in the intermediate regimes, and whether it is a fundamental or inescapble tension in the laws of physics, or simply a technical hurdle which imminent improvements in the state of the art are likely to overcome by engineering improvements or innovation.
+
+The answer to the first question, as discussed in earlier sections, seems to be that the operating speeds of a submarine tend to lie far below the probable levels where pumpjets are even remotely comparable to waterjets for efficiency.
+In later sections I will address further the acoustic comparisons in the speed ranges at 10kt and below.
+
+The answer to the second question seems to be that there the tension between acoustic and propulsive performance at intermediate speeds and loadings is driven by foundational physical requirements.
+That is not to say that optimal means of balancing those requirements will not still be found, and improved, as the art advances.
+However, it does mean that significantly adapting using known variables that can be changed to favour one design requirement are unlikely to yeild large improvements without some trade-off being made to the other.
+
+The reason for this trade-off occurring can be quite simply attributed to the nature of the shroud, or duct, which surrounds the the rotors of the propulsor, and the effect it has on the flow.
+If the duct acts to increase the pressure at the rotor, (which reduces cavitation) it necessarily must slow the flow at/around the rotor, which necessitates produceing a negative thrust on the duct.
+The system as a whole produces a positive thrust, becuase the impeller or rotor pushes a larger positive thrust, but the duct necessarily produces a negative thrust.
+At the speeds get lower, and the thrust required gets much much lower, this negative thrust becomes proportionally more significant, as described later.
+
+Put another way, the work required to slow the water down, then speed it up again, becomes a gradually greater penalty the the slower the water is coming in, and the less you actually want to speed it up.
+If the water is slowed enough by the duct to effectively stop, then the impeller must do the work of speeding it up (sucking it into the duct) before the duct slows it down, before the impeller raises its pressure, and the pressure accelerates it out of the nozzle afterwards.
+Compared to simply accelerating the water in one simple push, as a propeller blade does in an unconstrained flow, the efficiency penalty is very large.
+
+\subsection{The impact of duct loss on an ideal propeller}
+
+Derivations of the 'ideal' efficiencies are frequently made for propellers, and jets using basic conservation laws (momentum theory) which consider simply acceleration of some column of water as actuated by a disc representing the impeller or propeller.
+A fuller description of these equations can be found in  most standard texts \parencite[131]{lewis1988}.
+These methods are computationally simple, and allow simple upper-bounds to be be set for efficiencies achievable for given designs under certain conditions, without considering the portential impact of other more complex effects such as skin friction over the blades, potential flow separation, or cavitation.
+As such, they can be easily relied upon to constrain the overall plausible range of values that can be achieved from universally applicable inputs, such as the velocity of water going into a duct, and the velocity of water exiting a nozzle.
+
+Such methods are used in \cite[8-14]{wislicenus1973} to construct an 'ideal' jet efficiency that applies to waterjets and jetpumps.
+Whilst I won't repeat the full expansion here, but a fundamentally important early conclusion of the derivation is the distinction between 'useful work' and the total energy expended.
+since the rate of change in momentum is known to be necessarily equal to the force generated, and increasing momentum change (hence thrust) can be achieved either accelerating more mass, or accelerating mass to higher speeds.
+Given that energy embodied in a flow is proportional to it's velocity squared, but its momentum only to its velocity, a necessary and inevitable consequence is that some energy is always wasted in accelerating a fluid to a higher velocity than its surrounds.
+
+This is expressed in the equation which represents 'ideal jet efficiency', which is directly equivalent to 'ideal propeller efficiency' using momentum theory of propeller action:
+
+\begin{equation}
+\label{eq:3}
+\eta_j = \frac{1}{1+\frac{\Delta V}{2V_0}}
+\end{equation}
+
+The inevitable consequence of this relationshiop is that efficency approaches unity when $\Delta V$ (the change in water of the water between entering and exiting the system) relative to $V_0$ approaches zero.
+Intuitive embodiments of this relationship are (equivalently) that efficiency is maximum when the least force is exerted on any given bit of water, which might be when negligibly small work is being done (extreme low speed) or when such a large mass of water is acted on that it barely needs to be pushed at all.
+This relationship would lead to the conclusion that infinitely large propellers or jets are always desirable from an efficiency perspective.
+This of course isn't practical to build, and also doesn't hold when the effects of the drag on the propeller blades, or duct, are taken into account.
+Infinitely large systems would increase the total weight, as well as having significant surface draw when moving through the water, all of which are neglected in this 'ideal' consideration.
+
+In \cite[8-14]{wislicenus1973}, Wislicenus expands the relationship in the case of the jet to incorporate terms to represent some of the inevitable adjustments that would occur in the case of pump jets and waterjets, particulalry the duct head loss ($K$) (fraction of energy required to push the water through the duct), as well as the additional drag on the outside of the propulsor ($\Delta T$).
+This leads to an adjusted efficiency relationship, which still neglects any losses due to pump efficiency, or change in height of the jet (which is added for consideration for waterjets, but not relevant here).
+
+\begin{equation}
+\label{eq:4}
+\eta_j = \frac{1}{1 + \frac{\Delta V}{2V_0} + K\frac{V_0}{2 \Delta V}}(1 - \frac{\Delta T}{T})
+\end{equation}
+
+The conclusion of the analysis, and the plotting of these results for certain possible values of $K$ and $\Delta T$ is: "Its outstanding characteristic is that this efficiency approaches zero rather than unity for $\Delta V / V_0 = 0$, even for small duct losses".
+
+\begin{figure}
+\includegraphics[width=\textwidth]{PumpjetEfficiency.png}
+\caption{Ideal jet efficiencies sharply appoach zero at low values of $\Delta V / V_0$, where as ideal propeller efficiencies without a duct $K = 0$ approach unity. Credit: \cite{wislicenus1973}}
+\label{fig:PumpjetEfficiency.png}
+\end{figure}
+
+Wislicenus also acknowledges \cite{wislicenus1973} another author, Brandau, who makes a different assumption about the relationship between the duct loss and water velocity \parencite{brandau1967} in his derivation (along with another thorough discussion of potential and ideal efficiencies), who produces a different peak efficiency point, but retains a similar overall shape featuring efficiency falling very rapidly to zero at low values of $\Delta V / V_0$, and contrasting this with the ideal propeller, which approaches unity in this low regime.
+
+It's important to note at this point that whilst charts such as these are generally used with dimensionless ratios on the axes (such as $\Delta V / V_0$) to allow the most general solutions to be used to solve for any given particular vessel size, speed, or loading requirement which might be required, these  confusing to many readers unfamiliar with hydrodynamics.
+For the avoidance of any confusion, for a particular fixed vessel size and drag, and fixed propulsion system design smaller values of $\Delta V / V_0$ will correspond to to the propulsion system exerting less thrust, and also moving at a lower speed under self-propulsion, since the thust required (proportional to $\Delta V$ when operating at steady state) will decline faster than the vessel speed or water speed at intake, since the drag on the vessel scales with the square of the vessel speed.
+
+\begin{figure}
+\includegraphics[width=\textwidth]{EfficiencyBrandau.png}
+\caption{Alternative derivations of ideal jet efficiency also show sharply declining efficiency for low values of $\Delta V / V_0$ Credit: \cite{brandau1967}}
+\label{fig:EfficiencyBrandau.png}
+\end{figure}
+
+The structure of these overall curves is found to be overall quite similar for both pumpjets and waterjets, though the probably values for the losses incurred in the intake duct, or on the exterior of the shroud, are obviously different \parencite[13-14]{wislicenus1973}.
+Similar efficiency shapes can be found for discussions focussing explicitly on underwater pumpjets \parencite[13]{henderson1964}.
+It should be noted that the theoretical efficiency derived by these general equations is consistent both in derivation and results with those produced by other authors \parencite[227]{lewis1988}.
+
+Care should be taken, however, to note that different graphs for efficiency can be produced which have different variables on the x axis.
+For example, \cite[247]{mollard2011} a graph of ideal efficiency of a pump jet is shown using the speed of advance divided by the jet speed, or $V_1/V_2$.
+Given that the x axis used in the previous charts used $\Delta V / V_0$, which would be related to $V_2-V_1$ in this case, this has the effect of reversing the effective direction of the chart.
+In this case, for a fixed jet and vessel design, high-speed operations will be towards the left of the chart, and the precipitous decline in efficiency will occur as speed declines to zero, along with thrust, toward the right of the chart, as opposed to the left in earlier cases.
+
+\begin{figure}
+\includegraphics[width=\textwidth]{EfficiencyMollard.png}
+\caption{Other modern reference texts might demonstrate jet or propeller efficiencies with different x-axes, effectively reversing the way the graph should be understood for a single system operating at different speeds.
+Credit:  \cite{mollard2011}}
+\label{fig:EfficiencyMollard.png}
+\end{figure}
+
+\begin{figure}
+\includegraphics[width=\textwidth]{EfficiencyBulten.png}
+\caption{Another graph representing efficiency as a function of the ratio of jet velocity to incoming velocity, which shows efficiency declining at low thrust (to the right).
+The impact of increasing duct losses can be clearly seen, as opposed to the 'ideal' propeller without a duct.
+Credit: \cite{bulten2006numerical}}
+\label{fig:EfficiencyBulten.png}
+\end{figure}
+
+Similarly, it is very important to note that the very frequently used x-axis unit $J = V_a/nD$ when showing efficiency curves for both propellers and ducted propellers (such as used in earlier figures) are of a similar orientation to the chart shown in the figure above.
+The left side of the chart will correspond to the highest number of rotations for a given amount of progress through the water.
+For vessel of fixed load, and a propeller of fixed geomtry, this will correspond to a propeller facing the higest drag force due to the vessel moving at the highest speed.
+
+Readers may observe that the efficiency for all propellers also decline to zero at the other end of the spectrum, most sharply at a point where $J$ equals or exceeds the pitch ratio of the propeller.
+As discussed earlier, these correspond to cases where the propeller is slicing along its helical path without exerting any force to move water backwards, or when the slip ratio is zero.
+In this circumstance, the rotation of the propeller would not push any net water backwards, and simply act as a paddle-wheel needlessly spinning water.
+Such circumstances are unlikely to occur to a significant degree unless a vessel was decelerating, or experienced some other force to assist its propulsion.
+It is possible that at extremely low speeds a propeller could somewhat descend to the right of optimal advance ratio, when very dramatically reduced load was encounterd because of the extreme low speed, as might be suggested by the slight reduction in propulsive efficiency of the propeller below 5kt in the BMT study.
+The conclusion if design at no higher speeds was required that using a smaller propeller area, which accelerated slightly thicker slices of water with each blade, would re-optimise efficiency.
+These losses of efficiency, however, could never supercede those of a decelerating ducted propeller or jet, which has the shroud which necessarily imposes an \textbf{additional} drag to the watercolumn in excess of any other resistance faced by the rotor.
+This effect can be seen clearly even or an accelerating duct, as shown in \ref{fig:PropDuct.png}.
+
+\begin{figure}
+\includegraphics[width=\textwidth]{PropDuct.png}
+\caption{Even the addition of an accelerating ducts substantially reduces efficiency of propellers at very low loadings, or for constant loads, low speeds.
+Maximum attainable efficiency is also reduced.
+Their benefit lies in the bulging of the curve at low advance ratios (higher loads).
+Credit: \cite[6]{willemsen2013}}
+\label{fig:PropDuct.png}
+\end{figure}
+
+It should be clear from the above analysis that the efficiency of a pumpjet falling towards zero with a vessels waterspeed is an enduring and foundational result of basic physics.
+The addition of negative thrust on a decelerating duct, no matter how small, necessitates that efficiency must decline at some point with with blade loading, in contrast with an ideal propeller which rises to unity.
+This explains the overall shape given by the the mysterious unquantified chart cited by Andrew Davies \parencite{davies2017} and Access Economics \parencite{stanford2017} shown \ref{fig:PumpUp.png}, and also matches the shapes plotted by more recent studies as shown above, and also the shape presented in other modern publications studying the use of pumpjets \parencite{giles2010} and waterjets \parencite{fujisawa1995}.
+
+\begin{figure}
+\includegraphics[width=\textwidth]{PumpUp.png}
+\caption{The mysterious efficiency comparison chart which has been doing the rounds in the Australian debate probably quite reasonably reflects theoretical physical limits of 'ideal' pumpjets and propellers.
+Credit: \cite{stanford2017}}
+\label{fig:PumpUp.png}
+\end{figure}
+
+In terms of establishing plausible peaks and cross-over points, the study undertaken in \cite{giles2010} probably provides a sufficient indication as to the current state of the art in the relevant technologies.
+Whilst this effort focussed on a surface vessel rather than a   The study was undertaken with the backing of military customers with support and testing facilities (the US Navy), and involved two major companies who are world leaders in maritime engineering (BMT) and waterjet manufacture (Rolls Royce).
+Even if further technical advances through design refinement were possible, as the authors alude to, the cross-over point in terms of efficiency might be brought down to 25kt.
+The prospect that the crossover point lies anywhere near the transit speed of a convetional submarine (8-10kt) appears utterly implausible.
+In discussions with leading world experts involved in the development of advanced ducted propellers (including those involved in citations) I was told that a pumpjet could not be more efficient than a propeller below 18kt.
+
+With the optimum efficiency of a jet probably being no greater than 60\% or 70\% at its peak efficiency according to many theoretical and experimental studies \parencite{lu2015}, which would quite probably lie considerably higher in terms of waterspeed, these facts, along with the inevitable decline of efficiency towards zero as waterspeed also declines, a suitably constrained set of plausible efficiency curves may be investigated numerically, as will be done later in later sections.
+
+\subsection{An assessment of scope for improving the efficiency of a pumpjet at low speeds}
+Before continuing to assess the impact of the likely efficiency curves of pumpjets relative to propellers, it is worthwhile considering what the known literature about the theory of pumps, jets, and turbomachinery might tell us about the potential for a radical advance which make might substantially improve the shape of the curve at very low speeds.
+Whilst the field of propulsor design is no doubt complex, and I will not attempt to be exhaustive in detailing all of the potential design considerations which might make an improvement.
+However, the pursuit of reasonably comprehensive confidence is aided by the the scale of the gap which probably needs to be closed to make pumpjets comparably efficient to propellers at the relevant speeds.
+Minor adjustments and the improvement of small or marginal effects, of which there might be many, will clearly not help in shifting the efficiency peak of a jet from over 25kt to under 10kt.
+In other words, this gap will not be closed or noticeably reduced by 'tuning'.
+Quite substantial redesign of core characteristics (like mass flow) would be required, which would necessitate a cycle of redesign and tuning virtually all other aspect of the entire jet, as illustrated by numerous guides to the design of such systems (\cite{bruce1974}, \cite{furuya1988}, \cite{mccormick1963design}, \cite{henderson1964}).
+As such, I will only consider those parameters which could plausibly be altered quite substantially, where the consideration of larger changes make the likely limits or trade-offs quite clear.
+
+\subsubsection{Increase mass-flow by widening area of intake}
+Perhaps the most obvious means by which one might attain substantial increases in efficiency is to increase substantially the mass flow rate.
+This is derived from a fundamental relationship, as expressed earlier in the representation of ideal propeller and jet efficiency (prior to the consideration of any drag or other losses) which was given in the earlier equation.
+Since excessive velocity differences result in more energy expended than momentum changed, as given by momentum theory and basic mechanics, doing work on a large body of water, rather than working harder on more of it, is generally desirable from an efficiency perspective.
+Considering what the appropriate mass-flow should be at design speeds is generally the first thing that is specified to occur in all design processes for pumpjets, and good example of which is in \cite[6-7]{henderson1964}.
+
+As Henderson et al.
+note, the key constraint of this variable is that increasing the diameter of the pumpjet necessarily leads to an increase in the surface (including the external suface) of the shroud.
+The increased drag that this induces must be deducted from the thrust that the pumpjet produces, or effectively increasing $\Delta T$ as given in earlier equations.
+In this respect, there is "no free lunch" to be had in expanding the size of the intake.
+As Henderson et al.
+conclude, for this reason the diameter of a pumpjet tends to be about 15\% or 20\% smaller than that of a corresponding propeller, whereas the diameter to have equivalent mass-flow, highlighting the tension set up in pumpjet design with their earlier observation that "if the cavitation resistance of the blades is to be improved with no reduction in thrust and the same rate of flow through the propulsor, the pump jet rotor must be larger than that of the propeller." \parencite[1]{henderson1964}
+
+It should also noted that in order to substantially increase the opening area of the propeller, the difference between the velocity of movement at the blades nearer to the root (close to the hub) and the extremities of the blades increases substantially.
+In order for equivalent work to be done on all of the water distributed radially across area of the impeller, quite a different blade shape might be required, in order to provide a higher deflection of the water close to the root.
+In these areas, there is increase probability of flow separation occurring, where the velocity of movement of the water across the surface is insufficient to overcome the pressure gradient required to move over the blades.
+(In a pump, the pressure necessarily increases as work is done on it.)  Or, since there isn't enough pressure exerted on the whole of the water column to overcome the resistance imposed by the duct, not all of the water moves forward, and some starts to move backwards.
+This leads to flow separation, or stall occurring close to the blade roots, which dramatically diminishes the efficiency of the system.
+A further discussion of this effect can be found in \cite[15,27]{henderson1964},  \cite[807,8012]{bruce1974},  \cite[185]{wislicenus1973}, \cite[60]{mcbride1979} and many other related texts.
+This is a general problem that affects a range of axial-flow turbomachinery, particulalry operating in off-design conditions, and will present a general concern for the design of a new pumpjet \cite[185]{wislicenus1986}, \cite{li2013}.
+
+\begin{figure}
+\includegraphics[width=\textwidth]{stall.png}
+\caption{Flow separation, or stall, can occur when the flow over the blades of a pump slows too much, as shown in \cite{li2013}}
+\label{fig:stall.png}
+\end{figure}
+
+\begin{figure}
+\includegraphics[width=\textwidth]{backflow.png}
+\caption{Stall results in water flowing backwards near the hub of the blades at low flow rates, as shown in \cite{li2014rotating}}
+\label{fig:backflow.png}
+\end{figure}
+
+It is possible to limit the variation of deflection between the blade tips and roots by simply expanding the size of the hub.
+However, this necessarily reduces the effective frontal area that is available to receive the mass-flow, which necessitates further expansion of the outer diameter, and a very substantial increase in the weight of the system.
+In addition, expansion of the hub will necessarily increase the total surface area relative to the mass flow, including by increasing the amount of tapering required after the stators in order to prevent risk of cavitation occuring around the hub.
+
+Increasing the frontal area of the intake also is necessarily linked to the advance ratio of jet.
+Since low advance ratios correspond to higher shaft speed for a given forward velocity, increasing the thrust produced for a given rotation (by widening the jet area) necessarily results in higher advance ratios, since more work is done on the water for a given rotation.
+However, this also necessarily results in poorer cavitation performance of the jet, since higher blade velocities relative to the water flow result in necessarily increased propensity for cavitation.
+A detailed discussion of this can be found in \cite{gearhart1966selection}, but similar tensions can be readily found in most other pieces of literature, including \cite{henderson1964}, \cite{wislicenus1973}.
+
+Given the fundamental nature of the flow-rate and intake area to the design of pump-jets, and the great lenght of time for which they have been understood and discussed as a crucial design input for a waterjet, it is altogether implausible that some large increase in intake area can be achieved without incurring necessary and unavoidable tradeoffs.
+
+\subsubsection{Reduce degree of diffusion (i.e. switch to accelerating duct) to reduce negative thrust from duct}
+A further means of reducing the negative thrust produced by a duct may involve significantly reducing the degree of diffusion, or compression that occurs in advance of the rotor, since the process of diffusion necessarily slows the water and produces negative thrust.
+However, this amounts to transitioning towards an accelerating duct, or Kort Nozzle.
+Whilst nozzles of this type have been known to improve the cavitation characteristics of a propeller by enclosing the tips (which often tend to cavitate first), their ultimate function is to actually decrease the pressure at the blades of the propeller.
+Consequently, they necessarily increase the propensity of other types of cavitation to occur across the blade-face, including bubble, cloud and sheet cavitation.
+As such, these types of propellers are not considered for the military purposes which require an absolute minimum of noise to be created.
+Transitioning to an accelerating duct signify a shift away from a pump-jet, and necessarily be in direct tension with the need to minimise cavitation for acoustic considerations.
+
+In addition, whilst it is often demonstrated that for a given propeller under certain load conditions (high loading), the addition of an accelerating duct can considerably improve efficiency, all of these situations have generally been in the context of vessels which are designed for a relatively high thrust, low waterspeed situation, when propeller size is necessarily somewhat constrained.
+This is the case fishing vessels, or for tug boats, which need to impart significant thrust to a vessel dozens or hundreds of times their own size.
+Mounting a propeller the equivalent size of the vessels they mean to push, however, would be excessively heavy, arranging a mount and power drive that would keep the entire propeller submerged, and not have the effect of flipping the boat backwards when pushing hard, would be effectively impossible, without proportionally increasing the size of the tug.
+Furthermore, in most circumstances these propellers are still designed to operate with some cavitation, even if the extent of it is reduced by the presence of the duct \parencite{haimov2010}.
+
+In other words, the use of highly loaded propellers in such applications is actually particular to design constraints that are unlikely to be present for a submarine, which can position the propeller at the tail of its body, directly behind its centre of mass and drag, and consequenlty have freedom about the optimal size of the blades.
+It is well known that a duct cannot improve the performance of a low-loaded propeller which would be the situation for a slow-moving vessel with unconstrained propeller size.
+Authoritative references in the literature include \cite{oosterveld1970}, who calculates the minimum thrust coefficient $C_T$ (a measure of loading) for an accelerating duct to improve a propellers performance to be between 1 and 2, as shown in @AcceleratingIdeal.
+
+\begin{figure}
+\includegraphics[width=\textwidth]{DuctedEfficiency.png}
+\caption{Calculations of 'ideal' efficiencies of accelerating ducts (neglecting drag) in \cite{oosterveld1970} show that a duct can only improve the efficiency of a propeller if it is loaded above a certain degree, and that longer ducts produce better improvements in 'ideal' efficiencies}
+\label{fig:DuctedEfficiency.png}
+\end{figure}
+
+A parallel design change in the switch away from the degree of deceleration in the duct would be the reduction of the nozzle contraction after the rotor.
+For waterjets and pumpjets of all sorts, based on fundamental disc actuator theory, the degree of this contraction of the flow area has also been long identified as a means to increase pressure at the rotor, and hence decrease cavitation, but at the direct expense of efficiency \parencite[14]{van1996waterjet}.
+
+Numerous efforts to find suitably compromised duct shapes exist, but even advanced modern designs still exhibit declining efficiency at low speeds, as well as some cavitation during operations at design loads, which still tend to imply relatively high speeds, for example 27kt in \ref{fig:FineOptimised.png} .
+Consideration of accelerating duct designs for a military submarine would be extremely unlikely, and be a profound departure from any existing pumpjet designs.
+
+\begin{figure}
+\includegraphics[width=\textwidth]{FineOptimised.png}
+\caption{Extensive work has been done exploring compromise designs for decelerating ducted propellers (pumpjets), including with very modest deceleration in the intake, limited shroud taper, short duct lengths, or essentially incporate some of the advantages of accelerating ducts.
+Credit: \cite{abdel2010}}
+\label{fig:FineOptimised.png}
+\end{figure}
+
+\begin{figure}
+\includegraphics[width=\textwidth]{FineEfficiency.png}
+\caption{In experimental testing such designs still exhibit sharply falling effiency at high advance ratios, or low speed. Credit:  \cite{abdel2010}}
+\label{fig:FineEfficiency.png}
+\end{figure}
+
+\begin{figure}
+\includegraphics[width=\textwidth]{FineCavitation.png}
+\caption{TAt optimally efficient conditions (corresponding to 27kt) such designs still exhibit some cavitation. Credit: \cite{abdel2010}}
+\label{fig:FineCavitation.png}
+\end{figure}
+
+\subsubsection{Reduce shroud length to decrease drag on duct}
+The reduction of the length of the shroud in order to reduce drag presents another possible means to lower the duct resistance, as shown and discussed in \cite[21-25]{oosterveld1970}.
+The fundamental difficulty with greatly shortening a duct, however, is that it requires that any changes in the shape and with of the water column be accomplished much shorter distance for the duct to achieve its desired effect, which means that generally it has to achieve less of it.
+This means that extremely very low duct lenght achieve less of an efficiency improvement to the ideal efficiency of a propeller, in those circumstances where they can achieve an improvement  (accelerating ducts), as discussed above.
+
+If more aggressive shape changes are attemped in a limited space, the deflections of flow and pressure gradients become much more extreme.
+This leads to a far greater propensity for flow separation to occur.
+This can happen inside the duct ahead of the stator, particulalry when any degree of diffusion occurs in order to elevate static pressure at the blades, or any kind of direction change is required as in the case of waterjets.
+Unlike waterjets, pumpjets also have the potential for separation on the exterior surface, which is an additional concern in the design of duct \cite[13]{henderson1964}.
+Given the role of the duct is to change the speed of the flow at the impeller relative to what would occur in its absence (in particular to decelerate it) separation might tend to occur on different surfaces in different operating conditions, depending on the relative pressure gradients and velocity ratios they experience at different speeds and loads.
+This substantially complicates the challenge of designing a duct that performs well across a wide speed range.
+This is illustrated for the case of a waterjet in \cite[20]{bulten2006numerical}.
+Equivalent complexities occur in pumpjets operating in low speed ranges (2-6kt, where flow separation occurs ahead of the impeller, inside the shroud.
+In the case of an accelerating duct, this phenomenon is also observed at high advance ratios, or the equivalent of low speeds and low loads, as shown and discussed in some detail in \cite{willemsen2013}.
+
+\begin{figure}
+\includegraphics[width=\textwidth]{LowIVR.png}
+\caption{A an intake may experience risks of flow separation even at low speeds (Inlet Velocity Ratio) on some surfaces due to certain flow deflections being more severe.
+Source: \cite{bulten2006numerical}}
+\label{fig:LowIVR.png}
+\end{figure}
+
+\begin{figure}
+\includegraphics[width=\textwidth]{HighIVR.png}
+\caption{A an intake may experience risk of flow separation on different surfaces at high speeds (Inlet Velocity Ratio) where other flow deflections are more severe.
+Source: \cite{bulten2006numerical}}
+\label{fig:HighIVR.png}
+\end{figure}
+
+\begin{figure}
+\includegraphics[width=\textwidth]{DuctSeparation.png}
+\caption{Flow separation occurs at on the outside of an accelerating duct at high $J$ values, which corresponds to low loads, or speeds.
+For high $J$ values, the pressure reduction on the inside of the duct, creating positive duct thrust, can be seen.
+Source: \cite[73]{willemsen2013}}
+\label{fig:DuctSeparation.png}
+\end{figure}
+
+\begin{figure}
+\includegraphics[width=\textwidth]{NozzleDrag.png}
+\caption{Dramatically reducing the length of a duct relative to its diameter ($L/D$) as shown in \cite{oosterveld1970} can reduce its drag, which has the most significant impact at low thrusts or speeds.}
+\label{fig:NozzleDrag.png}
+\end{figure}
+
+In addition, the substantial reduction in the length of the shroud, both ahead of and behind the rotor, will also have the substantial effect of diminishing any acoustic sheilding which the duct may achieve.
+
+\subsubsection{Summary remarks on potential for redesign of pumpjet for low-speed conditions}
+Overall, while the parameter-space for the alteration of marine-propulsors is large, the parameters which would allow for very dramatic rather than minor changes to propulsor efficiency at low speed are relatively few.
+The major ones include increasing the width of the intake, decreasing the length of the duct, and decreasing the pressure inrease at the blades achieved by the duct.
+In short, it involves evolving the pumpjet back towards a propeller.
+None of these changes could be effected to a substantial degree without some inevitable trade-off in terms of the cavitation performance of the system at higher speeds.
+
+\section{Results of numerical simulation}
+
+\section{Acoustic Considerations}
+It is generally considered that the existence of cavitation of any sort will tend to dominate the acoustic signature of any vessel that doesn't produce any machinery noise by relying on all-electric drive as a conventional submarine would when dived (\cite{giles2010}, \cite{harvie1965construction}).
+Its quite possible that during the period when a submarine is snorting, or running its diesel engines, these sources of noise would be a more substantial and distinctive acoustic signature than small amounts of cavitation on the propulsion system, if it were to occur, owing in particular to the low freqency of diesel engines.
+
+Other sources of noise hydrodynamic noise, such as the turbulence produced in the boundary layer of the flow over the hull, or parts of the propulsor, will still contribute to overall noise and some signature, but it will tend to be vastly smaller in magnitude.
+This owes to the essentially incompressible nature of water.
+Without compression, even very relatively intense votices or tubulent flows will not propagate very much noise, if the water remains perfectly solid.
+All of the 'round and round' can't achieve much 'in and out', which is what results in propagated noise.
+This obviously doesn't hold true in compressible fluids, such as air, as one would observe noticing the intense noise produced by hand-dryers, aeroplanes, jet engines etc.
+It also doesn't hold true within a turbulent flow underwater.
+If you were to put a sensor inside a turbulent flow, the rapid movements would result in continual and significant pressure fluctuations, hence noise.
+But at a distance, these fluctuations tend to substantially cancel out.
+The larger the scale of the eddies and vortices, the further noise might propagate, and also the lower the frequencies of resultant noise might be, which makes noise propagate further.
+Likewise, larger caviation voids will tend to emit lower frequency sound aswell, as can be observed from the boiling of a kettle.
+
+However, when dived, it might be reasonably assumed that the complete elimination of all sources of cavitition will be the primary objective of acoustic signature management.
+This is neatly put in \cite[88]{gearhart1966}: "The occurence of cavitation in a propulsor results in adegradation of the propulsor operating characteristics if extensive amounts of cavitation exist, and in significant noise when minute amounts occur.
+As a result, the performance requirements of a submersible weapon system specify that the occurence or inception of cavitation be avoided below certain operating depths." One might consider that a submarines 'tactical silent speed' would most probably correspond to the speed at which it can operate safely below the point of any cavitation inception.
+
+It is crucial to understand that increasing depth necessarily increases the static pressure throughout a propulsion system.
+As such, the speed at which cavitation starts to occur, will necessarily be a function of depth.
+Approximately every 10m of water adds approximately the equivalent of an atmosphere of pressure.
+As such, the static pressure a system experiences at 10m depth will be half that experienced at 30m (since the air in the atmosphere comprises the first atmosphere of pressure).
+Consequently, diving deeper will necessarily increase the speeds one can achieve without any cavitation occurring \ref{fig:InceptionDepth.png}.
+
+\begin{figure}
+\includegraphics[width=\textwidth]{InceptionDepth.png}
+\caption{Increasing in depth improves the speed achievable before cavitation begins to occur, as shown in the case of a torpedo \parencite{suryanarayana2010cavitation}}
+\label{fig:InceptionDepth.png}
+\end{figure}
+
+It is also important to understand that the addition of a duct will necessarily to flow separation occuring on the inside of the duct opening at very low speeds which correspond to high advance ratios when load remains constant.
+Flow separation tends to lead to vortex shedding, which means that inconsistent flows (vortices) are then ingested into the blades of the impeller.
+This effect is well known and document.
+A good example is shown on the exterior of an \textbf{accelerating} duct at $J = .8$ in \cite{willemsen2013}.
+
+\begin{figure}
+\includegraphics[width=\textwidth]{VortexShedding.png}
+\caption{Vortex shedding following a flow separation leads to eddies being released unsteadily into the main flow stream, as shown here on the exterior of an accelerating duct.
+In the case of decelerating duct (pumpjet), duct shape is inverted, leading to the vortices being shed into the impeller blades.
+\parencite{willemsen2013}}
+\label{fig:VortexShedding.png}
+\end{figure}
+
+Whilst at very low power levels, or at good depths, it is quite plausible that this doesn't result in any cavitation, which would normally be a likely occurrence if the blades are operating closer to their design power.
+As such,
+
+\section{Conclusions}
+
+% \printbibliography
 

### 2018-01-02 pdfs/Untitled.Rmd

--- 2017-12-27 DraftExec.Rmd
@@
 date: "11/12/2017"
-output: bookdown::pdf_document2 
----
-
----
-references:
-
-- id: garland2005
-  title: Elementary Physics of Reactor Control Module - Fission Product Poisoning
-  author:
-  - family: Garland
-    given: Bill
-  volume: 3
-  URL: 'http://www.nuceng.ca/br_space/2017-09_4d03_6d03/misc_files/xenon_poisoning.pdf'
-  publisher: Nuclear Engineering (Originally from McMaster University)
-  type: article-journal
-  issued:
-    year: 2005
-
-- id: carlton2007
-  title: Marine Propellers and Propulsion
-  author:
-  - family: Carlton
-    given: John
-  URL: 'http://www.nuceng.ca/br_space/2017-09_4d03_6d03/misc_files/xenon_poisoning.pdf'
-  publisher: Elsevier
-  issued:
-    year: 2007
-
-- id: liang2014
-  title: The economics of slow steaming
-  author:
-  - family: Liang
-    given: Lee Hong
-  URL: 'http://www.seatrade-maritime.com/news/americas/the-economics-of-slow-steaming.html'
-  publisher: Seatrade Maritime News
-  issued:
-    year: 2014
-    month: 10
-    day: 7
-
-- id: sanguri2012
-  title: The Guide to Slow Steaming On Ships
-  author:
-  - family: Sanguri
-    given: Mohit
-  URL: 'https://www.marineinsight.com/wp-content/uploads/2013/01/The-guide-to-slow-steaming-on-ships.pdf'
-  publisher: Marine Insight
-  issued:
-    year: 2012
-    month: 12
-
-- id: WNA2017
-  title: Nuclear Powered Ships
-  author:
-  - family: 
-    given: 
-  URL: 'http://www.world-nuclear.org/information-library/non-power-nuclear-applications/transport/nuclear-powered-ships.aspx#ECSArticleLink0'
-  publisher: World Nuclear Association
-  issued:
-    year: 2017
-
-- id: WNA2009
-  title: Chernobyl Accident Appendix 1
-  author:
-  - family: 
-    given: 
-  URL: 'http://www.world-nuclear.org/information-library/safety-and-security/safety-of-plants/appendices/chernobyl-accident-appendix-1-sequence-of-events.aspx'
-  publisher: World Nuclear Association
-  issued:
-    year: 2009
-
-- id: patrick2012
-  title: The Driving Factor in the SEA 1000 choice
-  author:
-  - family: Patrick
-    given: Rex
-  URL: 'https://corporate.siemens.com.au/content/dam/internet/siemens-com-au/root/aunz-defence-solutions/apdr-october-2015-issue-future-submarine.pdf'
-  publisher: Asia-Pacific Defence Reporter
-  issued:
-    year: 2015
-    month: 10
-
-- id: hamilton1997
-  title: Jet Torque
-  author:
-  - family: Waterjets
-    given: Hamilton
-  volume: 8
-  URL: ''
-  publisher: Hamilton Waterjets
-  issued:
-    year: 1997
-    month: 3
+output: pdf_document
 ---
@@
 knitr::opts_chunk$set(echo = TRUE)
-library(ggplot2)
-library(knitr)
-
 ```
 
-# Contents
+## Executive Summary
 
-# Introduction
-
-The purpose of this paper is to investigate the suitability of the pump-jet selection for Australia's future fleet of submarines.
-In particular, the question of the relative efficiency, and quietness of pump-jets in comparison to open propellers is of particular importance.
-Much has been made of the significance of the pump-jet in the DCNS (now Naval Group) bid for the submarine, to the extent that it has been claimed that the pump-jet rendered propellers obsolete.
-
-Necessarily, this topic depends upon a considerable amount of physics, particularly around hydrodynamics, fluid mechanics, and turbomachinery, all of which might be difficult or daunting topics to usefully inform a public policy debate.
-However, in the case of a $50 billion dollar military acquisition, neglecting to engage with the technical matters that are so crucial to such a decision would be deeply foolish.
-To that end, this paper has a very specific ambition.
-It doesn't seek to advance the technical field with any new research or insight.
-Instead, it aims to make as much of the relvant physics as possible to this particular question comprehensible in laymans terms, and connect the essential concepts as directly as possible with the most important conclusions of public import.
-
-As such, it will tend to make frequent use of diagrams, illustrations, as well as simple prose, and relatively infrequently use mathematical equations, (unless of crucial importance), though some effort is made to point to references where the fuller and more formal derivations of these relationships can be found.
-
-It should be noted that the technical fields involved here are vast and rapidly evolving.
-The author makes no claim to be a world authority on the entirety of the topics which are touched.
-It is quite possible that there are other effects and phenomena which could be relevant which aren't mentioned here, and it's also true that additional advances in the state of the art are being made continually, and might not be known about in public literature.
-
-However, the nature of scientific discovery and technical progress is that many of best discoveries are actually built-upon and confirmed, rather than swept aside, by later developments.
-As such, plenty of foundational principals, in particular the conservation laws of energy and momentum and essential principals of mechanics and thermodynamics are just as true, right, and relevant as they were hundreds of years ago.
-Consequently, whilst we cannot always know the precise degree of advance in a particular field of engineering, we can still confidently know some of the fundamental bounds and constraints that will be inherent in that field.
-
-In this paper, since we cannot know the precise details of the particular state of the art in largely classified military programs, I will attempt to be particularly clear about those things for which we can have great confidence (often broad principals or relationships which determine trade-offs) and those things which we can't (the precise degrees or points and sensitivities).
-To that end, in this paper I've underaken modelling based upon the broad principals which we can have confidence, and conducted sensitivity testing based on a range of plausible values which seem realistic for those things about which we can have less confidence.
-I have also, in discussion with experts and through reading the available literature, attempted where possible to identify the most plausible of the uncertain values, in order to advance discussion.
-
-With the support of the principal sponsor of this paper, the relevant models may be found in a user-friendly format as a web-app to allow their further interrogation and testing for other plausible scenarios.
-In addition, for increased transparency, the relevant code and equations underlying are available on github.
-
-# Executive Summary
-
-#### The difference between nuclear and conventional for speed
 This paper was commissioned in order to investigate whether pumpjets could plausibly be as efficient, or more efficient than a suitably designed open propeller for a conventional submarine.
@@
 
-#### The importance of low speed operation for diesel-electric submarines
 The efficiency of the propulsion system at such low speeds is of great significance for a conventional submarine, since it must rely on batteries or other air-independent propulsion sources for power when entirely submerged.
@@
 
-#### What is cavitation
 Pumpjets have been widely adopted by navies operating nuclear submarines.
@@
 
-#### How does a pump-jet work
 A fundamental requirement of a pump-jet in order to avoid this cavitation is that the the working parts of the jet (the rotating blades inside it, known as the impeller) operate at a higher pressure than propeller blades operating in open water would.
@@
 
-#### The necessity of drag, and decelleration, induced by the duct
-The role of the shroud (or tunnel, mantel, duct) around the jet is to allow the pressure to be raised around the impeller, in a way that is not possible for an open propeller.
-By a fundamental requirements of physics, this actually requires that the water's incoming speed be **reduced** before it reaches the impeller.
+The role of the shroud (or tunnel, mantel, duct) around the jet is to allow the pressure to be so raised around the impeller, in a way that is not possible for an open propeller.
+By a fundamental requirements of physics, this actually requires that the water be **reduced** before it reaches the impeller.
 (The kinetic energy embodied in movement is converted into potential energy, or pressure.)  Whilst the duct narrowing at the nozzle also necessarily accelerates the water, (as the additinal pressure imparted by the impeller is converted back to kinetic energy) it's an essential feature of all pump-jets that the water flow is decelerated at the point of reaching the impeller.
@@
 
-#### What happens to a jet at low speed, in simple terms
 The problems for pump-jets arise when the water is already going slow, and you don't want it to go that much faster.
@@
 
-#### Why this isn't easily noticeable in normal circumstances
 It is for this reason that waterjets of any kind are known to have an efficiency curve that falls off towards zero as the net thrust they produce also diminishes to zero (which for a given vessel will correspond to water speed).
@@
 
-#### What the submarine requirement tells us about the duct
 Whist estimating the exact shape and level of the efficiency curve for a particular pump-jet and propeller is impossible without detailed knowledge of their design, the over all trends of their shapes in the extremes can be known from well-established principals.
@@
 
-#### Quantitative Conclusions
 In the production of this paper I have developed a computational model which maps the impact of different efficiency curves directly to the dived range and endurance of a submarine.
@@
 
-#### The acoustic advantage of jets at higher speed
 My review of the literature confirms that a pumpjet may produce a significant acoustic advantage in circumstances where an open propeller would experience any degree of cavitation.
@@
 
-#### Turbulence and flow separation at low speed
 However, at much lower speeds such as patrol speeds (2-4kt) it is most likely that a propeller will be able to operate well below the point of any cavitation inception, and would consequently also be extremely quiet.
 Moreover, a jet will necessarily incur substantially larger degrees and types of turbulence in order to produce net thrust in this regime.
-These likely include discontinuities and instability in the flow entering the duct and passing through the impeller and stator (flow separation) as the water is acelerated sufficient to be slowed, and then re-accelerated.
+These likely include discontinuities and instability in the flow entering the duct and passing through the impeller and stator (flow seperation) as the water is acelerated sufficient to be slowed, and then re-accelerated.
 In certain circumstances where resonances might arise these could be highly adverse to acoustic performance.
@@
 
-#### The acoustic question at low speed
-But in either case, given the necessarily raised levels of turblence generated by a jet than a propeller at very low speeds, the claim that the jet is quieter in this regime must rely entirely upon shielding effects from the shroud.
+But in either case, given the necessarily raised levels of turblence generated by a jet at very low speeds, the claim that the jet is quieter in this regime must rely entirely upon shielding effects from the shroud.
 These could be substantial in a directions perpendicular to the direction of travel, but would be much smaller when viewed from the aft or forward directions.
 It should also be noted that many underwater acoustic environments, sound tennds to reflect and bend in different directions as it propagates.
-As such, the claim that pump-jets are universally quieter than propellers should be treated with caution.
-In a variety of circumstances, including most operations at patrol speed, it may not always be true.
+As such, the claim that pump-jets are universally quieter than propellers should be
 
-#### The inevitable trade-off
 However, as a direct and necessary result of this higher tactical silent speed, some disadvantages will be incurred, owing to the additional drag induced by the shroud which prevents cavitation at the impeller.
@@
 
-#### The contradiction between stated requirement and chosen technology
-Given the substantial emphasis which was placed on overall range and endurance of the submarine, it is difficult to understand how one particular unique requirement would be elevated so high above the other strategic and tactical advantages afforded by propellers.
+Given that
 
-# Speed and Drag - Why very slow is very very $(very)^2$ economical
+When you click the **Knit** button a document will be generated that includes both content as well as the output of any embedded R code chunks within the document.
+You can embed an R code chunk like this:
 
-Perhaps the most important relationship to understand is the relationship between speed and drag as it pertains to submarines.
-This is immportant because it sets out the fundamental framework as it applies to any submarine, regardless of propulsion type.
-(It's also a pretty important for planes, cars, missiles, torpedoes, and basically everything else.)
+```{r cars}
+summary(cars)
+```
 
-Drag is the resistance that a fluid (air or water, in our case) gives to a body that is passing through it.
-Quite simply, it's a force that acts in the opposite direction.
-There are multiple sources of drag for different types of scenarios.
-For scenarios where an object is in contact with two different types of fluid (like a ship, on the ocean) or when the fluid doesn't really have contact with all of the object (supersonic flight, and supercavitating torpedoes) some more complex physics applies.
-For a fuller discussion of types of drag, see [@carlton2007]  But in the case of a submarine, which does it's business completely immersed in the ocean, the relevant physics is dominated the skin friction on the hull, which follows a very simple rule and relationship.
-The amount of drag ($F_D$) an objet experiences increases directly in proportion to the surface area $A$.
-For any given object of a certain (unchanging shape) there will be a constant coefficent ($C_D$) which reflects how aerodynamic or hydrodymanic the shape is.
-The drag is also directly proportional to the density of the fluid being moved through, $\rho$.
-(Air creates roughly one-thousandth the drag as water does on any given object at a given speed, since it's roughly one thousand times less dense.)
+## Including Plots
 
-But the most sensitive factor in this relationship is the speed at which the object moves through the liquid.
-The drag increases not with the speed ($v$), but with the **square** of the speed.
-This means if the speed doubles, the drag increases by four.
-If the speed triples, the drag increases by a factor of nine.
+You can also embed plots, for example:
 
-$$F_D = \frac{1}{2}\rho v^2C_DA$$
+```{r pressure, echo=FALSE}
+plot(pressure)
+```
 
-\begin{equation} \label{eq:1}
-\sum_{i=0}^{\infty} a_i x^i
-\end{equation}
-
-(Technical aside: This law applies wherever the the flow over the surface is turbulent.
-It is true that for very small objects, or very viscous fluids, or very slow movements a different apples called Stokes Equation, in the case wher Reynolds numbers are less than 1.
-Given that sea-water is not particularly viscous, and sumbarines not particularly small, Reynolds numbers are liekly to be much much greater than 1 (one or two thousand), even when moving at only one or two knots.
-Since it is unlikely that a significant proportion of the flow over the hull will be laminar, we'll use the drag equation in all modeling going forward when considering drag on the hull).
-
-It's crucial to understand, however, that drag is only a force, and doesn't directly inform us about how much energy is consumed, until we multiply it by the **distance** over which it is applied, not the time for which it is applied.
-To think about it simply, gravity exerts a force on you downwards all the time.
-But you don't expend any energy overcoming it when sitting still.
-If you climb stairs, the amount of gravitational potential you attain depends on how high up you climb, not how long you spend on the ladder or stair-case.
-
-This has a significant consequence for propulsion, since the amount of power (energy expended in a given time) required for thrust scales with the drag, multiplied by the distance covered in a given time.
-As such, power required (and fuel/battery consumption) scales with velocity **cubed** rather than velocity squared.
-This means that if you double your speed of travel, fuel consumption for a given period of time will increase by a factor of eight.
-Due to the increased speed of travel, the fuel required to cover a given distance will only quadruple.
-Hence the range that can be covered with a given amount of energy (assuming propulsion dominates energy requirements) tends to scale with the inverse of velocity squared, whereas the endurance (amount of time that can be spent travelling) scales with the inverse of velocity cubed.
-
-This has a profound impact on the operation and engineering of maritime vessels.
-Reducing velocity has such a substantial reduction on hull drag that going slower is almost always a reasonable means of conserving fuel overall.
-To the extent that time is non-critical, slower is always much much better.
-It's for this reason that during the financial crisis, many cargo shipping companies adopted the practice of slow steaming [@liang2014] in order to conserve fuel, despite this causing a range of possible new engineering issues which need to be accounted for in order to operate the engines at lower than normal power for sustained periods [@sanguri2012 pp.
-8-10].
-The fuel saving from operating even 30% slower means that around half the total energy is required for a given journey.
-Other incremental costs and inefficiences from operating 'off-design' are frequently outweight by such a dramatic reduction in overall power demand.
-
-This effect is of profound importance to understanding the operations of conventional submarines, which have extremely constrained energy stores when operating under the surface.
-Diesel fuel when burned has an energy density of approximately 45MJ/kg.
-In contrast, a lead-acid battery might have energy densities in the range of 0.08-0.14MJ/kg.
-With something like 400 times as much energy per kg embarked in diesel, operating on batteries imposes an extreme demand for economy on propulsive power.
-Happily for submarines, the square law for drag, and cubed law for power, allow an almost commensurate reduction in power demands to take place by slowing down to very slow speeds when submerged.
-Hypothetically, a conventional submarine might have a total range of 10,000nm from using it's diesel payload at 8kt.
-If an equal weight of lead-acid batteries as fuel were carried, the submarine could only travel about 25nm when submerged at the same speed in a single charge.
-But by travelling at 4kt, that quickly increases to 100nm, and 400nm at 2kt (neglecting hotel load here for simplicity).
-
-It's worthwhile pointing out that extreme demand for economy imposed by the poor energy density of batteries distinguishes submarines quite remarkably from other types of boat or ship design.
-In every other application, the propulsion system is designed around a particular speed and loading condition which it is optimally efficient for, and a band of plausible variation around this.
-For example, a cargo ship designed for operation at 22kt will have it's propeller and engine (propulsion system) very carefully designed around these speeds and the plausible levels of loading at those speeds.
-A top-speed which might be somewhat higher than 22kt would be calculated, but would be relatively unlikely to be used for any purpose other than an emergency.
-Lower speeds might be considered 15-18kt for the purposes of slow-steaming.
-However, the relative efficiency at 3kt is of little concern, since the drag at that speed will be under 2% of what it would be at 22kt, and the power required around 0.25%.
-Even if the efficiency was substantially worse, or better (by a factor of two or three even) the impact on the overall economics of the operation would be negligible when compared to marginal improvements at the design speed.
-
-The same consideration applies to other types of vessels, which might use water-jets.
-Fast pilot-boats, for instance, might spend quite a considerable time manouvering in and out of harbour or alongside at very low waterspeeds.
-However, their overall fuel consumption is still likely to be dominated by the fast section of their trip to a ship.
-If the in-harbour water-speed was 8kt, but 32kt was the optimum speed for the open-water section of their journey, the fuel requirement for a given distance could still be over ten times greater at 32kt.
-(Aside, the physics described above would say 16 times, but some reduction due to the reduction of surface area from planing might offset this.)  Consequently the owner/operator would be at least ten times as concerned about efficiency at 32kt as at 8kt, (unless the harbour transit was much much longer than the open water section), and have negligible concern about effiencies at 2kt or 3kt, which would rarely amount to more than a couple of percent of total propulsion energy demand.
-
-In stark contrast, a submarine operator might have less than 1% of the energy store available for the submerged movements at 2-3kt, and consequently be substantially **more** concerned about efficiency in this regime, particularly since this would be the regime in which all of their combat operations would be conducted.
-
-This imposes a constraint on the way in which literature on jets, propellers, and propulsion systems in general is read, since almost all of it is deeply concerned with how a particular system would operate at one particular design speed, and choosing the optimum system for that design speed and load.
-Considerable attention is given to how a propeller would work in off-design conditions relating to changes in load, such as in cases when a ship is more lightly laden, or a tug-boat is pushing a different sized ship.
-There is, however, a relative scarcity of attention given to the performance of a propulsion system operating at a dramatically different speed, with the same load.
-It also corresponds to scenarios where engineers and manufacturers seem from discussion to be far less confident in making quantitative claims about the performance of their own systems.
-(They are perfecly happy to make qualitative claims, "it would be alright", "nothing very bad would happen" etc.)
-
-Given that submarines almost by definition (since their hull shape doesn't change, and they must carry ballast to make up for any under-loading) are essentially always operating with the same load, but have to operate over an extremely wide variety of different speeds (a factor of four or more in variance) with severe efficiency concern at all of them, they represent a truely unique engineering question, seldom discussed in commercial applications.
-Consequently, there are relevantly few pieces of literature which address the present question directly.
-(Happily, not zero)  However, there are plenty of instances where the relevant physics and trade-offs are explained in depth to address related, but not identical questions.
-
-# The difference between nuclear and conventional propulsion
-
-Nuclear submarines vary quite remarkably from conventional submarines because of the means by which they generate their power.
-Because of the extremely high energy density in enriched uranium or plutonium, the reactors on board nuclear submarines generate an abundant supply of energy.
-Most submarine reactors are reportedly capable of generating between 25 to 50 megawatts (MW) of power, though Russian submarines have hundreds of megawatts of power available [@WNA2017].
-
-In contrast, the Collins Class Submarine's main motor is rated at less than 6MW, with designs for the future submarine appearing to be only slightly larger [@patrick2012].
-It is fair to say that in terms of maximum power ouput, nuclear submarines could have something like 5-10 times as much power at their disposal, and their **peak** output.
-
-The difference between the peak-powers of the submarines significantly understates how different the designs are, because the difference in the total amount of energy stored and available for use in a voyage or dive is vastly greater.
-A nuclear submarine might have literally millions of times more energy at its disposal, it is practically unlimited for all intents and purposes on a given voyage.
-Consequently, wheras a nuclear submarine might regularly conduct transits at or near it's peak power, a diesel electric submarine would probably transit at less than half the speed which it could manage at a sprint, (and use less than a quarter of the energy, as discussed earlier) and on patrol it might be operating at a tenth of the maximum speed, and use maybe just 1% of as much power on propulsion.
-In this situation, the amount of power that is drawn for lights, CO2 scrubbing, washing, cooking, heating, as well as the electronics driving the combat systems (the 'Hotel Load') might well become significant, and even be as larger or larger than what is required for propulsion.
-
-Consequently, the difference between the power output a conventional and a nuclear submarine might regularly operate at could be even quite a bit larger than the maximum amount of energy that they can deliver to their propulsion systems.
-
-Perhaps more significantly for power-design of nuclear submarines is a little-discussed phenomenon called Xenon poisoning, which affects nuclear reactors when they shut down or lower their power significantly.
-When Uranium or Plutonium atoms split (or fission) into two smaller isotopes atoms, a variety of different radio nucleides (unstable variants of atoms) are produced.
-Two of these are Xenon-135 and Iodine 135.
-Iodine is produced much more often, and decays into Xenon-135, which a half-life of about nine hours.
-This Xenon has a very special, perhaps unique role in reactors, since they very easily captures the free neutrons which cause the continued fission reactions in Uranium or Plutonium which drive the reactor.
-As such, Xenon is known as reactor 'poison' since it can kill the reactor's reactivity in very high doses.
-
-This high neutron-capture from Xenon means that it has a duplicitous relationship with the reactor's power level.
-At high levels of reactor power, lots of Xenon-135 and Iodine-135 are produced by the fission process.
-The high presence of Xenon reduced the reactor's reactivity.
-On the other hand, there are lots of neutrons available to 'burn off' the ass Xenon (which absobs the neturons to become a different isotope).
-This keeps the total level of Xenon in check.
-
-The situation becomes much more complicated when the reactor undergoes a sudden change in power output.
-If the power is lowered dramatically and suddenly, the production of Xenon continues quite rapidly for some time due to the decay of the large stock of Iodine-135.
-With less neutron flux available to 'burn off' the Xenon, the Xenon levels spike, and push down the reactor's reactivity.
-Unless the reactor is quickly raised back to relatively high power (~60%) quite quickly (an hour or less), the Xenon levels become so high that the reactor will have to be shut down, otherwise extreme (and dangerous) measures would be required in order to keep the reactor going.
-(This is essentially what lead to the Chernobyl Explosion [@WNA2009].)  A fuller discussion of Xenon poisoning can be found in [@garland2005], which demonstrates key concepts related to the poisoning effect shown in \ref{fig:ex2}.
-And you can see that in the [Introduction].
-
-The situation becomes much more complicated when the reactor undergoes a sudden change in power output.
-
-\begin{figure}
-\includegraphics[width=0.9\textwidth]{XenonPoison.png}
-\caption{A prototype of the Job Information dialog  }
-\label{fig:ex2}
-\end{figure}
-
-Consequently, nuclear reactors aren't well suited to rapid fluctuations in power, particularly dramatic reductions in power, as these can lead to instability in the reactor core.
-If the reactor is shut down in order to avoid such dangerous circumstances, it generally cannot be started again until Xenon levels have fallen again, which can take a couple of days.
-Obviously this is never desirable for a military vessel, and hence is avoided at almost all costs.
-
-\begin{figure}
-```{r, echo=F}
-kable(head(mtcars))
-```
-\caption{A table example}
-\label{fig:tbl1}
-\end{figure}
-
-This effect has a dramatic impact on nuclear submarine design, since much of their design is oriented around being able to comfortably disperse large amounts of excess power, rather than with conserving it.
-For the propulsion system, this actually makes having an inefficient propulsor at low speeds a considerable advantage.
-Since the reactor will likely need to dispose of excess power, particularly during ramp-down, an inefficient propulsion system actually provides a useful power sink.
-Since any excess power will have to be disposed of by some other means (normally by pumping more water to remove the power as the heat) inefficiency at low speeds has no penalty, and probably a marginal benefit, since it will reduce overall demand for additional systems.
-Provided the excess turbulence inside the pumpjet isn't too noisy, wasting energy through the propulsor is useful.
-
-It should also be noted that a nuclear reactor's aversion to sudden reductions in power would also have a substantial impact on the design of a submarine's combat system, and it's demand on the Hotel Load.
-For the same reason, a high Hotel Load, or power-hungry Combat System, could actually be advantageous, as it helps to set an elevated 'floor' for power requirements, reducing the scale of fluctuations in overall power demand from the reactor due to changes in propulsion speed.
-
-# Some essential concepts
-
-### Conservation of Energy
-
-This is perhaps one of the most fundamental and well-established principals in physics.
-The essential idea is that energy can move or change in form, but it isn't ever created or destroyed.
-Machines, plants and animals all derive their energy from a particular other source, which can be measured and evalutated to establish the limits of energy available.
-Plants collect energy from sunlight falling on their leaves.
-Humans (as well as combustion engines) capture the chemical potential energy in organic matter, and release it by combining it with oxygen.
-Hydro-electric power plants turn the gravitational potential energy of water stored at a height into electricity.
-
-Conservation of energy has a particularly relevant embodiment in fluid flows, which is given it's primary expression in Bernoullis equation.
-It says that the total energy in a connected body of fluid is constant, though it can change in form between kinetic energy (movement), gravitational potential energy (it being elevated) the heat energy in the fluid, and the pressure of the fluid.
-
-In different fluids, the dynamics of how energy moves between one and another change.
-For example, in gases, heating up a confined piece of gas will increase it's pressure, or if it is unconfined, increase its volume.
-This is particularly important for understanding gas turbines.
-In the case of a liquid, however, all the molecules are in close contact, and hence can't increase in volume or pressure substantially except by the creation of steam.
-As such, in the absence of large amounts of cavitation, the terms in the relationship which are most important for our consideration are the relationships between pressure, and kinetic energy, and gravity.
-
-In the case of most waterjets which propel surface vessels, the water is lifted from the bottom of the hull at the intake to the pump, which is generally incorporated inside the hull.
-This increase in gravitational potential coincides with a slowing down of the water relative to the vessel-speed.
-Since the jet in such vessels is generally ejected at the same height as the pump, this potential is never regained, and is technically a loss, however at high speeds such a loss is small relative to the total power ouput.
-
-#### Venturi Effect and Flow Diffusion
-In the case of submarines and torpedoes the water doesn't undertake a change in height, since intake and nozzle are generally all in line with the central axis of the submarine or torpedo.
-As a consequence, the key relationship in Bernoullis equations is the relationship between the liquids velocity, and its pressure.
-The consequence is that when water moves through a pipe (duct, or shroud) it's pressure is inversely related to the square of its velocity.
-This means that when a liquid is forced to to travel through a narrowing pipe, it's pressure necessarily decreases as the velocity increases.
-This is a simple embodiment of the Venturi effect.
-The inverse process is where a pipe increases in volume, and the flow is forced to slow down in order to fill the wider area, and the pressure correspondingly increases.
-This is a process called 'diffusion', and is important to achieving high pressure levels in many types of water pumps, including those which will be particularly relevant for pumpjets for watercraft propulsion [@hamilton1997].
-
-#### Energy waste and efficiency
-The conservation of energy also has useful implications for how the efficiency of systems is thought about.
-In particular, because energy is conserved, identifying inefficiencies in a system necessarily involves identifying where energy goes to doing tasks which aren't useful for the intended purpose.
-A perfectly efficient system won't do any work that isn't for the intended purpose.
-In the case of analysing the efficiency of propulsion systems, the relevant 'work' is almost always related to moving water backwards to produce thrust.
-Moving water in directions other than backwards, including random turbulent flows which wind producing heat rather than thust, are two examples of wasted energy.
-Noise in the water also reflects energy which is wasted.
-
-### Boundary Layer
-When water or any fluid flows with some speed relative to another solid surface nearby, there is some layer adjacent to the surface in which the speed of the fluid is diminished relative to the main flow.
-At a microscopic level, there are some molecules of the fluid on the surface which will be effectively static relative to the surface.
-The layer of fluid that joins the gap between the static surface, and the part of the flow which is moving at the full flow speed, is called the boundary layer.
-Exactly how thick the boundary layer is, and how the fluid moves in the boundary layer, is extremely important for consideration of efficiency of fluid flows over and around solid surfaces.
-
-### Turbulent and Laminar Flow
-
-### Pressure, Thrust, and Momentum
-
-### Cavitation
-
-### Flow separation
-
-# What tech are we talking about? The core relationships between pumps, jets, ducts, and propellers, and all the different things we call them, their parts, and combinations
-
-# Propulsive performance of the pumpjet, or decelerating ducted propeller
-
-## Efficiency Relationships, and various curves
-
-## Practical and recent evidence of efficiency relationships
-
-# Results of numerical simulation
-
-# Acoustic Considerations
-
-# Conclusions
-
-# References
+Note that the `echo = FALSE` parameter was added to the code chunk to prevent printing of the R code that generated the plot.

### 2018-01-02 DraftExec.Rnw
add plots

@@
 library(tidyverse)
+library(submarines)
 @
@@
 
+\section{App Outputs}
+
+Adjustable figures from App.
+
+<<compute, echo=F>>=
+
+df <- read_rds("extdata/default_subs_df.rds")
+
+df <- run_subs(df,
+             hotel = 150,
+             total.batt = 500,
+             batt.dens = 0.14,
+             patrol = 2.5,
+             max.speed = 18,
+             max.power = 7,
+             speed = 8,
+             power = 1500,
+             system = "jet",
+             method = "hotel match")
+
+eff_plot_df <- df %>%
+  filter(cat %in% c("eff.jet", "eff.prop")) %>%
+  mutate(cat = case_when(
+    cat ==  "eff.prop" ~ "Propeller",
+    cat ==  "eff.jet" ~ "Jet"
+  ))
+
+end_plot_df <- df %>%
+  filter(cat %in% c("endurance.prop.hour", "endurance.jet.hour")) %>%
+  mutate(cat = case_when(
+    cat ==  "endurance.prop.hour" ~ "Propeller",
+    cat ==  "endurance.jet.hour" ~ "Jet"
+  ))
+
+y_max_e <- roundUpNice(max(pull(end_plot_df, val)))
+
+range_plot_df <- df %>%
+  filter(cat %in% c("range.prop", "range.jet")) %>%
+  mutate(cat = case_when(
+    cat ==  "range.prop" ~ "Propeller",
+    cat ==  "range.jet" ~ "Jet"
+  ))
+
+y_max_r <- roundUpNice(max(pull(range_plot_df, val)))
+
+power_plot_df <- df %>%
+  filter(cat %in% c("hotel","power.mob.drawn.jet", "power.mob.req", "power.mob.drawn.prop")) %>%
+  mutate(cat = case_when(
+    cat == "hotel" ~ "Hotel Load",
+    cat ==  "power.mob.drawn.prop" ~ "Propulsion Power Drawn Propeller",
+    cat ==  "power.mob.drawn.jet" ~ "Propulsion Power Drawn Jet",
+    cat ==  "power.mob.req" ~ "Thrust Power Required"
+  ))
+
+y_max_p <- roundUpNice(max(pull(power_plot_df, val)))
+@
+
+\begin{figure}
+<<plot_eff, echo=FALSE>>=
+ggplot(eff_plot_df, aes(x = kts, y = val, col = cat))+
+        geom_line()+
+        theme_minimal() +
+        scale_y_continuous(breaks=seq(0, 1, 0.2),
+                           name = "Efficiency") +
+        scale_x_continuous(breaks=seq(0, 20, 2))
+@
+\caption{Efficiency}
+\label{fig:Efficiency_plot}
+\end{figure}
+
+\begin{figure}
+<<plot_end, echo=FALSE>>=
+ggplot(end_plot_df, aes(x = kts, y = val, col = cat))+
+          geom_line() +
+          theme_minimal() +
+          scale_y_continuous(breaks=seq(0, y_max_e, (y_max_e/10)),
+                             name = "hours") +
+          scale_x_continuous(breaks=seq(0, 20, 2)) +
+          geom_text(aes(label = round(val, digits = 1)), nudge_y = 2, angle = 45)
+@
+\caption{Endurance}
+\label{fig:Endurance_plot}
+\end{figure}
+
+\begin{figure}
+<<plot_range, echo=FALSE>>=
+ggplot(range_plot_df, aes(x = kts, y = val, col = cat))+
+          geom_line()+
+          theme_minimal()+
+          scale_y_continuous(breaks=seq(0, y_max_r, (y_max_r/10)),
+                             name = "nm") +
+          scale_x_continuous(breaks=seq(0, 20, 2)) +
+          geom_text(aes(label = round(val, digits = 1)), nudge_y = 2, angle = 45)
+@
+\caption{Range}
+\label{fig:Range_plot}
+\end{figure}
+
+\begin{figure}
+<<plot_power, echo=FALSE>>=
+        ggplot(power_plot_df, aes(x = kts, y = val, col = cat))+
+          geom_line() +
+          theme_minimal() +
+          scale_y_continuous(breaks=seq(0, y_max_p, (y_max_p/10)),
+                             name = "hours") +
+          scale_x_continuous(breaks=seq(0, 20, 2)) +
+          geom_text(aes(label = round(val, digits = 1)), nudge_y = 2, angle = 45)
+@
+\caption{Power}
+\label{fig:Power_plot}
+\end{figure}
+
 \section{Conclusions}

### 2018-01-02 DraftExec.Rnw
little fixes in bib etc. Still needs some work

@@
 The paper makes no effort to justify the selection of choice of jets on the efficiency of the propulsion system, instead focussing on improved acoustic performance, and potential machinery space-saving advantages.
-The the vessel under consideration would require 250 tonnes more fuel with jets find than with propellers, the jets would require more power to reach equal waterspeed to a conventionally propelled craft right up to 30 kt.
+The vessel under consideration would require 250 tonnes more fuel with jets find than with propellers, the jets would require more power to reach equal waterspeed to a conventionally propelled craft right up to 30 kt.
 The paper states: "This is an encouraging result for a first iteration as the AWJ-21™ arrangement has not been fully optimised and, hence, there is further prospect of reducing the shaft power for this form.
@@
 Moreover, recent literature seems to confirm that efficiency of pumpjets fall off precipitously towards zero the closer that the vessel speed comes to zero.
-This stronly suggests the existence of some fundamental constraint in the physics of the operation of jets, which recent advances in technology seems not to have overcome.
+This strongly suggests the existence of some fundamental constraint in the physics of the operation of jets, which recent advances in technology seems not to have overcome.
 To more firmly establish the extent to which adaptation or plausible improvments in technology could or could not change this trend, it is worthwhile considering the theoretical limits that are presented by the physics.
@@
 
-With the optimum efficiency of a jet probably being no greater than 60\% or 70\% at its peak efficiency according to many theoretical and experimental studies \parencite{lu2015}, which would quite probably lie considerably higher in terms of waterspeed, these facts, along with the inevitable decline of efficiency towards zero as waterspeed also declines, a suitably constrained set of plausible efficiency curves may be investigated numerically, as will be done later in later sections.
+With the optimum efficiency of a jet probably being no greater than 60\% or 70\% at its peak efficiency according to many theoretical and experimental studies \parencite{lu2016cfd}, which would quite probably lie considerably higher in terms of waterspeed, these facts, along with the inevitable decline of efficiency towards zero as waterspeed also declines, a suitably constrained set of plausible efficiency curves may be investigated numerically, as will be done later in later sections.
 
@@
 \begin{figure}
-\includegraphics[width=\textwidth]{DuctedEfficiency.png}
+\includegraphics[width=\textwidth]{AcceleratingIdeal.png}
 \caption{Calculations of 'ideal' efficiencies of accelerating ducts (neglecting drag) in \cite{oosterveld1970} show that a duct can only improve the efficiency of a propeller if it is loaded above a certain degree, and that longer ducts produce better improvements in 'ideal' efficiencies}
-\label{fig:DuctedEfficiency.png}
+\label{fig:AcceleratingIdeal.png}
 \end{figure}
@@
 However, when dived, it might be reasonably assumed that the complete elimination of all sources of cavitition will be the primary objective of acoustic signature management.
-This is neatly put in \cite[88]{gearhart1966}: "The occurence of cavitation in a propulsor results in adegradation of the propulsor operating characteristics if extensive amounts of cavitation exist, and in significant noise when minute amounts occur.
+This is neatly put in \cite[88]{gearhart1966selection}: "The occurence of cavitation in a propulsor results in adegradation of the propulsor operating characteristics if extensive amounts of cavitation exist, and in significant noise when minute amounts occur.
 As a result, the performance requirements of a submersible weapon system specify that the occurence or inception of cavitation be avoided below certain operating depths." One might consider that a submarines 'tactical silent speed' would most probably correspond to the speed at which it can operate safely below the point of any cavitation inception.

### 2018-01-02 DraftExec.tex
add plots
little fixes in bib etc. Still needs some work

@@
 
+\section{App Outputs}
+
+Adjustable figures from App.
+
+\begin{knitrout}
+\definecolor{shadecolor}{rgb}{0.969, 0.969, 0.969}\color{fgcolor}\begin{kframe}
+
+{\ttfamily\noindent\itshape\color{messagecolor}{\#\# in df}}\begin{verbatim}
+## # A tibble: 6 x 4
+##     kts   eff.jet  eff.prop hotel
+##   <dbl>     <dbl>     <dbl> <dbl>
+## 1   0.5 0.1728660 0.8902062   150
+## 2   1.0 0.2052115 0.8669509   150
+## 3   1.5 0.2356997 0.8463204   150
+## 4   2.0 0.2643992 0.8277603   150
+## 5   2.5 0.2913788 0.8108753   150
+## 6   3.0 0.3167071 0.7953734   150
+\end{verbatim}
+
+{\ttfamily\noindent\itshape\color{messagecolor}{\#\# speed.}}\begin{verbatim}
+## [1] 2.5
+\end{verbatim}
+
+{\ttfamily\noindent\itshape\color{messagecolor}{\#\# power.}}\begin{verbatim}
+## [1] 150
+\end{verbatim}
+
+{\ttfamily\noindent\itshape\color{messagecolor}{\#\# df.}}\begin{verbatim}
+## # A tibble: 6 x 3
+##     kts hotel       eff
+##   <dbl> <dbl>     <dbl>
+## 1   0.5   150 0.1728660
+## 2   1.0   150 0.2052115
+## 3   1.5   150 0.2356997
+## 4   2.0   150 0.2643992
+## 5   2.5   150 0.2913788
+## 6   3.0   150 0.3167071
+\end{verbatim}
+
+{\ttfamily\noindent\itshape\color{messagecolor}{\#\# index.}}\begin{verbatim}
+## [1] 5
+\end{verbatim}
+
+{\ttfamily\noindent\itshape\color{messagecolor}{\#\# eff.}}\begin{verbatim}
+##         5 
+## 0.2913788
+\end{verbatim}
+
+{\ttfamily\noindent\itshape\color{messagecolor}{\#\# const is 2.8}}\end{kframe}
+\end{knitrout}
+
+\begin{figure}
+\begin{knitrout}
+\definecolor{shadecolor}{rgb}{0.969, 0.969, 0.969}\color{fgcolor}
+
+{\centering \includegraphics[width=\maxwidth]{figures/plots-plot_eff-1}
+
+}
+
+\end{knitrout}
+\caption{Efficiency}
+\label{fig:Efficiency_plot}
+\end{figure}
+
+\begin{figure}
+\begin{knitrout}
+\definecolor{shadecolor}{rgb}{0.969, 0.969, 0.969}\color{fgcolor}
+
+{\centering \includegraphics[width=\maxwidth]{figures/plots-plot_end-1}
+
+}
+
+\end{knitrout}
+\caption{Endurance}
+\label{fig:Endurance_plot}
+\end{figure}
+
+\begin{figure}
+\begin{knitrout}
+\definecolor{shadecolor}{rgb}{0.969, 0.969, 0.969}\color{fgcolor}
+
+{\centering \includegraphics[width=\maxwidth]{figures/plots-plot_range-1}
+
+}
+
+\end{knitrout}
+\caption{Range}
+\label{fig:Range_plot}
+\end{figure}
+
+\begin{figure}
+\begin{knitrout}
+\definecolor{shadecolor}{rgb}{0.969, 0.969, 0.969}\color{fgcolor}
+
+{\centering \includegraphics[width=\maxwidth]{figures/plots-plot_power-1}
+
+}
+
+\end{knitrout}
+\caption{Power}
+\label{fig:Power_plot}
+\end{figure}
+
 \section{Conclusions}

### 2018-01-02 scripts/build df for paper.R
working on plots

@@ -1,0 +1,58 @@
+
+df <- read_rds("extdata/default_subs_df.rds")
+
+x <- list.files("extdata", pattern = "_input", full.names = T)
+
+# 1514864275.63078 %>% lubridate::as_datetime()
+# 1514865059.98424 %>% lubridate::as_datetime()
+# 1514865327.90148 %>% lubridate::as_datetime()
+# prop
+n <-  6
+x <- list.files("extdata", pattern = "_input_", full.names = T)  %>%
+  map2(letters[1:n], ~ read_rds(.x) %>%
+         as_tibble() %>%
+         mutate(id = .y)) %>%
+  bind_rows()
+
+x <- x %>%
+  mutate(kts = rep(df$kts, n))
+
+ggplot(x, aes(x = kts, y= value, col = id)) +
+  geom_line() +
+  ylim(0, 1)
+
+
+
+a <- lm(x %>% filter(id == "a") %>% pull(value) ~ poly(pull(df, kts), 4))$fitted.values
+b <- lm(x %>% filter(id == "b") %>% pull(value) ~ poly(pull(df, kts), 4))$fitted.values
+c <- lm(x %>% filter(id == "c") %>% pull(value) ~ poly(pull(df, kts), 3))$fitted.values
+e <- lm(x %>% filter(id == "d") %>% pull(value) ~ poly(pull(df, kts), 3))$fitted.values
+d <- lm(x %>% filter(id == "e") %>% pull(value) ~ poly(pull(df, kts), 4))$fitted.values
+f <- lm(x %>% filter(id == "f") %>% pull(value) ~ poly(pull(df, kts), 3))$fitted.values
+
+
+
+y <- x %>%
+  mutate(eff = c(a,
+                 b,
+                 c,
+                 d,
+                 e,
+                 f
+  )) %>%
+  mutate(System = case_when(
+    id == "a" ~ "Jet 1",
+    id == "b" ~ "Jet 2",
+    id == "c" ~ "Jet 3",
+    id == "d" ~ "Prop 1",
+    id == "e" ~ "Prop 2",
+    id == "f" ~ "Prop 3"
+  ))
+
+ggplot(y, aes(x = kts, y= eff, col = System)) +
+  geom_line() +
+  ylim(0, 1)
+
+plotly::ggplotly()
+
+y %>% write_rds("extdata/comb_eff_df.rds")

### 2018-01-02 scripts/build plots for paper.R

@@ -1,0 +1,175 @@
+
+library(stringr)
+
+df <- read_rds("extdata/comb_eff_df.rds")
+
+ # df <- read_rds("extdata/default_subs_df.rds")
+
+
+ggplot(df, aes(x = kts, y = eff, col = System)) +
+  geom_line() +
+  ylim(0, 1)
+
+
+df_1_j <- df %>%
+  filter(System == "Jet 1") %>%
+  rename(eff.jet = eff)
+df_1_p <- df %>%
+  filter(System == "Prop 1")
+df_1 <-  df_1_j %>%
+  mutate(eff.prop = pull(df_1_p, eff))
+
+df_2_j <- df %>%
+  filter(System == "Jet 2") %>%
+  rename(eff.jet = eff)
+df_2_p <- df %>%
+  filter(System == "Prop 2")
+df_2 <-  df_2_j %>%
+  mutate(eff.prop = pull(df_2_p, eff))
+
+df_3_j <- df %>%
+  filter(System == "Jet 3") %>%
+  rename(eff.jet = eff)
+df_3_p <- df %>%
+  filter(System == "Prop 3")
+df_3 <-  df_3_j %>%
+  mutate(eff.prop = pull(df_3_p, eff))
+
+
+ls <- list(df_1, df_2, df_3) %>%
+  map( ~ .x %>%
+         select(kts, eff.jet, eff.prop) %>%
+         run_subs(
+                  hotel = 150,
+                  total.batt = 500,
+                  batt.dens = 0.14,
+                  patrol = 2.5,
+                  max.speed = 18,
+                  max.power = 7,
+                  speed = 8,
+                  power = 1500,
+                  system = "jet",
+                  method = "hotel match")) %>%
+  set_names(c("Set 1", "Set 2", "Set 3"))
+
+
+
+
+produce_all_plots <- function(df) {
+  eff_plot_df <- df %>%
+    filter(cat %in% c("eff.jet", "eff.prop")) %>%
+    mutate(cat = case_when(
+      cat ==  "eff.prop" ~ "Propeller",
+      cat ==  "eff.jet" ~ "Jet"
+    ))
+
+
+  end_plot_df <- df %>%
+    filter(cat %in% c("endurance.prop.hour", "endurance.jet.hour")) %>%
+    mutate(cat = case_when(
+      cat ==  "endurance.prop.hour" ~ "Propeller",
+      cat ==  "endurance.jet.hour" ~ "Jet"
+    ))
+
[…]

### 2018-01-02 scripts/Inception.R
tiny change?

@@ -1,0 +1,31 @@
+library(dev)
+library(plotly)
+
+
+speed <- rep(seq(0.5,20,0.5),90)
+VP. <- c(rep(1.228*1000,40), rep(2.339*1000,40), rep(4.245*1000,40))
+VP <- rep(VP., 30)
+temp. <- c(rep(10,40), rep(20,40), rep(30,40))
+temp <- rep(temp., 30)
+depth <- c(rep(10,120), rep(20,120), rep(30,120), rep(40,120), rep(50,120), rep(60,120), rep(70,120), rep(80,120), rep(90,120), rep(100,120),
+           rep(110,120), rep(120,120), rep(130,120), rep(140,120), rep(150,120), rep(160,120), rep(170,120), rep(180,120), rep(190,120), rep(200,120),
+           rep(210,120), rep(220,120), rep(230,120), rep(240,120), rep(250,120), rep(260,120), rep(270,120), rep(280,120), rep(290,120), rep(300,120))
+
+df<- tibble(speed, depth, temp, VP)
+df2 <- df %>%
+  mutate(thrust = (speed^2)*35/40*1000,
+         pressure = (100 + depth*10)*1000,
+         disc.pressure = thrust/5,
+         peak.pressure.drop = disc.pressure*10,
+         min.stat.pressure = pressure - peak.pressure.drop,
+         min.stat.rel.VP = (min.stat.pressure - VP)/100000)
+
+plot <- df2 %>%
+  filter(temp == 20 & depth <= 100) %>%
+  ggplot(aes(x = speed, y = depth, fill = min.stat.rel.VP))+
+  geom_tile()+
+  scale_fill_gradientn(colours = rainbow(5))+
+  stat_contour(aes(z = min.stat.rel.VP), col = "black")+
+  stat_contour(data = df2 %>% filter(temp == 30 & depth <= 100), aes(z = min.stat.rel.VP), col = "red", linetype = 2)
+
+ggplotly(plot)

### 2018-01-03 scripts/Inception.R
graph finished

@@ -25,7 +25,11 @@ plot <- df2 %>%
   ggplot(aes(x = speed, y = depth, fill = min.stat.rel.VP))+
   geom_tile()+
   scale_fill_gradientn(colours = rainbow(5))+
+  scale_x_continuous(name ="speed (kt)")+
+  scale_y_continuous(name ="depth (m)")+
+  labs(title = "Cavitation Inception Equivalence Contours", fill = "Indicative \nExcess \nStatic Pressure \n(Atmospheres)")+
   stat_contour(aes(z = min.stat.rel.VP), col = "black")+
   stat_contour(data = df2 %>% filter(temp == 30 & depth <= 100), aes(z = min.stat.rel.VP), col = "red", linetype = 2)
+
 
 ggplotly(plot)

### 2018-01-03 scripts/build plots for paper.R
simplified and added proportional advantage for endurance plot

@@ -1,46 +1,43 @@
 
-library(stringr)
+library(submarines)
 
-df <- read_rds("extdata/comb_eff_df.rds")
+df <- read_csv(find::this("/Submarines/eff-data-2018-01-03.csv"))
 
  # df <- read_rds("extdata/default_subs_df.rds")
 
 
-ggplot(df, aes(x = kts, y = eff, col = System)) +
+ggplot(df, aes(x = kts, y = val, col = cat)) +
   geom_line() +
   ylim(0, 1)
 
+plotly::ggplotly()
 
-df_1_j <- df %>%
-  filter(System == "Jet 1") %>%
-  rename(eff.jet = eff)
-df_1_p <- df %>%
-  filter(System == "Prop 1")
-df_1 <-  df_1_j %>%
-  mutate(eff.prop = pull(df_1_p, eff))
+df <- df %>%
+  mutate(set = case_when(
+    cat %in% c("input_prop_1", "input_jet_1") ~ "Middle",
+    cat %in% c("input_prop_2", "input_jet_3") ~ "Close",
+    cat %in% c("input_prop_3", "input_jet_2") ~ "Far"
+  ))
 
-df_2_j <- df %>%
-  filter(System == "Jet 2") %>%
-  rename(eff.jet = eff)
-df_2_p <- df %>%
-  filter(System == "Prop 2")
-df_2 <-  df_2_j %>%
-  mutate(eff.prop = pull(df_2_p, eff))
-
-df_3_j <- df %>%
-  filter(System == "Jet 3") %>%
-  rename(eff.jet = eff)
-df_3_p <- df %>%
-  filter(System == "Prop 3")
-df_3 <-  df_3_j %>%
-  mutate(eff.prop = pull(df_3_p, eff))
+ls <- df %>%
+  split(.$set) %>%
+  map( ~ .x %>%
+         select(-set) %>%
+         mutate(cat = case_when(
+           str_detect(cat, "jet") ~ "eff.jet",
+           str_detect(cat, "prop") ~ "eff.prop"
+         )) %>%
+         spread(cat, val))
 
 
-ls <- list(df_1, df_2, df_3) %>%
+
+
+
+
+ls_comp <- ls %>%
   map( ~ .x %>%
          select(kts, eff.jet, eff.prop) %>%
-         run_subs(
-                  hotel = 150,
+         run_subs(hotel = 150,
                   total.batt = 500,
                   batt.dens = 0.14,
[…]

### 2018-01-04 DraftExec.Rnw
lots of work thurs

@@
 \usepackage[backend=bibtex, sorting=none, citestyle=authoryear]{biblatex}
-
+\usepackage[margin=0.8in]{geometry}
 \bibliography{references}
+
+\usepackage{draftwatermark}
+\SetWatermarkText{DRAFT}
+\SetWatermarkScale{5}
+\SetWatermarkLightness{0.95}
 
@@
 @
+
+Para
 
@@
 \begin{figure}
-\includegraphics[width=\textwidth]{SeparatedAerofoil.jpg}
+\includegraphics[width=0.5\textwidth]{SeparatedAerofoil.jpg}
 \caption{Flow separation occurring over the top of an aerofoil in a wind tunnel. Image courtesy of Deutsches Zentrum fuer Luft- und Raumfahrt e. V. (DLR)}
@@
 The vessel under consideration would require 250 tonnes more fuel with jets find than with propellers, the jets would require more power to reach equal waterspeed to a conventionally propelled craft right up to 30 kt.
+In the case of this analysis, modifications to the hullform in order to accommodate the jet (which are likely to be more substantial for a surface ship than a submarine) also have an impcat on the resistance offerd by the hull.
+This also has an interaction with the different shape of the propulsors, which alters the required level of delivered (effective) power to achieve a given speed, in addition to the considerations of how efficiently the propulsor generates that effective power.
 The paper states: "This is an encouraging result for a first iteration as the AWJ-21™ arrangement has not been fully optimised and, hence, there is further prospect of reducing the shaft power for this form.
 If the power could be reduced by 5\% (e.g.
-by fur ther improving the waterjet/hull fairing), then the AWJ-21™ vessel would be better at all speeds above 25 knots." \parencite{giles2010}
-
-\begin{figure}
-\includegraphics[width=\textwidth]{BMTPowering.png}
-\caption{Studies on submerged waterjets for military surface craft currently show that jets require more power than conventional propellers right up to 30kt \parencite{giles2010}}
-\label{fig:BMTPowering.png}
-\end{figure}
-
-In paper also provides specific curves for the efficiency of the propulsor as a function of vessel speed, which demonstrated that the waterjet efficiency was lower than that of conventional propellers throughout the speed range, but declined particularly precipitously below speeds of 10kt.
+by fur ther improving the waterjet/hull fairing), then the AWJ-21™ vessel would be better at all speeds above 25 knots."  This statement, however, is made in reference to the hullform needed to accommodate the jet, rather than the jet itself.
+\parencite{giles2010}
+
+The paper also provides specific curves for the efficiency of the propulsor as a function of vessel speed, which demonstrated that the waterjet efficiency was lower than that of conventional propellers throughout the speed range, but declined particularly precipitously below speeds of 10kt.
 At 5kt, the propeller was more than twice as efficient as the waterjet.
@@
 \begin{figure}
-\includegraphics[width=\textwidth]{LowIVR.png}
+\includegraphics[width=0.5\textwidth]{LowIVR.png}
 \caption{A an intake may experience risks of flow separation even at low speeds (Inlet Velocity Ratio) on some surfaces due to certain flow deflections being more severe.
@@
 
-\section{Results of numerical simulation}
+\section{Model Structures and Assumptions}
+
+\subsection{Mathematical Structure}
+
+The core of the model rests upon the assumption that drag increases with the square of the velocity of the submarine.
+This is a widely used assumption for fully submerged objects subject to turbulent flows \parencite[5]{wislicenus1973}, \parencite{davies2017}, as also discussed in Section \ref{Speed and Drag - Why very slow is very very $(very)^2$ economical}.
+Given that the amount of work done (energy consumed) by a system is equal to the distance over which it acts, which scales with speed also, we assume that the Effective Power $P_E$ (also known as Towing Power) required for propulsion to meet a given speed $v$ can be given by Equation \ref{eq:towpower}, where $C_d$ represents a drag constant.
+
+\begin{equation}
+\label{eq:towpower}
+P_E = C_dv^3
+\end{equation}
+
+The power drawn by the propulsion $P_D$ is assumed simply to be the propulsive efficiency at any given speed $\eta_v$ multiplied by the Effective Power, as given in Equation \ref{eq:powerdrawn}.
+Using a plausible efficiency assumption and total power required, Equation \ref{eq:powerdrawn} can be used to solve for the constant $C_d$.
+Other efficiency factors, such as the efficiency of electric motors or hull efficiency factors are assumed to be either relatively small, or essentially similar between submarines which might be conventionally powered but with a jet or propeller being the main point of difference.
+Hence the total power consumed by the submarine is given by Equation \ref{eq:totalpower}.
+
+\begin{equation}
+\label{eq:powerdrawn}
+P_D = \eta_vP_E
+\end{equation}
+
+\begin{equation}
+\label{eq:totalpower}
+P_T = P_D + H
+\end{equation}
+
+With this relationship defined across the speed range, it is elementary to calculate the time that a given speed could be sustained based on a finite energy store (endurance) and the distance covered in this time (range).
+This provides the fundamental structure that underlies the model which is used to assess the dived range and enurance of conventionally powered submarines, without using air independent propulsion (AIP).
+
+\subsection{Efficiency Curve Assumptions}
+
+In order to begin to model the likely impact of propulsion system choices on system peformance characteristics such as dived range and endurance, it is necessary to adopt some plausible curves indicating likely levels of efficiency across the operating speed range, in this case assumed to be up to 20kt.
+
+In this case, three possible curves have been selected for each system, representing a high, low, and central assumption for each, with peak efficiency for the propeller at about 65\% at near 5kt, and jet efficiency around 27\% at the sames speed.
+The central assumption has been taken in both cases from the study undertaken by BMT and Rolls Royce, as in \cite{giles2010}, and indicated in \ref{fig:BMTShape.png}, since this represents the most direct comparison made as a simple function of waterspeed, with comparable technologies.
+
+Variations from this central line have been made with consideration for some plausible trade-offs, which are intended to be indicative, but plausible.
+In the case of the high pumpjet assumption, it is assume that a substantially larger intake area has been chosen in order to achieve higher efficiencies at lower speeds, at the cost of peak ideal efficiency, which might be realised not too far above 20kt for such a design.
+The low assumption for the pumpjet is still represents higher efficiencies at equivalent speeds than those given in \cite{fujisawa1995} for a waterjet.
+It is assumed that for both the central and low pumpjet curves, peak efficiency would not be achieved until a speed considerably above 20kt.
+All of the efficiency curves reflect a smaller decline in efficiency than the indicative assumption used by Andrew Davies, namely that efficiency might reduce by a factor of four between transit and patrol speeds, if those speeds also differed by a factor of four \cite{davies2017}.
+
+The higher assumption curve used for the propeller approaches a maximum of just under 72\%.
+This performance would not be implausible if a specialised propeller was adopted specificially to deliver peak performance at around this speed.
+Advanced flexible composite propellers, such as those used by the German submarine builder TKMS \parencite{durrant2016}, are able to achieve very high efficiencies, as well as delayed onset of cavitation, by allowing the blades to deform in certain ways to adapt to their specific degree of loading \parencite{young2007hydroelastic2}.
+Such propellers have been observed in experiment to have propulsive efficiencies above 70\% \parencite{young2007hydroelastic}.
+In this case I have assumed that this design necessitates the trade-off of steadily declining efficiencies over 10kt, considerably below the speed for which most open propellers are optimised.
+Given the ability for composite propellers to adapt their shape as loads increase, it is actually likely that the performance curve in this case is actually far broader.
+However, assuming a trade-off in this case also facilitates a scenario whereby the efficiencies of a propeller and jet actually cross over at around 18kt, which is relevant to consider, since experts in discussion have ruled out the possibility that a jet could be more efficient at lower speeds than this.
+
+The lower propeller curve shows efficiency still increasing up to around 65\% at 20kt, which might be consistent with a propeller optimised for conventional ship speeds, but probably not likely for a specialised submarine propeller.
+For all propeller curves, sharply declining efficiency has been assumed at very low speeds, (2kt or less), in order to reflect the dip seen below 5kt in the efficiency curve seen in Figure \ref{fig:BMTEfficiency.png}, assuming that this is driven by the decline in efficiency shown in Figure \ref{fig:PropDuct.png} at advance ratios approaching 1.
+This has been exaccerbated beyond realistic levels in the case of the the low propeller assumption line for the sake of testing sensitivity more dramatically to some losses in this regime.
+These efficiency curve assumptions are shown in Figure \ref{fig:Efficiency_assumptions}.
+
+\begin{figure}
+<<plot_efficiency, echo=FALSE, warning=FALSE, message=F >>=
+df <- suppressMessages(read_csv(find::this("/Submarines/eff-data-2018-01-03.csv")))
+
+ # df <- read_rds("extdata/default_subs_df.rds")
+df <- df %>%
+  mutate(system = case_when(str_detect(cat, "jet") ~ "Pumpjet",
+                   str_detect(cat, "prop") ~ "Propeller"),
+         level = case_when(str_detect(cat, "1") ~ "Central",
+                            str_detect(cat, "2") ~ "High",
+                            str_detect(cat, "3") ~ "Low"))
+
+ggplot(df, aes(x = kts, y = val, col = level, linetype = system)) +
+  geom_line(size = 1) +
+  scale_color_manual(values = c("black", "blue", "red") )+
+  ylim(0, 1)+
+  labs(title = "Efficiency Curve Assumptions")+
+  scale_x_continuous(name ="speed (kt)")+
+  scale_y_continuous(name ="efficiency")
+@
+\caption{A range of possible efficiency curves for pumpjets and propellers to be modeled}
+\label{fig:Efficiency_assumptions}
+\end{figure}
+
+\subsection{Hotel Load Assumptions}
+It is also required that some assumptions are made about the hotel load, or the power which the submarine requires which doesn't go to propulsion.
+We might assume that this is in the order of 100kw, as suggested in \cite{martinez1995} though official sources of information are few.
+It is also likely that the actual hotel load varies considerably as different systems are switched on and off, some of which might be discretionary in combat scenarios (e.g.
+hot water for washing), and others (such as the combat systems) might have increased demand during crucial phases of combat operations.
+It is important to note that the combat system that has been selected for Australias FSM, the AN/BYG-1, is designed for a nuclear submarine, and is reportedly consumes considerably more power than other systems \cite{muir2015}, apparently even in the order of 100kW more.
+\cite{patrick2011}.
+Consequently, the actual hotel load experienced might vary considerably, and as such the model used in the app allows the hotel load assumption to be varied, in order to test sensitivity to this parameter.
+A flat 100kW hotel load is used as a starting assumption in the initial calculations in this paper, which as seen, produces realistic results.
+
+\subsection{Battery Assumptions}
+Since the purpose of the task at hand is to assess dived endurance only, this model focuses on a single energy store simulated by a total embarked mass (battery bank) and an energy density of those batteries.
+Since the SEA 1000 project has elected not to use lithium electric batteries, the energy density of the batteries can be assumed to be that of lead-acid batteries, which are known to be in the region of 30-50Wh/kg (Watt-hours per kilogram).
+Taking a central figure and converting this to Megajoules per kilogram (MJ/kg) we assume 0.14MJ/kg to be the effective energy density of lead acid batteries for our test submarine's purposes.
+This parameter can be adjusted easily for sensitivity testing.
+
+It is worthwhile considering that alternative types of battery may be on offer for submarines.
+In particular, lithium ion batteries were offered by TKMS in their bid for SEA 1000, and are being installed in Type 212 submarines operated by the German navy, as well as contracted for the new Norwegian submarines, as well as Israeli submarines.
+Precise energy density of these batteries is not publically available.
+For domestic uses such as handheld electronics or solar power applications, lithium ion batteries may have 3-4 times the energy density of lead-acid batteries.
+However, the actual useful density of lithium ion batteries can vary widely depending on different chemistries and designs.
+It might be assumed that for very large battery banks (such as submarines) where the stability of the battery bank is of critical importance, maximum density migth be sacrificed in favour of additional safety systems, such as increased monitoring, as well as spacing and insulation barriers between cells, in order to prevent thermal runaway in one cell leading to a catastrophic fire.
 
@@
 
+This relationship can be expressed in a general form in \ref{fig:InceptionContours}, which has been generated using an approxiate numeric model.
+As can be seen here, at very low speeds, modest changes in depths can achieve considerable additional surplus pressure in order to prevent cavitation.
+However at much higher speeds, sine the pressure diffential scales with the square of the velocity travelled, far greater additional depths are required to achieve sufficient excess static to supress cavitation for a give further speed increase.
+
+\begin{figure}
+<<plot_inception, echo=FALSE, warning=F, message=F>>=
+speed <- rep(seq(0.5,20,0.5),90)
+VP. <- c(rep(1.228*1000,40), rep(2.339*1000,40), rep(4.245*1000,40))
+VP <- rep(VP., 30)
+temp. <- c(rep(10,40), rep(20,40), rep(30,40))
+temp <- rep(temp., 30)
+depth <- c(rep(10,120), rep(20,120), rep(30,120), rep(40,120), rep(50,120), rep(60,120), rep(70,120), rep(80,120), rep(90,120), rep(100,120),
+           rep(110,120), rep(120,120), rep(130,120), rep(140,120), rep(150,120), rep(160,120), rep(170,120), rep(180,120), rep(190,120), rep(200,120),
+           rep(210,120), rep(220,120), rep(230,120), rep(240,120), rep(250,120), rep(260,120), rep(270,120), rep(280,120), rep(290,120), rep(300,120))
+
+df<- tibble(speed, depth, temp, VP)
+df2 <- df %>%
+  mutate(thrust = (speed^2)*35/40*1000,
+         pressure = (100 + depth*10)*1000,
+         disc.pressure = thrust/5,
+         peak.pressure.drop = disc.pressure*10,
+         min.stat.pressure = pressure - peak.pressure.drop,
+         min.stat.rel.VP = (min.stat.pressure - VP)/100000)
+
+df2 %>%
+  filter(temp == 20 & depth <= 100) %>%
+  ggplot(aes(x = speed, y = depth, fill = min.stat.rel.VP))+
+  geom_tile()+
+  scale_fill_gradientn(colours = rainbow(5))+
+  scale_x_continuous(name ="speed (kt)")+
+  scale_y_continuous(name ="depth (m)")+
+  labs(title = "Cavitation Inception Equivalence Contours", fill = "Indicative \nExcess \nStatic Pressure \n(Atmospheres)")+
+  stat_contour(aes(z = min.stat.rel.VP), col = "black")+
+  stat_contour(data = df2 %>% filter(temp == 30 & depth <= 100), aes(z = min.stat.rel.VP), col = "red", linetype = 2)
+
+@
+\caption{Contours showing plausible lines of plausible equivalent cavitation inception}
+\label{fig:InceptionContours}
+\end{figure}
+
 It is also important to understand that the addition of a duct will necessarily to flow separation occuring on the inside of the duct opening at very low speeds which correspond to high advance ratios when load remains constant.
@@
 
-Whilst at very low power levels, or at good depths, it is quite plausible that this doesn't result in any cavitation, which would normally be a likely occurrence if the blades are operating closer to their design power.
-As such,
+Whilst at very low power levels, or at good depths, it is quite plausible that this doesn't result in any cavitation, which would normally be a likely occurrence if the blades are operating closer to their design power and such such vortices were shed into the intake.
 
@@
 
-<<compute, echo=F>>=
+<<compute, echo=F, message=F, warning=FALSE>>=
 
@@
 \begin{figure}
-<<plot_eff, echo=FALSE>>=
-ggplot(eff_plot_df, aes(x = kts, y = val, col = cat))+
-        geom_line()+
-        theme_minimal() +
-        scale_y_continuous(breaks=seq(0, 1, 0.2),
-                           name = "Efficiency") +
-        scale_x_continuous(breaks=seq(0, 20, 2))
-@
-\caption{Efficiency}
-\label{fig:Efficiency_plot}
-\end{figure}
-
-\begin{figure}
 <<plot_end, echo=FALSE>>=
@@
 <<plot_power, echo=FALSE>>=
-        ggplot(power_plot_df, aes(x = kts, y = val, col = cat))+
-          geom_line() +
-          theme_minimal() +
-          scale_y_continuous(breaks=seq(0, y_max_p, (y_max_p/10)),
-                             name = "hours") +
-          scale_x_continuous(breaks=seq(0, 20, 2)) +
-          geom_text(aes(label = round(val, digits = 1)), nudge_y = 2, angle = 45)
+read_rds("extdata/rng_ploy_hotlx.rds")
 @
@@
 
-% \printbibliography
+\printbibliography
 

### 2018-01-04 DraftExec.tex
lots of work thurs

@@
 \maketitle
+
+Para
 
@@
 
-\section{Results of numerical simulation}
+\section{Model Structures and Assumptions}
+
+\subsection{Mathematical Structure}
+
+The core of the model rests upon the assumption that drag increases with the square of the velocity of the submarine.
+This is a widely used assumption for fully submerged objects subject to turbulent flows \parencite[5]{wislicenus1973}, \parencite{davies2017}, as also discussed in Section \ref{Speed and Drag - Why very slow is very very $(very)^2$ economical}.
+Given that the amount of work done (energy consumed) by a system is equal to the distance over which it acts, which scales with speed also, we assume that the Effective Power $P_E$ (also known as Towing Power) required for propulsion to meet a given speed $v$ can be given by Equation \ref{eq:towpower}, where $C_d$ represents a drag constant.
+
+\begin{equation}
+\label{eq:towpower}
+P_E = C_dv^3
+\end{equation}
+
+The power demanded by the propulsion is assumed simply to be the propulsive efficiency at any given speed $\eta_v$ multiplied by the Effective Power.
+Other efficiency factors, such as the efficiency of electric motors or hull efficiency factors are assumed to be either relatively small, or essentially similar between submarines which might be conventionally powered but with a jet or propeller being the main point of difference.
+Hence the total power consumed by the submarine is given by Equation \ref{eq:totalpower}.
+
+\begin{equation}
+\label{eq:totalpower}
+P_T = \eta_vC_dv^3 + H
+\end{equation}
+
+\subsection{Efficiency Curve Assumptions}
+
+In order to begin to model the likely impact of propulsion system choices on system peformance characteristics such as dived range and endurance, it is necessary to adopt some plausible curves indicating likely levels of efficiency across the operating speed range, in this case assumed to be up to 20kt.
+
+In this case, three possible curves have been selected for each system, representing a high, low, and central assumption for each, with peak efficiency for the propeller at about 65\% at near 5kt, and jet efficiency around 27\% at the sames speed.
+The central assumption has been taken in both cases from the study undertaken by BMT and Rolls Royce, as in \cite{giles2010}, and indicated in \ref{fig:BMTShape.png}, since this represents the most direct comparison made as a simple function of waterspeed, with comparable technologies.
+
+Variations from this central line have been made with consideration for some plausible trade-offs, which are intended to be indicative, but plausible.
+In the case of the high pumpjet assumption, it is assume that a substantially larger intake area has been chosen in order to achieve higher efficiencies at lower speeds, at the cost of peak ideal efficiency, which might be realised not too far above 20kt for such a design.
+The low assumption for the pumpjet is still represents higher efficiencies at equivalent speeds than those given in \cite{fujisawa1995} for a waterjet.
+It is assumed that for both the central and low pumpjet curves, peak efficiency would not be achieved until a speed considerably above 20kt.
+All of the efficiency curves reflect a smaller decline in efficiency than the indicative assumption used by Andrew Davies, namely that efficiency might reduce by a factor of four between transit and patrol speeds, if those speeds also differed by a factor of four \cite{davies2017}.
+
+The higher assumption curve used for the propeller approaches a maximum of just under 72\%.
+This performance would not be implausible if a specialised propeller was adopted specificially to deliver peak performance at around this speed.
+Advanced flexible composite propellers, such as those used by the German submarine builder TKMS \parencite{durrant2016}, are able to achieve very high efficiencies, as well as delayed onset of cavitation, by allowing the blades to deform in certain ways to adapt to their specific degree of loading \parencite{young2007hydroelastic2}.
+Such propellers have been observed in experiment to have propulsive efficiencies above 70\% \parencite{young2007hydroelastic}.
+In this case I have assumed that this design necessitates the trade-off of steadily declining efficiencies over 10kt, considerably below the speed for which most open propellers are optimised.
+Given the ability for composite propellers to adapt their shape as loads increase, it is actually likely that the performance curve in this case is actually far broader.
+However, assuming a trade-off in this case also facilitates a scenario whereby the efficiencies of a propeller and jet actually cross over at around 18kt, which is relevant to consider, since experts in discussion have ruled out the possibility that a jet could be more efficient at lower speeds than this.
+
+The lower propeller curve shows efficiency still increasing up to around 65\% at 20kt, which might be consistent with a propeller optimised for conventional ship speeds, but probably not likely for a specialised submarine propeller.
+For all propeller curves, sharply declining efficiency has been assumed at very low speeds, (2kt or less), in order to reflect the dip seen below 5kt in the efficiency curve seen in Figure \ref{fig:BMTEfficiency.png}, assuming that this is driven by the decline in efficiency shown in Figure \ref{fig:PropDuct.png} at advance ratios approaching 1.
+This has been exaccerbated beyond realistic levels in the case of the the low propeller assumption line for the sake of testing sensitivity more dramatically to some losses in this regime.
+These efficiency curve assumptions are shown in Figure \ref{fig:Efficiency_assumptions}.
+
+\begin{figure}
+\begin{knitrout}
+\definecolor{shadecolor}{rgb}{0.969, 0.969, 0.969}\color{fgcolor}
+
+{\centering \includegraphics[width=\maxwidth]{figures/plots-plot_efficiency-1}
+
+}
+
+\end{knitrout}
+\caption{A range of possible efficiency curves for pumpjets and propellers to be modeled}
+\label{fig:Efficiency_assumptions}
+\end{figure}
+
+It is also required that some assumptions are made about the hotel load, or the power which the submarine requires which doesn't go to propulsion.
+We might assume that this is in the order of 100kw, as suggested in \cite{martinez1995} though official sources of information are few.
+It is also likely that the actual hotel load varies considerably as different systems are switched on and off, some of which might be discretionary in combat scenarios (e.g.
+hot water for washing), and others (such as the combat systems) might have increased demand during crucial phases of combat operations.
+It is important to note that the combat system that has been selected for Australias FSM, the AN/BYG-1, is designed for a nuclear submarine, and is reportedly consumes considerably more power than other systems \cite{muir2015}, apparently even in the order of 100kW more.
+\cite{patrick2011}.
+Consequently, the actual hotel load experienced might vary considerably, and as such the model used in the app allows the hotel load assumption to be varied, in order to test sensitivity to this parameter.
+100kW hotel load is used as a starting assumption in the initial calculations in this paper.
 
@@
 
+This relationship can be expressed in a general form in \ref{fig:InceptionContours}, which has been generated using an approxiate numeric model.
+As can be seen here, at very low speeds, modest changes in depths can achieve considerable additional surplus pressure in order to prevent cavitation.
+However at much higher speeds, sine the pressure diffential scales with the square of the velocity travelled, far greater additional depths are required to achieve sufficient excess static to supress cavitation for a give further speed increase.
+
+\begin{figure}
+\begin{knitrout}
+\definecolor{shadecolor}{rgb}{0.969, 0.969, 0.969}\color{fgcolor}
+
+{\centering \includegraphics[width=\maxwidth]{figures/plots-plot_inception-1}
+
+}
+
+\end{knitrout}
+\caption{Contours showing plausible lines of plausible equivalent cavitation inception}
+\label{fig:InceptionContours}
+\end{figure}
+
 It is also important to understand that the addition of a duct will necessarily to flow separation occuring on the inside of the duct opening at very low speeds which correspond to high advance ratios when load remains constant.
@@
 \definecolor{shadecolor}{rgb}{0.969, 0.969, 0.969}\color{fgcolor}\begin{kframe}
-
-{\ttfamily\noindent\itshape\color{messagecolor}{\#\# in df}}\begin{verbatim}
+\begin{verbatim}
 ## # A tibble: 6 x 4
@@
 ## 6   3.0 0.3167071 0.7953734   150
-\end{verbatim}
-
-{\ttfamily\noindent\itshape\color{messagecolor}{\#\# speed.}}\begin{verbatim}
 ## [1] 2.5
-\end{verbatim}
-
-{\ttfamily\noindent\itshape\color{messagecolor}{\#\# power.}}\begin{verbatim}
 ## [1] 150
-\end{verbatim}
-
-{\ttfamily\noindent\itshape\color{messagecolor}{\#\# df.}}\begin{verbatim}
 ## # A tibble: 6 x 3
@@
 ## 6   3.0   150 0.3167071
-\end{verbatim}
-
-{\ttfamily\noindent\itshape\color{messagecolor}{\#\# index.}}\begin{verbatim}
 ## [1] 5
-\end{verbatim}
-
-{\ttfamily\noindent\itshape\color{messagecolor}{\#\# eff.}}\begin{verbatim}
 ##         5 
@@
 \end{verbatim}
-
-{\ttfamily\noindent\itshape\color{messagecolor}{\#\# const is 2.8}}\end{kframe}
+\end{kframe}
 \end{knitrout}
-
-\begin{figure}
-\begin{knitrout}
-\definecolor{shadecolor}{rgb}{0.969, 0.969, 0.969}\color{fgcolor}
-
-{\centering \includegraphics[width=\maxwidth]{figures/plots-plot_eff-1}
-
-}
-
-\end{knitrout}
-\caption{Efficiency}
-\label{fig:Efficiency_plot}
-\end{figure}
 

### 2018-01-04 scripts/build plots for paper.R

@@ -5,18 +5,31 @@ df <- read_csv(find::this("/Submarines/eff-data-2018-01-03.csv"))
 
  # df <- read_rds("extdata/default_subs_df.rds")
 
+df_temp <- df %>%
+  mutate(system = case_when(str_detect(cat, "jet") ~ "Pumpjet",
+                   str_detect(cat, "prop") ~ "Propeller"),
+         level = case_when(str_detect(cat, "1") ~ "Central",
+                            str_detect(cat, "2") ~ "High",
+                            str_detect(cat, "3") ~ "Low"))
 
-ggplot(df, aes(x = kts, y = val, col = cat)) +
-  geom_line() +
-  ylim(0, 1)
+ggplot(df_temp, aes(x = kts, y = val, col = cat)) +
+  geom_line(size = 1)
+
+ggplot(df_temp, aes(x = kts, y = val, col = level, linetype = system)) +
+  geom_line(size = 1) +
+  scale_color_manual(values = c("black", "blue", "red") )+
+  ylim(0, 1)+
+  labs(title = "Efficiency Curve Assumptions")+
+  scale_x_continuous(name ="speed (kt)")+
+  scale_y_continuous(name ="efficiency")
 
 plotly::ggplotly()
 
 df <- df %>%
   mutate(set = case_when(
-    cat %in% c("input_prop_1", "input_jet_1") ~ "Middle",
-    cat %in% c("input_prop_2", "input_jet_3") ~ "Close",
-    cat %in% c("input_prop_3", "input_jet_2") ~ "Far"
+    cat %in% c("input_prop_2", "input_jet_3") ~ "A",
+    cat %in% c("input_prop_3", "input_jet_2") ~ "B",
+    cat %in% c("input_prop_1", "input_jet_1") ~ "C"
   ))
 
 ls <- df %>%
@@ -37,27 +50,28 @@ ls <- df %>%
 ls_comp <- ls %>%
   map( ~ .x %>%
          select(kts, eff.jet, eff.prop) %>%
-         run_subs(hotel = 150,
+         run_subs(hotel = 100,
                   total.batt = 500,
                   batt.dens = 0.14,
                   patrol = 2.5,
                   max.speed = 18,
                   max.power = 7,
-                  speed = 8,
-                  power = 1500,
+                  speed = 20,
+                  power = 6500,
                   system = "jet",
[…]

### 2018-01-05 DraftExec.Rnw
massive day, all the important plots basically here

@@
 This has been exaccerbated beyond realistic levels in the case of the the low propeller assumption line for the sake of testing sensitivity more dramatically to some losses in this regime.
+It is unlikely that unless some other significant factor not discussed in this paper that a state-of-the-art submarine propeller would not be able to attain maximum efficiencies between 5 and 15kt.
+Hence, the low propeller curve is considered relatively unlikely, but useful for investigating sensitivity to errors or adjustments.
+
 These efficiency curve assumptions are shown in Figure \ref{fig:Efficiency_assumptions}.
@@
 Consequently, the actual hotel load experienced might vary considerably, and as such the model used in the app allows the hotel load assumption to be varied, in order to test sensitivity to this parameter.
-A flat 100kW hotel load is used as a starting assumption in the initial calculations in this paper, which as seen, produces realistic results.
+A flat 100kW hotel load is used as a starting assumption in the initial calculations in this paper, and some sensitivity analysis is given at half and double this power, to given an indication of the impact of plausible alternative energy demands dependent on operational circumstances, and choice of combat system.
 
@@
 Taking a central figure and converting this to Megajoules per kilogram (MJ/kg) we assume 0.14MJ/kg to be the effective energy density of lead acid batteries for our test submarine's purposes.
-This parameter can be adjusted easily for sensitivity testing.
+This parameter can be adjusted easily for sensitivity testing, and in our case this will be adjusted to half, and double this energy density, noting that lead-acid batteries become much slower to charge as their charge level passes 80\%, and consequently it may be the case that only 50\% of their available capacity might be genuinely available for tactical purposes \parencite{briggs2016}.
 
@@
 Precise energy density of these batteries is not publically available.
-For domestic uses such as handheld electronics or solar power applications, lithium ion batteries may have 3-4 times the energy density of lead-acid batteries.
+For domestic uses such as handheld electronics or solar power applications, lithium ion batteries may have 3-5 times the energy density of lead-acid batteries.
 However, the actual useful density of lithium ion batteries can vary widely depending on different chemistries and designs.
 It might be assumed that for very large battery banks (such as submarines) where the stability of the battery bank is of critical importance, maximum density migth be sacrificed in favour of additional safety systems, such as increased monitoring, as well as spacing and insulation barriers between cells, in order to prevent thermal runaway in one cell leading to a catastrophic fire.
+The realistic energy density advantage that might be obtained as been suggested to lying closer to a factor of two \parencite{greenfield2016}.
+
+The total embarked load of the batteries is not a well-known public figure, though it is likely to be several hundred tonnes in order to provide the likely desired levels of dived endurance.
+It has been suggested publically that the approximately 700 tonnes might be required for the Shortfin Barracuda \parencite{ohff2017}.
+This would appear to be quite a large fraction of the plausible payload of a submarine, even of over 4000 tonnes displacement, given that a great deal of the potential displacement will be taken by the thick steel involved in constructing a pressure vessel (probably at least a thousand tonnes per submarine) \parencite{jean2017}.
+500 tonnes of batteries has been assumed as a nominal figure for these purposes.
+Given that the mass of batteries embarked has only a direct and linear relationship with the dived range and endurance, no sensitivity testing will be undertaken on this variable.
+Proportional changes to the energy density of the batteries will have equivalent effect for the purposes of measuring the impact on performance.
+Consideration of likely impacts on space, and hull-resizing, and associated drag are beyond the scope of this paper.
+
+\subsection{Drag Coefficient of Hull}
+
+In the online version of the web-app, alternative reference points or known facts relating a given power output to a speed can be used in order to solve for this constant $C_d$.
+For instance, one might assume (as done by \cite{davies2017}) that the hotel load might match the propulsive demand at some given speed, for some given hotel load.
+However, given that the precise hotel load is generally not known, and highly variable, and there is potential for the different systems to have widely varying propulsive efficiencies at low speeds, this method is subject to very wide error margins.
+
+Given that any known power consumption level and speed match can be used as a reference, I have chosen instead to use the probable power consumptio at top speed, both of which we have some indication of, and given that propulsive efficiencies of the two systems are likely to be much closer at this speed, there will be less error in likely comparisons between different design concepts.
+It is known that the a early German designs for a plausible submarine for SEA 1000 included a 6MW electric motor \parencite{patrick2012}, and the Shortfin Barracuda of Naval Group is often referred to as having a 7MW electric motor \parencite{ohff2017} \parencite{coates2016}, with top speeds ">20kt".
+Given the nature of increasing drag at high speeds, it seems unlikely that top speeds will be very far above 20kt.
+Hence, in this model we have assumed that 20kt will be obtained with the ouput of 6.5MW.
+
+\section{Results and Discussion}
+
+\subsection{Power Demand}
+As part of the calculation of results for each pair of efficiency curves, the power requirements of each respective component (propulsion for the two respective systems, as well as effective power required, and hote load) can be plotted to ensure that realistic results are obtained, which is important for selecting plausible values for the constant $C_d$.
+For the central efficiency curves, we can show the resultant power curves for the the assumed values of battery load, energy density, and hotel load, as seen in Figure \ref{fig:power_curve}.
+
+\begin{figure}
+<<plot_power1, echo=FALSE, warning=FALSE, message=F >>=
+df <- read_rds("extdata/df_full_complete_fac.rds")
+df %>%
+  filter(var == "Power (kW)", pair %in% c("Central Selections")) %>%
+  ggplot(aes(x = as.numeric(kts), y = as.numeric(val), col = cat))+
+  scale_x_continuous(name ="speed (kt)")+
+  scale_y_continuous(name = "Power (kW)")+
+  labs(title = "Example Power Curves", col = "Power Demand")+
+  geom_line(size = 0.8)
+@
+\caption{With central assumptions, a pumpjet powered submarine reaches 20kt with 6.5MW of delivered power.  Hotel load matches propulsion power demand around or below 5kt.}
+\label{fig:power_curve}
+\end{figure}
+
+It can be seen here that one of Andrew Davies assumptions, regarding hotel load matching propulsion power demand at patrol speeds, might match quite closely with our assumptions at a patrol speed of 4kt for the case of a pumpjet.
+It also shows the remarkable impact of the square law on power demands, particularly their effect at low speeds.
+At just a few kt, an entire submarine might be propelled by a power level comfortably produced by a large family car.
+This gives further explains why it is that efficiencies of propulsion systems aren't routinely discussed at dramatically lowered speeds.
+Even for jets which have efficiencies which decline much faster than those we've assumed (for example in \cite{fujisawa1995}), the reduction in power demanded falls substantially faster than efficiency.
+Consequently, at low speeds nothing dramatic occurs to make the declining efficiencies of consequence to operators.
+It is often still more economical for such vessels to travel very slowly, well outside the ideal performance range of their propulsion systems, simply because of the drag on the hull.
+
+Overall, this chart is consistent with expectation, and confirms that a plausible value of $C_d$ has been selected under these assumptions.
+
+\subsection{Central Results}
+The plots comparing the resultant endurance and range calculations across the speed range considered are in Figures \ref{fig:central_extreme}, \ref{fig:central_jet}, \ref{fig:central_prop} and \ref{fig:bottom_top}.
+In all cases, the comparison for the central efficiency curves have been replotted centrally, in order to highlight varience under each of the different assumption comparisons.
+
+It is clear that under the central assumptions, the declining efficiency of the pumpjet has a very significant impact on range and endurance, particularly at speeds around 4-8kt.
+It might be noted that under the central assumption, the propeller system achieves 138 hours of dived endurance at 4kt where the pumpjet only manages 91 hours dived endurance at this speed.
+Consequently, at a likely patrol speed, the difference amounts of almost exactly two days in dived endurance.
+One might presume that such a difference is of some considerable tactical significance.
+In order to match the propeller's dived endurance, the pumpjet submarine would need to slow to 2.5kt.
+
+The phenomenon manifests itself in a very substantial reduction in the overall range that can be achieved submerged, and a difference between the speed at which maximum range can be achieved.
+Under central assumptions, the propeller driven submarine can reach over 540nm, at any speed between 4 and 5kt.
+The pump-jet driven submarine reaches over 360nm between 3 and 6kt.
+A difference 180nm dived range would presumeably be of considerable tactical significance.
+
+\begin{figure}
+<<plot_results1, echo=FALSE, warning=FALSE, message=F >>=
+df <- read_rds("extdata/df_full_complete_fac.rds")
+df %>%
+  filter( pair %in% c("Widest Separation", "Narrowest Separation", "Central Selections")) %>%
+  filter(var %in% c("Efficiency", "Endurance (hrs)", "Range (nm)")) %>%
+  #filter(var == "end_prop", pair %in% c("top", "bottom", "propup", "propdown", "jetup", "jetdown", "wide", "narrow", "mid") ) %>%
+  ggplot(aes(x = as.numeric(kts), y = as.numeric(val), linetype = as.factor(cat), col = level))+
+  geom_line(aes(group = cat))+
+  scale_linetype_manual(labels = c("Propeller", "Pumpjet" ), values = c(1, 2 ) )+
+  scale_colour_manual(values = c("black", "blue", "red") )+
+  scale_x_continuous(name ="speed (kt)")+
+  scale_y_continuous(name =element_blank())+
+  labs(linetype = "system")+
+  facet_grid(var ~ pair, scales = "free", switch =  "y")
+@
+\caption{The most extreme and central assumptions are compared}
+\label{fig:central_extreme}
+\end{figure}
+
+\begin{figure}
+<<plot_results2, echo=FALSE, warning=FALSE, message=F >>=
+df <- read_rds("extdata/df_full_complete_fac.rds")
+df %>%
+  filter( pair %in% c("Jet Lowered", "Jet Raised", "Central Selections")) %>%
+  filter(var %in% c("Efficiency", "Endurance (hrs)", "Range (nm)")) %>%
+  #filter(var == "end_prop", pair %in% c("top", "bottom", "propup", "propdown", "jetup", "jetdown", "wide", "narrow", "mid") ) %>%
+  ggplot(aes(x = as.numeric(kts), y = as.numeric(val), linetype = as.factor(cat), col = level))+
+  geom_line(aes(group = cat))+
+  scale_linetype_manual(labels = c("Propeller", "Pumpjet" ), values = c(1, 2 ) )+
+  scale_colour_manual(values = c("black", "blue", "red") )+
+  scale_x_continuous(name ="speed (kt)")+
+  scale_y_continuous(name =element_blank())+
+  labs(linetype = "system")+
+  facet_grid(var ~ pair, scales = "free", switch =  "y")
+@
+\caption{The impact of different jet efficiency assumptions compared to the central propeller curve shows wide variance}
+\label{fig:central_jet}
+\end{figure}
+
+\begin{figure}
+<<plot_results3, echo=FALSE, warning=FALSE, message=F >>=
+df <- read_rds("extdata/df_full_complete_fac.rds")
+df %>%
+  filter( pair %in% c("Propeller Lowered", "Propeller Raised", "Central Selections")) %>%
+  filter(var %in% c("Efficiency", "Endurance (hrs)", "Range (nm)")) %>%
+  #filter(var == "end_prop", pair %in% c("top", "bottom", "propup", "propdown", "jetup", "jetdown", "wide", "narrow", "mid") ) %>%
+  ggplot(aes(x = as.numeric(kts), y = as.numeric(val), linetype = as.factor(cat), col = level))+
+  geom_line(aes(group = cat))+
+  scale_linetype_manual(labels = c("Propeller", "Pumpjet" ), values = c(1, 2 ) )+
+  scale_colour_manual(values = c("black", "blue", "red") )+
+  scale_x_continuous(name ="speed (kt)")+
+  scale_y_continuous(name =element_blank())+
+  labs(linetype = "system")+
+  facet_grid(var ~ pair, scales = "free", switch =  "y")
+@
+\caption{The impact of different propeller efficiency assumptions compared to the central pumpjet curve shows less variance}
+\label{fig:central_prop}
+\end{figure}
+
+\begin{figure}
+<<plot_results4, echo=FALSE, warning=FALSE, message=F >>=
+df <- read_rds("extdata/df_full_complete_fac.rds")
+df %>%
+  filter( pair %in% c("Lowest Lines", "Highest Lines", "Central Selections")) %>%
+  filter(var %in% c("Efficiency", "Endurance (hrs)", "Range (nm)")) %>%
+  #filter(var == "end_prop", pair %in% c("top", "bottom", "propup", "propdown", "jetup", "jetdown", "wide", "narrow", "mid") ) %>%
+  ggplot(aes(x = as.numeric(kts), y = as.numeric(val), linetype = as.factor(cat), col = level))+
+  geom_line(aes(group = cat))+
+  scale_linetype_manual(labels = c("Propeller", "Pumpjet" ), values = c(1, 2 ) )+
+  scale_colour_manual(values = c("black", "blue", "red") )+
+  scale_x_continuous(name ="speed (kt)")+
+  scale_y_continuous(name =element_blank())+
+  labs(linetype = "system")+
+  facet_grid(var ~ pair, scales = "free", switch =  "y")
+@
+\caption{Comparing the lower and higher lines for both systems together}
+\label{fig:bottom_top}
+\end{figure}
+
+\subsection{Sensitivity of Curve Selection}
+As can be seen from the comprehensive comparison in Figures \ref{fig:central_extreme}, \ref{fig:central_jet}, \ref{fig:central_prop} and \ref{fig:bottom_top}, the shape of curves selected has very different levels of impact at different speeds.
+Key observations include how small the impact of efficiency variation to eithre curves tends to be at very low speeds (under 2kt).
+This is due to hotel load becoming clearly dominant at such speeds.
+The relative difference between the lines also diminishes somewhat towards higher speeds.
+More striking, however, is the very dramatic reduction in both range and endurance at higher speeds.
+At a speed of 20kt, a propeller driven sumbarine might manage to remain dived for 3.6 hours according to these assumptions, where a pumpjet driven submarine might manage just under 3 hours.
+A difference of about half an hour would presumeably be of less tactical significance.
+At any speed above 10kt, neither submarine would be able to remain dived for much more than 24 hours (25.4 hours for the propeller system, 18.3 for the jet).
+
+It should also be noted that these results, and those around 8kt, can also provide some insight into likely snorting intervals and indiscretion ratios.
+As given in \cite{patrick2012}, submarines tend to carry diesel power capacity that approaches their main motor power.
+Whilst the difference between the two systems still appears very significant at these speeds, it appears that if transits occur in the region of 8-10kt, snorts might occurr roughly daily, and last a number of hours.
+The indiscretion ratio during commute of the jet system would be significantly impaired given the substantially larger power demands at transit speeds, though a detailed calculation and comparison lies outside the scope of this paper.
+
+It seems extremely clear from these conclusions that the general results confirm very strongly that the impact of the propulsion system is likely to be of the greatest importance in the speed range 3-8kt, where propulsion power is more significant relative to hotel load, but excessive drag does not prohibit significant dives.
+
+A further feature of the results is that the relative difference in performance is much more sensitive to changes in the efficiency of the jet at the relevant speeds, owing to their larger proportional impact at low efficiencies.
+Whilst a smoothly declining curve from around or over 50\% at 20kt to zero at 0kt, the precise path taken, particularly the exact level around 3-8kt, is of great importance, substantially more than plausible variations in the propeller curve.
+This can be seen in an overlaid plot of the proportional differences in total power demand (hence endurance and range) for the different pairs of curves as in Figure \ref{fig:all_proportions}.
+
+\begin{figure}
+<<plot_proportions, echo=FALSE, warning=FALSE, message=F >>=
+df <- read_rds("extdata/df_full_complete_fac.rds")
+df %>%
+  filter(var == "Advantage Propeller %", pair %in% c("Lowest Lines", "Jet Lowered", "Propeller Lowered", "Widest Separation", "Central Selections", "Narrowest Separation", "Propeller Raised", "Jet Raised", "Highest Lines") ) %>%
+  ggplot(aes(x = as.numeric(kts), y = as.numeric(val), col = pair))+
+  geom_line()+
+  scale_x_continuous(name ="speed (kt)")+
+  labs(col = "Curves Compared")+
+  scale_y_continuous(name ="Percentage Propeller Advantage", breaks=seq(0, 120, 10))
+@
+\caption{The most proportional differences between the propeller and pumpjet performances for a range of assumptions}
+\label{fig:all_proportions}
+\end{figure}
+
+The proportional change in overall power requirement (directly determining both range and endurance) shows that our central scenario shows a propeller driven submarine would have more than 60\% advantage in terms of range and endurance around 5-6kt.
+It is interesting to note that the proportional change is greatest slightly above the speed where maximum range is achieved, and the point of peak advantage moves steadily towards higher speeds as the difference in efficiency between the systems increases.
+
+As can be seen, even compred to the central propeller curve, choosing a jet efficiency curve somewhat lower than the central assumption means that the peak difference immediately jumps to over 100\%, or effectively allow a propeller powered submarine to achieve double the range and endurance of a pumpjet powered submarine at those speeds.
+It is important to note that these speeds in include those speeds which might lie near the maximum speeds that could be sustained by Air Independent Propulsion (AIP) systems \parencite[100]{@stanford2017}.
+Consequently, it is likely that a conventional submarine that used pumpjets might only achieve a fraction of the dived range and endurance enhancements generally expected on propeller powered submarines which employ this technology.
+
+The proportional improvements will also be effective rough guides to the likely impact on indescretion ratios and overall range.\footnote{Additional considerations including the size and efficiency of the diesel engines and generators must be included for proper analyses of indescretion ratios, which will not be given here.}  Near patrol speed almost all scenarios show a difference approaching or exceeding 50\%, which will have a material impact on indescretion rates during all combat operations.
+
+Perhaps surprisingly, the impact at transit speeds is still quite significant, with no scenarios showing less than 20\% advantage around 8-10kt.
+The sensitivity to pumpjet efficiency here is also very high, with three scenarios showing an impact of over 70\%.
+Given the likelihood that a submarine's average speed over its range is likely to be dominated by its transit speed, this indicates that the choice of a pumpjet propulsion system will also result in a substantial reduction in the overall range of the submarine, quite probably in the range of between 25-50\%.
+This would require the embarkation of significantly larger volumes of diesel to meet minimum range requirements, and also contribute to higher running costs.
+
+\subsection{Sensitivity to Hotel Load}
+
+\subsection{Sensitivity to Battery Changes}
 

### 2018-01-05 DraftExec.tex
massive day, all the important plots basically here

@@
 
-The power demanded by the propulsion is assumed simply to be the propulsive efficiency at any given speed $\eta_v$ multiplied by the Effective Power.
+The power drawn by the propulsion $P_D$ is assumed simply to be the propulsive efficiency at any given speed $\eta_v$ multiplied by the Effective Power, as given in Equation \ref{eq:powerdrawn}.
+Using a plausible efficiency assumption and total power required, Equation \ref{eq:powerdrawn} can be used to solve for the constant $C_d$.
 Other efficiency factors, such as the efficiency of electric motors or hull efficiency factors are assumed to be either relatively small, or essentially similar between submarines which might be conventionally powered but with a jet or propeller being the main point of difference.
@@
 \begin{equation}
+\label{eq:powerdrawn}
+P_D = \eta_vP_E
+\end{equation}
+
+\begin{equation}
 \label{eq:totalpower}
-P_T = \eta_vC_dv^3 + H
+P_T = P_D + H
 \end{equation}
+
+With this relationship defined across the speed range, it is elementary to calculate the time that a given speed could be sustained based on a finite energy store (endurance) and the distance covered in this time (range).
+This provides the fundamental structure that underlies the model which is used to assess the dived range and enurance of conventionally powered submarines, without using air independent propulsion (AIP).
 
@@
 
+\subsection{Hotel Load Assumptions}
 It is also required that some assumptions are made about the hotel load, or the power which the submarine requires which doesn't go to propulsion.
@@
 Consequently, the actual hotel load experienced might vary considerably, and as such the model used in the app allows the hotel load assumption to be varied, in order to test sensitivity to this parameter.
-100kW hotel load is used as a starting assumption in the initial calculations in this paper.
+A flat 100kW hotel load is used as a starting assumption in the initial calculations in this paper, and some sensitivity analysis is given at half and double this power, to given an indication of the impact of plausible alternative energy demands dependent on operational circumstances, and choice of combat system.
+
+\subsection{Battery Assumptions}
+Since the purpose of the task at hand is to assess dived endurance only, this model focuses on a single energy store simulated by a total embarked mass (battery bank) and an energy density of those batteries.
+Since the SEA 1000 project has elected not to use lithium electric batteries, the energy density of the batteries can be assumed to be that of lead-acid batteries, which are known to be in the region of 30-50Wh/kg (Watt-hours per kilogram).
+Taking a central figure and converting this to Megajoules per kilogram (MJ/kg) we assume 0.14MJ/kg to be the effective energy density of lead acid batteries for our test submarine's purposes.
+This parameter can be adjusted easily for sensitivity testing, and in our case this will be adjusted to half, and double this energy density, noting that lead-acid batteries become much slower to charge as their charge level passes 80\%, and consequently it may be the case that only 50\% of their available capacity might be genuinely available for tactical purposes \parencite{briggs2016}.
+
+It is worthwhile considering that alternative types of battery may be on offer for submarines.
+In particular, lithium ion batteries were offered by TKMS in their bid for SEA 1000, and are being installed in Type 212 submarines operated by the German navy, as well as contracted for the new Norwegian submarines, as well as Israeli submarines.
+Precise energy density of these batteries is not publically available.
+For domestic uses such as handheld electronics or solar power applications, lithium ion batteries may have 3-5 times the energy density of lead-acid batteries.
+However, the actual useful density of lithium ion batteries can vary widely depending on different chemistries and designs.
+It might be assumed that for very large battery banks (such as submarines) where the stability of the battery bank is of critical importance, maximum density migth be sacrificed in favour of additional safety systems, such as increased monitoring, as well as spacing and insulation barriers between cells, in order to prevent thermal runaway in one cell leading to a catastrophic fire.
+The realistic energy density advantage that might be obtained as been suggested to lying closer to a factor of two \parencite{greenfield2016}.
+
+The total embarked load of the batteries is not a well-known public figure, though it is likely to be several hundred tonnes in order to provide the likely desired levels of dived endurance.
+It has been suggested publically that the approximately 700 tonnes might be required for the Shortfin Barracuda \parencite{ohff2017}.
+This would appear to be quite a large fraction of the plausible payload of a submarine, even of over 4000 tonnes displacement, given that a great deal of the potential displacement will be taken by the thick steel involved in constructing a pressure vessel (probably at least a thousand tonnes per submarine) \parencite{jean2017}.
+500 tonnes of batteries has been assumed as a nominal figure for these purposes.
+Given that the mass of batteries embarked has only a direct and linear relationship with the dived range and endurance, no sensitivity testing will be undertaken on this variable.
+Proportional changes to the energy density of the batteries will have equivalent effect for the purposes of measuring the impact on performance.
+Consideration of likely impacts on space, and hull-resizing, and associated drag are beyond the scope of this paper.
+
+\subsection{Drag Coefficient of Hull}
+
+In the online version of the web-app, alternative reference points or known facts relating a given power output to a speed can be used in order to solve for this constant $C_d$.
+For instance, one might assume (as done by \cite{davies2017}) that the hotel load might match the propulsive demand at some given speed, for some given hotel load.
+However, given that the precise hotel load is generally not known, and highly variable, and there is potential for the different systems to have widely varying propulsive efficiencies at low speeds, this method is subject to very wide error margins.
+
+Given that any known power consumption level and speed match can be used as a reference, I have chosen instead to use the probable power consumptio at top speed, both of which we have some indication of, and given that propulsive efficiencies of the two systems are likely to be much closer at this speed, there will be less error in likely comparisons between different design concepts.
+It is known that the a early German designs for a plausible submarine for SEA 1000 included a 6MW electric motor \parencite{patrick2012}, and the Shortfin Barracuda of Naval Group is often referred to as having a 7MW electric motor \parencite{ohff2017} \parencite{coates2016}, with top speeds ">20kt".
+Given the nature of increasing drag at high speeds, it seems unlikely that top speeds will be very far above 20kt.
+Hence, in this model we have assumed that 20kt will be obtained with the ouput of 6.5MW.
+
+\section{Results and Discussion}
+
+\subsection{Power Demand}
+As part of the calculation of results for each pair of efficiency curves, the power requirements of each respective component (propulsion for the two respective systems, as well as effective power required, and hote load) can be plotted to ensure that realistic results are obtained, which is important for selecting plausible values for the constant $C_d$.
+For the central efficiency curves, we can show the resultant power curves for the the assumed values of battery load, energy density, and hotel load, as seen in Figure \ref{fig:power_curve}.
+
+\begin{figure}
+\begin{knitrout}
+\definecolor{shadecolor}{rgb}{0.969, 0.969, 0.969}\color{fgcolor}
+
+{\centering \includegraphics[width=\maxwidth]{figures/plots-plot_power1-1}
+
+}
+
+\end{knitrout}
+\caption{With central assumptions, a pumpjet powered submarine reaches 20kt with 6.5MW of delivered power.  Hotel load matches propulsion power demand around or below 5kt.}
+\label{fig:power_curve}
+\end{figure}
+
+It can be seen here that one of Andrew Davies assumptions, regarding hotel load matching propulsion power demand at patrol speeds, might match quite closely with our assumptions at a patrol speed of 4kt for the case of a pumpjet.
+It also shows the remarkable impact of the square law on power demands, particularly their effect at low speeds.
+At just a few kt, an entire submarine might be propelled by a power level comfortably produced by a large family car.
+This gives further explains why it is that efficiencies of propulsion systems aren't routinely discussed at dramatically lowered speeds.
+Even for jets which have efficiencies which decline much faster than those we've assumed (for example in \cite{fujisawa1995}), the reduction in power demanded falls substantially faster than efficiency.
+Consequently, at low speeds nothing dramatic occurs to make the declining efficiencies of consequence to operators.
+It is often still more economical for such vessels to travel very slowly, well outside the ideal performance range of their propulsion systems, simply because of the drag on the hull.
+
+Overall, this chart is consistent with expectation, and confirms that a plausible value of $C_d$ has been selected under these assumptions.
+
+\subsection{Central Results}
+The plots comparing the resultant endurance and range calculations across the speed range considered are in Figures \ref{fig:central_extreme}, \ref{fig:central_jet}, \ref{fig:central_prop} and \ref{fig:bottom_top}.
+In all cases, the comparison for the central efficiency curves have been replotted centrally, in order to highlight varience under each of the different assumption comparisons.
+
+It is clear that under the central assumptions, the declining efficiency of the pumpjet has a very significant impact on range and endurance, particularly at speeds around 4-8kt.
+It might be noted that under the central assumption, the propeller system achieves 138 hours of dived endurance at 4kt where the pumpjet only manages 91 hours dived endurance at this speed.
+Consequently, at a likely patrol speed, the difference amounts of almost exactly two days in dived endurance.
+One might presume that such a difference is of some considerable tactical significance.
+In order to match the propeller's dived endurance, the pumpjet submarine would need to slow to 2.5kt.
+
+The phenomenon manifests itself in a very substantial reduction in the overall range that can be achieved submerged, and a difference between the speed at which maximum range can be achieved.
+Under central assumptions, the propeller driven submarine can reach over 540nm, at any speed between 4 and 5kt.
+The pump-jet driven submarine reaches over 360nm between 3 and 6kt.
+A difference 180nm dived range would presumeably be of considerable tactical significance.
+
+\begin{figure}
+\begin{knitrout}
+\definecolor{shadecolor}{rgb}{0.969, 0.969, 0.969}\color{fgcolor}
+
+{\centering \includegraphics[width=\maxwidth]{figures/plots-plot_results1-1}
+
+}
+
+\end{knitrout}
+\caption{The most extreme and central assumptions are compared}
+\label{fig:central_extreme}
+\end{figure}
+
+\begin{figure}
+\begin{knitrout}
+\definecolor{shadecolor}{rgb}{0.969, 0.969, 0.969}\color{fgcolor}
+
+{\centering \includegraphics[width=\maxwidth]{figures/plots-plot_results2-1}
+
+}
+
+\end{knitrout}
+\caption{The impact of different jet efficiency assumptions compared to the central propeller curve shows wide variance}
+\label{fig:central_jet}
+\end{figure}
+
+\begin{figure}
+\begin{knitrout}
+\definecolor{shadecolor}{rgb}{0.969, 0.969, 0.969}\color{fgcolor}
+
+{\centering \includegraphics[width=\maxwidth]{figures/plots-plot_results3-1}
+
+}
+
+\end{knitrout}
+\caption{The impact of different propeller efficiency assumptions compared to the central pumpjet curve shows less variance}
+\label{fig:central_prop}
+\end{figure}
+
+\begin{figure}
+\begin{knitrout}
+\definecolor{shadecolor}{rgb}{0.969, 0.969, 0.969}\color{fgcolor}
+
+{\centering \includegraphics[width=\maxwidth]{figures/plots-plot_results4-1}
+
+}
+
+\end{knitrout}
+\caption{Comparing the lower and higher lines for both systems together}
+\label{fig:bottom_top}
+\end{figure}
+
+\subsection{Sensitivity of Curve Selection}
+As can be seen from the comprehensive comparison in Figures \ref{fig:central_extreme}, \ref{fig:central_jet}, \ref{fig:central_prop} and \ref{fig:bottom_top}, the shape of curves selected has very different levels of impact at different speeds.
+Key observations include how small the impact of efficiency variation to eithre curves tends to be at very low speeds (under 2kt).
+This is due to hotel load becoming clearly dominant at such speeds.
+The relative difference between the lines also diminishes somewhat towards higher speeds.
+More striking, however, is the very dramatic reduction in both range and endurance at higher speeds.
+At a speed of 20kt, a propeller driven sumbarine might manage to remain dived for 3.6 hours according to these assumptions, where a pumpjet driven submarine might manage just under 3 hours.
+A difference of about half an hour would presumeably be of less tactical significance.
+At any speed above 10kt, neither submarine would be able to remain dived for much more than 24 hours (25.4 hours for the propeller system, 18.3 for the jet).
+
+It should also be noted that these results, and those around 8kt, can also provide some insight into likely snorting intervals and indiscretion ratios.
+As given in \cite{patrick2012}, submarines tend to carry diesel power capacity that approaches their main motor power.
+Whilst the difference between the two systems still appears very significant at these speeds, it appears that if transits occur in the region of 8-10kt, snorts might occurr roughly daily, and last a number of hours.
+The indiscretion ratio during commute of the jet system would be significantly impaired given the substantially larger power demands at transit speeds, though a detailed calculation and comparison lies outside the scope of this paper.
+
+It seems extremely clear from these conclusions that the general results confirm very strongly that the impact of the propulsion system is likely to be of the greatest importance in the speed range 3-8kt, where propulsion power is more significant relative to hotel load, but excessive drag does not prohibit significant dives.
+
+A further feature of the results is that the relative difference in performance is much more sensitive to changes in the efficiency of the jet at the relevant speeds, owing to their larger proportional impact at low efficiencies.
+Whilst a smoothly declining curve from around or over 50\% at 20kt to zero at 0kt, the precise path taken, particularly the exact level around 3-8kt, is of great importance, substantially more than plausible variations in the propeller curve.
+This can be seen in an overlaid plot of the proportional differences in total power demand (hence endurance and range) for the different pairs of curves as in Figure \ref{fig:all_proportions}.
+
+\begin{figure}
+\begin{knitrout}
+\definecolor{shadecolor}{rgb}{0.969, 0.969, 0.969}\color{fgcolor}
+
+{\centering \includegraphics[width=\maxwidth]{figures/plots-plot_proportions-1}
+
+}
+
+\end{knitrout}
+\caption{The most proportional differences between the propeller and pumpjet performances for a range of assumptions}
+\label{fig:all_proportions}
+\end{figure}
 
@@
 \begin{knitrout}
-\definecolor{shadecolor}{rgb}{0.969, 0.969, 0.969}\color{fgcolor}
-
-{\centering \includegraphics[width=\maxwidth]{figures/plots-plot_power-1}
-
-}
-
+\definecolor{shadecolor}{rgb}{0.969, 0.969, 0.969}\color{fgcolor}\begin{kframe}
+\begin{verbatim}
+## NULL
+\end{verbatim}
+\end{kframe}
 \end{knitrout}

### 2018-01-05 scripts/build plots for paper.R

@@ -27,9 +27,9 @@ plotly::ggplotly()
 
 df <- df %>%
   mutate(set = case_when(
-    cat %in% c("input_prop_2", "input_jet_3") ~ "A",
+    cat %in% c("input_prop_1", "input_jet_1") ~ "A",
     cat %in% c("input_prop_3", "input_jet_2") ~ "B",
-    cat %in% c("input_prop_1", "input_jet_1") ~ "C"
+    cat %in% c("input_prop_2", "input_jet_3") ~ "C"
   ))
 
 ls <- df %>%
@@ -50,9 +50,9 @@ ls <- df %>%
 ls_comp <- ls %>%
   map( ~ .x %>%
          select(kts, eff.jet, eff.prop) %>%
-         run_subs(hotel = 100,
+         run_subs(hotel = 50,
                   total.batt = 500,
-                  batt.dens = 0.14,
+                  batt.dens = 0.28,
                   patrol = 2.5,
                   max.speed = 18,
                   max.power = 7,
@@ -74,7 +74,6 @@ produce_all_plots <- function(df) {
       cat ==  "eff.jet" ~ "Pumpjet"
     ))
 
-
   end_plot_df <- df %>%
     filter(cat %in% c("endurance.prop.hour", "endurance.jet.hour")) %>%
     mutate(cat = case_when(
@@ -84,15 +83,28 @@ produce_all_plots <- function(df) {
 
   y_max_e <- roundUpNice(max(pull(end_plot_df, val)))
 
+  comp_plot_df <- df %>%
+    filter(cat %in% c("eff.jet", "eff.prop", "range.jet", "range.prop", "endurance.prop.hour", "endurance.jet.hour")) %>%
+    mutate(sys = case_when(
+      str_detect(cat,  "prop") ~ "Propeller",
+      str_detect(cat,  "jet") ~ "Pumpjet"),
+      cat = case_when(
+        str_detect(cat,  "end") ~ "end",
+        str_detect(cat,  "eff") ~ "eff",
+        str_detect(cat, "range") ~ "range"))
 
+  diff_plot_df <- comp_plot_df %>%
+    spread(sys, val) %>%
+    mutate(prop = (Propeller/Pumpjet -1) * 100,
+           diff = (Propeller - Pumpjet))
 
   end_prop_plot_df <- end_plot_df %>%
     spread(cat, val) %>%
[…]

### 2018-01-05 DraftExec.Rnw
cooking with gas

@@
 
+\begin{figure}
+<<plot_sens_hotel, echo=FALSE, warning=FALSE, message=F >>=
+df <- read_rds("extdata/df_full_diff_cornered.rds")
+df %>%
+  filter(test == "hotel") %>%
+  ggplot(aes(x = kts, y = val, col = pair))+
+  geom_line()+
+  labs(col = "Efficiency \nCurve Pair")+
+  scale_y_continuous(name = element_blank())+
+  facet_grid(var ~ as.factor(hotel), scales = "free")
+@
+\caption{The impact of the propulsion system efficiencies are very sensitive to changes in the hotel load, including as a proportion.}
+\label{fig:sens_hotel}
+\end{figure}
+
 \subsection{Sensitivity to Battery Changes}
+
+\begin{figure}
+<<plot_sens_battery, echo=FALSE, warning=FALSE, message=F >>=
+df <- read_rds("extdata/df_full_diff_cornered.rds")
+df %>%
+  filter(test == "battery") %>%
+  ggplot(aes(x = kts, y = val, col = pair))+
+  geom_line()+
+  labs(col = "Efficiency \nCurve Pair")+
+  scale_y_continuous(name = element_blank())+
+  facet_grid(var ~ as.factor(battery), scales = "free")
+@
+\caption{The impact of the propulsion system efficiencies are extremely sensitive to battery energy density (or load) though proportions remain fixed.}
+\label{fig:sens_battery}
+\end{figure}
+
+\subsection{The interaction of hotel load, battery energy density, and efficiency curves}
+
+\begin{figure}
+<<plot_sens_corners, echo=FALSE, warning=FALSE, message=F >>=
+df <- read_rds("extdata/df_full_diff_cornered.rds")
+df %>%
+  filter(test == "corners") %>%
+  ggplot(aes(x = kts, y = val, col = pair))+
+  geom_line()+
+  labs(col = "Efficiency \nCurve Pair")+
+  scale_y_continuous(name = element_blank())+
+  theme(legend.position="bottom")+
+  facet_grid(var ~ corner, scales = "free", switch = "y")
+@
+\caption{Specifying low density batteries and a high hotel load dramatically reduces the proportional and nominal impact of propulsion system efficiencies at low speeds}
+\label{fig:sens_corners}
+\end{figure}
 

### 2018-01-05 DraftExec.tex
cooking with gas

@@
 \label{fig:all_proportions}
+\end{figure}
+
+The proportional change in overall power requirement (directly determining both range and endurance) shows that our central scenario shows a propeller driven submarine would have more than 60\% advantage in terms of range and endurance around 5-6kt.
+It is interesting to note that the proportional change is greatest slightly above the speed where maximum range is achieved, and the point of peak advantage moves steadily towards higher speeds as the difference in efficiency between the systems increases.
+
+As can be seen, even compred to the central propeller curve, choosing a jet efficiency curve somewhat lower than the central assumption means that the peak difference immediately jumps to over 100\%, or effectively allow a propeller powered submarine to achieve double the range and endurance of a pumpjet powered submarine at those speeds.
+It is important to note that these speeds in include those speeds which might lie near the maximum speeds that could be sustained by Air Independent Propulsion (AIP) systems \parencite[100]{@stanford2017}.
+Consequently, it is likely that a conventional submarine that used pumpjets might only achieve a fraction of the dived range and endurance enhancements generally expected on propeller powered submarines which employ this technology.
+
+The proportional improvements will also be effective rough guides to the likely impact on indescretion ratios and overall range.\footnote{Additional considerations including the size and efficiency of the diesel engines and generators must be included for proper analyses of indescretion ratios, which will not be given here.}  Near patrol speed almost all scenarios show a difference approaching or exceeding 50\%, which will have a material impact on indescretion rates during all combat operations.
+
+Perhaps surprisingly, the impact at transit speeds is still quite significant, with no scenarios showing less than 20\% advantage around 8-10kt.
+The sensitivity to pumpjet efficiency here is also very high, with three scenarios showing an impact of over 70\%.
+Given the likelihood that a submarine's average speed over its range is likely to be dominated by its transit speed, this indicates that the choice of a pumpjet propulsion system will also result in a substantial reduction in the overall range of the submarine, quite probably in the range of between 25-50\%.
+This would require the embarkation of significantly larger volumes of diesel to meet minimum range requirements, and also contribute to higher running costs.
+
+\subsection{Sensitivity to Hotel Load}
+
+\begin{figure}
+\begin{knitrout}
+\definecolor{shadecolor}{rgb}{0.969, 0.969, 0.969}\color{fgcolor}
+
+{\centering \includegraphics[width=\maxwidth]{figures/plots-plot_sens_hotel-1}
+
+}
+
+\end{knitrout}
+\caption{The impact of the propulsion system efficiencies are very sensitive to changes in the hotel load, including as a proportion.}
+\label{fig:sens_hotel}
+\end{figure}
+
+\subsection{Sensitivity to Battery Changes}
+
+\begin{figure}
+\begin{knitrout}
+\definecolor{shadecolor}{rgb}{0.969, 0.969, 0.969}\color{fgcolor}
+
+{\centering \includegraphics[width=\maxwidth]{figures/plots-plot_sens_battery-1}
+
+}
+
+\end{knitrout}
+\caption{The impact of the propulsion system efficiencies are extremely sensitive to battery energy density (or load) though proportions remain fixed.}
+\label{fig:sens_battery}
+\end{figure}
+
+\subsection{The interaction of hotel load, battery energy density, and efficiency curves}
+
+\begin{figure}
+\begin{knitrout}
+\definecolor{shadecolor}{rgb}{0.969, 0.969, 0.969}\color{fgcolor}
+
+{\centering \includegraphics[width=\maxwidth]{figures/plots-plot_sens_corners-1}
+
+}
+
+\end{knitrout}
+\caption{Specifying low density batteries and a high hotel load dramatically reduces the proportional and nominal impact of propulsion system efficiencies at low speeds}
+\label{fig:sens_corners}
 \end{figure}

### 2018-01-05 scripts/build plots for paper.R

@@ -230,33 +230,50 @@ df_full <- df_full %>%
   bind_rows(df_filler)
 #df_full <- df_filler
 
-df_full %>% write_rds("extdata/df_full_diff_cornered.rds")
+df_full_cornered %>% write_rds("extdata/df_full_diff_cornered.rds")
 
 ######  Let's tru some plots  ######
 
-df_full %>%
+df_full_cornered$hotel <- factor(df_full_cornered$hotel, levels = c("50", "100", "200" ),
+                                  labels = c( "Hotel 50kW", "Hotel 100kW", "Hotel 200kW" ))
+df_full_cornered$battery <- factor(df_full_cornered$battery, levels = c("0.07", "0.14", "0.28" ),
+                                 labels = c( "Battery 0.07MJ/kg", "Battery 0.14MJ/kg", "Battery 0.28MJ/kg" ))
+
+df_full_cornered %>%
   filter(test == "hotel") %>%
   ggplot(aes(x = kts, y = val, col = pair))+
   geom_line()+
-  labs(title = "Hotel Comparison")+
+  labs(col = "Efficiency \nCurve Pair")+
+  scale_y_continuous(name = element_blank())+
   facet_grid(var ~ as.factor(hotel), scales = "free")
 
-df_full %>%
+df_full_cornered %>%
   filter(test == "battery") %>%
   ggplot(aes(x = kts, y = val, col = pair))+
   geom_line()+
-  labs(title = "Battery Comparison")+
+  labs(col = "Efficiency \nCurve Pair")+
+  scale_y_continuous(name = element_blank())+
   facet_grid(var ~ as.factor(battery), scales = "free")
 
 df_full_cornered <- df_full %>%
-  mutate(corner = paste0("a", battery, "and", hotel))
+  mutate(corner = paste0( hotel, " ", battery))
+
+df_full_cornered <- read_rds("extdata/df_full_diff_cornered.rds")
+
+df_full_cornered$var <- factor(df_full_cornered$var, levels =  c("end", "range", "prop"),
+                                 labels = c("Endurance (hrs)", "Range (nm)", "% Change"))
+df_full_cornered$corner <- factor(df_full_cornered$corner, levels = c("200 0.07", "200 0.28","50 0.07", "50 0.28", "100 0.14", "50 0.14",  "200 0.14", "100 0.07", "100 0.28" ),
[…]

### 2018-01-06 DraftExec.Rnw
more on acoustics

@@
 \section{Acoustic Considerations}
+
+\subsection{The importance of cavitation inception}
 It is generally considered that the existence of cavitation of any sort will tend to dominate the acoustic signature of any vessel that doesn't produce any machinery noise by relying on all-electric drive as a conventional submarine would when dived (\cite{giles2010}, \cite{harvie1965construction}).
@@
 
+\subsection{The impact of depth on cavitation}
 It is crucial to understand that increasing depth necessarily increases the static pressure throughout a propulsion system.
@@
 
+\subsection{The significance of flow separation}
 It is also important to understand that the addition of a duct will necessarily to flow separation occuring on the inside of the duct opening at very low speeds which correspond to high advance ratios when load remains constant.
@@
 Whilst at very low power levels, or at good depths, it is quite plausible that this doesn't result in any cavitation, which would normally be a likely occurrence if the blades are operating closer to their design power and such such vortices were shed into the intake.
-
-\section{App Outputs}
-
-Adjustable figures from App.
-
-<<compute, echo=F, message=F, warning=FALSE>>=
-
-df <- read_rds("extdata/default_subs_df.rds")
-
-df <- run_subs(df,
-             hotel = 150,
-             total.batt = 500,
-             batt.dens = 0.14,
-             patrol = 2.5,
-             max.speed = 18,
-             max.power = 7,
-             speed = 8,
-             power = 1500,
-             system = "jet",
-             method = "hotel match")
-
-eff_plot_df <- df %>%
-  filter(cat %in% c("eff.jet", "eff.prop")) %>%
-  mutate(cat = case_when(
-    cat ==  "eff.prop" ~ "Propeller",
-    cat ==  "eff.jet" ~ "Jet"
-  ))
-
-end_plot_df <- df %>%
-  filter(cat %in% c("endurance.prop.hour", "endurance.jet.hour")) %>%
-  mutate(cat = case_when(
-    cat ==  "endurance.prop.hour" ~ "Propeller",
-    cat ==  "endurance.jet.hour" ~ "Jet"
-  ))
-
-y_max_e <- roundUpNice(max(pull(end_plot_df, val)))
-
-range_plot_df <- df %>%
-  filter(cat %in% c("range.prop", "range.jet")) %>%
-  mutate(cat = case_when(
-    cat ==  "range.prop" ~ "Propeller",
-    cat ==  "range.jet" ~ "Jet"
-  ))
-
-y_max_r <- roundUpNice(max(pull(range_plot_df, val)))
-
-power_plot_df <- df %>%
-  filter(cat %in% c("hotel","power.mob.drawn.jet", "power.mob.req", "power.mob.drawn.prop")) %>%
-  mutate(cat = case_when(
-    cat == "hotel" ~ "Hotel Load",
-    cat ==  "power.mob.drawn.prop" ~ "Propulsion Power Drawn Propeller",
-    cat ==  "power.mob.drawn.jet" ~ "Propulsion Power Drawn Jet",
-    cat ==  "power.mob.req" ~ "Thrust Power Required"
-  ))
-
-y_max_p <- roundUpNice(max(pull(power_plot_df, val)))
-@
-
-\begin{figure}
-<<plot_end, echo=FALSE>>=
-ggplot(end_plot_df, aes(x = kts, y = val, col = cat))+
-          geom_line() +
-          theme_minimal() +
-          scale_y_continuous(breaks=seq(0, y_max_e, (y_max_e/10)),
-                             name = "hours") +
-          scale_x_continuous(breaks=seq(0, 20, 2)) +
-          geom_text(aes(label = round(val, digits = 1)), nudge_y = 2, angle = 45)
-@
-\caption{Endurance}
-\label{fig:Endurance_plot}
-\end{figure}
-
-\begin{figure}
-<<plot_range, echo=FALSE>>=
-ggplot(range_plot_df, aes(x = kts, y = val, col = cat))+
-          geom_line()+
-          theme_minimal()+
-          scale_y_continuous(breaks=seq(0, y_max_r, (y_max_r/10)),
-                             name = "nm") +
-          scale_x_continuous(breaks=seq(0, 20, 2)) +
-          geom_text(aes(label = round(val, digits = 1)), nudge_y = 2, angle = 45)
-@
-\caption{Range}
-\label{fig:Range_plot}
-\end{figure}
-
-\begin{figure}
-<<plot_power, echo=FALSE>>=
-read_rds("extdata/rng_ploy_hotlx.rds")
-@
-\caption{Power}
-\label{fig:Power_plot}
-\end{figure}
+However, given the inherently unpredictable and unsteady nation of such vortex formation and shedding, the this phenomenon would increase the risk of cavitation occurring unexpectedly in marginal conditions.
+In addition, in the absence of cavitation, the sudden and changing pressure gradients over the blades would dramatically increase the overall pressure fluctuations and turbulence level that occurs within the propulsor.
+This would necessarily increase the overall levels of noise generated.
+It is plausible that the noise is of sufficiently small level that it is not considered a significant compromise of stealth.
+In essence, the higher levels of noise lie below some threshold of significance.
+However, it is almost certainly untrue that that pumpjet actually generates less noise overall than a propeller at very low speeds.
+
+\subsection{The impact of duct shielding}
+The presence of the duct around the impeller may also have the effect of preventing the noise from being effectively radiated away from the propulsor.
+This may well result in substantial reductions in some types of noise being radiated away from the propulsor in certain directions, which may have given rise to the consistent qualification of 'radiated' being used to descrive the type of noise which differentiated the French submarine bid \parencite{stewart2016, davies2017, ohff2016}, including by the designers of the submarine \parencite{autret2016}.
+This is a secondary effect, quite separate to the avoidance of cavitation, which might be seen as an acoustic advantage in circumstances where cavitation might not be expected to occur on either type of propulsor.
+It is highly likely that this is the effect which is relied upon to sustain claims that a pumpjet is generally acoustically superior to a propeller, including at low speeds.
+
+It should be noted here that shielding effects might comprise the combination two distinct and seperate physically phenomenon: reflection and damping.
+Noise tends to be reflected very effectively off hard, smooth surfaces representing the boundary between mediums of very different density (such as steel and water).
+Changes in density which are gradual, often representing soft, squishy, or broken surfaces tend, to have the effect of absorbing sound, where the acoustic energy is converted to heat or degradation of the material, rather than reflecting it.
+Given the significance of increased drag on any surfaces within the propulsor, it is highly unlikely that any attempt at damping is made within the pumpjet.
+However, it should be noted that some pumpjets have been constructed with a hollow design, in order to reduce weight \parencite{harvie1965construction}, which would reduce the thickness of solid steel, though offer some potential opportunity for damping.
+Overall, damping is increased in its effectiveness at low frequencies by allowing it to occur over larger distances, corresponding to the sound's increased wavelenght, leading to increased thickness of the transition substance or damping material.
+Consequently, effective damping within the propulsor itself is unlikely to be attempted, or limited to medium and higher frequencies, probably suited to reducing noise from normal surface flow turbulence and very early onset cavitation.
+
+The ability of a thin layer of material to reflect noise is also highly frequecy dependent, and also dependent on mass of the shield layer.
+As the material becomes thinner and thinner, a hard shield layer will tend increasinly to transmit rather than reflect the noise, a phenomenon that increases as the amplitude (loudness) of the noise increases, and frequency decreases.
+Denser materials are more effective at reflecting sound, as more energy (pressure) is required to excite the same movement in the material.
+Consequently, the effectiveness of a shield depends on both its thickness and density.
+There is therefore some tension inherent in the design of a shield where thickness and total weight are constrained.
+A thicker layer solid steel will be very effective at reflecting sound, but have negligible absoption or damping.
+A hollow between two thin layers will offer some potential for a soft daping material to be inserted, but at the cost of reflection.
+
+It is most likely that little or no damping or absorbing layer is attempted within the body of the shroud, since increasing the duct thickness above hydrodynamically determined optimum profile is likely to have a detrimental impact on efficiency.
+Constructing especially thin layers, and still maintaining an optimal shape to the precision required for housing powerful turbomachinery like the impeller is also likely to involve considerable construction risk and complexity.
+Hence the shielding effect is most likely to dominated by the reflective characteristics of the duct, which would seek to internally reflect noise back into the turbulent flow, where sound waves would be increasingly convoluted, and somewhat absorbed through the eventual transition to small-scale turbulence and heat.
+
+The degree to which the total noise generated is reflected internally will also be shape-dependent.
+A more concave inner surface, consistent with higher degrees of diffusion in advance of the impeller, and greater nozzle narrowing after it, which is consistent with higher pressure elevation at the impeller (hence improved cavitation characteristics) but higher negative duct thrust, leading to even poorer efficiencies at low speed.
+Substantial alteration of duct shape on account of efficiency considerations at low speed are therefore necessarily going to have some impact on the degree of effective shielding that a duct can afford.
+
+\subsubsection{Directional Dependence}
+Two important qualifications must be made regarding shielding.
+The first is that its impact is highly directionally dependent.
+The pump jet requires openings at the front and back to allow water in and out.
+Only noise that is reflected internally sufficient to be largely absorbed is effectively shielded.
+The larger the areas of the intakes and outlets, the narrower the effective shielding will become.
+Consequently, design changes to make the pump more efficient at low speeds, which may include expanding both the inlet and outlent, and diminshing the difference between them, will necessarily be at the expense of diminishing effective width of the shielding 'belt' around the propulsor.
+
+When the submarine is travelling at high speeds, and there is significant flow through the propulsor, the impact of waterspeed through the propulsor might have the additional impact of diminishing the radiation of noise in the forwards direction, in the same way that sound tends tends not to travel well up-wind in air.
+However, at very low speeds this effect would be greatly diminished.
+
+It should be noted that in the underwater environment sound can be made to reflect in a variety of directions simply with differences in temperature, salinity, movement, as well as off plenty of nearby underwater objects, including the ocean floor and objects on it.
+This phenomenon generally serves to make the art of underwater acoustics extremely complex and unpredictable, and generally contributes to the difficulty of precisely locating a submarine, even when some signature is detected.
+However, it also means that direction-dependent reductions signature do not reliably translate to overall reduced probability of detection.
+
+\subsubsection{Frequency Dependence}
+It ought further to be noted that the effectiveness of acoustic shielding is also highly dependent on frequency.
+Much lower frequencies are very hard to reflect, and absorb, as can be easily observed in a house with a home-cinema, where lower frequencies emitted by the subwoofer will tend to affect many surrounding rooms, where higher frequency sounds might be substantially muffled by interceding walls.
+The same acoustic effects are at work in underwater acoustics.
+
+This has an important consequence for shielding for pumpjets at very low speeds because of the occurrence of flow separation and vortex shedding.
+In the absence of flow separation, and cavitation at very low speeds the most substantial sources of sound would be the turbulent flow over surfaces of in the propulsor.
+This noise will tend to be relatively white and high in its frequency owing to the very small length-scales of the eddies and vortices in such turbulence.
+The steel shroud would be very effective in reflecting such high frequency noise.
+Under acceleration or higher loads, the very instances of cavitation woud also be relatively high frequency, as the voids or cavities would be small at first.
+These sounds might also be very effectively reflected by the shroud, and consequently diminished.
+However, flow separation results in the creation of vortices of a dramatically larger scale than occurs with normal turbulence in a boundary later, and hence generate noise of a much lower frequency.
+In addition, as the flow rates slow even further, these vortices are likely to become larger in size, even if lesser in intensity, resulting in further reduction in the frequency of noise generated.
+Further to this effect, the rate at which the vortices shed from the surface and impact the impeller will also result in another acoustic signature, of even lower frequency.
+Again, at lower flow rates, these vortices will tend to develop and shed more slowly, lowering frequencies further.
+
+Consequently, whilst for certain types of noise which tend to become both louder and lower in frequency with increasing speed (cavitation), flow separation provides a contrasting characteristic, in that it is likely to decrease in frequecy with decreasing speed.
+As lower frequencies tend to propagate much further in open water, this could have a significand adverse impact on the total acoustic signature of the submarine, even if the flow separation doesn't lead to any cavitation.
+
+The foregoing consideration suggests that the acoustic signature of a jet-powered submarine might not reduce steadily with speed all the way to stationary movement.
+It is plausible that there could be some elevated propulsion signature at very low speeds due to flow separation, and that the submarine's optimal acoustic signature when moving is actually at some slightly higher speed, where smoother flow through the propulsor can be achieved.
+However, given that the amplitude, frequency, and combined impact of those two characteristics cannot be accurately evaluated without substantial further work and more detailed system knowledge, it is also possible that the increase in other noise (including hull flow noise and other machinery) at elevated speeds offsets the improvement in the smoothness of the pumpjet's operation.
+It is also possible that the overall level of noise radiated by the pumpjet, including under conditions when flow separation occurs, is still considered to be sufficiently low as to not be of significant tactical consequence.
+However it is highly unlikely that the accoustic signature of the pumpjet is actually superior to that of an open propeller at very low speeds, when cavitation occurs for neither system, but substantially increased turbulence including flow separation must neccessarily occur for the pumpjet.
 
 \section{Conclusions}
+This is a pretty bad result for the pumpjet.
 

### 2018-01-06 DraftExec.tex
more on acoustics

@@
 Whilst at very low power levels, or at good depths, it is quite plausible that this doesn't result in any cavitation, which would normally be a likely occurrence if the blades are operating closer to their design power and such such vortices were shed into the intake.
-
-\section{App Outputs}
-
-Adjustable figures from App.
-
-\begin{knitrout}
-\definecolor{shadecolor}{rgb}{0.969, 0.969, 0.969}\color{fgcolor}\begin{kframe}
-\begin{verbatim}
-## # A tibble: 6 x 4
-##     kts   eff.jet  eff.prop hotel
-##   <dbl>     <dbl>     <dbl> <dbl>
-## 1   0.5 0.1728660 0.8902062   150
-## 2   1.0 0.2052115 0.8669509   150
-## 3   1.5 0.2356997 0.8463204   150
-## 4   2.0 0.2643992 0.8277603   150
-## 5   2.5 0.2913788 0.8108753   150
-## 6   3.0 0.3167071 0.7953734   150
-## [1] 2.5
-## [1] 150
-## # A tibble: 6 x 3
-##     kts hotel       eff
-##   <dbl> <dbl>     <dbl>
-## 1   0.5   150 0.1728660
-## 2   1.0   150 0.2052115
-## 3   1.5   150 0.2356997
-## 4   2.0   150 0.2643992
-## 5   2.5   150 0.2913788
-## 6   3.0   150 0.3167071
-## [1] 5
-##         5 
-## 0.2913788
-\end{verbatim}
-\end{kframe}
-\end{knitrout}
-
-\begin{figure}
-\begin{knitrout}
-\definecolor{shadecolor}{rgb}{0.969, 0.969, 0.969}\color{fgcolor}
-
-{\centering \includegraphics[width=\maxwidth]{figures/plots-plot_end-1}
-
-}
-
-\end{knitrout}
-\caption{Endurance}
-\label{fig:Endurance_plot}
-\end{figure}
-
-\begin{figure}
-\begin{knitrout}
-\definecolor{shadecolor}{rgb}{0.969, 0.969, 0.969}\color{fgcolor}
-
-{\centering \includegraphics[width=\maxwidth]{figures/plots-plot_range-1}
-
-}
-
-\end{knitrout}
-\caption{Range}
-\label{fig:Range_plot}
-\end{figure}
-
-\begin{figure}
-\begin{knitrout}
-\definecolor{shadecolor}{rgb}{0.969, 0.969, 0.969}\color{fgcolor}\begin{kframe}
-\begin{verbatim}
-## NULL
-\end{verbatim}
-\end{kframe}
-\end{knitrout}
-\caption{Power}
-\label{fig:Power_plot}
-\end{figure}
+However, given the inherently unpredictable and unsteady nation of such vortex formation and shedding, the this phenomenon would increase the risk of cavitation occurring unexpectedly in marginal conditions.
+In addition, in the absence of cavitation, the sudden and changing pressure gradients over the blades would dramatically increase the overall pressure fluctuations and turbulence level that occurs within the propulsor.
+This would necessarily increase the overall levels of noise generated.
+It is plausible that the noise is of sufficiently small level that it is not considered a significant compromise of stealth.
+In essence, the higher levels of noise lie below some threshold of significance.
+However, it is almost certainly untrue that that pumpjet actually generates less noise overall than a propeller at very low speeds.
+
+\subsection{The impact of duct shielding}
+The presence of the duct around the impeller may also have the effect of preventing the noise from being effectively radiated away from the propulsor.
+This may well result in substantial reductions in some types of noise being radiated away from the propulsor in certain directions, which may have given rise to the consistent qualification of 'radiated' being used to descrive the type of noise which differentiated the French submarine bid \parencite{stewart2016, davies2017, ohff2016}, including by the designers of the submarine \parencite{autret2016}.
+This is a secondary effect, quite separate to the avoidance of cavitation, which might be seen as an acoustic advantage in circumstances where cavitation might not be expected to occur on either type of propulsor.
+It is highly likely that this is the effect which is relied upon to sustain claims that a pumpjet is generally acoustically superior to a propeller, including at low speeds.
+
+It should be noted here that shielding effects might comprise the combination two distinct and seperate physically phenomenon: reflection and damping.
+Noise tends to be reflected very effectively off hard, smooth surfaces representing the boundary between mediums of very different density (such as steel and water).
+Changes in density which are gradual, often representing soft, squishy, or broken surfaces tend, to have the effect of absorbing sound, where the acoustic energy is converted to heat or degradation of the material, rather than reflecting it.
+Given the significance of increased drag on any surfaces within the propulsor, it is highly unlikely that any attempt at damping is made within the pumpjet.
+However, it should be noted that some pumpjets have been constructed with a hollow design, in order to reduce weight \parencite{harvie1965construction}, which would reduce the thickness of solid steel, though offer some potential opportunity for damping.
+Overall, damping is increased in its effectiveness at low frequencies by allowing it to occur over larger distances, corresponding to the sound's increased wavelenght, leading to increased thickness of the transition substance or damping material.
+Consequently, effective damping within the propulsor itself is unlikely to be attempted, or limited to medium and higher frequencies, probably suited to reducing noise from normal surface flow turbulence and very early onset cavitation.
+
+The ability of a thin layer of material to reflect noise is also highly frequecy dependent, and also dependent on mass of the shield layer.
+As the material becomes thinner and thinner, a hard shield layer will tend increasinly to transmit rather than reflect the noise, a phenomenon that increases as the amplitude (loudness) of the noise increases, and frequency decreases.
+Denser materials are more effective at reflecting sound, as more energy (pressure) is required to excite the same movement in the material.
+Consequently, the effectiveness of a shield depends on both its thickness and density.
+There is therefore some tension inherent in the design of a shield where thickness and total weight are constrained.
+A thicker layer solid steel will be very effective at reflecting sound, but have negligible absoption or damping.
+A hollow between two thin layers will offer some potential for a soft daping material to be inserted, but at the cost of reflection.
+
+It is most likely that little or no damping or absorbing layer is attempted within the body of the shroud, since increasing the duct thickness above hydrodynamically determined optimum profile is likely to have a detrimental impact on efficiency.
+Constructing especially thin layers, and still maintaining an optimal shape to the precision required for housing powerful turbomachinery like the impeller is also likely to involve considerable construction risk and complexity.
+Hence the shielding effect is most likely to dominated by the reflective characteristics of the duct, which would seek to internally reflect noise back into the turbulent flow, where sound waves would be increasingly convoluted, and somewhat absorbed through the eventual transition to small-scale turbulence and heat.
+
+The degree to which the total noise generated is reflected internally will also be shape-dependent.
+A more concave inner surface, consistent with higher degrees of diffusion in advance of the impeller, and greater nozzle narrowing after it, which is consistent with higher pressure elevation at the impeller (hence improved cavitation characteristics) but higher negative duct thrust, leading to even poorer efficiencies at low speed.
+Substantial alteration of duct shape on account of efficiency considerations at low speed are therefore necessarily going to have some impact on the degree of effective shielding that a duct can afford.
+
+\subsubsection{Directional Dependence}
+Two important qualifications must be made regarding shielding.
+The first is that its impact is highly directionally dependent.
+The pump jet requires openings at the front and back to allow water in and out.
+Only noise that is reflected internally sufficient to be largely absorbed is effectively shielded.
+The larger the areas of the intakes and outlets, the narrower the effective shielding will become.
+Consequently, design changes to make the pump more efficient at low speeds, which may include expanding both the inlet and outlent, and diminshing the difference between them, will necessarily be at the expense of diminishing effective width of the shielding 'belt' around the propulsor.
+
+When the submarine is travelling at high speeds, and there is significant flow through the propulsor, the impact of waterspeed through the propulsor might have the additional impact of diminishing the radiation of noise in the forwards direction, in the same way that sound tends tends not to travel well up-wind in air.
+However, at very low speeds this effect would be greatly diminished.
+
+It should be noted that in the underwater environment sound can be made to reflect in a variety of directions simply with differences in temperature, salinity, movement, as well as off plenty of nearby underwater objects, including the ocean floor and objects on it.
+This phenomenon generally serves to make the art of underwater acoustics extremely complex and unpredictable, and generally contributes to the difficulty of precisely locating a submarine, even when some signature is detected.
+However, it also means that direction-dependent reductions signature do not reliably translate to overall reduced probability of detection.
+
+\subsubsection{Frequency Dependence}
+It ought further to be noted that the effectiveness of acoustic shielding is also highly dependent on frequency.
+Much lower frequencies are very hard to reflect, and absorb, as can be easily observed in a house with a home-cinema, where lower frequencies emitted by the subwoofer will tend to affect many surrounding rooms, where higher frequency sounds might be substantially muffled by interceding walls.
+The same acoustic effects are at work in underwater acoustics.
+
+This has an important consequence for shielding for pumpjets at very low speeds because of the occurrence of flow separation and vortex shedding.
+In the absence of flow separation, and cavitation at very low speeds the most substantial sources of sound would be the turbulent flow over surfaces of in the propulsor.
+This noise will tend to be relatively white and high in its frequency owing to the very small length-scales of the eddies and vortices in such turbulence.
+The steel shroud would be very effective in reflecting such high frequency noise.
+Under acceleration or higher loads, the very instances of cavitation woud also be relatively high frequency, as the voids or cavities would be small at first.
+These sounds might also be very effectively reflected by the shroud, and consequently diminished.
+However, flow separation results in the creation of vortices of a dramatically larger scale than occurs with normal turbulence in a boundary later, and hence generate noise of a much lower frequency.
+In addition, as the flow rates slow even further, these vortices are likely to become larger in size, even if lesser in intensity, resulting in further reduction in the frequency of noise generated.
+Further to this effect, the rate at which the vortices shed from the surface and impact the impeller will also result in another acoustic signature, of even lower frequency.
+Again, at lower flow rates, these vortices will tend to develop and shed more slowly, lowering frequencies further.
+
+Consequently, whilst for certain types of noise which tend to become both louder and lower in frequency with increasing speed (cavitation), flow separation provides a contrasting characteristic, in that it is likely to decrease in frequecy with decreasing speed.
+As lower frequencies tend to propagate much further in open water, this could have a significand adverse impact on the total acoustic signature of the submarine, even if the flow separation doesn't lead to any cavitation.
+
+The foregoing consideration suggests that the acoustic signature of a jet-powered submarine might not reduce steadily with speed all the way to stationary movement.
+It is plausible that there could be some elevated propulsion signature at very low speeds due to flow separation, and that the submarine's optimal acoustic signature when moving is actually at some slightly higher speed, where smoother flow through the propulsor can be achieved.
+However, given that the amplitude, frequency, and combined impact of those two characteristics cannot be accurately evaluated without substantial further work and more detailed system knowledge, it is also possible that the increase in other noise (including hull flow noise and other machinery) at elevated speeds offsets the improvement in the smoothness of the pumpjet's operation.
+It is also possible that the overall level of noise radiated by the pumpjet, including under conditions when flow separation occurs, is still considered to be sufficiently low as to not be of significant tactical consequence.
+However it is highly unlikely that the accoustic signature of the pumpjet is actually superior to that of an open propeller at very low speeds, when cavitation occurs for neither system, but substantially increased turbulence including flow separation must neccessarily occur for the pumpjet.
 
 \section{Conclusions}
+This is a pretty bad result for the pumpjet.
 

### 2018-01-07 DraftExec.Rnw
something going wrong... committing my work so far.

@@
 Hence, in this model we have assumed that 20kt will be obtained with the ouput of 6.5MW.
+
+Considerable confirmation that these assumptions are realistic can be made by assessing the endurance of the submarines under the model with other known reference points.
+For the central curves, we obtain a dived endurance betwen 60 and 100 hours for both systems at 5kt, reflecting nearly exactly the range of other public estimates of dived enurances endurances which other modern diesel-electric submarines operating only with batteries \parencite[3]{article{buckingham2008submarine} at around 4kt.
+Given that somewhat larger submarines might be somewhat assisted in this respect because their overall volume/surface ratio is slightly higher, this gives confirmation that the model is overall quite realistic in its magnitudes at the most important speeds.
+The sames sources also suggests that the same submarines achieve endurances of around 1-2 hours at 20kt, which also confirms that the relationship in our model linking differnt speeds is also in agreement with other models.
 
@@
 \subsection{Sensitivity to Hotel Load}
+As can be seen in Figure \ref{fig:sens_hotel} the difference between the two systems is very sensitive to changes in hotel load, both in proportional and nominal terms.
+The impact of halving the hotel load from 100kw to 50kw more than doubles the **difference** between a pumpjet and propeller dived endurances, from around two days to well over four days at speeds between 2.5 and 3.5kt.
+In the context of a submarine which might have a nominal dived requirement at patrol speeds in the order of several days, this suggests that extremely large gains can be made by reducing hotel load.
+Increasing the hotel load has the noticeable effect of pushing the speed of largest difference to higher speeds, and substantially reducing the total difference.
+If the hotel load was doubled to 200kW, the middle efficiency curve pair would still see a difference of nearly a day between the two systems, or around 19.5 hours at speeds of 4.5-5 kts.
+
+The difference between pumpjet and propeller propulsion systems in terms of range tends to be found at slightly higher speeds, though this point also moves towards higher speeds as hotel load increases.
+With a reduced hotel load of 50kw, the difference between the two systems in the middle scenario is about 370nm for speeds 3.5-4.5kt, substantially greater than the ~200nm around 4.5-5.5kt at a hotel load of 100kW.
+
+It's important to acknowledge that such substantial differences in hotel load merit serious consideration, both in terms system design, as well as for variance in operation of a submarine with a given system design.
+As has been discussed, the precise hotel load of submarines is not generally publically available, which justifies senitivity testing around a relatively wide potential range.
+The size of the crew required for different designs might vary considerably, as it also would for different missions (for example with the embarkation of a special forces congingent) which would necessarily have considerable impact on energy demand for life support, including sanitation, CO2 scrubbing, and cooking.
+In addition, it broadly understood that the AN/BYG-1 combat system consumes substantially more power than other alternatives, reportedly in the order of 100kW \parencite{patrick2011}.
+
+Furthermore, the hotel load may be varied considerably at the captain's discretion depending on operational requirements.
+Not all sensors or components of the combat system need to run at all times, for example during long transits in order to reach an operational area, or during a long dive in order to escape an operational area once a large strike has already been conducted.
+In wartime, when surfacing to snort within range of enemy airpower or close to naval bases might result in the sumbarine probably being sunk, the crew might willingly forego hot showers, hot meals, and climate-controlled air in order to increase their chances of survival.
+Consequently, it is perfectly plausible that changes of hotel load in the order of a factor of two for tactically significant time periods (days) achievable.
+The results of this sensitivity analysis show that the penalty inflicted by the pumpjet will be substantially greater in such circumstances.
+
+The proportional impact of the changes in hotel load area also substantial, and exhibit clearly the same trend of the impact being larger at higher speeds for higher hotel loads.
+For the efficiency curve pairs that comprise the central selections, decreasing the hotel load to 50kt would see the dived endurance and range of the submarine improved by the adoption of propellers 80\% between 4 and 5.5kt.
+A much higher hotel load, at 200kW, would see that peak difference over 40\% between 5.5 and 8kt.
+
+It is also clear that increasing the hotel load has a greatly diminised impact on the proportional differences between the systems at higher speeds.
+At very high speeds (e.g.
+15kt) changes in the hotel load have almost no impact in proportional terms to the difference, which lies at about 30\% in all cases for the central selection of curves.
+At 9kt, in the region where transit speed is likely to occur, increasing the hotel load by a factor of four (from 50kW to 200kW) only decreases the advantage of the propeller from 48\% to 38\%.
+Therefore one might say near certainty that any difference between efficiency around these patrol speeds is likely to have a substantial impact on the overall range of the submarine, with differences of much less than 30\% quite unlikely, irrespective of other changes in the submarine design or operation circumstance.
+
+It should also be noted that sensitivity testing has not been shown in this paper around alternative values of the drag coefficient $C_d$ which was solved for by matching peak power output and top speed.
+It is considered relatively unlikely that such dramatic changes in this constant are plausible.
+Furthermore, changing this constant will have a very similar impact to alerations in hotel load, in terms of its affect on the speed where greatest differences will occur and their probable magnitude.
+Consequently, consideration of the sensitivity to hotel load is likely to easily encompass the plausible extent, and nature, of the likely uncertainty in this factor.
 
@@
 df %>%
-  filter(test == "hotel") %>%
+  filter(test == "endurance") %>%
   ggplot(aes(x = kts, y = val, col = pair))+
@@
 \subsection{Sensitivity to Battery Changes}
+
+Changes in the effective battery energy density will have the same effect as having larger masses of battery stored.
+They simply directly impact the total amount of energy available for consumption before recharging batteries while snorkeling.
+There is therefore no shift in the proportional impact on range and endurance due to this factor, as can be seen in the lower graphs in Figure \ref{fig:sens_battery}.
+The nominal impact, however increases directly and linearly with the total energy stored.
+
+The results in nominal terms are striking in their magnitude.
+In terms of enduracy, doubling the stored electrical energy, for example by adopting lithium batteries, increase the **difference** between the two systems under the central efficiency curve assumptions to nearly four days (93 hours) at 4kt.
+Given that most modern, capable submarines are estimated to only be able to achieve dived endurances between 60 and 100 hours on batteries at 4kt \parencite{article{buckingham2008submarine} this cannot be dismissed in terms of its tactical significance.
+Even under the eficiency curve assumptions representing the narrowest spearation considered, the difference in this case would be 59 hours, almost equal to the estimated dived endurance of a Type214 operating without AIP.
+
+The in terms of submerged range, the same proportional impact is also extremely significant.
+Doubling the energy stored inceases the **difference** between two otherwise identical submarines to over 400nm under central assumptions at 4.5kt.
+That's a larger distance than the width of the Java Sea, the most central body of water in the Indonesian Archipelago, or most of the length of the Malacca Strait.
+The tactical significance of such distances are not likely to be lightly dismissed.
+If far smaller amounts of energy are embarked, the nominal differences reduce to perhaps 100nm, in direct proportion to the reduction of the submarine's dived endurance.
+This is still a non-trivial distance in terms of it's tactical signicance, and comprises a substantial fraction of the dived ranges of most non-AIP submarines.
+It still represents a significant difference in operational terms, equal approximately to the width of Hainan Island, or neraly twice the length of the Sunda Strait.
+
+The singificant range of different energy density values tested here reflect quite plausibly some of ranges of different energy levels available in operation.
+For instance, as has been discussed \parencite{briggs2016, greenfield2016}, lead acid batteries may well only be able to practically utilise about 50\% of their total energy stores for tactical operaions, unless operating in a very low-threat environment and undertaking very slow charging to reach the maximum full charge is possible.
+In addition, the plausible improvement in enery density of lithium batteries may well be in the order of a factor of two in practice \parencite{greenfield2016}, though theoretical limits of the chemistry suggest much higher may be possible \parencite{davies2016}.
+
+Only brief is given here to the impact of Air Independent Propulsion, which has substantially higher energy densities possible even than Lithium, and also amounts to the addtional embarkation of greater mass.
+The volumetric energy density of the feedstocks to modern AIP systems (liquid oxygen and methanol) is estimated to be approximately ten times greater than that of lead-acid batteries \parencite{davies2016}, around 4.5MJ/L.
+Given the greater mass embarked, and additional studies suggesting that at constant speeds they might extend the dived endurance of a modern submarine from around 60 hours to over 1000 hours (a factor of roughly sixteen) an approximate estimation of the the direction and magnitude of this impact may by scaling up energy density by this factor.
+This would amount to adding three further frames to the right of Figure \ref{fig:sens_battery}, with the energy density being doubled in each successive frame.
+This would plausibly (under the central line selections) amount to more than month of dived endurance **difference** between AIP inclusive submarines with propellers or pumpjets, amounting to thousands of nautical miles of continuously submerged travel.
+This difference would also now be material in terms of the overall range of the submarine amounting to a further 15-20\% reduction in the submarine's overall range.
+
+With such massive differences between the performance of the two systems emerging with the presence of the AIP, it is difficult to see how one could realistically consider a pumpjet as being competitive if a direct comparison were made.
 

### 2018-01-07 DraftExec.Rnw
something's not working again.  Just get "missing $ inserted." no line ref.  Aargh

@@
 In stark contrast, a submarine operator might have less than 1\% of the energy store available for the submerged movements at 2-3kt, and consequently be substantially \textbf{more} concerned about efficiency in this regime, particularly since this would be the regime in which all of their combat operations would be conducted.
-%
+
 This imposes a constraint on the way in which literature on jets, propellers, and propulsion systems in general is read, since almost all of it is deeply concerned with how a particular system would operate at one particular design speed, and choosing the optimum system for that design speed and load.
@@
 
+\subsection{Power and Energy}
 Nuclear submarines vary quite remarkably from conventional submarines because of the means by which they generate their power.
@@
 In contrast, the Collins Class Submarine's main motor is rated at less than 6MW, with designs for the future submarine appearing to be only slightly larger \parencite{patrick2012}.
-It is fair to say that in terms of maximum power ouput, nuclear submarines could have something like 5-10 times as much power at their disposal, and their \textbf{peak} output.
+It is fair to say that in terms of maximum power ouput, nuclear submarines could have something like 5-10 times as much power at their disposal, and their peak output.
 
@@
 
+\subsection{Xenon poisoning and low power limitations}
 Perhaps more significantly for power-design of nuclear submarines is a little-discussed phenomenon called Xenon poisoning, which affects nuclear reactors when they shut down or lower their power significantly.
@@
 
+\subsection{Recent Commentary}
+Some efforts have been made to 'debunk' such suggestions that nuclear submarines effectively have "unlimited" power \parencite{autret2016}, with the leaders of the DCNS's bid at the time making some effort to suggest that the needs and characteristics of nuclear and diesel-electric submarines "drives the same system design".
+They make three comments to justify this claim.
+The first is that the large submarine proposed by DCNS uses as much or more power for the hotel load at the most frequently used speeds.
+Assuming the authors are referring to speeds between patrol speeds and approaching transit speed, the subsequent analysis will show that this fact may be true, particulalry in the presence of the pumpjet, but on its own this doesn't constitute an argument worthy of any consideration.
+One might equally argue a glutton can be relied upon not to eat too much at dinner because he eats a dinner sized meal at afternoon tea.
+
+The authors of this piece also suggest that the high cost and weight of the nuclear reactor contributes to the desire to minimise hotel load.
+Given the way that the square law drives drag as discussed earlier, driving a cube law for power, there is no chance at all that the hotel load has any bearing on the reactor power required.
+The last few knots of top speed would demand ten times over the power required for hotel load, as can be seen in Figure \ref{fig:power_curve}.
+
+One more plausible argument is worth consideration, namely that a safety requirement to be able to operate on batteries without the reactor drives a need to minimise hotel load.
+For all the reasons discussed above, it's completely implausible that the reactor would ever be shut down volunarily during operations.
+One would think that the sorts of emergencies or accidents which would drive this are likely to also mean that the reactor can't be restarted.
+However, it is plausible that through the development of some instability, perhaps through an unknown fault or dramatic power reductions of the type which I described should be avoided, it might be the case that the reactor would need to be shut down, but could plausibly be restarted.
+However, as discussed above, the restart would need to occur either within minutes, or probably in a couple of days.
+
+This puts some serious constraint around type of contingency that needs to be planned for in a reactor shutdown event.
+It might be presumed that surfacing and remaining surfaced for a couple of days is likely to end combat operations for a submarine that was located in theatre.
+However, remaining submerged and continuing combat operations, even at low speeds will essentially necessitate replicating the full power demand of any diesel-electric submarine until such a time as the reactor can be restarted, especially since hotel load, in the case of the submarine they propose, amounts to more than propulsion at likely patrol speeds.
+As we can gather from other public estimates of submarine submerged endurance \parencite{buckingham2008submarine}, roughly a couple of days is a very significant fraction of the entire dived endurance of a diesel-electric submarine, perhaps 50\%.
+It is difficult to imagine that nuclear submarines carry nearly 50\% of the battery load of diesel-electric submarines, in addition to the reactor.
+
+Consequently, it seems much more likely that the sort of battery-requirements that are necessary for this kind of safety system are not closely linked to a full operational hotel load, but some substantially reduced survival load.
+As discussed in later sections, many of the most energy intensive systems in the hotel load that could readily be switched off in an emergency to save power, if saving power was what was required to get through the emergency, including the combat system.
+It is difficult to imagine what sort of emergency might require a reactor to be shut down, that wouldn't also justify shutting down essentially every other component of normal hotel load which could plausibly be delayed for a couple of days.
+
+If it was known that the reactor couldn't be restarted after the poison-out period, the mission need to be aborted in any case, and some kind of rescue sought.
+Even a hundred of tonnes of batteries embarked wouldn't change this reality, unless the submarine literally at the entrance of a home port.
+It seems much more likely that the safety requirement for batteries on a nuclear submarine is most likely to be framed around maintaining key life support systems (CO2 scrubbing for example) for sufficient time to enable some hope for a seach/rescue mission to rescure the crew, in the likely case that the kind of emergency which shut down the reactor was the kind of emergency which meant the submarine might not resurface, such as the case of the Russian nuclear submarine, the Kursk.
+
 \section{Some essential concepts}
@@
 
-Conservation of energy has a particularly relevant embodiment in fluid flows, which is given it's primary expression in Bernoullis equation.
-It says that the total energy in a connected body of fluid is constant, though it can change in form between kinetic energy (movement), gravitational potential energy (it being elevated) the heat energy in the fluid, and the pressure of the fluid.
+Conservation of energy has a particularly relevant embodiment in fluid flows, which is given it's primary expression in Bernoullis equation, as given in Equation \ref{eq:bernoullis}.
+It says that the total energy in a connected body of fluid is constant ($C$), though it can change in form between kinetic energy (movement, $\frac{1}{2}\rho v^2$), gravitational potential energy (it being elevated, $\rho gz$) the heat energy in the fluid (not shown, not relevant for incompbressible fluids), and the pressure of the fluid $p$.
+
+\begin{equation}
+\label{eq:bernoullis}
+\frac{1}{2}\rho v^2 + \rho gz + p = C
+\end{equation}
 
@@
 Since the jet in such vessels is generally ejected at the same height as the pump, this potential is never regained, and is technically a loss, however at high speeds such a loss is small relative to the total power ouput.
+In the case of submarine pumpjets, located on the end of the hull and parallel to the axis of the submarine, this effect (and the term $gz$) are not relevant.
 
@@
 In the case of submarines and torpedoes the water doesn't undertake a change in height, since intake and nozzle are generally all in line with the central axis of the submarine or torpedo.
-As a consequence, the key relationship in Bernoullis equations is the relationship between the liquids velocity, and its pressure.
+As a consequence, the key relationship in Bernoullis equation is the relationship between the liquids velocity, and its pressure.
 The consequence is that when water moves through a pipe (duct, or shroud) it's pressure is inversely related to the square of its velocity.
@@
 
+Flow separation tends to occur earlier for laminar boundary layers than for fully turbulent boundary layers.
+Consequently, some systems are able to become more efficient  by deliberately inducing turbulent, where the benefit from reduced friction in the laminar boundary layer is outweighed by the cost suction induced by the larger wake field created by earlier flow separation.
+This is the reason that golf balls have dimples, and tennis-balls have fur.
+
 \section{Basics of Ducted Propellers and Pumpjets}
@@
 \subsection{Modern literature and results}
-In addition, more modern works, including those which involve experimental validation of theoretical results also tend to show that restrictions on the efficiency of waterjets are enduring, and give us a clear starting-point for consideration of their probable orders of magnitude, and likely transition points indicating at what speeds pumpjets might become advantageously more efficient.
+In addition, more modern works, including those which involve experimental validation of theoretical results also clearly confirm that restrictions on the efficiency of waterjets are enduring, and give us a clear starting-point for consideration of their probable orders of magnitude, and likely transition points indicating at what speeds pumpjets might become advantageously more efficient.
+It should be noted however, that relatively little literature focuses on the performance of propulsive systems far far outside of their intended design speeds, which tends to be far higher than the patrol and transit speeds of diesel electric submarines.
+Fewer still tend to express their work in terms of ship-speed, instead adopting the convetion of using dimensionless ratios so that work done on one system can be readily translated into different combinations loads, pressures, speeds and sizes.
+However, from a few notable studies exist which provide important reference points.
+(Further interpretation of the what we can lean from the trends expressed in dimensionless coefficients will be discussed later.)
 
@@
 
-\subsection{Theoretical Limitations}
-In this section I will appraise some of the available literature, including some of the most canonical works in the design of pumpjets which is now declassified from US Navy funded research (and still extremely frequently cited by contemporary researchers).
-Before doing so, I'll attempt to summarise in simple language the mose essential conclusions in simple language.
-
-My broad conclusion is that the optimisation of a system for minimising cavitation, and hence improving acoustic performance, is a separate and independent design requirement to that of optimising a system for propulsive efficiency.
+\subsection{A simple theoretical explanation}
+
+Pumpjets are necessarily low in efficiency at very low speeds because they attempt to get the water to do multiple things.
+They don't just accelerate a water colum.
+They first slow it down (increasing its pressure), then add pressure to it, then accelerate it.
+When the total amount of energy that the incoming water has, and the total amount of energy you want to add to the water both get very small (as they do at very low speed) there simply isn't enough energy to do all things at once to all of the water.
+The energy penalty inflicted by using the additional devices becomes very large in proportion, and part of the water doesn't do all the necessary things for the jet to work.
+In essence, if you try to slow down something that's already going slowly, it tends to stop.
+That's exactly what happens in jets operating at very low flow-rates, when flow separation, also known as stall when it occurs between the blades, takes place.
+
+There are other some fundamental limitations on the peak efficiency that can be achieved by jets based on simple calculations of ideal efficiency from momentum theory, which are discussed in the next subsection.
+They also provide enduring, reliable limitations on the extent to which ingenuity or design changes might be able to improve the efficiency of jets at low speed.
+the the primary reason why jet efficiencies decline towards zero at low speed is to do with flow separation, which is what occurs when the assumptions inherent in momentum theory's analysis of an 'ideal' actuator disc start to break down.
+
+The conditions which lead to flow separation in a pumpjet can be relatively simply explained in terms of conservation of energy, and conservation of mass.
+Conservation of mass requires that as a pipe or duct expands, the flow through it necessarily slows down, in order to occupy the large volume, in proportion to the new cross-sectional area.
+Conservation of energy requires that it increases in pressure, and defines a very specific relationship between the pressure increase and the change in velocity, which is determined by the shape of the duct.
+That relationship says that the static pressure increase is inversely proportional change in velocity **squared**, as given by Bernoulli's equation, which was discussed in earlier sections.
+As a result, at high speeds, the total energy of the system is dominated by its velocity, but at very low speeds the static pressure is much more significant.
+
+It also means that for any given change in velocity, the static pressure changes also, but doesn't change as much as the velocity change.
+The consequences for this pumpjets are and inevitable, since the geometry of the duct is inevitably fixed.
+At very high flow rates, the expansion of the area ahead of the impeller creates an adverse pressure gradient, as the water must slow down to fill the greater area.
+This adverse pressure gradient is larger at higher speeds, the proportions aren't linear, hence the the speed is much much larger, due to the square relationship.
+
+At low speeds, the same effect leads to detrimental effects.
+The adverse pressure gradient is much reduced, but the velocity is much much $(much)^2$ reduced.
+The pressure gradient is necessarily constant across the entire area radial area of the duct, but the drag due to the boundary with the duct itself will necessarily be greater closer to the walls in the case of the intake of the duct.
+For this simple reason, as the flow rate decreases, some point is reached where the flow adjacent to the walls actually reverses, and the flow becomes separated.
+
+The same phenomenon is also prone to occur near roots of the blades of the rotor.
+Necessarily, the blades of the rotor are doing work on the fluid, so unless the impeller is in a region of considerable contraction, pressure will necessarily have to increase through the blades, providing an adverse pressure gradient, which again must be constant radially.
+Since closer to the hub the velocity of the blads through the water is lower, the blades tend often to also be longer and deflect the flow even more in order to do the same amount of work on the fluid.
+With the longer surface over which the fluid must travel, as well as greater angle of deflection, cavitation is prone to occur here too.
+The physics will be exactly analogous to that of the separation on the inside of the duct intake, except the pressure gradient will be provided by a different source (the rotor doing work), and the surface (the blade roots) will also be different.
+This will also mean that the direction of rotation of the vortices will be different, and the location also close to the middle of the jet, instead of the outside.
+
+As can be seen in figure \ref{fig:JetHighFlow.png}, the duct of a jet is design to operate by deflecting the flow away from the natural path that would occur for a propeller, which would accelerate the stream over its blades directly, with the flow necessarily contracting with the acceleration as shown in Figure \ref{fig:PropellerAction.png}.
+Instead, pumpjets tend to use some moderate expansion of the flow in advance of the impeller (diffusion) in order to raise its pressure, and reduce its velocity.
+(Allowing the blades to do work on the water at lower speeds and higher pressure is the essence of what reduces cavitation.)  The blades then do their work on the liquid, further raising its pressure.
+The duct allows this pressure then to be converted back to kinetic energy, as the water accelerates to move through the contracting area in the nozzle.
+This difference in the shape of the flow, and the consequent impacts on the static pressures, and demonstrated in Figure \ref{fig:FlowComparison.png} taken from \cite{mccormick1963design}.
+
+\begin{figure}
+\includegraphics[width=\textwidth]{FlowComparison.png}
+\caption{The effect of the duct is to deflect the flow away from the streamlines of an open propeller, elevating the static pressure at the stator blades by slowing the flow down.}
+\label{fig:FlowComparison.png}
+\end{figure}
+
+Consequently, the geometry of the duct defines a pressure gradient in the intake and through the nozzle.
+The greater the difference in the cross-sectional areas, the greater the total magintude of the pressure differential that will occur, as necessitated by conservation of mass.
+The same amount must go out as came in.
+As the area through which it passes changes, the velocity much change to that the flow rate remains constant, otherwise cavitation will occur.
+The rate at which the areas change (how rapidly they expand and contract along the axis) will determine the gradient.
+Making the pressure change as smooth as possible is an important aspect to jet design in order to minimise flow separation.
+
+\begin{figure}
+\includegraphics[width=\textwidth]{JetHighFlow.png}
+\caption{At high flow rates the water has sufficient velocity (indicated by arrow length) to overcome the adverse pressure gradients created by the duct shape, shown by the change from green to red.
+After the impeller has added further pressure to the water, the water accelerates out of the contracting nozzle through the favourable pressure gradient (red to green) to produce a jet, reaching equal pressure to outside at the exit}
+\label{fig:JetHighFlow.png}
+\end{figure}
+
+Figure \ref{fig:JetLowFlow.png} shows a jet operating at very low speed.
+The adverse pressure gradient combined with the drag forces of the boundary layer are sufficient to bring the flow to a stop and reverse it, as shown by the arrow size reducing, and changing direction in the diagram.
+This occurs very close to the inner surface near the opening of the duct, and potentially also between the blades of the rotor, beginning the roots \parencite[15]{henderson1964}.
+Flow separation necessarily sees additional energy expended in the kinetic energy of the additional vortices and turbulent flows created, which don't result in the creation of thrust, which demands a change in the net momentum of the liquid in a backwards direction.
+Consequently, the degree and extent that it occurs necessarily and inevitably reduces the efficiency of the propulsor.
+As the flow rates slow further, the areas of flow separation will become larger, effectively reducing the intake area of the jet, and further degrading efficiency.
+
+Flow separation also tend to lead to unsteady flows, with vortices periodically being shed into the main stream as they are speed up or change shape, and then ingested by the impeller and stator blades.
+As can be seen by the diagram in Figure \ref{fig:JetLowFlow.png}, regions abound which are adjacent to or behind flow blockages or reversing flow, where considerable turbulence and unsteady forces would exist.
+These effects also have acoustic consequences, which will be discussed later.
+
+\begin{figure}
+\includegraphics[width=\textwidth]{JetLowFlow.png}
+\caption{At low flow rates the water has insufficient velocity to overcome the adverse pressure gradients created by the duct shape, and flow separation begins to occur on the duct intake, and between the blades beginning near the roots (stall).
+Flow through the jet would be unsteady, with excessive energy expended creating vortices and turbulence that results in little thrust.}
+\label{fig:JetLowFlow.png}
+\end{figure}
+
+It is worthwhile noting that the diagrams in Figures \ref{fig:JetHighFlow.png} and \ref{fig:JetLowFlow.png} do not include a hub for the sake of simplicity.
+In practice, the cross-sectional are of the pumpjet naturally only includes that aread between the hub and the duct, and in the case of many pumpjets, particularly those on the ends of torpedoes or submarines, the hub may include/incorporate the tapering tail of the body, for example in Figures \ref{fig:PumpjetOutline.png} or \ref{fig:PumpjetWaterjet.png}.
+Particularly in Figure \ref{fig:PumpjetWaterjet.png}, the appearance of the inner side of the shroud to roughly follow contraction of the hub or body gives the optical appearance of very little diffusion or contraction of the flow occurring, as the one dimensional separation might seem to be roughly constant, or even expanding in places where I have earlier argued it generally contracts.
+It is important to consider the impact of the second dimension on the actual effective area.
+Since area of a disc follows $\pi r^2$, the impact of the contraction of the hub will be dramatically larger when it is large, than it will be when it is small.
+
+Consequently, the contraction of the outer shroud at the very end of the duct will dominate the final taper of the hub, and actually lead to substantial restriction of the flow.
+Similarly, where the body or hub is quite substantial and also tapering steadily, keeping the duct intake parallel to the axis of the body for a small distance, or even just tapering it proportionally less than the hub, can still result in a considerable expansion in the area available, and considerable diffusion.
+The example in Figure \ref{fig:PumpjetOutline.png} shows a case of extremely sharp flow diffusion on the intake, probably more than what is practially realistics to avoid flow separation, even at quite high speeds.
+It is important to recognise that this simple relationship with $r^2$ generally means that even moderate changes in terms of diameter, which is what we tend to observe when looking at images and diagrams can actually amount to much more substantial changes in velocity, and hence pressure, which drives the effects described in this section.
+A side-by-side comparison of the geometries and flow-lines likely in the case of pumpjet or propeller is give in Figure \ref{fig:Comparison.png}, from  to help illustrate this point.
+
+\begin{figure}
+\includegraphics[width=\textwidth]{Comparison.png}
+\caption{At the end of a tapering body the impact of the contracting hub will be much larger near the intake than near the tail, making the degree of diffusion ahead of the impeller, as well as contraction of the flow in the nozzle after it, much larger in impact than it appears by looking at the external shape.
+Credit \cite{gearhart1966selection}}
+\label{fig:Comparison.png}
+\end{figure}
+
+\begin{figure}
+\includegraphics[width=\textwidth]{Triomph.jpg}
+\caption{The outer shape of pumpjets on the aft end of torpedoes or submarines, such as this one on the French Triomphant class on which the Barracuda's jet will be based, may understate the degree of deceleration that the duct achieves.
+Credit \cite{reddit2016}}
+\label{fig:Triomph.jpg}
+\end{figure}
+
+\subsection{Trade-offs and Limitations on Improvements}
+
+Having established that the inevitable decline of a pumpjets efficiency towards lower speeds is driven by the increasing occurrence of flow separation, some consideration ought to be given to the potential scope for the effect to be improved or negated by design changes.
+Here I have found considerable help from well established literature, many including various old declassified technical reports funded by the US Navy, many of which are frequently cited in the most recent literature as laying out the enduring frameworks for the modern discipline.
+My reading on the matter which reminds me strongly of an adage recently quoted in Senate Estimates by General Manager for Submarines, Stephen Johnson: "There's no free lunch in physics."
+
+It is clear that the optimisation of a system for minimising cavitation, and hence improving acoustic performance, is a separate and independent design requirement to that of optimising a system for propulsive efficiency.
 These two design requirements can be found either to be in tension, or alignment, in different regimes.
@@
 However, in very low-speed, low-load environments, where cavitation effects are negligible, and larger areas need to be engaged achieve good mass-flow and efficiency, the two design requirements align relatively well in favour of an open propeller.
-In a variety of intermediate regimes, the two requirements find themselves in tension.
+In a variety of regimes (including at quite high speeds), the two requirements find themselves in tension and needing to be balanced \parencite{gearhart1966selection}.
 
@@
 The answer to the first question, as discussed in earlier sections, seems to be that the operating speeds of a submarine tend to lie far below the probable levels where pumpjets are even remotely comparable to waterjets for efficiency.
-In later sections I will address further the acoustic comparisons in the speed ranges at 10kt and below.
+In my research I have found no evidence of waterjets being advocated for speeds less than 30kt, and for pumpjets for speeds less than 27kt.
+Having consulted with leading academics in the field, I've had it strongly confirmed that a pumpjet will always be less efficient than a propeller at speeds under 18kt.
+In later sections I will address further the acoustic comparisons at low speeds where flow separation will begin to occur, for instance below 10kt.
 
@@
 If the water is slowed enough by the duct to effectively stop, then the impeller must do the work of speeding it up (sucking it into the duct) before the duct slows it down, before the impeller raises its pressure, and the pressure accelerates it out of the nozzle afterwards.
-Compared to simply accelerating the water in one simple push, as a propeller blade does in an unconstrained flow, the efficiency penalty is very large.
+Compared to simply accelerating the water in one simple push, as a propeller blade does in an unconstrained flow, the efficiency penalty is very large, as discussed earlier.
 
@@
 It's important to note at this point that whilst charts such as these are generally used with dimensionless ratios on the axes (such as $\Delta V / V_0$) to allow the most general solutions to be used to solve for any given particular vessel size, speed, or loading requirement which might be required, these  confusing to many readers unfamiliar with hydrodynamics.
-For the avoidance of any confusion, for a particular fixed vessel size and drag, and fixed propulsion system design smaller values of $\Delta V / V_0$ will correspond to to the propulsion system exerting less thrust, and also moving at a lower speed under self-propulsion, since the thust required (proportional to $\Delta V$ when operating at steady state) will decline faster than the vessel speed or water speed at intake, since the drag on the vessel scales with the square of the vessel speed.
+For the avoidance of any confusion, it is important to note that the value for $\Delta V / V_0$ for a given vessl and propulor may tend to increase with increasing waterspeed of the vessel, however one cannot rely upon this value diminishing towards zero at very low speeds.
+In an idealised circumstance, the flow rate increases directly with vessel speed.
+Given that the force produced scales with $\Delta V$, which at constant values of $\Delta V / V_0$ increase in proportion to (if, in the idealised circumstance, vessel speed is equal to $V_0$), an increase in thrust can theoretically match the $V_0^2$ requirement imposed by drag without any change in $\Delta V / V_0$.
+However, in practice the water encountered by the propulsor is slowed by the wake field, the impact of which is likely non-linear with vessel speed.
+In essence, equations developed by Wislicenus are used for the purposes generally of selecting the ideal propulor diameter, which can be made larger in the case of a propeller to access higher ideal efficiencies.
+This is an opportunity which is constrained in the case of jets because the efficiency diminishes at very low values of $\Delta V / V_0$.
+This is quite a different constraint to other "practical" considerations as some have suggested, such as weight, or space constraints.
+Open propellers may advantageously choose to lower their values of $\Delta V / V_0$ as low as 0.1 in order to obtain ideal efficiencies \parencite{wislicenus1973}.
 
@@
 
-The structure of these overall curves is found to be overall quite similar for both pumpjets and waterjets, though the probably values for the losses incurred in the intake duct, or on the exterior of the shroud, are obviously different \parencite[13-14]{wislicenus1973}.
+The structure of these curves is found to be overall quite similar for both pumpjets and waterjets, though the probably values for the losses incurred in the intake duct, or on the exterior of the shroud, are obviously different \parencite[13-14]{wislicenus1973}.
 Similar efficiency shapes can be found for discussions focussing explicitly on underwater pumpjets \parencite[13]{henderson1964}.
@@
 The core of the model rests upon the assumption that drag increases with the square of the velocity of the submarine.
-This is a widely used assumption for fully submerged objects subject to turbulent flows \parencite[5]{wislicenus1973}, \parencite{davies2017}, as also discussed in Section \ref{Speed and Drag - Why very slow is very very $(very)^2$ economical}.
+This is a widely used assumption for fully submerged objects subject to turbulent flows \parencite[5]{wislicenus1973}, \parencite{davies2017}, as also discussed in earlier sections.
 Given that the amount of work done (energy consumed) by a system is equal to the distance over which it acts, which scales with speed also, we assume that the Effective Power $P_E$ (also known as Towing Power) required for propulsion to meet a given speed $v$ can be given by Equation \ref{eq:towpower}, where $C_d$ represents a drag constant.
@@
 Precise energy density of these batteries is not publically available.
-For domestic uses such as handheld electronics or solar power applications, lithium ion batteries may have 3-5 times the energy density of lead-acid batteries.
+For domestic uses such as handheld electronics or solar power applications, lithium ion batteries may have 3-5 times the energy density of lead-acid batteries \parencite{oconnor2017}.
 However, the actual useful density of lithium ion batteries can vary widely depending on different chemistries and designs.
@@
 Considerable confirmation that these assumptions are realistic can be made by assessing the endurance of the submarines under the model with other known reference points.
-For the central curves, we obtain a dived endurance betwen 60 and 100 hours for both systems at 5kt, reflecting nearly exactly the range of other public estimates of dived enurances endurances which other modern diesel-electric submarines operating only with batteries \parencite[3]{article{buckingham2008submarine} at around 4kt.
+For the central curves, we obtain a dived endurance betwen 60 and 100 hours for both systems at 5kt, reflecting nearly exactly the range of other public estimates of dived enurances endurances which other modern diesel-electric submarines operating only with batteries \parencite[3]{buckingham2008submarine} at around 4kt.
 Given that somewhat larger submarines might be somewhat assisted in this respect because their overall volume/surface ratio is slightly higher, this gives confirmation that the model is overall quite realistic in its magnitudes at the most important speeds.
@@
 As can be seen, even compred to the central propeller curve, choosing a jet efficiency curve somewhat lower than the central assumption means that the peak difference immediately jumps to over 100\%, or effectively allow a propeller powered submarine to achieve double the range and endurance of a pumpjet powered submarine at those speeds.
-It is important to note that these speeds in include those speeds which might lie near the maximum speeds that could be sustained by Air Independent Propulsion (AIP) systems \parencite[100]{@stanford2017}.
+It is important to note that these speeds in include those speeds which might lie near the maximum speeds that could be sustained by Air Independent Propulsion (AIP) systems \parencite[100]{stanford2017}.
 Consequently, it is likely that a conventional submarine that used pumpjets might only achieve a fraction of the dived range and endurance enhancements generally expected on propeller powered submarines which employ this technology.
@@
 As can be seen in Figure \ref{fig:sens_hotel} the difference between the two systems is very sensitive to changes in hotel load, both in proportional and nominal terms.
-The impact of halving the hotel load from 100kw to 50kw more than doubles the **difference** between a pumpjet and propeller dived endurances, from around two days to well over four days at speeds between 2.5 and 3.5kt.
+The impact of halving the hotel load from 100kw to 50kw more than doubles the \textit{difference} between a pumpjet and propeller dived endurances, from around two days to well over four days at speeds between 2.5 and 3.5kt.
 In the context of a submarine which might have a nominal dived requirement at patrol speeds in the order of several days, this suggests that extremely large gains can be made by reducing hotel load.
@@
 df %>%
-  filter(test == "endurance") %>%
+  filter(test == "hotel") %>%
   ggplot(aes(x = kts, y = val, col = pair))+
@@
 The results in nominal terms are striking in their magnitude.
-In terms of enduracy, doubling the stored electrical energy, for example by adopting lithium batteries, increase the **difference** between the two systems under the central efficiency curve assumptions to nearly four days (93 hours) at 4kt.
-Given that most modern, capable submarines are estimated to only be able to achieve dived endurances between 60 and 100 hours on batteries at 4kt \parencite{article{buckingham2008submarine} this cannot be dismissed in terms of its tactical significance.
+In terms of enduracy, doubling the stored electrical energy, for example by adopting lithium batteries, increase the \textit{difference} between the two systems under the central efficiency curve assumptions to nearly four days (93 hours) at 4kt.
+Given that most modern, capable submarines are estimated to only be able to achieve dived endurances between 60 and 100 hours on batteries at 4kt \parencite{buckingham2008submarine} this cannot be dismissed in terms of its tactical significance.
 Even under the eficiency curve assumptions representing the narrowest spearation considered, the difference in this case would be 59 hours, almost equal to the estimated dived endurance of a Type214 operating without AIP.
@@
 The in terms of submerged range, the same proportional impact is also extremely significant.
-Doubling the energy stored inceases the **difference** between two otherwise identical submarines to over 400nm under central assumptions at 4.5kt.
+Doubling the energy stored inceases the \textit{difference} between two otherwise identical submarines to over 400nm under central assumptions at 4.5kt.
 That's a larger distance than the width of the Java Sea, the most central body of water in the Indonesian Archipelago, or most of the length of the Malacca Strait.
@@
 This would amount to adding three further frames to the right of Figure \ref{fig:sens_battery}, with the energy density being doubled in each successive frame.
-This would plausibly (under the central line selections) amount to more than month of dived endurance **difference** between AIP inclusive submarines with propellers or pumpjets, amounting to thousands of nautical miles of continuously submerged travel.
+This would plausibly (under the central line selections) amount to more than month of dived endurance \textit{difference} between AIP inclusive submarines with propellers or pumpjets, amounting to thousands of nautical miles of continuously submerged travel.
 This difference would also now be material in terms of the overall range of the submarine amounting to a further 15-20\% reduction in the submarine's overall range.
@@
 
+The precise point at which cavitation inception occurs, what type of cavitation it is, and where it occurs on the blades is a complex science, with considerable work being continually undertaken to validate theoretical models.
+A good overview of the some of the science can be found in reference \cite{kuiper1981cavitation}.
+Still today, similar techniques such as paint tests and visual observations are validate models and theory.
+However, despite the inherent complexity about predicting the precise point in time and space of onset, and the respective impact on efficiency and acoustics, several broad relationships are known to hold well, and can be relied upon to characterise the most significant trends that determine when cavitation will be more likely to occur, which is generally driven by the diffence between the vapour pressure of water and the local static pressure.
+
+\subsection{Impact of speed on cavitation}
+Perhaps the most significant determinant of the pressure deviations on the propulsion system is how hard the propellers have to push on the water at any given point.
+In general, the harder the propellers push, the larger is the required pull on some opposing edge or side.
+It is these pressure reductions, on the suction side of the blade, where sheet cavitation generally starts to occur.
+To think of it simply, the water isn't able to rush to fill the gap left behind by the propeller fast enough.
+However, at some point cavitation also occurs on the pressure side of the blade, as the impact of the leading edge breaking the water leads to sever local pressure drops just behind the leading edge.
+
+Whilst a range of very particular effects may be at work, perhaps the largest trend at work will be that pressure differentials across the blade must necessarily increase with increase thrust.
+The blade velocity relative to the oncoming stream is also a close indicator, which will be closely related for an open propeller.
+Consequently, since the thrust required for self-propulsion increases with velocity squared, the propensity for cavitation occur will necessarily tend to follow this relationship also.
+As such, modest changes in speed can have a big impact on the degree of probability of cavitation occurring.
+Halving your speed will roughly reduce local pressure fluctions by a factor of four.
+Consequently, even for non-specialised propellers, there is generaly some low speed for which no cavitation occurs, though almost all propellers in major surface vessels experience some cavitation at the speed they are designed to do their work at.
+
 \subsection{The impact of depth on cavitation}
@@
 
-This relationship can be expressed in a general form in \ref{fig:InceptionContours}, which has been generated using an approxiate numeric model.
+The interaction of these two relationships can be expressed in a general form in figure \ref{fig:InceptionContours}, which has been generated using an approxiate numeric model.
 As can be seen here, at very low speeds, modest changes in depths can achieve considerable additional surplus pressure in order to prevent cavitation.
@@
 
+It is noteworthy also that the vapour pressure of water is an important component in the equations that describe the conditions for cavitation to occur.
+It is generally given that when the local static pressure falls below the vapour pressure, cavitation occurs.
+Given that temperatures also affects the vapour pressure of water (warmer water is just a little closer to boiling) this does have an impact on likelihood of cavitation occurring.
+However, at plausible temperatures the impact is small.
+The difference between the vapour pressure of water at 10 degrees and 30 degrees is approximately 3 kilopascals.
+Diving just 10m deeper adds over 100 kilopascals to static pressure.
+In many circumstances, altering your speed by a couple of knots will have an even larger impact on local pressure reductions.
+Consequently, the impact of temperature on the inception of cavitation is essentially negligible.
+The difference between inception contours in 20 degree and 30 degree water are given by the difference between the black and red-dashed contours in Figure \ref{fig:InceptionContours}.
+
 \subsection{The significance of flow separation}
-It is also important to understand that the addition of a duct will necessarily to flow separation occuring on the inside of the duct opening at very low speeds which correspond to high advance ratios when load remains constant.
+The addition of a duct will necessarily lead to flow separation occuring on the inside of the duct opening at very low speeds which correspond to high advance ratios when load remains constant.
 Flow separation tends to lead to vortex shedding, which means that inconsistent flows (vortices) are then ingested into the blades of the impeller.

### 2018-01-07 DraftExec.tex
something's not working again.  Just get "missing $ inserted." no line ref.  Aargh

@@
 Hence, in this model we have assumed that 20kt will be obtained with the ouput of 6.5MW.
+
+Considerable confirmation that these assumptions are realistic can be made by assessing the endurance of the submarines under the model with other known reference points.
+For the central curves, we obtain a dived endurance betwen 60 and 100 hours for both systems at 5kt, reflecting nearly exactly the range of other public estimates of dived enurances endurances which other modern diesel-electric submarines operating only with batteries \parencite[3]{buckingham2008submarine} at around 4kt.
+Given that somewhat larger submarines might be somewhat assisted in this respect because their overall volume/surface ratio is slightly higher, this gives confirmation that the model is overall quite realistic in its magnitudes at the most important speeds.
+The sames sources also suggests that the same submarines achieve endurances of around 1-2 hours at 20kt, which also confirms that the relationship in our model linking differnt speeds is also in agreement with other models.
 
@@
 \subsection{Sensitivity to Hotel Load}
+As can be seen in Figure \ref{fig:sens_hotel} the difference between the two systems is very sensitive to changes in hotel load, both in proportional and nominal terms.
+The impact of halving the hotel load from 100kw to 50kw more than doubles the \textit{difference} between a pumpjet and propeller dived endurances, from around two days to well over four days at speeds between 2.5 and 3.5kt.
+In the context of a submarine which might have a nominal dived requirement at patrol speeds in the order of several days, this suggests that extremely large gains can be made by reducing hotel load.
+Increasing the hotel load has the noticeable effect of pushing the speed of largest difference to higher speeds, and substantially reducing the total difference.
+If the hotel load was doubled to 200kW, the middle efficiency curve pair would still see a difference of nearly a day between the two systems, or around 19.5 hours at speeds of 4.5-5 kts.
+
+The difference between pumpjet and propeller propulsion systems in terms of range tends to be found at slightly higher speeds, though this point also moves towards higher speeds as hotel load increases.
+With a reduced hotel load of 50kw, the difference between the two systems in the middle scenario is about 370nm for speeds 3.5-4.5kt, substantially greater than the ~200nm around 4.5-5.5kt at a hotel load of 100kW.
+
+It's important to acknowledge that such substantial differences in hotel load merit serious consideration, both in terms system design, as well as for variance in operation of a submarine with a given system design.
+As has been discussed, the precise hotel load of submarines is not generally publically available, which justifies senitivity testing around a relatively wide potential range.
+The size of the crew required for different designs might vary considerably, as it also would for different missions (for example with the embarkation of a special forces congingent) which would necessarily have considerable impact on energy demand for life support, including sanitation, CO2 scrubbing, and cooking.
+In addition, it broadly understood that the AN/BYG-1 combat system consumes substantially more power than other alternatives, reportedly in the order of 100kW \parencite{patrick2011}.
+
+Furthermore, the hotel load may be varied considerably at the captain's discretion depending on operational requirements.
+Not all sensors or components of the combat system need to run at all times, for example during long transits in order to reach an operational area, or during a long dive in order to escape an operational area once a large strike has already been conducted.
+In wartime, when surfacing to snort within range of enemy airpower or close to naval bases might result in the sumbarine probably being sunk, the crew might willingly forego hot showers, hot meals, and climate-controlled air in order to increase their chances of survival.
+Consequently, it is perfectly plausible that changes of hotel load in the order of a factor of two for tactically significant time periods (days) achievable.
+The results of this sensitivity analysis show that the penalty inflicted by the pumpjet will be substantially greater in such circumstances.
+
+The proportional impact of the changes in hotel load area also substantial, and exhibit clearly the same trend of the impact being larger at higher speeds for higher hotel loads.
+For the efficiency curve pairs that comprise the central selections, decreasing the hotel load to 50kt would see the dived endurance and range of the submarine improved by the adoption of propellers 80\% between 4 and 5.5kt.
+A much higher hotel load, at 200kW, would see that peak difference over 40\% between 5.5 and 8kt.
+
+It is also clear that increasing the hotel load has a greatly diminised impact on the proportional differences between the systems at higher speeds.
+At very high speeds (e.g.
+15kt) changes in the hotel load have almost no impact in proportional terms to the difference, which lies at about 30\% in all cases for the central selection of curves.
+At 9kt, in the region where transit speed is likely to occur, increasing the hotel load by a factor of four (from 50kW to 200kW) only decreases the advantage of the propeller from 48\% to 38\%.
+Therefore one might say near certainty that any difference between efficiency around these patrol speeds is likely to have a substantial impact on the overall range of the submarine, with differences of much less than 30\% quite unlikely, irrespective of other changes in the submarine design or operation circumstance.
+
+It should also be noted that sensitivity testing has not been shown in this paper around alternative values of the drag coefficient $C_d$ which was solved for by matching peak power output and top speed.
+It is considered relatively unlikely that such dramatic changes in this constant are plausible.
+Furthermore, changing this constant will have a very similar impact to alerations in hotel load, in terms of its affect on the speed where greatest differences will occur and their probable magnitude.
+Consequently, consideration of the sensitivity to hotel load is likely to easily encompass the plausible extent, and nature, of the likely uncertainty in this factor.
 
@@
 \subsection{Sensitivity to Battery Changes}
+
+Changes in the effective battery energy density will have the same effect as having larger masses of battery stored.
+They simply directly impact the total amount of energy available for consumption before recharging batteries while snorkeling.
+There is therefore no shift in the proportional impact on range and endurance due to this factor, as can be seen in the lower graphs in Figure \ref{fig:sens_battery}.
+The nominal impact, however increases directly and linearly with the total energy stored.
+
+The results in nominal terms are striking in their magnitude.
+In terms of enduracy, doubling the stored electrical energy, for example by adopting lithium batteries, increase the \textit{difference} between the two systems under the central efficiency curve assumptions to nearly four days (93 hours) at 4kt.
+Given that most modern, capable submarines are estimated to only be able to achieve dived endurances between 60 and 100 hours on batteries at 4kt \parencite{buckingham2008submarine} this cannot be dismissed in terms of its tactical significance.
+Even under the eficiency curve assumptions representing the narrowest spearation considered, the difference in this case would be 59 hours, almost equal to the estimated dived endurance of a Type214 operating without AIP.
+
+The in terms of submerged range, the same proportional impact is also extremely significant.
+Doubling the energy stored inceases the \textit{difference} between two otherwise identical submarines to over 400nm under central assumptions at 4.5kt.
+That's a larger distance than the width of the Java Sea, the most central body of water in the Indonesian Archipelago, or most of the length of the Malacca Strait.
+The tactical significance of such distances are not likely to be lightly dismissed.
+If far smaller amounts of energy are embarked, the nominal differences reduce to perhaps 100nm, in direct proportion to the reduction of the submarine's dived endurance.
+This is still a non-trivial distance in terms of it's tactical signicance, and comprises a substantial fraction of the dived ranges of most non-AIP submarines.
+It still represents a significant difference in operational terms, equal approximately to the width of Hainan Island, or neraly twice the length of the Sunda Strait.
+
+The singificant range of different energy density values tested here reflect quite plausibly some of ranges of different energy levels available in operation.
+For instance, as has been discussed \parencite{briggs2016, greenfield2016}, lead acid batteries may well only be able to practically utilise about 50\% of their total energy stores for tactical operaions, unless operating in a very low-threat environment and undertaking very slow charging to reach the maximum full charge is possible.
+In addition, the plausible improvement in enery density of lithium batteries may well be in the order of a factor of two in practice \parencite{greenfield2016}, though theoretical limits of the chemistry suggest much higher may be possible \parencite{davies2016}.
+
+Only brief is given here to the impact of Air Independent Propulsion, which has substantially higher energy densities possible even than Lithium, and also amounts to the addtional embarkation of greater mass.
+The volumetric energy density of the feedstocks to modern AIP systems (liquid oxygen and methanol) is estimated to be approximately ten times greater than that of lead-acid batteries \parencite{davies2016}, around 4.5MJ/L.
+Given the greater mass embarked, and additional studies suggesting that at constant speeds they might extend the dived endurance of a modern submarine from around 60 hours to over 1000 hours (a factor of roughly sixteen) an approximate estimation of the the direction and magnitude of this impact may by scaling up energy density by this factor.
+This would amount to adding three further frames to the right of Figure \ref{fig:sens_battery}, with the energy density being doubled in each successive frame.
+This would plausibly (under the central line selections) amount to more than month of dived endurance \textit{difference} between AIP inclusive submarines with propellers or pumpjets, amounting to thousands of nautical miles of continuously submerged travel.
+This difference would also now be material in terms of the overall range of the submarine amounting to a further 15-20\% reduction in the submarine's overall range.
+
+With such massive differences between the performance of the two systems emerging with the presence of the AIP, it is difficult to see how one could realistically consider a pumpjet as being competitive if a direct comparison were made.
 

### 2018-01-07 DraftExec.Rnw
for rosses eyes

@@
 
-The purpose of this paper is to investigate the suitability of the pump-jet selection for Australia's future fleet of submarines.
-In particular, the question of the relative efficiency, and quietness of pump-jets in comparison to open propellers is of particular importance.
-Much has been made of the significance of the pump-jet in the DCNS (now Naval Group) bid for the submarine, to the extent that it has been claimed that the pump-jet rendered propellers obsolete.
+It is worthwhile explaining briefly the origins and intent of this paper, how it came to be written, and how it ought to be read.
+
+The purpose is to investigate the suitability of the pump-jet selection for Australia's future fleet of submarines.
+Amongst this considerable controversy surrounding the submarines, this particular technology choice deserves special attention.
+Much has been made of the significance of the pump-jet in the DCNS (now Naval Group) bid for the submarine, to the extent that it has been claimed that the pump-jet rendered propellers obsolete \parencite{naval2018}, and features extremely prominently in the all the marketing and literature.
+It seems to be accepted knowledge that the proposal from Naval Group (formerly DCNS) was selected because of its acoustic superiority \parencite{stewart2016}, with the pumpjet apparently comprising the jewel in the crown.
+
+Having done my higher education in physics, and spent considerable time considering military systems running a small start-up defence company (Rubber Ducky Defence) I found the pumpjet choice particulaly curious.
+As is commonly quipped, including in defence circles, it can reliably assumed when it comes to technology choices "there's no free lunch in physics".
+Whilst all technologies steadily improve, and new ones continuously emerge, it's inevitably the case that significant technology switches are made, some trade-offs are necessary.
+In the case of the pumpjets, the trade-offs were obvious: propulsive efficiency.
+
+It's definitely true that efficiency isn't everything in military technology.
+For Rubber Ducky, I spent considerable time and money designing a small speed-boat and a medium-weight torpedo for naval combat, and in both cases I selected a waterjet or pumpjet for those purpose.
+Pumpjets, or some kind of "decelerating ducted propeller" (explained in detail later) suited very well the sorts of speeds and operational circumstances that they were required to operate in, which were well over 30kt.
+However, having made that selection, and on several occasions wrestled with adapting the boat to alternative purposes, (such as mine warfare, including the towing of heavy sweeps and decoys) I found myself continually balancing the inclination to design a better, more specialised tool for each job, with the desire to compromise performance consolidate capability into a multi-role, more flexible capability.
+The trade-offs inherent in propulsion systems, and the relative efficency of jets at high speeds and propellers at low speeds was a familiar encounter.
+
+The selection of the DCNS bid to design the future submarine attracted considerable attention and controversy on a number of fronts.
+The choice of the pumpjet, and the question of efficiency at low speeds, has been touched on briefly in various informed discussions \parencite{stanford2017} \parencite{davies2017}.
+However, it has remained the case that even extremely well-informed commentators reluctant to quantify and qualify the plausible scale of this trade-off, and what it might mean strategically and tactically for the submarine.
+Everyone knew it was an issue, but no-one confident enough in the scale (or even direction \parencite{davies2017}) of the impact to make it clear whether it mattered.
+Even the head of the progam to build the submarine acknowledged an improvement in dived endurance might be possible with conventional propellers \parencite{sbs2017}, which seemed a clear acknowledgment that some trade-off was acknowledged, even by the strongest advocates of the pumpjet.
+
+The full translation this initial intrigue into the full substance of this paper has come about due to the determination and interest of Gary Johnson.
+Australia is fortunate to have citizens so concerned with national interest to so substantially invest in research to advance public debate.
+Unless he had commissioned me to write it, there is no way that I would have been able to make the sustained and deliberate research which the topic required to reach a useful level of quantification, and confidence, to meaningfully inform debate.
+
+I should hasten to add that my expertise in this area is derived largely from an academic background which a different field of physics (particle physics), which enhanced only my aptitude for understanding and analysing physical concepts, paired with a substantial exposure to the relevant disciplines and industry from a commercial perspective as an entrepreneur developing new products.
+I don't claim to be an authority, but reasonably capable of grasping the works and words of those who are.
 
 Necessarily, this topic depends upon a considerable amount of physics, particularly around hydrodynamics, fluid mechanics, and turbomachinery, all of which might be difficult or daunting topics to usefully inform a public policy debate.
-However, in the case of a \$50 billion dollar military acquisition, neglecting to engage with the technical matters that are so crucial to such a decision would be deeply foolish.
-To that end, this paper has a very specific ambition.
-It doesn't seek to advance the technical field with any new research or insight.
-Instead, it aims to make as much of the relvant physics as possible to this particular question comprehensible in laymans terms, and connect the essential concepts as directly as possible with the most important conclusions of public import.
-
-As such, it will tend to make frequent use of diagrams, illustrations, as well as simple prose, and relatively infrequently use mathematical equations, (unless of crucial importance), though some effort is made to point to references where the fuller and more formal derivations of these relationships can be found.
+This paper doesn't seek to advance the technical field with any new research or insight.
+It aims only to bring together as neatly and simply as possible those concepts and results which are neccessary and sufficient to give the appropriate weighting to this question.
+
+As such, it will tend to make frequent use of diagrams, illustrations, as well as simple prose, and relatively infrequently use mathematical equations, (unless of crucial importance), though some effort is made to point to references where the fuller and more formal derivations of these relationships can be found, and more thorough discussions of those topics which are relevant, but not essential to the question at hand.
 
 It should be noted that the technical fields involved here are vast and rapidly evolving.
-The author makes no claim to be a world authority on the entirety of the topics which are touched.
 It is quite possible that there are other effects and phenomena which could be relevant which aren't mentioned here, and it's also true that additional advances in the state of the art are being made continually, and might not be known about in public literature.
@@
 As such, plenty of foundational principals, in particular the conservation laws of energy and momentum and essential principals of mechanics and thermodynamics are just as true, right, and relevant as they were hundreds of years ago.
-Consequently, whilst we cannot always know the precise degree of advance in a particular field of engineering, we can still confidently know some of the fundamental bounds and constraints that will be inherent in that field.
-
-In this paper, since we cannot know the precise details of the particular state of the art in largely classified military programs, I will attempt to be particularly clear about those things for which we can have great confidence (often broad principals or relationships which determine trade-offs) and those things which we can't (the precise degrees or points and sensitivities).
+Consequently, whilst we cannot always know the precise degree of advance in a particular field of engineering, we can still confidently know some of the fundamental bounds and constraints within which that field will operate.
+
+In this paper, since we cannot know the precise details of the particular state of the art in largely classified military programs, I will attempt to be particularly clear about those things for which we can have confidence (often broad principals or relationships which determine trade-offs) and those things which we can't (the precise embodiments in present design).
 To that end, in this paper I've underaken modelling based upon the broad principals which we can have confidence, and conducted sensitivity testing based on a range of plausible values which seem realistic for those things about which we can have less confidence.
@@
 
-Perhaps the most important relationship to understand is the relationship between speed and drag as it pertains to submarines.
-This is immportant because it sets out the fundamental framework as it applies to any submarine, regardless of propulsion type.
+Perhaps the most important relationship to understand to comprehend the operation of submarines is the relationship between speed and drag .
+This is important because it sets out the fundamental framework as it applies to any submarine, regardless of propulsion type.
 (It's also a pretty important for planes, cars, missiles, torpedoes, and basically everything else.)
@@
 Drag is the resistance that a fluid (air or water, in our case) gives to a body that is passing through it.
-Quite simply, it's a force that acts in the opposite direction.
+Quite simply, it's a force that acts in the opposite direction to movement.
 There are multiple sources of drag for different types of scenarios.
 For scenarios where an object is in contact with two different types of fluid (like a ship, on the ocean) or when the fluid doesn't really have contact with all of the object (supersonic flight, and supercavitating torpedoes) some more complex physics applies.
-For a fuller discussion of types of drag, see \cite{carlton2007}  But in the case of a submarine, which does it's business completely immersed in the ocean, the relevant physics is dominated the skin friction on the hull, which follows a very simple rule and relationship.
+For a fuller discussion of types of drag in the case of ships, see \cite{carlton2007}.
+But in the case of a submarine, which does it's business completely immersed in the ocean, the relevant physics is dominated the skin friction on the hull, which follows a very simple rule and relationship.
 The amount of drag ($F_D$) an objet experiences increases directly in proportion to the surface area $A$.
@@
 \footnote{Technical aside: This law applies wherever the the flow over the surface is turbulent.
-It is true that for very small objects, or very viscous fluids, or very slow movements a different apples called Stokes Equation, in the case wher Reynolds numbers are less than 1.
+It is true that for very small objects, or very viscous fluids, or very slow movements a different apples called Stokes Equation, in the case where Reynolds numbers are less than 1.
 Given that sea-water is not particularly viscous, and sumbarines not particularly small, Reynolds numbers are liekly to be much much greater than 1 (one or two thousand), even when moving at only one or two knots.
@@
 
-It's crucial to understand, however, that drag is only a force, and doesn't directly inform us about how much energy is consumed, until we multiply it by the \textbf{distance} over which it is applied, not the time for which it is applied.
+It's crucial to understand, however, that drag is only a force, and doesn't directly inform us about how much energy is consumed, until we multiply it by the \textbf{distance} over which it is applied (not the time for which it is applied).
 To think about it simply, gravity exerts a force on you downwards all the time.
@@
 At some local point for a moment in time, there is enough energy for the water present to boil.
-It tends to be on rough surfaces or in the presence of some impurity that cavitation will occur first (in the blackened part of the surface in this case.)  However, whilst the bubbles on the bottom can be seen plainly on the bottom, they collapse almost straight away again, and leave only a tiny bubble of stable steam ciculating in the water.
+It tends to be on rough surfaces or in the presence of some impurity that cavitation will occur first (in the blackened part of the surface in this case.)  However, whilst the bubbles on the bottom can be seen plainly on the bottom, they collapse almost straight away again, and leave only a tiny bubble of stable steam circulating in the water.
 It is the rapid expansion and collapse of these bubbles that causes the noise of a kettle, long before it has boiled.

### 2018-01-08 DraftExec.Rnw
next

@@
 
-The full translation this initial intrigue into the full substance of this paper has come about due to the determination and interest of Gary Johnson.
+The development of the full substance of this paper, has come about due to the determination and interest of Gary Johnson.
+Through mutual networks in industry and academia we discovered that we shared had both been intrigued by the technology choice, and considered it of potential significance to Australia's national interest.
 Australia is fortunate to have citizens so concerned with national interest to so substantially invest in research to advance public debate.
@@
 If the in-harbour water-speed was 8kt, but 32kt was the optimum speed for the open-water section of their journey, the fuel requirement for a given distance could still be over ten times greater at 32kt.
-(Aside, the physics described above would say 16 times, but some reduction due to the reduction of surface area from planing might offset this.)  Consequently the owner/operator would be at least ten times as concerned about efficiency at 32kt as at 8kt, (unless the harbour transit was much much longer than the open water section), and have negligible concern about effiencies at 2kt or 3kt, which would rarely amount to more than a couple of percent of total propulsion energy demand.
+\footnote{The physics described above would say 16 times, but some reduction due to the reduction of surface area from planing might offset this.}  Consequently the owner/operator would be at least ten times as concerned about efficiency at 32kt as at 8kt, (unless the harbour transit was much much longer than the open water section), and have negligible concern about effiencies at 2kt or 3kt, which would rarely amount to more than a couple of percent of total propulsion energy demand.
 
@@
 There is, however, a relative scarcity of attention given to the performance of a propulsion system operating at a dramatically different speed, with the same load.
-It also corresponds to scenarios where engineers and manufacturers seem from discussion to be far less confident in making quantitative claims about the performance of their own systems.
-(They are perfecly happy to make qualitative claims, "it would be alright", "nothing very bad would happen" etc.)
-
-Given that submarines almost by definition (since their hull shape doesn't change, and they must carry ballast to make up for any under-loading) are essentially always operating with the same load, but have to operate over an extremely wide variety of different speeds (a factor of four or more in variance) with severe efficiency concern at all of them, they represent a truely unique engineering question, seldom discussed in commercial applications.
+Even engineers and manufacturers of propulsion systems tend to be far less confident in making quantitative claims about the performance of their own systems in extreme off-design conditions, since normally decisions affecting system choice are made based on some threshold of performance change which occurs much closer to the nominal design speed or load.
+
+Given that submarines almost by definition (since their hull shape doesn't change, and they must carry ballast to make up for any under-loading) are essentially always operating with the same load (total drag force at any given speed), but have to operate over an extremely wide variety of different speeds (a factor of four or more in variance) with severe efficiency concern at all of them, they represent a truely unique engineering question, seldom discussed in commercial applications.
 Consequently, there are relevantly few pieces of literature which address the present question directly.
-(Happily, not zero)  However, there are plenty of instances where the relevant physics and trade-offs are explained in depth to address related, but not identical questions.
+However, there are plenty of instances where the relevant physics and trade-offs are explained in depth to address related, but not identical questions.
 
@@
 In contrast, the Collins Class Submarine's main motor is rated at less than 6MW, with designs for the future submarine appearing to be only slightly larger \parencite{patrick2012}.
-It is fair to say that in terms of maximum power ouput, nuclear submarines could have something like 5-10 times as much power at their disposal, and their peak output.
-
-The difference between the peak-powers of the submarines significantly understates how different the designs are, because the difference in the total amount of energy stored and available for use in a voyage or dive is vastly greater.
-A nuclear submarine might have literally millions of times more energy at its disposal, it is practically unlimited for all intents and purposes on a given voyage.
+It is fair to say that in terms of maximum power ouput, nuclear submarines could have something like 5-10 times as much power at their disposal than diesel-electric submarines would, even at their peak output.
+
+However, the difference between the peak-powers of the submarines significantly understates how different the designs are, because the difference in the total amount of energy stored and available for use in a voyage or dive is vastly greater.
+A nuclear submarine might have literally millions of times more energy at its disposal, it is widely regarded as practically unlimited for all intents and purposes on a given voyage.
 Consequently, wheras a nuclear submarine might regularly conduct transits at or near it's peak power, a diesel electric submarine would probably transit at less than half the speed which it could manage at a sprint, (and use less than a quarter of the energy, as discussed earlier) and on patrol it might be operating at a tenth of the maximum speed, and use maybe just 1\% of as much power on propulsion.
@@
 Consequently, the difference between the power output a conventional and a nuclear submarine might regularly operate at could be even quite a bit larger than the maximum amount of energy that they can deliver to their propulsion systems.
+This has considerable consequences for the design of all of the submarine systems, including the propulsion system.
 
@@
 Two of these are Xenon-135 and Iodine 135.
-Iodine is produced much more often, and decays into Xenon-135, which a half-life of about nine hours.
-This Xenon has a very special, perhaps unique role in reactors, since they very easily captures the free neutrons which cause the continued fission reactions in Uranium or Plutonium which drive the reactor.
+Iodine is produced much more often, and decays quickly into Xenon-135 (which has a half-life of about nine hours), and it is this process which produces the majority of Xenon in the reactor.
+This Xenon has a very special, perhaps unique role in reactors, since they very easily captures the free neutrons which trigger the continued fission reactions of Uranium or Plutonium which drive the reactor.
 As such, Xenon is known as reactor 'poison' since it can kill the reactor's reactivity in very high doses.
@@
 The high presence of Xenon reduced the reactor's reactivity.
-On the other hand, there are lots of neutrons available to 'burn off' the Xenon (which absobs the neturons to become a different isotope).
+But another process works in the opposite direction which offsets this, since there are lots of neutrons available to 'burn off' the Xenon (which absobs the neturons to become a different isotope).
 This keeps the total level of Xenon in check.
 
-The situation becomes much more complicated when the reactor undergoes a sudden change in power output.
+The situation changes suddenly when the reactor undergoes a sudden change in power output.
 If the power is lowered dramatically and suddenly, the production of Xenon continues quite rapidly for some time due to the decay of the large stock of Iodine-135.
@@
 Since the reactor will likely need to dispose of excess power, particularly during ramp-down, an inefficient propulsion system actually provides a useful power sink.
-Since any excess power will have to be disposed of by some other means (normally by pumping more water to remove the power as the heat) inefficiency at low speeds has no penalty, and probably a marginal benefit, since it will reduce overall demand for additional systems.
-Provided the excess turbulence inside the pumpjet isn't too noisy, wasting energy through the propulsor is useful.
+Since any excess power will have to be disposed of by some other means (normally by pumping more water to remove the power as the heat) inefficiency at low speeds has no penalty, and probably a marginal benefit, since it will reduce overall demand for additional systems such as pumps to remove the excess heat.
+Provided the excess turbulence inside the pumpjet isn't too noisy, wasting energy through the propulsor is useful whenever there is an ample surplus, which could be the case whenever operating noticeably below top speed.
+Whilst reactors can reach considerably lower power outputs safely in when there are low levels of Iodine and Xenon in the stocks are low (after a smooth ramp-down) the requirement to reduce power gradually means that when speed is reduced suddenly (relative to the 9-hour half life of Xenon 135), for example on arrival in theatre or a higher threat area, this high surplus power will be unavoidable.
 
@@
 They make three comments to justify this claim.
-The first is that the large submarine proposed by DCNS uses as much or more power for the hotel load at the most frequently used speeds.
-Assuming the authors are referring to speeds between patrol speeds and approaching transit speed, the subsequent analysis will show that this fact may be true, particulalry in the presence of the pumpjet, but on its own this doesn't constitute an argument worthy of any consideration.
+The first is that the large submarine proposed by DCNS for Australia uses as much or more power for the hotel load at the most frequently used speeds.
+Assuming the authors are referring to speeds between patrol speeds and approaching transit speed, the subsequent analysis will show that this fact may be true, particulalry in the presence of the pumpjet.
+However on its own this doesn't constitute an argument worthy of any consideration.
 One might equally argue a glutton can be relied upon not to eat too much at dinner because he eats a dinner sized meal at afternoon tea.
 
-The authors of this piece also suggest that the high cost and weight of the nuclear reactor contributes to the desire to minimise hotel load.
-Given the way that the square law drives drag as discussed earlier, driving a cube law for power, there is no chance at all that the hotel load has any bearing on the reactor power required.
+The authors of this piece seem to imply that the relatively high impact of hotel load at patrol speeds is linked to the high cost and weight of the nuclear reactor, and hence this contributes to the desire to minimise hotel load.
+Given the way that the square law drives drag as discussed earlier, driving a cube law for power, there is no chance at all that the hotel load has any bearing on the reactor power capacity required.
 The last few knots of top speed would demand ten times over the power required for hotel load, as can be seen in Figure \ref{fig:power_curve}.
@@
 For all the reasons discussed above, it's completely implausible that the reactor would ever be shut down volunarily during operations.
-One would think that the sorts of emergencies or accidents which would drive this are likely to also mean that the reactor can't be restarted.
+One would think that the sorts of emergencies or accidents which would cause a reactor are likely to also mean that the reactor can't be restarted.
 However, it is plausible that through the development of some instability, perhaps through an unknown fault or dramatic power reductions of the type which I described should be avoided, it might be the case that the reactor would need to be shut down, but could plausibly be restarted.
@@
 This puts some serious constraint around type of contingency that needs to be planned for in a reactor shutdown event.
-It might be presumed that surfacing and remaining surfaced for a couple of days is likely to end combat operations for a submarine that was located in theatre.
-However, remaining submerged and continuing combat operations, even at low speeds will essentially necessitate replicating the full power demand of any diesel-electric submarine until such a time as the reactor can be restarted, especially since hotel load, in the case of the submarine they propose, amounts to more than propulsion at likely patrol speeds.
+It might be presumed that surfacing and remaining surfaced for a couple of days is likely to constitute an end to combat operations for a submarine that was located in theatre, presumeably through enemy action.
+However, the implied alternative remaining submerged and continuing combat operations, even at low speeds will essentially necessitate replicating the full power demand of any diesel-electric submarine until such a time as the reactor can be restarted.
+This fact is emphasised by the DCNS claim that hotel load exceeds propulsive power at patrol speeds.
 As we can gather from other public estimates of submarine submerged endurance \parencite{buckingham2008submarine}, roughly a couple of days is a very significant fraction of the entire dived endurance of a diesel-electric submarine, perhaps 50\%.
-It is difficult to imagine that nuclear submarines carry nearly 50\% of the battery load of diesel-electric submarines, in addition to the reactor.
+It is difficult to imagine that nuclear submarines carry nearly 50\% of the battery load of equivalently sized diesel-electric submarines, in addition to the reactor, simply for safety purposes.
 
 Consequently, it seems much more likely that the sort of battery-requirements that are necessary for this kind of safety system are not closely linked to a full operational hotel load, but some substantially reduced survival load.
-As discussed in later sections, many of the most energy intensive systems in the hotel load that could readily be switched off in an emergency to save power, if saving power was what was required to get through the emergency, including the combat system.
+As discussed in later sections, many of the most energy intensive systems in the hotel load could readily be switched off in an emergency to save power, if saving power was what was required to get through the emergency, including the combat system.
 It is difficult to imagine what sort of emergency might require a reactor to be shut down, that wouldn't also justify shutting down essentially every other component of normal hotel load which could plausibly be delayed for a couple of days.
 
-If it was known that the reactor couldn't be restarted after the poison-out period, the mission need to be aborted in any case, and some kind of rescue sought.
-Even a hundred of tonnes of batteries embarked wouldn't change this reality, unless the submarine literally at the entrance of a home port.
+If it was known that the reactor couldn't be restarted after the poison-out period, the mission would need to be aborted in any case, and some kind of rescue sought.
+Even a hundred of tonnes of batteries embarked wouldn't change this reality, unless the submarine was literally at the entrance of a home port.
 It seems much more likely that the safety requirement for batteries on a nuclear submarine is most likely to be framed around maintaining key life support systems (CO2 scrubbing for example) for sufficient time to enable some hope for a seach/rescue mission to rescure the crew, in the likely case that the kind of emergency which shut down the reactor was the kind of emergency which meant the submarine might not resurface, such as the case of the Russian nuclear submarine, the Kursk.
+
+Claims that the power management imperatives of nuclear submarines and diesel electric submarines are even remotely similar at low speeds are generally implausible.
+Nuclear submarines will have unconstrained energy reserves, and a vast surplus of spare power capacity when ever operating below top speed.
+The dynamics of reactor management dictate that generally avoiding large fluctuation in power, and strictly avoiding sharp and dramtic reductions, will be dominant considerations in a nuclear submarine's power output.
+To put things simply, the characterisation that nuclear submarines have "power to burn" is generally sound, and would only ever be challenged when the submarine is travelling at flank speed, or the reactor is shut down due to a (presumeably near-catastrophic) emergency.
 
@@
 Conservation of energy has a particularly relevant embodiment in fluid flows, which is given it's primary expression in Bernoullis equation, as given in Equation \ref{eq:bernoullis}.
-It says that the total energy in a connected body of fluid is constant ($C$), though it can change in form between kinetic energy (movement, $\frac{1}{2}\rho v^2$), gravitational potential energy (it being elevated, $\rho gz$) the heat energy in the fluid (not shown, not relevant for incompbressible fluids), and the pressure of the fluid $p$.
+It says that the total energy in a connected body of fluid with no additional work being done on it is constant ($C$), though it can change in form between kinetic energy (movement, $\frac{1}{2}\rho v^2$), gravitational potential energy (it being elevated, $\rho gz$, relevant for liquids, less for gas) the heat energy in the fluid (not shown, not relevant for incompbressible fluids), and the pressure of the fluid $p$.
 
@@
 This means that when a liquid is forced to to travel through a narrowing pipe, it's pressure necessarily decreases as the velocity increases.
-This is a simple embodiment of the Venturi effect.
+This is what underlies the Venturi effect.
 
@@
 The inverse process is where a pipe increases in volume, and the flow is forced to slow down in order to fill the wider area, and the pressure correspondingly increases.
-This is a process called 'diffusion', and is important to achieving high pressure levels in many types of water pumps, including those which will be particularly relevant for pumpjets for watercraft propulsion \parencite{hamilton1997}.
+This is a process called 'diffusion', and is important to achieving high pressure levels in many types of water pumps, including those which will be particularly relevant for pumpjets for watercraft propulsion, particularly elevating static pressure levels in order to prevent cavitation \parencite{hamilton1997, mccormick1963design, bruce1974}.
 
@@
 Moving water in directions other than backwards, including random turbulent flows which wind producing heat rather than thust, are two examples of wasted energy.
-Noise in the water also reflects energy which is wasted.
+Noise in the water also reflects energy which is wasted, though the significance of the acoustic signature is not necessarily indicative of the magnitude of energy wasted.
+
+In general terms, energy waste in propulsion systems will result in turbulence (moving water in random chaotic patters, where the energy soon becomes smoothly distributed as heat) and acoustic energy that is propagated away.
+Flow separation incolves the creation of larger-scale vortices and eddies, which might be considered as a special case of turbulence, since the length-scale of the unsteady flows is much larger than that of normal turbulence, for instance in a boundary layer.
+
+On important consequences is that whilst increasing creation of noise, flow separation, or turbulence necessarily represent greater amounts of energy wasted, some increase in any given one does not neccessarily indicate the reduction of efficiency of the entire system, as larger amounts of energy may be being saved through reduced wastage through another means, or vice versa.
+For example, a squeaky join on a car might make a lot of noise, and it will represent some loss of kinetic energy.
+However, the brakes can be applied and absorb vastly larger amounts of energy, essentially silently.
+
+Furthermore, efficiency of a system is necessarily given by the a ratio of useful work to total energy expended.
+Some trends which show increasing levels of energy wasted might be outweighed by larger increases in the overall amount of productive work being achieved.
+Pumpjets, which generally tend to require highly turbulent flows, and even more turblent flows at higher speeds, still can become more efficient as speed inceases, as the amount of useful thrust produced increases even faster.
+So, despite the wasteful round-and-round movements increasing with speed, the increase in productive front-to-back movement improves even faster, and total efficiency rises.
 
@@
 
-Imagine what happens when something moves very fast through water.
+To try to conceptualise how it occurs in the context of propulsors, one might imagine what happens when something moves very fast through water.
 The front side of the object pushes the water forward, but on the back side the water has to push in to fill the space left behind.
@@
 (This is also described as water boiling at low pressure.)  However the gap doesn’t stay around for long.
-Soon the water catches up, and the bubble implodes with a pop, leaving only tiny bubbles as a result, which you can see in the wake of almost any boat or ship moving at speed.
-Whilst some energy turns to heat (the remaining steam in the tiny bubbles) quite a bit is propagated away as a sound-wave generated by the implosion.
+Soon the water catches up, and the bubble implodes with a pop, leaving only very tiny bubbles as a result, which you can see in the wake of almost any boat or ship moving at speed.
+Whilst some energy turns to heat (the remaining steam in the tiny bubbles) some is propagated away as a sound-wave generated by the implosion.
 
@@
 \subsection{Pressure, Thrust, and Momentum change}
-It is worthwhile spending just a moment to touch on the relationship between a few different concepts which are closely related.
+A few simple related concepts make up the foundations for many of the simplest and enduring theoretical models (Momentum Theory, still widely and routinely used to model 'ideal' propeller efficiencies by considering the propeller simple disc \parencite[169]{carlton2007}) and other quite intuitive concepts which migth be more readily grasped in understanding propulsors.
+
 Thrust is a force, measured in Newtons, which is generally used to refer to the forces acting on a vessel or object by its propulsion system to move it forward in the water.
@@
 Pressures are necessarily applied to some extent over all of the surfaces of any object, including due to the atmostpheric pressure of air.
-In considering the efficiency of propulsion systems for solid bodies, thrust is always the consequence of some \textbf{net} pressure difference on different sides of an object.
+In considering the efficiency of propulsion systems for solid bodies, thrust is always the consequence of some \textbf{net} pressure difference on different sides and surfaces of an object.
 For instance, on a propeller blade, the rear-facing side of the blade is the side that pushes the water backwards, and hence has a raised pressure on its surface.
@@
 In fact, this phenomenon is due to a the water-jet drawing adjacent water with it under-water, the movement of which exerts a negative pressure on the outside of the nozzle, conteracting the pressure exerted on the inside of the nozzle.
-This effect represents one of the important distinctions between the jets which release their water below or above the water-line.
+This effect represents one of the important distinctions between the jets which release their water below or above the water-line, and generally means that underwater jets have much lower jet velocity ratios than those released over water.
 
@@
 Put simply, the speed at which one speeds something up, or the increasing amount of something which begins to move, is always equivalent in magnitude to the thrust that is produced.
-Since momentum is a vector, this simple law requires that the momentum change has to net out to one particular direction in order to produce a net thrust.
+Since momentum is a vector, this simple law requires that the momentum change has to net out to one particular direction in order to produce a net thrust, which provides the core principal for Momentum Theory as it applies to propeller and jets.
 Circular movements necessarily are cancelling, and produce no thrust.
-It is useful to recall however that this principal is simply another complete and way of measuring the consequences of the net pressure differences discussed earlier.
-The total rate of change of momentum in the water-column which is induced as a conseqence of the pressure differentials on all sides of the propeller must necessarily be equivalent to the total net force that is produced.
+It is useful to recall however that this principal is simply another complete, alternative way of measuring the consequences of the net pressure differences discussed earlier.
+The total rate of change of momentum in the system which is induced as a conseqence of the pressure differentials on all sides of the propulsor must necessarily be equivalent to the total net force that is produced.
+For the garden hoze in the pool or bucket, the jet rapidly entering the liquid in one direction necessitates new flows of liquid in other directions, reducing net moment changed.
+This new liquid movement will exert pressure reductions on the nozzle surface surface, which reduce the net momentum change affected by the entire system.
+These net forces, and net momentum change are necessarily equivalent and equal in effect.
 Consequently literature seeking to optimise propulsion might discuss analysis referring to both of these effects or phenomena, and neither are in conflict.
@@
 
-In order for flow separation to occur, a fluid must be moving against an adverse pressure gradient, which simply means that the pressure is increasing in the direction of movement, which means the flow is slowing down as it moves.
+In order for flow separation to occur, a fluid must be moving against an adverse pressure gradient, which simply means that the static pressure (pressure in the rest frame of the fluid) is increasing in the direction of movement, which means the flow is slowing down as it moves, unless new work is being done on it.
 At the boundary layer, where the flow velocity is already reduced due to drag forces experienced near the surface, the adverse pressure gradient can be enough to reverse the flow altogether.
@@
 Consequently, some systems are able to become more efficient  by deliberately inducing turbulent, where the benefit from reduced friction in the laminar boundary layer is outweighed by the cost suction induced by the larger wake field created by earlier flow separation.
-This is the reason that golf balls have dimples, and tennis-balls have fur.
+This is the reason that golf balls have dimples, and tennis-balls have fur, to enable them to fly further.
+The impact of the increased surface area, and more turbulent flow over the surface, is exceeded by the reduced wake field ( trailing area where the flow is totally separted) behind the ball.
 
@@
 Between pumps, pumpjets, water-jets ducted or shrouded propellers, or impellers, there is plenty of grounds for some confusion  In this section, I aim to clarify in simple terms what the important differences between distinctly different systems are, and where some terms are used somewhat interchangeably without doing any great violence to the concepts underlying.
+Appreciating where the fundamental as opposed to superficial differences lie is important to understanding subsequently the limitations of different systems.
 
@@
 Perhaps the best starting point is the most basic, and oldest of the systems which we're considering: the basic screw propeller.
-A propeller uses a number of tilted blades attached to a central hub to sweep around disc in the water, and accelerate a column of water passing through the disc.
+A propeller uses a number of tilted blades attached to a central hub to sweep around a disc in the water, and accelerate a column of water passing through the disc.
 Due to the slant of the blades, water on the back sides of the blades is pushed astern, and reduced pressure on the front faces pulls more water from ahead to replace it.
@@
 A comprehensive description of the development of the propeller can be found in \cite{carlton2007}.
-Here I won't elaborate on beyond describing some essential features and characteristics with which one ought to be familiar, primarily for the purpose of comparing different propellers and their evolution into impellers of different designs.
+Here I won't elaborate on beyond describing some essential features and characteristics with which one ought to be familiar, primarily for the purpose of comparing different propellers and understanding their evolution into the systems such as pumpjets.
 
@@
 It is intuitive to think of as something akin to the angle-of-attack of the propeller blade to the water, however this is technically misleading as the movement of the water-column incoming to the blade, as well as the speed of rotation, also have a significant bearing on what the actual angle of attack of the blade ends up being.
-Whilst technically pitch is a distance (measured in meters), it is often expressed and used in formula as a ratio of the diameter of the propeller, to give it a unitless equivalent, as is common in the description of propellers and propulsion systems.
+Whilst technically pitch is a distance (measured in meters), it is often expressed and used in formula as a ratio of the diameter of the propeller, to give it a unitless equivalent, as is common in the description of many characteristics of propellers and propulsion systems.
 
@@
 
-The selection of the exact blade selection that is selected to define pitch must be specified as being at some fraction of the radius from the centre.
+The selection of the exact blade selection that is used to define pitch must be specified as being at some fraction of the radius from the centre, given the tendancy for the blades to twist.
 This distance is calculated to be the 'moment mean' or a technical derived effective average, which tends to lie between 0.6R and 0.7R.
@@
 To some extent, a vessel moving through water will always draw water in its wake along with it, which means that in the wake (where the propeller tends to operate to some extent or other) the water incident on the propeller will arrive at a lower speed than that implied by the ship speed \parencite[15]{MAN2017}.
-Another key distinction for the consideration of jets later on, is that the speed of advance does not necessarily incporate the movement of water due to the movement induced by the propeller (or duct) itself, which is a distinct effect to that of the wake field.
+Another key distinction for the consideration of jets later on is that the speed of advance does not necessarily incporate the movement of water due to the movement induced by the propeller (or duct) itself, which is a distinct effect to that of the wake field.
 In \cite[213]{lewis1988} $V_A$ is used specifically to denote the undisturbed stream velocity, since the role of certain ducts can be to accelerate or decelerate the stream prior to the water arriving at the blades of the propeller.
@@
 This might be the case when a eases it's power, and the momentum of the ship temporarily keeps the ship moving without the propeller exerting much or any net thrust, or where alternative propoulsion sources (sails, or other propellers) are keeping a speed higher than what the propeller actually has push backwards on water to maintain.
+Since drag due to friction scales with velocity squared, very low speed do correspond to increasing advance ratios for a given propeller size, as would increases in propeller size for a given speed and pitch ratio.
+In practice however, under self-propuslion conditions for open propellers, slowing down towards does not elevate the advance ratio far enough past beyond peak efficiencies for overall efficiency to fall very far, as can be seen in Figure \ref{fig:EfficiencyBulten.png}
 
@@
 A ducted propeller is a propeller in any kind of tubular shroud, casing or nozzle.
-Ducted propellers fall into broadly two categories, one of which is referred to frequently and interchangeably as a pump-jet.
-The two types are called the accelerating type (also known as a Kort Nozzle, after a company that made them popular in the 20th Century) and the decelerating type.
+Ducted propellers fall into two categories, one of which is referred to frequently and interchangeably as a pump-jet.
+The two types are called the accelerating type (also known as a Kort Nozzle, after a company that made them popular in the 20th Century) and the decelerating duct.
 The names describe the different effects that each type of nozzle have on the flow speed of the water at the point where it meets the propeller (which is also called an impeller, particularly when associated with decelerating ducts, or pumpjets).
@@
 A detailed description of the essential differences between the two can be easily demonstrated by diagram, which show a simple comparison an acelerating and decelerating duct, which have the shape of the aerofoil section which comprises the duct inverted.
-(Aside: It should be noted that some designs attempt to strike a balance between efficiency and cavitation performance of accelerating and decelerating ducts for certain purposes, and in such cases the duct shape might not be so clearly contrasted \parencite{abdel2010}.
-But for the purposes of explanation, as well as optimisation for military purposes, the distinction between the two will generally be quite clear.)
+\footnote{It should be noted that some designs attempt to strike a balance between efficiency and cavitation performance of accelerating and decelerating ducts for certain purposes, and in such cases the duct shape might not be so clearly contrasted \parencite{abdel2010}.
+But for the purposes of explanation, as well as optimisation for military purposes, the distinction between the two will generally be quite clear, and in practice ducts will still tend to decelerate the flow on approach to the impeller unless specifically shaped to narrow in the opening.}
 
@@
 In those circumstances, relatively small vessels need to exert considerable forces against substantial loads, at very low water-speeds.
-Also, in these circumstances, the practical size of a propeller might be constrained, since operating an optimally sized propeller might increase the complexity of machiner required to deliver the power to so low a hub in the water, and it might also be difficult to arrange for the thrust to be directed substantially against the centre of drag of the vessel.
-
-The effect of the nozzle in these circumstances actually increase the optimal efficiency of the propeller, when operating at very low water-speeds, since the pressure reduction on the inside of the nozzle induced by the movement of the water through the nozzle, actually produces a net thrust forward, since the interior face at the opening of the nozzle is necessarily angled forward.
+Also, in these circumstances, the practical size of a propeller might be constrained, since operating an optimally sized propeller might increase the complexity of machinery required to deliver the power to so low a hub in the water, and it might also be difficult to arrange for the thrust to be directed substantially against the centre of drag of the vessel.
+
+The effect of the nozzle in these circumstances actually increase the optimal efficiency of the propeller, when operating at very low water-speeds but high thrusts (low advance ratio), since the pressure reduction on the inside of the nozzle induced by the movement of the water through the nozzle actually produces a net thrust forward, since the interior face at the opening of the nozzle is necessarily angled forward.
 This has the effect of distributing the suction force across a larger area, which now includes the interior of the nozzle.
@@
 It is also around this vortex that blade tip cavitation tends to occur.
-By limiting the movement of water around the outside of the blade tip by having a tightly fitted shroud, losses due to this particulary type of tubulence, and the onset of this type of cavitation can be diminished.
+By limiting the movement of water around the outside of the blade tip by having a tightly fitted shroud, losses due to this particulary type of tubulence, and the onset of this type of cavitation, or the vortices that occur even when cavitation does not, can be diminished.
 
 However, it is important to note that the existence of an accelerating duct by no means eliminates all cavitation.
-In fact, given that the static pressure at the impeller is necessarily reduced by an accleratint duct, the propensity for cavitation other surface of the propeller is generally increased, and cavitation can and does still occur in ducted propellers near the blade tips \parencite{moulijn2015}.
+In fact, given that the static pressure at the impeller is necessarily reduced by an accleratint duct, the propensity for cavitation other surface of the propeller is generally increased.
+In many circumstance cavitation can and does still occur in ducted propellers near the blade tips at design loads \parencite{moulijn2015}.
 
@@
 At higher waterspeeds, the drag induced by the water-movement over the exterior of the duct begins to increase substantially, and the net thrust on the duct becomes negative.
-When the exterior of the duct experience relatively little water velocity, but the interior experiences a great deal more, the greatest advantages of an accelerating duct are realised.
+When the exterior of the duct experiences relatively little water velocity, but the interior experiences a great deal more, the greatest advantages of an accelerating duct are realised.
 Furthermore, in general these models do not incorporate the effect of flow separation, which is acknowledged as a possibility on both the exterior and interior of ducts, which can dramatically reduce efficiency performance (\cite[20]{oosterveld1970}, \cite[214]{lewis1988}).
@@
 
-In other definitive literature on the design of pumpjets \parencite{henderson1964} the difference between a pumpjet and a propeller is a described as being that "the stream of flow through the pumpjet is made to depart from the "natural" or free-stream sufrace that bounds the flow though a standard propeller."  According to the diagrams and the subsequent description it is clear that the definition embraces quite exactly the distinction between an accelerating duct, and a decelerating duct, namely that the flow is slowed at the point it encounters the impeller, where as in a free situation the free stream would contract at this point.
-
-The addition of some new part such as stator blades (to straighten out the flow) and other possible features which make the propeller system more like pump or turbomachinery components might be associated with systems more commonly referred to as pumpjets than ducted propellers.
+In other definitive literature on the design of pumpjets \parencite{henderson1964} the difference between a pumpjet and a propeller is a described as being that "the stream of flow through the pumpjet is made to depart from the "natural" or free-stream sufrace that bounds the flow though a standard propeller."  According to the diagrams and the subsequent description it is clear that the definition embraces quite exactly the distinction between an accelerating duct, and a decelerating duct, namely that the flow is slowed at the point it encounters the impeller, where as in a free situation the free stream would contract and accelerate at this point.
+
+The addition of some new parts such as stator blades (to straighten out the flow) and other possible features which make the propeller system more like pump or turbomachinery components might be associated with systems more commonly referred to as pumpjets than ducted propellers.
+Not all decelerating ducts have stators, but the impact of the duct on the stream will remain.
 The key concepts behind the operation of these systems, however, remains essentially similar in their nature and design intent.
+In the presence of a duct propellers might be advantageously shaped differently, including maintaing a larger chord lenght right to the shroud, rather than tapering as open propellers tend to lessen vorexs and cavitation on the extremities.
+However, there is not to my knowledge any single defining physical characteristic which the transition to the name 'impeller' implies, which woud not be effectively equivalent to specialised propellers for decelerating ducts.
+In systems tending to be described as pumpjets or waterjets hubs tend to be larger, blade-numbers higher, twist, skew and rake lower.
+No definitive transition point seems to be accepted as having great significance.
+Most commonly literature will refer to a 'rotor', a more general term which could encompass both propellers and impellers.
 
 \subsubsection{Waterjets}
-The use of the word waterjet, as opposed to pumpjet, signifies that the jet expelled from the shroud or pump is situated above the water, which is desirable in very fast-moving surface vessels which attempt to minimise their contact with the waters surface, including high-speed planing hull vessels, including pleasure craft.
+The use of the word waterjet, as opposed to pumpjet, has a tendency signify that the jet expelled from the shroud or pump is situated above the water, which is desirable in very fast-moving surface vessels which attempt to minimise their contact with the waters surface, including high-speed planing hull vessels, including pleasure craft.
 The physics of the two systems, however, are closely related, with the very noticeable exception of the absence of flow considerations for water on the exterior of the duct considerations for waterjets.
@@
 Many of the essential equations which govern the efficiencies and cavitation performance of pumpjets and waterjets are of the same or extremely similar form, as can be seen in detailed discussions given to both, by similar authors (\cite{wislicenus1973}, \cite{henderson1964}).
-There are, however, plenty of instances in recent literature where this distinction regarding the ejection of the jet above and below the waterline is not maintained, and waterjets and pumpjets seem to be used interchangeably \parencite{abdel2010}.
+There are, however, plenty of instances in recent literature where this distinction regarding the ejection of the jet above and below the waterline is not maintained, and waterjets and pumpjets seem to be used interchangeably \parencite{abdel2010}, or systems are described as 'submerged waterjets' \parencite{buckingham2008submarine}.
 
@@
 Recently Andrew Davies published a piece in the The Strategist (ASPI) arguing that the pump jet may actually save fuel at transit speeds (around 8kt) and overall improve the endurance of the submarine, despite likely having lower efficiency at low speeds such as patrol speed \parencite{davies2017}.
-This argument relies pump-jet efficiency still overtaking that of propellers at what is considered a relatively low speed for many applications.
+This argument relies on pump-jet efficiency still overtaking that of propellers at what is considered a relatively low speed for many applications.
 As Andrew acknowledges, further quantification of the likely efficiency is required, as the particular chart which he cites is not quantified.
@@
 
-There has also been questions put in parliamentary hearings regarding the pump jet \parencite{patrick2017} and responses from public officials to such questions indicate that some of the literature provided to support such lines of questions was dated, and that the state of the art may have advanced considerably since this time.
-Whilst the argument that the state of the art is much advanced is obviously true, it is the case with marine propulsion, as with most sciences, that new developments tend to confirm, rather than annul, many of the workings that have been previously undertaken.
+There has also been questions put in parliamentary hearings regarding the pump jet \parencite{patrick2017}, and responses from public officials to such questions indicate that some of the literature provided to support such lines of questions was dated, and that the state of the art may have advanced considerably since this time.
+Whilst the statement that the state of the art is much advanced is obviously true in the case with marine propulsion as with most sciences, it is also equally true that new advances tend to confirm, rather than annul, many of the works that have been previously undertaken.
 To the extent that modern literature still supports the overall constraints and bounds provided by some of the foundational earlier works, a great deal of useful information can still be obtained from earlier works.
 In particular, the foundational theories, often derived from well-established principals of conservation of momentum, energy, as applied by the likes of Bernoulli and Euler, are as well respected and deeply depended upon today as they were centuries earlier when they were first arrived at.
+It is noteworthy that some of oldest works, such as those by Wislicenus and Henderson, are still amongst the most frequently cited in the most recent of papers.
 
 \subsection{Modern literature and results}
-In addition, more modern works, including those which involve experimental validation of theoretical results also clearly confirm that restrictions on the efficiency of waterjets are enduring, and give us a clear starting-point for consideration of their probable orders of magnitude, and likely transition points indicating at what speeds pumpjets might become advantageously more efficient.
+Abundant modern works, including those which involve experimental validation of theoretical results also clearly confirm that restrictions on the efficiency of waterjets are enduring, and give us a clear starting-point for consideration of their probable orders of magnitude, and likely transition points indicating at what speeds pumpjets might become advantageously more efficient.
 It should be noted however, that relatively little literature focuses on the performance of propulsive systems far far outside of their intended design speeds, which tends to be far higher than the patrol and transit speeds of diesel electric submarines.
@@
 They also suggest that differences in efficiency could be very large, of a factor of two or more, including at speeds well over a submarine's transit speed.
-However, care must be taken making precise quantitative comparisons, since such experiements involved jets which expelled into air (waterjets) as opposed to remaining under water (pumpjets).
-Pumpjets would experience drag on the exterior of the shroud/duct, which waterjets would not, however waterjets are likely to have longer intake ducts, which might involve larger duct losses than pumpjets.
+However, care must be taken making precise quantitative comparisons, since such experiements involved jets which expelled into air (waterjets) as opposed to remaining under water (pumpjets) do have at least one significant difference in their makeup, though their internal pump mechanisms may often be similar.
+Specifically, pumpjets would experience drag on the exterior of the shroud/duct, which waterjets would not, however waterjets are likely to have longer intake ducts, which might involve larger duct losses than pumpjets.
 Given that these factors might affect efficiency, and the overall design of the nozzle considerably, more direct comparison with fully submerged jets should be sought to make more direct comparisons.
@@
 If the power could be reduced by 5\% (e.g.
-by fur ther improving the waterjet/hull fairing), then the AWJ-21™ vessel would be better at all speeds above 25 knots."  This statement, however, is made in reference to the hullform needed to accommodate the jet, rather than the jet itself.
-\parencite{giles2010}
+by fur ther improving the waterjet/hull fairing), then the AWJ-21™ vessel would be better at all speeds above 25 knots."  This statement, however, is made in reference to the hullform needed to accommodate the jet, rather than the jet propulsion efficiency itself \parencite{giles2010}.
+As the propulsive efficiency curves given in the study show, the propulsors themselves are considerably further from reaching parity, still separated by about 10\% at 25kt.
 
@@
 At 5kt, the propeller was more than twice as efficient as the waterjet.
+This study has been used to directly inform the 'central' efficiency curves assessed in this analysis.
 
@@
 Indeed, there seems to be considerable evidence that jets of most kinds suffer substantial efficiency penalties relative to propellers in this regime, and that this is understood and accepted as an inevitable design trade-off to be made.
-In the case of a submarine, which might spend nearly 100\% of its time travelling at 10kt or less, this disadvantage (as opposed to less than 35\%, as assumed for a surface ship) would become extremely substantial, which will be quantified in later sections.
+In the case of a submarine, which might spend nearly 100\% of its time travelling at 10kt or less, this disadvantage (as opposed to less than 35\%, as assumed for a surface ship in \cite{giles2010}) would become extremely substantial, which will be quantified in later sections.
 
@@
 To more firmly establish the extent to which adaptation or plausible improvments in technology could or could not change this trend, it is worthwhile considering the theoretical limits that are presented by the physics.
-This should further limit the range of error that might be made in estimating realistic bounds of performance for an unknown system.
+This should further limit the range of error that might be made in estimating realistic bounds of performance for a system which we do not have precise design specifications for.
 
@@
 They first slow it down (increasing its pressure), then add pressure to it, then accelerate it.
-When the total amount of energy that the incoming water has, and the total amount of energy you want to add to the water both get very small (as they do at very low speed) there simply isn't enough energy to do all things at once to all of the water.
+When the total amount of energy embodied in the incoming water stream, and the total amount of energy you want to add to the water both get very small (as they do at very low speed) there simply isn't enough energy to do all things at once to all of the water.
 The energy penalty inflicted by using the additional devices becomes very large in proportion, and part of the water doesn't do all the necessary things for the jet to work.
-In essence, if you try to slow down something that's already going slowly, it tends to stop.
-That's exactly what happens in jets operating at very low flow-rates, when flow separation, also known as stall when it occurs between the blades, takes place.
+In essence, if you try to slow down something that's already going slowly, it tends to stop  That's exactly what happens in jets operating at very low flow-rates, when flow separation, also known as stall when it occurs between the blades, begins to take place.
+When there isn't enough energy to move all the water through the pressure-hurdles imposed by the duct, only some of it moves, and the rest spins in place.
 
@@
 They also provide enduring, reliable limitations on the extent to which ingenuity or design changes might be able to improve the efficiency of jets at low speed.
-the the primary reason why jet efficiencies decline towards zero at low speed is to do with flow separation, which is what occurs when the assumptions inherent in momentum theory's analysis of an 'ideal' actuator disc start to break down.
+But the primary reason why jet efficiencies decline towards zero at low speed is to do with flow separation \parencite{fujisawa1995}, which is what occurs when the assumptions inherent in momentum theory's analysis of an 'ideal' actuator disc start to break down.
 
@@
 Conservation of energy requires that it increases in pressure, and defines a very specific relationship between the pressure increase and the change in velocity, which is determined by the shape of the duct.
-That relationship says that the static pressure increase is inversely proportional change in velocity **squared**, as given by Bernoulli's equation, which was discussed in earlier sections.
+That relationship says that the static pressure increase is inversely proportional change in velocity \textbf{squared}, as given by Bernoulli's equation, which was discussed in earlier sections.
 As a result, at high speeds, the total energy of the system is dominated by its velocity, but at very low speeds the static pressure is much more significant.
@@
 At very high flow rates, the expansion of the area ahead of the impeller creates an adverse pressure gradient, as the water must slow down to fill the greater area.
-This adverse pressure gradient is larger at higher speeds, the proportions aren't linear, hence the the speed is much much larger, due to the square relationship.
+This adverse pressure gradient is larger at higher speeds, the proportions aren't linear, hence the the speed is much much larger, due to the square relationship, and flow is consistent (though still turbulent) throughout the entire cross-sectional area of the duct.
 
@@
 
-The same phenomenon is also prone to occur near roots of the blades of the rotor.
-Necessarily, the blades of the rotor are doing work on the fluid, so unless the impeller is in a region of considerable contraction, pressure will necessarily have to increase through the blades, providing an adverse pressure gradient, which again must be constant radially.
-Since closer to the hub the velocity of the blads through the water is lower, the blades tend often to also be longer and deflect the flow even more in order to do the same amount of work on the fluid.
-With the longer surface over which the fluid must travel, as well as greater angle of deflection, cavitation is prone to occur here too.
+The same phenomenon is also prone to occur near roots of the blades of the rotor \parencite[15]{henderson1964}.
+Necessarily, the blades of the rotor are doing work on the fluid, so unless the impeller is in a region of considerable contraction, pressure will necessarily have to increase significantly through the blades, providing an adverse pressure gradient, which again must be constant radially.
+Since closer to the hub the velocity of the blades through the water is lower, the blades tend often to also be longer and deflect the flow even more in order to do the same amount of work on the fluid.
+With the longer surface over which the fluid must travel, as well as greater angle of deflection, flow separation is prone to occur here too.
 The physics will be exactly analogous to that of the separation on the inside of the duct intake, except the pressure gradient will be provided by a different source (the rotor doing work), and the surface (the blade roots) will also be different.
@@
 
-As can be seen in figure \ref{fig:JetHighFlow.png}, the duct of a jet is design to operate by deflecting the flow away from the natural path that would occur for a propeller, which would accelerate the stream over its blades directly, with the flow necessarily contracting with the acceleration as shown in Figure \ref{fig:PropellerAction.png}.
+As can be seen in figure \ref{fig:JetHighFlow.png}, the duct of a jet is designed to operate by deflecting the flow away from the natural path that would occur for a propeller, which would accelerate the stream over its blades directly, with the flow necessarily contracting with the acceleration as shown in Figure \ref{fig:PropellerAction.png}.
 Instead, pumpjets tend to use some moderate expansion of the flow in advance of the impeller (diffusion) in order to raise its pressure, and reduce its velocity.
@@
 The duct allows this pressure then to be converted back to kinetic energy, as the water accelerates to move through the contracting area in the nozzle.
-This difference in the shape of the flow, and the consequent impacts on the static pressures, and demonstrated in Figure \ref{fig:FlowComparison.png} taken from \cite{mccormick1963design}.
+This difference in the shape of the flow, and the consequent impacts on the static pressures, are demonstrated in Figure \ref{fig:FlowComparison.png} taken from \cite{mccormick1963design}.
 
@@
 The same amount must go out as came in.
-As the area through which it passes changes, the velocity much change to that the flow rate remains constant, otherwise cavitation will occur.
+As the area through which it passes changes, the velocity much change so that the flow rate remains constant, otherwise cavitation will occur.
 The rate at which the areas change (how rapidly they expand and contract along the axis) will determine the gradient.
@@
 
-Flow separation also tend to lead to unsteady flows, with vortices periodically being shed into the main stream as they are speed up or change shape, and then ingested by the impeller and stator blades.
+Flow separation also tends to lead to unsteady flows, with vortices periodically being shed into the main stream as they are speed up or change shape, and then ingested by the impeller and stator blades.
 As can be seen by the diagram in Figure \ref{fig:JetLowFlow.png}, regions abound which are adjacent to or behind flow blockages or reversing flow, where considerable turbulence and unsteady forces would exist.
-These effects also have acoustic consequences, which will be discussed later.
+These effects also have acoustic consequences, which will be discussed in Section \ref{Acoustic Considerations}.
 
@@
 The presence of the duct around the impeller may also have the effect of preventing the noise from being effectively radiated away from the propulsor.
-This may well result in substantial reductions in some types of noise being radiated away from the propulsor in certain directions, which may have given rise to the consistent qualification of 'radiated' being used to descrive the type of noise which differentiated the French submarine bid \parencite{stewart2016, davies2017, ohff2016}, including by the designers of the submarine \parencite{autret2016}.
+This may well result in substantial reductions in some types of noise being radiated away from the propulsor in certain directions, which may have given rise to the consistent qualification of 'radiated' being used to descrive the type of noise which differentiated the French submarine bid \parencite{stewart2016} \parencite{davies2017} \parencite{ohff2016}, including by the designers of the submarine \parencite{autret2016}.
 This is a secondary effect, quite separate to the avoidance of cavitation, which might be seen as an acoustic advantage in circumstances where cavitation might not be expected to occur on either type of propulsor.

### 2018-01-08 DraftExec.Rnw
bump

--- 2018-01-07 DraftExec.Rnw
@@
 
-Para
-
 \tableofcontents
@@
 The selection of the DCNS bid to design the future submarine attracted considerable attention and controversy on a number of fronts.
-The choice of the pumpjet, and the question of efficiency at low speeds, has been touched on briefly in various informed discussions \parencite{stanford2017} \parencite{davies2017}.
+The choice of the pumpjet, and the question of efficiency at low speeds, has been touched on briefly in various informed discussions (\cite{stanford2017}, \cite{davies2017}).
 However, it has remained the case that even extremely well-informed commentators reluctant to quantify and qualify the plausible scale of this trade-off, and what it might mean strategically and tactically for the submarine.
@@
 
-The conclusion of the analysis, and the plotting of these results for certain possible values of $K$ and $\Delta T$ is: "Its outstanding characteristic is that this efficiency approaches zero rather than unity for $\Delta V / V_0 = 0$, even for small duct losses".
+The conclusion of the analysis, and the plotting of these results for certain possible values of $K$ and $\Delta T$ is: "Its outstanding characteristic is that this efficiency approaches zero rather than unity for $\Delta V / V_0 = 0$ , even for small duct losses".
 
@@
 The presence of the duct around the impeller may also have the effect of preventing the noise from being effectively radiated away from the propulsor.
-This may well result in substantial reductions in some types of noise being radiated away from the propulsor in certain directions, which may have given rise to the consistent qualification of 'radiated' being used to descrive the type of noise which differentiated the French submarine bid \parencite{stewart2016, davies2017, ohff2016}, including by the designers of the submarine \parencite{autret2016}.
+This may well result in substantial reductions in some types of noise being radiated away from the propulsor in certain directions, which may have given rise to the consistent qualification of 'radiated' being used to descrive the type of noise which differentiated the French submarine bid (\cite{stewart2016}, \cite{davies2017}, \cite{ohff2016}), including by the designers of the submarine \parencite{autret2016}.
 This is a secondary effect, quite separate to the avoidance of cavitation, which might be seen as an acoustic advantage in circumstances where cavitation might not be expected to occur on either type of propulsor.

### 2018-01-08 DraftExec.Rnw
I'm at hotel load sensitivity

--- 2018-01-08 DraftExec.Rnw
@@
 In practice, the cross-sectional are of the pumpjet naturally only includes that aread between the hub and the duct, and in the case of many pumpjets, particularly those on the ends of torpedoes or submarines, the hub may include/incorporate the tapering tail of the body, for example in Figures \ref{fig:PumpjetOutline.png} or \ref{fig:PumpjetWaterjet.png}.
-Particularly in Figure \ref{fig:PumpjetWaterjet.png}, the appearance of the inner side of the shroud to roughly follow contraction of the hub or body gives the optical appearance of very little diffusion or contraction of the flow occurring, as the one dimensional separation might seem to be roughly constant, or even expanding in places where I have earlier argued it generally contracts.
+Particularly in Figure \ref{fig:PumpjetWaterjet.png}, the appearance of the inner side of the shroud to roughly follow contraction of the hub or body gives the optical appearance of very little diffusion or contraction of the flow occurring, as the one dimensional separation might seem to be roughly constant, or even expanding in places where I have earlier explained it generally contracts.
 It is important to consider the impact of the second dimension on the actual effective area.
-Since area of a disc follows $\pi r^2$, the impact of the contraction of the hub will be dramatically larger when it is large, than it will be when it is small.
+Since area of a disc follows $\pi r^2$, the impact of the contraction of the hub will be dramatically larger when it is large than it will be when it is small.
 
 Consequently, the contraction of the outer shroud at the very end of the duct will dominate the final taper of the hub, and actually lead to substantial restriction of the flow.
-Similarly, where the body or hub is quite substantial and also tapering steadily, keeping the duct intake parallel to the axis of the body for a small distance, or even just tapering it proportionally less than the hub, can still result in a considerable expansion in the area available, and considerable diffusion.
+Similarly, where the body or hub is quite substantial and also tapering steadily, keeping the duct intake parallel to the axis of the body for a small distance, or even just tapering it proportionally less than the hub, can still result in a considerable expansion in the area, and considerable diffusion.
 The example in Figure \ref{fig:PumpjetOutline.png} shows a case of extremely sharp flow diffusion on the intake, probably more than what is practially realistics to avoid flow separation, even at quite high speeds.
@@
 
-Having established that the inevitable decline of a pumpjets efficiency towards lower speeds is driven by the increasing occurrence of flow separation, some consideration ought to be given to the potential scope for the effect to be improved or negated by design changes.
+Having established that the inevitable decline of a pumpjets efficiency towards lower speeds is driven by the increasing occurrence of flow separation, some consideration ought to be given to the potential scope for the effect to be improved or negated by design changes, or otherwise mitigated with other improvements in efficiency.
 Here I have found considerable help from well established literature, many including various old declassified technical reports funded by the US Navy, many of which are frequently cited in the most recent literature as laying out the enduring frameworks for the modern discipline.
-My reading on the matter which reminds me strongly of an adage recently quoted in Senate Estimates by General Manager for Submarines, Stephen Johnson: "There's no free lunch in physics."
 
@@
 
-The answer to the second question seems to be that there the tension between acoustic and propulsive performance at intermediate speeds and loadings is driven by foundational physical requirements.
+The answer to the second question is that the tension between acoustic and propulsive performance at intermediate speeds and loadings is driven by foundational physical requirements.
 That is not to say that optimal means of balancing those requirements will not still be found, and improved, as the art advances.
@@
 The system as a whole produces a positive thrust, becuase the impeller or rotor pushes a larger positive thrust, but the duct necessarily produces a negative thrust.
-At the speeds get lower, and the thrust required gets much much lower, this negative thrust becomes proportionally more significant, as described later.
+As the speeds get lower, and the thrust required gets much much lower, this negative thrust becomes proportionally more significant, as described later.
 
 Put another way, the work required to slow the water down, then speed it up again, becomes a gradually greater penalty the the slower the water is coming in, and the less you actually want to speed it up.
-If the water is slowed enough by the duct to effectively stop, then the impeller must do the work of speeding it up (sucking it into the duct) before the duct slows it down, before the impeller raises its pressure, and the pressure accelerates it out of the nozzle afterwards.
-Compared to simply accelerating the water in one simple push, as a propeller blade does in an unconstrained flow, the efficiency penalty is very large, as discussed earlier.
+If the water is slowed enough by the duct to effectively stop, then the impeller must do the work of speeding it up (sucking it into the duct) before the duct slows it down, before the impeller raises its pressure, and the pressure accelerates it out of the nozzle afterwards, immediately reducing even the ideal efficiency to 50\% \parencite[277]{lewis1988}.
+Compared to simply accelerating the water in one simple push, as a propeller blade does in an unconstrained flow, the efficiency penalty is substantial, and larger in proportion at low speeds.
 
@@
 
-Derivations of the 'ideal' efficiencies are frequently made for propellers, and jets using basic conservation laws (momentum theory) which consider simply acceleration of some column of water as actuated by a disc representing the impeller or propeller.
+Derivations of the 'ideal' efficiencies are frequently made for propellers and jets using basic conservation laws (momentum theory) which consider simple acceleration of some column of water as actuated by a disc representing the impeller or propeller.
 A fuller description of these equations can be found in  most standard texts \parencite[131]{lewis1988}.
-These methods are computationally simple, and allow simple upper-bounds to be be set for efficiencies achievable for given designs under certain conditions, without considering the portential impact of other more complex effects such as skin friction over the blades, potential flow separation, or cavitation.
+These methods are computationally simple, and allow simple upper-bounds to be be set for efficiencies achievable for given designs under certain conditions, without considering the pottential impact of other more complex effects such as skin friction over the blades, potential flow separation, or cavitation.
 As such, they can be easily relied upon to constrain the overall plausible range of values that can be achieved from universally applicable inputs, such as the velocity of water going into a duct, and the velocity of water exiting a nozzle.
 
-Such methods are used in \cite[8-14]{wislicenus1973} to construct an 'ideal' jet efficiency that applies to waterjets and jetpumps.
-Whilst I won't repeat the full expansion here, but a fundamentally important early conclusion of the derivation is the distinction between 'useful work' and the total energy expended.
-since the rate of change in momentum is known to be necessarily equal to the force generated, and increasing momentum change (hence thrust) can be achieved either accelerating more mass, or accelerating mass to higher speeds.
+Such methods are used in \cite[8-14]{wislicenus1973} to construct an 'ideal' jet efficiency that applies to waterjets and pumpjets.
+Whilst I won't repeat the full expansion here, a fundamentally important early conclusion of the derivation is the distinction between 'useful work' and the total energy expended.
+Since the rate of change in momentum is known to be necessarily equal to the force generated, and increasing momentum change (hence thrust) can be achieved either accelerating more mass, or accelerating mass to higher speeds.
 Given that energy embodied in a flow is proportional to it's velocity squared, but its momentum only to its velocity, a necessary and inevitable consequence is that some energy is always wasted in accelerating a fluid to a higher velocity than its surrounds.
@@
 This relationship would lead to the conclusion that infinitely large propellers or jets are always desirable from an efficiency perspective.
-This of course isn't practical to build, and also doesn't hold when the effects of the drag on the propeller blades, or duct, are taken into account.
-Infinitely large systems would increase the total weight, as well as having significant surface draw when moving through the water, all of which are neglected in this 'ideal' consideration.
+This of course isn't practical to build, and also doesn't hold when taken to the etremes the effects of the drag on the propeller blades, or duct, are taken into account.
+Infinitely large systems would increase the total weight, as well as having significant surface drag when moving through the water, all of which are neglected in this 'ideal' consideration.
 
@@
 
-It's important to note at this point that whilst charts such as these are generally used with dimensionless ratios on the axes (such as $\Delta V / V_0$) to allow the most general solutions to be used to solve for any given particular vessel size, speed, or loading requirement which might be required, these  confusing to many readers unfamiliar with hydrodynamics.
-For the avoidance of any confusion, it is important to note that the value for $\Delta V / V_0$ for a given vessl and propulor may tend to increase with increasing waterspeed of the vessel, however one cannot rely upon this value diminishing towards zero at very low speeds.
+Charts such as these are generally used with dimensionless ratios on the axes (such as $\Delta V / V_0$) to allow the most general solutions to be used to solve for any given particular vessel size, speed, or loading requirement which might be required, these  confusing to many readers unfamiliar with hydrodynamics.
+For the avoidance of any confusion, it is important to note that the value for $\Delta V / V_0$ for a given vessl and propulor may tend to increase with increasing waterspeed of the vessel, however one cannot rely upon this particular value diminishing towards zero at very low speeds, as it does for related by different metrics.
 In an idealised circumstance, the flow rate increases directly with vessel speed.
 Given that the force produced scales with $\Delta V$, which at constant values of $\Delta V / V_0$ increase in proportion to (if, in the idealised circumstance, vessel speed is equal to $V_0$), an increase in thrust can theoretically match the $V_0^2$ requirement imposed by drag without any change in $\Delta V / V_0$.
-However, in practice the water encountered by the propulsor is slowed by the wake field, the impact of which is likely non-linear with vessel speed.
-In essence, equations developed by Wislicenus are used for the purposes generally of selecting the ideal propulor diameter, which can be made larger in the case of a propeller to access higher ideal efficiencies.
+However, in practice the water encountered by the propulsor is slowed by the wake field, and so the drag encountered by the vessel may increase faster than velocity of the water entering the intake, an effecnt which is likely non-linear with vessel speed.
+So whilst the trend with regards to vessel speed can be known with regards to the graphs in \cite{figLPumpjetEfficiency}, a direct translation of the x-axis used to vessel speed at any given point is difficult prove without additional system information.
+One might generally infer that the sorts of declines in efficiency observed as vessel speed discussed in recent literature correspond with movement towards very low values of $\Delta V / V_0$, though it is possible that alternative derivations have been used, or that the phenomenon known to occur in practice (such as flow separation) aren't adequately captured by such ideal equations derived from momentum theory in order to make a direct translation into real-world application.
+
+For context, the equations developed by Wislicenus are used for the purposes generally of selecting the ideal propulor diameter, which can be made larger in the case of a propeller to access higher ideal efficiencies.
 This is an opportunity which is constrained in the case of jets because the efficiency diminishes at very low values of $\Delta V / V_0$.
@@
 Open propellers may advantageously choose to lower their values of $\Delta V / V_0$ as low as 0.1 in order to obtain ideal efficiencies \parencite{wislicenus1973}.
+This still provides a hard limit on potential improvements that can be attained for a pump-jet that has non-zero duct loss.
 
@@
 For example, \cite[247]{mollard2011} a graph of ideal efficiency of a pump jet is shown using the speed of advance divided by the jet speed, or $V_1/V_2$.
-Given that the x axis used in the previous charts used $\Delta V / V_0$, which would be related to $V_2-V_1$ in this case, this has the effect of reversing the effective direction of the chart.
-In this case, for a fixed jet and vessel design, high-speed operations will be towards the left of the chart, and the precipitous decline in efficiency will occur as speed declines to zero, along with thrust, toward the right of the chart, as opposed to the left in earlier cases.
+Given that the x axis used in the previous charts used $\Delta V / V_0$, which would be related to $V_2-V_1$ in this case, this has the effect of reversing the effective direction of the chart and changing its shape.
+In this case, for a fixed jet and vessel design, high-speed operations will be towards the left of the chart, and the precipitous decline in efficiency will occur as speed declines to zero, along with thrust, toward the right of the chart, which can be known to correspond to waterspeeds declining to zero at $V_1/V_2=1$.
 
@@
 
-Similarly, it is very important to note that the very frequently used x-axis unit $J = V_a/nD$ when showing efficiency curves for both propellers and ducted propellers (such as used in earlier figures) are of a similar orientation to the chart shown in the figure above.
+These types of charts correspond more closely in orientation and trend that the very frequently used x-axis unit $J = V_a/nD$ when showing efficiency curves for both propellers and ducted propellers (such as used in Figure \ref{fig:OpenWater.png}) are of a similar orientation to the chart shown in the figure above.
 The left side of the chart will correspond to the highest number of rotations for a given amount of progress through the water.
@@
 It is possible that at extremely low speeds a propeller could somewhat descend to the right of optimal advance ratio, when very dramatically reduced load was encounterd because of the extreme low speed, as might be suggested by the slight reduction in propulsive efficiency of the propeller below 5kt in the BMT study.
-The conclusion if design at no higher speeds was required that using a smaller propeller area, which accelerated slightly thicker slices of water with each blade, would re-optimise efficiency.
 These losses of efficiency, however, could never supercede those of a decelerating ducted propeller or jet, which has the shroud which necessarily imposes an \textbf{additional} drag to the watercolumn in excess of any other resistance faced by the rotor.
@@
 
-It should be clear from the above analysis that the efficiency of a pumpjet falling towards zero with a vessels waterspeed is an enduring and foundational result of basic physics.
-The addition of negative thrust on a decelerating duct, no matter how small, necessitates that efficiency must decline at some point with with blade loading, in contrast with an ideal propeller which rises to unity.
-This explains the overall shape given by the the mysterious unquantified chart cited by Andrew Davies \parencite{davies2017} and Access Economics \parencite{stanford2017} shown \ref{fig:PumpUp.png}, and also matches the shapes plotted by more recent studies as shown above, and also the shape presented in other modern publications studying the use of pumpjets \parencite{giles2010} and waterjets \parencite{fujisawa1995}.
+It should be clear from the above analysis that the efficiency of a pumpjet falling towards zero with a vessel's waterspeed is an enduring and foundational result of basic physics.
+The addition of negative thrust on a decelerating duct, no matter how small, requires that efficiency must decline at some point with with blade loading, in contrast with an ideal propeller which rises to unity.
+This explains the overall shape given by the the mysterious unquantified chart cited by Andrew Davies \parencite{davies2017} and Access Economics \parencite{stanford2017} shown \ref{fig:PumpUp.png}, and is also consistent with the shapes plotted by more recent studies as shown above, and those presented in other modern publications studying the use of pumpjets \parencite{giles2010} and waterjets \parencite{fujisawa1995}.
 
@@
 In terms of establishing plausible peaks and cross-over points, the study undertaken in \cite{giles2010} probably provides a sufficient indication as to the current state of the art in the relevant technologies.
-Whilst this effort focussed on a surface vessel rather than a   The study was undertaken with the backing of military customers with support and testing facilities (the US Navy), and involved two major companies who are world leaders in maritime engineering (BMT) and waterjet manufacture (Rolls Royce).
+Whilst this effort focussed on a surface vessel rather than a submarine, the study was undertaken with the backing of military customers with support and testing facilities (the US Navy), and involved two major companies who are world leaders in maritime engineering (BMT) and waterjet manufacture (Rolls Royce).
 Even if further technical advances through design refinement were possible, as the authors alude to, the cross-over point in terms of efficiency might be brought down to 25kt.
-The prospect that the crossover point lies anywhere near the transit speed of a convetional submarine (8-10kt) appears utterly implausible.
-In discussions with leading world experts involved in the development of advanced ducted propellers (including those involved in citations) I was told that a pumpjet could not be more efficient than a propeller below 18kt.
-
-With the optimum efficiency of a jet probably being no greater than 60\% or 70\% at its peak efficiency according to many theoretical and experimental studies \parencite{lu2016cfd}, which would quite probably lie considerably higher in terms of waterspeed, these facts, along with the inevitable decline of efficiency towards zero as waterspeed also declines, a suitably constrained set of plausible efficiency curves may be investigated numerically, as will be done later in later sections.
+The claim that the crossover point lies anywhere near the transit speed of a convetional submarine (8-10kt) appears implausible.
+In discussions with leading world experts involved in the development of advanced ducted propellers (including those involved in citations) I was told that a pumpjet could not be more efficient than a propeller below 18kt, confirming this conclusion.
+
+With the optimum efficiency of a jet probably being no greater than 60\% or 70\% at its peak efficiency according to many theoretical and experimental studies \parencite{lu2016cfd}, which would quite probably lie considerably higher in terms of waterspeed even than 18kt.
+These facts, along with the inevitable decline of efficiency towards zero as waterspeed also declines, offer a suitably constrained set of plausible efficiency curves may be investigated numerically, as will be done later in later sections.
 
@@
 
-As Henderson et al.
-note, the key constraint of this variable is that increasing the diameter of the pumpjet necessarily leads to an increase in the surface (including the external suface) of the shroud.
+As noted in \cite{henderson1964}, the key constraint of this variable is that increasing the diameter of the pumpjet necessarily leads to an increase in the surface (including the external suface) of the shroud.
 The increased drag that this induces must be deducted from the thrust that the pumpjet produces, or effectively increasing $\Delta T$ as given in earlier equations.
@@
 However, this necessarily reduces the effective frontal area that is available to receive the mass-flow, which necessitates further expansion of the outer diameter, and a very substantial increase in the weight of the system.
-In addition, expansion of the hub will necessarily increase the total surface area relative to the mass flow, including by increasing the amount of tapering required after the stators in order to prevent risk of cavitation occuring around the hub.
+In addition, expansion of the hub will necessarily increase the total surface area relative to the mass flow, including by increasing the amount of tapering required after the stators in order to prevent risk of cavitation occuring around the tail the hub.
+Consequently, increasing the area is not plausible without also increasing duct loss to some degree, consistent with the earlier observations from \cite{henderson1964}.
 
@@
 However, this amounts to transitioning towards an accelerating duct, or Kort Nozzle.
-Whilst nozzles of this type have been known to improve the cavitation characteristics of a propeller by enclosing the tips (which often tend to cavitate first), their ultimate function is to actually decrease the pressure at the blades of the propeller.
+Whilst nozzles of this type have been known to improve the cavitation characteristics of a propeller by enclosing the tips (which often tend to cavitate first), their ultimate function is to actually decrease the pressure at the blades of the propeller under higher loadings.
 Consequently, they necessarily increase the propensity of other types of cavitation to occur across the blade-face, including bubble, cloud and sheet cavitation.
@@
 
-In addition, whilst it is often demonstrated that for a given propeller under certain load conditions (high loading), the addition of an accelerating duct can considerably improve efficiency, all of these situations have generally been in the context of vessels which are designed for a relatively high thrust, low waterspeed situation, when propeller size is necessarily somewhat constrained.
+In addition, whilst it is often demonstrated that for a given propeller under certain load conditions (high loading), the addition of an accelerating duct can considerably improve efficiency, all of these situations have generally been in the context of vessels which are designed for a relatively high thrust, low waterspeed situation, when propeller size is necessarily greatly constrained.
 This is the case fishing vessels, or for tug boats, which need to impart significant thrust to a vessel dozens or hundreds of times their own size.
-Mounting a propeller the equivalent size of the vessels they mean to push, however, would be excessively heavy, arranging a mount and power drive that would keep the entire propeller submerged, and not have the effect of flipping the boat backwards when pushing hard, would be effectively impossible, without proportionally increasing the size of the tug.
+Mounting a propeller the equivalent size of the vessels they mean to push, however, would be excessively heavy, arranging a mount and power drive that would keep the entire propeller submerged, and not have the effect of flipping the boat backwards or sideways when rotated hard, would be effectively impossible, without proportionally increasing the size of the tug.
 Furthermore, in most circumstances these propellers are still designed to operate with some cavitation, even if the extent of it is reduced by the presence of the duct \parencite{haimov2010}.
@@
 It is well known that a duct cannot improve the performance of a low-loaded propeller which would be the situation for a slow-moving vessel with unconstrained propeller size.
-Authoritative references in the literature include \cite{oosterveld1970}, who calculates the minimum thrust coefficient $C_T$ (a measure of loading) for an accelerating duct to improve a propellers performance to be between 1 and 2, as shown in @AcceleratingIdeal.
+Authoritative references in the literature include \cite{oosterveld1970}, who calculates the minimum thrust coefficient $C_T$ (a measure of loading) for an accelerating duct to improve a propellers performance to be between 1 and 2, as shown in \ref{fig:AcceleratingIdeal.png}.
 
@@
 The fundamental difficulty with greatly shortening a duct, however, is that it requires that any changes in the shape and with of the water column be accomplished much shorter distance for the duct to achieve its desired effect, which means that generally it has to achieve less of it.
-This means that extremely very low duct lenght achieve less of an efficiency improvement to the ideal efficiency of a propeller, in those circumstances where they can achieve an improvement  (accelerating ducts), as discussed above.
+This means that very low duct lenghts achieve less of an efficiency improvement to the ideal efficiency of a propeller, in those circumstances where they can achieve an improvement  (accelerating ducts at high loads), as discussed above.
 
@@
 Equivalent complexities occur in pumpjets operating in low speed ranges (2-6kt, where flow separation occurs ahead of the impeller, inside the shroud.
-In the case of an accelerating duct, this phenomenon is also observed at high advance ratios, or the equivalent of low speeds and low loads, as shown and discussed in some detail in \cite{willemsen2013}.
+In the case of an accelerating duct, this phenomenon is also observed on the outside of the shroud at high advance ratios (the equivalent of low speeds and low loads) as shown and discussed in some detail in \cite{willemsen2013}.
 
@@
 
-The power drawn by the propulsion $P_D$ is assumed simply to be the propulsive efficiency at any given speed $\eta_v$ multiplied by the Effective Power, as given in Equation \ref{eq:powerdrawn}.
+The power drawn by the propulsion $P_D$ is assumed simply to be the by the Effective Power divided by propulsive efficiency at any given speed $\eta_v$ as given in Equation \ref{eq:powerdrawn}.
 Using a plausible efficiency assumption and total power required, Equation \ref{eq:powerdrawn} can be used to solve for the constant $C_d$.
@@
 \label{eq:powerdrawn}
-P_D = \eta_vP_E
+P_D = \frac{P_E}{\eta_v}
 \end{equation}
@@
 The central assumption has been taken in both cases from the study undertaken by BMT and Rolls Royce, as in \cite{giles2010}, and indicated in \ref{fig:BMTShape.png}, since this represents the most direct comparison made as a simple function of waterspeed, with comparable technologies.
+The technologies naturally will not be identical.
+It is probable that the pumpjet of the submarine will be a purely axial-flow pump, as opposed to a mixed flow pump, but this difference in turbomachinery is unlikely to result in a overall loss of efficiency \parencite{hamilton1997}.
+The intake duct may be somewhat longer, increasing duct losses, but the degree of diffusion in the duct may be lower, reducing the adverse pressure gradient.
+It is probable that the the ship propeller would have five blades, where the submarine would likely have seven or nine.
+However, being  for a warship designed to travel at similar speeds to a nuclear submarine, with relatively unconstrained propeller size, and the pumpjet ejecting into the water, the technologies are likely to be far closer than any other publically available comparison.
+Overall, however, its considered that having an experimtally verified curve comparing very similar technologies as a direct function of waterspeed provides a far more certain basis for comparison than attempting to derive different curves \textit{ab initio} using theortical dimesnionless curves attempting to estimate how the proposed designs might correspond to them.
 
@@
 This has been exaccerbated beyond realistic levels in the case of the the low propeller assumption line for the sake of testing sensitivity more dramatically to some losses in this regime.
-It is unlikely that unless some other significant factor not discussed in this paper that a state-of-the-art submarine propeller would not be able to attain maximum efficiencies between 5 and 15kt.
-Hence, the low propeller curve is considered relatively unlikely, but useful for investigating sensitivity to errors or adjustments.
+It is unlikely that a state-of-the-art submarine propeller would not be able to attain maximum efficiencies between 5 and 15kt, unless some other significant factor not discussed in this paper affects propellers in these regimes.
+At very low speeds, some small but sudden changes in efficiency may be possible in either direction, particularly as complex effect with multiple overlaid effects take place, for example increasing laminar flow over the propeller blades leading to lower frictional losses, but potentially greater separation.
+However, on balance, the low propeller curve is considered relatively unlikely, but useful for investigating sensitivity to errors or adjustments.
+
+\begin{figure}
+\includegraphics[width=\textwidth]{LaminarProp.png}
+\caption{At very low speeds changes in regimes such as an increase in laminar flow over the blade section could lead to small changes in efficiency, perhaps explaining the distinct rise at low speeds for the properller at ~5kt in Figure \ref{fig:BMTEfficiency.png}}
+\label{fig:LaminarProp.png}
+\end{figure}
 
@@
 It is also required that some assumptions are made about the hotel load, or the power which the submarine requires which doesn't go to propulsion.
-We might assume that this is in the order of 100kw, as suggested in \cite{martinez1995} though official sources of information are few.
+We might assume that this is in the order of 100kw, as suggested in \cite{martinez1995} though official sources of information are scarce.
 It is also likely that the actual hotel load varies considerably as different systems are switched on and off, some of which might be discretionary in combat scenarios (e.g.
@@
 Consequently, the actual hotel load experienced might vary considerably, and as such the model used in the app allows the hotel load assumption to be varied, in order to test sensitivity to this parameter.
-A flat 100kW hotel load is used as a starting assumption in the initial calculations in this paper, and some sensitivity analysis is given at half and double this power, to given an indication of the impact of plausible alternative energy demands dependent on operational circumstances, and choice of combat system.
+A flat 100kW hotel load is used as a starting assumption in the initial calculations in this paper, and some sensitivity analysis is given at half and double this power, to give an indication of the impact of plausible alternative energy demands dependent on operational circumstances, and choice of combat system.
 
@@
 It might be assumed that for very large battery banks (such as submarines) where the stability of the battery bank is of critical importance, maximum density migth be sacrificed in favour of additional safety systems, such as increased monitoring, as well as spacing and insulation barriers between cells, in order to prevent thermal runaway in one cell leading to a catastrophic fire.
-The realistic energy density advantage that might be obtained as been suggested to lying closer to a factor of two \parencite{greenfield2016}.
+The realistic energy density advantage that might be obtained as been suggested to lying closer to a factor of two or slightly less \parencite{greenfield2016}.
 
@@
 For instance, one might assume (as done by \cite{davies2017}) that the hotel load might match the propulsive demand at some given speed, for some given hotel load.
-However, given that the precise hotel load is generally not known, and highly variable, and there is potential for the different systems to have widely varying propulsive efficiencies at low speeds, this method is subject to very wide error margins.
-
-Given that any known power consumption level and speed match can be used as a reference, I have chosen instead to use the probable power consumptio at top speed, both of which we have some indication of, and given that propulsive efficiencies of the two systems are likely to be much closer at this speed, there will be less error in likely comparisons between different design concepts.
+However, given that the precise hotel load is generally not known and highly variable, and the potential for the different systems to have widely varying propulsive efficiencies at low speeds, this method is subject to very wide error margins.
+
+Given that any known power consumption level and speed match can be used as a reference, I have chosen instead to use the probable power consumption at top speed, both of which we have some indication of, and given that propulsive efficiencies of the two systems are likely to be much closer at this speed, there will be less error in likely comparisons between different design concepts.
 It is known that the a early German designs for a plausible submarine for SEA 1000 included a 6MW electric motor \parencite{patrick2012}, and the Shortfin Barracuda of Naval Group is often referred to as having a 7MW electric motor \parencite{ohff2017} \parencite{coates2016}, with top speeds ">20kt".
@@
 The phenomenon manifests itself in a very substantial reduction in the overall range that can be achieved submerged, and a difference between the speed at which maximum range can be achieved.
-Under central assumptions, the propeller driven submarine can reach over 540nm, at any speed between 4 and 5kt.
+Under central assumptions, the propeller driven submarine can reach over 540nm at any speed between 4 and 5kt.
 The pump-jet driven submarine reaches over 360nm between 3 and 6kt.
-A difference 180nm dived range would presumeably be of considerable tactical significance.
+A difference of 180nm dived range would presumeably be of considerable tactical significance.
 
@@
 As can be seen from the comprehensive comparison in Figures \ref{fig:central_extreme}, \ref{fig:central_jet}, \ref{fig:central_prop} and \ref{fig:bottom_top}, the shape of curves selected has very different levels of impact at different speeds.
-Key observations include how small the impact of efficiency variation to eithre curves tends to be at very low speeds (under 2kt).
+Key observations include how small the impact of efficiency variation to either curves tends to be at very low speeds (under 2kt).
 This is due to hotel load becoming clearly dominant at such speeds.
@@
 A further feature of the results is that the relative difference in performance is much more sensitive to changes in the efficiency of the jet at the relevant speeds, owing to their larger proportional impact at low efficiencies.
-Whilst a smoothly declining curve from around or over 50\% at 20kt to zero at 0kt, the precise path taken, particularly the exact level around 3-8kt, is of great importance, substantially more than plausible variations in the propeller curve.
+Whilst a smoothly declining efficiency curve from around or over 50\% at 20kt to zero at 0kt might seem moderately well defined, the precise path taken, particularly the exact level around 3-8kt, is of great importance, substantially more than plausible variations in the propeller curve.
 This can be seen in an overlaid plot of the proportional differences in total power demand (hence endurance and range) for the different pairs of curves as in Figure \ref{fig:all_proportions}.
@@
 
-The proportional change in overall power requirement (directly determining both range and endurance) shows that our central scenario shows a propeller driven submarine would have more than 60\% advantage in terms of range and endurance around 5-6kt.
+The proportional change in overall power requirement (directly determining both range and endurance) shows that in the central scenario a propeller driven submarine would have more than 60\% advantage in terms of range and endurance around 5-6kt.
 It is interesting to note that the proportional change is greatest slightly above the speed where maximum range is achieved, and the point of peak advantage moves steadily towards higher speeds as the difference in efficiency between the systems increases.
 
-As can be seen, even compred to the central propeller curve, choosing a jet efficiency curve somewhat lower than the central assumption means that the peak difference immediately jumps to over 100\%, or effectively allow a propeller powered submarine to achieve double the range and endurance of a pumpjet powered submarine at those speeds.
-It is important to note that these speeds in include those speeds which might lie near the maximum speeds that could be sustained by Air Independent Propulsion (AIP) systems \parencite[100]{stanford2017}.
-Consequently, it is likely that a conventional submarine that used pumpjets might only achieve a fraction of the dived range and endurance enhancements generally expected on propeller powered submarines which employ this technology.
+As can be seen, even compared to the central propeller curve, choosing a jet efficiency curve somewhat lower than the central assumption means that the peak difference immediately jumps to over 100\%, or effectively allow a propeller powered submarine to achieve double the range and endurance of a pumpjet powered submarine at those speeds.
+It is interesting to note that these speeds in include those speeds which might lie near the maximum speeds that could be sustained by Air Independent Propulsion (AIP) systems \parencite[100]{stanford2017}.
+Consequently, it is likely that a conventional submarine that used pumpjets might only achieve a fraction of the dived range and endurance enhancements generally expected on propeller powered submarines which employ this technology, as will be discussed later.
 
@@
 The size of the crew required for different designs might vary considerably, as it also would for different missions (for example with the embarkation of a special forces congingent) which would necessarily have considerable impact on energy demand for life support, including sanitation, CO2 scrubbing, and cooking.
-In addition, it broadly understood that the AN/BYG-1 combat system consumes substantially more power than other alternatives, reportedly in the order of 100kW \parencite{patrick2011}.
+In addition, it is broadly understood that the AN/BYG-1 combat system consumes substantially more power than other alternatives, reportedly in the order of 100kW \parencite{patrick2011}.
 
@@
 In wartime, when surfacing to snort within range of enemy airpower or close to naval bases might result in the sumbarine probably being sunk, the crew might willingly forego hot showers, hot meals, and climate-controlled air in order to increase their chances of survival.
-Consequently, it is perfectly plausible that changes of hotel load in the order of a factor of two for tactically significant time periods (days) achievable.
+Consequently, it is perfectly plausible that changes of hotel load in the order of a factor of two for tactically significant time periods (days) is achievable.
 The results of this sensitivity analysis show that the penalty inflicted by the pumpjet will be substantially greater in such circumstances.

### 2018-01-08 DraftExec.Rnw
added reference connections to ones in could find...

@@
 
-\section{Introduction}
+\section{Introduction} \label{intro}
 
@@
 
-There are other some fundamental limitations on the peak efficiency that can be achieved by jets based on simple calculations of ideal efficiency from momentum theory, which are discussed in the next subsection.
+There are other some fundamental limitations on the peak efficiency that can be achieved by jets based on simple calculations of ideal efficiency from momentum theory, which are discussed in Section \ref{tradeoffs}.
 They also provide enduring, reliable limitations on the extent to which ingenuity or design changes might be able to improve the efficiency of jets at low speed.
@@
 As can be seen by the diagram in Figure \ref{fig:JetLowFlow.png}, regions abound which are adjacent to or behind flow blockages or reversing flow, where considerable turbulence and unsteady forces would exist.
-These effects also have acoustic consequences, which will be discussed in Section \ref{Acoustic Considerations}.
+These effects also have acoustic consequences, which will be discussed in Section \ref{acousticconsiderations}.
 
@@
 
-\subsection{Trade-offs and Limitations on Improvements}
+\subsection{Trade-offs and Limitations on Improvements} \label{tradeoffs}
 
@@
 
-\section{Acoustic Considerations}
+\section{Acoustic Considerations} \label{acousticconsiderations}
 

### 2018-01-08 DraftExec.Rnw
check if you like this style...

@@
 \bibliography{references}
+
+\usepackage{nameref}
 
@@
 \cite[24-25]{oosterveld1970} provides a more thorough derivation of the circumstances in which this can be reliably achieved, which generally include larger blade area ratios or with more rows of rotors for higher loading, or circumstances where the loading of the propeller is quite low.
-This creates an inevitable and necessary tension between optimisation of a duct for propulsive efficiency and reduction of cavitation, which will be returned to in subsequent section.
+This creates an inevitable and necessary tension between optimisation of a duct for propulsive efficiency and reduction of cavitation, which will be returned to in Section \ref{waterjets} (\nameref{waterjets}).
 
@@
 
-\subsubsection{Waterjets}
+\subsubsection{Waterjets} \label{waterjets}
+
 The use of the word waterjet, as opposed to pumpjet, has a tendency signify that the jet expelled from the shroud or pump is situated above the water, which is desirable in very fast-moving surface vessels which attempt to minimise their contact with the waters surface, including high-speed planing hull vessels, including pleasure craft.
@@
 
-There are other some fundamental limitations on the peak efficiency that can be achieved by jets based on simple calculations of ideal efficiency from momentum theory, which are discussed in Section \ref{tradeoffs}.
+There are other some fundamental limitations on the peak efficiency that can be achieved by jets based on simple calculations of ideal efficiency from momentum theory, which are discussed in Section \ref{tradeoffs} (\nameref{tradeoffs}).
 They also provide enduring, reliable limitations on the extent to which ingenuity or design changes might be able to improve the efficiency of jets at low speed.
@@
 As can be seen by the diagram in Figure \ref{fig:JetLowFlow.png}, regions abound which are adjacent to or behind flow blockages or reversing flow, where considerable turbulence and unsteady forces would exist.
-These effects also have acoustic consequences, which will be discussed in Section \ref{acousticconsiderations}.
+These effects also have acoustic consequences, which will be discussed in Section \ref{acousticconsiderations} (\nameref{acousticconsiderations}).
 

### 2018-01-08 DraftExec.Rnw
merge?

--- 2018-01-08 DraftExec.Rnw
@@
 
-The development of the full substance of this paper, has come about due to the determination and interest of Gary Johnson.
-Through mutual networks in industry and academia we discovered that we shared had both been intrigued by the technology choice, and considered it of potential significance to Australia's national interest.
+The development of the full substance of this paper has come about due to the determination and interest of Gary Johnson.
+Through mutual networks in industry and academia we discovered that we had both been intrigued by the technology choice, and considered it of potential significance to Australia's national interest.
 Australia is fortunate to have citizens so concerned with national interest to so substantially invest in research to advance public debate.
-Unless he had commissioned me to write it, there is no way that I would have been able to make the sustained and deliberate research which the topic required to reach a useful level of quantification, and confidence, to meaningfully inform debate.
+Unless he had commissioned me to write it, there is no way that I would have been able to make the sustained and deliberate research which the topic required to reach a useful level of quantification and confidence to meaningfully inform debate.
 
@@
 However, in practice the water encountered by the propulsor is slowed by the wake field, and so the drag encountered by the vessel may increase faster than velocity of the water entering the intake, an effecnt which is likely non-linear with vessel speed.
-So whilst the trend with regards to vessel speed can be known with regards to the graphs in \cite{figLPumpjetEfficiency}, a direct translation of the x-axis used to vessel speed at any given point is difficult prove without additional system information.
+So whilst the trend with regards to vessel speed can be known with regards to the graphs in \cite{fig:PumpjetEfficiency}, a direct translation of the x-axis used to vessel speed at any given point is difficult prove without additional system information.
 One might generally infer that the sorts of declines in efficiency observed as vessel speed discussed in recent literature correspond with movement towards very low values of $\Delta V / V_0$, though it is possible that alternative derivations have been used, or that the phenomenon known to occur in practice (such as flow separation) aren't adequately captured by such ideal equations derived from momentum theory in order to make a direct translation into real-world application.
@@
 Changes in the effective battery energy density will have the same effect as having larger masses of battery stored.
-They simply directly impact the total amount of energy available for consumption before recharging batteries while snorkeling.
+They equivalently impact the total amount of energy available for consumption before recharging batteries while snorkeling.
 There is therefore no shift in the proportional impact on range and endurance due to this factor, as can be seen in the lower graphs in Figure \ref{fig:sens_battery}.
@@
 Given that most modern, capable submarines are estimated to only be able to achieve dived endurances between 60 and 100 hours on batteries at 4kt \parencite{buckingham2008submarine} this cannot be dismissed in terms of its tactical significance.
-Even under the eficiency curve assumptions representing the narrowest spearation considered, the difference in this case would be 59 hours, almost equal to the estimated dived endurance of a Type214 operating without AIP.
-
-The in terms of submerged range, the same proportional impact is also extremely significant.
+Even under the efficiency curve assumptions representing the narrowest spearation considered, the difference in this case would be 59 hours, almost equal to the estimated dived endurance of a Type-214 operating without AIP.
+
+In terms of submerged range, the same proportional impact is also extremely significant.
 Doubling the energy stored inceases the \textit{difference} between two otherwise identical submarines to over 400nm under central assumptions at 4.5kt.
@@
 
-Only brief is given here to the impact of Air Independent Propulsion, which has substantially higher energy densities possible even than Lithium, and also amounts to the addtional embarkation of greater mass.
+Only brief consideration is given here to the impact of Air Independent Propulsion, which has substantially higher energy densities possible even than Lithium, and also amounts to the addtional embarkation of greater mass.
 The volumetric energy density of the feedstocks to modern AIP systems (liquid oxygen and methanol) is estimated to be approximately ten times greater than that of lead-acid batteries \parencite{davies2016}, around 4.5MJ/L.
-Given the greater mass embarked, and additional studies suggesting that at constant speeds they might extend the dived endurance of a modern submarine from around 60 hours to over 1000 hours (a factor of roughly sixteen) an approximate estimation of the the direction and magnitude of this impact may by scaling up energy density by this factor.
+Given the greater mass embarked, and additional studies suggesting that at constant speeds they might extend the dived endurance of a modern submarine from around 60 hours to over 1000 hours (a factor of roughly sixteen) \parencite{buckingham2008submarine} an approximate estimation of the the direction and magnitude of this impact may by scaling up energy density by this factor.
 This would amount to adding three further frames to the right of Figure \ref{fig:sens_battery}, with the energy density being doubled in each successive frame.
-This would plausibly (under the central line selections) amount to more than month of dived endurance \textit{difference} between AIP inclusive submarines with propellers or pumpjets, amounting to thousands of nautical miles of continuously submerged travel.
+This would plausibly (under the central line selections) amount to more than a month of dived endurance \textit{difference} between AIP inclusive submarines with propellers or pumpjets, amounting to thousands of nautical miles of continuously submerged travel.
 This difference would also now be material in terms of the overall range of the submarine amounting to a further 15-20\% reduction in the submarine's overall range.
@@
 \subsection{The interaction of hotel load, battery energy density, and efficiency curves}
+It is interesting to observe how the the combination of changes to the hotel load and battery density interact when altered together, as shown in Figure \ref{fig:sens_corners}.
+To the extent that each of these effects contribute to increasing or decreasing the relative difference between the systems, the two clearly combine in a multiplicative fashion.
+
+Where battery density and hotel load are both reduced, the proportional change has a much higher and sharper peak at around 5kt, reaching an 83\% peak at 4.5kt under central assumptions.
+For high hotel load and battery density, the porportional impact is lower at a higher speed, at 43\% at 7kt.
+However, it is intereting to note that the nominal impact of each of these different scenarios is somewhat more similar in magnitude for both endurance and range.
+For low hotel load and battery store, the endurance difference peaks at 54 hours at 3 kt, and range difference peaks at 190nm at 4kt.
+For raised hotel loads and battery stores, the difference is shifted to higher speeds and broadened somewhat, with a difference of 39hours at 4-5kt, and 200nm at 5-6kt.
+
+The most striking feature of this figure however is the impact on the submarine's performance when hotel load is lowered, and battery density is raised simultaneously.
+Here the two factors combined have a proportional impact of around 80\% improvement for the switch to propellers under central assumptions, and over 150\% in the scenario of widest separation.
+However, in nominal terms, this impact is even more significant, with central scenarios showing a \textit{difference} of over 200 hours between the two systesm under central assumptions at 3kt, and a range difference of over 750nm.
+These differences are extremely large, representing plausibly double the expected total ranges and endurances of existing modern diesel-electric submarines.
+Even if such large changes in hotel load and battery capacity are not simultaneously likely these sensitivitiesquite clearly demonstrate how even modest shifts in both hotel load and battery energy density might combine to have very significant impacts on the appearnce of the pump-jet and propeller comparison.
+For instance, a hotel load that reduced by 30\%, and a battery energy density increase of 40\% could plausibly have the impact of doubling the nominal differces between the performance of the two propulsive systems.
+
+This sort of extreme sensitivity ought to have a significant bearing on how the overall performance specifications of the submarine are defined, as seemingly incremental changes to hotel load and electrical energy stores are likely to substantially magnified, or muted, by the choice of propulsion system.
 
@@
 It is generally considered that the existence of cavitation of any sort will tend to dominate the acoustic signature of any vessel that doesn't produce any machinery noise by relying on all-electric drive as a conventional submarine would when dived (\cite{giles2010}, \cite{harvie1965construction}).
-Its quite possible that during the period when a submarine is snorting, or running its diesel engines, these sources of noise would be a more substantial and distinctive acoustic signature than small amounts of cavitation on the propulsion system, if it were to occur, owing in particular to the low freqency of diesel engines.
-
-Other sources of noise hydrodynamic noise, such as the turbulence produced in the boundary layer of the flow over the hull, or parts of the propulsor, will still contribute to overall noise and some signature, but it will tend to be vastly smaller in magnitude.
+Its quite possible that during the period when a submarine is snorting, or running its diesel engines, these sources of noise would be a more substantial and distinctive acoustic signature than small amounts of cavitation on the propulsion system if it were to occur, owing in particular to the low freqency of diesel engines.
+
+Other sources of noise hydrodynamic noise, such as the turbulence produced in the boundary layer of the flow over the hull, or parts of the propulsor, will still contribute to overall noise and some signature, but it will tend to be significant smaller in magnitude.
 This owes to the essentially incompressible nature of water.
@@
 This obviously doesn't hold true in compressible fluids, such as air, as one would observe noticing the intense noise produced by hand-dryers, aeroplanes, jet engines etc.
-It also doesn't hold true within a turbulent flow underwater.
+It also doesn't hold true \textit{within} a turbulent flow underwater.
 If you were to put a sensor inside a turbulent flow, the rapid movements would result in continual and significant pressure fluctuations, hence noise.
@@
 
-However, when dived, it might be reasonably assumed that the complete elimination of all sources of cavitition will be the primary objective of acoustic signature management.
-This is neatly put in \cite[88]{gearhart1966selection}: "The occurence of cavitation in a propulsor results in adegradation of the propulsor operating characteristics if extensive amounts of cavitation exist, and in significant noise when minute amounts occur.
-As a result, the performance requirements of a submersible weapon system specify that the occurence or inception of cavitation be avoided below certain operating depths." One might consider that a submarines 'tactical silent speed' would most probably correspond to the speed at which it can operate safely below the point of any cavitation inception.
+As a result, when dived, it might be reasonably assumed that the complete elimination of all sources of cavitition will be the primary objective of acoustic signature management.
+This is neatly put in \cite[88]{gearhart1966selection}: "The occurence of cavitation in a propulsor results in a degradation of the propulsor operating characteristics if extensive amounts of cavitation exist, and in significant noise when minute amounts occur.
+As a result, the performance requirements of a submersible weapon system specify that the occurence or inception of cavitation be avoided below certain operating depths." One might consider that a submarine's 'tactical silent speed' would most probably correspond to the speed at which it can operate safely below the point of any cavitation inception.
 
@@
 
-Whilst a range of very particular effects may be at work, perhaps the largest trend at work will be that pressure differentials across the blade must necessarily increase with increase thrust.
+Whilst a range of very particular and intricate phenomena may be at work, perhaps the largest and most dominant trend at work will be that pressure differentials across the blade must necessarily increase with increase thrust.
 The blade velocity relative to the oncoming stream is also a close indicator, which will be closely related for an open propeller.
@@
 \subsection{The impact of depth on cavitation}
-It is crucial to understand that increasing depth necessarily increases the static pressure throughout a propulsion system.
-As such, the speed at which cavitation starts to occur, will necessarily be a function of depth.
+It is also crucial to understand that increasing depth necessarily increases the static pressure throughout a propulsion system.
+As such, the speed at which cavitation starts to occur will necessarily be a function of depth.
 Approximately every 10m of water adds approximately the equivalent of an atmosphere of pressure.
@@
 As can be seen here, at very low speeds, modest changes in depths can achieve considerable additional surplus pressure in order to prevent cavitation.
-However at much higher speeds, sine the pressure diffential scales with the square of the velocity travelled, far greater additional depths are required to achieve sufficient excess static to supress cavitation for a give further speed increase.
+However at much higher speeds, since the pressure diffential scales with the square of the velocity travelled, far greater additional depths are required to achieve sufficient excess static to supress cavitation for a give further speed increase.
+Whilst certain system parameters such as the blade area and design (or duct design) will have a significant impact on which curve a particular system lies upon, the overall trend of a parabolic cavitation free "bucket" prevailing at low speeds and large depths will generally be robust.
+This is confirmed by the close resemblance between the curve shapes found by an experiment on a torpdo in Figure \ref{fig:InceptionDepth.png}, and the numerical model found in Figure \ref{fig:InceptionContours}.
 
@@
 It is noteworthy also that the vapour pressure of water is an important component in the equations that describe the conditions for cavitation to occur.
-It is generally given that when the local static pressure falls below the vapour pressure, cavitation occurs.
-Given that temperatures also affects the vapour pressure of water (warmer water is just a little closer to boiling) this does have an impact on likelihood of cavitation occurring.
+It is generally given that when the local static pressure falls below the vapour pressure, cavitation occurs, though known exceptions to this do exist.
+Given that temperatures also affects the vapour pressure of water (warmer water is just a little closer to boiling) this does have an impact on the cavitation inception point.
 However, at plausible temperatures the impact is small.
@@
 \subsection{The significance of flow separation}
-The addition of a duct will necessarily lead to flow separation occuring on the inside of the duct opening at very low speeds which correspond to high advance ratios when load remains constant.
+The addition of a duct will necessarily lead to flow separation occuring on the inside of the duct opening at some very low speeds which correspond to high advance ratios when load remains constant.
 Flow separation tends to lead to vortex shedding, which means that inconsistent flows (vortices) are then ingested into the blades of the impeller.
-This effect is well known and document.
+This effect is well known and documented.
 A good example is shown on the exterior of an \textbf{accelerating} duct at $J = .8$ in \cite{willemsen2013}.
@@
 
-Whilst at very low power levels, or at good depths, it is quite plausible that this doesn't result in any cavitation, which would normally be a likely occurrence if the blades are operating closer to their design power and such such vortices were shed into the intake.
-However, given the inherently unpredictable and unsteady nation of such vortex formation and shedding, the this phenomenon would increase the risk of cavitation occurring unexpectedly in marginal conditions.
+Whilst at very low power levels or at good depths, it is quite plausible that this doesn't result in any cavitation, which would normally be a likely occurrence if the blades are operating closer to their design power and such such vortices were shed into the intake.
+However, given the inherently unpredictable and unsteady nature of such vortex formation and shedding, this phenomenon would increase the risk of cavitation occurring unexpectedly in marginal conditions.
 In addition, in the absence of cavitation, the sudden and changing pressure gradients over the blades would dramatically increase the overall pressure fluctuations and turbulence level that occurs within the propulsor.
 This would necessarily increase the overall levels of noise generated.
-It is plausible that the noise is of sufficiently small level that it is not considered a significant compromise of stealth.
-In essence, the higher levels of noise lie below some threshold of significance.
+It is plausible that the noise is of sufficiently low level that it is not considered a significant compromise of stealth.
 However, it is almost certainly untrue that that pumpjet actually generates less noise overall than a propeller at very low speeds.
@@
 However, it should be noted that some pumpjets have been constructed with a hollow design, in order to reduce weight \parencite{harvie1965construction}, which would reduce the thickness of solid steel, though offer some potential opportunity for damping.
-Overall, damping is increased in its effectiveness at low frequencies by allowing it to occur over larger distances, corresponding to the sound's increased wavelenght, leading to increased thickness of the transition substance or damping material.
+Overall, damping is increased in its effectiveness at low frequencies by allowing it to occur over larger distances, corresponding to the sound's increased wavelength, leading to increased thickness of the transition substance or damping material.
 Consequently, effective damping within the propulsor itself is unlikely to be attempted, or limited to medium and higher frequencies, probably suited to reducing noise from normal surface flow turbulence and very early onset cavitation.
@@
 It is most likely that little or no damping or absorbing layer is attempted within the body of the shroud, since increasing the duct thickness above hydrodynamically determined optimum profile is likely to have a detrimental impact on efficiency.
-Constructing especially thin layers, and still maintaining an optimal shape to the precision required for housing powerful turbomachinery like the impeller is also likely to involve considerable construction risk and complexity.
+Constructing a large components from thing pieces of steel, and still maintaining an optimal shape to the precision required for housing powerful turbomachinery like the impeller is also likely to involve considerable construction risk and complexity.
 Hence the shielding effect is most likely to dominated by the reflective characteristics of the duct, which would seek to internally reflect noise back into the turbulent flow, where sound waves would be increasingly convoluted, and somewhat absorbed through the eventual transition to small-scale turbulence and heat.
@@
 The degree to which the total noise generated is reflected internally will also be shape-dependent.
-A more concave inner surface, consistent with higher degrees of diffusion in advance of the impeller, and greater nozzle narrowing after it, which is consistent with higher pressure elevation at the impeller (hence improved cavitation characteristics) but higher negative duct thrust, leading to even poorer efficiencies at low speed.
+A more concave inner surface, will increase the extent of internal reflection.
+Such a design would be consistent with higher degrees of diffusion in advance of the impeller, and greater nozzle narrowing after it, which is consistent with higher pressure elevation at the impeller (hence improved cavitation characteristics) but higher negative duct thrust, leading to even poorer efficiencies at low speed.
 Substantial alteration of duct shape on account of efficiency considerations at low speed are therefore necessarily going to have some impact on the degree of effective shielding that a duct can afford.
@@
 When the submarine is travelling at high speeds, and there is significant flow through the propulsor, the impact of waterspeed through the propulsor might have the additional impact of diminishing the radiation of noise in the forwards direction, in the same way that sound tends tends not to travel well up-wind in air.
-However, at very low speeds this effect would be greatly diminished.
-
-It should be noted that in the underwater environment sound can be made to reflect in a variety of directions simply with differences in temperature, salinity, movement, as well as off plenty of nearby underwater objects, including the ocean floor and objects on it.
-This phenomenon generally serves to make the art of underwater acoustics extremely complex and unpredictable, and generally contributes to the difficulty of precisely locating a submarine, even when some signature is detected.
+However, at very low speeds this effect would be greatly diminished, and possibly not relevant.
+
+It should further be noted that in the underwater environment sound can be made to reflect in a variety of directions simply with differences in temperature, salinity, movement of water bodies, as well as off plenty of nearby underwater objects, including the ocean floor and objects on it.
+This phenomenon generally serves to make the art of underwater acoustics complex and unpredictable, and generally contributes to the difficulty of precisely locating a submarine, even when some signature is detected.
 However, it also means that direction-dependent reductions signature do not reliably translate to overall reduced probability of detection.
@@
 This has an important consequence for shielding for pumpjets at very low speeds because of the occurrence of flow separation and vortex shedding.
-In the absence of flow separation, and cavitation at very low speeds the most substantial sources of sound would be the turbulent flow over surfaces of in the propulsor.
+In the absence of flow separation and cavitation at very low speeds the most substantial sources of sound would be the turbulent flow over surfaces of in the propulsor.
 This noise will tend to be relatively white and high in its frequency owing to the very small length-scales of the eddies and vortices in such turbulence.
@@
 These sounds might also be very effectively reflected by the shroud, and consequently diminished.
-However, flow separation results in the creation of vortices of a dramatically larger scale than occurs with normal turbulence in a boundary later, and hence generate noise of a much lower frequency.
+
+In contrast, flow separation results in the creation of vortices of a dramatically larger scale than occurs with normal turbulence in a boundary later, and hence would tend to generate noise of a much lower frequency.
 In addition, as the flow rates slow even further, these vortices are likely to become larger in size, even if lesser in intensity, resulting in further reduction in the frequency of noise generated.
-Further to this effect, the rate at which the vortices shed from the surface and impact the impeller will also result in another acoustic signature, of even lower frequency.
+The rate at which the vortices shed from the surface and impact the impeller will also result in another acoustic signature.
 Again, at lower flow rates, these vortices will tend to develop and shed more slowly, lowering frequencies further.
@@
 However, given that the amplitude, frequency, and combined impact of those two characteristics cannot be accurately evaluated without substantial further work and more detailed system knowledge, it is also possible that the increase in other noise (including hull flow noise and other machinery) at elevated speeds offsets the improvement in the smoothness of the pumpjet's operation.
-It is also possible that the overall level of noise radiated by the pumpjet, including under conditions when flow separation occurs, is still considered to be sufficiently low as to not be of significant tactical consequence.
+It is still quite likely the overall level of noise radiated by the pumpjet, including under conditions when flow separation occurs, is still considered to be sufficiently low in amplitude as to not be of significant tactical consequence.
 However it is highly unlikely that the accoustic signature of the pumpjet is actually superior to that of an open propeller at very low speeds, when cavitation occurs for neither system, but substantially increased turbulence including flow separation must neccessarily occur for the pumpjet.
 
-\section{Conclusions}
-This is a pretty bad result for the pumpjet.
+\section{Conclusion}
+
+The selection of the pumpjet propulsion system is quite probably the most significant new technology selection for Australia's future submarines.
+Unlike other new technolgoies such as Air Independent Propulsion and Lithium Ion Batteries, which are either in service or entering service in multiple other navies, no other conventionally powered submarines have adopted pump-jet technology, which is currently only used by nuclear submarines.
+
+Despite acknowledging that the pumpjet may have an adverse impact on dived endurance \parencite{sbs2017}, Naval Group justifies the superiority of the pumpjet in the following terms: "In a confrontation between two otherwise identical submarines, the one with pump jet propulsion always has the tactical advantage."  No further explanation of this superiority has been provided in detail, but it is widely held that the pumpjet offered acoustic advantages which were highly regarded by those Australians who were responsible for selecting the design partner \parencite{cameron2016}.
+
+The comparison between two otherwise identical submarines (though not necessarily limited to engaging just each other in combat), provides a useful and suitable framework for evaluating a single significant technology choice, where most other significantly novel or new technologies were apparently excluded from consideration \parencite{stanford2017}.
+The conclusion of this report is that the following statement can be made with confidence:
+\n
+\emph{In a comparison between two otherwise identical submarines, the one with the pump jet will alwayas have a lower dived endurance, a lower dived range, a worse indescretion ratio, a lower overall endurnace, and a lower overall range, than the one with a conventional propeller.
+This will confer a substantial tactical and strategic advantage on the conventionally propelled submarine in a very broad range of operational scenarios.}
+\n
+
+Whilst it is broadly acknowledged that propellers have superior efficiency at lower speeds, but jets might have higher efficiency at higher speeds, the precise point of cross-over seemed unclear in public debate.
+One informed commentator suggested that the crossover point was likely to be below or near the transit speed of the submarine, and that the pumpjet may offer enhanced range and endurance over the whole mission profile as a consequence\parencite{davies2017}.
+This report finds no evidence of a pumpjet of the type used on submarine every being recommended as superior for efficiency in general, and even closely related technologies for more commecial use not advocated below 25kt.
+During my research I encountered domain experts confident that a pumpjet could simply not feasibly be superior to a suitable propeller at speeds below 18kt.
+Evidence of propellers remaining generally superior to waterjets and pumpjets in the speed ranges up to and even over 30kt is abundant.
+Consequently, the possibility of some cross-over point leading to any kind of range or endurance advantage relevant for a conventional submarine can be safely ruled out.
+
+This report also explores in some detail the theoretical concepts which drive the limitations of pumpjet performance at lower speeds in comparison to propellers.
+Enduring and fundamental principals of momentum theory provide robust limitations on the circumstances where a propeller's 'ideal efficiency' might be improved by the addition of a duct/or shroud, one type of which constitutes the transition to a 'pumpjet'.
+These principals stronly support the the notion that pumpjets will not and indeed cannot suddenly overcome their efficiency disadvantage at low speeds by means of technical ingenuity or innovation.
+
+Furthmore, a particular hydrodynamic phenonomon called 'flow separation' can be identified as the primary cause of the pumpjet's declining efficiency at very low speeds.
+Put simply, the duct provides an additional hurdle or barrier though which the water must pass when moving through the propuslor.
+At very low speeds waterspeeds, this additional resistance is sufficient to stop some of the flow progressing through the jet, creating unsteady eddies and vortices which sap energy from the system, but produce no thrust, driving down efficiency to zero at very low flow-rates.
+
+The design of the pumpjet for military purposes, namely the reduction of acoustic signatures effectively mandates that this must be the way that pumpjets work.
+The act of slowing the water down is necessarily equivalent to the act of raising its pressure.
+It is this elevated pressure at the point where the blades to their work on the water which enables the blades to avoid cavitation (the rapid expansion and collapse of small voids or bubbles in the water) which is very large source of noise underwater.
+Maintaing this advantage to the same level necessitates a duct design which slows water down to the same extent.
+The act of slowing water down when it's already going slowly is what causes some of it to stop, which is when flow separation occurs.
+
+In order to assess the probable magnitude of the impact of this reduction in efficiency at low speeds, a numerical model was developed, based on known principles of physics, and publically available information relating to the likely technologies involved in the future submarine.
+The most important and variable assumptions - the efficiency curves of the pumpjet and propeller - were taken from a recent study lead by world-leading engineering companies, supported by the US and Royal Navies, comparing closely related propulsion systems \parencite{giles2010}.
+Knowing that the systems would not be identical to the propulsors on submarines (the precise design of which is obviously secret), additional curves varying from these assumptions were also modelled in all directions, and the combinations combined and tested.
+
+The model has been made available publically in the form of a web-app, so that alternative assumptions, including alternative efficiency curves may be tested by the general public.
+
+The model in its central assumptions was able to achieve very realistic results, which were consistent with other public estimates of performance parameters of conventional submarines and known characteristics \ref{buckingham2008submarine}.
+A variety of different assumptions regarding the hotel load, battery density and different combinations of efficiency curves were tested.
+The central assumption values showed that a submarine with a conventional propeller had a very substantial advantage in terms of range and endurance across the entire speed range.
+This advantage reached a proportional maxium at around 6kt of over 60\%.
+However, in nominal terms, the greatest differences occurred at slightly lower speeds.
+Under central assumptions, the propeller submarine had an advantage in submerged range of some 200nm at 5kt, and approximately a two-day advantage in dived endurance at 4kt.
+
+The full range of plausible efficiency curves exhibited a significant efficiency advantage right through transit speeds.
+Central assumptions suggested that this would result in the overall range of a submarine being enhanced by approximately 50\% with the use of propellers instead of a pumpjet.
+None of the scenarios suggested it was likely that this range advantage would be any less than 20\%.
+Given the significant emphasis that has been repeatedly placed the unique requirements for range and endurance for Australia's submarine, this raises substantial questions about suitability of this technology to meet Australia's needs.
+
+The model was used to test a wide range of different assumptions.
+A further key result of the study established the sensitity of the difference to different assumptions regarding hotel load, and battery energy density.
+This showed that the difference between the two systems was made substantially larger with smaller hotel loads, and larger stored electical energy (as might be achieved with lithium-ion batteries), and these two effects combined to give some particularly startling results.
+Circumstances where the \textit{difference} in dived range and endurance between the two systems in terms of range and endurance exceeded the \textit{entire} dived range and endurance achieved by many modern conventional submarines frequently arose.
+
+The model demonstrated clearly that all of the most pronounced advantages in dived range and endrance occurred at speeds around 3-7kt, generally within plausible speeds achievable with modern AIP systems.
+The addition of AIP was briefly considered which revealed that a difference in dived endurance of the order of one month, or thousands of nautical miles, would likely have emerged between the two propulsion systems.
+
+The consequences of flow separation also has acoustic consequences, which this report explores.
+The vortexes and eddies which are created by flow separation tend to be unsteady, and shed periodically into the main flow.
+This results in pressure fluctations inside the shroud and more intense, unsteady large-scale types of turbulence occurring as these vortices are ingested into the rotor blades.
+These effects are are likely to cause some accoustic signature which becomes lower in frequence as the submarine slows down further and the vortexes become larger, and shed more slowly.
+This is in contrast to early stages of cavitation, which would be high frequency at when bubble sizes are small, and move to lower fequences as speed increases and bubbles expand.
+Lower frequencies are of particular significance because they tend to propagate further, and would be less effectively shielded by the duct.
+
+It remains plausible that in the absence of cavitation the noise generated in the propulsor by such flow separation is not loud enough to be of significant tactical concern.
+However, it is extremely unlikely that at some very low speeds, where conventional propellers experience no cavitation and enjoy steady, smooth flows over the blades, that a pumpjet could actually have a lower acoustic signature, even in terms of radiated noise.
+Flow separtion demand that two or three times as much energy is wasted by a pumpjet creating and destroying these unsteady vortexes than a propeller would require to create equal thrust.
+Consequently, the claim that pumpjets are generally acoustically superior should be treated with some caution.
+This claim has strong grounds wherever a conventional propeller might experience cavitation, such as at higher speeds.
+But at some very low speeds it is unlikely to be true.
+
+Overall, it seems unlikely that the full range of consequences of pumpjet choice have been fully comprehended by the Australian government.
+The scale of the probable impact on range and endurance is quite probably so substantial that it is difficult to see how such a performance penalty is consistent with the government's stated aim of acquiring a regionally superior submarine.
 

### 2018-01-08 DraftExec.Rnw

@@
 
-There are other some fundamental limitations on the peak efficiency that can be achieved by jets based on simple calculations of ideal efficiency from momentum theory, which are discussed in the next subsection.
+There are other some fundamental limitations on the peak efficiency that can be achieved by jets based on simple calculations of ideal efficiency from momentum theory, which are discussed in Section \ref{tradeoffs} (\nameref{tradeoffs}).
 They also provide enduring, reliable limitations on the extent to which ingenuity or design changes might be able to improve the efficiency of jets at low speed.
@@
 As can be seen by the diagram in Figure \ref{fig:JetLowFlow.png}, regions abound which are adjacent to or behind flow blockages or reversing flow, where considerable turbulence and unsteady forces would exist.
-These effects also have acoustic consequences, which will be discussed in Section \ref{Acoustic Considerations}.
+These effects also have acoustic consequences, which will be discussed in Section \ref{acousticconsiderations} (\nameref{acousticconsiderations}).
 

### 2018-01-08 DraftExec.Rnw
add imgs and rnw

--- 2018-01-08 DraftExec.Rnw
@@
 
-\usepackage{nameref}
+%  For caption width
+\usepackage{graphicx,caption}
+% For Centering figs
+\usepackage[export]{adjustbox}
+% For split figures
+\usepackage{subcaption}
 
@@
 @
-
-Para
 
@@
 
-\begin{figure}
+\begin{figure}[h]
 \includegraphics[width=\textwidth]{XenonPoison.png}
@@
 
-\begin{figure}
-\includegraphics[width=\textwidth]{BoundaryLayer.png}
+\begin{figure}[h]
+\captionsetup{width=0.5\textwidth}
+\includegraphics[width=0.5\textwidth, center]{BoundaryLayer.png}
 \caption{The development of a boundary layer as shown in \parencite{mollard2011}}
@@
 
-\begin{figure}
-\includegraphics[width=\textwidth]{LaminarTurbulent.png}
+\begin{figure}[h]
+
+\begin{subfigure}{0.5\textwidth}
+\includegraphics[width=0.7\linewidth, center]{Laminar.png}
+\caption{Laminar}
+\end{subfigure}
+\begin{subfigure}{0.5\textwidth}
+\includegraphics[width=0.7\linewidth, center]{Turbulent.png}
+\caption{Turbulent}
+\end{subfigure}
+
 \caption{A simple comparison of laminar and turbulent flows from \parencite{NPdotnet2017}}
@@
 
-\begin{figure}
-\includegraphics[width=\textwidth]{EarlyCavitation.JPG}
-\caption{Cavitation bubbles form and collapse creating noise and some tiny bubbles long before boiling occurs}
+\begin{figure}[h]
+
+\begin{subfigure}{0.5\textwidth}
+\includegraphics[width=0.9\linewidth, height=5cm]{EarlyCavitation.JPG}
+\caption{}
+\end{subfigure}
+\begin{subfigure}{0.5\textwidth}
+\includegraphics[width=0.9\linewidth, height=5cm]{BoilingCavitation.JPG}
+\caption{}
+\end{subfigure}
+
+\caption{(a) Cavitation bubbles form and collapse creating noise and some tiny bubbles long before boiling occurs.
+(b) Only at very high temperatures do the bubbles endure at full-size in the water and reach the surface.}
 \label{fig:EarlyCavitation.JPG}
@@
 Even at higher temperature, the rougher parts of the surface provide the points where all the cavitation originate.
-
-\begin{figure}
-\includegraphics[width=\textwidth]{BoilingCavitation.JPG}
-\caption{Only at very high temperatures do the bubbles endure at full-size in the water and reach the surface}
-\label{fig:BoilingCavitation.JPG}
-\end{figure}
 
@@
 
-\begin{figure}
-\includegraphics[width=0.7\textwidth]{IncreasingCavitation.png}
+\begin{figure}[h]
+\captionsetup{width=0.7\textwidth}
+\includegraphics[width=0.7\textwidth, center]{IncreasingCavitation.png}
 \caption{Cavitation frequently occurs first at the extremities, then spreads inwards across the blad face, as shown in \parencite{mollard2011}}
@@
 
-\begin{figure}
-\includegraphics[width=0.7\textwidth]{CavitationTypes.png}
+\begin{figure}[h]
+\captionsetup{width=0.7\textwidth}
+\includegraphics[width=0.7\textwidth, center]{CavitationTypes.png}
 \caption{Different types of cavitation can occur on different parts of the propulsion system as shown in \parencite{carlton2007}}
@@
 
-\begin{figure}
-\includegraphics[width=0.5\textwidth]{SeparatedAerofoil.jpg}
+\begin{figure}[h]
+\captionsetup{width=0.5\textwidth}
+\includegraphics[width=0.5\textwidth, center]{SeparatedAerofoil.jpg}
 \caption{Flow separation occurring over the top of an aerofoil in a wind tunnel. Image courtesy of Deutsches Zentrum fuer Luft- und Raumfahrt e. V. (DLR)}
@@
 
-\begin{figure}
-\includegraphics[width=\textwidth]{FlowSeparation.png}
+\begin{figure}[h]
+\captionsetup{width=0.7\textwidth}
+\includegraphics[width=0.7\textwidth, center]{FlowSeparation.png}
 \caption{Flow separation involves flow reversal near a surface when a fluid is moving against an adverse pressure gradient, as shown in \parencite{mollard2011}}
@@
 
-\begin{figure}
-\includegraphics[width=\textwidth]{PropellerAction.png}
+\begin{figure}[h]
+\captionsetup{width=0.5\textwidth}
+\includegraphics[width=0.5\textwidth, center]{PropellerAction.png}
 \caption{Propellers generate thrust by accelerating a column of water, as shown in  \parencite[247]{mollard2011}}
@@
 
-\begin{figure}
-\includegraphics[width=\textwidth]{SkewRake.png}
+\begin{figure}[h]
+\captionsetup{width=0.5\textwidth}
+\includegraphics[width=0.5\textwidth, center]{SkewRake.png}
 \caption{The development of a boundary layer \parencite[262]{mollard2011}}
@@
 
-\begin{figure}
-\includegraphics[width=\textwidth]{HelicalLine.png}
+\begin{figure}[h]
+\captionsetup{width=0.5\textwidth}
+\includegraphics[width=0.5\textwidth, center]{HelicalLine.png}
 \caption{A blade section of a propeller traces a helical line around a cylinder as it is rotated \parencite{carlton2007}}
@@
 
-\begin{figure}
-\includegraphics[width=\textwidth]{PitchDefinition.png}
+\begin{figure}[h]
+\captionsetup{width=0.7\textwidth}
+\includegraphics[width=0.7\textwidth, center]{PitchDefinition.png}
 \caption{The definition of pitch is the distance traveled by a blade section along a cylinder, as given in \parencite{carlton2007}}
@@
 
-\begin{figure}
-\includegraphics[width=\textwidth]{CPPropeller.png}
+\begin{figure}[h]
+\captionsetup{width=0.8\textwidth}
+\includegraphics[width=0.8\textwidth, center]{CPPropeller.png}
 \caption{Controllable Pitch Propellers (CP Propellers) allow the pitch of the blades to be varied. Image credit: \parencite{mollard2011}}
@@
 
-\begin{figure}
-\includegraphics[width=\textwidth]{OpenWater.png}
+\begin{figure}[h]
+\captionsetup{width=0.8\textwidth}
+\includegraphics[width=0.8\textwidth]{OpenWater.png}
 \caption{A typical chart showing open water efficiency of a propeller, in this case a Wageningen B5-75. Image: MARIN}
@@
 
-\begin{figure}
-\includegraphics[width=\textwidth]{SlipRatio.png}
+\begin{figure}[h]
+\captionsetup{width=0.8\textwidth}
+\includegraphics[width=0.8\textwidth, center]{SlipRatio.png}
 \caption{The slip ratio of a propeller. Image credit: \cite{MAN2017}}
@@
 
-\begin{figure}
+\begin{figure}[h]
+\captionsetup{width=\textwidth}
 \includegraphics[width=\textwidth]{AccelDecel.png}
-\caption{he two types of ducted propeller, accelerating (left) and decelerating (right) as shown in \parencite[249]{mollard2011}}
+\caption{The two types of ducted propeller, accelerating (left) and decelerating (right) as shown in \parencite[249]{mollard2011}}
 \label{fig:AccelDecel.png}
@@
 
-\begin{figure}
-\includegraphics[width=\textwidth]{DuctStreamlines.png}
+\begin{figure}[h]
+\captionsetup{width=0.7\textwidth}
+\includegraphics[width=0.7\textwidth, center]{DuctStreamlines.png}
 \caption{The streamlines of flow through different duct types \parencite[214]{lewis1988}}
@@
 
-\begin{figure}
-\includegraphics[width=\textwidth]{CavitatingDuct.png}
+\begin{figure}[h]
+\captionsetup{width=0.7\textwidth}
+\includegraphics[width=0.7\textwidth, center]{CavitatingDuct.png}
 \caption{An example of the cavitation occurring at the blade tips of a ducted propeller. Image credit \cite{moulijn2015}}
@@
 
-\begin{figure}
-\includegraphics[width=\textwidth]{DuctedEfficiency.png}
+\begin{figure}[h]
+\captionsetup{width=0.7\textwidth}
+\includegraphics[width=0.7\textwidth, center]{DuctedEfficiency.png}
 \caption{Accelerating ducts produce a net positive thrust at low advance ratios, but a net negative thrust at higher advance ratios as the drag on the nozzle exterior increases.
@@
 \cite[24-25]{oosterveld1970} provides a more thorough derivation of the circumstances in which this can be reliably achieved, which generally include larger blade area ratios or with more rows of rotors for higher loading, or circumstances where the loading of the propeller is quite low.
-This creates an inevitable and necessary tension between optimisation of a duct for propulsive efficiency and reduction of cavitation, which will be returned to in Section \ref{waterjets} (\nameref{waterjets}).
+This creates an inevitable and necessary tension between optimisation of a duct for propulsive efficiency and reduction of cavitation, which will be returned to in Section \ref{waterjets}.
 
@@
 
-\begin{figure}
-\includegraphics[width=\textwidth]{PumpjetOutline.png}
+\begin{figure}[h]
+\captionsetup{width=0.4\textwidth}
+\includegraphics[width=0.4\textwidth, center]{PumpjetOutline.png}
 \caption{In \cite{carlton2007} the pump jet is described as a development of a classical ducted propeller, with the noticeable addition of a stator}
@@
 
-\begin{figure}
-\includegraphics[width=\textwidth]{PumpjetWaterjet.png}
-\caption{A comparison of a pumpjet and waterjet as given in \cite{wislicenus1973}}
+\begin{figure}[h]
+
+\begin{subfigure}{0.5\textwidth}
+\includegraphics[width=0.9\linewidth]{Pumpjet.png}
+\caption{}
+\end{subfigure}
+\begin{subfigure}{0.5\textwidth}
+\includegraphics[width=0.9\linewidth]{Waterjet.png}
+\caption{}
+\end{subfigure}
+
+\caption{A comparison of a pumpjet (a) and waterjet (b) as given in \cite{wislicenus1973}}
 \label{fig:PumpjetWaterjet.png}
@@
 
-\begin{figure}
-\includegraphics[width=\textwidth]{EfficiencyJapan.png}
+\begin{figure}[h]
+\captionsetup{width=0.7\textwidth}
+\includegraphics[width=0.7\textwidth, center]{EfficiencyJapan.png}
 \caption{A graph from a Japanese study of waterjet efficiency shows efficiency rapidly declining below craft velocities of 10m/s, in accordance with theory. \parencite{fujisawa1995}}
@@
 
-\begin{figure}
-\includegraphics[width=\textwidth]{EfficiencyTransport.png}
+\begin{figure}[h]
+\captionsetup{width=0.6\textwidth}
+\includegraphics[width=0.6\textwidth, center]{EfficiencyTransport.png}
 \caption{An Australian experimental study aimed at determining the crossover point for the efficiency of jets and propellers in the 20-30kt range revealed much higher power was required across the entire speed range \parencite{kamal2015}}
@@
 
-\begin{figure}
+\begin{figure}[h]
+\captionsetup{width=\textwidth}
 \includegraphics[width=\textwidth]{BMTShape.png}
@@
 
-\begin{figure}
-\includegraphics[width=\textwidth]{BMTEfficiency.png}
+\begin{figure}[p]
+
+\begin{subfigure}[b]{0.9\textwidth}
+\includegraphics[width=0.8\linewidth, center]{BMTEfficiency1.png}
+\end{subfigure}
+\begin{subfigure}[b]{0.9\textwidth}
+\includegraphics[width=0.8\linewidth, center]{BMTEfficiency2.png}
+\end{subfigure}
+
 \caption{Studies on submerged waterjets for military surface craft currently show that the efficiency of jets decline sharply below 10kt \parencite{giles2010}}
@@
 
-There are other some fundamental limitations on the peak efficiency that can be achieved by jets based on simple calculations of ideal efficiency from momentum theory, which are discussed in Section \ref{tradeoffs} (\nameref{tradeoffs}).
+There are other some fundamental limitations on the peak efficiency that can be achieved by jets based on simple calculations of ideal efficiency from momentum theory, which are discussed in Section \ref{tradeoffs}.
 They also provide enduring, reliable limitations on the extent to which ingenuity or design changes might be able to improve the efficiency of jets at low speed.
@@
 
-As can be seen in figure \ref{fig:JetHighFlow.png}, the duct of a jet is designed to operate by deflecting the flow away from the natural path that would occur for a propeller, which would accelerate the stream over its blades directly, with the flow necessarily contracting with the acceleration as shown in Figure \ref{fig:PropellerAction.png}.
+As can be seen in Figure \ref{fig:JetHighFlow.png}, the duct of a jet is designed to operate by deflecting the flow away from the natural path that would occur for a propeller, which would accelerate the stream over its blades directly, with the flow necessarily contracting with the acceleration as shown in Figure \ref{fig:PropellerAction.png}.
 Instead, pumpjets tend to use some moderate expansion of the flow in advance of the impeller (diffusion) in order to raise its pressure, and reduce its velocity.
@@
 \begin{figure}
-\includegraphics[width=\textwidth]{FlowComparison.png}
+\captionsetup{width=0.4\textwidth}
+\includegraphics[width=0.4\textwidth, center]{FlowComparison.png}
 \caption{The effect of the duct is to deflect the flow away from the streamlines of an open propeller, elevating the static pressure at the stator blades by slowing the flow down.}
@@
 
-\begin{figure}
-\includegraphics[width=\textwidth]{JetHighFlow.png}
+\begin{figure}[p]
+\captionsetup{width=1\textwidth}
+\includegraphics[width=1\textwidth]{JetHighFlow.png}
 \caption{At high flow rates the water has sufficient velocity (indicated by arrow length) to overcome the adverse pressure gradients created by the duct shape, shown by the change from green to red.
@@
 As can be seen by the diagram in Figure \ref{fig:JetLowFlow.png}, regions abound which are adjacent to or behind flow blockages or reversing flow, where considerable turbulence and unsteady forces would exist.
-These effects also have acoustic consequences, which will be discussed in Section \ref{acousticconsiderations} (\nameref{acousticconsiderations}).
-
-\begin{figure}
-\includegraphics[width=\textwidth]{JetLowFlow.png}
+These effects also have acoustic consequences, which will be discussed in Section \ref{acousticconsiderations}.
+
+\begin{figure}[p]
+\captionsetup{width=1\textwidth}
+\includegraphics[width=1\textwidth]{JetLowFlow.png}
 \caption{At low flow rates the water has insufficient velocity to overcome the adverse pressure gradients created by the duct shape, and flow separation begins to occur on the duct intake, and between the blades beginning near the roots (stall).
@@
 
-\begin{figure}
+\begin{figure}[p]
 \includegraphics[width=\textwidth]{Comparison.png}
@@
 
-\begin{figure}
-\includegraphics[width=\textwidth]{Triomph.jpg}
+\begin{figure}[p]
+\captionsetup{width=0.8\textwidth}
+\includegraphics[width=0.8\textwidth, center]{Triomph.jpg}
 \caption{The outer shape of pumpjets on the aft end of torpedoes or submarines, such as this one on the French Triomphant class on which the Barracuda's jet will be based, may understate the degree of deceleration that the duct achieves.
@@
 
-\begin{figure}
-\includegraphics[width=\textwidth]{PumpjetEfficiency.png}
+\begin{figure}[h]
+\captionsetup{width=0.6\textwidth}
+\includegraphics[width=0.7\textwidth, center]{PumpjetEfficiency.png}
 \caption{Ideal jet efficiencies sharply appoach zero at low values of $\Delta V / V_0$, where as ideal propeller efficiencies without a duct $K = 0$ approach unity. Credit: \cite{wislicenus1973}}
@@
 
-\begin{figure}
-\includegraphics[width=\textwidth]{EfficiencyBrandau.png}
+\begin{figure}[h]
+\captionsetup{width=0.5\textwidth}
+\includegraphics[width=0.5\textwidth, center]{EfficiencyBrandau.png}
 \caption{Alternative derivations of ideal jet efficiency also show sharply declining efficiency for low values of $\Delta V / V_0$ Credit: \cite{brandau1967}}
@@
 
-\begin{figure}
-\includegraphics[width=\textwidth]{EfficiencyMollard.png}
+\begin{figure}[h]
+\captionsetup{width=0.6\textwidth}
+\includegraphics[width=0.6\textwidth, center]{EfficiencyMollard.png}
 \caption{Other modern reference texts might demonstrate jet or propeller efficiencies with different x-axes, effectively reversing the way the graph should be understood for a single system operating at different speeds.
@@
 
-\begin{figure}
-\includegraphics[width=\textwidth]{EfficiencyBulten.png}
+\begin{figure}[h] \captionsetup{width=0.6\textwidth}
+\includegraphics[width=0.8\textwidth, center]{EfficiencyBulten.png}
 \caption{Another graph representing efficiency as a function of the ratio of jet velocity to incoming velocity, which shows efficiency declining at low thrust (to the right).
@@
 
-\begin{figure}
+\begin{figure}[h] \captionsetup{width=0.5\textwidth}
 \includegraphics[width=\textwidth]{PropDuct.png}
@@
 
-\begin{figure}
-\includegraphics[width=\textwidth]{PumpUp.png}
+\begin{figure}[h]
+\captionsetup{width=0.5\textwidth}
+\includegraphics[width=0.6\textwidth, center]{PumpUp.png}
 \caption{The mysterious efficiency comparison chart which has been doing the rounds in the Australian debate probably quite reasonably reflects theoretical physical limits of 'ideal' pumpjets and propellers.
@@
 
-\begin{figure}
-\includegraphics[width=\textwidth]{stall.png}
+\begin{figure}[h]
+\captionsetup{width=0.6\textwidth}
+\includegraphics[width=0.6\textwidth, center]{stall.png}
 \caption{Flow separation, or stall, can occur when the flow over the blades of a pump slows too much, as shown in \cite{li2013}}
@@
 
-\begin{figure}
-\includegraphics[width=\textwidth]{backflow.png}
+\begin{figure}[h]
+\captionsetup{width=0.8\textwidth}
+\includegraphics[width=0.8\textwidth, center]{backflow.png}
 \caption{Stall results in water flowing backwards near the hub of the blades at low flow rates, as shown in \cite{li2014rotating}}
@@
 
-\begin{figure}
+\begin{figure}[h] \captionsetup{width=0.5\textwidth}
 \includegraphics[width=\textwidth]{AcceleratingIdeal.png}
@@
 
-\begin{figure}
-\includegraphics[width=\textwidth]{FineOptimised.png}
-\caption{Extensive work has been done exploring compromise designs for decelerating ducted propellers (pumpjets), including with very modest deceleration in the intake, limited shroud taper, short duct lengths, or essentially incporate some of the advantages of accelerating ducts.
+\begin{figure}[h]
+
+\begin{subfigure}{0.5\textwidth}
+\includegraphics[width=0.9\linewidth]{FineOptimised.png}
+\caption{}
+\end{subfigure}
+\begin{subfigure}{0.5\textwidth}
+\includegraphics[width=0.9\linewidth]{FineCavitation.png}
+\caption{}
+\end{subfigure}
+
+\caption{(a) Extensive work has been done exploring compromise designs for decelerating ducted propellers (pumpjets), including with very modest deceleration in the intake, limited shroud taper, short duct lengths, or essentially incporate some of the advantages of accelerating ducts.
+(b) TAt optimally efficient conditions (corresponding to 27kt) such designs still exhibit some cavitation.
 Credit: \cite{abdel2010}}
@@
 
-\begin{figure}
-\includegraphics[width=\textwidth]{FineEfficiency.png}
+\begin{figure}[h]
+\captionsetup{width=0.7\textwidth}
+\includegraphics[width=0.7\textwidth, center]{FineEfficiency.png}
 \caption{In experimental testing such designs still exhibit sharply falling effiency at high advance ratios, or low speed. Credit:  \cite{abdel2010}}
 \label{fig:FineEfficiency.png}
-\end{figure}
-
-\begin{figure}
-\includegraphics[width=\textwidth]{FineCavitation.png}
-\caption{TAt optimally efficient conditions (corresponding to 27kt) such designs still exhibit some cavitation. Credit: \cite{abdel2010}}
-\label{fig:FineCavitation.png}
 \end{figure}
@@
 
-\begin{figure}
-\includegraphics[width=0.5\textwidth]{LowIVR.png}
-\caption{A an intake may experience risks of flow separation even at low speeds (Inlet Velocity Ratio) on some surfaces due to certain flow deflections being more severe.
+\begin{figure}[h]
+
+\begin{subfigure}{0.5\textwidth}
+\includegraphics[width=0.9\linewidth]{LowIVR.png}
+\caption{}
+\end{subfigure}
+\begin{subfigure}{0.5\textwidth}
+\includegraphics[width=0.9\linewidth]{HighIVR.png}
+\caption{}
+\end{subfigure}
+
+\caption{(a) A an intake may experience risks of flow separation even at low speeds (Inlet Velocity Ratio) on some surfaces due to certain flow deflections being more severe.
+(b) A an intake may experience risk of flow separation on different surfaces at high speeds (Inlet Velocity Ratio) where other flow deflections are more severe.
 Source: \cite{bulten2006numerical}}
@@
 
-\begin{figure}
-\includegraphics[width=\textwidth]{HighIVR.png}
-\caption{A an intake may experience risk of flow separation on different surfaces at high speeds (Inlet Velocity Ratio) where other flow deflections are more severe.
-Source: \cite{bulten2006numerical}}
-\label{fig:HighIVR.png}
-\end{figure}
-
-\begin{figure}
+\begin{figure}[h]
 \includegraphics[width=\textwidth]{DuctSeparation.png}
@@
 
-\begin{figure}
-\includegraphics[width=\textwidth]{NozzleDrag.png}
+\begin{figure}[h]
+\captionsetup{width=0.7\textwidth}
+\includegraphics[width=0.7\textwidth, center]{NozzleDrag.png}
 \caption{Dramatically reducing the length of a duct relative to its diameter ($L/D$) as shown in \cite{oosterveld1970} can reduce its drag, which has the most significant impact at low thrusts or speeds.}
@@
 
-\begin{figure}
-\includegraphics[width=\textwidth]{LaminarProp.png}
+\begin{figure}[h]
+\captionsetup{width=0.4\textwidth}
+\includegraphics[width=0.4\textwidth, center]{LaminarProp.png}
 \caption{At very low speeds changes in regimes such as an increase in laminar flow over the blade section could lead to small changes in efficiency, perhaps explaining the distinct rise at low speeds for the properller at ~5kt in Figure \ref{fig:BMTEfficiency.png}}
@@
 
-\begin{figure}
+\begin{figure}[p]
 <<plot_efficiency, echo=FALSE, warning=FALSE, message=F >>=
@@
 
-\begin{figure}
+\begin{figure}[p]
 <<plot_power1, echo=FALSE, warning=FALSE, message=F >>=
@@
 
-\begin{figure}
+\begin{figure}[p]
 <<plot_results1, echo=FALSE, warning=FALSE, message=F >>=
@@
 
-\begin{figure}
+\begin{figure}[p]
 <<plot_results2, echo=FALSE, warning=FALSE, message=F >>=
@@
 
-\begin{figure}
+\begin{figure}[p]
 <<plot_results3, echo=FALSE, warning=FALSE, message=F >>=
@@
 
-\begin{figure}
+\begin{figure}[p]
 <<plot_results4, echo=FALSE, warning=FALSE, message=F >>=
@@
 
-\begin{figure}
+\begin{figure}[p]
 <<plot_proportions, echo=FALSE, warning=FALSE, message=F >>=
@@
 \subsection{Sensitivity to Hotel Load}
+
 As can be seen in Figure \ref{fig:sens_hotel} the difference between the two systems is very sensitive to changes in hotel load, both in proportional and nominal terms.
@@
 
-The difference between pumpjet and propeller propulsion systems in terms of range tends to be found at slightly higher speeds, though this point also moves towards higher speeds as hotel load increases.
-With a reduced hotel load of 50kw, the difference between the two systems in the middle scenario is about 370nm for speeds 3.5-4.5kt, substantially greater than the ~200nm around 4.5-5.5kt at a hotel load of 100kW.
-
-It's important to acknowledge that such substantial differences in hotel load merit serious consideration, both in terms system design, as well as for variance in operation of a submarine with a given system design.
-As has been discussed, the precise hotel load of submarines is not generally publically available, which justifies senitivity testing around a relatively wide potential range.
-The size of the crew required for different designs might vary considerably, as it also would for different missions (for example with the embarkation of a special forces congingent) which would necessarily have considerable impact on energy demand for life support, including sanitation, CO2 scrubbing, and cooking.
-In addition, it is broadly understood that the AN/BYG-1 combat system consumes substantially more power than other alternatives, reportedly in the order of 100kW \parencite{patrick2011}.
-
-Furthermore, the hotel load may be varied considerably at the captain's discretion depending on operational requirements.
-Not all sensors or components of the combat system need to run at all times, for example during long transits in order to reach an operational area, or during a long dive in order to escape an operational area once a large strike has already been conducted.
-In wartime, when surfacing to snort within range of enemy airpower or close to naval bases might result in the sumbarine probably being sunk, the crew might willingly forego hot showers, hot meals, and climate-controlled air in order to increase their chances of survival.
-Consequently, it is perfectly plausible that changes of hotel load in the order of a factor of two for tactically significant time periods (days) is achievable.
-The results of this sensitivity analysis show that the penalty inflicted by the pumpjet will be substantially greater in such circumstances.
-
-The proportional impact of the changes in hotel load area also substantial, and exhibit clearly the same trend of the impact being larger at higher speeds for higher hotel loads.
-For the efficiency curve pairs that comprise the central selections, decreasing the hotel load to 50kt would see the dived endurance and range of the submarine improved by the adoption of propellers 80\% between 4 and 5.5kt.
-A much higher hotel load, at 200kW, would see that peak difference over 40\% between 5.5 and 8kt.
-
-It is also clear that increasing the hotel load has a greatly diminised impact on the proportional differences between the systems at higher speeds.
-At very high speeds (e.g.
-15kt) changes in the hotel load have almost no impact in proportional terms to the difference, which lies at about 30\% in all cases for the central selection of curves.
-At 9kt, in the region where transit speed is likely to occur, increasing the hotel load by a factor of four (from 50kW to 200kW) only decreases the advantage of the propeller from 48\% to 38\%.
-Therefore one might say near certainty that any difference between efficiency around these patrol speeds is likely to have a substantial impact on the overall range of the submarine, with differences of much less than 30\% quite unlikely, irrespective of other changes in the submarine design or operation circumstance.
-
-It should also be noted that sensitivity testing has not been shown in this paper around alternative values of the drag coefficient $C_d$ which was solved for by matching peak power output and top speed.
-It is considered relatively unlikely that such dramatic changes in this constant are plausible.
-Furthermore, changing this constant will have a very similar impact to alerations in hotel load, in terms of its affect on the speed where greatest differences will occur and their probable magnitude.
-Consequently, consideration of the sensitivity to hotel load is likely to easily encompass the plausible extent, and nature, of the likely uncertainty in this factor.
-
-\begin{figure}
+\begin{figure}[p]
 <<plot_sens_hotel, echo=FALSE, warning=FALSE, message=F >>=
@@
 
+The difference between pumpjet and propeller propulsion systems in terms of range tends to be found at slightly higher speeds, though this point also moves towards higher speeds as hotel load increases.
+With a reduced hotel load of 50kw, the difference between the two systems in the middle scenario is about 370nm for speeds 3.5-4.5kt, substantially greater than the ~200nm around 4.5-5.5kt at a hotel load of 100kW.
+
+It's important to acknowledge that such substantial differences in hotel load merit serious consideration, both in terms system design, as well as for variance in operation of a submarine with a given system design.
+As has been discussed, the precise hotel load of submarines is not generally publically available, which justifies senitivity testing around a relatively wide potential range.
+The size of the crew required for different designs might vary considerably, as it also would for different missions (for example with the embarkation of a special forces congingent) which would necessarily have considerable impact on energy demand for life support, including sanitation, CO2 scrubbing, and cooking.
+In addition, it is broadly understood that the AN/BYG-1 combat system consumes substantially more power than other alternatives, reportedly in the order of 100kW \parencite{patrick2011}.
+
+Furthermore, the hotel load may be varied considerably at the captain's discretion depending on operational requirements.
+Not all sensors or components of the combat system need to run at all times, for example during long transits in order to reach an operational area, or during a long dive in order to escape an operational area once a large strike has already been conducted.
+In wartime, when surfacing to snort within range of enemy airpower or close to naval bases might result in the sumbarine probably being sunk, the crew might willingly forego hot showers, hot meals, and climate-controlled air in order to increase their chances of survival.
+Consequently, it is perfectly plausible that changes of hotel load in the order of a factor of two for tactically significant time periods (days) is achievable.
+The results of this sensitivity analysis show that the penalty inflicted by the pumpjet will be substantially greater in such circumstances.
+
+The proportional impact of the changes in hotel load area also substantial, and exhibit clearly the same trend of the impact being larger at higher speeds for higher hotel loads.
+For the efficiency curve pairs that comprise the central selections, decreasing the hotel load to 50kt would see the dived endurance and range of the submarine improved by the adoption of propellers 80\% between 4 and 5.5kt.
+A much higher hotel load, at 200kW, would see that peak difference over 40\% between 5.5 and 8kt.
+
+It is also clear that increasing the hotel load has a greatly diminised impact on the proportional differences between the systems at higher speeds.
+At very high speeds (e.g.
+15kt) changes in the hotel load have almost no impact in proportional terms to the difference, which lies at about 30\% in all cases for the central selection of curves.
+At 9kt, in the region where transit speed is likely to occur, increasing the hotel load by a factor of four (from 50kW to 200kW) only decreases the advantage of the propeller from 48\% to 38\%.
+Therefore one might say near certainty that any difference between efficiency around these patrol speeds is likely to have a substantial impact on the overall range of the submarine, with differences of much less than 30\% quite unlikely, irrespective of other changes in the submarine design or operation circumstance.
+
+It should also be noted that sensitivity testing has not been shown in this paper around alternative values of the drag coefficient $C_d$ which was solved for by matching peak power output and top speed.
+It is considered relatively unlikely that such dramatic changes in this constant are plausible.
+Furthermore, changing this constant will have a very similar impact to alerations in hotel load, in terms of its affect on the speed where greatest differences will occur and their probable magnitude.
+Consequently, consideration of the sensitivity to hotel load is likely to easily encompass the plausible extent, and nature, of the likely uncertainty in this factor.
+
 \subsection{Sensitivity to Battery Changes}
@@
 
-The results in nominal terms are striking in their magnitude.
-In terms of enduracy, doubling the stored electrical energy, for example by adopting lithium batteries, increase the \textit{difference} between the two systems under the central efficiency curve assumptions to nearly four days (93 hours) at 4kt.
-Given that most modern, capable submarines are estimated to only be able to achieve dived endurances between 60 and 100 hours on batteries at 4kt \parencite{buckingham2008submarine} this cannot be dismissed in terms of its tactical significance.
-Even under the eficiency curve assumptions representing the narrowest spearation considered, the difference in this case would be 59 hours, almost equal to the estimated dived endurance of a Type214 operating without AIP.
-
-The in terms of submerged range, the same proportional impact is also extremely significant.
-Doubling the energy stored inceases the \textit{difference} between two otherwise identical submarines to over 400nm under central assumptions at 4.5kt.
-That's a larger distance than the width of the Java Sea, the most central body of water in the Indonesian Archipelago, or most of the length of the Malacca Strait.
-The tactical significance of such distances are not likely to be lightly dismissed.
-If far smaller amounts of energy are embarked, the nominal differences reduce to perhaps 100nm, in direct proportion to the reduction of the submarine's dived endurance.
-This is still a non-trivial distance in terms of it's tactical signicance, and comprises a substantial fraction of the dived ranges of most non-AIP submarines.
-It still represents a significant difference in operational terms, equal approximately to the width of Hainan Island, or neraly twice the length of the Sunda Strait.
-
-The singificant range of different energy density values tested here reflect quite plausibly some of ranges of different energy levels available in operation.
-For instance, as has been discussed \parencite{briggs2016, greenfield2016}, lead acid batteries may well only be able to practically utilise about 50\% of their total energy stores for tactical operaions, unless operating in a very low-threat environment and undertaking very slow charging to reach the maximum full charge is possible.
-In addition, the plausible improvement in enery density of lithium batteries may well be in the order of a factor of two in practice \parencite{greenfield2016}, though theoretical limits of the chemistry suggest much higher may be possible \parencite{davies2016}.
-
-Only brief is given here to the impact of Air Independent Propulsion, which has substantially higher energy densities possible even than Lithium, and also amounts to the addtional embarkation of greater mass.
-The volumetric energy density of the feedstocks to modern AIP systems (liquid oxygen and methanol) is estimated to be approximately ten times greater than that of lead-acid batteries \parencite{davies2016}, around 4.5MJ/L.
-Given the greater mass embarked, and additional studies suggesting that at constant speeds they might extend the dived endurance of a modern submarine from around 60 hours to over 1000 hours (a factor of roughly sixteen) an approximate estimation of the the direction and magnitude of this impact may by scaling up energy density by this factor.
-This would amount to adding three further frames to the right of Figure \ref{fig:sens_battery}, with the energy density being doubled in each successive frame.
-This would plausibly (under the central line selections) amount to more than month of dived endurance \textit{difference} between AIP inclusive submarines with propellers or pumpjets, amounting to thousands of nautical miles of continuously submerged travel.
-This difference would also now be material in terms of the overall range of the submarine amounting to a further 15-20\% reduction in the submarine's overall range.
-
-With such massive differences between the performance of the two systems emerging with the presence of the AIP, it is difficult to see how one could realistically consider a pumpjet as being competitive if a direct comparison were made.
-
-\begin{figure}
+\begin{figure}[p]
 <<plot_sens_battery, echo=FALSE, warning=FALSE, message=F >>=
@@
 
+The results in nominal terms are striking in their magnitude.
+In terms of enduracy, doubling the stored electrical energy, for example by adopting lithium batteries, increase the \textit{difference} between the two systems under the central efficiency curve assumptions to nearly four days (93 hours) at 4kt.
+Given that most modern, capable submarines are estimated to only be able to achieve dived endurances between 60 and 100 hours on batteries at 4kt \parencite{buckingham2008submarine} this cannot be dismissed in terms of its tactical significance.
+Even under the eficiency curve assumptions representing the narrowest spearation considered, the difference in this case would be 59 hours, almost equal to the estimated dived endurance of a Type214 operating without AIP.
+
+The in terms of submerged range, the same proportional impact is also extremely significant.
+Doubling the energy stored inceases the \textit{difference} between two otherwise identical submarines to over 400nm under central assumptions at 4.5kt.
+That's a larger distance than the width of the Java Sea, the most central body of water in the Indonesian Archipelago, or most of the length of the Malacca Strait.
+The tactical significance of such distances are not likely to be lightly dismissed.
+If far smaller amounts of energy are embarked, the nominal differences reduce to perhaps 100nm, in direct proportion to the reduction of the submarine's dived endurance.
+This is still a non-trivial distance in terms of it's tactical signicance, and comprises a substantial fraction of the dived ranges of most non-AIP submarines.
+It still represents a significant difference in operational terms, equal approximately to the width of Hainan Island, or neraly twice the length of the Sunda Strait.
+
+The singificant range of different energy density values tested here reflect quite plausibly some of ranges of different energy levels available in operation.
+For instance, as has been discussed \parencite{briggs2016, greenfield2016}, lead acid batteries may well only be able to practically utilise about 50\% of their total energy stores for tactical operaions, unless operating in a very low-threat environment and undertaking very slow charging to reach the maximum full charge is possible.
+In addition, the plausible improvement in enery density of lithium batteries may well be in the order of a factor of two in practice \parencite{greenfield2016}, though theoretical limits of the chemistry suggest much higher may be possible \parencite{davies2016}.
+
+Only brief is given here to the impact of Air Independent Propulsion, which has substantially higher energy densities possible even than Lithium, and also amounts to the addtional embarkation of greater mass.
+The volumetric energy density of the feedstocks to modern AIP systems (liquid oxygen and methanol) is estimated to be approximately ten times greater than that of lead-acid batteries \parencite{davies2016}, around 4.5MJ/L.
+Given the greater mass embarked, and additional studies suggesting that at constant speeds they might extend the dived endurance of a modern submarine from around 60 hours to over 1000 hours (a factor of roughly sixteen) an approximate estimation of the the direction and magnitude of this impact may by scaling up energy density by this factor.
+This would amount to adding three further frames to the right of Figure \ref{fig:sens_battery}, with the energy density being doubled in each successive frame.
+This would plausibly (under the central line selections) amount to more than month of dived endurance \textit{difference} between AIP inclusive submarines with propellers or pumpjets, amounting to thousands of nautical miles of continuously submerged travel.
+This difference would also now be material in terms of the overall range of the submarine amounting to a further 15-20\% reduction in the submarine's overall range.
+
+With such massive differences between the performance of the two systems emerging with the presence of the AIP, it is difficult to see how one could realistically consider a pumpjet as being competitive if a direct comparison were made.
+
 \subsection{The interaction of hotel load, battery energy density, and efficiency curves}
 
-\begin{figure}
-<<plot_sens_corners, echo=FALSE, warning=FALSE, message=F >>=
+Writing a paragraph here...
+
+\begin{figure}[p]
+<<plot_sens_corners, echo=FALSE, warning=FALSE, message=F>>=
 df <- read_rds("extdata/df_full_diff_cornered.rds")
 df %>%
+  mutate(corner = str_replace(corner, ", ", "\n")) %>%
   filter(test == "corners") %>%
@@
 
-\begin{figure}
-\includegraphics[width=\textwidth]{InceptionDepth.png}
+\begin{figure}[h]
+\captionsetup{width=0.7\textwidth}
+\includegraphics[width=0.7\textwidth, center]{InceptionDepth.png}
 \caption{Increasing in depth improves the speed achievable before cavitation begins to occur, as shown in the case of a torpedo \parencite{suryanarayana2010cavitation}}
@@
 
-\begin{figure}
+\begin{figure}[p]
 <<plot_inception, echo=FALSE, warning=F, message=F>>=
@@
 
-\begin{figure}
-\includegraphics[width=\textwidth]{VortexShedding.png}
+\begin{figure}[h]
+\captionsetup{width=0.5\textwidth}
+\includegraphics[width=0.6\textwidth, center]{VortexShedding.png}
 \caption{Vortex shedding following a flow separation leads to eddies being released unsteadily into the main flow stream, as shown here on the exterior of an accelerating duct.
@@
 The presence of the duct around the impeller may also have the effect of preventing the noise from being effectively radiated away from the propulsor.
-This may well result in substantial reductions in some types of noise being radiated away from the propulsor in certain directions, which may have given rise to the consistent qualification of 'radiated' being used to descrive the type of noise which differentiated the French submarine bid \parencite{stewart2016} \parencite{davies2017} \parencite{ohff2016}, including by the designers of the submarine \parencite{autret2016}.
+This may well result in substantial reductions in some types of noise being radiated away from the propulsor in certain directions, which may have given rise to the consistent qualification of 'radiated' being used to descrive the type of noise which differentiated the French submarine bid (\cite{stewart2016}, \cite{davies2017}, \cite{ohff2016}), including by the designers of the submarine \parencite{autret2016}.
 This is a secondary effect, quite separate to the avoidance of cavitation, which might be seen as an acoustic advantage in circumstances where cavitation might not be expected to occur on either type of propulsor.

### 2018-01-08 DraftExec.Rnw

--- 2018-01-08 DraftExec.Rnw
@@
 
-\subsection{Sensitivity to Battery Changes}
+The difference between pumpjet and propeller propulsion systems in terms of range tends to be found at slightly higher speeds, though this point also moves towards higher speeds as hotel load increases.
+With a reduced hotel load of 50kw, the difference between the two systems in the middle scenario is about 370nm for speeds 3.5-4.5kt, substantially greater than the ~200nm around 4.5-5.5kt at a hotel load of 100kW.
 
@@
 
-The singificant range of different energy density values tested here reflect quite plausibly some of ranges of different energy levels available in operation.
-For instance, as has been discussed \parencite{briggs2016, greenfield2016}, lead acid batteries may well only be able to practically utilise about 50\% of their total energy stores for tactical operaions, unless operating in a very low-threat environment and undertaking very slow charging to reach the maximum full charge is possible.
-In addition, the plausible improvement in enery density of lithium batteries may well be in the order of a factor of two in practice \parencite{greenfield2016}, though theoretical limits of the chemistry suggest much higher may be possible \parencite{davies2016}.
+It is also clear that increasing the hotel load has a greatly diminised impact on the proportional differences between the systems at higher speeds.
+At very high speeds (e.g.
+15kt) changes in the hotel load have almost no impact in proportional terms to the difference, which lies at about 30\% in all cases for the central selection of curves.
+At 9kt, in the region where transit speed is likely to occur, increasing the hotel load by a factor of four (from 50kW to 200kW) only decreases the advantage of the propeller from 48\% to 38\%.
+Therefore one might say near certainty that any difference between efficiency around these patrol speeds is likely to have a substantial impact on the overall range of the submarine, with differences of much less than 30\% quite unlikely, irrespective of other changes in the submarine design or operation circumstance.
 
@@
 
-With such massive differences between the performance of the two systems emerging with the presence of the AIP, it is difficult to see how one could realistically consider a pumpjet as being competitive if a direct comparison were made.
-
-\begin{figure}
+\subsection{Sensitivity to Battery Changes}
+
+Changes in the effective battery energy density will have the same effect as having larger masses of battery stored.
+They simply directly impact the total amount of energy available for consumption before recharging batteries while snorkeling.
+There is therefore no shift in the proportional impact on range and endurance due to this factor, as can be seen in the lower graphs in Figure \ref{fig:sens_battery}.
+The nominal impact, however increases directly and linearly with the total energy stored.
+
+\begin{figure}[p]
 <<plot_sens_battery, echo=FALSE, warning=FALSE, message=F >>=
@@
 
+The results in nominal terms are striking in their magnitude.
+In terms of enduracy, doubling the stored electrical energy, for example by adopting lithium batteries, increase the \textit{difference} between the two systems under the central efficiency curve assumptions to nearly four days (93 hours) at 4kt.
+Given that most modern, capable submarines are estimated to only be able to achieve dived endurances between 60 and 100 hours on batteries at 4kt \parencite{buckingham2008submarine} this cannot be dismissed in terms of its tactical significance.
+Even under the eficiency curve assumptions representing the narrowest spearation considered, the difference in this case would be 59 hours, almost equal to the estimated dived endurance of a Type214 operating without AIP.
+
+The in terms of submerged range, the same proportional impact is also extremely significant.
+Doubling the energy stored inceases the \textit{difference} between two otherwise identical submarines to over 400nm under central assumptions at 4.5kt.
+That's a larger distance than the width of the Java Sea, the most central body of water in the Indonesian Archipelago, or most of the length of the Malacca Strait.
+The tactical significance of such distances are not likely to be lightly dismissed.
+If far smaller amounts of energy are embarked, the nominal differences reduce to perhaps 100nm, in direct proportion to the reduction of the submarine's dived endurance.
+This is still a non-trivial distance in terms of it's tactical signicance, and comprises a substantial fraction of the dived ranges of most non-AIP submarines.
+It still represents a significant difference in operational terms, equal approximately to the width of Hainan Island, or neraly twice the length of the Sunda Strait.
+
+The singificant range of different energy density values tested here reflect quite plausibly some of ranges of different energy levels available in operation.
+For instance, as has been discussed \parencite{briggs2016, greenfield2016}, lead acid batteries may well only be able to practically utilise about 50\% of their total energy stores for tactical operaions, unless operating in a very low-threat environment and undertaking very slow charging to reach the maximum full charge is possible.
+In addition, the plausible improvement in enery density of lithium batteries may well be in the order of a factor of two in practice \parencite{greenfield2016}, though theoretical limits of the chemistry suggest much higher may be possible \parencite{davies2016}.
+
+Only brief is given here to the impact of Air Independent Propulsion, which has substantially higher energy densities possible even than Lithium, and also amounts to the addtional embarkation of greater mass.
+The volumetric energy density of the feedstocks to modern AIP systems (liquid oxygen and methanol) is estimated to be approximately ten times greater than that of lead-acid batteries \parencite{davies2016}, around 4.5MJ/L.
+Given the greater mass embarked, and additional studies suggesting that at constant speeds they might extend the dived endurance of a modern submarine from around 60 hours to over 1000 hours (a factor of roughly sixteen) an approximate estimation of the the direction and magnitude of this impact may by scaling up energy density by this factor.
+This would amount to adding three further frames to the right of Figure \ref{fig:sens_battery}, with the energy density being doubled in each successive frame.
+This would plausibly (under the central line selections) amount to more than month of dived endurance \textit{difference} between AIP inclusive submarines with propellers or pumpjets, amounting to thousands of nautical miles of continuously submerged travel.
+This difference would also now be material in terms of the overall range of the submarine amounting to a further 15-20\% reduction in the submarine's overall range.
+
+With such massive differences between the performance of the two systems emerging with the presence of the AIP, it is difficult to see how one could realistically consider a pumpjet as being competitive if a direct comparison were made.
+
 \subsection{The interaction of hotel load, battery energy density, and efficiency curves}
@@
 
-\begin{figure}
-<<plot_sens_corners, echo=FALSE, warning=FALSE, message=F >>=
+Writing a paragraph here...
+
+\begin{figure}[p]
+<<plot_sens_corners, echo=FALSE, warning=FALSE, message=F>>=
 df <- read_rds("extdata/df_full_diff_cornered.rds")
 df %>%
+  mutate(corner = str_replace(corner, ", ", "\n")) %>%
   filter(test == "corners") %>%

### 2018-01-08 DraftExec.Rnw
bump

@@
 \usepackage{subcaption}
+%  appendix
+\usepackage[toc,page]{appendix}
 
@@
 
-\maketitle
+\begin{titlepage}
+  \vspace*{\fill}
+  \begin{center}
+    {Efficiency}\\[0.5cm]
+    {Aidan Morrison}\\[0.4cm]
+    2018
+  \end{center}
+  \vspace*{\fill}
+\end{titlepage}
 
@@
 
+\newpage
+
 \tableofcontents
+
+\newpage
 
@@
 
-\section{Executive Summary}
-
-\subsection{The difference between nuclear and conventional for speed}
-This paper was commissioned in order to investigate whether pumpjets could plausibly be as efficient, or more efficient than a suitably designed open propeller for a conventional submarine.
-A crucial input to the investigation is a rough understanding of what the speeds of operation are likely to be for a submarine.
-Whereas nuclear submarines are reportedly capable of reaching speeds in excess of 30kt, and might transit long distances at such high speeds, conventional sumbarines are not thought of as being able to reach speeds far above 20kt in a sprint, and can only sustain speeds of 8-10kt for long-distance transits.
-Moreover, on patrol, a large portion of thier work is done at very low speeds, typically thought to be in the range of 2-4kt.
-
-\subsection{The importance of low speed operation for diesel-electric submarines}
-The efficiency of the propulsion system at such low speeds is of great significance for a conventional submarine, since it must rely on batteries or other air-independent propulsion sources for power when entirely submerged.
-Consequently, excessive energy consumption results in greatly reduced dived endurances.
-Since a submarine's position is vastly more likely to be discovered when it is on the surface operating its diesel engines, this ability to remain submerged for a long time is crucially important for combat operations, and in transiting through sensitive or contested areas.
-For nuclear submarines, which posess a practically infinite supply of energy from the nuclear reactor, efficiency at low speeds is of no concern.
-In fact, dispersing additional energy, (provided it can be done quietly) is probably advantageous for a nuclear submarine, since it will allow the nuclear reactor to avoid running at very low power levels, where the stability of the reactor is reduced.
-
-\subsection{What is cavitation}
-Pumpjets have been widely adopted by navies operating nuclear submarines.
-The principal advantage of a pumpjet relevant to submarines is their ability to avoid problems associated with cavitation, which is known to occur for propellers attempting to operate at high speeds.
-Cavitation, which is the rapid expansion and collapse of a bubble or void in the water, is particularly problematic for submarines, since it results in the creation of a great deal of noise which could be detected by an enemy.
-
-\subsection{How does a pump-jet work}
-A fundamental requirement of a pump-jet in order to avoid this cavitation is that the the working parts of the jet (the rotating blades inside it, known as the impeller) operate at a higher pressure than propeller blades operating in open water would.
-This means that the blades can turn at a lower speed relative to the water they are connected with, allowing a less violent action, which induces less cavitation.
-In this way, the jet does it's work less by directly acelerating the water, but by raising its pressure.
-This raised pressure is converted back to movement, which produces thrust, as it exits the jet and returns to the same pressure as the surrounding environment.
-
-\subsection{The necessity of drag, and decelleration, induced by the duct}
-The role of the shroud (or tunnel, mantel, duct) around the jet is to allow the pressure to be raised around the impeller, in a way that is not possible for an open propeller.
-By a fundamental requirements of physics, this actually requires that the water's incoming speed be \textbf{reduced} before it reaches the impeller.
-(The kinetic energy embodied in movement is converted into potential energy, or pressure.)  Whilst the duct narrowing at the nozzle also necessarily accelerates the water, (as the additinal pressure imparted by the impeller is converted back to kinetic energy) it's an essential feature of all pump-jets that the water flow is decelerated at the point of reaching the impeller.
-Consequently, an elementary form of the pump-jet is also termed the 'decelerating duct' applied to a propeller.
-Put simply, the water has to slow down to go fast again.
-
-\subsection{What happens to a jet at low speed, in simple terms}
-The problems for pump-jets arise when the water is already going slow, and you don't want it to go that much faster.
-This is the case when a jet designed for high-speed attempt to operate at a dramatically reduced speed.
-(Not to be confused with a vessel with very little water speed working very hard, as might be the case for a tug or barge.)  In this case, the slowing down and speeding up results in an unnecssary additional step which reduces greatly the efficiency of propulsive system.
-Or, put another way, overcomign the resistance of moving water through the shroud becomes much greater relative to the total thrus produced by the jet.
-
-\subsection{Why this isn't easily noticeable in normal circumstances}
-It is for this reason that waterjets of any kind are known to have an efficiency curve that falls off towards zero as the net thrust they produce also diminishes to zero (which for a given vessel will correspond to water speed).
-This doesn't mean that they don't still work at low speeds and produce some thrust.
-All it means is that far more power per unit of thrust will be required than might be at other speeds, as a higher fraction of energy is expended producing the turblence (random, round-and-round movement) inside and around the jet shroud than goes towards direct front-to-back acceleration of the water column, which produces thrust.
-
-\subsection{What the submarine requirement tells us about the duct}
-Whist estimating the exact shape and level of the efficiency curve for a particular pump-jet and propeller is impossible without detailed knowledge of their design, the over all trends of their shapes in the extremes can be known from well-established principals.
-Moreover, the particular demands of a pump-jet suited to a nuclear submarine, (eliminating all cavitation in as wide a range of operating circumstances as possible) considerably narrows the plausible range concerned.
-In order to minimise cavitation, the degree of pressure elvation at the impeller would be relatively high, or the total surfaces of the impeller blades much aslo become larger.
-Both of these design requirements necessitate changes to the duct or blade designs which would be in tension with overall efficiency, and most pronounced at the lowest of operating speeds.
-
-\subsection{Quantitative Conclusions}
-In the production of this paper I have developed a computational model which maps the impact of different efficiency curves directly to the dived range and endurance of a submarine.
-The most plausible scenarios I find include the reduction of dived endrance and range between 20\% and 50\%, or effectively halving time and distance that submarine may remain submerged for during combat operations at speeds around 3-5kt.
-
-\subsection{The acoustic advantage of jets at higher speed}
-My review of the literature confirms that a pumpjet may produce a significant acoustic advantage in circumstances where an open propeller would experience any degree of cavitation.
-Indeed, it has been remarked by naval researchers that pump-jets could be designed which would not cavitate past the point when the body they propel experienced caviation.
-As such, it seems perfectly plausible that pump-jets confer one advantage on a submarine, in that they can accelerate to a higher speed without cavitation occurring.
-This is known as a higher 'tactical silent speed'.
-
-\subsection{Turbulence and flow separation at low speed}
-However, at much lower speeds such as patrol speeds (2-4kt) it is most likely that a propeller will be able to operate well below the point of any cavitation inception, and would consequently also be extremely quiet.
-Moreover, a jet will necessarily incur substantially larger degrees and types of turbulence in order to produce net thrust in this regime.
-These likely include discontinuities and instability in the flow entering the duct and passing through the impeller and stator (flow separation) as the water is acelerated sufficient to be slowed, and then re-accelerated.
-In certain circumstances where resonances might arise these could be highly adverse to acoustic performance.
-However, assuming that by careful design such resonances can be eliminated (which we assume has been achieved) these effects would not result in any cavitation, and would only increase the noise attributable to turbulence in solid water, which is far less than for any cavitation.
-
-\subsection{The acoustic question at low speed}
-But in either case, given the necessarily raised levels of turblence generated by a jet than a propeller at very low speeds, the claim that the jet is quieter in this regime must rely entirely upon shielding effects from the shroud.
-These could be substantial in a directions perpendicular to the direction of travel, but would be much smaller when viewed from the aft or forward directions.
-It should also be noted that many underwater acoustic environments, sound tennds to reflect and bend in different directions as it propagates.
-As such, the claim that pump-jets are universally quieter than propellers should be treated with caution.
-In a variety of circumstances, including most operations at patrol speed, it may not always be true.
-
-\subsection{The inevitable trade-off}
-However, as a direct and necessary result of this higher tactical silent speed, some disadvantages will be incurred, owing to the additional drag induced by the shroud which prevents cavitation at the impeller.
-These are a lower dived endurance, a lower dived range, a lower overall endurance, a lower overall range, and a worse indescretion ratio.
-
-\subsection{The contradiction between stated requirement and chosen technology}
-Given the substantial emphasis which was placed on overall range and endurance of the submarine, it is difficult to understand how one particular unique requirement would be elevated so high above the other strategic and tactical advantages afforded by propellers.
-
-\section{Speed and Drag - Why very slow is very very $(very)^2$ economical}
+\newpage  \section{Executive Summary}
+
+The selection of the pumpjet propulsion system is quite probably the most significant new technology selection for Australia's future submarines.
+Unlike other new technolgoies such as Air Independent Propulsion and Lithium Ion Batteries, which are either in service or entering service in multiple other navies, no other conventionally powered submarines have adopted pump-jet technology, which is currently only used by nuclear submarines.
+
+Despite acknowledging that the pumpjet may have an adverse impact on dived endurance \parencite{sbs2017}, Naval Group justifies the superiority of the pumpjet in the following terms: "In a confrontation between two otherwise identical submarines, the one with pump jet propulsion always has the tactical advantage."  No further explanation of this superiority has been provided in detail, but it is widely held that the pumpjet offered acoustic advantages which were highly regarded by those Australians who were responsible for selecting the design partner \parencite{stewart2016}.
+
+The comparison between two otherwise identical submarines (though not necessarily limited to engaging just each other in combat), provides a useful and suitable framework for evaluating a single significant technology choice, where most other significantly novel or new technologies were apparently excluded from consideration \parencite{stanford2017}.
+The conclusion of this report is that the following statement can be made with confidence:
+\\
+\\
+\emph{In a comparison between two otherwise identical submarines, the one with the pump jet will alwayas have a lower dived endurance, a lower dived range, a worse indescretion ratio, a lower overall endurnace, and a lower overall range, than the one with a conventional propeller.
+This will confer a substantial tactical and strategic advantage on the conventionally propelled submarine in a very broad range of operational scenarios.}
+\\
+\\
+Whilst it is broadly acknowledged that propellers have superior efficiency at lower speeds, but jets might have higher efficiency at higher speeds, the precise point of cross-over seemed unclear in public debate.
+One informed commentator suggested that the crossover point was likely to be below or near the transit speed of the submarine, and that the pumpjet may offer enhanced range and endurance over the whole mission profile as a consequence \parencite{davies2017}.
+This report finds no evidence of a pumpjet of the type used on submarine every being recommended as superior for efficiency in general, and even closely related technologies for more commecial use not advocated below 25kt.
+During my research I encountered domain experts confident that a pumpjet could simply not feasibly be superior to a suitable propeller at speeds below 18kt.
+Evidence of propellers remaining generally superior to waterjets and pumpjets in the speed ranges up to and even over 30kt is abundant.
+Consequently, the possibility of some cross-over point leading to any kind of range or endurance advantage relevant for a conventional submarine can be safely ruled out.
+
+This report also explores in some detail the theoretical concepts which drive the limitations of pumpjet performance at lower speeds in comparison to propellers.
+Enduring and fundamental principals of momentum theory provide robust limitations on the circumstances where a propeller's 'ideal efficiency' might be improved by the addition of a duct/or shroud, one type of which constitutes the transition to a 'pumpjet'.
+These principals stronly support the the notion that pumpjets will not and indeed cannot suddenly overcome their efficiency disadvantage at low speeds by means of technical ingenuity or innovation.
+
+Furthmore, a particular hydrodynamic phenonomon called 'flow separation' can be identified as the primary cause of the pumpjet's declining efficiency at very low speeds.
+Put simply, the duct provides an additional hurdle or barrier though which the water must pass when moving through the propuslor.
+At very low speeds waterspeeds, this additional resistance is sufficient to stop some of the flow progressing through the jet, creating unsteady eddies and vortices which sap energy from the system, but produce no thrust, driving down efficiency to zero at very low flow-rates.
+
+The design of the pumpjet for military purposes, namely the reduction of acoustic signatures effectively mandates that this must be the way that pumpjets work.
+The act of slowing the water down is necessarily equivalent to the act of raising its pressure.
+It is this elevated pressure at the point where the blades to their work on the water which enables the blades to avoid cavitation (the rapid expansion and collapse of small voids or bubbles in the water) which is very large source of noise underwater.
+Maintaing this advantage to the same level necessitates a duct design which slows water down to the same extent.
+The act of slowing water down when it's already going slowly is what causes some of it to stop, which is when flow separation occurs.
+
+In order to assess the probable magnitude of the impact of this reduction in efficiency at low speeds, a numerical model was developed, based on known principles of physics, and publically available information relating to the likely technologies involved in the future submarine.
+The most important and variable assumptions - the efficiency curves of the pumpjet and propeller - were taken from a recent study lead by world-leading engineering companies, supported by the US and Royal Navies, comparing closely related propulsion systems  \parencite{giles2010}.
+Knowing that the systems would not be identical to the propulsors on submarines (the precise design of which is obviously secret), additional curves varying from these assumptions were also modelled in all directions, and the combinations combined and tested.
+
+The model has been made available publically in the form of a web-app, so that alternative assumptions, including alternative efficiency curves may be tested by the general public.
+
+The model in its central assumptions was able to achieve very realistic results, which were consistent with other public estimates of performance parameters of conventional submarines and known characteristics \ref{buckingham2008submarine}.
+A variety of different assumptions regarding the hotel load, battery density and different combinations of efficiency curves were tested.
+The central assumption values showed that a submarine with a conventional propeller had a very substantial advantage in terms of range and endurance across the entire speed range.
+This advantage reached a proportional maxium at around 6kt of over 60\%.
+However, in nominal terms, the greatest differences occurred at slightly lower speeds.
+Under central assumptions, the propeller submarine had an advantage in submerged range of some 200nm at 5kt, and approximately a two-day advantage in dived endurance at 4kt.
+
+The full range of plausible efficiency curves exhibited a significant efficiency advantage right through transit speeds.
+Central assumptions suggested that this would result in the overall range of a submarine being enhanced by approximately 50\% with the use of propellers instead of a pumpjet.
+None of the scenarios suggested it was likely that this range advantage would be any less than 20\%.
+Given the significant emphasis that has been repeatedly placed the unique requirements for range and endurance for Australia's submarine, this raises substantial questions about suitability of this technology to meet Australia's needs.
+
+The model was used to test a wide range of different assumptions.
+A further key result of the study established the sensitity of the difference to different assumptions regarding hotel load, and battery energy density.
+This showed that the difference between the two systems was made substantially larger with smaller hotel loads, and larger stored electical energy (as might be achieved with lithium-ion batteries), and these two effects combined to give some particularly startling results.
+Circumstances where the \textit{difference} in dived range and endurance between the two systems in terms of range and endurance exceeded the \textit{entire} dived range and endurance achieved by many modern conventional submarines frequently arose.
+
+The model demonstrated clearly that all of the most pronounced advantages in dived range and endrance occurred at speeds around 3-7kt, generally within plausible speeds achievable with modern AIP systems.
+The addition of AIP was briefly considered which revealed that a difference in dived endurance of the order of one month, or thousands of nautical miles, would likely have emerged between the two propulsion systems.
+
+The consequences of flow separation also has acoustic consequences, which this report explores.
+The vortexes and eddies which are created by flow separation tend to be unsteady, and shed periodically into the main flow.
+This results in pressure fluctations inside the shroud and more intense, unsteady large-scale types of turbulence occurring as these vortices are ingested into the rotor blades.
+These effects are are likely to cause some accoustic signature which becomes lower in frequence as the submarine slows down further and the vortexes become larger, and shed more slowly.
+This is in contrast to early stages of cavitation, which would be high frequency at when bubble sizes are small, and move to lower fequences as speed increases and bubbles expand.
+Lower frequencies are of particular significance because they tend to propagate further, and would be less effectively shielded by the duct.
+
+It remains plausible that in the absence of cavitation the noise generated in the propulsor by such flow separation is not loud enough to be of significant tactical concern.
+However, it is extremely unlikely that at some very low speeds, where conventional propellers experience no cavitation and enjoy steady, smooth flows over the blades, that a pumpjet could actually have a lower acoustic signature, even in terms of radiated noise.
+Flow separtion demand that two or three times as much energy is wasted by a pumpjet creating and destroying these unsteady vortexes than a propeller would require to create equal thrust.
+Consequently, the claim that pumpjets are generally acoustically superior should be treated with some caution.
+This claim has strong grounds wherever a conventional propeller might experience cavitation, such as at higher speeds.
+But at some very low speeds it is unlikely to be true.
+
+Overall, it seems unlikely that the full range of consequences of pumpjet choice have been fully comprehended by the Australian government.
+The scale of the probable impact on range and endurance is quite probably so substantial that it is difficult to see how such a performance penalty is consistent with the government's stated aim of acquiring a regionally superior submarine.
+
+\newpage  \section{Speed and Drag - Why very slow is very very $(very)^2$ economical}
 
@@
 
-\section{The difference between nuclear and conventional propulsion}
+\newpage  \section{The difference between nuclear and conventional propulsion}
 
@@
 
-\section{Some essential concepts}
-
-\subsection{Conservation of Energy}
-
-This is perhaps one of the most fundamental and well-established principals in physics.
-The essential idea is that energy can move or change in form, but it isn't ever created or destroyed.
-Machines, plants and animals all derive their energy from a particular other source, which can be measured and evalutated to establish the limits of energy available.
-Plants collect energy from sunlight falling on their leaves.
-Humans (as well as combustion engines) capture the chemical potential energy in organic matter, and release it by combining it with oxygen.
-Hydro-electric power plants turn the gravitational potential energy of water stored at a height into electricity.
-
-Conservation of energy has a particularly relevant embodiment in fluid flows, which is given it's primary expression in Bernoullis equation, as given in Equation \ref{eq:bernoullis}.
-It says that the total energy in a connected body of fluid with no additional work being done on it is constant ($C$), though it can change in form between kinetic energy (movement, $\frac{1}{2}\rho v^2$), gravitational potential energy (it being elevated, $\rho gz$, relevant for liquids, less for gas) the heat energy in the fluid (not shown, not relevant for incompbressible fluids), and the pressure of the fluid $p$.
-
-\begin{equation}
-\label{eq:bernoullis}
-\frac{1}{2}\rho v^2 + \rho gz + p = C
-\end{equation}
-
-In different fluids, the dynamics of how energy moves between one and another change.
-For example, in gases, heating up a confined piece of gas will increase it's pressure, or if it is unconfined, increase its volume.
-This is particularly important for understanding gas turbines.
-In the case of a liquid, however, all the molecules are in close contact, and hence can't increase in volume or pressure substantially except by the creation of steam.
-As such, in the absence of large amounts of cavitation, the terms in the relationship which are most important for our consideration are the relationships between pressure, and kinetic energy, and gravity.
-
-In the case of most waterjets which propel surface vessels, the water is lifted from the bottom of the hull at the intake to the pump, which is generally incorporated inside the hull.
-This increase in gravitational potential coincides with a slowing down of the water relative to the vessel-speed.
-Since the jet in such vessels is generally ejected at the same height as the pump, this potential is never regained, and is technically a loss, however at high speeds such a loss is small relative to the total power ouput.
-In the case of submarine pumpjets, located on the end of the hull and parallel to the axis of the submarine, this effect (and the term $gz$) are not relevant.
-
-\subsubsection{Venturi Effect}
-In the case of submarines and torpedoes the water doesn't undertake a change in height, since intake and nozzle are generally all in line with the central axis of the submarine or torpedo.
-As a consequence, the key relationship in Bernoullis equation is the relationship between the liquids velocity, and its pressure.
-The consequence is that when water moves through a pipe (duct, or shroud) it's pressure is inversely related to the square of its velocity.
-This means that when a liquid is forced to to travel through a narrowing pipe, it's pressure necessarily decreases as the velocity increases.
-This is what underlies the Venturi effect.
-
-\subsubsection{Flow Diffusion}
-The inverse process is where a pipe increases in volume, and the flow is forced to slow down in order to fill the wider area, and the pressure correspondingly increases.
-This is a process called 'diffusion', and is important to achieving high pressure levels in many types of water pumps, including those which will be particularly relevant for pumpjets for watercraft propulsion, particularly elevating static pressure levels in order to prevent cavitation \parencite{hamilton1997, mccormick1963design, bruce1974}.
-
-\subsubsection{Energy waste and efficiency}
-The conservation of energy also has useful implications for how the efficiency of systems is thought about.
-In particular, because energy is conserved, identifying inefficiencies in a system necessarily involves identifying where energy goes to doing tasks which aren't useful for the intended purpose.
-A perfectly efficient system won't do any work that isn't for the intended purpose.
-In the case of analysing the efficiency of propulsion systems, the relevant 'work' is almost always related to moving water backwards to produce thrust.
-Moving water in directions other than backwards, including random turbulent flows which wind producing heat rather than thust, are two examples of wasted energy.
-Noise in the water also reflects energy which is wasted, though the significance of the acoustic signature is not necessarily indicative of the magnitude of energy wasted.
-
-In general terms, energy waste in propulsion systems will result in turbulence (moving water in random chaotic patters, where the energy soon becomes smoothly distributed as heat) and acoustic energy that is propagated away.
-Flow separation incolves the creation of larger-scale vortices and eddies, which might be considered as a special case of turbulence, since the length-scale of the unsteady flows is much larger than that of normal turbulence, for instance in a boundary layer.
-
-On important consequences is that whilst increasing creation of noise, flow separation, or turbulence necessarily represent greater amounts of energy wasted, some increase in any given one does not neccessarily indicate the reduction of efficiency of the entire system, as larger amounts of energy may be being saved through reduced wastage through another means, or vice versa.
-For example, a squeaky join on a car might make a lot of noise, and it will represent some loss of kinetic energy.
-However, the brakes can be applied and absorb vastly larger amounts of energy, essentially silently.
-
-Furthermore, efficiency of a system is necessarily given by the a ratio of useful work to total energy expended.
-Some trends which show increasing levels of energy wasted might be outweighed by larger increases in the overall amount of productive work being achieved.
-Pumpjets, which generally tend to require highly turbulent flows, and even more turblent flows at higher speeds, still can become more efficient as speed inceases, as the amount of useful thrust produced increases even faster.
-So, despite the wasteful round-and-round movements increasing with speed, the increase in productive front-to-back movement improves even faster, and total efficiency rises.
-
-\subsection{Boundary Layer}
-When water or any fluid flows with some speed relative to another solid surface nearby, there is some layer adjacent to the surface in which the speed of the fluid is diminished relative to the main flow.
-At a microscopic level, there are some molecules of the fluid on the surface which will be effectively static relative to the surface.
-The layer of fluid that joins the gap between the static surface, and the part of the flow which is moving at the full flow speed, is called the boundary layer.
-Exactly how thick the boundary layer is, and how the fluid moves in the boundary layer, is extremely important for consideration of efficiency of fluid flows over and around solid surfaces.
-In particular, a boundary layer can be either turbulent or laminar in nature, and can transition to turbulent flow after a short distnace of laminar flow, as shown in Figure \ref{fig:BoundaryLayer.png}.
-
-\begin{figure}[h]
-\captionsetup{width=0.5\textwidth}
-\includegraphics[width=0.5\textwidth, center]{BoundaryLayer.png}
-\caption{The development of a boundary layer as shown in \parencite{mollard2011}}
-\label{fig:BoundaryLayer.png}
-\end{figure}
-
-\subsection{Turbulent and Laminar Flow}
-The way that fluids move relative to a surface can one of either two methods.
-In 'laminar flow' all the fluid moves in one direction in an smooth and orderly manner, with very little mixing between the layers of fluid travelling at different speeds.
-
-The alternative is 'turbulent flow', where the fluid moves around in unpredictable swirls and circles as well as moving overall in an predominant direction.
-There is considerable mixing between all the different layers, and the average speed remains relatively constant in the flow, with the exception of the flow immediately adjacent to the wall, or in the boundary layer.
-Becuase turbulence involves a lot of movement that is not overall in one productive direction and contributing to thrust, it necessarily leads to some loss of energy from the overall thrust.
-Once the swirls and movements become smaller and smaller, the energy winds up simply as heat in the fluid.
-
-\begin{figure}[h]
-
-\begin{subfigure}{0.5\textwidth}
-\includegraphics[width=0.7\linewidth, center]{Laminar.png}
-\caption{Laminar}
-\end{subfigure}
-\begin{subfigure}{0.5\textwidth}
-\includegraphics[width=0.7\linewidth, center]{Turbulent.png}
-\caption{Turbulent}
-\end{subfigure}
-
-\caption{A simple comparison of laminar and turbulent flows from \parencite{NPdotnet2017}}
-\label{fig:LaminarTurbulent.png}
-\end{figure}
-
-The likely point of transition between a flow being turbulent and laminar can be given by unitless number called the Reynolds Number, which is determined by a fluids velocity, viscoscity, density, and a characteristic length-scale over which the flow occurring.
-A fuller discussion of these important concepts can be found in \cite{NPdotnet2017}.
-
-Turbulent flows in gasses can be extremely noisy, since those fluids are compressible.
-Consequently, the rapid and intense ciruclar movements result in a lot of oscillatory compressions against the surrounding air.
-Consequently things like jet engines, vacuum cleaners, hand-dryers, and other devices that create rapid movements in gasses tend to be quite loud, including at some distance.
-
-In essentially incompressible fluids, the random changes in pressure and velocity which are involved in laminar flow tend not to generate nearly as much noise in the surrounding liquid outside the flow.
-Because the fluid is incompressible, all of the random 'round and round' movements don't amount to much 'in and out' movement, which is what creates the pressure waves which result in propagated noise.
-Consequently, whilst turbulent flows do inevitably generate some noise, the amount of noise generated is dramatically greater in the presence of cavitation, where a void opens up in the water, and vastly more 'in and out' movement is able to occur.
-
-\subsection{Cavitation}
-Cavitation is the rapid expansion and collapse of a void, or bubble in water.
-Put technically, caviation occurs when the local static pressure (pressure in the rest frame of the fluid) falls below the vapour pressure of the fluid (pressure at which the liquid will start to boil).
-
-To try to conceptualise how it occurs in the context of propulsors, one might imagine what happens when something moves very fast through water.
-The front side of the object pushes the water forward, but on the back side the water has to push in to fill the space left behind.
-If there’s not enough pressure to push the solid water in fast enough, a gap opens up, with just a few gaseous water molecules (steam) inside the cavity.
-(This is also described as water boiling at low pressure.)  However the gap doesn’t stay around for long.
-Soon the water catches up, and the bubble implodes with a pop, leaving only very tiny bubbles as a result, which you can see in the wake of almost any boat or ship moving at speed.
-Whilst some energy turns to heat (the remaining steam in the tiny bubbles) some is propagated away as a sound-wave generated by the implosion.
-
-A watching water boil in a glass kettle gives a quick and intuitive insight into the occurrence of cavitation.
-Quickly after the kettle starts heating, a considerable noise can be heard, which corresponds to the commencement of cavitation on heating element.
-At some local point for a moment in time, there is enough energy for the water present to boil.
-It tends to be on rough surfaces or in the presence of some impurity that cavitation will occur first (in the blackened part of the surface in this case.)  However, whilst the bubbles on the bottom can be seen plainly on the bottom, they collapse almost straight away again, and leave only a tiny bubble of stable steam circulating in the water.
-It is the rapid expansion and collapse of these bubbles that causes the noise of a kettle, long before it has boiled.
-
-\begin{figure}[h]
-
-\begin{subfigure}{0.5\textwidth}
-\includegraphics[width=0.9\linewidth, height=5cm]{EarlyCavitation.JPG}
-\caption{}
-\end{subfigure}
-\begin{subfigure}{0.5\textwidth}
-\includegraphics[width=0.9\linewidth, height=5cm]{BoilingCavitation.JPG}
-\caption{}
-\end{subfigure}
-
-\caption{(a) Cavitation bubbles form and collapse creating noise and some tiny bubbles long before boiling occurs.
-(b) Only at very high temperatures do the bubbles endure at full-size in the water and reach the surface.}
-\label{fig:EarlyCavitation.JPG}
-\end{figure}
-
-It is only when the water is all at a much higher temperature that the bubbles remain their full size for long enough to detatch and rise all the way to the surface, a process we typically think of as boiling water.
-It is worthwhile noting that the sound emitted at this stage is much softer and lower than the early onset of cavitation.
-Larger bubbles result in lower frequencies of sound being emitted.
-Even at higher temperature, the rougher parts of the surface provide the points where all the cavitation originate.
-
-Cavitation is extremely important in the study of ship propulsion, since it's occurrence in particular circumstances can lead to substantial losses of efficiency, as well as damage to the propeller and related appendages.
-It is particularly important for submarines, since the expansion and collapse of these bubbles tends to lead to the creation of noise which is often far larger, and more distincly characteristic than other turbulent disturbences in the water when no cavitation is present.
-Consequently, the onset of cavitation can be thought of as a distinct threshold in terms of the acoustic performance of propulsive system.
-
-Despite cavitation necessarily representing some energy being wasted generating unwanted noise in the water, some cavitation inevitably occurs around most propulsion systems operating at full power.
-In plenty of cases, the consequences for efficiency are relatively small, as they tend to be dominated by other efficiency considerations in imparting thrust.
-Put another way, the savings that can be gained by minimising wastage to turblence can outweigh the losses incurred by having some cavitation occur.
-In most cases, propellers are designed to work with a certain extent of cavitation for optimum efficiency under working loads \parencite{shin2015}.
-Cavitation also occurs to a considerable extent within the jets of most high-speed surface vessels, but doesn't necessarily have a particularly bad effect on their overall efficiency at their intended speed levels.
-It is generally only within very specialised military circumstances when the absolute avoidance of cavitation supercedes other concerns for efficiency, and require the elimination of all cavitation entirely \parencite{mollard2011}, \parencite{lewis1988}.
-These circumstances include the design of submarines and torpedoes.
-
-In certain special cases, particularly supercavitating propellers or surface-piercing drives, high levels of cavitation can be higly advantageous for efficiency.
-However these represent specialised designs, generally only for very high speed vessels, where acoustic concerns are negligible, and are of little concern for this particular endeavour.
-
-Cavitation can occur in a number of different ways at different points and forms on or around the propeller.
-It is common for cavitation to occur first near the out extremities of the propeller, where the blades are moving at the highest velocity relative to the water-flow.
-Cavitation tends to spread across the back of the blades in a sheet (sheet cavitation), since the back side of the blades are generally the areas of most depressed pressure, though it can also occur in places on the back side of the blades, where the dynamics of the water moving around the blade can produce points of near the edges of significantly reduced pressure.
-On surface vessels cavitation tends to occur most when the blades are closes to the surface, where the pressure is lowest.
-
-\begin{figure}[h]
-\captionsetup{width=0.7\textwidth}
-\includegraphics[width=0.7\textwidth, center]{IncreasingCavitation.png}
-\caption{Cavitation frequently occurs first at the extremities, then spreads inwards across the blad face, as shown in \parencite{mollard2011}}
-\label{fig:IncreasingCavitation.png}
-\end{figure}
-
-\begin{figure}[h]
-\captionsetup{width=0.7\textwidth}
-\includegraphics[width=0.7\textwidth, center]{CavitationTypes.png}
-\caption{Different types of cavitation can occur on different parts of the propulsion system as shown in \parencite{carlton2007}}
-\label{fig:CavitationTypes.png}
-\end{figure}
-
-\subsection{Pressure, Thrust, and Momentum change}
-A few simple related concepts make up the foundations for many of the simplest and enduring theoretical models (Momentum Theory, still widely and routinely used to model 'ideal' propeller efficiencies by considering the propeller simple disc \parencite[169]{carlton2007}) and other quite intuitive concepts which migth be more readily grasped in understanding propulsors.
-
-Thrust is a force, measured in Newtons, which is generally used to refer to the forces acting on a vessel or object by its propulsion system to move it forward in the water.
-Generating thrust is consequently a key objective of any propulsion system.
-
-Pressure is a measure of force divided by an area over which it is applied.
-Pressures are necessarily applied to some extent over all of the surfaces of any object, including due to the atmostpheric pressure of air.
-In considering the efficiency of propulsion systems for solid bodies, thrust is always the consequence of some \textbf{net} pressure difference on different sides and surfaces of an object.
-For instance, on a propeller blade, the rear-facing side of the blade is the side that pushes the water backwards, and hence has a raised pressure on its surface.
-Being generally backwards facing this acts to push the propeller forwards.
-This pressure acts equally to push water backwards, as well as to push the propeller forwards.
-Importantly, the forward-facing side of the propeller will expereince a lowered pressure, which also serves to pull water backwards, and suck the propeller forwards.
-It is the sum of all the pressure differences on the propeller that result in a net force.
-
-Importantly in hydrodynamics, it is important to consider the possibility of other forces being generated other surfaces which may have an impact on the movement of the vessel.
-In the presence of the hull, water being drawn by the propeller can cause pressure changes on other surface of the vessel which may assist or impede producing a net positive thrust.
-Submerged jets represent a particular case in point.
-A famous home experiment suggests that one compare the thrust (backwards pressure) experienced by someone holding a garden hose creating a jet of water when it is in the air, as opposed to submerged in a pool of water.
-When submerged, the backwards pressue appears dramatically reduced.
-In fact, this phenomenon is due to a the water-jet drawing adjacent water with it under-water, the movement of which exerts a negative pressure on the outside of the nozzle, conteracting the pressure exerted on the inside of the nozzle.
-This effect represents one of the important distinctions between the jets which release their water below or above the water-line, and generally means that underwater jets have much lower jet velocity ratios than those released over water.
-
-It is also an important law of physics that force or thrust is equal to the rate of change of momentum.
-Put simply, the speed at which one speeds something up, or the increasing amount of something which begins to move, is always equivalent in magnitude to the thrust that is produced.
-Since momentum is a vector, this simple law requires that the momentum change has to net out to one particular direction in order to produce a net thrust, which provides the core principal for Momentum Theory as it applies to propeller and jets.
-Circular movements necessarily are cancelling, and produce no thrust.
-It is useful to recall however that this principal is simply another complete, alternative way of measuring the consequences of the net pressure differences discussed earlier.
-The total rate of change of momentum in the system which is induced as a conseqence of the pressure differentials on all sides of the propulsor must necessarily be equivalent to the total net force that is produced.
-For the garden hoze in the pool or bucket, the jet rapidly entering the liquid in one direction necessitates new flows of liquid in other directions, reducing net moment changed.
-This new liquid movement will exert pressure reductions on the nozzle surface surface, which reduce the net momentum change affected by the entire system.
-These net forces, and net momentum change are necessarily equivalent and equal in effect.
-Consequently literature seeking to optimise propulsion might discuss analysis referring to both of these effects or phenomena, and neither are in conflict.
-
-Full and proper derivations of many of the thrust equations for different propulsion systems from these principals can be found in plenty of authoritative works, including \cite{lewis1988}, \cite{carlton2007} and \cite{mollard2011}.
-
-\subsection{Flow separation}
-Flow separation refers to a circumstance when the flow over a surface separates entirely from the surface, and a new eddy or votex is formed where the flow actually moves in the opposite direction to the dominant flow.
-It is of particular significance in hydrodynamics because it can lead to substantial losses of efficiency, as energy is diverted into the kinetic energy of the eddy or vortex, which is unproductive.
-It tends to occur near the surface of aerofoils with high angles of attack, or in the blades of turbomachinery including diffuser or mixed-flow pumps and ducted propellers operating in off-design conditions, including low flow rates \parencite{li2013}, parencite{bakker2006}, parencite{mcbride1979}, parencite{wislicenus1986}.
-
-\begin{figure}[h]
-\captionsetup{width=0.5\textwidth}
-\includegraphics[width=0.5\textwidth, center]{SeparatedAerofoil.jpg}
-\caption{Flow separation occurring over the top of an aerofoil in a wind tunnel. Image courtesy of Deutsches Zentrum fuer Luft- und Raumfahrt e. V. (DLR)}
-\label{fig:SeparatedAerofoil.jpg}
-\end{figure}
-
-In order for flow separation to occur, a fluid must be moving against an adverse pressure gradient, which simply means that the static pressure (pressure in the rest frame of the fluid) is increasing in the direction of movement, which means the flow is slowing down as it moves, unless new work is being done on it.
-At the boundary layer, where the flow velocity is already reduced due to drag forces experienced near the surface, the adverse pressure gradient can be enough to reverse the flow altogether.
-In this case, the fluid flows in reverse near the surface and forms a vortex, and the main flow becomes separated form the surface by the votex.
-
-\begin{figure}[h]
-\captionsetup{width=0.7\textwidth}
-\includegraphics[width=0.7\textwidth, center]{FlowSeparation.png}
-\caption{Flow separation involves flow reversal near a surface when a fluid is moving against an adverse pressure gradient, as shown in \parencite{mollard2011}}
-\label{fig:FlowSeparation.png}
-\end{figure}
-
-In addition to the necessary loss of efficiency, flow separation can often tend to result in unsteady flows, with the vortices periodically being shed into the flow \parencite[480]{mollard2011}.
-If this occurs ahead of the blades of a propeller or impeller, such disturbances to the flow can lead to instances of cavitation when otherwise a steady homogenous flow might be well below the cavitation inception point, and can be consequential for the acoustic performance of the system, particulaly if such shedding resonates with characteristic frequencies of any of the machinery.
-
-Flow separation tends to occur earlier for laminar boundary layers than for fully turbulent boundary layers.
-Consequently, some systems are able to become more efficient  by deliberately inducing turbulent, where the benefit from reduced friction in the laminar boundary layer is outweighed by the cost suction induced by the larger wake field created by earlier flow separation.
-This is the reason that golf balls have dimples, and tennis-balls have fur, to enable them to fly further.
-The impact of the increased surface area, and more turbulent flow over the surface, is exceeded by the reduced wake field ( trailing area where the flow is totally separted) behind the ball.
-
-\section{Basics of Ducted Propellers and Pumpjets}
+\newpage  \section{Basics of Ducted Propellers and Pumpjets}
 It's easy to be somewhat confused by the many different names which seem to be associated with related, or similar systems.
@@
 
-\section{Constraints on the efficiency of pumpjets at low speed}
+\newpage  \section{Constraints on the efficiency of pumpjets at low speed}
 
@@
 \label{fig:BMTShape.png}
+\end{figure}
+
+\begin{figure}
+
+\begin{subfigure}[b]{0.9\textwidth}
+\includegraphics[width=0.8\linewidth, center]{BMTEfficiency1.png}
+\end{subfigure}
+\begin{subfigure}[b]{0.9\textwidth}
+\includegraphics[width=0.8\linewidth, center]{BMTEfficiency2.png}
+\end{subfigure}
+
+\caption{Studies on submerged waterjets for military surface craft currently show that the efficiency of jets decline sharply below 10kt \parencite{giles2010}}
+\label{fig:BMTEfficiency.png}
 \end{figure}
@@
 
-\begin{figure}[p]
-
-\begin{subfigure}[b]{0.9\textwidth}
-\includegraphics[width=0.8\linewidth, center]{BMTEfficiency1.png}
-\end{subfigure}
-\begin{subfigure}[b]{0.9\textwidth}
-\includegraphics[width=0.8\linewidth, center]{BMTEfficiency2.png}
-\end{subfigure}
-
-\caption{Studies on submerged waterjets for military surface craft currently show that the efficiency of jets decline sharply below 10kt \parencite{giles2010}}
-\label{fig:BMTEfficiency.png}
-\end{figure}
-
 It appears from recent examples in the literature that there is no evidence of jets approaching the same efficiency as propellers at speeds as low as 10kt.
@@
 
-\begin{figure}[p]
+\begin{figure}[h]
 \captionsetup{width=1\textwidth}
@@
 
-\begin{figure}[p]
+\begin{figure}[h]
 \includegraphics[width=\textwidth]{Comparison.png}
@@
 
-\begin{figure}[p]
+\begin{figure}[h]
 \captionsetup{width=0.8\textwidth}
@@
 
-\begin{figure}[h]
+\begin{figure}
 \captionsetup{width=0.6\textwidth}
@@
 However, in practice the water encountered by the propulsor is slowed by the wake field, and so the drag encountered by the vessel may increase faster than velocity of the water entering the intake, an effecnt which is likely non-linear with vessel speed.
-So whilst the trend with regards to vessel speed can be known with regards to the graphs in \cite{fig:PumpjetEfficiency}, a direct translation of the x-axis used to vessel speed at any given point is difficult prove without additional system information.
+So whilst the trend with regards to vessel speed can be known with regards to the graphs in \cite{fig:PumpjetEfficiency.png}, a direct translation of the x-axis used to vessel speed at any given point is difficult prove without additional system information.
 One might generally infer that the sorts of declines in efficiency observed as vessel speed discussed in recent literature correspond with movement towards very low values of $\Delta V / V_0$, though it is possible that alternative derivations have been used, or that the phenomenon known to occur in practice (such as flow separation) aren't adequately captured by such ideal equations derived from momentum theory in order to make a direct translation into real-world application.
@@
 
-\section{Model Structures and Assumptions}
+\newpage  \section{Model Structures and Assumptions}
 
@@
 
-\begin{figure}[p]
+\begin{figure}
 <<plot_efficiency, echo=FALSE, warning=FALSE, message=F >>=
@@
 
-\section{Results and Discussion}
+\newpage  \section{Results and Discussion}
 
@@
 
-\begin{figure}[p]
+\begin{figure}
 <<plot_power1, echo=FALSE, warning=FALSE, message=F >>=
@@
 
-\begin{figure}[p]
+\begin{figure}
 <<plot_results1, echo=FALSE, warning=FALSE, message=F >>=
@@
 
-\begin{figure}[p]
+\begin{figure}
 <<plot_results2, echo=FALSE, warning=FALSE, message=F >>=
@@
 
-\begin{figure}[p]
+\begin{figure}
 <<plot_results3, echo=FALSE, warning=FALSE, message=F >>=
@@
 
-\begin{figure}[p]
+\begin{figure}
 <<plot_results4, echo=FALSE, warning=FALSE, message=F >>=
@@
 
-\begin{figure}[p]
+\begin{figure}
 <<plot_proportions, echo=FALSE, warning=FALSE, message=F >>=
@@
 
-\begin{figure}[p]
+\begin{figure}
 <<plot_sens_hotel, echo=FALSE, warning=FALSE, message=F >>=
@@
 
-\begin{figure}[p]
+\begin{figure}
 <<plot_sens_battery, echo=FALSE, warning=FALSE, message=F >>=
@@
 
-Writing a paragraph here...
-
-\begin{figure}[p]
+\begin{figure}
 <<plot_sens_corners, echo=FALSE, warning=FALSE, message=F>>=
@@
 
-\section{Acoustic Considerations} \label{acousticconsiderations}
+\newpage  \section{Acoustic Considerations} \label{acousticconsiderations}
 
@@
 
-\begin{figure}[p]
+\begin{figure}
 <<plot_inception, echo=FALSE, warning=F, message=F>>=
@@
 
-\section{Conclusion}
-
-The selection of the pumpjet propulsion system is quite probably the most significant new technology selection for Australia's future submarines.
-Unlike other new technolgoies such as Air Independent Propulsion and Lithium Ion Batteries, which are either in service or entering service in multiple other navies, no other conventionally powered submarines have adopted pump-jet technology, which is currently only used by nuclear submarines.
-
-Despite acknowledging that the pumpjet may have an adverse impact on dived endurance \parencite{sbs2017}, Naval Group justifies the superiority of the pumpjet in the following terms: "In a confrontation between two otherwise identical submarines, the one with pump jet propulsion always has the tactical advantage."  No further explanation of this superiority has been provided in detail, but it is widely held that the pumpjet offered acoustic advantages which were highly regarded by those Australians who were responsible for selecting the design partner \parencite{cameron2016}.
-
-The comparison between two otherwise identical submarines (though not necessarily limited to engaging just each other in combat), provides a useful and suitable framework for evaluating a single significant technology choice, where most other significantly novel or new technologies were apparently excluded from consideration \parencite{stanford2017}.
-The conclusion of this report is that the following statement can be made with confidence:
-\n
-\emph{In a comparison between two otherwise identical submarines, the one with the pump jet will alwayas have a lower dived endurance, a lower dived range, a worse indescretion ratio, a lower overall endurnace, and a lower overall range, than the one with a conventional propeller.
-This will confer a substantial tactical and strategic advantage on the conventionally propelled submarine in a very broad range of operational scenarios.}
-\n
-
-Whilst it is broadly acknowledged that propellers have superior efficiency at lower speeds, but jets might have higher efficiency at higher speeds, the precise point of cross-over seemed unclear in public debate.
-One informed commentator suggested that the crossover point was likely to be below or near the transit speed of the submarine, and that the pumpjet may offer enhanced range and endurance over the whole mission profile as a consequence\parencite{davies2017}.
-This report finds no evidence of a pumpjet of the type used on submarine every being recommended as superior for efficiency in general, and even closely related technologies for more commecial use not advocated below 25kt.
-During my research I encountered domain experts confident that a pumpjet could simply not feasibly be superior to a suitable propeller at speeds below 18kt.
-Evidence of propellers remaining generally superior to waterjets and pumpjets in the speed ranges up to and even over 30kt is abundant.
-Consequently, the possibility of some cross-over point leading to any kind of range or endurance advantage relevant for a conventional submarine can be safely ruled out.
-
-This report also explores in some detail the theoretical concepts which drive the limitations of pumpjet performance at lower speeds in comparison to propellers.
-Enduring and fundamental principals of momentum theory provide robust limitations on the circumstances where a propeller's 'ideal efficiency' might be improved by the addition of a duct/or shroud, one type of which constitutes the transition to a 'pumpjet'.
-These principals stronly support the the notion that pumpjets will not and indeed cannot suddenly overcome their efficiency disadvantage at low speeds by means of technical ingenuity or innovation.
-
-Furthmore, a particular hydrodynamic phenonomon called 'flow separation' can be identified as the primary cause of the pumpjet's declining efficiency at very low speeds.
-Put simply, the duct provides an additional hurdle or barrier though which the water must pass when moving through the propuslor.
-At very low speeds waterspeeds, this additional resistance is sufficient to stop some of the flow progressing through the jet, creating unsteady eddies and vortices which sap energy from the system, but produce no thrust, driving down efficiency to zero at very low flow-rates.
-
-The design of the pumpjet for military purposes, namely the reduction of acoustic signatures effectively mandates that this must be the way that pumpjets work.
-The act of slowing the water down is necessarily equivalent to the act of raising its pressure.
-It is this elevated pressure at the point where the blades to their work on the water which enables the blades to avoid cavitation (the rapid expansion and collapse of small voids or bubbles in the water) which is very large source of noise underwater.
-Maintaing this advantage to the same level necessitates a duct design which slows water down to the same extent.
-The act of slowing water down when it's already going slowly is what causes some of it to stop, which is when flow separation occurs.
-
-In order to assess the probable magnitude of the impact of this reduction in efficiency at low speeds, a numerical model was developed, based on known principles of physics, and publically available information relating to the likely technologies involved in the future submarine.
-The most important and variable assumptions - the efficiency curves of the pumpjet and propeller - were taken from a recent study lead by world-leading engineering companies, supported by the US and Royal Navies, comparing closely related propulsion systems \parencite{giles2010}.
-Knowing that the systems would not be identical to the propulsors on submarines (the precise design of which is obviously secret), additional curves varying from these assumptions were also modelled in all directions, and the combinations combined and tested.
-
-The model has been made available publically in the form of a web-app, so that alternative assumptions, including alternative efficiency curves may be tested by the general public.
-
-The model in its central assumptions was able to achieve very realistic results, which were consistent with other public estimates of performance parameters of conventional submarines and known characteristics \ref{buckingham2008submarine}.
-A variety of different assumptions regarding the hotel load, battery density and different combinations of efficiency curves were tested.
-The central assumption values showed that a submarine with a conventional propeller had a very substantial advantage in terms of range and endurance across the entire speed range.
-This advantage reached a proportional maxium at around 6kt of over 60\%.
-However, in nominal terms, the greatest differences occurred at slightly lower speeds.
-Under central assumptions, the propeller submarine had an advantage in submerged range of some 200nm at 5kt, and approximately a two-day advantage in dived endurance at 4kt.
-
-The full range of plausible efficiency curves exhibited a significant efficiency advantage right through transit speeds.
-Central assumptions suggested that this would result in the overall range of a submarine being enhanced by approximately 50\% with the use of propellers instead of a pumpjet.
-None of the scenarios suggested it was likely that this range advantage would be any less than 20\%.
-Given the significant emphasis that has been repeatedly placed the unique requirements for range and endurance for Australia's submarine, this raises substantial questions about suitability of this technology to meet Australia's needs.
-
-The model was used to test a wide range of different assumptions.
-A further key result of the study established the sensitity of the difference to different assumptions regarding hotel load, and battery energy density.
-This showed that the difference between the two systems was made substantially larger with smaller hotel loads, and larger stored electical energy (as might be achieved with lithium-ion batteries), and these two effects combined to give some particularly startling results.
-Circumstances where the \textit{difference} in dived range and endurance between the two systems in terms of range and endurance exceeded the \textit{entire} dived range and endurance achieved by many modern conventional submarines frequently arose.
-
-The model demonstrated clearly that all of the most pronounced advantages in dived range and endrance occurred at speeds around 3-7kt, generally within plausible speeds achievable with modern AIP systems.
-The addition of AIP was briefly considered which revealed that a difference in dived endurance of the order of one month, or thousands of nautical miles, would likely have emerged between the two propulsion systems.
-
-The consequences of flow separation also has acoustic consequences, which this report explores.
-The vortexes and eddies which are created by flow separation tend to be unsteady, and shed periodically into the main flow.
-This results in pressure fluctations inside the shroud and more intense, unsteady large-scale types of turbulence occurring as these vortices are ingested into the rotor blades.
-These effects are are likely to cause some accoustic signature which becomes lower in frequence as the submarine slows down further and the vortexes become larger, and shed more slowly.
-This is in contrast to early stages of cavitation, which would be high frequency at when bubble sizes are small, and move to lower fequences as speed increases and bubbles expand.
-Lower frequencies are of particular significance because they tend to propagate further, and would be less effectively shielded by the duct.
-
-It remains plausible that in the absence of cavitation the noise generated in the propulsor by such flow separation is not loud enough to be of significant tactical concern.
-However, it is extremely unlikely that at some very low speeds, where conventional propellers experience no cavitation and enjoy steady, smooth flows over the blades, that a pumpjet could actually have a lower acoustic signature, even in terms of radiated noise.
-Flow separtion demand that two or three times as much energy is wasted by a pumpjet creating and destroying these unsteady vortexes than a propeller would require to create equal thrust.
-Consequently, the claim that pumpjets are generally acoustically superior should be treated with some caution.
-This claim has strong grounds wherever a conventional propeller might experience cavitation, such as at higher speeds.
-But at some very low speeds it is unlikely to be true.
-
-Overall, it seems unlikely that the full range of consequences of pumpjet choice have been fully comprehended by the Australian government.
-The scale of the probable impact on range and endurance is quite probably so substantial that it is difficult to see how such a performance penalty is consistent with the government's stated aim of acquiring a regionally superior submarine.
+\newpage
 
@@
 
+\newpage
+
+\appendix
+\section{\\Some essential concepts}
+
+\subsection{Conservation of Energy}
+
+This is perhaps one of the most fundamental and well-established principals in physics.
+The essential idea is that energy can move or change in form, but it isn't ever created or destroyed.
+Machines, plants and animals all derive their energy from a particular other source, which can be measured and evalutated to establish the limits of energy available.
+Plants collect energy from sunlight falling on their leaves.
+Humans (as well as combustion engines) capture the chemical potential energy in organic matter, and release it by combining it with oxygen.
+Hydro-electric power plants turn the gravitational potential energy of water stored at a height into electricity.
+
+Conservation of energy has a particularly relevant embodiment in fluid flows, which is given it's primary expression in Bernoullis equation, as given in Equation \ref{eq:bernoullis}.
+It says that the total energy in a connected body of fluid with no additional work being done on it is constant ($C$), though it can change in form between kinetic energy (movement, $\frac{1}{2}\rho v^2$), gravitational potential energy (it being elevated, $\rho gz$, relevant for liquids, less for gas) the heat energy in the fluid (not shown, not relevant for incompbressible fluids), and the pressure of the fluid $p$.
+
+\begin{equation}
+\label{eq:bernoullis}
+\frac{1}{2}\rho v^2 + \rho gz + p = C
+\end{equation}
+
+In different fluids, the dynamics of how energy moves between one and another change.
+For example, in gases, heating up a confined piece of gas will increase it's pressure, or if it is unconfined, increase its volume.
+This is particularly important for understanding gas turbines.
+In the case of a liquid, however, all the molecules are in close contact, and hence can't increase in volume or pressure substantially except by the creation of steam.
+As such, in the absence of large amounts of cavitation, the terms in the relationship which are most important for our consideration are the relationships between pressure, and kinetic energy, and gravity.
+
+In the case of most waterjets which propel surface vessels, the water is lifted from the bottom of the hull at the intake to the pump, which is generally incorporated inside the hull.
+This increase in gravitational potential coincides with a slowing down of the water relative to the vessel-speed.
+Since the jet in such vessels is generally ejected at the same height as the pump, this potential is never regained, and is technically a loss, however at high speeds such a loss is small relative to the total power ouput.
+In the case of submarine pumpjets, located on the end of the hull and parallel to the axis of the submarine, this effect (and the term $gz$) are not relevant.
+
+\subsubsection{Venturi Effect}
+In the case of submarines and torpedoes the water doesn't undertake a change in height, since intake and nozzle are generally all in line with the central axis of the submarine or torpedo.
+As a consequence, the key relationship in Bernoullis equation is the relationship between the liquids velocity, and its pressure.
+The consequence is that when water moves through a pipe (duct, or shroud) it's pressure is inversely related to the square of its velocity.
+This means that when a liquid is forced to to travel through a narrowing pipe, it's pressure necessarily decreases as the velocity increases.
+This is what underlies the Venturi effect.
+
+\subsubsection{Flow Diffusion}
+The inverse process is where a pipe increases in volume, and the flow is forced to slow down in order to fill the wider area, and the pressure correspondingly increases.
+This is a process called 'diffusion', and is important to achieving high pressure levels in many types of water pumps, including those which will be particularly relevant for pumpjets for watercraft propulsion, particularly elevating static pressure levels in order to prevent cavitation \parencite{hamilton1997, mccormick1963design, bruce1974}.
+
+\subsubsection{Energy waste and efficiency}
+The conservation of energy also has useful implications for how the efficiency of systems is thought about.
+In particular, because energy is conserved, identifying inefficiencies in a system necessarily involves identifying where energy goes to doing tasks which aren't useful for the intended purpose.
+A perfectly efficient system won't do any work that isn't for the intended purpose.
+In the case of analysing the efficiency of propulsion systems, the relevant 'work' is almost always related to moving water backwards to produce thrust.
+Moving water in directions other than backwards, including random turbulent flows which wind producing heat rather than thust, are two examples of wasted energy.
+Noise in the water also reflects energy which is wasted, though the significance of the acoustic signature is not necessarily indicative of the magnitude of energy wasted.
+
+In general terms, energy waste in propulsion systems will result in turbulence (moving water in random chaotic patters, where the energy soon becomes smoothly distributed as heat) and acoustic energy that is propagated away.
+Flow separation incolves the creation of larger-scale vortices and eddies, which might be considered as a special case of turbulence, since the length-scale of the unsteady flows is much larger than that of normal turbulence, for instance in a boundary layer.
+
+On important consequences is that whilst increasing creation of noise, flow separation, or turbulence necessarily represent greater amounts of energy wasted, some increase in any given one does not neccessarily indicate the reduction of efficiency of the entire system, as larger amounts of energy may be being saved through reduced wastage through another means, or vice versa.
+For example, a squeaky join on a car might make a lot of noise, and it will represent some loss of kinetic energy.
+However, the brakes can be applied and absorb vastly larger amounts of energy, essentially silently.
+
+Furthermore, efficiency of a system is necessarily given by the a ratio of useful work to total energy expended.
+Some trends which show increasing levels of energy wasted might be outweighed by larger increases in the overall amount of productive work being achieved.
+Pumpjets, which generally tend to require highly turbulent flows, and even more turblent flows at higher speeds, still can become more efficient as speed inceases, as the amount of useful thrust produced increases even faster.
+So, despite the wasteful round-and-round movements increasing with speed, the increase in productive front-to-back movement improves even faster, and total efficiency rises.
+
+\subsection{Boundary Layer}
+When water or any fluid flows with some speed relative to another solid surface nearby, there is some layer adjacent to the surface in which the speed of the fluid is diminished relative to the main flow.
+At a microscopic level, there are some molecules of the fluid on the surface which will be effectively static relative to the surface.
+The layer of fluid that joins the gap between the static surface, and the part of the flow which is moving at the full flow speed, is called the boundary layer.
+Exactly how thick the boundary layer is, and how the fluid moves in the boundary layer, is extremely important for consideration of efficiency of fluid flows over and around solid surfaces.
+In particular, a boundary layer can be either turbulent or laminar in nature, and can transition to turbulent flow after a short distnace of laminar flow, as shown in Figure \ref{fig:BoundaryLayer.png}.
+
+\begin{figure}[h]
+\captionsetup{width=0.5\textwidth}
+\includegraphics[width=0.5\textwidth, center]{BoundaryLayer.png}
+\caption{The development of a boundary layer as shown in \parencite{mollard2011}}
+\label{fig:BoundaryLayer.png}
+\end{figure}
+
+\subsection{Turbulent and Laminar Flow}
+The way that fluids move relative to a surface can one of either two methods.
+In 'laminar flow' all the fluid moves in one direction in an smooth and orderly manner, with very little mixing between the layers of fluid travelling at different speeds.
+
+The alternative is 'turbulent flow', where the fluid moves around in unpredictable swirls and circles as well as moving overall in an predominant direction.
+There is considerable mixing between all the different layers, and the average speed remains relatively constant in the flow, with the exception of the flow immediately adjacent to the wall, or in the boundary layer.
+Becuase turbulence involves a lot of movement that is not overall in one productive direction and contributing to thrust, it necessarily leads to some loss of energy from the overall thrust.
+Once the swirls and movements become smaller and smaller, the energy winds up simply as heat in the fluid.
+
+\begin{figure}[h]
+
+\begin{subfigure}{0.5\textwidth}
+\includegraphics[width=0.7\linewidth, center]{Laminar.png}
+\caption{Laminar}
+\end{subfigure}
+\begin{subfigure}{0.5\textwidth}
+\includegraphics[width=0.7\linewidth, center]{Turbulent.png}
+\caption{Turbulent}
+\end{subfigure}
+
+\caption{A simple comparison of laminar and turbulent flows from \parencite{NPdotnet2017}}
+\label{fig:LaminarTurbulent.png}
+\end{figure}
+
+The likely point of transition between a flow being turbulent and laminar can be given by unitless number called the Reynolds Number, which is determined by a fluids velocity, viscoscity, density, and a characteristic length-scale over which the flow occurring.
+A fuller discussion of these important concepts can be found in \cite{NPdotnet2017}.
+
+Turbulent flows in gasses can be extremely noisy, since those fluids are compressible.
+Consequently, the rapid and intense ciruclar movements result in a lot of oscillatory compressions against the surrounding air.
+Consequently things like jet engines, vacuum cleaners, hand-dryers, and other devices that create rapid movements in gasses tend to be quite loud, including at some distance.
+
+In essentially incompressible fluids, the random changes in pressure and velocity which are involved in laminar flow tend not to generate nearly as much noise in the surrounding liquid outside the flow.
+Because the fluid is incompressible, all of the random 'round and round' movements don't amount to much 'in and out' movement, which is what creates the pressure waves which result in propagated noise.
+Consequently, whilst turbulent flows do inevitably generate some noise, the amount of noise generated is dramatically greater in the presence of cavitation, where a void opens up in the water, and vastly more 'in and out' movement is able to occur.
+
+\subsection{Cavitation}
+Cavitation is the rapid expansion and collapse of a void, or bubble in water.
+Put technically, caviation occurs when the local static pressure (pressure in the rest frame of the fluid) falls below the vapour pressure of the fluid (pressure at which the liquid will start to boil).
+
+To try to conceptualise how it occurs in the context of propulsors, one might imagine what happens when something moves very fast through water.
+The front side of the object pushes the water forward, but on the back side the water has to push in to fill the space left behind.
+If there’s not enough pressure to push the solid water in fast enough, a gap opens up, with just a few gaseous water molecules (steam) inside the cavity.
+(This is also described as water boiling at low pressure.)  However the gap doesn’t stay around for long.
+Soon the water catches up, and the bubble implodes with a pop, leaving only very tiny bubbles as a result, which you can see in the wake of almost any boat or ship moving at speed.
+Whilst some energy turns to heat (the remaining steam in the tiny bubbles) some is propagated away as a sound-wave generated by the implosion.
+
+A watching water boil in a glass kettle gives a quick and intuitive insight into the occurrence of cavitation.
+Quickly after the kettle starts heating, a considerable noise can be heard, which corresponds to the commencement of cavitation on heating element.
+At some local point for a moment in time, there is enough energy for the water present to boil.
+It tends to be on rough surfaces or in the presence of some impurity that cavitation will occur first (in the blackened part of the surface in this case.)  However, whilst the bubbles on the bottom can be seen plainly on the bottom, they collapse almost straight away again, and leave only a tiny bubble of stable steam circulating in the water.
+It is the rapid expansion and collapse of these bubbles that causes the noise of a kettle, long before it has boiled.
+
+\begin{figure}[h]
+
+\begin{subfigure}{0.5\textwidth}
+\includegraphics[width=0.9\linewidth, height=5cm]{EarlyCavitation.JPG}
+\caption{}
+\end{subfigure}
+\begin{subfigure}{0.5\textwidth}
+\includegraphics[width=0.9\linewidth, height=5cm]{BoilingCavitation.JPG}
+\caption{}
+\end{subfigure}
+
+\caption{(a) Cavitation bubbles form and collapse creating noise and some tiny bubbles long before boiling occurs.
+(b) Only at very high temperatures do the bubbles endure at full-size in the water and reach the surface.}
+\label{fig:EarlyCavitation.JPG}
+\end{figure}
+
+It is only when the water is all at a much higher temperature that the bubbles remain their full size for long enough to detatch and rise all the way to the surface, a process we typically think of as boiling water.
+It is worthwhile noting that the sound emitted at this stage is much softer and lower than the early onset of cavitation.
+Larger bubbles result in lower frequencies of sound being emitted.
+Even at higher temperature, the rougher parts of the surface provide the points where all the cavitation originate.
+
+Cavitation is extremely important in the study of ship propulsion, since it's occurrence in particular circumstances can lead to substantial losses of efficiency, as well as damage to the propeller and related appendages.
+It is particularly important for submarines, since the expansion and collapse of these bubbles tends to lead to the creation of noise which is often far larger, and more distincly characteristic than other turbulent disturbences in the water when no cavitation is present.
+Consequently, the onset of cavitation can be thought of as a distinct threshold in terms of the acoustic performance of propulsive system.
+
+Despite cavitation necessarily representing some energy being wasted generating unwanted noise in the water, some cavitation inevitably occurs around most propulsion systems operating at full power.
+In plenty of cases, the consequences for efficiency are relatively small, as they tend to be dominated by other efficiency considerations in imparting thrust.
+Put another way, the savings that can be gained by minimising wastage to turblence can outweigh the losses incurred by having some cavitation occur.
+In most cases, propellers are designed to work with a certain extent of cavitation for optimum efficiency under working loads \parencite{shin2015}.
+Cavitation also occurs to a considerable extent within the jets of most high-speed surface vessels, but doesn't necessarily have a particularly bad effect on their overall efficiency at their intended speed levels.
+It is generally only within very specialised military circumstances when the absolute avoidance of cavitation supercedes other concerns for efficiency, and require the elimination of all cavitation entirely \parencite{mollard2011}, \parencite{lewis1988}.
+These circumstances include the design of submarines and torpedoes.
+
+In certain special cases, particularly supercavitating propellers or surface-piercing drives, high levels of cavitation can be higly advantageous for efficiency.
+However these represent specialised designs, generally only for very high speed vessels, where acoustic concerns are negligible, and are of little concern for this particular endeavour.
+
+Cavitation can occur in a number of different ways at different points and forms on or around the propeller.
+It is common for cavitation to occur first near the out extremities of the propeller, where the blades are moving at the highest velocity relative to the water-flow.
+Cavitation tends to spread across the back of the blades in a sheet (sheet cavitation), since the back side of the blades are generally the areas of most depressed pressure, though it can also occur in places on the back side of the blades, where the dynamics of the water moving around the blade can produce points of near the edges of significantly reduced pressure.
+On surface vessels cavitation tends to occur most when the blades are closes to the surface, where the pressure is lowest.
+
+\begin{figure}[h]
+\captionsetup{width=0.7\textwidth}
+\includegraphics[width=0.7\textwidth, center]{IncreasingCavitation.png}
+\caption{Cavitation frequently occurs first at the extremities, then spreads inwards across the blad face, as shown in \parencite{mollard2011}}
+\label{fig:IncreasingCavitation.png}
+\end{figure}
+
+\begin{figure}[h]
+\captionsetup{width=0.7\textwidth}
+\includegraphics[width=0.7\textwidth, center]{CavitationTypes.png}
+\caption{Different types of cavitation can occur on different parts of the propulsion system as shown in \parencite{carlton2007}}
+\label{fig:CavitationTypes.png}
+\end{figure}
+
+\subsection{Pressure, Thrust, and Momentum change}
+A few simple related concepts make up the foundations for many of the simplest and enduring theoretical models (Momentum Theory, still widely and routinely used to model 'ideal' propeller efficiencies by considering the propeller simple disc \parencite[169]{carlton2007}) and other quite intuitive concepts which migth be more readily grasped in understanding propulsors.
+
+Thrust is a force, measured in Newtons, which is generally used to refer to the forces acting on a vessel or object by its propulsion system to move it forward in the water.
+Generating thrust is consequently a key objective of any propulsion system.
+
+Pressure is a measure of force divided by an area over which it is applied.
+Pressures are necessarily applied to some extent over all of the surfaces of any object, including due to the atmostpheric pressure of air.
+In considering the efficiency of propulsion systems for solid bodies, thrust is always the consequence of some \textbf{net} pressure difference on different sides and surfaces of an object.
+For instance, on a propeller blade, the rear-facing side of the blade is the side that pushes the water backwards, and hence has a raised pressure on its surface.
+Being generally backwards facing this acts to push the propeller forwards.
+This pressure acts equally to push water backwards, as well as to push the propeller forwards.
+Importantly, the forward-facing side of the propeller will expereince a lowered pressure, which also serves to pull water backwards, and suck the propeller forwards.
+It is the sum of all the pressure differences on the propeller that result in a net force.
+
+Importantly in hydrodynamics, it is important to consider the possibility of other forces being generated other surfaces which may have an impact on the movement of the vessel.
+In the presence of the hull, water being drawn by the propeller can cause pressure changes on other surface of the vessel which may assist or impede producing a net positive thrust.
+Submerged jets represent a particular case in point.
+A famous home experiment suggests that one compare the thrust (backwards pressure) experienced by someone holding a garden hose creating a jet of water when it is in the air, as opposed to submerged in a pool of water.
+When submerged, the backwards pressue appears dramatically reduced.
+In fact, this phenomenon is due to a the water-jet drawing adjacent water with it under-water, the movement of which exerts a negative pressure on the outside of the nozzle, conteracting the pressure exerted on the inside of the nozzle.
+This effect represents one of the important distinctions between the jets which release their water below or above the water-line, and generally means that underwater jets have much lower jet velocity ratios than those released over water.
+
+It is also an important law of physics that force or thrust is equal to the rate of change of momentum.
+Put simply, the speed at which one speeds something up, or the increasing amount of something which begins to move, is always equivalent in magnitude to the thrust that is produced.
+Since momentum is a vector, this simple law requires that the momentum change has to net out to one particular direction in order to produce a net thrust, which provides the core principal for Momentum Theory as it applies to propeller and jets.
+Circular movements necessarily are cancelling, and produce no thrust.
+It is useful to recall however that this principal is simply another complete, alternative way of measuring the consequences of the net pressure differences discussed earlier.
+The total rate of change of momentum in the system which is induced as a conseqence of the pressure differentials on all sides of the propulsor must necessarily be equivalent to the total net force that is produced.
+For the garden hoze in the pool or bucket, the jet rapidly entering the liquid in one direction necessitates new flows of liquid in other directions, reducing net moment changed.
+This new liquid movement will exert pressure reductions on the nozzle surface surface, which reduce the net momentum change affected by the entire system.
+These net forces, and net momentum change are necessarily equivalent and equal in effect.
+Consequently literature seeking to optimise propulsion might discuss analysis referring to both of these effects or phenomena, and neither are in conflict.
+
+Full and proper derivations of many of the thrust equations for different propulsion systems from these principals can be found in plenty of authoritative works, including \cite{lewis1988}, \cite{carlton2007} and \cite{mollard2011}.
+
+\subsection{Flow separation}
+Flow separation refers to a circumstance when the flow over a surface separates entirely from the surface, and a new eddy or votex is formed where the flow actually moves in the opposite direction to the dominant flow.
+It is of particular significance in hydrodynamics because it can lead to substantial losses of efficiency, as energy is diverted into the kinetic energy of the eddy or vortex, which is unproductive.
+It tends to occur near the surface of aerofoils with high angles of attack, or in the blades of turbomachinery including diffuser or mixed-flow pumps and ducted propellers operating in off-design conditions, including low flow rates \parencite{li2013}, parencite{bakker2006}, parencite{mcbride1979}, parencite{wislicenus1986}.
+
+\begin{figure}[h]
+\captionsetup{width=0.5\textwidth}
+\includegraphics[width=0.5\textwidth, center]{SeparatedAerofoil.jpg}
+\caption{Flow separation occurring over the top of an aerofoil in a wind tunnel. Image courtesy of Deutsches Zentrum fuer Luft- und Raumfahrt e. V. (DLR)}
+\label{fig:SeparatedAerofoil.jpg}
+\end{figure}
+
+In order for flow separation to occur, a fluid must be moving against an adverse pressure gradient, which simply means that the static pressure (pressure in the rest frame of the fluid) is increasing in the direction of movement, which means the flow is slowing down as it moves, unless new work is being done on it.
+At the boundary layer, where the flow velocity is already reduced due to drag forces experienced near the surface, the adverse pressure gradient can be enough to reverse the flow altogether.
+In this case, the fluid flows in reverse near the surface and forms a vortex, and the main flow becomes separated form the surface by the votex.
+
+\begin{figure}[h]
+\captionsetup{width=0.7\textwidth}
+\includegraphics[width=0.7\textwidth, center]{FlowSeparation.png}
+\caption{Flow separation involves flow reversal near a surface when a fluid is moving against an adverse pressure gradient, as shown in \parencite{mollard2011}}
+\label{fig:FlowSeparation.png}
+\end{figure}
+
+In addition to the necessary loss of efficiency, flow separation can often tend to result in unsteady flows, with the vortices periodically being shed into the flow \parencite[480]{mollard2011}.
+If this occurs ahead of the blades of a propeller or impeller, such disturbances to the flow can lead to instances of cavitation when otherwise a steady homogenous flow might be well below the cavitation inception point, and can be consequential for the acoustic performance of the system, particulaly if such shedding resonates with characteristic frequencies of any of the machinery.
+
+Flow separation tends to occur earlier for laminar boundary layers than for fully turbulent boundary layers.
+Consequently, some systems are able to become more efficient  by deliberately inducing turbulent, where the benefit from reduced friction in the laminar boundary layer is outweighed by the cost suction induced by the larger wake field created by earlier flow separation.
+This is the reason that golf balls have dimples, and tennis-balls have fur, to enable them to fly further.
+The impact of the increased surface area, and more turbulent flow over the surface, is exceeded by the reduced wake field ( trailing area where the flow is totally separted) behind the ball.
+
 \end{document}

### 2018-01-13 scripts/build plots for paper.R
Beginning of IRs calculation

@@ -1,5 +1,6 @@
 
 library(submarines)
+library(plotly)
 
 df <- read_csv(find::this("/Submarines/eff-data-2018-01-03.csv"))
 
@@ -50,9 +51,9 @@ ls <- df %>%
 ls_comp <- ls %>%
   map( ~ .x %>%
          select(kts, eff.jet, eff.prop) %>%
-         run_subs(hotel = 50,
+         run_subs(hotel = 100,
                   total.batt = 500,
-                  batt.dens = 0.28,
+                  batt.dens = 0.14,
                   patrol = 2.5,
                   max.speed = 18,
                   max.power = 7,
@@ -171,7 +172,8 @@ produce_all_plots <- function(df) {
     scale_x_continuous(breaks=seq(0, 20, 2))
 
   #list(eff = eff, end = end, rng = rng, end_p = end_p,  pwr =pwr)
-  list(comp_plot_df = comp_plot_df, end_prop_plot_df = end_prop_plot_df, range_prop_plot_df = range_prop_plot_df, diff_plot_df = diff_plot_df)
+  #list(comp_plot_df = comp_plot_df, end_prop_plot_df = end_prop_plot_df, range_prop_plot_df = range_prop_plot_df, diff_plot_df = diff_plot_df)
+  list(power_plot_df = power_plot_df, end_prop_plot_df = end_prop_plot_df, diff_plot_df=diff_plot_df)
 }
 
 
@@ -232,6 +234,55 @@ df_full <- df_full %>%
 
 df_full_cornered %>% write_rds("extdata/df_full_diff_cornered.rds")
 
+########## Time to do some indiscretion ratios (IR)  #############
+
+df_filler <- ls_plots$`Set 3`$power_plot_df
+
+df_filler <- df_filler %>%
+  mutate(pair = "wide",
+         hotel = 100,
+         battery = 0.14,
+         total.batt = 500)
+
+#df_full <- df_filler
+
+df_full <- df_full %>%
+  bind_rows(df_filler)
+
+### checking df ###
+df_full %>%
+  ggplot(aes(x = kts, y = val, col = pair, linetype = cat))+
+  geom_line()
+
+### calculations to get IRs ###
+df_IRs <- df_full %>%
+  mutate(energy.kJ = total.batt*1000*battery*1000,
+         charge.rate.kw = 5500)
+
+df_IRs <- df_IRs %>%
[…]

### 2018-01-15 DraftExec.Rnw
add
with all the indiscretion ratios

@@
 %  appendix
-\usepackage[toc,page]{appendix}
+\usepackage[titletoc,page]{appendix}
 
@@
 \begin{titlepage}
-  \vspace*{\fill}
     \begin{center}
-    {Efficiency}\\[0.5cm]
-    {Aidan Morrison}\\[0.4cm]
+        \vspace*{1cm}
+
+        \vspace{10cm}
+        \textbf{\huge Efficiency}
+
+        \vspace{0.5cm}
+
+        \vspace{1.5cm}
+
+        \textbf{Aidan Morrison}
+
+        \vspace{1.5cm}
         2018
+
+        \vfill
+
     \end{center}
-  \vspace*{\fill}
 \end{titlepage}
@@
 
+\subsection{Indiscretion Ratios}
+Indiscretion Ratios represent the fraction of time that a submarine must spend on the surface snorting (running diesel engines) in order to maintain a constant speed.
+In order to calculate the Indiscretion Ratio, it is necessary to know how much power the diesel engines are able to generate when running while snorting, and how much energy (net of any efficiency losses, hotel load, and power drawn to maintain propulsion) can be directed to recharging the batteries.
+Reliable information regarding the arrangement of engines, generators is not generally available in the public domain.
+However, from \cite{patrick2012}, it can be seen that credible submarine designs might include diesel engines that have a total capacity in the region of 85\% of the maximum power output of the main motor.
+Assuming that there are some necessary losses in the generation and recharging process, and plausibly some increase in the hotel load for the operation of additional systems, fur the purposes of this assessment it is assumed that while snorting approximately 5MW might be available to meet propulsion and normal hotel load requirements, and any surplus can be directed towards charging batteries.
+
+The fraction of time required to charge the batteries to a given level at a given speed, relative to the amount of time that the batteries could sustain the submarine at the same speed, then results in an indiscretion ratio.
+For similar assumptions to those used for central scenarios, including that hotel load is 100kW, 500tonnes of batteries are embarked, and the energy density of the batteries is that of lead-acid at 0.14MJ/kg, the results for the different efficiency curves considered for the two propulsion systems can be seen in Figure \ref{fig:IRs}.
+It should be noted that since the precise machinery choices for the likely submarine cannot be known with great confidence, the indiscretion ratio levels here at any given point have high uncertainties attached.
+However, the relative difference between indiscretion ratios at different speeds (the shape of the curve) as well as the difference between the two systems, should be closely resemblant of a real comparison of otherwise identical submarines, even if the level is somewhat shifted.
+
+The graph in \ref{fig:IRs} shows that very low indiscretion ratios, below 5\%, could be plausible for either system at very low speeds under 3kt.
+However, these very low indiscretion ratios do not last for a significant range of speeds for the pumpjet system.
+Whilst for a propeller driven submarine, such very low rates of indiscretion last up to around 6kt, pumpjet propelled submarines are likely to start to see their indiscretion ratios start rising significantly after 4kt.
+In the region of transit speeds, around 8-10kt, the difference between the two systems becomes clear and substantial.
+
+It should be noted that these plots only extend to 13kts, in contrast to previous plots which extend to 20kt.
+Since the propulsive power required to reach higher speeds can consume effectively all the power produced by the diesel engines, and then even exceed it at some speeds, the indiscretion ratios calculated by the method above become extremely large or negative, and don't assist in useful comparison.
+
+\begin{figure}
+<<plot_IR, echo=FALSE, warning=FALSE, message=F>>=
+df <- read_rds("extdata/df_IRs_100.rds")
+df %>%
+  filter(kts < 13.5) %>%
+  ggplot(aes(x = kts, y = IR, col = level, linetype = system))+
+  geom_line(size = 0.8)+
+  scale_color_manual(values = c("black", "blue", "red"))+
+  scale_y_continuous(breaks=seq(0, 1.0, 0.05), name = "Indiscretion Ratio") +
+  scale_x_continuous(breaks=seq(0, 13, 1), name ="speed (kt)")+
+  labs(linetype = "System", col = "Level")
+@
+\caption{Some calculation of plausible indiscretion ratios show that the pumpjet would substantially increase the amount of time spend snorting}
+\label{fig:IRs}
+\end{figure}
+
+\begin{figure}
+<<plot_IR_sens, echo=FALSE, warning=FALSE, message=F>>=
+df <- read_rds("extdata/df_IRs_hotels_long.rds")
+df %>%
+  filter(kts < 13.5) %>%
+  ggplot(aes(x = kts, y = value, col = pair))+
+  geom_line()+
+  scale_y_continuous(name = element_blank(),breaks=seq(0, 2.0, 0.1)) +
+  scale_x_continuous( name ="speed (kt)", breaks=seq(0, 14, 1))+
+  labs(col = "Efficiency \nCurve Pair")+
+  facet_grid(key ~ hotel, scales = "free", switch = "y")
+@
+\caption{The additional indiscretion ratio necessitated by a pumpjet, and this difference expressed as a proportion of the indiscretion ratio of a propeller driven submarine.
+Sensitivity to hotel load is not as great as for endurance or range calculations.}
+\label{fig:IRs_hotel}
+\end{figure}
+
+Figure \ref{fig:IRs_hotel} demonstrates the difference between the two otherwise idendtical submarines' indiscretion ratios as a function of speed, and also what this difference represents as a proportion of the indiscretion ratio at that speed for the submarine as a propeller.
+It can be seen here that the indiscretion ratio is generally somewhat less sensitive to variation in hotel load than calculations for range and endurance.
+(Indiscretion ratio is also independent of energy density of the batteries, and hence will not be compared.) However, increasing the hotel load does still have the impact of somewhat diminishing the peak difference between the two systems, and moving the point at which the proportional difference peaks to higher speeds.
+Where the penalty of the pumpjet on a submarine with a hotel load of 50kW might have the largest impact at 5kt, it the difference would probably be greatest at 6kt and 7kt for a sumbarine with hotel loads of 100kW and 200kW respectively.
+
+The magnitude of the difference is also substantial.
+For the central efficiency curve and hotel load assumptions, the pumpjet would cause a sumbarine to increase the time spent snorting for by over 60\% between 5kt and 7.5kt.
+One would expect that this would be of considerable tactical consequence.
+These graphs also demonstrate that the difference does not diminish at higher speeds, infact it begins to increase again above 10kt, as the power required for propulsion begins consume much of the power generated by the diesel engines, and efficiency difference strongly impacts the size of the diminishing surplus available for charging.
+It is also clear that the sensitivity to the selection of efficiency curves remains quite strong in this case.
+It is clearly plausible that in a range of circumstances the pumpjet could double the time spent snorting.
+However, the difference between the lines representing the pair of central efficiency curves, and the narrowest separation, is noticeably smaller, and the impact of chaning hotel load on these lines is also proportionally diminished.
+As such, it might be said with considerable confidence that the most likely impact of the pumpjet on indiscetion ratios at any speeds over 4kt will around 40\% or higher.
+It would be extremely unlikely that the penalty induced by the pumpjet at speeds of 5kt and above would be any less than 25\%.
+
 \newpage  \section{Acoustic Considerations} \label{acousticconsiderations}
@@
 
-\appendix
-\section{\\Some essential concepts}
+\begin{appendices}
+\section{Some essential concepts}
 
@@
 The impact of the increased surface area, and more turbulent flow over the surface, is exceeded by the reduced wake field ( trailing area where the flow is totally separted) behind the ball.
+\end{appendices}
 

### 2018-01-15 DraftExec.tex
with all the indiscretion ratios

@@
 
+%  For caption width
+\usepackage{graphicx,caption}
+% For Centering figs
+\usepackage[export]{adjustbox}
+% For split figures
+\usepackage{subcaption}
+%  appendix
+\usepackage[titletoc,page]{appendix}
+
 \usepackage{draftwatermark}
@@
 
-\maketitle
-
-Para
+\begin{titlepage}
+    \begin{center}
+        \vspace*{1cm}
+
+        \vspace{10cm}
+        \textbf{\huge Efficiency}
+
+        \vspace{0.5cm}
+
+        \vspace{1.5cm}
+
+        \textbf{Aidan Morrison}
+
+        \vspace{1.5cm}
+        2018
+
+        \vfill
+
+    \end{center}
+\end{titlepage}
+
+\newpage
 
@@
 
-\section{Introduction}
+\newpage
+
+\section{Introduction} \label{intro}
 
@@
 
+\newpage  \section{Basics of Ducted Propellers and Pumpjets}
+It's easy to be somewhat confused by the many different names which seem to be associated with related, or similar systems.
+Between pumps, pumpjets, water-jets ducted or shrouded propellers, or impellers, there is plenty of grounds for some confusion  In this section, I aim to clarify in simple terms what the important differences between distinctly different systems are, and where some terms are used somewhat interchangeably without doing any great violence to the concepts underlying.
+Appreciating where the fundamental as opposed to superficial differences lie is important to understanding subsequently the limitations of different systems.
+
+\subsection{Propellers}
+Perhaps the best starting point is the most basic, and oldest of the systems which we're considering: the basic screw propeller.
+A propeller uses a number of tilted blades attached to a central hub to sweep around a disc in the water, and accelerate a column of water passing through the disc.
+Due to the slant of the blades, water on the back sides of the blades is pushed astern, and reduced pressure on the front faces pulls more water from ahead to replace it.
+As such, it accelerates a column of water, which necessarily is contracted in circumference after the point of acceleration.
+
+\begin{figure}[h]
+\captionsetup{width=0.5\textwidth}
+\includegraphics[width=0.5\textwidth, center]{PropellerAction.png}
+\caption{Propellers generate thrust by accelerating a column of water, as shown in  \parencite[247]{mollard2011}}
+\label{fig:PropellerAction.png}
+\end{figure}
+
+A comprehensive description of the development of the propeller can be found in \cite{carlton2007}.
+Here I won't elaborate on beyond describing some essential features and characteristics with which one ought to be familiar, primarily for the purpose of comparing different propellers and understanding their evolution into the systems such as pumpjets.
+
+Most propellers will have between three and seven blades.
+In general they are shaped as an aerofoil, with the convex side being upstream, just as the convex side of a plane wing is above, in order to generate lower pressure and lift as it moves through the air.
+The blades tend to be twised so that they have a higher angle of attack closer to the hub, and lesser closer to the extremities, so that those faster moving sections push their respective parts of the water column at an overall similar speed.
+The are often also thinner towards the extremities, and will be swept backwards as if dragged by their rotation through the water (skew) and also dragged by the ship's movement throug the water (rake).
+Whilst the degree of all these characteristics is highly variable for different applications, these are a few of the commonly referred to characteristics which can be varied in order to optimise performance for any given application.
+
+\begin{figure}[h]
+\captionsetup{width=0.5\textwidth}
+\includegraphics[width=0.5\textwidth, center]{SkewRake.png}
+\caption{The development of a boundary layer \parencite[262]{mollard2011}}
+\label{fig:SkewRake.png}
+\end{figure}
+
+\subsubsection{Pitch}
+Perhaps one of the most important characteristics of a propeller is its pitch.
+The pitch represents the distance that a blade section would travel forward if it carved its way around a helix through one full rotation.
+It is intuitive to think of as something akin to the angle-of-attack of the propeller blade to the water, however this is technically misleading as the movement of the water-column incoming to the blade, as well as the speed of rotation, also have a significant bearing on what the actual angle of attack of the blade ends up being.
+Whilst technically pitch is a distance (measured in meters), it is often expressed and used in formula as a ratio of the diameter of the propeller, to give it a unitless equivalent, as is common in the description of many characteristics of propellers and propulsion systems.
+
+\begin{figure}[h]
+\captionsetup{width=0.5\textwidth}
+\includegraphics[width=0.5\textwidth, center]{HelicalLine.png}
+\caption{A blade section of a propeller traces a helical line around a cylinder as it is rotated \parencite{carlton2007}}
+\label{fig:HelicalLine.png}
+\end{figure}
+
+\begin{figure}[h]
+\captionsetup{width=0.7\textwidth}
+\includegraphics[width=0.7\textwidth, center]{PitchDefinition.png}
+\caption{The definition of pitch is the distance traveled by a blade section along a cylinder, as given in \parencite{carlton2007}}
+\label{fig:PitchDefinition.png}
+\end{figure}
+
+The selection of the exact blade selection that is used to define pitch must be specified as being at some fraction of the radius from the centre, given the tendancy for the blades to twist.
+This distance is calculated to be the 'moment mean' or a technical derived effective average, which tends to lie between 0.6R and 0.7R.
+A thorough discussion of pitch can be found in \cite[35-37]{carlton2007}.
+
+Pitch is of particular importance to the discussion the efficiency of propeller and propulsor design because its optimal choice tends to vary considerably with the different loads which a vessel is intended to operate at, which can also relate to a vessel's design speed.
+It is for this reason that a considerable number of vessels actually have variable pitch propellers.
+Whilst considerably more complex, expensive, and heavier than a traditional fixed-pitch propeller, the ability to vary the pitch of a propeller assists considerably, particularly when the amount of load (resistance) a vessel is expected to face varies considerably.
+Controllable pitch propellers also have advangates in terms of manouverability, since the pitch can be reversed and hence reverse thrust can be produced, without requiring the direction of the power coming from the engines to be reversed.
+This has advantages for ferries which undertake frequent docking manouvres in confined spaces (\cite{MAN2017}, \cite{lewis1988}).
+
+\begin{figure}[h]
+\captionsetup{width=0.8\textwidth}
+\includegraphics[width=0.8\textwidth, center]{CPPropeller.png}
+\caption{Controllable Pitch Propellers (CP Propellers) allow the pitch of the blades to be varied. Image credit: \parencite{mollard2011}}
+\label{fig:CPPropeller.png}
+\end{figure}
+
+\subsubsection{Pitch and the Advance Ratio}
+Pitch is also an extremely important concept to understand because it relates closely to the way that a propeller's 'open-water efficiency' $\eta_O$ is expressed in charts, as well as the advance ratio $J$ that usually comprises the x-axis in such charts.
+
+\begin{figure}[h]
+\captionsetup{width=0.8\textwidth}
+\includegraphics[width=0.8\textwidth]{OpenWater.png}
+\caption{A typical chart showing open water efficiency of a propeller, in this case a Wageningen B5-75. Image: MARIN}
+\label{fig:OpenWater.png}
+\end{figure}
+
+This chart also shows two other closely related dimensionless coefficients, the torque and thrust coefficients ($K_Q$ and $K_T$) which we shall not dwell on here.
+However, it is important to note that a number of different lines are expressed on this sort of chart, which represent different versions of the same propeller, with only the pitch (expressed in its dimensionless form as $P/D$) being different in each case.
+The x-axis, $J$, also warrants a clear definition:
+
+\begin{equation}
+\label{eq:2}
+J = \frac{V_a}{nD}
+\end{equation}
+
+Where $V_a$ is the speed of advance of the propeller, or the speed at which water arrives at the front-side of the propeller, and $n$ represents the rate of the propeller's rotation (revolutions per second) and $D$ is the propeller's diameter.
+For fullness of understanding it is also worthwhile to note here that $V_a$ does not correspond directly to the vessel's speed of travel, though they are related.
+To some extent, a vessel moving through water will always draw water in its wake along with it, which means that in the wake (where the propeller tends to operate to some extent or other) the water incident on the propeller will arrive at a lower speed than that implied by the ship speed \parencite[15]{MAN2017}.
+Another key distinction for the consideration of jets later on is that the speed of advance does not necessarily incporate the movement of water due to the movement induced by the propeller (or duct) itself, which is a distinct effect to that of the wake field.
+In \cite[213]{lewis1988} $V_A$ is used specifically to denote the undisturbed stream velocity, since the role of certain ducts can be to accelerate or decelerate the stream prior to the water arriving at the blades of the propeller.
+
+One may also note from the previous figure that for each line representing a given pitch, the efficiency falls precipitously towards zero around the point where $J$ reaches $P/D$.
+This is because this extreme of the curve represents a circumstance where the propeller is rotating, but not actually moving any water, since it is moving through the water at about that speed which corresponds to the exact helical path defined by the pitch being traced.
+This represents circumstances where a propeller encounters effectively zero load.
+This might be the case when a eases it's power, and the momentum of the ship temporarily keeps the ship moving without the propeller exerting much or any net thrust, or where alternative propoulsion sources (sails, or other propellers) are keeping a speed higher than what the propeller actually has push backwards on water to maintain.
+Since drag due to friction scales with velocity squared, very low speed do correspond to increasing advance ratios for a given propeller size, as would increases in propeller size for a given speed and pitch ratio.
+In practice however, under self-propuslion conditions for open propellers, slowing down towards does not elevate the advance ratio far enough past beyond peak efficiencies for overall efficiency to fall very far, as can be seen in Figure \ref{fig:EfficiencyBulten.png}
+
+A closely related concept here is the slip ratio, which is a measure of the difference between the pitch (the distance the blade section would travel if it traced a perfect helix for one rotation) and the actual distance that the blade did advance through the water.
+This is neatly defined and shown in a diagram in \ref{fig:SlipRatio.png}.
+
+\begin{figure}[h]
+\captionsetup{width=0.8\textwidth}
+\includegraphics[width=0.8\textwidth, center]{SlipRatio.png}
+\caption{The slip ratio of a propeller. Image credit: \cite{MAN2017}}
+\label{fig:SlipRatio.png}
+\end{figure}
+
+It can be seen from the diagram of the open-water efficiencies that there is some desirable non-zero slip ratio for a propeller.
+The other end of the spectrum, where efficiencies also fall off toward zero, represent circumstances where the propeller is simply spinning in the same place in the water.
+In this case the slip ratio is one.
+This represents a circumstance known as 'bollard pull' where a vessel is simply exerting force against an immovable object or infinite load.
+Whilst the very low efficiencies here could represent a case of a propeller that is simply too small for the task it is trying to achieve, it is actually a significant and important circumstance to consider for propellers on working vessels such as tugs or trawlers, which exert large forces at very low water-speeds.
+Measurement of the thrust generated in bollard pull, and the power required to achieve this thrust, is still an important metric for such vessels, but the definition of 'useful work' for normal propulsion requieres the thust to be excerted over some distance travelled.
+
+\subsection{Ducted Propellers}
+A ducted propeller is a propeller in any kind of tubular shroud, casing or nozzle.
+Ducted propellers fall into two categories, one of which is referred to frequently and interchangeably as a pump-jet.
+The two types are called the accelerating type (also known as a Kort Nozzle, after a company that made them popular in the 20th Century) and the decelerating duct.
+The names describe the different effects that each type of nozzle have on the flow speed of the water at the point where it meets the propeller (which is also called an impeller, particularly when associated with decelerating ducts, or pumpjets).
+
+A detailed description of the essential differences between the two can be easily demonstrated by diagram, which show a simple comparison an acelerating and decelerating duct, which have the shape of the aerofoil section which comprises the duct inverted.
+\footnote{It should be noted that some designs attempt to strike a balance between efficiency and cavitation performance of accelerating and decelerating ducts for certain purposes, and in such cases the duct shape might not be so clearly contrasted \parencite{abdel2010}.
+But for the purposes of explanation, as well as optimisation for military purposes, the distinction between the two will generally be quite clear, and in practice ducts will still tend to decelerate the flow on approach to the impeller unless specifically shaped to narrow in the opening.}
+
+\begin{figure}[h]
+\captionsetup{width=\textwidth}
+\includegraphics[width=\textwidth]{AccelDecel.png}
+\caption{The two types of ducted propeller, accelerating (left) and decelerating (right) as shown in \parencite[249]{mollard2011}}
+\label{fig:AccelDecel.png}
+\end{figure}
+
+A consideration of what happens to the flow as it moves through each type of duct helps to demonstrate how the acceleration and deceleration effects are achieved.
+In the case of an accelerating nozzle, the area through which the flow must pass contracts prior to the water arriving at the impeller.
+In the decellerating case, the available area contracts after the water has passed the impeller.
+In some cases, a degree of diffusion is actually incorporated ahead of the impeller, whereby the area actually of the flow actually expands ahead of the impeller, which slows down the flow and increases its pressure according to Benoulli's principal \parencite{wislicenus1973}.
+
+\begin{figure}[h]
+\captionsetup{width=0.7\textwidth}
+\includegraphics[width=0.7\textwidth, center]{DuctStreamlines.png}
+\caption{The streamlines of flow through different duct types \parencite[214]{lewis1988}}
+\label{fig:DuctStreamlines.png}
+\end{figure}
+
+\subsubsection{The Accelerating Duct}
+Accelerating ducts are most commonly used by vessels which have to operate at high loadings at very low speed, such as tugs, barges, or trawlers.
+In those circumstances, relatively small vessels need to exert considerable forces against substantial loads, at very low water-speeds.
+Also, in these circumstances, the practical size of a propeller might be constrained, since operating an optimally sized propeller might increase the complexity of machinery required to deliver the power to so low a hub in the water, and it might also be difficult to arrange for the thrust to be directed substantially against the centre of drag of the vessel.
+
+The effect of the nozzle in these circumstances actually increase the optimal efficiency of the propeller, when operating at very low water-speeds but high thrusts (low advance ratio), since the pressure reduction on the inside of the nozzle induced by the movement of the water through the nozzle actually produces a net thrust forward, since the interior face at the opening of the nozzle is necessarily angled forward.
+This has the effect of distributing the suction force across a larger area, which now includes the interior of the nozzle.
+An equivalent way of describing the same effect is that the nozzle causes the water to be accelerated more gradually, over a longer distance, by inducing it to move more quickly as it passes through to the entrance.
+As a broad generalisation, more gradual or gentle actions tend to involve lower overall losses to turbulence, and hence the efficiency of the propeller can be increased.
+
+A further means by which the efficiency of a propeller can be increased is through the reduction of the turbulence created by the blade-tips passing through the open (static) water at high speed.
+Since the two sides of the blades have low and high pressure on them respectively, at the tips where the two faces meet, there there is a tendancy for water to 'spill' from the high pressure side the the low pressure side, creating a vortex which leads to unwanted turbulence.
+It is also around this vortex that blade tip cavitation tends to occur.
+By limiting the movement of water around the outside of the blade tip by having a tightly fitted shroud, losses due to this particulary type of tubulence, and the onset of this type of cavitation, or the vortices that occur even when cavitation does not, can be diminished.
+
+However, it is important to note that the existence of an accelerating duct by no means eliminates all cavitation.
+In fact, given that the static pressure at the impeller is necessarily reduced by an accleratint duct, the propensity for cavitation other surface of the propeller is generally increased.
+In many circumstance cavitation can and does still occur in ducted propellers near the blade tips at design loads \parencite{moulijn2015}.
+
+\begin{figure}[h]
+\captionsetup{width=0.7\textwidth}
+\includegraphics[width=0.7\textwidth, center]{CavitatingDuct.png}
+\caption{An example of the cavitation occurring at the blade tips of a ducted propeller. Image credit \cite{moulijn2015}}
+\label{fig:CavitatingDuct.png}
+\end{figure}
+
+A detailed explanation of the means by which the ideal efficiency of propeller can be improved can be found in \cite[213-222]{lewis1988}, as well as other texts on marine propulsion, such as \cite{carlton2007}.
+A substantial work exploring the efficiencies of a range of accelerating and decelerating ducts can also be found in \cite{oosterveld1970}.
+A common feature of these extensive studies also show that for ducted propellers, including accelerating ducted propellers, the advantage in terms of efficiency is overall restricted to low waterspeeds.
+At higher waterspeeds, the drag induced by the water-movement over the exterior of the duct begins to increase substantially, and the net thrust on the duct becomes negative.
+When the exterior of the duct experiences relatively little water velocity, but the interior experiences a great deal more, the greatest advantages of an accelerating duct are realised.
+Furthermore, in general these models do not incorporate the effect of flow separation, which is acknowledged as a possibility on both the exterior and interior of ducts, which can dramatically reduce efficiency performance (\cite[20]{oosterveld1970}, \cite[214]{lewis1988}).
+
+\begin{figure}[h]
+\captionsetup{width=0.7\textwidth}
+\includegraphics[width=0.7\textwidth, center]{DuctedEfficiency.png}
+\caption{Accelerating ducts produce a net positive thrust at low advance ratios, but a net negative thrust at higher advance ratios as the drag on the nozzle exterior increases.
+Image credit:Open-water test results of Ka 3-65 screw series with nozzle no.
+19A \parencite{oosterveld1970}}
+\label{fig:DuctedEfficiency.png}
+\end{figure}
+
+\subsubsection{The Decelerating Duct or Pumpjet}
+Decelerating ducts necessarily produce a negative thrust, since the process of raising the pressure of the water about the impeller (equivalent to slowing it down) necessitate reducing its kinetic energy and consequently momentum.
+In a wide range of literature introducing the fundamentals of ducted propellers and water-jets, or discussing their more modern development there is universal acknowledgement that the choice of a decelerating duct is generally made for the avoidance of cavitation for specialised (generally military) purposes, rather than for the achievement of increased efficiency (\cite{carlton2007}, \cite{oosterveld1970}, \cite{mollard2011}, \cite{haimov2010}, \cite{lewis1988}).
+
+In order to achieve a reduction in cavitation, however, care must be taken in the design of a decelerating duct.
+Since the negative thrust produced by the duct must necessitate additional work being done by the impeller, there can be circumstances where the additional loading of the impeller out-weighs the impacts of elevating the pressure around it.
+\cite[24-25]{oosterveld1970} provides a more thorough derivation of the circumstances in which this can be reliably achieved, which generally include larger blade area ratios or with more rows of rotors for higher loading, or circumstances where the loading of the propeller is quite low.
+This creates an inevitable and necessary tension between optimisation of a duct for propulsive efficiency and reduction of cavitation, which will be returned to in Section \ref{waterjets}.
+
+It is perhaps most relevant at this point to further clarify the distinction between the different names for related or similar systems.
+In \cite[17]{carlton2007} pump jet is described as an "interesting development of the classical ducted propeller".
+Oosterveld, in his thesis on ducted propellers also states with regard to the decelerating duct "This ducted propeller system is the so called pumpjet" \parencite[8]{oosterveld1970}.
+Other authors give separate treatement to ducted propellers and pump jets, though the systems they describe are essentially similar in terms of composition, advantages, and physics \parencite[288]{lewis1988}.
+
+\begin{figure}[h]
+\captionsetup{width=0.4\textwidth}
+\includegraphics[width=0.4\textwidth, center]{PumpjetOutline.png}
+\caption{In \cite{carlton2007} the pump jet is described as a development of a classical ducted propeller, with the noticeable addition of a stator}
+\label{fig:PumpjetOutline.png}
+\end{figure}
+
+In other definitive literature on the design of pumpjets \parencite{henderson1964} the difference between a pumpjet and a propeller is a described as being that "the stream of flow through the pumpjet is made to depart from the "natural" or free-stream sufrace that bounds the flow though a standard propeller."  According to the diagrams and the subsequent description it is clear that the definition embraces quite exactly the distinction between an accelerating duct, and a decelerating duct, namely that the flow is slowed at the point it encounters the impeller, where as in a free situation the free stream would contract and accelerate at this point.
+
+The addition of some new parts such as stator blades (to straighten out the flow) and other possible features which make the propeller system more like pump or turbomachinery components might be associated with systems more commonly referred to as pumpjets than ducted propellers.
+Not all decelerating ducts have stators, but the impact of the duct on the stream will remain.
+The key concepts behind the operation of these systems, however, remains essentially similar in their nature and design intent.
+In the presence of a duct propellers might be advantageously shaped differently, including maintaing a larger chord lenght right to the shroud, rather than tapering as open propellers tend to lessen vorexs and cavitation on the extremities.
+However, there is not to my knowledge any single defining physical characteristic which the transition to the name 'impeller' implies, which woud not be effectively equivalent to specialised propellers for decelerating ducts.
+In systems tending to be described as pumpjets or waterjets hubs tend to be larger, blade-numbers higher, twist, skew and rake lower.
+No definitive transition point seems to be accepted as having great significance.
+Most commonly literature will refer to a 'rotor', a more general term which could encompass both propellers and impellers.
+
+\subsubsection{Waterjets} \label{waterjets}
+
+The use of the word waterjet, as opposed to pumpjet, has a tendency signify that the jet expelled from the shroud or pump is situated above the water, which is desirable in very fast-moving surface vessels which attempt to minimise their contact with the waters surface, including high-speed planing hull vessels, including pleasure craft.
+The physics of the two systems, however, are closely related, with the very noticeable exception of the absence of flow considerations for water on the exterior of the duct considerations for waterjets.
+However, the physics relating to the impeller and stator blades or (turbomachinery) designed to add energy to the flow are generally extremely closely linked, though a larger variety of diffeent pump types can be chosen for waterjets which might differ more significantly from the natural evolution of a propeller to an axial-flow pump.
+Many of the essential equations which govern the efficiencies and cavitation performance of pumpjets and waterjets are of the same or extremely similar form, as can be seen in detailed discussions given to both, by similar authors (\cite{wislicenus1973}, \cite{henderson1964}).
+There are, however, plenty of instances in recent literature where this distinction regarding the ejection of the jet above and below the waterline is not maintained, and waterjets and pumpjets seem to be used interchangeably \parencite{abdel2010}, or systems are described as 'submerged waterjets' \parencite{buckingham2008submarine}.
+
+\begin{figure}[h]
+
+\begin{subfigure}{0.5\textwidth}
+\includegraphics[width=0.9\linewidth]{Pumpjet.png}
+\caption{}
+\end{subfigure}
+\begin{subfigure}{0.5\textwidth}
+\includegraphics[width=0.9\linewidth]{Waterjet.png}
+\caption{}
+\end{subfigure}
+
+\caption{A comparison of a pumpjet (a) and waterjet (b) as given in \cite{wislicenus1973}}
+\label{fig:PumpjetWaterjet.png}
+\end{figure}
+
+\newpage  \section{Constraints on the efficiency of pumpjets at low speed}
+
+\subsection{Recent Commentary}
+There has been already some attention given to the probable efficiencies of pump jets at low speed.
+Recently Andrew Davies published a piece in the The Strategist (ASPI) arguing that the pump jet may actually save fuel at transit speeds (around 8kt) and overall improve the endurance of the submarine, despite likely having lower efficiency at low speeds such as patrol speed \parencite{davies2017}.
+This argument relies on pump-jet efficiency still overtaking that of propellers at what is considered a relatively low speed for many applications.
+As Andrew acknowledges, further quantification of the likely efficiency is required, as the particular chart which he cites is not quantified.
+This section aims to identify as far as is reasonably possible from public literature how the efficiency of a likely pump-jet for a submarine might perform in quantitative terms, relative to a propeller.
+
+There has also been questions put in parliamentary hearings regarding the pump jet \parencite{patrick2017}, and responses from public officials to such questions indicate that some of the literature provided to support such lines of questions was dated, and that the state of the art may have advanced considerably since this time.
+Whilst the statement that the state of the art is much advanced is obviously true in the case with marine propulsion as with most sciences, it is also equally true that new advances tend to confirm, rather than annul, many of the works that have been previously undertaken.
+To the extent that modern literature still supports the overall constraints and bounds provided by some of the foundational earlier works, a great deal of useful information can still be obtained from earlier works.
+In particular, the foundational theories, often derived from well-established principals of conservation of momentum, energy, as applied by the likes of Bernoulli and Euler, are as well respected and deeply depended upon today as they were centuries earlier when they were first arrived at.
+It is noteworthy that some of oldest works, such as those by Wislicenus and Henderson, are still amongst the most frequently cited in the most recent of papers.
+
+\subsection{Modern literature and results}
+Abundant modern works, including those which involve experimental validation of theoretical results also clearly confirm that restrictions on the efficiency of waterjets are enduring, and give us a clear starting-point for consideration of their probable orders of magnitude, and likely transition points indicating at what speeds pumpjets might become advantageously more efficient.
+It should be noted however, that relatively little literature focuses on the performance of propulsive systems far far outside of their intended design speeds, which tends to be far higher than the patrol and transit speeds of diesel electric submarines.
+Fewer still tend to express their work in terms of ship-speed, instead adopting the convetion of using dimensionless ratios so that work done on one system can be readily translated into different combinations loads, pressures, speeds and sizes.
+However, from a few notable studies exist which provide important reference points.
+(Further interpretation of the what we can lean from the trends expressed in dimensionless coefficients will be discussed later.)
+
+A 1995 Japanese experimental study on the performance of a waterjet demonstrated a close experimental fit to a theoretical relationship which sees the propulsive efficiency of the waterjet fall to zero with craft velocity, with significant declines in efficiency commencing just under 10m/s, or somewhere around 18kt \parencite{fujisawa1995}.
+
+\begin{figure}[h]
+\captionsetup{width=0.7\textwidth}
+\includegraphics[width=0.7\textwidth, center]{EfficiencyJapan.png}
+\caption{A graph from a Japanese study of waterjet efficiency shows efficiency rapidly declining below craft velocities of 10m/s, in accordance with theory. \parencite{fujisawa1995}}
+\label{fig:EfficiencyJapan.png}
+\end{figure}
+
+In 2015 an Australian team including staff from the Australian Maritime College and Incat published results of an experimental study comparing the use of waterjets and propellers for medium-speed ferries operating in the 20-30kt speed range \parencite{kamal2015}.
+This study explicitly set out to establish the cross-over point for where propellers become more efficient than jets, since Incat was interested in pioneering highly efficient medium-speed vessels, after having considerable success in very high speed catamarans.
+It is again acknowledged as given that at lower speeds propellers are more efficient than waterjets.
+The study found that propellers were considerably more efficient over the entire speed range tested, often by a factor of two in terms of transport efficiency.
+
+\begin{figure}[h]
+\captionsetup{width=0.6\textwidth}
+\includegraphics[width=0.6\textwidth, center]{EfficiencyTransport.png}
+\caption{An Australian experimental study aimed at determining the crossover point for the efficiency of jets and propellers in the 20-30kt range revealed much higher power was required across the entire speed range \parencite{kamal2015}}
+\label{fig:EfficiencyTransport.png}
+\end{figure}
+
+Both of these previous studies confirm the general and widespread assumption that jets are known to be more efficient at high speeds, and propellers at lower speeds.
+They also suggest that differences in efficiency could be very large, of a factor of two or more, including at speeds well over a submarine's transit speed.
+However, care must be taken making precise quantitative comparisons, since such experiements involved jets which expelled into air (waterjets) as opposed to remaining under water (pumpjets) do have at least one significant difference in their makeup, though their internal pump mechanisms may often be similar.
+Specifically, pumpjets would experience drag on the exterior of the shroud/duct, which waterjets would not, however waterjets are likely to have longer intake ducts, which might involve larger duct losses than pumpjets.
+Given that these factors might affect efficiency, and the overall design of the nozzle considerably, more direct comparison with fully submerged jets should be sought to make more direct comparisons.
+
+At IMARESTS 10th International Naval Engineering Conference, a paper was presented by BMT Defence and Rolls Royce \parencite{giles2010}, outlining a proposal for 'fully submerged waterjets' to be used on future surface ships in order to achieve better acoustic performance, particulalry for an anti-submarine warfare role.
+The jets comprise a mixed-flow pump coupled to an electric motor, and have been both extensively modelled and also tesed in demonstrator vessels.
+These may be a closer representation of the performance characteristics achievable by submerged pumpjets, since shape of the intake and shroud more closely resembles the shape of a pumpjet that might be found on a submarine.
+Importantly, this work focuses specifically on the performance of the jets as a function of ship-speed, including identifying possible crossover points.
+
+\begin{figure}[h]
+\captionsetup{width=\textwidth}
+\includegraphics[width=\textwidth]{BMTShape.png}
+\caption{The 'fully submerged waterjet' proposed by BMT and Rolls Royce for military surface craft may more closely represent the duct shape used on submarines than waterjets that release above the surface.
+\parencite{giles2010}}
+\label{fig:BMTShape.png}
+\end{figure}
+
+\begin{figure}
+
+\begin{subfigure}[b]{0.9\textwidth}
+\includegraphics[width=0.8\linewidth, center]{BMTEfficiency1.png}
+\end{subfigure}
+\begin{subfigure}[b]{0.9\textwidth}
+\includegraphics[width=0.8\linewidth, center]{BMTEfficiency2.png}
+\end{subfigure}
+
+\caption{Studies on submerged waterjets for military surface craft currently show that the efficiency of jets decline sharply below 10kt \parencite{giles2010}}
+\label{fig:BMTEfficiency.png}
+\end{figure}
+
+The paper makes no effort to justify the selection of choice of jets on the efficiency of the propulsion system, instead focussing on improved acoustic performance, and potential machinery space-saving advantages.
+The vessel under consideration would require 250 tonnes more fuel with jets find than with propellers, the jets would require more power to reach equal waterspeed to a conventionally propelled craft right up to 30 kt.
+In the case of this analysis, modifications to the hullform in order to accommodate the jet (which are likely to be more substantial for a surface ship than a submarine) also have an impcat on the resistance offerd by the hull.
+This also has an interaction with the different shape of the propulsors, which alters the required level of delivered (effective) power to achieve a given speed, in addition to the considerations of how efficiently the propulsor generates that effective power.
+The paper states: "This is an encouraging result for a first iteration as the AWJ-21™ arrangement has not been fully optimised and, hence, there is further prospect of reducing the shaft power for this form.
+If the power could be reduced by 5\% (e.g.
+by fur ther improving the waterjet/hull fairing), then the AWJ-21™ vessel would be better at all speeds above 25 knots."  This statement, however, is made in reference to the hullform needed to accommodate the jet, rather than the jet propulsion efficiency itself \parencite{giles2010}.
+As the propulsive efficiency curves given in the study show, the propulsors themselves are considerably further from reaching parity, still separated by about 10\% at 25kt.
+
+The paper also provides specific curves for the efficiency of the propulsor as a function of vessel speed, which demonstrated that the waterjet efficiency was lower than that of conventional propellers throughout the speed range, but declined particularly precipitously below speeds of 10kt.
+At 5kt, the propeller was more than twice as efficient as the waterjet.
+This study has been used to directly inform the 'central' efficiency curves assessed in this analysis.
+
+It appears from recent examples in the literature that there is no evidence of jets approaching the same efficiency as propellers at speeds as low as 10kt.
+Indeed, there seems to be considerable evidence that jets of most kinds suffer substantial efficiency penalties relative to propellers in this regime, and that this is understood and accepted as an inevitable design trade-off to be made.
+In the case of a submarine, which might spend nearly 100\% of its time travelling at 10kt or less, this disadvantage (as opposed to less than 35\%, as assumed for a surface ship in \cite{giles2010}) would become extremely substantial, which will be quantified in later sections.
+
+Moreover, recent literature seems to confirm that efficiency of pumpjets fall off precipitously towards zero the closer that the vessel speed comes to zero.
+This strongly suggests the existence of some fundamental constraint in the physics of the operation of jets, which recent advances in technology seems not to have overcome.
+To more firmly establish the extent to which adaptation or plausible improvments in technology could or could not change this trend, it is worthwhile considering the theoretical limits that are presented by the physics.
+This should further limit the range of error that might be made in estimating realistic bounds of performance for a system which we do not have precise design specifications for.
+
+\subsection{A simple theoretical explanation}
+
+Pumpjets are necessarily low in efficiency at very low speeds because they attempt to get the water to do multiple things.
+They don't just accelerate a water colum.
+They first slow it down (increasing its pressure), then add pressure to it, then accelerate it.
+When the total amount of energy embodied in the incoming water stream, and the total amount of energy you want to add to the water both get very small (as they do at very low speed) there simply isn't enough energy to do all things at once to all of the water.
+The energy penalty inflicted by using the additional devices becomes very large in proportion, and part of the water doesn't do all the necessary things for the jet to work.
+In essence, if you try to slow down something that's already going slowly, it tends to stop  That's exactly what happens in jets operating at very low flow-rates, when flow separation, also known as stall when it occurs between the blades, begins to take place.
+When there isn't enough energy to move all the water through the pressure-hurdles imposed by the duct, only some of it moves, and the rest spins in place.
+
+There are other some fundamental limitations on the peak efficiency that can be achieved by jets based on simple calculations of ideal efficiency from momentum theory, which are discussed in Section \ref{tradeoffs}.
+They also provide enduring, reliable limitations on the extent to which ingenuity or design changes might be able to improve the efficiency of jets at low speed.
+But the primary reason why jet efficiencies decline towards zero at low speed is to do with flow separation \parencite{fujisawa1995}, which is what occurs when the assumptions inherent in momentum theory's analysis of an 'ideal' actuator disc start to break down.
+
+The conditions which lead to flow separation in a pumpjet can be relatively simply explained in terms of conservation of energy, and conservation of mass.
+Conservation of mass requires that as a pipe or duct expands, the flow through it necessarily slows down, in order to occupy the large volume, in proportion to the new cross-sectional area.
+Conservation of energy requires that it increases in pressure, and defines a very specific relationship between the pressure increase and the change in velocity, which is determined by the shape of the duct.
+That relationship says that the static pressure increase is inversely proportional change in velocity \textbf{squared}, as given by Bernoulli's equation, which was discussed in earlier sections.
+As a result, at high speeds, the total energy of the system is dominated by its velocity, but at very low speeds the static pressure is much more significant.
+
+It also means that for any given change in velocity, the static pressure changes also, but doesn't change as much as the velocity change.
+The consequences for this pumpjets are and inevitable, since the geometry of the duct is inevitably fixed.
+At very high flow rates, the expansion of the area ahead of the impeller creates an adverse pressure gradient, as the water must slow down to fill the greater area.
+This adverse pressure gradient is larger at higher speeds, the proportions aren't linear, hence the the speed is much much larger, due to the square relationship, and flow is consistent (though still turbulent) throughout the entire cross-sectional area of the duct.
+
+At low speeds, the same effect leads to detrimental effects.
+The adverse pressure gradient is much reduced, but the velocity is much much $(much)^2$ reduced.
+The pressure gradient is necessarily constant across the entire area radial area of the duct, but the drag due to the boundary with the duct itself will necessarily be greater closer to the walls in the case of the intake of the duct.
+For this simple reason, as the flow rate decreases, some point is reached where the flow adjacent to the walls actually reverses, and the flow becomes separated.
+
+The same phenomenon is also prone to occur near roots of the blades of the rotor \parencite[15]{henderson1964}.
+Necessarily, the blades of the rotor are doing work on the fluid, so unless the impeller is in a region of considerable contraction, pressure will necessarily have to increase significantly through the blades, providing an adverse pressure gradient, which again must be constant radially.
+Since closer to the hub the velocity of the blades through the water is lower, the blades tend often to also be longer and deflect the flow even more in order to do the same amount of work on the fluid.
+With the longer surface over which the fluid must travel, as well as greater angle of deflection, flow separation is prone to occur here too.
+The physics will be exactly analogous to that of the separation on the inside of the duct intake, except the pressure gradient will be provided by a different source (the rotor doing work), and the surface (the blade roots) will also be different.
+This will also mean that the direction of rotation of the vortices will be different, and the location also close to the middle of the jet, instead of the outside.
+
+As can be seen in Figure \ref{fig:JetHighFlow.png}, the duct of a jet is designed to operate by deflecting the flow away from the natural path that would occur for a propeller, which would accelerate the stream over its blades directly, with the flow necessarily contracting with the acceleration as shown in Figure \ref{fig:PropellerAction.png}.
+Instead, pumpjets tend to use some moderate expansion of the flow in advance of the impeller (diffusion) in order to raise its pressure, and reduce its velocity.
+(Allowing the blades to do work on the water at lower speeds and higher pressure is the essence of what reduces cavitation.)  The blades then do their work on the liquid, further raising its pressure.
+The duct allows this pressure then to be converted back to kinetic energy, as the water accelerates to move through the contracting area in the nozzle.
+This difference in the shape of the flow, and the consequent impacts on the static pressures, are demonstrated in Figure \ref{fig:FlowComparison.png} taken from \cite{mccormick1963design}.
+
+\begin{figure}
+\captionsetup{width=0.4\textwidth}
+\includegraphics[width=0.4\textwidth, center]{FlowComparison.png}
+\caption{The effect of the duct is to deflect the flow away from the streamlines of an open propeller, elevating the static pressure at the stator blades by slowing the flow down.}
+\label{fig:FlowComparison.png}
+\end{figure}
+
+Consequently, the geometry of the duct defines a pressure gradient in the intake and through the nozzle.
+The greater the difference in the cross-sectional areas, the greater the total magintude of the pressure differential that will occur, as necessitated by conservation of mass.
+The same amount must go out as came in.
+As the area through which it passes changes, the velocity much change so that the flow rate remains constant, otherwise cavitation will occur.
+The rate at which the areas change (how rapidly they expand and contract along the axis) will determine the gradient.
+Making the pressure change as smooth as possible is an important aspect to jet design in order to minimise flow separation.
+
+\begin{figure}[h]
+\captionsetup{width=1\textwidth}
+\includegraphics[width=1\textwidth]{JetHighFlow.png}
+\caption{At high flow rates the water has sufficient velocity (indicated by arrow length) to overcome the adverse pressure gradients created by the duct shape, shown by the change from green to red.
+After the impeller has added further pressure to the water, the water accelerates out of the contracting nozzle through the favourable pressure gradient (red to green) to produce a jet, reaching equal pressure to outside at the exit}
+\label{fig:JetHighFlow.png}
+\end{figure}
+
+Figure \ref{fig:JetLowFlow.png} shows a jet operating at very low speed.
+The adverse pressure gradient combined with the drag forces of the boundary layer are sufficient to bring the flow to a stop and reverse it, as shown by the arrow size reducing, and changing direction in the diagram.
+This occurs very close to the inner surface near the opening of the duct, and potentially also between the blades of the rotor, beginning the roots \parencite[15]{henderson1964}.
+Flow separation necessarily sees additional energy expended in the kinetic energy of the additional vortices and turbulent flows created, which don't result in the creation of thrust, which demands a change in the net momentum of the liquid in a backwards direction.
+Consequently, the degree and extent that it occurs necessarily and inevitably reduces the efficiency of the propulsor.
+As the flow rates slow further, the areas of flow separation will become larger, effectively reducing the intake area of the jet, and further degrading efficiency.
+
+Flow separation also tends to lead to unsteady flows, with vortices periodically being shed into the main stream as they are speed up or change shape, and then ingested by the impeller and stator blades.
+As can be seen by the diagram in Figure \ref{fig:JetLowFlow.png}, regions abound which are adjacent to or behind flow blockages or reversing flow, where considerable turbulence and unsteady forces would exist.
+These effects also have acoustic consequences, which will be discussed in Section \ref{acousticconsiderations}.
+
+\begin{figure}[h]
+\captionsetup{width=1\textwidth}
+\includegraphics[width=1\textwidth]{JetLowFlow.png}
+\caption{At low flow rates the water has insufficient velocity to overcome the adverse pressure gradients created by the duct shape, and flow separation begins to occur on the duct intake, and between the blades beginning near the roots (stall).
+Flow through the jet would be unsteady, with excessive energy expended creating vortices and turbulence that results in little thrust.}
+\label{fig:JetLowFlow.png}
+\end{figure}
+
+It is worthwhile noting that the diagrams in Figures \ref{fig:JetHighFlow.png} and \ref{fig:JetLowFlow.png} do not include a hub for the sake of simplicity.
+In practice, the cross-sectional are of the pumpjet naturally only includes that aread between the hub and the duct, and in the case of many pumpjets, particularly those on the ends of torpedoes or submarines, the hub may include/incorporate the tapering tail of the body, for example in Figures \ref{fig:PumpjetOutline.png} or \ref{fig:PumpjetWaterjet.png}.
+Particularly in Figure \ref{fig:PumpjetWaterjet.png}, the appearance of the inner side of the shroud to roughly follow contraction of the hub or body gives the optical appearance of very little diffusion or contraction of the flow occurring, as the one dimensional separation might seem to be roughly constant, or even expanding in places where I have earlier explained it generally contracts.
+It is important to consider the impact of the second dimension on the actual effective area.
+Since area of a disc follows $\pi r^2$, the impact of the contraction of the hub will be dramatically larger when it is large than it will be when it is small.
+
+Consequently, the contraction of the outer shroud at the very end of the duct will dominate the final taper of the hub, and actually lead to substantial restriction of the flow.
+Similarly, where the body or hub is quite substantial and also tapering steadily, keeping the duct intake parallel to the axis of the body for a small distance, or even just tapering it proportionally less than the hub, can still result in a considerable expansion in the area, and considerable diffusion.
+The example in Figure \ref{fig:PumpjetOutline.png} shows a case of extremely sharp flow diffusion on the intake, probably more than what is practially realistics to avoid flow separation, even at quite high speeds.
+It is important to recognise that this simple relationship with $r^2$ generally means that even moderate changes in terms of diameter, which is what we tend to observe when looking at images and diagrams can actually amount to much more substantial changes in velocity, and hence pressure, which drives the effects described in this section.
+A side-by-side comparison of the geometries and flow-lines likely in the case of pumpjet or propeller is give in Figure \ref{fig:Comparison.png}, from  to help illustrate this point.
+
+\begin{figure}[h]
+\includegraphics[width=\textwidth]{Comparison.png}
+\caption{At the end of a tapering body the impact of the contracting hub will be much larger near the intake than near the tail, making the degree of diffusion ahead of the impeller, as well as contraction of the flow in the nozzle after it, much larger in impact than it appears by looking at the external shape.
+Credit \cite{gearhart1966selection}}
+\label{fig:Comparison.png}
+\end{figure}
+
+\begin{figure}[h]
+\captionsetup{width=0.8\textwidth}
+\includegraphics[width=0.8\textwidth, center]{Triomph.jpg}
+\caption{The outer shape of pumpjets on the aft end of torpedoes or submarines, such as this one on the French Triomphant class on which the Barracuda's jet will be based, may understate the degree of deceleration that the duct achieves.
+Credit \cite{reddit2016}}
+\label{fig:Triomph.jpg}
+\end{figure}
+
+\subsection{Trade-offs and Limitations on Improvements} \label{tradeoffs}
+
+Having established that the inevitable decline of a pumpjets efficiency towards lower speeds is driven by the increasing occurrence of flow separation, some consideration ought to be given to the potential scope for the effect to be improved or negated by design changes, or otherwise mitigated with other improvements in efficiency.
+Here I have found considerable help from well established literature, many including various old declassified technical reports funded by the US Navy, many of which are frequently cited in the most recent literature as laying out the enduring frameworks for the modern discipline.
+
+It is clear that the optimisation of a system for minimising cavitation, and hence improving acoustic performance, is a separate and independent design requirement to that of optimising a system for propulsive efficiency.
+These two design requirements can be found either to be in tension, or alignment, in different regimes.
+At high speeds, where cavitation tends to have severe impacts on propulsive efficiency of open propellers, and high mass-flow (which leads to good efficiency) can be achieved through compact ducts and pumps, pumpjets offer very good performance on both metrics.
+Hence, in this regime the two design requirements tend to align relatively well in favour of pump-jets.
+However, in very low-speed, low-load environments, where cavitation effects are negligible, and larger areas need to be engaged achieve good mass-flow and efficiency, the two design requirements align relatively well in favour of an open propeller.
+In a variety of regimes (including at quite high speeds), the two requirements find themselves in tension and needing to be balanced \parencite{gearhart1966selection}.
+
+For this particular consideration, two questions are of paricular importance.
+The first is where the clear transitions in the regimes will tend to lie, with respect to the probable requirements of a conventional submarine.
+The second is what drives the tension between those design requirements in the intermediate regimes, and whether it is a fundamental or inescapble tension in the laws of physics, or simply a technical hurdle which imminent improvements in the state of the art are likely to overcome by engineering improvements or innovation.
+
+The answer to the first question, as discussed in earlier sections, seems to be that the operating speeds of a submarine tend to lie far below the probable levels where pumpjets are even remotely comparable to waterjets for efficiency.
+In my research I have found no evidence of waterjets being advocated for speeds less than 30kt, and for pumpjets for speeds less than 27kt.
+Having consulted with leading academics in the field, I've had it strongly confirmed that a pumpjet will always be less efficient than a propeller at speeds under 18kt.
+In later sections I will address further the acoustic comparisons at low speeds where flow separation will begin to occur, for instance below 10kt.
+
+The answer to the second question is that the tension between acoustic and propulsive performance at intermediate speeds and loadings is driven by foundational physical requirements.
+That is not to say that optimal means of balancing those requirements will not still be found, and improved, as the art advances.
+However, it does mean that significantly adapting using known variables that can be changed to favour one design requirement are unlikely to yeild large improvements without some trade-off being made to the other.
+
+The reason for this trade-off occurring can be quite simply attributed to the nature of the shroud, or duct, which surrounds the the rotors of the propulsor, and the effect it has on the flow.
+If the duct acts to increase the pressure at the rotor, (which reduces cavitation) it necessarily must slow the flow at/around the rotor, which necessitates produceing a negative thrust on the duct.
+The system as a whole produces a positive thrust, becuase the impeller or rotor pushes a larger positive thrust, but the duct necessarily produces a negative thrust.
+As the speeds get lower, and the thrust required gets much much lower, this negative thrust becomes proportionally more significant, as described later.
+
+Put another way, the work required to slow the water down, then speed it up again, becomes a gradually greater penalty the the slower the water is coming in, and the less you actually want to speed it up.
+If the water is slowed enough by the duct to effectively stop, then the impeller must do the work of speeding it up (sucking it into the duct) before the duct slows it down, before the impeller raises its pressure, and the pressure accelerates it out of the nozzle afterwards, immediately reducing even the ideal efficiency to 50\% \parencite[277]{lewis1988}.
+Compared to simply accelerating the water in one simple push, as a propeller blade does in an unconstrained flow, the efficiency penalty is substantial, and larger in proportion at low speeds.
+
+\subsection{The impact of duct loss on an ideal propeller}
+
+Derivations of the 'ideal' efficiencies are frequently made for propellers and jets using basic conservation laws (momentum theory) which consider simple acceleration of some column of water as actuated by a disc representing the impeller or propeller.
+A fuller description of these equations can be found in  most standard texts \parencite[131]{lewis1988}.
+These methods are computationally simple, and allow simple upper-bounds to be be set for efficiencies achievable for given designs under certain conditions, without considering the pottential impact of other more complex effects such as skin friction over the blades, potential flow separation, or cavitation.
+As such, they can be easily relied upon to constrain the overall plausible range of values that can be achieved from universally applicable inputs, such as the velocity of water going into a duct, and the velocity of water exiting a nozzle.
+
+Such methods are used in \cite[8-14]{wislicenus1973} to construct an 'ideal' jet efficiency that applies to waterjets and pumpjets.
+Whilst I won't repeat the full expansion here, a fundamentally important early conclusion of the derivation is the distinction between 'useful work' and the total energy expended.
+Since the rate of change in momentum is known to be necessarily equal to the force generated, and increasing momentum change (hence thrust) can be achieved either accelerating more mass, or accelerating mass to higher speeds.
+Given that energy embodied in a flow is proportional to it's velocity squared, but its momentum only to its velocity, a necessary and inevitable consequence is that some energy is always wasted in accelerating a fluid to a higher velocity than its surrounds.
+
+This is expressed in the equation which represents 'ideal jet efficiency', which is directly equivalent to 'ideal propeller efficiency' using momentum theory of propeller action:
+
+\begin{equation}
+\label{eq:3}
+\eta_j = \frac{1}{1+\frac{\Delta V}{2V_0}}
+\end{equation}
+
+The inevitable consequence of this relationshiop is that efficency approaches unity when $\Delta V$ (the change in water of the water between entering and exiting the system) relative to $V_0$ approaches zero.
+Intuitive embodiments of this relationship are (equivalently) that efficiency is maximum when the least force is exerted on any given bit of water, which might be when negligibly small work is being done (extreme low speed) or when such a large mass of water is acted on that it barely needs to be pushed at all.
+This relationship would lead to the conclusion that infinitely large propellers or jets are always desirable from an efficiency perspective.
+This of course isn't practical to build, and also doesn't hold when taken to the etremes the effects of the drag on the propeller blades, or duct, are taken into account.
+Infinitely large systems would increase the total weight, as well as having significant surface drag when moving through the water, all of which are neglected in this 'ideal' consideration.
+
+In \cite[8-14]{wislicenus1973}, Wislicenus expands the relationship in the case of the jet to incorporate terms to represent some of the inevitable adjustments that would occur in the case of pump jets and waterjets, particulalry the duct head loss ($K$) (fraction of energy required to push the water through the duct), as well as the additional drag on the outside of the propulsor ($\Delta T$).
+This leads to an adjusted efficiency relationship, which still neglects any losses due to pump efficiency, or change in height of the jet (which is added for consideration for waterjets, but not relevant here).
+
+\begin{equation}
+\label{eq:4}
+\eta_j = \frac{1}{1 + \frac{\Delta V}{2V_0} + K\frac{V_0}{2 \Delta V}}(1 - \frac{\Delta T}{T})
+\end{equation}
+
+The conclusion of the analysis, and the plotting of these results for certain possible values of $K$ and $\Delta T$ is: "Its outstanding characteristic is that this efficiency approaches zero rather than unity for $\Delta V / V_0 = 0$, even for small duct losses".
+
+\begin{figure}
+\captionsetup{width=0.6\textwidth}
+\includegraphics[width=0.7\textwidth, center]{PumpjetEfficiency.png}
+\caption{Ideal jet efficiencies sharply appoach zero at low values of $\Delta V / V_0$, where as ideal propeller efficiencies without a duct $K = 0$ approach unity. Credit: \cite{wislicenus1973}}
+\label{fig:PumpjetEfficiency.png}
+\end{figure}
+
+Wislicenus also acknowledges \cite{wislicenus1973} another author, Brandau, who makes a different assumption about the relationship between the duct loss and water velocity \parencite{brandau1967} in his derivation (along with another thorough discussion of potential and ideal efficiencies), who produces a different peak efficiency point, but retains a similar overall shape featuring efficiency falling very rapidly to zero at low values of $\Delta V / V_0$, and contrasting this with the ideal propeller, which approaches unity in this low regime.
+
+Charts such as these are generally used with dimensionless ratios on the axes (such as $\Delta V / V_0$) to allow the most general solutions to be used to solve for any given particular vessel size, speed, or loading requirement which might be required, these  confusing to many readers unfamiliar with hydrodynamics.
+For the avoidance of any confusion, it is important to note that the value for $\Delta V / V_0$ for a given vessl and propulor may tend to increase with increasing waterspeed of the vessel, however one cannot rely upon this particular value diminishing towards zero at very low speeds, as it does for related by different metrics.
+In an idealised circumstance, the flow rate increases directly with vessel speed.
+Given that the force produced scales with $\Delta V$, which at constant values of $\Delta V / V_0$ increase in proportion to (if, in the idealised circumstance, vessel speed is equal to $V_0$), an increase in thrust can theoretically match the $V_0^2$ requirement imposed by drag without any change in $\Delta V / V_0$.
+However, in practice the water encountered by the propulsor is slowed by the wake field, and so the drag encountered by the vessel may increase faster than velocity of the water entering the intake, an effecnt which is likely non-linear with vessel speed.
+So whilst the trend with regards to vessel speed can be known with regards to the graphs in \cite{fig:PumpjetEfficiency.png}, a direct translation of the x-axis used to vessel speed at any given point is difficult prove without additional system information.
+One might generally infer that the sorts of declines in efficiency observed as vessel speed discussed in recent literature correspond with movement towards very low values of $\Delta V / V_0$, though it is possible that alternative derivations have been used, or that the phenomenon known to occur in practice (such as flow separation) aren't adequately captured by such ideal equations derived from momentum theory in order to make a direct translation into real-world application.
+
+For context, the equations developed by Wislicenus are used for the purposes generally of selecting the ideal propulor diameter, which can be made larger in the case of a propeller to access higher ideal efficiencies.
+This is an opportunity which is constrained in the case of jets because the efficiency diminishes at very low values of $\Delta V / V_0$.
+This is quite a different constraint to other "practical" considerations as some have suggested, such as weight, or space constraints.
+Open propellers may advantageously choose to lower their values of $\Delta V / V_0$ as low as 0.1 in order to obtain ideal efficiencies \parencite{wislicenus1973}.
+This still provides a hard limit on potential improvements that can be attained for a pump-jet that has non-zero duct loss.
+
+\begin{figure}[h]
+\captionsetup{width=0.5\textwidth}
+\includegraphics[width=0.5\textwidth, center]{EfficiencyBrandau.png}
+\caption{Alternative derivations of ideal jet efficiency also show sharply declining efficiency for low values of $\Delta V / V_0$ Credit: \cite{brandau1967}}
+\label{fig:EfficiencyBrandau.png}
+\end{figure}
+
+The structure of these curves is found to be overall quite similar for both pumpjets and waterjets, though the probably values for the losses incurred in the intake duct, or on the exterior of the shroud, are obviously different \parencite[13-14]{wislicenus1973}.
+Similar efficiency shapes can be found for discussions focussing explicitly on underwater pumpjets \parencite[13]{henderson1964}.
+It should be noted that the theoretical efficiency derived by these general equations is consistent both in derivation and results with those produced by other authors \parencite[227]{lewis1988}.
+
+Care should be taken, however, to note that different graphs for efficiency can be produced which have different variables on the x axis.
+For example, \cite[247]{mollard2011} a graph of ideal efficiency of a pump jet is shown using the speed of advance divided by the jet speed, or $V_1/V_2$.
+Given that the x axis used in the previous charts used $\Delta V / V_0$, which would be related to $V_2-V_1$ in this case, this has the effect of reversing the effective direction of the chart and changing its shape.
+In this case, for a fixed jet and vessel design, high-speed operations will be towards the left of the chart, and the precipitous decline in efficiency will occur as speed declines to zero, along with thrust, toward the right of the chart, which can be known to correspond to waterspeeds declining to zero at $V_1/V_2=1$.
+
+\begin{figure}[h]
+\captionsetup{width=0.6\textwidth}
+\includegraphics[width=0.6\textwidth, center]{EfficiencyMollard.png}
+\caption{Other modern reference texts might demonstrate jet or propeller efficiencies with different x-axes, effectively reversing the way the graph should be understood for a single system operating at different speeds.
+Credit:  \cite{mollard2011}}
+\label{fig:EfficiencyMollard.png}
+\end{figure}
+
+\begin{figure}[h] \captionsetup{width=0.6\textwidth}
+\includegraphics[width=0.8\textwidth, center]{EfficiencyBulten.png}
+\caption{Another graph representing efficiency as a function of the ratio of jet velocity to incoming velocity, which shows efficiency declining at low thrust (to the right).
+The impact of increasing duct losses can be clearly seen, as opposed to the 'ideal' propeller without a duct.
+Credit: \cite{bulten2006numerical}}
+\label{fig:EfficiencyBulten.png}
+\end{figure}
+
+These types of charts correspond more closely in orientation and trend that the very frequently used x-axis unit $J = V_a/nD$ when showing efficiency curves for both propellers and ducted propellers (such as used in Figure \ref{fig:OpenWater.png}) are of a similar orientation to the chart shown in the figure above.
+The left side of the chart will correspond to the highest number of rotations for a given amount of progress through the water.
+For vessel of fixed load, and a propeller of fixed geomtry, this will correspond to a propeller facing the higest drag force due to the vessel moving at the highest speed.
+
+Readers may observe that the efficiency for all propellers also decline to zero at the other end of the spectrum, most sharply at a point where $J$ equals or exceeds the pitch ratio of the propeller.
+As discussed earlier, these correspond to cases where the propeller is slicing along its helical path without exerting any force to move water backwards, or when the slip ratio is zero.
+In this circumstance, the rotation of the propeller would not push any net water backwards, and simply act as a paddle-wheel needlessly spinning water.
+Such circumstances are unlikely to occur to a significant degree unless a vessel was decelerating, or experienced some other force to assist its propulsion.
+It is possible that at extremely low speeds a propeller could somewhat descend to the right of optimal advance ratio, when very dramatically reduced load was encounterd because of the extreme low speed, as might be suggested by the slight reduction in propulsive efficiency of the propeller below 5kt in the BMT study.
+These losses of efficiency, however, could never supercede those of a decelerating ducted propeller or jet, which has the shroud which necessarily imposes an \textbf{additional} drag to the watercolumn in excess of any other resistance faced by the rotor.
+This effect can be seen clearly even or an accelerating duct, as shown in \ref{fig:PropDuct.png}.
+
+\begin{figure}[h] \captionsetup{width=0.5\textwidth}
+\includegraphics[width=\textwidth]{PropDuct.png}
+\caption{Even the addition of an accelerating ducts substantially reduces efficiency of propellers at very low loadings, or for constant loads, low speeds.
+Maximum attainable efficiency is also reduced.
+Their benefit lies in the bulging of the curve at low advance ratios (higher loads).
+Credit: \cite[6]{willemsen2013}}
+\label{fig:PropDuct.png}
+\end{figure}
+
+It should be clear from the above analysis that the efficiency of a pumpjet falling towards zero with a vessel's waterspeed is an enduring and foundational result of basic physics.
+The addition of negative thrust on a decelerating duct, no matter how small, requires that efficiency must decline at some point with with blade loading, in contrast with an ideal propeller which rises to unity.
+This explains the overall shape given by the the mysterious unquantified chart cited by Andrew Davies \parencite{davies2017} and Access Economics \parencite{stanford2017} shown \ref{fig:PumpUp.png}, and is also consistent with the shapes plotted by more recent studies as shown above, and those presented in other modern publications studying the use of pumpjets \parencite{giles2010} and waterjets \parencite{fujisawa1995}.
+
+\begin{figure}[h]
+\captionsetup{width=0.5\textwidth}
+\includegraphics[width=0.6\textwidth, center]{PumpUp.png}
+\caption{The mysterious efficiency comparison chart which has been doing the rounds in the Australian debate probably quite reasonably reflects theoretical physical limits of 'ideal' pumpjets and propellers.
+Credit: \cite{stanford2017}}
+\label{fig:PumpUp.png}
+\end{figure}
+
+In terms of establishing plausible peaks and cross-over points, the study undertaken in \cite{giles2010} probably provides a sufficient indication as to the current state of the art in the relevant technologies.
+Whilst this effort focussed on a surface vessel rather than a submarine, the study was undertaken with the backing of military customers with support and testing facilities (the US Navy), and involved two major companies who are world leaders in maritime engineering (BMT) and waterjet manufacture (Rolls Royce).
+Even if further technical advances through design refinement were possible, as the authors alude to, the cross-over point in terms of efficiency might be brought down to 25kt.
+The claim that the crossover point lies anywhere near the transit speed of a convetional submarine (8-10kt) appears implausible.
+In discussions with leading world experts involved in the development of advanced ducted propellers (including those involved in citations) I was told that a pumpjet could not be more efficient than a propeller below 18kt, confirming this conclusion.
+
+With the optimum efficiency of a jet probably being no greater than 60\% or 70\% at its peak efficiency according to many theoretical and experimental studies \parencite{lu2016cfd}, which would quite probably lie considerably higher in terms of waterspeed even than 18kt.
+These facts, along with the inevitable decline of efficiency towards zero as waterspeed also declines, offer a suitably constrained set of plausible efficiency curves may be investigated numerically, as will be done later in later sections.
+
+\subsection{An assessment of scope for improving the efficiency of a pumpjet at low speeds}
+Before continuing to assess the impact of the likely efficiency curves of pumpjets relative to propellers, it is worthwhile considering what the known literature about the theory of pumps, jets, and turbomachinery might tell us about the potential for a radical advance which make might substantially improve the shape of the curve at very low speeds.
+Whilst the field of propulsor design is no doubt complex, and I will not attempt to be exhaustive in detailing all of the potential design considerations which might make an improvement.
+However, the pursuit of reasonably comprehensive confidence is aided by the the scale of the gap which probably needs to be closed to make pumpjets comparably efficient to propellers at the relevant speeds.
+Minor adjustments and the improvement of small or marginal effects, of which there might be many, will clearly not help in shifting the efficiency peak of a jet from over 25kt to under 10kt.
+In other words, this gap will not be closed or noticeably reduced by 'tuning'.
+Quite substantial redesign of core characteristics (like mass flow) would be required, which would necessitate a cycle of redesign and tuning virtually all other aspect of the entire jet, as illustrated by numerous guides to the design of such systems (\cite{bruce1974}, \cite{furuya1988}, \cite{mccormick1963design}, \cite{henderson1964}).
+As such, I will only consider those parameters which could plausibly be altered quite substantially, where the consideration of larger changes make the likely limits or trade-offs quite clear.
+
+\subsubsection{Increase mass-flow by widening area of intake}
+Perhaps the most obvious means by which one might attain substantial increases in efficiency is to increase substantially the mass flow rate.
+This is derived from a fundamental relationship, as expressed earlier in the representation of ideal propeller and jet efficiency (prior to the consideration of any drag or other losses) which was given in the earlier equation.
+Since excessive velocity differences result in more energy expended than momentum changed, as given by momentum theory and basic mechanics, doing work on a large body of water, rather than working harder on more of it, is generally desirable from an efficiency perspective.
+Considering what the appropriate mass-flow should be at design speeds is generally the first thing that is specified to occur in all design processes for pumpjets, and good example of which is in \cite[6-7]{henderson1964}.
+
+As noted in \cite{henderson1964}, the key constraint of this variable is that increasing the diameter of the pumpjet necessarily leads to an increase in the surface (including the external suface) of the shroud.
+The increased drag that this induces must be deducted from the thrust that the pumpjet produces, or effectively increasing $\Delta T$ as given in earlier equations.
+In this respect, there is "no free lunch" to be had in expanding the size of the intake.
+As Henderson et al.
+conclude, for this reason the diameter of a pumpjet tends to be about 15\% or 20\% smaller than that of a corresponding propeller, whereas the diameter to have equivalent mass-flow, highlighting the tension set up in pumpjet design with their earlier observation that "if the cavitation resistance of the blades is to be improved with no reduction in thrust and the same rate of flow through the propulsor, the pump jet rotor must be larger than that of the propeller." \parencite[1]{henderson1964}
+
+It should also noted that in order to substantially increase the opening area of the propeller, the difference between the velocity of movement at the blades nearer to the root (close to the hub) and the extremities of the blades increases substantially.
+In order for equivalent work to be done on all of the water distributed radially across area of the impeller, quite a different blade shape might be required, in order to provide a higher deflection of the water close to the root.
+In these areas, there is increase probability of flow separation occurring, where the velocity of movement of the water across the surface is insufficient to overcome the pressure gradient required to move over the blades.
+(In a pump, the pressure necessarily increases as work is done on it.)  Or, since there isn't enough pressure exerted on the whole of the water column to overcome the resistance imposed by the duct, not all of the water moves forward, and some starts to move backwards.
+This leads to flow separation, or stall occurring close to the blade roots, which dramatically diminishes the efficiency of the system.
+A further discussion of this effect can be found in \cite[15,27]{henderson1964},  \cite[807,8012]{bruce1974},  \cite[185]{wislicenus1973}, \cite[60]{mcbride1979} and many other related texts.
+This is a general problem that affects a range of axial-flow turbomachinery, particulalry operating in off-design conditions, and will present a general concern for the design of a new pumpjet \cite[185]{wislicenus1986}, \cite{li2013}.
+
+\begin{figure}[h]
+\captionsetup{width=0.6\textwidth}
+\includegraphics[width=0.6\textwidth, center]{stall.png}
+\caption{Flow separation, or stall, can occur when the flow over the blades of a pump slows too much, as shown in \cite{li2013}}
+\label{fig:stall.png}
+\end{figure}
+
+\begin{figure}[h]
+\captionsetup{width=0.8\textwidth}
+\includegraphics[width=0.8\textwidth, center]{backflow.png}
+\caption{Stall results in water flowing backwards near the hub of the blades at low flow rates, as shown in \cite{li2014rotating}}
+\label{fig:backflow.png}
+\end{figure}
+
+It is possible to limit the variation of deflection between the blade tips and roots by simply expanding the size of the hub.
+However, this necessarily reduces the effective frontal area that is available to receive the mass-flow, which necessitates further expansion of the outer diameter, and a very substantial increase in the weight of the system.
+In addition, expansion of the hub will necessarily increase the total surface area relative to the mass flow, including by increasing the amount of tapering required after the stators in order to prevent risk of cavitation occuring around the tail the hub.
+Consequently, increasing the area is not plausible without also increasing duct loss to some degree, consistent with the earlier observations from \cite{henderson1964}.
+
+Increasing the frontal area of the intake also is necessarily linked to the advance ratio of jet.
+Since low advance ratios correspond to higher shaft speed for a given forward velocity, increasing the thrust produced for a given rotation (by widening the jet area) necessarily results in higher advance ratios, since more work is done on the water for a given rotation.
+However, this also necessarily results in poorer cavitation performance of the jet, since higher blade velocities relative to the water flow result in necessarily increased propensity for cavitation.
+A detailed discussion of this can be found in \cite{gearhart1966selection}, but similar tensions can be readily found in most other pieces of literature, including \cite{henderson1964}, \cite{wislicenus1973}.
+
+Given the fundamental nature of the flow-rate and intake area to the design of pump-jets, and the great lenght of time for which they have been understood and discussed as a crucial design input for a waterjet, it is altogether implausible that some large increase in intake area can be achieved without incurring necessary and unavoidable tradeoffs.
+
+\subsubsection{Reduce degree of diffusion (i.e. switch to accelerating duct) to reduce negative thrust from duct}
+A further means of reducing the negative thrust produced by a duct may involve significantly reducing the degree of diffusion, or compression that occurs in advance of the rotor, since the process of diffusion necessarily slows the water and produces negative thrust.
+However, this amounts to transitioning towards an accelerating duct, or Kort Nozzle.
+Whilst nozzles of this type have been known to improve the cavitation characteristics of a propeller by enclosing the tips (which often tend to cavitate first), their ultimate function is to actually decrease the pressure at the blades of the propeller under higher loadings.
+Consequently, they necessarily increase the propensity of other types of cavitation to occur across the blade-face, including bubble, cloud and sheet cavitation.
+As such, these types of propellers are not considered for the military purposes which require an absolute minimum of noise to be created.
+Transitioning to an accelerating duct signify a shift away from a pump-jet, and necessarily be in direct tension with the need to minimise cavitation for acoustic considerations.
+
+In addition, whilst it is often demonstrated that for a given propeller under certain load conditions (high loading), the addition of an accelerating duct can considerably improve efficiency, all of these situations have generally been in the context of vessels which are designed for a relatively high thrust, low waterspeed situation, when propeller size is necessarily greatly constrained.
+This is the case fishing vessels, or for tug boats, which need to impart significant thrust to a vessel dozens or hundreds of times their own size.
+Mounting a propeller the equivalent size of the vessels they mean to push, however, would be excessively heavy, arranging a mount and power drive that would keep the entire propeller submerged, and not have the effect of flipping the boat backwards or sideways when rotated hard, would be effectively impossible, without proportionally increasing the size of the tug.
+Furthermore, in most circumstances these propellers are still designed to operate with some cavitation, even if the extent of it is reduced by the presence of the duct \parencite{haimov2010}.
+
+In other words, the use of highly loaded propellers in such applications is actually particular to design constraints that are unlikely to be present for a submarine, which can position the propeller at the tail of its body, directly behind its centre of mass and drag, and consequenlty have freedom about the optimal size of the blades.
+It is well known that a duct cannot improve the performance of a low-loaded propeller which would be the situation for a slow-moving vessel with unconstrained propeller size.
+Authoritative references in the literature include \cite{oosterveld1970}, who calculates the minimum thrust coefficient $C_T$ (a measure of loading) for an accelerating duct to improve a propellers performance to be between 1 and 2, as shown in \ref{fig:AcceleratingIdeal.png}.
+
+\begin{figure}[h] \captionsetup{width=0.5\textwidth}
+\includegraphics[width=\textwidth]{AcceleratingIdeal.png}
+\caption{Calculations of 'ideal' efficiencies of accelerating ducts (neglecting drag) in \cite{oosterveld1970} show that a duct can only improve the efficiency of a propeller if it is loaded above a certain degree, and that longer ducts produce better improvements in 'ideal' efficiencies}
+\label{fig:AcceleratingIdeal.png}
+\end{figure}
+
+A parallel design change in the switch away from the degree of deceleration in the duct would be the reduction of the nozzle contraction after the rotor.
+For waterjets and pumpjets of all sorts, based on fundamental disc actuator theory, the degree of this contraction of the flow area has also been long identified as a means to increase pressure at the rotor, and hence decrease cavitation, but at the direct expense of efficiency \parencite[14]{van1996waterjet}.
+
+Numerous efforts to find suitably compromised duct shapes exist, but even advanced modern designs still exhibit declining efficiency at low speeds, as well as some cavitation during operations at design loads, which still tend to imply relatively high speeds, for example 27kt in \ref{fig:FineOptimised.png} .
+Consideration of accelerating duct designs for a military submarine would be extremely unlikely, and be a profound departure from any existing pumpjet designs.
+
+\begin{figure}[h]
+
+\begin{subfigure}{0.5\textwidth}
+\includegraphics[width=0.9\linewidth]{FineOptimised.png}
+\caption{}
+\end{subfigure}
+\begin{subfigure}{0.5\textwidth}
+\includegraphics[width=0.9\linewidth]{FineCavitation.png}
+\caption{}
+\end{subfigure}
+
+\caption{(a) Extensive work has been done exploring compromise designs for decelerating ducted propellers (pumpjets), including with very modest deceleration in the intake, limited shroud taper, short duct lengths, or essentially incporate some of the advantages of accelerating ducts.
+(b) TAt optimally efficient conditions (corresponding to 27kt) such designs still exhibit some cavitation.
+Credit: \cite{abdel2010}}
+\label{fig:FineOptimised.png}
+\end{figure}
+
+\begin{figure}[h]
+\captionsetup{width=0.7\textwidth}
+\includegraphics[width=0.7\textwidth, center]{FineEfficiency.png}
+\caption{In experimental testing such designs still exhibit sharply falling effiency at high advance ratios, or low speed. Credit:  \cite{abdel2010}}
+\label{fig:FineEfficiency.png}
+\end{figure}
+
+\subsubsection{Reduce shroud length to decrease drag on duct}
+The reduction of the length of the shroud in order to reduce drag presents another possible means to lower the duct resistance, as shown and discussed in \cite[21-25]{oosterveld1970}.
+The fundamental difficulty with greatly shortening a duct, however, is that it requires that any changes in the shape and with of the water column be accomplished much shorter distance for the duct to achieve its desired effect, which means that generally it has to achieve less of it.
+This means that very low duct lenghts achieve less of an efficiency improvement to the ideal efficiency of a propeller, in those circumstances where they can achieve an improvement  (accelerating ducts at high loads), as discussed above.
+
+If more aggressive shape changes are attemped in a limited space, the deflections of flow and pressure gradients become much more extreme.
+This leads to a far greater propensity for flow separation to occur.
+This can happen inside the duct ahead of the stator, particulalry when any degree of diffusion occurs in order to elevate static pressure at the blades, or any kind of direction change is required as in the case of waterjets.
+Unlike waterjets, pumpjets also have the potential for separation on the exterior surface, which is an additional concern in the design of duct \cite[13]{henderson1964}.
+Given the role of the duct is to change the speed of the flow at the impeller relative to what would occur in its absence (in particular to decelerate it) separation might tend to occur on different surfaces in different operating conditions, depending on the relative pressure gradients and velocity ratios they experience at different speeds and loads.
+This substantially complicates the challenge of designing a duct that performs well across a wide speed range.
+This is illustrated for the case of a waterjet in \cite[20]{bulten2006numerical}.
+Equivalent complexities occur in pumpjets operating in low speed ranges (2-6kt, where flow separation occurs ahead of the impeller, inside the shroud.
+In the case of an accelerating duct, this phenomenon is also observed on the outside of the shroud at high advance ratios (the equivalent of low speeds and low loads) as shown and discussed in some detail in \cite{willemsen2013}.
+
+\begin{figure}[h]
+
+\begin{subfigure}{0.5\textwidth}
+\includegraphics[width=0.9\linewidth]{LowIVR.png}
+\caption{}
+\end{subfigure}
+\begin{subfigure}{0.5\textwidth}
+\includegraphics[width=0.9\linewidth]{HighIVR.png}
+\caption{}
+\end{subfigure}
+
+\caption{(a) A an intake may experience risks of flow separation even at low speeds (Inlet Velocity Ratio) on some surfaces due to certain flow deflections being more severe.
+(b) A an intake may experience risk of flow separation on different surfaces at high speeds (Inlet Velocity Ratio) where other flow deflections are more severe.
+Source: \cite{bulten2006numerical}}
+\label{fig:LowIVR.png}
+\end{figure}
+
+\begin{figure}[h]
+\includegraphics[width=\textwidth]{DuctSeparation.png}
+\caption{Flow separation occurs at on the outside of an accelerating duct at high $J$ values, which corresponds to low loads, or speeds.
+For high $J$ values, the pressure reduction on the inside of the duct, creating positive duct thrust, can be seen.
+Source: \cite[73]{willemsen2013}}
+\label{fig:DuctSeparation.png}
+\end{figure}
+
+\begin{figure}[h]
+\captionsetup{width=0.7\textwidth}
+\includegraphics[width=0.7\textwidth, center]{NozzleDrag.png}
+\caption{Dramatically reducing the length of a duct relative to its diameter ($L/D$) as shown in \cite{oosterveld1970} can reduce its drag, which has the most significant impact at low thrusts or speeds.}
+\label{fig:NozzleDrag.png}
+\end{figure}
+
+In addition, the substantial reduction in the length of the shroud, both ahead of and behind the rotor, will also have the substantial effect of diminishing any acoustic sheilding which the duct may achieve.
+
+\subsubsection{Summary remarks on potential for redesign of pumpjet for low-speed conditions}
+Overall, while the parameter-space for the alteration of marine-propulsors is large, the parameters which would allow for very dramatic rather than minor changes to propulsor efficiency at low speed are relatively few.
+The major ones include increasing the width of the intake, decreasing the length of the duct, and decreasing the pressure inrease at the blades achieved by the duct.
+In short, it involves evolving the pumpjet back towards a propeller.
+None of these changes could be effected to a substantial degree without some inevitable trade-off in terms of the cavitation performance of the system at higher speeds.
+
+\newpage  \section{Model Structures and Assumptions}
+
+\subsection{Mathematical Structure}
+
+The core of the model rests upon the assumption that drag increases with the square of the velocity of the submarine.
+This is a widely used assumption for fully submerged objects subject to turbulent flows \parencite[5]{wislicenus1973}, \parencite{davies2017}, as also discussed in earlier sections.
+Given that the amount of work done (energy consumed) by a system is equal to the distance over which it acts, which scales with speed also, we assume that the Effective Power $P_E$ (also known as Towing Power) required for propulsion to meet a given speed $v$ can be given by Equation \ref{eq:towpower}, where $C_d$ represents a drag constant.
+
+\begin{equation}
+\label{eq:towpower}
+P_E = C_dv^3
+\end{equation}
+
+The power drawn by the propulsion $P_D$ is assumed simply to be the by the Effective Power divided by propulsive efficiency at any given speed $\eta_v$ as given in Equation \ref{eq:powerdrawn}.
+Using a plausible efficiency assumption and total power required, Equation \ref{eq:powerdrawn} can be used to solve for the constant $C_d$.
+Other efficiency factors, such as the efficiency of electric motors or hull efficiency factors are assumed to be either relatively small, or essentially similar between submarines which might be conventionally powered but with a jet or propeller being the main point of difference.
+Hence the total power consumed by the submarine is given by Equation \ref{eq:totalpower}.
+
+\begin{equation}
+\label{eq:powerdrawn}
+P_D = \frac{P_E}{\eta_v}
+\end{equation}
+
+\begin{equation}
+\label{eq:totalpower}
+P_T = P_D + H
+\end{equation}
+
+With this relationship defined across the speed range, it is elementary to calculate the time that a given speed could be sustained based on a finite energy store (endurance) and the distance covered in this time (range).
+This provides the fundamental structure that underlies the model which is used to assess the dived range and enurance of conventionally powered submarines, without using air independent propulsion (AIP).
+
+\subsection{Efficiency Curve Assumptions}
+
+In order to begin to model the likely impact of propulsion system choices on system peformance characteristics such as dived range and endurance, it is necessary to adopt some plausible curves indicating likely levels of efficiency across the operating speed range, in this case assumed to be up to 20kt.
+
+In this case, three possible curves have been selected for each system, representing a high, low, and central assumption for each, with peak efficiency for the propeller at about 65\% at near 5kt, and jet efficiency around 27\% at the sames speed.
+The central assumption has been taken in both cases from the study undertaken by BMT and Rolls Royce, as in \cite{giles2010}, and indicated in \ref{fig:BMTShape.png}, since this represents the most direct comparison made as a simple function of waterspeed, with comparable technologies.
+The technologies naturally will not be identical.
+It is probable that the pumpjet of the submarine will be a purely axial-flow pump, as opposed to a mixed flow pump, but this difference in turbomachinery is unlikely to result in a overall loss of efficiency \parencite{hamilton1997}.
+The intake duct may be somewhat longer, increasing duct losses, but the degree of diffusion in the duct may be lower, reducing the adverse pressure gradient.
+It is probable that the the ship propeller would have five blades, where the submarine would likely have seven or nine.
+However, being  for a warship designed to travel at similar speeds to a nuclear submarine, with relatively unconstrained propeller size, and the pumpjet ejecting into the water, the technologies are likely to be far closer than any other publically available comparison.
+Overall, however, its considered that having an experimtally verified curve comparing very similar technologies as a direct function of waterspeed provides a far more certain basis for comparison than attempting to derive different curves \textit{ab initio} using theortical dimesnionless curves attempting to estimate how the proposed designs might correspond to them.
+
+Variations from this central line have been made with consideration for some plausible trade-offs, which are intended to be indicative, but plausible.
+In the case of the high pumpjet assumption, it is assume that a substantially larger intake area has been chosen in order to achieve higher efficiencies at lower speeds, at the cost of peak ideal efficiency, which might be realised not too far above 20kt for such a design.
+The low assumption for the pumpjet is still represents higher efficiencies at equivalent speeds than those given in \cite{fujisawa1995} for a waterjet.
+It is assumed that for both the central and low pumpjet curves, peak efficiency would not be achieved until a speed considerably above 20kt.
+All of the efficiency curves reflect a smaller decline in efficiency than the indicative assumption used by Andrew Davies, namely that efficiency might reduce by a factor of four between transit and patrol speeds, if those speeds also differed by a factor of four \cite{davies2017}.
+
+The higher assumption curve used for the propeller approaches a maximum of just under 72\%.
+This performance would not be implausible if a specialised propeller was adopted specificially to deliver peak performance at around this speed.
+Advanced flexible composite propellers, such as those used by the German submarine builder TKMS \parencite{durrant2016}, are able to achieve very high efficiencies, as well as delayed onset of cavitation, by allowing the blades to deform in certain ways to adapt to their specific degree of loading \parencite{young2007hydroelastic2}.
+Such propellers have been observed in experiment to have propulsive efficiencies above 70\% \parencite{young2007hydroelastic}.
+In this case I have assumed that this design necessitates the trade-off of steadily declining efficiencies over 10kt, considerably below the speed for which most open propellers are optimised.
+Given the ability for composite propellers to adapt their shape as loads increase, it is actually likely that the performance curve in this case is actually far broader.
+However, assuming a trade-off in this case also facilitates a scenario whereby the efficiencies of a propeller and jet actually cross over at around 18kt, which is relevant to consider, since experts in discussion have ruled out the possibility that a jet could be more efficient at lower speeds than this.
+
+The lower propeller curve shows efficiency still increasing up to around 65\% at 20kt, which might be consistent with a propeller optimised for conventional ship speeds, but probably not likely for a specialised submarine propeller.
+For all propeller curves, sharply declining efficiency has been assumed at very low speeds, (2kt or less), in order to reflect the dip seen below 5kt in the efficiency curve seen in Figure \ref{fig:BMTEfficiency.png}, assuming that this is driven by the decline in efficiency shown in Figure \ref{fig:PropDuct.png} at advance ratios approaching 1.
+This has been exaccerbated beyond realistic levels in the case of the the low propeller assumption line for the sake of testing sensitivity more dramatically to some losses in this regime.
+It is unlikely that a state-of-the-art submarine propeller would not be able to attain maximum efficiencies between 5 and 15kt, unless some other significant factor not discussed in this paper affects propellers in these regimes.
+At very low speeds, some small but sudden changes in efficiency may be possible in either direction, particularly as complex effect with multiple overlaid effects take place, for example increasing laminar flow over the propeller blades leading to lower frictional losses, but potentially greater separation.
+However, on balance, the low propeller curve is considered relatively unlikely, but useful for investigating sensitivity to errors or adjustments.
+
+\begin{figure}[h]
+\captionsetup{width=0.4\textwidth}
+\includegraphics[width=0.4\textwidth, center]{LaminarProp.png}
+\caption{At very low speeds changes in regimes such as an increase in laminar flow over the blade section could lead to small changes in efficiency, perhaps explaining the distinct rise at low speeds for the properller at ~5kt in Figure \ref{fig:BMTEfficiency.png}}
+\label{fig:LaminarProp.png}
+\end{figure}
+
+These efficiency curve assumptions are shown in Figure \ref{fig:Efficiency_assumptions}.
+
+\begin{figure}
+\begin{knitrout}
+\definecolor{shadecolor}{rgb}{0.969, 0.969, 0.969}\color{fgcolor}
+
+{\centering \includegraphics[width=\maxwidth]{figures/plots-plot_efficiency-1}
+
+}
+
+\end{knitrout}
+\caption{A range of possible efficiency curves for pumpjets and propellers to be modeled}
+\label{fig:Efficiency_assumptions}
+\end{figure}
+
+\subsection{Hotel Load Assumptions}
+It is also required that some assumptions are made about the hotel load, or the power which the submarine requires which doesn't go to propulsion.
+We might assume that this is in the order of 100kw, as suggested in \cite{martinez1995} though official sources of information are scarce.
+It is also likely that the actual hotel load varies considerably as different systems are switched on and off, some of which might be discretionary in combat scenarios (e.g.
+hot water for washing), and others (such as the combat systems) might have increased demand during crucial phases of combat operations.
+It is important to note that the combat system that has been selected for Australias FSM, the AN/BYG-1, is designed for a nuclear submarine, and is reportedly consumes considerably more power than other systems \cite{muir2015}, apparently even in the order of 100kW more.
+\cite{patrick2011}.
+Consequently, the actual hotel load experienced might vary considerably, and as such the model used in the app allows the hotel load assumption to be varied, in order to test sensitivity to this parameter.
+A flat 100kW hotel load is used as a starting assumption in the initial calculations in this paper, and some sensitivity analysis is given at half and double this power, to give an indication of the impact of plausible alternative energy demands dependent on operational circumstances, and choice of combat system.
+
+\subsection{Battery Assumptions}
+Since the purpose of the task at hand is to assess dived endurance only, this model focuses on a single energy store simulated by a total embarked mass (battery bank) and an energy density of those batteries.
+Since the SEA 1000 project has elected not to use lithium electric batteries, the energy density of the batteries can be assumed to be that of lead-acid batteries, which are known to be in the region of 30-50Wh/kg (Watt-hours per kilogram).
+Taking a central figure and converting this to Megajoules per kilogram (MJ/kg) we assume 0.14MJ/kg to be the effective energy density of lead acid batteries for our test submarine's purposes.
+This parameter can be adjusted easily for sensitivity testing, and in our case this will be adjusted to half, and double this energy density, noting that lead-acid batteries become much slower to charge as their charge level passes 80\%, and consequently it may be the case that only 50\% of their available capacity might be genuinely available for tactical purposes \parencite{briggs2016}.
+
+It is worthwhile considering that alternative types of battery may be on offer for submarines.
+In particular, lithium ion batteries were offered by TKMS in their bid for SEA 1000, and are being installed in Type 212 submarines operated by the German navy, as well as contracted for the new Norwegian submarines, as well as Israeli submarines.
+Precise energy density of these batteries is not publically available.
+For domestic uses such as handheld electronics or solar power applications, lithium ion batteries may have 3-5 times the energy density of lead-acid batteries \parencite{oconnor2017}.
+However, the actual useful density of lithium ion batteries can vary widely depending on different chemistries and designs.
+It might be assumed that for very large battery banks (such as submarines) where the stability of the battery bank is of critical importance, maximum density migth be sacrificed in favour of additional safety systems, such as increased monitoring, as well as spacing and insulation barriers between cells, in order to prevent thermal runaway in one cell leading to a catastrophic fire.
+The realistic energy density advantage that might be obtained as been suggested to lying closer to a factor of two or slightly less \parencite{greenfield2016}.
+
+The total embarked load of the batteries is not a well-known public figure, though it is likely to be several hundred tonnes in order to provide the likely desired levels of dived endurance.
+It has been suggested publically that the approximately 700 tonnes might be required for the Shortfin Barracuda \parencite{ohff2017}.
+This would appear to be quite a large fraction of the plausible payload of a submarine, even of over 4000 tonnes displacement, given that a great deal of the potential displacement will be taken by the thick steel involved in constructing a pressure vessel (probably at least a thousand tonnes per submarine) \parencite{jean2017}.
+500 tonnes of batteries has been assumed as a nominal figure for these purposes.
+Given that the mass of batteries embarked has only a direct and linear relationship with the dived range and endurance, no sensitivity testing will be undertaken on this variable.
+Proportional changes to the energy density of the batteries will have equivalent effect for the purposes of measuring the impact on performance.
+Consideration of likely impacts on space, and hull-resizing, and associated drag are beyond the scope of this paper.
+
+\subsection{Drag Coefficient of Hull}
+
+In the online version of the web-app, alternative reference points or known facts relating a given power output to a speed can be used in order to solve for this constant $C_d$.
+For instance, one might assume (as done by \cite{davies2017}) that the hotel load might match the propulsive demand at some given speed, for some given hotel load.
+However, given that the precise hotel load is generally not known and highly variable, and the potential for the different systems to have widely varying propulsive efficiencies at low speeds, this method is subject to very wide error margins.
+
+Given that any known power consumption level and speed match can be used as a reference, I have chosen instead to use the probable power consumption at top speed, both of which we have some indication of, and given that propulsive efficiencies of the two systems are likely to be much closer at this speed, there will be less error in likely comparisons between different design concepts.
+It is known that the a early German designs for a plausible submarine for SEA 1000 included a 6MW electric motor \parencite{patrick2012}, and the Shortfin Barracuda of Naval Group is often referred to as having a 7MW electric motor \parencite{ohff2017} \parencite{coates2016}, with top speeds ">20kt".
+Given the nature of increasing drag at high speeds, it seems unlikely that top speeds will be very far above 20kt.
+Hence, in this model we have assumed that 20kt will be obtained with the ouput of 6.5MW.
+
+Considerable confirmation that these assumptions are realistic can be made by assessing the endurance of the submarines under the model with other known reference points.
+For the central curves, we obtain a dived endurance betwen 60 and 100 hours for both systems at 5kt, reflecting nearly exactly the range of other public estimates of dived enurances endurances which other modern diesel-electric submarines operating only with batteries \parencite[3]{buckingham2008submarine} at around 4kt.
+Given that somewhat larger submarines might be somewhat assisted in this respect because their overall volume/surface ratio is slightly higher, this gives confirmation that the model is overall quite realistic in its magnitudes at the most important speeds.
+The sames sources also suggests that the same submarines achieve endurances of around 1-2 hours at 20kt, which also confirms that the relationship in our model linking differnt speeds is also in agreement with other models.
+
+\newpage  \section{Results and Discussion}
+
+\subsection{Power Demand}
+As part of the calculation of results for each pair of efficiency curves, the power requirements of each respective component (propulsion for the two respective systems, as well as effective power required, and hote load) can be plotted to ensure that realistic results are obtained, which is important for selecting plausible values for the constant $C_d$.
+For the central efficiency curves, we can show the resultant power curves for the the assumed values of battery load, energy density, and hotel load, as seen in Figure \ref{fig:power_curve}.
+
+\begin{figure}
+\begin{knitrout}
+\definecolor{shadecolor}{rgb}{0.969, 0.969, 0.969}\color{fgcolor}
+
+{\centering \includegraphics[width=\maxwidth]{figures/plots-plot_power1-1}
+
+}
+
+\end{knitrout}
+\caption{With central assumptions, a pumpjet powered submarine reaches 20kt with 6.5MW of delivered power.  Hotel load matches propulsion power demand around or below 5kt.}
+\label{fig:power_curve}
+\end{figure}
+
+It can be seen here that one of Andrew Davies assumptions, regarding hotel load matching propulsion power demand at patrol speeds, might match quite closely with our assumptions at a patrol speed of 4kt for the case of a pumpjet.
+It also shows the remarkable impact of the square law on power demands, particularly their effect at low speeds.
+At just a few kt, an entire submarine might be propelled by a power level comfortably produced by a large family car.
+This gives further explains why it is that efficiencies of propulsion systems aren't routinely discussed at dramatically lowered speeds.
+Even for jets which have efficiencies which decline much faster than those we've assumed (for example in \cite{fujisawa1995}), the reduction in power demanded falls substantially faster than efficiency.
+Consequently, at low speeds nothing dramatic occurs to make the declining efficiencies of consequence to operators.
+It is often still more economical for such vessels to travel very slowly, well outside the ideal performance range of their propulsion systems, simply because of the drag on the hull.
+
+Overall, this chart is consistent with expectation, and confirms that a plausible value of $C_d$ has been selected under these assumptions.
+
+\subsection{Central Results}
+The plots comparing the resultant endurance and range calculations across the speed range considered are in Figures \ref{fig:central_extreme}, \ref{fig:central_jet}, \ref{fig:central_prop} and \ref{fig:bottom_top}.
+In all cases, the comparison for the central efficiency curves have been replotted centrally, in order to highlight varience under each of the different assumption comparisons.
+
+It is clear that under the central assumptions, the declining efficiency of the pumpjet has a very significant impact on range and endurance, particularly at speeds around 4-8kt.
+It might be noted that under the central assumption, the propeller system achieves 138 hours of dived endurance at 4kt where the pumpjet only manages 91 hours dived endurance at this speed.
+Consequently, at a likely patrol speed, the difference amounts of almost exactly two days in dived endurance.
+One might presume that such a difference is of some considerable tactical significance.
+In order to match the propeller's dived endurance, the pumpjet submarine would need to slow to 2.5kt.
+
+The phenomenon manifests itself in a very substantial reduction in the overall range that can be achieved submerged, and a difference between the speed at which maximum range can be achieved.
+Under central assumptions, the propeller driven submarine can reach over 540nm at any speed between 4 and 5kt.
+The pump-jet driven submarine reaches over 360nm between 3 and 6kt.
+A difference of 180nm dived range would presumeably be of considerable tactical significance.
+
+\begin{figure}
+\begin{knitrout}
+\definecolor{shadecolor}{rgb}{0.969, 0.969, 0.969}\color{fgcolor}
+
+{\centering \includegraphics[width=\maxwidth]{figures/plots-plot_results1-1}
+
+}
+
+\end{knitrout}
+\caption{The most extreme and central assumptions are compared}
+\label{fig:central_extreme}
+\end{figure}
+
+\begin{figure}
+\begin{knitrout}
+\definecolor{shadecolor}{rgb}{0.969, 0.969, 0.969}\color{fgcolor}
+
+{\centering \includegraphics[width=\maxwidth]{figures/plots-plot_results2-1}
+
+}
+
+\end{knitrout}
+\caption{The impact of different jet efficiency assumptions compared to the central propeller curve shows wide variance}
+\label{fig:central_jet}
+\end{figure}
+
+\begin{figure}
+\begin{knitrout}
+\definecolor{shadecolor}{rgb}{0.969, 0.969, 0.969}\color{fgcolor}
+
+{\centering \includegraphics[width=\maxwidth]{figures/plots-plot_results3-1}
+
+}
+
+\end{knitrout}
+\caption{The impact of different propeller efficiency assumptions compared to the central pumpjet curve shows less variance}
+\label{fig:central_prop}
+\end{figure}
+
+\begin{figure}
+\begin{knitrout}
+\definecolor{shadecolor}{rgb}{0.969, 0.969, 0.969}\color{fgcolor}
+
+{\centering \includegraphics[width=\maxwidth]{figures/plots-plot_results4-1}
+
+}
+
+\end{knitrout}
+\caption{Comparing the lower and higher lines for both systems together}
+\label{fig:bottom_top}
+\end{figure}
+
+\subsection{Sensitivity of Curve Selection}
+As can be seen from the comprehensive comparison in Figures \ref{fig:central_extreme}, \ref{fig:central_jet}, \ref{fig:central_prop} and \ref{fig:bottom_top}, the shape of curves selected has very different levels of impact at different speeds.
+Key observations include how small the impact of efficiency variation to either curves tends to be at very low speeds (under 2kt).
+This is due to hotel load becoming clearly dominant at such speeds.
+The relative difference between the lines also diminishes somewhat towards higher speeds.
+More striking, however, is the very dramatic reduction in both range and endurance at higher speeds.
+At a speed of 20kt, a propeller driven sumbarine might manage to remain dived for 3.6 hours according to these assumptions, where a pumpjet driven submarine might manage just under 3 hours.
+A difference of about half an hour would presumeably be of less tactical significance.
+At any speed above 10kt, neither submarine would be able to remain dived for much more than 24 hours (25.4 hours for the propeller system, 18.3 for the jet).
+
+It should also be noted that these results, and those around 8kt, can also provide some insight into likely snorting intervals and indiscretion ratios.
+As given in \cite{patrick2012}, submarines tend to carry diesel power capacity that approaches their main motor power.
+Whilst the difference between the two systems still appears very significant at these speeds, it appears that if transits occur in the region of 8-10kt, snorts might occurr roughly daily, and last a number of hours.
+The indiscretion ratio during commute of the jet system would be significantly impaired given the substantially larger power demands at transit speeds, though a detailed calculation and comparison lies outside the scope of this paper.
+
+It seems extremely clear from these conclusions that the general results confirm very strongly that the impact of the propulsion system is likely to be of the greatest importance in the speed range 3-8kt, where propulsion power is more significant relative to hotel load, but excessive drag does not prohibit significant dives.
+
+A further feature of the results is that the relative difference in performance is much more sensitive to changes in the efficiency of the jet at the relevant speeds, owing to their larger proportional impact at low efficiencies.
+Whilst a smoothly declining efficiency curve from around or over 50\% at 20kt to zero at 0kt might seem moderately well defined, the precise path taken, particularly the exact level around 3-8kt, is of great importance, substantially more than plausible variations in the propeller curve.
+This can be seen in an overlaid plot of the proportional differences in total power demand (hence endurance and range) for the different pairs of curves as in Figure \ref{fig:all_proportions}.
+
+\begin{figure}
+\begin{knitrout}
+\definecolor{shadecolor}{rgb}{0.969, 0.969, 0.969}\color{fgcolor}
+
+{\centering \includegraphics[width=\maxwidth]{figures/plots-plot_proportions-1}
+
+}
+
+\end{knitrout}
+\caption{The most proportional differences between the propeller and pumpjet performances for a range of assumptions}
+\label{fig:all_proportions}
+\end{figure}
+
+The proportional change in overall power requirement (directly determining both range and endurance) shows that in the central scenario a propeller driven submarine would have more than 60\% advantage in terms of range and endurance around 5-6kt.
+It is interesting to note that the proportional change is greatest slightly above the speed where maximum range is achieved, and the point of peak advantage moves steadily towards higher speeds as the difference in efficiency between the systems increases.
+
+As can be seen, even compared to the central propeller curve, choosing a jet efficiency curve somewhat lower than the central assumption means that the peak difference immediately jumps to over 100\%, or effectively allow a propeller powered submarine to achieve double the range and endurance of a pumpjet powered submarine at those speeds.
+It is interesting to note that these speeds in include those speeds which might lie near the maximum speeds that could be sustained by Air Independent Propulsion (AIP) systems \parencite[100]{stanford2017}.
+Consequently, it is likely that a conventional submarine that used pumpjets might only achieve a fraction of the dived range and endurance enhancements generally expected on propeller powered submarines which employ this technology, as will be discussed later.
+
+The proportional improvements will also be effective rough guides to the likely impact on indescretion ratios and overall range.\footnote{Additional considerations including the size and efficiency of the diesel engines and generators must be included for proper analyses of indescretion ratios, which will not be given here.}  Near patrol speed almost all scenarios show a difference approaching or exceeding 50\%, which will have a material impact on indescretion rates during all combat operations.
+
+Perhaps surprisingly, the impact at transit speeds is still quite significant, with no scenarios showing less than 20\% advantage around 8-10kt.
+The sensitivity to pumpjet efficiency here is also very high, with three scenarios showing an impact of over 70\%.
+Given the likelihood that a submarine's average speed over its range is likely to be dominated by its transit speed, this indicates that the choice of a pumpjet propulsion system will also result in a substantial reduction in the overall range of the submarine, quite probably in the range of between 25-50\%.
+This would require the embarkation of significantly larger volumes of diesel to meet minimum range requirements, and also contribute to higher running costs.
+
+\subsection{Sensitivity to Hotel Load}
+
+As can be seen in Figure \ref{fig:sens_hotel} the difference between the two systems is very sensitive to changes in hotel load, both in proportional and nominal terms.
+The impact of halving the hotel load from 100kw to 50kw more than doubles the \textit{difference} between a pumpjet and propeller dived endurances, from around two days to well over four days at speeds between 2.5 and 3.5kt.
+In the context of a submarine which might have a nominal dived requirement at patrol speeds in the order of several days, this suggests that extremely large gains can be made by reducing hotel load.
+Increasing the hotel load has the noticeable effect of pushing the speed of largest difference to higher speeds, and substantially reducing the total difference.
+If the hotel load was doubled to 200kW, the middle efficiency curve pair would still see a difference of nearly a day between the two systems, or around 19.5 hours at speeds of 4.5-5 kts.
+
+\begin{figure}
+\begin{knitrout}
+\definecolor{shadecolor}{rgb}{0.969, 0.969, 0.969}\color{fgcolor}
+
+{\centering \includegraphics[width=\maxwidth]{figures/plots-plot_sens_hotel-1}
+
+}
+
+\end{knitrout}
+\caption{The impact of the propulsion system efficiencies are very sensitive to changes in the hotel load, including as a proportion.}
+\label{fig:sens_hotel}
+\end{figure}
+
+The difference between pumpjet and propeller propulsion systems in terms of range tends to be found at slightly higher speeds, though this point also moves towards higher speeds as hotel load increases.
+With a reduced hotel load of 50kw, the difference between the two systems in the middle scenario is about 370nm for speeds 3.5-4.5kt, substantially greater than the ~200nm around 4.5-5.5kt at a hotel load of 100kW.
+
+Changes in the effective battery energy density will have the same effect as having larger masses of battery stored.
+They equivalently impact the total amount of energy available for consumption before recharging batteries while snorkeling.
+There is therefore no shift in the proportional impact on range and endurance due to this factor, as can be seen in the lower graphs in Figure \ref{fig:sens_battery}.
+The nominal impact, however increases directly and linearly with the total energy stored.
+
+The results in nominal terms are striking in their magnitude.
+In terms of enduracy, doubling the stored electrical energy, for example by adopting lithium batteries, increase the \textit{difference} between the two systems under the central efficiency curve assumptions to nearly four days (93 hours) at 4kt.
+Given that most modern, capable submarines are estimated to only be able to achieve dived endurances between 60 and 100 hours on batteries at 4kt \parencite{buckingham2008submarine} this cannot be dismissed in terms of its tactical significance.
+Even under the efficiency curve assumptions representing the narrowest spearation considered, the difference in this case would be 59 hours, almost equal to the estimated dived endurance of a Type-214 operating without AIP.
+
+In terms of submerged range, the same proportional impact is also extremely significant.
+Doubling the energy stored inceases the \textit{difference} between two otherwise identical submarines to over 400nm under central assumptions at 4.5kt.
+That's a larger distance than the width of the Java Sea, the most central body of water in the Indonesian Archipelago, or most of the length of the Malacca Strait.
+The tactical significance of such distances are not likely to be lightly dismissed.
+If far smaller amounts of energy are embarked, the nominal differences reduce to perhaps 100nm, in direct proportion to the reduction of the submarine's dived endurance.
+This is still a non-trivial distance in terms of it's tactical signicance, and comprises a substantial fraction of the dived ranges of most non-AIP submarines.
+It still represents a significant difference in operational terms, equal approximately to the width of Hainan Island, or neraly twice the length of the Sunda Strait.
+
+It is also clear that increasing the hotel load has a greatly diminised impact on the proportional differences between the systems at higher speeds.
+At very high speeds (e.g.
+15kt) changes in the hotel load have almost no impact in proportional terms to the difference, which lies at about 30\% in all cases for the central selection of curves.
+At 9kt, in the region where transit speed is likely to occur, increasing the hotel load by a factor of four (from 50kW to 200kW) only decreases the advantage of the propeller from 48\% to 38\%.
+Therefore one might say near certainty that any difference between efficiency around these patrol speeds is likely to have a substantial impact on the overall range of the submarine, with differences of much less than 30\% quite unlikely, irrespective of other changes in the submarine design or operation circumstance.
+
+Only brief consideration is given here to the impact of Air Independent Propulsion, which has substantially higher energy densities possible even than Lithium, and also amounts to the addtional embarkation of greater mass.
+The volumetric energy density of the feedstocks to modern AIP systems (liquid oxygen and methanol) is estimated to be approximately ten times greater than that of lead-acid batteries \parencite{davies2016}, around 4.5MJ/L.
+Given the greater mass embarked, and additional studies suggesting that at constant speeds they might extend the dived endurance of a modern submarine from around 60 hours to over 1000 hours (a factor of roughly sixteen) \parencite{buckingham2008submarine} an approximate estimation of the the direction and magnitude of this impact may by scaling up energy density by this factor.
+This would amount to adding three further frames to the right of Figure \ref{fig:sens_battery}, with the energy density being doubled in each successive frame.
+This would plausibly (under the central line selections) amount to more than a month of dived endurance \textit{difference} between AIP inclusive submarines with propellers or pumpjets, amounting to thousands of nautical miles of continuously submerged travel.
+This difference would also now be material in terms of the overall range of the submarine amounting to a further 15-20\% reduction in the submarine's overall range.
+
+\subsection{Sensitivity to Battery Changes}
+
+Changes in the effective battery energy density will have the same effect as having larger masses of battery stored.
+They simply directly impact the total amount of energy available for consumption before recharging batteries while snorkeling.
+There is therefore no shift in the proportional impact on range and endurance due to this factor, as can be seen in the lower graphs in Figure \ref{fig:sens_battery}.
+The nominal impact, however increases directly and linearly with the total energy stored.
+
+\begin{figure}
+\begin{knitrout}
+\definecolor{shadecolor}{rgb}{0.969, 0.969, 0.969}\color{fgcolor}
+
+{\centering \includegraphics[width=\maxwidth]{figures/plots-plot_sens_battery-1}
+
+}
+
+\end{knitrout}
+\caption{The impact of the propulsion system efficiencies are extremely sensitive to battery energy density (or load) though proportions remain fixed.}
+\label{fig:sens_battery}
+\end{figure}
+
+The results in nominal terms are striking in their magnitude.
+In terms of enduracy, doubling the stored electrical energy, for example by adopting lithium batteries, increase the \textit{difference} between the two systems under the central efficiency curve assumptions to nearly four days (93 hours) at 4kt.
+Given that most modern, capable submarines are estimated to only be able to achieve dived endurances between 60 and 100 hours on batteries at 4kt \parencite{buckingham2008submarine} this cannot be dismissed in terms of its tactical significance.
+Even under the eficiency curve assumptions representing the narrowest spearation considered, the difference in this case would be 59 hours, almost equal to the estimated dived endurance of a Type214 operating without AIP.
+
+The in terms of submerged range, the same proportional impact is also extremely significant.
+Doubling the energy stored inceases the \textit{difference} between two otherwise identical submarines to over 400nm under central assumptions at 4.5kt.
+That's a larger distance than the width of the Java Sea, the most central body of water in the Indonesian Archipelago, or most of the length of the Malacca Strait.
+The tactical significance of such distances are not likely to be lightly dismissed.
+If far smaller amounts of energy are embarked, the nominal differences reduce to perhaps 100nm, in direct proportion to the reduction of the submarine's dived endurance.
+This is still a non-trivial distance in terms of it's tactical signicance, and comprises a substantial fraction of the dived ranges of most non-AIP submarines.
+It still represents a significant difference in operational terms, equal approximately to the width of Hainan Island, or neraly twice the length of the Sunda Strait.
+
+The singificant range of different energy density values tested here reflect quite plausibly some of ranges of different energy levels available in operation.
+For instance, as has been discussed \parencite{briggs2016, greenfield2016}, lead acid batteries may well only be able to practically utilise about 50\% of their total energy stores for tactical operaions, unless operating in a very low-threat environment and undertaking very slow charging to reach the maximum full charge is possible.
+In addition, the plausible improvement in enery density of lithium batteries may well be in the order of a factor of two in practice \parencite{greenfield2016}, though theoretical limits of the chemistry suggest much higher may be possible \parencite{davies2016}.
+
+Only brief is given here to the impact of Air Independent Propulsion, which has substantially higher energy densities possible even than Lithium, and also amounts to the addtional embarkation of greater mass.
+The volumetric energy density of the feedstocks to modern AIP systems (liquid oxygen and methanol) is estimated to be approximately ten times greater than that of lead-acid batteries \parencite{davies2016}, around 4.5MJ/L.
+Given the greater mass embarked, and additional studies suggesting that at constant speeds they might extend the dived endurance of a modern submarine from around 60 hours to over 1000 hours (a factor of roughly sixteen) an approximate estimation of the the direction and magnitude of this impact may by scaling up energy density by this factor.
+This would amount to adding three further frames to the right of Figure \ref{fig:sens_battery}, with the energy density being doubled in each successive frame.
+This would plausibly (under the central line selections) amount to more than month of dived endurance \textit{difference} between AIP inclusive submarines with propellers or pumpjets, amounting to thousands of nautical miles of continuously submerged travel.
+This difference would also now be material in terms of the overall range of the submarine amounting to a further 15-20\% reduction in the submarine's overall range.
+
+With such massive differences between the performance of the two systems emerging with the presence of the AIP, it is difficult to see how one could realistically consider a pumpjet as being competitive if a direct comparison were made.
+
+\subsection{The interaction of hotel load, battery energy density, and efficiency curves}
+It is interesting to observe how the the combination of changes to the hotel load and battery density interact when altered together, as shown in Figure \ref{fig:sens_corners}.
+To the extent that each of these effects contribute to increasing or decreasing the relative difference between the systems, the two clearly combine in a multiplicative fashion.
+
+Where battery density and hotel load are both reduced, the proportional change has a much higher and sharper peak at around 5kt, reaching an 83\% peak at 4.5kt under central assumptions.
+For high hotel load and battery density, the porportional impact is lower at a higher speed, at 43\% at 7kt.
+However, it is intereting to note that the nominal impact of each of these different scenarios is somewhat more similar in magnitude for both endurance and range.
+For low hotel load and battery store, the endurance difference peaks at 54 hours at 3 kt, and range difference peaks at 190nm at 4kt.
+For raised hotel loads and battery stores, the difference is shifted to higher speeds and broadened somewhat, with a difference of 39hours at 4-5kt, and 200nm at 5-6kt.
+
+The most striking feature of this figure however is the impact on the submarine's performance when hotel load is lowered, and battery density is raised simultaneously.
+Here the two factors combined have a proportional impact of around 80\% improvement for the switch to propellers under central assumptions, and over 150\% in the scenario of widest separation.
+However, in nominal terms, this impact is even more significant, with central scenarios showing a \textit{difference} of over 200 hours between the two systesm under central assumptions at 3kt, and a range difference of over 750nm.
+These differences are extremely large, representing plausibly double the expected total ranges and endurances of existing modern diesel-electric submarines.
+Even if such large changes in hotel load and battery capacity are not simultaneously likely these sensitivitiesquite clearly demonstrate how even modest shifts in both hotel load and battery energy density might combine to have very significant impacts on the appearnce of the pump-jet and propeller comparison.
+For instance, a hotel load that reduced by 30\%, and a battery energy density increase of 40\% could plausibly have the impact of doubling the nominal differces between the performance of the two propulsive systems.
+
+This sort of extreme sensitivity ought to have a significant bearing on how the overall performance specifications of the submarine are defined, as seemingly incremental changes to hotel load and electrical energy stores are likely to substantially magnified, or muted, by the choice of propulsion system.
+
+\begin{figure}
+\begin{knitrout}
+\definecolor{shadecolor}{rgb}{0.969, 0.969, 0.969}\color{fgcolor}
+
+{\centering \includegraphics[width=\maxwidth]{figures/plots-plot_sens_corners-1}
+
+}
+
+\end{knitrout}
+\caption{Specifying low density batteries and a high hotel load dramatically reduces the proportional and nominal impact of propulsion system efficiencies at low speeds}
+\label{fig:sens_corners}
+\end{figure}
+
+\subsection{Indiscretion Ratios}
+Indiscretion Ratios represent the fraction of time that a submarine must spend on the surface snorting (running diesel engines) in order to maintain a constant speed.
+In order to calculate the Indiscretion Ratio, it is necessary to know how much power the diesel engines are able to generate when running while snorting, and how much energy (net of any efficiency losses, hotel load, and power drawn to maintain propulsion) can be directed to recharging the batteries.
+Reliable information regarding the arrangement of engines, generators is not generally available in the public domain.
+However, from \cite{patrick2012}, it can be seen that credible submarine designs might include diesel engines that have a total capacity in the region of 85\% of the maximum power output of the main motor.
+Assuming that there are some necessary losses in the generation and recharging process, and plausibly some increase in the hotel load for the operation of additional systems, fur the purposes of this assessment it is assumed that while snorting approximately 5MW might be available to meet propulsion and normal hotel load requirements, and any surplus can be directed towards charging batteries.
+
+The fraction of time required to charge the batteries to a given level at a given speed, relative to the amount of time that the batteries could sustain the submarine at the same speed, then results in an indiscretion ratio.
+For similar assumptions to those used for central scenarios, including that hotel load is 100kW, 500tonnes of batteries are embarked, and the energy density of the batteries is that of lead-acid at 0.14MJ/kg, the results for the different efficiency curves considered for the two propulsion systems can be seen in Figure \ref{fig:IRs}.
+It should be noted that since the precise machinery choices for the likely submarine cannot be known with great confidence, the indiscretion ratio levels here at any given point have high uncertainties attached.
+However, the relative difference between indiscretion ratios at different speeds (the shape of the curve) as well as the difference between the two systems, should be closely resemblant of a real comparison of otherwise identical submarines, even if the level is somewhat shifted.
+
+The graph in \ref{fig:IRs} shows that very low indiscretion ratios, below 5\%, could be plausible for either system at very low speeds under 3kt.
+However, these very low indiscretion ratios do not last for a significant range of speeds for the pumpjet system.
+Whilst for a propeller driven submarine, such very low rates of indiscretion last up to around 6kt, pumpjet propelled submarines are likely to start to see their indiscretion ratios start rising significantly after 4kt.
+In the region of transit speeds, around 8-10kt, the difference between the two systems becomes clear and substantial.
+
+It should be noted that these plots only extend to 13kts, in contrast to previous plots which extend to 20kt.
+Since the propulsive power required to reach higher speeds can consume effectively all the power produced by the diesel engines, and then even exceed it at some speeds, the indiscretion ratios calculated by the method above become extremely large or negative, and don't assist in useful comparison.
+
+\begin{figure}
+\begin{knitrout}
+\definecolor{shadecolor}{rgb}{0.969, 0.969, 0.969}\color{fgcolor}
+
+{\centering \includegraphics[width=\maxwidth]{figures/plots-plot_IR-1}
+
+}
+
+\end{knitrout}
+\caption{Some calculation of plausible indiscretion ratios show that the pumpjet would substantially increase the amount of time spend snorting}
+\label{fig:IRs}
+\end{figure}
+
+\begin{figure}
+\begin{knitrout}
+\definecolor{shadecolor}{rgb}{0.969, 0.969, 0.969}\color{fgcolor}
+
+{\centering \includegraphics[width=\maxwidth]{figures/plots-plot_IR_sens-1}
+
+}
+
+\end{knitrout}
+\caption{The additional indiscretion ratio necessitated by a pumpjet, and this difference expressed as a proportion of the indiscretion ratio of a propeller driven submarine.
+Sensitivity to hotel load is not as great as for endurance or range calculations.}
+\label{fig:IRs_hotel}
+\end{figure}
+
+Figure \ref{fig:IRs_hotel} demonstrates the difference between the two otherwise idendtical submarines' indiscretion ratios as a function of speed, and also what this difference represents as a proportion of the indiscretion ratio at that speed for the submarine as a propeller.
+It can be seen here that the indiscretion ratio is generally somewhat less sensitive to variation in hotel load than calculations for range and endurance.
+(Indiscretion ratio is also independent of energy density of the batteries, and hence will not be compared.) However, increasing the hotel load does still have the impact of somewhat diminishing the peak difference between the two systems, and moving the point at which the proportional difference peaks to higher speeds.
+Where the penalty of the pumpjet on a submarine with a hotel load of 50kW might have the largest impact at 5kt, it the difference would probably be greatest at 6kt and 7kt for a sumbarine with hotel loads of 100kW and 200kW respectively.
+
+The magnitude of the difference is also substantial.
+For the central efficiency curve and hotel load assumptions, the pumpjet would cause a sumbarine to increase the time spent snorting for by over 60\% between 5kt and 7.5kt.
+One would expect that this would be of considerable tactical consequence.
+These graphs also demonstrate that the difference does not diminish at higher speeds, infact it begins to increase again above 10kt, as the power required for propulsion begins consume much of the power generated by the diesel engines, and efficiency difference strongly impacts the size of the diminishing surplus available for charging.
+It is also clear that the sensitivity to the selection of efficiency curves remains quite strong in this case.
+It is clearly plausible that in a range of circumstances the pumpjet could double the time spent snorting.
+However, the difference between the lines representing the pair of central efficiency curves, and the narrowest separation, is noticeably smaller, and the impact of chaning hotel load on these lines is also proportionally diminished.
+As such, it might be said with considerable confidence that the most likely impact of the pumpjet on indiscetion ratios at any speeds over 4kt will around 40\% or higher.
+It would be extremely unlikely that the penalty induced by the pumpjet at speeds of 5kt and above would be any less than 25\%.
+
+\newpage  \section{Acoustic Considerations} \label{acousticconsiderations}
+
+\subsection{The importance of cavitation inception}
+It is generally considered that the existence of cavitation of any sort will tend to dominate the acoustic signature of any vessel that doesn't produce any machinery noise by relying on all-electric drive as a conventional submarine would when dived (\cite{giles2010}, \cite{harvie1965construction}).
+Its quite possible that during the period when a submarine is snorting, or running its diesel engines, these sources of noise would be a more substantial and distinctive acoustic signature than small amounts of cavitation on the propulsion system if it were to occur, owing in particular to the low freqency of diesel engines.
+
+Other sources of noise hydrodynamic noise, such as the turbulence produced in the boundary layer of the flow over the hull, or parts of the propulsor, will still contribute to overall noise and some signature, but it will tend to be significant smaller in magnitude.
+This owes to the essentially incompressible nature of water.
+Without compression, even very relatively intense votices or tubulent flows will not propagate very much noise, if the water remains perfectly solid.
+All of the 'round and round' can't achieve much 'in and out', which is what results in propagated noise.
+This obviously doesn't hold true in compressible fluids, such as air, as one would observe noticing the intense noise produced by hand-dryers, aeroplanes, jet engines etc.
+It also doesn't hold true \textit{within} a turbulent flow underwater.
+If you were to put a sensor inside a turbulent flow, the rapid movements would result in continual and significant pressure fluctuations, hence noise.
+But at a distance, these fluctuations tend to substantially cancel out.
+The larger the scale of the eddies and vortices, the further noise might propagate, and also the lower the frequencies of resultant noise might be, which makes noise propagate further.
+Likewise, larger caviation voids will tend to emit lower frequency sound aswell, as can be observed from the boiling of a kettle.
+
+As a result, when dived, it might be reasonably assumed that the complete elimination of all sources of cavitition will be the primary objective of acoustic signature management.
+This is neatly put in \cite[88]{gearhart1966selection}: "The occurence of cavitation in a propulsor results in a degradation of the propulsor operating characteristics if extensive amounts of cavitation exist, and in significant noise when minute amounts occur.
+As a result, the performance requirements of a submersible weapon system specify that the occurence or inception of cavitation be avoided below certain operating depths." One might consider that a submarine's 'tactical silent speed' would most probably correspond to the speed at which it can operate safely below the point of any cavitation inception.
+
+The precise point at which cavitation inception occurs, what type of cavitation it is, and where it occurs on the blades is a complex science, with considerable work being continually undertaken to validate theoretical models.
+A good overview of the some of the science can be found in reference \cite{kuiper1981cavitation}.
+Still today, similar techniques such as paint tests and visual observations are validate models and theory.
+However, despite the inherent complexity about predicting the precise point in time and space of onset, and the respective impact on efficiency and acoustics, several broad relationships are known to hold well, and can be relied upon to characterise the most significant trends that determine when cavitation will be more likely to occur, which is generally driven by the diffence between the vapour pressure of water and the local static pressure.
+
+\subsection{Impact of speed on cavitation}
+Perhaps the most significant determinant of the pressure deviations on the propulsion system is how hard the propellers have to push on the water at any given point.
+In general, the harder the propellers push, the larger is the required pull on some opposing edge or side.
+It is these pressure reductions, on the suction side of the blade, where sheet cavitation generally starts to occur.
+To think of it simply, the water isn't able to rush to fill the gap left behind by the propeller fast enough.
+However, at some point cavitation also occurs on the pressure side of the blade, as the impact of the leading edge breaking the water leads to sever local pressure drops just behind the leading edge.
+
+Whilst a range of very particular and intricate phenomena may be at work, perhaps the largest and most dominant trend at work will be that pressure differentials across the blade must necessarily increase with increase thrust.
+The blade velocity relative to the oncoming stream is also a close indicator, which will be closely related for an open propeller.
+Consequently, since the thrust required for self-propulsion increases with velocity squared, the propensity for cavitation occur will necessarily tend to follow this relationship also.
+As such, modest changes in speed can have a big impact on the degree of probability of cavitation occurring.
+Halving your speed will roughly reduce local pressure fluctions by a factor of four.
+Consequently, even for non-specialised propellers, there is generaly some low speed for which no cavitation occurs, though almost all propellers in major surface vessels experience some cavitation at the speed they are designed to do their work at.
+
+\subsection{The impact of depth on cavitation}
+It is also crucial to understand that increasing depth necessarily increases the static pressure throughout a propulsion system.
+As such, the speed at which cavitation starts to occur will necessarily be a function of depth.
+Approximately every 10m of water adds approximately the equivalent of an atmosphere of pressure.
+As such, the static pressure a system experiences at 10m depth will be half that experienced at 30m (since the air in the atmosphere comprises the first atmosphere of pressure).
+Consequently, diving deeper will necessarily increase the speeds one can achieve without any cavitation occurring \ref{fig:InceptionDepth.png}.
+
+\begin{figure}[h]
+\captionsetup{width=0.7\textwidth}
+\includegraphics[width=0.7\textwidth, center]{InceptionDepth.png}
+\caption{Increasing in depth improves the speed achievable before cavitation begins to occur, as shown in the case of a torpedo \parencite{suryanarayana2010cavitation}}
+\label{fig:InceptionDepth.png}
+\end{figure}
+
+The interaction of these two relationships can be expressed in a general form in figure \ref{fig:InceptionContours}, which has been generated using an approxiate numeric model.
+As can be seen here, at very low speeds, modest changes in depths can achieve considerable additional surplus pressure in order to prevent cavitation.
+However at much higher speeds, since the pressure diffential scales with the square of the velocity travelled, far greater additional depths are required to achieve sufficient excess static to supress cavitation for a give further speed increase.
+Whilst certain system parameters such as the blade area and design (or duct design) will have a significant impact on which curve a particular system lies upon, the overall trend of a parabolic cavitation free "bucket" prevailing at low speeds and large depths will generally be robust.
+This is confirmed by the close resemblance between the curve shapes found by an experiment on a torpdo in Figure \ref{fig:InceptionDepth.png}, and the numerical model found in Figure \ref{fig:InceptionContours}.
+
+\begin{figure}
+\begin{knitrout}
+\definecolor{shadecolor}{rgb}{0.969, 0.969, 0.969}\color{fgcolor}
+
+{\centering \includegraphics[width=\maxwidth]{figures/plots-plot_inception-1}
+
+}
+
+\end{knitrout}
+\caption{Contours showing plausible lines of plausible equivalent cavitation inception}
+\label{fig:InceptionContours}
+\end{figure}
+
+It is noteworthy also that the vapour pressure of water is an important component in the equations that describe the conditions for cavitation to occur.
+It is generally given that when the local static pressure falls below the vapour pressure, cavitation occurs, though known exceptions to this do exist.
+Given that temperatures also affects the vapour pressure of water (warmer water is just a little closer to boiling) this does have an impact on the cavitation inception point.
+However, at plausible temperatures the impact is small.
+The difference between the vapour pressure of water at 10 degrees and 30 degrees is approximately 3 kilopascals.
+Diving just 10m deeper adds over 100 kilopascals to static pressure.
+In many circumstances, altering your speed by a couple of knots will have an even larger impact on local pressure reductions.
+Consequently, the impact of temperature on the inception of cavitation is essentially negligible.
+The difference between inception contours in 20 degree and 30 degree water are given by the difference between the black and red-dashed contours in Figure \ref{fig:InceptionContours}.
+
+\subsection{The significance of flow separation}
+The addition of a duct will necessarily lead to flow separation occuring on the inside of the duct opening at some very low speeds which correspond to high advance ratios when load remains constant.
+Flow separation tends to lead to vortex shedding, which means that inconsistent flows (vortices) are then ingested into the blades of the impeller.
+This effect is well known and documented.
+A good example is shown on the exterior of an \textbf{accelerating} duct at $J = .8$ in \cite{willemsen2013}.
+
+\begin{figure}[h]
+\captionsetup{width=0.5\textwidth}
+\includegraphics[width=0.6\textwidth, center]{VortexShedding.png}
+\caption{Vortex shedding following a flow separation leads to eddies being released unsteadily into the main flow stream, as shown here on the exterior of an accelerating duct.
+In the case of decelerating duct (pumpjet), duct shape is inverted, leading to the vortices being shed into the impeller blades.
+\parencite{willemsen2013}}
+\label{fig:VortexShedding.png}
+\end{figure}
+
+Whilst at very low power levels or at good depths, it is quite plausible that this doesn't result in any cavitation, which would normally be a likely occurrence if the blades are operating closer to their design power and such such vortices were shed into the intake.
+However, given the inherently unpredictable and unsteady nature of such vortex formation and shedding, this phenomenon would increase the risk of cavitation occurring unexpectedly in marginal conditions.
+In addition, in the absence of cavitation, the sudden and changing pressure gradients over the blades would dramatically increase the overall pressure fluctuations and turbulence level that occurs within the propulsor.
+This would necessarily increase the overall levels of noise generated.
+It is plausible that the noise is of sufficiently low level that it is not considered a significant compromise of stealth.
+However, it is almost certainly untrue that that pumpjet actually generates less noise overall than a propeller at very low speeds.
+
+\subsection{The impact of duct shielding}
+The presence of the duct around the impeller may also have the effect of preventing the noise from being effectively radiated away from the propulsor.
+This may well result in substantial reductions in some types of noise being radiated away from the propulsor in certain directions, which may have given rise to the consistent qualification of 'radiated' being used to descrive the type of noise which differentiated the French submarine bid (\cite{stewart2016}, \cite{davies2017}, \cite{ohff2016}), including by the designers of the submarine \parencite{autret2016}.
+This is a secondary effect, quite separate to the avoidance of cavitation, which might be seen as an acoustic advantage in circumstances where cavitation might not be expected to occur on either type of propulsor.
+It is highly likely that this is the effect which is relied upon to sustain claims that a pumpjet is generally acoustically superior to a propeller, including at low speeds.
+
+It should be noted here that shielding effects might comprise the combination two distinct and seperate physically phenomenon: reflection and damping.
+Noise tends to be reflected very effectively off hard, smooth surfaces representing the boundary between mediums of very different density (such as steel and water).
+Changes in density which are gradual, often representing soft, squishy, or broken surfaces tend, to have the effect of absorbing sound, where the acoustic energy is converted to heat or degradation of the material, rather than reflecting it.
+Given the significance of increased drag on any surfaces within the propulsor, it is highly unlikely that any attempt at damping is made within the pumpjet.
+However, it should be noted that some pumpjets have been constructed with a hollow design, in order to reduce weight \parencite{harvie1965construction}, which would reduce the thickness of solid steel, though offer some potential opportunity for damping.
+Overall, damping is increased in its effectiveness at low frequencies by allowing it to occur over larger distances, corresponding to the sound's increased wavelength, leading to increased thickness of the transition substance or damping material.
+Consequently, effective damping within the propulsor itself is unlikely to be attempted, or limited to medium and higher frequencies, probably suited to reducing noise from normal surface flow turbulence and very early onset cavitation.
+
+The ability of a thin layer of material to reflect noise is also highly frequecy dependent, and also dependent on mass of the shield layer.
+As the material becomes thinner and thinner, a hard shield layer will tend increasinly to transmit rather than reflect the noise, a phenomenon that increases as the amplitude (loudness) of the noise increases, and frequency decreases.
+Denser materials are more effective at reflecting sound, as more energy (pressure) is required to excite the same movement in the material.
+Consequently, the effectiveness of a shield depends on both its thickness and density.
+There is therefore some tension inherent in the design of a shield where thickness and total weight are constrained.
+A thicker layer solid steel will be very effective at reflecting sound, but have negligible absoption or damping.
+A hollow between two thin layers will offer some potential for a soft daping material to be inserted, but at the cost of reflection.
+
+It is most likely that little or no damping or absorbing layer is attempted within the body of the shroud, since increasing the duct thickness above hydrodynamically determined optimum profile is likely to have a detrimental impact on efficiency.
+Constructing a large components from thing pieces of steel, and still maintaining an optimal shape to the precision required for housing powerful turbomachinery like the impeller is also likely to involve considerable construction risk and complexity.
+Hence the shielding effect is most likely to dominated by the reflective characteristics of the duct, which would seek to internally reflect noise back into the turbulent flow, where sound waves would be increasingly convoluted, and somewhat absorbed through the eventual transition to small-scale turbulence and heat.
+
+The degree to which the total noise generated is reflected internally will also be shape-dependent.
+A more concave inner surface, will increase the extent of internal reflection.
+Such a design would be consistent with higher degrees of diffusion in advance of the impeller, and greater nozzle narrowing after it, which is consistent with higher pressure elevation at the impeller (hence improved cavitation characteristics) but higher negative duct thrust, leading to even poorer efficiencies at low speed.
+Substantial alteration of duct shape on account of efficiency considerations at low speed are therefore necessarily going to have some impact on the degree of effective shielding that a duct can afford.
+
+\subsubsection{Directional Dependence}
+Two important qualifications must be made regarding shielding.
+The first is that its impact is highly directionally dependent.
+The pump jet requires openings at the front and back to allow water in and out.
+Only noise that is reflected internally sufficient to be largely absorbed is effectively shielded.
+The larger the areas of the intakes and outlets, the narrower the effective shielding will become.
+Consequently, design changes to make the pump more efficient at low speeds, which may include expanding both the inlet and outlent, and diminshing the difference between them, will necessarily be at the expense of diminishing effective width of the shielding 'belt' around the propulsor.
+
+When the submarine is travelling at high speeds, and there is significant flow through the propulsor, the impact of waterspeed through the propulsor might have the additional impact of diminishing the radiation of noise in the forwards direction, in the same way that sound tends tends not to travel well up-wind in air.
+However, at very low speeds this effect would be greatly diminished, and possibly not relevant.
+
+It should further be noted that in the underwater environment sound can be made to reflect in a variety of directions simply with differences in temperature, salinity, movement of water bodies, as well as off plenty of nearby underwater objects, including the ocean floor and objects on it.
+This phenomenon generally serves to make the art of underwater acoustics complex and unpredictable, and generally contributes to the difficulty of precisely locating a submarine, even when some signature is detected.
+However, it also means that direction-dependent reductions signature do not reliably translate to overall reduced probability of detection.
+
+\subsubsection{Frequency Dependence}
+It ought further to be noted that the effectiveness of acoustic shielding is also highly dependent on frequency.
+Much lower frequencies are very hard to reflect, and absorb, as can be easily observed in a house with a home-cinema, where lower frequencies emitted by the subwoofer will tend to affect many surrounding rooms, where higher frequency sounds might be substantially muffled by interceding walls.
+The same acoustic effects are at work in underwater acoustics.
+
+This has an important consequence for shielding for pumpjets at very low speeds because of the occurrence of flow separation and vortex shedding.
+In the absence of flow separation and cavitation at very low speeds the most substantial sources of sound would be the turbulent flow over surfaces of in the propulsor.
+This noise will tend to be relatively white and high in its frequency owing to the very small length-scales of the eddies and vortices in such turbulence.
+The steel shroud would be very effective in reflecting such high frequency noise.
+Under acceleration or higher loads, the very instances of cavitation woud also be relatively high frequency, as the voids or cavities would be small at first.
+These sounds might also be very effectively reflected by the shroud, and consequently diminished.
+
+In contrast, flow separation results in the creation of vortices of a dramatically larger scale than occurs with normal turbulence in a boundary later, and hence would tend to generate noise of a much lower frequency.
+In addition, as the flow rates slow even further, these vortices are likely to become larger in size, even if lesser in intensity, resulting in further reduction in the frequency of noise generated.
+The rate at which the vortices shed from the surface and impact the impeller will also result in another acoustic signature.
+Again, at lower flow rates, these vortices will tend to develop and shed more slowly, lowering frequencies further.
+
+Consequently, whilst for certain types of noise which tend to become both louder and lower in frequency with increasing speed (cavitation), flow separation provides a contrasting characteristic, in that it is likely to decrease in frequecy with decreasing speed.
+As lower frequencies tend to propagate much further in open water, this could have a significand adverse impact on the total acoustic signature of the submarine, even if the flow separation doesn't lead to any cavitation.
+
+The foregoing consideration suggests that the acoustic signature of a jet-powered submarine might not reduce steadily with speed all the way to stationary movement.
+It is plausible that there could be some elevated propulsion signature at very low speeds due to flow separation, and that the submarine's optimal acoustic signature when moving is actually at some slightly higher speed, where smoother flow through the propulsor can be achieved.
+However, given that the amplitude, frequency, and combined impact of those two characteristics cannot be accurately evaluated without substantial further work and more detailed system knowledge, it is also possible that the increase in other noise (including hull flow noise and other machinery) at elevated speeds offsets the improvement in the smoothness of the pumpjet's operation.
+It is still quite likely the overall level of noise radiated by the pumpjet, including under conditions when flow separation occurs, is still considered to be sufficiently low in amplitude as to not be of significant tactical consequence.
+However it is highly unlikely that the accoustic signature of the pumpjet is actually superior to that of an open propeller at very low speeds, when cavitation occurs for neither system, but substantially increased turbulence including flow separation must neccessarily occur for the pumpjet.
+
+\newpage
+
+\printbibliography
+
+\newpage
+
+\begin{appendices}
 \section{Some essential concepts}
@@
 The impact of the increased surface area, and more turbulent flow over the surface, is exceeded by the reduced wake field ( trailing area where the flow is totally separted) behind the ball.
-
-\section{Basics of Ducted Propellers and Pumpjets}
-It's easy to be somewhat confused by the many different names which seem to be associated with related, or similar systems.
-Between pumps, pumpjets, water-jets ducted or shrouded propellers, or impellers, there is plenty of grounds for some confusion  In this section, I aim to clarify in simple terms what the important differences between distinctly different systems are, and where some terms are used somewhat interchangeably without doing any great violence to the concepts underlying.
-Appreciating where the fundamental as opposed to superficial differences lie is important to understanding subsequently the limitations of different systems.
-
-\subsection{Propellers}
-Perhaps the best starting point is the most basic, and oldest of the systems which we're considering: the basic screw propeller.
-A propeller uses a number of tilted blades attached to a central hub to sweep around a disc in the water, and accelerate a column of water passing through the disc.
-Due to the slant of the blades, water on the back sides of the blades is pushed astern, and reduced pressure on the front faces pulls more water from ahead to replace it.
-As such, it accelerates a column of water, which necessarily is contracted in circumference after the point of acceleration.
-
-\begin{figure}
-\includegraphics[width=\textwidth]{PropellerAction.png}
-\caption{Propellers generate thrust by accelerating a column of water, as shown in  \parencite[247]{mollard2011}}
-\label{fig:PropellerAction.png}
-\end{figure}
-
-A comprehensive description of the development of the propeller can be found in \cite{carlton2007}.
-Here I won't elaborate on beyond describing some essential features and characteristics with which one ought to be familiar, primarily for the purpose of comparing different propellers and understanding their evolution into the systems such as pumpjets.
-
-Most propellers will have between three and seven blades.
-In general they are shaped as an aerofoil, with the convex side being upstream, just as the convex side of a plane wing is above, in order to generate lower pressure and lift as it moves through the air.
-The blades tend to be twised so that they have a higher angle of attack closer to the hub, and lesser closer to the extremities, so that those faster moving sections push their respective parts of the water column at an overall similar speed.
-The are often also thinner towards the extremities, and will be swept backwards as if dragged by their rotation through the water (skew) and also dragged by the ship's movement throug the water (rake).
-Whilst the degree of all these characteristics is highly variable for different applications, these are a few of the commonly referred to characteristics which can be varied in order to optimise performance for any given application.
-
-\begin{figure}
-\includegraphics[width=\textwidth]{SkewRake.png}
-\caption{The development of a boundary layer \parencite[262]{mollard2011}}
-\label{fig:SkewRake.png}
-\end{figure}
-
-\subsubsection{Pitch}
-Perhaps one of the most important characteristics of a propeller is its pitch.
-The pitch represents the distance that a blade section would travel forward if it carved its way around a helix through one full rotation.
-It is intuitive to think of as something akin to the angle-of-attack of the propeller blade to the water, however this is technically misleading as the movement of the water-column incoming to the blade, as well as the speed of rotation, also have a significant bearing on what the actual angle of attack of the blade ends up being.
-Whilst technically pitch is a distance (measured in meters), it is often expressed and used in formula as a ratio of the diameter of the propeller, to give it a unitless equivalent, as is common in the description of many characteristics of propellers and propulsion systems.
-
-\begin{figure}
-\includegraphics[width=\textwidth]{HelicalLine.png}
-\caption{A blade section of a propeller traces a helical line around a cylinder as it is rotated \parencite{carlton2007}}
-\label{fig:HelicalLine.png}
-\end{figure}
-
-\begin{figure}
-\includegraphics[width=\textwidth]{PitchDefinition.png}
-\caption{The definition of pitch is the distance traveled by a blade section along a cylinder, as given in \parencite{carlton2007}}
-\label{fig:PitchDefinition.png}
-\end{figure}
-
-The selection of the exact blade selection that is used to define pitch must be specified as being at some fraction of the radius from the centre, given the tendancy for the blades to twist.
-This distance is calculated to be the 'moment mean' or a technical derived effective average, which tends to lie between 0.6R and 0.7R.
-A thorough discussion of pitch can be found in \cite[35-37]{carlton2007}.
-
-Pitch is of particular importance to the discussion the efficiency of propeller and propulsor design because its optimal choice tends to vary considerably with the different loads which a vessel is intended to operate at, which can also relate to a vessel's design speed.
-It is for this reason that a considerable number of vessels actually have variable pitch propellers.
-Whilst considerably more complex, expensive, and heavier than a traditional fixed-pitch propeller, the ability to vary the pitch of a propeller assists considerably, particularly when the amount of load (resistance) a vessel is expected to face varies considerably.
-Controllable pitch propellers also have advangates in terms of manouverability, since the pitch can be reversed and hence reverse thrust can be produced, without requiring the direction of the power coming from the engines to be reversed.
-This has advantages for ferries which undertake frequent docking manouvres in confined spaces (\cite{MAN2017}, \cite{lewis1988}).
-
-\begin{figure}
-\includegraphics[width=\textwidth]{CPPropeller.png}
-\caption{Controllable Pitch Propellers (CP Propellers) allow the pitch of the blades to be varied. Image credit: \parencite{mollard2011}}
-\label{fig:CPPropeller.png}
-\end{figure}
-
-\subsubsection{Pitch and the Advance Ratio}
-Pitch is also an extremely important concept to understand because it relates closely to the way that a propeller's 'open-water efficiency' $\eta_O$ is expressed in charts, as well as the advance ratio $J$ that usually comprises the x-axis in such charts.
-
-\begin{figure}
-\includegraphics[width=\textwidth]{OpenWater.png}
-\caption{A typical chart showing open water efficiency of a propeller, in this case a Wageningen B5-75. Image: MARIN}
-\label{fig:OpenWater.png}
-\end{figure}
-
-This chart also shows two other closely related dimensionless coefficients, the torque and thrust coefficients ($K_Q$ and $K_T$) which we shall not dwell on here.
-However, it is important to note that a number of different lines are expressed on this sort of chart, which represent different versions of the same propeller, with only the pitch (expressed in its dimensionless form as $P/D$) being different in each case.
-The x-axis, $J$, also warrants a clear definition:
-
-\begin{equation}
-\label{eq:2}
-J = \frac{V_a}{nD}
-\end{equation}
-
-Where $V_a$ is the speed of advance of the propeller, or the speed at which water arrives at the front-side of the propeller, and $n$ represents the rate of the propeller's rotation (revolutions per second) and $D$ is the propeller's diameter.
-For fullness of understanding it is also worthwhile to note here that $V_a$ does not correspond directly to the vessel's speed of travel, though they are related.
-To some extent, a vessel moving through water will always draw water in its wake along with it, which means that in the wake (where the propeller tends to operate to some extent or other) the water incident on the propeller will arrive at a lower speed than that implied by the ship speed \parencite[15]{MAN2017}.
-Another key distinction for the consideration of jets later on is that the speed of advance does not necessarily incporate the movement of water due to the movement induced by the propeller (or duct) itself, which is a distinct effect to that of the wake field.
-In \cite[213]{lewis1988} $V_A$ is used specifically to denote the undisturbed stream velocity, since the role of certain ducts can be to accelerate or decelerate the stream prior to the water arriving at the blades of the propeller.
-
-One may also note from the previous figure that for each line representing a given pitch, the efficiency falls precipitously towards zero around the point where $J$ reaches $P/D$.
-This is because this extreme of the curve represents a circumstance where the propeller is rotating, but not actually moving any water, since it is moving through the water at about that speed which corresponds to the exact helical path defined by the pitch being traced.
-This represents circumstances where a propeller encounters effectively zero load.
-This might be the case when a eases it's power, and the momentum of the ship temporarily keeps the ship moving without the propeller exerting much or any net thrust, or where alternative propoulsion sources (sails, or other propellers) are keeping a speed higher than what the propeller actually has push backwards on water to maintain.
-Since drag due to friction scales with velocity squared, very low speed do correspond to increasing advance ratios for a given propeller size, as would increases in propeller size for a given speed and pitch ratio.
-In practice however, under self-propuslion conditions for open propellers, slowing down towards does not elevate the advance ratio far enough past beyond peak efficiencies for overall efficiency to fall very far, as can be seen in Figure \ref{fig:EfficiencyBulten.png}
-
-A closely related concept here is the slip ratio, which is a measure of the difference between the pitch (the distance the blade section would travel if it traced a perfect helix for one rotation) and the actual distance that the blade did advance through the water.
-This is neatly defined and shown in a diagram in \ref{fig:SlipRatio.png}.
-
-\begin{figure}
-\includegraphics[width=\textwidth]{SlipRatio.png}
-\caption{The slip ratio of a propeller. Image credit: \cite{MAN2017}}
-\label{fig:SlipRatio.png}
-\end{figure}
-
-It can be seen from the diagram of the open-water efficiencies that there is some desirable non-zero slip ratio for a propeller.
-The other end of the spectrum, where efficiencies also fall off toward zero, represent circumstances where the propeller is simply spinning in the same place in the water.
-In this case the slip ratio is one.
-This represents a circumstance known as 'bollard pull' where a vessel is simply exerting force against an immovable object or infinite load.
-Whilst the very low efficiencies here could represent a case of a propeller that is simply too small for the task it is trying to achieve, it is actually a significant and important circumstance to consider for propellers on working vessels such as tugs or trawlers, which exert large forces at very low water-speeds.
-Measurement of the thrust generated in bollard pull, and the power required to achieve this thrust, is still an important metric for such vessels, but the definition of 'useful work' for normal propulsion requieres the thust to be excerted over some distance travelled.
-
-\subsection{Ducted Propellers}
-A ducted propeller is a propeller in any kind of tubular shroud, casing or nozzle.
-Ducted propellers fall into two categories, one of which is referred to frequently and interchangeably as a pump-jet.
-The two types are called the accelerating type (also known as a Kort Nozzle, after a company that made them popular in the 20th Century) and the decelerating duct.
-The names describe the different effects that each type of nozzle have on the flow speed of the water at the point where it meets the propeller (which is also called an impeller, particularly when associated with decelerating ducts, or pumpjets).
-
-A detailed description of the essential differences between the two can be easily demonstrated by diagram, which show a simple comparison an acelerating and decelerating duct, which have the shape of the aerofoil section which comprises the duct inverted.
-\footnote{It should be noted that some designs attempt to strike a balance between efficiency and cavitation performance of accelerating and decelerating ducts for certain purposes, and in such cases the duct shape might not be so clearly contrasted \parencite{abdel2010}.
-But for the purposes of explanation, as well as optimisation for military purposes, the distinction between the two will generally be quite clear, and in practice ducts will still tend to decelerate the flow on approach to the impeller unless specifically shaped to narrow in the opening.}
-
-\begin{figure}
-\includegraphics[width=\textwidth]{AccelDecel.png}
-\caption{he two types of ducted propeller, accelerating (left) and decelerating (right) as shown in \parencite[249]{mollard2011}}
-\label{fig:AccelDecel.png}
-\end{figure}
-
-A consideration of what happens to the flow as it moves through each type of duct helps to demonstrate how the acceleration and deceleration effects are achieved.
-In the case of an accelerating nozzle, the area through which the flow must pass contracts prior to the water arriving at the impeller.
-In the decellerating case, the available area contracts after the water has passed the impeller.
-In some cases, a degree of diffusion is actually incorporated ahead of the impeller, whereby the area actually of the flow actually expands ahead of the impeller, which slows down the flow and increases its pressure according to Benoulli's principal \parencite{wislicenus1973}.
-
-\begin{figure}
-\includegraphics[width=\textwidth]{DuctStreamlines.png}
-\caption{The streamlines of flow through different duct types \parencite[214]{lewis1988}}
-\label{fig:DuctStreamlines.png}
-\end{figure}
-
-\subsubsection{The Accelerating Duct}
-Accelerating ducts are most commonly used by vessels which have to operate at high loadings at very low speed, such as tugs, barges, or trawlers.
-In those circumstances, relatively small vessels need to exert considerable forces against substantial loads, at very low water-speeds.
-Also, in these circumstances, the practical size of a propeller might be constrained, since operating an optimally sized propeller might increase the complexity of machinery required to deliver the power to so low a hub in the water, and it might also be difficult to arrange for the thrust to be directed substantially against the centre of drag of the vessel.
-
-The effect of the nozzle in these circumstances actually increase the optimal efficiency of the propeller, when operating at very low water-speeds but high thrusts (low advance ratio), since the pressure reduction on the inside of the nozzle induced by the movement of the water through the nozzle actually produces a net thrust forward, since the interior face at the opening of the nozzle is necessarily angled forward.
-This has the effect of distributing the suction force across a larger area, which now includes the interior of the nozzle.
-An equivalent way of describing the same effect is that the nozzle causes the water to be accelerated more gradually, over a longer distance, by inducing it to move more quickly as it passes through to the entrance.
-As a broad generalisation, more gradual or gentle actions tend to involve lower overall losses to turbulence, and hence the efficiency of the propeller can be increased.
-
-A further means by which the efficiency of a propeller can be increased is through the reduction of the turbulence created by the blade-tips passing through the open (static) water at high speed.
-Since the two sides of the blades have low and high pressure on them respectively, at the tips where the two faces meet, there there is a tendancy for water to 'spill' from the high pressure side the the low pressure side, creating a vortex which leads to unwanted turbulence.
-It is also around this vortex that blade tip cavitation tends to occur.
-By limiting the movement of water around the outside of the blade tip by having a tightly fitted shroud, losses due to this particulary type of tubulence, and the onset of this type of cavitation, or the vortices that occur even when cavitation does not, can be diminished.
-
-However, it is important to note that the existence of an accelerating duct by no means eliminates all cavitation.
-In fact, given that the static pressure at the impeller is necessarily reduced by an accleratint duct, the propensity for cavitation other surface of the propeller is generally increased.
-In many circumstance cavitation can and does still occur in ducted propellers near the blade tips at design loads \parencite{moulijn2015}.
-
-\begin{figure}
-\includegraphics[width=\textwidth]{CavitatingDuct.png}
-\caption{An example of the cavitation occurring at the blade tips of a ducted propeller. Image credit \cite{moulijn2015}}
-\label{fig:CavitatingDuct.png}
-\end{figure}
-
-A detailed explanation of the means by which the ideal efficiency of propeller can be improved can be found in \cite[213-222]{lewis1988}, as well as other texts on marine propulsion, such as \cite{carlton2007}.
-A substantial work exploring the efficiencies of a range of accelerating and decelerating ducts can also be found in \cite{oosterveld1970}.
-A common feature of these extensive studies also show that for ducted propellers, including accelerating ducted propellers, the advantage in terms of efficiency is overall restricted to low waterspeeds.
-At higher waterspeeds, the drag induced by the water-movement over the exterior of the duct begins to increase substantially, and the net thrust on the duct becomes negative.
-When the exterior of the duct experiences relatively little water velocity, but the interior experiences a great deal more, the greatest advantages of an accelerating duct are realised.
-Furthermore, in general these models do not incorporate the effect of flow separation, which is acknowledged as a possibility on both the exterior and interior of ducts, which can dramatically reduce efficiency performance (\cite[20]{oosterveld1970}, \cite[214]{lewis1988}).
-
-\begin{figure}
-\includegraphics[width=\textwidth]{DuctedEfficiency.png}
-\caption{Accelerating ducts produce a net positive thrust at low advance ratios, but a net negative thrust at higher advance ratios as the drag on the nozzle exterior increases.
-Image credit:Open-water test results of Ka 3-65 screw series with nozzle no.
-19A \parencite{oosterveld1970}}
-\label{fig:DuctedEfficiency.png}
-\end{figure}
-
-\subsubsection{The Decelerating Duct or Pumpjet}
-Decelerating ducts necessarily produce a negative thrust, since the process of raising the pressure of the water about the impeller (equivalent to slowing it down) necessitate reducing its kinetic energy and consequently momentum.
-In a wide range of literature introducing the fundamentals of ducted propellers and water-jets, or discussing their more modern development there is universal acknowledgement that the choice of a decelerating duct is generally made for the avoidance of cavitation for specialised (generally military) purposes, rather than for the achievement of increased efficiency (\cite{carlton2007}, \cite{oosterveld1970}, \cite{mollard2011}, \cite{haimov2010}, \cite{lewis1988}).
-
-In order to achieve a reduction in cavitation, however, care must be taken in the design of a decelerating duct.
-Since the negative thrust produced by the duct must necessitate additional work being done by the impeller, there can be circumstances where the additional loading of the impeller out-weighs the impacts of elevating the pressure around it.
-\cite[24-25]{oosterveld1970} provides a more thorough derivation of the circumstances in which this can be reliably achieved, which generally include larger blade area ratios or with more rows of rotors for higher loading, or circumstances where the loading of the propeller is quite low.
-This creates an inevitable and necessary tension between optimisation of a duct for propulsive efficiency and reduction of cavitation, which will be returned to in subsequent section.
-
-It is perhaps most relevant at this point to further clarify the distinction between the different names for related or similar systems.
-In \cite[17]{carlton2007} pump jet is described as an "interesting development of the classical ducted propeller".
-Oosterveld, in his thesis on ducted propellers also states with regard to the decelerating duct "This ducted propeller system is the so called pumpjet" \parencite[8]{oosterveld1970}.
-Other authors give separate treatement to ducted propellers and pump jets, though the systems they describe are essentially similar in terms of composition, advantages, and physics \parencite[288]{lewis1988}.
-
-\begin{figure}
-\includegraphics[width=\textwidth]{PumpjetOutline.png}
-\caption{In \cite{carlton2007} the pump jet is described as a development of a classical ducted propeller, with the noticeable addition of a stator}
-\label{fig:PumpjetOutline.png}
-\end{figure}
-
-In other definitive literature on the design of pumpjets \parencite{henderson1964} the difference between a pumpjet and a propeller is a described as being that "the stream of flow through the pumpjet is made to depart from the "natural" or free-stream sufrace that bounds the flow though a standard propeller."  According to the diagrams and the subsequent description it is clear that the definition embraces quite exactly the distinction between an accelerating duct, and a decelerating duct, namely that the flow is slowed at the point it encounters the impeller, where as in a free situation the free stream would contract and accelerate at this point.
-
-The addition of some new parts such as stator blades (to straighten out the flow) and other possible features which make the propeller system more like pump or turbomachinery components might be associated with systems more commonly referred to as pumpjets than ducted propellers.
-Not all decelerating ducts have stators, but the impact of the duct on the stream will remain.
-The key concepts behind the operation of these systems, however, remains essentially similar in their nature and design intent.
-In the presence of a duct propellers might be advantageously shaped differently, including maintaing a larger chord lenght right to the shroud, rather than tapering as open propellers tend to lessen vorexs and cavitation on the extremities.
-However, there is not to my knowledge any single defining physical characteristic which the transition to the name 'impeller' implies, which woud not be effectively equivalent to specialised propellers for decelerating ducts.
-In systems tending to be described as pumpjets or waterjets hubs tend to be larger, blade-numbers higher, twist, skew and rake lower.
-No definitive transition point seems to be accepted as having great significance.
-Most commonly literature will refer to a 'rotor', a more general term which could encompass both propellers and impellers.
-
-\subsubsection{Waterjets}
-The use of the word waterjet, as opposed to pumpjet, has a tendency signify that the jet expelled from the shroud or pump is situated above the water, which is desirable in very fast-moving surface vessels which attempt to minimise their contact with the waters surface, including high-speed planing hull vessels, including pleasure craft.
-The physics of the two systems, however, are closely related, with the very noticeable exception of the absence of flow considerations for water on the exterior of the duct considerations for waterjets.
-However, the physics relating to the impeller and stator blades or (turbomachinery) designed to add energy to the flow are generally extremely closely linked, though a larger variety of diffeent pump types can be chosen for waterjets which might differ more significantly from the natural evolution of a propeller to an axial-flow pump.
-Many of the essential equations which govern the efficiencies and cavitation performance of pumpjets and waterjets are of the same or extremely similar form, as can be seen in detailed discussions given to both, by similar authors (\cite{wislicenus1973}, \cite{henderson1964}).
-There are, however, plenty of instances in recent literature where this distinction regarding the ejection of the jet above and below the waterline is not maintained, and waterjets and pumpjets seem to be used interchangeably \parencite{abdel2010}, or systems are described as 'submerged waterjets' \parencite{buckingham2008submarine}.
-
-\begin{figure}
-\includegraphics[width=\textwidth]{PumpjetWaterjet.png}
-\caption{A comparison of a pumpjet and waterjet as given in \cite{wislicenus1973}}
-\label{fig:PumpjetWaterjet.png}
-\end{figure}
-
-\section{Constraints on the efficiency of pumpjets at low speed}
-
-\subsection{Recent Commentary}
-There has been already some attention given to the probable efficiencies of pump jets at low speed.
-Recently Andrew Davies published a piece in the The Strategist (ASPI) arguing that the pump jet may actually save fuel at transit speeds (around 8kt) and overall improve the endurance of the submarine, despite likely having lower efficiency at low speeds such as patrol speed \parencite{davies2017}.
-This argument relies on pump-jet efficiency still overtaking that of propellers at what is considered a relatively low speed for many applications.
-As Andrew acknowledges, further quantification of the likely efficiency is required, as the particular chart which he cites is not quantified.
-This section aims to identify as far as is reasonably possible from public literature how the efficiency of a likely pump-jet for a submarine might perform in quantitative terms, relative to a propeller.
-
-There has also been questions put in parliamentary hearings regarding the pump jet \parencite{patrick2017}, and responses from public officials to such questions indicate that some of the literature provided to support such lines of questions was dated, and that the state of the art may have advanced considerably since this time.
-Whilst the statement that the state of the art is much advanced is obviously true in the case with marine propulsion as with most sciences, it is also equally true that new advances tend to confirm, rather than annul, many of the works that have been previously undertaken.
-To the extent that modern literature still supports the overall constraints and bounds provided by some of the foundational earlier works, a great deal of useful information can still be obtained from earlier works.
-In particular, the foundational theories, often derived from well-established principals of conservation of momentum, energy, as applied by the likes of Bernoulli and Euler, are as well respected and deeply depended upon today as they were centuries earlier when they were first arrived at.
-It is noteworthy that some of oldest works, such as those by Wislicenus and Henderson, are still amongst the most frequently cited in the most recent of papers.
-
-\subsection{Modern literature and results}
-Abundant modern works, including those which involve experimental validation of theoretical results also clearly confirm that restrictions on the efficiency of waterjets are enduring, and give us a clear starting-point for consideration of their probable orders of magnitude, and likely transition points indicating at what speeds pumpjets might become advantageously more efficient.
-It should be noted however, that relatively little literature focuses on the performance of propulsive systems far far outside of their intended design speeds, which tends to be far higher than the patrol and transit speeds of diesel electric submarines.
-Fewer still tend to express their work in terms of ship-speed, instead adopting the convetion of using dimensionless ratios so that work done on one system can be readily translated into different combinations loads, pressures, speeds and sizes.
-However, from a few notable studies exist which provide important reference points.
-(Further interpretation of the what we can lean from the trends expressed in dimensionless coefficients will be discussed later.)
-
-A 1995 Japanese experimental study on the performance of a waterjet demonstrated a close experimental fit to a theoretical relationship which sees the propulsive efficiency of the waterjet fall to zero with craft velocity, with significant declines in efficiency commencing just under 10m/s, or somewhere around 18kt \parencite{fujisawa1995}.
-
-\begin{figure}
-\includegraphics[width=\textwidth]{EfficiencyJapan.png}
-\caption{A graph from a Japanese study of waterjet efficiency shows efficiency rapidly declining below craft velocities of 10m/s, in accordance with theory. \parencite{fujisawa1995}}
-\label{fig:EfficiencyJapan.png}
-\end{figure}
-
-In 2015 an Australian team including staff from the Australian Maritime College and Incat published results of an experimental study comparing the use of waterjets and propellers for medium-speed ferries operating in the 20-30kt speed range \parencite{kamal2015}.
-This study explicitly set out to establish the cross-over point for where propellers become more efficient than jets, since Incat was interested in pioneering highly efficient medium-speed vessels, after having considerable success in very high speed catamarans.
-It is again acknowledged as given that at lower speeds propellers are more efficient than waterjets.
-The study found that propellers were considerably more efficient over the entire speed range tested, often by a factor of two in terms of transport efficiency.
-
-\begin{figure}
-\includegraphics[width=\textwidth]{EfficiencyTransport.png}
-\caption{An Australian experimental study aimed at determining the crossover point for the efficiency of jets and propellers in the 20-30kt range revealed much higher power was required across the entire speed range \parencite{kamal2015}}
-\label{fig:EfficiencyTransport.png}
-\end{figure}
-
-Both of these previous studies confirm the general and widespread assumption that jets are known to be more efficient at high speeds, and propellers at lower speeds.
-They also suggest that differences in efficiency could be very large, of a factor of two or more, including at speeds well over a submarine's transit speed.
-However, care must be taken making precise quantitative comparisons, since such experiements involved jets which expelled into air (waterjets) as opposed to remaining under water (pumpjets) do have at least one significant difference in their makeup, though their internal pump mechanisms may often be similar.
-Specifically, pumpjets would experience drag on the exterior of the shroud/duct, which waterjets would not, however waterjets are likely to have longer intake ducts, which might involve larger duct losses than pumpjets.
-Given that these factors might affect efficiency, and the overall design of the nozzle considerably, more direct comparison with fully submerged jets should be sought to make more direct comparisons.
-
-At IMARESTS 10th International Naval Engineering Conference, a paper was presented by BMT Defence and Rolls Royce \parencite{giles2010}, outlining a proposal for 'fully submerged waterjets' to be used on future surface ships in order to achieve better acoustic performance, particulalry for an anti-submarine warfare role.
-The jets comprise a mixed-flow pump coupled to an electric motor, and have been both extensively modelled and also tesed in demonstrator vessels.
-These may be a closer representation of the performance characteristics achievable by submerged pumpjets, since shape of the intake and shroud more closely resembles the shape of a pumpjet that might be found on a submarine.
-Importantly, this work focuses specifically on the performance of the jets as a function of ship-speed, including identifying possible crossover points.
-
-\begin{figure}
-\includegraphics[width=\textwidth]{BMTShape.png}
-\caption{The 'fully submerged waterjet' proposed by BMT and Rolls Royce for military surface craft may more closely represent the duct shape used on submarines than waterjets that release above the surface.
-\parencite{giles2010}}
-\label{fig:BMTShape.png}
-\end{figure}
-
-The paper makes no effort to justify the selection of choice of jets on the efficiency of the propulsion system, instead focussing on improved acoustic performance, and potential machinery space-saving advantages.
-The vessel under consideration would require 250 tonnes more fuel with jets find than with propellers, the jets would require more power to reach equal waterspeed to a conventionally propelled craft right up to 30 kt.
-In the case of this analysis, modifications to the hullform in order to accommodate the jet (which are likely to be more substantial for a surface ship than a submarine) also have an impcat on the resistance offerd by the hull.
-This also has an interaction with the different shape of the propulsors, which alters the required level of delivered (effective) power to achieve a given speed, in addition to the considerations of how efficiently the propulsor generates that effective power.
-The paper states: "This is an encouraging result for a first iteration as the AWJ-21™ arrangement has not been fully optimised and, hence, there is further prospect of reducing the shaft power for this form.
-If the power could be reduced by 5\% (e.g.
-by fur ther improving the waterjet/hull fairing), then the AWJ-21™ vessel would be better at all speeds above 25 knots."  This statement, however, is made in reference to the hullform needed to accommodate the jet, rather than the jet propulsion efficiency itself \parencite{giles2010}.
-As the propulsive efficiency curves given in the study show, the propulsors themselves are considerably further from reaching parity, still separated by about 10\% at 25kt.
-
-The paper also provides specific curves for the efficiency of the propulsor as a function of vessel speed, which demonstrated that the waterjet efficiency was lower than that of conventional propellers throughout the speed range, but declined particularly precipitously below speeds of 10kt.
-At 5kt, the propeller was more than twice as efficient as the waterjet.
-This study has been used to directly inform the 'central' efficiency curves assessed in this analysis.
-
-\begin{figure}
-\includegraphics[width=\textwidth]{BMTEfficiency.png}
-\caption{Studies on submerged waterjets for military surface craft currently show that the efficiency of jets decline sharply below 10kt \parencite{giles2010}}
-\label{fig:BMTEfficiency.png}
-\end{figure}
-
-It appears from recent examples in the literature that there is no evidence of jets approaching the same efficiency as propellers at speeds as low as 10kt.
-Indeed, there seems to be considerable evidence that jets of most kinds suffer substantial efficiency penalties relative to propellers in this regime, and that this is understood and accepted as an inevitable design trade-off to be made.
-In the case of a submarine, which might spend nearly 100\% of its time travelling at 10kt or less, this disadvantage (as opposed to less than 35\%, as assumed for a surface ship in \cite{giles2010}) would become extremely substantial, which will be quantified in later sections.
-
-Moreover, recent literature seems to confirm that efficiency of pumpjets fall off precipitously towards zero the closer that the vessel speed comes to zero.
-This strongly suggests the existence of some fundamental constraint in the physics of the operation of jets, which recent advances in technology seems not to have overcome.
-To more firmly establish the extent to which adaptation or plausible improvments in technology could or could not change this trend, it is worthwhile considering the theoretical limits that are presented by the physics.
-This should further limit the range of error that might be made in estimating realistic bounds of performance for a system which we do not have precise design specifications for.
-
-\subsection{A simple theoretical explanation}
-
-Pumpjets are necessarily low in efficiency at very low speeds because they attempt to get the water to do multiple things.
-They don't just accelerate a water colum.
-They first slow it down (increasing its pressure), then add pressure to it, then accelerate it.
-When the total amount of energy embodied in the incoming water stream, and the total amount of energy you want to add to the water both get very small (as they do at very low speed) there simply isn't enough energy to do all things at once to all of the water.
-The energy penalty inflicted by using the additional devices becomes very large in proportion, and part of the water doesn't do all the necessary things for the jet to work.
-In essence, if you try to slow down something that's already going slowly, it tends to stop  That's exactly what happens in jets operating at very low flow-rates, when flow separation, also known as stall when it occurs between the blades, begins to take place.
-When there isn't enough energy to move all the water through the pressure-hurdles imposed by the duct, only some of it moves, and the rest spins in place.
-
-There are other some fundamental limitations on the peak efficiency that can be achieved by jets based on simple calculations of ideal efficiency from momentum theory, which are discussed in the next subsection.
-They also provide enduring, reliable limitations on the extent to which ingenuity or design changes might be able to improve the efficiency of jets at low speed.
-But the primary reason why jet efficiencies decline towards zero at low speed is to do with flow separation \parencite{fujisawa1995}, which is what occurs when the assumptions inherent in momentum theory's analysis of an 'ideal' actuator disc start to break down.
-
-The conditions which lead to flow separation in a pumpjet can be relatively simply explained in terms of conservation of energy, and conservation of mass.
-Conservation of mass requires that as a pipe or duct expands, the flow through it necessarily slows down, in order to occupy the large volume, in proportion to the new cross-sectional area.
-Conservation of energy requires that it increases in pressure, and defines a very specific relationship between the pressure increase and the change in velocity, which is determined by the shape of the duct.
-That relationship says that the static pressure increase is inversely proportional change in velocity \textbf{squared}, as given by Bernoulli's equation, which was discussed in earlier sections.
-As a result, at high speeds, the total energy of the system is dominated by its velocity, but at very low speeds the static pressure is much more significant.
-
-It also means that for any given change in velocity, the static pressure changes also, but doesn't change as much as the velocity change.
-The consequences for this pumpjets are and inevitable, since the geometry of the duct is inevitably fixed.
-At very high flow rates, the expansion of the area ahead of the impeller creates an adverse pressure gradient, as the water must slow down to fill the greater area.
-This adverse pressure gradient is larger at higher speeds, the proportions aren't linear, hence the the speed is much much larger, due to the square relationship, and flow is consistent (though still turbulent) throughout the entire cross-sectional area of the duct.
-
-At low speeds, the same effect leads to detrimental effects.
-The adverse pressure gradient is much reduced, but the velocity is much much $(much)^2$ reduced.
-The pressure gradient is necessarily constant across the entire area radial area of the duct, but the drag due to the boundary with the duct itself will necessarily be greater closer to the walls in the case of the intake of the duct.
-For this simple reason, as the flow rate decreases, some point is reached where the flow adjacent to the walls actually reverses, and the flow becomes separated.
-
-The same phenomenon is also prone to occur near roots of the blades of the rotor \parencite[15]{henderson1964}.
-Necessarily, the blades of the rotor are doing work on the fluid, so unless the impeller is in a region of considerable contraction, pressure will necessarily have to increase significantly through the blades, providing an adverse pressure gradient, which again must be constant radially.
-Since closer to the hub the velocity of the blades through the water is lower, the blades tend often to also be longer and deflect the flow even more in order to do the same amount of work on the fluid.
-With the longer surface over which the fluid must travel, as well as greater angle of deflection, flow separation is prone to occur here too.
-The physics will be exactly analogous to that of the separation on the inside of the duct intake, except the pressure gradient will be provided by a different source (the rotor doing work), and the surface (the blade roots) will also be different.
-This will also mean that the direction of rotation of the vortices will be different, and the location also close to the middle of the jet, instead of the outside.
-
-As can be seen in figure \ref{fig:JetHighFlow.png}, the duct of a jet is designed to operate by deflecting the flow away from the natural path that would occur for a propeller, which would accelerate the stream over its blades directly, with the flow necessarily contracting with the acceleration as shown in Figure \ref{fig:PropellerAction.png}.
-Instead, pumpjets tend to use some moderate expansion of the flow in advance of the impeller (diffusion) in order to raise its pressure, and reduce its velocity.
-(Allowing the blades to do work on the water at lower speeds and higher pressure is the essence of what reduces cavitation.)  The blades then do their work on the liquid, further raising its pressure.
-The duct allows this pressure then to be converted back to kinetic energy, as the water accelerates to move through the contracting area in the nozzle.
-This difference in the shape of the flow, and the consequent impacts on the static pressures, are demonstrated in Figure \ref{fig:FlowComparison.png} taken from \cite{mccormick1963design}.
-
-\begin{figure}
-\includegraphics[width=\textwidth]{FlowComparison.png}
-\caption{The effect of the duct is to deflect the flow away from the streamlines of an open propeller, elevating the static pressure at the stator blades by slowing the flow down.}
-\label{fig:FlowComparison.png}
-\end{figure}
-
-Consequently, the geometry of the duct defines a pressure gradient in the intake and through the nozzle.
-The greater the difference in the cross-sectional areas, the greater the total magintude of the pressure differential that will occur, as necessitated by conservation of mass.
-The same amount must go out as came in.
-As the area through which it passes changes, the velocity much change so that the flow rate remains constant, otherwise cavitation will occur.
-The rate at which the areas change (how rapidly they expand and contract along the axis) will determine the gradient.
-Making the pressure change as smooth as possible is an important aspect to jet design in order to minimise flow separation.
-
-\begin{figure}
-\includegraphics[width=\textwidth]{JetHighFlow.png}
-\caption{At high flow rates the water has sufficient velocity (indicated by arrow length) to overcome the adverse pressure gradients created by the duct shape, shown by the change from green to red.
-After the impeller has added further pressure to the water, the water accelerates out of the contracting nozzle through the favourable pressure gradient (red to green) to produce a jet, reaching equal pressure to outside at the exit}
-\label{fig:JetHighFlow.png}
-\end{figure}
-
-Figure \ref{fig:JetLowFlow.png} shows a jet operating at very low speed.
-The adverse pressure gradient combined with the drag forces of the boundary layer are sufficient to bring the flow to a stop and reverse it, as shown by the arrow size reducing, and changing direction in the diagram.
-This occurs very close to the inner surface near the opening of the duct, and potentially also between the blades of the rotor, beginning the roots \parencite[15]{henderson1964}.
-Flow separation necessarily sees additional energy expended in the kinetic energy of the additional vortices and turbulent flows created, which don't result in the creation of thrust, which demands a change in the net momentum of the liquid in a backwards direction.
-Consequently, the degree and extent that it occurs necessarily and inevitably reduces the efficiency of the propulsor.
-As the flow rates slow further, the areas of flow separation will become larger, effectively reducing the intake area of the jet, and further degrading efficiency.
-
-Flow separation also tends to lead to unsteady flows, with vortices periodically being shed into the main stream as they are speed up or change shape, and then ingested by the impeller and stator blades.
-As can be seen by the diagram in Figure \ref{fig:JetLowFlow.png}, regions abound which are adjacent to or behind flow blockages or reversing flow, where considerable turbulence and unsteady forces would exist.
-These effects also have acoustic consequences, which will be discussed in Section \ref{Acoustic Considerations}.
-
-\begin{figure}
-\includegraphics[width=\textwidth]{JetLowFlow.png}
-\caption{At low flow rates the water has insufficient velocity to overcome the adverse pressure gradients created by the duct shape, and flow separation begins to occur on the duct intake, and between the blades beginning near the roots (stall).
-Flow through the jet would be unsteady, with excessive energy expended creating vortices and turbulence that results in little thrust.}
-\label{fig:JetLowFlow.png}
-\end{figure}
-
-It is worthwhile noting that the diagrams in Figures \ref{fig:JetHighFlow.png} and \ref{fig:JetLowFlow.png} do not include a hub for the sake of simplicity.
-In practice, the cross-sectional are of the pumpjet naturally only includes that aread between the hub and the duct, and in the case of many pumpjets, particularly those on the ends of torpedoes or submarines, the hub may include/incorporate the tapering tail of the body, for example in Figures \ref{fig:PumpjetOutline.png} or \ref{fig:PumpjetWaterjet.png}.
-Particularly in Figure \ref{fig:PumpjetWaterjet.png}, the appearance of the inner side of the shroud to roughly follow contraction of the hub or body gives the optical appearance of very little diffusion or contraction of the flow occurring, as the one dimensional separation might seem to be roughly constant, or even expanding in places where I have earlier explained it generally contracts.
-It is important to consider the impact of the second dimension on the actual effective area.
-Since area of a disc follows $\pi r^2$, the impact of the contraction of the hub will be dramatically larger when it is large than it will be when it is small.
-
-Consequently, the contraction of the outer shroud at the very end of the duct will dominate the final taper of the hub, and actually lead to substantial restriction of the flow.
-Similarly, where the body or hub is quite substantial and also tapering steadily, keeping the duct intake parallel to the axis of the body for a small distance, or even just tapering it proportionally less than the hub, can still result in a considerable expansion in the area, and considerable diffusion.
-The example in Figure \ref{fig:PumpjetOutline.png} shows a case of extremely sharp flow diffusion on the intake, probably more than what is practially realistics to avoid flow separation, even at quite high speeds.
-It is important to recognise that this simple relationship with $r^2$ generally means that even moderate changes in terms of diameter, which is what we tend to observe when looking at images and diagrams can actually amount to much more substantial changes in velocity, and hence pressure, which drives the effects described in this section.
-A side-by-side comparison of the geometries and flow-lines likely in the case of pumpjet or propeller is give in Figure \ref{fig:Comparison.png}, from  to help illustrate this point.
-
-\begin{figure}
-\includegraphics[width=\textwidth]{Comparison.png}
-\caption{At the end of a tapering body the impact of the contracting hub will be much larger near the intake than near the tail, making the degree of diffusion ahead of the impeller, as well as contraction of the flow in the nozzle after it, much larger in impact than it appears by looking at the external shape.
-Credit \cite{gearhart1966selection}}
-\label{fig:Comparison.png}
-\end{figure}
-
-\begin{figure}
-\includegraphics[width=\textwidth]{Triomph.jpg}
-\caption{The outer shape of pumpjets on the aft end of torpedoes or submarines, such as this one on the French Triomphant class on which the Barracuda's jet will be based, may understate the degree of deceleration that the duct achieves.
-Credit \cite{reddit2016}}
-\label{fig:Triomph.jpg}
-\end{figure}
-
-\subsection{Trade-offs and Limitations on Improvements}
-
-Having established that the inevitable decline of a pumpjets efficiency towards lower speeds is driven by the increasing occurrence of flow separation, some consideration ought to be given to the potential scope for the effect to be improved or negated by design changes, or otherwise mitigated with other improvements in efficiency.
-Here I have found considerable help from well established literature, many including various old declassified technical reports funded by the US Navy, many of which are frequently cited in the most recent literature as laying out the enduring frameworks for the modern discipline.
-
-It is clear that the optimisation of a system for minimising cavitation, and hence improving acoustic performance, is a separate and independent design requirement to that of optimising a system for propulsive efficiency.
-These two design requirements can be found either to be in tension, or alignment, in different regimes.
-At high speeds, where cavitation tends to have severe impacts on propulsive efficiency of open propellers, and high mass-flow (which leads to good efficiency) can be achieved through compact ducts and pumps, pumpjets offer very good performance on both metrics.
-Hence, in this regime the two design requirements tend to align relatively well in favour of pump-jets.
-However, in very low-speed, low-load environments, where cavitation effects are negligible, and larger areas need to be engaged achieve good mass-flow and efficiency, the two design requirements align relatively well in favour of an open propeller.
-In a variety of regimes (including at quite high speeds), the two requirements find themselves in tension and needing to be balanced \parencite{gearhart1966selection}.
-
-For this particular consideration, two questions are of paricular importance.
-The first is where the clear transitions in the regimes will tend to lie, with respect to the probable requirements of a conventional submarine.
-The second is what drives the tension between those design requirements in the intermediate regimes, and whether it is a fundamental or inescapble tension in the laws of physics, or simply a technical hurdle which imminent improvements in the state of the art are likely to overcome by engineering improvements or innovation.
-
-The answer to the first question, as discussed in earlier sections, seems to be that the operating speeds of a submarine tend to lie far below the probable levels where pumpjets are even remotely comparable to waterjets for efficiency.
-In my research I have found no evidence of waterjets being advocated for speeds less than 30kt, and for pumpjets for speeds less than 27kt.
-Having consulted with leading academics in the field, I've had it strongly confirmed that a pumpjet will always be less efficient than a propeller at speeds under 18kt.
-In later sections I will address further the acoustic comparisons at low speeds where flow separation will begin to occur, for instance below 10kt.
-
-The answer to the second question is that the tension between acoustic and propulsive performance at intermediate speeds and loadings is driven by foundational physical requirements.
-That is not to say that optimal means of balancing those requirements will not still be found, and improved, as the art advances.
-However, it does mean that significantly adapting using known variables that can be changed to favour one design requirement are unlikely to yeild large improvements without some trade-off being made to the other.
-
-The reason for this trade-off occurring can be quite simply attributed to the nature of the shroud, or duct, which surrounds the the rotors of the propulsor, and the effect it has on the flow.
-If the duct acts to increase the pressure at the rotor, (which reduces cavitation) it necessarily must slow the flow at/around the rotor, which necessitates produceing a negative thrust on the duct.
-The system as a whole produces a positive thrust, becuase the impeller or rotor pushes a larger positive thrust, but the duct necessarily produces a negative thrust.
-As the speeds get lower, and the thrust required gets much much lower, this negative thrust becomes proportionally more significant, as described later.
-
-Put another way, the work required to slow the water down, then speed it up again, becomes a gradually greater penalty the the slower the water is coming in, and the less you actually want to speed it up.
-If the water is slowed enough by the duct to effectively stop, then the impeller must do the work of speeding it up (sucking it into the duct) before the duct slows it down, before the impeller raises its pressure, and the pressure accelerates it out of the nozzle afterwards, immediately reducing even the ideal efficiency to 50\% \parencite[277]{lewis1988}.
-Compared to simply accelerating the water in one simple push, as a propeller blade does in an unconstrained flow, the efficiency penalty is substantial, and larger in proportion at low speeds.
-
-\subsection{The impact of duct loss on an ideal propeller}
-
-Derivations of the 'ideal' efficiencies are frequently made for propellers and jets using basic conservation laws (momentum theory) which consider simple acceleration of some column of water as actuated by a disc representing the impeller or propeller.
-A fuller description of these equations can be found in  most standard texts \parencite[131]{lewis1988}.
-These methods are computationally simple, and allow simple upper-bounds to be be set for efficiencies achievable for given designs under certain conditions, without considering the pottential impact of other more complex effects such as skin friction over the blades, potential flow separation, or cavitation.
-As such, they can be easily relied upon to constrain the overall plausible range of values that can be achieved from universally applicable inputs, such as the velocity of water going into a duct, and the velocity of water exiting a nozzle.
-
-Such methods are used in \cite[8-14]{wislicenus1973} to construct an 'ideal' jet efficiency that applies to waterjets and pumpjets.
-Whilst I won't repeat the full expansion here, a fundamentally important early conclusion of the derivation is the distinction between 'useful work' and the total energy expended.
-Since the rate of change in momentum is known to be necessarily equal to the force generated, and increasing momentum change (hence thrust) can be achieved either accelerating more mass, or accelerating mass to higher speeds.
-Given that energy embodied in a flow is proportional to it's velocity squared, but its momentum only to its velocity, a necessary and inevitable consequence is that some energy is always wasted in accelerating a fluid to a higher velocity than its surrounds.
-
-This is expressed in the equation which represents 'ideal jet efficiency', which is directly equivalent to 'ideal propeller efficiency' using momentum theory of propeller action:
-
-\begin{equation}
-\label{eq:3}
-\eta_j = \frac{1}{1+\frac{\Delta V}{2V_0}}
-\end{equation}
-
-The inevitable consequence of this relationshiop is that efficency approaches unity when $\Delta V$ (the change in water of the water between entering and exiting the system) relative to $V_0$ approaches zero.
-Intuitive embodiments of this relationship are (equivalently) that efficiency is maximum when the least force is exerted on any given bit of water, which might be when negligibly small work is being done (extreme low speed) or when such a large mass of water is acted on that it barely needs to be pushed at all.
-This relationship would lead to the conclusion that infinitely large propellers or jets are always desirable from an efficiency perspective.
-This of course isn't practical to build, and also doesn't hold when taken to the etremes the effects of the drag on the propeller blades, or duct, are taken into account.
-Infinitely large systems would increase the total weight, as well as having significant surface drag when moving through the water, all of which are neglected in this 'ideal' consideration.
-
-In \cite[8-14]{wislicenus1973}, Wislicenus expands the relationship in the case of the jet to incorporate terms to represent some of the inevitable adjustments that would occur in the case of pump jets and waterjets, particulalry the duct head loss ($K$) (fraction of energy required to push the water through the duct), as well as the additional drag on the outside of the propulsor ($\Delta T$).
-This leads to an adjusted efficiency relationship, which still neglects any losses due to pump efficiency, or change in height of the jet (which is added for consideration for waterjets, but not relevant here).
-
-\begin{equation}
-\label{eq:4}
-\eta_j = \frac{1}{1 + \frac{\Delta V}{2V_0} + K\frac{V_0}{2 \Delta V}}(1 - \frac{\Delta T}{T})
-\end{equation}
-
-The conclusion of the analysis, and the plotting of these results for certain possible values of $K$ and $\Delta T$ is: "Its outstanding characteristic is that this efficiency approaches zero rather than unity for $\Delta V / V_0 = 0$, even for small duct losses".
-
-\begin{figure}
-\includegraphics[width=\textwidth]{PumpjetEfficiency.png}
-\caption{Ideal jet efficiencies sharply appoach zero at low values of $\Delta V / V_0$, where as ideal propeller efficiencies without a duct $K = 0$ approach unity. Credit: \cite{wislicenus1973}}
-\label{fig:PumpjetEfficiency.png}
-\end{figure}
-
-Wislicenus also acknowledges \cite{wislicenus1973} another author, Brandau, who makes a different assumption about the relationship between the duct loss and water velocity \parencite{brandau1967} in his derivation (along with another thorough discussion of potential and ideal efficiencies), who produces a different peak efficiency point, but retains a similar overall shape featuring efficiency falling very rapidly to zero at low values of $\Delta V / V_0$, and contrasting this with the ideal propeller, which approaches unity in this low regime.
-
-Charts such as these are generally used with dimensionless ratios on the axes (such as $\Delta V / V_0$) to allow the most general solutions to be used to solve for any given particular vessel size, speed, or loading requirement which might be required, these  confusing to many readers unfamiliar with hydrodynamics.
-For the avoidance of any confusion, it is important to note that the value for $\Delta V / V_0$ for a given vessl and propulor may tend to increase with increasing waterspeed of the vessel, however one cannot rely upon this particular value diminishing towards zero at very low speeds, as it does for related by different metrics.
-In an idealised circumstance, the flow rate increases directly with vessel speed.
-Given that the force produced scales with $\Delta V$, which at constant values of $\Delta V / V_0$ increase in proportion to (if, in the idealised circumstance, vessel speed is equal to $V_0$), an increase in thrust can theoretically match the $V_0^2$ requirement imposed by drag without any change in $\Delta V / V_0$.
-However, in practice the water encountered by the propulsor is slowed by the wake field, and so the drag encountered by the vessel may increase faster than velocity of the water entering the intake, an effecnt which is likely non-linear with vessel speed.
-So whilst the trend with regards to vessel speed can be known with regards to the graphs in \cite{figLPumpjetEfficiency}, a direct translation of the x-axis used to vessel speed at any given point is difficult prove without additional system information.
-One might generally infer that the sorts of declines in efficiency observed as vessel speed discussed in recent literature correspond with movement towards very low values of $\Delta V / V_0$, though it is possible that alternative derivations have been used, or that the phenomenon known to occur in practice (such as flow separation) aren't adequately captured by such ideal equations derived from momentum theory in order to make a direct translation into real-world application.
-
-For context, the equations developed by Wislicenus are used for the purposes generally of selecting the ideal propulor diameter, which can be made larger in the case of a propeller to access higher ideal efficiencies.
-This is an opportunity which is constrained in the case of jets because the efficiency diminishes at very low values of $\Delta V / V_0$.
-This is quite a different constraint to other "practical" considerations as some have suggested, such as weight, or space constraints.
-Open propellers may advantageously choose to lower their values of $\Delta V / V_0$ as low as 0.1 in order to obtain ideal efficiencies \parencite{wislicenus1973}.
-This still provides a hard limit on potential improvements that can be attained for a pump-jet that has non-zero duct loss.
-
-\begin{figure}
-\includegraphics[width=\textwidth]{EfficiencyBrandau.png}
-\caption{Alternative derivations of ideal jet efficiency also show sharply declining efficiency for low values of $\Delta V / V_0$ Credit: \cite{brandau1967}}
-\label{fig:EfficiencyBrandau.png}
-\end{figure}
-
-The structure of these curves is found to be overall quite similar for both pumpjets and waterjets, though the probably values for the losses incurred in the intake duct, or on the exterior of the shroud, are obviously different \parencite[13-14]{wislicenus1973}.
-Similar efficiency shapes can be found for discussions focussing explicitly on underwater pumpjets \parencite[13]{henderson1964}.
-It should be noted that the theoretical efficiency derived by these general equations is consistent both in derivation and results with those produced by other authors \parencite[227]{lewis1988}.
-
-Care should be taken, however, to note that different graphs for efficiency can be produced which have different variables on the x axis.
-For example, \cite[247]{mollard2011} a graph of ideal efficiency of a pump jet is shown using the speed of advance divided by the jet speed, or $V_1/V_2$.
-Given that the x axis used in the previous charts used $\Delta V / V_0$, which would be related to $V_2-V_1$ in this case, this has the effect of reversing the effective direction of the chart and changing its shape.
-In this case, for a fixed jet and vessel design, high-speed operations will be towards the left of the chart, and the precipitous decline in efficiency will occur as speed declines to zero, along with thrust, toward the right of the chart, which can be known to correspond to waterspeeds declining to zero at $V_1/V_2=1$.
-
-\begin{figure}
-\includegraphics[width=\textwidth]{EfficiencyMollard.png}
-\caption{Other modern reference texts might demonstrate jet or propeller efficiencies with different x-axes, effectively reversing the way the graph should be understood for a single system operating at different speeds.
-Credit:  \cite{mollard2011}}
-\label{fig:EfficiencyMollard.png}
-\end{figure}
-
-\begin{figure}
-\includegraphics[width=\textwidth]{EfficiencyBulten.png}
-\caption{Another graph representing efficiency as a function of the ratio of jet velocity to incoming velocity, which shows efficiency declining at low thrust (to the right).
-The impact of increasing duct losses can be clearly seen, as opposed to the 'ideal' propeller without a duct.
-Credit: \cite{bulten2006numerical}}
-\label{fig:EfficiencyBulten.png}
-\end{figure}
-
-These types of charts correspond more closely in orientation and trend that the very frequently used x-axis unit $J = V_a/nD$ when showing efficiency curves for both propellers and ducted propellers (such as used in Figure \ref{fig:OpenWater.png}) are of a similar orientation to the chart shown in the figure above.
-The left side of the chart will correspond to the highest number of rotations for a given amount of progress through the water.
-For vessel of fixed load, and a propeller of fixed geomtry, this will correspond to a propeller facing the higest drag force due to the vessel moving at the highest speed.
-
-Readers may observe that the efficiency for all propellers also decline to zero at the other end of the spectrum, most sharply at a point where $J$ equals or exceeds the pitch ratio of the propeller.
-As discussed earlier, these correspond to cases where the propeller is slicing along its helical path without exerting any force to move water backwards, or when the slip ratio is zero.
-In this circumstance, the rotation of the propeller would not push any net water backwards, and simply act as a paddle-wheel needlessly spinning water.
-Such circumstances are unlikely to occur to a significant degree unless a vessel was decelerating, or experienced some other force to assist its propulsion.
-It is possible that at extremely low speeds a propeller could somewhat descend to the right of optimal advance ratio, when very dramatically reduced load was encounterd because of the extreme low speed, as might be suggested by the slight reduction in propulsive efficiency of the propeller below 5kt in the BMT study.
-These losses of efficiency, however, could never supercede those of a decelerating ducted propeller or jet, which has the shroud which necessarily imposes an \textbf{additional} drag to the watercolumn in excess of any other resistance faced by the rotor.
-This effect can be seen clearly even or an accelerating duct, as shown in \ref{fig:PropDuct.png}.
-
-\begin{figure}
-\includegraphics[width=\textwidth]{PropDuct.png}
-\caption{Even the addition of an accelerating ducts substantially reduces efficiency of propellers at very low loadings, or for constant loads, low speeds.
-Maximum attainable efficiency is also reduced.
-Their benefit lies in the bulging of the curve at low advance ratios (higher loads).
-Credit: \cite[6]{willemsen2013}}
-\label{fig:PropDuct.png}
-\end{figure}
-
-It should be clear from the above analysis that the efficiency of a pumpjet falling towards zero with a vessel's waterspeed is an enduring and foundational result of basic physics.
-The addition of negative thrust on a decelerating duct, no matter how small, requires that efficiency must decline at some point with with blade loading, in contrast with an ideal propeller which rises to unity.
-This explains the overall shape given by the the mysterious unquantified chart cited by Andrew Davies \parencite{davies2017} and Access Economics \parencite{stanford2017} shown \ref{fig:PumpUp.png}, and is also consistent with the shapes plotted by more recent studies as shown above, and those presented in other modern publications studying the use of pumpjets \parencite{giles2010} and waterjets \parencite{fujisawa1995}.
-
-\begin{figure}
-\includegraphics[width=\textwidth]{PumpUp.png}
-\caption{The mysterious efficiency comparison chart which has been doing the rounds in the Australian debate probably quite reasonably reflects theoretical physical limits of 'ideal' pumpjets and propellers.
-Credit: \cite{stanford2017}}
-\label{fig:PumpUp.png}
-\end{figure}
-
-In terms of establishing plausible peaks and cross-over points, the study undertaken in \cite{giles2010} probably provides a sufficient indication as to the current state of the art in the relevant technologies.
-Whilst this effort focussed on a surface vessel rather than a submarine, the study was undertaken with the backing of military customers with support and testing facilities (the US Navy), and involved two major companies who are world leaders in maritime engineering (BMT) and waterjet manufacture (Rolls Royce).
-Even if further technical advances through design refinement were possible, as the authors alude to, the cross-over point in terms of efficiency might be brought down to 25kt.
-The claim that the crossover point lies anywhere near the transit speed of a convetional submarine (8-10kt) appears implausible.
-In discussions with leading world experts involved in the development of advanced ducted propellers (including those involved in citations) I was told that a pumpjet could not be more efficient than a propeller below 18kt, confirming this conclusion.
-
-With the optimum efficiency of a jet probably being no greater than 60\% or 70\% at its peak efficiency according to many theoretical and experimental studies \parencite{lu2016cfd}, which would quite probably lie considerably higher in terms of waterspeed even than 18kt.
-These facts, along with the inevitable decline of efficiency towards zero as waterspeed also declines, offer a suitably constrained set of plausible efficiency curves may be investigated numerically, as will be done later in later sections.
-
-\subsection{An assessment of scope for improving the efficiency of a pumpjet at low speeds}
-Before continuing to assess the impact of the likely efficiency curves of pumpjets relative to propellers, it is worthwhile considering what the known literature about the theory of pumps, jets, and turbomachinery might tell us about the potential for a radical advance which make might substantially improve the shape of the curve at very low speeds.
-Whilst the field of propulsor design is no doubt complex, and I will not attempt to be exhaustive in detailing all of the potential design considerations which might make an improvement.
-However, the pursuit of reasonably comprehensive confidence is aided by the the scale of the gap which probably needs to be closed to make pumpjets comparably efficient to propellers at the relevant speeds.
-Minor adjustments and the improvement of small or marginal effects, of which there might be many, will clearly not help in shifting the efficiency peak of a jet from over 25kt to under 10kt.
-In other words, this gap will not be closed or noticeably reduced by 'tuning'.
-Quite substantial redesign of core characteristics (like mass flow) would be required, which would necessitate a cycle of redesign and tuning virtually all other aspect of the entire jet, as illustrated by numerous guides to the design of such systems (\cite{bruce1974}, \cite{furuya1988}, \cite{mccormick1963design}, \cite{henderson1964}).
-As such, I will only consider those parameters which could plausibly be altered quite substantially, where the consideration of larger changes make the likely limits or trade-offs quite clear.
-
-\subsubsection{Increase mass-flow by widening area of intake}
-Perhaps the most obvious means by which one might attain substantial increases in efficiency is to increase substantially the mass flow rate.
-This is derived from a fundamental relationship, as expressed earlier in the representation of ideal propeller and jet efficiency (prior to the consideration of any drag or other losses) which was given in the earlier equation.
-Since excessive velocity differences result in more energy expended than momentum changed, as given by momentum theory and basic mechanics, doing work on a large body of water, rather than working harder on more of it, is generally desirable from an efficiency perspective.
-Considering what the appropriate mass-flow should be at design speeds is generally the first thing that is specified to occur in all design processes for pumpjets, and good example of which is in \cite[6-7]{henderson1964}.
-
-As noted in \cite{henderson1964}, the key constraint of this variable is that increasing the diameter of the pumpjet necessarily leads to an increase in the surface (including the external suface) of the shroud.
-The increased drag that this induces must be deducted from the thrust that the pumpjet produces, or effectively increasing $\Delta T$ as given in earlier equations.
-In this respect, there is "no free lunch" to be had in expanding the size of the intake.
-As Henderson et al.
-conclude, for this reason the diameter of a pumpjet tends to be about 15\% or 20\% smaller than that of a corresponding propeller, whereas the diameter to have equivalent mass-flow, highlighting the tension set up in pumpjet design with their earlier observation that "if the cavitation resistance of the blades is to be improved with no reduction in thrust and the same rate of flow through the propulsor, the pump jet rotor must be larger than that of the propeller." \parencite[1]{henderson1964}
-
-It should also noted that in order to substantially increase the opening area of the propeller, the difference between the velocity of movement at the blades nearer to the root (close to the hub) and the extremities of the blades increases substantially.
-In order for equivalent work to be done on all of the water distributed radially across area of the impeller, quite a different blade shape might be required, in order to provide a higher deflection of the water close to the root.
-In these areas, there is increase probability of flow separation occurring, where the velocity of movement of the water across the surface is insufficient to overcome the pressure gradient required to move over the blades.
-(In a pump, the pressure necessarily increases as work is done on it.)  Or, since there isn't enough pressure exerted on the whole of the water column to overcome the resistance imposed by the duct, not all of the water moves forward, and some starts to move backwards.
-This leads to flow separation, or stall occurring close to the blade roots, which dramatically diminishes the efficiency of the system.
-A further discussion of this effect can be found in \cite[15,27]{henderson1964},  \cite[807,8012]{bruce1974},  \cite[185]{wislicenus1973}, \cite[60]{mcbride1979} and many other related texts.
-This is a general problem that affects a range of axial-flow turbomachinery, particulalry operating in off-design conditions, and will present a general concern for the design of a new pumpjet \cite[185]{wislicenus1986}, \cite{li2013}.
-
-\begin{figure}
-\includegraphics[width=\textwidth]{stall.png}
-\caption{Flow separation, or stall, can occur when the flow over the blades of a pump slows too much, as shown in \cite{li2013}}
-\label{fig:stall.png}
-\end{figure}
-
-\begin{figure}
-\includegraphics[width=\textwidth]{backflow.png}
-\caption{Stall results in water flowing backwards near the hub of the blades at low flow rates, as shown in \cite{li2014rotating}}
-\label{fig:backflow.png}
-\end{figure}
-
-It is possible to limit the variation of deflection between the blade tips and roots by simply expanding the size of the hub.
-However, this necessarily reduces the effective frontal area that is available to receive the mass-flow, which necessitates further expansion of the outer diameter, and a very substantial increase in the weight of the system.
-In addition, expansion of the hub will necessarily increase the total surface area relative to the mass flow, including by increasing the amount of tapering required after the stators in order to prevent risk of cavitation occuring around the tail the hub.
-Consequently, increasing the area is not plausible without also increasing duct loss to some degree, consistent with the earlier observations from \cite{henderson1964}.
-
-Increasing the frontal area of the intake also is necessarily linked to the advance ratio of jet.
-Since low advance ratios correspond to higher shaft speed for a given forward velocity, increasing the thrust produced for a given rotation (by widening the jet area) necessarily results in higher advance ratios, since more work is done on the water for a given rotation.
-However, this also necessarily results in poorer cavitation performance of the jet, since higher blade velocities relative to the water flow result in necessarily increased propensity for cavitation.
-A detailed discussion of this can be found in \cite{gearhart1966selection}, but similar tensions can be readily found in most other pieces of literature, including \cite{henderson1964}, \cite{wislicenus1973}.
-
-Given the fundamental nature of the flow-rate and intake area to the design of pump-jets, and the great lenght of time for which they have been understood and discussed as a crucial design input for a waterjet, it is altogether implausible that some large increase in intake area can be achieved without incurring necessary and unavoidable tradeoffs.
-
-\subsubsection{Reduce degree of diffusion (i.e. switch to accelerating duct) to reduce negative thrust from duct}
-A further means of reducing the negative thrust produced by a duct may involve significantly reducing the degree of diffusion, or compression that occurs in advance of the rotor, since the process of diffusion necessarily slows the water and produces negative thrust.
-However, this amounts to transitioning towards an accelerating duct, or Kort Nozzle.
-Whilst nozzles of this type have been known to improve the cavitation characteristics of a propeller by enclosing the tips (which often tend to cavitate first), their ultimate function is to actually decrease the pressure at the blades of the propeller under higher loadings.
-Consequently, they necessarily increase the propensity of other types of cavitation to occur across the blade-face, including bubble, cloud and sheet cavitation.
-As such, these types of propellers are not considered for the military purposes which require an absolute minimum of noise to be created.
-Transitioning to an accelerating duct signify a shift away from a pump-jet, and necessarily be in direct tension with the need to minimise cavitation for acoustic considerations.
-
-In addition, whilst it is often demonstrated that for a given propeller under certain load conditions (high loading), the addition of an accelerating duct can considerably improve efficiency, all of these situations have generally been in the context of vessels which are designed for a relatively high thrust, low waterspeed situation, when propeller size is necessarily greatly constrained.
-This is the case fishing vessels, or for tug boats, which need to impart significant thrust to a vessel dozens or hundreds of times their own size.
-Mounting a propeller the equivalent size of the vessels they mean to push, however, would be excessively heavy, arranging a mount and power drive that would keep the entire propeller submerged, and not have the effect of flipping the boat backwards or sideways when rotated hard, would be effectively impossible, without proportionally increasing the size of the tug.
-Furthermore, in most circumstances these propellers are still designed to operate with some cavitation, even if the extent of it is reduced by the presence of the duct \parencite{haimov2010}.
-
-In other words, the use of highly loaded propellers in such applications is actually particular to design constraints that are unlikely to be present for a submarine, which can position the propeller at the tail of its body, directly behind its centre of mass and drag, and consequenlty have freedom about the optimal size of the blades.
-It is well known that a duct cannot improve the performance of a low-loaded propeller which would be the situation for a slow-moving vessel with unconstrained propeller size.
-Authoritative references in the literature include \cite{oosterveld1970}, who calculates the minimum thrust coefficient $C_T$ (a measure of loading) for an accelerating duct to improve a propellers performance to be between 1 and 2, as shown in \ref{fig:AcceleratingIdeal.png}.
-
-\begin{figure}
-\includegraphics[width=\textwidth]{AcceleratingIdeal.png}
-\caption{Calculations of 'ideal' efficiencies of accelerating ducts (neglecting drag) in \cite{oosterveld1970} show that a duct can only improve the efficiency of a propeller if it is loaded above a certain degree, and that longer ducts produce better improvements in 'ideal' efficiencies}
-\label{fig:AcceleratingIdeal.png}
-\end{figure}
-
-A parallel design change in the switch away from the degree of deceleration in the duct would be the reduction of the nozzle contraction after the rotor.
-For waterjets and pumpjets of all sorts, based on fundamental disc actuator theory, the degree of this contraction of the flow area has also been long identified as a means to increase pressure at the rotor, and hence decrease cavitation, but at the direct expense of efficiency \parencite[14]{van1996waterjet}.
-
-Numerous efforts to find suitably compromised duct shapes exist, but even advanced modern designs still exhibit declining efficiency at low speeds, as well as some cavitation during operations at design loads, which still tend to imply relatively high speeds, for example 27kt in \ref{fig:FineOptimised.png} .
-Consideration of accelerating duct designs for a military submarine would be extremely unlikely, and be a profound departure from any existing pumpjet designs.
-
-\begin{figure}
-\includegraphics[width=\textwidth]{FineOptimised.png}
-\caption{Extensive work has been done exploring compromise designs for decelerating ducted propellers (pumpjets), including with very modest deceleration in the intake, limited shroud taper, short duct lengths, or essentially incporate some of the advantages of accelerating ducts.
-Credit: \cite{abdel2010}}
-\label{fig:FineOptimised.png}
-\end{figure}
-
-\begin{figure}
-\includegraphics[width=\textwidth]{FineEfficiency.png}
-\caption{In experimental testing such designs still exhibit sharply falling effiency at high advance ratios, or low speed. Credit:  \cite{abdel2010}}
-\label{fig:FineEfficiency.png}
-\end{figure}
-
-\begin{figure}
-\includegraphics[width=\textwidth]{FineCavitation.png}
-\caption{TAt optimally efficient conditions (corresponding to 27kt) such designs still exhibit some cavitation. Credit: \cite{abdel2010}}
-\label{fig:FineCavitation.png}
-\end{figure}
-
-\subsubsection{Reduce shroud length to decrease drag on duct}
-The reduction of the length of the shroud in order to reduce drag presents another possible means to lower the duct resistance, as shown and discussed in \cite[21-25]{oosterveld1970}.
-The fundamental difficulty with greatly shortening a duct, however, is that it requires that any changes in the shape and with of the water column be accomplished much shorter distance for the duct to achieve its desired effect, which means that generally it has to achieve less of it.
-This means that very low duct lenghts achieve less of an efficiency improvement to the ideal efficiency of a propeller, in those circumstances where they can achieve an improvement  (accelerating ducts at high loads), as discussed above.
-
-If more aggressive shape changes are attemped in a limited space, the deflections of flow and pressure gradients become much more extreme.
-This leads to a far greater propensity for flow separation to occur.
-This can happen inside the duct ahead of the stator, particulalry when any degree of diffusion occurs in order to elevate static pressure at the blades, or any kind of direction change is required as in the case of waterjets.
-Unlike waterjets, pumpjets also have the potential for separation on the exterior surface, which is an additional concern in the design of duct \cite[13]{henderson1964}.
-Given the role of the duct is to change the speed of the flow at the impeller relative to what would occur in its absence (in particular to decelerate it) separation might tend to occur on different surfaces in different operating conditions, depending on the relative pressure gradients and velocity ratios they experience at different speeds and loads.
-This substantially complicates the challenge of designing a duct that performs well across a wide speed range.
-This is illustrated for the case of a waterjet in \cite[20]{bulten2006numerical}.
-Equivalent complexities occur in pumpjets operating in low speed ranges (2-6kt, where flow separation occurs ahead of the impeller, inside the shroud.
-In the case of an accelerating duct, this phenomenon is also observed on the outside of the shroud at high advance ratios (the equivalent of low speeds and low loads) as shown and discussed in some detail in \cite{willemsen2013}.
-
-\begin{figure}
-\includegraphics[width=0.5\textwidth]{LowIVR.png}
-\caption{A an intake may experience risks of flow separation even at low speeds (Inlet Velocity Ratio) on some surfaces due to certain flow deflections being more severe.
-Source: \cite{bulten2006numerical}}
-\label{fig:LowIVR.png}
-\end{figure}
-
-\begin{figure}
-\includegraphics[width=\textwidth]{HighIVR.png}
-\caption{A an intake may experience risk of flow separation on different surfaces at high speeds (Inlet Velocity Ratio) where other flow deflections are more severe.
-Source: \cite{bulten2006numerical}}
-\label{fig:HighIVR.png}
-\end{figure}
-
-\begin{figure}
-\includegraphics[width=\textwidth]{DuctSeparation.png}
-\caption{Flow separation occurs at on the outside of an accelerating duct at high $J$ values, which corresponds to low loads, or speeds.
-For high $J$ values, the pressure reduction on the inside of the duct, creating positive duct thrust, can be seen.
-Source: \cite[73]{willemsen2013}}
-\label{fig:DuctSeparation.png}
-\end{figure}
-
-\begin{figure}
-\includegraphics[width=\textwidth]{NozzleDrag.png}
-\caption{Dramatically reducing the length of a duct relative to its diameter ($L/D$) as shown in \cite{oosterveld1970} can reduce its drag, which has the most significant impact at low thrusts or speeds.}
-\label{fig:NozzleDrag.png}
-\end{figure}
-
-In addition, the substantial reduction in the length of the shroud, both ahead of and behind the rotor, will also have the substantial effect of diminishing any acoustic sheilding which the duct may achieve.
-
-\subsubsection{Summary remarks on potential for redesign of pumpjet for low-speed conditions}
-Overall, while the parameter-space for the alteration of marine-propulsors is large, the parameters which would allow for very dramatic rather than minor changes to propulsor efficiency at low speed are relatively few.
-The major ones include increasing the width of the intake, decreasing the length of the duct, and decreasing the pressure inrease at the blades achieved by the duct.
-In short, it involves evolving the pumpjet back towards a propeller.
-None of these changes could be effected to a substantial degree without some inevitable trade-off in terms of the cavitation performance of the system at higher speeds.
-
-\section{Model Structures and Assumptions}
-
-\subsection{Mathematical Structure}
-
-The core of the model rests upon the assumption that drag increases with the square of the velocity of the submarine.
-This is a widely used assumption for fully submerged objects subject to turbulent flows \parencite[5]{wislicenus1973}, \parencite{davies2017}, as also discussed in earlier sections.
-Given that the amount of work done (energy consumed) by a system is equal to the distance over which it acts, which scales with speed also, we assume that the Effective Power $P_E$ (also known as Towing Power) required for propulsion to meet a given speed $v$ can be given by Equation \ref{eq:towpower}, where $C_d$ represents a drag constant.
-
-\begin{equation}
-\label{eq:towpower}
-P_E = C_dv^3
-\end{equation}
-
-The power drawn by the propulsion $P_D$ is assumed simply to be the by the Effective Power divided by propulsive efficiency at any given speed $\eta_v$ as given in Equation \ref{eq:powerdrawn}.
-Using a plausible efficiency assumption and total power required, Equation \ref{eq:powerdrawn} can be used to solve for the constant $C_d$.
-Other efficiency factors, such as the efficiency of electric motors or hull efficiency factors are assumed to be either relatively small, or essentially similar between submarines which might be conventionally powered but with a jet or propeller being the main point of difference.
-Hence the total power consumed by the submarine is given by Equation \ref{eq:totalpower}.
-
-\begin{equation}
-\label{eq:powerdrawn}
-P_D = \frac{P_E}{\eta_v}
-\end{equation}
-
-\begin{equation}
-\label{eq:totalpower}
-P_T = P_D + H
-\end{equation}
-
-With this relationship defined across the speed range, it is elementary to calculate the time that a given speed could be sustained based on a finite energy store (endurance) and the distance covered in this time (range).
-This provides the fundamental structure that underlies the model which is used to assess the dived range and enurance of conventionally powered submarines, without using air independent propulsion (AIP).
-
-\subsection{Efficiency Curve Assumptions}
-
-In order to begin to model the likely impact of propulsion system choices on system peformance characteristics such as dived range and endurance, it is necessary to adopt some plausible curves indicating likely levels of efficiency across the operating speed range, in this case assumed to be up to 20kt.
-
-In this case, three possible curves have been selected for each system, representing a high, low, and central assumption for each, with peak efficiency for the propeller at about 65\% at near 5kt, and jet efficiency around 27\% at the sames speed.
-The central assumption has been taken in both cases from the study undertaken by BMT and Rolls Royce, as in \cite{giles2010}, and indicated in \ref{fig:BMTShape.png}, since this represents the most direct comparison made as a simple function of waterspeed, with comparable technologies.
-The technologies naturally will not be identical.
-It is probable that the pumpjet of the submarine will be a purely axial-flow pump, as opposed to a mixed flow pump, but this difference in turbomachinery is unlikely to result in a overall loss of efficiency \parencite{hamilton1997}.
-The intake duct may be somewhat longer, increasing duct losses, but the degree of diffusion in the duct may be lower, reducing the adverse pressure gradient.
-It is probable that the the ship propeller would have five blades, where the submarine would likely have seven or nine.
-However, being  for a warship designed to travel at similar speeds to a nuclear submarine, with relatively unconstrained propeller size, and the pumpjet ejecting into the water, the technologies are likely to be far closer than any other publically available comparison.
-Overall, however, its considered that having an experimtally verified curve comparing very similar technologies as a direct function of waterspeed provides a far more certain basis for comparison than attempting to derive different curves \textit{ab initio} using theortical dimesnionless curves attempting to estimate how the proposed designs might correspond to them.
-
-Variations from this central line have been made with consideration for some plausible trade-offs, which are intended to be indicative, but plausible.
-In the case of the high pumpjet assumption, it is assume that a substantially larger intake area has been chosen in order to achieve higher efficiencies at lower speeds, at the cost of peak ideal efficiency, which might be realised not too far above 20kt for such a design.
-The low assumption for the pumpjet is still represents higher efficiencies at equivalent speeds than those given in \cite{fujisawa1995} for a waterjet.
-It is assumed that for both the central and low pumpjet curves, peak efficiency would not be achieved until a speed considerably above 20kt.
-All of the efficiency curves reflect a smaller decline in efficiency than the indicative assumption used by Andrew Davies, namely that efficiency might reduce by a factor of four between transit and patrol speeds, if those speeds also differed by a factor of four \cite{davies2017}.
-
-The higher assumption curve used for the propeller approaches a maximum of just under 72\%.
-This performance would not be implausible if a specialised propeller was adopted specificially to deliver peak performance at around this speed.
-Advanced flexible composite propellers, such as those used by the German submarine builder TKMS \parencite{durrant2016}, are able to achieve very high efficiencies, as well as delayed onset of cavitation, by allowing the blades to deform in certain ways to adapt to their specific degree of loading \parencite{young2007hydroelastic2}.
-Such propellers have been observed in experiment to have propulsive efficiencies above 70\% \parencite{young2007hydroelastic}.
-In this case I have assumed that this design necessitates the trade-off of steadily declining efficiencies over 10kt, considerably below the speed for which most open propellers are optimised.
-Given the ability for composite propellers to adapt their shape as loads increase, it is actually likely that the performance curve in this case is actually far broader.
-However, assuming a trade-off in this case also facilitates a scenario whereby the efficiencies of a propeller and jet actually cross over at around 18kt, which is relevant to consider, since experts in discussion have ruled out the possibility that a jet could be more efficient at lower speeds than this.
-
-The lower propeller curve shows efficiency still increasing up to around 65\% at 20kt, which might be consistent with a propeller optimised for conventional ship speeds, but probably not likely for a specialised submarine propeller.
-For all propeller curves, sharply declining efficiency has been assumed at very low speeds, (2kt or less), in order to reflect the dip seen below 5kt in the efficiency curve seen in Figure \ref{fig:BMTEfficiency.png}, assuming that this is driven by the decline in efficiency shown in Figure \ref{fig:PropDuct.png} at advance ratios approaching 1.
-This has been exaccerbated beyond realistic levels in the case of the the low propeller assumption line for the sake of testing sensitivity more dramatically to some losses in this regime.
-It is unlikely that a state-of-the-art submarine propeller would not be able to attain maximum efficiencies between 5 and 15kt, unless some other significant factor not discussed in this paper affects propellers in these regimes.
-At very low speeds, some small but sudden changes in efficiency may be possible in either direction, particularly as complex effect with multiple overlaid effects take place, for example increasing laminar flow over the propeller blades leading to lower frictional losses, but potentially greater separation.
-However, on balance, the low propeller curve is considered relatively unlikely, but useful for investigating sensitivity to errors or adjustments.
-
-\begin{figure}
-\includegraphics[width=\textwidth]{LaminarProp.png}
-\caption{At very low speeds changes in regimes such as an increase in laminar flow over the blade section could lead to small changes in efficiency, perhaps explaining the distinct rise at low speeds for the properller at ~5kt in Figure \ref{fig:BMTEfficiency.png}}
-\label{fig:LaminarProp.png}
-\end{figure}
-
-These efficiency curve assumptions are shown in Figure \ref{fig:Efficiency_assumptions}.
-
-\begin{figure}
-\begin{knitrout}
-\definecolor{shadecolor}{rgb}{0.969, 0.969, 0.969}\color{fgcolor}
-
-{\centering \includegraphics[width=\maxwidth]{figures/plots-plot_efficiency-1}
-
-}
-
-\end{knitrout}
-\caption{A range of possible efficiency curves for pumpjets and propellers to be modeled}
-\label{fig:Efficiency_assumptions}
-\end{figure}
-
-\subsection{Hotel Load Assumptions}
-It is also required that some assumptions are made about the hotel load, or the power which the submarine requires which doesn't go to propulsion.
-We might assume that this is in the order of 100kw, as suggested in \cite{martinez1995} though official sources of information are scarce.
-It is also likely that the actual hotel load varies considerably as different systems are switched on and off, some of which might be discretionary in combat scenarios (e.g.
-hot water for washing), and others (such as the combat systems) might have increased demand during crucial phases of combat operations.
-It is important to note that the combat system that has been selected for Australias FSM, the AN/BYG-1, is designed for a nuclear submarine, and is reportedly consumes considerably more power than other systems \cite{muir2015}, apparently even in the order of 100kW more.
-\cite{patrick2011}.
-Consequently, the actual hotel load experienced might vary considerably, and as such the model used in the app allows the hotel load assumption to be varied, in order to test sensitivity to this parameter.
-A flat 100kW hotel load is used as a starting assumption in the initial calculations in this paper, and some sensitivity analysis is given at half and double this power, to give an indication of the impact of plausible alternative energy demands dependent on operational circumstances, and choice of combat system.
-
-\subsection{Battery Assumptions}
-Since the purpose of the task at hand is to assess dived endurance only, this model focuses on a single energy store simulated by a total embarked mass (battery bank) and an energy density of those batteries.
-Since the SEA 1000 project has elected not to use lithium electric batteries, the energy density of the batteries can be assumed to be that of lead-acid batteries, which are known to be in the region of 30-50Wh/kg (Watt-hours per kilogram).
-Taking a central figure and converting this to Megajoules per kilogram (MJ/kg) we assume 0.14MJ/kg to be the effective energy density of lead acid batteries for our test submarine's purposes.
-This parameter can be adjusted easily for sensitivity testing, and in our case this will be adjusted to half, and double this energy density, noting that lead-acid batteries become much slower to charge as their charge level passes 80\%, and consequently it may be the case that only 50\% of their available capacity might be genuinely available for tactical purposes \parencite{briggs2016}.
-
-It is worthwhile considering that alternative types of battery may be on offer for submarines.
-In particular, lithium ion batteries were offered by TKMS in their bid for SEA 1000, and are being installed in Type 212 submarines operated by the German navy, as well as contracted for the new Norwegian submarines, as well as Israeli submarines.
-Precise energy density of these batteries is not publically available.
-For domestic uses such as handheld electronics or solar power applications, lithium ion batteries may have 3-5 times the energy density of lead-acid batteries \parencite{oconnor2017}.
-However, the actual useful density of lithium ion batteries can vary widely depending on different chemistries and designs.
-It might be assumed that for very large battery banks (such as submarines) where the stability of the battery bank is of critical importance, maximum density migth be sacrificed in favour of additional safety systems, such as increased monitoring, as well as spacing and insulation barriers between cells, in order to prevent thermal runaway in one cell leading to a catastrophic fire.
-The realistic energy density advantage that might be obtained as been suggested to lying closer to a factor of two or slightly less \parencite{greenfield2016}.
-
-The total embarked load of the batteries is not a well-known public figure, though it is likely to be several hundred tonnes in order to provide the likely desired levels of dived endurance.
-It has been suggested publically that the approximately 700 tonnes might be required for the Shortfin Barracuda \parencite{ohff2017}.
-This would appear to be quite a large fraction of the plausible payload of a submarine, even of over 4000 tonnes displacement, given that a great deal of the potential displacement will be taken by the thick steel involved in constructing a pressure vessel (probably at least a thousand tonnes per submarine) \parencite{jean2017}.
-500 tonnes of batteries has been assumed as a nominal figure for these purposes.
-Given that the mass of batteries embarked has only a direct and linear relationship with the dived range and endurance, no sensitivity testing will be undertaken on this variable.
-Proportional changes to the energy density of the batteries will have equivalent effect for the purposes of measuring the impact on performance.
-Consideration of likely impacts on space, and hull-resizing, and associated drag are beyond the scope of this paper.
-
-\subsection{Drag Coefficient of Hull}
-
-In the online version of the web-app, alternative reference points or known facts relating a given power output to a speed can be used in order to solve for this constant $C_d$.
-For instance, one might assume (as done by \cite{davies2017}) that the hotel load might match the propulsive demand at some given speed, for some given hotel load.
-However, given that the precise hotel load is generally not known and highly variable, and the potential for the different systems to have widely varying propulsive efficiencies at low speeds, this method is subject to very wide error margins.
-
-Given that any known power consumption level and speed match can be used as a reference, I have chosen instead to use the probable power consumption at top speed, both of which we have some indication of, and given that propulsive efficiencies of the two systems are likely to be much closer at this speed, there will be less error in likely comparisons between different design concepts.
-It is known that the a early German designs for a plausible submarine for SEA 1000 included a 6MW electric motor \parencite{patrick2012}, and the Shortfin Barracuda of Naval Group is often referred to as having a 7MW electric motor \parencite{ohff2017} \parencite{coates2016}, with top speeds ">20kt".
-Given the nature of increasing drag at high speeds, it seems unlikely that top speeds will be very far above 20kt.
-Hence, in this model we have assumed that 20kt will be obtained with the ouput of 6.5MW.
-
-Considerable confirmation that these assumptions are realistic can be made by assessing the endurance of the submarines under the model with other known reference points.
-For the central curves, we obtain a dived endurance betwen 60 and 100 hours for both systems at 5kt, reflecting nearly exactly the range of other public estimates of dived enurances endurances which other modern diesel-electric submarines operating only with batteries \parencite[3]{buckingham2008submarine} at around 4kt.
-Given that somewhat larger submarines might be somewhat assisted in this respect because their overall volume/surface ratio is slightly higher, this gives confirmation that the model is overall quite realistic in its magnitudes at the most important speeds.
-The sames sources also suggests that the same submarines achieve endurances of around 1-2 hours at 20kt, which also confirms that the relationship in our model linking differnt speeds is also in agreement with other models.
-
-\section{Results and Discussion}
-
-\subsection{Power Demand}
-As part of the calculation of results for each pair of efficiency curves, the power requirements of each respective component (propulsion for the two respective systems, as well as effective power required, and hote load) can be plotted to ensure that realistic results are obtained, which is important for selecting plausible values for the constant $C_d$.
-For the central efficiency curves, we can show the resultant power curves for the the assumed values of battery load, energy density, and hotel load, as seen in Figure \ref{fig:power_curve}.
-
-\begin{figure}
-\begin{knitrout}
-\definecolor{shadecolor}{rgb}{0.969, 0.969, 0.969}\color{fgcolor}
-
-{\centering \includegraphics[width=\maxwidth]{figures/plots-plot_power1-1}
-
-}
-
-\end{knitrout}
-\caption{With central assumptions, a pumpjet powered submarine reaches 20kt with 6.5MW of delivered power.  Hotel load matches propulsion power demand around or below 5kt.}
-\label{fig:power_curve}
-\end{figure}
-
-It can be seen here that one of Andrew Davies assumptions, regarding hotel load matching propulsion power demand at patrol speeds, might match quite closely with our assumptions at a patrol speed of 4kt for the case of a pumpjet.
-It also shows the remarkable impact of the square law on power demands, particularly their effect at low speeds.
-At just a few kt, an entire submarine might be propelled by a power level comfortably produced by a large family car.
-This gives further explains why it is that efficiencies of propulsion systems aren't routinely discussed at dramatically lowered speeds.
-Even for jets which have efficiencies which decline much faster than those we've assumed (for example in \cite{fujisawa1995}), the reduction in power demanded falls substantially faster than efficiency.
-Consequently, at low speeds nothing dramatic occurs to make the declining efficiencies of consequence to operators.
-It is often still more economical for such vessels to travel very slowly, well outside the ideal performance range of their propulsion systems, simply because of the drag on the hull.
-
-Overall, this chart is consistent with expectation, and confirms that a plausible value of $C_d$ has been selected under these assumptions.
-
-\subsection{Central Results}
-The plots comparing the resultant endurance and range calculations across the speed range considered are in Figures \ref{fig:central_extreme}, \ref{fig:central_jet}, \ref{fig:central_prop} and \ref{fig:bottom_top}.
-In all cases, the comparison for the central efficiency curves have been replotted centrally, in order to highlight varience under each of the different assumption comparisons.
-
-It is clear that under the central assumptions, the declining efficiency of the pumpjet has a very significant impact on range and endurance, particularly at speeds around 4-8kt.
-It might be noted that under the central assumption, the propeller system achieves 138 hours of dived endurance at 4kt where the pumpjet only manages 91 hours dived endurance at this speed.
-Consequently, at a likely patrol speed, the difference amounts of almost exactly two days in dived endurance.
-One might presume that such a difference is of some considerable tactical significance.
-In order to match the propeller's dived endurance, the pumpjet submarine would need to slow to 2.5kt.
-
-The phenomenon manifests itself in a very substantial reduction in the overall range that can be achieved submerged, and a difference between the speed at which maximum range can be achieved.
-Under central assumptions, the propeller driven submarine can reach over 540nm at any speed between 4 and 5kt.
-The pump-jet driven submarine reaches over 360nm between 3 and 6kt.
-A difference of 180nm dived range would presumeably be of considerable tactical significance.
-
-\begin{figure}
-\begin{knitrout}
-\definecolor{shadecolor}{rgb}{0.969, 0.969, 0.969}\color{fgcolor}
-
-{\centering \includegraphics[width=\maxwidth]{figures/plots-plot_results1-1}
-
-}
-
-\end{knitrout}
-\caption{The most extreme and central assumptions are compared}
-\label{fig:central_extreme}
-\end{figure}
-
-\begin{figure}
-\begin{knitrout}
-\definecolor{shadecolor}{rgb}{0.969, 0.969, 0.969}\color{fgcolor}
-
-{\centering \includegraphics[width=\maxwidth]{figures/plots-plot_results2-1}
-
-}
-
-\end{knitrout}
-\caption{The impact of different jet efficiency assumptions compared to the central propeller curve shows wide variance}
-\label{fig:central_jet}
-\end{figure}
-
-\begin{figure}
-\begin{knitrout}
-\definecolor{shadecolor}{rgb}{0.969, 0.969, 0.969}\color{fgcolor}
-
-{\centering \includegraphics[width=\maxwidth]{figures/plots-plot_results3-1}
-
-}
-
-\end{knitrout}
-\caption{The impact of different propeller efficiency assumptions compared to the central pumpjet curve shows less variance}
-\label{fig:central_prop}
-\end{figure}
-
-\begin{figure}
-\begin{knitrout}
-\definecolor{shadecolor}{rgb}{0.969, 0.969, 0.969}\color{fgcolor}
-
-{\centering \includegraphics[width=\maxwidth]{figures/plots-plot_results4-1}
-
-}
-
-\end{knitrout}
-\caption{Comparing the lower and higher lines for both systems together}
-\label{fig:bottom_top}
-\end{figure}
-
-\subsection{Sensitivity of Curve Selection}
-As can be seen from the comprehensive comparison in Figures \ref{fig:central_extreme}, \ref{fig:central_jet}, \ref{fig:central_prop} and \ref{fig:bottom_top}, the shape of curves selected has very different levels of impact at different speeds.
-Key observations include how small the impact of efficiency variation to either curves tends to be at very low speeds (under 2kt).
-This is due to hotel load becoming clearly dominant at such speeds.
-The relative difference between the lines also diminishes somewhat towards higher speeds.
-More striking, however, is the very dramatic reduction in both range and endurance at higher speeds.
-At a speed of 20kt, a propeller driven sumbarine might manage to remain dived for 3.6 hours according to these assumptions, where a pumpjet driven submarine might manage just under 3 hours.
-A difference of about half an hour would presumeably be of less tactical significance.
-At any speed above 10kt, neither submarine would be able to remain dived for much more than 24 hours (25.4 hours for the propeller system, 18.3 for the jet).
-
-It should also be noted that these results, and those around 8kt, can also provide some insight into likely snorting intervals and indiscretion ratios.
-As given in \cite{patrick2012}, submarines tend to carry diesel power capacity that approaches their main motor power.
-Whilst the difference between the two systems still appears very significant at these speeds, it appears that if transits occur in the region of 8-10kt, snorts might occurr roughly daily, and last a number of hours.
-The indiscretion ratio during commute of the jet system would be significantly impaired given the substantially larger power demands at transit speeds, though a detailed calculation and comparison lies outside the scope of this paper.
-
-It seems extremely clear from these conclusions that the general results confirm very strongly that the impact of the propulsion system is likely to be of the greatest importance in the speed range 3-8kt, where propulsion power is more significant relative to hotel load, but excessive drag does not prohibit significant dives.
-
-A further feature of the results is that the relative difference in performance is much more sensitive to changes in the efficiency of the jet at the relevant speeds, owing to their larger proportional impact at low efficiencies.
-Whilst a smoothly declining efficiency curve from around or over 50\% at 20kt to zero at 0kt might seem moderately well defined, the precise path taken, particularly the exact level around 3-8kt, is of great importance, substantially more than plausible variations in the propeller curve.
-This can be seen in an overlaid plot of the proportional differences in total power demand (hence endurance and range) for the different pairs of curves as in Figure \ref{fig:all_proportions}.
-
-\begin{figure}
-\begin{knitrout}
-\definecolor{shadecolor}{rgb}{0.969, 0.969, 0.969}\color{fgcolor}
-
-{\centering \includegraphics[width=\maxwidth]{figures/plots-plot_proportions-1}
-
-}
-
-\end{knitrout}
-\caption{The most proportional differences between the propeller and pumpjet performances for a range of assumptions}
-\label{fig:all_proportions}
-\end{figure}
-
-The proportional change in overall power requirement (directly determining both range and endurance) shows that in the central scenario a propeller driven submarine would have more than 60\% advantage in terms of range and endurance around 5-6kt.
-It is interesting to note that the proportional change is greatest slightly above the speed where maximum range is achieved, and the point of peak advantage moves steadily towards higher speeds as the difference in efficiency between the systems increases.
-
-As can be seen, even compared to the central propeller curve, choosing a jet efficiency curve somewhat lower than the central assumption means that the peak difference immediately jumps to over 100\%, or effectively allow a propeller powered submarine to achieve double the range and endurance of a pumpjet powered submarine at those speeds.
-It is interesting to note that these speeds in include those speeds which might lie near the maximum speeds that could be sustained by Air Independent Propulsion (AIP) systems \parencite[100]{stanford2017}.
-Consequently, it is likely that a conventional submarine that used pumpjets might only achieve a fraction of the dived range and endurance enhancements generally expected on propeller powered submarines which employ this technology, as will be discussed later.
-
-The proportional improvements will also be effective rough guides to the likely impact on indescretion ratios and overall range.\footnote{Additional considerations including the size and efficiency of the diesel engines and generators must be included for proper analyses of indescretion ratios, which will not be given here.}  Near patrol speed almost all scenarios show a difference approaching or exceeding 50\%, which will have a material impact on indescretion rates during all combat operations.
-
-Perhaps surprisingly, the impact at transit speeds is still quite significant, with no scenarios showing less than 20\% advantage around 8-10kt.
-The sensitivity to pumpjet efficiency here is also very high, with three scenarios showing an impact of over 70\%.
-Given the likelihood that a submarine's average speed over its range is likely to be dominated by its transit speed, this indicates that the choice of a pumpjet propulsion system will also result in a substantial reduction in the overall range of the submarine, quite probably in the range of between 25-50\%.
-This would require the embarkation of significantly larger volumes of diesel to meet minimum range requirements, and also contribute to higher running costs.
-
-\subsection{Sensitivity to Hotel Load}
-As can be seen in Figure \ref{fig:sens_hotel} the difference between the two systems is very sensitive to changes in hotel load, both in proportional and nominal terms.
-The impact of halving the hotel load from 100kw to 50kw more than doubles the \textit{difference} between a pumpjet and propeller dived endurances, from around two days to well over four days at speeds between 2.5 and 3.5kt.
-In the context of a submarine which might have a nominal dived requirement at patrol speeds in the order of several days, this suggests that extremely large gains can be made by reducing hotel load.
-Increasing the hotel load has the noticeable effect of pushing the speed of largest difference to higher speeds, and substantially reducing the total difference.
-If the hotel load was doubled to 200kW, the middle efficiency curve pair would still see a difference of nearly a day between the two systems, or around 19.5 hours at speeds of 4.5-5 kts.
-
-The difference between pumpjet and propeller propulsion systems in terms of range tends to be found at slightly higher speeds, though this point also moves towards higher speeds as hotel load increases.
-With a reduced hotel load of 50kw, the difference between the two systems in the middle scenario is about 370nm for speeds 3.5-4.5kt, substantially greater than the ~200nm around 4.5-5.5kt at a hotel load of 100kW.
-
-It's important to acknowledge that such substantial differences in hotel load merit serious consideration, both in terms system design, as well as for variance in operation of a submarine with a given system design.
-As has been discussed, the precise hotel load of submarines is not generally publically available, which justifies senitivity testing around a relatively wide potential range.
-The size of the crew required for different designs might vary considerably, as it also would for different missions (for example with the embarkation of a special forces congingent) which would necessarily have considerable impact on energy demand for life support, including sanitation, CO2 scrubbing, and cooking.
-In addition, it is broadly understood that the AN/BYG-1 combat system consumes substantially more power than other alternatives, reportedly in the order of 100kW \parencite{patrick2011}.
-
-Furthermore, the hotel load may be varied considerably at the captain's discretion depending on operational requirements.
-Not all sensors or components of the combat system need to run at all times, for example during long transits in order to reach an operational area, or during a long dive in order to escape an operational area once a large strike has already been conducted.
-In wartime, when surfacing to snort within range of enemy airpower or close to naval bases might result in the sumbarine probably being sunk, the crew might willingly forego hot showers, hot meals, and climate-controlled air in order to increase their chances of survival.
-Consequently, it is perfectly plausible that changes of hotel load in the order of a factor of two for tactically significant time periods (days) is achievable.
-The results of this sensitivity analysis show that the penalty inflicted by the pumpjet will be substantially greater in such circumstances.
-
-The proportional impact of the changes in hotel load area also substantial, and exhibit clearly the same trend of the impact being larger at higher speeds for higher hotel loads.
-For the efficiency curve pairs that comprise the central selections, decreasing the hotel load to 50kt would see the dived endurance and range of the submarine improved by the adoption of propellers 80\% between 4 and 5.5kt.
-A much higher hotel load, at 200kW, would see that peak difference over 40\% between 5.5 and 8kt.
-
-It is also clear that increasing the hotel load has a greatly diminised impact on the proportional differences between the systems at higher speeds.
-At very high speeds (e.g.
-15kt) changes in the hotel load have almost no impact in proportional terms to the difference, which lies at about 30\% in all cases for the central selection of curves.
-At 9kt, in the region where transit speed is likely to occur, increasing the hotel load by a factor of four (from 50kW to 200kW) only decreases the advantage of the propeller from 48\% to 38\%.
-Therefore one might say near certainty that any difference between efficiency around these patrol speeds is likely to have a substantial impact on the overall range of the submarine, with differences of much less than 30\% quite unlikely, irrespective of other changes in the submarine design or operation circumstance.
-
-It should also be noted that sensitivity testing has not been shown in this paper around alternative values of the drag coefficient $C_d$ which was solved for by matching peak power output and top speed.
-It is considered relatively unlikely that such dramatic changes in this constant are plausible.
-Furthermore, changing this constant will have a very similar impact to alerations in hotel load, in terms of its affect on the speed where greatest differences will occur and their probable magnitude.
-Consequently, consideration of the sensitivity to hotel load is likely to easily encompass the plausible extent, and nature, of the likely uncertainty in this factor.
-
-\begin{figure}
-\begin{knitrout}
-\definecolor{shadecolor}{rgb}{0.969, 0.969, 0.969}\color{fgcolor}
-
-{\centering \includegraphics[width=\maxwidth]{figures/plots-plot_sens_hotel-1}
-
-}
-
-\end{knitrout}
-\caption{The impact of the propulsion system efficiencies are very sensitive to changes in the hotel load, including as a proportion.}
-\label{fig:sens_hotel}
-\end{figure}
-
-\subsection{Sensitivity to Battery Changes}
-
-Changes in the effective battery energy density will have the same effect as having larger masses of battery stored.
-They simply directly impact the total amount of energy available for consumption before recharging batteries while snorkeling.
-There is therefore no shift in the proportional impact on range and endurance due to this factor, as can be seen in the lower graphs in Figure \ref{fig:sens_battery}.
-The nominal impact, however increases directly and linearly with the total energy stored.
-
-The results in nominal terms are striking in their magnitude.
-In terms of enduracy, doubling the stored electrical energy, for example by adopting lithium batteries, increase the \textit{difference} between the two systems under the central efficiency curve assumptions to nearly four days (93 hours) at 4kt.
-Given that most modern, capable submarines are estimated to only be able to achieve dived endurances between 60 and 100 hours on batteries at 4kt \parencite{buckingham2008submarine} this cannot be dismissed in terms of its tactical significance.
-Even under the eficiency curve assumptions representing the narrowest spearation considered, the difference in this case would be 59 hours, almost equal to the estimated dived endurance of a Type214 operating without AIP.
-
-The in terms of submerged range, the same proportional impact is also extremely significant.
-Doubling the energy stored inceases the \textit{difference} between two otherwise identical submarines to over 400nm under central assumptions at 4.5kt.
-That's a larger distance than the width of the Java Sea, the most central body of water in the Indonesian Archipelago, or most of the length of the Malacca Strait.
-The tactical significance of such distances are not likely to be lightly dismissed.
-If far smaller amounts of energy are embarked, the nominal differences reduce to perhaps 100nm, in direct proportion to the reduction of the submarine's dived endurance.
-This is still a non-trivial distance in terms of it's tactical signicance, and comprises a substantial fraction of the dived ranges of most non-AIP submarines.
-It still represents a significant difference in operational terms, equal approximately to the width of Hainan Island, or neraly twice the length of the Sunda Strait.
-
-The singificant range of different energy density values tested here reflect quite plausibly some of ranges of different energy levels available in operation.
-For instance, as has been discussed \parencite{briggs2016, greenfield2016}, lead acid batteries may well only be able to practically utilise about 50\% of their total energy stores for tactical operaions, unless operating in a very low-threat environment and undertaking very slow charging to reach the maximum full charge is possible.
-In addition, the plausible improvement in enery density of lithium batteries may well be in the order of a factor of two in practice \parencite{greenfield2016}, though theoretical limits of the chemistry suggest much higher may be possible \parencite{davies2016}.
-
-Only brief is given here to the impact of Air Independent Propulsion, which has substantially higher energy densities possible even than Lithium, and also amounts to the addtional embarkation of greater mass.
-The volumetric energy density of the feedstocks to modern AIP systems (liquid oxygen and methanol) is estimated to be approximately ten times greater than that of lead-acid batteries \parencite{davies2016}, around 4.5MJ/L.
-Given the greater mass embarked, and additional studies suggesting that at constant speeds they might extend the dived endurance of a modern submarine from around 60 hours to over 1000 hours (a factor of roughly sixteen) an approximate estimation of the the direction and magnitude of this impact may by scaling up energy density by this factor.
-This would amount to adding three further frames to the right of Figure \ref{fig:sens_battery}, with the energy density being doubled in each successive frame.
-This would plausibly (under the central line selections) amount to more than month of dived endurance \textit{difference} between AIP inclusive submarines with propellers or pumpjets, amounting to thousands of nautical miles of continuously submerged travel.
-This difference would also now be material in terms of the overall range of the submarine amounting to a further 15-20\% reduction in the submarine's overall range.
-
-With such massive differences between the performance of the two systems emerging with the presence of the AIP, it is difficult to see how one could realistically consider a pumpjet as being competitive if a direct comparison were made.
-
-\begin{figure}
-\begin{knitrout}
-\definecolor{shadecolor}{rgb}{0.969, 0.969, 0.969}\color{fgcolor}
-
-{\centering \includegraphics[width=\maxwidth]{figures/plots-plot_sens_battery-1}
-
-}
-
-\end{knitrout}
-\caption{The impact of the propulsion system efficiencies are extremely sensitive to battery energy density (or load) though proportions remain fixed.}
-\label{fig:sens_battery}
-\end{figure}
-
-\subsection{The interaction of hotel load, battery energy density, and efficiency curves}
-
-\begin{figure}
-\begin{knitrout}
-\definecolor{shadecolor}{rgb}{0.969, 0.969, 0.969}\color{fgcolor}
-
-{\centering \includegraphics[width=\maxwidth]{figures/plots-plot_sens_corners-1}
-
-}
-
-\end{knitrout}
-\caption{Specifying low density batteries and a high hotel load dramatically reduces the proportional and nominal impact of propulsion system efficiencies at low speeds}
-\label{fig:sens_corners}
-\end{figure}
-
-\section{Acoustic Considerations}
-
-\subsection{The importance of cavitation inception}
-It is generally considered that the existence of cavitation of any sort will tend to dominate the acoustic signature of any vessel that doesn't produce any machinery noise by relying on all-electric drive as a conventional submarine would when dived (\cite{giles2010}, \cite{harvie1965construction}).
-Its quite possible that during the period when a submarine is snorting, or running its diesel engines, these sources of noise would be a more substantial and distinctive acoustic signature than small amounts of cavitation on the propulsion system, if it were to occur, owing in particular to the low freqency of diesel engines.
-
-Other sources of noise hydrodynamic noise, such as the turbulence produced in the boundary layer of the flow over the hull, or parts of the propulsor, will still contribute to overall noise and some signature, but it will tend to be vastly smaller in magnitude.
-This owes to the essentially incompressible nature of water.
-Without compression, even very relatively intense votices or tubulent flows will not propagate very much noise, if the water remains perfectly solid.
-All of the 'round and round' can't achieve much 'in and out', which is what results in propagated noise.
-This obviously doesn't hold true in compressible fluids, such as air, as one would observe noticing the intense noise produced by hand-dryers, aeroplanes, jet engines etc.
-It also doesn't hold true within a turbulent flow underwater.
-If you were to put a sensor inside a turbulent flow, the rapid movements would result in continual and significant pressure fluctuations, hence noise.
-But at a distance, these fluctuations tend to substantially cancel out.
-The larger the scale of the eddies and vortices, the further noise might propagate, and also the lower the frequencies of resultant noise might be, which makes noise propagate further.
-Likewise, larger caviation voids will tend to emit lower frequency sound aswell, as can be observed from the boiling of a kettle.
-
-However, when dived, it might be reasonably assumed that the complete elimination of all sources of cavitition will be the primary objective of acoustic signature management.
-This is neatly put in \cite[88]{gearhart1966selection}: "The occurence of cavitation in a propulsor results in adegradation of the propulsor operating characteristics if extensive amounts of cavitation exist, and in significant noise when minute amounts occur.
-As a result, the performance requirements of a submersible weapon system specify that the occurence or inception of cavitation be avoided below certain operating depths." One might consider that a submarines 'tactical silent speed' would most probably correspond to the speed at which it can operate safely below the point of any cavitation inception.
-
-The precise point at which cavitation inception occurs, what type of cavitation it is, and where it occurs on the blades is a complex science, with considerable work being continually undertaken to validate theoretical models.
-A good overview of the some of the science can be found in reference \cite{kuiper1981cavitation}.
-Still today, similar techniques such as paint tests and visual observations are validate models and theory.
-However, despite the inherent complexity about predicting the precise point in time and space of onset, and the respective impact on efficiency and acoustics, several broad relationships are known to hold well, and can be relied upon to characterise the most significant trends that determine when cavitation will be more likely to occur, which is generally driven by the diffence between the vapour pressure of water and the local static pressure.
-
-\subsection{Impact of speed on cavitation}
-Perhaps the most significant determinant of the pressure deviations on the propulsion system is how hard the propellers have to push on the water at any given point.
-In general, the harder the propellers push, the larger is the required pull on some opposing edge or side.
-It is these pressure reductions, on the suction side of the blade, where sheet cavitation generally starts to occur.
-To think of it simply, the water isn't able to rush to fill the gap left behind by the propeller fast enough.
-However, at some point cavitation also occurs on the pressure side of the blade, as the impact of the leading edge breaking the water leads to sever local pressure drops just behind the leading edge.
-
-Whilst a range of very particular effects may be at work, perhaps the largest trend at work will be that pressure differentials across the blade must necessarily increase with increase thrust.
-The blade velocity relative to the oncoming stream is also a close indicator, which will be closely related for an open propeller.
-Consequently, since the thrust required for self-propulsion increases with velocity squared, the propensity for cavitation occur will necessarily tend to follow this relationship also.
-As such, modest changes in speed can have a big impact on the degree of probability of cavitation occurring.
-Halving your speed will roughly reduce local pressure fluctions by a factor of four.
-Consequently, even for non-specialised propellers, there is generaly some low speed for which no cavitation occurs, though almost all propellers in major surface vessels experience some cavitation at the speed they are designed to do their work at.
-
-\subsection{The impact of depth on cavitation}
-It is crucial to understand that increasing depth necessarily increases the static pressure throughout a propulsion system.
-As such, the speed at which cavitation starts to occur, will necessarily be a function of depth.
-Approximately every 10m of water adds approximately the equivalent of an atmosphere of pressure.
-As such, the static pressure a system experiences at 10m depth will be half that experienced at 30m (since the air in the atmosphere comprises the first atmosphere of pressure).
-Consequently, diving deeper will necessarily increase the speeds one can achieve without any cavitation occurring \ref{fig:InceptionDepth.png}.
-
-\begin{figure}
-\includegraphics[width=\textwidth]{InceptionDepth.png}
-\caption{Increasing in depth improves the speed achievable before cavitation begins to occur, as shown in the case of a torpedo \parencite{suryanarayana2010cavitation}}
-\label{fig:InceptionDepth.png}
-\end{figure}
-
-The interaction of these two relationships can be expressed in a general form in figure \ref{fig:InceptionContours}, which has been generated using an approxiate numeric model.
-As can be seen here, at very low speeds, modest changes in depths can achieve considerable additional surplus pressure in order to prevent cavitation.
-However at much higher speeds, sine the pressure diffential scales with the square of the velocity travelled, far greater additional depths are required to achieve sufficient excess static to supress cavitation for a give further speed increase.
-
-\begin{figure}
-\begin{knitrout}
-\definecolor{shadecolor}{rgb}{0.969, 0.969, 0.969}\color{fgcolor}
-
-{\centering \includegraphics[width=\maxwidth]{figures/plots-plot_inception-1}
-
-}
-
-\end{knitrout}
-\caption{Contours showing plausible lines of plausible equivalent cavitation inception}
-\label{fig:InceptionContours}
-\end{figure}
-
-It is noteworthy also that the vapour pressure of water is an important component in the equations that describe the conditions for cavitation to occur.
-It is generally given that when the local static pressure falls below the vapour pressure, cavitation occurs.
-Given that temperatures also affects the vapour pressure of water (warmer water is just a little closer to boiling) this does have an impact on likelihood of cavitation occurring.
-However, at plausible temperatures the impact is small.
-The difference between the vapour pressure of water at 10 degrees and 30 degrees is approximately 3 kilopascals.
-Diving just 10m deeper adds over 100 kilopascals to static pressure.
-In many circumstances, altering your speed by a couple of knots will have an even larger impact on local pressure reductions.
-Consequently, the impact of temperature on the inception of cavitation is essentially negligible.
-The difference between inception contours in 20 degree and 30 degree water are given by the difference between the black and red-dashed contours in Figure \ref{fig:InceptionContours}.
-
-\subsection{The significance of flow separation}
-The addition of a duct will necessarily lead to flow separation occuring on the inside of the duct opening at very low speeds which correspond to high advance ratios when load remains constant.
-Flow separation tends to lead to vortex shedding, which means that inconsistent flows (vortices) are then ingested into the blades of the impeller.
-This effect is well known and document.
-A good example is shown on the exterior of an \textbf{accelerating} duct at $J = .8$ in \cite{willemsen2013}.
-
-\begin{figure}
-\includegraphics[width=\textwidth]{VortexShedding.png}
-\caption{Vortex shedding following a flow separation leads to eddies being released unsteadily into the main flow stream, as shown here on the exterior of an accelerating duct.
-In the case of decelerating duct (pumpjet), duct shape is inverted, leading to the vortices being shed into the impeller blades.
-\parencite{willemsen2013}}
-\label{fig:VortexShedding.png}
-\end{figure}
-
-Whilst at very low power levels, or at good depths, it is quite plausible that this doesn't result in any cavitation, which would normally be a likely occurrence if the blades are operating closer to their design power and such such vortices were shed into the intake.
-However, given the inherently unpredictable and unsteady nation of such vortex formation and shedding, the this phenomenon would increase the risk of cavitation occurring unexpectedly in marginal conditions.
-In addition, in the absence of cavitation, the sudden and changing pressure gradients over the blades would dramatically increase the overall pressure fluctuations and turbulence level that occurs within the propulsor.
-This would necessarily increase the overall levels of noise generated.
-It is plausible that the noise is of sufficiently small level that it is not considered a significant compromise of stealth.
-In essence, the higher levels of noise lie below some threshold of significance.
-However, it is almost certainly untrue that that pumpjet actually generates less noise overall than a propeller at very low speeds.
-
-\subsection{The impact of duct shielding}
-The presence of the duct around the impeller may also have the effect of preventing the noise from being effectively radiated away from the propulsor.
-This may well result in substantial reductions in some types of noise being radiated away from the propulsor in certain directions, which may have given rise to the consistent qualification of 'radiated' being used to descrive the type of noise which differentiated the French submarine bid \parencite{stewart2016} \parencite{davies2017} \parencite{ohff2016}, including by the designers of the submarine \parencite{autret2016}.
-This is a secondary effect, quite separate to the avoidance of cavitation, which might be seen as an acoustic advantage in circumstances where cavitation might not be expected to occur on either type of propulsor.
-It is highly likely that this is the effect which is relied upon to sustain claims that a pumpjet is generally acoustically superior to a propeller, including at low speeds.
-
-It should be noted here that shielding effects might comprise the combination two distinct and seperate physically phenomenon: reflection and damping.
-Noise tends to be reflected very effectively off hard, smooth surfaces representing the boundary between mediums of very different density (such as steel and water).
-Changes in density which are gradual, often representing soft, squishy, or broken surfaces tend, to have the effect of absorbing sound, where the acoustic energy is converted to heat or degradation of the material, rather than reflecting it.
-Given the significance of increased drag on any surfaces within the propulsor, it is highly unlikely that any attempt at damping is made within the pumpjet.
-However, it should be noted that some pumpjets have been constructed with a hollow design, in order to reduce weight \parencite{harvie1965construction}, which would reduce the thickness of solid steel, though offer some potential opportunity for damping.
-Overall, damping is increased in its effectiveness at low frequencies by allowing it to occur over larger distances, corresponding to the sound's increased wavelenght, leading to increased thickness of the transition substance or damping material.
-Consequently, effective damping within the propulsor itself is unlikely to be attempted, or limited to medium and higher frequencies, probably suited to reducing noise from normal surface flow turbulence and very early onset cavitation.
-
-The ability of a thin layer of material to reflect noise is also highly frequecy dependent, and also dependent on mass of the shield layer.
-As the material becomes thinner and thinner, a hard shield layer will tend increasinly to transmit rather than reflect the noise, a phenomenon that increases as the amplitude (loudness) of the noise increases, and frequency decreases.
-Denser materials are more effective at reflecting sound, as more energy (pressure) is required to excite the same movement in the material.
-Consequently, the effectiveness of a shield depends on both its thickness and density.
-There is therefore some tension inherent in the design of a shield where thickness and total weight are constrained.
-A thicker layer solid steel will be very effective at reflecting sound, but have negligible absoption or damping.
-A hollow between two thin layers will offer some potential for a soft daping material to be inserted, but at the cost of reflection.
-
-It is most likely that little or no damping or absorbing layer is attempted within the body of the shroud, since increasing the duct thickness above hydrodynamically determined optimum profile is likely to have a detrimental impact on efficiency.
-Constructing especially thin layers, and still maintaining an optimal shape to the precision required for housing powerful turbomachinery like the impeller is also likely to involve considerable construction risk and complexity.
-Hence the shielding effect is most likely to dominated by the reflective characteristics of the duct, which would seek to internally reflect noise back into the turbulent flow, where sound waves would be increasingly convoluted, and somewhat absorbed through the eventual transition to small-scale turbulence and heat.
-
-The degree to which the total noise generated is reflected internally will also be shape-dependent.
-A more concave inner surface, consistent with higher degrees of diffusion in advance of the impeller, and greater nozzle narrowing after it, which is consistent with higher pressure elevation at the impeller (hence improved cavitation characteristics) but higher negative duct thrust, leading to even poorer efficiencies at low speed.
-Substantial alteration of duct shape on account of efficiency considerations at low speed are therefore necessarily going to have some impact on the degree of effective shielding that a duct can afford.
-
-\subsubsection{Directional Dependence}
-Two important qualifications must be made regarding shielding.
-The first is that its impact is highly directionally dependent.
-The pump jet requires openings at the front and back to allow water in and out.
-Only noise that is reflected internally sufficient to be largely absorbed is effectively shielded.
-The larger the areas of the intakes and outlets, the narrower the effective shielding will become.
-Consequently, design changes to make the pump more efficient at low speeds, which may include expanding both the inlet and outlent, and diminshing the difference between them, will necessarily be at the expense of diminishing effective width of the shielding 'belt' around the propulsor.
-
-When the submarine is travelling at high speeds, and there is significant flow through the propulsor, the impact of waterspeed through the propulsor might have the additional impact of diminishing the radiation of noise in the forwards direction, in the same way that sound tends tends not to travel well up-wind in air.
-However, at very low speeds this effect would be greatly diminished.
-
-It should be noted that in the underwater environment sound can be made to reflect in a variety of directions simply with differences in temperature, salinity, movement, as well as off plenty of nearby underwater objects, including the ocean floor and objects on it.
-This phenomenon generally serves to make the art of underwater acoustics extremely complex and unpredictable, and generally contributes to the difficulty of precisely locating a submarine, even when some signature is detected.
-However, it also means that direction-dependent reductions signature do not reliably translate to overall reduced probability of detection.
-
-\subsubsection{Frequency Dependence}
-It ought further to be noted that the effectiveness of acoustic shielding is also highly dependent on frequency.
-Much lower frequencies are very hard to reflect, and absorb, as can be easily observed in a house with a home-cinema, where lower frequencies emitted by the subwoofer will tend to affect many surrounding rooms, where higher frequency sounds might be substantially muffled by interceding walls.
-The same acoustic effects are at work in underwater acoustics.
-
-This has an important consequence for shielding for pumpjets at very low speeds because of the occurrence of flow separation and vortex shedding.
-In the absence of flow separation, and cavitation at very low speeds the most substantial sources of sound would be the turbulent flow over surfaces of in the propulsor.
-This noise will tend to be relatively white and high in its frequency owing to the very small length-scales of the eddies and vortices in such turbulence.
-The steel shroud would be very effective in reflecting such high frequency noise.
-Under acceleration or higher loads, the very instances of cavitation woud also be relatively high frequency, as the voids or cavities would be small at first.
-These sounds might also be very effectively reflected by the shroud, and consequently diminished.
-However, flow separation results in the creation of vortices of a dramatically larger scale than occurs with normal turbulence in a boundary later, and hence generate noise of a much lower frequency.
-In addition, as the flow rates slow even further, these vortices are likely to become larger in size, even if lesser in intensity, resulting in further reduction in the frequency of noise generated.
-Further to this effect, the rate at which the vortices shed from the surface and impact the impeller will also result in another acoustic signature, of even lower frequency.
-Again, at lower flow rates, these vortices will tend to develop and shed more slowly, lowering frequencies further.
-
-Consequently, whilst for certain types of noise which tend to become both louder and lower in frequency with increasing speed (cavitation), flow separation provides a contrasting characteristic, in that it is likely to decrease in frequecy with decreasing speed.
-As lower frequencies tend to propagate much further in open water, this could have a significand adverse impact on the total acoustic signature of the submarine, even if the flow separation doesn't lead to any cavitation.
-
-The foregoing consideration suggests that the acoustic signature of a jet-powered submarine might not reduce steadily with speed all the way to stationary movement.
-It is plausible that there could be some elevated propulsion signature at very low speeds due to flow separation, and that the submarine's optimal acoustic signature when moving is actually at some slightly higher speed, where smoother flow through the propulsor can be achieved.
-However, given that the amplitude, frequency, and combined impact of those two characteristics cannot be accurately evaluated without substantial further work and more detailed system knowledge, it is also possible that the increase in other noise (including hull flow noise and other machinery) at elevated speeds offsets the improvement in the smoothness of the pumpjet's operation.
-It is also possible that the overall level of noise radiated by the pumpjet, including under conditions when flow separation occurs, is still considered to be sufficiently low as to not be of significant tactical consequence.
-However it is highly unlikely that the accoustic signature of the pumpjet is actually superior to that of an open propeller at very low speeds, when cavitation occurs for neither system, but substantially increased turbulence including flow separation must neccessarily occur for the pumpjet.
-
-\section{Conclusions}
-This is a pretty bad result for the pumpjet.
-
-\printbibliography
+\end{appendices}
 

### 2018-01-15 scripts/build plots for paper.R

@@ -51,7 +51,7 @@ ls <- df %>%
 ls_comp <- ls %>%
   map( ~ .x %>%
          select(kts, eff.jet, eff.prop) %>%
-         run_subs(hotel = 100,
+         run_subs(hotel = 200,
                   total.batt = 500,
                   batt.dens = 0.14,
                   patrol = 2.5,
@@ -236,11 +236,11 @@ df_full_cornered %>% write_rds("extdata/df_full_diff_cornered.rds")
 
 ########## Time to do some indiscretion ratios (IR)  #############
 
-df_filler <- ls_plots$`Set 3`$power_plot_df
+df_filler <- ls_plots$`Set 1`$power_plot_df
 
 df_filler <- df_filler %>%
-  mutate(pair = "wide",
-         hotel = 100,
+  mutate(pair = "mid",
+         hotel = 200,
          battery = 0.14,
          total.batt = 500)
 
@@ -252,31 +252,78 @@ df_full <- df_full %>%
 ### checking df ###
 df_full %>%
   ggplot(aes(x = kts, y = val, col = pair, linetype = cat))+
-  geom_line()
+  geom_line()+
+  facet_grid(.~hotel, scales = "free")
 
 ### calculations to get IRs ###
 df_IRs <- df_full %>%
   mutate(energy.kJ = total.batt*1000*battery*1000,
-         charge.rate.kw = 5500)
+         charge.rate.kw = 5000)
 
 df_IRs <- df_IRs %>%
   filter(cat %in% c("Propulsion Power Drawn Jet", "Propulsion Power Drawn Propeller"))
 
 df_IRs <- df_IRs %>%
   mutate(endurance.hrs = energy.kJ/(val + hotel)/3600,
-         charge.time.hrs = energy.kJ/charge.rate.kw/3600,
-         IR = charge.time.hrs/endurance.hrs)
+         charge.time.hrs = energy.kJ/(charge.rate.kw-(val + hotel))/3600,
+         IR = charge.time.hrs/endurance.hrs,
+         cat = case_when(str_detect(cat, "Jet") == TRUE ~ "Pumpjet",
+                         str_detect(cat, "Prop") == TRUE ~ "Propeller"),
+         level = case_when(cat == "Pumpjet" & pair == "wide" ~ "Low",
+                           cat == "Propeller" & pair == "wide" ~ "High",
[…]

### 2018-01-21 DraftExec.Rnw
axial mixed update, and john and jim edits

@@
 
-It is worthwhile explaining briefly the origins and intent of this paper, how it came to be written, and how it ought to be read.
-
-The purpose is to investigate the suitability of the pump-jet selection for Australia's future fleet of submarines.
+The purpose of this paper is to investigate the suitability of the pumpjet propulsor technology selection for Australia's future fleet of submarines.
 Amongst this considerable controversy surrounding the submarines, this particular technology choice deserves special attention.
-Much has been made of the significance of the pump-jet in the DCNS (now Naval Group) bid for the submarine, to the extent that it has been claimed that the pump-jet rendered propellers obsolete \parencite{naval2018}, and features extremely prominently in the all the marketing and literature.
+Much has been made of the significance of the pumpjet in the DCNS (now Naval Group) bid for the submarine, to the extent that it has been claimed that the pumpjet rendered propellers obsolete \parencite{naval2018}, and features extremely prominently in the all the marketing and literature.
 It seems to be accepted knowledge that the proposal from Naval Group (formerly DCNS) was selected because of its acoustic superiority \parencite{stewart2016}, with the pumpjet apparently comprising the jewel in the crown.
 
-Having done my higher education in physics, and spent considerable time considering military systems running a small start-up defence company (Rubber Ducky Defence) I found the pumpjet choice particulaly curious.
-As is commonly quipped, including in defence circles, it can reliably assumed when it comes to technology choices "there's no free lunch in physics".
-Whilst all technologies steadily improve, and new ones continuously emerge, it's inevitably the case that significant technology switches are made, some trade-offs are necessary.
-In the case of the pumpjets, the trade-offs were obvious: propulsive efficiency.
+The selection of the DCNS bid to design the future submarine attracted considerable attention and controversy on a number of fronts.
+The choice of the pumpjet, and the question of efficiency at low speeds, has been touched on briefly in various informed discussions \parencite{stanford2017} \parencite{davies2017}.
+However, it has remained the case that even extremely well-informed commentators reluctant to quantify and qualify the plausible scale of this trade-off, and what it might mean strategically and tactically for the submarine.
+Everyone knew it was an issue, but no-one confident enough in the scale of the impact (or even at transit speeds the direction of the impact \parencite{davies2017}) to make it clear whether it mattered.
+Even the head of the progam to build the submarine acknowledged an improvement in dived endurance would be possible with conventional propellers \parencite{sbs2017}, which seemed a clear acknowledgment that some trade-off was acknowledged, even by the strongest advocates of the pumpjet.
+But despite this, the chiefs of DCNS and SEA 1000 both remain adamant that the pumpjet was an "integral part of our bid" \parencite{guillou2017} and that it can be efficient "across the entire speed range" \parencite{sammut2017}.
+
+Having done my higher education in physics, and spent considerable time considering military maritime technology running a small start-up defence company I found seemingly unresolved series of exchanges regarding the pumpjet choice particulaly curious.
+As is commonly quipped, including in defence circles, it can be reliably assumed when it comes to technology choices "there's no free lunch in physics".
+Whilst all technologies steadily improve, and new ones continuously emerge, it's inevitably the case that when significant technology switches are made, some trade-offs are necessary.
+In the case of the pumpjets, the trade-offs what might have been expected appeared obvious, and probably unavoidable to some extent to all those experienced in maritime systems: it would have lower propulsive efficiency, particularly at low speeds.
 
 It's definitely true that efficiency isn't everything in military technology.
-For Rubber Ducky, I spent considerable time and money designing a small speed-boat and a medium-weight torpedo for naval combat, and in both cases I selected a waterjet or pumpjet for those purpose.
+In recent years I was heavily involved in designing a small speed-boat and a medium-weight torpedo for naval combat, and in both cases I selected a waterjet or pumpjet for those purpose.
 Pumpjets, or some kind of "decelerating ducted propeller" (explained in detail later) suited very well the sorts of speeds and operational circumstances that they were required to operate in, which were well over 30kt.
@@
 
-The selection of the DCNS bid to design the future submarine attracted considerable attention and controversy on a number of fronts.
-The choice of the pumpjet, and the question of efficiency at low speeds, has been touched on briefly in various informed discussions \parencite{stanford2017} \parencite{davies2017}.
-However, it has remained the case that even extremely well-informed commentators reluctant to quantify and qualify the plausible scale of this trade-off, and what it might mean strategically and tactically for the submarine.
-Everyone knew it was an issue, but no-one confident enough in the scale (or even direction \parencite{davies2017}) of the impact to make it clear whether it mattered.
-Even the head of the progam to build the submarine acknowledged an improvement in dived endurance might be possible with conventional propellers \parencite{sbs2017}, which seemed a clear acknowledgment that some trade-off was acknowledged, even by the strongest advocates of the pumpjet.
-
-The development of the full substance of this paper has come about due to the determination and interest of Gary Johnson.
+The development of this line of inquiry into of the full substance of this paper has come about not due to my own intrigue, but largely due to the determination and interest of Gary Johnson.
 Through mutual networks in industry and academia we discovered that we had both been intrigued by the technology choice, and considered it of potential significance to Australia's national interest.
 Australia is fortunate to have citizens so concerned with national interest to so substantially invest in research to advance public debate.
-Unless he had commissioned me to write it, there is no way that I would have been able to make the sustained and deliberate research which the topic required to reach a useful level of quantification and confidence to meaningfully inform debate.
-
-I should hasten to add that my expertise in this area is derived largely from an academic background which a different field of physics (particle physics), which enhanced only my aptitude for understanding and analysing physical concepts, paired with a substantial exposure to the relevant disciplines and industry from a commercial perspective as an entrepreneur developing new products.
+Unless he had commissioned me to write it, there is no way that I would have been able to make the sustained and deliberate inquiry which the topic required to reach a useful level of quantification and confidence to meaningfully inform debate.
+
+I should hasten to add that my expertise in this area is derived largely from an academic background whithin a different field of physics (particle physics), which enhanced only my aptitude for understanding and analysing physical concepts, paired with a substantial exposure to the relevant disciplines and industry from a commercial perspective as an entrepreneur developing new products.
 I don't claim to be an authority, but reasonably capable of grasping the works and words of those who are.
@@
 It is quite possible that there are other effects and phenomena which could be relevant which aren't mentioned here, and it's also true that additional advances in the state of the art are being made continually, and might not be known about in public literature.
-
-However, the nature of scientific discovery and technical progress is that many of best discoveries are actually built-upon and confirmed, rather than swept aside, by later developments.
-As such, plenty of foundational principals, in particular the conservation laws of energy and momentum and essential principals of mechanics and thermodynamics are just as true, right, and relevant as they were hundreds of years ago.
+If other subject matter experts have additional or better insights into the topic than those presented here, I would welcome their further comment and contribution to debate.
+
+However, the existence of unknown developments and advances does not mean that realistic approximate bounds and limits of technologies cannot be found from public inquiry.
+Scientific discovery and technical tends to confirm many of the best discoveries are actually built-upon and confirmed, rather than simply sweep them aside  As such, plenty of foundational principals, in particular the conservation laws of energy and momentum and essential principals of mechanics and thermodynamics are just as true, right, and relevant as they were hundreds of years ago.
 Consequently, whilst we cannot always know the precise degree of advance in a particular field of engineering, we can still confidently know some of the fundamental bounds and constraints within which that field will operate.
+My research into this topic has confirmed that this is certainly the case in the relevant arts of hydrodynamics and turbomachinery, where a number of canonical works from the middle of last century are still very frequently referenced in the most modern literature for outining the foundational principals.
 
@@
 
-The selection of the pumpjet propulsion system is quite probably the most significant new technology selection for Australia's future submarines.
-Unlike other new technolgoies such as Air Independent Propulsion and Lithium Ion Batteries, which are either in service or entering service in multiple other navies, no other conventionally powered submarines have adopted pump-jet technology, which is currently only used by nuclear submarines.
+The choice of the pumpjet propulsion system is quite probably the most significant new technology inclusion for Australia's future submarines.
+Unlike other modern technolgoies such as Air Independent Propulsion and Lithium Ion Batteries, which are either in service or entering service in multiple other navies, no other navies have have adopted pumpjet technology for their conventional sumbarines, which is currently only used by nuclear submarines.
 
@@
 \\
-\emph{In a comparison between two otherwise identical submarines, the one with the pump jet will alwayas have a lower dived endurance, a lower dived range, a worse indescretion ratio, a lower overall endurnace, and a lower overall range, than the one with a conventional propeller.
+\emph{In a comparison between two otherwise identical submarines, the one with the pump jet will always have a lower dived endurance, a lower dived range, a worse indiscretion ratio, a lower overall endurace, and a lower overall range, than the one with a conventional propeller.
 This will confer a substantial tactical and strategic advantage on the conventionally propelled submarine in a very broad range of operational scenarios.}
@@
 One informed commentator suggested that the crossover point was likely to be below or near the transit speed of the submarine, and that the pumpjet may offer enhanced range and endurance over the whole mission profile as a consequence \parencite{davies2017}.
-This report finds no evidence of a pumpjet of the type used on submarine every being recommended as superior for efficiency in general, and even closely related technologies for more commecial use not advocated below 25kt.
+This report finds no evidence of a pumpjet of the type used on submarine ever being recommended as superior for efficiency in general, and even closely related technologies for more commecial use not advocated below 25kt.
 During my research I encountered domain experts confident that a pumpjet could simply not feasibly be superior to a suitable propeller at speeds below 18kt.
@@
 Consequently, the possibility of some cross-over point leading to any kind of range or endurance advantage relevant for a conventional submarine can be safely ruled out.
-
-This report also explores in some detail the theoretical concepts which drive the limitations of pumpjet performance at lower speeds in comparison to propellers.
+The basis for claims made by public officials to the contrary, such as those by Rear Admiral Sammut \parencite{sammut2017}, must be called into question.
+
+This report also explores in some detail the theoretical concepts that drive the limitations of pumpjet performance at lower speeds in comparison to propellers.
 Enduring and fundamental principals of momentum theory provide robust limitations on the circumstances where a propeller's 'ideal efficiency' might be improved by the addition of a duct/or shroud, one type of which constitutes the transition to a 'pumpjet'.
@@
 
-Furthmore, a particular hydrodynamic phenonomon called 'flow separation' can be identified as the primary cause of the pumpjet's declining efficiency at very low speeds.
-Put simply, the duct provides an additional hurdle or barrier though which the water must pass when moving through the propuslor.
-At very low speeds waterspeeds, this additional resistance is sufficient to stop some of the flow progressing through the jet, creating unsteady eddies and vortices which sap energy from the system, but produce no thrust, driving down efficiency to zero at very low flow-rates.
+Furthmore, a particular hydrodynamic phenonemon called 'flow separation' can be identified as the primary cause of the pumpjet's declining efficiency at very low speeds.
+Put simply, the duct provides an additional hurdle or barrier though which the water must pass when moving through the propulsor.
+At very low waterspeeds, this additional resistance is sufficient to stop some of the flow progressing through the jet, creating unsteady eddies and vortices which sap energy from the system, but produce no thrust, driving down efficiency to zero at very low flow-rates.
 
@@
 It is this elevated pressure at the point where the blades to their work on the water which enables the blades to avoid cavitation (the rapid expansion and collapse of small voids or bubbles in the water) which is very large source of noise underwater.
-Maintaing this advantage to the same level necessitates a duct design which slows water down to the same extent.
+Maintaining this advantage to the same level necessitates a duct design which slows water down to the same extent.
 The act of slowing water down when it's already going slowly is what causes some of it to stop, which is when flow separation occurs.
@@
 
-The model has been made available publically in the form of a web-app, so that alternative assumptions, including alternative efficiency curves may be tested by the general public.
-
-The model in its central assumptions was able to achieve very realistic results, which were consistent with other public estimates of performance parameters of conventional submarines and known characteristics \ref{buckingham2008submarine}.
+The model has been made available publically in the form of a web-app, so that alternative assumptions, including alternative efficiency curves may be tested by the interested parties.
+
+The model in its central assumptions was able to achieve very realistic results, which were consistent with other public estimates of performance parameters of conventional submarines and their known characteristics \ref{buckingham2008submarine}.
 A variety of different assumptions regarding the hotel load, battery density and different combinations of efficiency curves were tested.
 The central assumption values showed that a submarine with a conventional propeller had a very substantial advantage in terms of range and endurance across the entire speed range.
-This advantage reached a proportional maxium at around 6kt of over 60\%.
+This advantage reached a proportional maximum at around 6kt of over 60\%.
 However, in nominal terms, the greatest differences occurred at slightly lower speeds.
@@
 The model was used to test a wide range of different assumptions.
-A further key result of the study established the sensitity of the difference to different assumptions regarding hotel load, and battery energy density.
+A further key result of the study established the sensitivity of the difference to a range of assumptions regarding hotel load, and battery energy density.
 This showed that the difference between the two systems was made substantially larger with smaller hotel loads, and larger stored electical energy (as might be achieved with lithium-ion batteries), and these two effects combined to give some particularly startling results.
@@
 
-The model demonstrated clearly that all of the most pronounced advantages in dived range and endrance occurred at speeds around 3-7kt, generally within plausible speeds achievable with modern AIP systems.
+The model demonstrated clearly that all of the most pronounced advantages in dived range and endurance occurred at speeds around 3-7kt, generally within plausible speeds achievable with modern AIP systems.
 The addition of AIP was briefly considered which revealed that a difference in dived endurance of the order of one month, or thousands of nautical miles, would likely have emerged between the two propulsion systems.
@@
 The vortexes and eddies which are created by flow separation tend to be unsteady, and shed periodically into the main flow.
-This results in pressure fluctations inside the shroud and more intense, unsteady large-scale types of turbulence occurring as these vortices are ingested into the rotor blades.
-These effects are are likely to cause some accoustic signature which becomes lower in frequence as the submarine slows down further and the vortexes become larger, and shed more slowly.
-This is in contrast to early stages of cavitation, which would be high frequency at when bubble sizes are small, and move to lower fequences as speed increases and bubbles expand.
+This results in pressure fluctuations inside the shroud and more intense, unsteady large-scale types of turbulence occurring as these vortices are ingested into the rotor blades.
+These effects are are likely to cause some accoustic signature which becomes lower in frequency as the submarine slows down further and the vortexes become larger, and shed more slowly.
+This is in contrast to early stages of cavitation, which would be high frequency at when bubble sizes are small, and move to lower fequencies as speed increases and bubbles expand.
 Lower frequencies are of particular significance because they tend to propagate further, and would be less effectively shielded by the duct.
@@
 However, it is extremely unlikely that at some very low speeds, where conventional propellers experience no cavitation and enjoy steady, smooth flows over the blades, that a pumpjet could actually have a lower acoustic signature, even in terms of radiated noise.
-Flow separtion demand that two or three times as much energy is wasted by a pumpjet creating and destroying these unsteady vortexes than a propeller would require to create equal thrust.
+Flow separation demands that two or three times as much energy is wasted by a pumpjet creating and destroying these unsteady vortexes than a propeller would require to create equal thrust.
 Consequently, the claim that pumpjets are generally acoustically superior should be treated with some caution.
@@
 With something like 400 times as much energy per kg embarked in diesel, operating on batteries imposes an extreme demand for economy on propulsive power.
-Happily for submarines, the square law for drag, and cubed law for power, allow an almost commensurate reduction in power demands to take place by slowing down to very slow speeds when submerged.
+Fortuitously for submarines, the square law for drag, and cubed law for power, allow an almost commensurate reduction in power demands to take place by slowing down to very slow speeds when submerged.
 Hypothetically, a conventional submarine might have a total range of 10,000nm from using it's diesel payload at 8kt.
@@
 However, it is plausible that through the development of some instability, perhaps through an unknown fault or dramatic power reductions of the type which I described should be avoided, it might be the case that the reactor would need to be shut down, but could plausibly be restarted.
-However, as discussed above, the restart would need to occur either within minutes, or probably in a couple of days.
+However, as discussed above, the restart would need to occur either within minutes, or probably wait for a couple of days.
 
@@
 A ducted propeller is a propeller in any kind of tubular shroud, casing or nozzle.
-Ducted propellers fall into two categories, one of which is referred to frequently and interchangeably as a pump-jet.
+Ducted propellers fall into two categories, one of which is referred to frequently and interchangeably as a pumpjet.
 The two types are called the accelerating type (also known as a Kort Nozzle, after a company that made them popular in the 20th Century) and the decelerating duct.
@@
 The key concepts behind the operation of these systems, however, remains essentially similar in their nature and design intent.
-In the presence of a duct propellers might be advantageously shaped differently, including maintaing a larger chord lenght right to the shroud, rather than tapering as open propellers tend to lessen vorexs and cavitation on the extremities.
+In the presence of a duct propellers might be advantageously shaped differently, including maintaing a larger chord lenght right to the shroud, rather than tapering as open propellers tend to in order to lessen vortices and cavitation on the extremities.
 However, there is not to my knowledge any single defining physical characteristic which the transition to the name 'impeller' implies, which woud not be effectively equivalent to specialised propellers for decelerating ducts.
@@
 
+\subsection{Pump Types} \label{pumptypes}
+There are broadly three categories of pump types for water which are used for propulsion.
+The first is called an axial-flow.
+In such a pump, the water has work done on in by blades which push the flow in the same direction to the axis of the blades rotation.
+This is what a decelerating ducted propeller comprises.
+The blades of the rotor push the water from the front, to the back, in the same direction as the craft moves, by blades that turn about an axis that is also oriented in that direction.
+These types of pumps are generally best for handling very large flows of water, with relatively less head (energy added) per amount of water flowing through, and are consequently generally desired for propulsion systems, where moving water at high velocity, rather than elevating it to high pressure, is the end objective of the system.
+
+The contrasting system is that of a radial or centrifugal pump.
+In one of these systems, work is done on the water by spinning outwards, from close to the axis of rotation to the extremity.
+In such a case, the axis of the rotor is actually perpendicular to the direction of flow of the water exiting the pump.
+The pumps are generally favoured for pump systems which need to add quite a large amount of head, mostly in the form of pressure, to a given amount of flow.
+Consequently they are desirable for fire pumps, and or other systems which have to raise water to a high pressure, including lifting water up to great heights.
+However, centrifugal pumps were originally the type of pump used in the first waterjets created by Hamilton \parencite{hamilton1997}, and are still used in some waterjet applications today, for example in ouboard waterjets \parencite{yamaha2018}.
+
+The third major class of pumps are called mixed-flow pumps, because these types of pumps involve the water having work done on it whilst flowing in both the axial and radial directions.
+In overall appearance, they tend to reasonably closely resemble that of an axial flow pump, since the axis of the rotor is parallel to both the direction of water intake and outlet.
+However, the water is also accelerated in an outwards direction by the impeller, by means of the hub expanding (and the outer shroud also expanding) in that section where the rotor does work on the water flow.
+The outwards flow is subsequently deflected back towards the axis by the shape of the shroud contracting towards the nozzle, and the hub tapers to a point in this section also.
+The main difference between these can be seen in Figure \ref{fig:AxialMixedHub.png}.
+
+\begin{figure}[h]
+
+\begin{subfigure}{0.5\textwidth}
+\includegraphics[width=0.9\linewidth]{AxialMixed.png}
+\caption{}
+\end{subfigure}
+\begin{subfigure}{0.5\textwidth}
+\includegraphics[width=0.9\linewidth]{MixedHub.png}
+\caption{}
+\end{subfigure}
+
+\caption{Mixed flow pumps push the water partly outwards as the water flows through the impeller blades, in a part-way step towards a fully centrifugal or radial flow, as shown in (a) from \cite{carlton2007}.
+Different hub shapes can induce different levels of radial flow or axial flow in the pump, as seen in (b) from \cite{hamilton1997}}
+\label{fig:AxialMixedHub.png}
+\end{figure}
+
+The physics of of the fluid flows through the impeller of a mixed-flow pump are quite different to that of a pure-axial pump (the flows must be modeled in three rather than two dimensions), and blade design necessarily changes somewhat to accomodate this type of change.
+However, the core limitations and physics surrounding duct design before and after the impeller remain in place for mixed-flow pumpjets, with the primary additional consideration being that the pump section of the duct will tend to need to be somewhat more bulbous in shape to accommodate the larger hub.
+As such, differences in the pump efficiency (how well the impellers add energy to the flow) at any particular flow rate might not represent a large impact on the overall efficiency of the propulsor, as the losses due intake duct and nozzle will comprise separate and independent considerations, quite independent of the efficiency of the pump design.
+
+Mixed flow pumps also can experience flow separation or stall in just the same way that this occurs in pure axial-flow pumps \ref{li2013}, and the similar design considerations (such as optimal flow rate) impact the extent to which this occurs.
+Similarly, the design considerations for reduction of cavitation in mixed-flow pumps tend to be similar, including increasing blade area, and similar efficiency trade-offs are also generally found \parencite{hamilton1997}
+
+Mixed flow pumps can have a variety of different degrees of radial flow over the impeller, and a variety of different designs are found in the pumps in waterjets which vary from pure axial to mixed-flow designs.
+This is because the optimal efficiency of mixed-flow pumps tend to lie between the optimal efficiencies of axial and centrifugal pumps, across an intermediate range of flow requirements.
+Many waterjet manufacturers offer a combination of both pure axial and mixed-flow designs for various applications, and often considerations other than efficiency, including machinery size and shape, can be even more significant than pure efficiency considerations.
+Detailed discussions of the differences between pure axial-flow and mixed-flow pumps can be found in \cite{wislicenus1986}.
+
 \newpage  \section{Constraints on the efficiency of pumpjets at low speed}
@@
 Recently Andrew Davies published a piece in the The Strategist (ASPI) arguing that the pump jet may actually save fuel at transit speeds (around 8kt) and overall improve the endurance of the submarine, despite likely having lower efficiency at low speeds such as patrol speed \parencite{davies2017}.
-This argument relies on pump-jet efficiency still overtaking that of propellers at what is considered a relatively low speed for many applications.
+This argument relies on pumpjet efficiency still overtaking that of propellers at what is considered a relatively low speed for many applications.
 As Andrew acknowledges, further quantification of the likely efficiency is required, as the particular chart which he cites is not quantified.
-This section aims to identify as far as is reasonably possible from public literature how the efficiency of a likely pump-jet for a submarine might perform in quantitative terms, relative to a propeller.
+This section aims to identify as far as is reasonably possible from public literature how the efficiency of a likely pumpjet for a submarine might perform in quantitative terms, relative to a propeller.
 
@@
 \begin{figure}[h]
-\captionsetup{width=1\textwidth}
-\includegraphics[width=1\textwidth]{JetHighFlow.png}
+\captionsetup{width=.8\textwidth}
+\includegraphics[width=.8\textwidth]{JetHighFlow.png}
 \caption{At high flow rates the water has sufficient velocity (indicated by arrow length) to overcome the adverse pressure gradients created by the duct shape, shown by the change from green to red.
@@
 \begin{figure}[h]
-\captionsetup{width=1\textwidth}
-\includegraphics[width=1\textwidth]{JetLowFlow.png}
+\captionsetup{width=.8\textwidth}
+\includegraphics[width=.8\textwidth]{JetLowFlow.png}
 \caption{At low flow rates the water has insufficient velocity to overcome the adverse pressure gradients created by the duct shape, and flow separation begins to occur on the duct intake, and between the blades beginning near the roots (stall).
@@
 At high speeds, where cavitation tends to have severe impacts on propulsive efficiency of open propellers, and high mass-flow (which leads to good efficiency) can be achieved through compact ducts and pumps, pumpjets offer very good performance on both metrics.
-Hence, in this regime the two design requirements tend to align relatively well in favour of pump-jets.
+Hence, in this regime the two design requirements tend to align relatively well in favour of pumpjets.
 However, in very low-speed, low-load environments, where cavitation effects are negligible, and larger areas need to be engaged achieve good mass-flow and efficiency, the two design requirements align relatively well in favour of an open propeller.
@@
 Open propellers may advantageously choose to lower their values of $\Delta V / V_0$ as low as 0.1 in order to obtain ideal efficiencies \parencite{wislicenus1973}.
-This still provides a hard limit on potential improvements that can be attained for a pump-jet that has non-zero duct loss.
+This still provides a hard limit on potential improvements that can be attained for a pumpjet that has non-zero duct loss.
 
@@
 
-Given the fundamental nature of the flow-rate and intake area to the design of pump-jets, and the great lenght of time for which they have been understood and discussed as a crucial design input for a waterjet, it is altogether implausible that some large increase in intake area can be achieved without incurring necessary and unavoidable tradeoffs.
+Given the fundamental nature of the flow-rate and intake area to the design of pumpjets, and the great lenght of time for which they have been understood and discussed as a crucial design input for a waterjet, it is altogether implausible that some large increase in intake area can be achieved without incurring necessary and unavoidable tradeoffs.
 
@@
 As such, these types of propellers are not considered for the military purposes which require an absolute minimum of noise to be created.
-Transitioning to an accelerating duct signify a shift away from a pump-jet, and necessarily be in direct tension with the need to minimise cavitation for acoustic considerations.
+Transitioning to an accelerating duct signify a shift away from a pumpjet, and necessarily be in direct tension with the need to minimise cavitation for acoustic considerations.
 
@@
 
-In order to begin to model the likely impact of propulsion system choices on system peformance characteristics such as dived range and endurance, it is necessary to adopt some plausible curves indicating likely levels of efficiency across the operating speed range, in this case assumed to be up to 20kt.
-
-In this case, three possible curves have been selected for each system, representing a high, low, and central assumption for each, with peak efficiency for the propeller at about 65\% at near 5kt, and jet efficiency around 27\% at the sames speed.
+In order to begin to model the likely impact of propulsion system choices on system peformance characteristics such as dived range and endurance, it is necessary to adopt some plausible curves indicating likely levels of efficiency across the operating speed range, in this case assumed to be up to 20kt.In this case, three possible curves have been selected for each system, representing a high, low, and central assumption for each, with peak efficiency for the propeller at about 65\% at near 5kt, and jet efficiency around 27\% at the sames speed.
+
 The central assumption has been taken in both cases from the study undertaken by BMT and Rolls Royce, as in \cite{giles2010}, and indicated in \ref{fig:BMTShape.png}, since this represents the most direct comparison made as a simple function of waterspeed, with comparable technologies.
-The technologies naturally will not be identical.
-It is probable that the pumpjet of the submarine will be a purely axial-flow pump, as opposed to a mixed flow pump, but this difference in turbomachinery is unlikely to result in a overall loss of efficiency \parencite{hamilton1997}.
-The intake duct may be somewhat longer, increasing duct losses, but the degree of diffusion in the duct may be lower, reducing the adverse pressure gradient.
+It is probable that the pumpjet of the submarine will be a purely axial-flow pump, as opposed to a mixed-flow pump, but this difference in turbomachinery is unlikely to result in a substantial loss of propulsor efficiency, as discussed in Section \ref{pumptypes}.
+Having the outflow of the waterjet released entirely underneath the surface (as opposed to just above it, in the case of most waterjets) certainly removes the most substantive difference that might drive significant changes in the flow-rates, diameters, and other significant determinants of the machinery characteristics that might normally distinguish different types of pumps.
+In addition, the intake duct may be somewhat longer, increasing duct losses, but the degree of diffusion in the duct may be lower, reducing the adverse pressure gradient.
+In other words, it is plausible that different individual components may incur quite different levels of certain types of losses, but it is unlikely that the overall efficiency profile of the system as a whole is dramatically altered assuming competent design, since an enclosed intake controlling the waterspeed at the impeller, and contracting nozzle beyond will remain common features of both.
+Indeed, it is difficult to imagine why the US Navy, having already attempted a pure axial-flow pumpjet directly in place of propellers for similar ship for the same purpose (reduction of cavitation-induced noise for Anti-Submarine Warfare) in 1965 \parencite{harvie1965construction}, would support a study 40 years later using a pump type that was expected to have substantially reduced efficiency at the intended operating speeds.
+
+Small design difference between the propellers used on submarines and surface ships may also exsit.
 It is probable that the the ship propeller would have five blades, where the submarine would likely have seven or nine.
-However, being  for a warship designed to travel at similar speeds to a nuclear submarine, with relatively unconstrained propeller size, and the pumpjet ejecting into the water, the technologies are likely to be far closer than any other publically available comparison.
-Overall, however, its considered that having an experimtally verified curve comparing very similar technologies as a direct function of waterspeed provides a far more certain basis for comparison than attempting to derive different curves \textit{ab initio} using theortical dimesnionless curves attempting to estimate how the proposed designs might correspond to them.
-
-Variations from this central line have been made with consideration for some plausible trade-offs, which are intended to be indicative, but plausible.
-In the case of the high pumpjet assumption, it is assume that a substantially larger intake area has been chosen in order to achieve higher efficiencies at lower speeds, at the cost of peak ideal efficiency, which might be realised not too far above 20kt for such a design.
+However, being for a warship designed to travel at similar speeds to those that a nuclear submarine might, with relatively unconstrained propeller size, the impact of the propeller difference here are is still likely to be very small relative to the contrast with jets.
+
+Overall, however, its considered that having an experimentally verified curve comparing very similar technologies as a direct function of waterspeed, on vessels attempting to optimise for the same characteristic (reduced noise induced by cavitation) provides a far more certain basis for comparison than attempting to derive efficiency curves \textit{ab initio} using theortical dimensionless factors corresponding and our best estimate of a more likely precise design.
+
+Variations from this central line have been made with consideration for some plausible trade-offs errors, which are intended to be indicative of magnitude possible unknowns in system design and their likely impact.
+In the case of the 'high' pumpjet assumption, it is assume that a substantially larger intake area has been chosen in order to achieve higher efficiencies at lower speeds, at the cost of peak ideal efficiency, which might be realised not too far above 20kt for such a design.
 The low assumption for the pumpjet is still represents higher efficiencies at equivalent speeds than those given in \cite{fujisawa1995} for a waterjet.
@@
 It is worthwhile considering that alternative types of battery may be on offer for submarines.
-In particular, lithium ion batteries were offered by TKMS in their bid for SEA 1000, and are being installed in Type 212 submarines operated by the German navy, as well as contracted for the new Norwegian submarines, as well as Israeli submarines.
+In particular, lithium ion batteries have been selected by Norway in a selection process which happend almost concurrently with Australia's \parencite{ohff201702}, and would probably have been on offer to Australia if they had been sought.
 Precise energy density of these batteries is not publically available.
@@
 Under central assumptions, the propeller driven submarine can reach over 540nm at any speed between 4 and 5kt.
-The pump-jet driven submarine reaches over 360nm between 3 and 6kt.
+The pumpjet driven submarine reaches over 360nm between 3 and 6kt.
 A difference of 180nm dived range would presumeably be of considerable tactical significance.
@@
 These differences are extremely large, representing plausibly double the expected total ranges and endurances of existing modern diesel-electric submarines.
-Even if such large changes in hotel load and battery capacity are not simultaneously likely these sensitivitiesquite clearly demonstrate how even modest shifts in both hotel load and battery energy density might combine to have very significant impacts on the appearnce of the pump-jet and propeller comparison.
+Even if such large changes in hotel load and battery capacity are not simultaneously likely these sensitivitiesquite clearly demonstrate how even modest shifts in both hotel load and battery energy density might combine to have very significant impacts on the appearnce of the pumpjet and propeller comparison.
 For instance, a hotel load that reduced by 30\%, and a battery energy density increase of 40\% could plausibly have the impact of doubling the nominal differces between the performance of the two propulsive systems.
@@
 One would expect that this would be of considerable tactical consequence.
-These graphs also demonstrate that the difference does not diminish at higher speeds, infact it begins to increase again above 10kt, as the power required for propulsion begins consume much of the power generated by the diesel engines, and efficiency difference strongly impacts the size of the diminishing surplus available for charging.
+These graphs also demonstrate that the difference does not diminish at higher speeds, infact it begins to increase again above 10kt, as the power required for propulsion begins to consume much of the power generated by the diesel engines, and the efficiency difference strongly impacts the size of the diminishing surplus available for charging.
 It is also clear that the sensitivity to the selection of efficiency curves remains quite strong in this case.

### 2018-01-24 DraftExec.Rnw
added one more ref to AFR

@@
 The choice of the pumpjet propulsion system is quite probably the most significant new technology inclusion for Australia's future submarines.
-Unlike other modern technolgoies such as Air Independent Propulsion and Lithium Ion Batteries, which are either in service or entering service in multiple other navies, no other navies have have adopted pumpjet technology for their conventional sumbarines, which is currently only used by nuclear submarines.
+Unlike other modern technolgoies such as Air Independent Propulsion and Lithium Ion Batteries, which are either in service or entering service in multiple other navies, no other navies have have adopted pumpjet technology for their conventional sumbarines, which is currently only used by nuclear submarines \parencite(coorey2016).
 

### 2018-01-27 scripts/build plots for paper.R
couple of minor graphics/caption tweaks, centering images etc.

@@ -278,9 +278,10 @@ df_IRs <- df_IRs %>%
 colnames(df_IRs)[2] <- "system"
 
 df_IRs %>% write_rds("extdata/df_IRs_hotels.rds")
+df_IRs <- read_rds("extdata/df_IRs_hotels.rds")
 
 plot <- df_IRs %>%
-  filter(kts < 13.5) %>%
+  filter(kts < 13.5, level == "Central", pair == "mid", hotel == 100) %>%
   ggplot(aes(x = kts, y = IR, col = level, linetype = system))+
   geom_line(size = 0.8)+
   scale_color_manual(values = c("black", "blue", "red"))+
@@ -310,6 +311,8 @@ df_IRs_long$key <- factor(df_IRs_long$key, levels = c("diff", "prop.diff"),
                            labels = c("Difference in Indiscretion Ratio", "Proportional Difference"))
 
 df_IRs_long %>% write_rds("extdata/df_IRs_hotels_long.rds")
+
+df_IRs_long <-  read_rds("extdata/df_IRs_hotels_long.rds")
 
 plot <- df_IRs_long %>%
   filter(kts < 13.5) %>%
@@ -488,7 +491,21 @@ df %>%
   facet_grid(var ~ pair, scales = "free", switch =  "y")
 
 
+df_full_complete_fac <- read_rds("extdata/df_full_complete_fac.rds")
 
+plot <- df_full_complete_fac %>%
+  filter( pair %in% c("Central Selections")) %>%
+  filter(var %in% c("Efficiency", "Endurance (hrs)", "Range (nm)")) %>%
+  #filter(var == "end_prop", pair %in% c("top", "bottom", "propup", "propdown", "jetup", "jetdown", "wide", "narrow", "mid") ) %>%
+  ggplot(aes(x = as.numeric(kts), y = as.numeric(val), linetype = as.factor(cat), col = level))+
+  geom_line(aes(group = cat))+
+  scale_linetype_manual(labels = c("Propeller", "Pumpjet" ), values = c(1, 2 ) )+
+  scale_colour_manual(values = c("black", "blue", "red") )+
+  scale_x_continuous(name ="speed (kt)")+
+  #scale_y_continuous(name =element_blank())+
+  #labs(linetype = "system")+
+  facet_grid(var ~ ., scales = "free", switch =  "y")
+ggplotly(plot)
 
 # plotly
 

### 2018-01-27 DraftExec.Rnw
couple of minor graphics/caption tweaks, centering images etc.
Run through the Rstudio spell checker for full document.  Now can't compile.  Wish I knew why.

--- 2018-01-21 DraftExec.Rnw
@@
 The purpose of this paper is to investigate the suitability of the pumpjet propulsor technology selection for Australia's future fleet of submarines.
-Amongst this considerable controversy surrounding the submarines, this particular technology choice deserves special attention.
-Much has been made of the significance of the pumpjet in the DCNS (now Naval Group) bid for the submarine, to the extent that it has been claimed that the pumpjet rendered propellers obsolete \parencite{naval2018}, and features extremely prominently in the all the marketing and literature.
+Amongst the controversy surrounding the submarines, this particular technology choice deserves special attention.
+Much has been made of the significance of the pumpjet in the DCNS (now Naval Group) bid for the submarine, to the extent that it has been claimed that the pumpjet rendered propellers obsolete \parencite{naval2018}.
+The pumpjet featured extremely prominently in the all the marketing and literature during the CEP \parenctite{costello2016}.
 It seems to be accepted knowledge that the proposal from Naval Group (formerly DCNS) was selected because of its acoustic superiority \parencite{stewart2016}, with the pumpjet apparently comprising the jewel in the crown.
@@
 The selection of the DCNS bid to design the future submarine attracted considerable attention and controversy on a number of fronts.
-The choice of the pumpjet, and the question of efficiency at low speeds, has been touched on briefly in various informed discussions \parencite{stanford2017} \parencite{davies2017}.
-However, it has remained the case that even extremely well-informed commentators reluctant to quantify and qualify the plausible scale of this trade-off, and what it might mean strategically and tactically for the submarine.
-Everyone knew it was an issue, but no-one confident enough in the scale of the impact (or even at transit speeds the direction of the impact \parencite{davies2017}) to make it clear whether it mattered.
-Even the head of the progam to build the submarine acknowledged an improvement in dived endurance would be possible with conventional propellers \parencite{sbs2017}, which seemed a clear acknowledgment that some trade-off was acknowledged, even by the strongest advocates of the pumpjet.
+The choice of the pumpjet, and the question of efficiency at low speeds, has been touched on briefly in various discussion \parencite{stanford2017} \parencite{davies2017}.
+However, it has remained the case that even extremely well-informed commentators remain reluctant to quantify and qualify the plausible scale of this trade-off, and what it might mean strategically and tactically for the submarine.
+Everyone seems to acknowledge a trade-off exists, no-one seems confident enough to nominate the scale of the impact (or even at transit speeds the direction of the impact \parencite{davies2017}) to make it clear whether it mattered.
+Even the head of the program to build the submarine acknowledged an improvement in dived endurance would be possible with conventional propellers \parencite{sbs2017}, which seemed a clear acknowledgment that some trade-off was present, and was worthy of consideration.
 But despite this, the chiefs of DCNS and SEA 1000 both remain adamant that the pumpjet was an "integral part of our bid" \parencite{guillou2017} and that it can be efficient "across the entire speed range" \parencite{sammut2017}.
 
-Having done my higher education in physics, and spent considerable time considering military maritime technology running a small start-up defence company I found seemingly unresolved series of exchanges regarding the pumpjet choice particulaly curious.
+As someone with some formal training in physics, and having considerable time considering military maritime technology running a small start-up defence company, I found this seemingly unresolved series of exchanges regarding the pumpjet choice particularly intriguing.
 As is commonly quipped, including in defence circles, it can be reliably assumed when it comes to technology choices "there's no free lunch in physics".
-Whilst all technologies steadily improve, and new ones continuously emerge, it's inevitably the case that when significant technology switches are made, some trade-offs are necessary.
-In the case of the pumpjets, the trade-offs what might have been expected appeared obvious, and probably unavoidable to some extent to all those experienced in maritime systems: it would have lower propulsive efficiency, particularly at low speeds.
-
-It's definitely true that efficiency isn't everything in military technology.
-In recent years I was heavily involved in designing a small speed-boat and a medium-weight torpedo for naval combat, and in both cases I selected a waterjet or pumpjet for those purpose.
-Pumpjets, or some kind of "decelerating ducted propeller" (explained in detail later) suited very well the sorts of speeds and operational circumstances that they were required to operate in, which were well over 30kt.
-However, having made that selection, and on several occasions wrestled with adapting the boat to alternative purposes, (such as mine warfare, including the towing of heavy sweeps and decoys) I found myself continually balancing the inclination to design a better, more specialised tool for each job, with the desire to compromise performance consolidate capability into a multi-role, more flexible capability.
-The trade-offs inherent in propulsion systems, and the relative efficency of jets at high speeds and propellers at low speeds was a familiar encounter.
-
-The development of this line of inquiry into of the full substance of this paper has come about not due to my own intrigue, but largely due to the determination and interest of Gary Johnson.
+Whilst all technologies steadily improve, and new ones continuously emerge, it's inevitably the case that when significant technology switches are made, some trade-offs are necessary in one respect or another.
+In the case of the pumpjets, which are claimed to be superior from an acoustic perspective when accelerating to higher speeds, the existence of some trade-off in propulsive efficiency (particularly at low speeds), appeared almost inevitable, potentially quite significant in scale.
+Waterjets and pumpjets are widely held in maritime engineering circles to have their greatest advantage at very high speeds, but to be generally undesirable for low and medium speeds.
+Confirming the validity of that view, and quantifying the scale of the impact, seemed to be an important and obvious 'common-sense' check for the credibility of the design concept put forward by the French Naval Group, and adopted by the Australian Government.
+
+It is important also to acknowledge that assessing trade-offs requires more than a detailed understanding about one specific characteristic.
+It's true that efficiency isn't everything in military technology, at least not in all circumstances.
+In recent years, I was heavily involved in designing a small speed-boat and a medium-weight torpedo for naval combat, and in both cases I selected a waterjet or pumpjet for those purposes, which generally operated at high speeds.
+However, for some roles and operations (including survey or mine-countermeasures) other systems may have been more efficient.
+Balancing other desirable characteristics is an important and complex process.
+In the case of submarines, the circumstances which lead to cavitation, and other drivers of acoustic performance are obviously closely related.
+Consequently, the concurrent consideration of the link between propulsion systems and acoustic performance was a necessary feature of a paper attempting to shed light on this topic.
+In addition, the propagation of efficiency comparisons through to a variety of particular metrics of direct importance to submarine operations and tactics (dived range, endurance, and indiscretion ratio) are also necessary and important to allow meaningful assessment of the trade-offs with other considerations to take place.
+
+I am very glad to acknowledge that the development of this line of inquiry into of the full substance of this paper has come about not due to my own intrigue, but largely due to the support of Gary Johnston.
 Through mutual networks in industry and academia we discovered that we had both been intrigued by the technology choice, and considered it of potential significance to Australia's national interest.
-Australia is fortunate to have citizens so concerned with national interest to so substantially invest in research to advance public debate.
-Unless he had commissioned me to write it, there is no way that I would have been able to make the sustained and deliberate inquiry which the topic required to reach a useful level of quantification and confidence to meaningfully inform debate.
-
-I should hasten to add that my expertise in this area is derived largely from an academic background whithin a different field of physics (particle physics), which enhanced only my aptitude for understanding and analysing physical concepts, paired with a substantial exposure to the relevant disciplines and industry from a commercial perspective as an entrepreneur developing new products.
-I don't claim to be an authority, but reasonably capable of grasping the works and words of those who are.
+Australia is fortunate to have citizens so concerned with national interest to so substantially invest their own resources in research to advance public debate.
+Unless he had commissioned me to write this report, it is unlikely that I would have been able to make the sustained and deliberate inquiry which the topic required to reach the scope and depth required to meaningfully inform debate.
+
+I should hasten to add that my expertise in this area is derived largely from an academic background within a different field of physics (particle physics), which enhanced only my aptitude for understanding and analysing physical concepts, paired with a substantial exposure to the relevant disciplines and industry from a commercial perspective as an entrepreneur developing new products.
+I don't claim to be a global authority on this topic, but reasonably capable of grasping the works and words of those who are, which I have sought to do in my research at some length.
 
@@
 This paper doesn't seek to advance the technical field with any new research or insight.
-It aims only to bring together as neatly and simply as possible those concepts and results which are neccessary and sufficient to give the appropriate weighting to this question.
+It aims only to bring together as neatly and simply as possible those concepts and results which are necessary and sufficient to give the appropriate weighting to this question.
 
@@
 However, the existence of unknown developments and advances does not mean that realistic approximate bounds and limits of technologies cannot be found from public inquiry.
-Scientific discovery and technical tends to confirm many of the best discoveries are actually built-upon and confirmed, rather than simply sweep them aside  As such, plenty of foundational principals, in particular the conservation laws of energy and momentum and essential principals of mechanics and thermodynamics are just as true, right, and relevant as they were hundreds of years ago.
+Scientific discovery and technical progress tends to confirm and build upon many of the best previous discoveries and concepts, rather than simply sweeping them aside.
+As such, plenty of foundational principals, in particular the conservation laws of energy and momentum, as well as many essential principals of mechanics and thermodynamics, are just as true, right, and relevant today as they were hundreds of years ago.
 Consequently, whilst we cannot always know the precise degree of advance in a particular field of engineering, we can still confidently know some of the fundamental bounds and constraints within which that field will operate.
-My research into this topic has confirmed that this is certainly the case in the relevant arts of hydrodynamics and turbomachinery, where a number of canonical works from the middle of last century are still very frequently referenced in the most modern literature for outining the foundational principals.
+My research into this topic has confirmed that this is certainly the case in the relevant arts of hydrodynamics and turbomachinery, where a number of canonical works from the middle of last century are still very frequently referenced in the most modern literature for outlining the foundational principals.
 
 In this paper, since we cannot know the precise details of the particular state of the art in largely classified military programs, I will attempt to be particularly clear about those things for which we can have confidence (often broad principals or relationships which determine trade-offs) and those things which we can't (the precise embodiments in present design).
-To that end, in this paper I've underaken modelling based upon the broad principals which we can have confidence, and conducted sensitivity testing based on a range of plausible values which seem realistic for those things about which we can have less confidence.
+To that end, in this paper I've undertaken modelling based upon the broad principals which we can have confidence, and conducted sensitivity testing based on a range of plausible values which seem realistic for those things about which we can have less confidence.
 I have also, in discussion with experts and through reading the available literature, attempted where possible to identify the most plausible of the uncertain values, in order to advance discussion.
@@
 With the support of the principal sponsor of this paper, the relevant models may be found in a user-friendly format as a web-app to allow their further interrogation and testing for other plausible scenarios.
-In addition, for increased transparency, the relevant code and equations underlying are available on github.
+In addition, for increased transparency, the relevant code and equations underlying are available on GitHub.
 
@@
 The choice of the pumpjet propulsion system is quite probably the most significant new technology inclusion for Australia's future submarines.
-Unlike other modern technolgoies such as Air Independent Propulsion and Lithium Ion Batteries, which are either in service or entering service in multiple other navies, no other navies have have adopted pumpjet technology for their conventional sumbarines, which is currently only used by nuclear submarines.
+Unlike other modern technologies such as Air Independent Propulsion and Lithium Ion Batteries, which are either in service or entering service in multiple other navies, no other navies have adopted pumpjet technology for their conventional submarines, which is currently only used by nuclear submarines \parencite(coorey2016).
 
@@
 \\
-\emph{In a comparison between two otherwise identical submarines, the one with the pump jet will always have a lower dived endurance, a lower dived range, a worse indiscretion ratio, a lower overall endurace, and a lower overall range, than the one with a conventional propeller.
+\emph{In a comparison between two otherwise identical submarines, the one with the pump jet will always have a lower dived endurance, a lower dived range, a worse indiscretion ratio, a lower overall endurance, and a lower overall range, than the one with a conventional propeller.
 This will confer a substantial tactical and strategic advantage on the conventionally propelled submarine in a very broad range of operational scenarios.}
@@
 One informed commentator suggested that the crossover point was likely to be below or near the transit speed of the submarine, and that the pumpjet may offer enhanced range and endurance over the whole mission profile as a consequence \parencite{davies2017}.
-This report finds no evidence of a pumpjet of the type used on submarine ever being recommended as superior for efficiency in general, and even closely related technologies for more commecial use not advocated below 25kt.
-During my research I encountered domain experts confident that a pumpjet could simply not feasibly be superior to a suitable propeller at speeds below 18kt.
+This report finds no evidence of a pumpjet of the type used on submarine ever being recommended as superior for efficiency in general, and even closely related technologies for more commercial use not advocated below 25kt.
+During my research, I encountered one domain expert who was confident that a pumpjet could simply not feasibly be superior to a suitable propeller at speeds below 18kt.
 Evidence of propellers remaining generally superior to waterjets and pumpjets in the speed ranges up to and even over 30kt is abundant.
@@
 This report also explores in some detail the theoretical concepts that drive the limitations of pumpjet performance at lower speeds in comparison to propellers.
-Enduring and fundamental principals of momentum theory provide robust limitations on the circumstances where a propeller's 'ideal efficiency' might be improved by the addition of a duct/or shroud, one type of which constitutes the transition to a 'pumpjet'.
-These principals stronly support the the notion that pumpjets will not and indeed cannot suddenly overcome their efficiency disadvantage at low speeds by means of technical ingenuity or innovation.
-
-Furthmore, a particular hydrodynamic phenonemon called 'flow separation' can be identified as the primary cause of the pumpjet's declining efficiency at very low speeds.
+Enduring and fundamental principles of momentum theory provide robust limitations on the circumstances where a propeller's 'ideal efficiency' might be improved by the addition of a duct/or shroud, one type of which constitutes the transition to a 'pumpjet'.
+These principles strongly support the notion that pumpjets will not and indeed cannot suddenly overcome their efficiency disadvantage at low speeds by means of technical ingenuity or innovation.
+
+Furthermore, a particular hydrodynamic phenomenon called 'flow separation' can be identified as the primary cause of the pumpjet's declining efficiency at very low speeds.
 Put simply, the duct provides an additional hurdle or barrier though which the water must pass when moving through the propulsor.
-At very low waterspeeds, this additional resistance is sufficient to stop some of the flow progressing through the jet, creating unsteady eddies and vortices which sap energy from the system, but produce no thrust, driving down efficiency to zero at very low flow-rates.
+At very low water-speeds, this additional resistance is sufficient to stop some of the flow progressing through the jet, creating unsteady eddies and vortices which sap energy from the system, but produce no thrust, driving down efficiency to zero at very low flow-rates.
 
@@
 
-In order to assess the probable magnitude of the impact of this reduction in efficiency at low speeds, a numerical model was developed, based on known principles of physics, and publically available information relating to the likely technologies involved in the future submarine.
+In order to assess the probable magnitude of the impact of this reduction in efficiency at low speeds, a numerical model was developed, based on known principles of physics, and publicly available information relating to the likely technologies involved in the future submarine.
 The most important and variable assumptions - the efficiency curves of the pumpjet and propeller - were taken from a recent study lead by world-leading engineering companies, supported by the US and Royal Navies, comparing closely related propulsion systems  \parencite{giles2010}.
-Knowing that the systems would not be identical to the propulsors on submarines (the precise design of which is obviously secret), additional curves varying from these assumptions were also modelled in all directions, and the combinations combined and tested.
-
-The model has been made available publically in the form of a web-app, so that alternative assumptions, including alternative efficiency curves may be tested by the interested parties.
+Knowing that the systems would not be identical to the propulsors on submarines (the precise design of which is obviously secret), additional curves varying from these assumptions were also modeled in all directions, and the combinations combined and tested.
+
+The model has been made available publicly in the form of a web-app, so that alternative assumptions, including alternative efficiency curves may be tested by the interested parties.
 
@@
 A further key result of the study established the sensitivity of the difference to a range of assumptions regarding hotel load, and battery energy density.
-This showed that the difference between the two systems was made substantially larger with smaller hotel loads, and larger stored electical energy (as might be achieved with lithium-ion batteries), and these two effects combined to give some particularly startling results.
+This showed that the difference between the two systems was made substantially larger with smaller hotel loads, and larger stored electrical energy (as might be achieved with lithium-ion batteries), and these two effects combined to give some particularly startling results.
 Circumstances where the \textit{difference} in dived range and endurance between the two systems in terms of range and endurance exceeded the \textit{entire} dived range and endurance achieved by many modern conventional submarines frequently arose.
@@
 
-The consequences of flow separation also has acoustic consequences, which this report explores.
+Flow separation also has acoustic consequences, which this report explores.
 The vortexes and eddies which are created by flow separation tend to be unsteady, and shed periodically into the main flow.
 This results in pressure fluctuations inside the shroud and more intense, unsteady large-scale types of turbulence occurring as these vortices are ingested into the rotor blades.
-These effects are are likely to cause some accoustic signature which becomes lower in frequency as the submarine slows down further and the vortexes become larger, and shed more slowly.
-This is in contrast to early stages of cavitation, which would be high frequency at when bubble sizes are small, and move to lower fequencies as speed increases and bubbles expand.
+These effects are likely to cause some acoustic signature which becomes lower in frequency as the submarine slows down further and the vortexes become larger, and shed more slowly.
+This is in contrast to early stages of cavitation, which would be high frequency at when bubble sizes are small, and move to lower frequencies as speed increases and bubbles expand.
 Lower frequencies are of particular significance because they tend to propagate further, and would be less effectively shielded by the duct.
@@
 But in the case of a submarine, which does it's business completely immersed in the ocean, the relevant physics is dominated the skin friction on the hull, which follows a very simple rule and relationship.
-The amount of drag ($F_D$) an objet experiences increases directly in proportion to the surface area $A$.
-For any given object of a certain (unchanging shape) there will be a constant coefficent ($C_D$) which reflects how aerodynamic or hydrodymanic the shape is.
+The amount of drag ($F_D$) an object experiences increases directly in proportion to the surface area $A$.
+For any given object of a certain (unchanging shape) there will be a constant coefficient ($C_D$) which reflects how aerodynamic or hydrodynamic the shape is.
 The drag is also directly proportional to the density of the fluid being moved through, $\rho$.
@@
 It is true that for very small objects, or very viscous fluids, or very slow movements a different apples called Stokes Equation, in the case where Reynolds numbers are less than 1.
-Given that sea-water is not particularly viscous, and sumbarines not particularly small, Reynolds numbers are liekly to be much much greater than 1 (one or two thousand), even when moving at only one or two knots.
+Given that sea-water is not particularly viscous, and submarines not particularly small, Reynolds numbers are likely to be much much greater than 1 (one or two thousand), even when moving at only one or two knots.
 Since it is unlikely that a significant proportion of the flow over the hull will be laminar, we'll use the drag equation in all modeling going forward when considering drag on the hull.}
@@
 The fuel saving from operating even 30\% slower means that around half the total energy is required for a given journey.
-Other incremental costs and inefficiences from operating 'off-design' are frequently outweight by such a dramatic reduction in overall power demand.
+Other incremental costs and inefficiencies from operating 'off-design' are frequently outweigh by such a dramatic reduction in overall power demand.
 
@@
 The same consideration applies to other types of vessels, which might use water-jets.
-Fast pilot-boats, for instance, might spend quite a considerable time manouvering in and out of harbour or alongside at very low waterspeeds.
+Fast pilot-boats, for instance, might spend quite a considerable time manoeuvering in and out of harbour or alongside at very low water-speeds.
 However, their overall fuel consumption is still likely to be dominated by the fast section of their trip to a ship.
 If the in-harbour water-speed was 8kt, but 32kt was the optimum speed for the open-water section of their journey, the fuel requirement for a given distance could still be over ten times greater at 32kt.
-\footnote{The physics described above would say 16 times, but some reduction due to the reduction of surface area from planing might offset this.}  Consequently the owner/operator would be at least ten times as concerned about efficiency at 32kt as at 8kt, (unless the harbour transit was much much longer than the open water section), and have negligible concern about effiencies at 2kt or 3kt, which would rarely amount to more than a couple of percent of total propulsion energy demand.
+\footnote{The physics described above would say 16 times, but some reduction due to the reduction of surface area from planing might offset this.}  Consequently the owner/operator would be at least ten times as concerned about efficiency at 32kt as at 8kt, (unless the harbour transit was much much longer than the open water section), and have negligible concern about efficiency at 2kt or 3kt, which would rarely amount to more than a couple of percent of total propulsion energy demand.
 
@@
 
-Given that submarines almost by definition (since their hull shape doesn't change, and they must carry ballast to make up for any under-loading) are essentially always operating with the same load (total drag force at any given speed), but have to operate over an extremely wide variety of different speeds (a factor of four or more in variance) with severe efficiency concern at all of them, they represent a truely unique engineering question, seldom discussed in commercial applications.
+Given that submarines almost by definition (since their hull shape doesn't change, and they must carry ballast to make up for any under-loading) are essentially always operating with the same load (total drag force at any given speed), but have to operate over an extremely wide variety of different speeds (a factor of four or more in variance) with severe efficiency concern at all of them, they represent a truly unique engineering question, seldom discussed in commercial applications.
 Consequently, there are relevantly few pieces of literature which address the present question directly.
@@
 In contrast, the Collins Class Submarine's main motor is rated at less than 6MW, with designs for the future submarine appearing to be only slightly larger \parencite{patrick2012}.
-It is fair to say that in terms of maximum power ouput, nuclear submarines could have something like 5-10 times as much power at their disposal than diesel-electric submarines would, even at their peak output.
+It is fair to say that in terms of maximum power output, nuclear submarines could have something like 5-10 times as much power at their disposal than diesel-electric submarines would, even at their peak output.
 
@@
 A nuclear submarine might have literally millions of times more energy at its disposal, it is widely regarded as practically unlimited for all intents and purposes on a given voyage.
-Consequently, wheras a nuclear submarine might regularly conduct transits at or near it's peak power, a diesel electric submarine would probably transit at less than half the speed which it could manage at a sprint, (and use less than a quarter of the energy, as discussed earlier) and on patrol it might be operating at a tenth of the maximum speed, and use maybe just 1\% of as much power on propulsion.
+Consequently, whereas a nuclear submarine might regularly conduct transits at or near it's peak power, a diesel electric submarine would probably transit at less than half the speed which it could manage at a sprint, (and use less than a quarter of the energy, as discussed earlier) and on patrol it might be operating at a tenth of the maximum speed, and use maybe just 1\% of as much power on propulsion.
 In this situation, the amount of power that is drawn for lights, CO2 scrubbing, washing, cooking, heating, as well as the electronics driving the combat systems (the 'Hotel Load') might well become significant, and even be as larger or larger than what is required for propulsion.
@@
 The high presence of Xenon reduced the reactor's reactivity.
-But another process works in the opposite direction which offsets this, since there are lots of neutrons available to 'burn off' the Xenon (which absobs the neturons to become a different isotope).
+But another process works in the opposite direction which offsets this, since there are lots of neutrons available to 'burn off' the Xenon (which absorbs the neutrons to become a different isotope).
 This keeps the total level of Xenon in check.
@@
 The first is that the large submarine proposed by DCNS for Australia uses as much or more power for the hotel load at the most frequently used speeds.
-Assuming the authors are referring to speeds between patrol speeds and approaching transit speed, the subsequent analysis will show that this fact may be true, particulalry in the presence of the pumpjet.
+Assuming the authors are referring to speeds between patrol speeds and approaching transit speed, the subsequent analysis will show that this fact may be true, particularly in the presence of the pumpjet.
 However on its own this doesn't constitute an argument worthy of any consideration.
@@
 One more plausible argument is worth consideration, namely that a safety requirement to be able to operate on batteries without the reactor drives a need to minimise hotel load.
-For all the reasons discussed above, it's completely implausible that the reactor would ever be shut down volunarily during operations.
+For all the reasons discussed above, it's completely implausible that the reactor would ever be shut down voluntarily during operations.
 One would think that the sorts of emergencies or accidents which would cause a reactor are likely to also mean that the reactor can't be restarted.
@@
 This puts some serious constraint around type of contingency that needs to be planned for in a reactor shutdown event.
-It might be presumed that surfacing and remaining surfaced for a couple of days is likely to constitute an end to combat operations for a submarine that was located in theatre, presumeably through enemy action.
+It might be presumed that surfacing and remaining surfaced for a couple of days is likely to constitute an end to combat operations for a submarine that was located in theatre, presumably through enemy action.
 However, the implied alternative remaining submerged and continuing combat operations, even at low speeds will essentially necessitate replicating the full power demand of any diesel-electric submarine until such a time as the reactor can be restarted.
@@
 Even a hundred of tonnes of batteries embarked wouldn't change this reality, unless the submarine was literally at the entrance of a home port.
-It seems much more likely that the safety requirement for batteries on a nuclear submarine is most likely to be framed around maintaining key life support systems (CO2 scrubbing for example) for sufficient time to enable some hope for a seach/rescue mission to rescure the crew, in the likely case that the kind of emergency which shut down the reactor was the kind of emergency which meant the submarine might not resurface, such as the case of the Russian nuclear submarine, the Kursk.
+It seems much more likely that the safety requirement for batteries on a nuclear submarine is most likely to be framed around maintaining key life support systems (CO2 scrubbing for example) for sufficient time to enable some hope for a search/rescue mission to rescue the crew, in the likely case that the kind of emergency which shut down the reactor was the kind of emergency which meant the submarine might not resurface, such as the case of the Russian nuclear submarine, the Kursk.
 
@@
 Nuclear submarines will have unconstrained energy reserves, and a vast surplus of spare power capacity when ever operating below top speed.
-The dynamics of reactor management dictate that generally avoiding large fluctuation in power, and strictly avoiding sharp and dramtic reductions, will be dominant considerations in a nuclear submarine's power output.
-To put things simply, the characterisation that nuclear submarines have "power to burn" is generally sound, and would only ever be challenged when the submarine is travelling at flank speed, or the reactor is shut down due to a (presumeably near-catastrophic) emergency.
+The dynamics of reactor management dictate that generally avoiding large fluctuation in power, and strictly avoiding sharp and dramatic reductions, will be dominant considerations in a nuclear submarine's power output.
+To put things simply, the characterisation that nuclear submarines have "power to burn" is generally sound, and would only ever be challenged when the submarine is travelling at flank speed, or the reactor is shut down due to a (presumably near-catastrophic) emergency.
 
@@
 In general they are shaped as an aerofoil, with the convex side being upstream, just as the convex side of a plane wing is above, in order to generate lower pressure and lift as it moves through the air.
-The blades tend to be twised so that they have a higher angle of attack closer to the hub, and lesser closer to the extremities, so that those faster moving sections push their respective parts of the water column at an overall similar speed.
-The are often also thinner towards the extremities, and will be swept backwards as if dragged by their rotation through the water (skew) and also dragged by the ship's movement throug the water (rake).
+The blades tend to be twisted so that they have a higher angle of attack closer to the hub, and lesser closer to the extremities, so that those faster moving sections push their respective parts of the water column at an overall similar speed.
+The are often also thinner towards the extremities, and will be swept backwards as if dragged by their rotation through the water (skew) and also dragged by the ship's movement through the water (rake).
 Whilst the degree of all these characteristics is highly variable for different applications, these are a few of the commonly referred to characteristics which can be varied in order to optimise performance for any given application.
@@
 It is intuitive to think of as something akin to the angle-of-attack of the propeller blade to the water, however this is technically misleading as the movement of the water-column incoming to the blade, as well as the speed of rotation, also have a significant bearing on what the actual angle of attack of the blade ends up being.
-Whilst technically pitch is a distance (measured in meters), it is often expressed and used in formula as a ratio of the diameter of the propeller, to give it a unitless equivalent, as is common in the description of many characteristics of propellers and propulsion systems.
+Whilst technically pitch is a distance (measured in meters), it is often expressed and used in formula as a ratio of the diameter of the propeller, to give it a unit-less equivalent, as is common in the description of many characteristics of propellers and propulsion systems.
 
@@
 
-The selection of the exact blade selection that is used to define pitch must be specified as being at some fraction of the radius from the centre, given the tendancy for the blades to twist.
+The selection of the exact blade selection that is used to define pitch must be specified as being at some fraction of the radius from the centre, given the tendency for the blades to twist.
 This distance is calculated to be the 'moment mean' or a technical derived effective average, which tends to lie between 0.6R and 0.7R.
@@
 Whilst considerably more complex, expensive, and heavier than a traditional fixed-pitch propeller, the ability to vary the pitch of a propeller assists considerably, particularly when the amount of load (resistance) a vessel is expected to face varies considerably.
-Controllable pitch propellers also have advangates in terms of manouverability, since the pitch can be reversed and hence reverse thrust can be produced, without requiring the direction of the power coming from the engines to be reversed.
-This has advantages for ferries which undertake frequent docking manouvres in confined spaces (\cite{MAN2017}, \cite{lewis1988}).
+Controllable pitch propellers also have advantages in terms of manoeuverability, since the pitch can be reversed and hence reverse thrust can be produced, without requiring the direction of the power coming from the engines to be reversed.
+This has advantages for ferries which undertake frequent docking manoeuvres in confined spaces (\cite{MAN2017}, \cite{lewis1988}).
 
@@
 To some extent, a vessel moving through water will always draw water in its wake along with it, which means that in the wake (where the propeller tends to operate to some extent or other) the water incident on the propeller will arrive at a lower speed than that implied by the ship speed \parencite[15]{MAN2017}.
-Another key distinction for the consideration of jets later on is that the speed of advance does not necessarily incporate the movement of water due to the movement induced by the propeller (or duct) itself, which is a distinct effect to that of the wake field.
+Another key distinction for the consideration of jets later on is that the speed of advance does not necessarily incorporate the movement of water due to the movement induced by the propeller (or duct) itself, which is a distinct effect to that of the wake field.
 In \cite[213]{lewis1988} $V_A$ is used specifically to denote the undisturbed stream velocity, since the role of certain ducts can be to accelerate or decelerate the stream prior to the water arriving at the blades of the propeller.
@@
 This represents circumstances where a propeller encounters effectively zero load.
-This might be the case when a eases it's power, and the momentum of the ship temporarily keeps the ship moving without the propeller exerting much or any net thrust, or where alternative propoulsion sources (sails, or other propellers) are keeping a speed higher than what the propeller actually has push backwards on water to maintain.
+This might be the case when a eases it's power, and the momentum of the ship temporarily keeps the ship moving without the propeller exerting much or any net thrust, or where alternative propulsion sources (sails, or other propellers) are keeping a speed higher than what the propeller actually has push backwards on water to maintain.
 Since drag due to friction scales with velocity squared, very low speed do correspond to increasing advance ratios for a given propeller size, as would increases in propeller size for a given speed and pitch ratio.
-In practice however, under self-propuslion conditions for open propellers, slowing down towards does not elevate the advance ratio far enough past beyond peak efficiencies for overall efficiency to fall very far, as can be seen in Figure \ref{fig:EfficiencyBulten.png}
+In practice however, under self-propulsion conditions for open propellers, slowing down towards does not elevate the advance ratio far enough past beyond peak efficiencies for overall efficiency to fall very far, as can be seen in Figure \ref{fig:EfficiencyBulten.png}
 
@@
 Whilst the very low efficiencies here could represent a case of a propeller that is simply too small for the task it is trying to achieve, it is actually a significant and important circumstance to consider for propellers on working vessels such as tugs or trawlers, which exert large forces at very low water-speeds.
-Measurement of the thrust generated in bollard pull, and the power required to achieve this thrust, is still an important metric for such vessels, but the definition of 'useful work' for normal propulsion requieres the thust to be excerted over some distance travelled.
+Measurement of the thrust generated in bollard pull, and the power required to achieve this thrust, is still an important metric for such vessels, but the definition of 'useful work' for normal propulsion requires the thrust to be exerted over some distance traveled.
 
@@
 
-A detailed description of the essential differences between the two can be easily demonstrated by diagram, which show a simple comparison an acelerating and decelerating duct, which have the shape of the aerofoil section which comprises the duct inverted.
+A detailed description of the essential differences between the two can be easily demonstrated by diagram, which show a simple comparison an accelerating and decelerating duct, which have the shape of the aerofoil section which comprises the duct inverted.
 \footnote{It should be noted that some designs attempt to strike a balance between efficiency and cavitation performance of accelerating and decelerating ducts for certain purposes, and in such cases the duct shape might not be so clearly contrasted \parencite{abdel2010}.
@@
 In the case of an accelerating nozzle, the area through which the flow must pass contracts prior to the water arriving at the impeller.
-In the decellerating case, the available area contracts after the water has passed the impeller.
+In the decelerating case, the available area contracts after the water has passed the impeller.
 In some cases, a degree of diffusion is actually incorporated ahead of the impeller, whereby the area actually of the flow actually expands ahead of the impeller, which slows down the flow and increases its pressure according to Benoulli's principal \parencite{wislicenus1973}.
@@
 A further means by which the efficiency of a propeller can be increased is through the reduction of the turbulence created by the blade-tips passing through the open (static) water at high speed.
-Since the two sides of the blades have low and high pressure on them respectively, at the tips where the two faces meet, there there is a tendancy for water to 'spill' from the high pressure side the the low pressure side, creating a vortex which leads to unwanted turbulence.
+Since the two sides of the blades have low and high pressure on them respectively, at the tips where the two faces meet, there there is a tendency for water to 'spill' from the high pressure side the the low pressure side, creating a vortex which leads to unwanted turbulence.
 It is also around this vortex that blade tip cavitation tends to occur.
-By limiting the movement of water around the outside of the blade tip by having a tightly fitted shroud, losses due to this particulary type of tubulence, and the onset of this type of cavitation, or the vortices that occur even when cavitation does not, can be diminished.
+By limiting the movement of water around the outside of the blade tip by having a tightly fitted shroud, losses due to this particular type of turbulence, and the onset of this type of cavitation, or the cortices that occur even when cavitation does not, can be diminished.
 
 However, it is important to note that the existence of an accelerating duct by no means eliminates all cavitation.
-In fact, given that the static pressure at the impeller is necessarily reduced by an accleratint duct, the propensity for cavitation other surface of the propeller is generally increased.
+In fact, given that the static pressure at the impeller is necessarily reduced by an accelerating duct, the propensity for cavitation other surface of the propeller is generally increased.
 In many circumstance cavitation can and does still occur in ducted propellers near the blade tips at design loads \parencite{moulijn2015}.
@@
 A substantial work exploring the efficiencies of a range of accelerating and decelerating ducts can also be found in \cite{oosterveld1970}.
-A common feature of these extensive studies also show that for ducted propellers, including accelerating ducted propellers, the advantage in terms of efficiency is overall restricted to low waterspeeds.
-At higher waterspeeds, the drag induced by the water-movement over the exterior of the duct begins to increase substantially, and the net thrust on the duct becomes negative.
+A common feature of these extensive studies also show that for ducted propellers, including accelerating ducted propellers, the advantage in terms of efficiency is overall restricted to low water-speeds.
+At higher water-speeds, the drag induced by the water-movement over the exterior of the duct begins to increase substantially, and the net thrust on the duct becomes negative.
 When the exterior of the duct experiences relatively little water velocity, but the interior experiences a great deal more, the greatest advantages of an accelerating duct are realised.
@@
 Oosterveld, in his thesis on ducted propellers also states with regard to the decelerating duct "This ducted propeller system is the so called pumpjet" \parencite[8]{oosterveld1970}.
-Other authors give separate treatement to ducted propellers and pump jets, though the systems they describe are essentially similar in terms of composition, advantages, and physics \parencite[288]{lewis1988}.
+Other authors give separate treatment to ducted propellers and pump jets, though the systems they describe are essentially similar in terms of composition, advantages, and physics \parencite[288]{lewis1988}.
 
@@
 
-In other definitive literature on the design of pumpjets \parencite{henderson1964} the difference between a pumpjet and a propeller is a described as being that "the stream of flow through the pumpjet is made to depart from the "natural" or free-stream sufrace that bounds the flow though a standard propeller."  According to the diagrams and the subsequent description it is clear that the definition embraces quite exactly the distinction between an accelerating duct, and a decelerating duct, namely that the flow is slowed at the point it encounters the impeller, where as in a free situation the free stream would contract and accelerate at this point.
+In other definitive literature on the design of pumpjets \parencite{henderson1964} the difference between a pumpjet and a propeller is a described as being that "the stream of flow through the pumpjet is made to depart from the "natural" or free-stream surface that bounds the flow though a standard propeller."  According to the diagrams and the subsequent description it is clear that the definition embraces quite exactly the distinction between an accelerating duct, and a decelerating duct, namely that the flow is slowed at the point it encounters the impeller, where as in a free situation the free stream would contract and accelerate at this point.
 
@@
 The key concepts behind the operation of these systems, however, remains essentially similar in their nature and design intent.
-In the presence of a duct propellers might be advantageously shaped differently, including maintaing a larger chord lenght right to the shroud, rather than tapering as open propellers tend to in order to lessen vortices and cavitation on the extremities.
-However, there is not to my knowledge any single defining physical characteristic which the transition to the name 'impeller' implies, which woud not be effectively equivalent to specialised propellers for decelerating ducts.
+In the presence of a duct propellers might be advantageously shaped differently, including maintaining a larger chord length right to the shroud, rather than tapering as open propellers tend to in order to lessen vortices and cavitation on the extremities.
+However, there is not to my knowledge any single defining physical characteristic which the transition to the name 'impeller' implies, which would not be effectively equivalent to specialised propellers for decelerating ducts.
 In systems tending to be described as pumpjets or waterjets hubs tend to be larger, blade-numbers higher, twist, skew and rake lower.
@@
 The physics of the two systems, however, are closely related, with the very noticeable exception of the absence of flow considerations for water on the exterior of the duct considerations for waterjets.
-However, the physics relating to the impeller and stator blades or (turbomachinery) designed to add energy to the flow are generally extremely closely linked, though a larger variety of diffeent pump types can be chosen for waterjets which might differ more significantly from the natural evolution of a propeller to an axial-flow pump.
+However, the physics relating to the impeller and stator blades or (turbomachinery) designed to add energy to the flow are generally extremely closely linked, though a larger variety of different pump types can be chosen for waterjets which might differ more significantly from the natural evolution of a propeller to an axial-flow pump.
 Many of the essential equations which govern the efficiencies and cavitation performance of pumpjets and waterjets are of the same or extremely similar form, as can be seen in detailed discussions given to both, by similar authors (\cite{wislicenus1973}, \cite{henderson1964}).
@@
 Consequently they are desirable for fire pumps, and or other systems which have to raise water to a high pressure, including lifting water up to great heights.
-However, centrifugal pumps were originally the type of pump used in the first waterjets created by Hamilton \parencite{hamilton1997}, and are still used in some waterjet applications today, for example in ouboard waterjets \parencite{yamaha2018}.
+However, centrifugal pumps were originally the type of pump used in the first waterjets created by Hamilton \parencite{hamilton1997}, and are still used in some waterjet applications today, for example in outboard waterjets \parencite{yamaha2018}.
 
@@
 
-The physics of of the fluid flows through the impeller of a mixed-flow pump are quite different to that of a pure-axial pump (the flows must be modeled in three rather than two dimensions), and blade design necessarily changes somewhat to accomodate this type of change.
+The physics of of the fluid flows through the impeller of a mixed-flow pump are quite different to that of a pure-axial pump (the flows must be modeled in three rather than two dimensions), and blade design necessarily changes somewhat to accommodate this type of change.
 However, the core limitations and physics surrounding duct design before and after the impeller remain in place for mixed-flow pumpjets, with the primary additional consideration being that the pump section of the duct will tend to need to be somewhat more bulbous in shape to accommodate the larger hub.
@@
 It should be noted however, that relatively little literature focuses on the performance of propulsive systems far far outside of their intended design speeds, which tends to be far higher than the patrol and transit speeds of diesel electric submarines.
-Fewer still tend to express their work in terms of ship-speed, instead adopting the convetion of using dimensionless ratios so that work done on one system can be readily translated into different combinations loads, pressures, speeds and sizes.
+Fewer still tend to express their work in terms of ship-speed, instead adopting the convention of using dimensionless ratios so that work done on one system can be readily translated into different combinations loads, pressures, speeds and sizes.
 However, from a few notable studies exist which provide important reference points.
@@
 They also suggest that differences in efficiency could be very large, of a factor of two or more, including at speeds well over a submarine's transit speed.
-However, care must be taken making precise quantitative comparisons, since such experiements involved jets which expelled into air (waterjets) as opposed to remaining under water (pumpjets) do have at least one significant difference in their makeup, though their internal pump mechanisms may often be similar.
+However, care must be taken making precise quantitative comparisons, since such experiments involved jets which expelled into air (waterjets) as opposed to remaining under water (pumpjets) do have at least one significant difference in their makeup, though their internal pump mechanisms may often be similar.
 Specifically, pumpjets would experience drag on the exterior of the shroud/duct, which waterjets would not, however waterjets are likely to have longer intake ducts, which might involve larger duct losses than pumpjets.
@@
 
-At IMARESTS 10th International Naval Engineering Conference, a paper was presented by BMT Defence and Rolls Royce \parencite{giles2010}, outlining a proposal for 'fully submerged waterjets' to be used on future surface ships in order to achieve better acoustic performance, particulalry for an anti-submarine warfare role.
-The jets comprise a mixed-flow pump coupled to an electric motor, and have been both extensively modelled and also tesed in demonstrator vessels.
+At IMARESTS 10th International Naval Engineering Conference, a paper was presented by BMT Defence and Rolls Royce \parencite{giles2010}, outlining a proposal for 'fully submerged waterjets' to be used on future surface ships in order to achieve better acoustic performance, particularly for an anti-submarine warfare role.
+The jets comprise a mixed-flow pump coupled to an electric motor, and have been both extensively modeled and also tested in demonstrator vessels.
 These may be a closer representation of the performance characteristics achievable by submerged pumpjets, since shape of the intake and shroud more closely resembles the shape of a pumpjet that might be found on a submarine.
@@
 
-The paper makes no effort to justify the selection of choice of jets on the efficiency of the propulsion system, instead focussing on improved acoustic performance, and potential machinery space-saving advantages.
-The vessel under consideration would require 250 tonnes more fuel with jets find than with propellers, the jets would require more power to reach equal waterspeed to a conventionally propelled craft right up to 30 kt.
-In the case of this analysis, modifications to the hullform in order to accommodate the jet (which are likely to be more substantial for a surface ship than a submarine) also have an impcat on the resistance offerd by the hull.
+The paper makes no effort to justify the selection of choice of jets on the efficiency of the propulsion system, instead focusing on improved acoustic performance, and potential machinery space-saving advantages.
+The vessel under consideration would require 250 tonnes more fuel with jets find than with propellers, the jets would require more power to reach equal water-speed to a conventionally propelled craft right up to 30 kt.
+In the case of this analysis, modifications to the hull-form in order to accommodate the jet (which are likely to be more substantial for a surface ship than a submarine) also have an impact on the resistance offered by the hull.
 This also has an interaction with the different shape of the propulsors, which alters the required level of delivered (effective) power to achieve a given speed, in addition to the considerations of how efficiently the propulsor generates that effective power.
@@
 If the power could be reduced by 5\% (e.g.
-by fur ther improving the waterjet/hull fairing), then the AWJ-21™ vessel would be better at all speeds above 25 knots."  This statement, however, is made in reference to the hullform needed to accommodate the jet, rather than the jet propulsion efficiency itself \parencite{giles2010}.
+by fur ther improving the waterjet/hull fairing), then the AWJ-21™ vessel would be better at all speeds above 25 knots."  This statement, however, is made in reference to the hull-form needed to accommodate the jet, rather than the jet propulsion efficiency itself \parencite{giles2010}.
 As the propulsive efficiency curves given in the study show, the propulsors themselves are considerably further from reaching parity, still separated by about 10\% at 25kt.
@@
 This strongly suggests the existence of some fundamental constraint in the physics of the operation of jets, which recent advances in technology seems not to have overcome.
-To more firmly establish the extent to which adaptation or plausible improvments in technology could or could not change this trend, it is worthwhile considering the theoretical limits that are presented by the physics.
+To more firmly establish the extent to which adaptation or plausible improvements in technology could or could not change this trend, it is worthwhile considering the theoretical limits that are presented by the physics.
 This should further limit the range of error that might be made in estimating realistic bounds of performance for a system which we do not have precise design specifications for.
@@
 Pumpjets are necessarily low in efficiency at very low speeds because they attempt to get the water to do multiple things.
-They don't just accelerate a water colum.
+They don't just accelerate a water column.
 They first slow it down (increasing its pressure), then add pressure to it, then accelerate it.
@@
 Consequently, the geometry of the duct defines a pressure gradient in the intake and through the nozzle.
-The greater the difference in the cross-sectional areas, the greater the total magintude of the pressure differential that will occur, as necessitated by conservation of mass.
+The greater the difference in the cross-sectional areas, the greater the total magnitude of the pressure differential that will occur, as necessitated by conservation of mass.
 The same amount must go out as came in.
@@
 \captionsetup{width=.8\textwidth}
-\includegraphics[width=.8\textwidth]{JetHighFlow.png}
+\includegraphics[width=.8\textwidth, center]{JetHighFlow.png}
 \caption{At high flow rates the water has sufficient velocity (indicated by arrow length) to overcome the adverse pressure gradients created by the duct shape, shown by the change from green to red.
@@
 \captionsetup{width=.8\textwidth}
-\includegraphics[width=.8\textwidth]{JetLowFlow.png}
+\includegraphics[width=.8\textwidth, center]{JetLowFlow.png}
 \caption{At low flow rates the water has insufficient velocity to overcome the adverse pressure gradients created by the duct shape, and flow separation begins to occur on the duct intake, and between the blades beginning near the roots (stall).
@@
 It is worthwhile noting that the diagrams in Figures \ref{fig:JetHighFlow.png} and \ref{fig:JetLowFlow.png} do not include a hub for the sake of simplicity.
-In practice, the cross-sectional are of the pumpjet naturally only includes that aread between the hub and the duct, and in the case of many pumpjets, particularly those on the ends of torpedoes or submarines, the hub may include/incorporate the tapering tail of the body, for example in Figures \ref{fig:PumpjetOutline.png} or \ref{fig:PumpjetWaterjet.png}.
+In practice, the cross-sectional are of the pumpjet naturally only includes that area between the hub and the duct, and in the case of many pumpjets, particularly those on the ends of torpedoes or submarines, the hub may include/incorporate the tapering tail of the body, for example in Figures \ref{fig:PumpjetOutline.png} or \ref{fig:PumpjetWaterjet.png}.
 Particularly in Figure \ref{fig:PumpjetWaterjet.png}, the appearance of the inner side of the shroud to roughly follow contraction of the hub or body gives the optical appearance of very little diffusion or contraction of the flow occurring, as the one dimensional separation might seem to be roughly constant, or even expanding in places where I have earlier explained it generally contracts.
@@
 Similarly, where the body or hub is quite substantial and also tapering steadily, keeping the duct intake parallel to the axis of the body for a small distance, or even just tapering it proportionally less than the hub, can still result in a considerable expansion in the area, and considerable diffusion.
-The example in Figure \ref{fig:PumpjetOutline.png} shows a case of extremely sharp flow diffusion on the intake, probably more than what is practially realistics to avoid flow separation, even at quite high speeds.
+The example in Figure \ref{fig:PumpjetOutline.png} shows a case of extremely sharp flow diffusion on the intake, probably more than what is practically realistic to avoid flow separation, even at quite high speeds.
 It is important to recognise that this simple relationship with $r^2$ generally means that even moderate changes in terms of diameter, which is what we tend to observe when looking at images and diagrams can actually amount to much more substantial changes in velocity, and hence pressure, which drives the effects described in this section.
@@
 
-For this particular consideration, two questions are of paricular importance.
+For this particular consideration, two questions are of particular importance.
 The first is where the clear transitions in the regimes will tend to lie, with respect to the probable requirements of a conventional submarine.
-The second is what drives the tension between those design requirements in the intermediate regimes, and whether it is a fundamental or inescapble tension in the laws of physics, or simply a technical hurdle which imminent improvements in the state of the art are likely to overcome by engineering improvements or innovation.
+The second is what drives the tension between those design requirements in the intermediate regimes, and whether it is a fundamental or inescapable tension in the laws of physics, or simply a technical hurdle which imminent improvements in the state of the art are likely to overcome by engineering improvements or innovation.
 
@@
 That is not to say that optimal means of balancing those requirements will not still be found, and improved, as the art advances.
-However, it does mean that significantly adapting using known variables that can be changed to favour one design requirement are unlikely to yeild large improvements without some trade-off being made to the other.
+However, it does mean that significantly adapting using known variables that can be changed to favour one design requirement are unlikely to yield large improvements without some trade-off being made to the other.
 
 The reason for this trade-off occurring can be quite simply attributed to the nature of the shroud, or duct, which surrounds the the rotors of the propulsor, and the effect it has on the flow.
-If the duct acts to increase the pressure at the rotor, (which reduces cavitation) it necessarily must slow the flow at/around the rotor, which necessitates produceing a negative thrust on the duct.
-The system as a whole produces a positive thrust, becuase the impeller or rotor pushes a larger positive thrust, but the duct necessarily produces a negative thrust.
+If the duct acts to increase the pressure at the rotor, (which reduces cavitation) it necessarily must slow the flow at/around the rotor, which necessitates producing a negative thrust on the duct.
+The system as a whole produces a positive thrust, because the impeller or rotor pushes a larger positive thrust, but the duct necessarily produces a negative thrust.
 As the speeds get lower, and the thrust required gets much much lower, this negative thrust becomes proportionally more significant, as described later.
@@
 A fuller description of these equations can be found in  most standard texts \parencite[131]{lewis1988}.
-These methods are computationally simple, and allow simple upper-bounds to be be set for efficiencies achievable for given designs under certain conditions, without considering the pottential impact of other more complex effects such as skin friction over the blades, potential flow separation, or cavitation.
+These methods are computationally simple, and allow simple upper-bounds to be be set for efficiencies achievable for given designs under certain conditions, without considering the potential impact of other more complex effects such as skin friction over the blades, potential flow separation, or cavitation.
 As such, they can be easily relied upon to constrain the overall plausible range of values that can be achieved from universally applicable inputs, such as the velocity of water going into a duct, and the velocity of water exiting a nozzle.
@@
 
-The inevitable consequence of this relationshiop is that efficency approaches unity when $\Delta V$ (the change in water of the water between entering and exiting the system) relative to $V_0$ approaches zero.
+The inevitable consequence of this relationship is that efficiency approaches unity when $\Delta V$ (the change in water of the water between entering and exiting the system) relative to $V_0$ approaches zero.
 Intuitive embodiments of this relationship are (equivalently) that efficiency is maximum when the least force is exerted on any given bit of water, which might be when negligibly small work is being done (extreme low speed) or when such a large mass of water is acted on that it barely needs to be pushed at all.
 This relationship would lead to the conclusion that infinitely large propellers or jets are always desirable from an efficiency perspective.
-This of course isn't practical to build, and also doesn't hold when taken to the etremes the effects of the drag on the propeller blades, or duct, are taken into account.
+This of course isn't practical to build, and also doesn't hold when taken to the extremes the effects of the drag on the propeller blades, or duct, are taken into account.
 Infinitely large systems would increase the total weight, as well as having significant surface drag when moving through the water, all of which are neglected in this 'ideal' consideration.
 
-In \cite[8-14]{wislicenus1973}, Wislicenus expands the relationship in the case of the jet to incorporate terms to represent some of the inevitable adjustments that would occur in the case of pump jets and waterjets, particulalry the duct head loss ($K$) (fraction of energy required to push the water through the duct), as well as the additional drag on the outside of the propulsor ($\Delta T$).
+In \cite[8-14]{wislicenus1973}, Wislicenus expands the relationship in the case of the jet to incorporate terms to represent some of the inevitable adjustments that would occur in the case of pump jets and waterjets, particularly the duct head loss ($K$) (fraction of energy required to push the water through the duct), as well as the additional drag on the outside of the propulsor ($\Delta T$).
 This leads to an adjusted efficiency relationship, which still neglects any losses due to pump efficiency, or change in height of the jet (which is added for consideration for waterjets, but not relevant here).
@@
 \includegraphics[width=0.7\textwidth, center]{PumpjetEfficiency.png}
-\caption{Ideal jet efficiencies sharply appoach zero at low values of $\Delta V / V_0$, where as ideal propeller efficiencies without a duct $K = 0$ approach unity. Credit: \cite{wislicenus1973}}
+\caption{Ideal jet efficiencies sharply approach zero at low values of $\Delta V / V_0$, where as ideal propeller efficiencies without a duct $K = 0$ approach unity. Credit: \cite{wislicenus1973}}
 \label{fig:PumpjetEfficiency.png}
@@
 Charts such as these are generally used with dimensionless ratios on the axes (such as $\Delta V / V_0$) to allow the most general solutions to be used to solve for any given particular vessel size, speed, or loading requirement which might be required, these  confusing to many readers unfamiliar with hydrodynamics.
-For the avoidance of any confusion, it is important to note that the value for $\Delta V / V_0$ for a given vessl and propulor may tend to increase with increasing waterspeed of the vessel, however one cannot rely upon this particular value diminishing towards zero at very low speeds, as it does for related by different metrics.
+For the avoidance of any confusion, it is important to note that the value for $\Delta V / V_0$ for a given vessel and propulsor may tend to increase with increasing water-speed of the vessel, however one cannot rely upon this particular value diminishing towards zero at very low speeds, as it does for related by different metrics.
 In an idealised circumstance, the flow rate increases directly with vessel speed.
 Given that the force produced scales with $\Delta V$, which at constant values of $\Delta V / V_0$ increase in proportion to (if, in the idealised circumstance, vessel speed is equal to $V_0$), an increase in thrust can theoretically match the $V_0^2$ requirement imposed by drag without any change in $\Delta V / V_0$.
-However, in practice the water encountered by the propulsor is slowed by the wake field, and so the drag encountered by the vessel may increase faster than velocity of the water entering the intake, an effecnt which is likely non-linear with vessel speed.
+However, in practice the water encountered by the propulsor is slowed by the wake field, and so the drag encountered by the vessel may increase faster than velocity of the water entering the intake, an effect which is likely non-linear with vessel speed.
 So whilst the trend with regards to vessel speed can be known with regards to the graphs in \cite{fig:PumpjetEfficiency.png}, a direct translation of the x-axis used to vessel speed at any given point is difficult prove without additional system information.
@@
 
-For context, the equations developed by Wislicenus are used for the purposes generally of selecting the ideal propulor diameter, which can be made larger in the case of a propeller to access higher ideal efficiencies.
+For context, the equations developed by Wislicenus are used for the purposes generally of selecting the ideal propulsor diameter, which can be made larger in the case of a propeller to access higher ideal efficiencies.
 This is an opportunity which is constrained in the case of jets because the efficiency diminishes at very low values of $\Delta V / V_0$.
@@
 The structure of these curves is found to be overall quite similar for both pumpjets and waterjets, though the probably values for the losses incurred in the intake duct, or on the exterior of the shroud, are obviously different \parencite[13-14]{wislicenus1973}.
-Similar efficiency shapes can be found for discussions focussing explicitly on underwater pumpjets \parencite[13]{henderson1964}.
+Similar efficiency shapes can be found for discussions focusing explicitly on underwater pumpjets \parencite[13]{henderson1964}.
 It should be noted that the theoretical efficiency derived by these general equations is consistent both in derivation and results with those produced by other authors \parencite[227]{lewis1988}.
@@
 Given that the x axis used in the previous charts used $\Delta V / V_0$, which would be related to $V_2-V_1$ in this case, this has the effect of reversing the effective direction of the chart and changing its shape.
-In this case, for a fixed jet and vessel design, high-speed operations will be towards the left of the chart, and the precipitous decline in efficiency will occur as speed declines to zero, along with thrust, toward the right of the chart, which can be known to correspond to waterspeeds declining to zero at $V_1/V_2=1$.
+In this case, for a fixed jet and vessel design, high-speed operations will be towards the left of the chart, and the precipitous decline in efficiency will occur as speed declines to zero, along with thrust, toward the right of the chart, which can be known to correspond to water-speeds declining to zero at $V_1/V_2=1$.
 
@@
 The left side of the chart will correspond to the highest number of rotations for a given amount of progress through the water.
-For vessel of fixed load, and a propeller of fixed geomtry, this will correspond to a propeller facing the higest drag force due to the vessel moving at the highest speed.
+For vessel of fixed load, and a propeller of fixed geometry, this will correspond to a propeller facing the highest drag force due to the vessel moving at the highest speed.
 
@@
 Such circumstances are unlikely to occur to a significant degree unless a vessel was decelerating, or experienced some other force to assist its propulsion.
-It is possible that at extremely low speeds a propeller could somewhat descend to the right of optimal advance ratio, when very dramatically reduced load was encounterd because of the extreme low speed, as might be suggested by the slight reduction in propulsive efficiency of the propeller below 5kt in the BMT study.
-These losses of efficiency, however, could never supercede those of a decelerating ducted propeller or jet, which has the shroud which necessarily imposes an \textbf{additional} drag to the watercolumn in excess of any other resistance faced by the rotor.
+It is possible that at extremely low speeds a propeller could somewhat descend to the right of optimal advance ratio, when very dramatically reduced load was encountered because of the extreme low speed, as might be suggested by the slight reduction in propulsive efficiency of the propeller below 5kt in the BMT study.
+These losses of efficiency, however, could never supersede those of a decelerating ducted propeller or jet, which has the shroud which necessarily imposes an \textbf{additional} drag to the water-column in excess of any other resistance faced by the rotor.
 This effect can be seen clearly even or an accelerating duct, as shown in \ref{fig:PropDuct.png}.
@@
 
-It should be clear from the above analysis that the efficiency of a pumpjet falling towards zero with a vessel's waterspeed is an enduring and foundational result of basic physics.
+It should be clear from the above analysis that the efficiency of a pumpjet falling towards zero with a vessel's water-speed is an enduring and foundational result of basic physics.
 The addition of negative thrust on a decelerating duct, no matter how small, requires that efficiency must decline at some point with with blade loading, in contrast with an ideal propeller which rises to unity.
-This explains the overall shape given by the the mysterious unquantified chart cited by Andrew Davies \parencite{davies2017} and Access Economics \parencite{stanford2017} shown \ref{fig:PumpUp.png}, and is also consistent with the shapes plotted by more recent studies as shown above, and those presented in other modern publications studying the use of pumpjets \parencite{giles2010} and waterjets \parencite{fujisawa1995}.
+This explains the overall shape given by the the mysterious quantified chart cited by Andrew Davies \parencite{davies2017} and Access Economics \parencite{stanford2017} shown \ref{fig:PumpUp.png}, and is also consistent with the shapes plotted by more recent studies as shown above, and those presented in other modern publications studying the use of pumpjets \parencite{giles2010} and waterjets \parencite{fujisawa1995}.
 
@@
 In terms of establishing plausible peaks and cross-over points, the study undertaken in \cite{giles2010} probably provides a sufficient indication as to the current state of the art in the relevant technologies.
-Whilst this effort focussed on a surface vessel rather than a submarine, the study was undertaken with the backing of military customers with support and testing facilities (the US Navy), and involved two major companies who are world leaders in maritime engineering (BMT) and waterjet manufacture (Rolls Royce).
-Even if further technical advances through design refinement were possible, as the authors alude to, the cross-over point in terms of efficiency might be brought down to 25kt.
-The claim that the crossover point lies anywhere near the transit speed of a convetional submarine (8-10kt) appears implausible.
+Whilst this effort focused on a surface vessel rather than a submarine, the study was undertaken with the backing of military customers with support and testing facilities (the US Navy), and involved two major companies who are world leaders in maritime engineering (BMT) and waterjet manufacture (Rolls Royce).
+Even if further technical advances through design refinement were possible, as the authors allude to, the cross-over point in terms of efficiency might be brought down to 25kt.
+The claim that the crossover point lies anywhere near the transit speed of a conventional submarine (8-10kt) appears implausible.
 In discussions with leading world experts involved in the development of advanced ducted propellers (including those involved in citations) I was told that a pumpjet could not be more efficient than a propeller below 18kt, confirming this conclusion.
 
-With the optimum efficiency of a jet probably being no greater than 60\% or 70\% at its peak efficiency according to many theoretical and experimental studies \parencite{lu2016cfd}, which would quite probably lie considerably higher in terms of waterspeed even than 18kt.
-These facts, along with the inevitable decline of efficiency towards zero as waterspeed also declines, offer a suitably constrained set of plausible efficiency curves may be investigated numerically, as will be done later in later sections.
+With the optimum efficiency of a jet probably being no greater than 60\% or 70\% at its peak efficiency according to many theoretical and experimental studies \parencite{lu2016cfd}, which would quite probably lie considerably higher in terms of water-speed even than 18kt.
+These facts, along with the inevitable decline of efficiency towards zero as water-speed also declines, offer a suitably constrained set of plausible efficiency curves may be investigated numerically, as will be done later in later sections.
 
@@
 
-As noted in \cite{henderson1964}, the key constraint of this variable is that increasing the diameter of the pumpjet necessarily leads to an increase in the surface (including the external suface) of the shroud.
+As noted in \cite{henderson1964}, the key constraint of this variable is that increasing the diameter of the pumpjet necessarily leads to an increase in the surface (including the external surface) of the shroud.
 The increased drag that this induces must be deducted from the thrust that the pumpjet produces, or effectively increasing $\Delta T$ as given in earlier equations.
@@
 A further discussion of this effect can be found in \cite[15,27]{henderson1964},  \cite[807,8012]{bruce1974},  \cite[185]{wislicenus1973}, \cite[60]{mcbride1979} and many other related texts.
-This is a general problem that affects a range of axial-flow turbomachinery, particulalry operating in off-design conditions, and will present a general concern for the design of a new pumpjet \cite[185]{wislicenus1986}, \cite{li2013}.
+This is a general problem that affects a range of axial-flow turbomachinery, particularly operating in off-design conditions, and will present a general concern for the design of a new pumpjet \cite[185]{wislicenus1986}, \cite{li2013}.
 
@@
 However, this necessarily reduces the effective frontal area that is available to receive the mass-flow, which necessitates further expansion of the outer diameter, and a very substantial increase in the weight of the system.
-In addition, expansion of the hub will necessarily increase the total surface area relative to the mass flow, including by increasing the amount of tapering required after the stators in order to prevent risk of cavitation occuring around the tail the hub.
+In addition, expansion of the hub will necessarily increase the total surface area relative to the mass flow, including by increasing the amount of tapering required after the stators in order to prevent risk of cavitation occurring around the tail the hub.
 Consequently, increasing the area is not plausible without also increasing duct loss to some degree, consistent with the earlier observations from \cite{henderson1964}.
@@
 
-Given the fundamental nature of the flow-rate and intake area to the design of pumpjets, and the great lenght of time for which they have been understood and discussed as a crucial design input for a waterjet, it is altogether implausible that some large increase in intake area can be achieved without incurring necessary and unavoidable tradeoffs.
+Given the fundamental nature of the flow-rate and intake area to the design of pumpjets, and the great length of time for which they have been understood and discussed as a crucial design input for a waterjet, it is altogether implausible that some large increase in intake area can be achieved without incurring necessary and unavoidable trade-offs.
 
@@
 
-In addition, whilst it is often demonstrated that for a given propeller under certain load conditions (high loading), the addition of an accelerating duct can considerably improve efficiency, all of these situations have generally been in the context of vessels which are designed for a relatively high thrust, low waterspeed situation, when propeller size is necessarily greatly constrained.
+In addition, whilst it is often demonstrated that for a given propeller under certain load conditions (high loading), the addition of an accelerating duct can considerably improve efficiency, all of these situations have generally been in the context of vessels which are designed for a relatively high thrust, low water-speed situation, when propeller size is necessarily greatly constrained.
 This is the case fishing vessels, or for tug boats, which need to impart significant thrust to a vessel dozens or hundreds of times their own size.
@@
 
-In other words, the use of highly loaded propellers in such applications is actually particular to design constraints that are unlikely to be present for a submarine, which can position the propeller at the tail of its body, directly behind its centre of mass and drag, and consequenlty have freedom about the optimal size of the blades.
+In other words, the use of highly loaded propellers in such applications is actually particular to design constraints that are unlikely to be present for a submarine, which can position the propeller at the tail of its body, directly behind its centre of mass and drag, and consequently have freedom about the optimal size of the blades.
 It is well known that a duct cannot improve the performance of a low-loaded propeller which would be the situation for a slow-moving vessel with unconstrained propeller size.
@@
 
-\begin{figure}[h] \captionsetup{width=0.5\textwidth}
+\begin{figure}[h] \captionsetup{width=0.8\textwidth}
 \includegraphics[width=\textwidth]{AcceleratingIdeal.png}
@@
 
-\caption{(a) Extensive work has been done exploring compromise designs for decelerating ducted propellers (pumpjets), including with very modest deceleration in the intake, limited shroud taper, short duct lengths, or essentially incporate some of the advantages of accelerating ducts.
-(b) TAt optimally efficient conditions (corresponding to 27kt) such designs still exhibit some cavitation.
+\caption{(a) Extensive work has been done exploring compromise designs for decelerating ducted propellers (pumpjets), including with very modest deceleration in the intake, limited shroud taper, short duct lengths, or essentially incorporate some of the advantages of accelerating ducts.
+(b) At optimally efficient conditions (corresponding to 27kt) such designs still exhibit some cavitation.
 Credit: \cite{abdel2010}}
@@
 The fundamental difficulty with greatly shortening a duct, however, is that it requires that any changes in the shape and with of the water column be accomplished much shorter distance for the duct to achieve its desired effect, which means that generally it has to achieve less of it.
-This means that very low duct lenghts achieve less of an efficiency improvement to the ideal efficiency of a propeller, in those circumstances where they can achieve an improvement  (accelerating ducts at high loads), as discussed above.
-
-If more aggressive shape changes are attemped in a limited space, the deflections of flow and pressure gradients become much more extreme.
+This means that very low duct lengths achieve less of an efficiency improvement to the ideal efficiency of a propeller, in those circumstances where they can achieve an improvement  (accelerating ducts at high loads), as discussed above.
+
+If more aggressive shape changes are attempted in a limited space, the deflections of flow and pressure gradients become much more extreme.
 This leads to a far greater propensity for flow separation to occur.
-This can happen inside the duct ahead of the stator, particulalry when any degree of diffusion occurs in order to elevate static pressure at the blades, or any kind of direction change is required as in the case of waterjets.
+This can happen inside the duct ahead of the stator, particularly when any degree of diffusion occurs in order to elevate static pressure at the blades, or any kind of direction change is required as in the case of waterjets.
 Unlike waterjets, pumpjets also have the potential for separation on the exterior surface, which is an additional concern in the design of duct \cite[13]{henderson1964}.
@@
 
-In addition, the substantial reduction in the length of the shroud, both ahead of and behind the rotor, will also have the substantial effect of diminishing any acoustic sheilding which the duct may achieve.
+In addition, the substantial reduction in the length of the shroud, both ahead of and behind the rotor, will also have the substantial effect of diminishing any acoustic shielding which the duct may achieve.
 
@@
 Overall, while the parameter-space for the alteration of marine-propulsors is large, the parameters which would allow for very dramatic rather than minor changes to propulsor efficiency at low speed are relatively few.
-The major ones include increasing the width of the intake, decreasing the length of the duct, and decreasing the pressure inrease at the blades achieved by the duct.
+The major ones include increasing the width of the intake, decreasing the length of the duct, and decreasing the pressure increase at the blades achieved by the duct.
 In short, it involves evolving the pumpjet back towards a propeller.
@@
 With this relationship defined across the speed range, it is elementary to calculate the time that a given speed could be sustained based on a finite energy store (endurance) and the distance covered in this time (range).
-This provides the fundamental structure that underlies the model which is used to assess the dived range and enurance of conventionally powered submarines, without using air independent propulsion (AIP).
+This provides the fundamental structure that underlies the model which is used to assess the dived range and endurance of conventionally powered submarines, without using air independent propulsion (AIP).
 
@@
 
-In order to begin to model the likely impact of propulsion system choices on system peformance characteristics such as dived range and endurance, it is necessary to adopt some plausible curves indicating likely levels of efficiency across the operating speed range, in this case assumed to be up to 20kt.In this case, three possible curves have been selected for each system, representing a high, low, and central assumption for each, with peak efficiency for the propeller at about 65\% at near 5kt, and jet efficiency around 27\% at the sames speed.
-
-The central assumption has been taken in both cases from the study undertaken by BMT and Rolls Royce, as in \cite{giles2010}, and indicated in \ref{fig:BMTShape.png}, since this represents the most direct comparison made as a simple function of waterspeed, with comparable technologies.
+In order to begin to model the likely impact of propulsion system choices on system performance characteristics such as dived range and endurance, it is necessary to adopt some plausible curves indicating likely levels of efficiency across the operating speed range, in this case assumed to be up to 20kt.In this case, three possible curves have been selected for each system, representing a high, low, and central assumption for each, with peak efficiency for the propeller at about 65\% at near 5kt, and jet efficiency around 27\% at the sames speed.
+
+The central assumption has been taken in both cases from the study undertaken by BMT and Rolls Royce, as in \cite{giles2010}, and indicated in \ref{fig:BMTShape.png}, since this represents the most direct comparison made as a simple function of water-speed, with comparable technologies.
 It is probable that the pumpjet of the submarine will be a purely axial-flow pump, as opposed to a mixed-flow pump, but this difference in turbomachinery is unlikely to result in a substantial loss of propulsor efficiency, as discussed in Section \ref{pumptypes}.
@@
 In addition, the intake duct may be somewhat longer, increasing duct losses, but the degree of diffusion in the duct may be lower, reducing the adverse pressure gradient.
-In other words, it is plausible that different individual components may incur quite different levels of certain types of losses, but it is unlikely that the overall efficiency profile of the system as a whole is dramatically altered assuming competent design, since an enclosed intake controlling the waterspeed at the impeller, and contracting nozzle beyond will remain common features of both.
+In other words, it is plausible that different individual components may incur quite different levels of certain types of losses, but it is unlikely that the overall efficiency profile of the system as a whole is dramatically altered assuming competent design, since an enclosed intake controlling the water-speed at the impeller, and contracting nozzle beyond will remain common features of both.
 Indeed, it is difficult to imagine why the US Navy, having already attempted a pure axial-flow pumpjet directly in place of propellers for similar ship for the same purpose (reduction of cavitation-induced noise for Anti-Submarine Warfare) in 1965 \parencite{harvie1965construction}, would support a study 40 years later using a pump type that was expected to have substantially reduced efficiency at the intended operating speeds.
 
-Small design difference between the propellers used on submarines and surface ships may also exsit.
+Small design difference between the propellers used on submarines and surface ships may also exist.
 It is probable that the the ship propeller would have five blades, where the submarine would likely have seven or nine.
@@
 
-Overall, however, its considered that having an experimentally verified curve comparing very similar technologies as a direct function of waterspeed, on vessels attempting to optimise for the same characteristic (reduced noise induced by cavitation) provides a far more certain basis for comparison than attempting to derive efficiency curves \textit{ab initio} using theortical dimensionless factors corresponding and our best estimate of a more likely precise design.
+Overall, however, its considered that having an experimentally verified curve comparing very similar technologies as a direct function of water-speed, on vessels attempting to optimise for the same characteristic (reduced noise induced by cavitation) provides a far more certain basis for comparison than attempting to derive efficiency curves \textit{ab initio} using theoretical dimensionless factors corresponding and our best estimate of a more likely precise design.
 
@@
 The higher assumption curve used for the propeller approaches a maximum of just under 72\%.
-This performance would not be implausible if a specialised propeller was adopted specificially to deliver peak performance at around this speed.
+This performance would not be implausible if a specialised propeller was adopted specifically to deliver peak performance at around this speed.
 Advanced flexible composite propellers, such as those used by the German submarine builder TKMS \parencite{durrant2016}, are able to achieve very high efficiencies, as well as delayed onset of cavitation, by allowing the blades to deform in certain ways to adapt to their specific degree of loading \parencite{young2007hydroelastic2}.
@@
 For all propeller curves, sharply declining efficiency has been assumed at very low speeds, (2kt or less), in order to reflect the dip seen below 5kt in the efficiency curve seen in Figure \ref{fig:BMTEfficiency.png}, assuming that this is driven by the decline in efficiency shown in Figure \ref{fig:PropDuct.png} at advance ratios approaching 1.
-This has been exaccerbated beyond realistic levels in the case of the the low propeller assumption line for the sake of testing sensitivity more dramatically to some losses in this regime.
+This has been exacerbated beyond realistic levels in the case of the the low propeller assumption line for the sake of testing sensitivity more dramatically to some losses in this regime.
 It is unlikely that a state-of-the-art submarine propeller would not be able to attain maximum efficiencies between 5 and 15kt, unless some other significant factor not discussed in this paper affects propellers in these regimes.
@@
 \includegraphics[width=0.4\textwidth, center]{LaminarProp.png}
-\caption{At very low speeds changes in regimes such as an increase in laminar flow over the blade section could lead to small changes in efficiency, perhaps explaining the distinct rise at low speeds for the properller at ~5kt in Figure \ref{fig:BMTEfficiency.png}}
+\caption{At very low speeds changes in regimes such as an increase in laminar flow over the blade section could lead to small changes in efficiency, perhaps explaining the distinct rise at low speeds for the propeller at ~5kt in Figure \ref{fig:BMTEfficiency.png}}
 \label{fig:LaminarProp.png}
@@
 hot water for washing), and others (such as the combat systems) might have increased demand during crucial phases of combat operations.
-It is important to note that the combat system that has been selected for Australias FSM, the AN/BYG-1, is designed for a nuclear submarine, and is reportedly consumes considerably more power than other systems \cite{muir2015}, apparently even in the order of 100kW more.
+It is important to note that the combat system that has been selected for Australia's FSM, the AN/BYG-1, is designed for a nuclear submarine, and is reportedly consumes considerably more power than other systems \cite{muir2015}, apparently even in the order of 100kW more.
 \cite{patrick2011}.
@@
 Since the SEA 1000 project has elected not to use lithium electric batteries, the energy density of the batteries can be assumed to be that of lead-acid batteries, which are known to be in the region of 30-50Wh/kg (Watt-hours per kilogram).
-Taking a central figure and converting this to Megajoules per kilogram (MJ/kg) we assume 0.14MJ/kg to be the effective energy density of lead acid batteries for our test submarine's purposes.
+Taking a central figure and converting this to Mega-joules per kilogram (MJ/kg) we assume 0.14MJ/kg to be the effective energy density of lead acid batteries for our test submarine's purposes.
 This parameter can be adjusted easily for sensitivity testing, and in our case this will be adjusted to half, and double this energy density, noting that lead-acid batteries become much slower to charge as their charge level passes 80\%, and consequently it may be the case that only 50\% of their available capacity might be genuinely available for tactical purposes \parencite{briggs2016}.
@@
 It is worthwhile considering that alternative types of battery may be on offer for submarines.
-In particular, lithium ion batteries have been selected by Norway in a selection process which happend almost concurrently with Australia's \parencite{ohff201702}, and would probably have been on offer to Australia if they had been sought.
-Precise energy density of these batteries is not publically available.
+In particular, lithium ion batteries have been selected by Norway in a selection process which happened almost concurrently with Australia's \parencite{ohff201702}, and would probably have been on offer to Australia if they had been sought.
+Precise energy density of these batteries is not publicly available.
 For domestic uses such as handheld electronics or solar power applications, lithium ion batteries may have 3-5 times the energy density of lead-acid batteries \parencite{oconnor2017}.
 However, the actual useful density of lithium ion batteries can vary widely depending on different chemistries and designs.
-It might be assumed that for very large battery banks (such as submarines) where the stability of the battery bank is of critical importance, maximum density migth be sacrificed in favour of additional safety systems, such as increased monitoring, as well as spacing and insulation barriers between cells, in order to prevent thermal runaway in one cell leading to a catastrophic fire.
+It might be assumed that for very large battery banks (such as submarines) where the stability of the battery bank is of critical importance, maximum density might be sacrificed in favour of additional safety systems, such as increased monitoring, as well as spacing and insulation barriers between cells, in order to prevent thermal runaway in one cell leading to a catastrophic fire.
 The realistic energy density advantage that might be obtained as been suggested to lying closer to a factor of two or slightly less \parencite{greenfield2016}.
@@
 The total embarked load of the batteries is not a well-known public figure, though it is likely to be several hundred tonnes in order to provide the likely desired levels of dived endurance.
-It has been suggested publically that the approximately 700 tonnes might be required for the Shortfin Barracuda \parencite{ohff2017}.
+It has been suggested publicly that the approximately 700 tonnes might be required for the Shortfin Barracuda \parencite{ohff2017}.
 This would appear to be quite a large fraction of the plausible payload of a submarine, even of over 4000 tonnes displacement, given that a great deal of the potential displacement will be taken by the thick steel involved in constructing a pressure vessel (probably at least a thousand tonnes per submarine) \parencite{jean2017}.
@@
 Given the nature of increasing drag at high speeds, it seems unlikely that top speeds will be very far above 20kt.
-Hence, in this model we have assumed that 20kt will be obtained with the ouput of 6.5MW.
+Hence, in this model we have assumed that 20kt will be obtained with the output of 6.5MW.
 
 Considerable confirmation that these assumptions are realistic can be made by assessing the endurance of the submarines under the model with other known reference points.
-For the central curves, we obtain a dived endurance betwen 60 and 100 hours for both systems at 5kt, reflecting nearly exactly the range of other public estimates of dived enurances endurances which other modern diesel-electric submarines operating only with batteries \parencite[3]{buckingham2008submarine} at around 4kt.
+For the central curves, we obtain a dived endurance between 60 and 100 hours for both systems at 5kt, reflecting nearly exactly the range of other public estimates of dived endurances  which other modern diesel-electric submarines operating only with batteries \parencite[3]{buckingham2008submarine} at around 4kt.
 Given that somewhat larger submarines might be somewhat assisted in this respect because their overall volume/surface ratio is slightly higher, this gives confirmation that the model is overall quite realistic in its magnitudes at the most important speeds.
-The sames sources also suggests that the same submarines achieve endurances of around 1-2 hours at 20kt, which also confirms that the relationship in our model linking differnt speeds is also in agreement with other models.
+The sames sources also suggests that the same submarines achieve endurances of around 1-2 hours at 20kt, which also confirms that the relationship in our model linking different speeds is also in agreement with other models.
 
@@
 \subsection{Power Demand}
-As part of the calculation of results for each pair of efficiency curves, the power requirements of each respective component (propulsion for the two respective systems, as well as effective power required, and hote load) can be plotted to ensure that realistic results are obtained, which is important for selecting plausible values for the constant $C_d$.
+As part of the calculation of results for each pair of efficiency curves, the power requirements of each respective component (propulsion for the two respective systems, as well as effective power required, and hotel load) can be plotted to ensure that realistic results are obtained, which is important for selecting plausible values for the constant $C_d$.
 For the central efficiency curves, we can show the resultant power curves for the the assumed values of battery load, energy density, and hotel load, as seen in Figure \ref{fig:power_curve}.
@@
 The plots comparing the resultant endurance and range calculations across the speed range considered are in Figures \ref{fig:central_extreme}, \ref{fig:central_jet}, \ref{fig:central_prop} and \ref{fig:bottom_top}.
-In all cases, the comparison for the central efficiency curves have been replotted centrally, in order to highlight varience under each of the different assumption comparisons.
+In all cases, the comparison for the central efficiency curves have been re-plotted centrally, in order to highlight variance under each of the different assumption comparisons.
 
@@
 The pumpjet driven submarine reaches over 360nm between 3 and 6kt.
-A difference of 180nm dived range would presumeably be of considerable tactical significance.
+A difference of 180nm dived range would presumably be of considerable tactical significance.
 
@@
 More striking, however, is the very dramatic reduction in both range and endurance at higher speeds.
-At a speed of 20kt, a propeller driven sumbarine might manage to remain dived for 3.6 hours according to these assumptions, where a pumpjet driven submarine might manage just under 3 hours.
-A difference of about half an hour would presumeably be of less tactical significance.
+At a speed of 20kt, a propeller driven submarine might manage to remain dived for 3.6 hours according to these assumptions, where a pumpjet driven submarine might manage just under 3 hours.
+A difference of about half an hour would presumably be of less tactical significance.
 At any speed above 10kt, neither submarine would be able to remain dived for much more than 24 hours (25.4 hours for the propeller system, 18.3 for the jet).
@@
 As given in \cite{patrick2012}, submarines tend to carry diesel power capacity that approaches their main motor power.
-Whilst the difference between the two systems still appears very significant at these speeds, it appears that if transits occur in the region of 8-10kt, snorts might occurr roughly daily, and last a number of hours.
+Whilst the difference between the two systems still appears very significant at these speeds, it appears that if transits occur in the region of 8-10kt, snorts might occur roughly daily, and last a number of hours.
 The indiscretion ratio during commute of the jet system would be significantly impaired given the substantially larger power demands at transit speeds, though a detailed calculation and comparison lies outside the scope of this paper.
@@
 
-The proportional improvements will also be effective rough guides to the likely impact on indescretion ratios and overall range.\footnote{Additional considerations including the size and efficiency of the diesel engines and generators must be included for proper analyses of indescretion ratios, which will not be given here.}  Near patrol speed almost all scenarios show a difference approaching or exceeding 50\%, which will have a material impact on indescretion rates during all combat operations.
+The proportional improvements will also be effective rough guides to the likely impact on indiscretion ratios and overall range.\footnote{Additional considerations including the size and efficiency of the diesel engines and generators must be included for proper analyses of indiscretion ratios, which will not be given here.}  Near patrol speed almost all scenarios show a difference approaching or exceeding 50\%, which will have a material impact on indiscretion rates during all combat operations.
 
@@
 The results in nominal terms are striking in their magnitude.
-In terms of enduracy, doubling the stored electrical energy, for example by adopting lithium batteries, increase the \textit{difference} between the two systems under the central efficiency curve assumptions to nearly four days (93 hours) at 4kt.
+In terms of endurance, doubling the stored electrical energy, for example by adopting lithium batteries, increase the \textit{difference} between the two systems under the central efficiency curve assumptions to nearly four days (93 hours) at 4kt.
 Given that most modern, capable submarines are estimated to only be able to achieve dived endurances between 60 and 100 hours on batteries at 4kt \parencite{buckingham2008submarine} this cannot be dismissed in terms of its tactical significance.
-Even under the efficiency curve assumptions representing the narrowest spearation considered, the difference in this case would be 59 hours, almost equal to the estimated dived endurance of a Type-214 operating without AIP.
+Even under the efficiency curve assumptions representing the narrowest separation considered, the difference in this case would be 59 hours, almost equal to the estimated dived endurance of a Type-214 operating without AIP.
 
 In terms of submerged range, the same proportional impact is also extremely significant.
-Doubling the energy stored inceases the \textit{difference} between two otherwise identical submarines to over 400nm under central assumptions at 4.5kt.
+Doubling the energy stored increases the \textit{difference} between two otherwise identical submarines to over 400nm under central assumptions at 4.5kt.
 That's a larger distance than the width of the Java Sea, the most central body of water in the Indonesian Archipelago, or most of the length of the Malacca Strait.
@@
 If far smaller amounts of energy are embarked, the nominal differences reduce to perhaps 100nm, in direct proportion to the reduction of the submarine's dived endurance.
-This is still a non-trivial distance in terms of it's tactical signicance, and comprises a substantial fraction of the dived ranges of most non-AIP submarines.
-It still represents a significant difference in operational terms, equal approximately to the width of Hainan Island, or neraly twice the length of the Sunda Strait.
-
-It is also clear that increasing the hotel load has a greatly diminised impact on the proportional differences between the systems at higher speeds.
+This is still a non-trivial distance in terms of it's tactical significance, and comprises a substantial fraction of the dived ranges of most non-AIP submarines.
+It still represents a significant difference in operational terms, equal approximately to the width of Hainan Island, or nearly twice the length of the Sunda Strait.
+
+It is also clear that increasing the hotel load has a greatly diminished impact on the proportional differences between the systems at higher speeds.
 At very high speeds (e.g.
@@
 
-Only brief consideration is given here to the impact of Air Independent Propulsion, which has substantially higher energy densities possible even than Lithium, and also amounts to the addtional embarkation of greater mass.
-The volumetric energy density of the feedstocks to modern AIP systems (liquid oxygen and methanol) is estimated to be approximately ten times greater than that of lead-acid batteries \parencite{davies2016}, around 4.5MJ/L.
+Only brief consideration is given here to the impact of Air Independent Propulsion, which has substantially higher energy densities possible even than Lithium, and also amounts to the additional embarkation of greater mass.
+The  energy density of the feed-stocks to modern AIP systems (liquid oxygen and methanol) is estimated to be approximately ten times greater than that of lead-acid batteries \parencite{davies2016}, around 4.5MJ/L.
 Given the greater mass embarked, and additional studies suggesting that at constant speeds they might extend the dived endurance of a modern submarine from around 60 hours to over 1000 hours (a factor of roughly sixteen) \parencite{buckingham2008submarine} an approximate estimation of the the direction and magnitude of this impact may by scaling up energy density by this factor.
@@
 The results in nominal terms are striking in their magnitude.
-In terms of enduracy, doubling the stored electrical energy, for example by adopting lithium batteries, increase the \textit{difference} between the two systems under the central efficiency curve assumptions to nearly four days (93 hours) at 4kt.
+In terms of endurance, doubling the stored electrical energy, for example by adopting lithium batteries, increase the \textit{difference} between the two systems under the central efficiency curve assumptions to nearly four days (93 hours) at 4kt.
 Given that most modern, capable submarines are estimated to only be able to achieve dived endurances between 60 and 100 hours on batteries at 4kt \parencite{buckingham2008submarine} this cannot be dismissed in terms of its tactical significance.
-Even under the eficiency curve assumptions representing the narrowest spearation considered, the difference in this case would be 59 hours, almost equal to the estimated dived endurance of a Type214 operating without AIP.
+Even under the efficiency curve assumptions representing the narrowest separation considered, the difference in this case would be 59 hours, almost equal to the estimated dived endurance of a Type214 operating without AIP.
 
 The in terms of submerged range, the same proportional impact is also extremely significant.
-Doubling the energy stored inceases the \textit{difference} between two otherwise identical submarines to over 400nm under central assumptions at 4.5kt.
+Doubling the energy stored increases the \textit{difference} between two otherwise identical submarines to over 400nm under central assumptions at 4.5kt.
 That's a larger distance than the width of the Java Sea, the most central body of water in the Indonesian Archipelago, or most of the length of the Malacca Strait.
@@
 If far smaller amounts of energy are embarked, the nominal differences reduce to perhaps 100nm, in direct proportion to the reduction of the submarine's dived endurance.
-This is still a non-trivial distance in terms of it's tactical signicance, and comprises a substantial fraction of the dived ranges of most non-AIP submarines.
-It still represents a significant difference in operational terms, equal approximately to the width of Hainan Island, or neraly twice the length of the Sunda Strait.
-
-The singificant range of different energy density values tested here reflect quite plausibly some of ranges of different energy levels available in operation.
-For instance, as has been discussed \parencite{briggs2016, greenfield2016}, lead acid batteries may well only be able to practically utilise about 50\% of their total energy stores for tactical operaions, unless operating in a very low-threat environment and undertaking very slow charging to reach the maximum full charge is possible.
-In addition, the plausible improvement in enery density of lithium batteries may well be in the order of a factor of two in practice \parencite{greenfield2016}, though theoretical limits of the chemistry suggest much higher may be possible \parencite{davies2016}.
-
-Only brief is given here to the impact of Air Independent Propulsion, which has substantially higher energy densities possible even than Lithium, and also amounts to the addtional embarkation of greater mass.
-The volumetric energy density of the feedstocks to modern AIP systems (liquid oxygen and methanol) is estimated to be approximately ten times greater than that of lead-acid batteries \parencite{davies2016}, around 4.5MJ/L.
+This is still a non-trivial distance in terms of it's tactical significance, and comprises a substantial fraction of the dived ranges of most non-AIP submarines.
+It still represents a significant difference in operational terms, equal approximately to the width of Hainan Island, or nearly twice the length of the Sunda Strait.
+
+The significant range of different energy density values tested here reflect quite plausibly some of ranges of different energy levels available in operation.
+For instance, as has been discussed \parencite{briggs2016, greenfield2016}, lead acid batteries may well only be able to practically utilise about 50\% of their total energy stores for tactical operations, unless operating in a very low-threat environment and undertaking very slow charging to reach the maximum full charge is possible.
+In addition, the plausible improvement in energy density of lithium batteries may well be in the order of a factor of two in practice \parencite{greenfield2016}, though theoretical limits of the chemistry suggest much higher may be possible \parencite{davies2016}.
+
+Only brief is given here to the impact of Air Independent Propulsion, which has substantially higher energy densities possible even than Lithium, and also amounts to the additional embarkation of greater mass.
+The volumetric energy density of the feed-stocks to modern AIP systems (liquid oxygen and methanol) is estimated to be approximately ten times greater than that of lead-acid batteries \parencite{davies2016}, around 4.5MJ/L.
 Given the greater mass embarked, and additional studies suggesting that at constant speeds they might extend the dived endurance of a modern submarine from around 60 hours to over 1000 hours (a factor of roughly sixteen) an approximate estimation of the the direction and magnitude of this impact may by scaling up energy density by this factor.
@@
 Where battery density and hotel load are both reduced, the proportional change has a much higher and sharper peak at around 5kt, reaching an 83\% peak at 4.5kt under central assumptions.
-For high hotel load and battery density, the porportional impact is lower at a higher speed, at 43\% at 7kt.
-However, it is intereting to note that the nominal impact of each of these different scenarios is somewhat more similar in magnitude for both endurance and range.
+For high hotel load and battery density, the proportional impact is lower at a higher speed, at 43\% at 7kt.
+However, it is interesting to note that the nominal impact of each of these different scenarios is somewhat more similar in magnitude for both endurance and range.
 For low hotel load and battery store, the endurance difference peaks at 54 hours at 3 kt, and range difference peaks at 190nm at 4kt.
@@
 Here the two factors combined have a proportional impact of around 80\% improvement for the switch to propellers under central assumptions, and over 150\% in the scenario of widest separation.
-However, in nominal terms, this impact is even more significant, with central scenarios showing a \textit{difference} of over 200 hours between the two systesm under central assumptions at 3kt, and a range difference of over 750nm.
+However, in nominal terms, this impact is even more significant, with central scenarios showing a \textit{difference} of over 200 hours between the two systems under central assumptions at 3kt, and a range difference of over 750nm.
 These differences are extremely large, representing plausibly double the expected total ranges and endurances of existing modern diesel-electric submarines.
-Even if such large changes in hotel load and battery capacity are not simultaneously likely these sensitivitiesquite clearly demonstrate how even modest shifts in both hotel load and battery energy density might combine to have very significant impacts on the appearnce of the pumpjet and propeller comparison.
-For instance, a hotel load that reduced by 30\%, and a battery energy density increase of 40\% could plausibly have the impact of doubling the nominal differces between the performance of the two propulsive systems.
+Even if such large changes in hotel load and battery capacity are not simultaneously likely these sensitivities quite clearly demonstrate how even modest shifts in both hotel load and battery energy density might combine to have very significant impacts on the appearance of the pumpjet and propeller comparison.
+For instance, a hotel load that reduced by 30\%, and a battery energy density increase of 40\% could plausibly have the impact of doubling the nominal differences between the performance of the two propulsive systems.
 
@@
 
-Figure \ref{fig:IRs_hotel} demonstrates the difference between the two otherwise idendtical submarines' indiscretion ratios as a function of speed, and also what this difference represents as a proportion of the indiscretion ratio at that speed for the submarine as a propeller.
+Figure \ref{fig:IRs_hotel} demonstrates the difference between the two otherwise identical submarines' indiscretion ratios as a function of speed, and also what this difference represents as a proportion of the indiscretion ratio at that speed for the submarine as a propeller.
 It can be seen here that the indiscretion ratio is generally somewhat less sensitive to variation in hotel load than calculations for range and endurance.
 (Indiscretion ratio is also independent of energy density of the batteries, and hence will not be compared.) However, increasing the hotel load does still have the impact of somewhat diminishing the peak difference between the two systems, and moving the point at which the proportional difference peaks to higher speeds.
-Where the penalty of the pumpjet on a submarine with a hotel load of 50kW might have the largest impact at 5kt, it the difference would probably be greatest at 6kt and 7kt for a sumbarine with hotel loads of 100kW and 200kW respectively.
+Where the penalty of the pumpjet on a submarine with a hotel load of 50kW might have the largest impact at 5kt, it the difference would probably be greatest at 6kt and 7kt for a submarine with hotel loads of 100kW and 200kW respectively.
 
 The magnitude of the difference is also substantial.
-For the central efficiency curve and hotel load assumptions, the pumpjet would cause a sumbarine to increase the time spent snorting for by over 60\% between 5kt and 7.5kt.
+For the central efficiency curve and hotel load assumptions, the pumpjet would cause a submarine to increase the time spent snorting for by over 60\% between 5kt and 7.5kt.
 One would expect that this would be of considerable tactical consequence.
-These graphs also demonstrate that the difference does not diminish at higher speeds, infact it begins to increase again above 10kt, as the power required for propulsion begins to consume much of the power generated by the diesel engines, and the efficiency difference strongly impacts the size of the diminishing surplus available for charging.
+These graphs also demonstrate that the difference does not diminish at higher speeds, in fact it begins to increase again above 10kt, as the power required for propulsion begins to consume much of the power generated by the diesel engines, and the efficiency difference strongly impacts the size of the diminishing surplus available for charging.
 It is also clear that the sensitivity to the selection of efficiency curves remains quite strong in this case.
 It is clearly plausible that in a range of circumstances the pumpjet could double the time spent snorting.
-However, the difference between the lines representing the pair of central efficiency curves, and the narrowest separation, is noticeably smaller, and the impact of chaning hotel load on these lines is also proportionally diminished.
-As such, it might be said with considerable confidence that the most likely impact of the pumpjet on indiscetion ratios at any speeds over 4kt will around 40\% or higher.
+However, the difference between the lines representing the pair of central efficiency curves, and the narrowest separation, is noticeably smaller, and the impact of changing hotel load on these lines is also proportionally diminished.
+As such, it might be said with considerable confidence that the most likely impact of the pumpjet on indiscretion ratios at any speeds over 4kt will around 40\% or higher.
 It would be extremely unlikely that the penalty induced by the pumpjet at speeds of 5kt and above would be any less than 25\%.
@@
 It is generally considered that the existence of cavitation of any sort will tend to dominate the acoustic signature of any vessel that doesn't produce any machinery noise by relying on all-electric drive as a conventional submarine would when dived (\cite{giles2010}, \cite{harvie1965construction}).
-Its quite possible that during the period when a submarine is snorting, or running its diesel engines, these sources of noise would be a more substantial and distinctive acoustic signature than small amounts of cavitation on the propulsion system if it were to occur, owing in particular to the low freqency of diesel engines.
+Its quite possible that during the period when a submarine is snorting, or running its diesel engines, these sources of noise would be a more substantial and distinctive acoustic signature than small amounts of cavitation on the propulsion system if it were to occur, owing in particular to the low frequency of diesel engines.
 
 Other sources of noise hydrodynamic noise, such as the turbulence produced in the boundary layer of the flow over the hull, or parts of the propulsor, will still contribute to overall noise and some signature, but it will tend to be significant smaller in magnitude.
-This owes to the essentially incompressible nature of water.
-Without compression, even very relatively intense votices or tubulent flows will not propagate very much noise, if the water remains perfectly solid.
+This owes to the essentially in-compressible nature of water.
+Without compression, even very relatively intense votices or turbulent flows will not propagate very much noise, if the water remains perfectly solid.
 All of the 'round and round' can't achieve much 'in and out', which is what results in propagated noise.
@@
 The larger the scale of the eddies and vortices, the further noise might propagate, and also the lower the frequencies of resultant noise might be, which makes noise propagate further.
-Likewise, larger caviation voids will tend to emit lower frequency sound aswell, as can be observed from the boiling of a kettle.
-
-As a result, when dived, it might be reasonably assumed that the complete elimination of all sources of cavitition will be the primary objective of acoustic signature management.
-This is neatly put in \cite[88]{gearhart1966selection}: "The occurence of cavitation in a propulsor results in a degradation of the propulsor operating characteristics if extensive amounts of cavitation exist, and in significant noise when minute amounts occur.
-As a result, the performance requirements of a submersible weapon system specify that the occurence or inception of cavitation be avoided below certain operating depths." One might consider that a submarine's 'tactical silent speed' would most probably correspond to the speed at which it can operate safely below the point of any cavitation inception.
+Likewise, larger cavitation voids will tend to emit lower frequency sound as well, as can be observed from the boiling of a kettle.
+
+As a result, when dived, it might be reasonably assumed that the complete elimination of all sources of cavitation will be the primary objective of acoustic signature management.
+This is neatly put in \cite[88]{gearhart1966selection}: "The occurrence of cavitation in a propulsor results in a degradation of the propulsor operating characteristics if extensive amounts of cavitation exist, and in significant noise when minute amounts occur.
+As a result, the performance requirements of a submersible weapon system specify that the occurrence or inception of cavitation be avoided below certain operating depths." One might consider that a submarine's 'tactical silent speed' would most probably correspond to the speed at which it can operate safely below the point of any cavitation inception.
 
@@
 Still today, similar techniques such as paint tests and visual observations are validate models and theory.
-However, despite the inherent complexity about predicting the precise point in time and space of onset, and the respective impact on efficiency and acoustics, several broad relationships are known to hold well, and can be relied upon to characterise the most significant trends that determine when cavitation will be more likely to occur, which is generally driven by the diffence between the vapour pressure of water and the local static pressure.
+However, despite the inherent complexity about predicting the precise point in time and space of onset, and the respective impact on efficiency and acoustics, several broad relationships are known to hold well, and can be relied upon to characterise the most significant trends that determine when cavitation will be more likely to occur, which is generally driven by the difference between the vapour pressure of water and the local static pressure.
 
@@
 As such, modest changes in speed can have a big impact on the degree of probability of cavitation occurring.
-Halving your speed will roughly reduce local pressure fluctions by a factor of four.
-Consequently, even for non-specialised propellers, there is generaly some low speed for which no cavitation occurs, though almost all propellers in major surface vessels experience some cavitation at the speed they are designed to do their work at.
+Halving your speed will roughly reduce local pressure fluctuations by a factor of four.
+Consequently, even for non-specialised propellers, there is generally some low speed for which no cavitation occurs, though almost all propellers in major surface vessels experience some cavitation at the speed they are designed to do their work at.
 
@@
 
-The interaction of these two relationships can be expressed in a general form in figure \ref{fig:InceptionContours}, which has been generated using an approxiate numeric model.
+The interaction of these two relationships can be expressed in a general form in figure \ref{fig:InceptionContours}, which has been generated using an approximate numeric model.
 As can be seen here, at very low speeds, modest changes in depths can achieve considerable additional surplus pressure in order to prevent cavitation.
-However at much higher speeds, since the pressure diffential scales with the square of the velocity travelled, far greater additional depths are required to achieve sufficient excess static to supress cavitation for a give further speed increase.
+However at much higher speeds, since the pressure differential scales with the square of the velocity traveled, far greater additional depths are required to achieve sufficient excess static to suppress cavitation for a give further speed increase.
 Whilst certain system parameters such as the blade area and design (or duct design) will have a significant impact on which curve a particular system lies upon, the overall trend of a parabolic cavitation free "bucket" prevailing at low speeds and large depths will generally be robust.
-This is confirmed by the close resemblance between the curve shapes found by an experiment on a torpdo in Figure \ref{fig:InceptionDepth.png}, and the numerical model found in Figure \ref{fig:InceptionContours}.
+This is confirmed by the close resemblance between the curve shapes found by an experiment on a torpedo in Figure \ref{fig:InceptionDepth.png}, and the numerical model found in Figure \ref{fig:InceptionContours}.
 
@@
 However, at plausible temperatures the impact is small.
-The difference between the vapour pressure of water at 10 degrees and 30 degrees is approximately 3 kilopascals.
-Diving just 10m deeper adds over 100 kilopascals to static pressure.
+The difference between the vapour pressure of water at 10 degrees and 30 degrees is approximately 3 kilo-pascals.
+Diving just 10m deeper adds over 100 kilo pascals to static pressure.
 In many circumstances, altering your speed by a couple of knots will have an even larger impact on local pressure reductions.
@@
 \subsection{The significance of flow separation}
-The addition of a duct will necessarily lead to flow separation occuring on the inside of the duct opening at some very low speeds which correspond to high advance ratios when load remains constant.
+The addition of a duct will necessarily lead to flow separation occurring on the inside of the duct opening at some very low speeds which correspond to high advance ratios when load remains constant.
 Flow separation tends to lead to vortex shedding, which means that inconsistent flows (vortices) are then ingested into the blades of the impeller.
@@
 The presence of the duct around the impeller may also have the effect of preventing the noise from being effectively radiated away from the propulsor.
-This may well result in substantial reductions in some types of noise being radiated away from the propulsor in certain directions, which may have given rise to the consistent qualification of 'radiated' being used to descrive the type of noise which differentiated the French submarine bid (\cite{stewart2016}, \cite{davies2017}, \cite{ohff2016}), including by the designers of the submarine \parencite{autret2016}.
+This may well result in substantial reductions in some types of noise being radiated away from the propulsor in certain directions, which may have given rise to the consistent qualification of 'radiated' being used to describe the type of noise which differentiated the French submarine bid (\cite{stewart2016}, \cite{davies2017}, \cite{ohff2016}), including by the designers of the submarine \parencite{autret2016}.
 This is a secondary effect, quite separate to the avoidance of cavitation, which might be seen as an acoustic advantage in circumstances where cavitation might not be expected to occur on either type of propulsor.
@@
 
-It should be noted here that shielding effects might comprise the combination two distinct and seperate physically phenomenon: reflection and damping.
+It should be noted here that shielding effects might comprise the combination two distinct and separate physically phenomenon: reflection and damping.
 Noise tends to be reflected very effectively off hard, smooth surfaces representing the boundary between mediums of very different density (such as steel and water).
@@
 
-The ability of a thin layer of material to reflect noise is also highly frequecy dependent, and also dependent on mass of the shield layer.
-As the material becomes thinner and thinner, a hard shield layer will tend increasinly to transmit rather than reflect the noise, a phenomenon that increases as the amplitude (loudness) of the noise increases, and frequency decreases.
+The ability of a thin layer of material to reflect noise is also highly frequency dependent, and also dependent on mass of the shield layer.
+As the material becomes thinner and thinner, a hard shield layer will tend increasingly to transmit rather than reflect the noise, a phenomenon that increases as the amplitude (loudness) of the noise increases, and frequency decreases.
 Denser materials are more effective at reflecting sound, as more energy (pressure) is required to excite the same movement in the material.
@@
 There is therefore some tension inherent in the design of a shield where thickness and total weight are constrained.
-A thicker layer solid steel will be very effective at reflecting sound, but have negligible absoption or damping.
-A hollow between two thin layers will offer some potential for a soft daping material to be inserted, but at the cost of reflection.
-
-It is most likely that little or no damping or absorbing layer is attempted within the body of the shroud, since increasing the duct thickness above hydrodynamically determined optimum profile is likely to have a detrimental impact on efficiency.
+A thicker layer solid steel will be very effective at reflecting sound, but have negligible absorption or damping.
+A hollow between two thin layers will offer some potential for a soft damping material to be inserted, but at the cost of reflection.
+
+It is most likely that little or no damping or absorbing layer is attempted within the body of the shroud, since increasing the duct thickness above hydro-dynamically determined optimum profile is likely to have a detrimental impact on efficiency.
 Constructing a large components from thing pieces of steel, and still maintaining an optimal shape to the precision required for housing powerful turbomachinery like the impeller is also likely to involve considerable construction risk and complexity.
@@
 The larger the areas of the intakes and outlets, the narrower the effective shielding will become.
-Consequently, design changes to make the pump more efficient at low speeds, which may include expanding both the inlet and outlent, and diminshing the difference between them, will necessarily be at the expense of diminishing effective width of the shielding 'belt' around the propulsor.
-
-When the submarine is travelling at high speeds, and there is significant flow through the propulsor, the impact of waterspeed through the propulsor might have the additional impact of diminishing the radiation of noise in the forwards direction, in the same way that sound tends tends not to travel well up-wind in air.
+Consequently, design changes to make the pump more efficient at low speeds, which may include expanding both the inlet and outlet, and diminishing the difference between them, will necessarily be at the expense of diminishing effective width of the shielding 'belt' around the propulsor.
+
+When the submarine is travelling at high speeds, and there is significant flow through the propulsor, the impact of water-speed through the propulsor might have the additional impact of diminishing the radiation of noise in the forwards direction, in the same way that sound tends tends not to travel well up-wind in air.
 However, at very low speeds this effect would be greatly diminished, and possibly not relevant.
@@
 It ought further to be noted that the effectiveness of acoustic shielding is also highly dependent on frequency.
-Much lower frequencies are very hard to reflect, and absorb, as can be easily observed in a house with a home-cinema, where lower frequencies emitted by the subwoofer will tend to affect many surrounding rooms, where higher frequency sounds might be substantially muffled by interceding walls.
+Much lower frequencies are very hard to reflect, and absorb, as can be easily observed in a house with a home-cinema, where lower frequencies emitted by the sub-woofer will tend to affect many surrounding rooms, where higher frequency sounds might be substantially muffled by interceding walls.
 The same acoustic effects are at work in underwater acoustics.
@@
 The steel shroud would be very effective in reflecting such high frequency noise.
-Under acceleration or higher loads, the very instances of cavitation woud also be relatively high frequency, as the voids or cavities would be small at first.
+Under acceleration or higher loads, the very instances of cavitation would also be relatively high frequency, as the voids or cavities would be small at first.
 These sounds might also be very effectively reflected by the shroud, and consequently diminished.
@@
 
-Consequently, whilst for certain types of noise which tend to become both louder and lower in frequency with increasing speed (cavitation), flow separation provides a contrasting characteristic, in that it is likely to decrease in frequecy with decreasing speed.
-As lower frequencies tend to propagate much further in open water, this could have a significand adverse impact on the total acoustic signature of the submarine, even if the flow separation doesn't lead to any cavitation.
+Consequently, whilst for certain types of noise which tend to become both louder and lower in frequency with increasing speed (cavitation), flow separation provides a contrasting characteristic, in that it is likely to decrease in frequency with decreasing speed.
+As lower frequencies tend to propagate much further in open water, this could have a significant adverse impact on the total acoustic signature of the submarine, even if the flow separation doesn't lead to any cavitation.
 
@@
 It is plausible that there could be some elevated propulsion signature at very low speeds due to flow separation, and that the submarine's optimal acoustic signature when moving is actually at some slightly higher speed, where smoother flow through the propulsor can be achieved.
-However, given that the amplitude, frequency, and combined impact of those two characteristics cannot be accurately evaluated without substantial further work and more detailed system knowledge, it is also possible that the increase in other noise (including hull flow noise and other machinery) at elevated speeds offsets the improvement in the smoothness of the pumpjet's operation.
+However, given that the amplitude, frequency, and combined impact of those two characteristics cannot be accurately evaluated without substantial further work and more detailed system knowledge, it is also possible that the increase in other noise (including hull flow noise and other machinery) at elevated speeds offsets the improvement in the smoothness of the pumpjets operation.
 It is still quite likely the overall level of noise radiated by the pumpjet, including under conditions when flow separation occurs, is still considered to be sufficiently low in amplitude as to not be of significant tactical consequence.
-However it is highly unlikely that the accoustic signature of the pumpjet is actually superior to that of an open propeller at very low speeds, when cavitation occurs for neither system, but substantially increased turbulence including flow separation must neccessarily occur for the pumpjet.
+However it is highly unlikely that the acoustic signature of the pumpjet is actually superior to that of an open propeller at very low speeds, when cavitation occurs for neither system, but substantially increased turbulence including flow separation must necessarily occur for the pumpjet.
 
@@
 The essential idea is that energy can move or change in form, but it isn't ever created or destroyed.
-Machines, plants and animals all derive their energy from a particular other source, which can be measured and evalutated to establish the limits of energy available.
+Machines, plants and animals all derive their energy from a particular other source, which can be measured and evaluated to establish the limits of energy available.
 Plants collect energy from sunlight falling on their leaves.
@@
 Conservation of energy has a particularly relevant embodiment in fluid flows, which is given it's primary expression in Bernoullis equation, as given in Equation \ref{eq:bernoullis}.
-It says that the total energy in a connected body of fluid with no additional work being done on it is constant ($C$), though it can change in form between kinetic energy (movement, $\frac{1}{2}\rho v^2$), gravitational potential energy (it being elevated, $\rho gz$, relevant for liquids, less for gas) the heat energy in the fluid (not shown, not relevant for incompbressible fluids), and the pressure of the fluid $p$.
+It says that the total energy in a connected body of fluid with no additional work being done on it is constant ($C$), though it can change in form between kinetic energy (movement, $\frac{1}{2}\rho v^2$), gravitational potential energy (it being elevated, $\rho gz$, relevant for liquids, less for gas) the heat energy in the fluid (not shown, not relevant for incompressible fluids), and the pressure of the fluid $p$.
 
@@
 This increase in gravitational potential coincides with a slowing down of the water relative to the vessel-speed.
-Since the jet in such vessels is generally ejected at the same height as the pump, this potential is never regained, and is technically a loss, however at high speeds such a loss is small relative to the total power ouput.
+Since the jet in such vessels is generally ejected at the same height as the pump, this potential is never regained, and is technically a loss, however at high speeds such a loss is small relative to the total power output.
 In the case of submarine pumpjets, located on the end of the hull and parallel to the axis of the submarine, this effect (and the term $gz$) are not relevant.
@@
 In the case of submarines and torpedoes the water doesn't undertake a change in height, since intake and nozzle are generally all in line with the central axis of the submarine or torpedo.
-As a consequence, the key relationship in Bernoullis equation is the relationship between the liquids velocity, and its pressure.
+As a consequence, the key relationship in Bernoulli's equation is the relationship between the liquids velocity, and its pressure.
 The consequence is that when water moves through a pipe (duct, or shroud) it's pressure is inversely related to the square of its velocity.
@@
 In the case of analysing the efficiency of propulsion systems, the relevant 'work' is almost always related to moving water backwards to produce thrust.
-Moving water in directions other than backwards, including random turbulent flows which wind producing heat rather than thust, are two examples of wasted energy.
+Moving water in directions other than backwards, including random turbulent flows which wind producing heat rather than thrust, are two examples of wasted energy.
 Noise in the water also reflects energy which is wasted, though the significance of the acoustic signature is not necessarily indicative of the magnitude of energy wasted.
@@
 In general terms, energy waste in propulsion systems will result in turbulence (moving water in random chaotic patters, where the energy soon becomes smoothly distributed as heat) and acoustic energy that is propagated away.
-Flow separation incolves the creation of larger-scale vortices and eddies, which might be considered as a special case of turbulence, since the length-scale of the unsteady flows is much larger than that of normal turbulence, for instance in a boundary layer.
-
-On important consequences is that whilst increasing creation of noise, flow separation, or turbulence necessarily represent greater amounts of energy wasted, some increase in any given one does not neccessarily indicate the reduction of efficiency of the entire system, as larger amounts of energy may be being saved through reduced wastage through another means, or vice versa.
+Flow separation involves the creation of larger-scale vortices and eddies, which might be considered as a special case of turbulence, since the length-scale of the unsteady flows is much larger than that of normal turbulence, for instance in a boundary layer.
+
+On important consequences is that whilst increasing creation of noise, flow separation, or turbulence necessarily represent greater amounts of energy wasted, some increase in any given one does not necessarily indicate the reduction of efficiency of the entire system, as larger amounts of energy may be being saved through reduced wastage through another means, or vice versa.
 For example, a squeaky join on a car might make a lot of noise, and it will represent some loss of kinetic energy.
@@
 Some trends which show increasing levels of energy wasted might be outweighed by larger increases in the overall amount of productive work being achieved.
-Pumpjets, which generally tend to require highly turbulent flows, and even more turblent flows at higher speeds, still can become more efficient as speed inceases, as the amount of useful thrust produced increases even faster.
+Pumpjets, which generally tend to require highly turbulent flows, and even more turbulent flows at higher speeds, still can become more efficient as speed increases, as the amount of useful thrust produced increases even faster.
 So, despite the wasteful round-and-round movements increasing with speed, the increase in productive front-to-back movement improves even faster, and total efficiency rises.
@@
 Exactly how thick the boundary layer is, and how the fluid moves in the boundary layer, is extremely important for consideration of efficiency of fluid flows over and around solid surfaces.
-In particular, a boundary layer can be either turbulent or laminar in nature, and can transition to turbulent flow after a short distnace of laminar flow, as shown in Figure \ref{fig:BoundaryLayer.png}.
+In particular, a boundary layer can be either turbulent or laminar in nature, and can transition to turbulent flow after a short distance of laminar flow, as shown in Figure \ref{fig:BoundaryLayer.png}.
 
@@
 There is considerable mixing between all the different layers, and the average speed remains relatively constant in the flow, with the exception of the flow immediately adjacent to the wall, or in the boundary layer.
-Becuase turbulence involves a lot of movement that is not overall in one productive direction and contributing to thrust, it necessarily leads to some loss of energy from the overall thrust.
+Because turbulence involves a lot of movement that is not overall in one productive direction and contributing to thrust, it necessarily leads to some loss of energy from the overall thrust.
 Once the swirls and movements become smaller and smaller, the energy winds up simply as heat in the fluid.
@@
 
-The likely point of transition between a flow being turbulent and laminar can be given by unitless number called the Reynolds Number, which is determined by a fluids velocity, viscoscity, density, and a characteristic length-scale over which the flow occurring.
+The likely point of transition between a flow being turbulent and laminar can be given by unit-less number called the Reynolds Number, which is determined by a fluids velocity, viscosity, density, and a characteristic length-scale over which the flow occurring.
 A fuller discussion of these important concepts can be found in \cite{NPdotnet2017}.
@@
 Turbulent flows in gasses can be extremely noisy, since those fluids are compressible.
-Consequently, the rapid and intense ciruclar movements result in a lot of oscillatory compressions against the surrounding air.
+Consequently, the rapid and intense circular movements result in a lot of oscillatory compressions against the surrounding air.
 Consequently things like jet engines, vacuum cleaners, hand-dryers, and other devices that create rapid movements in gasses tend to be quite loud, including at some distance.
@@
 Cavitation is the rapid expansion and collapse of a void, or bubble in water.
-Put technically, caviation occurs when the local static pressure (pressure in the rest frame of the fluid) falls below the vapour pressure of the fluid (pressure at which the liquid will start to boil).
+Put technically, cavitation occurs when the local static pressure (pressure in the rest frame of the fluid) falls below the vapour pressure of the fluid (pressure at which the liquid will start to boil).
 
@@
 
-It is only when the water is all at a much higher temperature that the bubbles remain their full size for long enough to detatch and rise all the way to the surface, a process we typically think of as boiling water.
+It is only when the water is all at a much higher temperature that the bubbles remain their full size for long enough to detach and rise all the way to the surface, a process we typically think of as boiling water.
 It is worthwhile noting that the sound emitted at this stage is much softer and lower than the early onset of cavitation.
@@
 Cavitation is extremely important in the study of ship propulsion, since it's occurrence in particular circumstances can lead to substantial losses of efficiency, as well as damage to the propeller and related appendages.
-It is particularly important for submarines, since the expansion and collapse of these bubbles tends to lead to the creation of noise which is often far larger, and more distincly characteristic than other turbulent disturbences in the water when no cavitation is present.
+It is particularly important for submarines, since the expansion and collapse of these bubbles tends to lead to the creation of noise which is often far larger, and more distinctly characteristic than other turbulent disturbances in the water when no cavitation is present.
 Consequently, the onset of cavitation can be thought of as a distinct threshold in terms of the acoustic performance of propulsive system.
@@
 In plenty of cases, the consequences for efficiency are relatively small, as they tend to be dominated by other efficiency considerations in imparting thrust.
-Put another way, the savings that can be gained by minimising wastage to turblence can outweigh the losses incurred by having some cavitation occur.
+Put another way, the savings that can be gained by minimising wastage to turbulence can outweigh the losses incurred by having some cavitation occur.
 In most cases, propellers are designed to work with a certain extent of cavitation for optimum efficiency under working loads \parencite{shin2015}.
 Cavitation also occurs to a considerable extent within the jets of most high-speed surface vessels, but doesn't necessarily have a particularly bad effect on their overall efficiency at their intended speed levels.
-It is generally only within very specialised military circumstances when the absolute avoidance of cavitation supercedes other concerns for efficiency, and require the elimination of all cavitation entirely \parencite{mollard2011}, \parencite{lewis1988}.
+It is generally only within very specialised military circumstances when the absolute avoidance of cavitation supersedes other concerns for efficiency, and require the elimination of all cavitation entirely \parencite{mollard2011}, \parencite{lewis1988}.
 These circumstances include the design of submarines and torpedoes.
 
-In certain special cases, particularly supercavitating propellers or surface-piercing drives, high levels of cavitation can be higly advantageous for efficiency.
-However these represent specialised designs, generally only for very high speed vessels, where acoustic concerns are negligible, and are of little concern for this particular endeavour.
+In certain special cases, particularly supercavitating propellers or surface-piercing drives, high levels of cavitation can be highly advantageous for efficiency.
+However these represent specialised designs, generally only for very high speed vessels, where acoustic concerns are negligible, and are of little concern for this particular endeavor.
 
@@
 \includegraphics[width=0.7\textwidth, center]{IncreasingCavitation.png}
-\caption{Cavitation frequently occurs first at the extremities, then spreads inwards across the blad face, as shown in \parencite{mollard2011}}
+\caption{Cavitation frequently occurs first at the extremities, then spreads inwards across the blade face, as shown in \parencite{mollard2011}}
 \label{fig:IncreasingCavitation.png}
@@
 \subsection{Pressure, Thrust, and Momentum change}
-A few simple related concepts make up the foundations for many of the simplest and enduring theoretical models (Momentum Theory, still widely and routinely used to model 'ideal' propeller efficiencies by considering the propeller simple disc \parencite[169]{carlton2007}) and other quite intuitive concepts which migth be more readily grasped in understanding propulsors.
+A few simple related concepts make up the foundations for many of the simplest and enduring theoretical models (Momentum Theory, still widely and routinely used to model 'ideal' propeller efficiencies by considering the propeller simple disc \parencite[169]{carlton2007}) and other quite intuitive concepts which might be more readily grasped in understanding propulsors.
 
@@
 Pressure is a measure of force divided by an area over which it is applied.
-Pressures are necessarily applied to some extent over all of the surfaces of any object, including due to the atmostpheric pressure of air.
+Pressures are necessarily applied to some extent over all of the surfaces of any object, including due to the atmospheric pressure of air.
 In considering the efficiency of propulsion systems for solid bodies, thrust is always the consequence of some \textbf{net} pressure difference on different sides and surfaces of an object.
@@
 This pressure acts equally to push water backwards, as well as to push the propeller forwards.
-Importantly, the forward-facing side of the propeller will expereince a lowered pressure, which also serves to pull water backwards, and suck the propeller forwards.
+Importantly, the forward-facing side of the propeller will experience a lowered pressure, which also serves to pull water backwards, and suck the propeller forwards.
 It is the sum of all the pressure differences on the propeller that result in a net force.
@@
 A famous home experiment suggests that one compare the thrust (backwards pressure) experienced by someone holding a garden hose creating a jet of water when it is in the air, as opposed to submerged in a pool of water.
-When submerged, the backwards pressue appears dramatically reduced.
-In fact, this phenomenon is due to a the water-jet drawing adjacent water with it under-water, the movement of which exerts a negative pressure on the outside of the nozzle, conteracting the pressure exerted on the inside of the nozzle.
+When submerged, the backwards pressure appears dramatically reduced.
+In fact, this phenomenon is due to a the water-jet drawing adjacent water with it under-water, the movement of which exerts a negative pressure on the outside of the nozzle, counteracting the pressure exerted on the inside of the nozzle.
 This effect represents one of the important distinctions between the jets which release their water below or above the water-line, and generally means that underwater jets have much lower jet velocity ratios than those released over water.
@@
 It is useful to recall however that this principal is simply another complete, alternative way of measuring the consequences of the net pressure differences discussed earlier.
-The total rate of change of momentum in the system which is induced as a conseqence of the pressure differentials on all sides of the propulsor must necessarily be equivalent to the total net force that is produced.
-For the garden hoze in the pool or bucket, the jet rapidly entering the liquid in one direction necessitates new flows of liquid in other directions, reducing net moment changed.
+The total rate of change of momentum in the system which is induced as a consequence of the pressure differentials on all sides of the propulsor must necessarily be equivalent to the total net force that is produced.
+For the garden hose in the pool or bucket, the jet rapidly entering the liquid in one direction necessitates new flows of liquid in other directions, reducing net moment changed.
 This new liquid movement will exert pressure reductions on the nozzle surface surface, which reduce the net momentum change affected by the entire system.
@@
 \subsection{Flow separation}
-Flow separation refers to a circumstance when the flow over a surface separates entirely from the surface, and a new eddy or votex is formed where the flow actually moves in the opposite direction to the dominant flow.
+Flow separation refers to a circumstance when the flow over a surface separates entirely from the surface, and a new eddy or vortex is formed where the flow actually moves in the opposite direction to the dominant flow.
 It is of particular significance in hydrodynamics because it can lead to substantial losses of efficiency, as energy is diverted into the kinetic energy of the eddy or vortex, which is unproductive.
@@
 At the boundary layer, where the flow velocity is already reduced due to drag forces experienced near the surface, the adverse pressure gradient can be enough to reverse the flow altogether.
-In this case, the fluid flows in reverse near the surface and forms a vortex, and the main flow becomes separated form the surface by the votex.
+In this case, the fluid flows in reverse near the surface and forms a vortex, and the main flow becomes separated form the surface by the vortex.
 
@@
 In addition to the necessary loss of efficiency, flow separation can often tend to result in unsteady flows, with the vortices periodically being shed into the flow \parencite[480]{mollard2011}.
-If this occurs ahead of the blades of a propeller or impeller, such disturbances to the flow can lead to instances of cavitation when otherwise a steady homogenous flow might be well below the cavitation inception point, and can be consequential for the acoustic performance of the system, particulaly if such shedding resonates with characteristic frequencies of any of the machinery.
+If this occurs ahead of the blades of a propeller or impeller, such disturbances to the flow can lead to instances of cavitation when otherwise a steady homogeneous flow might be well below the cavitation inception point, and can be consequential for the acoustic performance of the system, particularly if such shedding resonates with characteristic frequencies of any of the machinery.
 
@@
 This is the reason that golf balls have dimples, and tennis-balls have fur, to enable them to fly further.
-The impact of the increased surface area, and more turbulent flow over the surface, is exceeded by the reduced wake field ( trailing area where the flow is totally separted) behind the ball.
+The impact of the increased surface area, and more turbulent flow over the surface, is exceeded by the reduced wake field ( trailing area where the flow is totally separated) behind the ball.
 \end{appendices}

### 2018-01-28 DraftExec.Rnw
With boxes in place

--- 2018-01-24 DraftExec.Rnw
@@
 \usepackage[titletoc,page]{appendix}
+\setlength{\parindent}{0em}
+
+\usepackage{tikz,lipsum,lmodern}
+\usepackage[most]{tcolorbox}
 
@@
 \SetWatermarkText{DRAFT}
-\SetWatermarkScale{5}
+\SetWatermarkScale{2}
 \SetWatermarkLightness{0.95}
@@
         \vspace{10cm}
-        \textbf{\huge Efficiency}
+        \textbf{\huge A comparison of pumpjets and propellers for non-nuclear submarine propulsion}
 
@@
         \vspace{1.5cm}
-        2018
+        January 2018
 
@@
 \newpage
+\setlength{\parskip}{1em}
 
@@
 Unlike other modern technolgoies such as Air Independent Propulsion and Lithium Ion Batteries, which are either in service or entering service in multiple other navies, no other navies have have adopted pumpjet technology for their conventional sumbarines, which is currently only used by nuclear submarines \parencite(coorey2016).
-
-Despite acknowledging that the pumpjet may have an adverse impact on dived endurance \parencite{sbs2017}, Naval Group justifies the superiority of the pumpjet in the following terms: "In a confrontation between two otherwise identical submarines, the one with pump jet propulsion always has the tactical advantage."  No further explanation of this superiority has been provided in detail, but it is widely held that the pumpjet offered acoustic advantages which were highly regarded by those Australians who were responsible for selecting the design partner \parencite{stewart2016}.
-
-The comparison between two otherwise identical submarines (though not necessarily limited to engaging just each other in combat), provides a useful and suitable framework for evaluating a single significant technology choice, where most other significantly novel or new technologies were apparently excluded from consideration \parencite{stanford2017}.
-The conclusion of this report is that the following statement can be made with confidence:
-\\
-\\
-\emph{In a comparison between two otherwise identical submarines, the one with the pump jet will always have a lower dived endurance, a lower dived range, a worse indiscretion ratio, a lower overall endurace, and a lower overall range, than the one with a conventional propeller.
-This will confer a substantial tactical and strategic advantage on the conventionally propelled submarine in a very broad range of operational scenarios.}
-\\
-\\
+Questions have been raised about the efficiency of pumpjets at low speeds \parencite{davies2017}.
+Propulsive efficiency at low speeds might be very important for conventional submarines since it will largely determine how long they can remain underwater before sufracing to recharge their batteries, but it would be of little or no concern to a nuclear submarine, which has effectively unlimited energy stores.
+
+Despite acknowledging that the pumpjet may have an adverse impact on dived endurance \parencite{sbs2017}, Naval Group justifies the superiority of the pumpjet in the following terms: "In a contest between two otherwise identical submarines, the one with pump jet propulsion always has the tactical advantage."\parencite{guillou2017}  No further explanation of this superiority has been provided in detail.
+This report does not attempt to engage with specific tactical encounters such as this, however, it is widely held that the pumpjet offered acoustic advantages which were highly regarded by those Australians who were responsible for selecting the design partner \parencite{stewart2016}.
+Commentary from the designers of Shortfin Barracuda Block 1A seem to confirm that it is an acoustic advantage at elevated speeds which drive this tactical advantage when facing another submarine in combat \parencite{autret2016}.
+
+The comparison between two "otherwise identical" submarines (though not necessarily limited to engaging just each other in combat), provides a useful and suitable framework for evaluating a potential trade-off in terms of other parameters, such as propulsive efficiency.
+This is particularly the case given that most other significantly novel or new technologies which might dramatically impact propulsive performance were apparently not included for serious consideration \parencite[10]{stanford2017}.
+Whilst many of the pages of this report are concerned with establishing the probable magnitude of the impact of efficiency at a range of speeds, the existence and direction of an efficiency difference between the two systems was quite firmly established at all the speeds most important for sumbarine operations.
+This allows for a surprisingly simple and strong qualitative conclusion to be reached in terms of the "otherwise identical" submarine comparison.
+
+\begin{tcolorbox}[width=0.8\textwidth, center, colback=blue!5!white,colframe=blue!75!black, title= The Conclusive Qualitative Comparison]
+In a comparison between two otherwise identical submarines, the one with the pumpjet will always have a lower dived endurance, a lower dived range, a worse indiscretion ratio, a lower overall endurace, and a lower overall range, than the one with a propeller.
+This will confer a substantial tactical and strategic advantage on the conventionally propelled submarine in a very broad range of operational scenarios.
+\end{tcolorbox}
+
 Whilst it is broadly acknowledged that propellers have superior efficiency at lower speeds, but jets might have higher efficiency at higher speeds, the precise point of cross-over seemed unclear in public debate.
@@
 
+\begin{figure}[h]
+
+\begin{subfigure}{0.5\textwidth}
+\includegraphics[width=0.9\linewidth]{HiFlowSmall.png}
+\caption{}
+\end{subfigure}
+\begin{subfigure}{0.5\textwidth}
+\includegraphics[width=0.9\linewidth]{LoFlowSmall.png}
+\caption{}
+\end{subfigure}
+
+\caption{(a) At high speeds, the intake of a pumpjet slows the waterflow and raises its pressure, and it subsequently accelerates through the contracting nozzle to form an efficient jet (b) At low speeds, the pressure gradient created by the duct causes some of the waterflow to stop and reverse.
+This 'flow separation' wastes energy spinning water in circles, and dramatically reduces the system's efficiency.}
+\label{fig:HiLoSmall.png}
+\end{figure}
+
 In order to assess the probable magnitude of the impact of this reduction in efficiency at low speeds, a numerical model was developed, based on known principles of physics, and publically available information relating to the likely technologies involved in the future submarine.
@@
 
+\begin{tcolorbox}[width=0.9\textwidth, center, colback=blue!5!white,colframe=blue!75!black, title= Quantitative comparison under central assumptions]
+\begin{center}
+\begin{tabular}{ |c||c|c|c| }
+ \hline
+ \multicolumn{4}{|c|}{4kt Comparison Table} \\
+ \hline
+ Propulsion System & Submerged Range & Submerged Endurance &Indiscretion Ratio\\
+ \hline
+ Propeller &554nm & 138 hours & 2.9\%\\
+ Pumpjet & 366nm & 91 hours & 4.4\%\\
+ \hline
+\end{tabular}
+
+\vspace{6mm}
+\begin{tabular}{ |c||c|c|c| }
+ \hline
+ \multicolumn{4}{|c|}{8kt Comparison Table} \\
+ \hline
+ Propulsion System & Submerged Range & Submerged Endurance & Indiscretion Ratio\\
+ \hline
+ Propeller &357nm & 45 hours & 9.6\%\\
+ Pumpjet & 237nm & 30 hours & 15.1\%\\
+ \hline
+\end{tabular}
+\vspace{6mm}
+
+\begin{tabular}{ |p{3cm}|p{11cm}| }
+ \hline
+ \multicolumn{2}{|c|}{Assumption Table for "otherwise identical" submarines} \\
+ \hline
+ Batteries & 500 tonnes lead-acid, 0.14MJ/kg density, 100\% discharge\\
+ Power Reference & 20kt reached by pumpjet submarine with 6.5MW power delivered\\
+ Efficiency Curve & Central lines in Figure \ref{fig:Efficiency_assumptions} modeled on those in Figure \ref{fig:BMTEfficiency.png}\\
+ Diesel Generators & 5MW delivered for effective re-charging\\
+ Hotel Load & Constant 100kW\\
+ AIP Technology & None\\
+ \hline
+\end{tabular}
+
+\end{center}
+\end{tcolorbox}
+
 The full range of plausible efficiency curves exhibited a significant efficiency advantage right through transit speeds.
@@
 
-Perhaps the most important relationship to understand to comprehend the operation of submarines is the relationship between speed and drag .
+\subsection{They physical relationship}
+Perhaps the most important relationship to comprehend in the operation of submarines is the relationship between speed and drag .
 This is important because it sets out the fundamental framework as it applies to any submarine, regardless of propulsion type.
@@
 
+\begin{tcolorbox}[width=0.8\textwidth, center, colback=blue!5!white,colframe=blue!75!black, title= Quick Summary: Sped and Drag]
+Drag directly determines the energy required to cover a given distance.
+The physics relevant to submarines require that drag increases with speed squared, and the power required by a vessel to reach a given speed scales with speed cubed.
+Consequently, the energy \textit{required} to move a vessel a given distance can vary massively over different speeds, often much more than the efficiency of any given system \textit{delivering that energy} will tend to vary.
+\\
+\\
+This means that most vessels will be able to go much further by travelling extremely slowly, almost regardless of particulars of just how they are designed and what propulsion systems they use.
+It also means that almost all the attention of propulsion system design focuses on 'design speed', generally pretty close to the top speed, since at very low speed so little thrust (hence fuel) is required that efficiency is of little concern.
+\\
+\\
+For conventional submarines, however, the situation is very different.
+They already adopt very low speeds (8-10kt) for transit to maximise the range they can get from their diesel.
+Even most commercial shipping, including bulk carriers generally go much faster (16-24kt).
+When a submarine dives, it relies on its batteries, which might store less than 1\% of the energy that's stored in its diesel tanks, and consequently has to slow down much more again (~4kt) to get meaningful range before returning to the surface to recharge.
+Whilst much higher speeds are possible even when dived, up to ~20kt, this has such a dramatic impact on the power demanded that it can't be sustained for useful periods of time, and are consequently only useful in an emergency, or very particular rapidly evolving tactical circumstances.
+\\
+\\
+Since a conventional submarine already has to slow down to a crawl to get good dived range and endurance, the efficiency of the propulsion system now becomes of very great importance, as poor efficiency here will have a signifcant impact on how soon they have to return to the surface.
+Nuclear submarines have effectively unlimited energy stores and don't require air to use it, and consequently don't share this concern.
+They might also also spend much more of their dived time at much higher speeds as a result.
+\end{tcolorbox}
+
+\subsection{Practical implications for submarines}
 This has a profound impact on the operation and engineering of maritime vessels.
@@
 
+\begin{tcolorbox}[width=0.8\textwidth, center, colback=blue!5!white,colframe=blue!75!black, title= The other reason reactors are different]
+Nuclear reactors have an enormous amount of energy stored in a small space, so for submarines they may not need refuelling for many years, if every.
+However, they vary from other power plants in another crucial way, that can have downsides to their operations.
+\\
+\\
+Most people take an internal combustion engine's ability to vary its power output for granted.
+Simply open the throttle, and more power is produced, ease it right off and it drops again.
+The engines that drive our cars and trucks are used to going from full power to almost no power every time we change gears, or stop in traffic.
+Other than a tiny bit more long-term wear on the engine, there's little consequence to large power fluctuations.
+\\
+\\
+Nuclear reactors don't enjoy the same luxury.
+The fissioning of Uranium or Plutonium atoms leaves a range of other unstable nuclear isotopes as by-products, which in turn decay to other nuclear isotopes, some of which absorb neutrons in the reactor core, or have other impacts on the 'reactivity' of the reactor (how fast it goes).
+As a consequence, nuclear reactors can't freely oscilate thier power levels, as some of these by-products following sudden changes (particulalry dramatic reductions in power) can produce dangerous situations where the reactor can become unstable.
+(It was following such a exercise in reducing power that the Chernobyl explosion occurred.)
+\\
+\\
+The physics of these decay chains mean that nuclear reactors, including those in submarines generally avoid sudden changes in power.
+Large reductions in power have to be taken in staged steps or a steady ramp-down over many hours, or more likely a couple of days.
+Furthermore, if the reactor is ever completely shut-down, it generally can't be started safely again for some time, most likely a couple of days.
+For obvious reasons, an operating sumbarine will strictly avoid anything that might risk requiring a shutdown.
+\\
+\\
+The result is that nuclear submarines whenever operating at reduced speeds tend to have an abundance of spare power to dispose of.
+The normal desire for energy 'efficiencies' are thrown into reverse, since it's undesirable to reduce the power of the reactor very far, and dangerous to reduce it fast.
+Whatever excess energy isn't sucked up by hotel load and propulsion has to be dispersed by additional systems, generally pumping water to dispose of the excess energy as heat.
+As such, in nuclear sumbarine design trading-off propulsive efficiency at low speeds is an easy, or even desirable choice.
+\end{tcolorbox}
+
 \subsection{Recent Commentary}
@@
 
-One may also note from the previous figure that for each line representing a given pitch, the efficiency falls precipitously towards zero around the point where $J$ reaches $P/D$.
+One may also note from Figure \ref{fig:OpenWater.png} that for each line representing a given pitch, the efficiency falls precipitously towards zero around the point where $J$ reaches $P/D$.
 This is because this extreme of the curve represents a circumstance where the propeller is rotating, but not actually moving any water, since it is moving through the water at about that speed which corresponds to the exact helical path defined by the pitch being traced.
@@
 
+\begin{tcolorbox}[width=0.8\textwidth, center, colback=blue!5!white,colframe=blue!75!black, title= The difference between a pumpjet and a propeller]
+In it simplest form, a pumpjet can be thought of as a special type of ducted propeller, or propeller contained inside a shroud.
+The impact of the shroud in a pumpjet has the special effect of \textit{decelerating} the water at the point that it reaches the propeller.
+This has the dual effect of raising the pressure of the water at the point that the rotor has to do its work on it, which is how pumpjets are able to reduce or eliminate cavitation.
+\\
+\\
+This type of duct shape contrasts with the system most commonly associated with ducted propellers on tug-boats and trawlers, known as accelerating ducts or 'Kort Nozzles'.
+This type of duct performs essentially the opposite function to the decelerating duct, contracting in advance of the propeller and acclerating the flow at the stage that it meets propeller.
+This effect accentuates the normal contraction of the water-column that occurs in open propellers, and can improve the efficiency of small propellers attempting to exert a large force at low speeds (such as tugs and trawlers) where an optimally sized propeller for such a role would be impractically large for the vessel.
+In this circumstance the suction force on the inside of the duct helps generate a positive thrust that is greater than the negative drag on the exterior of the duct.
+\end{tcolorbox}
+
 \subsubsection{Waterjets} \label{waterjets}
@@
 
-\begin{figure}
-
-\begin{subfigure}[b]{0.9\textwidth}
-\includegraphics[width=0.8\linewidth, center]{BMTEfficiency1.png}
-\end{subfigure}
-\begin{subfigure}[b]{0.9\textwidth}
+\begin{figure}[h]
+\captionsetup{width = 0.9\textwidth}
 \includegraphics[width=0.8\linewidth, center]{BMTEfficiency2.png}
-\end{subfigure}
-
 \caption{Studies on submerged waterjets for military surface craft currently show that the efficiency of jets decline sharply below 10kt \parencite{giles2010}}
@@
 \end{figure}
+
+\begin{tcolorbox}[width=0.8\textwidth, center, colback=blue!5!white,colframe=blue!75!black, title= Why pumjet efficiency declines at low speed]
+Pumpjets have lower efficiency at lower speeds because the shroud forces the stream of water to flow in a very different pattern to what it would naturally.
+By expanding and decelerating the flow at the point where it meets the rotor (where it would contract and accelerate without the duct) the duct forces some of the flows kinetic energy to be converted back to pressure.
+\\
+\\
+This has a terrible impact on efficiency when the incoming water doesn't have very much kinetic energy, and not a great deal more needs to be added to it, as happens when a vessel is moving slowly.
+The effect of slowing down water that's already going slowly means that some of the water stops, and in fact turns around and creates a vortex, known as 'flow separation'.
+The effective size of the water-colum being accelerated reduces and efficiency declines, as more of the system's energy is absorbed in spinning the water around in circles, rather than accelerating it backwards.
+\end{tcolorbox}
 
@@
 
+\begin{tcolorbox}[width=0.8\textwidth, center, colback=blue!5!white,colframe=blue!75!black, title= Can we just 'tune' the pumpjet to be more efficient at low speeds?]
+A effects different effect and factors working in different directions make the process of re-designing a pumpjet quite complex, and bode poorly for a high-efficiency, low-speed pumpjet.
+Generally increasing the mass-flow, by increasing duct size, might be assumed to be a first step.
+However this increases the size of the shroud, and efficiencies can decline further, since the duct inevitably produces some negative thrust.
+Moderating the 'deceleration' by making the shape change more gradual could reduce some flow separation, but make the duct much longer, and increase overall friction drag from the surfaces.
+Easy wins are not obvious.
+\\
+\\
+Furthermore, with peak efficiencies for pumpjets generally currently well above 25kt, only quite radical design changes could be expected to have the sort of effect that might make a big enough difference (good efficiency below 10kt), which aren't really within the scope of 'tuning'.
+The radical redesigns which might look capable of having a significant impact start to look very much like evolving right back to a propeller, which defeats the acoustic purposes of having a pumpjet in the first place.
+\end{tcolorbox}
+
 \subsubsection{Increase mass-flow by widening area of intake}
@@
 
-\caption{(a) A an intake may experience risks of flow separation even at low speeds (Inlet Velocity Ratio) on some surfaces due to certain flow deflections being more severe.
+\caption{(a) An intake may experience risks of flow separation even at low speeds (Inlet Velocity Ratio) on some surfaces due to certain flow deflections being more severe.
 (b) A an intake may experience risk of flow separation on different surfaces at high speeds (Inlet Velocity Ratio) where other flow deflections are more severe.
@@
 
+\begin{tcolorbox}[width=0.8\textwidth, center, colback=blue!5!white,colframe=blue!75!black, title= What about hotel load and the system as a whole?]
+The non-propulsive power demand of the submarine (hotel load) has an enormous impact on a conventional submarine's performance underwater.
+As extra power is drained as time ticks by, underwater endurance and range decline substantially with higher hotel loads.
+\\
+\\
+Hotel load also has a significant impact on the \textit{difference} between the pumpjet and propeller affect the performance of the whole system.
+By driving the speed at which the propulsion becomes the dominant power demand higher, an extra-large hotel load reduces the overall difference between the two identical submarines, and pushes the peak proportional difference to higher speeds, where endurance and range is naturally lower due to drag increasing.
+\\
+\\
+The inverse relationship leads to even more striking results, particularly if large energy stores (AIP or lithium-ion batteries) are available.
+If the hotel load is very low (either because it was designed that way, or discretionary components are shut-down in a high-threat sitation to maximise endurance) then the penalty inflicted by the pumpjet becomes extremely large, quite probably close to a factor of two, amounting to around or over a week of endurance at 3kt, and many hundreds of nautical miles at slightly higher speeds.
+\\
+\\
+The necessary consequence of this dynamic is that selecting a power-hungry combat system such as the AN/BYG-1 in conjunction with the pumpjet and lead-acid batteries for energy when dived will virtually ensure that many substantially smaller conventional sumbarines operating in the Asia Pacific in the coming decades will easily outclass the Shortfin Barracuda in terms of dived range and endurance.
+\end{tcolorbox}
+
 \begin{figure}

### 2018-01-28 DraftExec.tex
With boxes in place

@@
 This sort of extreme sensitivity ought to have a significant bearing on how the overall performance specifications of the submarine are defined, as seemingly incremental changes to hotel load and electrical energy stores are likely to substantially magnified, or muted, by the choice of propulsion system.
+
+\begin{tcolorbox}[width=0.8\textwidth, center, colback=blue!5!white,colframe=blue!75!black, title= What about hotel load and the system as a whole?]
+The non-propulsive power demand of the submarine (hotel load) has an enormous impact on a conventional submarine's performance underwater.
+As extra power is drained as time ticks by, underwater endurance and range decline substantially with higher hotel loads.
+\\
+\\
+Hotel load also has a significant impact on the \textit{difference} between the pumpjet and propeller affect the performance of the whole system.
+By driving the speed at which the propulsion becomes the dominant power demand higher, an extra-large hotel load reduces the overall difference between the two identical submarines, and pushes the peak proportional difference to higher speeds, where endurance and range is naturally lower due to drag increasing.
+\\
+\\
+The inverse relationship leads to even more striking results, particularly if large energy stores (AIP or lithium-ion batteries) are available.
+If the hotel load is very low (either because it was designed that way, or discretionary components are shut-down in a high-threat sitation to maximise endurance) then the penalty inflicted by the pumpjet becomes extremely large, quite probably close to a factor of two, amounting to around or over a week of endurance at 3kt, and many hundreds of nautical miles at slightly higher speeds.
+\\
+\\
+The necessary consequence of this dynamic is that selecting a power-hungry combat system such as the AN/BYG-1 in conjunction with the pumpjet and lead-acid batteries for energy when dived will virtually ensure that many substantially smaller conventional sumbarines operating in the Asia Pacific in the coming decades will easily outclass the Shortfin Barracuda in terms of dived range and endurance.
+\end{tcolorbox}
 

### 2018-02-04 DraftExec.Rnw
fix
with the battery discharge qualification

@@
 
-\usepackage[backend=bibtex, sorting=none, citestyle=authoryear]{biblatex}
+\usepackage[backend=bibtex, sorting=none, citestyle=numeric]{biblatex}
 \usepackage[margin=0.8in]{geometry}
@@
 \SetWatermarkText{DRAFT}
-\SetWatermarkScale{2}
+\SetWatermarkScale{1.5}
 \SetWatermarkLightness{0.95}
@@
 Even the head of the progam to build the submarine acknowledged an improvement in dived endurance would be possible with conventional propellers \parencite{sbs2017}, which seemed a clear acknowledgment that some trade-off was acknowledged, even by the strongest advocates of the pumpjet.
-But despite this, the chiefs of DCNS and SEA 1000 both remain adamant that the pumpjet was an "integral part of our bid" \parencite{guillou2017} and that it can be efficient "across the entire speed range" \parencite{sammut2017}.
+But despite this, the chiefs of DCNS and SEA 1000 both remain adamant that the pumpjet was an ``integral part of our bid" \parencite{guillou2017} and that it can be efficient ``across the entire speed range" \parencite{sammut2017}.
 
 Having done my higher education in physics, and spent considerable time considering military maritime technology running a small start-up defence company I found seemingly unresolved series of exchanges regarding the pumpjet choice particulaly curious.
-As is commonly quipped, including in defence circles, it can be reliably assumed when it comes to technology choices "there's no free lunch in physics".
+As is commonly quipped, including in defence circles, it can be reliably assumed when it comes to technology choices ``there's no free lunch in physics".
 Whilst all technologies steadily improve, and new ones continuously emerge, it's inevitably the case that when significant technology switches are made, some trade-offs are necessary.
@@
 In recent years I was heavily involved in designing a small speed-boat and a medium-weight torpedo for naval combat, and in both cases I selected a waterjet or pumpjet for those purpose.
-Pumpjets, or some kind of "decelerating ducted propeller" (explained in detail later) suited very well the sorts of speeds and operational circumstances that they were required to operate in, which were well over 30kt.
+Pumpjets, or some kind of ``decelerating ducted propeller" (explained in detail later) suited very well the sorts of speeds and operational circumstances that they were required to operate in, which were well over 30kt.
 However, having made that selection, and on several occasions wrestled with adapting the boat to alternative purposes, (such as mine warfare, including the towing of heavy sweeps and decoys) I found myself continually balancing the inclination to design a better, more specialised tool for each job, with the desire to compromise performance consolidate capability into a multi-role, more flexible capability.
@@
 The choice of the pumpjet propulsion system is quite probably the most significant new technology inclusion for Australia's future submarines.
-Unlike other modern technolgoies such as Air Independent Propulsion and Lithium Ion Batteries, which are either in service or entering service in multiple other navies, no other navies have have adopted pumpjet technology for their conventional sumbarines, which is currently only used by nuclear submarines \parencite(coorey2016).
+Unlike other modern technolgoies such as Air Independent Propulsion and Lithium Ion Batteries, which are either in service or entering service in multiple other navies, no other navies have have adopted pumpjet technology for their conventional sumbarines, which is currently only used by nuclear submarines \parencite{coorey2016}.
 Questions have been raised about the efficiency of pumpjets at low speeds \parencite{davies2017}.
@@
 
-Despite acknowledging that the pumpjet may have an adverse impact on dived endurance \parencite{sbs2017}, Naval Group justifies the superiority of the pumpjet in the following terms: "In a contest between two otherwise identical submarines, the one with pump jet propulsion always has the tactical advantage."\parencite{guillou2017}  No further explanation of this superiority has been provided in detail.
+Despite acknowledging that the pumpjet may have an adverse impact on dived endurance \parencite{sbs2017}, Naval Group justifies the superiority of the pumpjet in the following terms: ``In a contest between two otherwise identical submarines, the one with pump jet propulsion always has the tactical advantage."\parencite{guillou2017}  No further explanation of this superiority has been provided in detail.
 This report does not attempt to engage with specific tactical encounters such as this, however, it is widely held that the pumpjet offered acoustic advantages which were highly regarded by those Australians who were responsible for selecting the design partner \parencite{stewart2016}.
@@
 
-The comparison between two "otherwise identical" submarines (though not necessarily limited to engaging just each other in combat), provides a useful and suitable framework for evaluating a potential trade-off in terms of other parameters, such as propulsive efficiency.
+The comparison between two ``otherwise identical" submarines (though not necessarily limited to engaging just each other in combat), provides a useful and suitable framework for evaluating a potential trade-off in terms of other parameters, such as propulsive efficiency.
 This is particularly the case given that most other significantly novel or new technologies which might dramatically impact propulsive performance were apparently not included for serious consideration \parencite[10]{stanford2017}.
 Whilst many of the pages of this report are concerned with establishing the probable magnitude of the impact of efficiency at a range of speeds, the existence and direction of an efficiency difference between the two systems was quite firmly established at all the speeds most important for sumbarine operations.
-This allows for a surprisingly simple and strong qualitative conclusion to be reached in terms of the "otherwise identical" submarine comparison.
+This allows for a surprisingly simple and strong qualitative conclusion to be reached in terms of the ``otherwise identical" submarine comparison.
 
@@
 This report also explores in some detail the theoretical concepts that drive the limitations of pumpjet performance at lower speeds in comparison to propellers.
-Enduring and fundamental principals of momentum theory provide robust limitations on the circumstances where a propeller's 'ideal efficiency' might be improved by the addition of a duct/or shroud, one type of which constitutes the transition to a 'pumpjet'.
+Enduring and fundamental principals of momentum theory provide robust limitations on the circumstances where a propeller's `ideal efficiency' might be improved by the addition of a duct/or shroud, one type of which constitutes the transition to a `pumpjet'.
 These principals stronly support the the notion that pumpjets will not and indeed cannot suddenly overcome their efficiency disadvantage at low speeds by means of technical ingenuity or innovation.
 
-Furthmore, a particular hydrodynamic phenonemon called 'flow separation' can be identified as the primary cause of the pumpjet's declining efficiency at very low speeds.
+Furthmore, a particular hydrodynamic phenonemon called `flow separation' can be identified as the primary cause of the pumpjet's declining efficiency at very low speeds.
 Put simply, the duct provides an additional hurdle or barrier though which the water must pass when moving through the propulsor.
@@
 \caption{(a) At high speeds, the intake of a pumpjet slows the waterflow and raises its pressure, and it subsequently accelerates through the contracting nozzle to form an efficient jet (b) At low speeds, the pressure gradient created by the duct causes some of the waterflow to stop and reverse.
-This 'flow separation' wastes energy spinning water in circles, and dramatically reduces the system's efficiency.}
+This `flow separation' wastes energy spinning water in circles, and dramatically reduces the system's efficiency.}
 \label{fig:HiLoSmall.png}
@@
  \hline
- \multicolumn{2}{|c|}{Assumption Table for "otherwise identical" submarines} \\
+ \multicolumn{2}{|c|}{Assumption Table for ``otherwise identical" submarines} \\
  \hline
@@
 This means that most vessels will be able to go much further by travelling extremely slowly, almost regardless of particulars of just how they are designed and what propulsion systems they use.
-It also means that almost all the attention of propulsion system design focuses on 'design speed', generally pretty close to the top speed, since at very low speed so little thrust (hence fuel) is required that efficiency is of little concern.
+It also means that almost all the attention of propulsion system design focuses on `design speed', generally pretty close to the top speed, since at very low speed so little thrust (hence fuel) is required that efficiency is of little concern.
 \\
@@
 The fuel saving from operating even 30\% slower means that around half the total energy is required for a given journey.
-Other incremental costs and inefficiences from operating 'off-design' are frequently outweight by such a dramatic reduction in overall power demand.
+Other incremental costs and inefficiences from operating `off-design' are frequently outweight by such a dramatic reduction in overall power demand.
 
@@
 Consequently, wheras a nuclear submarine might regularly conduct transits at or near it's peak power, a diesel electric submarine would probably transit at less than half the speed which it could manage at a sprint, (and use less than a quarter of the energy, as discussed earlier) and on patrol it might be operating at a tenth of the maximum speed, and use maybe just 1\% of as much power on propulsion.
-In this situation, the amount of power that is drawn for lights, CO2 scrubbing, washing, cooking, heating, as well as the electronics driving the combat systems (the 'Hotel Load') might well become significant, and even be as larger or larger than what is required for propulsion.
+In this situation, the amount of power that is drawn for lights, CO2 scrubbing, washing, cooking, heating, as well as the electronics driving the combat systems (the `Hotel Load') might well become significant, and even be as larger or larger than what is required for propulsion.
 
@@
 This Xenon has a very special, perhaps unique role in reactors, since they very easily captures the free neutrons which trigger the continued fission reactions of Uranium or Plutonium which drive the reactor.
-As such, Xenon is known as reactor 'poison' since it can kill the reactor's reactivity in very high doses.
+As such, Xenon is known as reactor `poison' since it can kill the reactor's reactivity in very high doses.
 
@@
 The high presence of Xenon reduced the reactor's reactivity.
-But another process works in the opposite direction which offsets this, since there are lots of neutrons available to 'burn off' the Xenon (which absobs the neturons to become a different isotope).
+But another process works in the opposite direction which offsets this, since there are lots of neutrons available to `burn off' the Xenon (which absobs the neturons to become a different isotope).
 This keeps the total level of Xenon in check.
@@
 If the power is lowered dramatically and suddenly, the production of Xenon continues quite rapidly for some time due to the decay of the large stock of Iodine-135.
-With less neutron flux available to 'burn off' the Xenon, the Xenon levels spike, and push down the reactor's reactivity.
+With less neutron flux available to `burn off' the Xenon, the Xenon levels spike, and push down the reactor's reactivity.
 Unless the reactor is quickly raised back to relatively high power (~60\%) quite quickly (an hour or less), the Xenon levels become so high that the reactor will have to be shut down, otherwise extreme (and dangerous) measures would be required in order to keep the reactor going.
@@
 It should also be noted that a nuclear reactor's aversion to sudden reductions in power would also have a substantial impact on the design of a submarine's combat system, and it's demand on the Hotel Load.
-For the same reason, a high Hotel Load, or power-hungry Combat System, could actually be advantageous, as it helps to set an elevated 'floor' for power requirements, reducing the scale of fluctuations in overall power demand from the reactor due to changes in propulsion speed.
+For the same reason, a high Hotel Load, or power-hungry Combat System, could actually be advantageous, as it helps to set an elevated `floor' for power requirements, reducing the scale of fluctuations in overall power demand from the reactor due to changes in propulsion speed.
 
@@
 Nuclear reactors don't enjoy the same luxury.
-The fissioning of Uranium or Plutonium atoms leaves a range of other unstable nuclear isotopes as by-products, which in turn decay to other nuclear isotopes, some of which absorb neutrons in the reactor core, or have other impacts on the 'reactivity' of the reactor (how fast it goes).
+The fissioning of Uranium or Plutonium atoms leaves a range of other unstable nuclear isotopes as by-products, which in turn decay to other nuclear isotopes, some of which absorb neutrons in the reactor core, or have other impacts on the `reactivity' of the reactor (how fast it goes).
 As a consequence, nuclear reactors can't freely oscilate thier power levels, as some of these by-products following sudden changes (particulalry dramatic reductions in power) can produce dangerous situations where the reactor can become unstable.
@@
 The result is that nuclear submarines whenever operating at reduced speeds tend to have an abundance of spare power to dispose of.
-The normal desire for energy 'efficiencies' are thrown into reverse, since it's undesirable to reduce the power of the reactor very far, and dangerous to reduce it fast.
+The normal desire for energy `efficiencies' are thrown into reverse, since it's undesirable to reduce the power of the reactor very far, and dangerous to reduce it fast.
 Whatever excess energy isn't sucked up by hotel load and propulsion has to be dispersed by additional systems, generally pumping water to dispose of the excess energy as heat.
@@
 \subsection{Recent Commentary}
-Some efforts have been made to 'debunk' such suggestions that nuclear submarines effectively have "unlimited" power \parencite{autret2016}, with the leaders of the DCNS's bid at the time making some effort to suggest that the needs and characteristics of nuclear and diesel-electric submarines "drives the same system design".
+Some efforts have been made to `debunk' such suggestions that nuclear submarines effectively have ``unlimited" power \parencite{autret2016}, with the leaders of the DCNS's bid at the time making some effort to suggest that the needs and characteristics of nuclear and diesel-electric submarines ``drives the same system design".
 They make three comments to justify this claim.
@@
 The dynamics of reactor management dictate that generally avoiding large fluctuation in power, and strictly avoiding sharp and dramtic reductions, will be dominant considerations in a nuclear submarine's power output.
-To put things simply, the characterisation that nuclear submarines have "power to burn" is generally sound, and would only ever be challenged when the submarine is travelling at flank speed, or the reactor is shut down due to a (presumeably near-catastrophic) emergency.
+To put things simply, the characterisation that nuclear submarines have ``power to burn" is generally sound, and would only ever be challenged when the submarine is travelling at flank speed, or the reactor is shut down due to a (presumeably near-catastrophic) emergency.
 
@@
 The selection of the exact blade selection that is used to define pitch must be specified as being at some fraction of the radius from the centre, given the tendancy for the blades to twist.
-This distance is calculated to be the 'moment mean' or a technical derived effective average, which tends to lie between 0.6R and 0.7R.
+This distance is calculated to be the `moment mean' or a technical derived effective average, which tends to lie between 0.6R and 0.7R.
 A thorough discussion of pitch can be found in \cite[35-37]{carlton2007}.
@@
 \subsubsection{Pitch and the Advance Ratio}
-Pitch is also an extremely important concept to understand because it relates closely to the way that a propeller's 'open-water efficiency' $\eta_O$ is expressed in charts, as well as the advance ratio $J$ that usually comprises the x-axis in such charts.
+Pitch is also an extremely important concept to understand because it relates closely to the way that a propeller's `open-water efficiency' $\eta_O$ is expressed in charts, as well as the advance ratio $J$ that usually comprises the x-axis in such charts.
 
@@
 In this case the slip ratio is one.
-This represents a circumstance known as 'bollard pull' where a vessel is simply exerting force against an immovable object or infinite load.
+This represents a circumstance known as `bollard pull' where a vessel is simply exerting force against an immovable object or infinite load.
 Whilst the very low efficiencies here could represent a case of a propeller that is simply too small for the task it is trying to achieve, it is actually a significant and important circumstance to consider for propellers on working vessels such as tugs or trawlers, which exert large forces at very low water-speeds.
-Measurement of the thrust generated in bollard pull, and the power required to achieve this thrust, is still an important metric for such vessels, but the definition of 'useful work' for normal propulsion requieres the thust to be excerted over some distance travelled.
+Measurement of the thrust generated in bollard pull, and the power required to achieve this thrust, is still an important metric for such vessels, but the definition of `useful work' for normal propulsion requieres the thust to be excerted over some distance travelled.
 
@@
 A further means by which the efficiency of a propeller can be increased is through the reduction of the turbulence created by the blade-tips passing through the open (static) water at high speed.
-Since the two sides of the blades have low and high pressure on them respectively, at the tips where the two faces meet, there there is a tendancy for water to 'spill' from the high pressure side the the low pressure side, creating a vortex which leads to unwanted turbulence.
+Since the two sides of the blades have low and high pressure on them respectively, at the tips where the two faces meet, there there is a tendancy for water to `spill' from the high pressure side the the low pressure side, creating a vortex which leads to unwanted turbulence.
 It is also around this vortex that blade tip cavitation tends to occur.
@@
 It is perhaps most relevant at this point to further clarify the distinction between the different names for related or similar systems.
-In \cite[17]{carlton2007} pump jet is described as an "interesting development of the classical ducted propeller".
-Oosterveld, in his thesis on ducted propellers also states with regard to the decelerating duct "This ducted propeller system is the so called pumpjet" \parencite[8]{oosterveld1970}.
+In \cite[17]{carlton2007} pump jet is described as an ``interesting development of the classical ducted propeller".
+Oosterveld, in his thesis on ducted propellers also states with regard to the decelerating duct ``This ducted propeller system is the so called pumpjet" \parencite[8]{oosterveld1970}.
 Other authors give separate treatement to ducted propellers and pump jets, though the systems they describe are essentially similar in terms of composition, advantages, and physics \parencite[288]{lewis1988}.
@@
 
-In other definitive literature on the design of pumpjets \parencite{henderson1964} the difference between a pumpjet and a propeller is a described as being that "the stream of flow through the pumpjet is made to depart from the "natural" or free-stream sufrace that bounds the flow though a standard propeller."  According to the diagrams and the subsequent description it is clear that the definition embraces quite exactly the distinction between an accelerating duct, and a decelerating duct, namely that the flow is slowed at the point it encounters the impeller, where as in a free situation the free stream would contract and accelerate at this point.
+In other definitive literature on the design of pumpjets \parencite{henderson1964} the difference between a pumpjet and a propeller is a described as being that ``the stream of flow through the pumpjet is made to depart from the ``natural" or free-stream sufrace that bounds the flow though a standard propeller."  According to the diagrams and the subsequent description it is clear that the definition embraces quite exactly the distinction between an accelerating duct, and a decelerating duct, namely that the flow is slowed at the point it encounters the impeller, where as in a free situation the free stream would contract and accelerate at this point.
 
@@
 In the presence of a duct propellers might be advantageously shaped differently, including maintaing a larger chord lenght right to the shroud, rather than tapering as open propellers tend to in order to lessen vortices and cavitation on the extremities.
-However, there is not to my knowledge any single defining physical characteristic which the transition to the name 'impeller' implies, which woud not be effectively equivalent to specialised propellers for decelerating ducts.
+However, there is not to my knowledge any single defining physical characteristic which the transition to the name `impeller' implies, which woud not be effectively equivalent to specialised propellers for decelerating ducts.
 In systems tending to be described as pumpjets or waterjets hubs tend to be larger, blade-numbers higher, twist, skew and rake lower.
 No definitive transition point seems to be accepted as having great significance.
-Most commonly literature will refer to a 'rotor', a more general term which could encompass both propellers and impellers.
+Most commonly literature will refer to a `rotor', a more general term which could encompass both propellers and impellers.
 
@@
 \\
-This type of duct shape contrasts with the system most commonly associated with ducted propellers on tug-boats and trawlers, known as accelerating ducts or 'Kort Nozzles'.
+This type of duct shape contrasts with the system most commonly associated with ducted propellers on tug-boats and trawlers, known as accelerating ducts or `Kort Nozzles'.
 This type of duct performs essentially the opposite function to the decelerating duct, contracting in advance of the propeller and acclerating the flow at the stage that it meets propeller.
@@
 Many of the essential equations which govern the efficiencies and cavitation performance of pumpjets and waterjets are of the same or extremely similar form, as can be seen in detailed discussions given to both, by similar authors (\cite{wislicenus1973}, \cite{henderson1964}).
-There are, however, plenty of instances in recent literature where this distinction regarding the ejection of the jet above and below the waterline is not maintained, and waterjets and pumpjets seem to be used interchangeably \parencite{abdel2010}, or systems are described as 'submerged waterjets' \parencite{buckingham2008submarine}.
+There are, however, plenty of instances in recent literature where this distinction regarding the ejection of the jet above and below the waterline is not maintained, and waterjets and pumpjets seem to be used interchangeably \parencite{abdel2010}, or systems are described as `submerged waterjets' \parencite{buckingham2008submarine}.
 
@@
 
-At IMARESTS 10th International Naval Engineering Conference, a paper was presented by BMT Defence and Rolls Royce \parencite{giles2010}, outlining a proposal for 'fully submerged waterjets' to be used on future surface ships in order to achieve better acoustic performance, particulalry for an anti-submarine warfare role.
+At IMARESTS 10th International Naval Engineering Conference, a paper was presented by BMT Defence and Rolls Royce \parencite{giles2010}, outlining a proposal for `fully submerged waterjets' to be used on future surface ships in order to achieve better acoustic performance, particulalry for an anti-submarine warfare role.
 The jets comprise a mixed-flow pump coupled to an electric motor, and have been both extensively modelled and also tesed in demonstrator vessels.
@@
 \includegraphics[width=\textwidth]{BMTShape.png}
-\caption{The 'fully submerged waterjet' proposed by BMT and Rolls Royce for military surface craft may more closely represent the duct shape used on submarines than waterjets that release above the surface.
+\caption{The `fully submerged waterjet' proposed by BMT and Rolls Royce for military surface craft may more closely represent the duct shape used on submarines than waterjets that release above the surface.
 \parencite{giles2010}}
@@
 This also has an interaction with the different shape of the propulsors, which alters the required level of delivered (effective) power to achieve a given speed, in addition to the considerations of how efficiently the propulsor generates that effective power.
-The paper states: "This is an encouraging result for a first iteration as the AWJ-21™ arrangement has not been fully optimised and, hence, there is further prospect of reducing the shaft power for this form.
+The paper states: ``This is an encouraging result for a first iteration as the AWJ-21™ arrangement has not been fully optimised and, hence, there is further prospect of reducing the shaft power for this form.
 If the power could be reduced by 5\% (e.g.
@@
 At 5kt, the propeller was more than twice as efficient as the waterjet.
-This study has been used to directly inform the 'central' efficiency curves assessed in this analysis.
+This study has been used to directly inform the `central' efficiency curves assessed in this analysis.
 
@@
 They also provide enduring, reliable limitations on the extent to which ingenuity or design changes might be able to improve the efficiency of jets at low speed.
-But the primary reason why jet efficiencies decline towards zero at low speed is to do with flow separation \parencite{fujisawa1995}, which is what occurs when the assumptions inherent in momentum theory's analysis of an 'ideal' actuator disc start to break down.
+But the primary reason why jet efficiencies decline towards zero at low speed is to do with flow separation \parencite{fujisawa1995}, which is what occurs when the assumptions inherent in momentum theory's analysis of an `ideal' actuator disc start to break down.
 
@@
 This has a terrible impact on efficiency when the incoming water doesn't have very much kinetic energy, and not a great deal more needs to be added to it, as happens when a vessel is moving slowly.
-The effect of slowing down water that's already going slowly means that some of the water stops, and in fact turns around and creates a vortex, known as 'flow separation'.
+The effect of slowing down water that's already going slowly means that some of the water stops, and in fact turns around and creates a vortex, known as `flow separation'.
 The effective size of the water-colum being accelerated reduces and efficiency declines, as more of the system's energy is absorbed in spinning the water around in circles, rather than accelerating it backwards.
@@
 
-Derivations of the 'ideal' efficiencies are frequently made for propellers and jets using basic conservation laws (momentum theory) which consider simple acceleration of some column of water as actuated by a disc representing the impeller or propeller.
+Derivations of the `ideal' efficiencies are frequently made for propellers and jets using basic conservation laws (momentum theory) which consider simple acceleration of some column of water as actuated by a disc representing the impeller or propeller.
 A fuller description of these equations can be found in  most standard texts \parencite[131]{lewis1988}.
@@
 
-Such methods are used in \cite[8-14]{wislicenus1973} to construct an 'ideal' jet efficiency that applies to waterjets and pumpjets.
-Whilst I won't repeat the full expansion here, a fundamentally important early conclusion of the derivation is the distinction between 'useful work' and the total energy expended.
+Such methods are used in \cite[8-14]{wislicenus1973} to construct an `ideal' jet efficiency that applies to waterjets and pumpjets.
+Whilst I won't repeat the full expansion here, a fundamentally important early conclusion of the derivation is the distinction between `useful work' and the total energy expended.
 Since the rate of change in momentum is known to be necessarily equal to the force generated, and increasing momentum change (hence thrust) can be achieved either accelerating more mass, or accelerating mass to higher speeds.
@@
 
-This is expressed in the equation which represents 'ideal jet efficiency', which is directly equivalent to 'ideal propeller efficiency' using momentum theory of propeller action:
+This is expressed in the equation which represents `ideal jet efficiency', which is directly equivalent to `ideal propeller efficiency' using momentum theory of propeller action:
 
@@
 This of course isn't practical to build, and also doesn't hold when taken to the etremes the effects of the drag on the propeller blades, or duct, are taken into account.
-Infinitely large systems would increase the total weight, as well as having significant surface drag when moving through the water, all of which are neglected in this 'ideal' consideration.
+Infinitely large systems would increase the total weight, as well as having significant surface drag when moving through the water, all of which are neglected in this `ideal' consideration.
 
@@
 
-The conclusion of the analysis, and the plotting of these results for certain possible values of $K$ and $\Delta T$ is: "Its outstanding characteristic is that this efficiency approaches zero rather than unity for $\Delta V / V_0 = 0$, even for small duct losses".
+The conclusion of the analysis, and the plotting of these results for certain possible values of $K$ and $\Delta T$ is: ``Its outstanding characteristic is that this efficiency approaches zero rather than unity for $\Delta V / V_0 = 0$, even for small duct losses".
 
@@
 This is an opportunity which is constrained in the case of jets because the efficiency diminishes at very low values of $\Delta V / V_0$.
-This is quite a different constraint to other "practical" considerations as some have suggested, such as weight, or space constraints.
+This is quite a different constraint to other ``practical" considerations as some have suggested, such as weight, or space constraints.
 Open propellers may advantageously choose to lower their values of $\Delta V / V_0$ as low as 0.1 in order to obtain ideal efficiencies \parencite{wislicenus1973}.
@@
 \caption{Another graph representing efficiency as a function of the ratio of jet velocity to incoming velocity, which shows efficiency declining at low thrust (to the right).
-The impact of increasing duct losses can be clearly seen, as opposed to the 'ideal' propeller without a duct.
+The impact of increasing duct losses can be clearly seen, as opposed to the `ideal' propeller without a duct.
 Credit: \cite{bulten2006numerical}}
@@
 \includegraphics[width=0.6\textwidth, center]{PumpUp.png}
-\caption{The mysterious efficiency comparison chart which has been doing the rounds in the Australian debate probably quite reasonably reflects theoretical physical limits of 'ideal' pumpjets and propellers.
+\caption{The mysterious efficiency comparison chart which has been doing the rounds in the Australian debate probably quite reasonably reflects theoretical physical limits of `ideal' pumpjets and propellers.
 Credit: \cite{stanford2017}}
@@
 Minor adjustments and the improvement of small or marginal effects, of which there might be many, will clearly not help in shifting the efficiency peak of a jet from over 25kt to under 10kt.
-In other words, this gap will not be closed or noticeably reduced by 'tuning'.
+In other words, this gap will not be closed or noticeably reduced by `tuning'.
 Quite substantial redesign of core characteristics (like mass flow) would be required, which would necessitate a cycle of redesign and tuning virtually all other aspect of the entire jet, as illustrated by numerous guides to the design of such systems (\cite{bruce1974}, \cite{furuya1988}, \cite{mccormick1963design}, \cite{henderson1964}).
@@
 
-\begin{tcolorbox}[width=0.8\textwidth, center, colback=blue!5!white,colframe=blue!75!black, title= Can we just 'tune' the pumpjet to be more efficient at low speeds?]
-A effects different effect and factors working in different directions make the process of re-designing a pumpjet quite complex, and bode poorly for a high-efficiency, low-speed pumpjet.
+\begin{tcolorbox}[width=0.8\textwidth, center, colback=blue!5!white,colframe=blue!75!black, title= Can we just `tune' the pumpjet to be more efficient at low speeds?]
+Several different factors working in different directions make the process of re-designing a pumpjet quite complex, and bode poorly for a high-efficiency, low-speed redesign.
 Generally increasing the mass-flow, by increasing duct size, might be assumed to be a first step.
 However this increases the size of the shroud, and efficiencies can decline further, since the duct inevitably produces some negative thrust.
-Moderating the 'deceleration' by making the shape change more gradual could reduce some flow separation, but make the duct much longer, and increase overall friction drag from the surfaces.
+Moderating the `deceleration' by making the shape change more gradual could reduce some flow separation, but make the duct much longer, and increase overall friction drag from the surfaces.
 Easy wins are not obvious.
@@
 \\
-Furthermore, with peak efficiencies for pumpjets generally currently well above 25kt, only quite radical design changes could be expected to have the sort of effect that might make a big enough difference (good efficiency below 10kt), which aren't really within the scope of 'tuning'.
+Furthermore, with peak efficiencies for pumpjets generally currently well above 25kt, only quite radical design changes could be expected to have the sort of effect that might make a big enough difference (good efficiency below 10kt), which aren't really within the scope of `tuning'.
 The radical redesigns which might look capable of having a significant impact start to look very much like evolving right back to a propeller, which defeats the acoustic purposes of having a pumpjet in the first place.
@@
 The increased drag that this induces must be deducted from the thrust that the pumpjet produces, or effectively increasing $\Delta T$ as given in earlier equations.
-In this respect, there is "no free lunch" to be had in expanding the size of the intake.
+In this respect, there is ``no free lunch" to be had in expanding the size of the intake.
 As Henderson et al.
-conclude, for this reason the diameter of a pumpjet tends to be about 15\% or 20\% smaller than that of a corresponding propeller, whereas the diameter to have equivalent mass-flow, highlighting the tension set up in pumpjet design with their earlier observation that "if the cavitation resistance of the blades is to be improved with no reduction in thrust and the same rate of flow through the propulsor, the pump jet rotor must be larger than that of the propeller." \parencite[1]{henderson1964}
+conclude, for this reason the diameter of a pumpjet tends to be about 15\% or 20\% smaller than that of a corresponding propeller, whereas the diameter to have equivalent mass-flow, highlighting the tension set up in pumpjet design with their earlier observation that ``if the cavitation resistance of the blades is to be improved with no reduction in thrust and the same rate of flow through the propulsor, the pump jet rotor must be larger than that of the propeller." \parencite[1]{henderson1964}
 
@@
 \includegraphics[width=\textwidth]{AcceleratingIdeal.png}
-\caption{Calculations of 'ideal' efficiencies of accelerating ducts (neglecting drag) in \cite{oosterveld1970} show that a duct can only improve the efficiency of a propeller if it is loaded above a certain degree, and that longer ducts produce better improvements in 'ideal' efficiencies}
+\caption{Calculations of `ideal' efficiencies of accelerating ducts (neglecting drag) in \cite{oosterveld1970} show that a duct can only improve the efficiency of a propeller if it is loaded above a certain degree, and that longer ducts produce better improvements in `ideal' efficiencies}
 \label{fig:AcceleratingIdeal.png}
@@
 Variations from this central line have been made with consideration for some plausible trade-offs errors, which are intended to be indicative of magnitude possible unknowns in system design and their likely impact.
-In the case of the 'high' pumpjet assumption, it is assume that a substantially larger intake area has been chosen in order to achieve higher efficiencies at lower speeds, at the cost of peak ideal efficiency, which might be realised not too far above 20kt for such a design.
+In the case of the `high' pumpjet assumption, it is assume that a substantially larger intake area has been chosen in order to achieve higher efficiencies at lower speeds, at the cost of peak ideal efficiency, which might be realised not too far above 20kt for such a design.
 The low assumption for the pumpjet is still represents higher efficiencies at equivalent speeds than those given in \cite{fujisawa1995} for a waterjet.
@@
 In particular, lithium ion batteries have been selected by Norway in a selection process which happend almost concurrently with Australia's \parencite{ohff201702}, and would probably have been on offer to Australia if they had been sought.
-Precise energy density of these batteries is not publically available.
+Precise energy density of these batteries is not readily publically available.
 For domestic uses such as handheld electronics or solar power applications, lithium ion batteries may have 3-5 times the energy density of lead-acid batteries \parencite{oconnor2017}.
@@
 
+Furthermore, batteries of most types tend to have higher levels of internal resistance at higher discharge rates, resulting in energy being lost to heat, which reduces their effective capacity.
+As such, there are pracitcal limits to how quickly a batterie's energy can be withdrawn, and some decline in effective capacity as the discharge rate increases.
+In the case of lead-acid batteries, the very high rate of discharge associated with the top-speeds of a submarine very substantially reduces their effective capacity.
+In the modelling for this paper, no specific consideration for the effects of internal resistance is given, and constant capacity is assumed for all speeds.
+Given the that the most substantial results of this paper are at relatively low speeds, perhaps half maxium, and consequently an eigth the maxium discharge rate from the batteries, this will have a negligible effect on the most significant results.
+However, the modeled endurances and ranges at top speed may be substantially overstated in practice for a submarine for lead-acid batteries.
+
 \subsection{Drag Coefficient of Hull}
@@
 Given that any known power consumption level and speed match can be used as a reference, I have chosen instead to use the probable power consumption at top speed, both of which we have some indication of, and given that propulsive efficiencies of the two systems are likely to be much closer at this speed, there will be less error in likely comparisons between different design concepts.
-It is known that the a early German designs for a plausible submarine for SEA 1000 included a 6MW electric motor \parencite{patrick2012}, and the Shortfin Barracuda of Naval Group is often referred to as having a 7MW electric motor \parencite{ohff2017} \parencite{coates2016}, with top speeds ">20kt".
+It is known that the a early German designs for a plausible submarine for SEA 1000 included a 6MW electric motor \parencite{patrick2012}, and the Shortfin Barracuda of Naval Group is often referred to as having a 7MW electric motor \parencite{ohff2017} \parencite{coates2016}, with top speeds ``>20kt".
 Given the nature of increasing drag at high speeds, it seems unlikely that top speeds will be very far above 20kt.
@@
 Without compression, even very relatively intense votices or tubulent flows will not propagate very much noise, if the water remains perfectly solid.
-All of the 'round and round' can't achieve much 'in and out', which is what results in propagated noise.
+All of the `round and round' can't achieve much `in and out', which is what results in propagated noise.
 This obviously doesn't hold true in compressible fluids, such as air, as one would observe noticing the intense noise produced by hand-dryers, aeroplanes, jet engines etc.
@@
 As a result, when dived, it might be reasonably assumed that the complete elimination of all sources of cavitition will be the primary objective of acoustic signature management.
-This is neatly put in \cite[88]{gearhart1966selection}: "The occurence of cavitation in a propulsor results in a degradation of the propulsor operating characteristics if extensive amounts of cavitation exist, and in significant noise when minute amounts occur.
-As a result, the performance requirements of a submersible weapon system specify that the occurence or inception of cavitation be avoided below certain operating depths." One might consider that a submarine's 'tactical silent speed' would most probably correspond to the speed at which it can operate safely below the point of any cavitation inception.
+This is neatly put in \cite[88]{gearhart1966selection}: ``The occurence of cavitation in a propulsor results in a degradation of the propulsor operating characteristics if extensive amounts of cavitation exist, and in significant noise when minute amounts occur.
+As a result, the performance requirements of a submersible weapon system specify that the occurence or inception of cavitation be avoided below certain operating depths." One might consider that a submarine's `tactical silent speed' would most probably correspond to the speed at which it can operate safely below the point of any cavitation inception.
 
@@
 However at much higher speeds, since the pressure diffential scales with the square of the velocity travelled, far greater additional depths are required to achieve sufficient excess static to supress cavitation for a give further speed increase.
-Whilst certain system parameters such as the blade area and design (or duct design) will have a significant impact on which curve a particular system lies upon, the overall trend of a parabolic cavitation free "bucket" prevailing at low speeds and large depths will generally be robust.
+Whilst certain system parameters such as the blade area and design (or duct design) will have a significant impact on which curve a particular system lies upon, the overall trend of a parabolic cavitation free ``bucket" prevailing at low speeds and large depths will generally be robust.
 This is confirmed by the close resemblance between the curve shapes found by an experiment on a torpdo in Figure \ref{fig:InceptionDepth.png}, and the numerical model found in Figure \ref{fig:InceptionContours}.
@@
 The presence of the duct around the impeller may also have the effect of preventing the noise from being effectively radiated away from the propulsor.
-This may well result in substantial reductions in some types of noise being radiated away from the propulsor in certain directions, which may have given rise to the consistent qualification of 'radiated' being used to descrive the type of noise which differentiated the French submarine bid (\cite{stewart2016}, \cite{davies2017}, \cite{ohff2016}), including by the designers of the submarine \parencite{autret2016}.
+This may well result in substantial reductions in some types of noise being radiated away from the propulsor in certain directions, which may have given rise to the consistent qualification of `radiated' being used to descrive the type of noise which differentiated the French submarine bid (\cite{stewart2016}, \cite{davies2017}, \cite{ohff2016}), including by the designers of the submarine \parencite{autret2016}.
 This is a secondary effect, quite separate to the avoidance of cavitation, which might be seen as an acoustic advantage in circumstances where cavitation might not be expected to occur on either type of propulsor.
@@
 
+\begin{tcolorbox}[width=0.8\textwidth, center, colback=blue!5!white,colframe=blue!75!black, title= Exactly when and how are pumpjets quieter than propellers?]
+It's generally accepted that the most important acoustic consideration is preventing cavitation from occurring.
+Whilst all turbulence and other disturbances underwater create some noise, becuase water is incompressible, quite a lot of `round and round' movement (as in turbulence) still results in relatively little `in and out' movement, which is what tends to result in pressure waves propagating away from the sources.
+When gaps open up in the water and subsequently collapse, as is the case in cavitation, the noise produced is much louder.
+\\
+\\
+Pumpjets have an acoustic advantage because they can avoid cavitation in circumstances where an open propeller might begin to cavitate, in particular at higher speeds.
+However, at lower speeds such as patrol speeds (and also at greater depths) propellers can easily avoid cavitation because they don't need turn very fast in the water.
+Consequently, the avoidance of cavitation in these circumstances offers no advantage.
+In addition, at very low speeds pumpjets experience increasing levels of flow separation, which creates unsteady vortices before and between the blades of the rotor.
+This degrades the acoustic performance of the jet, particularly at low frequencies, which are harder to shield, and tend to carry much further.
+At some very low speed, a well-designed propeller will necessarily be much quieter than a jet, though both may still be much quieter than if cavitation occurred.
+\\
+\\
+This means that the selection of a propulsion system for acoustic advantage is highly dependent on the probable speed profiles of submarine's operations.
+For a nuclear submarine, which can operate at higher speeds (above 15 or 20kt) for as long as they like, the ability to remain relatively quiet at these speeds is extremely desirable.
+It would allow them to undertake fast transits, or stalk enemy shipping, while remaining very quiet.
+Diesel electric submarines simply cannot sustain these sorts of speeds for any more than a few hours, and hence they are unlikely to use them unless in an emergency or critical tactical manoeuvre.
+Almost all of their sensitive or combative operations are likely to be at very low speeds, and hence being as quiet as possible in absolute terms at low speeds is most likely to be advantageous.
+The selection of a pumpjet on a diesel-electric submarine on acoustic grounds is consequently a strange choice, as it amounts to elevating a tactical scenario which demands a short, high-speed burst above all other circumstances where a long, silent dive at low speed is required.
+\end{tcolorbox}
+
 \subsubsection{Directional Dependence}
@@
 The larger the areas of the intakes and outlets, the narrower the effective shielding will become.
-Consequently, design changes to make the pump more efficient at low speeds, which may include expanding both the inlet and outlent, and diminshing the difference between them, will necessarily be at the expense of diminishing effective width of the shielding 'belt' around the propulsor.
+Consequently, design changes to make the pump more efficient at low speeds, which may include expanding both the inlet and outlent, and diminshing the difference between them, will necessarily be at the expense of diminishing effective width of the shielding `belt' around the propulsor.
 
@@
 
-It should further be noted that in the underwater environment sound can be made to reflect in a variety of directions simply with differences in temperature, salinity, movement of water bodies, as well as off plenty of nearby underwater objects, including the ocean floor and objects on it.
+It should further be noted that in the underwater environment sound can be made to refract in a variety of directions simply with differences in temperature, salinity, movement of water bodies, as well as off plenty of nearby underwater objects, including the ocean floor and objects on it.
 This phenomenon generally serves to make the art of underwater acoustics complex and unpredictable, and generally contributes to the difficulty of precisely locating a submarine, even when some signature is detected.
@@
 The inverse process is where a pipe increases in volume, and the flow is forced to slow down in order to fill the wider area, and the pressure correspondingly increases.
-This is a process called 'diffusion', and is important to achieving high pressure levels in many types of water pumps, including those which will be particularly relevant for pumpjets for watercraft propulsion, particularly elevating static pressure levels in order to prevent cavitation \parencite{hamilton1997, mccormick1963design, bruce1974}.
+This is a process called `diffusion', and is important to achieving high pressure levels in many types of water pumps, including those which will be particularly relevant for pumpjets for watercraft propulsion, particularly elevating static pressure levels in order to prevent cavitation \parencite{hamilton1997, mccormick1963design, bruce1974}.
 
@@
 A perfectly efficient system won't do any work that isn't for the intended purpose.
-In the case of analysing the efficiency of propulsion systems, the relevant 'work' is almost always related to moving water backwards to produce thrust.
+In the case of analysing the efficiency of propulsion systems, the relevant `work' is almost always related to moving water backwards to produce thrust.
 Moving water in directions other than backwards, including random turbulent flows which wind producing heat rather than thust, are two examples of wasted energy.
@@
 The way that fluids move relative to a surface can one of either two methods.
-In 'laminar flow' all the fluid moves in one direction in an smooth and orderly manner, with very little mixing between the layers of fluid travelling at different speeds.
-
-The alternative is 'turbulent flow', where the fluid moves around in unpredictable swirls and circles as well as moving overall in an predominant direction.
+In `laminar flow' all the fluid moves in one direction in an smooth and orderly manner, with very little mixing between the layers of fluid travelling at different speeds.
+
+The alternative is `turbulent flow', where the fluid moves around in unpredictable swirls and circles as well as moving overall in an predominant direction.
 There is considerable mixing between all the different layers, and the average speed remains relatively constant in the flow, with the exception of the flow immediately adjacent to the wall, or in the boundary layer.
@@
 In essentially incompressible fluids, the random changes in pressure and velocity which are involved in laminar flow tend not to generate nearly as much noise in the surrounding liquid outside the flow.
-Because the fluid is incompressible, all of the random 'round and round' movements don't amount to much 'in and out' movement, which is what creates the pressure waves which result in propagated noise.
-Consequently, whilst turbulent flows do inevitably generate some noise, the amount of noise generated is dramatically greater in the presence of cavitation, where a void opens up in the water, and vastly more 'in and out' movement is able to occur.
+Because the fluid is incompressible, all of the random `round and round' movements don't amount to much `in and out' movement, which is what creates the pressure waves which result in propagated noise.
+Consequently, whilst turbulent flows do inevitably generate some noise, the amount of noise generated is dramatically greater in the presence of cavitation, where a void opens up in the water, and vastly more `in and out' movement is able to occur.
 
@@
 \subsection{Pressure, Thrust, and Momentum change}
-A few simple related concepts make up the foundations for many of the simplest and enduring theoretical models (Momentum Theory, still widely and routinely used to model 'ideal' propeller efficiencies by considering the propeller simple disc \parencite[169]{carlton2007}) and other quite intuitive concepts which migth be more readily grasped in understanding propulsors.
+A few simple related concepts make up the foundations for many of the simplest and enduring theoretical models (Momentum Theory, still widely and routinely used to model `ideal' propeller efficiencies by considering the propeller simple disc \parencite[169]{carlton2007}) and other quite intuitive concepts which migth be more readily grasped in understanding propulsors.
 

### 2018-02-11 DraftExec.Rnw
latest

@@
 Whilst all technologies steadily improve, and new ones continuously emerge, it's inevitably the case that when significant technology switches are made, some trade-offs are necessary.
-In the case of the pumpjets, the trade-offs what might have been expected appeared obvious, and probably unavoidable to some extent to all those experienced in maritime systems: it would have lower propulsive efficiency, particularly at low speeds.
+In the case of the pumpjets, the trade-offs that might have been expected appeared obvious, and probably unavoidable to some extent to all those experienced in maritime systems: it would have lower propulsive efficiency, particularly at low speeds.
 

### 2018-05-21 DraftExec.tex
with ross acknowledged :-)
bump

@@
 
-\usepackage{draftwatermark}
-\SetWatermarkText{DRAFT}
-\SetWatermarkScale{1.5}
-\SetWatermarkLightness{0.95}
+% for cover page
+\usepackage{geometry}
+\usepackage{xcolor}
+\usepackage{graphicx}
+
+% \usepackage{draftwatermark}
+% \SetWatermarkText{DRAFT}
+% \SetWatermarkScale{1.5}
+% \SetWatermarkLightness{0.95}
 
@@
 \author{Aidan Morrison}
+
+% \SweaveOpts{concordance=TRUE}
+\definecolor{titlepagecolor}{cmyk}{0, 0, 0, .30}
+\definecolor{namecolor}{cmyk}{0, 0, 0, .80}
 \IfFileExists{upquote.sty}{\usepackage{upquote}}{}
@@
 
+%-----------------------------------------------------------------
+
+% ----------------------------------------------------------------
 \begin{titlepage}
-    \begin{center}
-        \vspace*{1cm}
-
-        \vspace{10cm}
-        \textbf{\huge A comparison of pumpjets and propellers for non-nuclear submarine propulsion}
-
-        \vspace{0.5cm}
-
-        \vspace{1.5cm}
-
+\newgeometry{left=7.5cm} %defines the geometry for the titlepage
+\pagecolor{titlepagecolor}
+\noindent
+
+{\huge \textsf{A comparison of pumpjets and}}
+\vskip\baselineskip
+{\huge \textsf{propellers for non-nuclear}}
+\vskip\baselineskip
+{\huge \textsf{submarine propulsion}}
+\vskip\baselineskip
+\noindent
 \textbf{Aidan Morrison}
-
-        \vspace{1.5cm}
-        January 2018
+\vskip\baselineskip
+\textsf{January 2018}
+\vskip\baselineskip
+\vskip\baselineskip
 
 \vfill
-
-    \end{center}
+\noindent
+
+\includegraphics[width=6cm]{tl_logo.png}\\[-1em]
+\color{white}
+\makebox[0pt][l]{\rule{1.3\textwidth}{1pt}}
+\par
+\noindent
+\textbf{\textsf{Trendlock}} \textcolor{namecolor}{\textsf{Consulting}}
+
 \end{titlepage}
+\restoregeometry % restores the geometry
+\nopagecolor% Use this to restore the color pages to white
+
+\newpage
 
@@
 Unless he had commissioned me to write it, there is no way that I would have been able to make the sustained and deliberate inquiry which the topic required to reach a useful level of quantification and confidence to meaningfully inform debate.
+In addition, the support of my colleague from Trendlock, Ross Ireland, was invaluable in producing the extensive modelling required.
 

### 2018-06-05 DraftExec.Rnw
with ross acknowledged :-)
latest, with typos for senate

@@
 
+% \SweaveOpts{concordance=TRUE}
+\definecolor{titlepagecolor}{cmyk}{0, 0, 0, .30}
+\definecolor{namecolor}{cmyk}{0, 0, 0, .80}
+
 \begin{document}
 
+%-----------------------------------------------------------------
+
+% ----------------------------------------------------------------
 \begin{titlepage}
-    \begin{center}
-        \vspace*{1cm}
-
-        \vspace{10cm}
-        \textbf{\huge A comparison of pumpjets and propellers for non-nuclear submarine propulsion}
-
-        \vspace{0.5cm}
-
-        \vspace{1.5cm}
-
+\newgeometry{left=7.5cm} %defines the geometry for the titlepage
+\pagecolor{titlepagecolor}
+\noindent
+
+{\huge \textsf{A comparison of pumpjets and}}
+\vskip\baselineskip
+{\huge \textsf{propellers for non-nuclear}}
+\vskip\baselineskip
+{\huge \textsf{submarine propulsion}}
+\vskip\baselineskip
+\noindent
 \textbf{Aidan Morrison}
-
-        \vspace{1.5cm}
-        January 2018
+\vskip\baselineskip
+\textsf{January 2018}
+\vskip\baselineskip
+\vskip\baselineskip
 
 \vfill
-
-    \end{center}
+\noindent
+
+\includegraphics[width=6cm]{tl_logo.png}\\[-1em]
+\color{white}
+\makebox[0pt][l]{\rule{1.3\textwidth}{1pt}}
+\par
+\noindent
+\textbf{\textsf{Trendlock}} \textcolor{namecolor}{\textsf{Consulting}}
+
 \end{titlepage}
+\restoregeometry % restores the geometry
+\nopagecolor% Use this to restore the color pages to white
+
+\newpage
 
@@
 The purpose of this paper is to investigate the suitability of the pumpjet propulsor technology selection for Australia's future fleet of submarines.
-Amongst this considerable controversy surrounding the submarines, this particular technology choice deserves special attention.
+Amongst the considerable controversy surrounding the submarines, this particular technology choice deserves special attention.
 Much has been made of the significance of the pumpjet in the DCNS (now Naval Group) bid for the submarine, to the extent that it has been claimed that the pumpjet rendered propellers obsolete \parencite{naval2018}, and features extremely prominently in the all the marketing and literature.
@@
 However, it has remained the case that even extremely well-informed commentators reluctant to quantify and qualify the plausible scale of this trade-off, and what it might mean strategically and tactically for the submarine.
-Everyone knew it was an issue, but no-one confident enough in the scale of the impact (or even at transit speeds the direction of the impact \parencite{davies2017}) to make it clear whether it mattered.
+Everyone knew it was an issue, but no-one was confident enough on the scale of the impact (or even at transit speeds the direction of the impact \parencite{davies2017}) to make it clear whether it mattered.
 Even the head of the progam to build the submarine acknowledged an improvement in dived endurance would be possible with conventional propellers \parencite{sbs2017}, which seemed a clear acknowledgment that some trade-off was acknowledged, even by the strongest advocates of the pumpjet.
@@
 
-Having done my higher education in physics, and spent considerable time considering military maritime technology running a small start-up defence company I found seemingly unresolved series of exchanges regarding the pumpjet choice particulaly curious.
+Having done my higher education in physics, and spent considerable time considering military maritime technology running a small start-up defence company I found this seemingly unresolved series of exchanges regarding the pumpjet choice particulaly curious.
 As is commonly quipped, including in defence circles, it can be reliably assumed when it comes to technology choices ``there's no free lunch in physics".
@@
 Pumpjets, or some kind of ``decelerating ducted propeller" (explained in detail later) suited very well the sorts of speeds and operational circumstances that they were required to operate in, which were well over 30kt.
-However, having made that selection, and on several occasions wrestled with adapting the boat to alternative purposes, (such as mine warfare, including the towing of heavy sweeps and decoys) I found myself continually balancing the inclination to design a better, more specialised tool for each job, with the desire to compromise performance consolidate capability into a multi-role, more flexible capability.
+However, having made that selection, and on several occasions wrestled with adapting the boat to alternative purposes, (such as mine warfare, including the towing of heavy sweeps and decoys) I found myself continually balancing the inclination to design a better, more specialised tool for each job, with the desire to compromise performance and consolidate capability into a multi-role, more flexible capability.
 The trade-offs inherent in propulsion systems, and the relative efficency of jets at high speeds and propellers at low speeds was a familiar encounter.
@@
 However, the existence of unknown developments and advances does not mean that realistic approximate bounds and limits of technologies cannot be found from public inquiry.
-Scientific discovery and technical tends to confirm many of the best discoveries are actually built-upon and confirmed, rather than simply sweep them aside  As such, plenty of foundational principals, in particular the conservation laws of energy and momentum and essential principals of mechanics and thermodynamics are just as true, right, and relevant as they were hundreds of years ago.
+Scientific discovery and technical progress tends to confirm and build upon many of the best discoveries, rather than simply sweep them aside.
+As such, plenty of foundational principals, in particular the conservation laws of energy and momentum and essential principals of mechanics and thermodynamics are just as true, right, and relevant as they were hundreds of years ago.
 Consequently, whilst we cannot always know the precise degree of advance in a particular field of engineering, we can still confidently know some of the fundamental bounds and constraints within which that field will operate.
@@
 The act of slowing the water down is necessarily equivalent to the act of raising its pressure.
-It is this elevated pressure at the point where the blades to their work on the water which enables the blades to avoid cavitation (the rapid expansion and collapse of small voids or bubbles in the water) which is very large source of noise underwater.
+It is this elevated pressure at the point where the blades do their work on the water which enables the blades to avoid cavitation (the rapid expansion and collapse of small voids or bubbles in the water) which is very large source of noise underwater.
 Maintaining this advantage to the same level necessitates a duct design which slows water down to the same extent.
@@
 
-The model in its central assumptions was able to achieve very realistic results, which were consistent with other public estimates of performance parameters of conventional submarines and their known characteristics \ref{buckingham2008submarine}.
+The model in its central assumptions was able to achieve very realistic results, which were consistent with other public estimates of performance parameters of conventional submarines and their known characteristics \parencite{buckingham2008submarine}.
 A variety of different assumptions regarding the hotel load, battery density and different combinations of efficiency curves were tested.
@@
 
-The full range of plausible efficiency curves exhibited a significant efficiency advantage right through transit speeds.
+The full range of plausible efficiency curves exhibited a significant efficiency advantage right through to transit speeds.
 Central assumptions suggested that this would result in the overall range of a submarine being enhanced by approximately 50\% with the use of propellers instead of a pumpjet.
 None of the scenarios suggested it was likely that this range advantage would be any less than 20\%.
-Given the significant emphasis that has been repeatedly placed the unique requirements for range and endurance for Australia's submarine, this raises substantial questions about suitability of this technology to meet Australia's needs.
+Given the significant emphasis that has been repeatedly placed on the unique requirements for range and endurance for Australia's submarine, this raises substantial questions about the suitability of this technology to meet Australia's needs.
 
@@
 This showed that the difference between the two systems was made substantially larger with smaller hotel loads, and larger stored electical energy (as might be achieved with lithium-ion batteries), and these two effects combined to give some particularly startling results.
-Circumstances where the \textit{difference} in dived range and endurance between the two systems in terms of range and endurance exceeded the \textit{entire} dived range and endurance achieved by many modern conventional submarines frequently arose.
+Circumstances where the \textit{difference} in dived range and endurance between the two systems exceeded the \textit{entire} dived range and endurance achieved by many modern conventional submarines frequently arose.
 
@@
 
-\subsection{They physical relationship}
+\subsection{The physical relationship}
 Perhaps the most important relationship to comprehend in the operation of submarines is the relationship between speed and drag .
@@
 For a fuller discussion of types of drag in the case of ships, see \cite{carlton2007}.
-But in the case of a submarine, which does it's business completely immersed in the ocean, the relevant physics is dominated the skin friction on the hull, which follows a very simple rule and relationship.
+But in the case of a submarine, which does it's business completely immersed in the ocean, the relevant physics is dominated by the skin friction on the hull, which follows a very simple rule and relationship.
 The amount of drag ($F_D$) an objet experiences increases directly in proportion to the surface area $A$.
-For any given object of a certain (unchanging shape) there will be a constant coefficent ($C_D$) which reflects how aerodynamic or hydrodymanic the shape is.
+For any given object of a certain (unchanging) shape there will be a constant coefficent ($C_D$) which reflects how aerodynamic or hydrodymanic the shape is.
 The drag is also directly proportional to the density of the fluid being moved through, $\rho$.
@@
 \footnote{Technical aside: This law applies wherever the the flow over the surface is turbulent.
-It is true that for very small objects, or very viscous fluids, or very slow movements a different apples called Stokes Equation, in the case where Reynolds numbers are less than 1.
+It is true that for very small objects, or very viscous fluids, or very slow movements a different rule applies called Stokes Equation, in the case where Reynolds numbers are less than 1.
 Given that sea-water is not particularly viscous, and sumbarines not particularly small, Reynolds numbers are liekly to be much much greater than 1 (one or two thousand), even when moving at only one or two knots.
@@
 
-Consequently, the difference between the power output a conventional and a nuclear submarine might regularly operate at could be even quite a bit larger than the maximum amount of energy that they can deliver to their propulsion systems.
+Consequently, the difference between the power output a conventional and a nuclear submarine might regularly operate at could be quite a bit larger than the maximum amount of energy that they can deliver to their propulsion systems.
 This has considerable consequences for the design of all of the submarine systems, including the propulsion system.
@@
 It's easy to be somewhat confused by the many different names which seem to be associated with related, or similar systems.
-Between pumps, pumpjets, water-jets ducted or shrouded propellers, or impellers, there is plenty of grounds for some confusion  In this section, I aim to clarify in simple terms what the important differences between distinctly different systems are, and where some terms are used somewhat interchangeably without doing any great violence to the concepts underlying.
+Between pumps, pumpjets, water-jets, ducted or shrouded propellers, or impellers, there is plenty of grounds for some confusion.
+In this section, I aim to clarify in simple terms what the important differences between distinctly different systems are, and where some terms are used somewhat interchangeably without doing any great violence to the concepts underlying.
 Appreciating where the fundamental as opposed to superficial differences lie is important to understanding subsequently the limitations of different systems.
@@
 \begin{tcolorbox}[width=0.8\textwidth, center, colback=blue!5!white,colframe=blue!75!black, title= The difference between a pumpjet and a propeller]
-In it simplest form, a pumpjet can be thought of as a special type of ducted propeller, or propeller contained inside a shroud.
+In its simplest form, a pumpjet can be thought of as a special type of ducted propeller, or propeller contained inside a shroud.
 The impact of the shroud in a pumpjet has the special effect of \textit{decelerating} the water at the point that it reaches the propeller.
@@
 There are broadly three categories of pump types for water which are used for propulsion.
-The first is called an axial-flow.
+The first is called an axial-flow pump.
 In such a pump, the water has work done on in by blades which push the flow in the same direction to the axis of the blades rotation.
@@
 \subsection{Recent Commentary}
-There has been already some attention given to the probable efficiencies of pump jets at low speed.
+There has already been some attention given to the probable efficiencies of pump jets at low speed.
 Recently Andrew Davies published a piece in the The Strategist (ASPI) arguing that the pump jet may actually save fuel at transit speeds (around 8kt) and overall improve the endurance of the submarine, despite likely having lower efficiency at low speeds such as patrol speed \parencite{davies2017}.
@@
 
-There has also been questions put in parliamentary hearings regarding the pump jet \parencite{patrick2017}, and responses from public officials to such questions indicate that some of the literature provided to support such lines of questions was dated, and that the state of the art may have advanced considerably since this time.
+There have also been questions put in parliamentary hearings regarding the pump jet \parencite{patrick2017}, and responses from public officials to such questions indicate that some of the literature provided to support such lines of questions was dated, and that the state of the art may have advanced considerably since this time.
 Whilst the statement that the state of the art is much advanced is obviously true in the case with marine propulsion as with most sciences, it is also equally true that new advances tend to confirm, rather than annul, many of the works that have been previously undertaken.
@@
 The energy penalty inflicted by using the additional devices becomes very large in proportion, and part of the water doesn't do all the necessary things for the jet to work.
-In essence, if you try to slow down something that's already going slowly, it tends to stop  That's exactly what happens in jets operating at very low flow-rates, when flow separation, also known as stall when it occurs between the blades, begins to take place.
+In essence, if you try to slow down something that's already going slowly, it tends to stop.
+That's exactly what happens in jets operating at very low flow-rates, when flow separation, also known as stall when it occurs between the blades, begins to take place.
 When there isn't enough energy to move all the water through the pressure-hurdles imposed by the duct, only some of it moves, and the rest spins in place.
@@
 It also means that for any given change in velocity, the static pressure changes also, but doesn't change as much as the velocity change.
-The consequences for this pumpjets are and inevitable, since the geometry of the duct is inevitably fixed.
+The consequences for pumpjets are inevitable, since the geometry of the duct is inevitably fixed.
 At very high flow rates, the expansion of the area ahead of the impeller creates an adverse pressure gradient, as the water must slow down to fill the greater area.
-This adverse pressure gradient is larger at higher speeds, the proportions aren't linear, hence the the speed is much much larger, due to the square relationship, and flow is consistent (though still turbulent) throughout the entire cross-sectional area of the duct.
+This adverse pressure gradient is larger at higher speeds, but the proportions aren't linear, hence the the speed is much much larger, due to the square relationship, and flow is still consistent (though also turbulent) throughout the entire cross-sectional area of the duct.
 
@@
 \\
-Hotel load also has a significant impact on the \textit{difference} between the pumpjet and propeller affect the performance of the whole system.
+Hotel load also has a significant impact on the \textit{difference} between the pumpjet and propeller performance when considering the whole system.
 By driving the speed at which the propulsion becomes the dominant power demand higher, an extra-large hotel load reduces the overall difference between the two identical submarines, and pushes the peak proportional difference to higher speeds, where endurance and range is naturally lower due to drag increasing.
@@
 However, from \cite{patrick2012}, it can be seen that credible submarine designs might include diesel engines that have a total capacity in the region of 85\% of the maximum power output of the main motor.
-Assuming that there are some necessary losses in the generation and recharging process, and plausibly some increase in the hotel load for the operation of additional systems, fur the purposes of this assessment it is assumed that while snorting approximately 5MW might be available to meet propulsion and normal hotel load requirements, and any surplus can be directed towards charging batteries.
+Assuming that there are some necessary losses in the generation and recharging process, and plausibly some increase in the hotel load for the operation of additional systems, for the purposes of this assessment it is assumed that while snorting approximately 5MW might be available to meet propulsion and normal hotel load requirements, and any surplus can be directed towards charging batteries.
 
@@
 Pumpjets have an acoustic advantage because they can avoid cavitation in circumstances where an open propeller might begin to cavitate, in particular at higher speeds.
-However, at lower speeds such as patrol speeds (and also at greater depths) propellers can easily avoid cavitation because they don't need turn very fast in the water.
+However, at lower speeds such as patrol speeds (and also at greater depths) propellers can easily avoid cavitation because they don't need to turn very fast in the water.
 Consequently, the avoidance of cavitation in these circumstances offers no advantage.

### 2018-08-31 Second.Rnw
first take

\documentclass{article}

\usepackage[backend=bibtex, sorting=none, citestyle=numeric]{biblatex}
\usepackage[margin=0.8in]{geometry}
\bibliography{references}

%  For caption width
\usepackage{graphicx,caption}
% For Centering figs
\usepackage[export]{adjustbox}
% For split figures
\usepackage{subcaption}
%  appendix
\usepackage[titletoc,page]{appendix}
\setlength{\parindent}{0em}

\usepackage{tikz,lipsum,lmodern}
\usepackage[most]{tcolorbox}

% for cover page
\usepackage{geometry}
\usepackage{xcolor}
\usepackage{graphicx}

% \usepackage{draftwatermark}
% \SetWatermarkText{DRAFT}
% \SetWatermarkScale{1.5}
% \SetWatermarkLightness{0.95}

\title{Efficiency}
\author{Aidan Morrison}

% \SweaveOpts{concordance=TRUE}
\definecolor{titlepagecolor}{cmyk}{0, 0, 0, .30}
\definecolor{namecolor}{cmyk}{0, 0, 0, .80}

\begin{document}

%-----------------------------------------------------------------

% ----------------------------------------------------------------
\begin{titlepage}
\newgeometry{left=7.5cm} %defines the geometry for the titlepage
\pagecolor{titlepagecolor}
\noindent

{\huge \textsf{A response to the Government response}}
\vskip\baselineskip
{\huge \textsf{(Or lack thereof)}}
\vskip\baselineskip
{\huge \textsf{on the question of the pumpjet}}
\vskip\baselineskip
\noindent
\textbf{Aidan Morrison}
\vskip\baselineskip
\textsf{August 2018}
\vskip\baselineskip
\vskip\baselineskip

\vfill
\noindent

\includegraphics[width=6cm]{tl_logo.png}\\[-1em]
\color{white}
\makebox[0pt][l]{\rule{1.3\textwidth}{1pt}}
\par
\noindent
\textbf{\textsf{Trendlock}} \textcolor{namecolor}{\textsf{Consulting}}

\end{titlepage}
\restoregeometry % restores the geometry
\nopagecolor% Use this to restore the color pages to white

\newpage

<<setup, include=FALSE, cache=FALSE, echo=FALSE>>=
opts_chunk$set(fig.path='figures/plots-', fig.align='center', fig.show='hold', eval=T, echo=F)
options(replace.assign=TRUE,width=80)
library(tidyverse)
library(submarines)
@

\newpage

\tableofcontents

\newpage
\setlength{\parskip}{1em}

\section{Introduction} \label{intro}

The pumpjet propulsion system is a key feature of Naval Group’s Shortfin Barracuda submarine concept, and was apparently a key reason why the concept was selected by the Australian Government as the basis for the design of Australia’s new submarine under project SEA 1000.  Naval Group’s official website, reinforced by comments from officials representing both Naval Group and the Australian Government,  claims that the pumpjet system offers acoustic advantages which will make the Shortfin Barracuda much more effective than a submarine driven by a conventional propeller.

However, over a number of years now the pumpjet choice has attracted sustained attention and criticism, mostly around its efficiency at low speeds, where a diesel-electric submarine would necessarily spend a great deal of its operating time.  I was commissioned by Gary Johnston to investigate this topic in detail and produce a report, which concluded that a very substantial reduction in dived range and endurance (as well as overall range) would necessarily result from the pumpjet choice.  Furthermore, that paper challenged the assumption that the pumpjet would be acoustically superior at low speeds, where open propellers would have no cavitation, and cause much less turbulence than the equivalent jet.

That paper, and a subsequent blog post on the topic which made the paper publically available, attracted the interest of the Senate Economics Reference Committee Inquiry into the Future of Naval Shipbuilding in Australia.  Senator Chris Ketter asked written questions on notice to the Department of Defence requesting responses to the paper specifically, and related comments from public officials, with a return date of 3rd April 2018.  I was also invited to provide evidence to the committee in a public hearing on the 7th of June 2018.  In that public hearing, I stood by the conclusions of my substantive paper, which was: "between two otherwise identical submarines, the one with the pump jet will always have a lower dived endurance, a lower dived range, a worse indiscretion ratio, a lower overall endurance and a lower overall range than the one with the propeller."  In the presence of both the outgoing acting CEO of Naval Group Australia Brent Clark, and General Manager of Submarines (for Defence) Stephen Johnson, I pointed out that that statement was yet to be challenged, and that I'd be very surprised if anyone with technical competence or professional integrity sought to challenge it.

In their own evidence during the course of the day, neither of those two did choose to challenge it.  Senator Kim Carr requested that defence produce a written response to that report for the committee prior to their handing down of their finaly report for the Inquiry, and Defence agreed to do their best in the time available, with the head of Capability and Sustainment Group (CASG) Kim Gillis noting that "we do hold that expertise", whilst making clear his reluctance to make a response clear, with comments such as "I regularly get correspondence from the public saying that almost every decision that we've made for the last 50 years on a defence matter is wrong, because somebody has an opinion".  In the final report of the committee, this request, and defence's acceptance of it was noted, but no acknowledgement was made of whether any response was actually received, or what it contained.

In the August edition of Australian Defence Magazine a major article was featured on the pumpjet question, referencing both my testimony in the hearing and the paper.  Comment for the article was sought from CASG, the Royal Australian Navy (RAN) and Naval Group, and none was forthcoming.  However, some substantial comments were provided by Dr David Kershaw, the chief of DSTG's Maritime Division, who was also Science and Technology advisor for a time on SEA 1000, the Future Submarine Project.

This paper has been commissioned by Gary Johnston of Submarines for Australia to provide a summary of the responses that have been received from official sources (in government, and Naval Group) to the earlier major report, and respond to them.  Only two substantive arguments emerge seeking to engage with the technical arguments made in the earlier paper.  The first is that the large size of the submarine now makes the pumpjet viable, where as for smaller submarines the large weight of the pumpjet would make it unviable, and that the pumpjet would be generally well-suited to smaller sumbarines as well were it not for this weight consideration.  The second is that interaction between the pumpjet and the hull obviate the evidence I have produced showing that pumpjet efficiency declines at low speeds, which are claimed to neglect considering such interactions.  This paper seeks to consider those two lines of reasoning in sufficient detail to suitably discuss the circumstances in which such arguments might be made.

\begin{appendices}
\section{Extracts from public statements}

\subsection{Hansard Element}

CHAIR: Senator, Hanson, you had a couple of final questions.

Senator HANSON: The one reason for selecting the Shortfin Barracuda was stated to be that it has a very quiet pump-jet propulsion system, as used on nuclear submarines. Engineering experts said at the time that pump-jets were inefficient for conventional submarines, but Naval Group said that propellers were now obsolete. Recently, however, Naval Group said that propellers may be used instead. Just two weeks ago Naval Group said that we may have to go back to propellers. So, are we going with pump-jet propulsion or are we going back to propellers?

Vice Adm. Griggs: The Commonwealth intends to incorporate a pump-jet into the Future Submarine, and the design is proceeding along on that basis.

Senator HANSON: When the choice of Naval Group was announced, the Minister for Defence Industry stated that all 12 submarines would be built in Australia. At the same time, then President Hollande announced the creation of 4,000 jobs in France, while his defence minister stated on French TV that the first two submarines would be built in France. Can the minister explain the discrepancy and assure the Senate that all submarines will be built in Australia.

Senator Payne: All the submarines will be built in Australia.

\end{appendices}

\end{document}

### 2018-09-03 Second.Rnw
With the first hansard ref all done

@@
 
-The pumpjet propulsion system is a key feature of Naval Group’s Shortfin Barracuda submarine concept, and was apparently a key reason why the concept was selected by the Australian Government as the basis for the design of Australia’s new submarine under project SEA 1000.
-Naval Group’s official website, reinforced by comments from officials representing both Naval Group and the Australian Government,  claims that the pumpjet system offers acoustic advantages which will make the Shortfin Barracuda much more effective than a submarine driven by a conventional propeller.
+The pumpjet propulsion system is a key feature of Naval Group's Shortfin Barracuda submarine concept, and was apparently a key reason why the concept was selected by the Australian Government as the basis for the design of Australia's new submarine under project SEA 1000.
+Naval Group's official website, reinforced by comments from officials representing both Naval Group and the Australian Government,  claims that the pumpjet system offers acoustic advantages which will make the Shortfin Barracuda much more effective than a submarine driven by a conventional propeller.
 
 However, over a number of years now the pumpjet choice has attracted sustained attention and criticism, mostly around its efficiency at low speeds, where a diesel-electric submarine would necessarily spend a great deal of its operating time.
-I was commissioned by Gary Johnston to investigate this topic in detail and produce a report, which concluded that a very substantial reduction in dived range and endurance (as well as overall range) would necessarily result from the pumpjet choice.
+I was commissioned by Gary Johnston to investigate this topic in detail and produce a report \parencite{morrison2018}, which concluded that a very substantial reduction in dived range and endurance (as well as overall range) would necessarily result from the pumpjet choice.
 Furthermore, that paper challenged the assumption that the pumpjet would be acoustically superior at low speeds, where open propellers would have no cavitation, and cause much less turbulence than the equivalent jet.
@@
 I was also invited to provide evidence to the committee in a public hearing on the 7th of June 2018.
-In that public hearing, I stood by the conclusions of my substantive paper, which was: "between two otherwise identical submarines, the one with the pump jet will always have a lower dived endurance, a lower dived range, a worse indiscretion ratio, a lower overall endurance and a lower overall range than the one with the propeller."  In the presence of both the outgoing acting CEO of Naval Group Australia Brent Clark, and General Manager of Submarines (for Defence) Stephen Johnson, I pointed out that that statement was yet to be challenged, and that I'd be very surprised if anyone with technical competence or professional integrity sought to challenge it.
+In that public hearing, I stood by the conclusions of my substantive paper, which was: ``between two otherwise identical submarines, the one with the pump jet will always have a lower dived endurance, a lower dived range, a worse indiscretion ratio, a lower overall endurance and a lower overall range than the one with the propeller."  In the presence of both the outgoing acting CEO of Naval Group Australia Brent Clark, and General Manager of Submarines (for Defence) Stephen Johnson, I pointed out that that statement was yet to be challenged, and that I'd be very surprised if anyone with technical competence or professional integrity sought to challenge it.
 
 In their own evidence during the course of the day, neither of those two did choose to challenge it.
-Senator Kim Carr requested that defence produce a written response to that report for the committee prior to their handing down of their finaly report for the Inquiry, and Defence agreed to do their best in the time available, with the head of Capability and Sustainment Group (CASG) Kim Gillis noting that "we do hold that expertise", whilst making clear his reluctance to make a response clear, with comments such as "I regularly get correspondence from the public saying that almost every decision that we've made for the last 50 years on a defence matter is wrong, because somebody has an opinion".
-In the final report of the committee, this request, and defence's acceptance of it was noted, but no acknowledgement was made of whether any response was actually received, or what it contained.
+Senator Kim Carr requested that defence produce a written response to that report for the committee prior to their handing down of their finaly report for the Inquiry, and Defence agreed to do their best in the time available, with the head of Capability and Sustainment Group (CASG) Kim Gillis noting that ``we do hold that expertise", whilst making clear his reluctance to make a response clear, with comments such as "I regularly get correspondence from the public saying that almost every decision that we've made for the last 50 years on a defence matter is wrong, because somebody has an opinion".
+In the final report of the committee, this request, and defence's acceptance of it was noted, but no acknowledgement was made of whether any response was actually received, or what it contained \parencite{SERC2018final}.
 
-In the August edition of Australian Defence Magazine a major article was featured on the pumpjet question, referencing both my testimony in the hearing and the paper.
+In the August edition of Australian Defence Magazine a major article was featured on the pumpjet question \parencite{kerr2018}, referencing both my testimony in the hearing and the paper.
 Comment for the article was sought from CASG, the Royal Australian Navy (RAN) and Naval Group, and none was forthcoming.
@@
 
+\section{A brief chronology of comment}
+
+The following gives a brief overview of the significant known public comments regarding the pumpjet that have been made by official representatives of either the Commonwealth of Australia, or Naval Group, formally known as DCNS.
+Only brief quotations are selected here, with references being made to the full excerpts either in Appendices or References.
+
+\begin{tcolorbox}[width=0.9\textwidth, center, colback=red!5!white,colframe=red!75!black, title= \textbf{28 May 2012 - Senate Estimates}]
+\textit{Senator Johnston directly challenges Rear Admiral Justin Moffit (then director of SEA 1000) about the pumpjet proposal.  Full exchange in \ref{SENMoffit}.}
+\\
+\\
+\textbf{Senator JOHNSTON:} I am told that there are no pump jets on any conventional class submarines in the world.
+\\
+\textbf{Rear Adm. Moffitt:} That is quite right. It might be a very easy task for us to write it off. I do not know, because I have not done-
+\\
+\textbf{Senator JOHNSTON:} The French and the Russians have tried. No-one has used them. They are not being used on anything other than nuclear powered submarines-particularly the Americans.
+\\
+\textbf{Rear Adm. Moffitt:} If it does not look like it makes sense, I will rule it out. But I cannot give you the answer to that yet because we have not done the work.
+\\
+\textbf{Senator JOHNSTON:} Who had the idea to look at pump jets? Where did that come from? Did it come from capability group saying that we had better had a look at that?
+\\
+\textbf{Rear Adm. Moffitt:} I do not know the answer to that. I doubt that there is a clear answer.
+
+\end{tcolorbox}
+
+\newpage
+
+\printbibliography
+
+\newpage
+
 \begin{appendices}
+
 \section{Extracts from public statements}
 
-\subsection{Hansard Element}
+\subsection{Hansard Extract, Senate Estimates - 28th May 2012 \cite{SE2012moffit}}\label{SENMoffit}
 
-CHAIR: Senator, Hanson, you had a couple of final questions.
+\textbf{Senator JOHNSTON:} There were announcements made about these various studies.
+You have explained propulsion and energy storage.
+I get that.
+Signatures and stealth performance is self evident.
+You have touched on combat systems and the hydrodynamics of the propellers.
+But pump jets is the one that has the big question mark over it.
+Why are we looking at pump jets?
 
-Senator HANSON: The one reason for selecting the Shortfin Barracuda was stated to be that it has a very quiet pump-jet propulsion system, as used on nuclear submarines.
-Engineering experts said at the time that pump-jets were inefficient for conventional submarines, but Naval Group said that propellers were now obsolete.
-Recently, however, Naval Group said that propellers may be used instead.
-Just two weeks ago Naval Group said that we may have to go back to propellers.
-So, are we going with pump-jet propulsion or are we going back to propellers?
+\textbf{Rear Adm. Moffitt:}  Why are we looking at it?
 
-Vice Adm. Griggs: The Commonwealth intends to incorporate a pump-jet into the Future Submarine, and the design is proceeding along on that basis.
+\textbf{Senator JOHNSTON:} Why are we looking at pump jets on a conventionally powered submarine?
 
-Senator HANSON: When the choice of Naval Group was announced, the Minister for Defence Industry stated that all 12 submarines would be built in Australia.
-At the same time, then President Hollande announced the creation of 4,000 jobs in France, while his defence minister stated on French TV that the first two submarines would be built in France.
-Can the minister explain the discrepancy and assure the Senate that all submarines will be built in Australia.
+\textbf{Rear Adm. Moffitt:}  We are not specifically ruling in or out anything at the moment. Pump jets are in use in submarines and we will-
 
-Senator Payne: All the submarines will be built in Australia.
+\textbf{Senator JOHNSTON:} Mainly nuclear submarines with a speed far and away beyond 25 knots.
+
+\textbf{Rear Adm.
+Moffitt:}  That is true, as I understand it.
+The issue is that we need to understand that, because we have not got much knowledge of it, before we rule it in or out.
+There are many new technologies, including pump jets, AIP and lithium batteries, that have emerged since we executed the Collins program about which we know relatively little and wish to become more educated about than we are today before we recommend to government that they be included or discounted.
+We just do not know enough to give a sensible answer to government at the moment.
+I am not saying that pump jets are in.
+Nor am I saying that they are out.
+We just do not know enough yet.
+
+\textbf{Senator JOHNSTON:} What is our methodology to get that information? Who are we asking? How much is all that costing?
+
+\textbf{Rear Adm. Moffitt:}  We will be relying fairly heavily on DSTO to help us with that. We will also be fairly keen to get some advice from the US Navy on that as well.
+
+\textbf{Senator JOHNSTON:} I am told that there are no pump jets on any conventional class submarines in the world.
+
+\textbf{Rear Adm. Moffitt:}  That is quite right. It might be a very easy task for us to write it off. I do not know, because I have not done-
+
+\textbf{Senator JOHNSTON:} The French and the Russians have tried. No-one has used them. They are not being used on anything other than nuclear powered submarines—particularly the Americans.
+
+\textbf{Rear Adm. Moffitt:}  If it does not look like it makes sense, I will rule it out. But I cannot give you the answer to that yet because we have not done the work.
+
+\textbf{Senator JOHNSTON:} Who had the idea to look at pump jets? Where did that come from? Did it come from capability group saying that we had better had a look at that?
+
+\textbf{Rear Adm.
+Moffitt:}  I do not know the answer to that.
+I doubt that there is a clear answer.
+It is a submarine technology that we think that we need to know more about than we do today.
+I do not anticipate spending a substantial amount of money to get the understanding that I am looking for—I may not have to spend anything at all.
 

### 2018-09-05 Second.Rnw
some QONs included

@@
 
+\begin{tcolorbox}[width=0.9\textwidth, center, colback=blue!5!white,colframe=blue!75!black, title= \textbf{May 2016 onwards - DCNS/Naval Group Website for Shortfin Barracuda}]
+\textit{The official website for the submarine claims the advantages of the pumpjet in specific terms, drawing comparisons with "obsolete propeller technology" and "otherwise identical submarines".
+Last accessed 4 Sept 2018 \parencite{naval2018}.}
+\\
+\\
+``Pump jet propulsion means the Shortfin Barracuda can move more quietly than submarines with obsolete propeller technology.
+In a confrontation between two otherwise identical submarines, the one with pump jet propulsion always has the tactical advantage."
+
+\end{tcolorbox}
+
+\begin{tcolorbox}[width=0.9\textwidth, center, colback=blue!5!white,colframe=blue!75!black, title= \textbf{4 October 2017 - Jean-Michel Billig, Executive Director Australian submarine Program, Naval Group}]
+\textit{In comments to journalists at the PACIFIC 2017 trade show, the Naval Group executive in charge of the Australian Program speculates that convetional propellers may be used on the Shortfin Barracuda \parencite{sbs2017}.
+Subsequent reports quote Naval Group shifting responsibility for the propulsion choice to the Australian Government.}
+\\
+\\
+``The vessels may end up with conventional propellers as well as air independent propulsion, which helps to increase underwater endurance, Mr Billig said."
+\\
+\\
+``However decisions on forms of propulsion are ultimately a matter for the Australian Government."\parencite{iggulden2017}
+
+\end{tcolorbox}
+
+\begin{tcolorbox}[width=0.9\textwidth, center, colback=black!5!white,colframe=black!75!black, title= \textbf{3 November 2017 - Department of Defence, in response to Question on Notice from Senator Xenophon from October 2017}]
+\textit{As quoted in an ASPI Strategist blog post\parencite{davies2017}, and also by Senator Patrick in \ref{SENjohnson}, the Department of Defence argued in defence of the pump jet's efficiency.}
+\\
+\\
+'A [pump jet] can provide higher propulsive efficiency across the speed range of a submarine, including the patrol speed of conventional submarines'.
+
+\end{tcolorbox}
+
+\begin{tcolorbox}[width=0.9\textwidth, center, colback=red!5!white,colframe=red!75!black, title= \textbf{25 Oct 2017 - Senate Estimates}]
+\textit{Senator Hanson directly challenges Rear Admiral Gregory Sammut (then director of SEA 1000) about the pumpjet proposal.  Full exchange in \ref{SENhanson}.}
+\\
+\\
+\textbf{Rear Adm. Sammut:}   I think the point that we're trying to make here is that the proposal
+offered by France included a pump jet.
+
+\textbf{Senator Hanson:}   So it is pump jet? So you're going pump jet?
+
+\textbf{Rear Adm. Sammut:}   The proposal—
+
+\textbf{Senator Hanson:}   This is why you've chosen them as the design—to go with the pump
+jet. So it's not propellers, is it?
+
+\textbf{Rear Adm. Sammut:}   The proposal offered by France included a pump jet, and that's the
+proposal that we've accepted. A pump jet remains on offer, and that's what we intend to use
+for the Future Submarine.
+
+\textbf{Senator Hanson:}   And that's why you chose France's Naval Group to—
+
+\textbf{Rear Adm. Sammut:}   No, that is not, alone, why we chose France....
+\\
+\\
+\textit{Extensive remarks from Rear Adm. Sammut, and further questions from Senator Hanson about whether the proposal involves converting a nuclear submarine to a diesel sumbarine}
+\\
+\\
+\textbf{Rear Adm. Sammut:}  ...  The effort is not to convert an
+existing boat into a diesel-electric submarine; we're designing a submarine using many of the
+learnings and references that come out of the Barracuda design. A good example there would
+be something like hull diameter. If you're going to change the diameter of the hull of a
+submarine, you're into a whole new set of calculations about the structural integrity and
+strength of the submarine. If we can utilise the work that France has already done in having a
+hull diameter that's very similar to Barracuda, we can leverage all of that design work, which
+has not only gone into the Barracuda but also has been tested by France in their methods of
+design and certification to get to this point.
+
+\end{tcolorbox}
+
+\begin{tcolorbox}[width=0.9\textwidth, center, colback=red!5!white,colframe=red!75!black, title= \textbf{15 Dec 2017 - Senate Estimates}]
+\textit{Senator Rex Patrick asks Stephen Johnson (General Manager Submarines) about the pumpjet proposal, and produces documents suggesting they are less efficient at low speed.
+The Defence Chief Scientist, Dr Zelinsky, and Chief of Defence Force, Air Chief Marshal Binskin, and Head of Capability and Sustainment Group, Kim Gillis are also involved in the discussion, defending the choice of the pumpjet.
+Senator Patrick also specifically raises the concern about the impact of the weight of the pumpjet at the extremity of the submarine.
+Full exchange in \ref{SENjohnson}.}
+\\
+\\
+\textbf{Senator PATRICK:} Mr Johnson, there's been a lot of talk about pump jets.
+I've now tabled a document which has come from the United States, from Connecticut.
+You'd be very familiar with the organisation there.
+That basically describes the characteristics of pump jets, their pros and cons, as any piece of engineering equipment has.
+The document—and there's also another document that I've taken from an ASPI site—shows the efficiency characteristics of a pump jet.
+Both the US document—now declassified—and the document from the ASPI website suggest that, at low speed, pump jets are relatively inefficient compared to, say, a propeller.
+That may well be the reason why we see a pump jet on a Mark 48 torpedo but generally not on merchant ships and so forth.
+Would you care to comment on that?
+
+\textbf{Mr Johnson:} As we said just a little earlier, the design process—whether it is a propeller or propulsion—is very much keyed to the criteria that you look for.
+The first thing I said was speed.
+No matter what kind of propeller or propulsion pump jet you design, it has to be optimised for a particular speed.
+There are classes of surface ships that look to optimise cross-ocean transit.
+Physics usually keeps you around 17 to 18 knots, and then there is the cost of maintenance and the cost of repair and the fact that there's no particular stealth requirement.
+Typically, you will find that those ships will use an open-tip propeller.
+However, other ships, cruise ships in particular, use an electric shrouded pod, which is a variation of a pump-jet technology.
+Of course, the technology gets changed over time, so the older a document is, the more it's dated to that point in time.
+\\
+\\
+...
+\\
+\\
+\textbf{Mr Johnson:} We were talking about optimising for the speed and the characteristics of the pump jet.
+The decision-making process for submarines is a little more complicated because there's an acoustic performance aspect as well as the pure efficiency of the blade-propulsor-pump jet arrangement.
+Without going into details, it's a decision that we make with Navy, but I'm confident, having done this a few times already, that we are going to find this pump-jet approach to be the optimum for our particular circumstances for this particular submarine.
+\\
+\\
+...
+\\
+\\
+\textbf{Senator PATRICK:} Okay.
+I'll go back to you, Mr Johnson.
+Noting that you're saying that this is based on something that hasn't been to sea yet - it will go to sea on the Barracuda; presumably that's the submarine you're referring to-there's still a risk that's sitting there.
+As the guardian of the taxpayer, I know where you sit.
+I have a responsibility from the taxpayers' perspective to ask: do you have a fallback position if the data that comes from the French trials, or other trials that you might do, presents you with data that's inconsistent with the operational profile that you intend for the submarine?
+
+\textbf{Mr Johnson:} Yes.
+We will understand the performance data early enough in the design that we can accommodate whatever is required.
+And then, just to reassure you, it's really going to be an adjustment to the design of the pump jet, not a decision to leave it.
+Without going into too much detail, the curve that you're referring to is about the efficiency of the propeller, not about its acoustic performance.
+As you know, there's nothing free in physics, so those are different-shaped curves.
+What we seek to do, as Chief of the Defence Force said, is to make the right balance of multiple dimensions of this now.
+
+\end{tcolorbox}
+
+\begin{tcolorbox}[width=0.9\textwidth, center, colback=black!5!white,colframe=black!75!black, title= \textbf{9 February 2018 - Department of Defence, in response to Question on Notice from Senator Rex Patrick in December 2017}]
+\textit{Senator Patrick inquires about the nature, cost, and conclusions of earlier research into pumpjets undertaken as part of \$214 million worth of ``de-risking" studies already undertaken, as well as the schedule and cost of subsequent research and development of the pumpjet.
+Prior research consisted of a single 99 page report obtained via literature review and discussion with experts at a cost of $68,825.
+Ongoing research included $1.1million spent to date to develop a generic submarine and pumpjet design, compuational and analytic models, and validation of these models through experimental studies.
+The development and test program is not scheduled to commence until the last quarter of 2018.
+Full questions and responses can be found in \ref{QON148}, \ref{QON150}, \ref{QON151}}
+\\
+\\
+``The study concluded that pumpjets have the potential to deliver increased propulsive efficiency and lower shaft torque requirements, relative to an open propeller.
+This should lead to beneficial platform systems and operational impacts to increase vessel speed and/or submerged endurance.
+The study developed a parametric model of pumpjet thrust and efficiency that has been used as the basis for more recent studies into pumpjet performance."
+\\
+\\
+``Specifications for the pump jet are being developed as design of the Future Submarine proceeds.
+Initial inputs will be developed in the last quarter of 2018, which marks the beginning of the development and test program for the pump jet.
+Development, testing, and assurance will be conducted by several parties including Naval Group, Direction générale de l’armement (DGA), and DSTG.
+Testing and assurance activities will take place in France (including at the premises of the DGA Techniques Hydrodynamiques in Val de Reuil) and Australia.
+Testing will be largely completed by CDR."
+
+\end{tcolorbox}
+
 \newpage
@@
 
+\newpage
+\subsection{Hansard Extract, Senate Estimates - 25th October 2017 \cite{SEN2017hanson}}\label{SENhanson}
+
+\textbf{Senator Hanson:}   I have a question about the Shortfin Barracuda subs. Just to get it on the record, is it correct that the cost of the contract is \$50 billion?
+
+\textbf{Rear Adm.
+Sammut:}   The cost of the program is \$50 billion in acquisition.
+That is not the cost of the contract per se; that is the costs that will be required to establish the submarine capability in full, which extends to not only the acquisition of the submarines but the
+establishment of the facilities necessary to operate the submarine; the establishment of a shipyard to be able to build those submarines; wharfs; training centres; test and integration facilities for the combat system; the acquisition of the combat system; and the establishment
+of a sovereign capability in Australia as well.
+
+\textbf{Senator Hanson:}   Is that for a total of 12 submarines?
+
+\textbf{Rear Adm. Sammut:}   That's for the submarine capability, which comprises, as part of
+that, 12 submarines.
+
+\textbf{Senator Hanson:}   When was the contract signed?
+
+\textbf{Rear Adm. Sammut:}   There will be several contracts signed for the delivery of the Future
+Submarine. The first contract was signed in September last year. That was known as the
+design and mobilisation contract, under which we have commenced design activities and
+activities in terms of mobilising both Naval Group and Lockheed Martin, along with the
+Commonwealth, to commence that design process, which will continue over the next several
+years. By way of example, we have commenced the design of the submarine itself. We've also
+established a resident project office in Cherbourg. We have commenced basing our people in
+France. We've started the development of our engineering plans, the establishment of an
+integrated master schedule and so on.
+
+\textbf{Senator Hanson:}   At what time prior to that did you do investigations around other
+submarines—how many months prior to signing that contract? How much time and effort did
+the Defence Force put into looking at other subs that were available prior to the Shortfin
+Barracuda?
+
+\textbf{Rear Adm. Sammut:}   We commenced the competitive evaluation process in 2015 and
+work on other options had been proceeding from about 2011. So there was about four years of
+work looking at various options that included military off the shelf, modified military off the
+shelf, and evolved Collins.
+
+\textbf{Senator Hanson:}   With regard to the Shortfin Barracuda, you have signed a contract
+with Naval Group, which is a French firm. Did you specifically go looking for a diesel
+submarine? Or what were you looking for?
+
+\textbf{Rear Adm. Sammut:}   The competitive evaluation process was for a conventionally
+powered submarine, a diesel-electric submarine, not any other form of submarine.
+\textbf{Senator Hanson:}   Was it based on a pump jet sub?
+
+\textbf{Rear Adm. Sammut:}   What we were looking for was based on the best solution that the
+three international participants within the competitive evaluation process could offer to meet
+Australia's capability requirements. We did not specify to any of the participants the form of
+solution it should take, the form of the submarine. It was up to those international participants
+to present their design and, more importantly, their design capabilities to meet Australia's
+capability requirements for its Future Submarine.
+
+\textbf{Senator Hanson:}   You didn't want a nuclear sub, that was completely out of the
+equation?
+
+\textbf{Rear Adm. Sammut:}   That's correct, a nuclear submarine was not in consideration. The
+government policy was that we were to acquire conventionally powered submarines, the
+Future Submarine.
+
+\textbf{Senator Hanson:}   So you were looking at either a pump jet or a diesel?
+
+\textbf{Rear Adm. Sammut:}   I think you're confusing, Senator, a pump jet as being a nuclearpowered
+submarine, versus what is in fact a conventionally powered submarine. A pump jet is
+a form of propulsion. It's not a form of power for the submarine. It is a form of propulsion that
+fits to the back of the submarine and, rather than taking the form of a conventional propeller,
+it takes the form of a shrouded rotor with a set of stators in there that, in operation, creates
+thrust through the water.
+
+\textbf{Senator Hanson:}   All right. How are the Shortfin Barracuda run—on what?
+
+\textbf{Rear Adm. Sammut:}   I think the point that we're trying to make here is that the proposal
+offered by France included a pump jet.
+
+\textbf{Senator Hanson:}   So it is pump jet? So you're going pump jet?
+
+\textbf{Rear Adm. Sammut:}   The proposal—
+
+\textbf{Senator Hanson:}   This is why you've chosen them as the design—to go with the pump
+jet. So it's not propellers, is it?
+
+\textbf{Rear Adm. Sammut:}   The proposal offered by France included a pump jet, and that's the
+proposal that we've accepted. A pump jet remains on offer, and that's what we intend to use
+for the Future Submarine.
+
+\textbf{Senator Hanson:}   And that's why you chose France's Naval Group to—
+
+\textbf{Rear Adm. Sammut:}   No, that is not, alone, why we chose France. There are a number of
+things. I am on record as providing testimony at the Senate of the number of things we
+considered in the selection of Naval Group as our international partner. There were a range of
+factors, and they were all over the capability of the international partner to deliver a regionally
+superior submarine and one which Australia would have the sovereign capacity to operate and
+sustain throughout its life. There were a number of criteria in mind.
+We looked at the capabilities. That includes the platform capabilities. We looked at the
+ability of the international partners to work with the combat system integrator of our choice.
+Those things in particular pertain to the capability of the boat. We looked at rough order of
+magnitude costs and schedules to ensure the participants understood the scale of the program
+that we were embarking upon and could give us insights into the manner in which they would
+manage costs and schedules throughout the program. We looked at matters concerning
+program management. We looked at the design and the way that safety is incorporated into
+the design. It's important that that was understood from the outset because, whilst the
+international partner would be designing the submarine, it would be the Commonwealth that
+would be responsible for certifying the submarine as safe and fit for operations, and we
+needed to make sure that the certifying process would match the design philosophy.
+We looked at methods of sustainment. We looked at the manner in which each of the
+participants would support the establishment of a sovereign sustainment capability in
+Australia. Of course, that was a key lesson from the Collins program. Even though we built
+the submarines in Australia, it took us some time to establish the sovereign capacity to sustain
+those boats. Indeed, having done that now is the reason why we are able to deliver beyond
+international benchmark availability. We also looked into the way in which data would be
+transferred to Australia, the manner in which intellectual property would also be managed in
+the program. We looked at the ability of the participants to manage risks. Again, throughout
+this complex program, that would be a very important factor. I would say that we didn't just
+pick a submarine with a pump jet; we picked an international partner that could meet a range
+of requirements to achieve not only a regionally superior submarine but the sovereign
+capacity to operate and sustain that submarine and to deliver that within the time frame that
+would be necessary.
+I'd like to point out that that process was peer reviewed and it was peer reviewed by an
+expert advisory panel. The expert advisory panel found that the process was sound and that
+the participants had been treated fairly. I might add that the expert advisory panel also met
+with each of the participants on several occasions to ensure that they were satisfied that the
+process was being run competently and soundly. I'd also say that the results of our work were
+peer reviewed by former program managers from the United States, who made no
+recommendations to change the course and the manner in which we were conducting the
+competitive evaluation process and indeed came to the conclusion that the manner in which it
+was being conducted would indeed lead to appropriate recommendations on the selection of
+an international partner.
+
+\textbf{Senator Hanson:}   That is correct: you did get advice from the United States, from US
+Navy secretary Don Winter, retired US Navy Vice Admiral Paul Sullivan and Rear Admiral
+Thomas Eccles. They were actually on this based on sound experience. But the fact is that
+America had not launched a diesel submarine for 55 years.
+
+\textbf{Rear Adm. Sammut:}   This was about a submarine program. These are people with a lot of
+experience in submarine design. Might I also add that we have a lot of experience in the
+operation of diesel-electric submarines ourselves. We understand many of the factors around
+battery technology. I'd just like to add—
+
+\textbf{Senator Hanson:}   Sorry, I have limited time with this. I just want to ask you—
+
+\textbf{Senator GALLACHER:} Surely courtesy on both sides is required here.
+
+\textbf{CHAIR:} Senator Hanson has the call.
+
+\textbf{Senator Hanson:}   I agree, but I need to ask these questions. I just have to say: you're
+actually going to put pump jets into a sub. Is it true that pump-jet submarines can stay
+underwater for only 20 minutes?
+
+\textbf{Rear Adm. Sammut:}   No, I do not know what you're referring to about pump-jet
+submarines remaining underwater for 20 minutes. Can I be clear again: a pump jet is a form
+of propeller. It has no bearing on how deep a submarine can go or how long it can stay
+underwater. They are matters that pertain to the platform design of the submarine.
+
+\textbf{Senator Hanson:}   What about a diesel sub? We are going diesel.
+
+\textbf{Rear Adm. Sammut:}   We are going to buy diesel-electric submarines, and I can assure
+you they are going to stay underwater for much longer than 20 minutes.
+
+\textbf{Senator Hanson:}   Three days? What's the period of time?
+
+\textbf{Rear Adm. Sammut:}   I am not going to reveal what our requirements are for submerged
+endurance. That is classified.
+
+\textbf{Senator Hanson:}   I don't think it's classified.
+
+CHAIR: You don't get to make the call on what's classified and unclassified in this
+conversation.
+
+\textbf{Senator Hanson:}   So you can't disclose knowledge of how long a diesel sub can stay
+underwater compared to a nuclear sub?
+
+Mr Johnson: A diesel-electric submarine can stay underwater depending on the way it's
+designed. That refers to matters about the size of the battery, the manner in which the hotel
+load, which is the power that is consumed by the submarine when it's operating, is used and
+so forth. There are a number of factors that go into the period of time that a diesel-electric
+submarine can remain submerged.
+
+\textbf{Senator Hanson:}   This Barracuda—there is still one on the slipways in France that
+hasn't even been tested. We're basing the design on that which has not been tested. Do you
+believe that is in our interest?
+
+\textbf{Rear Adm. Sammut:}   We have selected Naval Group as competent designers of dieselelectric
+submarines and indeed ocean-going submarines. We made that assessment on the
+basis of the competitive evaluation process, which, as I've endeavoured to explain, was a very
+thorough process. We have no reason to believe that France is not capable of designing a
+submarine to meet our requirements.
+
+\textbf{Senator Hanson:}   You don't believe we may be making the same mistakes as we did
+with the Collins class when that was first designed, which cost us billions of dollars?
+
+\textbf{Rear Adm. Sammut:}   No, I don't, because we've learned many lessons from the Collins
+program and implemented those as we've gone forward with the Future Submarine.
+
+\textbf{Senator Hanson:}   Is the Barracuda designed as a nuclear sub—that is, basically is a
+nuclear sub?
+
+\textbf{Rear Adm. Sammut:}   The French Barracuda, the submarine that they will be employing
+in their navy, is a nuclear powered submarine. The submarine that we're working on with
+Naval Group is a diesel-electric submarine.
+
+\textbf{Senator Hanson:}   But it is still the Shortfin Barracuda, which is a nuclear sub, so we're
+taking a nuclear sub and we're going to make it a diesel sub.
+
+\textbf{Rear Adm. Sammut:}   No, we are designing a submarine that will have a certain lineage in
+the work that France has done in designing both the Barracuda and conventional submarines.
+That's quite natural in any submarine design and build process. If you looked at any nation
+designing and building a submarine, they will always refer back to the lineage of designs
+they've done. What we have the benefit of doing here with France is leveraging the design
+work that they've done in the Barracuda and a number of conventional submarines that
+they've designed, built and delivered to other navies, and using that as the basis for the design
+of the Future Submarine, which I said, is a conventionally-powered submarine. You cannot
+convert a nuclear submarine to conventional power. That has often been stated in the media,
+and I understand most recently in the Insight Economics report. The effort is not to convert an
+existing boat into a diesel-electric submarine; we're designing a submarine using many of the
+learnings and references that come out of the Barracuda design. A good example there would
+be something like hull diameter. If you're going to change the diameter of the hull of a
+submarine, you're into a whole new set of calculations about the structural integrity and
+strength of the submarine. If we can utilise the work that France has already done in having a
+hull diameter that's very similar to Barracuda, we can leverage all of that design work, which
+has not only gone into the Barracuda but also has been tested by France in their methods of
+design and certification to get to this point.
+
+\newpage
+\subsection{Hansard Extract, Senate Estimates - 15th December 2017 \cite{SEN2017johnson}}\label{SENjohnson}
+
+\textbf{Senator KITCHING:} Are the future subs going to have pump jet propulsion?
+
+\textbf{Mr Johnson:} Yes, they will. Given you're interested, I might take a moment or two to explain.
+
+\textbf{Senator KITCHING:} Yes, sure. Thank you.
+
+\textbf{Mr Johnson:} The method of designing for either a propeller system or a propulser system, a pump jet, is very similar.
+It starts with key parameters: how fast do we want to go, how big, what's the weight and diameter of the vessel that we're going to power, is it a surface ship or a submarine—although the calculations are really different—and then what's the shape of that hull?
+All of those things then establish the particular characteristics of a propeller or a propulser pump jet design.
+The submarine that we're building is well suited, and the requirements that we have for performance are stringent that the pump jet technology is the optimum approach.
+Since it's a new design—contrary to misunderstandings in the press, this is a new design—we're in a good position to do those calculations for the pump jet from day one and to balance the submarine's designs so that it performs very well with that pump jet technology.
+That's been our plan since the selection of what was then DCNS and is now Naval Group France.
+
+.
+.
+.
+
+\textbf{Senator PATRICK:} I go back to pump jets, if that's possible. Maybe the Chief Defence Scientist could come to the table if he's about. He might be able to help.
+
+\textbf{Senator Payne:} Dr Zelinsky will be here somewhere.
+
+\textbf{Senator PATRICK:} Mr Johnson, there's been a lot of talk about pump jets.
+I've now tabled a document which has come from the United States, from Connecticut.
+You'd be very familiar with the organisation there.
+That basically describes the characteristics of pump jets, their pros and cons, as any piece of engineering equipment has.
+The document—and there's also another document that I've taken from an ASPI site—shows the efficiency characteristics of a pump jet.
+Both the US document—now declassified—and the document from the ASPI website suggest that, at low speed, pump jets are relatively inefficient compared to, say, a propeller.
+That may well be the reason why we see a pump jet on a Mark 48 torpedo but generally not on merchant ships and so forth.
+Would you care to comment on that?
+
+\textbf{Mr Johnson:} As we said just a little earlier, the design process—whether it is a propeller or propulsion—is very much keyed to the criteria that you look for.
+The first thing I said was speed.
+No matter what kind of propeller or propulsion pump jet you design, it has to be optimised for a particular speed.
+There are classes of surface ships that look to optimise cross-ocean transit.
+Physics usually keeps you around 17 to 18 knots, and then there is the cost of maintenance and the cost of repair and the fact that there's no particular stealth requirement.
+Typically, you will find that those ships will use an open-tip propeller.
+However, other ships, cruise ships in particular, use an electric shrouded pod, which is a variation of a pump-jet technology.
+Of course, the technology gets changed over time, so the older a document is, the more it's dated to that point in time.
+
+\textbf{Air Chief Marshal Binskin:} Can I just confirm—we've got two documents and I'm trying to make sure we've baselined ourselves here. We've got one document, which is the 1963 document.
+
+\textbf{Senator PATRICK:} That's correct.
+
+\textbf{Air Chief Marshal Binskin:} We haven't got the other document yet.
+
+\textbf{Senator PATRICK:} It's just a diagram.
+
+\textbf{Air Chief Marshal Binskin:} Thanks for that.
+
+\textbf{Senator Payne:} I'm sure the officials can speak in the general, Senator, in response to your questions, but, if your questions go to the detail of that original larger document, you can hardly expect them to speak to the specifics of it.
+
+\textbf{Senator PATRICK:} Sure.
+
+\textbf{Mr Johnson:} We were talking about optimising for the speed and the characteristics of the pump jet.
+The decision-making process for submarines is a little more complicated because there's an acoustic performance aspect as well as the pure efficiency of the blade-propulsor-pump jet arrangement.
+Without going into details, it's a decision that we make with Navy, but I'm confident, having done this a few times already, that we are going to find this pump-jet approach to be the optimum for our particular circumstances for this particular submarine.
+
+\textbf{Senator PATRICK:} The general summary of that larger document is that pump jets are most efficient at high speed.
+That's where you get the benefit because there's the cost of the propulsor and the weight of the propulsor.
+The shroud, for example, is great for reducing cavitation at high speed, which is typically what a nuclear powered submarine might do in transiting from one place to another.
+Indeed, to my knowledge, the only submarines that are fitted with pump jets, other than one Agosta 70 French submarine it was trialled on and one Russian submarine, are nuclear powered submarines.
+So the data in the public domain suggests that the advantages of the pump jet occur at high speed.
+The diagram that you're looking at shows efficiency at low speed, and that's where a submarine spends most of its time in an operational area.
+A question was asked by Senator Xenophon back in October about this efficiency problem.
+The answer came back that 'a submarine propulsor can provide higher propulsive efficiency than a propeller across the speed range of a submarine, including the patrol speed'.
+I don't want you to go to the technical details or anything that's classified, but are you saying about the diagram that I've presented, which is in the public domain, that there's now new technology that would be contrary to that?
+
+\textbf{Air Chief Marshal Binskin:} No.
+I think what Mr Johnson's saying is that all aspects of our concept of operations will be factored into the submarine design in order for it to be best able to conduct the operations we need.
+So I don't think we should be talking in any more detail than that here in an open forum.
+
+\textbf{Senator PATRICK:} Let me tell you my concern.
+
+\textbf{Air Chief Marshal Binskin:} I think I understand your concern.
+
+\textbf{Senator PATRICK:} Actually, it's a concern that relates to risk in some sense.
+When you put this really heavy pump jet at the back end of a submarine and then at some later stage you find that in actual fact the pump jet doesn't perform as you might have expected, noting what is in the public domain, to change that out, noting it's a big, heavy thing at the extremities of the submarine, will have significant impact on the design.
+So my question to you is: are you relying on data that is based on a real in-situ propulsor somewhere or are you relying on modelling data to make this decision?
+
+\textbf{Mr Johnson:} The decision to use a pump jet was carefully made.
+It includes a wider understanding of their use in other applications—leaving that open.
+We have already accessed a relatively large scale model, so its physical data, and we will have the advantage of slightly different but very similar arrangements when the French attack submarine goes to sea.
+The risk that you note is certainly a valid concern, but the information that we have is across a wide spectrum of available sources, some not available to you, and the fact is that we will have this very early full-size, full-scale performance data of a different propulser, but nevertheless one from the same team in a very similar hull shape.
+At this point, the Commonwealth should be very comfortable with the calculated decision that we're making.
+
+\textbf{Senator PATRICK:} Thank you very much, and I will come back to you briefly on that.
+Back in 2014, Senator Johnston was making inquiries as to the way in which the Commonwealth was spending \$214 million of de-risking money for their submarines.
+In amongst that money was a study on pump jets.
+I will refer you to the Hansard of 28 May 2012, where it makes it clear that DSTO were doing some studies on pump jets.
+Are you aware of those studies and the outcomes of those studies?
+
+\textbf{Dr Zelinsky:} I'm aware that we have been studying propulsers, propellers, various forms of propulsion for submarines, but I cannot, at this stage, give you the outcomes of that study from five years ago.
+I'd have to take that on notice.
+Subsequently, as we have gone through the competitive evaluation process, we have provided work and studies to that.
+I am sure our work has matured and our work does underpin the work of CASG.
+
+\textbf{Mr Gillis:} This is a 1963 document, which took a number of years to be declassified to this point.
+As I think most of the people here know, submarine technology and what we decide, and how and why we decide it, is, by its nature, classified.
+We want to be really careful that we don't go into a point where the money and great effort that's been undertaken by our Defence Science and Technology Group, the work we've done with our partners, is then put on any sort of public forum for debating.
+
+\textbf{Senator PATRICK:} I wasn't after the detail.
+
+\textbf{Mr Gillis:} The issue we have is that we have some of the best scientists in the world; we have some of the most experienced submarine program managers in the world working at it.
+We partner around the world to ensure we get the best possible outcome.
+As the head of procurement, I trust that those people are making good, sound decisions not based on very dated data; they are based on very up-to-date data.
+And as you saw, it's a \$200 million-plus investment by the government, and one of the aspects was propulsion.
+I think you should be comfortable that we have invested the money.
+The issue is I have no intention of putting that sort of information in the public domain.
+
+\textbf{Senator PATRICK:} Sure.
+All I'm after is just some indication from the chief scientist, noting there's a diagram on the table, as to whether or not he's comfortable.
+That is the question.
+I'm not asking him to go to the detail.
+
+\textbf{CHAIR:} So is your question essentially: is the chief scientist comfortable that the science backing up the decision of Defence is sound?
+
+\textbf{Senator PATRICK:} Yes.
+
+\textbf{Dr Zelinsky:} Yes.
+
+\textbf{CHAIR:} Thank you.
+
+\textbf{Senator PATRICK:} Okay.
+I'll go back to you, Mr Johnson.
+Noting that you're saying that this is based on something that hasn't been to sea yet—it will go to sea on the Barracuda; presumably that's the submarine you're referring to—there's still a risk that's sitting there.
+As the guardian of the taxpayer, I know where you sit.
+I have a responsibility from the taxpayers' perspective to ask: do you have a fallback position if the data that comes from the French trials, or other trials that you might do, presents you with data that's inconsistent with the operational profile that you intend for the submarine?
+
+\textbf{Mr Johnson:} Yes.
+We will understand the performance data early enough in the design that we can accommodate whatever is required.
+And then, just to reassure you, it's really going to be an adjustment to the design of the pump jet, not a decision to leave it.
+Without going into too much detail, the curve that you're referring to is about the efficiency of the propeller, not about its acoustic performance.
+As you know, there's nothing free in physics, so those are different-shaped curves.
+What we seek to do, as Chief of the Defence Force said, is to make the right balance of multiple dimensions of this now.
+
+\textbf{CHAIR:} Senator, did you want a private briefing to go into the detail?
+
+\textbf{Senator PATRICK:} No, I'm satisfied with that; I'm satisfied with the detail.
+
+\textbf{Mr Johnson:} One other reassurance: we will run scaled model testing, and that will start relatively soon.
+It's a standard approach.
+It's been used by the French navy and the US navy for a long time.
+We're going to march right down the path, and we will do it with the chief scientist.
+
+\newpage
+\subsection{Answer to Question on Notice}\label{QON148}
+
+\begin{figure}[h]
+\includegraphics[width=0.8\textwidth, center]{QON148.png}
+\end{figure}
+
+\newpage
+\subsection{Answer to Question on Notice}\label{QON150}
+
+\begin{figure}[h]
+\includegraphics[width=0.7\textwidth, center]{QON150.png}
+\end{figure}
+
+\newpage
+\subsection{Answer to Question on Notice}\label{QON151}
+\begin{figure}[h]
+\includegraphics[width=0.7\textwidth, center]{QON151.png}
+\end{figure}
+
+\begin{figure}[h]
+\includegraphics[width=0.7\textwidth, center]{QON151p2.png}
+\end{figure}
+
+\newpage
+\subsection{Answer to Question on Notice}\label{QON1}
+\begin{figure}[h]
+\includegraphics[width=0.7\textwidth, center]{QON1.png}
+\end{figure}
+
+\newpage
+\subsection{Answer to Question on Notice}\label{QON2}
+\begin{figure}[h]
+\includegraphics[width=0.7\textwidth, center]{QON2.png}
+\end{figure}
+
+\newpage
+\subsection{Answer to Question on Notice}\label{QON3}
+\begin{figure}[h]
+\includegraphics[width=0.7\textwidth, center]{QON3.png}
+\end{figure}
+
 \end{appendices}

### 2018-09-05 Second.Rnw
more more

@@
 
+\begin{tcolorbox}[width=0.9\textwidth, center, colback=blue!5!white,colframe=blue!75!black, title= \textbf{10 Feb 2016 - Sean Costello, CEO DCNS Australia}]
+\textit{In a blog post\parencite{costello2016} for the ASPI Strategist prior to the announcement in April, Sean Costello describes the significance of the pump jet technology to their bid.}
+\\
+\\
+``As a complete submarine power, we understand conventional propulsion, which is why we also understand that propulsion is but one part of the submarine puzzle.
+In designing to the Australian requirement, it should come as no surprise then that the conventionally powered Shortfin Barracuda Block 1A is only 5\% lighter than its nuclear cousin."
+\\
+\\
+``There’s no better example of what I’m talking about than the offer from France to transfer to Australia sovereign control and use of pump-jet propulsion technology for the Shortfin Barracuda—technology resident only in France, the UK and the USA.
+Technology born from the French SSBN program a generation ago.
+The stealth and hydrodynamic performances of pump jet propulsion are of course classified and in Australia known only to DCNS and the Australian Government."
+
+\end{tcolorbox}
+
+\begin{tcolorbox}[width=0.9\textwidth, center, colback=blue!5!white,colframe=blue!75!black, title= \textbf{8 Apr 2016 - Sean Costello, CEO DCNS Australia, and Gerard Autret, Chief Naval Architect of the Shortfin Barracuda}]
+\textit{In a blog post\parencite{autret2016} for the ASPI Strategist prior to the announcement in April, Sean Costello describes a number of the major features and design approaches for the Shortfin Barracuda, including relating the choice of the pumpjet closely to an accoustic requirement to close with another submarine at higher speed.}
+\\
+\\
+``The challenge for any attack submarine is to maintain a nearly silent acoustic signature at a speed necessary to manoeuvre within weapons range of the target.
+The nuclear Barracuda is designed to reduce radiated noise when operating at a speed sufficient in order to close a threat submarine undetected.
+Of particular importance is the pump-jet propulsor, which combines a rotor and stator within a duct to significantly reduce the level of radiated noise through the effects of wake harmonisation and avoidance of cavitation."
+\end{tcolorbox}
+
 \begin{tcolorbox}[width=0.9\textwidth, center, colback=blue!5!white,colframe=blue!75!black, title= \textbf{May 2016 onwards - DCNS/Naval Group Website for Shortfin Barracuda}]
@@
 
+\end{tcolorbox}
+
+\begin{tcolorbox}[width=0.9\textwidth, center, colback=blue!5!white,colframe=blue!75!black, title= \textbf{25 Oct 2017 - Hevre Guillou, CEO Naval Group}]
+\textit{In an interview that was published in a post\parencite{guillou2017} on the ASPI Strategist the global CEO of Naval group responds to a question from Brendan Nicholson asking whether the Australian submarine will be fitted with a pump jet, following Jean-Michel Billig's comments.}
+\\
+\\
+``Naval Group is the manufacturer of both conventional propeller-driven submarines and larger nuclear submarines with the propulsor.
+The propulsor’s inclusion means the future submarine will move more quietly than submarines with obsolete propeller technology.
+In a contest between two otherwise identical submarines, the one with the pump jet always has the tactical advantage.
+The propulsor was an integral part of our bid.
+The Department of Defence says it intends to use the pump jet propulsion system."
+\end{tcolorbox}
+
+\begin{tcolorbox}[width=0.9\textwidth, center, colback=black!5!white,colframe=black!75!black, title= \textbf{31 Oct 2017 - Greg Sammut, head of SEA 1000}]
+\textit{In an interview that was published in a post\parencite{sammut2017} on the ASPI Strategist the leader of the future submarine project for the RAN responds to a question from Brendan Nicholson asking whether there is any doubt about using the pump-jet.}
+\\
+\\
+``We are using pump-jet propulsion that was offered by France.
+Some claims about the pump jet are misleading.
+It’s not purely for use in nuclear submarines.
+It can be efficient across the entire speed range, taking account of the submarine’s size, the speed required and stealth.
+It needs to be tuned to the hull of the submarine.
+The hull needs to be tuned to the pump jet."
 \end{tcolorbox}
@@
 \textit{Senator Patrick inquires about the nature, cost, and conclusions of earlier research into pumpjets undertaken as part of \$214 million worth of ``de-risking" studies already undertaken, as well as the schedule and cost of subsequent research and development of the pumpjet.
-Prior research consisted of a single 99 page report obtained via literature review and discussion with experts at a cost of $68,825.
-Ongoing research included $1.1million spent to date to develop a generic submarine and pumpjet design, compuational and analytic models, and validation of these models through experimental studies.
+Prior research consisted of a single 99 page report obtained via literature review and discussion with experts at a cost of \$68,825.
+Ongoing research included \$1.1million spent to date to develop a generic submarine and pumpjet design, compuational and analytic models, and validation of these models through experimental studies.
 The development and test program is not scheduled to commence until the last quarter of 2018.
@@
 Testing will be largely completed by CDR."
+
+\end{tcolorbox}
+
+\begin{tcolorbox}[width=0.9\textwidth, center, colback=black!5!white,colframe=black!75!black, title= \textbf{3 April 2018 - Department of Defence, in response to Question on Notice from Senator Chris Ketter in December 2018}]
+\textit{Senator Ketter asks Question 1, referencing specifically Greg Sammut in \parencite{sammut2017} and my paper \parencite{morrison2018} asks specifically whether Defence maintians that the pumpjet can be efficient across the entire speed range.
+Full question can be found in \ref{QON1}.}
+\\
+\\
+Question: Does defence still maintain that a pump-jet can be efficient across the entire speed range?
+\\
+\\
+Answer: ``The efficiency of pump-jets designed for submarine use cannot be determined through comparisons with water-jets and ducted propellers, which are different in their operation from pump-jets.
+Moreover, the global efficiency of submarine pump-jets cannot be derived from ‘open water’ tests of pump-jets, where the pump-jet is assessed independently of the hull of the submarine.
+A correctly designed pump-jet, suited to the aft hull shape of the submarine, achieves better optimisation of efficiency and stealth than any propeller, even at low speeds."
+
+\end{tcolorbox}
+
+\begin{tcolorbox}[width=0.9\textwidth, center, colback=black!5!white,colframe=black!75!black, title= \textbf{3 April 2018 - Department of Defence, in response to Question on Notice from Senator Chris Ketter in December 2018}]
+\textit{Senator Ketter asks Question 2, inquiring as to how Defence assesses Naval Group's opinion regarding the pumpjet's suitability, following Jean-Michel Billig's comments.
+Full questions and responses  can be found in \ref{QON2}.}
+\\
+\\
+Question: Is Defence still confident that Naval Group still hold the opinion that the pump-jet is wellsuited to diesel-electric submarines?
+\\
+\\
+Answer: ``Defence intends to use the pump jet propulsion system for the Future Submarine that wasoffered by Naval Group as part of their Competitive Evaluation Proposal and remains onoffer to Australia."
+
+\end{tcolorbox}
+
+\begin{tcolorbox}[width=0.9\textwidth, center, colback=black!5!white,colframe=black!75!black, title= \textbf{3 April 2018 - Department of Defence, in response to Question on Notice from Senator Chris Ketter in December 2018}]
+\textit{Senator Ketter asks Question 3, which focuses specifically on whether Defence believes the pumpjet is acoustically superior across the entire speed range, or just at higher speeds, referencing my paper \parencite{morrison2018}, Herve Guillou \parencite{guillou2017} and Costello and Autret \parencite{autret2016}.
+Full question and response can be found in \ref{QON3}.}
+\\
+\\
+Question: Is Defence of the belief that the pump-jet is quieter than propellers at all operating speeds, or is it the case, as suggested by Mr Morrison, that at some low speeds propellers are acoustically superior, and only at higher speeds is the pump-jet acoustically superior?
+\\
+\\
+Answer: ``See response to Senate Economics Reference Committee Question on Notice, Q1 – Pump Jet Efficiency Across the Entire Speed Range."
+
+\end{tcolorbox}
+
+\begin{tcolorbox}[width=0.9\textwidth, center, colback=blue!5!white,colframe=blue!75!black, title= \textbf{17 May 2018 - Brent Clark, Acting CEO Naval Group Australia}]
+\textit{In an interview that was published in a post\parencite{guillou2017} on the ASPI Strategist the Autralian CEO of Naval group explains to Brendan Nicholson why Naval Group is recommending the pumpjet for the Shortfin Barracuda, and answers some critics.}
+\\
+\\
+The company building the Navy’s 12 new submarines says they'll be the world’s first conventional boats big enough to effectively use the company's pump jet propulsion system.
+French Naval Group Australia says that its propulsor will prove much quieter than a propeller on the 4,700‑tonne submarine it has dubbed the 'Shortfin Barracuda'.
+\\
+\\
+Naval Group Australia's interim CEO, Brent Clark, tells The Strategist that the decision about the best propulsion system for the submarine will be made by the Defence Department, but that he hadn’t had any indication that it would chose anything but a propulsor.
+'We've had every indication that they like what was offered.'
+\\
+\\
+The propulsor that Naval Group is recommending for its Shortfin Barracuda consists of a propeller within a duct to reduce noise and the water disturbance known as cavitation, which can give away a boat’s position to hunters.
+Clark says he’s read many negative comments about propulsors, but some critics used very old studies to try and justify their argument.
+\\
+\\
+"A lot of people don’t fully comprehend, I think, that yes, a propulsor on a 1,500-tonne coastal submarine would make no sense.
+But we’re not building a 1,500-tonne coastal submarine.
+We’re building quite a large submarine.
+And this is the first time that a conventional submarine's been large enough to take a propulsor."
 
@@
 
+\newpage
+\subsection{Hansard Extract, Senate Economic References Committee Public Hearing - 7th June 2018 \cite{SEN2018morrison}}\label{SENmorrison}
+
+\textbf{Mr Morrison:} I would firstly like to thank and acknowledge two supporters: Mr Gary Johnston from Submarines for Australia, without whose support I wouldn't have been able to put together the substantial report that you have today; and my partner in Trendlock, the analytics consulting company that we used to deliver the report, whose help was invaluable.
+
+In my opening remarks I would like to briefly restate the core conclusions only, and then I'd like to contrast and compare those with some other public statements that have been made on the record by Naval Group officials and officials of the Australian government.
+The key conclusions that I arrived at were that pump jets have a far lower efficiency than propellers at a low speed of travel in contrast to high speeds, where jets tend to become more efficient.
+The crossover does not lie near or anywhere under the probable transit speed of eight to 10 knots of a diesel-electric submarine.
+One world-renowned expert I discussed the matter with said it was impossible that a pump jet could be designed to be more efficient than a well-designed propeller under 18 knots.
+
+One of the main reasons for the precipitous decline in propulsive efficiency at low speeds is directly linked to the mechanisms that enable the acoustic advantage to be achieved at high speeds.
+The duct has the effect of slowing the water down where the blades do their work on it.
+This has the effect of raising the pressure, which is what reduces the phenomenon of cavitation.
+Consequently, redesigning or modifying a pump jet to improve propulsive efficiency at low speeds cannot be done without necessitating some trade-off in terms of acoustic advantage achieved through the elimination of cavitation at higher speed.
+Very importantly, I also concluded that the extent of the efficiency declines at low speeds over the operational profile of a diesel-electric submarine will be substantial.
+An improvement in the order of 50 per cent or more in dive range or endurance could be achieved through switching back to conventional propellers.
+This could amount to some days dived on station or hundreds of nautical miles.
+In my report, I made the conclusive comparison that, between two otherwise identical submarines, the one with the pump jet will always have a lower dived endurance, a lower dived range, a worse indiscretion ratio, a lower overall endurance and a lower overall range than the one with the propeller.
+This statement, so far as I know, has yet to be challenged, and I'd be very surprised if anyone with technical competence or professional integrity sought to challenge it.
+
+In contrast, I would like to quickly turn to a couple of other statements that have been made on the public record.
+The first is from the official Naval Group website, which says about the Shortfin Barracuda:
+
+``Pump jet propulsion means the Shortfin Barracuda can move more quietly than submarines with obsolete propeller technology."
+
+This is a very interesting statement because at face values it reads as if to say that pump jets are now so quiet that propellers have become obsolete.
+I know that that interpretation has been made, and I think it has been made before this committee or another Senate committee.
+Part of my further research that you have before you today indicated that the phenomenon that leads to increased turbulence at very low speeds is also likely to produce much more noise than a propeller would at very low speeds.
+This is a phenomenon called flow separation.
+It actually seems to me very surprising that the claim would be made that at very low speeds a pump jet would indeed be quieter than a well-designed propeller.
+Read more literally, it actually sounds like a different statement might be made, for the pump jet is only quieter when compared with an obsolete propeller.
+This could actually be the statement that was intended to be made, in which case it is perfectly consistent with the research that I have done and put before you today.
+
+It might be very strange to try to figure out—and it was to me—why it would be useful to have a submarine that was only quieter at relatively high speeds and actually conferred disadvantages at all other speeds, including acoustically, with a significant trade-off in terms of overall range.
+It seemed like it would be quite a strange tactical circumstance that would require such a thing.
+In particular, and this is something else that needs emphasising, high speed for a submarine is actually still low speed for basically any other surface ship.
+Even very slow cargo ships moving at 18 or 20 knots move much faster than a submarine would be able to sustain for any significant period of time.
+Warships, of course, can easily go much, much faster, so it's difficult to see what that particular circumstance would be.
+I would like to emphasise that the committee should pay particular attention to what that circumstance might be.
+
+Again, on the Naval Group public website, it is stated that:
+
+``In a confrontation between two otherwise identical submarines, the one with pump jet propulsion always has the tactical advantage."
+
+This is a very interesting statement as well.
+It certainly does not hold against a nuclear submarine.
+It wouldn't hold against any kind of surface ship.
+It doesn't hold in a strategic sense, either, because the necessity to snort more frequently on the way in and the lower dived endurance on station would mean that you'd be more exposed more quickly after any encounter.
+However, tactically, once the encounter has already happened and a target has been acquired, it is plausible that this statement is still true.
+It is quite clear that, since some considerable time ago, DCNS has been very, very well aware of what the advantages and disadvantages are of pump jet propulsion.
+There are also a number of other statements on the public record that also concur and fit extremely well with the research that I've done.
+In particular, statements by Jean-Michel Billig, who said back in October in 2017 that they may switch back to propellers and even insert AIP for higher underwater endurance, are perfectly consistent with the research that I've done.
+As well, more recently, the statements by Brent Clark on the ASPI Strategist blog published only a couple of weeks ago suggesting that a pump jet would not be suitable for a small submarine—but this submarine, being big enough, is perfectly consistent, as I pointed out in my public reply, with the research that I have already conducted.
+It seems to me that there is no contradiction clearly put by any representative of Naval Group to the research that I have presented on pump jets.
+
+I would like to compare that briefly to a couple of statements that have been made by representatives of the Commonwealth.
+Perhaps the first that is worth turning to is that by Rear Admiral Greg Sammut:
+
+``It can be efficient across the entire speed range, taking account of the submarine’s size, the speed required and stealth.
+It needs to be tuned to the hull of the submarine.
+The hull needs to be tuned to the pump jet."
+
+This was a statement made in October 2017, also published in an ASPI Strategist blog post his is a strange statement, because the first part—that it can be efficient across the entire speed range—taken on its own does not seem to be possible to be true.
+It seems to be completely false.
+Taking into account the submarine's size, the speed requirement and stealth, it seems to suggest that only if the admiral meant that he didn't actually want it to be efficient, and therefore could treat it as efficient—that seems to be the only way that it could be reconciled with reality and the physics that I've discovered.
+
+The subsequent statements are even more peculiar - that it needs to be tuned to the hull of the submarine.
+As far as I can tell, the hull of the submarine proposed is so close in length and diameter to the one that's actually already being built by the French for the nuclear version, it seems very, very, strange indeed to make a tuning when such a small adjustment would be required.
+In contrast, if what was actually meant was that the pump jet should be tuned to the speed, the degree of the change in the speed of the operational profile would necessitate such a drastic overhaul to the design of the propulsion system that the idea of tuning seems quite thoroughly out of place.
+This is, to put it simply, as if someone bought an efficient hybrid car like a Prius and then decided to fit it with tank tracks from a dozer or a battle tank, and then argued they would tune the tank tracks back to be as efficient as the Prius was in the first place.
+It goes radically against what the technology was intended for in the first place.
+The idea of tuning is completely odd to any technical reading of the story.
+
+I'd like to look at one further public statement that was made.
+That is from Stephen Johnson.
+This was made in relation to an answer that was given to this committee some time ago.
+When challenged very specifically about efficiency by Senator Patrick, he made this statement:
+
+``No matter what kind of propeller or propulsion pump jet you design, it has to be optimised for a particular speed.
+There are classes of surface ships that look to optimise cross ocean transit.
+Physics usually keeps you around 17 to 18 knots, and then there is the cost of maintenance and the cost of repair and the fact that there's no particular stealth requirement.
+Typically, you will find that those ships will use an open-tip propeller.
+However, other ships, cruise ships in particular, use an electric shrouded pod, which is a variation of a pump-jet technology."
+
+This also struck me as quite strange, because in my research I went to some lengths to make sure that I understood all the different variants of different types of propulsors, including propellers, ducted propellers, shrouds of various sorts and the way the different names are used.
+So far as I can tell, despite the fact that cruise ships typically travel substantially faster than would be possible for any diesel and electric submarine, this is particularly strange because I can't find any evidence of any propulsor used on a cruise ship that does indeed include a shroud.
+It seems to me that some confusion might have occurred about what a propulsor actually is in this case.
+To clarify, my best understanding is that 'propulsor' is a perfectly general term that can be applied to basically any basic pump-jet propulsor or any other device to push something through the water.
+It is sometimes used just to differentiate novel, different or contrasting systems.
+
+I wonder whether here Mr Johnson has actually been confused with the Azipods that are installed on large cruise ships.
+They do use an electric motor, that's the only similarity I can find with the pump jet, and are indeed open to propellers and are quite similar to the propulsion systems used on the LHD.
+That is the best explanation that I can find so far for these comments.
+I certainly can't find any evidence of any shrouds being used on cruise-ship propulsion.
+Moreover, I can't find any evidence of pump jets of this sort being used on any ship, for which economic travel is a main priority, operating below the speeds of about 25 or 30 knots.
+So it's very interesting to wonder what exactly Mr Johnson was referring to in this case.
+
+I will rest my opening statement there, but there are other statements as well and other matters to do with the way the procurement process and design process looks to be unfolding that seem concerning from the readings of the physics that I have done so far.
+I'm happy to have some further questions.
+
+CHAIR: Thank you. I just want to place on record that since the commencement of the public hearing this morning Senator Reynolds has been joining us over the telephone line. She is now here with us.
+
+Senator PATRICK: Mr Morrison, firstly I will point out that Mr Johnson is in the back of the room, so he'll have an opportunity later today to rebut anything that you've said.
+
+Mr Morrison: Excellent.
+
+CHAIR: We'll certainly give him that opportunity. In terms of your credentials—you haven't explained. Have you got engineering or scientific credentials?
+
+Mr Morrison: I have a physics background.
+I studied theoretical physics at the ANU and got first-class honours in theoretical particle physics.
+I commenced a PhD at the University of Melbourne, which I didn't finish.
+I went to start a small defence technology company, designing small boats and torpedos for maritime warfare—a swarming concept.
+We used pump jets.
+We selected pump jets for the torpedo and water jets for the boat, so I became familiar with the particular systems and the trade-offs involved in those during that time.
+I had considerable discussions with manufacturers, naval architects et cetera during that process.
+I would like to point out that we selected pump jets for both those applications.
+I'm a fan of pump jets per se.
+They're very good.
+They're a go-to in many circumstances.
+It is a very peculiar, quite odd, situation to employ them on a conventional submarine.
+
+Senator PATRICK: I just want to make sure that I've understood you correctly.
+You are saying that physics suggests that the pump jet is not efficient at the low speeds that a conventional submarine typically operates at in an operational area?
+
+Mr Morrison: That's right.
+
+Senator PATRICK: Of course, they use pump jets on nuclear powered submarines.
+Your paper says that's not a problem because they have no problems in respect of generating power, and you actually even say it's helpful.
+
+Mr Morrison: Yes. Would you like me to clarify that point?
+
+Senator PATRICK: Yes.
+
+Mr Morrison: Nuclear reactors have a tendency to have different instabilities build up when they try to oscillate their power outputs radically.
+There are different decaying isotopes, xenon in particular, that make the reactor less stable after it has dramatically reduced its pressure.
+They prefer to operate in a steady state.
+They certainly wouldn't like to go down to very low power and then very high power and then very low power.
+That makes them somewhat more difficult to manage.
+So, to the extent of keeping a floor, a minimum level of energy required, they wouldn't need to pull the energy down below.
+That would be useful.
+They would reduce the extent that they would have to run other pumps or other systems to be able to take the excess energy out of the reactor.
+
+Senator PATRICK: To your knowledge, are there any other conventional submarines—clearly, there are nuclear submarines that use a propulsor, or pump jets, but are you aware of any other conventional submarines that use them?
+
+Mr Morrison: There are none in service at all.
+I'm aware of a prototype or demonstrator that was built once on a Russian submarine, a Kilo class submarine.
+I've seen a photo of that, but as far as I can tell it was never put into much active service.
+
+Senator PATRICK: The bottom line for you is that it's problematic for a conventional submarine for all the reasons that you mentioned in terms of 'efficiency is king'.
+It affects indiscretion ratios and it exposes a submarine, because it will have to snort more often.
+
+Mr Morrison: That's right.
+It seems like a genuinely strange choice.
+Once you take out the superpowers that a nuclear reactor confers on a submarine, leaving the pump jet is a massive hindrance to everything else about a diesel-electric submarine.
+The impact on the indiscretion ratio range, dive and endurance seems quite substantial.
+
+Senator PATRICK: I tabled a document on 15 December, and it was pointed out that it was a 1963 declassified document from the US Navy.
+There were some criticisms levelled at me because of the age of the document.
+How reliable is the physics?
+Are you satisfied that the physics hasn't changed substantially, such that the information in the public domain could be wrong?
+
+Mr Morrison: I'm absolutely satisfied.
+I have done quite a bit of research.
+As I mention in my blog post, I have referred to many more modern papers as well.
+The field of hydrodynamic propulsion does evolve; it does change.
+However, it tends to build on and confirm some of the most foundational and fundamental principles on which the conclusions that I am sure your paper and a few others that I have here with me, which are from, as you say, the sixties and seventies, are built.
+It is particularly important when you are reading more modern literature that you find typically authors who wrote during those times who still consistently turn up in the references; they are still referred to.
+They are not dismissed in any regard.
+They are referred to as, I suppose, 'the greats' who wrote the books on these topics.
+They produced the full discussions of a comprehensive method for designing a pump jet.
+Even at that time, it was clear that pump jets never were intended - and I can give you some quotes, if you like, from certain old papers - for those very low speeds at all.
+In fact, they were really only advantageous at quite high speeds where cavitation was a severe problem.
+
+Senator PATRICK: What you're saying is akin to saying that we shouldn't reject Einstein's theory of relativity just because it's old?
+
+Mr Morrison: That's absolutely right.
+It's foundational, in fact.
+I can see nothing—and I have discussed this with other current experts—that has overturned the books, so to speak, on these matters.
+So I think some of those old foundational papers—and, indeed, I think I quoted one where it is surmised in a sentence—reached the conclusion: 'It's not good at low speeds'—almost as simply as that.
+They don't go back and re-labour and reprove what has been so long and well established in the annals of physics.
+
+Senator PATRICK: You say that, acoustically, the pump jet is actually noisier at low speeds?
+
+Mr Morrison: I think that's highly probable.
+I won't make the absolutely categorical claim, as I've done with the efficiency, but I think that is extremely highly probably because, as you can see from my efficiency research, a pump-jet would be producing something like twice as much energy release to turbulence at those very low speeds as what a propeller would be.
+It's hard to see how something discharging twice as much energy into the water can still remain quieter.
+Indeed, in this case, the sorts of turbulence are very large scale vortices.
+This phenomenon called flow separation and the periodic vortex shedding that might occur into the rotors, that's a relatively low-frequency sound; it's not the sort of high-frequency sound that might be associated with the very, very early onset of cavitation.
+For very high-frequency sounds, the shielding from the duct is likely to be effective—still only in the directions that are perpendicular to the access of the shaft, but will be somewhat effective.
+At the lower frequencies, I think there is good reason to think the shielding will not be as effective.
+I won't make the absolutely categorical claim that it has to be much noisier, but it seems difficult to believe how it is superior.
+I've never heard anyone explicitly make that claim, that it is actually acoustically superior at low speeds to a state-of-the-art propeller.
+I would be very surprised if that was the case.
+
+Senator PATRICK: Chair, I would like to table a paper that was produced by DCNS in 2012 at a UDT conference? Will you permit me to do that?
+
+CHAIR: Is there any objection to that document being tabled? No? That's okay.
+
+Senator PATRICK: I would like the witness to have a look at it.
+I think the company was called DCN then, and it's their engineering area.
+Could you look on page 8 of this document?
+I've highlighted the relevant text.
+This is about the French Scorpene submarine and the permanent magnet motor.
+It says: 'The main advantages in terms of signatures of a permanent magnet motor is the ability to maintain a very low turns per knot, even with a high output, thanks to the high torques that are available for that particular motor.
+This low TPK reduces significantly the propeller cavitation apparition risks.
+For instance, within conventional operating depths, so 150 metres and higher'—which I presume is French for deeper—'no cavitation is expected whatever speed may be.' That tends to suggest to me that the French designed a propeller back in 2002 that does not cavitate at any speed provided you are below 150 metres.
+Is that consistent with your view on propellers versus pump-jets?
+
+Mr Morrison: Yes, that's absolutely consistent.
+I think there's a diagram that I've drawn in my report, in the acoustics section.
+It's absolutely the case that raising the static pressure, which is what happens when you go very deep—and it's also what the shroud of the duct attempts to achieve to the water flow in a pump-jet—is what reduces the onset of cavitation, which is the principal cause of all the noise.
+Going deep simply does reduce or eliminate that.
+I think that, in this case, it is quite probable that at more modest speeds you'd be able to go quite shallow, potentially up to almost all of the routinely-used speeds of a diesel-electric submarine.
+The reason why you would want to be able to go in very shallow water at very high speeds for a very short period of time—it seems like a very narrow and a very obscure tactical advantage indeed, and that seems to be confirmed by the limitation of the superiority to a contest against another diesel-electric submarine only.
+It's hard to imagine how, against any other target, it could be useful.
+
+Senator PATRICK: You've got an engineering background.
+You said you've worked on some projects.
+What do you see is the risk, if you were to categorise the risk of Defence achieving the publicly-stated performance expectations of the pump-jet—that is, efficiency across the entire speed range and acoustically quiet?
+
+Mr Morrison: If they mean 'efficiency' to be anything close to a propeller across the entire speed range, I think achieving efficiency is impossible.
+I don't think they'll achieve that at all.
+I'd like to point out as well that there are a couple of other comments that seem particularly at odds regarding the engineering process and the design process that's happening here.
+It was stated by Rear Admiral Greg Sammut in the same post I referred to earlier.
+He said:
+``We did not specify that the boat should have or not have air-independent propulsion or lithium-ion batteries, rather that it met our requirements.
+We'll continue to look at all applicable technologies as we go through the design process."
+
+I think Brent Clark, in his very recent post on the ASPI Strategist, said the decision about the best propulsion for the submarine will be made by the Defence department.
+He is indicating quite clearly the final decision is yet to be made.
+And yet, still as of October, Admiral Sammut was saying:
+
+``Work on establishing the dimensions of the submarine is complete. That pre-sizing—the initial length, diameter, displacement and other key parameters …"
+
+It seems very clear that the size of the hull, the total displacement, the volume of the hull is now fixed. It seems very clear as well that the requirements are fixed, one would think.
+
+However, that sits greatly at odds with the idea that the technology surrounding propulsion, particularly pump-jets, is not fixed.
+The reason for that is quite simple.
+As I've described through my research, the difference in efficiency is sufficiently large that to meet the same fixed requirement would differ the amount of fuel and batteries required in the order of hundreds of tonnes—hundreds of tonnes—for this sized submarine.
+It seems very difficult to imagine how, for a submarine, which basically has to be perfectly neutrally buoyant in all its operations, we could add or subtract hundreds of tonnes in the same fixed displacement without having a catastrophic engineering failure.
+I'm talking about Super Seasprite style: build the air frame, set the requirements.
+They simply do not go together whichever way you put them.
+The risk that you simply have a submarine that doesn't float or doesn't sink—I'm not sure which is worse—is abundantly clear from these.
+The only alternative that I can see is that they actually haven't fixed the requirements.
+They've designed the hull and they'll have whatever performance turns out of whatever technologies they put into it in whatever volumes seem suitable.
+
+Senator PATRICK: I think you've partly answered this, but what do you see the risk is for the submarine project of proceeding with the pump-jet, noting you say it's impossible to achieve everything that they want to achieve and particularly noting the weight, the size and the fact that it's positioned on the arced extremity of the submarine?
+
+Mr Morrison: I think it adds exactly the point I've made: This is a very, very major change.
+I didn't even mention that; it's in addition to the additional total volume of fuel and batteries required.
+Locating something that large and heavy, the shroud will be many tonnes, tens of tonnes, at the rear end.
+Quite frankly, it looks incomprehensible.
+I don't know how you could consider saying that you've just won a concept about a submarine if you haven't defined the propulsion system or the type of battery or whether it has got AIP.
+There is simply nothing more important in the engineering of the whole thing than fixing those sorts of things in place.
+It seems extremely odd to me that they are saying they're still considering these things and they've fixed the size of the submarine and they're making progress on the design, because all these things will have such a radical impact on the amount of weight embarked and the balance of the whole thing.
+It seems like an incomprehensible sequence of events to have happening at this stage.
+
+Senator PATRICK: Thank you. That's all I have.
+
+Senator KIM CARR: Mr Morrison, it's a very important point you make.
+You're going to the very fundamental design questions that are at the heart of a major acquisition for this country.
+I'm just wondering what are you seeking from this committee in terms of the propositions you're putting before us?
+
+Mr Morrison: Look, I'm just offering my evidence here.
+To me it seems that the pump-jet is strange.
+I find it difficult to believe that this is the primary design feature that we have knowingly and deliberately selected.
+Whatever has led to that, I just suggest this committee needs to get to the bottom of why it was selected because it doesn't seem consistent with building a world-class submarine and it seems like we could be heading on a track that could lead to a real procurement mess.
+I would just direct your focus of figuring out how we got here, because if the forces that put such a strange engineering choice into the submarine are still at work then there could be worse to come in the future and it doesn't bode well for a successful procurement.
+
+Senator KIM CARR: Given where we are in the procurement process, I think it is incumbent upon us to determine what we do with your evidence.
+What are you actually recommending we do?
+For instance, if I might be so bold: I'm not certain this committee would have the technical expertise to evaluate what you're saying, and the Department of Defence clearly challenges what you're saying, so where do we go to evaluate what you're proposing?
+
+Mr Morrison: So far as I can tell, this is not controversial. Whilst it seems to be very technical physics, the paper that you -
+
+Senator KIM CARR: I'm not certain the Department of Defence would agree with your proposition that it's not controversial.
+
+Mr Morrison: I'd be very interested to see that.
+I've had discussions with other concerned physicists and people in the naval community about this, and I've heard of discussions had with the Department of Defence and some of the basis for their thinking that pump jets are as efficient, plausibly, over the whole speed range.
+I'm quite confident that further investigation needs to be done of that.
+I've been pointed to a particular chart, and I think they've misread it; I think they haven't been able to interpret it.
+So, I think you should dig hard.
+I'm quite confident of the research.
+I think that my paper to you is a carefully and meticulously researched statement of what is relatively obvious.
+I'm not aware of anyone in the marine engineering community seriously proposing that you would ever select a pump jet for something designed to operate anywhere near, let alone below, 10 knots.
+It is quite a strange proposition, so I think you should investigate it very, very seriously indeed.
+
+Senator KIM CARR: You're seeking from this committee further investigation of your claims. Is that the essence of your recommendation to us?
+
+Mr Morrison: Yes, and I would also strongly suggest that you carefully investigate the basis for the anti-submarine-warfare superiority claim.
+To me, that seems the last plausible thread of logic on which you could justify a submarine choice with a pump jet, and that's to say last plausible and not categorically and obviously untrue.
+There are gross doubts that I have about that as well: how you could ever acquire such a small target passively; how you would measure the range that you were at and know that you would be able to close in the short distance that you're able to; whether you'd be able to keep track of your target when you accelerate to those high speeds—there are a whole bunch of reasons which look almost as dubious as the proposition about pump jets being efficient.
+
+I'm actually seeking to do some more research on that topic exactly, because it seems perfectly clear that that is the only situation in which there's a plausible case that the pump jet might be of any value, and I think your focus should very much dwell there.
+I'd like to get some more resources to conduct a thorough study of that, and I think that it should be challenged very directly with Defence—whether we're buying a submarine for all our strategic needs or whether we have just selected one particular scenario to justify a choice that's based on some other justification.
+I can't see a logical justification for having made this particular design choice other than that.
+That is the last plausible reason there might be, and it's only barely plausible.
+I think it should be investigated very thoroughly.
+
+Senator KIM CARR: You're saying you're undertaking further research. What's the state of that research?
+
+Mr Morrison: It's very early.
+I'm basically seeking further financial support to be able to do the same sort of sustained and thorough investigation as in the paper that you've got before you today.
+I haven't been able to get that yet, but I think it would be worthwhile.
+I think it would be in the interests of the Commonwealth to determine whether there is serious merit in that statement.
+I wouldn't make the sort of categorical and strong claims I've made about efficiency about the anti-submarine scenario without doing that sort of research.
+I like to take the academia, the literature and the physics very seriously.
+But I think that sort of thing should be done quite urgently because I think there have been some very strange, potentially misleading—perhaps deliberately misleading—statements made about the superiority of the pump jet.
+
+Senator KIM CARR: This committee is due to report on 27 June. I'm not certain we'd have a lot more time to -
+
+Senator PATRICK: Senator Carr, if I may, just to kick this along a bit—it's a very good thought you're having. Do you have a security clearance?
+
+Mr Morrison: No.
+
+Senator PATRICK: That would preclude you from getting in.
+My thought was you could perhaps go inside, work with Defence and then pop out and say either, 'I'm not convinced,' or, 'I am convinced,' keeping it at an unclassified level.
+
+Mr Morrison: The rules of physics are not classified or classifiable, and I think that's -
+
+Senator KIM CARR: I'm interested in pursuing what you're saying.
+As a citizen, I think you're entitled to put views which go to the heart of this procurement.
+I'm not saying that you're right.
+Please don't misinterpret me—I'm not qualified to say that you're correct, but I am interested in what you're saying because of the seriousness of the propositions you're advancing.
+I am trying to come to grips with what we do with what you're saying.
+We've only got limited time.
+I'm clearly of the view we're about to face a general election, and it may well be that there will be a different set of circumstances after that election.
+In terms of this immediate report that's required to go to the parliament, we need to find a way to evaluate what you're saying, given that this committee, in my view, doesn't internally have the capacity to technically evaluate your claims.
+
+Mr Morrison: Might I suggest, as a way to trying to seek validation, you could see whether anyone else would contradict or oppose some of the simple statements that I've made.
+
+Senator KIM CARR: We'll ask the Department of Defence but it will not surprise me if they tell me there's a technical reason why you can't be right.
+
+Mr Morrison: I'd be amazed to see that.
+Having been challenged, I would be amazed if someone wants to restate—and I've made some careful statements in my report.
+You can ask them whether I'm wrong and see what they come back with.
+I'd need to be quite badly wrong on most of this.
+
+Senator KIM CARR: Would you accept it on technical grounds, if they can demonstrate that you're wrong in theory?
+
+Mr Morrison: The whole thing has got to be wrong.
+It's got to be utterly wrong.
+I think this is worthwhile.
+If I was to go to a particular page of my report—on page 32, if you like, there's a chart from a paper that was done not too many years ago by BMT Defence Services—you can see the graph there—that shows the efficiency curve of a fully submerged water jet and a propeller going down to low speeds.
+Just look at the shape of the curve and the gap between those.
+If the department comes back and says I'm wrong on some technicality, some small thing, which means the lower curve is not exactly right and it needs to be shifted up a tiny bit, that would simply not matter at all.
+The conclusions I am putting to you would still be right.
+They need to argue that the curve literally does not decline anything like that point, and I don't think anyone would try to make that case.
+
+Senator KIM CARR: I think that's a reasonable proposition. We may well seek from the department a critique of your paper -
+
+Mr Morrison: I'd be interested if they say it's not exactly the right design.
+
+Senator KIM CARR: in the normal peer review process.
+
+Mr Morrison: I am very happy to submit to my peers on this matter. I'm really quite confident. Or you might consider that this justifies further inquiry or extending this inquiry or a Senate inquiry.
+
+Senator KIM CARR: That's right. But it may well be that the department agrees that it requires further inquiry. Is that what you're suggesting to us?
+
+Mr Morrison: I think that when they check their numbers, yes, they might agree.
+
+Senator KIM CARR: I look forward to that. Thank you very much.
+
+CHAIR: Thank you very much, Mr Morrison.
+
+...
+
 \end{appendices}

### 2018-09-09 Second.Rnw
more

@@
 
-That paper, and a subsequent blog post on the topic which made the paper publically available, attracted the interest of the Senate Economics Reference Committee Inquiry into the Future of Naval Shipbuilding in Australia.
+That paper, and a subsequent blog post \parencite{morrison2018a} on the topic which made the paper publically available, attracted the interest of the Senate Economics Reference Committee Inquiry into the Future of Naval Shipbuilding in Australia.
 Senator Chris Ketter asked written questions on notice to the Department of Defence requesting responses to the paper specifically, and related comments from public officials, with a return date of 3rd April 2018.
@@
 \begin{tcolorbox}[width=0.9\textwidth, center, colback=red!5!white,colframe=red!75!black, title= \textbf{28 May 2012 - Senate Estimates}]
-\textit{Senator Johnston directly challenges Rear Admiral Justin Moffit (then director of SEA 1000) about the pumpjet proposal.  Full exchange in \ref{SENMoffit}.}
+\textit{Senator Johnston directly challenges Rear Admiral Justin Moffit (then director of SEA 1000) about the pumpjet proposal.  Full exchange in \ref{SENMoffit}. Emphasis added.}
 \\
@@
 \\
-\textbf{Rear Adm. Moffitt:} If it does not look like it makes sense, I will rule it out. But I cannot give you the answer to that yet because we have not done the work.
+\textbf{Rear Adm. Moffitt:} \textbf{If it does not look like it makes sense, I will rule it out. But I cannot give you the answer to that yet because we have not done the work.}
 \\
@@
 
+This encounter shows that the viability of pumpjets was challenged some time ago.
+It is noteworthy at the time that RADM Moffit makes no attempt to defend the consideration, acknowledging even that it might be easy for them to "write it off".
+It is also noteworthy the tone of confidence that Senator Johnston's challenge has, driving directly for an explanation of who had the original idea, which RADM Moffit was unable or unwilling to answer.
+
 \begin{tcolorbox}[width=0.9\textwidth, center, colback=blue!5!white,colframe=blue!75!black, title= \textbf{10 Feb 2016 - Sean Costello, CEO DCNS Australia}]
-\textit{In a blog post\parencite{costello2016} for the ASPI Strategist prior to the announcement in April, Sean Costello describes the significance of the pump jet technology to their bid.}
+\textit{In a blog post\parencite{costello2016} for the ASPI Strategist prior to the announcement in April, Sean Costello describes the significance of the pump jet technology to their bid.
+Emphasis added.}
 \\
@@
 ``As a complete submarine power, we understand conventional propulsion, which is why we also understand that propulsion is but one part of the submarine puzzle.
-In designing to the Australian requirement, it should come as no surprise then that the conventionally powered Shortfin Barracuda Block 1A is only 5\% lighter than its nuclear cousin."
-\\
-\\
-``There’s no better example of what I’m talking about than the offer from France to transfer to Australia sovereign control and use of pump-jet propulsion technology for the Shortfin Barracuda—technology resident only in France, the UK and the USA.
+In designing to the Australian requirement, it should come as no surprise then that \textbf{the conventionally powered Shortfin Barracuda Block 1A is only 5\% lighter than its nuclear cousin.}"
+\\
+\\
+``There's no better example of what I'm talking about than the \textbf{offer from France to transfer to Australia sovereign control and use of pump-jet propulsion technology for the Shortfin Barracuda}-technology resident only in France, the UK and the USA.
 Technology born from the French SSBN program a generation ago.
@@
 
+The significance of the pumpjet in Frances bid was clearly acknowledged well before the public announcement of DCNS's success.
+It also provides a clear indication that the size of the submarine already appeared to be quite well-known, and was only intended to be a very minor modification from the nuclear Barracuda.
+For such a small variation in displacement, it might be very reasonably expected that the length of the submarine would vary (the easiest modification), but the diameter would not need to.
+These comments also display quite clearly that the pumpjet technology in France's possession is recognised as common with technology in the US and the UK.
+This rules out the likelihood that some special or unique pumpjet technology, better suited just to diesel-electric submarines, had been pioneered by France alone.
+
 \begin{tcolorbox}[width=0.9\textwidth, center, colback=blue!5!white,colframe=blue!75!black, title= \textbf{8 Apr 2016 - Sean Costello, CEO DCNS Australia, and Gerard Autret, Chief Naval Architect of the Shortfin Barracuda}]
@@
 ``The challenge for any attack submarine is to maintain a nearly silent acoustic signature at a speed necessary to manoeuvre within weapons range of the target.
-The nuclear Barracuda is designed to reduce radiated noise when operating at a speed sufficient in order to close a threat submarine undetected.
-Of particular importance is the pump-jet propulsor, which combines a rotor and stator within a duct to significantly reduce the level of radiated noise through the effects of wake harmonisation and avoidance of cavitation."
+The nuclear Barracuda is designed to \textbf{reduce radiated noise when operating at a speed sufficient in order to close a threat submarine undetected.
+Of particular importance is the pump-jet propulsor}, which combines a rotor and stator within a duct to significantly reduce the level of radiated noise through the effects of wake harmonisation and avoidance of cavitation."
 \end{tcolorbox}
+
+This comment provides clear context for the importance of the pump-jet.
+Consistent with the physics outlined in Reference \parencite{morrison2018} the acoustic advantage is only emphasised by DCNS as being significant at higher speeds.
+Whether it is viable for a diesel-electric submarine (as opposed to the nuclear one mentioned here) to maintain such speeds to close with many types of targets is left unclear.
 
@@
 \textit{The official website for the submarine claims the advantages of the pumpjet in specific terms, drawing comparisons with "obsolete propeller technology" and "otherwise identical submarines".
-Last accessed 4 Sept 2018 \parencite{naval2018}.}
+Last accessed 4 Sept 2018 \parencite{naval2018}.
+Emphasis added.}
 \\
@@
 ``Pump jet propulsion means the Shortfin Barracuda can move more quietly than submarines with obsolete propeller technology.
-In a confrontation between two otherwise identical submarines, the one with pump jet propulsion always has the tactical advantage."
+In a confrontation between two \textbf{otherwise identical submarines}, the one with pump jet propulsion always has the \textbf{tactical} advantage."
 
@@
 
+The official explanation of the designer and builders website makes a carefully worded explanation of the superiority of the pumpjet, by limiting the comparison to a "tactical" encounter with an "otherwise identical" submarine.
+As discussed in \ref{SENmorrison}, this is the only \textit{possible} target that an elevated speed such as could be achieved by a diesel-electric submarine might assist in closing with.
+The submarine would be at a substantial "strategic" disadvantage in such an encounter, since its lower dived endurance, dived range, and poorer indiscretion ratio dramatically increases the probability that the pumpjet propelled submarine would wind up being the hunted, rather than the hunter.
+More significantly, the speeds achieved would not allow the submarine to close with any other target from behind such as almost all surface ships, or nuclear submarines.
+This statement exhibits extreme care in the wording in order to be plausibly defensible, and is aguably misleading.
+
 \begin{tcolorbox}[width=0.9\textwidth, center, colback=blue!5!white,colframe=blue!75!black, title= \textbf{4 October 2017 - Jean-Michel Billig, Executive Director Australian submarine Program, Naval Group}]
-\textit{In comments to journalists at the PACIFIC 2017 trade show, the Naval Group executive in charge of the Australian Program speculates that convetional propellers may be used on the Shortfin Barracuda \parencite{sbs2017}.
+\textit{In comments to journalists at the PACIFIC 2017 trade show, the Naval Group executive in charge of the Australian Program acknowledges that convetional propellers may be used on the Shortfin Barracuda \parencite{sbs2017}.
 Subsequent reports quote Naval Group shifting responsibility for the propulsion choice to the Australian Government.}
@@
 \\
-``The vessels may end up with conventional propellers as well as air independent propulsion, which helps to increase underwater endurance, Mr Billig said."
+(Interview transcribed from Reference \parencite{lane2017}.)
+Journalist: Does the pumpjet technology that is on Suffren remain viable for the Australian submarine?
+Jean-Michel Billig: It could be viable...
+Journalist: So it's not mandated, its...
+Jean-Michel Billig: What is driving the choice of technology is basically the operational requirement...
+Journalist: So we could end up with conventional propeller?
+Jean-Michel Billig: Yes, yes, we could, we could, we could also end up with something different
 \\
@@
 ``However decisions on forms of propulsion are ultimately a matter for the Australian Government."\parencite{iggulden2017}
-
 \end{tcolorbox}
 
+This crucial concession reveals a very different level of confidence in the pumpjet being suitable for the Australian diesel-electric submarine.
+It also gives an important glimpse into the processes that both selected the winning bidder, as well as the processes for the selection of key technology components.
+By revealing that the pumpjet has not yet been firmly decided upon, and that the technology choice is a product of the operational requirement, it makes it clear that the operational requirement has not yet been determined with sufficient precision to make the suitable technology choice clear to a ship-builder with ample experience in both technologies.
+Moreover, it also reveals that this was also the case at the conlcusion of the CEP, and remained so well over a year after.
+
 \begin{tcolorbox}[width=0.9\textwidth, center, colback=blue!5!white,colframe=blue!75!black, title= \textbf{25 Oct 2017 - Hevre Guillou, CEO Naval Group}]
-\textit{In an interview that was published in a post\parencite{guillou2017} on the ASPI Strategist the global CEO of Naval group responds to a question from Brendan Nicholson asking whether the Australian submarine will be fitted with a pump jet, following Jean-Michel Billig's comments.}
+\textit{In an interview that was published in a post\parencite{guillou2017} on the ASPI Strategist the global CEO of Naval group responds to a question from Brendan Nicholson asking whether the Australian submarine will be fitted with a pump jet, following Jean-Michel Billig's comments.
+Emphasis added.}
 \\
@@
 ``Naval Group is the manufacturer of both conventional propeller-driven submarines and larger nuclear submarines with the propulsor.
-The propulsor’s inclusion means the future submarine will move more quietly than submarines with obsolete propeller technology.
-In a contest between two otherwise identical submarines, the one with the pump jet always has the tactical advantage.
-The propulsor was an integral part of our bid.
-The Department of Defence says it intends to use the pump jet propulsion system."
+The propulsor's inclusion means the future submarine will move more quietly than submarines with obsolete propeller technology.
+\textbf{In a contest between two otherwise identical submarines, the one with the pump jet always has the tactical advantage.
+The propulsor was an integral part of our bid.} The Department of Defence says it intends to use the pump jet propulsion system."
 \end{tcolorbox}
+
+The verbatim quotation of the carefully worded comparison between ``otherwise identical" submarines by the global CEO more than a year after they first appeared on the Australian DCNS website demonstrate a high degree of preparation, and continued careful word selection on the topic of the pumpjet.
+This statement also further reveals nuances about the CEP selection process, simultaneously declaring that the propulsor (i.e.
+pumpjet) was ``integral" to their bid, whilst also deferring to what the Department of Defence ``intends" do use.
+This exposes a clear tension between how 'integral' the pumpjet was to the bid, and Defence's freedom to choose otherwise.
+It also exposes a dramatic divergence between the level of emphasis that is given to the pumpjet as an important key feature of the proposal in public promotion (as highlighted in \cite{lane2017}) and a Defence decision which seemed to choose only to have the option of including the pumpjet, with the real decision to be deferreed.
+
+\begin{tcolorbox}[width=0.9\textwidth, center, colback=red!5!white,colframe=red!75!black, title= \textbf{25 Oct 2017 - Senate Estimates}]
+\textit{Senator Hanson directly challenges Rear Admiral Gregory Sammut (then director of SEA 1000) about the pumpjet proposal.  Full exchange in \ref{SENhanson}.}
+\\
+\\
+\textbf{Senator Hanson:}   This is why you've chosen them as the design-to go with the pump
+jet. So it's not propellers, is it?
+
+\textbf{Rear Adm. Sammut:}   The proposal offered by France included a pump jet, and that's the
+proposal that we've accepted. \textbf{A pump jet remains on offer, and that's what we intend to use
+for the Future Submarine.}
+
+\textbf{Senator Hanson:}   And that's why you chose France's Naval Group to-
+
+\textbf{Rear Adm. Sammut:}   No, that is not, alone, why we chose France. There are a number of
+things. I am on record as providing testimony at the Senate of the number of things we
+considered in the selection of Naval Group as our international partner. There were a range of
+factors, and they were all over the capability of the international partner to deliver a regionally
+superior submarine and one which Australia would have the sovereign capacity to operate and
+sustain throughout its life. There were a number of criteria in mind.
+We looked at the capabilities. That includes the platform capabilities. We looked at the
+ability of the international partners to work with the combat system integrator of our choice.
+Those things in particular pertain to the capability of the boat. We looked at rough order of
+magnitude costs and schedules to ensure the participants understood the scale of the program
+that we were embarking upon and could give us insights into the manner in which they would
+manage costs and schedules throughout the program. We looked at matters concerning
+program management. \textbf{We looked at the design and the way that safety is incorporated into
+the design. It's important that that was understood from the outset because, whilst the
+international partner would be designing the submarine, it would be the Commonwealth that
+would be responsible for certifying the submarine as safe and fit for operations, and we
+needed to make sure that the certifying process would match the design philosophy.}
+We looked at methods of sustainment. \textbf{We looked at the manner in which each of the
+participants would support the establishment of a sovereign sustainment capability in
+Australia.} Of course, that was a key lesson from the Collins program. Even though we built
+the submarines in Australia, it took us some time to establish the sovereign capacity to sustain
+those boats. Indeed, having done that now is the reason why we are able to deliver beyond
+international benchmark availability. \textbf{We also looked into the way in which data would be
+transferred to Australia, the manner in which intellectual property would also be managed in
+the program.} We looked at the ability of the participants to manage risks. Again, throughout
+this complex program, that would be a very important factor. I would say that we didn't just
+pick a submarine with a pump jet; we picked an international partner that could meet a range
+of requirements to achieve not only a regionally superior submarine but the sovereign
+capacity to operate and sustain that submarine and to deliver that within the time frame that
+would be necessary.
+\\
+\\
+...
+\\
+\\
+\textbf{Rear Adm. Sammut:}  ...  The effort is not to convert an
+existing boat into a diesel-electric submarine; we're designing a submarine using many of the
+learnings and references that come out of the Barracuda design. \textbf{A good example there would
+be something like hull diameter. If you're going to change the diameter of the hull of a
+submarine, you're into a whole new set of calculations about the structural integrity and
+strength of the submarine.} If we can utilise the work that France has already done in having a
+hull diameter that's very similar to Barracuda, \textbf{we can leverage all of that design work, which
+has not only gone into the Barracuda but also has been tested by France in their methods of
+design and certification to get to this point.}
+
+\end{tcolorbox}
+This dialogue in the Senate Estimates received media coverage highlighting Senator Hanson's apparent lack of understanding about the difference between a submarine's power source (nuclear or diesel) and propulsion (propeller or pumpjet).
+However, she succeeded in applying sustained pressure to the question of the pumpjet, and the significance that it had in the selection of DCNS as a design partner.
+This elicited a committment to the pumpjet that appeared hesitant in its wording, which noted that the pump jet ``remains on offer", which is what ``we intend to use".
+
+RADM goes to length to explain that the pumpjet was not ``alone" the reason for selecting DCNS as a partner, and in doing so reveals important insights into Defence's thinking and process regarding the decision.
+Crucially, he reveals that the 'diameter of the hull' is an important aspect of France's design reference in the Barracuda, pointing out that changing the diameter of the hull incurrs a ``whole new set of calculations about the structural integrity and strength of the hull."  Together with remarks about the ``way that safety is incorporated into the design" and concerns about the alignment of the certifying process with the ``design philosophy" it would appear that the hull diameter actually emerges as plausibly one of the most significant distinguishing features in the French bid, perhaps alongside the pumpjet.
+
+This is alarming for two reasons.
+The first is that it contradicts subsequent explanations regarding pumpjet efficiency being incorrectly assessed by outsiders because they failed to account for how how the hull would be 'tuned' to the pumpjet and the pumpjet 'tuned' to the hull.
+If the hull diameter is already fixed and known as being identical to that of the nuclear Barracuda, then no adjustment ought to be necessary to a good working pumpjet design that is used to propel this platform on the hull's account.
+The performance of the pumpjet on the relevant hull would be well understood by the manufacturer offering it, and presumeably available to the Australian government who were looking to acquire it.
+
+The second is that is reveals a stunning naivety in terms of basic understanding of the heirachy of engineering challenges.
+It may be true that there are very many complex and challenging engineering calculations involved in building a submarine.
+However, it is certainly not true that scaling up a pressure hull to a new diameter to meet the same fixed static load is amongst the most difficult, risky or complex.
+Whilst extending the length of a submarine is undoubtedly even easier, adjusting the diameter still to meet a structural integrity requirement to meet a known static pressure (from depth) is a trivially simple calculation in comparison to the complexities of, for instance, re-designing the duct intake of blade arrangments of a pumpjet to maximise efficiency at a different speed whilst simultaneously attempting to avoid cavitation.
+It was quite simply and successfully accomplished when the Collins Class submarine was developed by scaling up both the length and diameter of it's smaller predecessor.
+
+That this ranks as a significant consideration in the mind of the chief of this program should be the cause of considerable alarm to those taking an interest the project.
+Were it known to the competitors that their ability to achieve this trivial engineering requirement was in question, they would no-doubt be highly offended.
+Any contender that was not considered competent to design a pressure hull of a new diameter should never have been invited to participate.
+RADM Sammut should be called upon to clarify what he meant by ``we needed to make sure that the certifying process would match the design philosophy" because if indeed he meant that Defence had already decided that their own certifying process already precluded scaled-up diameter hulls, this represents a very bold departure from what might be considered normal risk assessment processes for major engineering projects, including those used in designing and certifying the Collins.
+Furthermore, if the advantage was considered so substantial, he should be called upon to substantiate how significant increased difficulty in certification would have been, with respect to cost and schedule.
+
+Amongst RADM Sammut's additional claims regarding the reason for choosing DCNS/Naval Group, are futher claims regarding the establishment of 'sovereign capability' and 'data transfer' to Australia.
+This has significant bearing on the way that subsequent comments made by Defence and DCNS regarding the state of their understanding of pumpjets, and the method for their development for the Australian submarine.
 
@@
 Some claims about the pump jet are misleading.
-It’s not purely for use in nuclear submarines.
-It can be efficient across the entire speed range, taking account of the submarine’s size, the speed required and stealth.
+It's not purely for use in nuclear submarines.
+It can be efficient across the entire speed range, taking account of the submarine's size, the speed required and stealth.
 It needs to be tuned to the hull of the submarine.
-The hull needs to be tuned to the pump jet."
+The hull needs to be tuned to the pump jet.
+We did not specify that the boat should have or not have air-independent propulsion or lithium-ion batteries, rather that it met our requirements."
 \end{tcolorbox}
+
+This comment confirms that the official policy is still to use the pumpjet, implying that the decision has indeed been made already.
+This is at odds with the statement made only weeks earlier by Jean-Michel Billig.
+This is also the first noteable reference to the inter-relationship between the hull and the pumpjet, implying that the pumpjet will indeed be modified in order to suit the hull, and that the hull will be modified to suit the pumpjet.
+It also implies that the same pumpjet as France is fitting to the Suffren (i.e.
+nuclear barracuda) will not simply be inlcuded 'as is' on the Australian non-nuclear 'Shortfin' version.
+Given that the length and diameter of the hull were announced as being already decided, and seemingly very similar to the nuclear version, it raises the very serious question as to whether/why the changed shape of the hull would necessitate modification of the complex pumpjet.
+As pointed out in the authors testimony to the Senate \ref{SENmorrison}, the emphasis of this relationship, above the vastly different operational speed profile of the nuclear and non-nuclear submarines which would affect the propulsion system profoundly, raises serious questions about the origin of the comment.
 
@@
 
-\begin{tcolorbox}[width=0.9\textwidth, center, colback=red!5!white,colframe=red!75!black, title= \textbf{25 Oct 2017 - Senate Estimates}]
-\textit{Senator Hanson directly challenges Rear Admiral Gregory Sammut (then director of SEA 1000) about the pumpjet proposal.  Full exchange in \ref{SENhanson}.}
-\\
-\\
-\textbf{Rear Adm. Sammut:}   I think the point that we're trying to make here is that the proposal
-offered by France included a pump jet.
-
-\textbf{Senator Hanson:}   So it is pump jet? So you're going pump jet?
-
-\textbf{Rear Adm. Sammut:}   The proposal—
-
-\textbf{Senator Hanson:}   This is why you've chosen them as the design—to go with the pump
-jet. So it's not propellers, is it?
-
-\textbf{Rear Adm. Sammut:}   The proposal offered by France included a pump jet, and that's the
-proposal that we've accepted. A pump jet remains on offer, and that's what we intend to use
-for the Future Submarine.
-
-\textbf{Senator Hanson:}   And that's why you chose France's Naval Group to—
-
-\textbf{Rear Adm. Sammut:}   No, that is not, alone, why we chose France....
-\\
-\\
-\textit{Extensive remarks from Rear Adm. Sammut, and further questions from Senator Hanson about whether the proposal involves converting a nuclear submarine to a diesel sumbarine}
-\\
-\\
-\textbf{Rear Adm. Sammut:}  ...  The effort is not to convert an
-existing boat into a diesel-electric submarine; we're designing a submarine using many of the
-learnings and references that come out of the Barracuda design. A good example there would
-be something like hull diameter. If you're going to change the diameter of the hull of a
-submarine, you're into a whole new set of calculations about the structural integrity and
-strength of the submarine. If we can utilise the work that France has already done in having a
-hull diameter that's very similar to Barracuda, we can leverage all of that design work, which
-has not only gone into the Barracuda but also has been tested by France in their methods of
-design and certification to get to this point.
-
-\end{tcolorbox}
+This corroborates that the comments made by RADM Sammut regarding efficiency were also made by the Department, under due consideration, to a Question on Notice.
 
@@
 That basically describes the characteristics of pump jets, their pros and cons, as any piece of engineering equipment has.
-The document—and there's also another document that I've taken from an ASPI site—shows the efficiency characteristics of a pump jet.
-Both the US document—now declassified—and the document from the ASPI website suggest that, at low speed, pump jets are relatively inefficient compared to, say, a propeller.
+The document-and there's also another document that I've taken from an ASPI site-shows the efficiency characteristics of a pump jet.
+Both the US document-now declassified-and the document from the ASPI website suggest that, at low speed, pump jets are relatively inefficient compared to, say, a propeller.
 That may well be the reason why we see a pump jet on a Mark 48 torpedo but generally not on merchant ships and so forth.
@@
 
-\textbf{Mr Johnson:} As we said just a little earlier, the design process—whether it is a propeller or propulsion—is very much keyed to the criteria that you look for.
+\textbf{Mr Johnson:} As we said just a little earlier, the design process-whether it is a propeller or propulsion-is very much keyed to the criteria that you look for.
 The first thing I said was speed.
-No matter what kind of propeller or propulsion pump jet you design, it has to be optimised for a particular speed.
-There are classes of surface ships that look to optimise cross-ocean transit.
+\textbf{No matter what kind of propeller or propulsion pump jet you design, it has to be optimised for a particular speed.} There are classes of surface ships that look to optimise cross-ocean transit.
 Physics usually keeps you around 17 to 18 knots, and then there is the cost of maintenance and the cost of repair and the fact that there's no particular stealth requirement.
 Typically, you will find that those ships will use an open-tip propeller.
-However, other ships, cruise ships in particular, use an electric shrouded pod, which is a variation of a pump-jet technology.
-Of course, the technology gets changed over time, so the older a document is, the more it's dated to that point in time.
+\textbf{However, other ships, cruise ships in particular, use an electric shrouded pod, which is a variation of a pump-jet technology.} Of course, the technology gets changed over time, so the older a document is, the more it's dated to that point in time.
 \\
@@
 The decision-making process for submarines is a little more complicated because there's an acoustic performance aspect as well as the pure efficiency of the blade-propulsor-pump jet arrangement.
-Without going into details, it's a decision that we make with Navy, but I'm confident, having done this a few times already, that we are going to find this pump-jet approach to be the optimum for our particular circumstances for this particular submarine.
+\textbf{Without going into details, it's a decision that we make with Navy, but I'm confident, having done this a few times already, that we are going to find this pump-jet approach to be the optimum for our particular circumstances for this particular submarine.}
 \\
@@
 \textbf{Mr Johnson:} Yes.
-We will understand the performance data early enough in the design that we can accommodate whatever is required.
-And then, just to reassure you, it's really going to be an adjustment to the design of the pump jet, not a decision to leave it.
-Without going into too much detail, the curve that you're referring to is about the efficiency of the propeller, not about its acoustic performance.
+\textbf{We will understand the performance data early enough in the design that we can accommodate whatever is required.
+And then, just to reassure you, it's really going to be an adjustment to the design of the pump jet, not a decision to leave it.} Without going into too much detail, the curve that you're referring to is about the efficiency of the propeller, not about its acoustic performance.
 As you know, there's nothing free in physics, so those are different-shaped curves.
@@
 \end{tcolorbox}
+
+This testimony contradicts the assurance provided in an October interview with RADM Sammut.
+It is clear from multiple parts of Mr Johnson's statement that the decision on whether to use the pumpjet or not, has not yet been taken, since he is ``confident...
+that we are going to find this pump-jet approach to be the optimum...".
+Whilst acknowledging that he doesn't yet have the performance data from the nuclear Barracuda class (which is not yet in the water), he gives assurance that it will be obtained early enough to ``accommodate whateve is required", which he gives assurance is only going to be adjustments to the design of the pumpjet.
+This suggests that it is still not even known whether the pump-jet that was an 'integral' part of the bid was offered in an unmodified state, or whether it was known by France that it would need to be modified from existing designs, or whether what was offered was actually only a skill-set and capability for the design of a new pumpjet.
+
+In responding to the evidence produced by Senator Patrick that pumpjets may have an efficiency issue at low speeds, Mr Johnson refers to an 'electric shrouded pod' which is used on cruise ships, which he claims is a modification of a pumpjet.
+Nothing about this section of his testimony seems to have any reference of connection to reality.
+Cross-ocean transit regularly occur at higher speeds than 17-18kt, as do cruise ships, which he mentions.
+These speeds are essentially unreacheable (or only extremely briefly sustainable) by a diesel electric submarine.
+I can find no evidence of any cruise ships having any kind of shround around the tips of their propellers, or bearing any resemblance or relationship to a pumpjet.
+As I pointed out in Mr Johnson's presence in my later testimony to the Senate on a later occasion (which he could not rebut or answer in any way), it appears that Mr Johnson may have confused electric Azipods, which are used on large ocean-going ships such as cruise ships, and the Navy's new LHDs, with pumpjets.
+Azipods have open-tipped propellers, and are not remotely related to pumpjets.
+So far as I can discern, this particular part of the testimony was made-up entirely impromptu without any knowledge or care for the truth of the facts.
+The degree of confidence that we can have in his assurance that the pump-jet will be the optimum approach for the submarine should be considered in light of this.
 
@@
 Initial inputs will be developed in the last quarter of 2018, which marks the beginning of the development and test program for the pump jet.
-Development, testing, and assurance will be conducted by several parties including Naval Group, Direction générale de l’armement (DGA), and DSTG.
+Development, testing, and assurance will be conducted by several parties including Naval Group, Direction générale de l'armement (DGA), and DSTG.
 Testing and assurance activities will take place in France (including at the premises of the DGA Techniques Hydrodynamiques in Val de Reuil) and Australia.
@@
 Answer: ``The efficiency of pump-jets designed for submarine use cannot be determined through comparisons with water-jets and ducted propellers, which are different in their operation from pump-jets.
-Moreover, the global efficiency of submarine pump-jets cannot be derived from ‘open water’ tests of pump-jets, where the pump-jet is assessed independently of the hull of the submarine.
+Moreover, the global efficiency of submarine pump-jets cannot be derived from ‘open water' tests of pump-jets, where the pump-jet is assessed independently of the hull of the submarine.
 A correctly designed pump-jet, suited to the aft hull shape of the submarine, achieves better optimisation of efficiency and stealth than any propeller, even at low speeds."
@@
 \end{tcolorbox}
+
+This answer to a question again rests on the claim that the influence of the hull means that comparisons and studies such as those referred to in \parencite{morrison2018} can be ignored.
+This is a question which will be given a substantial reply in subsequent sections.
+
+However, the response also avoids directly answering the question which specifically refers to efficiency in isolation.
+By only claiming that the pumpjet ``achieves better optimisation of efficiency \textit{and stealth}, even at low speeds, the response avoids directly contradicting the evidence that pumpjet efficiency inevitably declines dramatically at low speeds.
 
@@
 \\
-Question: Is Defence still confident that Naval Group still hold the opinion that the pump-jet is wellsuited to diesel-electric submarines?
-\\
-\\
-Answer: ``Defence intends to use the pump jet propulsion system for the Future Submarine that wasoffered by Naval Group as part of their Competitive Evaluation Proposal and remains onoffer to Australia."
+Question: Is Defence still confident that Naval Group still hold the opinion that the pump-jet is well suited to diesel-electric submarines?
+\\
+\\
+Answer: ``Defence intends to use the pump jet propulsion system for the Future Submarine that was offered by Naval Group as part of their Competitive Evaluation Proposal and remains on offer to Australia."
 
 \end{tcolorbox}
+
+This constitutes point-blank refusal to answer the question that was asked.
+Given the earlier carefully worded comments of the most senior Naval Group officials, and careful placement of ultimate responsibility for the choice with the Commonwealth, such a refusal to answer a question is alarming.
+If the opinion that the pump-jet was overall well suited to diesel-electric submarines was held by Naval Group, it would be only natural to expect that that would be confidently communicated to the Department of Defence, given that they had assessed their bid which it was 'integral' to.
+Provided that opinion was communicated and hadn't changed, it would be easy and expected that Defence would give a confident answer to the question.
+The possibility that Naval Group offered a proposal which included technologies were known to be poorly suited together would constitute a breach of trust, and possibly fraud.
+
+Overall, this response casts serious doubt on the level of communication and advice which has flowed between Naval Group and the Department of Defence regarding the pumpjet to date.
+As a ship-builder which had built both pumpjets and propellers, to both diesel-electric as well as nuclear submarines, there could be no excuse for Naval Group to not have a well-informed opinion on the matter.
+For Defence not to have obtained that opinion from DCNS at the time of considering the bid would be a matter of incomprehensible oversight, and raises the question of whether deliberate neglect may have played a role.
 
@@
 Answer: ``See response to Senate Economics Reference Committee Question on Notice, Q1 – Pump Jet Efficiency Across the Entire Speed Range."
-
 \end{tcolorbox}
+The response to Q1 - a question specifically and only asking a question about \textit{efficiency} of the pumpjet, avoids contradicting the evidence directly by conflating the question with acoustics.
+To avoid the answering a question that is specifically about acoustics by referring to that same answer which incorporates efficiency is nothing more than a point-blank refusal to answer either question directly.
 
@@
 \\
-The company building the Navy’s 12 new submarines says they'll be the world’s first conventional boats big enough to effectively use the company's pump jet propulsion system.
+The company building the Navy's 12 new submarines says they'll be the world's first conventional boats big enough to effectively use the company's pump jet propulsion system.
 French Naval Group Australia says that its propulsor will prove much quieter than a propeller on the 4,700‑tonne submarine it has dubbed the 'Shortfin Barracuda'.
@@
 \\
-Naval Group Australia's interim CEO, Brent Clark, tells The Strategist that the decision about the best propulsion system for the submarine will be made by the Defence Department, but that he hadn’t had any indication that it would chose anything but a propulsor.
+Naval Group Australia's interim CEO, Brent Clark, tells The Strategist that the decision about the best propulsion system for the submarine will be made by the Defence Department, but that he hadn't had any indication that it would chose anything but a propulsor.
 'We've had every indication that they like what was offered.'
@@
 \\
-The propulsor that Naval Group is recommending for its Shortfin Barracuda consists of a propeller within a duct to reduce noise and the water disturbance known as cavitation, which can give away a boat’s position to hunters.
-Clark says he’s read many negative comments about propulsors, but some critics used very old studies to try and justify their argument.
-\\
-\\
-"A lot of people don’t fully comprehend, I think, that yes, a propulsor on a 1,500-tonne coastal submarine would make no sense.
-But we’re not building a 1,500-tonne coastal submarine.
-We’re building quite a large submarine.
+The propulsor that Naval Group is recommending for its Shortfin Barracuda consists of a propeller within a duct to reduce noise and the water disturbance known as cavitation, which can give away a boat's position to hunters.
+Clark says he's read many negative comments about propulsors, but some critics used very old studies to try and justify their argument.
+\\
+\\
+"A lot of people don't fully comprehend, I think, that yes, a propulsor on a 1,500-tonne coastal submarine would make no sense.
+But we're not building a 1,500-tonne coastal submarine.
+We're building quite a large submarine.
 And this is the first time that a conventional submarine's been large enough to take a propulsor."
@@
 \end{tcolorbox}
+
+Five months after the previous Senate hearing, this blog post confirms that the final choice on whether the pump-jet is to be included is still yet to be made, and Naval Group is still clear that the decision rests with Defence, rather than being assumed as tied to their bid.
+
+This is also the first time that the argument that the size of the submarine is a crucial enabling difference that makes the pumpjet viable is made.
+It is the closest that Naval Group has come from making a firm statement of faith in the suitability of the technology for the Australian diesel-elctric submarine.
+As I discussed in another blog post in ASPI Strategist \parencite{morrison2018b} shortly afterwards, this is a remarkeably concession that is consistent with the physics that I outlined, since the inefficiency of the pumpjet would likely drive demand for additional fuel and batteries (hence size) to meet requirements, which is more of a concession of the detrimental impact of the pumpjet on the submarine.
+This particular blog post was acknowledged on twitter by one former US Navy submariner and submarine enthusiast as an ``Awesome article about pump jet efficiencies and viability".
+Naval Group was tagged in the re-tweet, which was subsequently re-tweeted and liked by naval experts, publications, commentators and authors from around the world.
+
+\begin{tcolorbox}[width=0.9\textwidth, center, colback=red!5!white,colframe=red!75!black, title= \textbf{07 Jun 2018 - Senate Economic References Committee}]
+\textit{The author presents evidence to the final public hearing of this committee's Inquiry into the Future of Australia's Naval Shipbuilding Industry, challenging the pumpjet choice.
+Brent Clark, the outgoing CEO Naval Group Australia, Stephen Johnson, the General Manager Submarines from Defence, as well as Kim Gillis, Head of Capability and Sustainment Group at Defence, are all invited to respond to author's evidence.
+Full extracts of all discussion on the pumpjet topic are in  \ref{SENmorrison}.}
+\\
+\\
+\textbf{Mr Morrison:} ...In my report, I made the conclusive comparison that, between two otherwise identical submarines, the one with the pump jet will always have a lower dived endurance, a lower dived range, a worse indiscretion ratio, a lower overall endurance and a lower overall range than the one with the propeller.
+This statement, so far as I know, has yet to be challenged, and \textbf{I'd be very surprised if anyone with technical competence or professional integrity sought to challenge it.}
+...
+
+I'd like to look at one further public statement that was made.
+That is from Stephen Johnson.
+This was made in relation to an answer that was given to this committee some time ago.
+When challenged very specifically about efficiency by Senator Patrick, he made this statement:
+
+``No matter what kind of propeller or propulsion pump jet you design, it has to be optimised for a particular speed.
+There are classes of surface ships that look to optimise cross ocean transit.
+Physics usually keeps you around 17 to 18 knots, and then there is the cost of maintenance and the cost of repair and the fact that there's no particular stealth requirement.
+Typically, you will find that those ships will use an open-tip propeller.
+However, other ships, cruise ships in particular, use an electric shrouded pod, which is a variation of a pump-jet technology."
+...
+\textbf{I wonder whether here Mr Johnson has actually been confused with the Azipods that are installed on large cruise ships.
+They do use an electric motor, that's the only similarity I can find with the pump jet, and are indeed open to propellers and are quite similar to the propulsion systems used on the LHD.
+That is the best explanation that I can find so far for these comments.
+I certainly can't find any evidence of any shrouds being used on cruise-ship propulsion.
+Moreover, I can't find any evidence of pump jets of this sort being used on any ship, for which economic travel is a main priority, operating below the speeds of about 25 or 30 knots.
+So it's very interesting to wonder what exactly Mr Johnson was referring to in this case.}
+\\
+\\
+...
+\\
+\\
+\textbf{Senator PATRICK:} Firstly, it's probably proper that we give you an opportunity to respond to Mr Morrison's evidence. That's probably proper, Chair.
+
+\textbf{Mr Clark:} Thank you, Senator Patrick.
+I believe this is the first time we've sat opposite each other in a committee, so let the fun begin!
+From my perspective, \textbf{I don't intend to offer any commentary on the previous speaker.} Suffice to say that \textbf{the information that we've been supplying of a classified nature to the Australian government is held by the Australian government and was supplied by the French government.} The Australian government and the Australian Department of Defence obviously have access to classified research from the United States Navy, the Royal Navy and the French Navy on this matter, and I don't intend to get into a debate on this topic.
+\\
+\\
+\end{tcolorbox}
+
+Naval Group here steadfastly refuses to offer a clear opinion on the pumpjet.
+Only months earlier Defence was asked a specific question on notice about whether they thought Naval Group maintained that a pumpjet was suitable for a diesel-electric submarine.
+Clarke's interview on the Strategist, which seemed to be strongest statement of faith from Naval group in the suitability of the pumpjet for a conventional submarine (perhaps in an effort to patch up the noteable refusal to answer the question) had been answered publically by the same person who had just appeared and he was being asked to respond to.
+It might have been expected that Mr Clark would at least attempt to demonstrate conviction about the merits of the proposal.
+Instead, he described the information that they were supplying as a classified, government-to-government exchange, as if he were an impartial bystander.
+No assurance of the proposals superiority, suitability, or viability in any respect was made.
+
+Observers in the room or viewers of the video record may also observe that Mr Clark seemed to cut his statement short quite spontaneously at the point when he reached descriptions of engineering accomplishments surrounding the propulsion system, and there is no apparent record of the remainder of his statement actually being tabled as he said he would.
+
+\begin{tcolorbox}[width=0.9\textwidth, center, colback=red!5!white,colframe=red!75!black, title= \textbf{07 Jun 2018 - Senate Economic References Committee - CONTINUED}]
+
+\textbf{Senator PATRICK:} Sure.
+I have one more question, and this is to give you an opportunity to respond to Mr Morrison's evidence, Mr Johnson.
+He contradicted your evidence about cruise ships having pump jets.
+\textbf{He said that you might have meant Azipods?}
+
+\textbf{Mr Johnson:} \textbf{I don't remember that clearly enough.} I do remember that you had produced a document, and I was trying to answer questions without the benefit of where we were going.
+
+\textbf{Senator PATRICK:} Just giving you an opportunity to-
+
+\textbf{Mr Johnson:} I don't have a clear idea of exactly the choice of words I used, but I think the concept remains the same: propellers and propulsers are designed for very, very specific purposes, and if you change the purpose you will change your choice of design.
+
+\textbf{Senator PATRICK:} Thank you.
+
+\textbf{Senator KIM CARR:} Mr Johnson, you sat through the evidence this morning. I noted you at the back of the room when Mr Morrison made his presentation-is that correct?
+
+\textbf{Mr Johnson:} \textbf{Yes, I was here this morning. I wasn't necessarily paying attention.}
+
+\textbf{Senator KIM CARR:} You weren't? That's a big thing to admit on the Hansard!
+
+Senator Reynolds interjecting-
+
+Senator Kim Carr interjecting-
+
+\textbf{Mr Gillis:} I will support my colleague: not all of this is riveting!
+
+\textbf{Senator KIM CARR:} Is that right?
+
+\textbf{Senator REYNOLDS:} Mr Gillis!
+\\
+\\
+...
+\\
+\\
+\textbf{Senator KIM CARR:} I contend that we don't have the technical expertise to evaluate the report.
+It may well be that we recommend that further work be done on it, but I'd want to see your response to this-to give us a critique of it; your perspective as to the report.
+He's made some assertions to the committee.
+You no doubt will get the Hansard pretty quickly, if the secretary can make sure that that's available to the officers.
+
+\textbf{Mr Gillis:} The only caveat I'll put on any response we make is that the one thing we do really well is detailed engineering analysis.
+When you have somebody's report-\textbf{and I have not read the report yet; we weren't actually given a copy.} To actually go through it-and it looks like something which is yea thick-with serious analysis is going to take us time.
+We can only do what we can do.
+
+\textbf{Senator KIM CARR:} If anyone comes to me and says that fundamentally you've got it wrong on a technical matter such as a propulsion system, (1) we have to take notice of that, but (2) I'd be surprised, frankly, given the level of technical expertise available.
+
+\textbf{Mr Gillis:} Senator, that's one.
+I regularly get correspondence from the public saying that almost every decision that we've made for the last 50 years on a defence matter is wrong, because somebody has an opinion.
+
+\textbf{Senator KIM CARR:} Not since the fall of Singapore, surely!
+
+\textbf{Mr Gillis:} Almost since the fall of Singapore.
+\textbf{There are many people out there with opinions, and the vast majority-I would say over 99 per cent of them-you can dismiss within the first page of reading them because they are fundamentally wrong.}
+
+\textbf{Senator KIM CARR:} We clearly don't do that, because Mr Morrison wouldn't have got a hearing at this committee if we'd dismissed it. I'd like an assessment from you.
+
+\textbf{Mr Gillis:} I accept that.
+We will provide the committee with the best possible advice we can in the shortest possible time frame to meet your requirements so that you can be better informed, because we do hold that expertise.
+\\
+\\
+...
+\\
+\\
+\textbf{Mr Johnson:} Without contemporary knowledge or without understanding necessarily the intricacy.
+For example, the shaft horsepower-the power required to move even a very big ship at very low speeds-is astonishingly small.
+If you look at inefficiencies, you look at a curve and say, 'That must be a problem.' It's not.
+It's like: 'Hey, you paid twice what you needed to for that.
+\textbf{It was a penny.
+It cost you two pennies.
+That's 100 per cent over.}' So some of this is out of context with what we're doing, which will make it hard to give you an answer of empirical-
+
+\end{tcolorbox}
+
+This exchange demonstrates a clearly contemptuous attitude towards the evidence I had presented, which is clearly pervasive and supported in the higest levels of the defence department.
+The General Manager Submarines, Stephen Johnson, claims he simply cannot recall what he said about pumpjets being related to pods on cruise ships, despite being physically present in the same room when I pointedly quoted and contradicted him.
+This being pointed out, he claimed to simply not to be paying attention, which won the support of the Head of Capapbility and Sustainment Group, Kim Gillis.
+Despite the obvious shock and exclamation of the senators, Mr Gillis went on to attempt to dismiss my evidence as just another opinion which could probably to be dismissed along with 99\% of the others.
+
+This level of ambivalence towards external criticism from a such a senior official on such a topic as this should be alarming.
+The topic of the pumpjet had such sustained criticism and interest from various quarters, including multiple questions from senators from all sides of government over six years, reports in the ABC, ASPI Strategist, Defence Connect, Insight Economics, and many others.
+My paper was commissioned by a concerned citizen who wanted to have a better investigation into the physics underlying the controversy which had already received considerable attention, albeit with confusing results.
+
+Such an attitude raises questions about whether Mr Gillis's evidence is even given honestly.
+For instance, Mr Gillis' claim that he had not read the report, since ``we weren't actually  given a copy" warrants serious attention.
+My paper was made publically available with the blog-post summarising it on the 14th of February, and as well as in my reply to Brend Clark on  the 24th of May.
+Moreover, his department had received written Questions on Notice from Senator Chris Ketter, acting on behalf of this very committee, specifically requesting written responses to my paper with a return date of 3 April.
+His department did indeed responded to prior to the hearing, as noted above with responses which blatantly refused to answer the questions.
+Furthermore this particular hearing was also especially long in the making, since it was once postponed on account of Defence refusing to make officials available to attend.
+My presence as a witness was requested in February when the proposed date was still in May, and the list of witnesses was provided publically circulated during the weeks prior to the hearing.
+
+In such circumstances, claims to be ignorant of my report, and an attempt to lump it with the other 99\% of dissenting correspondence, seem about as reasonable as Mr Johnson's claim that he simply wasn't paying attention when he was being quoted and refuted before the committee earlier in the day.
+That this is the type of responses being given by the most senior officials who are responsible for the largest defence procurement in Australia's history should be alarming.
+
+\begin{tcolorbox}[width=0.9\textwidth, center, colback=black!5!white,colframe=black!75!black, title= \textbf{August 2018 - Dr David Kershaw, in Australian Defence Magazine}]
+\textit{Julian Kerr includes extensive quotation from Dr David Kershaw in an article titled "Future Submarine: Pumpjet propulsor technology choice still not well explained" \parencite{kerr2018}.
+Dr Kershaw is the current Chief of Maritime Division, a position he held since May 2016.
+He was previously the Science and Technology Advisor for SEA 1000 between July 2011 and July 2014.}
+\\
+\\
+As explained by Dr Kershaw, there was a degree of misunderstanding concerning pumpjet design and performance.
+To propel a surface ship or submarine involved overcoming resistance to push the vessel through the water with a suitably-designed propulsor.
+Surface ships need to overcome wave-making resistance, which varies with speed.
+Their propulsors therefore need to be optimised for the required operating speed; when operating at off-design speeds, efficiency drops off.
+\textbf{Since submerged submarines do not have to overcome wave resistance, the efficiency of a propulsor can be considered independent of speed} and the aim is to design an efficient propulsor.
+In other words, selecting the propulsor for a submarine is driven by the types of mission it is expected to undertake and where the boat will operate, Dr Kershaw stated.
+\\
+\\
+``The question of propeller versus pump jet really comes down to the acoustic signature; a well-designed pump jet is considerably quieter than an equivalent propeller." he said.
+"One of the advantages of a well-designed pump jet is that it can smooth out the water flow that is coming off the back of a submarine before the flow hits the moving rotor and there's less unsteady flow, and thus less noise.
+``This smoothing-out also increases the propulsive efficiency; a well-designed pump-jet can have efficiencies equal to, or better than, an equivalent propeller.
+You just have to have the capabilities and facilities to be able to design an efficient pumpjet and we know that France can do that."
+Echoing RADM Sammut, Dr Kershaw stressed that the pumpjet on a submarine must be designed in conjunction with the hullform to which it was to be attached.
+``Its' not like a propeller where you can design your propeller open loop i.e.
+without considering the interaction with the hull, and then attache the propeller to the hull.
+\textbf{The water flow over the hull to the pump jet is so integral you've got to design it as a pair, and the hull and the pump jet have to be tuned together}, it's a total system that you're building."
+\\
+\\
+``One of the issues with pumpjets and the reason they're not used on diesel-electric submarines is they're much heavier than a conventional propeller, although they have a smaller diameter.
+\textbf{``As the size of the submarine increases} (the Future Submarine's 4,700 tonne submerged displacement is significantly heavier than the 3,407 tonnes of the Collins-class it will replace) \textbf{you start to get to the point where a pump jet can become sensible."}
+\end{tcolorbox}
+
+The first feature of this passage is Dr Kershaws's telling remarks about wave-making resistance, and how in the special submarine scenario the efficiency of a propulsor can be considered independent of speed.
+This represents a pure, brazen, and obvious falsehood, justified by a seemingly-relevant truth.
+It's akin to arguing 'camels have to cary two heavy humps on their backs.
+Horses don't have to carry these two heavy humps, which is why they can fly.'
+
+Whilst it is true that deeply submerged bodies do not need to overcome wave-making resistance, all bodies in contact with any fluid must also overcome skin friction, wich means that resistance from this source scales with \textit{velocity squared} in almost all circumstances, including for ships.
+This is a most fundamental and basic assumption which is absolutely intrinsic to all basic marine propulsion, and should also be understood by essentially anyone who has passed first-year engineering or physics courses.
+(It's known) (For instance, see Pages 9 and 10 of Reference \parencite{MAN2017}, or Chapter 12 of Reference \parencite{Carlton2007}, or Chapter 3 of Reference \parencite{Mollard2011}.)
+
+Apart from it being quite an elementary law of phyiscs (the Drag Law), referred to and relied upon in posts such those by Andrew Davies \parencite{davies2017}, any substantial reading of the literature focussed on the design of pumpjets for fully submerged objects (such as torpedoes or submarines) will also make it abundantly clear, and the special distinction with wave-making resistance (which is crucial for fast-vessels in the process of reaching the plane) is also abundantly clear, and has been well known for a long time \parencite{wislicenus1973} \parencite{henderson1964}.
+
+That the Science and Technology Advisor on SEA 1000, and current Head of Maritime Division at DST could make such an gross error is extremely shocking.
+The lack of understanding demonstrated strongly contradict the assumption that he is competent to speak or advise on this topic.
+Whilst his expertise is most probably actually not that closely related to propulsion systems or hydrodynamics (his online bio suggests his torpedo-related experience is actually closer to the computer systems and signal processing involved \parencite{DSTG2018}) it would seem extremely strange for a competent physicist to such public statements outside their area of expertise, if they weren't confident in their own understanding.
+
+The other revealing thing about Dr David Kershaw's remarks here is the clear connection with comments made by RADM Sammut.
+He gives the most detailed description of the interaction between the hull and the pumpjet thus far, arriving at the same conclusion that they have to be 'tuned' together which has been previously used by RADM Sammut.
+(This argument will be addressed fully in a later chapter.)  An apparent conclusion is that RADM may have relied upon advice from Dr David Kershaw, (which given the ineptitute of earlier comments relating to hydrodynamics is gravely concerning), or that both of them have relied upon advice from another common source.
+
+The same conclusion might be reached here regarding the comments about submarine size, and a connection to comments made by Brent Clark for the ASPI Strategist \parencite{clark2018}.
+The comments by Dr Kershaw cast the argument in a very different light, suggesting that it is the weight of the pumpjet itself (as opposed to the weight of the extra fuel and batteries it demanded) was what made the pumpjet only viable for a very large diesel-electric submarine than anyone previous had built.
+This argument will be dealt with in a subsequent chapter.
+However, if it is indeed the case that the former CEO of Naval Group Australia, Brent Clark, was relying on Dr Kershaw for advice regarding reasons why the pumpjet was viable for the Australian future submarine, this raises the question as to why Mr Clark doesn't have his own advice from within Naval Group.
+This further throws into doubt that Naval Group still do hold the opinion that the pumpjet is well suited to a diesel-electric submarine.
+
+\begin{tcolorbox}[width=0.9\textwidth, center, colback=black!5!white,colframe=black!75!black, title= \textbf{August 2018 - Dr David Kershaw, in Australian Defence Magazine - CONTINUED}]
+Although pump jet propulsion would be new to the RAN, the technology was well-known to DSTG.
+"We've invested heavily in our hydrodynamics and hydroacoustics program over the past 20 years and this is includes the computational models as well as experimental facilities that confirm modelling results," Dr Kershaw said to \textit{ADM}.
+\textbf{We've been able to develop our analysis on a generic pumpjet design and thus confirm our understanding of pumpjet performance; you basically build it up from first principles.
+"Our analysis at the moment is all around analysing current options but we will be looking at what's ahead in the future and enable Australia to develop the sovereign capability to take best advantage of pumpjet technology."}
+\\
+\\
+So will pump jet technology definitely be used for the Future Submarine?
+``That is what is currently in the design \textbf{and that's what a lot of our science and technology program is about; having the capability to do the independent analysis on the technologyies} and capabilities planned for the Future Submarine, including the pump jet program," Dr Kershaw said to \textit{ADM}.
+\textbf{Discussions were currently underway with Naval Group and the French government regarding closer collaboration on a range of technologies and this would include pump jets, Dr Kershaw disclosed.}
+The program was still being worked on, and on the Australian side would involve a government-only team to comply with French security regulations.
+``One thing I would like to emphasis is that Australia's geographic circumstances require a larger conventional submarine than those currently produced for export to other parts of the world.
+The size of the Future Submarine means a pumpjet is a real game changer for optimising the performance of the submarine in its intended role, and ensuring regional superiority in undersea warfare."
+Dr Kershaw concluded.
+
+\end{tcolorbox}
+
+These comments are extremely important because the illuminate the true state of DSTG's research, and the degree of confidence they have in the pumpjet.
+It is consistent with responses given to a question on notice from Rex Patrick \ref{QON150}, where it was revealed that 1.1 million was spent on ``development of a \textit{generic submarine and pumpjet design}, computational and analytics models, and validation of these models through experimental studies".
+The admission that DSTG is still working on a 'generic' design also casts into new light their repeated insistence that the performance of a pumpjet cannot be assessed unless it is exactly tuned carefully with the hull, and vice versa.
+It is plainly obvious that if they are still working on 'generic' models, rather than the particular one which is proposed by France, they are in no better position to make positive assessment that contradicts findings in Reference \parencite{morrison2018}.
+By their own assessment, according to the insistence of Dr Kershaw and RADM Sammut, they are in no position to assess the performance of the pumpjet that will be used on the future submarine, given they haven't been able to 'tune' it to the hull shape of the future submarine, and 'tune' the hullshape to the pumjet.
+They are still working with 'generic' designs.
+
+As I have already put before the Senate \ref{SENmorrison}, I find that idea that such 'tuning' could so dramatically alter the performance of the pumjet to be completely implausible.
+I would support the idea that the kind of generic modelling described would be quite sufficient to reveal results that are consistent with the results I produced from a basic literature review (that pumpjet efficiency declines towards zero at very low water-speeds) and safely rule-out the suitability of a pumpjet for diesel electric submarines.
+If the results of the modelling done by DSTG to date could confidently contradict such a finding, one would expect that in Senate hearings and in responses to Questions on Notice, a very confident and clear-cut refution of the claims made in my report and evidence would be able to be made.
+This has been conspicuously absent from all responses, including this one.
+Instead, it is claimed that the 'tuning' with the hull, nullifies the result.
+If instead, the DSTG studies are actually consistent with my findings, then DSTG and Defence are knowingly supressing such results, or naively hoping that the French design, and the finer 'tuning' between the pumpjet and the hull can avhieve such a dramatically diferent result that the results of 'generic' tests can be ignored.
+Either situation would have terrible consequences for the program, and if the former is true it is likely that the Parliament has been deliberately misled on multiple occasions.
+
+These comments also illuminate a complete failure on the part of Naval Group to supply sufficient information to the Australian government regarding the pumpjet.
+Instead of relying on Naval Group's own modelling of the pumpjet that is being fitted to the nuclear Barracuda (which will have the same hull diameter as the Australian Shortfin Barracuda according to RADM Sammut), or even the proven, in-service pumpjet used on the larger French SSBNs, the Australian Defence Science and Technology Group is having to develop their own analysis on a ``generic pumpjet design" to confirm their own understanding through a process which they ``build up from first principles".
+The non-existent state of information transfer is further confirmed by the admission that \textit{discussions} are ``currently underway" for closer collaboration on this technology.
+
+That DSTG is still working up from first principles to develop the capability to do ``independent analysis on the technologies" that we have accepted in a $50 billion dollar bid from a supposedly trusted design partner should come to all observers as a shocking revelation.
+It is a deeply humiliating position for Australia, as a customer and an ally, to be so denied access to sufficient information to clearly and confidently understand the suitability of this technology.
+
+It is particularly shocking in light of the fact that the program head, RADM Sammut explicitly told Senate Estimates that, alongside the pumpjet, ``the way in which data would be transferred to Australia, the manner in which intellectual property would also be managed in the program" was part of the reason the partner was selected.
+
+It is even more shocking in the light of the fact that the CEO of Naval Group Australia, Sean Costello's, publically boasted about the significance of the ``offer from France to transfer to Australia sovereign control and use of pump-jet propulsion technology" \parencite{costello2016}, and his simultaneous claim (prior to the conclusion of the CEP) in the same that the performance data of the pumpjet are known by Australian government.
+
+It is still more shocking in the light of the then outgoing acting CEO, Brent Clark's further confimration before a Senate Inquiry \ref{SENmorrison} ``we've been supplying" classified documents on this topic to the Australian Government.
+
+In contrast to the assurance of full technology transfer, the comments by Dr Kershaw in this article reveals the true state of information flow from DCNS.
+Apparently, there is still not sufficent information to be confident about the use of the pumpjet.
+Still, a confident decision on whether the pumpjet will even be used cannot be made.
+Consistent with the statements of Mr Johnson \ref{SENjohnson}, the information about the pumpjet performance of the pumpjet on the nuclear Barracuda has not yet arrived.
+Nor, it would appear, is sufficient theoretical modelling, or scale model testing, or any other confirmatory data sufficient on this pumpjet for DSTG to be clearly persuaded of the suitability of the pumpjet for our submarine.
+Whilst simultaneously claiming that efficiency and performance of the pumpjet can't be assessed without the precise form of the hull and the pumpjet being 'tuned' together, they admit that they are having to model a 'generic' pumpjet, building up from first principals, in order to have sufficient confidence in their own 'indepenent analysis' of the technologies.
+
+It can be concluded firmly from this interview that the promise of technology transfer promised by Sean Costello has not been made good, even to a sufficient extent to allow the government to confidently decide on its suitability for the submarine, and assess Naval Groups confidence in the suitability sufficient to answer parliamentary Questions on Notice.
+Such gravity of such a failure is of profound consequence to the well-being of such a major project, given the significance of the pumpjet to the overall weight, balance and performance of the submarine as I outlined before the Senate \ref{SEMmorrison}.
+
+It also throws into doubt the integrity of the statements made by RADM Sammut before Senate Estimates.
+In the days and weeks immediately prior to his testimony \ref{SEMsammut}, Jean-Michel Billig had triggerd the controversy around the pumpjet at PACIFIC 2017, the largest Naval Show and conference in Australia, which occurs only each two years.
+This had lead to substantial coverage on the ABC, Defence Connect, and ASPI Strategist, and almost certainly contributed to the line of questioning that Senator Hanson doggedly pursued about the pumpjet that day.
+An interview where he faced further questions on the pumpjet's inclusion with Brendan Nicholson was published just days later\parencite{sammut2018}.
+It is inconceivable that the significance of the pumpjet question, and the state of the Commonwealth's confidence in its inclusion in the design was not front of mind at the time when he was being questioned by Senator Hanson on the 25th of October 2017.
+And he still chose to include the ``the way in which data would be transferred to Australia" as a reason for selecting the French, alongside the pumpjet, to deflect Senator Hanson's suggestion that the pumpjet alone was the main reason for selecting the French.
+
+If Naval Group's level of information transfer was proceeding as promised, then RADM Sammut had accepted (as being superior to other bids proposals) a sequence of data and technology transfer which would leave DSTG relying on their own 'generic' models in trying to assess the suitability of the pumpjet, and Defence unable to answer confidently whether Naval Group even had confidence in the suitability of the technology two years after the conclusion of the evaluation process.
+Unless every other competitor had also promised to withold critical information to assess the suitability of a technology that was 'integral' to their bid, there can be no plausible defence of such a decision as competent.
+Alternatively, if Naval group had promised more fluid information transfer, and this information had subsequently been withheld, RADM Sammut was actually concealing that a crucial break-down in relations with the major contractor had already occurred, with potentially disasterous consequences for the future of the submarine program.
+The only other viable alternative is that information transfer was not really something that commonwealth looked at in considering the selection of DCNS as a design partner, and RADM was knowingly misleading the Parliament.
+
+\begin{tcolorbox}[width=0.9\textwidth, center, colback=black!5!white,colframe=black!75!black, title= \textbf{August 2018 - RADM Sammut, in Australia Pacific Defence Reporter}]
+\textit{Kym Bergmann includes extensive quotation from RADM Greg Sammut in an article titled "SEA 1000 Meeting Major Milestones" \parencite{bergmann2018}.
+The following is an extract on the technologies relevant to propulsion.}
+\\
+\\
+Sammut confirmed that \textbf{the current thinking is
+for a pump jet} system to drive the submarines
+through the water. Navy is of the view that
+there are different characteristics to propulsion
+systems using propellers, ducted propellers
+and pump jets. \textbf{The performance of a pump
+jet needs to be assessed in conjunction with
+the characteristics of the submarine hull} and
+not in isolation, because they are matched -
+something that is overlooked in much of the
+current literature. \textbf{Pump jets are also stealthier
+than propellers across the entire speed range of
+the submarine} - something particularly important
+for the RAN because the submarines have to
+remain undetected during very lengthy transits
+and not only during the time on station.
+\\
+\\
+An advantage that Naval Group brings to this is
+that they have a great deal of experience with pump
+jet technology, including for the highly successful
+French SSBN fleet. This activity is also being
+supported by our Defence Science \& Technology
+Group (DSTG) and will result in scale models
+being tank tested in France at a highly classified
+facility that the author visited in 2017. On the other
+hand, \textbf{the first batch of submarines will use lead acid
+batteries rather than lithium ion because the
+RAN does not want to be introducing a massive
+number of new technologies simultaneously.}
+\\
+\\
+The Navy \textbf{is continuing to examine Air
+Independent Propulsion (AIP)} but has not yet
+found systems that are compatible with Australia's
+operational profile. This is because of our enormous
+transit distances, where the submarine might be
+better off carrying more fuel and food rather than
+a relatively cumbersome AIP system - though that
+could change.
+\end{tcolorbox}
+
+These series of comments by RADM Sammut are remarkeably consistent with other comments of his own and government sources, and wholly inconsitent with a competent and sound engineering and design process.
+
+It is inconceivable in an engineering project of this scale that after two years of work, and the conclusion October 2017 of the length, diameter and displacement of the submarine \parencite{sbs2017} that the type of propulsion is not yet known with certainty, and the inclusion (or not) of AIP is also not yet known with certainty.
+As discussed in my senate testimony \ref{SENmorrison} either of these technologies will alter load (of batteries, fuel, and AIP reactants) of the order of hundreds of tonnes, in order to meet a fixed performance requirement.
+Making such alterations within a hull that has its displacement already fixed could lead to catastrophic failure, essentially compromising the submarine's ability to sink or float.
+
+Only three possibilities exist which can explain why such a statement has been made.
+
+One possibility is that a firm decision actually already has been reached on the inclusion or exclusion of these technologies, and RADM Sammut is not willing to reveal it.
+It should be noted here that the 'current thinking' wording might be interpreted as something of a retreat from earlier 'we are using' statements made in late October 2017 \cite{sammut2017}.
+The possibility that a pumpjet has already actually been ruled out must at least be considered.
+In this scenario, it is almost certainly the case that the public and Parliament were misled on multiple occasions.
+
+A second possibility is that RADM Sammut's comments are genuine and true, no firm decision about technologies except the exclusion of lithium and the size of the hull have been made, in which case the future submarine project could be headed towards catastrophic failure in exactly the same trajectory of the Super-Seasprite, by allowing massive changes to the included capabilities and technologies after the selection of a fixed platform of finite capacity.
+
+The third possibility is that the performance specifications actually haven't been fixed either, and defence is planning on accepting whatever level of performance results from the combination of fuel, batteries, oxidants that fits in the defined hull-space in a neutrally buoyant and balanced arrangement driving whichever propulsor (propeller or pumpjet) is ultimately decided upon.
+That such an approach to project management could be taken on a project like this is almost unthinkable from an engineering and risk-management perspective, not to mention cost-effectiveness.
+
+However, such a possibility throws into sharp relieve some tensions between statements made by the government and Naval Group on multiple occasions.
+RADM Sammut had previously stated that ``We did not specify that the boat should have or not have air-independent propulsion or lithium-ion batteries, rather that it met our requirements." \cite{sammut2017}.
+Weeks earlier, Jean-Michel Billig had cited said ``What is driving the choice of technology is basically the operational requirement" when challenged as to whether the pumpjet was actually mandated \cite{lane2017}.
+With both parties pleading ignorance of what technologies would ultimately be included, and both citing 'requirements', it does seem to imply that the requirements had not yet been settled.
+
+If it was the case that the requirements had not been determined, then is likely that the Competitive Evaluation Process was entirely fraudulent.
+If performance requirements weren't actually given to the competitors in order that their proposed solutions to meet them could be fairly and properly assessed, then it is extremely difficult to see how they could be `treated fairly' as RADM Sammut claimed \ref{SENhanson}, and the credibility or conscience of the advisory panel who claimed they were should be seriously re-examined.
+If the requirements have changed substantially since the CEP, then the CEP ultimately served no purpose.
+A vast number of public statements made by politicians and officials regarding the significance of Australia's unique 'requirements', particularly for long range, should be critically re-examined.
+It is likely that numerous statements made citing these requirements, and implying that they existed, were known, and were communicated to other parties including throughout and subsequent to the CEP, would be found to be misleading of the public, and the Parliament.
 
@@
 
-\textbf{Senator JOHNSTON:} The French and the Russians have tried. No-one has used them. They are not being used on anything other than nuclear powered submarines—particularly the Americans.
+\textbf{Senator JOHNSTON:} The French and the Russians have tried. No-one has used them. They are not being used on anything other than nuclear powered submarines-particularly the Americans.
 
@@
 It is a submarine technology that we think that we need to know more about than we do today.
-I do not anticipate spending a substantial amount of money to get the understanding that I am looking for—I may not have to spend anything at all.
+I do not anticipate spending a substantial amount of money to get the understanding that I am looking for-I may not have to spend anything at all.
 
@@
 \textbf{Senator Hanson:}   At what time prior to that did you do investigations around other
-submarines—how many months prior to signing that contract? How much time and effort did
+submarines-how many months prior to signing that contract? How much time and effort did
 the Defence Force put into looking at other subs that were available prior to the Shortfin
@@
 
-\textbf{Senator Hanson:}   All right. How are the Shortfin Barracuda run—on what?
+\textbf{Senator Hanson:}   All right. How are the Shortfin Barracuda run-on what?
 
@@
 
-\textbf{Rear Adm. Sammut:}   The proposal—
-
-\textbf{Senator Hanson:}   This is why you've chosen them as the design—to go with the pump
+\textbf{Rear Adm. Sammut:}   The proposal-
+
+\textbf{Senator Hanson:}   This is why you've chosen them as the design-to go with the pump
 jet. So it's not propellers, is it?
@@
 
-\textbf{Senator Hanson:}   And that's why you chose France's Naval Group to—
+\textbf{Senator Hanson:}   And that's why you chose France's Naval Group to-
 
@@
 operation of diesel-electric submarines ourselves. We understand many of the factors around
-battery technology. I'd just like to add—
-
-\textbf{Senator Hanson:}   Sorry, I have limited time with this. I just want to ask you—
+battery technology. I'd just like to add-
+
+\textbf{Senator Hanson:}   Sorry, I have limited time with this. I just want to ask you-
 
@@
 
-\textbf{Senator Hanson:}   This Barracuda—there is still one on the slipways in France that
+\textbf{Senator Hanson:}   This Barracuda-there is still one on the slipways in France that
 hasn't even been tested. We're basing the design on that which has not been tested. Do you
@@
 
-\textbf{Senator Hanson:}   Is the Barracuda designed as a nuclear sub—that is, basically is a
+\textbf{Senator Hanson:}   Is the Barracuda designed as a nuclear sub-that is, basically is a
 nuclear sub?
@@
 \textbf{Mr Johnson:} The method of designing for either a propeller system or a propulser system, a pump jet, is very similar.
-It starts with key parameters: how fast do we want to go, how big, what's the weight and diameter of the vessel that we're going to power, is it a surface ship or a submarine—although the calculations are really different—and then what's the shape of that hull?
+It starts with key parameters: how fast do we want to go, how big, what's the weight and diameter of the vessel that we're going to power, is it a surface ship or a submarine-although the calculations are really different-and then what's the shape of that hull?
 All of those things then establish the particular characteristics of a propeller or a propulser pump jet design.
 The submarine that we're building is well suited, and the requirements that we have for performance are stringent that the pump jet technology is the optimum approach.
-Since it's a new design—contrary to misunderstandings in the press, this is a new design—we're in a good position to do those calculations for the pump jet from day one and to balance the submarine's designs so that it performs very well with that pump jet technology.
+Since it's a new design-contrary to misunderstandings in the press, this is a new design-we're in a good position to do those calculations for the pump jet from day one and to balance the submarine's designs so that it performs very well with that pump jet technology.
 That's been our plan since the selection of what was then DCNS and is now Naval Group France.
@@
 
-\textbf{Mr Johnson:} As we said just a little earlier, the design process—whether it is a propeller or propulsion—is very much keyed to the criteria that you look for.
+\textbf{Mr Johnson:} As we said just a little earlier, the design process-whether it is a propeller or propulsion-is very much keyed to the criteria that you look for.
 The first thing I said was speed.
@@
 
-\textbf{Air Chief Marshal Binskin:} Can I just confirm—we've got two documents and I'm trying to make sure we've baselined ourselves here. We've got one document, which is the 1963 document.
+\textbf{Air Chief Marshal Binskin:} Can I just confirm-we've got two documents and I'm trying to make sure we've baselined ourselves here. We've got one document, which is the 1963 document.
 
@@
 \textbf{Mr Johnson:} The decision to use a pump jet was carefully made.
-It includes a wider understanding of their use in other applications—leaving that open.
+It includes a wider understanding of their use in other applications-leaving that open.
 We have already accessed a relatively large scale model, so its physical data, and we will have the advantage of slightly different but very similar arrangements when the French attack submarine goes to sea.
@@
 I'll go back to you, Mr Johnson.
-Noting that you're saying that this is based on something that hasn't been to sea yet—it will go to sea on the Barracuda; presumably that's the submarine you're referring to—there's still a risk that's sitting there.
+Noting that you're saying that this is based on something that hasn't been to sea yet-it will go to sea on the Barracuda; presumably that's the submarine you're referring to-there's still a risk that's sitting there.
 As the guardian of the taxpayer, I know where you sit.
@@
 
-It might be very strange to try to figure out—and it was to me—why it would be useful to have a submarine that was only quieter at relatively high speeds and actually conferred disadvantages at all other speeds, including acoustically, with a significant trade-off in terms of overall range.
+It might be very strange to try to figure out-and it was to me-why it would be useful to have a submarine that was only quieter at relatively high speeds and actually conferred disadvantages at all other speeds, including acoustically, with a significant trade-off in terms of overall range.
 It seemed like it would be quite a strange tactical circumstance that would require such a thing.
@@
 In particular, statements by Jean-Michel Billig, who said back in October in 2017 that they may switch back to propellers and even insert AIP for higher underwater endurance, are perfectly consistent with the research that I've done.
-As well, more recently, the statements by Brent Clark on the ASPI Strategist blog published only a couple of weeks ago suggesting that a pump jet would not be suitable for a small submarine—but this submarine, being big enough, is perfectly consistent, as I pointed out in my public reply, with the research that I have already conducted.
+As well, more recently, the statements by Brent Clark on the ASPI Strategist blog published only a couple of weeks ago suggesting that a pump jet would not be suitable for a small submarine-but this submarine, being big enough, is perfectly consistent, as I pointed out in my public reply, with the research that I have already conducted.
 It seems to me that there is no contradiction clearly put by any representative of Naval Group to the research that I have presented on pump jets.
@@
 
-``It can be efficient across the entire speed range, taking account of the submarine’s size, the speed required and stealth.
+``It can be efficient across the entire speed range, taking account of the submarine's size, the speed required and stealth.
 It needs to be tuned to the hull of the submarine.
@@
 
-This was a statement made in October 2017, also published in an ASPI Strategist blog post his is a strange statement, because the first part—that it can be efficient across the entire speed range—taken on its own does not seem to be possible to be true.
+This was a statement made in October 2017, also published in an ASPI Strategist blog post his is a strange statement, because the first part-that it can be efficient across the entire speed range-taken on its own does not seem to be possible to be true.
 It seems to be completely false.
-Taking into account the submarine's size, the speed requirement and stealth, it seems to suggest that only if the admiral meant that he didn't actually want it to be efficient, and therefore could treat it as efficient—that seems to be the only way that it could be reconciled with reality and the physics that I've discovered.
+Taking into account the submarine's size, the speed requirement and stealth, it seems to suggest that only if the admiral meant that he didn't actually want it to be efficient, and therefore could treat it as efficient-that seems to be the only way that it could be reconciled with reality and the physics that I've discovered.
 
@@
 
-CHAIR: Thank you. I just want to place on record that since the commencement of the public hearing this morning Senator Reynolds has been joining us over the telephone line. She is now here with us.
-
-Senator PATRICK: Mr Morrison, firstly I will point out that Mr Johnson is in the back of the room, so he'll have an opportunity later today to rebut anything that you've said.
-
-Mr Morrison: Excellent.
-
-CHAIR: We'll certainly give him that opportunity. In terms of your credentials—you haven't explained. Have you got engineering or scientific credentials?
-
-Mr Morrison: I have a physics background.
+\textbf{CHAIR:} Thank you.
+I just want to place on record that since the commencement of the public hearing this morning Senator Reynolds has been joining us over the telephone line.
+She is now here with us.
+
+\textbf{Senator PATRICK:} Mr Morrison, firstly I will point out that Mr Johnson is in the back of the room, so he'll have an opportunity later today to rebut anything that you've said.
+
+\textbf{Mr Morrison:} Excellent.
+
+\textbf{CHAIR:} We'll certainly give him that opportunity. In terms of your credentials-you haven't explained. Have you got engineering or scientific credentials?
+
+\textbf{Mr Morrison:} I have a physics background.
 I studied theoretical physics at the ANU and got first-class honours in theoretical particle physics.
 I commenced a PhD at the University of Melbourne, which I didn't finish.
-I went to start a small defence technology company, designing small boats and torpedos for maritime warfare—a swarming concept.
+I went to start a small defence technology company, designing small boats and torpedos for maritime warfare-a swarming concept.
 We used pump jets.
@@
 
-Senator PATRICK: I just want to make sure that I've understood you correctly.
+\textbf{Senator PATRICK:} I just want to make sure that I've understood you correctly.
 You are saying that physics suggests that the pump jet is not efficient at the low speeds that a conventional submarine typically operates at in an operational area?
 
-Mr Morrison: That's right.
-
-Senator PATRICK: Of course, they use pump jets on nuclear powered submarines.
+\textbf{Mr Morrison:} That's right.
+
+\textbf{Senator PATRICK:} Of course, they use pump jets on nuclear powered submarines.
 Your paper says that's not a problem because they have no problems in respect of generating power, and you actually even say it's helpful.
 
-Mr Morrison: Yes. Would you like me to clarify that point?
-
-Senator PATRICK: Yes.
-
-Mr Morrison: Nuclear reactors have a tendency to have different instabilities build up when they try to oscillate their power outputs radically.
+\textbf{Mr Morrison:} Yes. Would you like me to clarify that point?
+
+\textbf{Senator PATRICK:} Yes.
+
+\textbf{Mr Morrison:} Nuclear reactors have a tendency to have different instabilities build up when they try to oscillate their power outputs radically.
 There are different decaying isotopes, xenon in particular, that make the reactor less stable after it has dramatically reduced its pressure.
@@
 
-Senator PATRICK: To your knowledge, are there any other conventional submarines—clearly, there are nuclear submarines that use a propulsor, or pump jets, but are you aware of any other conventional submarines that use them?
-
-Mr Morrison: There are none in service at all.
+\textbf{Senator PATRICK:} To your knowledge, are there any other conventional submarines-clearly, there are nuclear submarines that use a propulsor, or pump jets, but are you aware of any other conventional submarines that use them?
+
+\textbf{Mr Morrison:} There are none in service at all.
 I'm aware of a prototype or demonstrator that was built once on a Russian submarine, a Kilo class submarine.
@@
 
-Senator PATRICK: The bottom line for you is that it's problematic for a conventional submarine for all the reasons that you mentioned in terms of 'efficiency is king'.
+\textbf{Senator PATRICK:} The bottom line for you is that it's problematic for a conventional submarine for all the reasons that you mentioned in terms of 'efficiency is king'.
 It affects indiscretion ratios and it exposes a submarine, because it will have to snort more often.
 
-Mr Morrison: That's right.
+\textbf{Mr Morrison:} That's right.
 It seems like a genuinely strange choice.
@@
 
-Senator PATRICK: I tabled a document on 15 December, and it was pointed out that it was a 1963 declassified document from the US Navy.
+\textbf{Senator PATRICK:} I tabled a document on 15 December, and it was pointed out that it was a 1963 declassified document from the US Navy.
 There were some criticisms levelled at me because of the age of the document.
@@
 
-Mr Morrison: I'm absolutely satisfied.
+\textbf{Mr Morrison:} I'm absolutely satisfied.
 I have done quite a bit of research.
@@
 
-Senator PATRICK: What you're saying is akin to saying that we shouldn't reject Einstein's theory of relativity just because it's old?
-
-Mr Morrison: That's absolutely right.
+\textbf{Senator PATRICK:} What you're saying is akin to saying that we shouldn't reject Einstein's theory of relativity just because it's old?
+
+\textbf{Mr Morrison:} That's absolutely right.
 It's foundational, in fact.
-I can see nothing—and I have discussed this with other current experts—that has overturned the books, so to speak, on these matters.
-So I think some of those old foundational papers—and, indeed, I think I quoted one where it is surmised in a sentence—reached the conclusion: 'It's not good at low speeds'—almost as simply as that.
+I can see nothing-and I have discussed this with other current experts-that has overturned the books, so to speak, on these matters.
+So I think some of those old foundational papers-and, indeed, I think I quoted one where it is surmised in a sentence-reached the conclusion: 'It's not good at low speeds'-almost as simply as that.
 They don't go back and re-labour and reprove what has been so long and well established in the annals of physics.
 
-Senator PATRICK: You say that, acoustically, the pump jet is actually noisier at low speeds?
-
-Mr Morrison: I think that's highly probable.
+\textbf{Senator PATRICK:} You say that, acoustically, the pump jet is actually noisier at low speeds?
+
+\textbf{Mr Morrison:} I think that's highly probable.
 I won't make the absolutely categorical claim, as I've done with the efficiency, but I think that is extremely highly probably because, as you can see from my efficiency research, a pump-jet would be producing something like twice as much energy release to turbulence at those very low speeds as what a propeller would be.
@@
 This phenomenon called flow separation and the periodic vortex shedding that might occur into the rotors, that's a relatively low-frequency sound; it's not the sort of high-frequency sound that might be associated with the very, very early onset of cavitation.
-For very high-frequency sounds, the shielding from the duct is likely to be effective—still only in the directions that are perpendicular to the access of the shaft, but will be somewhat effective.
+For very high-frequency sounds, the shielding from the duct is likely to be effective-still only in the directions that are perpendicular to the access of the shaft, but will be somewhat effective.
 At the lower frequencies, I think there is good reason to think the shielding will not be as effective.
@@
 
-Senator PATRICK: Chair, I would like to table a paper that was produced by DCNS in 2012 at a UDT conference? Will you permit me to do that?
-
-CHAIR: Is there any objection to that document being tabled? No? That's okay.
-
-Senator PATRICK: I would like the witness to have a look at it.
+\textbf{Senator PATRICK:} Chair, I would like to table a paper that was produced by DCNS in 2012 at a UDT conference? Will you permit me to do that?
+
+\textbf{CHAIR:} Is there any objection to that document being tabled? No? That's okay.
+
+\textbf{Senator PATRICK:} I would like the witness to have a look at it.
 I think the company was called DCN then, and it's their engineering area.
@@
 This low TPK reduces significantly the propeller cavitation apparition risks.
-For instance, within conventional operating depths, so 150 metres and higher'—which I presume is French for deeper—'no cavitation is expected whatever speed may be.' That tends to suggest to me that the French designed a propeller back in 2002 that does not cavitate at any speed provided you are below 150 metres.
+For instance, within conventional operating depths, so 150 metres and higher'-which I presume is French for deeper-'no cavitation is expected whatever speed may be.' That tends to suggest to me that the French designed a propeller back in 2002 that does not cavitate at any speed provided you are below 150 metres.
 Is that consistent with your view on propellers versus pump-jets?
 
-Mr Morrison: Yes, that's absolutely consistent.
+\textbf{Mr Morrison:} Yes, that's absolutely consistent.
 I think there's a diagram that I've drawn in my report, in the acoustics section.
-It's absolutely the case that raising the static pressure, which is what happens when you go very deep—and it's also what the shroud of the duct attempts to achieve to the water flow in a pump-jet—is what reduces the onset of cavitation, which is the principal cause of all the noise.
+It's absolutely the case that raising the static pressure, which is what happens when you go very deep-and it's also what the shroud of the duct attempts to achieve to the water flow in a pump-jet-is what reduces the onset of cavitation, which is the principal cause of all the noise.
 Going deep simply does reduce or eliminate that.
 I think that, in this case, it is quite probable that at more modest speeds you'd be able to go quite shallow, potentially up to almost all of the routinely-used speeds of a diesel-electric submarine.
-The reason why you would want to be able to go in very shallow water at very high speeds for a very short period of time—it seems like a very narrow and a very obscure tactical advantage indeed, and that seems to be confirmed by the limitation of the superiority to a contest against another diesel-electric submarine only.
+The reason why you would want to be able to go in very shallow water at very high speeds for a very short period of time-it seems like a very narrow and a very obscure tactical advantage indeed, and that seems to be confirmed by the limitation of the superiority to a contest against another diesel-electric submarine only.
 It's hard to imagine how, against any other target, it could be useful.
 
-Senator PATRICK: You've got an engineering background.
+\textbf{Senator PATRICK:} You've got an engineering background.
 You said you've worked on some projects.
-What do you see is the risk, if you were to categorise the risk of Defence achieving the publicly-stated performance expectations of the pump-jet—that is, efficiency across the entire speed range and acoustically quiet?
-
-Mr Morrison: If they mean 'efficiency' to be anything close to a propeller across the entire speed range, I think achieving efficiency is impossible.
+What do you see is the risk, if you were to categorise the risk of Defence achieving the publicly-stated performance expectations of the pump-jet-that is, efficiency across the entire speed range and acoustically quiet?
+
+\textbf{Mr Morrison:} If they mean 'efficiency' to be anything close to a propeller across the entire speed range, I think achieving efficiency is impossible.
 I don't think they'll achieve that at all.
@@
 
-``Work on establishing the dimensions of the submarine is complete. That pre-sizing—the initial length, diameter, displacement and other key parameters …"
+``Work on establishing the dimensions of the submarine is complete. That pre-sizing-the initial length, diameter, displacement and other key parameters …"
 
@@
 The reason for that is quite simple.
-As I've described through my research, the difference in efficiency is sufficiently large that to meet the same fixed requirement would differ the amount of fuel and batteries required in the order of hundreds of tonnes—hundreds of tonnes—for this sized submarine.
+As I've described through my research, the difference in efficiency is sufficiently large that to meet the same fixed requirement would differ the amount of fuel and batteries required in the order of hundreds of tonnes-hundreds of tonnes-for this sized submarine.
 It seems very difficult to imagine how, for a submarine, which basically has to be perfectly neutrally buoyant in all its operations, we could add or subtract hundreds of tonnes in the same fixed displacement without having a catastrophic engineering failure.
@@
 They simply do not go together whichever way you put them.
-The risk that you simply have a submarine that doesn't float or doesn't sink—I'm not sure which is worse—is abundantly clear from these.
+The risk that you simply have a submarine that doesn't float or doesn't sink-I'm not sure which is worse-is abundantly clear from these.
 The only alternative that I can see is that they actually haven't fixed the requirements.
@@
 
-Senator PATRICK: I think you've partly answered this, but what do you see the risk is for the submarine project of proceeding with the pump-jet, noting you say it's impossible to achieve everything that they want to achieve and particularly noting the weight, the size and the fact that it's positioned on the arced extremity of the submarine?
-
-Mr Morrison: I think it adds exactly the point I've made: This is a very, very major change.
+\textbf{Senator PATRICK:} I think you've partly answered this, but what do you see the risk is for the submarine project of proceeding with the pump-jet, noting you say it's impossible to achieve everything that they want to achieve and particularly noting the weight, the size and the fact that it's positioned on the arced extremity of the submarine?
+
+\textbf{Mr Morrison:} I think it adds exactly the point I've made: This is a very, very major change.
 I didn't even mention that; it's in addition to the additional total volume of fuel and batteries required.
@@
 
-Senator PATRICK: Thank you. That's all I have.
-
-Senator KIM CARR: Mr Morrison, it's a very important point you make.
+\textbf{Senator PATRICK:} Thank you. That's all I have.
+
+\textbf{Senator KIM CARR:} Mr Morrison, it's a very important point you make.
 You're going to the very fundamental design questions that are at the heart of a major acquisition for this country.
@@
 
-Mr Morrison: Look, I'm just offering my evidence here.
+\textbf{Mr Morrison:} Look, I'm just offering my evidence here.
 To me it seems that the pump-jet is strange.
@@
 
-Senator KIM CARR: Given where we are in the procurement process, I think it is incumbent upon us to determine what we do with your evidence.
+\textbf{Senator KIM CARR:} Given where we are in the procurement process, I think it is incumbent upon us to determine what we do with your evidence.
 What are you actually recommending we do?
@@
 
-Mr Morrison: So far as I can tell, this is not controversial. Whilst it seems to be very technical physics, the paper that you -
-
-Senator KIM CARR: I'm not certain the Department of Defence would agree with your proposition that it's not controversial.
-
-Mr Morrison: I'd be very interested to see that.
+\textbf{Mr Morrison:} So far as I can tell, this is not controversial. Whilst it seems to be very technical physics, the paper that you -
+
+\textbf{Senator KIM CARR:} I'm not certain the Department of Defence would agree with your proposition that it's not controversial.
+
+\textbf{Mr Morrison:} I'd be very interested to see that.
 I've had discussions with other concerned physicists and people in the naval community about this, and I've heard of discussions had with the Department of Defence and some of the basis for their thinking that pump jets are as efficient, plausibly, over the whole speed range.
@@
 
-Senator KIM CARR: You're seeking from this committee further investigation of your claims. Is that the essence of your recommendation to us?
-
-Mr Morrison: Yes, and I would also strongly suggest that you carefully investigate the basis for the anti-submarine-warfare superiority claim.
+\textbf{Senator KIM CARR:} You're seeking from this committee further investigation of your claims. Is that the essence of your recommendation to us?
+
+\textbf{Mr Morrison:} Yes, and I would also strongly suggest that you carefully investigate the basis for the anti-submarine-warfare superiority claim.
 To me, that seems the last plausible thread of logic on which you could justify a submarine choice with a pump jet, and that's to say last plausible and not categorically and obviously untrue.
-There are gross doubts that I have about that as well: how you could ever acquire such a small target passively; how you would measure the range that you were at and know that you would be able to close in the short distance that you're able to; whether you'd be able to keep track of your target when you accelerate to those high speeds—there are a whole bunch of reasons which look almost as dubious as the proposition about pump jets being efficient.
+There are gross doubts that I have about that as well: how you could ever acquire such a small target passively; how you would measure the range that you were at and know that you would be able to close in the short distance that you're able to; whether you'd be able to keep track of your target when you accelerate to those high speeds-there are a whole bunch of reasons which look almost as dubious as the proposition about pump jets being efficient.
 
 I'm actually seeking to do some more research on that topic exactly, because it seems perfectly clear that that is the only situation in which there's a plausible case that the pump jet might be of any value, and I think your focus should very much dwell there.
-I'd like to get some more resources to conduct a thorough study of that, and I think that it should be challenged very directly with Defence—whether we're buying a submarine for all our strategic needs or whether we have just selected one particular scenario to justify a choice that's based on some other justification.
+I'd like to get some more resources to conduct a thorough study of that, and I think that it should be challenged very directly with Defence-whether we're buying a submarine for all our strategic needs or whether we have just selected one particular scenario to justify a choice that's based on some other justification.
 I can't see a logical justification for having made this particular design choice other than that.
@@
 
-Senator KIM CARR: You're saying you're undertaking further research. What's the state of that research?
-
-Mr Morrison: It's very early.
+\textbf{Senator KIM CARR:} You're saying you're undertaking further research. What's the state of that research?
+
+\textbf{Mr Morrison:} It's very early.
 I'm basically seeking further financial support to be able to do the same sort of sustained and thorough investigation as in the paper that you've got before you today.
@@
 I like to take the academia, the literature and the physics very seriously.
-But I think that sort of thing should be done quite urgently because I think there have been some very strange, potentially misleading—perhaps deliberately misleading—statements made about the superiority of the pump jet.
-
-Senator KIM CARR: This committee is due to report on 27 June. I'm not certain we'd have a lot more time to -
-
-Senator PATRICK: Senator Carr, if I may, just to kick this along a bit—it's a very good thought you're having. Do you have a security clearance?
-
-Mr Morrison: No.
-
-Senator PATRICK: That would preclude you from getting in.
+But I think that sort of thing should be done quite urgently because I think there have been some very strange, potentially misleading-perhaps deliberately misleading-statements made about the superiority of the pump jet.
+
+\textbf{Senator KIM CARR:} This committee is due to report on 27 June. I'm not certain we'd have a lot more time to -
+
+\textbf{Senator PATRICK:} Senator Carr, if I may, just to kick this along a bit-it's a very good thought you're having. Do you have a security clearance?
+
+\textbf{Mr Morrison:} No.
+
+\textbf{Senator PATRICK:} That would preclude you from getting in.
 My thought was you could perhaps go inside, work with Defence and then pop out and say either, 'I'm not convinced,' or, 'I am convinced,' keeping it at an unclassified level.
 
-Mr Morrison: The rules of physics are not classified or classifiable, and I think that's -
-
-Senator KIM CARR: I'm interested in pursuing what you're saying.
+\textbf{Mr Morrison:} The rules of physics are not classified or classifiable, and I think that's -
+
+\textbf{Senator KIM CARR:} I'm interested in pursuing what you're saying.
 As a citizen, I think you're entitled to put views which go to the heart of this procurement.
 I'm not saying that you're right.
-Please don't misinterpret me—I'm not qualified to say that you're correct, but I am interested in what you're saying because of the seriousness of the propositions you're advancing.
+Please don't misinterpret me-I'm not qualified to say that you're correct, but I am interested in what you're saying because of the seriousness of the propositions you're advancing.
 I am trying to come to grips with what we do with what you're saying.
@@
 
-Mr Morrison: Might I suggest, as a way to trying to seek validation, you could see whether anyone else would contradict or oppose some of the simple statements that I've made.
-
-Senator KIM CARR: We'll ask the Department of Defence but it will not surprise me if they tell me there's a technical reason why you can't be right.
-
-Mr Morrison: I'd be amazed to see that.
-Having been challenged, I would be amazed if someone wants to restate—and I've made some careful statements in my report.
+\textbf{Mr Morrison:} Might I suggest, as a way to trying to seek validation, you could see whether anyone else would contradict or oppose some of the simple statements that I've made.
+
+\textbf{Senator KIM CARR:} We'll ask the Department of Defence but it will not surprise me if they tell me there's a technical reason why you can't be right.
+
+\textbf{Mr Morrison:} I'd be amazed to see that.
+Having been challenged, I would be amazed if someone wants to restate-and I've made some careful statements in my report.
 You can ask them whether I'm wrong and see what they come back with.
@@
 
-Senator KIM CARR: Would you accept it on technical grounds, if they can demonstrate that you're wrong in theory?
-
-Mr Morrison: The whole thing has got to be wrong.
+\textbf{Senator KIM CARR:} Would you accept it on technical grounds, if they can demonstrate that you're wrong in theory?
+
+\textbf{Mr Morrison:} The whole thing has got to be wrong.
 It's got to be utterly wrong.
 I think this is worthwhile.
-If I was to go to a particular page of my report—on page 32, if you like, there's a chart from a paper that was done not too many years ago by BMT Defence Services—you can see the graph there—that shows the efficiency curve of a fully submerged water jet and a propeller going down to low speeds.
+If I was to go to a particular page of my report-on page 32, if you like, there's a chart from a paper that was done not too many years ago by BMT Defence Services-you can see the graph there-that shows the efficiency curve of a fully submerged water jet and a propeller going down to low speeds.
 Just look at the shape of the curve and the gap between those.
@@
 
-Senator KIM CARR: I think that's a reasonable proposition. We may well seek from the department a critique of your paper -
-
-Mr Morrison: I'd be interested if they say it's not exactly the right design.
-
-Senator KIM CARR: in the normal peer review process.
-
-Mr Morrison: I am very happy to submit to my peers on this matter. I'm really quite confident. Or you might consider that this justifies further inquiry or extending this inquiry or a Senate inquiry.
-
-Senator KIM CARR: That's right. But it may well be that the department agrees that it requires further inquiry. Is that what you're suggesting to us?
-
-Mr Morrison: I think that when they check their numbers, yes, they might agree.
-
-Senator KIM CARR: I look forward to that. Thank you very much.
-
-CHAIR: Thank you very much, Mr Morrison.
+\textbf{Senator KIM CARR:} I think that's a reasonable proposition. We may well seek from the department a critique of your paper -
+
+\textbf{Mr Morrison:} I'd be interested if they say it's not exactly the right design.
+
+\textbf{Senator KIM CARR:} in the normal peer review process.
+
+\textbf{Mr Morrison:} I am very happy to submit to my peers on this matter.
+I'm really quite confident.
+Or you might consider that this justifies further inquiry or extending this inquiry or a Senate inquiry.
+
+\textbf{Senator KIM CARR:} That's right. But it may well be that the department agrees that it requires further inquiry. Is that what you're suggesting to us?
+
+\textbf{Mr Morrison:} I think that when they check their numbers, yes, they might agree.
+
+\textbf{Senator KIM CARR:} I look forward to that. Thank you very much.
+
+\textbf{CHAIR:} Thank you very much, Mr Morrison.
 
@@
 
+\textbf{CHAIR:} I now welcome Mr Brent Clark from Naval Group Australia.
+Information on parliamentary privilege and the protection of witnesses and evidence has been provided to you.
+I invite you to make a brief opening statement and then we'll open it up for questions.
+
+\textbf{Mr Clark:} Good morning, everybody.
+I'm the special advisor on the Future Submarine program and to the Naval Group Australia chairman, Jean-Michel Billig.
+I was formally the interim chief executive of Naval Group Australia.
+It is always a pleasure to appear before this committee to provide you with an update on the Future Submarine program from the perspective of Naval Group Australia.
+Naval Group's work on the program has been proceeding at pace since my last appearance here almost 12 months ago.
+What I intend to do today is bring you up to speed with our efforts to deliver Australia a regionally superior submarine that will be built, operated and sustained with Australian sovereignty.
+
+Australia's \$1.6 trillion economy is dependent on secure sea lanes.
+As such, Australia needs the best possible submarine to protect our trade and support our maritime security.
+Working in collaboration with the Commonwealth and Lockheed Martin Australia, that is exactly what Naval Group will deliver.
+The Future Submarine program is the largest defence program in Australia's history and a truly national endeavour that will maximise Australian industry involvement and create thousands of jobs for many generations of Australians for decades to come.
+
+Naval Group, in partnership with Combat System Integration, Lockheed Martin Australia and the Australian government, is building 12 regionally superior diesel powered submarines for the Royal Australian Navy in order to maintain Australia's regional maritime superiority and sovereignty.
+Submarines are one of the most complex artefacts on the planet.
+Each submarine will be made up of at least one million parts, compared to a car which has 3,000 parts.
+As a result, we need to take our time.
+This program cannot be rushed.
+We need to get it right.
+I can tell you today that Naval Group is more than halfway through the design and mobilisation contract which focuses on early design activities in coordination with Lockheed Martin Australia and our other Australian partners.
+
+The Future Submarine is currently progressing through concept design on schedule.
+This has involved successful work to establish the initial length, diameter and displacement of the submarine ahead of the subsequent design phases.
+We are working with the Commonwealth on the components and systems that will be required for the submarine.
+The first stage of the submarine design, known as the functional analysis, was completed on schedule in October 2017, culminating in a successful preliminary system requirements review.
+We have finalised our Australian Industry Capability Plan, which will detail our early work to honour our commitments to maximise Australian industry involvement, particularly including SMEs.
+
+As at the start of May 2018, 813 companies had engaged with Naval Group nationally through the Industry Capability Network, an online database which gives companies access to projects large and small.
+Suppliers for the top five pieces of equipment, including the main motor, diesel generators, switchboards, batteries and weapons discharge systems are currently being reviewed.
+We will work with the suppliers of this equipment to ensure that there is a full technology transfer and that Australian industry is able to maintain and support this equipment.
+
+To give you some numbers to work with, the Naval Group workforce has already grown to about 90 people at our Adelaide headquarters, we have six engineers currently in France and we'll be recruiting another 31 engineers to transfer to France this year.
+That number will grow to about 100 Australian engineers in France over the next few years.
+Our workforce is expected to grow to about 155 by the end of this year.
+Also by the end of 2018, we're working with the Commonwealth towards turning the first sod of soil and laying the foundations for the new submarine construction yard at Osborne in Adelaide's northern precinct.
+Over the next three years, the construction sheds, which will house the equipment to build the submarines, and the construction halls for the submarines will go up.
+The establishment of the infrastructure in this time frame will be required to ensure that construction of the submarines can begin from 2022-23.
+We envisage building up the Naval Group Australia workforce to around 1,500 employees by around 2028-29.
+Overall, the program is expected to generate an annual average of around 2,800 jobs in Australia.
+
+The first submarine will commence service in the early 2030s, with the last to be constructed some time in the late 2040s and sustainment to continue to the end of the 2080s.
+Over the next four years, our partners, Lockheed Martin, will establish the combat system testing and integration facility to test the capabilities that will contribute to the submarines' lethality and regional superiority.
+Naval Group will build a land based propulsion and testing facility to derisk the submarine's propulsion system, which is a sophisticated piece of engineering that attempts to reduce the noise signature of the diesel-electric propulsion system.
+
+At this point, if the chair is happy, I'll table the remainder of my speech so that we can get to the questions.
+
+\textbf{CHAIR:} Thank you very much, Mr Clark.
+
+\textbf{Senator PATRICK:} Firstly, it's probably proper that we give you an opportunity to respond to Mr Morrison's evidence. That's probably proper, Chair.
+
+\textbf{Mr Clark:} Thank you, Senator Patrick.
+I believe this is the first time we've sat opposite each other in a committee, so let the fun begin!
+From my perspective, I don't intend to offer any commentary on the previous speaker.
+Suffice to say that the information that we've been supplying of a classified nature to the Australian government is held by the Australian government and was supplied by the French government.
+The Australian government and the Australian Department of Defence obviously have access to classified research from the United States Navy, the Royal Navy and the French Navy on this matter, and I don't intend to get into a debate on this topic.
+
+\textbf{Senator PATRICK:} Thank you. I might just step back. Mr Morrison suggested that there had been a Russian commercial submarine that had been fitted with a pump-jet.
+
+\textbf{Mr Clark:} Yes.
+
+\textbf{Senator PATRICK:} Can you confirm that the French fitted, for a trial, a pump-jet on to an Agosta 70 class submarine?
+
+\textbf{Mr Clark:} Yes they did, many years ago.
+
+\textbf{Senator PATRICK:} DCNS obviously has nuclear customers, being Brazil and, of course, the French Navy, and a range of conventional customers.
+Has DCNS offered or put a propulsor on any conventional submarine?
+
+\textbf{Mr Clark:} Can I clarify one thing?
+Naval Group does not have a nuclear customer in Brazil.
+We are assisting the Brazilians to create their own sovereignty in their own country, but France has not committed to supply any kind of nuclear technology to the Brazilians.
+
+\textbf{Senator PATRICK:} So they're doing the reactor-
+
+\textbf{Mr Clark:} Correct.
+
+\textbf{Senator PATRICK:} but you're doing the submarine?
+
+\textbf{Mr Clark:} We're assisting them in terms of designing their submarine.
+
+\textbf{Senator PATRICK:} Are you assisting them with a pump-jet?
+
+\textbf{Mr Clark:} Not at this stage.
+Pump-jet technology, like much of the technology on this submarine, is incredibly sensitive.
+The DGA, the French authorities, are the design authority for the pump-jet for Naval Group and they are working with the Commonwealth of Australia.
+This is not a technology that's offered to anyone.
+This is a technology that is only offered to certain allies.
+
+...
+
+\textbf{Senator PATRICK:} Sure.
+I have one more question, and this is to give you an opportunity to respond to Mr Morrison's evidence, Mr Johnson.
+He contradicted your evidence about cruise ships having pump jets.
+He said that you might have meant Azipods?
+
+\textbf{Mr Johnson:} I don't remember that clearly enough. I do remember that you had produced a document, and I was trying to answer questions without the benefit of where we were going.
+
+\textbf{Senator PATRICK:} Just giving you an opportunity to-
+
+\textbf{Mr Johnson:} I don't have a clear idea of exactly the choice of words I used, but I think the concept remains the same: propellers and propulsers are designed for very, very specific purposes, and if you change the purpose you will change your choice of design.
+
+\textbf{Senator PATRICK:} Thank you.
+
+\textbf{Senator KIM CARR:} Mr Johnson, you sat through the evidence this morning. I noted you at the back of the room when Mr Morrison made his presentation-is that correct?
+
+\textbf{Mr Johnson:} Yes, I was here this morning. I wasn't necessarily paying attention.
+
+\textbf{Senator KIM CARR:} You weren't? That's a big thing to admit on the Hansard!
+
+Senator Reynolds interjecting-
+
+Senator Kim Carr interjecting-
+
+\textbf{Mr Gillis:} I will support my colleague: not all of this is riveting!
+
+\textbf{Senator KIM CARR:} Is that right?
+
+\textbf{Senator REYNOLDS:} Mr Gillis!
+
+\textbf{Senator KIM CARR:} Mr Chesworth, maybe this is a point where it will be easier than you thought, with this level of humour!
+What's your response to the suggestion that we get a critique from the department as to the paper that's been presented to this committee?
+Is that possible?
+
+\textbf{Mr Johnson:} I'm sure it's possible, yes.
+
+\textbf{Senator KIM CARR:} We have a deadline in terms of the committee's reporting date. If you were conscience of that, would it-
+
+\textbf{Mr Johnson:} What would that deadline be?
+
+\textbf{CHAIR:} 27 June.
+
+\textbf{Mr Johnson:} Surely you would want something before that?
+
+\textbf{Senator KIM CARR:} Of course. That's the point. I'm not as humorous as you fellows, but it would look pretty bloody stupid after the report's handed in, wouldn't it?
+
+\textbf{Mr Johnson:} I've only had a chance to look through it. First of all, we welcome all ideas. I think I don't want to say much more than we'll do our very best to support you.
+
+\textbf{Senator KIM CARR:} The reason I raise this is that I think citizens are entitled to put a view about matters of this importance.
+
+\textbf{Mr Gillis:} Absolutely.
+
+\textbf{Senator KIM CARR:} I contend that we don't have the technical expertise to evaluate the report.
+It may well be that we recommend that further work be done on it, but I'd want to see your response to this-to give us a critique of it; your perspective as to the report.
+He's made some assertions to the committee.
+You no doubt will get the Hansard pretty quickly, if the secretary can make sure that that's available to the officers.
+
+\textbf{Mr Gillis:} The only caveat I'll put on any response we make is that the one thing we do really well is detailed engineering analysis.
+When you have somebody's report-and I have not read the report yet; we weren't actually given a copy.
+To actually go through it-and it looks like something which is yea thick-with serious analysis is going to take us time.
+We can only do what we can do.
+
+\textbf{Senator KIM CARR:} If anyone comes to me and says that fundamentally you've got it wrong on a technical matter such as a propulsion system, (1) we have to take notice of that, but (2) I'd be surprised, frankly, given the level of technical expertise available.
+
+\textbf{Mr Gillis:} Senator, that's one.
+I regularly get correspondence from the public saying that almost every decision that we've made for the last 50 years on a defence matter is wrong, because somebody has an opinion.
+
+\textbf{Senator KIM CARR:} Not since the fall of Singapore, surely!
+
+\textbf{Mr Gillis:} Almost since the fall of Singapore.
+There are many people out there with opinions, and the vast majority-I would say over 99 per cent of them-you can dismiss within the first page of reading them because they are fundamentally wrong.
+
+\textbf{Senator KIM CARR:} We clearly don't do that, because Mr Morrison wouldn't have got a hearing at this committee if we'd dismissed it. I'd like an assessment from you.
+
+\textbf{Mr Gillis:} I accept that.
+We will provide the committee with the best possible advice we can in the shortest possible time frame to meet your requirements so that you can be better informed, because we do hold that expertise.
+
+\textbf{Senator KIM CARR:} Thank you.
+There was a reference to advice that you have from various navies-obviously, classified advice.
+Now, you won't be able to draw on that, but we need to know whether or not it's consistent with the advice that you have in terms of contemporary-that's the point.
+
+\textbf{Mr Johnson:} The intended plan is to use a propulsor. Is that what you're referring to?
+
+\textbf{Senator KIM CARR:} The point of the solution-
+
+\textbf{Mr Johnson:} That's well supported across-
+
+\textbf{Senator KIM CARR:} It may be that we seek an in camera, private briefing, from you.
+
+\textbf{Mr Johnson:} That's exactly where I was headed.
+That use is-I think it is fair to say-well supported.
+I don't want you to misunderstand me, because the very first thing I said was, 'Well, of course, we will.' But this report also talks about things-and I'll give you one example-like batteries and battery efficiencies.
+Those are not things we want out in the public-dive and endurance and those sorts of things.
+So the beauty of a paper such as this, which comes from outside of the establishment, is that there are sometimes good ideas in new papers.
+The disadvantage is that sometimes they are created and constructed with a misunderstanding-
+
+\textbf{Senator KIM CARR:} Or without contemporary knowledge.
+
+\textbf{Mr Johnson:} Without contemporary knowledge or without understanding necessarily the intricacy.
+For example, the shaft horsepower-the power required to move even a very big ship at very low speeds-is astonishingly small.
+If you look at inefficiencies, you look at a curve and say, 'That must be a problem.' It's not.
+It's like: 'Hey, you paid twice what you needed to for that.
+It was a penny.
+It cost you two pennies.
+That's 100 per cent over.' So some of this is out of context with what we're doing, which will make it hard to give you an answer of empirical-
+
+\textbf{Mr Gillis:} One of the things that we learned during the competitive evaluation process and through all of the studies I've done on submarines-and I have nowhere near the life experience that the admiral here has-is that every country in the world is highly sensitive about the performance of their submarines, what technology they use and how they do it.
+
+\textbf{Senator KIM CARR:} So they should.
+
+\textbf{Mr Gillis:} So it is almost that the vast majority of external papers are written in a vacuum of information about what's really out there.
+That's the problem.
+The same thing happened on the joint strike fighter with stealth technology.
+We had 10 years of commentators saying the thing would never fly.
+It would never meet stealth requirements.
+It would never do those things.
+The only thing I haven't heard in the three years since I've been here was after the JSF landed here in Australia: they seem to have shut up, because it exists.
+So I just counsel that sometimes, especially on very highly sensitive defence materiel, people will use 'cousin Google' to get a lot of information.
+I'm not referring to this individual, but I see a lot of that stuff.
+
+\textbf{Senator KIM CARR:} I don't think this fellow meets that criteria.
+
+\textbf{Mr Gillis:} No, I don't. We have to value the input that we get from the public, and I look forward to reading the paper.
+
+\textbf{CHAIR:} Our time is nearly out, so I will call on Senator Reynolds.
+
+\subsection{Australian Defence Magazine, August 2018 Extract}
+
+An extract from "Future Submarine: Pumpjet propulsor technology choice still not well explained" \cite{kerr2018}, article commencing Page 20.
+
+As explained by Dr Kershaw, there was a degree of misunderstanding concerning pumpjet design and performance.
+To propel a surface ship or submarine involved overcoming resistance to push the vessel through the water with a suitably-designed propulsor.
+Surface ships need to overcome wave-making resistance, which varies with speed.
+Their propulsors therefore need to be optimised for the required operating speed; when operating at off-design speeds, efficiency drops off.
+Since submerged submarines do not have to overcome wave resistance, the efficiency  of a propulsor can be considered independent of speed and the aim is to design an efficient propulsor.
+In other words, selecting the propulsor for a submarine is driven by the types of mission it is expected to undertake and where the boat will operate, Dr Kershaw stated.
+
+``The question of propeller versus pump jet really comes down to the acoustic signature; a well-designed pump jet is considerably quieter than an equivalent propeller." he said.
+"One of the advantages of a well-designed pump jet is that it can smooth out the water flow that is coming off the back of a submarine before the flow hits the moving rotor and there's less unsteady flow, and thus less noise.
+``This smoothing-out also increases the propulsive efficiency; a well-designed pump-jet can have efficiencies equal to, or better than, an equivalent propeller.
+You just have to have the capabilities and facilities to be able to design an efficient pumpjet and we know that France can do that."
+Echoing RADM Sammut, Dr Kershaw stressed that the pumpjet on a submarine must be designed in conjunction with the hullform to which it was to be attached.
+``Its' not like a propeller where you can design your propeller open loop i.e.
+without considering the interaction with the hull, and then attache the propeller to the hull.
+The water flow over the hull to the pump jet is so integral you've got to design it as a pair, and the hull and the pump jet have to be tuned together, it's a total system that you're building."
+
+``One of the issues with pumpjets and the reason they're not used on diesel-electric submarines is they're much heavier than a conventional propeller, although they have a smaller diameter.
+``As the size of the submarine increases (the Future Submarine's 4,700 tonne submerged displacement is significantly heavier than the 3,407 tonnes of the Collins-class it will replace) you start to get to the point where a pump jet can become sensible."
+
+Although pump jet propulsion would be new to the RAN, the technology was well-known to DSTG.
+"We've invested heavily in our hydrodynamics and hydroacoustics program over the past 20 years and this is includes the computational models as well as experimental facilities that confirm modelling results," Dr Kershaw said to \textit{ADM}.
+We've been able to develop our analysis on a generic pumpjet design and thus confirm our understanding of pumpjet performance; you basically build it up from first principles.
+"Our analysis at the moment is all around analysing current options but we will be looking at what's ahead in the future and enable Australia to develop the sovereign capability to take best advantage of pumpjet technology."
+
+\textbf{Classified information}
+
+It was worth noting, however, that modern pumpjets were primarily employed on military submarines and torpedoes and information was very limited and tightly held, making it difficult to review performance predictions.
+``DST though, is in the position where we can access certain privileged information, so we can get validation of our analysis capabilities," Dr Kershaw noted.
+Pumpjet design had advanced over many years with a particular forcus on the characteristics of all the components, although much of that was classified.
+Some propellers were highly specialised and highly complex. Pump jets had complexities in other aspects but the moving rotor was not as complex as a normal propeller, benefitting sustainment.
+
+So will pump jet technology definitely be used for the Future Submarine?
+``That is what is currently in the design  and that's what a lot of our science and technology program is about; having the capability to do the independent analysis on the technologyies and capabilities planned for the Future Submarine, including the pump jet program," Dr Kershaw said to \textit{ADM}.
+Discussions were currently underway with Naval Group and the French government regarding closer collaboration on a range of technologies and this would include pump jets, Dr Kershaw disclosed.
+The program was still being worked on, and on the Australian side would involve a government-only team to comply with French security regulations.
+``One thing I would like to emphasis is that Australia's geographic circumstances require a larger conventional submarine than those currently produced for export to other parts of the world.
+The size of the Future Submarine means a pumpjet is a real game changer for optimising the performance of the submarine in its intended role, and ensuring regional superiority in undersea warfare."
+Dr Kershaw concluded.
+
 \end{appendices}

### 2018-09-09 Second.Rnw
some more

@@
 Journalist: Does the pumpjet technology that is on Suffren remain viable for the Australian submarine?
-Jean-Michel Billig: It could be viable...
-Journalist: So it's not mandated, its...
-Jean-Michel Billig: What is driving the choice of technology is basically the operational requirement...
-Journalist: So we could end up with conventional propeller?
-Jean-Michel Billig: Yes, yes, we could, we could, we could also end up with something different
+
+\textbf{Jean-Michel Billig:} It could be viable...
+
+\textbf{Journalist:} So it's not mandated, its...
+
+\textbf{Jean-Michel Billig:} What is driving the choice of technology is basically the operational requirement...
+
+\textbf{Journalist:} So we could end up with conventional propeller?
+
+\textbf{Jean-Michel Billig:} Yes, yes, we could, we could, we could also end up with something different
 \\
@@
 
-That DSTG is still working up from first principles to develop the capability to do ``independent analysis on the technologies" that we have accepted in a $50 billion dollar bid from a supposedly trusted design partner should come to all observers as a shocking revelation.
+That DSTG is still working up from first principles to develop the capability to do ``independent analysis on the technologies" that we have accepted in a \$50 billion dollar bid from a supposedly trusted design partner should come to all observers as a shocking revelation.
 It is a deeply humiliating position for Australia, as a customer and an ally, to be so denied access to sufficient information to clearly and confidently understand the suitability of this technology.
@@
 
-These series of comments by RADM Sammut are remarkeably consistent with other comments of his own and government sources, and wholly inconsitent with a competent and sound engineering and design process.
+These series of comments by RADM Sammut are consistent with other comments of his own and other government sources, and wholly inconsitent with a competent and sound engineering and design process.
 
@@
 One possibility is that a firm decision actually already has been reached on the inclusion or exclusion of these technologies, and RADM Sammut is not willing to reveal it.
-It should be noted here that the 'current thinking' wording might be interpreted as something of a retreat from earlier 'we are using' statements made in late October 2017 \cite{sammut2017}.
+It should be noted here that the 'current thinking' wording might be interpreted as something of a retreat from earlier 'we are using' statements made in late October 2017 \parencite{sammut2017}.
 The possibility that a pumpjet has already actually been ruled out must at least be considered.
@@
 However, such a possibility throws into sharp relieve some tensions between statements made by the government and Naval Group on multiple occasions.
-RADM Sammut had previously stated that ``We did not specify that the boat should have or not have air-independent propulsion or lithium-ion batteries, rather that it met our requirements." \cite{sammut2017}.
-Weeks earlier, Jean-Michel Billig had cited said ``What is driving the choice of technology is basically the operational requirement" when challenged as to whether the pumpjet was actually mandated \cite{lane2017}.
+RADM Sammut had previously stated that ``We did not specify that the boat should have or not have air-independent propulsion or lithium-ion batteries, rather that it met our requirements." \parencite{sammut2017}.
+Weeks earlier, Jean-Michel Billig had cited said ``What is driving the choice of technology is basically the operational requirement" when challenged as to whether the pumpjet was actually mandated \parencite{lane2017}.
 With both parties pleading ignorance of what technologies would ultimately be included, and both citing 'requirements', it does seem to imply that the requirements had not yet been settled.
@@
 If it was the case that the requirements had not been determined, then is likely that the Competitive Evaluation Process was entirely fraudulent.
-If performance requirements weren't actually given to the competitors in order that their proposed solutions to meet them could be fairly and properly assessed, then it is extremely difficult to see how they could be `treated fairly' as RADM Sammut claimed \ref{SENhanson}, and the credibility or conscience of the advisory panel who claimed they were should be seriously re-examined.
+If all the competitors weren't actually given the same performance requirements, which were the real requirements for the submarine which the government intended to procure, then it is extremely difficult to see how they could be `treated fairly' as RADM Sammut claimed \ref{SENhanson}.
+Moreover, credibility or conscience of the advisory panel who claimed they were should be seriously re-examined.
 If the requirements have changed substantially since the CEP, then the CEP ultimately served no purpose.
@@
 It is likely that numerous statements made citing these requirements, and implying that they existed, were known, and were communicated to other parties including throughout and subsequent to the CEP, would be found to be misleading of the public, and the Parliament.
+
+The explanation given regarding AIP adds weight to the bizarre notion that a back-to-front design process is indeed taking place.
+The enormous transit distances cited have been well known for decades, and presumeably the requirement to meet these had been distilled into specific ranges and endurances long before the commencement of the CEP.
+It's true that large amounts of fuel and food might consequently be required.
+But the idea that we 'might be better off carrying more food and fuel' betrays either either a lack of confidence that the volumes which were presumeably propososed by all contenders alongside an AIP system were sufficient, or a casual aspiration to substantially enlarge the range and endurance of the submarine even further.
+It would be relatively simple at the very early stages of concept design (prior to the diameter and legnth of the submarine being fixed) to incorporate both sufficient fuel and food for a specific overall range and endurance requirement, as well as an AIP system to enhance the dived endurance, without compromising the overall range and endurance.
+That this was not done, and instead the dimensions of the hull were fixed, whilst Navy continued to search for a system which might fit within this pre-defined platform envelope does not remotely resemble a competent strategy to acquire a capability to meet a fixed requirement.
+
+Worryingly, it adds confirmation that RADM Sammut might genuinely believe that the hull-diameter (and its resemblance of a pre-existing design) ought to be a pre-determined feature of a submarine design, to which the performance and capability of the submarine is subordinate.
+Such a view, including held on the grounds of safety as suggested by RADM Sammut himself \ref{SENhanson}, is such a bizarre understanding of program risk and the heirachy of engineering challenges that it should call into question the competence of a program manager, and the technical advisors who he depends upon.
+
+The competence of RADM Sammut's program management and the technical advisors on whom he depends is further marred by the comment about lithium-ion batteries being now firmly excluded from the design because the ``RAN does not want to be introducing a massive number of new technologies simultaneously".
+This is telling because lead-acid batteries are known to have a relatively high internal resistance, which increases as they become hot, which occurs during levels of faster discharge.
+This means that the efficiency of the batteries falls substantially at higher rates of discharge, reducing substantially the total amount of stored energy that they can usefully recover at very high discharge rates.
+Whilst this is generally true for all batteries, Lithium-ion batteries are known to have substantially lower internal resistance, and consequently have dramatically improved performance (in terms efficiency) at higher rates of discharge when  compared to lead-acid batteries.
+It is for this reason that in submarines, lithium-ion batteries are acknowledged as having the most significant improvement over lead-acid at high speeds \parencite{briggs2016}.
+Given that pumpjets advantages are widely acknowledged to be at higher speeds, including implicitly by Naval Group\parencite{autret2016}, lithium-ion batteries should really be considered an enabling technology for pumpjets on diesel-electric submarines.
+Without them, the pumpjet efficiency will sap performance at low speeds, and at high speeds (where pumpjet efficiency improves, albeit not as high as propellers) the battery efficiency will hobble performance.
+This effect, not accounted for in the modelling in reference \parencite{morrison2018}, will probably result in the maximum endurance of a pumpjet-submarine in a sprint at 20kt from over 2 hours (as that modelling suggested) to under one hour.
+With lithium ion batteries, it is plausible that slightly over 2 hours at this speed could be sustained.
+
+As I have also pointed out
 

### 2018-09-14 Second.Rnw
a couple more bits

@@
 It is for this reason that in submarines, lithium-ion batteries are acknowledged as having the most significant improvement over lead-acid at high speeds \parencite{briggs2016}.
-Given that pumpjets advantages are widely acknowledged to be at higher speeds, including implicitly by Naval Group\parencite{autret2016}, lithium-ion batteries should really be considered an enabling technology for pumpjets on diesel-electric submarines.
-Without them, the pumpjet efficiency will sap performance at low speeds, and at high speeds (where pumpjet efficiency improves, albeit not as high as propellers) the battery efficiency will hobble performance.
-This effect, not accounted for in the modelling in reference \parencite{morrison2018}, will probably result in the maximum endurance of a pumpjet-submarine in a sprint at 20kt from over 2 hours (as that modelling suggested) to under one hour.
-With lithium ion batteries, it is plausible that slightly over 2 hours at this speed could be sustained.
-
-As I have also pointed out
+Given that pumpjets advantages are widely acknowledged to be at higher speeds, including implicitly by Naval Group\parencite{autret2016}, lithium-ion batteries should really be considered an enabling technology for pumpjets on diesel-electric submarines, at least if the case is to be made that priorisation of that extremely high-speed sprint capability should be made above all other performance characteristics, which I have argued is an obscure and implausible case to be made \ref{SENmorrison}  Without them, the pumpjet efficiency will sap performance at low speeds, and at high speeds (where pumpjet efficiency improves, albeit not as high as propellers) the battery efficiency will hobble performance.
+This effect, not accounted for in the modelling in reference \parencite{morrison2018}, will probably result in the maximum endurance of a pumpjet-submarine in a sprint at 20kt from over 2 hours (as that modelling suggested) to well under one hour.
+With lithium ion batteries, it is plausible that around 2 hours at this speed might be sustained.
+The exclusion of lithium batteries \textit{because} of the pumpjet's inclusion further indicates that RADM Sammut remains completely unaware of such interactions between technologies.
 
 \newpage
+
+\section{The weight and size argument}
+Both Dr David Kershaw, Chief of Maritime Division at DSTG\parencite{kerr2018}, as well as Mr Brent Clark\parencite{clark2018} have both made arugments to the effect of suggesting that it was the Australian submarine's very large size which is what made the pumpjet viable, where it might not be for smaller submarines.
+I responded directly to Clark's comments arguing that the size necessitated by the extra fuel and batteries a pumpjet demanded could hardly be counted as an advantage \parencite{morrison2018b}.
+However, Dr Kershaw's argument reveals a different as to why it was that size was important.
+His comments make it clear that it is actually the large weight of the pumpjet \textit{itself} rather than the additional fuel and batteries it required, that was significant.
+
+This argument implies that a smaller submarine couldn't adequately provide sufficient buoyancy or balance to support the heavy pumpjet.
+If the submarine couldn't provide such buoyancy or balance, the consequences would be catastophic failure, specifically the submarine would tend to either tail-stand, or sink.
+It would be unthinkable for a competent manufacturer, conscious of the weight of the pumpjet and such risks, to neglect to account for such risks.
+Weights and balances are known and predictable quantities, which could most easily be accounted for in the design of a new submarine.
+
+However, Dr Kershaw's argument is most comprehensively debunked by the known fact that pumpjets have indeed been successfully retro-fitted to existing smaller diesel-electric submarines.
+As one would expect, none of the catastrophic failures eventuated.
+It is reported that the Russian Kilo Class submarine B-871 'Alrosa' has and does take part in exercises, and is remains in active service.
+However, given that no other conventional submarines have since been designed with or converted to pumpjet propulsion, it is difficult to conclude that the benefits of a pumpjet on this type of submarine outweigh the costs.
+It is trivial, however, to conclude that the weight of the pumpjet itself is not prohibitive for a smaller submarine, including  one smaller than the Collins Class, such as the Kilo.
+
+Furthermore, the French also fitted a pumpjet to an Agosta 70 class submarine for a trial, as confirmed by Brent Clark in the Senate \ref{SENmorrison}.
+This submarine is even smaller than the Kilo, around 1500 tonnes, exactly the type of 'coastal' submarine that Brent Clark said a pumpjet wouldn't make sense for \parencite{clark2018}.
+One would assume however that if the weight of the pumpjet itself was likely to cause the submarine to tail-stand or sink, such a trial would not have even been attempted.
+
+Further physical arguments could be made as to how the weight of the pumpjet ought to scale with the size of submarine.
+There are obvious reasons why the width/height (which must be equal, since the jet is cylindrical/conical) ought to scale in proportion to the diameter of the hull, in order to engage a proportionally larger water-column to produce the thrust necessary.
+Similarly, the length  of the shround is likely to remain relatively fixed in proportion to the diameter for a given performance requirement.
+It is unlikely that the proportional thickness of the blades or shroud could be greatly diminished at larger scales without structural integrity and hydrodynamic qualities of the shroud and rotor blades being compromised.
+All of these relationships strongly suggest that Dr Kershaw's implied suggestion that the weight of the pumpjet does not scale consistently with the weight of the submarine is completely unfounded, probably thoughtless, or otherwise  deliberately misleading.
+However, given the clear-cut evidence that the type of failure that an over-weight propulsor would cause have been easily avoided on \textit{existing} smaller conventional submarines, such arguments wont be expanded in detail here.
+It is abundantly clear that the arguments are entirely baseless, and demonstrably so.
+
+It is difficult to imagine what sort of circumstance would lead both an acting CEO of the contracted design partner, who was aware of the experience of the French experiment with the Agosta, as well as a very senior defence scientist to make such demonstrably false claims in public.
+In the case of Dr Kershaw, it can be seen that he deliberately and repeatedly (at the conclusion of his interview  \parencite{kerr2018}) drew attention to the size being significant.
+His concluding remarks for the interview were that ``The size of the Future Submarine means a pumpjet is a real game changer for optimising the performance of the submarine in its intended role...".
+To deliberately mount an argument that is so simply ridiculed and dismissed as being so significant seems to be a bizarre choice for a supposedly competent scientist to make.
+If stronger arguments were available, one would assume that they would be made instead.
+With over a million dollars already spent on modelling and experiments, one would assume that if the results of that research could confidently refute the conclusions my own research \parencite{morrison2018}, such a desperate and embarrassing statement would not need to be made.
+
+In a clear parallel with the argument that the performance of propulsion systems on submerged bodies is independed of speed because they don't face wave-making resistance, this line of reasoning involves the presentation of a seemingly relevant, technical, yet understandable truth which is used to justify a completely blatant technical falsehood.
+Even if Dr Kershaw is not familiar with this area of science and speaking outside his competence, it is difficult to see how such a pattern of argument occuring so consistently by accident.
+For Dr Kershaw to genuinely believe in the arguments that he is making here is almost unthinkable.
+For him to casually or accidentally make such careless arguments also seems implausibly reckless.
+A more likely explanation is that the results of the research already conducted by the government broadly support the conclusions of my own research, and this is well known by Dr Kershaw, and  that these unfortunate statements are being made in a desparate attempt to deflect attention in order to avoid admitting as much.
+
+\newpage
+
+\section{The interaction between the hull and the propulsor}
+The argument that the 'hull needs to be tuned to pumpjet' and \textit{vice versa} has been used to dismiss critcism of the pumpjet repeatedly by RADM Sammut \parencite{sammut2017} \parencite{bergmann2018}, and also by Dr David Kershaw \parencite{kerr2018}.
+It has also been referred to in an answer to a Question on Notice given by Senator Ketter\ref{QON1}.
+The answer to the Question on Notice, and also Dr Kershaw's comments comprise clearly the more detailed explanations, which  will be referred to in principal in this section.
+However, the reference of RADM Sammut to this line of argument on multiple occasions, almost a year separated both before and after the publication of my paper\parencite{morrison2018} and my senate evidence \ref{SENmorrison} clearly indicates that it is a significant, perhaps the most significant line of defence of the  pumpjet in the program chief's mind.
+The written response supplied by Defence to the Senate Economic References Committee also refers clearly to this argument \parencite{defence2018}.
+
+As has already been noted, there are glaring contradictions from a project-management perspective which oppose this argument.
+In particular, given that we are, according to RADM Sammut accept the pumpjet that was offered by France, and our submarine supposedly has a near identical hull diameter and  length  to the nuclear Barracuda, it bizarre that any adjustment at all of either the aft section of the submarine, or the pumpjet itself, ought to be contemplated.
+The major advantage that the Naval Group proposal \textit{ought} to have is that those potentially complicated engineering challenges can be safely and entirely avoided.
+Furthermore, the idea that the hull is what necessitates the tuning, as opposed to the speed profile, is almost incomprehensibly naive, as I pointed out in my evidence given to the Senate \ref{SENmorrison}.
+It is noteworthy that no officials from Naval Group have ever alluded to this 'tuning' argument, or any adjustments to be made to the pumpjet or the hull on account of one another.
+However, it seems pertinent here to give a reasonable discussion of the relevant physics nonetheless, in order that the public might fully assess the sort of argument being made by our most senior defence officials.
+
+It is true that the hull does indeed change the flow of incoming water to a propulsor.
+The area behind the vessel where the water is on being slightly dragged along with the ship (or the water is moving less quickly past the vessel, in the vessel's reference frame) is known as the wake.
+Since any aft-mounted propulsor has to operate in the wake-field, it is true that wake is of importance to the design of the propolsor.
+What is not true, however, is that this is grounds for disregarding open-water efficiency, or that it was neglected from consideration in the preparation of my report.
+
+In reality, the operation of a propulsor in the wake-field is an absolutely standard consideration for deriving propulsive efficiencies, as can be seen in any number of standard texts \parencite{Carlton2007}, \parencite{Mollard2011}, \parencite{MAN2017}, \parencite{lewis1988}.
+
+There is a ratio which is commonly discussed, relative rotative efficiency (denoted $\eta_R$), which has as it's precise purpose to describe the ratio between the efficiency of the propulsor in open water ($\eta_O$), and the efficiency of the propulsor in the wake of the hull ($\eta_B$).
+
+It is widely acknowledged (for example in all of references \parencite{Carlton2007}, \parencite{Mollard2011}, \parencite{MAN2017}, \parencite{lewis1988}) that relative rotative efficiency generally lies around 1, and does not vary greatly.
+For instance section 16.3.2, page 411 of reference \parencite{Mollard2011} says ``$\eta_R$ ranges typically from 0.95 to 1.05 and is often assumed to be unity for preliminary design purposes." \footnote{Observers may note that efficiencies are generally thought to range between 0 and 1.
+The possibility that $\eta_R$ could be greater than one is opened up by the fact that it is actually a \textit{relative} efficiency, one efficiency divided by another.
+Some propellers can actually work marginally better in the wake field than outside of it.}
+
+In clear contrast however, the open-water efficiency $\eta_O$ of a propulsor varies between 0\% and 75\%, depending crucially on the advance ratio $J$, which is highly speed dependent due to the changing resistance of the vessel at different speeds.
+In \parencite{morrison2018} it was shown that it was most probable a pumpjet's open-water efficiency could be around half that of a propeller at speeds at or below 5kt.
+It is consequently vastly more important to give detailed consideration to early in the design and selection of a propulsor, particularly for a diesel-electric submarine.
+Unless defence is proposing that a pumpjet can have a relative rotative efficiency of more than ~\%200, there is no feasibly means that this line of argument can sustain their dismissal of my conclusion.
+
+A particularly desparate line of argument might lead one to argue that such arguments do not apply to submarines propulsors due to their great level of depth and the unique shape of their hull.
+This would not be true.
+Owing to their generally cylindical shape and generally conical afterbody, submarines tend to have wake fields are much less complex and intense in the degree of their variation that surface ships.
+This is because ships bodies also have wave-making effects involved as well as a hull that tapers above the propulosor, creating an uneven effect with different heights.
+In contrast, wake fields of propellers have a relatively even gradient in wake fraction (the intensity of waterspeed difference due to the hull) in a radial direction, with only the control surfaces ahead of the propulsor causing substantial variations.
+Furthermore, at very low speeds, where pumpjets were shown to have the lowest efficiecies, the effect of the wake field would be the least compared to the open-water case, since the size of the boundary layer about the aft of the hull would be substantially diminished.
+On absolutely all accounts, the interaction between the hull of a submarine and propulsor at low speeds cannot be credibly argued to have any impact which would overturn the conclusion reached about pumpjets having clearly and substantially inferior efficiency at low speed, as argued in \parencite{morrison2018}.
+
+As such, the reason that my earlier paper did not explicitly discuss relative rotative efficiency and the impact of the wake field, this was due to it being so obviously and demonstrably of no bearing whatsoever on my conclusions, and I had assume that as much would immediately clear to any technically competent reviewer.
+
+The use of this argument displays a degree of ignorance of the physics, as clearly described in numerous and publically available sources, that it is difficult to account for by incompetence alone.
+A substantial measure of contempt for public intelligence must be required in order to appeal to such trivially simple statements of truth in order to support conclusions which can be so trivially demonstrated to be completely false.
+
+To use a simple analogy to, imagine that you came to buy a coffee from a your favourite cafe, only to find that they weren't selling any coffee that day, and wouldn't be selling any more indefinitely.
+Their coffee maching had been removed, and they were only selling cold-drip, coffee, stored in little bottles in the fridge.
+Overcoming initial shock (since you weren't aware of any other cafes who would make such a bizarre decision as to not sell espresso-coffee as a substantial part of their offering), you made some tentative inquiries as to why.
+At first, you're told vaguely that it was to do with health and safety.
+This seemed a little odd, since there obvious benefits to everyone's enjoyment and presumeably the business, which you point out to the cafe owner, and you ask how the health and safety risk could really be so large as to outweigh that.
+With just a little annoyance at having to explain, the cafe owner briefly says something about temperature being the problem.
+You look at the empty spot on the bench where the coffee machine used to be, and speculate that perhaps there really was some serious fault with the machine, maybe even a flaw in its design requireing it to be recalled.
+Perhaps it was running uncontrollably hot, and might have risked melting componentry, creating electrocution risk, or even a risk of explosion if the boiler rose to an elevated pressure.
+That would explain the safety.
+Or perhaps there was some emerging science about carcinogens being produced in the high-temperature extraction of espresso, which cold-drip might presumeably avoid?
+In anycase, it sounded like it could be very technical, and you decided to trust your very experienced and seemingly intelligent barista and forget about your lingering doubts as to why no other cafes were following suit.
+Then later you find a notice posted up at the front of the cafe, explaining fully the rationale behind the abandmonment of the coffee machine, and the connection with temperature.
+In it, it is explained that all espresso-based coffees were hot, those with milk around 65 or 70 degrees, and long blacks or fresh espresso potentially above 90 degrees.
+If consumed quickly or accidentally spilled such drinks could potentially cause burns.
+Therefore in the interest of health and safety no more hot drinks would be served, and cold-drip would be the only coffee available.
+
+You read the statement again, not quite believing that a hazard so obviously known and successfully navigated by millions of coffee drinkers every day, could be the cause of such a decision.
+Althought it's true the owner didn't lie about it being a health and safety concern, and can't dispute the facts about the coffee temperatures being above 65 degrees, you still conclude that owner is either insane, or trying to disguise some other reason for abandoning the espresso coffee.
 

### 2018-09-14 Second.Rnw
with first picture in
four  pics

@@
 Some propellers can actually work marginally better in the wake field than outside of it.}
+
+\begin{figure}[h]
+\captionsetup{width=0.7\textwidth}
+\includegraphics[width=0.7\textwidth, center]{roteff1.png}
+\caption{One text's explanations of the variance between open-water efficiency and behind a hull from page 152 of \parencite{lewis1988}}
+\label{fig:roteff1.png}
+\end{figure}
+
+\begin{figure}[h]
+\captionsetup{width=0.7\textwidth}
+\includegraphics[width=0.7\textwidth, center]{roteff2.png}
+\caption{One text's explanations of the variance between open-water efficiency and behind a hull from page 131 of \parencite{Carlton2007}}
+\label{fig:roteff2.png}
+\end{figure}
+
+\begin{figure}[h]
+\captionsetup{width=0.7\textwidth}
+\includegraphics[width=0.7\textwidth, center]{roteff3.png}
+\caption{One text's explanations of the variance between open-water efficiency and behind a hull from page 14 of \parencite{MAN2017}}
+\label{fig:roteff3.png}
+\end{figure}
+
+\begin{figure}[h]
+\captionsetup{width=0.7\textwidth}
+\includegraphics[width=0.7\textwidth, center]{roteff4.png}
+\caption{One text's explanations of the variance between open-water efficiency and behind a hull from page 411 of \parencite{Mollard2011}}
+\label{fig:roteff4.png}
+\end{figure}
 

### 2018-09-15 Second.Rnw
with some drag stuff

@@
 
+\section{The drag resistance of underwater bodies}
+It has been asserted a number of times, both in reference \parencite{kerr2018} and \parencite{defence2018} that it is only surface vessels who experience substantial changes in the drag they experience at different speeds, owing to the wave-making resistance that they encounter.
+
+\begin{quotation}
+``Since submerged submarines do not have to overcome wave resistance, the efficiency of a propulsor can be considered independent of speed..."\\ - \textit{Dr David Kershaw, Chief of Maritime DSTG} \parencite{kerr2018}
+\label{resistancequotekershaw}
+\end{quotation}
+
+\begin{quotation}
+``The efficiency of a propulsor, whether it is a pumpjet or propeller, for a submerged submarine is related to the total submerged resistance and is essentially constant over speed" \\ - \textit{Department of Defence} \parencite{defence2018}
+\label{resistancequotedefence}
+\end{quotation}
+
+\begin{quotation}
+``6.
+There are two primary components of resistance to forward motion for a surface ship, the underwater drag created by the hull form and the wave making resistance.
+The underwater drag is comprised mainly of frictional resistance and is a form of drag due to the viscosity of the water and the shape of the body.
+Also, when a vessel is on the surface it will generate surface waves.
+The wave-making resistance is the resistance caused by the generation of these surface waves and is a function of the speed.
+\\
+7.
+Deeply submerged submarines do not experience this wave-making resistance.
+This means that the propulsive efficiency for a submerged body will remain constant over the speed range; whereas the propulsive efficiency drops off for a surface ship as it moves away from the design speed.
+This is independent of the type of propulsor.
+Therefore, the propulsive efficiency of a pumpjet designed for a submarine does not vary significantly with the speed of the boat."  \\ - \textit{Department of Defence} \parencite{defence2018}
+\label{resistancequotedefence2}
+\end{quotation}
+
+These statements again use a seemingly relevant truth, in order to distract or obscure a subsequent falsehood.
+It is indeed true that surface ships do encounter wavemaking resistance, and that this varies substantially as a function of speed, and submarines do not encounter this  type of resistance.
+It is true that skin friction, (which is the shear component of the drag forces of any body that moves through a viscous fluid which act along its sides) as well as pressure resistance (the forces normal to the hull, which are non-cancelling on the nose and the tail) comprises the majority of the drag that applies to a fully submerged submarine.
+However, to claim that these forces are constant with speed is to claim the most bizarre and obtuse denial of the most basic and well-established physics in this respect.
+These forces scale with velocity squared.
+As such, speed plays the dominant role in considering the efficiency of propulsors on submerged bodies, since they also have their efficiency fall away at off-design conditions.
+
+It is simply not credible that the Department of Defence can genuinely hold such a utterly  naive belief that is so clearly contradicted by known experience and the most basic and abundant literature on this topic.
+Were it the case that the resistance faced by a fully-submerged body was ``essentially constant over speed" then a submarine or torpedo could reach any speed that it liked by applying the same force, and would have unlimited speed.
+This would mean that even a diesel submarine would also have unlimited range by virtue of it's ability to reach any speed  that it liked.
+This assertion  by defence is so utterly and blatantly ridiculous that there is no plausible way that it could be genuinely held to be a true by any one of even moderate technical competence.
+It is difficult to even believe that it could be held with any confidence by even moderately intelligent adults of little or no technical background.
+It is difficult to avoid reaching the conlcusion that the statement was given in a state of extreme contempt for the intelligence of the intended reader, as it constitutes without any doubt a deliberate lie and active attempt to mislead.
+
+Almost every available source or text makes clear that the drag in water generally scales with velocity squared, including submerged vessels.
+
+\begin{figure}[h]
+\captionsetup{width=0.7\textwidth}
+\includegraphics[width=0.7\textwidth, center]{drag1.png}
+\caption{Taken from the appendix of \textbf(Fundamentals of Submarine Concept  Design) by  CAPT Harry A.
+Jackson USN (Ret.) \parencite{jackson1992}.
+Note that the $V^2$ which indicates the relationship with velocity squared relationship}
+\label{fig:drag1.png}
+\end{figure}
+
 \section{The weight and size argument}
@@
 \includegraphics[width=0.7\textwidth, center]{roteff3.png}
-\caption{One text's explanations of the variance between open-water efficiency and behind a hull from page 14 of \parencite{MAN2017}}
+\caption{One text's desciption of the various types of efficiency from page 14 of \parencite{MAN2017}}
 \label{fig:roteff3.png}
+\end{figure}
+
+\begin{figure}[h]
+\captionsetup{width=0.7\textwidth}
+\includegraphics[width=0.7\textwidth, center]{roteff5.png}
+\caption{One text's explanations of the variance between open-water efficiency and behind a hull from page 16 of \parencite{MAN2017}}
+\label{fig:roteff5.png}
 \end{figure}

### 2018-09-17 Second.Rnw
big commit, getting close!

@@
 However, over a number of years now the pumpjet choice has attracted sustained attention and criticism, mostly around its efficiency at low speeds, where a diesel-electric submarine would necessarily spend a great deal of its operating time.
-I was commissioned by Gary Johnston to investigate this topic in detail and produce a report \parencite{morrison2018}, which concluded that a very substantial reduction in dived range and endurance (as well as overall range) would necessarily result from the pumpjet choice.
+I was commissioned by Gary Johnston of Submarines for Australia to investigate this topic in detail and produce a report \parencite{morrison2018}, which concluded that a very substantial reduction in dived range and endurance (as well as overall range) would necessarily result from the pumpjet choice.
 Furthermore, that paper challenged the assumption that the pumpjet would be acoustically superior at low speeds, where open propellers would have no cavitation, and cause much less turbulence than the equivalent jet.
@@
 I was also invited to provide evidence to the committee in a public hearing on the 7th of June 2018.
-In that public hearing, I stood by the conclusions of my substantive paper, which was: ``between two otherwise identical submarines, the one with the pump jet will always have a lower dived endurance, a lower dived range, a worse indiscretion ratio, a lower overall endurance and a lower overall range than the one with the propeller."  In the presence of both the outgoing acting CEO of Naval Group Australia Brent Clark, and General Manager of Submarines (for Defence) Stephen Johnson, I pointed out that that statement was yet to be challenged, and that I'd be very surprised if anyone with technical competence or professional integrity sought to challenge it.
+In that public hearing, I stood by the conclusion of my substantive paper, which was: ``between two otherwise identical submarines, the one with the pump jet will always have a lower dived endurance, a lower dived range, a worse indiscretion ratio, a lower overall endurance and a lower overall range than the one with the propeller."  In the presence of both the outgoing acting CEO of Naval Group Australia Brent Clark, and General Manager of Submarines (for Defence) Stephen Johnson, I pointed out that that statement was yet to be challenged, and that I'd be very surprised if anyone with technical competence or professional integrity sought to challenge it.
 
 In their own evidence during the course of the day, neither of those two did choose to challenge it.
-Senator Kim Carr requested that defence produce a written response to that report for the committee prior to their handing down of their finaly report for the Inquiry, and Defence agreed to do their best in the time available, with the head of Capability and Sustainment Group (CASG) Kim Gillis noting that ``we do hold that expertise", whilst making clear his reluctance to make a response clear, with comments such as "I regularly get correspondence from the public saying that almost every decision that we've made for the last 50 years on a defence matter is wrong, because somebody has an opinion".
-In the final report of the committee, this request, and defence's acceptance of it was noted, but no acknowledgement was made of whether any response was actually received, or what it contained \parencite{SERC2018final}.
-
-In the August edition of Australian Defence Magazine a major article was featured on the pumpjet question \parencite{kerr2018}, referencing both my testimony in the hearing and the paper.
+Senator Kim Carr requested that defence produce a written response to that report for the committee prior to their handing down of their finaly report for the Inquiry, and Defence agreed to do their best in the time available, with the head of Capability and Sustainment Group (CASG) Kim Gillis noting that ``we do hold that expertise", and Senator Carr emphasising that ``it would look pretty bloody stupid after the report's handed in''.
+Despite these exhortations for timeliness, in the final report of the committee, this request, and defence's acceptance of it was noted, but no acknowledgement was made of whether any response was actually received, or what it contained \parencite{SERC2018final}.
+
+In the August edition of Australian Defence Magazine a feature article on the titled `Future Submarine: Pumpjet propulsor technology choice still not well explained' \parencite{kerr2018} appeared, referencing both my testimony in the hearing and the paper.
 Comment for the article was sought from CASG, the Royal Australian Navy (RAN) and Naval Group, and none was forthcoming.
@@
 
-This paper has been commissioned by Gary Johnston of Submarines for Australia to provide a summary of the responses that have been received from official sources (in government, and Naval Group) to the earlier major report, and respond to them.
-Only two substantive arguments emerge seeking to engage with the technical arguments made in the earlier paper.
+This paper has been commissioned by Gary Johnston to provide a summary of the responses that have been received from official sources (in government, and Naval Group) to the earlier major report, and respond to them.
+At the time of its original commissioning, two substantive arguments emerge seeking to engage with the technical arguments made in the earlier paper.
 The first is that the large size of the submarine now makes the pumpjet viable, where as for smaller submarines the large weight of the pumpjet would make it unviable, and that the pumpjet would be generally well-suited to smaller sumbarines as well were it not for this weight consideration.
 The second is that interaction between the pumpjet and the hull obviate the evidence I have produced showing that pumpjet efficiency declines at low speeds, which are claimed to neglect considering such interactions.
-This paper seeks to consider those two lines of reasoning in sufficient detail to suitably discuss the circumstances in which such arguments might be made.
+
+In early September, the Department of Defence gave a six-page written response to my paper, a full three months after the initial request.
+The two substantive arguments already identified were repeated, but the weight of their argument actually came to rest on another assumption.
+This assumption was mentioned already by Dr Kershaw in the ADM article, but it was so completely at odds with the most basic physics that I was inclined to dismiss it as perhaps a clumsily worded statement made in an interview compounded by a little journalistic mis-quotation, rather than a substantive argument.
+The re-appearance of this argument, which was spelled out and repeated multiple times in a written document that was prepared by Defence over the course of three months, well after any nominal deadline for timeliness had elapsed, dispelled all illusions that the argument was included by accident.
+As such, I have also addressed it here as the first of three substantial arguments made against my earlier conclusions regarding pumpjet efficiency, which has quite significantly altered the conclusions of this paper, which seeks to consider all three lines of reasoning in sufficient detail to suitably discuss the circumstances in which such arguments might be made.
+
+This paper commences with a chronologically sequenced history of the most significant comments that have been made by public officials or Naval Group/DCNS with respect to the pumpjet, highlighting the most important implications of each in their context.
+These are followed by three sections, addressing in more detail the three substantive arguments against my conclusions.
+Some discussion developing the final conclusions follows, followed by appendices which include more substantial extracts or originals sources where appropriate, for reference.
 
@@
 Only brief quotations are selected here, with references being made to the full excerpts either in Appendices or References.
+Any emphasis in-text has been added.
 
@@
 This encounter shows that the viability of pumpjets was challenged some time ago.
-It is noteworthy at the time that RADM Moffit makes no attempt to defend the consideration, acknowledging even that it might be easy for them to "write it off".
+It is noteworthy at the time that RADM Moffit makes no attempt to defend the consideration, acknowledging even that it might be easy for them to ``write it off".
 It is also noteworthy the tone of confidence that Senator Johnston's challenge has, driving directly for an explanation of who had the original idea, which RADM Moffit was unable or unwilling to answer.
@@
 This rules out the likelihood that some special or unique pumpjet technology, better suited just to diesel-electric submarines, had been pioneered by France alone.
+This comment also directly implies that information regarding the performance of the pumpjet had already been given to the Australian  government, \textit{prior} to the conclusion of the Competitive Evaluation Process (CEP).
 
@@
 Consistent with the physics outlined in Reference \parencite{morrison2018} the acoustic advantage is only emphasised by DCNS as being significant at higher speeds.
-Whether it is viable for a diesel-electric submarine (as opposed to the nuclear one mentioned here) to maintain such speeds to close with many types of targets is left unclear.
+Whether it is viable for a diesel-electric submarine (as opposed to the nuclear one mentioned here) to maintain such speeds to close with many types of targets is left unclear, or at least unmentioned.
+As later discussed in my evidence to the Senate \ref{SENmorrison}, this is consistent with DCNS being well aware with the advantages and disadvantages of the pumpjet, whilst not making verifyably false claims.
 
 \begin{tcolorbox}[width=0.9\textwidth, center, colback=blue!5!white,colframe=blue!75!black, title= \textbf{May 2016 onwards - DCNS/Naval Group Website for Shortfin Barracuda}]
-\textit{The official website for the submarine claims the advantages of the pumpjet in specific terms, drawing comparisons with "obsolete propeller technology" and "otherwise identical submarines".
+\textit{The official website for the submarine claims the advantages of the pumpjet in specific terms, drawing comparisons with ``obsolete propeller technology" and ``otherwise identical submarines".
 Last accessed 4 Sept 2018 \parencite{naval2018}.
@@
 
-The official explanation of the designer and builders website makes a carefully worded explanation of the superiority of the pumpjet, by limiting the comparison to a "tactical" encounter with an "otherwise identical" submarine.
-As discussed in \ref{SENmorrison}, this is the only \textit{possible} target that an elevated speed such as could be achieved by a diesel-electric submarine might assist in closing with.
-The submarine would be at a substantial "strategic" disadvantage in such an encounter, since its lower dived endurance, dived range, and poorer indiscretion ratio dramatically increases the probability that the pumpjet propelled submarine would wind up being the hunted, rather than the hunter.
+The official explanation of the designer and builder's website makes a carefully worded explanation of the superiority of the pumpjet, by limiting the comparison to a ``tactical" encounter with an ``otherwise identical" submarine.
+As discussed in \ref{SENmorrison}, this is the only \textit{possible} target that an elevated speed - such as could be achieved by a diesel-electric submarine - might assist in closing with.
+The submarine would be at a substantial `strategic' disadvantage in such an encounter, since its lower dived endurance, dived range, and poorer indiscretion ratio dramatically increases the probability that the pumpjet propelled submarine would wind up being the hunted, rather than the hunter.
 More significantly, the speeds achieved would not allow the submarine to close with any other target from behind such as almost all surface ships, or nuclear submarines.
@@
 (Interview transcribed from Reference \parencite{lane2017}.)
-Journalist: Does the pumpjet technology that is on Suffren remain viable for the Australian submarine?
+
+\textbf{Journalist:} Does the pumpjet technology that is on Suffren remain viable for the Australian submarine?
 
@@
 This crucial concession reveals a very different level of confidence in the pumpjet being suitable for the Australian diesel-electric submarine.
-It also gives an important glimpse into the processes that both selected the winning bidder, as well as the processes for the selection of key technology components.
-By revealing that the pumpjet has not yet been firmly decided upon, and that the technology choice is a product of the operational requirement, it makes it clear that the operational requirement has not yet been determined with sufficient precision to make the suitable technology choice clear to a ship-builder with ample experience in both technologies.
-Moreover, it also reveals that this was also the case at the conlcusion of the CEP, and remained so well over a year after.
+It also gives an important glimpse into the processes that selected the winning bidder, and key technology components, which reveals a clear contradictions between what would have been previously foregone conclusions.
+By revealing that the pumpjet has not yet been firmly decided upon, it is directly implied that France's exclusive offer to transfer the pumpjet technology could not have been the real distinguishing feature that elevated that bid above the others.
+Simultaneously, the statement that technology choice is a product of the operational requirement makes it clearly implies that the operational requirement has not yet been determined with sufficient precision to make the suitable technology.
+As a ship-builder with ample experience in both technologies, uniquely positioned as having fitted both pumpjets and propellers to both diesel and nuclear submarines, it is remakeable that the performance information transferred prior or during the CEP was not sufficient to allow a decision  on the technology to be made.
+Moreover, it also reveals that this was not only the case at the conlcusion of the CEP, and remained so well over a year after.
 
@@
 \textbf{In a contest between two otherwise identical submarines, the one with the pump jet always has the tactical advantage.
-The propulsor was an integral part of our bid.} The Department of Defence says it intends to use the pump jet propulsion system."
+The propulsor was an integral part of our bid.
+The Department of Defence says it intends to use the pump jet propulsion system.}"
 \end{tcolorbox}
@@
 pumpjet) was ``integral" to their bid, whilst also deferring to what the Department of Defence ``intends" do use.
-This exposes a clear tension between how 'integral' the pumpjet was to the bid, and Defence's freedom to choose otherwise.
-It also exposes a dramatic divergence between the level of emphasis that is given to the pumpjet as an important key feature of the proposal in public promotion (as highlighted in \cite{lane2017}) and a Defence decision which seemed to choose only to have the option of including the pumpjet, with the real decision to be deferreed.
+This further exposes a clear tension between how 'integral' the pumpjet was to the bid, and Defence's freedom to choose otherwise.
+It also exposes a dramatic divergence between the level of emphasis that is given to the pumpjet as an important key feature of the proposal in public promotion (as highlighted in \cite{lane2017}) and a Defence decision which seemed to choose only to have the option of including the pumpjet, with the real decision to be deferred.
 
@@
 
-RADM goes to length to explain that the pumpjet was not ``alone" the reason for selecting DCNS as a partner, and in doing so reveals important insights into Defence's thinking and process regarding the decision.
-Crucially, he reveals that the 'diameter of the hull' is an important aspect of France's design reference in the Barracuda, pointing out that changing the diameter of the hull incurrs a ``whole new set of calculations about the structural integrity and strength of the hull."  Together with remarks about the ``way that safety is incorporated into the design" and concerns about the alignment of the certifying process with the ``design philosophy" it would appear that the hull diameter actually emerges as plausibly one of the most significant distinguishing features in the French bid, perhaps alongside the pumpjet.
+RADM Sammut goes to length to explain that the pumpjet was not ``alone" the reason for selecting DCNS as a partner, and in doing so reveals important insights into Defence's thinking and process regarding the decision.
+Crucially, he reveals that the `diameter of the hull' is an important aspect of France's design reference in the Barracuda, pointing out that changing the diameter of the hull incurrs a ``whole new set of calculations about the structural integrity and strength of the hull."  Together with remarks about the ``way that safety is incorporated into the design" and concerns about the alignment of the certifying process with the ``design philosophy" it would appear that the hull diameter actually emerges as plausibly one of the most significant distinguishing features in the French bid, perhaps alongside the pumpjet, and aligns with emphasis given to the exact diameter of the pumpjet remaining unchanged in media report earlier in the year.\parencite{kerr2017}
 
 This is alarming for two reasons.
-The first is that it contradicts subsequent explanations regarding pumpjet efficiency being incorrectly assessed by outsiders because they failed to account for how how the hull would be 'tuned' to the pumpjet and the pumpjet 'tuned' to the hull.
+The first is that it contradicts subsequent explanations regarding pumpjet efficiency being incorrectly assessed by outsiders because they failed to account for how how the hull would be `tuned' to the pumpjet and the pumpjet `tuned' to the hull.
 If the hull diameter is already fixed and known as being identical to that of the nuclear Barracuda, then no adjustment ought to be necessary to a good working pumpjet design that is used to propel this platform on the hull's account.
 The performance of the pumpjet on the relevant hull would be well understood by the manufacturer offering it, and presumeably available to the Australian government who were looking to acquire it.
+Further discussion of the hull-propulsor interaction is given in Section \ref{hullpropulsor}.
 
@@
 However, it is certainly not true that scaling up a pressure hull to a new diameter to meet the same fixed static load is amongst the most difficult, risky or complex.
-Whilst extending the length of a submarine is undoubtedly even easier, adjusting the diameter still to meet a structural integrity requirement to meet a known static pressure (from depth) is a trivially simple calculation in comparison to the complexities of, for instance, re-designing the duct intake of blade arrangments of a pumpjet to maximise efficiency at a different speed whilst simultaneously attempting to avoid cavitation.
-It was quite simply and successfully accomplished when the Collins Class submarine was developed by scaling up both the length and diameter of it's smaller predecessor.
-
-That this ranks as a significant consideration in the mind of the chief of this program should be the cause of considerable alarm to those taking an interest the project.
-Were it known to the competitors that their ability to achieve this trivial engineering requirement was in question, they would no-doubt be highly offended.
-Any contender that was not considered competent to design a pressure hull of a new diameter should never have been invited to participate.
+Whilst extending the length of a submarine is undoubtedly even easier, adjusting the diameter still to meet a structural integrity requirement to meet a known static pressure (driven by the dive depth) is a relatively simple calculation in comparison to the complexities of many other calculations involved in submarine design.
+For instance, re-designing the duct intake of blade arrangments of a pumpjet to maximise efficiency at a different speed whilst simultaneously attempting to avoid cavitation would be vastly more complex and uncertain given the unstable difficult-to-model dynamics of turbulence and cavitation involved.
+In contrast, scaling up a hull-diameter was quite simply and successfully accomplished when the Collins Class submarine was developed by scaling up both the length and diameter of it's smaller Swedish predecessor.
+In 2006, DSTO commissioned a report which was dedicated to proposition of a different hull diameter being used for submarine used to replace Collins \parencite{joubert2006} in order to achieve lower hydrodynamic drag.
+If such a development path was actually thought to impose  such significant challenges on the engineering and certification process as RADM Sammut's remarks imply, there is no way that in 2006 when the replacement was presumed to be undertaken for a smaller number of new submarines which would enter service in 2026 (as opposed to mid-2030s as currently planned) that such an undertaking would have been seriously considered.
+
+That this ranks as a significant consideration in the mind of the chief of this program should be the cause of considerable alarm to those taking an interest the project, as it would appear that he is quite alone in this belief that this really is a significant or risky hurdle in a project of this scale.
+Were it known to the competitors that their ability to achieve this simple engineering requirement was in question, one might assume they would be highly offended, since any contender that was not considered competent to design a pressure hull of a new diameter would presumeably never have been invited to participate.
 RADM Sammut should be called upon to clarify what he meant by ``we needed to make sure that the certifying process would match the design philosophy" because if indeed he meant that Defence had already decided that their own certifying process already precluded scaled-up diameter hulls, this represents a very bold departure from what might be considered normal risk assessment processes for major engineering projects, including those used in designing and certifying the Collins.
-Furthermore, if the advantage was considered so substantial, he should be called upon to substantiate how significant increased difficulty in certification would have been, with respect to cost and schedule.
-
-Amongst RADM Sammut's additional claims regarding the reason for choosing DCNS/Naval Group, are futher claims regarding the establishment of 'sovereign capability' and 'data transfer' to Australia.
-This has significant bearing on the way that subsequent comments made by Defence and DCNS regarding the state of their understanding of pumpjets, and the method for their development for the Australian submarine.
+It would also have essentially precluded both the Germans and Japanese from the outset.
+Furthermore, if the advantage was considered so substantial, he should be called upon to substantiate how significant increased difficulty in certification would have been, with respect to cost and schedule, noting that the French proposal had a larger cost and longer budget than either of the other two contenders.
+
+Amongst RADM Sammut's additional claims regarding the reason for choosing DCNS/Naval Group, include remarks regarding the establishment of 'sovereign capability' and 'data transfer' to Australia.
+One would assume that this meant that the French had demonstrated a willingness to supply complete and timely information on key technologies, and that RADM Sammut's exisiting experience with the French supplying such information on the pumpjet (as Sean Costello had claimed had taken place during and prior to the conclusion of the CEP) supported this view.
 
@@
 This comment confirms that the official policy is still to use the pumpjet, implying that the decision has indeed been made already.
-This is at odds with the statement made only weeks earlier by Jean-Michel Billig.
+This is at odds with the statement made only a few weeks earlier by Jean-Michel Billig.
 This is also the first noteable reference to the inter-relationship between the hull and the pumpjet, implying that the pumpjet will indeed be modified in order to suit the hull, and that the hull will be modified to suit the pumpjet.
-It also implies that the same pumpjet as France is fitting to the Suffren (i.e.
-nuclear barracuda) will not simply be inlcuded 'as is' on the Australian non-nuclear 'Shortfin' version.
-Given that the length and diameter of the hull were announced as being already decided, and seemingly very similar to the nuclear version, it raises the very serious question as to whether/why the changed shape of the hull would necessitate modification of the complex pumpjet.
-As pointed out in the authors testimony to the Senate \ref{SENmorrison}, the emphasis of this relationship, above the vastly different operational speed profile of the nuclear and non-nuclear submarines which would affect the propulsion system profoundly, raises serious questions about the origin of the comment.
+This contradicts the notion that the same pumpjet as France is fitting to the Suffren (i.e.
+nuclear barracuda) will simply be inlcuded 'as is' on the Australian non-nuclear 'Shortfin' version.
+Given that the length and diameter of the hull were announced as being already decided and practically identical to the nuclear version, it raises the very serious question as to whether/why the changed shape of the hull would necessitate modification of the complex pumpjet.
+As pointed out in the my testimony to the Senate \ref{SENmorrison}, the emphasis of this relationship, above the vastly different operational speed profile of the nuclear and non-nuclear submarines which would affect the propulsion system profoundly, raises serious questions about the origin of the comment.
 
@@
 \\
-'A [pump jet] can provide higher propulsive efficiency across the speed range of a submarine, including the patrol speed of conventional submarines'.
+`A [pump jet] can provide higher propulsive efficiency across the speed range of a submarine, including the patrol speed of conventional submarines'.
 
@@
 This testimony contradicts the assurance provided in an October interview with RADM Sammut.
-It is clear from multiple parts of Mr Johnson's statement that the decision on whether to use the pumpjet or not, has not yet been taken, since he is ``confident...
-that we are going to find this pump-jet approach to be the optimum...".
-Whilst acknowledging that he doesn't yet have the performance data from the nuclear Barracuda class (which is not yet in the water), he gives assurance that it will be obtained early enough to ``accommodate whateve is required", which he gives assurance is only going to be adjustments to the design of the pumpjet.
-This suggests that it is still not even known whether the pump-jet that was an 'integral' part of the bid was offered in an unmodified state, or whether it was known by France that it would need to be modified from existing designs, or whether what was offered was actually only a skill-set and capability for the design of a new pumpjet.
-
-In responding to the evidence produced by Senator Patrick that pumpjets may have an efficiency issue at low speeds, Mr Johnson refers to an 'electric shrouded pod' which is used on cruise ships, which he claims is a modification of a pumpjet.
+It is clear from multiple parts of Mr Johnson's statement that the decision on whether to use the pumpjet or not has not yet been taken, since he is ``confident...
+that we are \textit{going to find} this pump-jet approach to be the optimum...".
+Whilst acknowledging that he doesn't yet have the performance data from the nuclear Barracuda class (which is not yet in the water), he claims that it will be obtained early enough to ``accommodate whatever is required", which he gives assurance is only going to be adjustments to the design of the pumpjet.
+This suggests that it is still not even known whether the pump-jet that was an `integral' part of the bid was offered in an unmodified state, or whether it was known by France that it would need to be modified from existing designs, or whether what was offered was actually only a skill-set and capability for the design of a new pumpjet.
+
+In responding to the evidence produced by Senator Patrick that pumpjets may have an efficiency issue at low speeds, Mr Johnson refers to an `electric shrouded pod' which is used on cruise ships, which he claims is a modification of a pumpjet.
 Nothing about this section of his testimony seems to have any reference of connection to reality.
-Cross-ocean transit regularly occur at higher speeds than 17-18kt, as do cruise ships, which he mentions.
+Cross-ocean transits regularly occur at higher speeds than 17-18kt, including by cruise ships, which he mentions.
 These speeds are essentially unreacheable (or only extremely briefly sustainable) by a diesel electric submarine.
-I can find no evidence of any cruise ships having any kind of shround around the tips of their propellers, or bearing any resemblance or relationship to a pumpjet.
-As I pointed out in Mr Johnson's presence in my later testimony to the Senate on a later occasion (which he could not rebut or answer in any way), it appears that Mr Johnson may have confused electric Azipods, which are used on large ocean-going ships such as cruise ships, and the Navy's new LHDs, with pumpjets.
+As pointed out in my evidence to Senate Economic References Committee \ref{SENmorrison} I can find no evidence of any cruise ships having any kind of shround around the tips of their propellers, or bearing any resemblance or relationship to a pumpjet.
+In the same session, whilst in Mr Johnson's presence, it appears that Mr Johnson may have confused electric Azipods, which are used on large ocean-going ships such as cruise ships, and the Navy's new LHDs, with pumpjets.
 Azipods have open-tipped propellers, and are not remotely related to pumpjets.
-So far as I can discern, this particular part of the testimony was made-up entirely impromptu without any knowledge or care for the truth of the facts.
+So far as I can discern, this particular part of the testimony was made-up entirely impromptu without any knowledge or care for the truth of the matter.
 The degree of confidence that we can have in his assurance that the pump-jet will be the optimum approach for the submarine should be considered in light of this.
@@
 Initial inputs will be developed in the last quarter of 2018, which marks the beginning of the development and test program for the pump jet.
-Development, testing, and assurance will be conducted by several parties including Naval Group, Direction générale de l'armement (DGA), and DSTG.
+Development, testing, and assurance will be conducted by several parties including Naval Group, Direction generale de l'armement (DGA), and DSTG.
 Testing and assurance activities will take place in France (including at the premises of the DGA Techniques Hydrodynamiques in Val de Reuil) and Australia.
@@
 \end{tcolorbox}
+
+These responses represent another chance for the Department of Defence to articulate a considered opinion with regards to the merit of the pumpjet.
+On this occasion, the mention of efficiency is accompanied by no specific claim with regards to the plausible speeds at which this might apply.
+Instead, a mention of shaft torque (force per degree of rotation) requirements is made.
+Given that pumpjets are known to be smaller in diameter and rotate at a higher speed, the requirement of lower torque is a trivial observation.
+
+The statement that `inputs for the pump jet' will be developed in the last quarter of 2018 seem to imply that the pumpjet system is actually being designed ground-up, with its own `development and test program'.
+Given that the hull diameter is known to be exactly the same \parencite{kerr2017}, and the length barely altered from the nuclear Barracuda, that the pumpjet is being altered at all on account of the hull, as directly implied by RADM Sammut \parencite{sammut2017}, is extremely surprising.
+As discussed in my evidence to the Senate \ref{SENmorrison}, the prospect that a pumpjet even could be effectively `tuned' to match such a radically different operating profile with regards to speed is extremely odd.
+I likened it to being ``as if someone bought an efficient hybrid car like a Prius and then decided to fit it with tank tracks from a dozer or a battle tank, and then argued they would tune the tank tracks back to be as efficient as the Prius was in the first place.
+It goes radically against what the technology was intended for in the first place."
+
+Perhaps more remarkable is what this implies about the value (or eventuality) of pumpjet technology being transferred from France to Australia, and how that sits with statements made by RADM Sammut in particular.
+DCNS originally boasted about their intention to transfer the technology to Australia \parencite{costello2016}.
+RADM Sammut has since claimed that the way that ``data would be transferred to Australia" was amongst the reasons for his selection of DCNS as a design partner, whilst simultaneously citing the importance of the hull-diameter remaining identical to the nuclear submarine from which ours would be derived, in order to avoid ``a whole new set of calculations".
+Only days later, he refers the need for the jet to be ``tuned to the hull" and the hull would be ``tuned to the pump jet'.
+In contrast to the `tuning' statement, this answer describes a ground-up development process from inputs which are yet to be determined, which raises a serious question as what value there is in having a design partner who intends to transfer that technology from near-complete submarine with an identical diameter hull.
+One would assume that if such a ground-up development process were to be embarked upon which would necessarily have a feed-back loop with the final shaping of the hull, this would necessarily substantially heighten project risk.
+One would further assume that substantial work have been done to establish the suitability of the technology in advance, and the viability of reaching sufficient performance to meet the likely inputs in relatively tight development spiral with limited impacts on other systems or capabilities.
 
@@
 Answer: ``The efficiency of pump-jets designed for submarine use cannot be determined through comparisons with water-jets and ducted propellers, which are different in their operation from pump-jets.
-Moreover, the global efficiency of submarine pump-jets cannot be derived from ‘open water' tests of pump-jets, where the pump-jet is assessed independently of the hull of the submarine.
+Moreover, the global efficiency of submarine pump-jets cannot be derived from `open water' tests of pump-jets, where the pump-jet is assessed independently of the hull of the submarine.
 A correctly designed pump-jet, suited to the aft hull shape of the submarine, achieves better optimisation of efficiency and stealth than any propeller, even at low speeds."
@@
 This answer to a question again rests on the claim that the influence of the hull means that comparisons and studies such as those referred to in \parencite{morrison2018} can be ignored.
-This is a question which will be given a substantial reply in subsequent sections.
+This is a question which will be given a substantial reply in subsequent sections, but even a quick reading superficial reading reveals alarming levels of over-reach in the answer.
+(For intance, a type of ducted propeller known as a decelerating ducted propeller actually \textit{is} a pumpjet, by definition.
+An entire chapter (5) of the paper to which Defence is supposedly responding \parencite{morrison2018} discusses this at some length.)
 
 However, the response also avoids directly answering the question which specifically refers to efficiency in isolation.
-By only claiming that the pumpjet ``achieves better optimisation of efficiency \textit{and stealth}, even at low speeds, the response avoids directly contradicting the evidence that pumpjet efficiency inevitably declines dramatically at low speeds.
+By only claiming that the pumpjet ``achieves better optimisation of efficiency \textit{and stealth}, even at low speeds", the response avoids directly contradicting the evidence that pumpjet efficiency inevitably declines dramatically at low speeds, which is the core contention of the paper in question.
 
@@
 
-This constitutes point-blank refusal to answer the question that was asked.
+This constitutes a point-blank refusal to answer the question that was asked.
 Given the earlier carefully worded comments of the most senior Naval Group officials, and careful placement of ultimate responsibility for the choice with the Commonwealth, such a refusal to answer a question is alarming.
-If the opinion that the pump-jet was overall well suited to diesel-electric submarines was held by Naval Group, it would be only natural to expect that that would be confidently communicated to the Department of Defence, given that they had assessed their bid which it was 'integral' to.
+If the opinion that the pump-jet was overall well suited to diesel-electric submarines was held by Naval Group, it would be only natural to expect that that would be confidently communicated to the Department of Defence, given that they had assessed their bid which it was `integral' to.
 Provided that opinion was communicated and hadn't changed, it would be easy and expected that Defence would give a confident answer to the question.
-The possibility that Naval Group offered a proposal which included technologies were known to be poorly suited together would constitute a breach of trust, and possibly fraud.
-
-Overall, this response casts serious doubt on the level of communication and advice which has flowed between Naval Group and the Department of Defence regarding the pumpjet to date.
-As a ship-builder which had built both pumpjets and propellers, to both diesel-electric as well as nuclear submarines, there could be no excuse for Naval Group to not have a well-informed opinion on the matter.
-For Defence not to have obtained that opinion from DCNS at the time of considering the bid would be a matter of incomprehensible oversight, and raises the question of whether deliberate neglect may have played a role.
+If Naval Group offered a proposal which included technologies were known to be poorly suited together would constitute at the minimum a severe breach of trust, and quite probably fraud.
+
+Overall, this response casts serious doubt on the type of communication and advice which has flowed between Naval Group and the Department of Defence regarding the pumpjet to date.
+As a ship-builder which had designed and manufactured both pumpjets and propellers, and fitted them to both diesel-electric as well as nuclear submarines, there could be no excuse for Naval Group to not have a well-informed opinion on this matter.
+For Defence not to have obtained that opinion from DCNS at the time of considering the bid would be a matter of incomprehensible oversight.
+It is made all the more difficult to believe given the scrutiny applied to the matter in the Senate in 2012 \ref{SENmoffit}, and the program director's assurance that the work was being done to assess its suitability.
+
+Naval Group certainly said that the performance of the pumpjet was known to the Government prior to the close of the CEP \parencite{costello2016}.
+It is a matter of substantial public interest to determine whether that performance information given at the time was sufficient to communicate Naval Group's opinion on the suitability of the pumpjet to a diesel-electric submarine, or allow Defence to confidently form their own opinion based on this information.
+If Defence came away with a positive opinion of the pumpjet's suitability whilst Naval Group actually privately maintained a negative one, (and elected not to avail Defence of their false impression for commercial reasons) this would amount to a gross betrayal of trust and quite possibly fraud on the part of the Naval Group and the French Government.
+It would also demonstrate an catastrophic level of incompetence on the part of Defence, given their stated efforts to assess the pumpjet prior to this point.
+If Defence has since become aware of Naval Group's lack of confidence in the pump jet's suitability, this may explain the defence response to this question, which explicitly refers to the pump jet propulsion being offered `as part of their Competitive Evaluation Proposal' and `remains on offer to Australia', which could be read to deflect responsibility for technology choice to Naval Group, implying they would or ought to withdraw the offer if it was now known to be inappropriate.
+
+In such a scenario Defence is also acting to shelter Naval Group from accusations of fraud as well as the likely legal proceedings that would be open to the Commonwealth if information provided in the CEP was found to be fraudulent.
+It's probable that this constitutes misleading parliament, but of far greater concern is the possibility that the necessary corrective action to avoid a procurement disaster of unprecedent scale is not yet being taken by Defence.
+Why that could be the case is also a matter of substantial public interest.
+
+It is possible to believe that Defence's own embarrassment at having been unable to detect the insincerity in Naval Group's offer of the pumpjet, and inability to reach their own technical conclusions quickly enough to detect the fraud, may motivate the present actions.
+This amount to a disasterous failure on the part of the public service to act with integrity and in the national interest.
+However, if Defence had truely been un-wittingly deceived, it is difficult to see how they could not recover their pride (and regain a necessary measure of respect from their build partner) by quickly adopting a more combative stance, calling out the deceptions on the part contractors part, and quickly exacting whatever contractual reparations they might be entitled to.
+If the contractual reparations were not sufficiently strong, the Commonwealth's freedom to not proceed to a build contract (from the initial design contract) or re-introduce competition into the design process as many have suggested should occur in any case \parencite{stanford2017}, would have provided ample commercial leverage to reset the relationship in a manner which was better aligned with the Commonwealth's interest.
+
+One must at this point give consideration to the possibly that Defence had not been entirely unwittingly deceived, and was actually complicit in their own deception, and it's present upkeep.
+This would mean that the Department of Defence and the Navy, or at least a few of very senior people within those organisations, had to conspired allow a proposal which was known to be inappropriate to the Commonwealth's needs to appear victorious in the CEP,  presumeably in order to favour a bidder for whom no more defensible explanation could be found.
+It appears unlikely that there could be any decent motive for such a course of action.
+However, one must evaluate the probabilities of the necessary oversights for such an eventuatlity occurring by accident or by arrangement.
+If the pumpjet was known by DCNS to be an unsuitable proposition for a diesel-electric submarine, a seemingly unlikely degree of naive incompetence would be required for Defence to fail to ascertain as much during the Competitive Evaluation Process.
+To remain genuinely unwitting for two subsequent years of design development would seem even more so.
+The possibility of some corruption influencing the selection of Australia's \$50bn future submarine contract must at least be considered.
 
@@
 
-This exchange demonstrates a clearly contemptuous attitude towards the evidence I had presented, which is clearly pervasive and supported in the higest levels of the defence department.
+This exchange demonstrates a dismissive attitude towards the evidence I had presented, which borders on contempt.
+The attitude is clearly pervasive and supported in the higest levels of the defence department.
 The General Manager Submarines, Stephen Johnson, claims he simply cannot recall what he said about pumpjets being related to pods on cruise ships, despite being physically present in the same room when I pointedly quoted and contradicted him.
@@
 It's akin to arguing 'camels have to cary two heavy humps on their backs.
-Horses don't have to carry these two heavy humps, which is why they can fly.'
+Horses don't have to carry these two heavy humps, which is why they can fly.'  It also directly contradicts the claim made by Stephen Johnson before the Senate, that ``No matter what kind of propeller or propulsion pump jet  you design, it has to be optimised for a particular speed.".
 
@@
 
-Apart from it being quite an elementary law of phyiscs (the Drag Law), referred to and relied upon in posts such those by Andrew Davies \parencite{davies2017}, any substantial reading of the literature focussed on the design of pumpjets for fully submerged objects (such as torpedoes or submarines) will also make it abundantly clear, and the special distinction with wave-making resistance (which is crucial for fast-vessels in the process of reaching the plane) is also abundantly clear, and has been well known for a long time \parencite{wislicenus1973} \parencite{henderson1964}.
+Apart from it being quite an elementary law of phyiscs (the Drag Law), referred to and relied upon in posts such those by Andrew Davies \parencite{davies2017} commenting on this exact question, any substantial reading of the literature focussed on the design of pumpjets for fully submerged objects (such as torpedoes or submarines) will also make it abundantly clear, and the special distinction with wave-making resistance (which is crucial for fast-vessels in the process of reaching the plane) is also abundantly clear, and has been well known for a long time \parencite{wislicenus1973} \parencite{henderson1964}.
 
@@
 
-These series of comments by RADM Sammut are consistent with other comments of his own and other government sources, and wholly inconsitent with a competent and sound engineering and design process.
+These series of comments by RADM Sammut are consistent with other comments of his own and other government sources, and wholly inconsistent with a competent and sound engineering and design process.
 
@@
 That such an approach to project management could be taken on a project like this is almost unthinkable from an engineering and risk-management perspective, not to mention cost-effectiveness.
-
-However, such a possibility throws into sharp relieve some tensions between statements made by the government and Naval Group on multiple occasions.
+However, in light of the evidence available, the third possibility may warrant quite serious consideration.
+
+Such a possibility throws into sharp relieve some tensions between statements made by the government and Naval Group on multiple occasions.
 RADM Sammut had previously stated that ``We did not specify that the boat should have or not have air-independent propulsion or lithium-ion batteries, rather that it met our requirements." \parencite{sammut2017}.
@@
 
-If it was the case that the requirements had not been determined, then is likely that the Competitive Evaluation Process was entirely fraudulent.
+If it was the case that the requirements had not been determined, then is likely that the Competitive Evaluation Process was essentially fraudulent.
 If all the competitors weren't actually given the same performance requirements, which were the real requirements for the submarine which the government intended to procure, then it is extremely difficult to see how they could be `treated fairly' as RADM Sammut claimed \ref{SENhanson}.
@@
 It is for this reason that in submarines, lithium-ion batteries are acknowledged as having the most significant improvement over lead-acid at high speeds \parencite{briggs2016}.
-Given that pumpjets advantages are widely acknowledged to be at higher speeds, including implicitly by Naval Group\parencite{autret2016}, lithium-ion batteries should really be considered an enabling technology for pumpjets on diesel-electric submarines, at least if the case is to be made that priorisation of that extremely high-speed sprint capability should be made above all other performance characteristics, which I have argued is an obscure and implausible case to be made \ref{SENmorrison}  Without them, the pumpjet efficiency will sap performance at low speeds, and at high speeds (where pumpjet efficiency improves, albeit not as high as propellers) the battery efficiency will hobble performance.
-This effect, not accounted for in the modelling in reference \parencite{morrison2018}, will probably result in the maximum endurance of a pumpjet-submarine in a sprint at 20kt from over 2 hours (as that modelling suggested) to well under one hour.
+Given that pumpjets advantages are widely acknowledged to be at higher speeds, including implicitly by Naval Group\parencite{autret2016}, lithium-ion batteries should really be considered an enabling technology for pumpjets on diesel-electric submarines.
+(At least if the case is to be made that priorisation of that extremely high-speed sprint capability should be made above all other performance characteristics, which I have argued is the only technically defensible case to be made, whilst is an obscure and implausible case to be made in terms of useful operational capability \ref{SENmorrison})  Without lithium-ion batteries, the pumpjet efficiency will sap performance at low speeds, and at high speeds (where pumpjet efficiency improves, albeit not as high as propellers) the battery efficiency will hobble performance.
+This effect, not accounted for in the modelling in reference \parencite{morrison2018}, will probably result in reducing the maximum endurance of a pumpjet-submarine in a sprint at 20kt from over 2 hours (as that modelling suggested) to well under one hour.
 With lithium ion batteries, it is plausible that around 2 hours at this speed might be sustained.
-The exclusion of lithium batteries \textit{because} of the pumpjet's inclusion further indicates that RADM Sammut remains completely unaware of such interactions between technologies.
+The exclusion of lithium batteries \textit{because} of the pumpjet's inclusion further indicates that RADM Sammut remains, or is pretending to remain, completely unaware of such basic interactions between technologies.
+
+\begin{tcolorbox}[width=0.9\textwidth, center, colback=black!5!white,colframe=black!75!black, title= \textbf{Sept 2018 - Department of Defence, Report for the Senate Economic References Committee}]
+\textit{After being requested to produce a response to my paper (Reference \parencite{morrison2018}) by the Senate Economics References Committee in a public hearing on 7 June 2018, Defence produces a six-page report including the following.
+It should be noted that it was explicitly requested by Senator Carr that this report be produced \text{it} to the committee's final report from the inquiry the Future of Naval Ship Building Industry on the 27th of June.
+Instead of the three weeks requested, Defence took about three months to deliver this report to the Senate Committee.
+The full response is in }
+\\
+\\
+\begin{itemize}
+\item The Department maintains the view that this conclusion is incorrect in the case of submerged submarine for the following reasons:
+\begin{enumerate}
+\item The efficiency of a propulsor, whether it is a pumpjet or propeller, for a submerged submarine \textbf{is related to the total submerged resistance and is essentially constant over speed}, noting the submerged resistance of a submarine is dependent on the total skin friction and the shape of the submarine.
+\item The efficiency of a propulsor, again whether it is a pump jet or propeller, for a vessel at the surface is related to the total underwater drag submerged resistance plus the wave making resistance, which means efficiency varies with speed for a surfaced vessel.
+A propulsor for a surface vessel will have been designed for a specific speed, with propulsor efficiency falling away at off-design speeds.
+\item The efficiency of a well-designed submarine pumpjet will be similar to, or better than, a well-designed submarine propeller across the range of operating speeds.
+\item Propulsors are designed for specific purposes. Pumpjets offer specific advantages for submarines, especially with respects to the acoustic signature.
+\end{enumerate}
+\item There are two primary components of resistance to forward motion for a surface ship, the underwater drag created by the hull form and the wave making resistance.
+The underwater drag is comprised mainly of frictional resistance and is a form of drag due to the viscosity of the water and the shape of the body.
+Also, when a vessel is on the surface it will generate surface waves.
+The wave-making resistance is the resistance caused by the generation of these surface waves and is a function of the speed.
+\item \textbf{Deeply submerged submarines do not experience this wave-making resistance.
+This means that the propulsive efficiency for a submerged body will remain constant over the speed range; whereas the propulsive efficiency drops off for a surface ship as it moves away from the design speed.} This is independent of the type of propulsor.
+Therefore, the propulsive efficiency of a pumpjet designed for a submarine does not vary significantly with the speed of the boat.
+\end{itemize}
+...
+\\
+\\
+An accelerating duct will generate forward thrust, especially at low speeds, hence its use on ducted propellers on surface ships operating at low speeds (tugs and trawlers).
+Decelerating ducts on the other hand reduces internal velocity, thus improving the cavitation performance.
+For a pumpjet, the duct and the stators contribute to improve the regularity of the wake around the aft part of the submarine .
+\textbf{For a submarine, an accelerating duct is attractive} as it generates additional forward thrust, further increasing efficiency.
+
+\end{tcolorbox}
+
+The assertion that the resistance of a submerged body is constant with speed is such an easily demonstrable falsehood (as discussed in more detail in the subsequent section) that one can quickly dismiss the core of their argument against my conclusion, which depends on this false assumption.
+The competence or intention of the authors is further exacerbated by numerous additional remarks.
+Perhaps noteworthy is the bizarre suggestion that an accelerating duct might be desirable for a submarine.
+As was discussed in some length in my paper, accelerating ducts have the effect of reducing the static pressure of the water at the point that they reach the blades, which actually increases the amount of cavitation, and makes them very noisy.
+They are never considered to be remotely desirable for military submarines.
+It would appear that the author didn't even read my paper, nor any other relevant literature (for instance Reference \parencite{jackson1992}), or was simply intending to confuse and mislead an entirely non-technical audience.
+
+\begin{figure}[h]
+\captionsetup{width=0.7\textwidth}
+\includegraphics[width=0.7\textwidth, center]{kortvspump.png}
+\caption{A diagram taken from \underline(Fundamentals of Submarine Concept  Design) by  CAPT Harry A.
+Jackson USN (Ret.) \parencite{jackson1992}, illustrating the comparison between an accelerating ducted propeller (Kort Nozzle) and a pumpjet (which is a decelerating ducted propeller).}
+\label{fig:drag1.png}
+\end{figure}
 
@@
 It is indeed true that surface ships do encounter wavemaking resistance, and that this varies substantially as a function of speed, and submarines do not encounter this  type of resistance.
-It is true that skin friction, (which is the shear component of the drag forces of any body that moves through a viscous fluid which act along its sides) as well as pressure resistance (the forces normal to the hull, which are non-cancelling on the nose and the tail) comprises the majority of the drag that applies to a fully submerged submarine.
+It is true that skin friction, (which is the shear component of the drag forces of any body that moves through a viscous fluid which act along its sides) as well as pressure resistance (the forces normal to the hull, which are non-cancelling on the nose and the tail) comprises essentially all of the drag that applies to a fully submerged submarine, as shown in Figure \ref{fig:drag4.png}.
 However, to claim that these forces are constant with speed is to claim the most bizarre and obtuse denial of the most basic and well-established physics in this respect.
-These forces scale with velocity squared.
+These forces scale with velocity squared, a fact which has been widely known and accepted since Froude conducted his skin-friction tests in the 1870s.
+\footnote{It should be noted that our understanding of skin friction has improved somewhat since the 1870s.
+In particular, the distiction between laminar and turbulent flows and the transition between these two regimes as demonstrated by Osborne Reynolds in the 1880s was significant, since it was discovered that laminar flows exerted a lower skin friction, and it is the turbulent flows which scale with velocity squared.
+However, whilst some effort is made to attmept to achieve some laminar flow over bow-mounted sensors in submarines at certain speeds, it is widely accepted that such a large vessel as a submarine is almost entirely enveloped by turbulent flows at essentially all operating speeds, and consequently the $V^2$ relationship has been essentially universally accepted as the dominant relationship for this purpose, including for the caluclation of scalar drag coefficients (which represent the impact of the shape and smoothness of the hull) to which much design effort and literature is devoted.}  The results of those tests are shown in Figure \ref{fig:dragfroude.png}.
 As such, speed plays the dominant role in considering the efficiency of propulsors on submerged bodies, since they also have their efficiency fall away at off-design conditions.
@@
 \captionsetup{width=0.7\textwidth}
+\includegraphics[width=0.7\textwidth, center]{drag4.png}
+\caption{A breakdown of the components of resistance taken from Reference \parencite{Mollard2011} does indeed reveal that deeply submerged submarines encounter no wave-making pressure, but still encounter frictional resistance (which scales with velocity squared).}
+\label{fig:drag4.png}
+\end{figure}
+
+\begin{figure}[h]
+\captionsetup{width=0.7\textwidth}
+\includegraphics[width=0.7\textwidth, center]{dragfroude.png}
+\caption{A table showing the results of Froude's original experiments used to measure skin friction of various surfaces in the 1870s, taken from reference \parencite{lewis1988}.
+Note that the exponent of velocity is determined consistently to be about 2.
+}
+\label{fig:dragfroude.png}
+\end{figure}
+
+\begin{figure}[h]
+\captionsetup{width=0.7\textwidth}
 \includegraphics[width=0.7\textwidth, center]{drag1.png}
-\caption{Taken from the appendix of \textbf(Fundamentals of Submarine Concept  Design) by  CAPT Harry A.
+\caption{Taken from the appendix of \underline(Fundamentals of Submarine Concept  Design) by  CAPT Harry A.
 Jackson USN (Ret.) \parencite{jackson1992}.
@@
 \label{fig:drag1.png}
+\end{figure}
+
+\begin{figure}[h]
+\captionsetup{width=0.7\textwidth}
+\includegraphics[width=0.7\textwidth, center]{drag3.png}
+\caption{Taken from Research Paper commissioned by Department of Navy (US) titled \underline{Elements of the drag of underwater bodies} \parencite{granville1976}.
+Note that the $U^2$ which indicates the relationship with velocity squared relationship}
+\label{fig:drag3.png}
+\end{figure}
+
+\begin{figure}[h]
+\captionsetup{width=0.7\textwidth}
+\includegraphics[width=0.7\textwidth, center]{drag2.png}
+\caption{Taken from conference paper by Professor Volker Bertram titled \underline{Submarine Hull Design} \parencite{bertram2012}, same equation can also be found in published textbook by same author \parencite{bertram2000}.
+Note that the $V^2$ which indicates the relationship with velocity squared relationship}
+\label{fig:drag2.png}
 \end{figure}
@@
 
-\section{The interaction between the hull and the propulsor}
+\section{The interaction between the hull and the propulsor}\label{hullpropulsor}
+
 The argument that the 'hull needs to be tuned to pumpjet' and \textit{vice versa} has been used to dismiss critcism of the pumpjet repeatedly by RADM Sammut \parencite{sammut2017} \parencite{bergmann2018}, and also by Dr David Kershaw \parencite{kerr2018}.
@@
 As has already been noted, there are glaring contradictions from a project-management perspective which oppose this argument.
-In particular, given that we are, according to RADM Sammut accept the pumpjet that was offered by France, and our submarine supposedly has a near identical hull diameter and  length  to the nuclear Barracuda, it bizarre that any adjustment at all of either the aft section of the submarine, or the pumpjet itself, ought to be contemplated.
+In particular, given that we are, according to RADM Sammut accept the pumpjet that was offered by France, and our submarine supposedly has a near identical hull diameter and  length  to the nuclear Barracuda, it is peculiar that any adjustment at all of either the aft section of the submarine, or the pumpjet itself, ought to be contemplated.
 The major advantage that the Naval Group proposal \textit{ought} to have is that those potentially complicated engineering challenges can be safely and entirely avoided.
@@
 
+\subsection{The impact of the wake-field}
+
 It is true that the hull does indeed change the flow of incoming water to a propulsor.
 The area behind the vessel where the water is on being slightly dragged along with the ship (or the water is moving less quickly past the vessel, in the vessel's reference frame) is known as the wake.
-Since any aft-mounted propulsor has to operate in the wake-field, it is true that wake is of importance to the design of the propolsor.
+Since any aft-mounted propulsor has to operate in the wake, or wake-field as it is often called, it is true that wake is of importance to the design of the propolsor.
 What is not true, however, is that this is grounds for disregarding open-water efficiency, or that it was neglected from consideration in the preparation of my report.
@@
 In reality, the operation of a propulsor in the wake-field is an absolutely standard consideration for deriving propulsive efficiencies, as can be seen in any number of standard texts \parencite{Carlton2007}, \parencite{Mollard2011}, \parencite{MAN2017}, \parencite{lewis1988}.
-
 There is a ratio which is commonly discussed, relative rotative efficiency (denoted $\eta_R$), which has as it's precise purpose to describe the ratio between the efficiency of the propulsor in open water ($\eta_O$), and the efficiency of the propulsor in the wake of the hull ($\eta_B$).
@@
 The possibility that $\eta_R$ could be greater than one is opened up by the fact that it is actually a \textit{relative} efficiency, one efficiency divided by another.
-Some propellers can actually work marginally better in the wake field than outside of it.}
+Some propellers can actually work marginally better in the wake field than outside of it.}  The argument that the efficiency of pumpjets cannot be  assessed using their open-water efficiency is also blatantly false.
+The 'open-water' condition is can be usefully thought of as the normal or 'ideal' efficiency of the propulsor, when the stream of incoming water is undisturbed, and consequently is still a very functional maximum efficiency of a system who's purpose is to smooth out the inconsistency in the flow.
+No higher efficiency could be reasonably expected to be achieved through such  smoothing than would be acheived by beginning with a smooth stream in the first place.
+Consequently, one can have considerable confidence that the extent to which this 'smoothing' effect might assist a pumpjet, relative to a propeller, is bounded by the extent to which it actually impedes a propellers performance.
+As is amply demonstrated by abundant sources, this is a very small extent in all cases, and may actually be to the slight advantage of a propeller in some cases  where a desirable pre-swirl already exists.
 
@@
 
-In clear contrast however, the open-water efficiency $\eta_O$ of a propulsor varies between 0\% and 75\%, depending crucially on the advance ratio $J$, which is highly speed dependent due to the changing resistance of the vessel at different speeds.
+\begin{figure}[h]
+\captionsetup{width=0.7\textwidth}
+\includegraphics[width=0.7\textwidth, center]{roteff6.png}
+\caption{A paper from named \underline{Hydrodynamic Support in the Design of Submarines} also uses the same equations and relationships such as relative rotative efficiency to describe propulsion efficiency as that used for ships also uses \parencite{terwisga1988}}
+\label{fig:roteff6.png}
+\end{figure}
+
+\subsection{Turbulence and Flow Separation Re-Capped}
+It is noteworthy that absolutely no effort is made to engage in a discussion of turbulent or laminar in the boundary layer, nor does Defence make any mention of `flow separation' which are all important to the degradation of pumpjet efficiency at very low speeds.
+A basic understanding of these concepts would be irreconcilable with the view that resistance underwater is constant with respect to speed.
+In Reference \parencite{morrison2018} there is a substantial description  of these effects, particularly in Appendix 3 and Section 5, which have been apparently ignored by the Department of Defence.
+Some basic re-capping is suitable here.
+
+When any solid surface moves through a fluid (or \textit{vice versa}) there is always some boundary adjacent to surface, where the the fluid is actually non-stationary, and moves along partly with the solid.
+This is intuitively named the `boundary layer'.
+There are two different types of boundary layer.
+A \textit{laminar} boundary layer has all the layers of the fluid in the boundary moving in parallel, with the speed of each layer reducing further away from the surface.
+Such  a type of flow tends to only happen over very small distances, or at very low speeds, or in very viscous fluids.
+For the majority of the surface consideration of a submarine, the boundary layer is actually turbulent, which means that water in the boundary layer is tumbling and turning in a confused and unpredictable pattern.
+Whilst \textit{on average} the boundary layer moves closer to the surface speed closer to it, and slower further away from it, at any local point on a tiny lengthscale the movement of the water will be chaotic, and altogether very non-parallel.
+Turbulent boundary layers tend to be much thicker, and become even thicker at higher velocities, resulting in larger amounts of energy being sapped producing the faster, more intense chaotic swirling motions that comprise turblent flows.
+
+This distinction is important in understanding the limitations of a pumpjets claim to make the incoming stream of water less inconsistent, or more regular.
+The duct intake, and any stator blades in advance of the rotor can be effective in making the flow more homogenous, but in doing so they simultaneously make the flow more turbulent, since the necessarily involve passing the water over more surfaces, rather than less.
+This can be advantageous to the extent that very large-scale and intense vortices might be broken down into smaller-scale vortices, which appears more homogenous when assessed on the length-scale relevant to the blades of the rotors.
+This could indeed have the effect of reducing the likelihood of very large pressure fluctuations impacting the blades, reducing the probability of cavitation occurring or invoking some kind of resonant beats within the propulsor.
+\footnote{It is this effect that is likely referred to in comments made by Professor Joubert in a paper commissioned by DSTO in 2004 \parencite{joubert2004}.
+However, with competently designed appendages and fairings, the kind of unsteady vortex shedding which would see such large-scale eddies enter the propulsor should be safely avoided.
+Substantial effort has been devoted to developing the smoothest possible wake-field for submarines for this purpose.
+\parencite{toxopeus2014}}
+
+However, it must be noted that the in making the flow more homogenous, the flow is also made more homogenously turbulent, and the total level of energy which is wasted to  small-scale chaotic turbulent movements necessarily increases, as it necessarily must to with any additional surface.
+The circumstances where a pumpjet is able to attain higher efficiency are those where the total amount of additional productive energy (thrust, produced by front to back acceleration) added to the water column dramatically exceeds the losses incurred by the additional turbulence (round and round acceleration).
+This occurs at high speeds.
+
+It should be noted that a pumpjet's duct also has the effect of decelerating the incoming flow, which increases its static pressure.\footnote{In fact, a pumpjet is by definition a decelerating ducted propeller, contrary to claim that they are so different as to be incomparable made \ref{QON1}.} Whilst this has the desirable effect of further reducing the likelyhood of cavitation occurring, it also creates an `adverse pressure gradient' in the intake.
+This means that the water must move towards an area of higher pressure, agains the direction that the pressure would naturally push it.
+Whilst at high speeds the water has enough momentum to easily overcome this resistance, at very low speeds there isn't enough energy in the flow for the entire water column to move consistently in this direction, and part of the flow simply turns around and moves in the opposite direction, in a phenomenon known as `flow separation'.
+This occurs against a surface (in this case just inside the pumpjet intake), where the boundary layer has to move more slowly than the remainder of the stream, and in  this case it is the first part to reverse direction.
+Whilst the entirety of the boundary layer, (or indeed the entire flow) may actually be turbulent, with many very small-scale vortices, this phenomenon creates new large-scale macroscopic vortices, which could form, grow and shed in an unsteady creating exactly the same risk of perodic shedding (i.e.
+beats) impacting the rotor that a poorly designed mast or appendage might have.\footnote{A cylindrical mast, for instance, could set up a Karman Vortex Street at certain speeds.} Whilst it might be presumed the shedding of vortices created by flow separation into the pumpjet at low speeds has been adequately managed or mitigated to the extent that it does not risk the creation of cavitation on the blades, it would most certainly lead to the very substantial loss of efficiency through creation of these vortices.
+Since open propellers do not have a duct which creates an adverse pressure gradient, they do not face this problem at low speeds.
+In fact, at very low speeds more substantial parts of the blade may actually have laminar flow over them, even further enhancing efficiency.
+
+\begin{figure}[h]
+
+\begin{subfigure}{0.5\textwidth}
+\includegraphics[width=0.9\linewidth]{HiFlowSmall.png}
+\caption{}
+\end{subfigure}
+\begin{subfigure}{0.5\textwidth}
+\includegraphics[width=0.9\linewidth]{LoFlowSmall.png}
+\caption{}
+\end{subfigure}
+
+\caption{(a) At high speeds, the intake of a pumpjet slows the waterflow and raises its pressure, and it subsequently accelerates through the contracting nozzle to form an efficient jet (b) At low speeds, the pressure gradient created by the duct causes some of the waterflow to stop and reverse.
+This `flow separation' wastes energy spinning water in circles, and dramatically reduces the system's efficiency.}
+\label{fig:HiLoSmall.png}
+\end{figure}
+
+In clear contrast to the trivial changes that the relative rotative efficiency $\eta_R$ incurs,  the open-water efficiency $\eta_O$ of a propulsor varies between 0\% and 75\%, depending crucially on the advance ratio $J$, which is highly speed dependent due to the changing resistance of the vessel at different speeds.
 In \parencite{morrison2018} it was shown that it was most probable a pumpjet's open-water efficiency could be around half that of a propeller at speeds at or below 5kt.
@@
 As such, the reason that my earlier paper did not explicitly discuss relative rotative efficiency and the impact of the wake field, this was due to it being so obviously and demonstrably of no bearing whatsoever on my conclusions, and I had assume that as much would immediately clear to any technically competent reviewer.
-
 The use of this argument displays a degree of ignorance of the physics, as clearly described in numerous and publically available sources, that it is difficult to account for by incompetence alone.
@@
 
-To use a simple analogy to, imagine that you came to buy a coffee from a your favourite cafe, only to find that they weren't selling any coffee that day, and wouldn't be selling any more indefinitely.
+Imagine that you came to buy a coffee from a your favourite cafe, only to find that they weren't selling any coffee that day, and wouldn't be selling any more indefinitely.
 Their coffee maching had been removed, and they were only selling cold-drip, coffee, stored in little bottles in the fridge.
 Overcoming initial shock (since you weren't aware of any other cafes who would make such a bizarre decision as to not sell espresso-coffee as a substantial part of their offering), you made some tentative inquiries as to why.
+
 At first, you're told vaguely that it was to do with health and safety.
@@
 In anycase, it sounded like it could be very technical, and you decided to trust your very experienced and seemingly intelligent barista and forget about your lingering doubts as to why no other cafes were following suit.
+
 Then later you find a notice posted up at the front of the cafe, explaining fully the rationale behind the abandmonment of the coffee machine, and the connection with temperature.
@@
 
+\newpage
+\subsection{Defence Submission to Senate Economic References Committee}\label{def2018}
+
+\begin{figure}[h]
+\includegraphics[width=0.9\textwidth, center]{def1.png}
+\end{figure}
+
+\begin{figure}[h]
+\includegraphics[width=0.9\textwidth, center]{def2.png}
+\end{figure}
+
+\begin{figure}[h]
+\includegraphics[width=0.9\textwidth, center]{def3.png}
+\end{figure}
+
+\begin{figure}[h]
+\includegraphics[width=0.9\textwidth, center]{def4.png}
+\end{figure}
+
+\begin{figure}[h]
+\includegraphics[width=0.9\textwidth, center]{def5.png}
+\end{figure}
+
+\begin{figure}[h]
+\includegraphics[width=0.9\textwidth, center]{def6.png}
+\end{figure}
+
 \end{appendices}

### 2018-09-18 Second.Rnw
big commit

@@
 I was also invited to provide evidence to the committee in a public hearing on the 7th of June 2018.
-In that public hearing, I stood by the conclusion of my substantive paper, which was: ``between two otherwise identical submarines, the one with the pump jet will always have a lower dived endurance, a lower dived range, a worse indiscretion ratio, a lower overall endurance and a lower overall range than the one with the propeller."  In the presence of both the outgoing acting CEO of Naval Group Australia Brent Clark, and General Manager of Submarines (for Defence) Stephen Johnson, I pointed out that that statement was yet to be challenged, and that I'd be very surprised if anyone with technical competence or professional integrity sought to challenge it.
+In that public hearing \ref{SENmorrison}, I stood by the conclusion of my substantive paper, which was: ``between two otherwise identical submarines, the one with the pump jet will always have a lower dived endurance, a lower dived range, a worse indiscretion ratio, a lower overall endurance and a lower overall range than the one with the propeller."  In the presence of both the outgoing acting CEO of Naval Group Australia Brent Clark, and General Manager of Submarines (for Defence) Stephen Johnson, I pointed out that that statement was yet to be challenged, and that I'd be very surprised if anyone with technical competence or professional integrity sought to challenge it.
 
 In their own evidence during the course of the day, neither of those two did choose to challenge it.
-Senator Kim Carr requested that defence produce a written response to that report for the committee prior to their handing down of their finaly report for the Inquiry, and Defence agreed to do their best in the time available, with the head of Capability and Sustainment Group (CASG) Kim Gillis noting that ``we do hold that expertise", and Senator Carr emphasising that ``it would look pretty bloody stupid after the report's handed in''.
-Despite these exhortations for timeliness, in the final report of the committee, this request, and defence's acceptance of it was noted, but no acknowledgement was made of whether any response was actually received, or what it contained \parencite{SERC2018final}.
+Senator Kim Carr requested that defence produce a written response to that report for the committee prior to their handing down of their finaly report for the Inquiry, and Defence agreed to do their best in the time available, with the head of Capability and Sustainment Group (CASG) Kim Gillis assuring the committee that ``We  will  provide  the  committee  with  the  best  possible  advice  wecan  in  the  shortest  possible  time  frame  to  meet  your  requirements  so  that  you  can  be  betterinformed, because we do hold that expertise." Senator Carr emphasised the importance of timeliness that ``it would look pretty bloody stupid after the report's handed in''.
+Despite these exhortations, in the final report of the committee this request, and defence's acceptance of it was noted, but no acknowledgement was made of whether any response was actually received, or what it contained \parencite{SERC2018final}.
 
@@
 
+\subsection{RADM Rowan Moffit, Head of SEA 1000, before the Senate}
+
 \begin{tcolorbox}[width=0.9\textwidth, center, colback=red!5!white,colframe=red!75!black, title= \textbf{28 May 2012 - Senate Estimates}]
@@
 It is also noteworthy the tone of confidence that Senator Johnston's challenge has, driving directly for an explanation of who had the original idea, which RADM Moffit was unable or unwilling to answer.
+
+\subsection{Naval Group on ASPI Strategist and own website}
 
@@
 
+\subsection{Naval Group at PACIFIC 2017 and in ASPI Strategist}
 \begin{tcolorbox}[width=0.9\textwidth, center, colback=blue!5!white,colframe=blue!75!black, title= \textbf{4 October 2017 - Jean-Michel Billig, Executive Director Australian submarine Program, Naval Group}]
@@
 
+\subsection{RADM Sammut, Head of SEA 1000, to Senate and ASPI Strategist}
 \begin{tcolorbox}[width=0.9\textwidth, center, colback=red!5!white,colframe=red!75!black, title= \textbf{25 Oct 2017 - Senate Estimates}]
@@
 
+\subsection{Department of Defence, to Senator Xenophon's Question on Notice}
 \begin{tcolorbox}[width=0.9\textwidth, center, colback=black!5!white,colframe=black!75!black, title= \textbf{3 November 2017 - Department of Defence, in response to Question on Notice from Senator Xenophon from October 2017}]
@@
 
+\subsection{Stephen Johnson, General Manager Submarines, before the Senate}
 \begin{tcolorbox}[width=0.9\textwidth, center, colback=red!5!white,colframe=red!75!black, title= \textbf{15 Dec 2017 - Senate Estimates}]
@@
 
+\subsection{Department of Defence, to Senator Patrick's Questions on Notice}
 \begin{tcolorbox}[width=0.9\textwidth, center, colback=black!5!white,colframe=black!75!black, title= \textbf{9 February 2018 - Department of Defence, in response to Question on Notice from Senator Rex Patrick in December 2017}]
@@
 
+\subsection{Department of Defence, to Senator Ketter's Questions on Notice}
 \begin{tcolorbox}[width=0.9\textwidth, center, colback=black!5!white,colframe=black!75!black, title= \textbf{3 April 2018 - Department of Defence, in response to Question on Notice from Senator Chris Ketter in December 2018}]
@@
 For Defence not to have obtained that opinion from DCNS at the time of considering the bid would be a matter of incomprehensible oversight.
-It is made all the more difficult to believe given the scrutiny applied to the matter in the Senate in 2012 \ref{SENmoffit}, and the program director's assurance that the work was being done to assess its suitability.
+It is made all the more difficult to believe given the scrutiny applied to the matter in the Senate in 2012 \ref{SENMoffit}, and the program director's assurance that the work was being done to assess its suitability.
 
@@
 It is possible to believe that Defence's own embarrassment at having been unable to detect the insincerity in Naval Group's offer of the pumpjet, and inability to reach their own technical conclusions quickly enough to detect the fraud, may motivate the present actions.
-This amount to a disasterous failure on the part of the public service to act with integrity and in the national interest.
-However, if Defence had truely been un-wittingly deceived, it is difficult to see how they could not recover their pride (and regain a necessary measure of respect from their build partner) by quickly adopting a more combative stance, calling out the deceptions on the part contractors part, and quickly exacting whatever contractual reparations they might be entitled to.
+This would  amount to a disasterous failure on the part of the public service to act with integrity and in the national interest.
+However, if Defence had truely been un-wittingly deceived, it is difficult to see how they could not recover their pride (and regain a necessary measure of respect from their design partner) by quickly adopting a more combative stance, calling out the deceptions on the part contractors part, and quickly exacting whatever contractual reparations they might be entitled to.
 If the contractual reparations were not sufficiently strong, the Commonwealth's freedom to not proceed to a build contract (from the initial design contract) or re-introduce competition into the design process as many have suggested should occur in any case \parencite{stanford2017}, would have provided ample commercial leverage to reset the relationship in a manner which was better aligned with the Commonwealth's interest.
 
-One must at this point give consideration to the possibly that Defence had not been entirely unwittingly deceived, and was actually complicit in their own deception, and it's present upkeep.
-This would mean that the Department of Defence and the Navy, or at least a few of very senior people within those organisations, had to conspired allow a proposal which was known to be inappropriate to the Commonwealth's needs to appear victorious in the CEP,  presumeably in order to favour a bidder for whom no more defensible explanation could be found.
-It appears unlikely that there could be any decent motive for such a course of action.
-However, one must evaluate the probabilities of the necessary oversights for such an eventuatlity occurring by accident or by arrangement.
+One must at this point give consideration to the possibly that Defence had not been entirely unwittingly deceived, and was actually complicit in their own deception, including its present upkeep.
+This would mean that the Department of Defence and the Navy, or at least a few of very senior people within those organisations, had conspired allow a proposal which was known to be inappropriate to the Commonwealth's needs to appear victorious in the CEP,  presumeably in order to favour a bidder for whom no more defensible explanation could be found.
+It appears unlikely that there could be any decent motive for such a course of action, and would lead invariably to the grave conclusion that some kind of corruption has occurred.
+In the circumstances, one must evaluate the probabilities of the necessary oversights for such an eventuatlity occurring by accident against the occurring by arrangement.
 If the pumpjet was known by DCNS to be an unsuitable proposition for a diesel-electric submarine, a seemingly unlikely degree of naive incompetence would be required for Defence to fail to ascertain as much during the Competitive Evaluation Process.
@@
 \\
-Answer: ``See response to Senate Economics Reference Committee Question on Notice, Q1 – Pump Jet Efficiency Across the Entire Speed Range."
+Answer: ``See response to Senate Economics Reference Committee Question on Notice, Q1 - Pump Jet Efficiency Across the Entire Speed Range."
 \end{tcolorbox}
 The response to Q1 - a question specifically and only asking a question about \textit{efficiency} of the pumpjet, avoids contradicting the evidence directly by conflating the question with acoustics.
-To avoid the answering a question that is specifically about acoustics by referring to that same answer which incorporates efficiency is nothing more than a point-blank refusal to answer either question directly.
-
+To avoid the answering a question that is specifically about acoustics by referring to that same answer which incorporates efficiency amounts to nothing more than a point-blank refusal to answer either question directly.
+
+\subsection{Brent Clark, Acting CEO Naval Group Australia, in ASPI Strategist}
 \begin{tcolorbox}[width=0.9\textwidth, center, colback=blue!5!white,colframe=blue!75!black, title= \textbf{17 May 2018 - Brent Clark, Acting CEO Naval Group Australia}]
@@
 The company building the Navy's 12 new submarines says they'll be the world's first conventional boats big enough to effectively use the company's pump jet propulsion system.
-French Naval Group Australia says that its propulsor will prove much quieter than a propeller on the 4,700‑tonne submarine it has dubbed the 'Shortfin Barracuda'.
+French Naval Group Australia says that its propulsor will prove much quieter than a propeller on the 4,700-tonne submarine it has dubbed the `Shortfin Barracuda'.
 \\
@@
 Naval Group Australia's interim CEO, Brent Clark, tells The Strategist that the decision about the best propulsion system for the submarine will be made by the Defence Department, but that he hadn't had any indication that it would chose anything but a propulsor.
-'We've had every indication that they like what was offered.'
+`We've had every indication that they like what was offered.'
 \\
@@
 \\
-"A lot of people don't fully comprehend, I think, that yes, a propulsor on a 1,500-tonne coastal submarine would make no sense.
+``A lot of people don't fully comprehend, I think, that yes, a propulsor on a 1,500-tonne coastal submarine would make no sense.
 But we're not building a 1,500-tonne coastal submarine.
@@
 
-This is also the first time that the argument that the size of the submarine is a crucial enabling difference that makes the pumpjet viable is made.
-It is the closest that Naval Group has come from making a firm statement of faith in the suitability of the technology for the Australian diesel-elctric submarine.
-As I discussed in another blog post in ASPI Strategist \parencite{morrison2018b} shortly afterwards, this is a remarkeably concession that is consistent with the physics that I outlined, since the inefficiency of the pumpjet would likely drive demand for additional fuel and batteries (hence size) to meet requirements, which is more of a concession of the detrimental impact of the pumpjet on the submarine.
+This is also the first appearance of the argument that the size of the submarine is a crucial factor that makes the pumpjet viable.
+It is the closest that Naval Group has come to making a firm statement of faith in the suitability of the technology for the Australian diesel-elctric submarine.
+As I discussed in another blog post in ASPI Strategist \parencite{morrison2018b} shortly afterwards, this is a remarkeably concession that is consistent with the physics that I outlined, since the inefficiency of the pumpjet would likely drive demand for additional fuel and batteries (hence size) to meet requirements.
+As such, this admission is more of a concession of the detrimental impact of the pumpjet on the submarine.
 This particular blog post was acknowledged on twitter by one former US Navy submariner and submarine enthusiast as an ``Awesome article about pump jet efficiencies and viability".
@@
 
+\subsection{Morrison Senate evidence and Clark, Johnson, Gillis responses}
 \begin{tcolorbox}[width=0.9\textwidth, center, colback=red!5!white,colframe=red!75!black, title= \textbf{07 Jun 2018 - Senate Economic References Committee}]
@@
 Naval Group here steadfastly refuses to offer a clear opinion on the pumpjet.
-Only months earlier Defence was asked a specific question on notice about whether they thought Naval Group maintained that a pumpjet was suitable for a diesel-electric submarine.
+Only months earlier Defence was asked a specific question on notice about whether they thought Naval Group maintained the opinion that a pumpjet was suitable for a diesel-electric submarine.
 Clarke's interview on the Strategist, which seemed to be strongest statement of faith from Naval group in the suitability of the pumpjet for a conventional submarine (perhaps in an effort to patch up the noteable refusal to answer the question) had been answered publically by the same person who had just appeared and he was being asked to respond to.
@@
 \textbf{Senator REYNOLDS:} Mr Gillis!
-\\
-\\
+
 ...
-\\
-\\
+\end{tcolorbox}
+
+\begin{tcolorbox}[width=0.9\textwidth, center, colback=red!5!white,colframe=red!75!black, title= \textbf{07 Jun 2018 - Senate Economic References Committee - CONTINUED}]
+
 \textbf{Senator KIM CARR:} I contend that we don't have the technical expertise to evaluate the report.
@@
 Despite the obvious shock and exclamation of the senators, Mr Gillis went on to attempt to dismiss my evidence as just another opinion which could probably to be dismissed along with 99\% of the others.
-
-This level of ambivalence towards external criticism from a such a senior official on such a topic as this should be alarming.
+This is a sustained and deliberate effort to adopt a dismissive attitude in the face of serious criticism.
+
+This level of ambivalence towards external criticism from a such a senior official on such a topic as this is worth noting.
 The topic of the pumpjet had such sustained criticism and interest from various quarters, including multiple questions from senators from all sides of government over six years, reports in the ABC, ASPI Strategist, Defence Connect, Insight Economics, and many others.
@@
 
-Such an attitude raises questions about whether Mr Gillis's evidence is even given honestly.
+Such an openly dismissive attitude, exemplified by Mr Gillis' ``not all of this is rivetting" raises questions about whether Mr Gillis's evidence is even given genuinely and honestly.
 For instance, Mr Gillis' claim that he had not read the report, since ``we weren't actually  given a copy" warrants serious attention.
@@
 Moreover, his department had received written Questions on Notice from Senator Chris Ketter, acting on behalf of this very committee, specifically requesting written responses to my paper with a return date of 3 April.
-His department did indeed responded to prior to the hearing, as noted above with responses which blatantly refused to answer the questions.
-Furthermore this particular hearing was also especially long in the making, since it was once postponed on account of Defence refusing to make officials available to attend.
-My presence as a witness was requested in February when the proposed date was still in May, and the list of witnesses was provided publically circulated during the weeks prior to the hearing.
-
-In such circumstances, claims to be ignorant of my report, and an attempt to lump it with the other 99\% of dissenting correspondence, seem about as reasonable as Mr Johnson's claim that he simply wasn't paying attention when he was being quoted and refuted before the committee earlier in the day.
-That this is the type of responses being given by the most senior officials who are responsible for the largest defence procurement in Australia's history should be alarming.
-
+His department most probably responded to prior to the hearing, as noted above, with responses which blatantly refused to answer the questions.
+\footnote{However, I was not made aware either of the questions or their answers prior to hearing.} Furthermore this particular hearing was also especially long in the making, since it was once postponed on account of Defence refusing to make officials available to attend.\parencite{adm2018}  My presence as a witness was requested in February when the proposed date was still in May, and the list of witnesses was publically circulated during the weeks prior to the hearing.
+In such circumstances, claims to be essentially ignorant of my report and pretend it could probably be dismissed after the reading of the first page, should be examined critically.
+
+Mr Johnson's comments regarding propulsion power at low ship speeds further demonstrates a cavalier attitude towards the truth.
+He is absolutely right to say that the shaft horse-power required to move a large vessel at low speeds is astonishingly small.
+This is because the resistance of a mody moving through a fluid scales generally with velocity squared, as was discussed at length in my paper, since it comprised one of the `intricacies' which I considered particularly important to this matter.
+His blunt description of paying `two pennies' when you could pay only one, however, betrays a staggering misunderstanding of the energy densities relevant to conventional submarines.
+His remark holds well in the case of any air-breathing surface ships burning diesel, or nuclear submarine.
+However, once dived a conventional submarine is forced to rely on batteries, and amounts of energy which were previously a penny immediately become more like a pound.
+Or actually, taking the energy density of diesel at 45MJ/kg, and lead-acid batteries at 0.13MJ/kg, the truth is closer to four pounds.
+
+In contrast to Mr Johnson's inference that my report might be made `without contemporary knowledge or without understanding nececessarily the intricacy', his own comments on powering requirements of ships, just like the propulsive pods for ships, demonstrate a complete disregard or ignorance of the most basic and fundamental concepts relevant.
+
+\subsection{Dr David Kershaw, Chief of Martime Division DSTG, in Australian Defence Magazine}
 \begin{tcolorbox}[width=0.9\textwidth, center, colback=black!5!white,colframe=black!75!black, title= \textbf{August 2018 - Dr David Kershaw, in Australian Defence Magazine}]
-\textit{Julian Kerr includes extensive quotation from Dr David Kershaw in an article titled "Future Submarine: Pumpjet propulsor technology choice still not well explained" \parencite{kerr2018}.
+\textit{Julian Kerr includes extensive quotation from Dr David Kershaw in an article titled ``Future Submarine: Pumpjet propulsor technology choice still not well explained" \parencite{kerr2018}.
 Dr Kershaw is the current Chief of Maritime Division, a position he held since May 2016.
@@
 ``The question of propeller versus pump jet really comes down to the acoustic signature; a well-designed pump jet is considerably quieter than an equivalent propeller." he said.
-"One of the advantages of a well-designed pump jet is that it can smooth out the water flow that is coming off the back of a submarine before the flow hits the moving rotor and there's less unsteady flow, and thus less noise.
-``This smoothing-out also increases the propulsive efficiency; a well-designed pump-jet can have efficiencies equal to, or better than, an equivalent propeller.
-You just have to have the capabilities and facilities to be able to design an efficient pumpjet and we know that France can do that."
+``One of the advantages of a well-designed pump jet is that it can smooth out the water flow that is coming off the back of a submarine before the flow hits the moving rotor and there's less unsteady flow, and thus less noise.
+\textbf{``This smoothing-out also increases the propulsive efficiency;} a well-designed pump-jet can have efficiencies equal to, or better than, an equivalent propeller.
+\textbf{You just have to have the capabilities and facilities to be able to design an efficient pumpjet and we know that France can do that."}
 Echoing RADM Sammut, Dr Kershaw stressed that the pumpjet on a submarine must be designed in conjunction with the hullform to which it was to be attached.
@@
 It's akin to arguing 'camels have to cary two heavy humps on their backs.
-Horses don't have to carry these two heavy humps, which is why they can fly.'  It also directly contradicts the claim made by Stephen Johnson before the Senate, that ``No matter what kind of propeller or propulsion pump jet  you design, it has to be optimised for a particular speed.".
+Pigs don't have to carry these two heavy humps, which is why they can fly.'  It also directly contradicts the claim made by Stephen Johnson before the Senate, that ``No matter what kind of propeller or propulsion pump jet  you design, it has to be optimised for a particular speed.".
 
@@
 This is a most fundamental and basic assumption which is absolutely intrinsic to all basic marine propulsion, and should also be understood by essentially anyone who has passed first-year engineering or physics courses.
-(It's known) (For instance, see Pages 9 and 10 of Reference \parencite{MAN2017}, or Chapter 12 of Reference \parencite{Carlton2007}, or Chapter 3 of Reference \parencite{Mollard2011}.)
-
-Apart from it being quite an elementary law of phyiscs (the Drag Law), referred to and relied upon in posts such those by Andrew Davies \parencite{davies2017} commenting on this exact question, any substantial reading of the literature focussed on the design of pumpjets for fully submerged objects (such as torpedoes or submarines) will also make it abundantly clear, and the special distinction with wave-making resistance (which is crucial for fast-vessels in the process of reaching the plane) is also abundantly clear, and has been well known for a long time \parencite{wislicenus1973} \parencite{henderson1964}.
+(For instance, see Pages 9 and 10 of Reference \parencite{MAN2017}, or Chapter 12 of Reference \parencite{carlton2007}, or Chapter 3 of Reference \parencite{mollard2011}.) Apart from it being quite an elementary law of phyiscs (the Drag Law), referred to and relied upon in posts such those by Andrew Davies \parencite{davies2017} commenting on this exact question, any substantial reading of the literature focussed on the design of pumpjets for fully submerged objects (such as torpedoes or submarines) will also make it abundantly clear, and the special distinction with wave-making resistance (which is crucial for fast-vessels in the process of reaching the plane) is also abundantly clear, and has been well known for a long time \parencite{wislicenus1973} \parencite{henderson1964}.
+Further discussion is given in a subsequent section of this paper.
 
 That the Science and Technology Advisor on SEA 1000, and current Head of Maritime Division at DST could make such an gross error is extremely shocking.
-The lack of understanding demonstrated strongly contradict the assumption that he is competent to speak or advise on this topic.
-Whilst his expertise is most probably actually not that closely related to propulsion systems or hydrodynamics (his online bio suggests his torpedo-related experience is actually closer to the computer systems and signal processing involved \parencite{DSTG2018}) it would seem extremely strange for a competent physicist to such public statements outside their area of expertise, if they weren't confident in their own understanding.
+Whilst it might be tempting to write it off as a careless wording or some journalistic misquotation, subsequent correspondence from the Department of Defence rules this out as a possibility.
+The lack of understanding demonstrated strongly contradicts the assumption that he is competent to speak or advise on this topic.
+Whilst his expertise is most probably actually not that closely related to propulsion systems or hydrodynamics (his online bio suggests his torpedo-related experience is actually closer to the computer systems and signal processing involved \parencite{DSTG2018}) it would seem extremely strange for a competent physicist to make such public statements outside their area of expertise, if they weren't confident in their own understanding.
 
 The other revealing thing about Dr David Kershaw's remarks here is the clear connection with comments made by RADM Sammut.
-He gives the most detailed description of the interaction between the hull and the pumpjet thus far, arriving at the same conclusion that they have to be 'tuned' together which has been previously used by RADM Sammut.
+He gives the most detailed description of the interaction between the hull and the pumpjet thus far, arriving at the same conclusion that they have to be `tuned' together - the same wording that has been previously used by RADM Sammut.
 (This argument will be addressed fully in a later chapter.)  An apparent conclusion is that RADM may have relied upon advice from Dr David Kershaw, (which given the ineptitute of earlier comments relating to hydrodynamics is gravely concerning), or that both of them have relied upon advice from another common source.
@@
 Although pump jet propulsion would be new to the RAN, the technology was well-known to DSTG.
-"We've invested heavily in our hydrodynamics and hydroacoustics program over the past 20 years and this is includes the computational models as well as experimental facilities that confirm modelling results," Dr Kershaw said to \textit{ADM}.
+``We've invested heavily in our hydrodynamics and hydroacoustics program over the past 20 years and this is includes the computational models as well as experimental facilities that confirm modelling results," Dr Kershaw said to \textit{ADM}.
 \textbf{We've been able to develop our analysis on a generic pumpjet design and thus confirm our understanding of pumpjet performance; you basically build it up from first principles.
-"Our analysis at the moment is all around analysing current options but we will be looking at what's ahead in the future and enable Australia to develop the sovereign capability to take best advantage of pumpjet technology."}
+``Our analysis at the moment is all around analysing current options but we will be looking at what's ahead in the future and enable Australia to develop the sovereign capability to take best advantage of pumpjet technology."}
 \\
@@
 So will pump jet technology definitely be used for the Future Submarine?
-``That is what is currently in the design \textbf{and that's what a lot of our science and technology program is about; having the capability to do the independent analysis on the technologyies} and capabilities planned for the Future Submarine, including the pump jet program," Dr Kershaw said to \textit{ADM}.
+``That is what is currently in the design \textbf{and that's what a lot of our science and technology program is about; having the capability to do the independent analysis on the technologies} and capabilities planned for the Future Submarine, including the pump jet program," Dr Kershaw said to \textit{ADM}.
 \textbf{Discussions were currently underway with Naval Group and the French government regarding closer collaboration on a range of technologies and this would include pump jets, Dr Kershaw disclosed.}
@@
 ``One thing I would like to emphasis is that Australia's geographic circumstances require a larger conventional submarine than those currently produced for export to other parts of the world.
-The size of the Future Submarine means a pumpjet is a real game changer for optimising the performance of the submarine in its intended role, and ensuring regional superiority in undersea warfare."
+\textbf{The size of the Future Submarine means a pumpjet is a real game changer for optimising the performance of the submarine in its intended role, and ensuring regional superiority in undersea warfare."}
 Dr Kershaw concluded.
@@
 
-These comments are extremely important because the illuminate the true state of DSTG's research, and the degree of confidence they have in the pumpjet.
+These comments are extremely important because they illuminate the true state of DSTG's research, and the degree of confidence they have in the pumpjet.
 It is consistent with responses given to a question on notice from Rex Patrick \ref{QON150}, where it was revealed that 1.1 million was spent on ``development of a \textit{generic submarine and pumpjet design}, computational and analytics models, and validation of these models through experimental studies".
-The admission that DSTG is still working on a 'generic' design also casts into new light their repeated insistence that the performance of a pumpjet cannot be assessed unless it is exactly tuned carefully with the hull, and vice versa.
-It is plainly obvious that if they are still working on 'generic' models, rather than the particular one which is proposed by France, they are in no better position to make positive assessment that contradicts findings in Reference \parencite{morrison2018}.
-By their own assessment, according to the insistence of Dr Kershaw and RADM Sammut, they are in no position to assess the performance of the pumpjet that will be used on the future submarine, given they haven't been able to 'tune' it to the hull shape of the future submarine, and 'tune' the hullshape to the pumjet.
-They are still working with 'generic' designs.
-
-As I have already put before the Senate \ref{SENmorrison}, I find that idea that such 'tuning' could so dramatically alter the performance of the pumjet to be completely implausible.
-I would support the idea that the kind of generic modelling described would be quite sufficient to reveal results that are consistent with the results I produced from a basic literature review (that pumpjet efficiency declines towards zero at very low water-speeds) and safely rule-out the suitability of a pumpjet for diesel electric submarines.
-If the results of the modelling done by DSTG to date could confidently contradict such a finding, one would expect that in Senate hearings and in responses to Questions on Notice, a very confident and clear-cut refution of the claims made in my report and evidence would be able to be made.
+The admission that DSTG is still working on a `generic' design also casts into new light their repeated insistence that the performance of a pumpjet cannot be assessed unless it is exactly tuned carefully with the hull, and vice versa.
+It is plainly obvious that if they are still working on `generic' models, rather than the particular one which is proposed by France, they are in no better position to make positive assessment that contradicts findings in Reference \parencite{morrison2018}.
+By their own assessment, according to the insistence of Dr Kershaw and RADM Sammut, they are in no position to assess the performance of the pumpjet either given they haven't been able to do the necessary `tuning' the hull and pumpjet.
+Instead they are still working with `generic' designs.
+
+As I have already put before the Senate \ref{SENmorrison}, I find that idea that such `tuning' could so dramatically alter the performance of the pumjet to be completely implausible.\footnote{In my evidence is said it was: ``as if someone bought an efficient hybrid car like a Prius and then decided to fit it with tank tracks from a dozer or a battle tank, and then argued they would tune the tank tracks back to be as efficient as the Prius was in the first place.
+It goes radically against what the technology was intended for in the first place".}  Rather, I would support the idea that the kind of generic modelling described would be quite sufficient to reveal results that are consistent with the results I produced from a basic literature review (that pumpjet efficiency declines towards zero at very low water-speeds) and safely rule-out the suitability of a pumpjet for diesel electric submarines.
+If the results of the modelling done by DSTG to date could confidently contradict such a finding, one would expect that in Senate hearings and in responses to Questions on Notice, a very confident and clear-cut statement to that effect would be made.
 This has been conspicuously absent from all responses, including this one.
-Instead, it is claimed that the 'tuning' with the hull, nullifies the result.
-If instead, the DSTG studies are actually consistent with my findings, then DSTG and Defence are knowingly supressing such results, or naively hoping that the French design, and the finer 'tuning' between the pumpjet and the hull can avhieve such a dramatically diferent result that the results of 'generic' tests can be ignored.
-Either situation would have terrible consequences for the program, and if the former is true it is likely that the Parliament has been deliberately misled on multiple occasions.
+Instead, the response claims that the future `tuning' with the hull nullifies the result, claiming the result can't be properly known yet, rather than claiming they have a different or contradictory result.
+
+Given the ample resources of Defence and Naval Group, this insistence that the result cannot yet actually be known, rather than that they know better and reached a different result and raises the question of what they actually do know, and how different it actually is to my result.
+If the DSTG studies are actually consistent with my findings, then DSTG and Defence are knowingly supressing such results, or naively hoping that the French design, and the finer `tuning' between the pumpjet and the hull can avhieve such a dramatically diferent result that the results of `generic' tests can be ignored.
+Either situation would have terrible consequences for the program, and if the former is true it is almost certain that the Parliament has been deliberately misled on multiple occasions.
 
@@
 Apparently, there is still not sufficent information to be confident about the use of the pumpjet.
-Still, a confident decision on whether the pumpjet will even be used cannot be made.
-Consistent with the statements of Mr Johnson \ref{SENjohnson}, the information about the pumpjet performance of the pumpjet on the nuclear Barracuda has not yet arrived.
-Nor, it would appear, is sufficient theoretical modelling, or scale model testing, or any other confirmatory data sufficient on this pumpjet for DSTG to be clearly persuaded of the suitability of the pumpjet for our submarine.
+A decision on whether the pumpjet will even be used still cannot be made.
+Consistent with the statements of Mr Johnson \ref{SENjohnson}, the information about the pumpjet performance on the nuclear Barracuda has not yet arrived.
+Nor, it would appear, is sufficient theoretical modelling, or scale model testing, or any other confirmatory data sufficient on this pumpjet for DSTG to be clearly persuaded of the suitability of the pumpjet for our submarine, which will have an almost-identical hull.
 Whilst simultaneously claiming that efficiency and performance of the pumpjet can't be assessed without the precise form of the hull and the pumpjet being 'tuned' together, they admit that they are having to model a 'generic' pumpjet, building up from first principals, in order to have sufficient confidence in their own 'indepenent analysis' of the technologies.
 
-It can be concluded firmly from this interview that the promise of technology transfer promised by Sean Costello has not been made good, even to a sufficient extent to allow the government to confidently decide on its suitability for the submarine, and assess Naval Groups confidence in the suitability sufficient to answer parliamentary Questions on Notice.
-Such gravity of such a failure is of profound consequence to the well-being of such a major project, given the significance of the pumpjet to the overall weight, balance and performance of the submarine as I outlined before the Senate \ref{SEMmorrison}.
+It can be concluded firmly from this interview that the promise of technology transfer made by Sean Costello when he was CEO of DCNS Australia has not been made good, even to a sufficient extent for the Australian government to assess the suitability of the technology.
+Even more worrying, it would seem insufficient to even assess Naval Groups opinion about the suitability of the technology sufficient to answer parliamentary Questions on Notice.
+The gravity of such a failure is of profound consequence to the well-being of such a major project, given the significance of the pumpjet to the overall weight, balance and performance of the submarine as I outlined before the Senate \ref{SENmorrison}.
 
 It also throws into doubt the integrity of the statements made by RADM Sammut before Senate Estimates.
-In the days and weeks immediately prior to his testimony \ref{SEMsammut}, Jean-Michel Billig had triggerd the controversy around the pumpjet at PACIFIC 2017, the largest Naval Show and conference in Australia, which occurs only each two years.
-This had lead to substantial coverage on the ABC, Defence Connect, and ASPI Strategist, and almost certainly contributed to the line of questioning that Senator Hanson doggedly pursued about the pumpjet that day.
-An interview where he faced further questions on the pumpjet's inclusion with Brendan Nicholson was published just days later\parencite{sammut2018}.
+In the days and weeks immediately prior to his testimony \ref{SENsammut}, Jean-Michel Billig had triggerd the controversy around the pumpjet at PACIFIC 2017, the largest Naval Show and conference in Australia, which occurs only each two years.
+This had lead to substantial coverage on the ABC\parencite{lane2017}\parencite{iggulden2017}, Defence Connect\parencite{mcmahon2017a}\parencite{mcmahon2017b}, and ASPI Strategist\parencite{davies2017}, and almost certainly contributed to the line of questioning that Senator Hanson doggedly pursued about the pumpjet that day.
+An interview where he faced further questions on the pumpjet's inclusion with Brendan Nicholson was published just days later\parencite{sammut2017}.
 It is inconceivable that the significance of the pumpjet question, and the state of the Commonwealth's confidence in its inclusion in the design was not front of mind at the time when he was being questioned by Senator Hanson on the 25th of October 2017.
@@
 
-If Naval Group's level of information transfer was proceeding as promised, then RADM Sammut had accepted (as being superior to other bids proposals) a sequence of data and technology transfer which would leave DSTG relying on their own 'generic' models in trying to assess the suitability of the pumpjet, and Defence unable to answer confidently whether Naval Group even had confidence in the suitability of the technology two years after the conclusion of the evaluation process.
-Unless every other competitor had also promised to withold critical information to assess the suitability of a technology that was 'integral' to their bid, there can be no plausible defence of such a decision as competent.
-Alternatively, if Naval group had promised more fluid information transfer, and this information had subsequently been withheld, RADM Sammut was actually concealing that a crucial break-down in relations with the major contractor had already occurred, with potentially disasterous consequences for the future of the submarine program.
-The only other viable alternative is that information transfer was not really something that commonwealth looked at in considering the selection of DCNS as a design partner, and RADM was knowingly misleading the Parliament.
-
+If Naval Group's level of information transfer was proceeding as promised, then RADM Sammut had accepted (as being superior to other bids proposals, and sufficienct for the Commonwealth's purposes) a sequence of data and technology transfer which would leave DSTG relying on their own `generic' models in trying to assess the suitability of the pumpjet, and Defence unable to answer confidently whether Naval Group even had confidence in the suitability of the technology two years after the conclusion of the evaluation process.
+Unless every other competitor had also promised to withold critical information to assess the suitability of a technology that was `integral' to their bid during the bid process and for two years after, there can be no plausible defence of such a decision.
+Alternatively, if Naval Group had promised more fluid information transfer, and this information had subsequently been withheld, RADM Sammut was actually concealing that a crucial break-down in relations with the major contractor had already occurred, with potentially disasterous consequences for the future of the submarine program.
+The only other viable alternative is that information transfer was not really something that commonwealth looked at in considering the selection of DCNS as a design partner, and RADM was knowingly misleading the Parliament when he was answering questions for in a Senate Estimates hearing.
+
+\subsection{RADM Greg Sammut, Head of SEA 1000, in Australian Pacific Defence Reporter}
 \begin{tcolorbox}[width=0.9\textwidth, center, colback=black!5!white,colframe=black!75!black, title= \textbf{August 2018 - RADM Sammut, in Australia Pacific Defence Reporter}]
@@
 
-These series of comments by RADM Sammut are consistent with other comments of his own and other government sources, and wholly inconsistent with a competent and sound engineering and design process.
+These series of comments by RADM Sammut are consistent with other comments of his own and other government sources, but extremely incongruous with what one would expect in a design phase for a properly managed project.
 
@@
 One possibility is that a firm decision actually already has been reached on the inclusion or exclusion of these technologies, and RADM Sammut is not willing to reveal it.
-It should be noted here that the 'current thinking' wording might be interpreted as something of a retreat from earlier 'we are using' statements made in late October 2017 \parencite{sammut2017}.
+It should be noted here that the `current thinking' wording might be interpreted as something of a retreat from earlier `we are using' statements made in late October 2017 \parencite{sammut2017}.
 The possibility that a pumpjet has already actually been ruled out must at least be considered.
@@
 Weeks earlier, Jean-Michel Billig had cited said ``What is driving the choice of technology is basically the operational requirement" when challenged as to whether the pumpjet was actually mandated \parencite{lane2017}.
-With both parties pleading ignorance of what technologies would ultimately be included, and both citing 'requirements', it does seem to imply that the requirements had not yet been settled.
+With both parties pleading ignorance of what technologies would ultimately be included, and both citing `requirements', it does seem to imply that the requirements had not yet been settled.
+This should be a matter of significant public interest.
 
@@
 If all the competitors weren't actually given the same performance requirements, which were the real requirements for the submarine which the government intended to procure, then it is extremely difficult to see how they could be `treated fairly' as RADM Sammut claimed \ref{SENhanson}.
-Moreover, credibility or conscience of the advisory panel who claimed they were should be seriously re-examined.
-If the requirements have changed substantially since the CEP, then the CEP ultimately served no purpose.
+(The credibility of conscience of the supposedly independent expert advisory panel who claimed they were should be seriously re-examined.)  If the requirements have changed substantially since the CEP, then the CEP ultimately served no purpose in selecting a partner who would best meet them.
 A vast number of public statements made by politicians and officials regarding the significance of Australia's unique 'requirements', particularly for long range, should be critically re-examined.
@@
 It's true that large amounts of fuel and food might consequently be required.
-But the idea that we 'might be better off carrying more food and fuel' betrays either either a lack of confidence that the volumes which were presumeably propososed by all contenders alongside an AIP system were sufficient, or a casual aspiration to substantially enlarge the range and endurance of the submarine even further.
+But the idea that we `might be better off carrying more food and fuel' betrays either either a lack of confidence that the volumes which were presumeably propososed by all contenders alongside an AIP system were sufficient, or a casual aspiration to substantially enlarge the range and endurance of the submarine even further than  what the requirements demand.
 It would be relatively simple at the very early stages of concept design (prior to the diameter and legnth of the submarine being fixed) to incorporate both sufficient fuel and food for a specific overall range and endurance requirement, as well as an AIP system to enhance the dived endurance, without compromising the overall range and endurance.
@@
 Worryingly, it adds confirmation that RADM Sammut might genuinely believe that the hull-diameter (and its resemblance of a pre-existing design) ought to be a pre-determined feature of a submarine design, to which the performance and capability of the submarine is subordinate.
-Such a view, including held on the grounds of safety as suggested by RADM Sammut himself \ref{SENhanson}, is such a bizarre understanding of program risk and the heirachy of engineering challenges that it should call into question the competence of a program manager, and the technical advisors who he depends upon.
+Such a view, including held on the grounds of safety as suggested by RADM Sammut himself \ref{SENhanson}, is such a bizarre understanding of program risk, the heirachy of engineering challenges and normal spiral of platform design that it should call into question his competence as a project manager, and the technical advisors whom he depends upon.
 
@@
 Whilst this is generally true for all batteries, Lithium-ion batteries are known to have substantially lower internal resistance, and consequently have dramatically improved performance (in terms efficiency) at higher rates of discharge when  compared to lead-acid batteries.
-It is for this reason that in submarines, lithium-ion batteries are acknowledged as having the most significant improvement over lead-acid at high speeds \parencite{briggs2016}.
-Given that pumpjets advantages are widely acknowledged to be at higher speeds, including implicitly by Naval Group\parencite{autret2016}, lithium-ion batteries should really be considered an enabling technology for pumpjets on diesel-electric submarines.
-(At least if the case is to be made that priorisation of that extremely high-speed sprint capability should be made above all other performance characteristics, which I have argued is the only technically defensible case to be made, whilst is an obscure and implausible case to be made in terms of useful operational capability \ref{SENmorrison})  Without lithium-ion batteries, the pumpjet efficiency will sap performance at low speeds, and at high speeds (where pumpjet efficiency improves, albeit not as high as propellers) the battery efficiency will hobble performance.
+It is for this reason that in submarines, lithium-ion batteries are acknowledged as having the most significant improvement over lead-acid during bursts at high speed \parencite{briggs2016}.
+Given that pumpjets advantages are widely acknowledged to be at higher speeds, including implicitly by Naval Group\parencite{autret2016}, lithium-ion batteries should really be considered an enabling technology for pumpjets on diesel-electric submarines.\footnote{At least this is the case is if one chooses to prioritise a extremely high-speed sprint capability above virtually all other performance characteristics.
+I have argued this is the only technically defensible case to be made for the pumpjet, whilst such a tactical scenario is likely to be sufficiently obscure and unlikely for other technical reasons that it makes the argument dubious in the extreme.
+\ref{SENmorrison}}  Without lithium-ion batteries, the pumpjet efficiency will sap performance at low speeds, and at high speeds (where pumpjet efficiency improves, albeit not as high as propellers) the battery efficiency will hobble performance.
 This effect, not accounted for in the modelling in reference \parencite{morrison2018}, will probably result in reducing the maximum endurance of a pumpjet-submarine in a sprint at 20kt from over 2 hours (as that modelling suggested) to well under one hour.
@@
 
+\subsection{Department of Defence, for Senate Economic References Committee}
 \begin{tcolorbox}[width=0.9\textwidth, center, colback=black!5!white,colframe=black!75!black, title= \textbf{Sept 2018 - Department of Defence, Report for the Senate Economic References Committee}]
@@
 Instead of the three weeks requested, Defence took about three months to deliver this report to the Senate Committee.
-The full response is in }
+The full response is in \ref{def2018}.}
 \\
@@
 However, to claim that these forces are constant with speed is to claim the most bizarre and obtuse denial of the most basic and well-established physics in this respect.
-These forces scale with velocity squared, a fact which has been widely known and accepted since Froude conducted his skin-friction tests in the 1870s.
-\footnote{It should be noted that our understanding of skin friction has improved somewhat since the 1870s.
+These forces scale with velocity squared, a fact which has been widely known and accepted since Froude conducted his skin-friction tests in the 1870s.\footnote{It should be noted that our understanding of skin friction has improved somewhat since the 1870s.
 In particular, the distiction between laminar and turbulent flows and the transition between these two regimes as demonstrated by Osborne Reynolds in the 1880s was significant, since it was discovered that laminar flows exerted a lower skin friction, and it is the turbulent flows which scale with velocity squared.
@@
 As such, speed plays the dominant role in considering the efficiency of propulsors on submerged bodies, since they also have their efficiency fall away at off-design conditions.
+
+\begin{tcolorbox}[width=0.9\textwidth, center, colback=green!5!white,colframe=green!75!black, title= \textbf{How a scientist feels about being told that because underwater bodies don't make waves, the resistance they encounter is constant with speed}]
+
+You know how camels have to carry two big heavy humps on their back?  Yeah, well pigs don't have to carry those two big heavy humps!  That's why pigs can fly.
+
+\end{tcolorbox}
 
@@
 This assertion  by defence is so utterly and blatantly ridiculous that there is no plausible way that it could be genuinely held to be a true by any one of even moderate technical competence.
-It is difficult to even believe that it could be held with any confidence by even moderately intelligent adults of little or no technical background.
-It is difficult to avoid reaching the conlcusion that the statement was given in a state of extreme contempt for the intelligence of the intended reader, as it constitutes without any doubt a deliberate lie and active attempt to mislead.
+It is difficult to even believe that it could be held with any confidence by even moderately intelligent adults of little or no technical background who took a mild interest in researching the topic.
+Unless such an extremely level of incompetence pervades every defence official and scientist who were involved in preparing this statement (which was requested directly from Kim Gillis, Head of CASG and Stephen Johnson, General Manager Submarines) is difficult to avoid reaching the conlcusion that the statement was given in a state of extreme contempt for the intelligence of the intended reader, and constitutes a deliberate lie and active attempt to mislead.
 
@@
 \includegraphics[width=0.7\textwidth, center]{drag4.png}
-\caption{A breakdown of the components of resistance taken from Reference \parencite{Mollard2011} does indeed reveal that deeply submerged submarines encounter no wave-making pressure, but still encounter frictional resistance (which scales with velocity squared).}
+\caption{A breakdown of the components of resistance taken from Reference \parencite{mollard2011} does indeed reveal that deeply submerged submarines encounter no wave-making pressure, but still encounter frictional resistance (which scales with velocity squared).}
 \label{fig:drag4.png}
@@
 I responded directly to Clark's comments arguing that the size necessitated by the extra fuel and batteries a pumpjet demanded could hardly be counted as an advantage \parencite{morrison2018b}.
-However, Dr Kershaw's argument reveals a different as to why it was that size was important.
+However, Dr Kershaw's argument reveals a different takon on why size is important.
 His comments make it clear that it is actually the large weight of the pumpjet \textit{itself} rather than the additional fuel and batteries it required, that was significant.
@@
 
-However, Dr Kershaw's argument is most comprehensively debunked by the known fact that pumpjets have indeed been successfully retro-fitted to existing smaller diesel-electric submarines.
+However, Dr Kershaw's argument is most comprehensively debunked by the known fact that pumpjets have indeed been successfully retro-fitted to existing smaller diesel-electric submarines, without even necessitating a new ground-up design.
 As one would expect, none of the catastrophic failures eventuated.
 It is reported that the Russian Kilo Class submarine B-871 'Alrosa' has and does take part in exercises, and is remains in active service.
-However, given that no other conventional submarines have since been designed with or converted to pumpjet propulsion, it is difficult to conclude that the benefits of a pumpjet on this type of submarine outweigh the costs.
+However, given that no other conventional submarines have since been designed with or converted to pumpjet propulsion, either in the Russian Navy or any of their export customers, it is difficult to conclude that the benefits of a pumpjet on this type of submarine outweigh the costs.
 It is trivial, however, to conclude that the weight of the pumpjet itself is not prohibitive for a smaller submarine, including  one smaller than the Collins Class, such as the Kilo.
+
+\begin{figure}[h]
+\captionsetup{width=0.7\textwidth}
+\includegraphics[width=0.7\textwidth, center]{alrosa1.png}
+\caption{A `1500 tonne coastal submarine', the Russian Kilo-Class B-871 `Alrosa' with pumpjet fitted.
+Twitter user @AlexLuck9 explains to another why only one of the 22 Kilo-class submarines currently serving in the Russian Navy currently have such a propulsion system.
+The thread originated with a poster from Twitter user @CovertShores, showing that the Alrosa is part of the Russian Black Sea fleet, where Krasnodar is another in this fleet depicted without a pumpjet.
+Within such a restricted body of water such as the Black Sea, the restricted range and endurance imposed by the pumpjet might be considered a tolerable handicap.}
+\label{fig:alrosa1.png}
+\end{figure}
+
+\begin{figure}[h]
+\captionsetup{width=0.7\textwidth}
+\includegraphics[width=0.7\textwidth, center]{alrosa2.png}
+\caption{Despite the heavy pumpjet, the small Russian submarine B-871 has no trouble remaining afloat, and level, during low-speed operations.
+Image Credit: Vitaliy Kostrichenko, as seen on shipspotting.com}
+\label{fig:alrosa2.png}
+\end{figure}
+
+\begin{figure}[h]
+\captionsetup{width=0.7\textwidth}
+\includegraphics[width=0.7\textwidth, center]{alrosa3.png}
+\caption{B-871 seen here with the pumpjet just visible while docked.  Image courtesy of Pakistan Defence as seen in \parencite{pk2018}.}
+\label{fig:alrosa3.png}
+\end{figure}
 
@@
 
-In a clear parallel with the argument that the performance of propulsion systems on submerged bodies is independed of speed because they don't face wave-making resistance, this line of reasoning involves the presentation of a seemingly relevant, technical, yet understandable truth which is used to justify a completely blatant technical falsehood.
-Even if Dr Kershaw is not familiar with this area of science and speaking outside his competence, it is difficult to see how such a pattern of argument occuring so consistently by accident.
+In a clear parallel with the argument that the performance of propulsion systems on submerged bodies is independed of speed because they don't face wave-making resistance, this line of reasoning involves the presentation of a seemingly relevant, comprehensible technical truth which is used to justify a completely blatant technical falsehood.
+Even if Dr Kershaw is not familiar with this area of science and speaking outside his competence, it is difficult to see how such a pattern of argument would occur so consistently by accident.
 For Dr Kershaw to genuinely believe in the arguments that he is making here is almost unthinkable.
 For him to casually or accidentally make such careless arguments also seems implausibly reckless.
-A more likely explanation is that the results of the research already conducted by the government broadly support the conclusions of my own research, and this is well known by Dr Kershaw, and  that these unfortunate statements are being made in a desparate attempt to deflect attention in order to avoid admitting as much.
+An alternative explanation is that the results of the research already conducted by the government broadly support the conclusions of my own research, and this is well known by Dr Kershaw, and  that these unfortunate statements are being made in a desparate attempt to deflect attention in order to avoid admitting this fact.
 
@@
 
-The argument that the 'hull needs to be tuned to pumpjet' and \textit{vice versa} has been used to dismiss critcism of the pumpjet repeatedly by RADM Sammut \parencite{sammut2017} \parencite{bergmann2018}, and also by Dr David Kershaw \parencite{kerr2018}.
+The argument that the `hull needs to be tuned to pumpjet' and \textit{vice versa} has been used to dismiss critcism of the pumpjet repeatedly by RADM Sammut \parencite{sammut2017} \parencite{bergmann2018}, and also by Dr David Kershaw \parencite{kerr2018}.
 It has also been referred to in an answer to a Question on Notice given by Senator Ketter\ref{QON1}.
@@
 
-As has already been noted, there are glaring contradictions from a project-management perspective which oppose this argument.
-In particular, given that we are, according to RADM Sammut accept the pumpjet that was offered by France, and our submarine supposedly has a near identical hull diameter and  length  to the nuclear Barracuda, it is peculiar that any adjustment at all of either the aft section of the submarine, or the pumpjet itself, ought to be contemplated.
+As has already been noted, there are glaring contradictions from a project-management perspective inherent in this argument.
+In particular, given that we are, according to RADM Sammut accepting the pumpjet that was offered by France, and our submarine supposedly has a near identical hull diameter and  length  to the nuclear Barracuda, it is peculiar that any adjustment at all of either the aft section of the submarine, or the pumpjet itself, ought to be contemplated.
 The major advantage that the Naval Group proposal \textit{ought} to have is that those potentially complicated engineering challenges can be safely and entirely avoided.
 Furthermore, the idea that the hull is what necessitates the tuning, as opposed to the speed profile, is almost incomprehensibly naive, as I pointed out in my evidence given to the Senate \ref{SENmorrison}.
-It is noteworthy that no officials from Naval Group have ever alluded to this 'tuning' argument, or any adjustments to be made to the pumpjet or the hull on account of one another.
+It is noteworthy that no officials from Naval Group have ever alluded to this `tuning' argument, or any adjustments to be made to the pumpjet or the hull on account of one another.
 However, it seems pertinent here to give a reasonable discussion of the relevant physics nonetheless, in order that the public might fully assess the sort of argument being made by our most senior defence officials.
 
-\subsection{The impact of the wake-field}
+In \parencite{defence2018} the following argument was made, which builds on that in \ref{QON1}:
+
+\begin{quotation}
+All propulsion devices need to be matched and optimised to the hull with which they will operate.
+Many devices can be designed and tested in an `open water' condition (i.e.
+without the interaction with the hull).
+They are then matched and optimised to the hull and operating conditions.
+However, pumpjets cannot be developed from open water testing.
+The interaction between the aft part of the submarine and the propulsor is much more important than the interaction between a propeller and the hull of a surface ship for the computation of the propulsive efficiency.
+Pumpjet design is fully integrated with hull form design to optimise performance.
+\end{quotation}
+
+This statement appears to be made in almost complete ignorance of the degree to which the hull-propulsor interactions are routinely and comprehensively considered in essentially all propulsor design, including for propellers and pumpjets.
+In almost any design text \parencite{carlton2007}, \parencite{mollard2011}, \parencite{MAN2017}, \parencite{lewis1988}, including those written specifically for submarines \parencite{jackson1992}, it can be seen that the the various efficiency co-efficients for efficiencyare generally amalgamated into a `quasi-propulsive coefficient', called either $QPC$ or $\eta_D$.
+
+\begin{equation}
+\label{QPD}
+\eta_D = \eta_H\eta_O\eta_R
+\end{equation}
+
+The open-water efficiency, $\eta_O$ was dwelt on substantially in my earlier report because this efficiency has the greatest degree of possible variance, (from zero, to around \%75 in some cases) and this variance occurs directly and necessarily with speed.
+The remaining two coefficients describe the interaction between hull and the propulsor.
+$\eta_R$ is the called the relative rotative efficiency, and describes the degree to which the rotating propulsor is more or less efficient operating in a wake, where the water is has an inconsistent velocity across the disc of the rotor.
+$\eta_H$ describes net effect that the propulsor has on the hull by drawing water around its aft end at different speeds and altering the pressure forces on the hull, accounting for the potential for the propulsor to regain some of the energy inerent in the wake behind the hull.
+Both of these effects can be shown to be most significanly driven by the hull shape and the arrangement of the propeller with respect to the hull (its relative diameter, and proximity), and be relatively constant with speed.
+As will be shown in this section, they could plausibly only vary a few percent each, which would have no impact in overturning the collapsing open-water efficiency of a pumpjet operating so far outside of design-conditions, such as a pumpjet at very low speed.
+
+To argue that none of these concepts have any useful bearing or meaning when applied to pumpjets is a bizarre, seemingly childish argument.
+A pumpjet is actually by definition a particuar form of a ducted propeller, specifically one in which the duct is designed to \textit{decelerate} the flow ahead of it the rotor\parencite{oosterveld1970}.
+Essentially all of those same concepts apply, regardless of whether the pumpjet appears occurs on the aft end of a body of revolution, or suspended beneath the hull of a ship.\footnote{It is actually water-jets which are hardest to test in an open-water condition, since their intake is necessarily a tunnel connecting to a hull.}
+
+In the case of a pumpjet mounted on the aft end of a body of revolution such as a torpedo or submarine, for the most important elements in determining efficiency (such as mass flow rate) control lies in the designers hands to adapt the shape of the shroud to that of the hub in order to maintain equivalent volumes between the hub and the shroud, and preserve equivalent streamlines.
+The ratio between the area of the inlet, and the area of the outlet, and the ratio between the velocities of the water stream at the inlet and outlet of the shroud, are fundamental parameters in pumpjet design with regards to efficiency, as can be seen from a reading of Reference \parencite{henderson1964}, or more recently Reference \parencite{abdel2010}.
+And whilst in the discussion of different propulsion mechanisms alternative definitions and nomenclatures may be used in various forms of literature (for instance, to discuss the pump efficiency as a component of the jet efficiency, or account for different non-ideal effects in building up from an ideal model, as seen in Reference \parencite{wislicenus1973}, or because of the focus of that particular study) all will allow for a method of accounting for velocity variance in the incoming flow, and the effect of propulsors action on the boundary layer or wake ahead of it.
+Ultimately, all marine propulsion theory can be related to the acceleration of water, and despite the differences in the different means by which that acceleration occurs, there are natural bounds and limits to the extent to which the impact of accelerated water flow around a hull become incomparable due to the methods of that acceleration occurring.
+
+\begin{tcolorbox}[width=0.9\textwidth, center, colback=green!5!white,colframe=green!75!black, title= \textbf{How a scientist feels about being told the pumpjet efficiency can't be assessed because it depends on the hull}]
+
+Bruce: I'd like to buy some light, breatheable shoes to go running with.
+
+James: Here, have these double-padded, thick-soled, waterpoof, steelcapped, all-weather, leather work-boots.
+
+Bruce: But I don't think those shoes would perform well for running... aren't they uncomfortable?
+
+James: Don't you know that shoes have to operate closely in conjunction with socks?
+Sometimes socks can get crinkled when running, leading to discomfort.
+The thickly padded sole in this boot is designed to aleviate this discomfort, which will help you run faster too!
+
+James: Hmm... actually I've just talked to a few people with similar boots, they all agree they'd be uncomfortable running.
+
+Bruce: How could they know?
+Their shoes could be similar maybe, but not quite the same.
+Besides, they'd all have different types of socks!
+Don't you know that boots have to operate closely in conjunction with socks?
+My boots are special, they're designed to operate \textit{especially} closely with socks.
+It's really impossible to assess them separately.
+
+James: Ok, but I just noticed I'm already wearing the same socks as you, can you tell me how this sock-boot pair works when running?
+
+Bruce: What??!
+oh...
+Hey - didn't I just tell you, boots have to operate in close conjunction with socks!
+Unlike shoes, with boots you can't just choose a pair that suit your purpose and get some socks that match.
+Boots are different!
+The two have to be designed as a pair.
+It's impossible to know how they'll perform without designing them together so they can be tuned to one another, and so we'll have to design your socks and boots together at the same time!
+
+\end{tcolorbox}
+
+\subsection{Relative rotative efficiency, $\eta_R$}
 
 It is true that the hull does indeed change the flow of incoming water to a propulsor.
-The area behind the vessel where the water is on being slightly dragged along with the ship (or the water is moving less quickly past the vessel, in the vessel's reference frame) is known as the wake.
-Since any aft-mounted propulsor has to operate in the wake, or wake-field as it is often called, it is true that wake is of importance to the design of the propolsor.
+The area behind the vessel, where the water is being slightly dragged along with the ship (or the water is moving less quickly past the vessel, in the vessel's reference frame) is known as the wake.
+Since any aft-mounted propulsor has to operate in the wake, (or wake-field as it is often called) it is true that wake is of importance to the design of the propolsor.
 What is not true, however, is that this is grounds for disregarding open-water efficiency, or that it was neglected from consideration in the preparation of my report.
 
-In reality, the operation of a propulsor in the wake-field is an absolutely standard consideration for deriving propulsive efficiencies, as can be seen in any number of standard texts \parencite{Carlton2007}, \parencite{Mollard2011}, \parencite{MAN2017}, \parencite{lewis1988}.
-There is a ratio which is commonly discussed, relative rotative efficiency (denoted $\eta_R$), which has as it's precise purpose to describe the ratio between the efficiency of the propulsor in open water ($\eta_O$), and the efficiency of the propulsor in the wake of the hull ($\eta_B$).
-
-It is widely acknowledged (for example in all of references \parencite{Carlton2007}, \parencite{Mollard2011}, \parencite{MAN2017}, \parencite{lewis1988}) that relative rotative efficiency generally lies around 1, and does not vary greatly.
-For instance section 16.3.2, page 411 of reference \parencite{Mollard2011} says ``$\eta_R$ ranges typically from 0.95 to 1.05 and is often assumed to be unity for preliminary design purposes." \footnote{Observers may note that efficiencies are generally thought to range between 0 and 1.
+In reality, the operation of a propulsor in the wake-field is an absolutely standard consideration for deriving propulsive efficiencies, as can be seen in any number of standard texts \parencite{carlton2007}, \parencite{mollard2011}, \parencite{MAN2017}, \parencite{lewis1988}.
+There is a ratio which is commonly discussed called relative rotative efficiency (denoted $\eta_R$), which is defined as being the ratio between the efficiency of the propulsor in open water ($\eta_O$), and the efficiency of the propulsor in the wake of the hull ($\eta_B$).
+
+It is widely acknowledged (for example in all of references \parencite{carlton2007}, \parencite{mollard2011}, \parencite{MAN2017}, \parencite{lewis1988}) that relative rotative efficiency generally lies around 1, and does not vary greatly.
+For instance section 16.3.2, page 411 of reference \parencite{mollard2011} says ``$\eta_R$ ranges typically from 0.95 to 1.05 and is often assumed to be unity for preliminary design purposes."\footnote{Observers may note that efficiencies are generally thought to range between 0 and 1.
 The possibility that $\eta_R$ could be greater than one is opened up by the fact that it is actually a \textit{relative} efficiency, one efficiency divided by another.
-Some propellers can actually work marginally better in the wake field than outside of it.}  The argument that the efficiency of pumpjets cannot be  assessed using their open-water efficiency is also blatantly false.
+Some propellers can actually work marginally better in the wake field than outside of it if the hull actually induces an advantageous pre-swirl.}
+
+The argument that the efficiency of pumpjets cannot be  assessed using their open-water efficiency is also blatantly false.
 The 'open-water' condition is can be usefully thought of as the normal or 'ideal' efficiency of the propulsor, when the stream of incoming water is undisturbed, and consequently is still a very functional maximum efficiency of a system who's purpose is to smooth out the inconsistency in the flow.
-No higher efficiency could be reasonably expected to be achieved through such  smoothing than would be acheived by beginning with a smooth stream in the first place.
+In the absence of an advantageous pre-swirl, no higher efficiency could be reasonably expected to be achieved through such smoothing than would be acheived by beginning with a smooth stream in the first place.
 Consequently, one can have considerable confidence that the extent to which this 'smoothing' effect might assist a pumpjet, relative to a propeller, is bounded by the extent to which it actually impedes a propellers performance.
@@
 \captionsetup{width=0.7\textwidth}
-\includegraphics[width=0.7\textwidth, center]{roteff1.png}
+\includegraphics[width=0.4\textwidth, center]{roteff1.png}
 \caption{One text's explanations of the variance between open-water efficiency and behind a hull from page 152 of \parencite{lewis1988}}
@@
 \captionsetup{width=0.7\textwidth}
-\includegraphics[width=0.7\textwidth, center]{roteff2.png}
-\caption{One text's explanations of the variance between open-water efficiency and behind a hull from page 131 of \parencite{Carlton2007}}
+\includegraphics[width=0.4\textwidth, center]{roteff2.png}
+\caption{One text's explanations of the variance between open-water efficiency and behind a hull from page 131 of \parencite{carlton2007}}
 \label{fig:roteff2.png}
@@
 \captionsetup{width=0.7\textwidth}
-\includegraphics[width=0.7\textwidth, center]{roteff3.png}
+\includegraphics[width=0.4\textwidth, center]{roteff3.png}
 \caption{One text's desciption of the various types of efficiency from page 14 of \parencite{MAN2017}}
@@
 \captionsetup{width=0.7\textwidth}
-\includegraphics[width=0.7\textwidth, center]{roteff5.png}
+\includegraphics[width=0.4\textwidth, center]{roteff5.png}
 \caption{One text's explanations of the variance between open-water efficiency and behind a hull from page 16 of \parencite{MAN2017}}
@@
 \includegraphics[width=0.7\textwidth, center]{roteff4.png}
-\caption{One text's explanations of the variance between open-water efficiency and behind a hull from page 411 of \parencite{Mollard2011}}
+\caption{One text's explanations of the variance between open-water efficiency and behind a hull from page 411 of \parencite{mollard2011}}
 \label{fig:roteff4.png}
@@
 \end{figure}
+
+\subsection{Hull Efficiency, $\eta_H$}
+Hull efficiency $\eta_H$ is used almost universally to describe the impact of of the propulsor on the hull, which is used both in the discussion of ships and submarines.
+
+Hull efficiency, like relative rotative efficiency, is not a true efficiency in that it is a ratio of two effects which can work in different directions, and consequently can be greater than one.
+The two effects are the impact of the wake fraction on the propulsor, as well as the factor known as the 'thrust deduction' caused by the impact of the propulsor on the hull.
+
+The thrust deduction is comprised of three components \parencite{mollard2011}.
+The first two result in increase in the drag experienced by the hull due to the effect of the propulsor drawing water more quickly across the aft section of the hull than would naturally occur if the vessel were under tow.
+The first component is the increase in frictional resistance, and the second an increase in pressure resistance according to Bernouli (which requires that as speed increases, pressure reduces, sucking the ship backwards).
+The third is that this pressure reduction may reduce the degree of flow separation that occurs on the aft section of the hull created by the adverse pressure gradient that occurs as the water flow slows down.
+This can have an offsetting effect reducing drag somewhat, however the overall combination of these three is essentially always that more thrust is required from the propulsor than the resistance of the hull under tow would require.
+
+The wake fraction represents the average proportal difference between the speed of water that would flow into the propulsor (due to the wake unmodified by the propulsors operation), relative to the vessel speed.
+Since water being dragged along with the ship represents pure energy waste, the extent to which the propulsor can gain thrust by accelerating such water back to stationary state (relative to the ocean clear of the vessel) rather than re-accelating it to a new backwards velocity represents the effective re-capture of energy that was previously lost to the wake.
+
+These effects are well-know and understood, and comprise essential steps in the powering relationships for all surface ships, contrary to the Defence assertion that propellers can be designed without reference to the hull-propulsor interaction, as can be seen from a reading of a number of texts \parencite{carlton2007}, \parencite{mollard2011}, \parencite{MAN2017}, \parencite{lewis1988}.
+The identical concepts are also used for the design of submarines \parencite{jackson1992}.
+
+Hull efficiency tends to be dominated by the shape of the hull, and positioning of the propulsor relative to the hull, and significant changes tend to have counter-acting effects in terms of wake-fraction and thrust deduction.
+Consider a hull with a very blunt (as opposed to tapering) aft section.
+Such a vessel would draw quite a bit of water essentially with it directly behind its stern.
+If the propulsor was mounted very close to the blunt end of the hull (as opposed to further away on an extended shaft) the wake fraction would be very large, which would drive up hull efficiency.
+Countering that effect, the propulsor would be attempting to suck water directly off the blunt aft end section, which would create a significant backwards suck retarding the advance of the hull, increasing the thrust deduction and reducing hull efficiency.
+In contrast, a vessel with a long, gradualy tapering aft and a propulor extended well away from the hull would have a very low wake fraction, but also a very small thrust deduction.
+Those two contrasting vessels will also necessarily require different amounts of thrust to be moved through the water, owing to their contrasting shapes, which (perhaps counterintuitively) is not actually reflected in hull efficiency $\eta_H$, which purely reflects the interaction between the propulsor and the hull.
+
+The way that the wake-fraction and thrust deduction tend to scale together as a product of differing hull geometries in the case of submarines is amply demonstrated in the paper by \underline{Fundamentals of submarine concept design} by CAPT Harry A.
+Jackson \parencite{jackson1992}.
+A third plot, reproduced here in Figure \ref{fig:hulleff.png} demonstrates how these two values combine to produce the hull efficiency.
+As can be clearly seen, the size of the propeler relative to the hull also has an impact, and alters the hull efficiency by approximately 15\% when the propulsor size is halved or doubled.\footnote{The propulor size also impacts open water efficiency, so an global optimum may be constrained away from the hull-efficiency optimum.}  Given that one of the key features of the pumpjet is that they are smaller in total diameter to a comparable propeller, (and hence accelerate a smaller water-column to greater speeds) this provides a useful indication of the likley degree of variance that might be expected.
+However, the size differences are unlikely to be nearly as large as a factor of two.
+Reference \parencite{henderson1964}, titled \underline{A Method for the Design of Pumpjets} indicates on page that 15-20\% less diameter than an equivalent propeller is a suitable starting point for the optimisation of the mass-flow rate (which will be crucial for determining open-water efficiency, or its analogue).
+As such, it would be reasonable to conclude that something of the order of 3\% is more likely to be the plausible degree of variation in the hull efficiency that might occur due to the selection of a pumpjet over a propeller.
+
+\begin{figure}[h]
+
+\begin{subfigure}{0.5\textwidth}
+\includegraphics[width=0.9\linewidth]{hulleff1.png}
+\caption{}
+\end{subfigure}
+\begin{subfigure}{0.5\textwidth}
+\includegraphics[width=0.9\linewidth]{hulleff2.png}
+\caption{}
+\end{subfigure}
+\caption{Dramatic changes in propeller size on a submarine (doubling/halving) would only produce variance of appoximately 15\% in hull efficiency, according to Reference \parencite{jackson1992}.
+A pumpjet, which be smaller in diameter than a propeller by about \%20, would likely only make a few percentage points of difference due to changed flow over the hull ahead of it.}
+\label{fig:hulleff.png}
+\end{figure}
+
+\subsection{Open water efficiency, $\eta_0$}
+In clear contrast to the trivial changes that the relative rotative efficiency $\eta_R$ incurs, the open-water efficiency $\eta_O$ of a propulsor varies between 0\% and 75\%, depending crucially on the advance ratio $J$, which is highly speed dependent due to the changing resistance of the vessel at different speeds.
+In \parencite{morrison2018} it was shown that it was most probable a pumpjet's open-water efficiency could be around half that of a propeller at speeds at or below 5kt.
+It is consequently vastly more important to give detailed consideration to early in the design and selection of a propulsor, particularly for a diesel-electric submarine.
+Unless defence is proposing that a pumpjet can have a relative rotative efficiency or hull efficiency of more than ~200\%, neither of which are remotely plausible, there is no feasibly means that this line of argument can sustain their dismissal of my conclusion.
+
+As can be seen in any number of texts, the open-water efficiency $\eta_0$ of a propulsor goes through a wide range from zero, to a peak, then back to zero, as a function of a variable called $J$.
+$J$ is known as the advance ratio, and can intuitively be thought of as how much the propulsor moves forward through the water that it's operating in per a given degree of rotation.
+Since the resistance that water exerts on a body \textit{does} change with speed (it scales with $V^2$), contrary to defence's claims, $J$ actually has a counter-intuitive relationship with speed.
+Low values of $J$ correspond to conditions where a given vessel is actually moving very fast, since the high resistance encountered means that it must actually turn around \textit{more} times, relative to the distance it advances.
+It is actually at some high value of $J$ that corresponds to a value of extremely slow waterspeed, approaching stationarity.
+In such as cases, since the resistance is extremely low, the propeller actually advances the maximum distance that it can through the water for a given revolution, akin to a corkscrew advancing through a cork, where there is no `slippage' throught the water.
+As can be seen from the shape of the curves shown for an open propeller in Figure \ref{fig:OpenWater.png}, efficiency falls off to zero at such a point where $J$ is equal to the $P/D$ label corresponding to each line.
+Without labouring the mathematical proofs, this is because $J$ being equal to $P/D$ is essentially the definition of the corkscrew condition being met.\footnote{Pitch is defined as a distance per revolutions, so $P/D$ expresses this as unitless fraction of the diameter.
+The distance is determined by the path that the blade angle would trace in a helix, corkscrew fashion, but since blades tend to have some degree of twist, which means that this angle varies depending on which point along the blade (radially) is chosen.
+Though conventions do exist, a small degree of variance is necessarily introduced, and consequently efficiency falls to zero close to, but not necessarily precisely on the $P/D = J$ condition being met.} When $J = 0$, the opposition condition is true, and the resistance is so high that the propulsor is unable to move through any water, and is spinning in one place, egg-beater style.
+As one would expect, almost all operations, and the optimal operating condition, occurs somewhere in between.
+
+It is important to point out here that these plots are carefully constructed in order to ensure that all the values represented are unitless, so that a scale-invariance can be applied, so that scale-models in tank-tests can be use to meaningfully inform the development of a propulsor for a variety of vessel sizes and speeds.
+A direct, and desirable corollary of this design is that is conceptual equivalence between different methods of increasing the resistance encountered by any given propeller.
+Specifically, using the same propeller to push a much bigger and heavier hull will push the operating scenario to lower $J$ values in exactly the same way that demanding the propeller to push the original vessel much faster will.
+Conversely, fitting an extremely large propeller to a small load will allow the propeller to advance in nearly a corkscrew fashion (high $J$), just as requiring only an extremely low speed (and hence very low drag resistance) will achieve the same effect.
+
+This equivalence leads to the common misunderstanding that accelerating ducted propellers are well suited to low-speed operations, such as those undertaken by tugs and trawlers.
+This is not the case.
+In reality, accelerating ducts are suited to propellers that are highly loaded (low $J$), which is the same end of the spectrum that a any propulsion system moves if it attempt to push a given vessel to higher speeds.
+In reality, tugs and trawlers operate at unusually high $J$ values because their propeller size is constrained.
+If a tug were to chose an ideally sized propeller to push around a 200,000 tonne bulk carrier, as it's routine job might be, it would naturally choose a similar sized propeller to that used by the 200,000 tonne ship it was attempting to push.
+However, such a propeller would probably sink the tug, and if not the torque require to rotate it would likely cause it to capsize.
+With a constrained propeller size, a tug under load actually experience exactly the same propeller conditions that an over-powered speedboat, attempting speeds higher than its propeller was designed for, would experience.
+In general, vessels which have high speed as their primary design consideration do not have such a propeller size constraint, and can choose to have an optimal propeller design which would be `lightly' loaded in comparison to a tugs, even when travelling at high speeds.
+Of course, at lower speeds, the propeller would still be far less loaded again.
+
+Two other lines are generally present in these graphs, which are the lines which indicate $K_T$, which represent the thrust coefficient, and $K_Q$, which represents the torque coefficient.
+As can be seen in Figure \ref{fig:OpenWater.png}, for any given propeller geometry ($P/D$ value) the thrust coeffient $K_T$ reaches zero at the same point that the efficiency $\eta_0$ does at a high $J$ value.
+This corresponds to zero waterspeed, when no effective forward thust is produced by the propulsor.
+It will be noted at this point that the lines corresponding to $K_Q$ have not quite reached zero.
+Some small amount of torque (force directed to rotating the rotor) is still applied, but no thrust results.
+Consequently efficiency at this point is zero.
+It should also be noted at this point that the unitless nature of these coefficients demands mathematical constructions which are non-linear.
+As such, there is no directly relationship between $J$, $K_Q$ or $K_T$ which can be used to determine a corresponding united measurements (e.g.
+waterspeed, rpm, power etc) unless \textit{all} of the necessary assumptions are made and equations solved, and similarly fractions or ratios of coefficients hold to their conceptual analogues.
+($K_T$ of 0.5 is not half a the thrust of a $K_T$ of 1.0).
+However, at the extremities and boundaries one may usefully describe what is happening.
+Important to note is that all efficiencies approach zero at $J = 0$, since this describes the egg-beater situation, where the resistance is so great that no movement can be achieved (and hence no `work' done), such as the 'bollard-pull' condition of attempting to move an immoveable object.
+Thrust and torque are non-zero, but the failure to advance any further means that efficiency is zero.
+Open-water efficiency measures the ability for a propulsor to convert rotational force into linear thust to achieve the greatest movement in that linear direction.
+Without movement, this doesn't occur.
+
+The difference that applying the two different types of nozzle (accelerating and decelerating, as shown in Figure \ref{fig:nozza.png}) has on the efficiency of rotors is illustrated by comparing Figure \ref{fig:OpenWater.png} to Figures \ref{fig:nozz19b} and \ref{fig:nozz32b} which show the corresponding open-water efficiency plots.
+The first distinguishing feature that should be noted is the point at which the efficiency curves fall to zero at higher $J$.
+Instead of falling almost exactly at a $J$ value equal to the $P/D$ value, as in the case of an open propeller, an accelerating duct shifts this point to lower $J$ values, and a decelerating duct to higher $J$ values.
+This directly corresponds to the impact that the duct has on the water speed that arrives at the \textit{rotor}, versus the water speed that arrives at the entry to the duct of the propulsor.
+Since the rotor in the accelerating duct is actually pushing through water moving \textit{faster} than the water arriving at the mouth of the duct, this condition where forwards thrust (and waterspeed) fall to zero occurs when the propulsor is actually advancing through less water per revolution of the rotor.
+In the decelerating case, the opposite happens.
+Since the water is slowed down before the rotor acts on it, the propulsor as a whole is actually advancing even further through the water than the `corkscrew' condition would apply as assessd by the revolutions of the rotor.
+This is a general feature of the contrasting nature of decelerating and accelerating ducts.
+
+Unchanged in all three diagrams, however, is the coincidence of $K_T$ and $\eta_0$ reaching zero at the same point, corresponding to going so slowly as to come to a stop.
+Likewise, at infinite load (or conceptually, infinite speed) the efficiency also falls to zero, as it must when the egg-beater condition occurs.
+Either no advance is gained for the next rotations, or infinite rotations must occur for to reach the next advance increment.
+
+Naturally we can assume that these particular shrouds and propellers don't exactly represent those to be used on a future submarine, but some of the most significant features are reliably illustrative of the general contrasts between unshrouded rotors, and the propulsors resulting from shrouding them in  decelerating and decelerating ducts.
+It might well be noted that both of the ducted propellers have slightly lower peak efficiency than the open one.
+The shape of the efficiency curve for the accelerating case is broadly similar to the unshrouded version, whereas the decelerating curve has a clearly elongated, flattened top, being able to undertake a large increase in loading without detrimental impacts on the performance of the propulsor, largely due to the elevated pressure alleviating the increasing occurrence of cavitation.
+
+However, the most significant difference between the two different curves is the addition of a fourth set of lines at the bottom of the graph, denoted $K_{TD}$.
+These represent the thust that is produced by the duct itself.
+As defined in Reference \parencite{oosterveld1970} the $\eta_0$ line represents only the effect of the thrust generated by the rotor.
+In order to determine the efficiency of the propulsor as a whole the impact of the duct itself must be taken into account.
+In the case of the accelerating duct, it can be see that the the duct can have a positive thrust, which can be very significant relative to the actual thrust produced by the rotor, and which remains positive for the majority of the optimal $J$ values available to the propulsor.
+This has the effect of substantially `arching' the back of a whole-system efficiency curve, allowing relatively good efficiencies to be achieved at very high loadings, or low $J$ values.
+For any given propeller curve it can be seen that at some high $J$ value, corresponding to low loading (or low speed, for a given propulsor geometry and vessel) the shroud still has a net negative impact, as the additional drag overwhelmes the benefits achieved by the distribution of pressure across a larger surface area.
+
+The decelerating case represents a very different case.
+In this scenario, the net force applied by the duct is negative for the majority of the useful $J$ values, and clearly remains negative right up to the highest $J$ values, right up to the stationary condition.
+If this effect were to be integrated into a whole-system efficiency curve, this would have the effect of significantly crimping the brow of the curve on the right hand side.
+This would mean that as the a vessel slowed and the propulsor unloaded (increased $J$), instead of climbing ever-higher until falling sharply only at extremely low speeds the when the corkscrew condition was almost met (as in the case for an open propeller) efficiency would begin to decline sooner, from non-trival speeds.
+This is the impact that a decelerating shroud has at very low speeds: when it attempts to decelerate a flow that is already going very slowly, and the imposed resistance - desirably at high speeds in preventing cavitation - disrupts the flow excessively and wastes energy creating vortices and turbulence.
+
+\begin{figure}[h]
+\captionsetup{width=0.8\textwidth}
+\includegraphics[width=0.8\textwidth]{OpenWater.png}
+\caption{A typical chart showing open water efficiency $\eta_0$ of a propeller.
+Note that $\eta_0$ falls to zero at a $J$ value which approximately equals $P/D$.
+Image Courtesy: MARIN, as shown in reference \parencite{carlton2007}.}
+\label{fig:OpenWater.png}
+\end{figure}
+
+\begin{figure}[h]
+
+\begin{subfigure}{0.5\textwidth}
+\includegraphics[width=0.9\linewidth]{nozz19a.png}
+\caption{An accelerating duct shape}
+\end{subfigure}
+\begin{subfigure}{0.5\textwidth}
+\includegraphics[width=0.9\linewidth]{nozz32a.png}
+\caption{A decelerating duct shape}
+\end{subfigure}
+\caption{Where an accelerating duct (a) contracts in advance of the rotor, accelerating the flow before it arrives, a decelerating duct shown around an illustrative rotor (b), also known as a pumpjet, contracts after the rotor, slowing the flow in advance of the pumpjet and forcing acceleration to occur afterwards.
+Images as shown in Reference \parencite{oosterveld1970}.
+}
+\label{fig:nozza.png}
+\end{figure}
+
+\begin{figure}[h]
+\captionsetup{width=0.8\textwidth}
+\includegraphics[width=0.9\linewidth]{nozz19b.png}
+\caption{Open-water test results for an accelerating duct illustrate a similar $\eta_0$ as an open propeller, additional thust is derived from the duct itself $K_{TD}$ at low $J$ values, enhancing efficiency at high speeds.
+Images as shown in Reference \parencite{oosterveld1970}.}
+\label{fig:nozz19b.png}
+\end{figure}
+
+\begin{figure}[h]
+\captionsetup{width=0.8\textwidth}
+\includegraphics[width=0.9\linewidth]{nozz32b.png}
+\caption{Open-water test results for a decelerating duct exihibits a flatter top for the rotor's efficiency, owing to reduced cavitation as speed increases.
+However, the duct imposes a substantial negative thrust at higher $J$ values, degrading efficiency at lower speeds.
+Images as shown in Reference \parencite{oosterveld1970}.}
+\label{fig:nozz32b.png}
+\end{figure}
+
+% A particularly desparate line of argument might lead one to argue that such arguments do not apply to submarines propulsors due to their great level of depth and the unique shape of their hull.
+This would not be true.
+Owing to their generally cylindical shape and generally conical afterbody, submarines tend to have wake fields are much less complex and intense in the degree of their variation that surface ships.
+This is because ships bodies also have wave-making effects involved as well as a hull that tapers above the propulosor, creating an uneven effect with different heights.
+In contrast, wake fields of submarines have a relatively even gradient in wake fraction (the intensity of waterspeed difference due to the hull) in a radial direction, with only the control surfaces ahead of the propulsor causing substantial variations.
+Furthermore, at very low speeds, where pumpjets were shown to have the lowest efficiecies, the effect of the wake field would be the least compared to the open-water case, since the size of the boundary layer about the aft of the hull would be substantially diminished.
+On absolutely all accounts, the interaction between the hull of a submarine and propulsor at low speeds cannot be credibly argued to have any impact which would overturn the conclusion reached about pumpjets having clearly and substantially inferior efficiency at low speed, as argued in \parencite{morrison2018}.
+
+As such, the reason that my earlier paper did not explicitly discuss relative rotative efficiency and the impact of the wake field, this was due to it being so obviously and demonstrably of no bearing whatsoever on my conclusions.
+Perhaps I had falsely assumed that as much would immediately clear to any technically competent reviewer, who would readily grasp that open-water efficiency is the primary parameter which determines the efficiency of a propulsor as it pertains to the loading of the propulsor, and consequently speed.
+The use of this argument displays a substantial degree of ignorance of the relevant physics, as clearly described in numerous and publically available sources.
 
@@
 Whilst \textit{on average} the boundary layer moves closer to the surface speed closer to it, and slower further away from it, at any local point on a tiny lengthscale the movement of the water will be chaotic, and altogether very non-parallel.
-Turbulent boundary layers tend to be much thicker, and become even thicker at higher velocities, resulting in larger amounts of energy being sapped producing the faster, more intense chaotic swirling motions that comprise turblent flows.
-
-This distinction is important in understanding the limitations of a pumpjets claim to make the incoming stream of water less inconsistent, or more regular.
-The duct intake, and any stator blades in advance of the rotor can be effective in making the flow more homogenous, but in doing so they simultaneously make the flow more turbulent, since the necessarily involve passing the water over more surfaces, rather than less.
+Turbulent boundary layers tend to be much thicker, and become even thicker at higher velocities, resulting in larger amounts of energy being sapped producing the faster, more intensely chaotic swirling motions that comprise turblent flows.
+
+This distinction is important in understanding the limitations of a pumpjet's claim to make the incoming stream of water less inconsistent, or more regular.
+The duct intake and any stator blades in advance of the rotor can be effective in making the flow more homogenous as assessed on a macroscopic scale, but in doing so they simultaneously make the flow more turbulent at a smaller scale, since the necessarily involve passing the water over more surfaces, rather than less.
 This can be advantageous to the extent that very large-scale and intense vortices might be broken down into smaller-scale vortices, which appears more homogenous when assessed on the length-scale relevant to the blades of the rotors.
@@
 \footnote{It is this effect that is likely referred to in comments made by Professor Joubert in a paper commissioned by DSTO in 2004 \parencite{joubert2004}.
-However, with competently designed appendages and fairings, the kind of unsteady vortex shedding which would see such large-scale eddies enter the propulsor should be safely avoided.
-Substantial effort has been devoted to developing the smoothest possible wake-field for submarines for this purpose.
-\parencite{toxopeus2014}}
+However, with competently designed appendages and fairings, the kind of unsteady vortex shedding which would see such large-scale eddies shed into the propulsor should be safely avoided.
+Substantial effort has been devoted to developing the smoothest possible wake-field for submarines for this purpose.\parencite{toxopeus2014}}
 
@@
 The circumstances where a pumpjet is able to attain higher efficiency are those where the total amount of additional productive energy (thrust, produced by front to back acceleration) added to the water column dramatically exceeds the losses incurred by the additional turbulence (round and round acceleration).
-This occurs at high speeds.
+This occurs at high speeds, which necessitate a high level of turublence which a propeller is unable to avoid.
 
 It should be noted that a pumpjet's duct also has the effect of decelerating the incoming flow, which increases its static pressure.\footnote{In fact, a pumpjet is by definition a decelerating ducted propeller, contrary to claim that they are so different as to be incomparable made \ref{QON1}.} Whilst this has the desirable effect of further reducing the likelyhood of cavitation occurring, it also creates an `adverse pressure gradient' in the intake.
-This means that the water must move towards an area of higher pressure, agains the direction that the pressure would naturally push it.
+This means that the water must move towards an area of higher pressure, against the direction that the pressure would naturally push it.
 Whilst at high speeds the water has enough momentum to easily overcome this resistance, at very low speeds there isn't enough energy in the flow for the entire water column to move consistently in this direction, and part of the flow simply turns around and moves in the opposite direction, in a phenomenon known as `flow separation'.
-This occurs against a surface (in this case just inside the pumpjet intake), where the boundary layer has to move more slowly than the remainder of the stream, and in  this case it is the first part to reverse direction.
+This occurs ajacent to surface (in this case just inside the pumpjet intake), where the boundary layer has to move more slowly than the remainder of the stream, and in  this case it is the first part to reverse direction.
 Whilst the entirety of the boundary layer, (or indeed the entire flow) may actually be turbulent, with many very small-scale vortices, this phenomenon creates new large-scale macroscopic vortices, which could form, grow and shed in an unsteady creating exactly the same risk of perodic shedding (i.e.
@@
 Since open propellers do not have a duct which creates an adverse pressure gradient, they do not face this problem at low speeds.
-In fact, at very low speeds more substantial parts of the blade may actually have laminar flow over them, even further enhancing efficiency.
+In fact, at very low speeds more substantial parts of a propeller blade may actually have laminar flow over them, even further enhancing efficiency.
 
@@
 
-In clear contrast to the trivial changes that the relative rotative efficiency $\eta_R$ incurs,  the open-water efficiency $\eta_O$ of a propulsor varies between 0\% and 75\%, depending crucially on the advance ratio $J$, which is highly speed dependent due to the changing resistance of the vessel at different speeds.
-In \parencite{morrison2018} it was shown that it was most probable a pumpjet's open-water efficiency could be around half that of a propeller at speeds at or below 5kt.
-It is consequently vastly more important to give detailed consideration to early in the design and selection of a propulsor, particularly for a diesel-electric submarine.
-Unless defence is proposing that a pumpjet can have a relative rotative efficiency of more than ~\%200, there is no feasibly means that this line of argument can sustain their dismissal of my conclusion.
-
-A particularly desparate line of argument might lead one to argue that such arguments do not apply to submarines propulsors due to their great level of depth and the unique shape of their hull.
-This would not be true.
-Owing to their generally cylindical shape and generally conical afterbody, submarines tend to have wake fields are much less complex and intense in the degree of their variation that surface ships.
-This is because ships bodies also have wave-making effects involved as well as a hull that tapers above the propulosor, creating an uneven effect with different heights.
-In contrast, wake fields of propellers have a relatively even gradient in wake fraction (the intensity of waterspeed difference due to the hull) in a radial direction, with only the control surfaces ahead of the propulsor causing substantial variations.
-Furthermore, at very low speeds, where pumpjets were shown to have the lowest efficiecies, the effect of the wake field would be the least compared to the open-water case, since the size of the boundary layer about the aft of the hull would be substantially diminished.
-On absolutely all accounts, the interaction between the hull of a submarine and propulsor at low speeds cannot be credibly argued to have any impact which would overturn the conclusion reached about pumpjets having clearly and substantially inferior efficiency at low speed, as argued in \parencite{morrison2018}.
-
-As such, the reason that my earlier paper did not explicitly discuss relative rotative efficiency and the impact of the wake field, this was due to it being so obviously and demonstrably of no bearing whatsoever on my conclusions, and I had assume that as much would immediately clear to any technically competent reviewer.
-The use of this argument displays a degree of ignorance of the physics, as clearly described in numerous and publically available sources, that it is difficult to account for by incompetence alone.
-A substantial measure of contempt for public intelligence must be required in order to appeal to such trivially simple statements of truth in order to support conclusions which can be so trivially demonstrated to be completely false.
-
-Imagine that you came to buy a coffee from a your favourite cafe, only to find that they weren't selling any coffee that day, and wouldn't be selling any more indefinitely.
+\section{Discussion}
+Having given due consideration to the three signiciant arguments that have been made in response to my paper, one can only conclude that either a degree of incompetence hitherto unthinkeable pervades defence, or alternatively the arguments are disingenuous.
+
+% \begin{tcolorbox}[width=0.9\textwidth, center, colback=green!5!white,colframe=green!75!black, title= \textbf{How a scientist feels about the pumpjet smoothing argument}]
+% Imagine that you came to buy a coffee from a your favourite cafe, only to find that they weren't selling any coffee that day, and wouldn't be selling any more indefinitely.
 Their coffee maching had been removed, and they were only selling cold-drip, coffee, stored in little bottles in the fridge.
 Overcoming initial shock (since you weren't aware of any other cafes who would make such a bizarre decision as to not sell espresso-coffee as a substantial part of their offering), you made some tentative inquiries as to why.
-
-At first, you're told vaguely that it was to do with health and safety.
+%
+% At first, you're told vaguely that it was to do with health and safety.
 This seemed a little odd, since there obvious benefits to everyone's enjoyment and presumeably the business, which you point out to the cafe owner, and you ask how the health and safety risk could really be so large as to outweigh that.
@@
 In anycase, it sounded like it could be very technical, and you decided to trust your very experienced and seemingly intelligent barista and forget about your lingering doubts as to why no other cafes were following suit.
-
-Then later you find a notice posted up at the front of the cafe, explaining fully the rationale behind the abandmonment of the coffee machine, and the connection with temperature.
+%
+% Then later you find a notice posted up at the front of the cafe, explaining fully the rationale behind the abandmonment of the coffee machine, and the connection with temperature.
 In it, it is explained that all espresso-based coffees were hot, those with milk around 65 or 70 degrees, and long blacks or fresh espresso potentially above 90 degrees.
@@
 Therefore in the interest of health and safety no more hot drinks would be served, and cold-drip would be the only coffee available.
-
-You read the statement again, not quite believing that a hazard so obviously known and successfully navigated by millions of coffee drinkers every day, could be the cause of such a decision.
-Althought it's true the owner didn't lie about it being a health and safety concern, and can't dispute the facts about the coffee temperatures being above 65 degrees, you still conclude that owner is either insane, or trying to disguise some other reason for abandoning the espresso coffee.
+%
+% You read the statement again, not quite believing that a hazard so obviously known and successfully navigated by millions of coffee drinkers every day, could be the cause of such a decision.
+Althought it's true the owner didn't lie about it being a health and safety concern, and you can't dispute the facts about the coffee temperatures being above 65 degrees, you still conclude that owner is either insane, or trying to disguise some other reason for abandoning the espresso coffee.
+% \end{tcolorbox}
 

### 2018-09-20 Second.Rnw
with the crucial float barriers

@@
 \setlength{\parindent}{0em}
+
+\usepackage{placeins}
 
@@
 
+\FloatBarrier
+
 \newpage
@@
 
+\FloatBarrier
+
 \section{The weight and size argument}
+
 Both Dr David Kershaw, Chief of Maritime Division at DSTG\parencite{kerr2018}, as well as Mr Brent Clark\parencite{clark2018} have both made arugments to the effect of suggesting that it was the Australian submarine's very large size which is what made the pumpjet viable, where it might not be for smaller submarines.
@@
 An alternative explanation is that the results of the research already conducted by the government broadly support the conclusions of my own research, and this is well known by Dr Kershaw, and  that these unfortunate statements are being made in a desparate attempt to deflect attention in order to avoid admitting this fact.
+
+\begin{tcolorbox}[width=0.9\textwidth, center, colback=green!5!white,colframe=green!75!black, title= \textbf{How a scientist feels being told that the pumpjet is a gamechanger for our diesel-electric submarine because it is so large, where it wouldn't make sense on a smaller one}]
+
+Bruce: I'm not sure about using these fat nobbly tires on my road-bike for the cycle tour... No-one else is using such tyres on a road-bike, I think they'd be slow and inefficient.
+
+James: Don't be silly!
+Everyone else would love to use fat nobbly tyres on a road-bike, the only reason they can't is that those cyclists themselves are just so slim and athletic, and can't use the nobbly fat nobbly tyes.
+Bruce, don't you realise that you're the \textit{only} cyclist every attempting this tour who weights 200kg, roughly three times what most road-cyclistst weigh?
+If a 70kg rider attempted those speeds with fat knobbly tyres, they wouldn't have the weight to stabilise the forces involved, and they'd be thrown off their bicycle in a catastophic crash which would probably be fatal.
+But for fat cyclists such as yourself, fat nobbly tires are a real game-changer!
+
+Bruce:  What about that one lonely guy over there riding on fat knobbly tyres.
+He looks like he weighs about 70kg, and he's not having a catastopic crash, he's just going really slow and looks tired.
+Maybe that's why he's sticking to the footpath.
+
+James: Oh...
+
+Bruce: And didn't you also tell me that you tried riding with fat nobbly tires once years ago?  How did you survive the catastophic crash?
+
+\end{tcolorbox}
+
+\FloatBarrier
 
@@
 
+\FloatBarrier
+
 \section{Discussion}
-Having given due consideration to the three signiciant arguments that have been made in response to my paper, one can only conclude that either a degree of incompetence hitherto unthinkeable pervades defence, or alternatively the arguments are disingenuous.
 
@@
 % \end{tcolorbox}
+
+\FloatBarrier
 
@@
 \newpage
+
 \subsection{Answer to Question on Notice}\label{QON148}
@@
 \begin{figure}[h]
-\includegraphics[width=0.8\textwidth, center]{QON148.png}
+\includegraphics[width=0.7\textwidth, center]{QON148.png}
 \end{figure}
@@
 \newpage
+
 \subsection{Answer to Question on Notice}\label{QON150}
@@
 \newpage
+
 \subsection{Answer to Question on Notice}\label{QON151}
+
 \begin{figure}[h]
@@
 \newpage
+
 \subsection{Answer to Question on Notice}\label{QON1}
+
 \begin{figure}[h]
@@
 \newpage
+
 \subsection{Answer to Question on Notice}\label{QON2}
+
 \begin{figure}[h]
@@
 \newpage
+
 \subsection{Answer to Question on Notice}\label{QON3}
+
 \begin{figure}[h]
@@
 \newpage
+
 \subsection{Hansard Extract, Senate Economic References Committee Public Hearing - 7th June 2018 \cite{SEN2018morrison}}\label{SENmorrison}

### 2018-09-24 Second.Rnw
Final Draft

@@
 
-% \usepackage{draftwatermark}
-% \SetWatermarkText{DRAFT}
-% \SetWatermarkScale{1.5}
-% \SetWatermarkLightness{0.95}
+\usepackage{draftwatermark}
+\SetWatermarkText{DRAFT}
+\SetWatermarkScale{1.5}
+\SetWatermarkLightness{0.95}
 
@@
 
+\section{Executive Summary}
+In January 2018, I produced a detailed report\parencite{morrison2018}, commissioned by Gary Johnston from Submarines for Australia, in order to examine the question of the efficiency of a pumpjet.
+Pumpjets, whilst common on nuclear submarines which can operate at high speeds (above 25kt) for sustained periods, are generally not used on diesel-electric submarine, which need to operate at very low speeds (around 10kt during transit, and under 5kt during combat operations).
+Considerable emphasis had been placed on the significancw of the pumpjet proposed in successful French bid made by DCNS (now Naval Group) for Australia's Future Submarine.
+I concluded that an inevitable characateristic of pumpjets - directly necessitated by the characteristics which gave them good acoustic performance at high speeds - was very low propulsive performance (efficiency) at low speeds.
+This would mean that a diesel-electric submarine would suffer poorer range and endurance, and significantly worse dived range and endurance, as well as a worse indiscretion ratio due to the use of pumpjets.
+Given the profound importance of these characteristics to all diesel-electric submarines, aswell as elevated importance of range and endurance to the Australian requirement, it would appear that a pumpjet can be safely regarded as entirely unsuitable for Australia's Future Submarine.
+
+On June 7 2018, I appeared as a witness before the Senate Economic References Committee \parencite{SEN2018morrison}, where I defended and expanded upon my results, including in the presence of senior officials from Defence and Naval Group, who steadfastly refused to comment.
+The Committee directed Defence to provide a written response to my paper prior to the conclusion of the Inquiry Into the Future Of Naval Shipbuilding Industy in Australia on 27th of June, which they finally produced in mid September.
+Submarines for Australia had already commissioned me to write a subsequent report in reply to the responses that had appeared in responses to Questions on Notice in Parliament, and also in the media.
+With Defence's submission finally received, this paper is in a position to comprehensively assess the totalality of the arguments put by Defence against the conclusions of my previous paper.
+
+With the receipt of the written report it has become clear that Defence has no credible basis for rejecting the findings of my research, and instead has disgraced their technical credibility with utterly ridiculous misrepresentations of the relevant physics.
+Their claim that pumpjets can be efficient across the entire speed range are instead based on the assumption that the resistance faced by fully submerged bodies is ``essentially constant over speed"\parencite{defence2018} \parencite{kerr2018}.
+This is simply untrue.
+It is a bizarre, irrational claim with no basis whatsoever in physics.
+It is frankly bewildering that such a claim could be made, given how easily it's falsehood can be established by even moderate research, or simple logic.
+If it were true, then submarines and torpedoes would be able to elevate their speed without increased resistance, and consequently reach any speed they might choose without requiring larger motors, or more fuel, and hence attain virtually unlimited range.
+
+The truth is that the resistance of submerged bodies in the real world scales with velocity squared.
+This exponential increase in resistance explains why conventional submarines slow to extremely low speeds in order to conserve energy when dived, and has been used by other respected physicists discussing this topic \parencite{davies2017} as well as essentially all scientific literature involved hydrodynamics.
+It also explains why a propulsor, which must faces one hundred times as much resistance when travelling at 30kt as it does at 3kt, behaves very differently in these regimes, and consequently has a different efficiency.
+For Defence, assisted by DSTG, to claim that the relationshp between drag resistance and speed under-water is constant is an extremely shocking revelation.
+In the engineering/technical community, it would be just as strange to hear that a dentist with teenage children believes in the tooth-fairy.
+Not only is is their job to know otherwise, but abundant life experience would have continually proven otherwise.
+
+An unavoidable conclusion is that claims made by Defence to Parliament in response to Questions on Notice claiming that the pumpjet can be efficient across the entire speed range are entirely baseless and extremely misleading.
+
+Such a revelation necessitates a further question regarding whether such misleading claims where made out of complete ignorance and incompetence, or whether they were deliberately intended to deceive.
+As can be seen from the chronology of public comments which this report details, it is difficult to see how Defence could hold such a naive and embarrassing view after having their attention drawn to the opposing evidence multiple times, and devoted so much resources to understanding the relevant technologies.
+In 2012 Senator Johnston challenged RADM Moffit directly as to why research money was even being directed towards pumpjets, given how obviously inappropriate they were for our use, but RADM Moffit insisted they would still do the work required.
+In 2017 Defence revealed that a 99 page report was commissioned on pumpjets as part of \$214 million in 'de-risking' research, and that over a million dollars had subsequently been spent on pumpjet research specifically.
+
+Notwithstanding the improbability that the incompetence of Defence is so completely pervasive that all of this research failed to yeild even a basic understanding of hydrodynamics, the probability that the necessary degree of ignorance has occurred through arrangement rather than accident can be most firmly established by three simple propositions, which are made obvious through a reading of this report:
+
+\begin{enumerate}
+\item  \textbf{The French know the answer.}  As one of only two countries in the world who have designed and produced both pumpjets and propellers, nuclear and conventional submarines, and fitted and operated each type of propulsor on both types of submarines, there can be no doubt whatsoever that Naval Group and the French Government do have well-formed opinion about the performance of pumpjets on diesel-electric submarines.
+They would have extensive operational performance data from their in-service SSBNS which use the pumpjet.
+They would have extremely detailed modelling, and tank-test results from the pumpjet from the almost-complete Barracuda class, which has an identical hull diameter to the proposed Shortfin Barracuda.
+The would have extensive operational performance data from their own Rubis Class nuclear submarines, which still use propellers, as well as the results of earlier tests, confirmed to have occurred by Naval Group Officials, where a pumpjet was fitted to a diesel-electric Agosta Class submarine.
+That they have adequate performance data from all the necessary reference points to assess the likely performance of the pumpjet on our own is entirely beyond dispute.
+
+\item  \textbf{The French gave the answer to the Australians.} It would naturally be a matter of profound neglect on the part of Australian officials to fail to ascertain the necessary performance data - both acoustic and propulsive efficiency - from the French during a Competitive Evaluation Process.
+Any doubt that this occurred, however, can be reliably dismissed on account of public statements made by DCNS during the CEP claiming proudly that ``The stealth and hydrodynamic performances of pump jet propulsion are of course classified and in Australia known only to DCNS and the Australian Government."\parencite{costello2016}.
+Given RADM Sammut's repeated statements \parencite{SEN2016sammut} \parencite{SEN2017hanson} to Senate Estimates that  ``the way in which technical data would be transferred to Australia and the way in which intellectual property would be managed" was one of the reasons that DCNS was selected as design partner, and DCNS's clear pride in  ``the offer from France to transfer to Australia sovereign control and use of pump-jet propulsion technology" there can also be no doubt that after the conclusion of the CEP and the commencement of design, absolutely nothing would have prevented Australia seeking further performance details or reference points, or even assistance in interpretation in the case that the information provided in the CEP was found to be insufficient.
+
+\item \textbf{The Australians can't produce the answer, and are instead denying one exists.} Instead of countering my report by claiming that they have examined test results provided by the French, (or from their own research and modelling) which demonstrates that pumpjet efficiency does not decline at low speeds as I have concluded, Defence is now disgracing themselves with a series of entirely implausible arguments which amount to arguing that no adequate reference point really exists, and therefore the answer cannot be known.
+In simple terms, those arguments are:
+
+\begin{itemize}
+
+\item \textbf{The hull and the pumpjet need to be `tuned' to each other.} This is the longest-running argument, which has been made repeatedly by RADM Sammut in sequential years \parencite{sammut2017}, \parencite{bergmann2018}, in answers to Questions on Notice \ref{QON1}, and by Chief of Maritime at DSTG \parencite{kerr2018} as well as the final response to my paper.\parencite{defence2018}.
+It constitutes an attempt to deny that other nuclear submarines, including the nuclear Barracuda, serve as an adequate reference point to assess the probable performance of the pumpjet on the Shortfin Barracuda.
+It requires an enormous contortion of logic order to argue that the performance of the pumpjet we will have can't be known on account of the hull.
+It would be equally fair to argue that one cannot know how a pair of shoes would perform in a given circumstance without simultaneously tuning them to a pair of socks.
+Whilst it is true that the two have to work in close conjunction with each other and a particularly poor match might impair performance, it is necessary (and sufficient) during concept design to assume that a competent (or even ideal) match is made in order to assess the overall suitability of the system.
+The implausibility of this argument is further cemented by just how much emphasis has been given to the fact that the Shortfin Barracuda will have \textit{exactly the same} diameter hull as the nuclear cousin from which it is derived \ref{SEN2017hanson}, \parencite{kerr2017}, which ought to make adjustments to the pumpjet on account of the hull entirely unwarranted.
+
+\item \textbf{The weight of the pumpjet makes it inappropriate for smaller diesel-electric submarines, but not for larger ones.} This argument has only occurred recently, first made by former CEO of Naval Group Australia Brent Clark \parencite{clark2018}, (to which I responded\parencite{morrison2018b}), and then later explained in more detail by Chief of Maritime at DSTG Dr David Kershaw\parencite{kerr2018}.
+It is an attempt to deny that any other non-nuclear submarine, including every operating class in existence (none of which pumpjets) constitute an adequate reference point for assessing the likely performance of the pumpjet on our future non-nuclear submarine.
+It is demonstrably ridiculous, since if a smaller submarine could not maintain proper balance and buoyancy with a heavy pumpjet attached to its tail, then the the submarine would be doomed to either tail-stand or sink, which would amount to a catastrophic failure if ever such a submarine were to detach from its moorings.
+That this is not the case is amply demonstrated by the fact that both the French and Russians have succeeded in fitting pumpjets to small submarines, and such catastrophes were avoided.
+The Russian test boat, B-871 `Alrosa' still operates regularly in the Black Sea.\footnote{The Black Sea is an inland sea about half the size of NSW, which shares borders only with Romania, Bulgaria, Georgia, Ukraine and Turkey.
+Russia operates a fleet of seven Kilo Class Submarines in the Black Sea, mostly out of Sevastopol, the port which it aquired control of through the 2014 annexation of Crimea.
+The Black Sea might thus be described appropriately as a `Russian Lake', where the endurance and range penalties imposed on Alrosa by the pumpjet might be tolerated.
+None of the Kilo Class Submarines produced since Alrosa have a pumpjet, and none of the 15 Russian Kilo Class submarines operating in the Northern, Pacific, or Baltic Fleets have pumpjets.}
+
+\item \textbf{The resistance of submerged bodies is constant with speed, unlike that of ships who encounter wave-making resistance.}  This is the most recent, and the most utterly and demonstrably false claim, made both by Dr David Kershaw of DSTG in an interview with Australian Defence Magazine \parencite{kerr2018}, and in Defence's considered written submission to the Senate Economic References Committee \parencite{defence2018}.
+It constitutes a clearly desperate and severely embarrassing attempt to argue the no propulsion system of any sort as tested or demonstrated on a surface vessel serves as any kind of reference point for its performance on a submarine.
+It is true that ships have the \textit{additional} resistance from wave-making resistance, but for most displacement vessels operating at design speed (and certainly all submarines) pressure drag and skin friction on the submerged part of the hull tend to dominate, both of which scale with velocity squared.
+Consequently, the `advance ratio' (how far forward through the water a propeller moves for a given degree of rotation) changes enormously with speed, for a submarine just as it does for a surface ship, and efficiency consequently changes along with this.
+
+\end{itemize}
+
+\end{enumerate}
+
+As noted by Andrew Davies when he contemplated the possibility that `Naval Group is pulling the wool over the eyes of a clueless Commonwealth' with the pump jet, extraordinary claims require extraordinary evidence \parencite{davies2017}.
+There can be no doubt that such extraordinary evidence of cluelessness has now been supplied by the Department of Defence themselves.
+But this evidence, revealing impression of cluelessness so complete, is difficult to reconcile with the adamant insistence of competence and confidence in the choice, given by multiple senior officials in the face of sustained questions backed up by substantial evidence.
+It is so extraordinary that it demands the question of whether the appointed officials who represent the Commonwealth are indeed `clueless' or whether officials in the Navy and Department of Defence are actually conspiring with Naval Group and the French Government to pull the wool over the eyes of a clueless Parliament, as well as the properly elected members of the executive who ultimately signed off on the purchase, and the Australian public at large.
+
+The gravity of such a conclusion is difficult to understate, given the likely nature of the motives which would attend to such a decision.
+One would presume that one would not advance into the public arena a submarine design which was known to be unviable, and lead to deep embarassment and ridicule if subjected to thorough scrutiny, unless the real motive for the design or partner choice was even less capable of withstanding public scrutiny.
+The possibility that this is the case must be considered very seriously, given the difficulty of reconciling recent events with the counterfactual proposition:
+
+After the initial publication of my first report, three Questions on Notice were asked by Senator Ketter, specifically referencing my report \ref{QON1}.
+Whilst Defence avoiding answering all of them directly, one of them was particularly brazenly side-stepped.
+Asked ``Is Defence still confident that Naval Group still hold the opinion that the pump-jet is well suited to diesel-electric submarines?" Defence merely reiterated their intent to use the pumpjets as offered, strictly avoiding mentioning anyone's opinion, either their own or Naval Group's.
+If the answer actually was what we would all assume it should be, that should be a very easy answer to directly.
+A simple `yes' would do just fine.
+If the answer is `no', then we must assume that at the time of writing (presumeably before the return date of 3 April, though Defence isn't always timely) that Defence an inkling (at the very least) of what Naval Group undoubtedly knew all along - that pumpjets are entirely unsuitable for diesel-electric submarines.
+In avoiding answering the question directly, Defence was not just misleading the parliament, but clearly acting to shelter Naval Group from subsequent accusations of fraudulent or misleading conduct during the CEP.
+
+Giving Defence the most generous possible interpretation, one might allow the possibility that Defence were just coming to grips with being completely hood-winked during the CEP by their chosen design-partner, and perhaps they felt that their wrath at being deliberately deceived ought not be publicly vented, or at least not at this stage.
+One might, at a stretch, allow that maintaining the opportunity for the partner to save-face was still a useful bargaining-chip that our Defence chiefs to use in order to bring a potentially treacherous contractor to heel, along with a range of contractual penalties, as well as the commercial leverage the Commonwealth would have by being able to re-introduce competition into the design process as many have suggested ought be done in any case \parencite{stanford2017}, or not proceed to a build contract on completion of the design.
+
+The state of commercial negotiatons between the Commonwealth and Naval Group, as revealed in the Senate Economic References Committee \parencite{SEN2018morrison}, reveal that Defence was moving on exactly the opposite tack - shocking Senators with the revelation that Defence was actually moving instead to contract immediately with DCNS for the full construction of all 12 submarines, to the complete exclusion of the Australian Submarine Corporation, within coming weeks, effectively wiping out the opportunity for any competitive tension at any stage of the ~\$50 billion project.
+
+In light of these events, I find the most plausible explanation to be the worst:
+\begin{itemize}
+\item The propulsive performance of pumpjets do indeed decline to very low efficiencies at low speeds, making them entirely unsuitable for a diesel-electric submarine which operates almost entirely at such speeds, broadly in line with the findings of my first report.
+\item Some group of individuals, including senior French officials from Naval Group as well as the Australian Government, were aware of this fact at the time of the Competitive Evaluation Process, and remains so today.
+\item That group has deliberatly cultivated the impression that pump-jets would confer a substantial technological edge to Australia's Future Submarine, knowing that would not be the case, either by virtue of the pump-jets not ultimately being used, or because if they were they would confer a disadvantage.
+\item That impression was necessary in order to defend of a choice which selected a design partner who's proposal was both more expensive, and slower to deliver that their competitors, for which no more defensible reason could be found.
+\end{itemize}
+
+It appears that the pumpjet idea was simply a lead-character used to construct the fiction that France's vaunted heritage as a designer of nuclear submarines offers a substantial technical advantage to a non-nuclear submarine.
+There is almost no chance that it will be included in the detailed design.
+In reality, without a nuclear reactor, there is barely any relevant technology or experience (other than the combat system, mandated from a 3rd party) which is usefully transferrable.
+The extent to which the bottom of the barrel needed to be scraped was illustrated when RADM Sammut was challenged on the benefit of the nuclear design, other than pump-jets \ref{SENhanson}, and he was forced to refer to the `diameter of the hull' as being a key learning, a prospect which would be much more easily subjected to public ridicule had it been more substantially promoted.\footnote{Amongst the many engineering challenges presented in designing a new submarine, ensuring the structural integrity of the hull at a new diameter is one of the most well-understood and predictable.
+The Collins Class was evolved from a submarine of smaller hull diameter, and this aspect of the design and development encountered no problems.
+In 2006, DSTO commissioned a paper suggesting that the Collins Class replacement be designed at a wider hull, which didn't correspond to that of the Barracuda.}  In contrast, those countries whos own navies relied entirely upon diesel-electric submarines would likely boast significant advantages in technologies which are genuine game-changers for diesel-electric submarines, such as advanced composite propellers, Air-Independent Propulsion, and Lithium-Ion batteries, which we are now told won't be included in the Future Submarine \textit{because} of the pumpjet\parencite{bergmann2018}.
+
+Australians have a right to know what the true motive of Defence was in choosing the French as a design partner.
+There can be no doubt that the pumpjet played no legitimate part in it.
+Any claims to the contrary are made either out of obscene incompetence, as Defence's latest response might have us believe, or more likely are made with the intent of being misleading.
+
 \section{Introduction} \label{intro}
@@
 This statement exhibits extreme care in the wording in order to be plausibly defensible, and is aguably misleading.
+
+\subsection{RADM Gregory Sammut, Head of SEA 1000 before the Senate}
+
+\begin{tcolorbox}[width=0.9\textwidth, center, colback=red!5!white,colframe=red!75!black, title= \textbf{6 May 2016 - Senate Estimates}]
+\textit{Senator Conroy asks Rear Admiral Greg Sammut (then director of SEA 1000) about the propulsion system of the proposed submarine.  Full exchange in \ref{SENsammut2}. Emphasis added.}
+
+\textbf{Senator CONROY:}  The French do not have this propulsion system on a conventional submarine anywhere, do they?
+
+\textbf{Rear Adm. Sammut:}  They do not have a nuclear reactor on a conventional submarine.
+
+\textbf{Senator CONROY:}  No, the propulsion system they are proposing for the Shortfin Barracuda is not powered by diesel; it is only nuclear powered.
+
+\textbf{Rear Adm. Sammut:}  DCNS has proposed to us a conventionally-powered submarine that includes—I think you are trying to get to the form of propeller we are talking about here, Senator.
+
+\textbf{Senator CONROY:}  Yes.
+
+\textbf{Rear Adm.
+Sammut:}  It is a pump jet, which is what DCNS has said is part of their proposal.
+That is just one element of the entire propulsion system.
+We were discussing before the fact that there are many elements to the propulsion system.
+There is the propeller, there is the motor that drives the propeller and the diesels that charge the battery that drives the motor.
+All of those factors combined lead to the propulsion system that is fitted to a submarine .
+When we consider all of those elements we look at what each contributes to the likely noise signature of the submarine and make an assessment about the overall performance of the submarine as far as its stealth is concerned.
+
+\textbf{Senator CONROY:}  I have read that the signature of the nuclear-powered sub with the propulsion system propeller that we are taking is very good at high speed.
+But at the speeds that a conventional sub will go-which by definition are considerably less than the speeds a nuclear-powered one achieves-there is a difference in the signature.
+Could you comment on that?
+
+\textbf{Rear Adm. Sammut:}  \textbf{Only to say that there is an impact across all speed ranges of the propulsion system. It does not purely pertain to high speed.}
+
+\textbf{Senator CONROY:}  But what I have read is that the propulsion system is, if you like, designed and attuned to perform against the nuclear-powered aspect, and that they swap the powering out and keep the same propulsion system, but they have not adapted it in any specific way.
+I am inviting you to comment.
+The two parts work beautifully together, but the conventional power that we plan to use does not create as successful a signature at the lower speeds—so above 20 knots versus below 20 knots.
+Would you like to comment?
+
+\textbf{Rear Adm.
+Sammut:}  Only to say what I said before.
+There is an advantage across all the speed ranges in the technologies that are being proposed by DCNS.
+\textbf{You are right: there has to be a degree of adaption of any propeller that is designed for a submarine , whether it is a normal propeller or whether it is a pump jet.} That always has to take place.
+The principles that underpin the quietness of a pump jet apply in any case.
+
+\end{tcolorbox}
+
+Only weeks after the announcement of the success of the DCNS bid to be the design partner for SEA 1000, specific questions about the speed-suitability of the propulsion system are being raised by senators.
+In this particular line of questioning, Senator Conroy implies that the acoustic advantages of the pumpjet may not apply as well at the speeds that a conventional submarine could operate at, which is consistent with findings in my earlier report that the acoustic advantage would essentially be constrained to circumstances where high speeds would cause cavitation on an open propeller.
+In his responses, RADM Sammut claims that the advantages apply at all speeds.
+He also concedes that some modification of the pumpjet will be required for the purpose, in response to a question about the suitability of the speed.
+He is not, however, specific about whether it is the speed requirement, or the shape of the hull, which drives this need for adjustment, though the context would seem to imply the former.
+He also here clearly discusses a pumpjet as being a special type of propeller, rather than a completely unrelated system which bears no comparison.
 
@@
 
-The open-water efficiency, $\eta_O$ was dwelt on substantially in my earlier report because this efficiency has the greatest degree of possible variance, (from zero, to around \%75 in some cases) and this variance occurs directly and necessarily with speed.
+The open-water efficiency, $\eta_O$ was dwelt on substantially in my earlier report because this efficiency has the greatest degree of possible variance, (from zero, to around 75\% in some cases) and this variance occurs directly and necessarily with speed.
 The remaining two coefficients describe the interaction between hull and the propulsor.
@@
 The argument that the efficiency of pumpjets cannot be  assessed using their open-water efficiency is also blatantly false.
-The 'open-water' condition is can be usefully thought of as the normal or 'ideal' efficiency of the propulsor, when the stream of incoming water is undisturbed, and consequently is still a very functional maximum efficiency of a system who's purpose is to smooth out the inconsistency in the flow.
+The `open-water' condition is can be usefully thought of as the normal or `ideal' efficiency of the propulsor, when the stream of incoming water is undisturbed, and consequently is still a very functional maximum efficiency of a system who's purpose is to smooth out the inconsistency in the flow.
 In the absence of an advantageous pre-swirl, no higher efficiency could be reasonably expected to be achieved through such smoothing than would be acheived by beginning with a smooth stream in the first place.
-Consequently, one can have considerable confidence that the extent to which this 'smoothing' effect might assist a pumpjet, relative to a propeller, is bounded by the extent to which it actually impedes a propellers performance.
+Consequently, one can have considerable confidence that the extent to which this `smoothing' effect might assist a pumpjet, relative to a propeller, is bounded by the extent to which it actually impedes a propellers performance.
 As is amply demonstrated by abundant sources, this is a very small extent in all cases, and may actually be to the slight advantage of a propeller in some cases where a desirable pre-swirl already exists.
@@
 
-Hull efficiency, like relative rotative efficiency, is not a true efficiency in that it is a ratio of two effects which can work in different directions, and consequently can be greater than one.
-The two effects are the impact of the wake fraction on the propulsor, as well as the factor known as the 'thrust deduction' caused by the impact of the propulsor on the hull.
+Hull efficiency, like relative rotative efficiency, is not a true efficiency in that it is a ratio of two effects, either of which can be greater than the other, and consequently can be greater than one.
+The two effects are the impact of the wake fraction on the propulsor, as well as the factor known as the `thrust deduction' caused by the impact of the propulsor on the hull.
 
@@
 A third plot, reproduced here in Figure \ref{fig:hulleff.png} demonstrates how these two values combine to produce the hull efficiency.
-As can be clearly seen, the size of the propeler relative to the hull also has an impact, and alters the hull efficiency by approximately 15\% when the propulsor size is halved or doubled.\footnote{The propulor size also impacts open water efficiency, so an global optimum may be constrained away from the hull-efficiency optimum.}  Given that one of the key features of the pumpjet is that they are smaller in total diameter to a comparable propeller, (and hence accelerate a smaller water-column to greater speeds) this provides a useful indication of the likley degree of variance that might be expected.
+As can be clearly seen, the size of the propeller relative to the hull also has an impact, and alters the hull efficiency by approximately 15\% when the propulsor size is halved or doubled.\footnote{The propulor size also impacts open water efficiency, so an global optimum may be constrained away from the hull-efficiency optimum.}  Given that one of the key features of the pumpjet is that they are smaller in total diameter to a comparable propeller, (and hence accelerate a smaller water-column to greater speeds) this provides a useful indication of the likley degree of variance that might be expected.
 However, the size differences are unlikely to be nearly as large as a factor of two.
@@
 This is not the case.
-In reality, accelerating ducts are suited to propellers that are highly loaded (low $J$), which is the same end of the spectrum that a any propulsion system moves if it attempt to push a given vessel to higher speeds.
+In reality, accelerating ducts are suited to propellers that are highly loaded (low $J$), which is the same end of the spectrum that a any propulsion system moves in if it attempts to push a given vessel to higher speeds.
 In reality, tugs and trawlers operate at unusually high $J$ values because their propeller size is constrained.
@@
 This distinction is important in understanding the limitations of a pumpjet's claim to make the incoming stream of water less inconsistent, or more regular.
-The duct intake and any stator blades in advance of the rotor can be effective in making the flow more homogenous as assessed on a macroscopic scale, but in doing so they simultaneously make the flow more turbulent at a smaller scale, since the necessarily involve passing the water over more surfaces, rather than less.
+The duct intake and any stator blades in advance of the rotor can be effective in making the flow more homogenous as assessed on a macroscopic scale, but in doing so they simultaneously make the flow more turbulent at a smaller scale, since it necessarily involve passing the water over more surfaces, rather than less.
 This can be advantageous to the extent that very large-scale and intense vortices might be broken down into smaller-scale vortices, which appears more homogenous when assessed on the length-scale relevant to the blades of the rotors.
@@
 \footnote{It is this effect that is likely referred to in comments made by Professor Joubert in a paper commissioned by DSTO in 2004 \parencite{joubert2004}.
-However, with competently designed appendages and fairings, the kind of unsteady vortex shedding which would see such large-scale eddies shed into the propulsor should be safely avoided.
+However, with competently designed appendages and fairings, the kind of unsteady vortex shedding from appendages which would see such large-scale eddies shed into the propulsor should be safely avoided.
 Substantial effort has been devoted to developing the smoothest possible wake-field for submarines for this purpose.\parencite{toxopeus2014}}
@@
 \section{Discussion}
+
+Having considered in some detail the nature of the three substantive arguments given in opposition to my findings, it is clear that none actually provide any credible evidence to oppose my conclusions.
+However, a clear similarit or parallel between the three can be found.
+Each of them respectively actually comprise an argument against an obvious reference point being used to assess the efficiency of the pumpjet.
+
+The argument that resistance is constant for underwater bodies, since they encounter no wave-making resistance, seeks to oppose any use of propulsive data derived from propulsors which might be used on ships, including for highly comparable systems (waterjets), or even directly comparable systems (pumpjets, or decelerating ducted propellers).
+
+The argument that the size of the submarine is a game-changer, attempts to dismiss the possibility that any experience with other diesel-electric submarines constitutes and adequate reference-point.
+
+The argument that the hull and the pumpjet need to be tuned together is clearly an argument that is intended to dismiss the possibility than any other nuclear submarine, including the Barracuda from which the Shortfin Barracuda is derived, serves as an adequate reference point.
+
+It is indeed remarkeable that none of the arguments made actually involve pointing to a better reference point than what I have been able to use from open-source literature, or announce that any results from Naval Group directly contradict the findings of my earlier report.
+It is a relatively simple proposition that Naval Group, as one of only two countries in the world to have fitted both pumpjets and propellers to both diesel and nuclear submarines, has in its possession the best data possible to serve as reference-points for our submarine.
+Detailed modelling, and tank testing from the pumpjet on the Barracuda would certainly provide the best reference, but experience from the in-service Rubis Class (which use propellers) as well as the in-service SSBNs would doubtless also provide adequate data.
+
+As I have argued, I think that there actually are adquate reference points in open-source literature regarding comparable propulsion systems, designed for comparable purposes, employed in relevant and comparable, in order to gain what is really only a very broad, but clear conlcusion: that propulsive efficiency declines inevitably and dramatically at low speeds.
+
+Taking it as given that the French do have posession of adquate of far more detailed performance data, derived from even closer reference points than what I can glean from public literature, one must ask whether that data actually confirms, or contradicts my findings.
+Given public claims from DCNS during the CEP that information regarding both the hydrodynmic and acoustic performance of the pumpjet were in Australian government hands in Australia \parencite{costello2016} there can be little doubt Australian officials have access to such information.
+Indeed, it would be an inexcuseable oversight if they didn't obtain that information during the CEP.
+
+This poses a seemingly irreconcileable tension between the desparate, implausible arguments currently being made by Defence, and the possibility that they are being made because Defence genuinely doesn't know better, and genuinely believes them to be true.
+It seems far more probable that Defence actually is in posession of clear, conclusive evidence which broadly supports the conclusions of my report, and is now deliberately concealing that evidence, and attempting to excuse why they have not drawn the obvious conclusions that any sensible person normally would from that evidence - that a pumpjet is entirely unsuited to a diesel-electric submarine.
 
@@
 
+\subsection{Hansard Extract, Senate Estimates - 6th May 2016 \cite{SEN2016sammut}}\label{SENsammut2}
+
+\textbf{Senator CONROY:}  The joint media release, Minister, states, 'Subject to discussions on commercial matters, the design of the Future Submarine» with DCNS will begin this year,' which I think is what Admiral Sammut is indicating.
+Is it fair to say that the process so far, up until the announcement, was about identifying a design partner—not a specific submarine design—and that the specifics, such as cost, schedule and performance, are still to be worked out?
+I am just going off the press release that you issued, Minister.
+
+\textbf{Senator Payne:}  As you said in your previous question to Admiral Sammut, we are at the beginning of the design process, not the end.
+So, in selecting DCNS, the evaluation process identified that proposition, which they submitted, as the one which best met the capability requirements that Australia has for a regionally superior submarine .
+We will now proceed to work with the DCNS, as Admiral Sammut has outlined, to develop that design.
+
+\textbf{Senator CONROY:}  There has been a lot of robust public debate about the decision.
+I want to offer Admiral Sammut the opportunity to outline to the committee, and to the broader public who pay attention to these things, the basis on which the decision was made.
+We have established that we do not have the finalised costs, we do not have a finalised production schedule and we do not have a mature design, so what were the factors that you based your decision around in choosing the French submarine ?
+
+\textbf{Rear Adm.
+Sammut:}  The consideration was around a range of factors.
+But, first and foremost, it was the ability to work with an international partner that would develop with us and help us to deliver a regionally superior submarine , and one over which we would have the sovereign capacity to operate and sustain throughout its life.
+With those principal requirements in mind, we considered a range of criteria, such as the capabilities offered—that is, the platform capabilities offered by the participants.
+We did look at their capacity to work with our combat system integrator to support the integration of the combat system into the submarine .
+These matters go to the heart of the capability that the future submarine will have in the future.
+
+We did look at rough order magnitude costs and schedules to ensure that the participants understood the scale of the program that we were embarking upon and could give us insights into the manner in which they would manage costs and schedule throughout this long program.
+
+Of course, we looked at matters concerning program management and we looked at a design and the way that safety is included in the design process, noting that, at the end of the day, even though Australia will be working with an international partner to design the submarine , Australia will have responsibility for certifying the submarine as safe and fit for operations, so we need to understand the safety philosophy.
+
+We looked at the methods of sustainment and the matter in which each of the participants would support us in establishing a sovereign sustainment capability from the outset of the program.
+Of course, this was a key lesson from the Collins program and many international submarine programs as well.
+Connected with that was looking at the potential commercial arrangements in the government-to-government agreements that would support such a complex program over its long life.
+That went to matters such as the way in which technical data would be transferred to Australia and the way in which intellectual property would be managed.
+
+We also looked at the participants' ability to manage risk, again, throughout this complex program.
+That was a very comprehensive process.
+It was conducted with a large team working within the Future submarine program.
+The evaluation of those proposals took place over a long period of time and in accordance with the schedule that we set so that we could complete a very thorough and comprehensive evaluation of the very high-quality proposals we received from each of the participants.
+That process, of course, was overviewed by an expert advisory panel, which reported that the process was indeed sound and was defensible from an accountability and probity perspective, and that all participants had been treated fairly and equitably in the process.
+
+We also invited peer review of our evaluation.
+External review of that, which has been announced, included review by experienced submarine program managers from the United States, who also concluded that our process was very sound and would properly inform the selection of the most suitable international partner.
+They made no recommendations to change the course of the process or the manner in which it was being conducted.
+We also had external review of our analysis conducted by RAND Corporation, and we also invited the US Navy to look at the manner in which we were assessing costs and the veracity of the ROM costs that were being provided to us.
+On the basis of that very thorough process, we believe we have come up with a firm recommendation to government, which government has made its decision on.
+That gives us confidence also now that we are best placed to enter contracts with DCNS for the design of the submarine and to work with us on the construction of that submarine as well.
+
+\textbf{Senator CONROY:}  Okay.
+I want you to pretend that I know absolutely nothing about what I am going to ask next, which should be easy because it is true.
+It is very easy for him to fake this, because it is true.
+There has been a lot of debate in the papers about the propulsion system—the French versus anybody else's.
+Could you tell us about the propulsion system, because there has been a lot of adverse discussion about it.
+I am interested in your perspective.
+Without revealing anything that you should not, are you able to explain to the committee?
+
+\textbf{Rear Adm.
+Sammut:}  We looked at the propulsion system, of course, offered by all participants.
+The propulsion system includes a range of systems in the submarine designed to propel it through the water and to do so quietly.
+Propulsion for any submarine , of course, has a large bearing on its stealth—in other words, how quietly the submarine moves through the water.
+We considered all of the proposals presented by the participants using a range of subject matter experts.
+It included assessment by the Defence Science and Technology Group, and we looked not only at what was proposed by participants, in terms of what sort of propeller they would put on the submarine and what sort of drive train they would have, but also at what their response was in terms of the performance of those systems.
+In the final analysis, we decided that the DCNS proposal best met our requirements for stealth based on our understanding of what would be required of a regionally superior submarine operating in the 2030s and beyond.
+
+\textbf{Senator CONROY:}  Thank you.
+I have been reading—as you would hope and expect—as much of the commentary as I can keep up with on this.
+The current French propulsion system is propelled by a nuclear engine.
+Is that the current Barracuda?
+
+\textbf{Rear Adm. Sammut:}  The submarine they are designing and building for themselves is a nuclear-powered submarine .
+
+\textbf{Senator CONROY:}  So it is a nuclear reactor that drives the propulsion system?
+
+\textbf{Rear Adm. Sammut:}  That is correct.
+
+\textbf{Senator CONROY:}  I think the Prime Minister said this, but can I just confirm from the expert panel:  there was no consideration, and this was not a factor in any way whatsoever, about this mooted change to a nuclear submarine halfway through the process.
+I am inviting you to confirm that this bore no part whatsoever in the thinking of your group, you or any of the other partners.
+
+\textbf{Rear Adm. Sammut:}  I can confirm that categorically.
+
+\textbf{Senator CONROY:}  So if at some stage over the life build of the 12 we ended up with a nuclear sub as part of this process, it would be a mystery to you?
+
+\textbf{Rear Adm.
+Sammut:}  The competitive evaluation process was about the acquisition of a future submarine that was conventionally powered.
+That was purely the consideration upon which the proposals were assessed.
+
+\textbf{Senator CONROY:}  I understand the propulsion system of the French proposal is rumoured—again, if there are areas you cannot disclose, please say so—to work brilliantly quietly with the nuclear reactor at high speeds.
+It is very quiet.
+Is that correct?
+
+\textbf{Rear Adm.
+Sammut:}  The performance of all submarines operating at high speeds is largely about the manner in which the propeller works, somewhat independently of what is supplying power to the propeller.
+
+\textbf{Senator CONROY:}  But I understand one of the proud claims from DCNS of its nuclear-powered Barracuda is that at high speed with its nuclear-powered propulsion system it is very quiet in a submarine underwater signature sense.
+
+\textbf{Rear Adm.
+Sammut:}  The French would have goals, of course, for the noise performance of its submarines.
+I should add that it is not related just to the propulsion system but to the way the hull is designed, the flow of the water over the hull at those high speeds and how that is managed to reduce flow noise that emanates from the submarine at high speeds.
+
+\textbf{Senator CONROY:}  The French do not have this propulsion system on a conventional submarine anywhere, do they?
+
+\textbf{Rear Adm. Sammut:}  They do not have a nuclear reactor on a conventional submarine.
+
+\textbf{Senator CONROY:}  No, the propulsion system they are proposing for the Shortfin Barracuda is not powered by diesel; it is only nuclear powered.
+
+\textbf{Rear Adm. Sammut:}  DCNS has proposed to us a conventionally-powered submarine that includes—I think you are trying to get to the form of propeller we are talking about here, Senator.
+
+\textbf{Senator CONROY:}  Yes.
+
+\textbf{Rear Adm.
+Sammut:}  It is a pump jet, which is what DCNS has said is part of their proposal.
+That is just one element of the entire propulsion system.
+We were discussing before the fact that there are many elements to the propulsion system.
+There is the propeller, there is the motor that drives the propeller and the diesels that charge the battery that drives the motor.
+All of those factors combined lead to the propulsion system that is fitted to a submarine .
+When we consider all of those elements we look at what each contributes to the likely noise signature of the submarine and make an assessment about the overall performance of the submarine as far as its stealth is concerned.
+
+\textbf{Senator CONROY:}  I have read that the signature of the nuclear-powered sub with the propulsion system propeller that we are taking is very good at high speed.
+But at the speeds that a conventional sub will go—which by definition are considerably less than the speeds a nuclear-powered one achieves—there is a difference in the signature.
+Could you comment on that?
+
+\textbf{Rear Adm. Sammut:}  Only to say that there is an impact across all speed ranges of the propulsion system. It does not purely pertain to high speed.
+
+\textbf{Senator CONROY:}  But what I have read is that the propulsion system is, if you like, designed and attuned to perform against the nuclear-powered aspect, and that they swap the powering out and keep the same propulsion system, but they have not adapted it in any specific way.
+I am inviting you to comment.
+The two parts work beautifully together, but the conventional power that we plan to use does not create as successful a signature at the lower speeds—so above 20 knots versus below 20 knots.
+Would you like to comment?
+
+\textbf{Rear Adm.
+Sammut:}  Only to say what I said before.
+There is an advantage across all the speed ranges in the technologies that are being proposed by DCNS.
+You are right: there has to be a degree of adaption of any propeller that is designed for a submarine , whether it is a normal propeller or whether it is a pump jet.
+That always has to take place.
+The principles that underpin the quietness of a pump jet apply in any case.
+
+\newpage
+
 \subsection{Answer to Question on Notice}\label{QON148}

### 2018-09-26 Second.Rnw
exec summary proofed

@@
 
-{\huge \textsf{A response to the Government response}}
+{\huge \textsf{A reply to the Government's response}}
 \vskip\baselineskip
-{\huge \textsf{(Or lack thereof)}}
-\vskip\baselineskip
-{\huge \textsf{on the question of the pumpjet}}
+{\huge \textsf{to the question of the pumpjet}}
 \vskip\baselineskip
@@
 \vskip\baselineskip
-\textsf{August 2018}
+\textsf{September 2018}
 \vskip\baselineskip
@@
 \textbf{\textsf{Trendlock}} \textcolor{namecolor}{\textsf{Consulting}}
+\vskip\baselineskip
+\textsf{Prepared for Submarines for Australia}
 
@@
 In January 2018, I produced a detailed report\parencite{morrison2018}, commissioned by Gary Johnston from Submarines for Australia, in order to examine the question of the efficiency of a pumpjet.
-Pumpjets, whilst common on nuclear submarines which can operate at high speeds (above 25kt) for sustained periods, are generally not used on diesel-electric submarine, which need to operate at very low speeds (around 10kt during transit, and under 5kt during combat operations).
-Considerable emphasis had been placed on the significancw of the pumpjet proposed in successful French bid made by DCNS (now Naval Group) for Australia's Future Submarine.
-I concluded that an inevitable characateristic of pumpjets - directly necessitated by the characteristics which gave them good acoustic performance at high speeds - was very low propulsive performance (efficiency) at low speeds.
-This would mean that a diesel-electric submarine would suffer poorer range and endurance, and significantly worse dived range and endurance, as well as a worse indiscretion ratio due to the use of pumpjets.
-Given the profound importance of these characteristics to all diesel-electric submarines, aswell as elevated importance of range and endurance to the Australian requirement, it would appear that a pumpjet can be safely regarded as entirely unsuitable for Australia's Future Submarine.
-
-On June 7 2018, I appeared as a witness before the Senate Economic References Committee \parencite{SEN2018morrison}, where I defended and expanded upon my results, including in the presence of senior officials from Defence and Naval Group, who steadfastly refused to comment.
-The Committee directed Defence to provide a written response to my paper prior to the conclusion of the Inquiry Into the Future Of Naval Shipbuilding Industy in Australia on 27th of June, which they finally produced in mid September.
+Pumpjets, whilst common on nuclear submarines which can operate at high speeds (above 25kt) for sustained periods, are generally not used on diesel-electric submarines, which need to operate at very low speeds (around 10kt during transit, and under 5kt during combat operations).
+Considerable emphasis had been placed on the significance of the pumpjet proposed in the successful French bid made by DCNS (now Naval Group) for Australia's Future Submarine.
+I concluded that an inevitable characteristic of pumpjets - directly necessitated by the characteristics which gave them good acoustic performance at high speeds - was very low propulsive performance (efficiency) at low speeds.
+This would mean that a diesel-electric submarine would suffer poorer range and endurance, and significantly worse dived range and endurance, as well as a worse indiscretion ratio due to the use of a pumpjet.
+Given the profound importance of these characteristics to all diesel-electric submarines, as well as the elevated importance of range and endurance to the Australian requirement, it would appear that a pumpjet can be safely regarded as entirely unsuitable for Australia's Future Submarine.
+
+On June 7 2018, I appeared as a witness before the Senate Economic References Committee \parencite{SEN2018morrison}, where I defended and expanded upon my results.
+My evidence was presented in the presence of senior officials from Defence and Naval Group, who steadfastly refused to comment.
+The Committee directed Defence to provide a written response to my paper prior to the conclusion of the Inquiry into the Future of Australia's Naval Shipbuilding Industry (scheduled for the 27th of June) which they finally produced in mid September.
 Submarines for Australia had already commissioned me to write a subsequent report in reply to the responses that had appeared in responses to Questions on Notice in Parliament, and also in the media.
-With Defence's submission finally received, this paper is in a position to comprehensively assess the totalality of the arguments put by Defence against the conclusions of my previous paper.
-
-With the receipt of the written report it has become clear that Defence has no credible basis for rejecting the findings of my research, and instead has disgraced their technical credibility with utterly ridiculous misrepresentations of the relevant physics.
-Their claim that pumpjets can be efficient across the entire speed range are instead based on the assumption that the resistance faced by fully submerged bodies is ``essentially constant over speed"\parencite{defence2018} \parencite{kerr2018}.
+With Defence's submission finally received, this paper is in a position to comprehensively assess the totality of the arguments put by Defence against the conclusions of my previous paper.
+
+With the receipt of the written report it has become clear that Defence has no credible basis for rejecting the findings of my research.
+Their arguments are so completely incongruous with the most essential and frankly obvious physical principles, it appears that Defences has quite thoroughly disgraced their technical credibility in making this response.
+
+Their claim that pumpjets can be efficient across the entire speed range, including patrol speed, are instead based on the assumption that the efficiency of propulsors operating on fully-submerged bodies (like submarines) is ``essentially constant over speed"\parencite{defence2018} \parencite{kerr2018}.
+This is supported by the argument that efficiency relates to resistance, and that resistance faced by underwater bodies does not vary with speed, since it is only wave-making resistance (encountered by vessels on the surface) that is the speed-dependent component of resistance.
+
 This is simply untrue.
@@
 
-The truth is that the resistance of submerged bodies in the real world scales with velocity squared.
-This exponential increase in resistance explains why conventional submarines slow to extremely low speeds in order to conserve energy when dived, and has been used by other respected physicists discussing this topic \parencite{davies2017} as well as essentially all scientific literature involved hydrodynamics.
-It also explains why a propulsor, which must faces one hundred times as much resistance when travelling at 30kt as it does at 3kt, behaves very differently in these regimes, and consequently has a different efficiency.
-For Defence, assisted by DSTG, to claim that the relationshp between drag resistance and speed under-water is constant is an extremely shocking revelation.
-In the engineering/technical community, it would be just as strange to hear that a dentist with teenage children believes in the tooth-fairy.
-Not only is is their job to know otherwise, but abundant life experience would have continually proven otherwise.
+The truth is that the resistance of submerged bodies in almost all relevant real-world situations scales with velocity squared.
+(Wave-making resistance is in \textit{addition} to this.) This exponential increase in resistance explains why conventional submarines slow to extremely low speeds in order to conserve energy when dived, and has been used by other respected physicists discussing this topic \parencite{davies2017} as well as essentially all scientific literature involving hydrodynamics.
+It also explains why a propulsor, which must faces one hundred times as much resistance when travelling at 30kt as it does at 3kt on the same vessel, functions differently in these regimes, and consequently has a different efficiency.
+
+For Defence, assisted by Defence Science and Technology Group (DSTG), to believe that the relationship between drag resistance and speed under-water is anywhere near constant is an extremely shocking revelation.
+In the scientific community, it would be just as strange to hear that a dentist with teenage children believes in the tooth-fairy.
+Not only is is their job to know better, but abundant life experience would have continually proven otherwise.
 
@@
 Such a revelation necessitates a further question regarding whether such misleading claims where made out of complete ignorance and incompetence, or whether they were deliberately intended to deceive.
-As can be seen from the chronology of public comments which this report details, it is difficult to see how Defence could hold such a naive and embarrassing view after having their attention drawn to the opposing evidence multiple times, and devoted so much resources to understanding the relevant technologies.
-In 2012 Senator Johnston challenged RADM Moffit directly as to why research money was even being directed towards pumpjets, given how obviously inappropriate they were for our use, but RADM Moffit insisted they would still do the work required.
-In 2017 Defence revealed that a 99 page report was commissioned on pumpjets as part of \$214 million in 'de-risking' research, and that over a million dollars had subsequently been spent on pumpjet research specifically.
-
-Notwithstanding the improbability that the incompetence of Defence is so completely pervasive that all of this research failed to yeild even a basic understanding of hydrodynamics, the probability that the necessary degree of ignorance has occurred through arrangement rather than accident can be most firmly established by three simple propositions, which are made obvious through a reading of this report:
+As can be seen from the chronology of public comments which this report details, it is difficult to see how Defence could hold such a naive and embarrassing view after having their attention drawn to the opposing evidence multiple times, and devoted considerable resources to understanding the relevant technologies.
+In 2012 Senator Johnston challenged RADM Moffit directly as to why research money was even being directed towards pumpjets, given how obviously inappropriate they were for our use, but RADM Moffit insisted they would still do the work required to better their understanding\parencite{SE2012moffit}.
+In 2017 Defence revealed that a 99 page report was commissioned on pumpjets as part of \$214 million in `de-risking' research, and that over a million dollars had subsequently been spent on pumpjet research specifically.
+
+Notwithstanding the improbability that the incompetence of Defence is so completely pervasive that all of this research failed to yield even a basic understanding of hydrodynamics, the probability that the necessary degree of ignorance has occurred through arrangement rather than accident can be most firmly established by three simple propositions:
 
 \begin{enumerate}
-\item  \textbf{The French know the answer.}  As one of only two countries in the world who have designed and produced both pumpjets and propellers, nuclear and conventional submarines, and fitted and operated each type of propulsor on both types of submarines, there can be no doubt whatsoever that Naval Group and the French Government do have well-formed opinion about the performance of pumpjets on diesel-electric submarines.
+\item  \textbf{The French know the answer.}  As one of only two countries in the world who have designed and produced both pumpjets and propellers, nuclear and conventional submarines, and fitted and operated each type of propulsor on both types of submarines, there can be no doubt whatsoever that Naval Group and the French Government do have a well-formed opinion about the performance of pumpjets on diesel-electric submarines.
 They would have extensive operational performance data from their in-service SSBNS which use the pumpjet.
 They would have extremely detailed modelling, and tank-test results from the pumpjet from the almost-complete Barracuda class, which has an identical hull diameter to the proposed Shortfin Barracuda.
-The would have extensive operational performance data from their own Rubis Class nuclear submarines, which still use propellers, as well as the results of earlier tests, confirmed to have occurred by Naval Group Officials, where a pumpjet was fitted to a diesel-electric Agosta Class submarine.
-That they have adequate performance data from all the necessary reference points to assess the likely performance of the pumpjet on our own is entirely beyond dispute.
+They would have extensive operational performance data from their own Rubis Class nuclear submarines, which still use propellers, as well as the results of earlier tests, confirmed to have occurred by Naval Group Officials\parencite{SEN2018morrison}, where a pumpjet was fitted to a diesel-electric Agosta Class submarine.
+That they have adequate performance data from all the necessary reference points to assess the likely performance of the proposed pumpjet on our own submarine is entirely beyond dispute.
 
@@
 Any doubt that this occurred, however, can be reliably dismissed on account of public statements made by DCNS during the CEP claiming proudly that ``The stealth and hydrodynamic performances of pump jet propulsion are of course classified and in Australia known only to DCNS and the Australian Government."\parencite{costello2016}.
-Given RADM Sammut's repeated statements \parencite{SEN2016sammut} \parencite{SEN2017hanson} to Senate Estimates that  ``the way in which technical data would be transferred to Australia and the way in which intellectual property would be managed" was one of the reasons that DCNS was selected as design partner, and DCNS's clear pride in  ``the offer from France to transfer to Australia sovereign control and use of pump-jet propulsion technology" there can also be no doubt that after the conclusion of the CEP and the commencement of design, absolutely nothing would have prevented Australia seeking further performance details or reference points, or even assistance in interpretation in the case that the information provided in the CEP was found to be insufficient.
+Given RADM Sammut's repeated statements \parencite{SEN2016sammut} \parencite{SEN2017hanson} to Senate Estimates that  ``the way in which technical data would be transferred to Australia and the way in which intellectual property would be managed" was one of the reasons that DCNS was selected as design partner, and DCNS's clear pride in  ``the offer from France to transfer to Australia sovereign control and use of pump-jet propulsion technology"\parencite{costello2016} there can also be no doubt that after the conclusion of the CEP and the commencement of design, absolutely nothing would have prevented Australia seeking further performance details or reference points, or even assistance in interpretation in the case that the information provided in the CEP was found to be in any way insufficient or unclear.
 
@@
 
-\item \textbf{The hull and the pumpjet need to be `tuned' to each other.} This is the longest-running argument, which has been made repeatedly by RADM Sammut in sequential years \parencite{sammut2017}, \parencite{bergmann2018}, in answers to Questions on Notice \ref{QON1}, and by Chief of Maritime at DSTG \parencite{kerr2018} as well as the final response to my paper.\parencite{defence2018}.
+\item \textbf{The hull and the pumpjet need to be `tuned' to each other.} This is the longest-running argument, which has been made repeatedly by RADM Sammut in sequential years\parencite{sammut2017}\parencite{bergmann2018}, in answers to Questions on Notice[\ref{QON1}], and by Chief of Maritime at DSTG \parencite{kerr2018} as well as the final response to my paper\parencite{defence2018}.
 It constitutes an attempt to deny that other nuclear submarines, including the nuclear Barracuda, serve as an adequate reference point to assess the probable performance of the pumpjet on the Shortfin Barracuda.
-It requires an enormous contortion of logic order to argue that the performance of the pumpjet we will have can't be known on account of the hull.
+It requires an enormous contortion of logic in order to argue that the approximate performance of the pumpjet we will have can't be known on account of the hull.
 It would be equally fair to argue that one cannot know how a pair of shoes would perform in a given circumstance without simultaneously tuning them to a pair of socks.
-Whilst it is true that the two have to work in close conjunction with each other and a particularly poor match might impair performance, it is necessary (and sufficient) during concept design to assume that a competent (or even ideal) match is made in order to assess the overall suitability of the system.
-The implausibility of this argument is further cemented by just how much emphasis has been given to the fact that the Shortfin Barracuda will have \textit{exactly the same} diameter hull as the nuclear cousin from which it is derived \ref{SEN2017hanson}, \parencite{kerr2017}, which ought to make adjustments to the pumpjet on account of the hull entirely unwarranted.
-
-\item \textbf{The weight of the pumpjet makes it inappropriate for smaller diesel-electric submarines, but not for larger ones.} This argument has only occurred recently, first made by former CEO of Naval Group Australia Brent Clark \parencite{clark2018}, (to which I responded\parencite{morrison2018b}), and then later explained in more detail by Chief of Maritime at DSTG Dr David Kershaw\parencite{kerr2018}.
-It is an attempt to deny that any other non-nuclear submarine, including every operating class in existence (none of which pumpjets) constitute an adequate reference point for assessing the likely performance of the pumpjet on our future non-nuclear submarine.
+Whilst it is true that the two have to work in close conjunction with each other and a particularly poor match might impair performance, it is necessary (and sufficient) during concept design to assume that a competent (normally ideal) match is made in order to assess the overall suitability of the system.
+The implausibility of this argument is further cemented by just how much emphasis has been given to the fact that the Shortfin Barracuda will have \textit{exactly the same} diameter hull as the nuclear cousin from which it is derived\parencite{SEN2017hanson}\parencite{kerr2017}, which ought to make adjustments to the pumpjet on account of the hull entirely unwarranted.
+
+\item \textbf{The weight of the pumpjet makes it inappropriate for smaller diesel-electric submarines, but not for larger ones.} This argument has only occurred recently, first made by former CEO of Naval Group Australia Brent Clark\parencite{clark2018}, (to which I responded\parencite{morrison2018b}), and then later explained in more detail by Chief of Maritime at DSTG Dr David Kershaw\parencite{kerr2018}.
+It is an attempt to deny that any other non-nuclear submarines, including every operating class in existence (none of which have adopted pumpjets) constitute an adequate reference point for assessing the likely performance of the pumpjet on our future non-nuclear submarine.
 It is demonstrably ridiculous, since if a smaller submarine could not maintain proper balance and buoyancy with a heavy pumpjet attached to its tail, then the the submarine would be doomed to either tail-stand or sink, which would amount to a catastrophic failure if ever such a submarine were to detach from its moorings.
@@
 The Russian test boat, B-871 `Alrosa' still operates regularly in the Black Sea.\footnote{The Black Sea is an inland sea about half the size of NSW, which shares borders only with Romania, Bulgaria, Georgia, Ukraine and Turkey.
-Russia operates a fleet of seven Kilo Class Submarines in the Black Sea, mostly out of Sevastopol, the port which it aquired control of through the 2014 annexation of Crimea.
+Russia operates a fleet of seven Kilo Class Submarines in the Black Sea, mostly out of Sevastopol, the port which it acquired control of through the 2014 annexation of Crimea.
 The Black Sea might thus be described appropriately as a `Russian Lake', where the endurance and range penalties imposed on Alrosa by the pumpjet might be tolerated.
@@
 
-\item \textbf{The resistance of submerged bodies is constant with speed, unlike that of ships who encounter wave-making resistance.}  This is the most recent, and the most utterly and demonstrably false claim, made both by Dr David Kershaw of DSTG in an interview with Australian Defence Magazine \parencite{kerr2018}, and in Defence's considered written submission to the Senate Economic References Committee \parencite{defence2018}.
-It constitutes a clearly desperate and severely embarrassing attempt to argue the no propulsion system of any sort as tested or demonstrated on a surface vessel serves as any kind of reference point for its performance on a submarine.
-It is true that ships have the \textit{additional} resistance from wave-making resistance, but for most displacement vessels operating at design speed (and certainly all submarines) pressure drag and skin friction on the submerged part of the hull tend to dominate, both of which scale with velocity squared.
-Consequently, the `advance ratio' (how far forward through the water a propeller moves for a given degree of rotation) changes enormously with speed, for a submarine just as it does for a surface ship, and efficiency consequently changes along with this.
+\item \textbf{The resistance of submerged bodies is constant with speed, unlike that of ships who encounter wave-making resistance.}  This is the most recent, and the most utterly and demonstrably false argument, made both by Dr David Kershaw of DSTG in an interview with Australian Defence Magazine\parencite{kerr2018}, and in Defence's considered written submission to the Senate Economic References Committee\parencite{defence2018}.
+It constitutes a clearly desperate and severely embarrassing attempt to dismiss the volumous literature assessing the efficiency of propulsors as a function of `advance ratio' ($J$, a unitless metric designed specifically for the purpose of making comparisons possible across vessels encountering different levels of resistance from any source including speed) simply because such literature is often associated with ships.
+It is true that ships have the \textit{additional} resistance of wave-making resistance that submarines do not, but for most displacement vessels operating at design speed (and certainly all submarines) pressure drag and skin friction on the submerged part of the hull tend to dominate, both of which scale with velocity squared.
+Consequently, the `advance ratio' (how far forward through the water a propeller moves for a given degree of rotation) changes substantially with speed, for a submarine just as it does for a surface ship, and efficiency consequently changes along with this.
 
@@
 
-As noted by Andrew Davies when he contemplated the possibility that `Naval Group is pulling the wool over the eyes of a clueless Commonwealth' with the pump jet, extraordinary claims require extraordinary evidence \parencite{davies2017}.
-There can be no doubt that such extraordinary evidence of cluelessness has now been supplied by the Department of Defence themselves.
-But this evidence, revealing impression of cluelessness so complete, is difficult to reconcile with the adamant insistence of competence and confidence in the choice, given by multiple senior officials in the face of sustained questions backed up by substantial evidence.
-It is so extraordinary that it demands the question of whether the appointed officials who represent the Commonwealth are indeed `clueless' or whether officials in the Navy and Department of Defence are actually conspiring with Naval Group and the French Government to pull the wool over the eyes of a clueless Parliament, as well as the properly elected members of the executive who ultimately signed off on the purchase, and the Australian public at large.
-
-The gravity of such a conclusion is difficult to understate, given the likely nature of the motives which would attend to such a decision.
-One would presume that one would not advance into the public arena a submarine design which was known to be unviable, and lead to deep embarassment and ridicule if subjected to thorough scrutiny, unless the real motive for the design or partner choice was even less capable of withstanding public scrutiny.
-The possibility that this is the case must be considered very seriously, given the difficulty of reconciling recent events with the counterfactual proposition:
-
-After the initial publication of my first report, three Questions on Notice were asked by Senator Ketter, specifically referencing my report \ref{QON1}.
+As noted by Andrew Davies when he contemplated the possibility that ``Naval Group is pulling the wool over the eyes of a clueless Commonwealth" with the pump jet, extraordinary claims require extraordinary evidence\parencite{davies2017}.
+There can be no doubt that such extraordinary evidence of cluelessness has now been supplied by the Department of Defence.
+There can also be no doubt that Naval Group's proposal in the Competitive Evaluation Process (CEP) to which the pumpjet was ``integral"\parencite{guillou2017} was disingenuous at best, and quite possibly fraudulent.
+
+However, the impression of cluelessness is so severe, it is difficult to reconcile with the adamant insistence of competence and confidence in the pumpjet choice, given by multiple senior officials in the face of sustained questions backed up by substantial evidence on numerous occasions.\footnote{When Senator Carr was confirming in a public hearing with Kim Gillis (Head of Capability and Sustainment Group), that he was able to review my report and respond to my earlier evidence quickly before the close of the inquiry, Gillis replied ``The only caveat I'll put on any response we make is that the one thing we do really well is detailed engineering analysis."} It is so extraordinary that it demands the question of whether the appointed officials who represent the Commonwealth are indeed `clueless'.
+The alternative proposition is that officials in the Navy and Department of Defence are complicit or co-operating with Naval Group and the French Government to pull the wool over the eyes of a clueless Parliament, including the Cabinet (who signed off on the decision), and the Australian public at large.
+
+The gravity of reaching such a conclusion is difficult to understate, given the likely nature of the motives which would attend to such a decision.
+One would presume that one would not advance into the public arena a submarine design which was known to be nonviable, and lead to deep embarrassment and ridicule if subjected to thorough scrutiny, unless the real motive for the design or partner choice was even less capable of withstanding public scrutiny.
+The possibility that this is the case must be considered very seriously, given the difficulty of reconciling recent events with the counter-factual proposition:
+
+After the initial publication of my first report, three Questions on Notice were asked by Senator Ketter, specifically referencing my report[\ref{QON1}, \ref{QON2}, \ref{QON3}].
 Whilst Defence avoiding answering all of them directly, one of them was particularly brazenly side-stepped.
-Asked ``Is Defence still confident that Naval Group still hold the opinion that the pump-jet is well suited to diesel-electric submarines?" Defence merely reiterated their intent to use the pumpjets as offered, strictly avoiding mentioning anyone's opinion, either their own or Naval Group's.
-If the answer actually was what we would all assume it should be, that should be a very easy answer to directly.
-A simple `yes' would do just fine.
-If the answer is `no', then we must assume that at the time of writing (presumeably before the return date of 3 April, though Defence isn't always timely) that Defence an inkling (at the very least) of what Naval Group undoubtedly knew all along - that pumpjets are entirely unsuitable for diesel-electric submarines.
-In avoiding answering the question directly, Defence was not just misleading the parliament, but clearly acting to shelter Naval Group from subsequent accusations of fraudulent or misleading conduct during the CEP.
-
-Giving Defence the most generous possible interpretation, one might allow the possibility that Defence were just coming to grips with being completely hood-winked during the CEP by their chosen design-partner, and perhaps they felt that their wrath at being deliberately deceived ought not be publicly vented, or at least not at this stage.
-One might, at a stretch, allow that maintaining the opportunity for the partner to save-face was still a useful bargaining-chip that our Defence chiefs to use in order to bring a potentially treacherous contractor to heel, along with a range of contractual penalties, as well as the commercial leverage the Commonwealth would have by being able to re-introduce competition into the design process as many have suggested ought be done in any case \parencite{stanford2017}, or not proceed to a build contract on completion of the design.
-
-The state of commercial negotiatons between the Commonwealth and Naval Group, as revealed in the Senate Economic References Committee \parencite{SEN2018morrison}, reveal that Defence was moving on exactly the opposite tack - shocking Senators with the revelation that Defence was actually moving instead to contract immediately with DCNS for the full construction of all 12 submarines, to the complete exclusion of the Australian Submarine Corporation, within coming weeks, effectively wiping out the opportunity for any competitive tension at any stage of the ~\$50 billion project.
+Asked ``Is Defence still confident that Naval Group still hold the opinion that the pump-jet is well suited to diesel-electric submarines?" Defence merely reiterated their intent is to use the pumpjets as offered, strictly avoiding mentioning anyone's opinion, either their own or Naval Group's.
+If the answer actually was what we would all assume it should be, that would have been a very easy question to answer directly.
+A simple `yes' would be sufficient.
+If the answer was `no', then we must assume that at the time of writing (presumably before the return date of 3 April, though Defence isn't always timely) that Defence had an inkling (at the very least) of what Naval Group undoubtedly knew all along - that pumpjets are entirely unsuitable for diesel-electric submarines.
+In avoiding answering the question directly, Defence was not just misleading the Parliament, but clearly acting to shelter Naval Group from subsequent questions about fraudulent or misleading conduct during the CEP.
+
+Giving Defence the most generous possible interpretation, one might allow the possibility that Defence were just coming to grips with being completely hood-winked during the CEP by their chosen design-partner, and perhaps felt that their wrath at being deliberately deceived ought not be publicly vented, or at least not at this stage.
+One might, at a stretch, allow that maintaining the opportunity for the partner to save-face was still a useful bargaining-chip that our Defence chiefs to use in order to bring a potentially treacherous contractor to heel.
+(Presumeably in addition to a range of contractual penalties, as well as the commercial leverage the Commonwealth would have by being able to re-introduce competition into the design process as many have suggested ought be done in any case \parencite{stanford2017}, or simply not proceed to a build contract on completion of the design.)
+
+The state of commercial negotiations between the Commonwealth and Naval Group, as revealed in the Senate Economic References Committee \parencite{SEN2018morrison}, show that Defence was moving in exactly the opposite direction - shocking Senators with the revelation that Defence was looking instead to contract immediately with DCNS for the full construction of all 12 submarines, to the complete exclusion of the Australian Submarine Corporation, within coming months as part of the Strategic Partnering Agreement.
+This would effectively wipe out the opportunity for any competitive tension at any stage of the \$79 billion five-decade project within the first two years of a five-year design phase.
 
@@
 \begin{itemize}
-\item The propulsive performance of pumpjets do indeed decline to very low efficiencies at low speeds, making them entirely unsuitable for a diesel-electric submarine which operates almost entirely at such speeds, broadly in line with the findings of my first report.
+\item The propulsive performance of pumpjets do indeed decline to very low efficiencies at low speeds, making them completely unsuitable for diesel-electric submarines which are forced to operate almost entirely at such speeds.
+This is particularly the case for a submarine required to operate over large distances, such as that demanded by the Australian requirement.
 \item Some group of individuals, including senior French officials from Naval Group as well as the Australian Government, were aware of this fact at the time of the Competitive Evaluation Process, and remains so today.
-\item That group has deliberatly cultivated the impression that pump-jets would confer a substantial technological edge to Australia's Future Submarine, knowing that would not be the case, either by virtue of the pump-jets not ultimately being used, or because if they were they would confer a disadvantage.
-\item That impression was necessary in order to defend of a choice which selected a design partner who's proposal was both more expensive, and slower to deliver that their competitors, for which no more defensible reason could be found.
+\item That group has deliberately cultivated the belief that pumpjet technology, descending directly from French experience with nuclear submarines which no other contender could claim,  would confer a substantial technological edge to Australia's Future Submarine.
+\item This belief was known by that group to be false - either because pumpjets would not ultimately be used or their use would be generally disadvantageous - at the same time that is was central to justifying a choice of design partner who's proposal was substantially more expensive and slower to deliver, for which no more defensible reason could be found.
 \end{itemize}
 
-It appears that the pumpjet idea was simply a lead-character used to construct the fiction that France's vaunted heritage as a designer of nuclear submarines offers a substantial technical advantage to a non-nuclear submarine.
+In summary, I conclude that the pumpjet was a fraudulent, rather than incomptent proposal, used to construct the fiction that France's vaunted heritage as a designer of nuclear submarines offers a substantial technical advantage to a non-nuclear submarine.
 There is almost no chance that it will be included in the detailed design.
-In reality, without a nuclear reactor, there is barely any relevant technology or experience (other than the combat system, mandated from a 3rd party) which is usefully transferrable.
-The extent to which the bottom of the barrel needed to be scraped was illustrated when RADM Sammut was challenged on the benefit of the nuclear design, other than pump-jets \ref{SENhanson}, and he was forced to refer to the `diameter of the hull' as being a key learning, a prospect which would be much more easily subjected to public ridicule had it been more substantially promoted.\footnote{Amongst the many engineering challenges presented in designing a new submarine, ensuring the structural integrity of the hull at a new diameter is one of the most well-understood and predictable.
+In reality, without a nuclear reactor, there is relatively little relevant nuclear-related technology or experience (other than the combat system, mandated from a 3rd party) which is usefully transferable.
+The extent to which the bottom of the barrel needed to be scraped to maintain this fiction was illustrated when RADM Sammut was challenged on the benefit of the nuclear design after facing pressure on the pumpjet, and was forced to refer to the `diameter of the hull' as being a key learning[\ref{SENhanson}].
+Given the relatively more accessible engineering concepts involved, such claim would be much more easily subjected to public ridicule had it been more substantially promoted, as the pumpjet was.\footnote{Among the many engineering challenges presented in designing a new submarine, ensuring the structural integrity of the hull at a new diameter is one of the most well-understood and predictable.
 The Collins Class was evolved from a submarine of smaller hull diameter, and this aspect of the design and development encountered no problems.
-In 2006, DSTO commissioned a paper suggesting that the Collins Class replacement be designed at a wider hull, which didn't correspond to that of the Barracuda.}  In contrast, those countries whos own navies relied entirely upon diesel-electric submarines would likely boast significant advantages in technologies which are genuine game-changers for diesel-electric submarines, such as advanced composite propellers, Air-Independent Propulsion, and Lithium-Ion batteries, which we are now told won't be included in the Future Submarine \textit{because} of the pumpjet\parencite{bergmann2018}.
-
-Australians have a right to know what the true motive of Defence was in choosing the French as a design partner.
-There can be no doubt that the pumpjet played no legitimate part in it.
-Any claims to the contrary are made either out of obscene incompetence, as Defence's latest response might have us believe, or more likely are made with the intent of being misleading.
+In 2006, DSTO commissioned a paper suggesting that the Collins Class replacement be designed at a wider hull, which didn't correspond to that of the Barracuda.\parencite{joubert2006}}  In contrast, those countries whose own navies relied entirely upon diesel-electric submarines would likely boast significant advantages in technologies which are genuine game-changers for diesel-electric submarines, such as advanced composite propellers, Air-Independent Propulsion, and Lithium-Ion batteries, which we are now told won't be included in the Future Submarine \textit{because} of the pumpjet\parencite{bergmann2018}.
+To that extent, it is almost certain that the pumpjet proposal will not only impose a substantial cost on Australian taxpayers (tens of billions of dollars), but also substantially degrade Australia's future military capability.
+
+In light of these considerations, Defence should be required to explain again what their reasons were for choosing the French as a design partner, as there can now be no doubt that the pumpjet played no legitimate part in it.
+Any claims to the contrary are made either out of incompetence, as Defence's latest response might have us believe, or are made with the deliberate intent of misleading.
 
@@
 
-That paper, and a subsequent blog post \parencite{morrison2018a} on the topic which made the paper publically available, attracted the interest of the Senate Economics Reference Committee Inquiry into the Future of Naval Shipbuilding in Australia.
+That paper, and a subsequent blog post \parencite{morrison2018a} on the topic which made the paper publicly available, attracted the interest of the Senate Economics Reference Committee Inquiry into the Future of Naval Shipbuilding in Australia.
 Senator Chris Ketter asked written questions on notice to the Department of Defence requesting responses to the paper specifically, and related comments from public officials, with a return date of 3rd April 2018.
@@
 In their own evidence during the course of the day, neither of those two did choose to challenge it.
-Senator Kim Carr requested that defence produce a written response to that report for the committee prior to their handing down of their finaly report for the Inquiry, and Defence agreed to do their best in the time available, with the head of Capability and Sustainment Group (CASG) Kim Gillis assuring the committee that ``We  will  provide  the  committee  with  the  best  possible  advice  wecan  in  the  shortest  possible  time  frame  to  meet  your  requirements  so  that  you  can  be  betterinformed, because we do hold that expertise." Senator Carr emphasised the importance of timeliness that ``it would look pretty bloody stupid after the report's handed in''.
-Despite these exhortations, in the final report of the committee this request, and defence's acceptance of it was noted, but no acknowledgement was made of whether any response was actually received, or what it contained \parencite{SERC2018final}.
+Senator Kim Carr requested that defence produce a written response to that report for the committee prior to their handing down of their final report for the Inquiry, and Defence agreed to do their best in the time available, with the head of Capability and Sustainment Group (CASG) Kim Gillis assuring the committee that ``We  will  provide  the  committee  with  the  best  possible  advice  we can  in  the  shortest  possible  time  frame  to  meet  your  requirements  so  that  you  can  be  better informed, because we do hold that expertise." Senator Carr emphasised the importance of timeliness that ``it would look pretty bloody stupid after the report's handed in''.
+Despite these exhortations, in the final report of the committee this request, and Defence's acceptance of it was noted, but no acknowledgement was made of whether any response was actually received, or what it contained \parencite{SERC2018final}.
 
@@
 Comment for the article was sought from CASG, the Royal Australian Navy (RAN) and Naval Group, and none was forthcoming.
-However, some substantial comments were provided by Dr David Kershaw, the chief of DSTG's Maritime Division, who was also Science and Technology advisor for a time on SEA 1000, the Future Submarine Project.
+However, some substantial comments were provided by Dr David Kershaw, the chief of DSTG's Maritime Division, who was also Science and Technology Advisor for a time on SEA 1000, the Future Submarine Project.
 
@@
 At the time of its original commissioning, two substantive arguments emerge seeking to engage with the technical arguments made in the earlier paper.
-The first is that the large size of the submarine now makes the pumpjet viable, where as for smaller submarines the large weight of the pumpjet would make it unviable, and that the pumpjet would be generally well-suited to smaller sumbarines as well were it not for this weight consideration.
+The first is that the large size of the submarine now makes the pumpjet viable, where as for smaller submarines the large weight of the pumpjet would make it nonviable, and that the pumpjet would be generally well-suited to smaller submarines as well were it not for this weight consideration.
 The second is that interaction between the pumpjet and the hull obviate the evidence I have produced showing that pumpjet efficiency declines at low speeds, which are claimed to neglect considering such interactions.
@@
 \begin{tcolorbox}[width=0.9\textwidth, center, colback=blue!5!white,colframe=blue!75!black, title= \textbf{8 Apr 2016 - Sean Costello, CEO DCNS Australia, and Gerard Autret, Chief Naval Architect of the Shortfin Barracuda}]
-\textit{In a blog post\parencite{autret2016} for the ASPI Strategist prior to the announcement in April, Sean Costello describes a number of the major features and design approaches for the Shortfin Barracuda, including relating the choice of the pumpjet closely to an accoustic requirement to close with another submarine at higher speed.}
+\textit{In a blog post\parencite{autret2016} for the ASPI Strategist prior to the announcement in April, Sean Costello describes a number of the major features and design approaches for the Shortfin Barracuda, including relating the choice of the pumpjet closely to an acoustic requirement to close with another submarine at higher speed.}
 \\
@@
 Whether it is viable for a diesel-electric submarine (as opposed to the nuclear one mentioned here) to maintain such speeds to close with many types of targets is left unclear, or at least unmentioned.
-As later discussed in my evidence to the Senate \ref{SENmorrison}, this is consistent with DCNS being well aware with the advantages and disadvantages of the pumpjet, whilst not making verifyably false claims.
+As later discussed in my evidence to the Senate \ref{SENmorrison}, this is consistent with DCNS being well aware with the advantages and disadvantages of the pumpjet, whilst not making verifiably false claims.
 
@@
 More significantly, the speeds achieved would not allow the submarine to close with any other target from behind such as almost all surface ships, or nuclear submarines.
-This statement exhibits extreme care in the wording in order to be plausibly defensible, and is aguably misleading.
+This statement exhibits extreme care in the wording in order to be plausibly defensible, and is arguably misleading.
 
@@
 \begin{tcolorbox}[width=0.9\textwidth, center, colback=blue!5!white,colframe=blue!75!black, title= \textbf{4 October 2017 - Jean-Michel Billig, Executive Director Australian submarine Program, Naval Group}]
-\textit{In comments to journalists at the PACIFIC 2017 trade show, the Naval Group executive in charge of the Australian Program acknowledges that convetional propellers may be used on the Shortfin Barracuda \parencite{sbs2017}.
+\textit{In comments to journalists at the PACIFIC 2017 trade show, the Naval Group executive in charge of the Australian Program acknowledges that conventional propellers may be used on the Shortfin Barracuda \parencite{sbs2017}.
 Subsequent reports quote Naval Group shifting responsibility for the propulsion choice to the Australian Government.}
@@
 Simultaneously, the statement that technology choice is a product of the operational requirement makes it clearly implies that the operational requirement has not yet been determined with sufficient precision to make the suitable technology.
-As a ship-builder with ample experience in both technologies, uniquely positioned as having fitted both pumpjets and propellers to both diesel and nuclear submarines, it is remakeable that the performance information transferred prior or during the CEP was not sufficient to allow a decision  on the technology to be made.
-Moreover, it also reveals that this was not only the case at the conlcusion of the CEP, and remained so well over a year after.
-
-\begin{tcolorbox}[width=0.9\textwidth, center, colback=blue!5!white,colframe=blue!75!black, title= \textbf{25 Oct 2017 - Hevre Guillou, CEO Naval Group}]
+As a ship-builder with ample experience in both technologies, uniquely positioned as having fitted both pumpjets and propellers to both diesel and nuclear submarines, it is remarkable that the performance information transferred prior or during the CEP was not sufficient to allow a decision  on the technology to be made.
+Moreover, it also reveals that this was not only the case at the conclusion of the CEP, and remained so well over a year after.
+
+\begin{tcolorbox}[width=0.9\textwidth, center, colback=blue!5!white,colframe=blue!75!black, title= \textbf{25 Oct 2017 - Herve Guillou, CEO Naval Group}]
 \textit{In an interview that was published in a post\parencite{guillou2017} on the ASPI Strategist the global CEO of Naval group responds to a question from Brendan Nicholson asking whether the Australian submarine will be fitted with a pump jet, following Jean-Michel Billig's comments.
@@
 However, she succeeded in applying sustained pressure to the question of the pumpjet, and the significance that it had in the selection of DCNS as a design partner.
-This elicited a committment to the pumpjet that appeared hesitant in its wording, which noted that the pump jet ``remains on offer", which is what ``we intend to use".
+This elicited a commitment to the pumpjet that appeared hesitant in its wording, which noted that the pump jet ``remains on offer", which is what ``we intend to use".
 
 RADM Sammut goes to length to explain that the pumpjet was not ``alone" the reason for selecting DCNS as a partner, and in doing so reveals important insights into Defence's thinking and process regarding the decision.
-Crucially, he reveals that the `diameter of the hull' is an important aspect of France's design reference in the Barracuda, pointing out that changing the diameter of the hull incurrs a ``whole new set of calculations about the structural integrity and strength of the hull."  Together with remarks about the ``way that safety is incorporated into the design" and concerns about the alignment of the certifying process with the ``design philosophy" it would appear that the hull diameter actually emerges as plausibly one of the most significant distinguishing features in the French bid, perhaps alongside the pumpjet, and aligns with emphasis given to the exact diameter of the pumpjet remaining unchanged in media report earlier in the year.\parencite{kerr2017}
+Crucially, he reveals that the `diameter of the hull' is an important aspect of France's design reference in the Barracuda, pointing out that changing the diameter of the hull incurs a ``whole new set of calculations about the structural integrity and strength of the hull."  Together with remarks about the ``way that safety is incorporated into the design" and concerns about the alignment of the certifying process with the ``design philosophy" it would appear that the hull diameter actually emerges as plausibly one of the most significant distinguishing features in the French bid, perhaps alongside the pumpjet, and aligns with emphasis given to the exact diameter of the pumpjet remaining unchanged in media report earlier in the year.\parencite{kerr2017}
 
@@
 If the hull diameter is already fixed and known as being identical to that of the nuclear Barracuda, then no adjustment ought to be necessary to a good working pumpjet design that is used to propel this platform on the hull's account.
-The performance of the pumpjet on the relevant hull would be well understood by the manufacturer offering it, and presumeably available to the Australian government who were looking to acquire it.
+The performance of the pumpjet on the relevant hull would be well understood by the manufacturer offering it, and presumably available to the Australian government who were looking to acquire it.
 Further discussion of the hull-propulsor interaction is given in Section \ref{hullpropulsor}.
 
-The second is that is reveals a stunning naivety in terms of basic understanding of the heirachy of engineering challenges.
+The second is that is reveals a stunning naivety in terms of basic understanding of the hierarchy of engineering challenges.
 It may be true that there are very many complex and challenging engineering calculations involved in building a submarine.
-However, it is certainly not true that scaling up a pressure hull to a new diameter to meet the same fixed static load is amongst the most difficult, risky or complex.
+However, it is certainly not true that scaling up a pressure hull to a new diameter to meet the same fixed static load is among the most difficult, risky or complex.
 Whilst extending the length of a submarine is undoubtedly even easier, adjusting the diameter still to meet a structural integrity requirement to meet a known static pressure (driven by the dive depth) is a relatively simple calculation in comparison to the complexities of many other calculations involved in submarine design.
-For instance, re-designing the duct intake of blade arrangments of a pumpjet to maximise efficiency at a different speed whilst simultaneously attempting to avoid cavitation would be vastly more complex and uncertain given the unstable difficult-to-model dynamics of turbulence and cavitation involved.
+For instance, re-designing the duct intake of blade arrangements of a pumpjet to maximise efficiency at a different speed whilst simultaneously attempting to avoid cavitation would be vastly more complex and uncertain given the unstable difficult-to-model dynamics of turbulence and cavitation involved.
 In contrast, scaling up a hull-diameter was quite simply and successfully accomplished when the Collins Class submarine was developed by scaling up both the length and diameter of it's smaller Swedish predecessor.
@@
 That this ranks as a significant consideration in the mind of the chief of this program should be the cause of considerable alarm to those taking an interest the project, as it would appear that he is quite alone in this belief that this really is a significant or risky hurdle in a project of this scale.
-Were it known to the competitors that their ability to achieve this simple engineering requirement was in question, one might assume they would be highly offended, since any contender that was not considered competent to design a pressure hull of a new diameter would presumeably never have been invited to participate.
+Were it known to the competitors that their ability to achieve this simple engineering requirement was in question, one might assume they would be highly offended, since any contender that was not considered competent to design a pressure hull of a new diameter would presumably never have been invited to participate.
 RADM Sammut should be called upon to clarify what he meant by ``we needed to make sure that the certifying process would match the design philosophy" because if indeed he meant that Defence had already decided that their own certifying process already precluded scaled-up diameter hulls, this represents a very bold departure from what might be considered normal risk assessment processes for major engineering projects, including those used in designing and certifying the Collins.
@@
 
-Amongst RADM Sammut's additional claims regarding the reason for choosing DCNS/Naval Group, include remarks regarding the establishment of 'sovereign capability' and 'data transfer' to Australia.
-One would assume that this meant that the French had demonstrated a willingness to supply complete and timely information on key technologies, and that RADM Sammut's exisiting experience with the French supplying such information on the pumpjet (as Sean Costello had claimed had taken place during and prior to the conclusion of the CEP) supported this view.
+Among RADM Sammut's additional claims regarding the reason for choosing DCNS/Naval Group, include remarks regarding the establishment of 'sovereign capability' and 'data transfer' to Australia.
+One would assume that this meant that the French had demonstrated a willingness to supply complete and timely information on key technologies, and that RADM Sammut's existing experience with the French supplying such information on the pumpjet (as Sean Costello had claimed had taken place during and prior to the conclusion of the CEP) supported this view.
 
@@
 This is at odds with the statement made only a few weeks earlier by Jean-Michel Billig.
-This is also the first noteable reference to the inter-relationship between the hull and the pumpjet, implying that the pumpjet will indeed be modified in order to suit the hull, and that the hull will be modified to suit the pumpjet.
+This is also the first notable reference to the inter-relationship between the hull and the pumpjet, implying that the pumpjet will indeed be modified in order to suit the hull, and that the hull will be modified to suit the pumpjet.
 This contradicts the notion that the same pumpjet as France is fitting to the Suffren (i.e.
-nuclear barracuda) will simply be inlcuded 'as is' on the Australian non-nuclear 'Shortfin' version.
+nuclear barracuda) will simply be included 'as is' on the Australian non-nuclear 'Shortfin' version.
 Given that the length and diameter of the hull were announced as being already decided and practically identical to the nuclear version, it raises the very serious question as to whether/why the changed shape of the hull would necessitate modification of the complex pumpjet.
@@
 Cross-ocean transits regularly occur at higher speeds than 17-18kt, including by cruise ships, which he mentions.
-These speeds are essentially unreacheable (or only extremely briefly sustainable) by a diesel electric submarine.
-As pointed out in my evidence to Senate Economic References Committee \ref{SENmorrison} I can find no evidence of any cruise ships having any kind of shround around the tips of their propellers, or bearing any resemblance or relationship to a pumpjet.
+These speeds are essentially unreachable (or only extremely briefly sustainable) by a diesel electric submarine.
+As pointed out in my evidence to Senate Economic References Committee \ref{SENmorrison} I can find no evidence of any cruise ships having any kind of shroud around the tips of their propellers, or bearing any resemblance or relationship to a pumpjet.
 In the same session, whilst in Mr Johnson's presence, it appears that Mr Johnson may have confused electric Azipods, which are used on large ocean-going ships such as cruise ships, and the Navy's new LHDs, with pumpjets.
@@
 Prior research consisted of a single 99 page report obtained via literature review and discussion with experts at a cost of \$68,825.
-Ongoing research included \$1.1million spent to date to develop a generic submarine and pumpjet design, compuational and analytic models, and validation of these models through experimental studies.
+Ongoing research included \$1.1million spent to date to develop a generic submarine and pumpjet design, computational and analytic models, and validation of these models through experimental studies.
 The development and test program is not scheduled to commence until the last quarter of 2018.
@@
 \begin{tcolorbox}[width=0.9\textwidth, center, colback=black!5!white,colframe=black!75!black, title= \textbf{3 April 2018 - Department of Defence, in response to Question on Notice from Senator Chris Ketter in December 2018}]
-\textit{Senator Ketter asks Question 1, referencing specifically Greg Sammut in \parencite{sammut2017} and my paper \parencite{morrison2018} asks specifically whether Defence maintians that the pumpjet can be efficient across the entire speed range.
+\textit{Senator Ketter asks Question 1, referencing specifically Greg Sammut in \parencite{sammut2017} and my paper \parencite{morrison2018} asks specifically whether Defence maintains that the pumpjet can be efficient across the entire speed range.
 Full question can be found in \ref{QON1}.}
@@
 This is a question which will be given a substantial reply in subsequent sections, but even a quick reading superficial reading reveals alarming levels of over-reach in the answer.
-(For intance, a type of ducted propeller known as a decelerating ducted propeller actually \textit{is} a pumpjet, by definition.
+(For instance, a type of ducted propeller known as a decelerating ducted propeller actually \textit{is} a pumpjet, by definition.
 An entire chapter (5) of the paper to which Defence is supposedly responding \parencite{morrison2018} discusses this at some length.)
@@
 In such a scenario Defence is also acting to shelter Naval Group from accusations of fraud as well as the likely legal proceedings that would be open to the Commonwealth if information provided in the CEP was found to be fraudulent.
-It's probable that this constitutes misleading parliament, but of far greater concern is the possibility that the necessary corrective action to avoid a procurement disaster of unprecedent scale is not yet being taken by Defence.
+It's probable that this constitutes misleading parliament, but of far greater concern is the possibility that the necessary corrective action to avoid a procurement disaster of unprecedented scale is not yet being taken by Defence.
 Why that could be the case is also a matter of substantial public interest.
@@
 It is possible to believe that Defence's own embarrassment at having been unable to detect the insincerity in Naval Group's offer of the pumpjet, and inability to reach their own technical conclusions quickly enough to detect the fraud, may motivate the present actions.
-This would  amount to a disasterous failure on the part of the public service to act with integrity and in the national interest.
-However, if Defence had truely been un-wittingly deceived, it is difficult to see how they could not recover their pride (and regain a necessary measure of respect from their design partner) by quickly adopting a more combative stance, calling out the deceptions on the part contractors part, and quickly exacting whatever contractual reparations they might be entitled to.
+This would  amount to a disastrous failure on the part of the public service to act with integrity and in the national interest.
+However, if Defence had truly been un-wittingly deceived, it is difficult to see how they could not recover their pride (and regain a necessary measure of respect from their design partner) by quickly adopting a more combative stance, calling out the deceptions on the part contractors part, and quickly exacting whatever contractual reparations they might be entitled to.
 If the contractual reparations were not sufficiently strong, the Commonwealth's freedom to not proceed to a build contract (from the initial design contract) or re-introduce competition into the design process as many have suggested should occur in any case \parencite{stanford2017}, would have provided ample commercial leverage to reset the relationship in a manner which was better aligned with the Commonwealth's interest.
@@
 One must at this point give consideration to the possibly that Defence had not been entirely unwittingly deceived, and was actually complicit in their own deception, including its present upkeep.
-This would mean that the Department of Defence and the Navy, or at least a few of very senior people within those organisations, had conspired allow a proposal which was known to be inappropriate to the Commonwealth's needs to appear victorious in the CEP,  presumeably in order to favour a bidder for whom no more defensible explanation could be found.
+This would mean that the Department of Defence and the Navy, or at least a few of very senior people within those organisations, had conspired allow a proposal which was known to be inappropriate to the Commonwealth's needs to appear victorious in the CEP,  presumably in order to favour a bidder for whom no more defensible explanation could be found.
 It appears unlikely that there could be any decent motive for such a course of action, and would lead invariably to the grave conclusion that some kind of corruption has occurred.
-In the circumstances, one must evaluate the probabilities of the necessary oversights for such an eventuatlity occurring by accident against the occurring by arrangement.
+In the circumstances, one must evaluate the probabilities of the necessary oversights for such an eventuality occurring by accident against the occurring by arrangement.
 If the pumpjet was known by DCNS to be an unsuitable proposition for a diesel-electric submarine, a seemingly unlikely degree of naive incompetence would be required for Defence to fail to ascertain as much during the Competitive Evaluation Process.
@@
 \begin{tcolorbox}[width=0.9\textwidth, center, colback=blue!5!white,colframe=blue!75!black, title= \textbf{17 May 2018 - Brent Clark, Acting CEO Naval Group Australia}]
-\textit{In an interview that was published in a post\parencite{guillou2017} on the ASPI Strategist the Autralian CEO of Naval group explains to Brendan Nicholson why Naval Group is recommending the pumpjet for the Shortfin Barracuda, and answers some critics.}
+\textit{In an interview that was published in a post\parencite{guillou2017} on the ASPI Strategist the Australian CEO of Naval group explains to Brendan Nicholson why Naval Group is recommending the pumpjet for the Shortfin Barracuda, and answers some critics.}
 \\
@@
 This is also the first appearance of the argument that the size of the submarine is a crucial factor that makes the pumpjet viable.
-It is the closest that Naval Group has come to making a firm statement of faith in the suitability of the technology for the Australian diesel-elctric submarine.
-As I discussed in another blog post in ASPI Strategist \parencite{morrison2018b} shortly afterwards, this is a remarkeably concession that is consistent with the physics that I outlined, since the inefficiency of the pumpjet would likely drive demand for additional fuel and batteries (hence size) to meet requirements.
+It is the closest that Naval Group has come to making a firm statement of faith in the suitability of the technology for the Australian diesel-electric submarine.
+As I discussed in another blog post in ASPI Strategist \parencite{morrison2018b} shortly afterwards, this is a remarkably concession that is consistent with the physics that I outlined, since the inefficiency of the pumpjet would likely drive demand for additional fuel and batteries (hence size) to meet requirements.
 As such, this admission is more of a concession of the detrimental impact of the pumpjet on the submarine.
@@
 Only months earlier Defence was asked a specific question on notice about whether they thought Naval Group maintained the opinion that a pumpjet was suitable for a diesel-electric submarine.
-Clarke's interview on the Strategist, which seemed to be strongest statement of faith from Naval group in the suitability of the pumpjet for a conventional submarine (perhaps in an effort to patch up the noteable refusal to answer the question) had been answered publically by the same person who had just appeared and he was being asked to respond to.
+Clarke's interview on the Strategist, which seemed to be strongest statement of faith from Naval group in the suitability of the pumpjet for a conventional submarine (perhaps in an effort to patch up the notable refusal to answer the question) had been answered publicly by the same person who had just appeared and he was being asked to respond to.
 It might have been expected that Mr Clark would at least attempt to demonstrate conviction about the merits of the proposal.
@@
 This exchange demonstrates a dismissive attitude towards the evidence I had presented, which borders on contempt.
-The attitude is clearly pervasive and supported in the higest levels of the defence department.
+The attitude is clearly pervasive and supported in the highest levels of the defence department.
 The General Manager Submarines, Stephen Johnson, claims he simply cannot recall what he said about pumpjets being related to pods on cruise ships, despite being physically present in the same room when I pointedly quoted and contradicted him.
-This being pointed out, he claimed to simply not to be paying attention, which won the support of the Head of Capapbility and Sustainment Group, Kim Gillis.
+This being pointed out, he claimed to simply not to be paying attention, which won the support of the Head of Capability and Sustainment Group, Kim Gillis.
 Despite the obvious shock and exclamation of the senators, Mr Gillis went on to attempt to dismiss my evidence as just another opinion which could probably to be dismissed along with 99\% of the others.
@@
 
-Such an openly dismissive attitude, exemplified by Mr Gillis' ``not all of this is rivetting" raises questions about whether Mr Gillis's evidence is even given genuinely and honestly.
+Such an openly dismissive attitude, exemplified by Mr Gillis' ``not all of this is riveting" raises questions about whether Mr Gillis's evidence is even given genuinely and honestly.
 For instance, Mr Gillis' claim that he had not read the report, since ``we weren't actually  given a copy" warrants serious attention.
-My paper was made publically available with the blog-post summarising it on the 14th of February, and as well as in my reply to Brend Clark on  the 24th of May.
+My paper was made publicly available with the blog-post summarising it on the 14th of February, and as well as in my reply to Brent Clark on  the 24th of May.
 Moreover, his department had received written Questions on Notice from Senator Chris Ketter, acting on behalf of this very committee, specifically requesting written responses to my paper with a return date of 3 April.
 His department most probably responded to prior to the hearing, as noted above, with responses which blatantly refused to answer the questions.
-\footnote{However, I was not made aware either of the questions or their answers prior to hearing.} Furthermore this particular hearing was also especially long in the making, since it was once postponed on account of Defence refusing to make officials available to attend.\parencite{adm2018}  My presence as a witness was requested in February when the proposed date was still in May, and the list of witnesses was publically circulated during the weeks prior to the hearing.
+\footnote{However, I was not made aware either of the questions or their answers prior to hearing.} Furthermore this particular hearing was also especially long in the making, since it was once postponed on account of Defence refusing to make officials available to attend.\parencite{adm2018}  My presence as a witness was requested in February when the proposed date was still in May, and the list of witnesses was publicly circulated during the weeks prior to the hearing.
 In such circumstances, claims to be essentially ignorant of my report and pretend it could probably be dismissed after the reading of the first page, should be examined critically.
@@
 He is absolutely right to say that the shaft horse-power required to move a large vessel at low speeds is astonishingly small.
-This is because the resistance of a mody moving through a fluid scales generally with velocity squared, as was discussed at length in my paper, since it comprised one of the `intricacies' which I considered particularly important to this matter.
+This is because the resistance of a body moving through a fluid scales generally with velocity squared, as was discussed at length in my paper, since it comprised one of the `intricacies' which I considered particularly important to this matter.
 His blunt description of paying `two pennies' when you could pay only one, however, betrays a staggering misunderstanding of the energy densities relevant to conventional submarines.
@@
 
-In contrast to Mr Johnson's inference that my report might be made `without contemporary knowledge or without understanding nececessarily the intricacy', his own comments on powering requirements of ships, just like the propulsive pods for ships, demonstrate a complete disregard or ignorance of the most basic and fundamental concepts relevant.
-
-\subsection{Dr David Kershaw, Chief of Martime Division DSTG, in Australian Defence Magazine}
+In contrast to Mr Johnson's inference that my report might be made `without contemporary knowledge or without understanding necessarily the intricacy', his own comments on powering requirements of ships, just like the propulsive pods for ships, demonstrate a complete disregard or ignorance of the most basic and fundamental concepts relevant.
+
+\subsection{Dr David Kershaw, Chief of Maritime Division DSTG, in Australian Defence Magazine}
 \begin{tcolorbox}[width=0.9\textwidth, center, colback=black!5!white,colframe=black!75!black, title= \textbf{August 2018 - Dr David Kershaw, in Australian Defence Magazine}]
@@
 \textbf{You just have to have the capabilities and facilities to be able to design an efficient pumpjet and we know that France can do that."}
-Echoing RADM Sammut, Dr Kershaw stressed that the pumpjet on a submarine must be designed in conjunction with the hullform to which it was to be attached.
+Echoing RADM Sammut, Dr Kershaw stressed that the pumpjet on a submarine must be designed in conjunction with the hull-form to which it was to be attached.
 ``Its' not like a propeller where you can design your propeller open loop i.e.
@@
 This represents a pure, brazen, and obvious falsehood, justified by a seemingly-relevant truth.
-It's akin to arguing 'camels have to cary two heavy humps on their backs.
+It's akin to arguing 'camels have to carry two heavy humps on their backs.
 Pigs don't have to carry these two heavy humps, which is why they can fly.'  It also directly contradicts the claim made by Stephen Johnson before the Senate, that ``No matter what kind of propeller or propulsion pump jet  you design, it has to be optimised for a particular speed.".
 
-Whilst it is true that deeply submerged bodies do not need to overcome wave-making resistance, all bodies in contact with any fluid must also overcome skin friction, wich means that resistance from this source scales with \textit{velocity squared} in almost all circumstances, including for ships.
+Whilst it is true that deeply submerged bodies do not need to overcome wave-making resistance, all bodies in contact with any fluid must also overcome skin friction, which means that resistance from this source scales with \textit{velocity squared} in almost all circumstances, including for ships.
 This is a most fundamental and basic assumption which is absolutely intrinsic to all basic marine propulsion, and should also be understood by essentially anyone who has passed first-year engineering or physics courses.
-(For instance, see Pages 9 and 10 of Reference \parencite{MAN2017}, or Chapter 12 of Reference \parencite{carlton2007}, or Chapter 3 of Reference \parencite{mollard2011}.) Apart from it being quite an elementary law of phyiscs (the Drag Law), referred to and relied upon in posts such those by Andrew Davies \parencite{davies2017} commenting on this exact question, any substantial reading of the literature focussed on the design of pumpjets for fully submerged objects (such as torpedoes or submarines) will also make it abundantly clear, and the special distinction with wave-making resistance (which is crucial for fast-vessels in the process of reaching the plane) is also abundantly clear, and has been well known for a long time \parencite{wislicenus1973} \parencite{henderson1964}.
+(For instance, see Pages 9 and 10 of Reference \parencite{MAN2017}, or Chapter 12 of Reference \parencite{carlton2007}, or Chapter 3 of Reference \parencite{mollard2011}.) Apart from it being quite an elementary law of physics (the Drag Law), referred to and relied upon in posts such those by Andrew Davies \parencite{davies2017} commenting on this exact question, any substantial reading of the literature focused on the design of pumpjets for fully submerged objects (such as torpedoes or submarines) will also make it abundantly clear, and the special distinction with wave-making resistance (which is crucial for fast-vessels in the process of reaching the plane) is also abundantly clear, and has been well known for a long time \parencite{wislicenus1973} \parencite{henderson1964}.
 Further discussion is given in a subsequent section of this paper.
@@
 He gives the most detailed description of the interaction between the hull and the pumpjet thus far, arriving at the same conclusion that they have to be `tuned' together - the same wording that has been previously used by RADM Sammut.
-(This argument will be addressed fully in a later chapter.)  An apparent conclusion is that RADM may have relied upon advice from Dr David Kershaw, (which given the ineptitute of earlier comments relating to hydrodynamics is gravely concerning), or that both of them have relied upon advice from another common source.
+(This argument will be addressed fully in a later chapter.)  An apparent conclusion is that RADM may have relied upon advice from Dr David Kershaw, (which given the ineptitude of earlier comments relating to hydrodynamics is gravely concerning), or that both of them have relied upon advice from another common source.
 
@@
 Although pump jet propulsion would be new to the RAN, the technology was well-known to DSTG.
-``We've invested heavily in our hydrodynamics and hydroacoustics program over the past 20 years and this is includes the computational models as well as experimental facilities that confirm modelling results," Dr Kershaw said to \textit{ADM}.
+``We've invested heavily in our hydrodynamics and hydro-acoustics program over the past 20 years and this is includes the computational models as well as experimental facilities that confirm modelling results," Dr Kershaw said to \textit{ADM}.
 \textbf{We've been able to develop our analysis on a generic pumpjet design and thus confirm our understanding of pumpjet performance; you basically build it up from first principles.
@@
 These comments are extremely important because they illuminate the true state of DSTG's research, and the degree of confidence they have in the pumpjet.
-It is consistent with responses given to a question on notice from Rex Patrick \ref{QON150}, where it was revealed that 1.1 million was spent on ``development of a \textit{generic submarine and pumpjet design}, computational and analytics models, and validation of these models through experimental studies".
+It is consistent with responses given to a question on notice from Rex Patrick \ref{QON150}, where it was revealed that 1.1 million was spent on ``development of a \textit{generic submarine and pumpjet design}, computational and analytic models, and validation of these models through experimental studies".
 The admission that DSTG is still working on a `generic' design also casts into new light their repeated insistence that the performance of a pumpjet cannot be assessed unless it is exactly tuned carefully with the hull, and vice versa.
@@
 
-As I have already put before the Senate \ref{SENmorrison}, I find that idea that such `tuning' could so dramatically alter the performance of the pumjet to be completely implausible.\footnote{In my evidence is said it was: ``as if someone bought an efficient hybrid car like a Prius and then decided to fit it with tank tracks from a dozer or a battle tank, and then argued they would tune the tank tracks back to be as efficient as the Prius was in the first place.
+As I have already put before the Senate \ref{SENmorrison}, I find that idea that such `tuning' could so dramatically alter the performance of the pumpjet to be completely implausible.\footnote{In my evidence is said it was: ``as if someone bought an efficient hybrid car like a Prius and then decided to fit it with tank tracks from a dozer or a battle tank, and then argued they would tune the tank tracks back to be as efficient as the Prius was in the first place.
 It goes radically against what the technology was intended for in the first place".}  Rather, I would support the idea that the kind of generic modelling described would be quite sufficient to reveal results that are consistent with the results I produced from a basic literature review (that pumpjet efficiency declines towards zero at very low water-speeds) and safely rule-out the suitability of a pumpjet for diesel electric submarines.
@@
 Given the ample resources of Defence and Naval Group, this insistence that the result cannot yet actually be known, rather than that they know better and reached a different result and raises the question of what they actually do know, and how different it actually is to my result.
-If the DSTG studies are actually consistent with my findings, then DSTG and Defence are knowingly supressing such results, or naively hoping that the French design, and the finer `tuning' between the pumpjet and the hull can avhieve such a dramatically diferent result that the results of `generic' tests can be ignored.
+If the DSTG studies are actually consistent with my findings, then DSTG and Defence are knowingly suppressing such results, or naively hoping that the French design, and the finer `tuning' between the pumpjet and the hull can achieve such a dramatically different result that the results of `generic' tests can be ignored.
 Either situation would have terrible consequences for the program, and if the former is true it is almost certain that the Parliament has been deliberately misled on multiple occasions.
@@
 
-It is even more shocking in the light of the fact that the CEO of Naval Group Australia, Sean Costello's, publically boasted about the significance of the ``offer from France to transfer to Australia sovereign control and use of pump-jet propulsion technology" \parencite{costello2016}, and his simultaneous claim (prior to the conclusion of the CEP) in the same that the performance data of the pumpjet are known by Australian government.
-
-It is still more shocking in the light of the then outgoing acting CEO, Brent Clark's further confimration before a Senate Inquiry \ref{SENmorrison} ``we've been supplying" classified documents on this topic to the Australian Government.
+It is even more shocking in the light of the fact that the CEO of Naval Group Australia, Sean Costello's, publicly boasted about the significance of the ``offer from France to transfer to Australia sovereign control and use of pump-jet propulsion technology" \parencite{costello2016}, and his simultaneous claim (prior to the conclusion of the CEP) in the same that the performance data of the pumpjet are known by Australian government.
+
+It is still more shocking in the light of the then outgoing acting CEO, Brent Clark's further confirmation before a Senate Inquiry \ref{SENmorrison} ``we've been supplying" classified documents on this topic to the Australian Government.
 
 In contrast to the assurance of full technology transfer, the comments by Dr Kershaw in this article reveals the true state of information flow from DCNS.
-Apparently, there is still not sufficent information to be confident about the use of the pumpjet.
+Apparently, there is still not sufficient information to be confident about the use of the pumpjet.
 A decision on whether the pumpjet will even be used still cannot be made.
@@
 Nor, it would appear, is sufficient theoretical modelling, or scale model testing, or any other confirmatory data sufficient on this pumpjet for DSTG to be clearly persuaded of the suitability of the pumpjet for our submarine, which will have an almost-identical hull.
-Whilst simultaneously claiming that efficiency and performance of the pumpjet can't be assessed without the precise form of the hull and the pumpjet being 'tuned' together, they admit that they are having to model a 'generic' pumpjet, building up from first principals, in order to have sufficient confidence in their own 'indepenent analysis' of the technologies.
+Whilst simultaneously claiming that efficiency and performance of the pumpjet can't be assessed without the precise form of the hull and the pumpjet being 'tuned' together, they admit that they are having to model a 'generic' pumpjet, building up from first principals, in order to have sufficient confidence in their own 'independent analysis' of the technologies.
 
@@
 It also throws into doubt the integrity of the statements made by RADM Sammut before Senate Estimates.
-In the days and weeks immediately prior to his testimony \ref{SENsammut}, Jean-Michel Billig had triggerd the controversy around the pumpjet at PACIFIC 2017, the largest Naval Show and conference in Australia, which occurs only each two years.
+In the days and weeks immediately prior to his testimony \ref{SENsammut}, Jean-Michel Billig had triggered the controversy around the pumpjet at PACIFIC 2017, the largest Naval Show and conference in Australia, which occurs only each two years.
 This had lead to substantial coverage on the ABC\parencite{lane2017}\parencite{iggulden2017}, Defence Connect\parencite{mcmahon2017a}\parencite{mcmahon2017b}, and ASPI Strategist\parencite{davies2017}, and almost certainly contributed to the line of questioning that Senator Hanson doggedly pursued about the pumpjet that day.
@@
 If Naval Group's level of information transfer was proceeding as promised, then RADM Sammut had accepted (as being superior to other bids proposals, and sufficienct for the Commonwealth's purposes) a sequence of data and technology transfer which would leave DSTG relying on their own `generic' models in trying to assess the suitability of the pumpjet, and Defence unable to answer confidently whether Naval Group even had confidence in the suitability of the technology two years after the conclusion of the evaluation process.
-Unless every other competitor had also promised to withold critical information to assess the suitability of a technology that was `integral' to their bid during the bid process and for two years after, there can be no plausible defence of such a decision.
-Alternatively, if Naval Group had promised more fluid information transfer, and this information had subsequently been withheld, RADM Sammut was actually concealing that a crucial break-down in relations with the major contractor had already occurred, with potentially disasterous consequences for the future of the submarine program.
+Unless every other competitor had also promised to withhold critical information to assess the suitability of a technology that was `integral' to their bid during the bid process and for two years after, there can be no plausible defence of such a decision.
+Alternatively, if Naval Group had promised more fluid information transfer, and this information had subsequently been withheld, RADM Sammut was actually concealing that a crucial break-down in relations with the major contractor had already occurred, with potentially disastrous consequences for the future of the submarine program.
 The only other viable alternative is that information transfer was not really something that commonwealth looked at in considering the selection of DCNS as a design partner, and RADM was knowingly misleading the Parliament when he was answering questions for in a Senate Estimates hearing.
@@
 The explanation given regarding AIP adds weight to the bizarre notion that a back-to-front design process is indeed taking place.
-The enormous transit distances cited have been well known for decades, and presumeably the requirement to meet these had been distilled into specific ranges and endurances long before the commencement of the CEP.
+The enormous transit distances cited have been well known for decades, and presumably the requirement to meet these had been distilled into specific ranges and endurances long before the commencement of the CEP.
 It's true that large amounts of fuel and food might consequently be required.
-But the idea that we `might be better off carrying more food and fuel' betrays either either a lack of confidence that the volumes which were presumeably propososed by all contenders alongside an AIP system were sufficient, or a casual aspiration to substantially enlarge the range and endurance of the submarine even further than  what the requirements demand.
-It would be relatively simple at the very early stages of concept design (prior to the diameter and legnth of the submarine being fixed) to incorporate both sufficient fuel and food for a specific overall range and endurance requirement, as well as an AIP system to enhance the dived endurance, without compromising the overall range and endurance.
+But the idea that we `might be better off carrying more food and fuel' betrays either either a lack of confidence that the volumes which were presumably proposed by all contenders alongside an AIP system were sufficient, or a casual aspiration to substantially enlarge the range and endurance of the submarine even further than  what the requirements demand.
+It would be relatively simple at the very early stages of concept design (prior to the diameter and length of the submarine being fixed) to incorporate both sufficient fuel and food for a specific overall range and endurance requirement, as well as an AIP system to enhance the dived endurance, without compromising the overall range and endurance.
 That this was not done, and instead the dimensions of the hull were fixed, whilst Navy continued to search for a system which might fit within this pre-defined platform envelope does not remotely resemble a competent strategy to acquire a capability to meet a fixed requirement.
@@
 Worryingly, it adds confirmation that RADM Sammut might genuinely believe that the hull-diameter (and its resemblance of a pre-existing design) ought to be a pre-determined feature of a submarine design, to which the performance and capability of the submarine is subordinate.
-Such a view, including held on the grounds of safety as suggested by RADM Sammut himself \ref{SENhanson}, is such a bizarre understanding of program risk, the heirachy of engineering challenges and normal spiral of platform design that it should call into question his competence as a project manager, and the technical advisors whom he depends upon.
+Such a view, including held on the grounds of safety as suggested by RADM Sammut himself \ref{SENhanson}, is such a bizarre understanding of program risk, the hierarchy of engineering challenges and normal spiral of platform design that it should call into question his competence as a project manager, and the technical advisors whom he depends upon.
 
@@
 These statements again use a seemingly relevant truth, in order to distract or obscure a subsequent falsehood.
-It is indeed true that surface ships do encounter wavemaking resistance, and that this varies substantially as a function of speed, and submarines do not encounter this  type of resistance.
+It is indeed true that surface ships do encounter wave-making resistance, and that this varies substantially as a function of speed, and submarines do not encounter this  type of resistance.
 It is true that skin friction, (which is the shear component of the drag forces of any body that moves through a viscous fluid which act along its sides) as well as pressure resistance (the forces normal to the hull, which are non-cancelling on the nose and the tail) comprises essentially all of the drag that applies to a fully submerged submarine, as shown in Figure \ref{fig:drag4.png}.
@@
 These forces scale with velocity squared, a fact which has been widely known and accepted since Froude conducted his skin-friction tests in the 1870s.\footnote{It should be noted that our understanding of skin friction has improved somewhat since the 1870s.
-In particular, the distiction between laminar and turbulent flows and the transition between these two regimes as demonstrated by Osborne Reynolds in the 1880s was significant, since it was discovered that laminar flows exerted a lower skin friction, and it is the turbulent flows which scale with velocity squared.
-However, whilst some effort is made to attmept to achieve some laminar flow over bow-mounted sensors in submarines at certain speeds, it is widely accepted that such a large vessel as a submarine is almost entirely enveloped by turbulent flows at essentially all operating speeds, and consequently the $V^2$ relationship has been essentially universally accepted as the dominant relationship for this purpose, including for the caluclation of scalar drag coefficients (which represent the impact of the shape and smoothness of the hull) to which much design effort and literature is devoted.}  The results of those tests are shown in Figure \ref{fig:dragfroude.png}.
+In particular, the distinction between laminar and turbulent flows and the transition between these two regimes as demonstrated by Osborne Reynolds in the 1880s was significant, since it was discovered that laminar flows exerted a lower skin friction, and it is the turbulent flows which scale with velocity squared.
+However, whilst some effort is made to attempt to achieve some laminar flow over bow-mounted sensors in submarines at certain speeds, it is widely accepted that such a large vessel as a submarine is almost entirely enveloped by turbulent flows at essentially all operating speeds, and consequently the $V^2$ relationship has been essentially universally accepted as the dominant relationship for this purpose, including for the calculation of scalar drag coefficients (which represent the impact of the shape and smoothness of the hull) to which much design effort and literature is devoted.}  The results of those tests are shown in Figure \ref{fig:dragfroude.png}.
 As such, speed plays the dominant role in considering the efficiency of propulsors on submerged bodies, since they also have their efficiency fall away at off-design conditions.
@@
 It is difficult to even believe that it could be held with any confidence by even moderately intelligent adults of little or no technical background who took a mild interest in researching the topic.
-Unless such an extremely level of incompetence pervades every defence official and scientist who were involved in preparing this statement (which was requested directly from Kim Gillis, Head of CASG and Stephen Johnson, General Manager Submarines) is difficult to avoid reaching the conlcusion that the statement was given in a state of extreme contempt for the intelligence of the intended reader, and constitutes a deliberate lie and active attempt to mislead.
+Unless such an extremely level of incompetence pervades every defence official and scientist who were involved in preparing this statement (which was requested directly from Kim Gillis, Head of CASG and Stephen Johnson, General Manager Submarines) is difficult to avoid reaching the conclusion that the statement was given in a state of extreme contempt for the intelligence of the intended reader, and constitutes a deliberate lie and active attempt to mislead.
 
@@
 
-Both Dr David Kershaw, Chief of Maritime Division at DSTG\parencite{kerr2018}, as well as Mr Brent Clark\parencite{clark2018} have both made arugments to the effect of suggesting that it was the Australian submarine's very large size which is what made the pumpjet viable, where it might not be for smaller submarines.
+Both Dr David Kershaw, Chief of Maritime Division at DSTG\parencite{kerr2018}, as well as Mr Brent Clark\parencite{clark2018} have both made arguments to the effect of suggesting that it was the Australian submarine's very large size which is what made the pumpjet viable, where it might not be for smaller submarines.
 I responded directly to Clark's comments arguing that the size necessitated by the extra fuel and batteries a pumpjet demanded could hardly be counted as an advantage \parencite{morrison2018b}.
-However, Dr Kershaw's argument reveals a different takon on why size is important.
+However, Dr Kershaw's argument reveals a different take on on why size is important.
 His comments make it clear that it is actually the large weight of the pumpjet \textit{itself} rather than the additional fuel and batteries it required, that was significant.
@@
 This argument implies that a smaller submarine couldn't adequately provide sufficient buoyancy or balance to support the heavy pumpjet.
-If the submarine couldn't provide such buoyancy or balance, the consequences would be catastophic failure, specifically the submarine would tend to either tail-stand, or sink.
+If the submarine couldn't provide such buoyancy or balance, the consequences would be catastrophic failure, specifically the submarine would tend to either tail-stand, or sink.
 It would be unthinkable for a competent manufacturer, conscious of the weight of the pumpjet and such risks, to neglect to account for such risks.
@@
 There are obvious reasons why the width/height (which must be equal, since the jet is cylindrical/conical) ought to scale in proportion to the diameter of the hull, in order to engage a proportionally larger water-column to produce the thrust necessary.
-Similarly, the length  of the shround is likely to remain relatively fixed in proportion to the diameter for a given performance requirement.
+Similarly, the length  of the shroud is likely to remain relatively fixed in proportion to the diameter for a given performance requirement.
 It is unlikely that the proportional thickness of the blades or shroud could be greatly diminished at larger scales without structural integrity and hydrodynamic qualities of the shroud and rotor blades being compromised.
@@
 
-In a clear parallel with the argument that the performance of propulsion systems on submerged bodies is independed of speed because they don't face wave-making resistance, this line of reasoning involves the presentation of a seemingly relevant, comprehensible technical truth which is used to justify a completely blatant technical falsehood.
+In a clear parallel with the argument that the performance of propulsion systems on submerged bodies is independent of speed because they don't face wave-making resistance, this line of reasoning involves the presentation of a seemingly relevant, comprehensible technical truth which is used to justify a completely blatant technical falsehood.
 Even if Dr Kershaw is not familiar with this area of science and speaking outside his competence, it is difficult to see how such a pattern of argument would occur so consistently by accident.
@@
 For him to casually or accidentally make such careless arguments also seems implausibly reckless.
-An alternative explanation is that the results of the research already conducted by the government broadly support the conclusions of my own research, and this is well known by Dr Kershaw, and  that these unfortunate statements are being made in a desparate attempt to deflect attention in order to avoid admitting this fact.
-
-\begin{tcolorbox}[width=0.9\textwidth, center, colback=green!5!white,colframe=green!75!black, title= \textbf{How a scientist feels being told that the pumpjet is a gamechanger for our diesel-electric submarine because it is so large, where it wouldn't make sense on a smaller one}]
-
-Bruce: I'm not sure about using these fat nobbly tires on my road-bike for the cycle tour... No-one else is using such tyres on a road-bike, I think they'd be slow and inefficient.
+An alternative explanation is that the results of the research already conducted by the government broadly support the conclusions of my own research, and this is well known by Dr Kershaw, and  that these unfortunate statements are being made in a desperate attempt to deflect attention in order to avoid admitting this fact.
+
+\begin{tcolorbox}[width=0.9\textwidth, center, colback=green!5!white,colframe=green!75!black, title= \textbf{How a scientist feels being told that the pumpjet is a game-changer for our diesel-electric submarine because it is so large, where it wouldn't make sense on a smaller one}]
+
+Bruce: I'm not sure about using these fat knobbly tires on my road-bike for the cycle tour... No-one else is using such tyres on a road-bike, I think they'd be slow and inefficient.
 
 James: Don't be silly!
-Everyone else would love to use fat nobbly tyres on a road-bike, the only reason they can't is that those cyclists themselves are just so slim and athletic, and can't use the nobbly fat nobbly tyes.
-Bruce, don't you realise that you're the \textit{only} cyclist every attempting this tour who weights 200kg, roughly three times what most road-cyclistst weigh?
-If a 70kg rider attempted those speeds with fat knobbly tyres, they wouldn't have the weight to stabilise the forces involved, and they'd be thrown off their bicycle in a catastophic crash which would probably be fatal.
-But for fat cyclists such as yourself, fat nobbly tires are a real game-changer!
+Everyone else would love to use fat knobbly tyres on a road-bike, the only reason they can't is that those cyclists themselves are just so slim and athletic, and can't use the knobbly fat knobbly tyres.
+Bruce, don't you realise that you're the \textit{only} cyclist every attempting this tour who weights 200kg, roughly three times what most road-cyclists weigh?
+If a 70kg rider attempted those speeds with fat knobbly tyres, they wouldn't have the weight to stabilise the forces involved, and they'd be thrown off their bicycle in a catastrophic crash which would probably be fatal.
+But for fat cyclists such as yourself, fat knobbly tires are a real game-changer!
 
 Bruce:  What about that one lonely guy over there riding on fat knobbly tyres.
-He looks like he weighs about 70kg, and he's not having a catastopic crash, he's just going really slow and looks tired.
+He looks like he weighs about 70kg, and he's not having a catastrophic crash, he's just going really slow and looks tired.
 Maybe that's why he's sticking to the footpath.
@@
 
-Bruce: And didn't you also tell me that you tried riding with fat nobbly tires once years ago?  How did you survive the catastophic crash?
+Bruce: And didn't you also tell me that you tried riding with fat knobbly tires once years ago?  How did you survive the catastrophic crash?
 
@@
 
-The argument that the `hull needs to be tuned to pumpjet' and \textit{vice versa} has been used to dismiss critcism of the pumpjet repeatedly by RADM Sammut \parencite{sammut2017} \parencite{bergmann2018}, and also by Dr David Kershaw \parencite{kerr2018}.
+The argument that the `hull needs to be tuned to pumpjet' and \textit{vice versa} has been used to dismiss criticism of the pumpjet repeatedly by RADM Sammut \parencite{sammut2017} \parencite{bergmann2018}, and also by Dr David Kershaw \parencite{kerr2018}.
 It has also been referred to in an answer to a Question on Notice given by Senator Ketter\ref{QON1}.
@@
 This statement appears to be made in almost complete ignorance of the degree to which the hull-propulsor interactions are routinely and comprehensively considered in essentially all propulsor design, including for propellers and pumpjets.
-In almost any design text \parencite{carlton2007}, \parencite{mollard2011}, \parencite{MAN2017}, \parencite{lewis1988}, including those written specifically for submarines \parencite{jackson1992}, it can be seen that the the various efficiency co-efficients for efficiencyare generally amalgamated into a `quasi-propulsive coefficient', called either $QPC$ or $\eta_D$.
+In almost any design text \parencite{carlton2007}, \parencite{mollard2011}, \parencite{MAN2017}, \parencite{lewis1988}, including those written specifically for submarines \parencite{jackson1992}, it can be seen that the the various efficiency co-efficients for efficiency are generally amalgamated into a `quasi-propulsive coefficient', called either $QPC$ or $\eta_D$.
 
@@
 $\eta_R$ is the called the relative rotative efficiency, and describes the degree to which the rotating propulsor is more or less efficient operating in a wake, where the water is has an inconsistent velocity across the disc of the rotor.
-$\eta_H$ describes net effect that the propulsor has on the hull by drawing water around its aft end at different speeds and altering the pressure forces on the hull, accounting for the potential for the propulsor to regain some of the energy inerent in the wake behind the hull.
-Both of these effects can be shown to be most significanly driven by the hull shape and the arrangement of the propeller with respect to the hull (its relative diameter, and proximity), and be relatively constant with speed.
+$\eta_H$ describes net effect that the propulsor has on the hull by drawing water around its aft end at different speeds and altering the pressure forces on the hull, accounting for the potential for the propulsor to regain some of the energy inherent in the wake behind the hull.
+Both of these effects can be shown to be most significantly driven by the hull shape and the arrangement of the propeller with respect to the hull (its relative diameter, and proximity), and be relatively constant with speed.
 As will be shown in this section, they could plausibly only vary a few percent each, which would have no impact in overturning the collapsing open-water efficiency of a pumpjet operating so far outside of design-conditions, such as a pumpjet at very low speed.
@@
 To argue that none of these concepts have any useful bearing or meaning when applied to pumpjets is a bizarre, seemingly childish argument.
-A pumpjet is actually by definition a particuar form of a ducted propeller, specifically one in which the duct is designed to \textit{decelerate} the flow ahead of it the rotor\parencite{oosterveld1970}.
+A pumpjet is actually by definition a particular form of a ducted propeller, specifically one in which the duct is designed to \textit{decelerate} the flow ahead of it the rotor\parencite{oosterveld1970}.
 Essentially all of those same concepts apply, regardless of whether the pumpjet appears occurs on the aft end of a body of revolution, or suspended beneath the hull of a ship.\footnote{It is actually water-jets which are hardest to test in an open-water condition, since their intake is necessarily a tunnel connecting to a hull.}
@@
 
-Bruce: I'd like to buy some light, breatheable shoes to go running with.
-
-James: Here, have these double-padded, thick-soled, waterpoof, steelcapped, all-weather, leather work-boots.
+Bruce: I'd like to buy some light, breathable shoes to go running with.
+
+James: Here, have these double-padded, thick-soled, water-poof, steel-capped, all-weather, leather work-boots.
 
@@
 Sometimes socks can get crinkled when running, leading to discomfort.
-The thickly padded sole in this boot is designed to aleviate this discomfort, which will help you run faster too!
+The thickly padded sole in this boot is designed to alleviate this discomfort, which will help you run faster too!
 
@@
 The area behind the vessel, where the water is being slightly dragged along with the ship (or the water is moving less quickly past the vessel, in the vessel's reference frame) is known as the wake.
-Since any aft-mounted propulsor has to operate in the wake, (or wake-field as it is often called) it is true that wake is of importance to the design of the propolsor.
+Since any aft-mounted propulsor has to operate in the wake, (or wake-field as it is often called) it is true that wake is of importance to the design of the propulsor.
 What is not true, however, is that this is grounds for disregarding open-water efficiency, or that it was neglected from consideration in the preparation of my report.
@@
 The `open-water' condition is can be usefully thought of as the normal or `ideal' efficiency of the propulsor, when the stream of incoming water is undisturbed, and consequently is still a very functional maximum efficiency of a system who's purpose is to smooth out the inconsistency in the flow.
-In the absence of an advantageous pre-swirl, no higher efficiency could be reasonably expected to be achieved through such smoothing than would be acheived by beginning with a smooth stream in the first place.
+In the absence of an advantageous pre-swirl, no higher efficiency could be reasonably expected to be achieved through such smoothing than would be achieved by beginning with a smooth stream in the first place.
 Consequently, one can have considerable confidence that the extent to which this `smoothing' effect might assist a pumpjet, relative to a propeller, is bounded by the extent to which it actually impedes a propellers performance.
@@
 \includegraphics[width=0.4\textwidth, center]{roteff3.png}
-\caption{One text's desciption of the various types of efficiency from page 14 of \parencite{MAN2017}}
+\caption{One text's description of the various types of efficiency from page 14 of \parencite{MAN2017}}
 \label{fig:roteff3.png}
@@
 The first two result in increase in the drag experienced by the hull due to the effect of the propulsor drawing water more quickly across the aft section of the hull than would naturally occur if the vessel were under tow.
-The first component is the increase in frictional resistance, and the second an increase in pressure resistance according to Bernouli (which requires that as speed increases, pressure reduces, sucking the ship backwards).
+The first component is the increase in frictional resistance, and the second an increase in pressure resistance according to Bernoulli (which requires that as speed increases, pressure reduces, sucking the ship backwards).
 The third is that this pressure reduction may reduce the degree of flow separation that occurs on the aft section of the hull created by the adverse pressure gradient that occurs as the water flow slows down.
@@
 
-The wake fraction represents the average proportal difference between the speed of water that would flow into the propulsor (due to the wake unmodified by the propulsors operation), relative to the vessel speed.
-Since water being dragged along with the ship represents pure energy waste, the extent to which the propulsor can gain thrust by accelerating such water back to stationary state (relative to the ocean clear of the vessel) rather than re-accelating it to a new backwards velocity represents the effective re-capture of energy that was previously lost to the wake.
+The wake fraction represents the average proportional difference between the speed of water that would flow into the propulsor (due to the wake unmodified by the propulsors operation), relative to the vessel speed.
+Since water being dragged along with the ship represents pure energy waste, the extent to which the propulsor can gain thrust by accelerating such water back to stationary state (relative to the ocean clear of the vessel) rather than re-accelerating it to a new backwards velocity represents the effective re-capture of energy that was previously lost to the wake.
 
@@
 Countering that effect, the propulsor would be attempting to suck water directly off the blunt aft end section, which would create a significant backwards suck retarding the advance of the hull, increasing the thrust deduction and reducing hull efficiency.
-In contrast, a vessel with a long, gradualy tapering aft and a propulor extended well away from the hull would have a very low wake fraction, but also a very small thrust deduction.
-Those two contrasting vessels will also necessarily require different amounts of thrust to be moved through the water, owing to their contrasting shapes, which (perhaps counterintuitively) is not actually reflected in hull efficiency $\eta_H$, which purely reflects the interaction between the propulsor and the hull.
+In contrast, a vessel with a long, gradually tapering aft and a propulsor extended well away from the hull would have a very low wake fraction, but also a very small thrust deduction.
+Those two contrasting vessels will also necessarily require different amounts of thrust to be moved through the water, owing to their contrasting shapes, which (perhaps counter-intuitively) is not actually reflected in hull efficiency $\eta_H$, which purely reflects the interaction between the propulsor and the hull.
 
@@
 A third plot, reproduced here in Figure \ref{fig:hulleff.png} demonstrates how these two values combine to produce the hull efficiency.
-As can be clearly seen, the size of the propeller relative to the hull also has an impact, and alters the hull efficiency by approximately 15\% when the propulsor size is halved or doubled.\footnote{The propulor size also impacts open water efficiency, so an global optimum may be constrained away from the hull-efficiency optimum.}  Given that one of the key features of the pumpjet is that they are smaller in total diameter to a comparable propeller, (and hence accelerate a smaller water-column to greater speeds) this provides a useful indication of the likley degree of variance that might be expected.
+As can be clearly seen, the size of the propeller relative to the hull also has an impact, and alters the hull efficiency by approximately 15\% when the propulsor size is halved or doubled.\footnote{The propulsor size also impacts open water efficiency, so an global optimum may be constrained away from the hull-efficiency optimum.}  Given that one of the key features of the pumpjet is that they are smaller in total diameter to a comparable propeller, (and hence accelerate a smaller water-column to greater speeds) this provides a useful indication of the likely degree of variance that might be expected.
 However, the size differences are unlikely to be nearly as large as a factor of two.
@@
 \end{subfigure}
-\caption{Dramatic changes in propeller size on a submarine (doubling/halving) would only produce variance of appoximately 15\% in hull efficiency, according to Reference \parencite{jackson1992}.
+\caption{Dramatic changes in propeller size on a submarine (doubling/halving) would only produce variance of approximately 15\% in hull efficiency, according to Reference \parencite{jackson1992}.
 A pumpjet, which be smaller in diameter than a propeller by about \%20, would likely only make a few percentage points of difference due to changed flow over the hull ahead of it.}
@@
 $J$ is known as the advance ratio, and can intuitively be thought of as how much the propulsor moves forward through the water that it's operating in per a given degree of rotation.
-Since the resistance that water exerts on a body \textit{does} change with speed (it scales with $V^2$), contrary to defence's claims, $J$ actually has a counter-intuitive relationship with speed.
+Since the resistance that water exerts on a body \textit{does} change with speed (it scales with $V^2$), contrary to Defence's claims, $J$ actually has a counter-intuitive relationship with speed.
 Low values of $J$ correspond to conditions where a given vessel is actually moving very fast, since the high resistance encountered means that it must actually turn around \textit{more} times, relative to the distance it advances.
-It is actually at some high value of $J$ that corresponds to a value of extremely slow waterspeed, approaching stationarity.
-In such as cases, since the resistance is extremely low, the propeller actually advances the maximum distance that it can through the water for a given revolution, akin to a corkscrew advancing through a cork, where there is no `slippage' throught the water.
+It is actually at some high value of $J$ that corresponds to a value of extremely slow water-speed, approaching stationarity.
+In such as cases, since the resistance is extremely low, the propeller actually advances the maximum distance that it can through the water for a given revolution, akin to a corkscrew advancing through a cork, where there is no `slippage' through the water.
 As can be seen from the shape of the curves shown for an open propeller in Figure \ref{fig:OpenWater.png}, efficiency falls off to zero at such a point where $J$ is equal to the $P/D$ label corresponding to each line.
@@
 Two other lines are generally present in these graphs, which are the lines which indicate $K_T$, which represent the thrust coefficient, and $K_Q$, which represents the torque coefficient.
-As can be seen in Figure \ref{fig:OpenWater.png}, for any given propeller geometry ($P/D$ value) the thrust coeffient $K_T$ reaches zero at the same point that the efficiency $\eta_0$ does at a high $J$ value.
-This corresponds to zero waterspeed, when no effective forward thust is produced by the propulsor.
+As can be seen in Figure \ref{fig:OpenWater.png}, for any given propeller geometry ($P/D$ value) the thrust coefficient $K_T$ reaches zero at the same point that the efficiency $\eta_0$ does at a high $J$ value.
+This corresponds to zero water-speed, when no effective forward thrust is produced by the propulsor.
 It will be noted at this point that the lines corresponding to $K_Q$ have not quite reached zero.
@@
 As such, there is no directly relationship between $J$, $K_Q$ or $K_T$ which can be used to determine a corresponding united measurements (e.g.
-waterspeed, rpm, power etc) unless \textit{all} of the necessary assumptions are made and equations solved, and similarly fractions or ratios of coefficients hold to their conceptual analogues.
+water-speed, rpm, power etc) unless \textit{all} of the necessary assumptions are made and equations solved, and similarly fractions or ratios of coefficients hold to their conceptual analogues.
 ($K_T$ of 0.5 is not half a the thrust of a $K_T$ of 1.0).
 However, at the extremities and boundaries one may usefully describe what is happening.
-Important to note is that all efficiencies approach zero at $J = 0$, since this describes the egg-beater situation, where the resistance is so great that no movement can be achieved (and hence no `work' done), such as the 'bollard-pull' condition of attempting to move an immoveable object.
+Important to note is that all efficiencies approach zero at $J = 0$, since this describes the egg-beater situation, where the resistance is so great that no movement can be achieved (and hence no `work' done), such as the 'bollard-pull' condition of attempting to move an immovable object.
 Thrust and torque are non-zero, but the failure to advance any further means that efficiency is zero.
-Open-water efficiency measures the ability for a propulsor to convert rotational force into linear thust to achieve the greatest movement in that linear direction.
+Open-water efficiency measures the ability for a propulsor to convert rotational force into linear thrust to achieve the greatest movement in that linear direction.
 Without movement, this doesn't occur.
@@
 This directly corresponds to the impact that the duct has on the water speed that arrives at the \textit{rotor}, versus the water speed that arrives at the entry to the duct of the propulsor.
-Since the rotor in the accelerating duct is actually pushing through water moving \textit{faster} than the water arriving at the mouth of the duct, this condition where forwards thrust (and waterspeed) fall to zero occurs when the propulsor is actually advancing through less water per revolution of the rotor.
+Since the rotor in the accelerating duct is actually pushing through water moving \textit{faster} than the water arriving at the mouth of the duct, this condition where forwards thrust (and water-speed) fall to zero occurs when the propulsor is actually advancing through less water per revolution of the rotor.
 In the decelerating case, the opposite happens.
-Since the water is slowed down before the rotor acts on it, the propulsor as a whole is actually advancing even further through the water than the `corkscrew' condition would apply as assessd by the revolutions of the rotor.
+Since the water is slowed down before the rotor acts on it, the propulsor as a whole is actually advancing even further through the water than the `corkscrew' condition would apply as assessed by the revolutions of the rotor.
 This is a general feature of the contrasting nature of decelerating and accelerating ducts.
@@
 
-Naturally we can assume that these particular shrouds and propellers don't exactly represent those to be used on a future submarine, but some of the most significant features are reliably illustrative of the general contrasts between unshrouded rotors, and the propulsors resulting from shrouding them in  decelerating and decelerating ducts.
+Naturally we can assume that these particular shrouds and propellers don't exactly represent those to be used on a future submarine, but some of the most significant features are reliably illustrative of the general contrasts between un-shrouded rotors, and the propulsors resulting from shrouding them in  decelerating and decelerating ducts.
 It might well be noted that both of the ducted propellers have slightly lower peak efficiency than the open one.
-The shape of the efficiency curve for the accelerating case is broadly similar to the unshrouded version, whereas the decelerating curve has a clearly elongated, flattened top, being able to undertake a large increase in loading without detrimental impacts on the performance of the propulsor, largely due to the elevated pressure alleviating the increasing occurrence of cavitation.
+The shape of the efficiency curve for the accelerating case is broadly similar to the un-shrouded version, whereas the decelerating curve has a clearly elongated, flattened top, being able to undertake a large increase in loading without detrimental impacts on the performance of the propulsor, largely due to the elevated pressure alleviating the increasing occurrence of cavitation.
 
 However, the most significant difference between the two different curves is the addition of a fourth set of lines at the bottom of the graph, denoted $K_{TD}$.
-These represent the thust that is produced by the duct itself.
+These represent the thrust that is produced by the duct itself.
 As defined in Reference \parencite{oosterveld1970} the $\eta_0$ line represents only the effect of the thrust generated by the rotor.
@@
 This has the effect of substantially `arching' the back of a whole-system efficiency curve, allowing relatively good efficiencies to be achieved at very high loadings, or low $J$ values.
-For any given propeller curve it can be seen that at some high $J$ value, corresponding to low loading (or low speed, for a given propulsor geometry and vessel) the shroud still has a net negative impact, as the additional drag overwhelmes the benefits achieved by the distribution of pressure across a larger surface area.
+For any given propeller curve it can be seen that at some high $J$ value, corresponding to low loading (or low speed, for a given propulsor geometry and vessel) the shroud still has a net negative impact, as the additional drag overwhelms the benefits achieved by the distribution of pressure across a larger surface area.
 
@@
 If this effect were to be integrated into a whole-system efficiency curve, this would have the effect of significantly crimping the brow of the curve on the right hand side.
-This would mean that as the a vessel slowed and the propulsor unloaded (increased $J$), instead of climbing ever-higher until falling sharply only at extremely low speeds the when the corkscrew condition was almost met (as in the case for an open propeller) efficiency would begin to decline sooner, from non-trival speeds.
+This would mean that as the a vessel slowed and the propulsor unloaded (increased $J$), instead of climbing ever-higher until falling sharply only at extremely low speeds the when the corkscrew condition was almost met (as in the case for an open propeller) efficiency would begin to decline sooner, from non-trivial speeds.
 This is the impact that a decelerating shroud has at very low speeds: when it attempts to decelerate a flow that is already going very slowly, and the imposed resistance - desirably at high speeds in preventing cavitation - disrupts the flow excessively and wastes energy creating vortices and turbulence.
@@
 \includegraphics[width=0.9\linewidth]{nozz19b.png}
-\caption{Open-water test results for an accelerating duct illustrate a similar $\eta_0$ as an open propeller, additional thust is derived from the duct itself $K_{TD}$ at low $J$ values, enhancing efficiency at high speeds.
+\caption{Open-water test results for an accelerating duct illustrate a similar $\eta_0$ as an open propeller, additional thrust is derived from the duct itself $K_{TD}$ at low $J$ values, enhancing efficiency at high speeds.
 Images as shown in Reference \parencite{oosterveld1970}.}
@@
 \includegraphics[width=0.9\linewidth]{nozz32b.png}
-\caption{Open-water test results for a decelerating duct exihibits a flatter top for the rotor's efficiency, owing to reduced cavitation as speed increases.
+\caption{Open-water test results for a decelerating duct exhibits a flatter top for the rotor's efficiency, owing to reduced cavitation as speed increases.
 However, the duct imposes a substantial negative thrust at higher $J$ values, degrading efficiency at lower speeds.
@@
 Perhaps I had falsely assumed that as much would immediately clear to any technically competent reviewer, who would readily grasp that open-water efficiency is the primary parameter which determines the efficiency of a propulsor as it pertains to the loading of the propulsor, and consequently speed.
-The use of this argument displays a substantial degree of ignorance of the relevant physics, as clearly described in numerous and publically available sources.
+The use of this argument displays a substantial degree of ignorance of the relevant physics, as clearly described in numerous and publicly available sources.
 
@@
 For the majority of the surface consideration of a submarine, the boundary layer is actually turbulent, which means that water in the boundary layer is tumbling and turning in a confused and unpredictable pattern.
-Whilst \textit{on average} the boundary layer moves closer to the surface speed closer to it, and slower further away from it, at any local point on a tiny lengthscale the movement of the water will be chaotic, and altogether very non-parallel.
-Turbulent boundary layers tend to be much thicker, and become even thicker at higher velocities, resulting in larger amounts of energy being sapped producing the faster, more intensely chaotic swirling motions that comprise turblent flows.
+Whilst \textit{on average} the boundary layer moves closer to the surface speed closer to it, and slower further away from it, at any local point on a tiny length-scale the movement of the water will be chaotic, and altogether very non-parallel.
+Turbulent boundary layers tend to be much thicker, and become even thicker at higher velocities, resulting in larger amounts of energy being sapped producing the faster, more intensely chaotic swirling motions that comprise turbulent flows.
 
 This distinction is important in understanding the limitations of a pumpjet's claim to make the incoming stream of water less inconsistent, or more regular.
-The duct intake and any stator blades in advance of the rotor can be effective in making the flow more homogenous as assessed on a macroscopic scale, but in doing so they simultaneously make the flow more turbulent at a smaller scale, since it necessarily involve passing the water over more surfaces, rather than less.
-This can be advantageous to the extent that very large-scale and intense vortices might be broken down into smaller-scale vortices, which appears more homogenous when assessed on the length-scale relevant to the blades of the rotors.
+The duct intake and any stator blades in advance of the rotor can be effective in making the flow more homogeneous as assessed on a macroscopic scale, but in doing so they simultaneously make the flow more turbulent at a smaller scale, since it necessarily involve passing the water over more surfaces, rather than less.
+This can be advantageous to the extent that very large-scale and intense vortices might be broken down into smaller-scale vortices, which appears more homogeneous when assessed on the length-scale relevant to the blades of the rotors.
 This could indeed have the effect of reducing the likelihood of very large pressure fluctuations impacting the blades, reducing the probability of cavitation occurring or invoking some kind of resonant beats within the propulsor.
@@
 
-However, it must be noted that the in making the flow more homogenous, the flow is also made more homogenously turbulent, and the total level of energy which is wasted to small-scale chaotic turbulent movements necessarily increases, as it necessarily must to with any additional surface.
+However, it must be noted that the in making the flow more homogeneous, the flow is also made more homogeneously turbulent, and the total level of energy which is wasted to small-scale chaotic turbulent movements necessarily increases, as it necessarily must to with any additional surface.
 The circumstances where a pumpjet is able to attain higher efficiency are those where the total amount of additional productive energy (thrust, produced by front to back acceleration) added to the water column dramatically exceeds the losses incurred by the additional turbulence (round and round acceleration).
-This occurs at high speeds, which necessitate a high level of turublence which a propeller is unable to avoid.
-
-It should be noted that a pumpjet's duct also has the effect of decelerating the incoming flow, which increases its static pressure.\footnote{In fact, a pumpjet is by definition a decelerating ducted propeller, contrary to claim that they are so different as to be incomparable made \ref{QON1}.} Whilst this has the desirable effect of further reducing the likelyhood of cavitation occurring, it also creates an `adverse pressure gradient' in the intake.
+This occurs at high speeds, which necessitate a high level of turbulence which a propeller is unable to avoid.
+
+It should be noted that a pumpjet's duct also has the effect of decelerating the incoming flow, which increases its static pressure.\footnote{In fact, a pumpjet is by definition a decelerating ducted propeller, contrary to claim that they are so different as to be incomparable made \ref{QON1}.} Whilst this has the desirable effect of further reducing the likelihood of cavitation occurring, it also creates an `adverse pressure gradient' in the intake.
 This means that the water must move towards an area of higher pressure, against the direction that the pressure would naturally push it.
 Whilst at high speeds the water has enough momentum to easily overcome this resistance, at very low speeds there isn't enough energy in the flow for the entire water column to move consistently in this direction, and part of the flow simply turns around and moves in the opposite direction, in a phenomenon known as `flow separation'.
-This occurs ajacent to surface (in this case just inside the pumpjet intake), where the boundary layer has to move more slowly than the remainder of the stream, and in  this case it is the first part to reverse direction.
-Whilst the entirety of the boundary layer, (or indeed the entire flow) may actually be turbulent, with many very small-scale vortices, this phenomenon creates new large-scale macroscopic vortices, which could form, grow and shed in an unsteady creating exactly the same risk of perodic shedding (i.e.
+This occurs adjacent to surface (in this case just inside the pumpjet intake), where the boundary layer has to move more slowly than the remainder of the stream, and in  this case it is the first part to reverse direction.
+Whilst the entirety of the boundary layer, (or indeed the entire flow) may actually be turbulent, with many very small-scale vortices, this phenomenon creates new large-scale macroscopic vortices, which could form, grow and shed in an unsteady creating exactly the same risk of periodic shedding (i.e.
 beats) impacting the rotor that a poorly designed mast or appendage might have.\footnote{A cylindrical mast, for instance, could set up a Karman Vortex Street at certain speeds.} Whilst it might be presumed the shedding of vortices created by flow separation into the pumpjet at low speeds has been adequately managed or mitigated to the extent that it does not risk the creation of cavitation on the blades, it would most certainly lead to the very substantial loss of efficiency through creation of these vortices.
@@
 
-\caption{(a) At high speeds, the intake of a pumpjet slows the waterflow and raises its pressure, and it subsequently accelerates through the contracting nozzle to form an efficient jet (b) At low speeds, the pressure gradient created by the duct causes some of the waterflow to stop and reverse.
+\caption{(a) At high speeds, the intake of a pumpjet slows the water-flow and raises its pressure, and it subsequently accelerates through the contracting nozzle to form an efficient jet (b) At low speeds, the pressure gradient created by the duct causes some of the water-flow to stop and reverse.
 This `flow separation' wastes energy spinning water in circles, and dramatically reduces the system's efficiency.}
@@
 Having considered in some detail the nature of the three substantive arguments given in opposition to my findings, it is clear that none actually provide any credible evidence to oppose my conclusions.
-However, a clear similarit or parallel between the three can be found.
+However, a clear similarity or parallel between the three can be found.
 Each of them respectively actually comprise an argument against an obvious reference point being used to assess the efficiency of the pumpjet.
@@
 
-It is indeed remarkeable that none of the arguments made actually involve pointing to a better reference point than what I have been able to use from open-source literature, or announce that any results from Naval Group directly contradict the findings of my earlier report.
+It is indeed remarkable that none of the arguments made actually involve pointing to a better reference point than what I have been able to use from open-source literature, or announce that any results from Naval Group directly contradict the findings of my earlier report.
 It is a relatively simple proposition that Naval Group, as one of only two countries in the world to have fitted both pumpjets and propellers to both diesel and nuclear submarines, has in its possession the best data possible to serve as reference-points for our submarine.
@@
 
-As I have argued, I think that there actually are adquate reference points in open-source literature regarding comparable propulsion systems, designed for comparable purposes, employed in relevant and comparable, in order to gain what is really only a very broad, but clear conlcusion: that propulsive efficiency declines inevitably and dramatically at low speeds.
-
-Taking it as given that the French do have posession of adquate of far more detailed performance data, derived from even closer reference points than what I can glean from public literature, one must ask whether that data actually confirms, or contradicts my findings.
-Given public claims from DCNS during the CEP that information regarding both the hydrodynmic and acoustic performance of the pumpjet were in Australian government hands in Australia \parencite{costello2016} there can be little doubt Australian officials have access to such information.
-Indeed, it would be an inexcuseable oversight if they didn't obtain that information during the CEP.
-
-This poses a seemingly irreconcileable tension between the desparate, implausible arguments currently being made by Defence, and the possibility that they are being made because Defence genuinely doesn't know better, and genuinely believes them to be true.
-It seems far more probable that Defence actually is in posession of clear, conclusive evidence which broadly supports the conclusions of my report, and is now deliberately concealing that evidence, and attempting to excuse why they have not drawn the obvious conclusions that any sensible person normally would from that evidence - that a pumpjet is entirely unsuited to a diesel-electric submarine.
+As I have argued, I think that there actually are adequate reference points in open-source literature regarding comparable propulsion systems, designed for comparable purposes, employed in relevant and comparable, in order to gain what is really only a very broad, but clear conclusion: that propulsive efficiency declines inevitably and dramatically at low speeds.
+
+Taking it as given that the French do have possession of adequate of far more detailed performance data, derived from even closer reference points than what I can glean from public literature, one must ask whether that data actually confirms, or contradicts my findings.
+Given public claims from DCNS during the CEP that information regarding both the hydrodynamic and acoustic performance of the pumpjet were in Australian government hands in Australia \parencite{costello2016} there can be little doubt Australian officials have access to such information.
+Indeed, it would be an inexcusable oversight if they didn't obtain that information during the CEP.
+
+This poses a seemingly irreconcilable tension between the desperate, implausible arguments currently being made by Defence, and the possibility that they are being made because Defence genuinely doesn't know better, and genuinely believes them to be true.
+It seems far more probable that Defence actually is in possession of clear, conclusive evidence which broadly supports the conclusions of my report, and is now deliberately concealing that evidence, and attempting to excuse why they have not drawn the obvious conclusions that any sensible person normally would from that evidence - that a pumpjet is entirely unsuited to a diesel-electric submarine.
 
@@
 
-\textbf{Rear Adm. Sammut:}   We have selected Naval Group as competent designers of dieselelectric
+\textbf{Rear Adm. Sammut:}   We have selected Naval Group as competent designers of diesel electric
 submarines and indeed ocean-going submarines. We made that assessment on the
@@
 ``DST though, is in the position where we can access certain privileged information, so we can get validation of our analysis capabilities," Dr Kershaw noted.
-Pumpjet design had advanced over many years with a particular forcus on the characteristics of all the components, although much of that was classified.
+Pumpjet design had advanced over many years with a particular focus on the characteristics of all the components, although much of that was classified.
 Some propellers were highly specialised and highly complex. Pump jets had complexities in other aspects but the moving rotor was not as complex as a normal propeller, benefitting sustainment.
@@
 So will pump jet technology definitely be used for the Future Submarine?
-``That is what is currently in the design  and that's what a lot of our science and technology program is about; having the capability to do the independent analysis on the technologyies and capabilities planned for the Future Submarine, including the pump jet program," Dr Kershaw said to \textit{ADM}.
+``That is what is currently in the design  and that's what a lot of our science and technology program is about; having the capability to do the independent analysis on the technologies and capabilities planned for the Future Submarine, including the pump jet program," Dr Kershaw said to \textit{ADM}.
 Discussions were currently underway with Naval Group and the French government regarding closer collaboration on a range of technologies and this would include pump jets, Dr Kershaw disclosed.

### 2018-09-26 Second.Rnw
Final Draft

@@
 However, over a number of years now the pumpjet choice has attracted sustained attention and criticism, mostly around its efficiency at low speeds, where a diesel-electric submarine would necessarily spend a great deal of its operating time.
-I was commissioned by Gary Johnston of Submarines for Australia to investigate this topic in detail and produce a report \parencite{morrison2018}, which concluded that a very substantial reduction in dived range and endurance (as well as overall range) would necessarily result from the pumpjet choice.
+I was commissioned by Gary Johnston of Submarines for Australia to investigate this topic in detail and produce a report\parencite{morrison2018}, which concluded that a very substantial reduction in dived range and endurance (as well as overall range) would necessarily result from the pumpjet choice.
 Furthermore, that paper challenged the assumption that the pumpjet would be acoustically superior at low speeds, where open propellers would have no cavitation, and cause much less turbulence than the equivalent jet.
 
-That paper, and a subsequent blog post \parencite{morrison2018a} on the topic which made the paper publicly available, attracted the interest of the Senate Economics Reference Committee Inquiry into the Future of Naval Shipbuilding in Australia.
+That paper, and a subsequent blog post\parencite{morrison2018a} on the topic which made the paper publicly available, attracted the interest of the Senate Economics Reference Committee Inquiry into the Future of Naval Shipbuilding in Australia.
 Senator Chris Ketter asked written questions on notice to the Department of Defence requesting responses to the paper specifically, and related comments from public officials, with a return date of 3rd April 2018.
 I was also invited to provide evidence to the committee in a public hearing on the 7th of June 2018.
-In that public hearing \ref{SENmorrison}, I stood by the conclusion of my substantive paper, which was: ``between two otherwise identical submarines, the one with the pump jet will always have a lower dived endurance, a lower dived range, a worse indiscretion ratio, a lower overall endurance and a lower overall range than the one with the propeller."  In the presence of both the outgoing acting CEO of Naval Group Australia Brent Clark, and General Manager of Submarines (for Defence) Stephen Johnson, I pointed out that that statement was yet to be challenged, and that I'd be very surprised if anyone with technical competence or professional integrity sought to challenge it.
+In that public hearing[\ref{SENmorrison}] I stood by the conclusion of my substantive paper, which was: ``between two otherwise identical submarines, the one with the pump jet will always have a lower dived endurance, a lower dived range, a worse indiscretion ratio, a lower overall endurance and a lower overall range than the one with the propeller."  In the presence of both the outgoing acting CEO of Naval Group Australia Brent Clark, and General Manager of Submarines (for Defence) Stephen Johnson, I pointed out that that statement was yet to be challenged, and that I'd be very surprised if anyone with technical competence or professional integrity sought to challenge it.
 
@@
 Senator Kim Carr requested that defence produce a written response to that report for the committee prior to their handing down of their final report for the Inquiry, and Defence agreed to do their best in the time available, with the head of Capability and Sustainment Group (CASG) Kim Gillis assuring the committee that ``We  will  provide  the  committee  with  the  best  possible  advice  we can  in  the  shortest  possible  time  frame  to  meet  your  requirements  so  that  you  can  be  better informed, because we do hold that expertise." Senator Carr emphasised the importance of timeliness that ``it would look pretty bloody stupid after the report's handed in''.
-Despite these exhortations, in the final report of the committee this request, and Defence's acceptance of it was noted, but no acknowledgement was made of whether any response was actually received, or what it contained \parencite{SERC2018final}.
-
-In the August edition of Australian Defence Magazine a feature article on the titled `Future Submarine: Pumpjet propulsor technology choice still not well explained' \parencite{kerr2018} appeared, referencing both my testimony in the hearing and the paper.
+Despite these exhortations, in the final report of the committee this request, and Defence's acceptance of it was noted, but no acknowledgement was made of whether any response was actually received, or what it contained\parencite{SERC2018final}.
+
+In the August edition of Australian Defence Magazine a feature article titled `Future Submarine: Pumpjet propulsor technology choice still not well explained'\parencite{kerr2018} appeared, referencing both my testimony in the hearing and the paper.
 Comment for the article was sought from CASG, the Royal Australian Navy (RAN) and Naval Group, and none was forthcoming.
@@
 This assumption was mentioned already by Dr Kershaw in the ADM article, but it was so completely at odds with the most basic physics that I was inclined to dismiss it as perhaps a clumsily worded statement made in an interview compounded by a little journalistic mis-quotation, rather than a substantive argument.
-The re-appearance of this argument, which was spelled out and repeated multiple times in a written document that was prepared by Defence over the course of three months, well after any nominal deadline for timeliness had elapsed, dispelled all illusions that the argument was included by accident.
-As such, I have also addressed it here as the first of three substantial arguments made against my earlier conclusions regarding pumpjet efficiency, which has quite significantly altered the conclusions of this paper, which seeks to consider all three lines of reasoning in sufficient detail to suitably discuss the circumstances in which such arguments might be made.
+The re-appearance of this argument, which was spelled out and repeated multiple times in a written document that was prepared by Defence over the course of three months (well after any nominal deadline for timeliness had elapsed) dispelled all illusions that the argument was included by accident.
+As such, I have also addressed it here as one of three substantial arguments made against my earlier conclusions regarding pumpjet efficiency.
+This third argument coming from a document so cafeully and deliberately prepared has had a material impact on the strength of the conclusions of this paper, which seeks to consider all three lines of reasoning in sufficient detail to suitably discuss the circumstances in which such arguments might be made.
 
@@
 
+\newpage
 \section{A brief chronology of comment}
@@
 \begin{tcolorbox}[width=0.9\textwidth, center, colback=red!5!white,colframe=red!75!black, title= \textbf{28 May 2012 - Senate Estimates}]
-\textit{Senator Johnston directly challenges Rear Admiral Justin Moffit (then director of SEA 1000) about the pumpjet proposal.  Full exchange in \ref{SENMoffit}. Emphasis added.}
+\textit{Senator Johnston directly challenges Rear Admiral Justin Moffit (then director of SEA 1000) about the pumpjet proposal.  Full exchange in \ref{SENMoffit}.}
 \\
@@
 \begin{tcolorbox}[width=0.9\textwidth, center, colback=blue!5!white,colframe=blue!75!black, title= \textbf{10 Feb 2016 - Sean Costello, CEO DCNS Australia}]
-\textit{In a blog post\parencite{costello2016} for the ASPI Strategist prior to the announcement in April, Sean Costello describes the significance of the pump jet technology to their bid.
-Emphasis added.}
+\textit{In a blog post\parencite{costello2016} for the ASPI Strategist prior to the announcement in April, Sean Costello describes the significance of the pump jet technology to their bid.}
 \\
@@
 Technology born from the French SSBN program a generation ago.
-The stealth and hydrodynamic performances of pump jet propulsion are of course classified and in Australia known only to DCNS and the Australian Government."
+\textbf{The stealth and hydrodynamic performances of pump jet propulsion are of course classified and in Australia known only to DCNS and the Australian Government."}
 
@@
 This rules out the likelihood that some special or unique pumpjet technology, better suited just to diesel-electric submarines, had been pioneered by France alone.
-This comment also directly implies that information regarding the performance of the pumpjet had already been given to the Australian  government, \textit{prior} to the conclusion of the Competitive Evaluation Process (CEP).
+This comment also directly implies that information regarding the performance of the pumpjet, both its acoustic performance and propulsive performance, had already been given to the Australian  government, \textit{prior} to the conclusion of the Competitive Evaluation Process (CEP).
 
 \begin{tcolorbox}[width=0.9\textwidth, center, colback=blue!5!white,colframe=blue!75!black, title= \textbf{8 Apr 2016 - Sean Costello, CEO DCNS Australia, and Gerard Autret, Chief Naval Architect of the Shortfin Barracuda}]
-\textit{In a blog post\parencite{autret2016} for the ASPI Strategist prior to the announcement in April, Sean Costello describes a number of the major features and design approaches for the Shortfin Barracuda, including relating the choice of the pumpjet closely to an acoustic requirement to close with another submarine at higher speed.}
+\textit{In a blog post\parencite{autret2016} for the ASPI Strategist prior to the announcement in April, Sean Costello and Gerard Autret describe a number of the major features and design approaches for the Shortfin Barracuda, including relating the choice of the pumpjet closely to an acoustic requirement to close with another submarine at higher speed.}
 \\
@@
 This comment provides clear context for the importance of the pump-jet.
-Consistent with the physics outlined in Reference \parencite{morrison2018} the acoustic advantage is only emphasised by DCNS as being significant at higher speeds.
+Consistent with the physics outlined in my earlier paper \parencite{morrison2018} the acoustic advantage is only emphasised by DCNS as being significant at higher speeds.
 Whether it is viable for a diesel-electric submarine (as opposed to the nuclear one mentioned here) to maintain such speeds to close with many types of targets is left unclear, or at least unmentioned.
-As later discussed in my evidence to the Senate \ref{SENmorrison}, this is consistent with DCNS being well aware with the advantages and disadvantages of the pumpjet, whilst not making verifiably false claims.
+As later discussed in my evidence to the Senate[\ref{SENmorrison}], this is consistent with DCNS being well aware with the advantages and disadvantages of the pumpjet, whilst not making verifiably false claims.
 
@@
 The official explanation of the designer and builder's website makes a carefully worded explanation of the superiority of the pumpjet, by limiting the comparison to a ``tactical" encounter with an ``otherwise identical" submarine.
-As discussed in \ref{SENmorrison}, this is the only \textit{possible} target that an elevated speed - such as could be achieved by a diesel-electric submarine - might assist in closing with.
+As discussed in[\ref{SENmorrison}], this is the only \textit{possible} target that an elevated speed - such as could be achieved by a diesel-electric submarine - might assist in closing with.
 The submarine would be at a substantial `strategic' disadvantage in such an encounter, since its lower dived endurance, dived range, and poorer indiscretion ratio dramatically increases the probability that the pumpjet propelled submarine would wind up being the hunted, rather than the hunter.
@@
 
+Read literally, the comparison with `obsolete' propeller technology implies that the pumpjet is only quieter than obsolete propellers, rather than so much more quiet that all propellers that propellers in general are obsolete.
+This is consistent with my earlier findings.
+This is also consistent with there being care taken not to make any false claims, whilst still giving a very confident first-impression.
+
 \subsection{RADM Gregory Sammut, Head of SEA 1000 before the Senate}
@@
 
-\textbf{Rear Adm. Sammut:}  DCNS has proposed to us a conventionally-powered submarine that includes—I think you are trying to get to the form of propeller we are talking about here, Senator.
+\textbf{Rear Adm. Sammut:}  DCNS has proposed to us a conventionally-powered submarine that includes-I think you are trying to get to the form of propeller we are talking about here, Senator.
 
@@
 I am inviting you to comment.
-The two parts work beautifully together, but the conventional power that we plan to use does not create as successful a signature at the lower speeds—so above 20 knots versus below 20 knots.
+The two parts work beautifully together, but the conventional power that we plan to use does not create as successful a signature at the lower speeds-so above 20 knots versus below 20 knots.
 Would you like to comment?
@@
 \begin{tcolorbox}[width=0.9\textwidth, center, colback=blue!5!white,colframe=blue!75!black, title= \textbf{4 October 2017 - Jean-Michel Billig, Executive Director Australian submarine Program, Naval Group}]
-\textit{In comments to journalists at the PACIFIC 2017 trade show, the Naval Group executive in charge of the Australian Program acknowledges that conventional propellers may be used on the Shortfin Barracuda \parencite{sbs2017}.
+\textit{In comments to journalists at the PACIFIC 2017 trade show, the Naval Group executive in charge of the Australian program acknowledges that conventional propellers may be used on the Shortfin Barracuda \parencite{sbs2017}.
 Subsequent reports quote Naval Group shifting responsibility for the propulsion choice to the Australian Government.}
@@
 \textbf{Jean-Michel Billig:} Yes, yes, we could, we could, we could also end up with something different
+\\
+\\
+(End transcription from interview)
 \\
@@
 This crucial concession reveals a very different level of confidence in the pumpjet being suitable for the Australian diesel-electric submarine.
-It also gives an important glimpse into the processes that selected the winning bidder, and key technology components, which reveals a clear contradictions between what would have been previously foregone conclusions.
+It also gives an important glimpse into the processes that selected the winning bidder, and key technology components, which reveals a clear contradiction between what would have been previously foregone conclusions.
 By revealing that the pumpjet has not yet been firmly decided upon, it is directly implied that France's exclusive offer to transfer the pumpjet technology could not have been the real distinguishing feature that elevated that bid above the others.
-Simultaneously, the statement that technology choice is a product of the operational requirement makes it clearly implies that the operational requirement has not yet been determined with sufficient precision to make the suitable technology.
+Simultaneously, the statement that technology choice is a product of the operational requirement makes it clearly implies that the operational requirement has not yet been determined with sufficient precision to make the suitable technology choice clear.
 As a ship-builder with ample experience in both technologies, uniquely positioned as having fitted both pumpjets and propellers to both diesel and nuclear submarines, it is remarkable that the performance information transferred prior or during the CEP was not sufficient to allow a decision  on the technology to be made.
@@
 \begin{tcolorbox}[width=0.9\textwidth, center, colback=blue!5!white,colframe=blue!75!black, title= \textbf{25 Oct 2017 - Herve Guillou, CEO Naval Group}]
-\textit{In an interview that was published in a post\parencite{guillou2017} on the ASPI Strategist the global CEO of Naval group responds to a question from Brendan Nicholson asking whether the Australian submarine will be fitted with a pump jet, following Jean-Michel Billig's comments.
+\textit{In an interview that was published in a post\parencite{guillou2017} on the ASPI Strategist the global CEO of Naval Group responds to a question from Brendan Nicholson asking whether the Australian submarine will be fitted with a pump jet, following Jean-Michel Billig's comments.
 Emphasis added.}
@@
 pumpjet) was ``integral" to their bid, whilst also deferring to what the Department of Defence ``intends" do use.
-This further exposes a clear tension between how 'integral' the pumpjet was to the bid, and Defence's freedom to choose otherwise.
+This further exposes a clear tension between how `integral' the pumpjet was to the bid, and Defence's freedom to choose otherwise.
 It also exposes a dramatic divergence between the level of emphasis that is given to the pumpjet as an important key feature of the proposal in public promotion (as highlighted in \cite{lane2017}) and a Defence decision which seemed to choose only to have the option of including the pumpjet, with the real decision to be deferred.
@@
 The first is that it contradicts subsequent explanations regarding pumpjet efficiency being incorrectly assessed by outsiders because they failed to account for how how the hull would be `tuned' to the pumpjet and the pumpjet `tuned' to the hull.
-If the hull diameter is already fixed and known as being identical to that of the nuclear Barracuda, then no adjustment ought to be necessary to a good working pumpjet design that is used to propel this platform on the hull's account.
+If the hull diameter is already fixed and known as being identical to that of the nuclear Barracuda, then no adjustment ought to be necessary to a good working pumpjet design that is used to propel such a similar platform - at least not on the hull's account.
 The performance of the pumpjet on the relevant hull would be well understood by the manufacturer offering it, and presumably available to the Australian government who were looking to acquire it.
@@
 However, it is certainly not true that scaling up a pressure hull to a new diameter to meet the same fixed static load is among the most difficult, risky or complex.
-Whilst extending the length of a submarine is undoubtedly even easier, adjusting the diameter still to meet a structural integrity requirement to meet a known static pressure (driven by the dive depth) is a relatively simple calculation in comparison to the complexities of many other calculations involved in submarine design.
-For instance, re-designing the duct intake of blade arrangements of a pumpjet to maximise efficiency at a different speed whilst simultaneously attempting to avoid cavitation would be vastly more complex and uncertain given the unstable difficult-to-model dynamics of turbulence and cavitation involved.
+Whilst extending the length of a submarine is undoubtedly even easier, adjusting the diameter to meet a known static pressure (driven by the dive depth) is a relatively simple calculation in comparison to the complexities of many other calculations involved in submarine design.
+For instance, re-designing the duct intake of blade arrangements of a pumpjet to maximise efficiency at a different speed whilst simultaneously attempting to avoid cavitation would be vastly more complex and uncertain given the famously unstable, difficult-to-model dynamics of turbulence and cavitation involved.
 In contrast, scaling up a hull-diameter was quite simply and successfully accomplished when the Collins Class submarine was developed by scaling up both the length and diameter of it's smaller Swedish predecessor.
-In 2006, DSTO commissioned a report which was dedicated to proposition of a different hull diameter being used for submarine used to replace Collins \parencite{joubert2006} in order to achieve lower hydrodynamic drag.
+In 2006, DSTO commissioned a report which was dedicated to the proposition of a different hull diameter being used for submarine used to replace Collins \parencite{joubert2006} in order to achieve lower hydrodynamic drag.
 If such a development path was actually thought to impose  such significant challenges on the engineering and certification process as RADM Sammut's remarks imply, there is no way that in 2006 when the replacement was presumed to be undertaken for a smaller number of new submarines which would enter service in 2026 (as opposed to mid-2030s as currently planned) that such an undertaking would have been seriously considered.
@@
 It would also have essentially precluded both the Germans and Japanese from the outset.
-Furthermore, if the advantage was considered so substantial, he should be called upon to substantiate how significant increased difficulty in certification would have been, with respect to cost and schedule, noting that the French proposal had a larger cost and longer budget than either of the other two contenders.
-
-Among RADM Sammut's additional claims regarding the reason for choosing DCNS/Naval Group, include remarks regarding the establishment of 'sovereign capability' and 'data transfer' to Australia.
+Furthermore, if the advantage was considered so substantial, he should be called upon to substantiate how significant the increased difficulty in certification would have been, with respect to cost and schedule, noting that the French proposal had a larger cost and schedule than other contenders.
+
+Among RADM Sammut's additional claims regarding the reason for choosing DCNS/Naval Group, include remarks regarding the establishment of `sovereign capability' and `data transfer' to Australia.
 One would assume that this meant that the French had demonstrated a willingness to supply complete and timely information on key technologies, and that RADM Sammut's existing experience with the French supplying such information on the pumpjet (as Sean Costello had claimed had taken place during and prior to the conclusion of the CEP) supported this view.
@@
 This contradicts the notion that the same pumpjet as France is fitting to the Suffren (i.e.
-nuclear barracuda) will simply be included 'as is' on the Australian non-nuclear 'Shortfin' version.
+nuclear Barracuda) will simply be included `as is' on the Australian non-nuclear `Shortfin' version.
 Given that the length and diameter of the hull were announced as being already decided and practically identical to the nuclear version, it raises the very serious question as to whether/why the changed shape of the hull would necessitate modification of the complex pumpjet.
-As pointed out in the my testimony to the Senate \ref{SENmorrison}, the emphasis of this relationship, above the vastly different operational speed profile of the nuclear and non-nuclear submarines which would affect the propulsion system profoundly, raises serious questions about the origin of the comment.
+As pointed out in the my testimony to the Senate[\ref{SENmorrison}], the emphasis of this relationship, above the vastly different operational speed profile of the nuclear and non-nuclear submarines which would affect the propulsion system profoundly, raises serious questions about the origin of the comment.
 
@@
 \\
-\\
 ...
-\\
 \\
@@
 In responding to the evidence produced by Senator Patrick that pumpjets may have an efficiency issue at low speeds, Mr Johnson refers to an `electric shrouded pod' which is used on cruise ships, which he claims is a modification of a pumpjet.
-Nothing about this section of his testimony seems to have any reference of connection to reality.
+Very little about this section of his testimony seems to have any reference of connection to reality.
 Cross-ocean transits regularly occur at higher speeds than 17-18kt, including by cruise ships, which he mentions.
 These speeds are essentially unreachable (or only extremely briefly sustainable) by a diesel electric submarine.
-As pointed out in my evidence to Senate Economic References Committee \ref{SENmorrison} I can find no evidence of any cruise ships having any kind of shroud around the tips of their propellers, or bearing any resemblance or relationship to a pumpjet.
-In the same session, whilst in Mr Johnson's presence, it appears that Mr Johnson may have confused electric Azipods, which are used on large ocean-going ships such as cruise ships, and the Navy's new LHDs, with pumpjets.
+As pointed out in my evidence to Senate Economic References Committee[\ref{SENmorrison}] I can find no evidence of any cruise ships having any kind of shroud around the tips of their propellers, or bearing any resemblance or relationship to a pumpjet.
+In the same session, whilst in Mr Johnson's presence, I pointed out that it appears that Mr Johnson may have confused electric Azipods, which are used on large ocean-going ships such as cruise ships, and the Navy's new LHDs, with pumpjets.
 Azipods have open-tipped propellers, and are not remotely related to pumpjets.
@@
 
-This answer to a question again rests on the claim that the influence of the hull means that comparisons and studies such as those referred to in \parencite{morrison2018} can be ignored.
-This is a question which will be given a substantial reply in subsequent sections, but even a quick reading superficial reading reveals alarming levels of over-reach in the answer.
+This answer to a question again rests on the claim that the influence of the hull means that comparisons and studies such as those referred to in Reference \parencite{morrison2018} can be ignored.
+This is a question which will be given a substantial reply in subsequent sections, but even a quick superficial reading reveals alarming levels of over-reach in the answer.
 (For instance, a type of ducted propeller known as a decelerating ducted propeller actually \textit{is} a pumpjet, by definition.
@@
 For Defence not to have obtained that opinion from DCNS at the time of considering the bid would be a matter of incomprehensible oversight.
-It is made all the more difficult to believe given the scrutiny applied to the matter in the Senate in 2012 \ref{SENMoffit}, and the program director's assurance that the work was being done to assess its suitability.
+It is made all the more difficult to believe given the scrutiny applied to the matter in the Senate in 2012[\ref{SENMoffit}], and the program director's assurance that the work was being done to assess its suitability.
 
@@
 If Defence came away with a positive opinion of the pumpjet's suitability whilst Naval Group actually privately maintained a negative one, (and elected not to avail Defence of their false impression for commercial reasons) this would amount to a gross betrayal of trust and quite possibly fraud on the part of the Naval Group and the French Government.
-It would also demonstrate an catastrophic level of incompetence on the part of Defence, given their stated efforts to assess the pumpjet prior to this point.
-If Defence has since become aware of Naval Group's lack of confidence in the pump jet's suitability, this may explain the defence response to this question, which explicitly refers to the pump jet propulsion being offered `as part of their Competitive Evaluation Proposal' and `remains on offer to Australia', which could be read to deflect responsibility for technology choice to Naval Group, implying they would or ought to withdraw the offer if it was now known to be inappropriate.
+It would also demonstrate a catastrophic level of incompetence on the part of Defence, given their stated efforts to assess the pumpjet prior to this point.
+If Defence has since become aware of Naval Group's lack of confidence in the pump jet's suitability, this may explain the defence response to this question, which explicitly refers to the pump jet propulsion being offered ``as part of their Competitive Evaluation Proposal" and ``remains on offer to Australia", which could be read to deflect responsibility for technology choice to Naval Group, implying they would or ought to withdraw the offer if it was now known to be inappropriate.
 
@@
 One must at this point give consideration to the possibly that Defence had not been entirely unwittingly deceived, and was actually complicit in their own deception, including its present upkeep.
-This would mean that the Department of Defence and the Navy, or at least a few of very senior people within those organisations, had conspired allow a proposal which was known to be inappropriate to the Commonwealth's needs to appear victorious in the CEP,  presumably in order to favour a bidder for whom no more defensible explanation could be found.
-It appears unlikely that there could be any decent motive for such a course of action, and would lead invariably to the grave conclusion that some kind of corruption has occurred.
+This would mean that the Department of Defence and the Navy, or at least a few senior people within those organisations, had conspired to allow a proposal which was known to be inappropriate to the Commonwealth's needs to appear victorious in the CEP,  presumably in order to favour a bidder for whom no more defensible explanation could be found.
+It appears unlikely that there could be any decent motive for such a course of action.
 In the circumstances, one must evaluate the probabilities of the necessary oversights for such an eventuality occurring by accident against the occurring by arrangement.
@@
 \begin{tcolorbox}[width=0.9\textwidth, center, colback=black!5!white,colframe=black!75!black, title= \textbf{3 April 2018 - Department of Defence, in response to Question on Notice from Senator Chris Ketter in December 2018}]
-\textit{Senator Ketter asks Question 3, which focuses specifically on whether Defence believes the pumpjet is acoustically superior across the entire speed range, or just at higher speeds, referencing my paper \parencite{morrison2018}, Herve Guillou \parencite{guillou2017} and Costello and Autret \parencite{autret2016}.
-Full question and response can be found in \ref{QON3}.}
+\textit{Senator Ketter asks Question 3, which focuses specifically on whether Defence believes the pumpjet is acoustically superior across the entire speed range, or just at higher speeds, referencing my paper\parencite{morrison2018}, Herve Guillou\parencite{guillou2017} and Costello and Autret\parencite{autret2016}.
+Full question and response can be found in Appendix \ref{QON3}.}
 \\
@@
 \\
-The company building the Navy's 12 new submarines says they'll be the world's first conventional boats big enough to effectively use the company's pump jet propulsion system.
-French Naval Group Australia says that its propulsor will prove much quieter than a propeller on the 4,700-tonne submarine it has dubbed the `Shortfin Barracuda'.
+\textbf{The company building the Navy's 12 new submarines says they'll be the world's first conventional boats big enough to effectively use the company's pump jet propulsion system.} French Naval Group Australia says that its propulsor will prove much quieter than a propeller on the 4,700-tonne submarine it has dubbed the `Shortfin Barracuda'.
 \\
@@
 \\
-``A lot of people don't fully comprehend, I think, that yes, a propulsor on a 1,500-tonne coastal submarine would make no sense.
+\textbf{``A lot of people don't fully comprehend, I think, that yes, a propulsor on a 1,500-tonne coastal submarine would make no sense.
 But we're not building a 1,500-tonne coastal submarine.
 We're building quite a large submarine.
-And this is the first time that a conventional submarine's been large enough to take a propulsor."
+And this is the first time that a conventional submarine's been large enough to take a propulsor."}
 
@@
 As such, this admission is more of a concession of the detrimental impact of the pumpjet on the submarine.
-This particular blog post was acknowledged on twitter by one former US Navy submariner and submarine enthusiast as an ``Awesome article about pump jet efficiencies and viability".
+My reply post was acknowledged on twitter by one former US Navy submariner and submarine enthusiast as an ``Awesome article about pump jet efficiencies and viability".
 Naval Group was tagged in the re-tweet, which was subsequently re-tweeted and liked by naval experts, publications, commentators and authors from around the world.
@@
 For example, the shaft horsepower-the power required to move even a very big ship at very low speeds-is astonishingly small.
-If you look at inefficiencies, you look at a curve and say, 'That must be a problem.' It's not.
-It's like: 'Hey, you paid twice what you needed to for that.
+If you look at inefficiencies, you look at a curve and say, `That must be a problem.' It's not.
+It's like: `Hey, you paid twice what you needed to for that.
 \textbf{It was a penny.
@@
 The attitude is clearly pervasive and supported in the highest levels of the defence department.
-The General Manager Submarines, Stephen Johnson, claims he simply cannot recall what he said about pumpjets being related to pods on cruise ships, despite being physically present in the same room when I pointedly quoted and contradicted him.
+The General Manager Submarines, Stephen Johnson, claims he simply cannot recall what he said about pumpjets being related to pods on cruise ships, despite being physically present in the same room when I pointedly quoted and contradicted him that morning.
 This being pointed out, he claimed to simply not to be paying attention, which won the support of the Head of Capability and Sustainment Group, Kim Gillis.
 Despite the obvious shock and exclamation of the senators, Mr Gillis went on to attempt to dismiss my evidence as just another opinion which could probably to be dismissed along with 99\% of the others.
-This is a sustained and deliberate effort to adopt a dismissive attitude in the face of serious criticism.
 
@@
 Moreover, his department had received written Questions on Notice from Senator Chris Ketter, acting on behalf of this very committee, specifically requesting written responses to my paper with a return date of 3 April.
-His department most probably responded to prior to the hearing, as noted above, with responses which blatantly refused to answer the questions.
-\footnote{However, I was not made aware either of the questions or their answers prior to hearing.} Furthermore this particular hearing was also especially long in the making, since it was once postponed on account of Defence refusing to make officials available to attend.\parencite{adm2018}  My presence as a witness was requested in February when the proposed date was still in May, and the list of witnesses was publicly circulated during the weeks prior to the hearing.
+His department most probably responded to prior to the hearing, as noted above, with responses which blatantly refused to answer the questions.\footnote{However, I was not made aware either of the questions or their answers prior to hearing.} Furthermore this particular hearing was also especially long in the making, since it was once postponed on account of Defence refusing to make officials available to attend.\parencite{adm2018}  My presence as a witness was requested in February when the proposed date was still in May, and the list of witnesses was publicly circulated during the weeks prior to the hearing.
 In such circumstances, claims to be essentially ignorant of my report and pretend it could probably be dismissed after the reading of the first page, should be examined critically.
@@
 
-In contrast to Mr Johnson's inference that my report might be made `without contemporary knowledge or without understanding necessarily the intricacy', his own comments on powering requirements of ships, just like the propulsive pods for ships, demonstrate a complete disregard or ignorance of the most basic and fundamental concepts relevant.
-
 \subsection{Dr David Kershaw, Chief of Maritime Division DSTG, in Australian Defence Magazine}
 \begin{tcolorbox}[width=0.9\textwidth, center, colback=black!5!white,colframe=black!75!black, title= \textbf{August 2018 - Dr David Kershaw, in Australian Defence Magazine}]
-\textit{Julian Kerr includes extensive quotation from Dr David Kershaw in an article titled ``Future Submarine: Pumpjet propulsor technology choice still not well explained" \parencite{kerr2018}.
-Dr Kershaw is the current Chief of Maritime Division, a position he held since May 2016.
+\textit{Julian Kerr includes extensive quotation from Dr David Kershaw in an article titled ``Future Submarine: Pumpjet propulsor technology choice still not well explained"\parencite{kerr2018}.
+Dr Kershaw is the current Chief of Maritime Division at DSTG, a position he held since May 2016.
 He was previously the Science and Technology Advisor for SEA 1000 between July 2011 and July 2014.}
@@
 This represents a pure, brazen, and obvious falsehood, justified by a seemingly-relevant truth.
-It's akin to arguing 'camels have to carry two heavy humps on their backs.
+It's akin to arguing `camels have to carry two heavy humps on their backs.
 Pigs don't have to carry these two heavy humps, which is why they can fly.'  It also directly contradicts the claim made by Stephen Johnson before the Senate, that ``No matter what kind of propeller or propulsion pump jet  you design, it has to be optimised for a particular speed.".
@@
 
-The same conclusion might be reached here regarding the comments about submarine size, and a connection to comments made by Brent Clark for the ASPI Strategist \parencite{clark2018}.
+The same conclusion might be reached here regarding the comments about submarine size, and a connection to comments made by Brent Clark for the ASPI Strategist\parencite{clark2018}.
 The comments by Dr Kershaw cast the argument in a very different light, suggesting that it is the weight of the pumpjet itself (as opposed to the weight of the extra fuel and batteries it demanded) was what made the pumpjet only viable for a very large diesel-electric submarine than anyone previous had built.
@@
 These comments are extremely important because they illuminate the true state of DSTG's research, and the degree of confidence they have in the pumpjet.
-It is consistent with responses given to a question on notice from Rex Patrick \ref{QON150}, where it was revealed that 1.1 million was spent on ``development of a \textit{generic submarine and pumpjet design}, computational and analytic models, and validation of these models through experimental studies".
+It is consistent with responses given to a question on notice from Rex Patrick[\ref{QON150}], where it was revealed that 1.1 million was spent on ``development of a \textit{generic submarine and pumpjet design}, computational and analytic models, and validation of these models through experimental studies".
 The admission that DSTG is still working on a `generic' design also casts into new light their repeated insistence that the performance of a pumpjet cannot be assessed unless it is exactly tuned carefully with the hull, and vice versa.
 It is plainly obvious that if they are still working on `generic' models, rather than the particular one which is proposed by France, they are in no better position to make positive assessment that contradicts findings in Reference \parencite{morrison2018}.
-By their own assessment, according to the insistence of Dr Kershaw and RADM Sammut, they are in no position to assess the performance of the pumpjet either given they haven't been able to do the necessary `tuning' the hull and pumpjet.
-Instead they are still working with `generic' designs.
-
-As I have already put before the Senate \ref{SENmorrison}, I find that idea that such `tuning' could so dramatically alter the performance of the pumpjet to be completely implausible.\footnote{In my evidence is said it was: ``as if someone bought an efficient hybrid car like a Prius and then decided to fit it with tank tracks from a dozer or a battle tank, and then argued they would tune the tank tracks back to be as efficient as the Prius was in the first place.
+
+As I have already put before the Senate[\ref{SENmorrison}], I find that idea that such `tuning' could so dramatically alter the performance of the pumpjet to be completely implausible.\footnote{In my evidence is said it was: ``as if someone bought an efficient hybrid car like a Prius and then decided to fit it with tank tracks from a dozer or a battle tank, and then argued they would tune the tank tracks back to be as efficient as the Prius was in the first place.
 It goes radically against what the technology was intended for in the first place".}  Rather, I would support the idea that the kind of generic modelling described would be quite sufficient to reveal results that are consistent with the results I produced from a basic literature review (that pumpjet efficiency declines towards zero at very low water-speeds) and safely rule-out the suitability of a pumpjet for diesel electric submarines.
@@
 
-Given the ample resources of Defence and Naval Group, this insistence that the result cannot yet actually be known, rather than that they know better and reached a different result and raises the question of what they actually do know, and how different it actually is to my result.
+Given the ample resources of Defence and Naval Group, this insistence that the result cannot yet actually be known, rather than that they know better and reached a different result raises the question of what they actually do know, and how different it actually is to my result.
 If the DSTG studies are actually consistent with my findings, then DSTG and Defence are knowingly suppressing such results, or naively hoping that the French design, and the finer `tuning' between the pumpjet and the hull can achieve such a dramatically different result that the results of `generic' tests can be ignored.
@@
 In contrast to the assurance of full technology transfer, the comments by Dr Kershaw in this article reveals the true state of information flow from DCNS.
-Apparently, there is still not sufficient information to be confident about the use of the pumpjet.
-A decision on whether the pumpjet will even be used still cannot be made.
-Consistent with the statements of Mr Johnson \ref{SENjohnson}, the information about the pumpjet performance on the nuclear Barracuda has not yet arrived.
+Apparently, there is still not sufficient information to be confident about the use of the pumpjet, or else such information is being supressed, and consequently decision on whether the pumpjet will even be used still either cannot be made.
+Alternatively, one might conclude the decision is being deliberately averted and delayed.
+
+The former view is consistent with some of the statements of Mr Johnson[\ref{SENjohnson}], that the information about the pumpjet performance on the nuclear Barracuda has not yet arrived.
 Nor, it would appear, is sufficient theoretical modelling, or scale model testing, or any other confirmatory data sufficient on this pumpjet for DSTG to be clearly persuaded of the suitability of the pumpjet for our submarine, which will have an almost-identical hull.
-Whilst simultaneously claiming that efficiency and performance of the pumpjet can't be assessed without the precise form of the hull and the pumpjet being 'tuned' together, they admit that they are having to model a 'generic' pumpjet, building up from first principals, in order to have sufficient confidence in their own 'independent analysis' of the technologies.
+Whilst simultaneously claiming that efficiency and performance of the pumpjet can't be assessed without the precise form of the hull and the pumpjet being `tuned' together, they admit that they are having to model a `generic' pumpjet, building up from first principals, in order to have sufficient confidence in their own `independent analysis' of the technologies.
 
@@
 Even more worrying, it would seem insufficient to even assess Naval Groups opinion about the suitability of the technology sufficient to answer parliamentary Questions on Notice.
-The gravity of such a failure is of profound consequence to the well-being of such a major project, given the significance of the pumpjet to the overall weight, balance and performance of the submarine as I outlined before the Senate \ref{SENmorrison}.
-
-It also throws into doubt the integrity of the statements made by RADM Sammut before Senate Estimates.
-In the days and weeks immediately prior to his testimony \ref{SENsammut}, Jean-Michel Billig had triggered the controversy around the pumpjet at PACIFIC 2017, the largest Naval Show and conference in Australia, which occurs only each two years.
+The gravity of such a failure is of profound consequence to the well-being of such a major project, given the significance of the pumpjet to the overall weight, balance and performance of the submarine as I outlined before the Senate[\ref{SENmorrison}].
+
+This justifies renewed attention to statements made by RADM Sammut before Senate Estimates.
+In the days and weeks immediately prior to his testimony \ref{SENhanson}, Jean-Michel Billig had triggered the controversy around the pumpjet at PACIFIC 2017, the largest Naval Show and conference in Australia, which occurs only each two years.
 This had lead to substantial coverage on the ABC\parencite{lane2017}\parencite{iggulden2017}, Defence Connect\parencite{mcmahon2017a}\parencite{mcmahon2017b}, and ASPI Strategist\parencite{davies2017}, and almost certainly contributed to the line of questioning that Senator Hanson doggedly pursued about the pumpjet that day.
@@
 It is inconceivable in an engineering project of this scale that after two years of work, and the conclusion October 2017 of the length, diameter and displacement of the submarine \parencite{sbs2017} that the type of propulsion is not yet known with certainty, and the inclusion (or not) of AIP is also not yet known with certainty.
-As discussed in my senate testimony \ref{SENmorrison} either of these technologies will alter load (of batteries, fuel, and AIP reactants) of the order of hundreds of tonnes, in order to meet a fixed performance requirement.
+As discussed in my senate testimony \ref{SENmorrison} either of these technologies will alter the load (of batteries, fuel, and AIP reactants) of the order of hundreds of tonnes, in order to meet a fixed performance requirement.
 Making such alterations within a hull that has its displacement already fixed could lead to catastrophic failure, essentially compromising the submarine's ability to sink or float.
 
-Only three possibilities exist which can explain why such a statement has been made.
+Only three possibilities exist which can explain why such a indeterminate statement has been made.
 
@@
 That such an approach to project management could be taken on a project like this is almost unthinkable from an engineering and risk-management perspective, not to mention cost-effectiveness.
+Defining the requirements should always come first.
 However, in light of the evidence available, the third possibility may warrant quite serious consideration.
 
-Such a possibility throws into sharp relieve some tensions between statements made by the government and Naval Group on multiple occasions.
+Such a possibility throws into sharp relief some tensions between statements made by the government and Naval Group on multiple occasions.
 RADM Sammut had previously stated that ``We did not specify that the boat should have or not have air-independent propulsion or lithium-ion batteries, rather that it met our requirements." \parencite{sammut2017}.
@@
 If all the competitors weren't actually given the same performance requirements, which were the real requirements for the submarine which the government intended to procure, then it is extremely difficult to see how they could be `treated fairly' as RADM Sammut claimed \ref{SENhanson}.
-(The credibility of conscience of the supposedly independent expert advisory panel who claimed they were should be seriously re-examined.)  If the requirements have changed substantially since the CEP, then the CEP ultimately served no purpose in selecting a partner who would best meet them.
-A vast number of public statements made by politicians and officials regarding the significance of Australia's unique 'requirements', particularly for long range, should be critically re-examined.
+If the requirements have changed substantially since the CEP, then the CEP ultimately served no purpose in selecting a partner who would best meet them.
+A vast number of public statements made by politicians and officials regarding the significance of Australia's unique `requirements', particularly for long range, should be critically re-examined.
 It is likely that numerous statements made citing these requirements, and implying that they existed, were known, and were communicated to other parties including throughout and subsequent to the CEP, would be found to be misleading of the public, and the Parliament.
@@
 It's true that large amounts of fuel and food might consequently be required.
-But the idea that we `might be better off carrying more food and fuel' betrays either either a lack of confidence that the volumes which were presumably proposed by all contenders alongside an AIP system were sufficient, or a casual aspiration to substantially enlarge the range and endurance of the submarine even further than  what the requirements demand.
+But the idea that we `might be better off carrying more food and fuel' betrays either either a lack of confidence that the volumes which were presumably proposed by all contenders (alongside an AIP system, if so proposed) were sufficient, or a casual aspiration to substantially enlarge the range and endurance of the submarine even further than  what the requirements demand.
 It would be relatively simple at the very early stages of concept design (prior to the diameter and length of the submarine being fixed) to incorporate both sufficient fuel and food for a specific overall range and endurance requirement, as well as an AIP system to enhance the dived endurance, without compromising the overall range and endurance.
@@
 Worryingly, it adds confirmation that RADM Sammut might genuinely believe that the hull-diameter (and its resemblance of a pre-existing design) ought to be a pre-determined feature of a submarine design, to which the performance and capability of the submarine is subordinate.
-Such a view, including held on the grounds of safety as suggested by RADM Sammut himself \ref{SENhanson}, is such a bizarre understanding of program risk, the hierarchy of engineering challenges and normal spiral of platform design that it should call into question his competence as a project manager, and the technical advisors whom he depends upon.
+This would indeed represent a bizarre understanding of program risk, the hierarchy of engineering challenges and normal spiral of platform design.
 
 The competence of RADM Sammut's program management and the technical advisors on whom he depends is further marred by the comment about lithium-ion batteries being now firmly excluded from the design because the ``RAN does not want to be introducing a massive number of new technologies simultaneously".
-This is telling because lead-acid batteries are known to have a relatively high internal resistance, which increases as they become hot, which occurs during levels of faster discharge.
+This comment is significant because lead-acid batteries are known to have a relatively high internal resistance, which increases as they become hot, which occurs during levels of faster discharge.
 This means that the efficiency of the batteries falls substantially at higher rates of discharge, reducing substantially the total amount of stored energy that they can usefully recover at very high discharge rates.
-Whilst this is generally true for all batteries, Lithium-ion batteries are known to have substantially lower internal resistance, and consequently have dramatically improved performance (in terms efficiency) at higher rates of discharge when  compared to lead-acid batteries.
+Whilst this is generally true for all batteries, Lithium-ion batteries are known to have substantially lower internal resistance, and consequently have dramatically improved performance (in terms of efficiency) at higher rates of discharge when  compared to lead-acid batteries.
 It is for this reason that in submarines, lithium-ion batteries are acknowledged as having the most significant improvement over lead-acid during bursts at high speed \parencite{briggs2016}.
-Given that pumpjets advantages are widely acknowledged to be at higher speeds, including implicitly by Naval Group\parencite{autret2016}, lithium-ion batteries should really be considered an enabling technology for pumpjets on diesel-electric submarines.\footnote{At least this is the case is if one chooses to prioritise a extremely high-speed sprint capability above virtually all other performance characteristics.
-I have argued this is the only technically defensible case to be made for the pumpjet, whilst such a tactical scenario is likely to be sufficiently obscure and unlikely for other technical reasons that it makes the argument dubious in the extreme.
+
+Given that pumpjets' advantages are widely acknowledged to be at higher speeds, including implicitly by Naval Group\parencite{autret2016}, lithium-ion batteries should really be considered an enabling technology for pumpjets on diesel-electric submarines.\footnote{At least this is the case is if one chooses to prioritise a extremely high-speed sprint capability above virtually all other performance characteristics.
+I have argued this is the only technically defensible case to be made for the pumpjet, whilst such a tactical scenario is likely to be sufficiently obscure and unlikely for other technical reasons that it still makes the argument dubious in the extreme.
 \ref{SENmorrison}}  Without lithium-ion batteries, the pumpjet efficiency will sap performance at low speeds, and at high speeds (where pumpjet efficiency improves, albeit not as high as propellers) the battery efficiency will hobble performance.
@@
 The exclusion of lithium batteries \textit{because} of the pumpjet's inclusion further indicates that RADM Sammut remains, or is pretending to remain, completely unaware of such basic interactions between technologies.
+If the due resources and determination that one would expect were applied during the CEP when assessing the alternative pre-concept proposals, this ought not to be the case.
 
@@
 
-The assertion that the resistance of a submerged body is constant with speed is such an easily demonstrable falsehood (as discussed in more detail in the subsequent section) that one can quickly dismiss the core of their argument against my conclusion, which depends on this false assumption.
+The assertion that the resistance of a submerged body is constant with speed is such an easily demonstrable falsehood (as discussed in more detail in a subsequent section) that one can quickly dismiss the core of their argument against my conclusion, which depends on this false assumption.
 The competence or intention of the authors is further exacerbated by numerous additional remarks.
@@
 This would mean that even a diesel submarine would also have unlimited range by virtue of it's ability to reach any speed  that it liked.
-This assertion  by defence is so utterly and blatantly ridiculous that there is no plausible way that it could be genuinely held to be a true by any one of even moderate technical competence.
+This assertion  by defence is so ridiculous that there is no plausible way that it could be genuinely held to be a true by any one of even moderate technical competence.
 It is difficult to even believe that it could be held with any confidence by even moderately intelligent adults of little or no technical background who took a mild interest in researching the topic.
-Unless such an extremely level of incompetence pervades every defence official and scientist who were involved in preparing this statement (which was requested directly from Kim Gillis, Head of CASG and Stephen Johnson, General Manager Submarines) is difficult to avoid reaching the conclusion that the statement was given in a state of extreme contempt for the intelligence of the intended reader, and constitutes a deliberate lie and active attempt to mislead.
+(Google `Drag Law' and read the Wikipedia article.)  Unless such an extremely level of incompetence pervades every defence official and scientist who were involved in preparing this statement (which was requested directly from Kim Gillis, Head of CASG and Stephen Johnson, General Manager Submarines) is difficult to avoid reaching the conclusion that the statement was given in a state of extreme contempt for the intelligence of the intended reader, and constitutes a deliberate lie and active attempt to mislead.
 
@@
 Both Dr David Kershaw, Chief of Maritime Division at DSTG\parencite{kerr2018}, as well as Mr Brent Clark\parencite{clark2018} have both made arguments to the effect of suggesting that it was the Australian submarine's very large size which is what made the pumpjet viable, where it might not be for smaller submarines.
-I responded directly to Clark's comments arguing that the size necessitated by the extra fuel and batteries a pumpjet demanded could hardly be counted as an advantage \parencite{morrison2018b}.
+I responded directly to Clark's comments arguing that the size necessitated by the extra fuel and batteries a pumpjet demanded could hardly be counted as an advantage\parencite{morrison2018b}.
 However, Dr Kershaw's argument reveals a different take on on why size is important.
@@
 As one would expect, none of the catastrophic failures eventuated.
-It is reported that the Russian Kilo Class submarine B-871 'Alrosa' has and does take part in exercises, and is remains in active service.
+It is reported that the Russian Kilo Class submarine B-871 `Alrosa' has and does take part in exercises, and is remains in active service.
 However, given that no other conventional submarines have since been designed with or converted to pumpjet propulsion, either in the Russian Navy or any of their export customers, it is difficult to conclude that the benefits of a pumpjet on this type of submarine outweigh the costs.
@@
 \begin{figure}[h]
-\captionsetup{width=0.7\textwidth}
+\captionsetup{width=0.8\textwidth}
 \includegraphics[width=0.7\textwidth, center]{alrosa1.png}
@@
 
-Furthermore, the French also fitted a pumpjet to an Agosta 70 class submarine for a trial, as confirmed by Brent Clark in the Senate \ref{SENmorrison}.
-This submarine is even smaller than the Kilo, around 1500 tonnes, exactly the type of 'coastal' submarine that Brent Clark said a pumpjet wouldn't make sense for \parencite{clark2018}.
+Furthermore, the French also fitted a pumpjet to an Agosta 70 class submarine for a trial, as confirmed by Brent Clark in the Senate[\ref{SENmorrison}].
+This submarine is even smaller than the Kilo, around 1500 tonnes, exactly the type of 'coastal' submarine that Brent Clark said a pumpjet wouldn't make sense for\parencite{clark2018}.
 One would assume however that if the weight of the pumpjet itself was likely to cause the submarine to tail-stand or sink, such a trial would not have even been attempted.
@@
 It is difficult to imagine what sort of circumstance would lead both an acting CEO of the contracted design partner, who was aware of the experience of the French experiment with the Agosta, as well as a very senior defence scientist to make such demonstrably false claims in public.
-In the case of Dr Kershaw, it can be seen that he deliberately and repeatedly (at the conclusion of his interview  \parencite{kerr2018}) drew attention to the size being significant.
+In the case of Dr Kershaw, it can be seen that he deliberately and repeatedly (at the conclusion of his interview\parencite{kerr2018}) drew attention to the size being significant.
 His concluding remarks for the interview were that ``The size of the Future Submarine means a pumpjet is a real game changer for optimising the performance of the submarine in its intended role...".
@@
 In a clear parallel with the argument that the performance of propulsion systems on submerged bodies is independent of speed because they don't face wave-making resistance, this line of reasoning involves the presentation of a seemingly relevant, comprehensible technical truth which is used to justify a completely blatant technical falsehood.
-Even if Dr Kershaw is not familiar with this area of science and speaking outside his competence, it is difficult to see how such a pattern of argument would occur so consistently by accident.
+Even if Dr Kershaw is not familiar with this area of science and is speaking outside his competence, it is difficult to see how such a pattern of argument would occur so consistently by accident.
 For Dr Kershaw to genuinely believe in the arguments that he is making here is almost unthinkable.
@@
 
-Bruce: I'm not sure about using these fat knobbly tires on my road-bike for the cycle tour... No-one else is using such tyres on a road-bike, I think they'd be slow and inefficient.
-
-James: Don't be silly!
+\textbf{Bruce:} I'm not sure about using these fat knobbly tires on my road-bike for the cycle tour... No-one else is using such tyres on a road-bike, I think they'd be slow and inefficient.
+
+\textbf{James:} Don't be silly!
 Everyone else would love to use fat knobbly tyres on a road-bike, the only reason they can't is that those cyclists themselves are just so slim and athletic, and can't use the knobbly fat knobbly tyres.
-Bruce, don't you realise that you're the \textit{only} cyclist every attempting this tour who weights 200kg, roughly three times what most road-cyclists weigh?
+Bruce, don't you realise that you're the \textit{only} cyclist ever attempting this tour who weights 200kg, roughly three times what most road-cyclists weigh?
 If a 70kg rider attempted those speeds with fat knobbly tyres, they wouldn't have the weight to stabilise the forces involved, and they'd be thrown off their bicycle in a catastrophic crash which would probably be fatal.
@@
 
-Bruce:  What about that one lonely guy over there riding on fat knobbly tyres.
+\textbf{Bruce:}  What about that one lonely guy over there riding on fat knobbly tyres.
 He looks like he weighs about 70kg, and he's not having a catastrophic crash, he's just going really slow and looks tired.
@@
 
-James: Oh...
-
-Bruce: And didn't you also tell me that you tried riding with fat knobbly tires once years ago?  How did you survive the catastrophic crash?
+\textbf{James:} Oh...
+
+\textbf{Bruce:} And didn't you also tell me that you tried riding with fat knobbly tires once years ago?  How did you survive the catastrophic crash?
 
@@
 
-The argument that the `hull needs to be tuned to pumpjet' and \textit{vice versa} has been used to dismiss criticism of the pumpjet repeatedly by RADM Sammut \parencite{sammut2017} \parencite{bergmann2018}, and also by Dr David Kershaw \parencite{kerr2018}.
-It has also been referred to in an answer to a Question on Notice given by Senator Ketter\ref{QON1}.
-The answer to the Question on Notice, and also Dr Kershaw's comments comprise clearly the more detailed explanations, which  will be referred to in principal in this section.
-However, the reference of RADM Sammut to this line of argument on multiple occasions, almost a year separated both before and after the publication of my paper\parencite{morrison2018} and my senate evidence \ref{SENmorrison} clearly indicates that it is a significant, perhaps the most significant line of defence of the  pumpjet in the program chief's mind.
-The written response supplied by Defence to the Senate Economic References Committee also refers clearly to this argument \parencite{defence2018}.
+The argument that the `hull needs to be tuned to pumpjet' and \textit{vice versa} has been used to dismiss criticism of the pumpjet repeatedly by RADM Sammut\parencite{sammut2017}\parencite{bergmann2018}, and also by Dr David Kershaw \parencite{kerr2018}.
+It has also been referred to in an answer to a Question on Notice given by Senator Ketter[\ref{QON1}].
+The answer to the Question on Notice, Dr Kershaw's comments and the Defence response to my paper\parencite{defence2018} comprise clearly the more detailed explanations, which  will be referred to in principal in this section.
+However, the reference of RADM Sammut to this line of argument on multiple occasions, almost a year separated - both before and after the publication of my paper\parencite{morrison2018} and my senate evidence[\ref{SENmorrison}] - clearly indicates that it is a significant, perhaps the most significant line of defence of the  pumpjet in the program chief's mind.
 
@@
 The major advantage that the Naval Group proposal \textit{ought} to have is that those potentially complicated engineering challenges can be safely and entirely avoided.
-Furthermore, the idea that the hull is what necessitates the tuning, as opposed to the speed profile, is almost incomprehensibly naive, as I pointed out in my evidence given to the Senate \ref{SENmorrison}.
+Furthermore, the idea that the hull is what necessitates the tuning, as opposed to the speed profile, is almost incomprehensibly naive, as I pointed out in my evidence given to the Senate[\ref{SENmorrison}].
 It is noteworthy that no officials from Naval Group have ever alluded to this `tuning' argument, or any adjustments to be made to the pumpjet or the hull on account of one another.
@@
 
-In \parencite{defence2018} the following argument was made, which builds on that in \ref{QON1}:
+In \parencite{defence2018} the following argument was made, which builds on that in Appendix \ref{QON1}:
 
@@
 $\eta_H$ describes net effect that the propulsor has on the hull by drawing water around its aft end at different speeds and altering the pressure forces on the hull, accounting for the potential for the propulsor to regain some of the energy inherent in the wake behind the hull.
-Both of these effects can be shown to be most significantly driven by the hull shape and the arrangement of the propeller with respect to the hull (its relative diameter, and proximity), and be relatively constant with speed.
+Both of these effects can be shown to be most significantly driven by the hull shape and the arrangement of the propeller with respect to the hull (its relative diameter, and proximity).
 As will be shown in this section, they could plausibly only vary a few percent each, which would have no impact in overturning the collapsing open-water efficiency of a pumpjet operating so far outside of design-conditions, such as a pumpjet at very low speed.
 
-To argue that none of these concepts have any useful bearing or meaning when applied to pumpjets is a bizarre, seemingly childish argument.
+To argue that none of these concepts have any useful bearing or meaning when applied to pumpjets is a bizarre argument.
 A pumpjet is actually by definition a particular form of a ducted propeller, specifically one in which the duct is designed to \textit{decelerate} the flow ahead of it the rotor\parencite{oosterveld1970}.
@@
 
-In the case of a pumpjet mounted on the aft end of a body of revolution such as a torpedo or submarine, for the most important elements in determining efficiency (such as mass flow rate) control lies in the designers hands to adapt the shape of the shroud to that of the hub in order to maintain equivalent volumes between the hub and the shroud, and preserve equivalent streamlines.
-The ratio between the area of the inlet, and the area of the outlet, and the ratio between the velocities of the water stream at the inlet and outlet of the shroud, are fundamental parameters in pumpjet design with regards to efficiency, as can be seen from a reading of Reference \parencite{henderson1964}, or more recently Reference \parencite{abdel2010}.
-And whilst in the discussion of different propulsion mechanisms alternative definitions and nomenclatures may be used in various forms of literature (for instance, to discuss the pump efficiency as a component of the jet efficiency, or account for different non-ideal effects in building up from an ideal model, as seen in Reference \parencite{wislicenus1973}, or because of the focus of that particular study) all will allow for a method of accounting for velocity variance in the incoming flow, and the effect of propulsors action on the boundary layer or wake ahead of it.
-Ultimately, all marine propulsion theory can be related to the acceleration of water, and despite the differences in the different means by which that acceleration occurs, there are natural bounds and limits to the extent to which the impact of accelerated water flow around a hull become incomparable due to the methods of that acceleration occurring.
+% In the case of a pumpjet mounted on the aft end of a body of revolution such as a torpedo or submarine, for the most important elements in determining efficiency (such as mass flow rate) control lies in the designers hands to adapt the shape of the shroud to that of the hub in order to maintain equivalent volumes between the hub and the shroud, and preserve equivalent streamlines.
+The ratio between the area of the inlet, and the area of the outlet, and the ratio between the velocities of the water stream at the inlet and outlet of the shroud, are fundamental parameters in pumpjet design with regards to efficiency, as can be seen from a reading of Reference \parencite{henderson1964}, or Reference \parencite{abdel2010}.
+And whilst in the discussion of different propulsion mechanisms alternative definitions and nomenclatures may be used in various forms of literature (for instance, to discuss the pump efficiency as a component of the jet efficiency, or account for different non-ideal effects in building up from an ideal model as seen in Reference \parencite{wislicenus1973}, or because of the focus of that particular study) all will allow for a method of accounting for velocity variance in the incoming flow, and the effect of propulsors action on the boundary layer or wake ahead of it.
+% Ultimately, all marine propulsion theory can be related to the acceleration of water, and despite the differences in the different means by which that acceleration occurs, there are natural bounds and limits to the extent to which the impact of accelerated water flow around a hull become incomparable due to the methods of that acceleration occurring.
 
@@
 
-Bruce: I'd like to buy some light, breathable shoes to go running with.
-
-James: Here, have these double-padded, thick-soled, water-poof, steel-capped, all-weather, leather work-boots.
-
-Bruce: But I don't think those shoes would perform well for running... aren't they uncomfortable?
-
-James: Don't you know that shoes have to operate closely in conjunction with socks?
+\textbf{Bruce:} I'd like to buy some light, breathable shoes to go running with.
+
+\textbf{James:} Here, have these double-padded, thick-soled, water-poof, steel-capped, all-weather, leather work-boots.
+
+\textbf{Bruce:} But I don't think those shoes would perform well for running... aren't they uncomfortable?
+
+\textbf{James:} Don't you know that shoes have to operate closely in conjunction with socks?
 Sometimes socks can get crinkled when running, leading to discomfort.
@@
 
-James: Hmm... actually I've just talked to a few people with similar boots, they all agree they'd be uncomfortable running.
-
-Bruce: How could they know?
+\textbf{James:} Hmm... actually I've just talked to a few people with similar boots, they all agree they'd be uncomfortable running.
+
+\textbf{Bruce:} How could they know?
 Their shoes could be similar maybe, but not quite the same.
@@
 
-James: Ok, but I just noticed I'm already wearing the same socks as you, can you tell me how this sock-boot pair works when running?
-
-Bruce: What??!
+\textbf{James:} Ok, but I just noticed I'm already wearing the same socks as you, can you tell me how this sock-boot pair works when running?
+
+\textbf{Bruce:} What??!
 oh...
@@
 
-\begin{figure}[h]
-\captionsetup{width=0.7\textwidth}
-\includegraphics[width=0.4\textwidth, center]{roteff1.png}
-\caption{One text's explanations of the variance between open-water efficiency and behind a hull from page 152 of \parencite{lewis1988}}
+% \begin{figure}[h]
+% \captionsetup{width=0.7\textwidth}
+% \includegraphics[width=0.4\textwidth, center]{roteff1.png}
+% \caption{One text's explanations of the variance between open-water efficiency and behind a hull from page 152 of \parencite{lewis1988}}
+% \label{fig:roteff1.png}
+% \end{figure}
+%
+% \begin{figure}[h]
+% \captionsetup{width=0.7\textwidth}
+% \includegraphics[width=0.4\textwidth, center]{roteff2.png}
+% \caption{One text's explanations of the variance between open-water efficiency and behind a hull from page 131 of \parencite{carlton2007}}
+% \label{fig:roteff2.png}
+% \end{figure}
+
+\begin{figure}[h]
+
+\begin{subfigure}{0.5\textwidth}
+\includegraphics[width=0.9\linewidth]{roteff1.png}
+\caption{}
+\end{subfigure}
+\begin{subfigure}{0.5\textwidth}
+\includegraphics[width=0.9\linewidth]{roteff2.png}
+\caption{}
+\end{subfigure}
+\caption{Two texts' explanations of the variance between open-water efficiency and behind a hull from (a) page 152 of \parencite{lewis1988} and (b) page 131 of \parencite{carlton2007}}
 \label{fig:roteff1.png}
-\end{figure}
-
-\begin{figure}[h]
-\captionsetup{width=0.7\textwidth}
-\includegraphics[width=0.4\textwidth, center]{roteff2.png}
-\caption{One text's explanations of the variance between open-water efficiency and behind a hull from page 131 of \parencite{carlton2007}}
-\label{fig:roteff2.png}
 \end{figure}
@@
 \caption{Dramatic changes in propeller size on a submarine (doubling/halving) would only produce variance of approximately 15\% in hull efficiency, according to Reference \parencite{jackson1992}.
-A pumpjet, which be smaller in diameter than a propeller by about \%20, would likely only make a few percentage points of difference due to changed flow over the hull ahead of it.}
+A pumpjet, which be smaller in diameter than a propeller by about 20\%, would likely only make a few percentage points of difference due to changed flow over the hull ahead of it.}
 \label{fig:hulleff.png}
@@
 \subsection{Open water efficiency, $\eta_0$}
-In clear contrast to the trivial changes that the relative rotative efficiency $\eta_R$ incurs, the open-water efficiency $\eta_O$ of a propulsor varies between 0\% and 75\%, depending crucially on the advance ratio $J$, which is highly speed dependent due to the changing resistance of the vessel at different speeds.
-In \parencite{morrison2018} it was shown that it was most probable a pumpjet's open-water efficiency could be around half that of a propeller at speeds at or below 5kt.
+In clear contrast to the trivial changes that the relative rotative efficiency $\eta_R$ incurs, the open-water efficiency $\eta_O$ of a propulsor varies between 0\% and maximum efficiency (up to 75\%), depending crucially on the advance ratio $J$, which is highly speed dependent due to the changing resistance of the vessel at different speeds.
+In Reference \parencite{morrison2018} it was shown that it was most probable a pumpjet's open-water efficiency could be around half that of a propeller at speeds at or below 5kt.
 It is consequently vastly more important to give detailed consideration to early in the design and selection of a propulsor, particularly for a diesel-electric submarine.
@@
 As can be seen in any number of texts, the open-water efficiency $\eta_0$ of a propulsor goes through a wide range from zero, to a peak, then back to zero, as a function of a variable called $J$.
-$J$ is known as the advance ratio, and can intuitively be thought of as how much the propulsor moves forward through the water that it's operating in per a given degree of rotation.
+$J$ is known as the advance ratio, and can intuitively be thought of as how much the propulsor moves forward through the water that it's operating in per a given degree of rotation, and is defined as $\frac{V_a}{nD}$.
 Since the resistance that water exerts on a body \textit{does} change with speed (it scales with $V^2$), contrary to Defence's claims, $J$ actually has a counter-intuitive relationship with speed.
@@
 It is important to point out here that these plots are carefully constructed in order to ensure that all the values represented are unitless, so that a scale-invariance can be applied, so that scale-models in tank-tests can be use to meaningfully inform the development of a propulsor for a variety of vessel sizes and speeds.
-A direct, and desirable corollary of this design is that is conceptual equivalence between different methods of increasing the resistance encountered by any given propeller.
+A direct, and desirable corollary of this design is that there is a conceptual equivalence between different methods of increasing the resistance encountered by any given propeller.
 Specifically, using the same propeller to push a much bigger and heavier hull will push the operating scenario to lower $J$ values in exactly the same way that demanding the propeller to push the original vessel much faster will.
-Conversely, fitting an extremely large propeller to a small load will allow the propeller to advance in nearly a corkscrew fashion (high $J$), just as requiring only an extremely low speed (and hence very low drag resistance) will achieve the same effect.
+Conversely, fitting an extremely large propeller to a small load will allow the propeller to advance in nearly a corkscrew fashion (high $J$), just as requiring only an extremely low speed (and hence very low drag resistance) on a given hull will achieve the same effect.
 
@@
 As such, there is no directly relationship between $J$, $K_Q$ or $K_T$ which can be used to determine a corresponding united measurements (e.g.
-water-speed, rpm, power etc) unless \textit{all} of the necessary assumptions are made and equations solved, and similarly fractions or ratios of coefficients hold to their conceptual analogues.
+water-speed, rpm, power etc) unless \textit{all} of the necessary assumptions are made and equations solved, and similarly fractions or ratios of coefficients do not hold to their conceptual analogues.
 ($K_T$ of 0.5 is not half a the thrust of a $K_T$ of 1.0).
@@
 
-The difference that applying the two different types of nozzle (accelerating and decelerating, as shown in Figure \ref{fig:nozza.png}) has on the efficiency of rotors is illustrated by comparing Figure \ref{fig:OpenWater.png} to Figures \ref{fig:nozz19b} and \ref{fig:nozz32b} which show the corresponding open-water efficiency plots.
+The difference that applying the two different types of nozzle (accelerating and decelerating, as shown in Figure \ref{fig:nozza.png}) has on the efficiency of rotors is illustrated by comparing Figure \ref{fig:OpenWater.png} to Figures \ref{fig:nozz19b.png} and \ref{fig:nozz32b.png} which show the corresponding open-water efficiency plots.
 The first distinguishing feature that should be noted is the point at which the efficiency curves fall to zero at higher $J$.
@@
 Likewise, at infinite load (or conceptually, infinite speed) the efficiency also falls to zero, as it must when the egg-beater condition occurs.
-Either no advance is gained for the next rotations, or infinite rotations must occur for to reach the next advance increment.
-
-Naturally we can assume that these particular shrouds and propellers don't exactly represent those to be used on a future submarine, but some of the most significant features are reliably illustrative of the general contrasts between un-shrouded rotors, and the propulsors resulting from shrouding them in  decelerating and decelerating ducts.
+Either no advance is gained for the next rotations, or infinite rotations must occur to reach the next advance increment.
+
+Naturally we can assume that these particular shrouds and propellers don't exactly represent those to be used on a future submarine, but some of the most significant features are reliably illustrative of the general contrasts between un-shrouded rotors, and the propulsors resulting from shrouding them in  accelerating and decelerating ducts.
 It might well be noted that both of the ducted propellers have slightly lower peak efficiency than the open one.
@@
 As such, the reason that my earlier paper did not explicitly discuss relative rotative efficiency and the impact of the wake field, this was due to it being so obviously and demonstrably of no bearing whatsoever on my conclusions.
-Perhaps I had falsely assumed that as much would immediately clear to any technically competent reviewer, who would readily grasp that open-water efficiency is the primary parameter which determines the efficiency of a propulsor as it pertains to the loading of the propulsor, and consequently speed.
+Perhaps I had falsely assumed that as much would immediately clear to any technically competent reviewer, who would readily grasp that open-water efficiency is the primary parameter which determines the efficiency of a propulsor as it pertains to the loading of the propulsor (changing values of $J$), and consequently speed.
 The use of this argument displays a substantial degree of ignorance of the relevant physics, as clearly described in numerous and publicly available sources.
@@
 \subsection{Turbulence and Flow Separation Re-Capped}
-It is noteworthy that absolutely no effort is made to engage in a discussion of turbulent or laminar in the boundary layer, nor does Defence make any mention of `flow separation' which are all important to the degradation of pumpjet efficiency at very low speeds.
+It is noteworthy that absolutely no effort is made to engage in a discussion of turbulent or laminar flow in the boundary layer, nor does Defence make any mention of `flow separation' which are all important to the degradation of pumpjet efficiency at very low speeds.
 A basic understanding of these concepts would be irreconcilable with the view that resistance underwater is constant with respect to speed.
@@
 
+\begin{figure}[h]
+\captionsetup{width=0.7\textwidth}
+\includegraphics[width=0.7\textwidth, center]{FlowSeparation.png}
+\caption{Flow separation involves flow reversal near a surface when a fluid is moving against an adverse pressure gradient, as shown in \parencite{mollard2011}}
+\label{fig:FlowSeparation.png}
+\end{figure}
+
 This distinction is important in understanding the limitations of a pumpjet's claim to make the incoming stream of water less inconsistent, or more regular.
-The duct intake and any stator blades in advance of the rotor can be effective in making the flow more homogeneous as assessed on a macroscopic scale, but in doing so they simultaneously make the flow more turbulent at a smaller scale, since it necessarily involve passing the water over more surfaces, rather than less.
+The duct intake and any stator blades in advance of the rotor can be effective in making the flow more homogeneous as assessed on a macroscopic scale, but in doing so they simultaneously make the flow more turbulent at a smaller scale, since it necessarily involves passing the water over more surfaces, rather than less.
 This can be advantageous to the extent that very large-scale and intense vortices might be broken down into smaller-scale vortices, which appears more homogeneous when assessed on the length-scale relevant to the blades of the rotors.
@@
 The circumstances where a pumpjet is able to attain higher efficiency are those where the total amount of additional productive energy (thrust, produced by front to back acceleration) added to the water column dramatically exceeds the losses incurred by the additional turbulence (round and round acceleration).
-This occurs at high speeds, which necessitate a high level of turbulence which a propeller is unable to avoid.
+This occurs at high speeds, when a great deal of thrust is added to the water-column, which is moving at very high speeds which make high levels of turbulence unavoidable to either type of propulsor.
 
@@
 Whilst at high speeds the water has enough momentum to easily overcome this resistance, at very low speeds there isn't enough energy in the flow for the entire water column to move consistently in this direction, and part of the flow simply turns around and moves in the opposite direction, in a phenomenon known as `flow separation'.
-This occurs adjacent to surface (in this case just inside the pumpjet intake), where the boundary layer has to move more slowly than the remainder of the stream, and in  this case it is the first part to reverse direction.
-Whilst the entirety of the boundary layer, (or indeed the entire flow) may actually be turbulent, with many very small-scale vortices, this phenomenon creates new large-scale macroscopic vortices, which could form, grow and shed in an unsteady creating exactly the same risk of periodic shedding (i.e.
+This occurs adjacent to a surface (in this case just inside the pumpjet intake), where the boundary layer has to move more slowly than the remainder of the stream, and in  this case it is the first part to reverse direction as shown in Figure \ref{fig:FlowSeparation.png}.
+Whilst the entirety of the boundary layer, (or indeed the entire flow) may actually be turbulent, with many very small-scale vortices, this phenomenon creates new large-scale macroscopic vortices, which could form, grow and shed in an unsteady manner creating exactly the same risk of periodic shedding (i.e.
 beats) impacting the rotor that a poorly designed mast or appendage might have.\footnote{A cylindrical mast, for instance, could set up a Karman Vortex Street at certain speeds.} Whilst it might be presumed the shedding of vortices created by flow separation into the pumpjet at low speeds has been adequately managed or mitigated to the extent that it does not risk the creation of cavitation on the blades, it would most certainly lead to the very substantial loss of efficiency through creation of these vortices.
@@
 However, a clear similarity or parallel between the three can be found.
-Each of them respectively actually comprise an argument against an obvious reference point being used to assess the efficiency of the pumpjet.
+Each of them respectively actually comprise an argument against an obvious and eligible reference point being used to assess the efficiency of the pumpjet.
 
 The argument that resistance is constant for underwater bodies, since they encounter no wave-making resistance, seeks to oppose any use of propulsive data derived from propulsors which might be used on ships, including for highly comparable systems (waterjets), or even directly comparable systems (pumpjets, or decelerating ducted propellers).
-
-The argument that the size of the submarine is a game-changer, attempts to dismiss the possibility that any experience with other diesel-electric submarines constitutes and adequate reference-point.
+This is extremely strange, given how the nature of propulsive tests, and the carefully unitless dimensions created are created specifically for the purpose of making the results transferrable to different situations, including facing different levels of resistance without the source of that resistance being specified.
+
+The argument that the size of the submarine is a game-changer, attempts to dismiss the possibility that any experience with other diesel-electric submarines constitutes and adequate reference-point, which is strange in the extreme given Naval Group's own experience with such experiments.
 
@@
 
-As I have argued, I think that there actually are adequate reference points in open-source literature regarding comparable propulsion systems, designed for comparable purposes, employed in relevant and comparable, in order to gain what is really only a very broad, but clear conclusion: that propulsive efficiency declines inevitably and dramatically at low speeds.
-
-Taking it as given that the French do have possession of adequate of far more detailed performance data, derived from even closer reference points than what I can glean from public literature, one must ask whether that data actually confirms, or contradicts my findings.
-Given public claims from DCNS during the CEP that information regarding both the hydrodynamic and acoustic performance of the pumpjet were in Australian government hands in Australia \parencite{costello2016} there can be little doubt Australian officials have access to such information.
+As I have argued, I think that there actually are adequate reference points in open-source literature regarding comparable propulsion systems, designed for comparable purposes, employed in relevant and comparable conditions, to gain the necessary level of accuracy to support a broad, but clear conclusion: that propulsive efficiency in pumpjets declines inevitably and substantially at low speeds in a way that propellers do not.
+
+Taking it as given that the French do have possession of adequate and far more detailed performance data, derived from even closer reference points than what I can glean from public literature, one must ask whether that data actually confirms, or contradicts my findings.
+Given public claims from DCNS during the CEP that information regarding both the hydrodynamic and acoustic performance of the pumpjet were in Australian government hands in Australia\parencite{costello2016} there can be little doubt Australian officials have access to such information.
 Indeed, it would be an inexcusable oversight if they didn't obtain that information during the CEP.
 
-This poses a seemingly irreconcilable tension between the desperate, implausible arguments currently being made by Defence, and the possibility that they are being made because Defence genuinely doesn't know better, and genuinely believes them to be true.
-It seems far more probable that Defence actually is in possession of clear, conclusive evidence which broadly supports the conclusions of my report, and is now deliberately concealing that evidence, and attempting to excuse why they have not drawn the obvious conclusions that any sensible person normally would from that evidence - that a pumpjet is entirely unsuited to a diesel-electric submarine.
+This poses a seemingly irreconcilable tension between the desperate, implausible arguments currently being made by Defence, and the almost certainty that they have quite definitive and reliable results readily available to them.
+It is difficult to accept as likely that such desperate and embarrassing arguments are being made because Defence genuinely doesn't know better, if the means to know better are actually so close at hand.
+It seems far more probable that Defence actually is in possession of clear, conclusive evidence which broadly supports the conclusions of my report, and is now deliberately concealing that evidence, and attempting to excuse why they have not drawn the obvious conclusions that any sensible person normally would: that a pumpjet is entirely unsuited to a diesel-electric submarine.
 
@@
 
+\newpage
+
 \begin{figure}[h]

### 2018-09-30 Second.Rnw
Final final, about to send

@@
 \usepackage{placeins}
+\usepackage{wrapfig}
 
@@
 
-\usepackage{draftwatermark}
-\SetWatermarkText{DRAFT}
-\SetWatermarkScale{1.5}
-\SetWatermarkLightness{0.95}
+% \usepackage{draftwatermark}
+% \SetWatermarkText{DRAFT}
+% \SetWatermarkScale{1.5}
+% \SetWatermarkLightness{0.95}
 
@@
 This would mean that a diesel-electric submarine would suffer poorer range and endurance, and significantly worse dived range and endurance, as well as a worse indiscretion ratio due to the use of a pumpjet.
-Given the profound importance of these characteristics to all diesel-electric submarines, as well as the elevated importance of range and endurance to the Australian requirement, it would appear that a pumpjet can be safely regarded as entirely unsuitable for Australia's Future Submarine.
+Given the profound importance of these characteristics to all diesel-electric submarines, as well as the elevated importance of range and endurance to the Australian requirement, it appears from these results that a pumpjet can be safely regarded as entirely unsuitable for Australia's Future Submarine.
 
@@
 The Committee directed Defence to provide a written response to my paper prior to the conclusion of the Inquiry into the Future of Australia's Naval Shipbuilding Industry (scheduled for the 27th of June) which they finally produced in mid September.
-Submarines for Australia had already commissioned me to write a subsequent report in reply to the responses that had appeared in responses to Questions on Notice in Parliament, and also in the media.
+Submarines for Australia had already commissioned me to write a subsequent report in reply to the responses that had appeared in answers to Questions on Notice in Parliament, and also in the media.
 With Defence's submission finally received, this paper is in a position to comprehensively assess the totality of the arguments put by Defence against the conclusions of my previous paper.
@@
 With the receipt of the written report it has become clear that Defence has no credible basis for rejecting the findings of my research.
-Their arguments are so completely incongruous with the most essential and frankly obvious physical principles, it appears that Defences has quite thoroughly disgraced their technical credibility in making this response.
-
-Their claim that pumpjets can be efficient across the entire speed range, including patrol speed, are instead based on the assumption that the efficiency of propulsors operating on fully-submerged bodies (like submarines) is ``essentially constant over speed"\parencite{defence2018} \parencite{kerr2018}.
+Their arguments are so completely incongruous with the most essential (and relatively obvious) physical principles, it appears that Defences has quite thoroughly disgraced their technical credibility in making this response.
+
+Their claim that pumpjets can be efficient across the entire speed range, including patrol speed, are based on the assumption that the efficiency of propulsors operating on fully-submerged bodies (like submarines) is ``essentially constant over speed"\parencite{defence2018} \parencite{kerr2018}.
 This is supported by the argument that efficiency relates to resistance, and that resistance faced by underwater bodies does not vary with speed, since it is only wave-making resistance (encountered by vessels on the surface) that is the speed-dependent component of resistance.
@@
 \item  \textbf{The French know the answer.}  As one of only two countries in the world who have designed and produced both pumpjets and propellers, nuclear and conventional submarines, and fitted and operated each type of propulsor on both types of submarines, there can be no doubt whatsoever that Naval Group and the French Government do have a well-formed opinion about the performance of pumpjets on diesel-electric submarines.
-They would have extensive operational performance data from their in-service SSBNS which use the pumpjet.
+They would have extensive operational performance data from their in-service SSBNs which use the pumpjet.
 They would have extremely detailed modelling, and tank-test results from the pumpjet from the almost-complete Barracuda class, which has an identical hull diameter to the proposed Shortfin Barracuda.
@@
 After the initial publication of my first report, three Questions on Notice were asked by Senator Ketter, specifically referencing my report[\ref{QON1}, \ref{QON2}, \ref{QON3}].
-Whilst Defence avoiding answering all of them directly, one of them was particularly brazenly side-stepped.
+Whilst Defence avoided answering all of them directly, one of them was particularly brazenly side-stepped.
 Asked ``Is Defence still confident that Naval Group still hold the opinion that the pump-jet is well suited to diesel-electric submarines?" Defence merely reiterated their intent is to use the pumpjets as offered, strictly avoiding mentioning anyone's opinion, either their own or Naval Group's.
@@
 
-Giving Defence the most generous possible interpretation, one might allow the possibility that Defence were just coming to grips with being completely hood-winked during the CEP by their chosen design-partner, and perhaps felt that their wrath at being deliberately deceived ought not be publicly vented, or at least not at this stage.
+Giving Defence the most generous possible interpretation, one might allow the possibility that Defence was just coming to grips with being completely hood-winked during the CEP by their chosen design-partner, and perhaps felt that their wrath at being deliberately deceived ought not be publicly vented, or at least not at this stage.
+News that negotiations have largely broken-down\parencite{greene2018} could well be considered in this light.
 One might, at a stretch, allow that maintaining the opportunity for the partner to save-face was still a useful bargaining-chip that our Defence chiefs to use in order to bring a potentially treacherous contractor to heel.
-(Presumeably in addition to a range of contractual penalties, as well as the commercial leverage the Commonwealth would have by being able to re-introduce competition into the design process as many have suggested ought be done in any case \parencite{stanford2017}, or simply not proceed to a build contract on completion of the design.)
-
-The state of commercial negotiations between the Commonwealth and Naval Group, as revealed in the Senate Economic References Committee \parencite{SEN2018morrison}, show that Defence was moving in exactly the opposite direction - shocking Senators with the revelation that Defence was looking instead to contract immediately with DCNS for the full construction of all 12 submarines, to the complete exclusion of the Australian Submarine Corporation, within coming months as part of the Strategic Partnering Agreement.
-This would effectively wipe out the opportunity for any competitive tension at any stage of the \$79 billion five-decade project within the first two years of a five-year design phase.
+(Presumably in addition to a range of contractual penalties, as well as the commercial leverage the Commonwealth would have by being able to re-introduce competition into the design process as many have suggested ought be done in any case \parencite{stanford2017}, or simply not proceed to a build contract on completion of the design.)
+
+However, at least at the time of the final public hearing for the Inquiry into the Future of Naval Shipbuilding in Australia \parencite{SEN2018morrison}, it appeared that Defence was moving in exactly the opposite direction.
+Senators on the committee were deeply concerned with establishing just how early it was during or after the CEP process that Defence agreed (with Naval Group) that they would be the \textit{builder} of the submarines, not just the design partner as previously assumed, and that Australian Submarine Corporation was to be consumed by DCNS.\footnote{On the 20th of June 2017 the Acting CEO of DCNS, Brent Clark stated that ``from our perspective, ASC will be consumed by DCNS."}  A further topic of substantial interest was whether the Strategic Partnering Agreement (SPA) that was currently being negotiated would commit the Commonwealth to use DCNS as the \textit{builder} of \textit{all 12} submarines.
+It was clear to anyone in the room that Defence's responses implied that it was very likely that this would be the case.
+
+Taken together, these observations paint a very different picture of the intended direction of Defence's relationship with DCNS.
+Defence appeared to be excluding two options which might have been assumed as likely or even default courses of action for a government seeking to minimise risk and retain some sovereign control over a crucial defence project:  either that the chosen foreign design partner would be expected to work with a local builder (one would assume ASC) to deliver the build, or that allocation of the actual build contract (or contracts, in the case they proceed in blocks) be left until a later stage when the detailed design (or preceding block of submarines) was closer to being completed satisfactorily.
+As members of the committee were clearly conscious, the apparent course taken by defence would effectively wipe out the opportunity for any competitive tension (or external technical scrutiny) at any stage of the \$79 billion five-decade project within the first two years of a five-year detailed design phase.
+Why Defence might be contemplating eliminating, rather than exercising opportunities to increase competitive tension at a time when it was becoming clear that the pumpjet proposal was disingenuous is extremely unclear.
 
@@
 
-In summary, I conclude that the pumpjet was a fraudulent, rather than incomptent proposal, used to construct the fiction that France's vaunted heritage as a designer of nuclear submarines offers a substantial technical advantage to a non-nuclear submarine.
-There is almost no chance that it will be included in the detailed design.
-In reality, without a nuclear reactor, there is relatively little relevant nuclear-related technology or experience (other than the combat system, mandated from a 3rd party) which is usefully transferable.
+In summary, I conclude that the pumpjet was a fraudulent, rather than incompetent proposal, used to construct the fiction that France's heritage as a designer of nuclear submarines offers a substantial technical advantage to a non-nuclear submarine.
+In reality, without a nuclear reactor, there is essentially no nuclear-related technology or experience which is usefully transferable to a diesel-electric submarine.
 The extent to which the bottom of the barrel needed to be scraped to maintain this fiction was illustrated when RADM Sammut was challenged on the benefit of the nuclear design after facing pressure on the pumpjet, and was forced to refer to the `diameter of the hull' as being a key learning[\ref{SENhanson}].
-Given the relatively more accessible engineering concepts involved, such claim would be much more easily subjected to public ridicule had it been more substantially promoted, as the pumpjet was.\footnote{Among the many engineering challenges presented in designing a new submarine, ensuring the structural integrity of the hull at a new diameter is one of the most well-understood and predictable.
+Given the relatively more accessible engineering concepts involved, such a claim would be much more easily subjected to public ridicule had it been more substantially promoted, as the pumpjet was.\footnote{Among the many engineering challenges presented in designing a new submarine, ensuring the structural integrity of the hull at a new diameter is one of the most well-understood and predictable.
 The Collins Class was evolved from a submarine of smaller hull diameter, and this aspect of the design and development encountered no problems.
@@
 In light of these considerations, Defence should be required to explain again what their reasons were for choosing the French as a design partner, as there can now be no doubt that the pumpjet played no legitimate part in it.
+The use of a pumpjet will inevitably cause a diesel-electric submarine to have poorer endurance and range, particularly dived endurance and range, and a poorer indiscretion ratio than one with open-tipped propellers.
 Any claims to the contrary are made either out of incompetence, as Defence's latest response might have us believe, or are made with the deliberate intent of misleading.
+
+\newpage
 
@@
 As such, I have also addressed it here as one of three substantial arguments made against my earlier conclusions regarding pumpjet efficiency.
-This third argument coming from a document so cafeully and deliberately prepared has had a material impact on the strength of the conclusions of this paper, which seeks to consider all three lines of reasoning in sufficient detail to suitably discuss the circumstances in which such arguments might be made.
+This third argument coming from a document so carefully and deliberately prepared has had a material impact on the strength of the conclusions of this paper, which seeks to consider all three lines of reasoning in sufficient detail to suitably discuss the circumstances in which such arguments might be made.
 
@@
 \newpage
-\section{A brief chronology of comment}
+\section{A chronology of comment}
 
@@
 \subsection{Department of Defence, to Senator Ketter's Questions on Notice}
-\begin{tcolorbox}[width=0.9\textwidth, center, colback=black!5!white,colframe=black!75!black, title= \textbf{3 April 2018 - Department of Defence, in response to Question on Notice from Senator Chris Ketter in December 2018}]
+\begin{tcolorbox}[width=0.9\textwidth, center, colback=black!5!white,colframe=black!75!black, title= \textbf{3 April 2018 - Department of Defence, in response to Question on Notice from Senator Chris Ketter with return date of 3 April 2018}]
 \textit{Senator Ketter asks Question 1, referencing specifically Greg Sammut in \parencite{sammut2017} and my paper \parencite{morrison2018} asks specifically whether Defence maintains that the pumpjet can be efficient across the entire speed range.
@@
 
-\begin{tcolorbox}[width=0.9\textwidth, center, colback=black!5!white,colframe=black!75!black, title= \textbf{3 April 2018 - Department of Defence, in response to Question on Notice from Senator Chris Ketter in December 2018}]
+\begin{tcolorbox}[width=0.9\textwidth, center, colback=black!5!white,colframe=black!75!black, title= \textbf{3 April 2018 - Department of Defence, in response to Question on Notice from Senator Chris Ketter with return date 3 April 2018}]
 \textit{Senator Ketter asks Question 2, inquiring as to how Defence assesses Naval Group's opinion regarding the pumpjet's suitability, following Jean-Michel Billig's comments.
@@
 
+\subsection{Department of Defence, to Senator Patrick's Questions on Notice}
+\begin{tcolorbox}[width=0.9\textwidth, center, colback=black!5!white,colframe=black!75!black, title= \textbf{Sept 2018 - Department of Defence, in response to Question on Notice from Senator Rex Patrick with return date 18 July 2018}]
+\textit{Senator Patrick asks a series of questions regarding the technical readiness level (TRL) of various technologies in the Future Submarine Program, and which ones were included or excluded in the French pre-concept design in the CEP.
+Full questions and answers can be found in Appendix \ref{QON827}.}
+\\
+\\
+\textbf{Question:} What is the Technological Readiness Level (TRL) of the propulsor being advanced for inclusion in the future submarine design.
+
+\textbf{Answer:} The technology to be used in the propulsor for the Future Submarine is assessed to be at TRL 9 noting it has been in operational use since the end of the 1990s on French submarines.
+The propulsor for the Future Submarine will be based on this technology.
+\\
+\\
+\textbf{Question:} Has an AIP been ruled out as a Future Submarine System in the early design work.
+
+\textbf{Answer:} At this point of the design process, increased battery and recharging capacity is proving more effective than the addition of AIP in meeting Australia's submarine mission requirements.
+
+\end{tcolorbox}
+
+This response to Questions on Notice clearly demonstrates that the technology being referred to for the pumpjet is indeed the in-service design from the Triomphant Class of SSBNs in the French Navy.
+There is no conceivable way that any technology at a higher level of readiness than TRL 7 would not have detailed performance data available, across the full range of operating conditions\parencite{DSTG2018a}.
+Alternatively, if the proposal is actually to design a new pumpjet ground-up for the radically different operating speed profile, then this answer is almost certainly misleading, as the actual propulsor to be used is probably at TRL 2, or at best TRL 3.
+However, as I have argued before the Senate[\ref{SENmorrison}] and in Reference \parencite{morrison2018}, any attempt to develop a pumpjet that is as efficient as a propeller at low speeds is essentially impossible.
+The elevated pressure (and lower speed) of water at the rotor which allows the superior acoustic performance at high speed \textit{necessitates} the decline in efficiency at low speeds.
+
+The comment regarding AIP also confirms that RADM Sammut's words in the APDR are not simply careless statements to the media or poor reporting, but represent a considered position of Defence - that AIP is still being considered, (i.e.
+not ruled out) but doesn't currently appear superior in meeting `mission requirements'.
+Another clear implication of this statement is that the mission requirements do exist, and are fixed.
+If they are not so, then this statement would be most severely misleading.
+This response also seems to confirm that a back-to-front design process is indeed in place, whereby the size of the hull is fixed, whilst there is still a large enough portion of the available displacement unallocated that one could consider an AIP system (several hundreds of tonnes) as one of a plurality of options to include.
+That Defence requires an \textit{increased} level of batteries in order to meet the requirement can only mean that the previous nominal load of batteries was insufficienct to meet the requirement, or that the plan now is to very substantially exceed the requirement.
+
+As was discussed earlier in this section, the scale of the impact in terms of performance and displacement of major technology choices such as the type of batteries, type of propulsor, and inclusion of AIP is so large that their status would need to be determined very early in pre-concept design, well in advance of the final hull dimensions being finalised.
+Otherwise the resultant variation in displacement/performance will result either in a potential engineering catastrophe (of Super-Seasprite proportions, i.e.
+the vessel doesn't sink and/or float), or a performance outcome that will not even remotely resemble the performance specified in the requirement.
+
 \FloatBarrier
@@
 
-\begin{tcolorbox}[width=0.9\textwidth, center, colback=green!5!white,colframe=green!75!black, title= \textbf{How a scientist feels about being told that because underwater bodies don't make waves, the resistance they encounter is constant with speed}]
+\begin{tcolorbox}[width=0.9\textwidth, center, colback=olive!5!white,colframe=olive!75!black, title= \textbf{How a scientist feels about being told that because underwater bodies don't make waves, the resistance they encounter is constant with speed}]
 
@@
 \begin{figure}[h]
-\captionsetup{width=0.7\textwidth}
+\captionsetup{width=0.9\textwidth}
 \includegraphics[width=0.7\textwidth, center]{drag4.png}
@@
 \begin{figure}[h]
-\captionsetup{width=0.7\textwidth}
+\captionsetup{width=0.9\textwidth}
 \includegraphics[width=0.7\textwidth, center]{dragfroude.png}
@@
 \begin{figure}[h]
-\captionsetup{width=0.7\textwidth}
+\captionsetup{width=0.9\textwidth}
 \includegraphics[width=0.7\textwidth, center]{drag1.png}
@@
 \begin{figure}[h]
-\captionsetup{width=0.7\textwidth}
+\captionsetup{width=0.9\textwidth}
 \includegraphics[width=0.7\textwidth, center]{drag3.png}
@@
 \begin{figure}[h]
-\captionsetup{width=0.7\textwidth}
+\captionsetup{width=0.9\textwidth}
 \includegraphics[width=0.7\textwidth, center]{drag2.png}
@@
 
-\begin{tcolorbox}[width=0.9\textwidth, center, colback=green!5!white,colframe=green!75!black, title= \textbf{How a scientist feels being told that the pumpjet is a game-changer for our diesel-electric submarine because it is so large, where it wouldn't make sense on a smaller one}]
+\begin{tcolorbox}[width=0.9\textwidth, center, colback=olive!5!white,colframe=olive!75!black, title= \textbf{How a scientist feels being told that the pumpjet is a game-changer for our diesel-electric submarine because it is so large, where it wouldn't make sense on a smaller one}]
 
@@
 
-\begin{tcolorbox}[width=0.9\textwidth, center, colback=green!5!white,colframe=green!75!black, title= \textbf{How a scientist feels about being told the pumpjet efficiency can't be assessed because it depends on the hull}]
+\begin{tcolorbox}[width=0.9\textwidth, center, colback=olive!5!white,colframe=olive!75!black, title= \textbf{How a scientist feels about being told the pumpjet efficiency can't be assessed because it depends on the hull}]
 
@@
 It seems far more probable that Defence actually is in possession of clear, conclusive evidence which broadly supports the conclusions of my report, and is now deliberately concealing that evidence, and attempting to excuse why they have not drawn the obvious conclusions that any sensible person normally would: that a pumpjet is entirely unsuited to a diesel-electric submarine.
+
+\newpage
+
+\section{About the Author}
+
+\begin{wrapfigure}{r}{0.35\textwidth}
+  \begin{center}
+    \includegraphics[width=0.28\textwidth]{aidan.jpg}
+  \end{center}
+  \caption{Aidan Morrison}
+\end{wrapfigure}
+
+Aidan Morrison is a physicist, entrepreneur and data-scientist who lives in Sydney with his wife Constance.
+He graduated from the Australian National University with First Class Honours in Theoretical Particle Physics, and commenced a PhD in Experimental Particle Physics, doing work on the ATLAS Experiment at CERN searching for the Higgs Boson.
+
+He left his PhD incomplete in order to found Rubber Ducky Defence, a start-up dedicated to developing small naval platforms and weapons for swarm warfare, and chose waterjets and pumpjets respectively for the  semi-submersible Swarm Combat Vessel and medium-weight torpedo involved.
+Rubber Ducky responded to Requests for Proposal (RFPs) for SEA 1778 and JP 1770, by adapting this vessel to meet a requirement for deployable mine-countermeasures and rapid environmental assessment.
+This requirement included towing and deploying a range of fully-submerged sensors, decoys, remotely and autonomously operated underwater vehicles in a variety of speeds to meet the specified conceptual operations.
+Endurance and range were crucial considerations in responding to the Preliminary Function and Performance Specifications for these two projects.
+A detailed proposal evolved from this concept and submitted to DSTO's Concept Technology Demonstrator program in 2013 was acknowledged as ``probably a world-class development".
+
+His other accomplishments include co-founding a bar-restaurant in Randwick called Bat Country, and developing a predictive forecasting tool for hospitality businesses using machine-learning techniques under Trendlock, which also undertakes additional consulting services.
+He has also lead a team of economic analysts at boutique think-tank, and currently works as a data-scientist at Tridant, most recently using machine-learning and statistical techniques to implement predictive maintenance routines for heavy industrial machinery.
 
@@
 
+\newpage
+
+\begin{flushleft}
+
 \printbibliography
+
+\end{flushleft}
+
+\FloatBarrier
 
@@
 
+\FloatBarrier
+
+\newpage
+
+\subsection{Answer to Question on Notice}\label{QON827}
+
+\begin{figure}[h]
+\includegraphics[width=0.7\textwidth, center]{QON827a.png}
+\end{figure}
+
+\newpage
+
+\begin{figure}[h]
+\includegraphics[width=0.7\textwidth, center]{QON827b.png}
+\end{figure}
+
+\FloatBarrier
 \end{appendices}
